### 2012-01-04 intro.tex
Intro

\section{Introduction}
\label{sec:intro}

% Talks about how AUV missions are traditionally planned. Why they might have
% been adequate in the past and are no longer so. How can deliberation
% (i.e. planning help) and how can it contribute to new ways of observing the
% ocean.

### 2012-01-04 related.tex
related work

\section{Related Work}
\label{sec:related}

% Refer to past work in AUV autonomy as well as robotic autonomy in
% general. This is general purpose and should be a good literature survey
% for robotics and AI.

### 2012-01-04 conclude.tex
updates

\section{Conclusion}

% Final parting words summarizing the contribution of deliberation towards
% vehicle autonomy.

### 2012-01-04 trunk/arch.tex
updates

\section{\texttt{T-REX} Architecture}
\label{sec:arch}

% Gives a high-level overview of T-REX, the general design principles and how
% these principles aid in software engineering. Show T-REX block diagram.

### 2012-01-04 trunk/concepts.tex
updates

\section{Key Concepts}
\label{sec:concepts}

% This is the core of the chapter and talks about Planning and
% Execution. Subsections with details on EUROPA as well as on T-REX
% synchronization which will have to be different from the sutton
% book. Most of our effort will be on this section with multiple
% sub-sections.

### 2012-01-04 trunk/future.tex
update

\section{Future Work}
\label{sec:future}

% Where do we see all this going esp. in the context of embedded AI Planning
% for marine robots? This can be (and should be)
% speculative. Shore-side/onboard autonomy coordination and control of
% multiple vehicles is an important result of this activity.

### 2012-01-04 results.tex
update

\section{Experimental Results}
\label{sec:results}

% Show some of the results of the use of T-REX to convince people that is for
% real. CANON will be the predominant driver of the work.

### 2012-01-04 setobook.tex
update

\documentclass[12pt]{svmono}

\input{preamble}

\usepackage{graphicx}

\usepackage[agsm]{harvard}

\title{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
\author{}
\date{\small Monterey Bay Aquarium Research Institute \\
7700 Sandholdt Rd.\\
Moss Landing,  CA, 95039\\
{\tt }}

\begin{document}

\maketitle

\begin{abstract}

  We describe a general purpose Artificial Intelligence based control
  architecture that implements the sense-plan-act paradigm. The
  Teleo-Reactive EXecutive (T-REX) framework is deliberates about
  future states, plans for actions and executes activities while
  monitoring plans for anomalous conditions. It is unusual in being
  operational and in routine use for operational oceanography,
  providing a new tool for science to sample and observe the dynamic
  coastal ocean. Plans are no longer scripted a priori but synthesized
  onboard with high-level directives instead of low level
  commands. Further, the architecture uses multiple control loops for
  a 'divide and conquer' problem solving strategy allowing for
  incremental computational model building, robust and focused failure
  recovery, ease of software development and ability to use legacy or
  non-native computational paradigms. Vehicle adaptation and sampling
  occurs in situ with additional modules which can be selectively used
  depending on the application in focus. Abstraction in problem
  solving allows different applications to be programmed relatively
  easily, with little to no changes to the core search engine thereby
  making software engineering sustainable. The representational
  ability to deal with time and resources coupled with Machine
  Learning techniques for event detection, allows balancing shorter
  term benefits with longer term needs, an important need as AUV
  hardware becomes more robust allowing persistent ocean sampling and
  observation.

%  % Traditional approaches to commanding marine robots have relied on
%  %  pre-scripted plans with a priori defined waypoints. These have
%  %  served the oceanographic community well in enabling robots to
%  %  cost-effectively perform surveys traditionally undertaken by
%  %  ship-board observations. However, in dynamic coastal environments
%  %  where is substantial interaction between atmospheric, oceanographic,
%  %  estuarine/riverine and land-sea interaction processes requires a
%  %  robot to be adaptive and receptive to opportunistic
%  %  science. Sustained use of such autonomous platforms in military
%  %  applications have also pushed.

\end{abstract}

\input{intro.tex}

\input{related.tex}

\input{arch.tex}

\input{concepts.tex}

\input{results.tex}

\input{future.tex}

\input{conclde.tex}

\section{Acknowledgements}

The authors are funded by a block grant from the David and Lucile
Packard Foundation to MBARI.

\bibliography{references}

\end{document}

### 2012-01-09 trunk/setobook.tex
update

--- 2012-01-04 setobook.tex
@@
   We describe a general purpose Artificial Intelligence based control
-  architecture that implements the sense-plan-act paradigm. The
-  Teleo-Reactive EXecutive (T-REX) framework is deliberates about
-  future states, plans for actions and executes activities while
-  monitoring plans for anomalous conditions. It is unusual in being
-  operational and in routine use for operational oceanography,
-  providing a new tool for science to sample and observe the dynamic
-  coastal ocean. Plans are no longer scripted a priori but synthesized
+  architecture that incorporates in-situ decision making for
+  adaptation and control for autonomous underwater vehicles
+  (AUVs). The Teleo-Reactive EXecutive (\texttt{T-REX}) framework
+  deliberates about future states, plans for actions and executes
+  generated activities while monitoring plans for anomalous
+  conditions. Plans are no longer scripted a priori but synthesized
   onboard with high-level directives instead of low level
@@
   ability to deal with time and resources coupled with Machine
-  Learning techniques for event detection, allows balancing shorter
+  Learning techniques for event detection allows balancing shorter
   term benefits with longer term needs, an important need as AUV
   hardware becomes more robust allowing persistent ocean sampling and
-  observation.
+  observation. \rx is in operational and routine use at MBARI,
+  providing scientists a new tool to sample and observe the dynamic
+  coastal ocean.
 

### 2012-01-12 trunk/intro.tex
updates

\section{Introduction}
\label{sec:intro}

% Talks about how AUV missions are traditionally planned. Why they might have
% been adequate in the past and are no longer so. How can deliberation
% (i.e. planning help) and how can it contribute to new ways of observing the
% ocean.

Autonomous platforms in the marine environment have had a substantial
and immediate impact for both civil and military
applications. Autonomous Underwater Vehicles (AUVs) in particular have
altered the way oceanographers are able to reach beyond the surface
and make observations which are not only at higher resolution but have
altered both the discipline and impacted the engineering methodologies
behind Sampling, Control and Robotics itself \comment{need higher
  level citation here}. Yet challenges remain, principally in making
such robotic devices more attune to the environment, to sample large
swaths at the Meso-scale ($> 50$ km\textsuperscript{2}) and doing so
systematically and adaptively. Further, over the last two decades
substantial improvements in hardware including propulsion, battery
technologies and hardware have not been kept abreast of in software
especially in algorithmic methods in Sampling and Control and in
systematic engineering and software engineering development.

This is one important reason that periodic calls towards sustained
exploration and sampling \cite{rudnick03} are issued by scientists for
whom the scope and scale of observations have become increasingly
tractable using autonomous platforms both below and on the
surface. The complexity of oceanic processes and the chronic
under-sampling of the coastal oceans especially of dynamic features
spread over large spatial scales (from meters to kilometers in extent)
with dynamic biological activity across the temporal spectrum (in the
order of hours to weeks), has made it challenging for robotic
platforms to autonomously track and sample. A prominent example of
such a process is coastal algal blooms which are patchy and could
cover large coastal zones. Persistent observation of such dynamic
events dictates that our robotic assets track and sample such patches
which can evolve rapidly due to inherent bio-geochemical activity,
advection as well as diffusion with the water mass it resides in.

At the Monterey Bay Aquarium Research Institute (MBARI), a unique
inter-disciplinary collaboration between biologists, ecologists,
geneticists and roboticists on the CANON (Controlled Agile Novel
Observation Network) \cite{canon} has driven autonomous system
exploration and control. Fundamental questions related to sampling
strategies have driven the use cases for advanced concepts in
autonomy, specifically to Artificial Intelligence (AI) Planning and
Execution. Current work is also looking at how other fields in Machine
Intelligence especially Machine Learning can be leveraged to impact
the science of engineering systems in marine robotics and as a
consequence oceanography itself.

In this chapter, we propose an alternative view to decision-making
using in-situ automated Planning \cite{ghallab04} and more important
demonstrate that while earlier control paradigms primarily driven by
\emph{reactive} approaches to control can be augmented to not only
provide targeted observation and sampling, but also a more nuanced and
balanced consideration of mission objectives, environmental conditions
and available resources. Planning (or deliberation or projection in
action space) is important to balance current needs of a robot with
future desires or goals. And to do such planning in the context of
time and available resources in-situ on the robot. Planning without
time or action planning in of itself is not adequate; dealing with
time ensures that the balance between now and the future is handled
systematically. Yet, plans have to be tied to robotic action and in
the real-world which is not static. Further, plans project future
state which is not only likely to change, but is based on substantial
uncertainty\footnote{It is important to remember that these principles
  are broad much as Dwight Eisenhower is reputed to have said
  ``\emph{In preparing for battle I have always found that plans are
    useless, but planning is indispensable}'' and ``\emph{Failing to
    plan is planning to fail}''}. Modeling of the marine environment
to provide cues is hampered by poor understanding of the worlds ocean
and poor predictive skill of ocean models and often little to no
availability of a priori synoptic views of the survey area.

AUV control architectures have primarily been driven by reactive
Subsumption-based approaches \cite{brooks86}. They have been adequate
to doing routine straight-line survey tracts which have played a
valuable role in the acceptance of AUVs as a new tool for ocean
science and DoD applications. However, scaling of problem domains in
addition to dealing with event-response and discovery in the
water-column challenges \emph{how} controllers are written in the
first place, maintained and upgraded over the life-cycle of the
platform. Dealing with off-nominal endogenous or exogenous conditions
including sensor failures stretches how AUVs are currently being
utilized; one consequence for instance, is any off-nominal condition
results in a \emph{fail-safe} surface mode to radio for help. Further
these approaches, words to the contrary are not only hard to maintain,
are not 'adaptive' enough. Despite vigorous defense by Brooks early on
\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
require a model if not of the world then of the \emph{physics} of the
vehicle when it interacts with the world.

Persistent and early legitimate criticism of deliberation on the other
hand also well articulated by Brooks
\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
was targeting the performance of in-situ deliberation. Terrestrial
robotic platforms were well-known to frequently stop and ``plan''. In
the marine robotics world, such a situation would simply not be
permissible given the need to make continuous observations without
stopping an under-articulated robotic platform such as an AUV. However
algorithmic and implementational advances by the AI community over the
years has produced a range of planners with demonstrated embedded
capabilities \comment{cite}. Our work has been influenced by and in
turn influenced a wide range of activities in the field of AI Planning
and Plan Execution with demonstrated real-world and highly visible
capabilities in the Space domain \cite{mus98, rajan00, ai-chang04,
  bresina05} for NASA. Lessons learned have been applied to the world
of marine robotics which we bring together in this chapter. Important
lessons that were derived from this rich legacy involved the following:

\begin{itemize}

\item Projecting plans (or using \emph{deliberation}) can and should
  balance near-term objectives with end-term or even evolving goals.

\item Dealing with time and resources is essential in real-world
  planning.

\item A rich representation that can deal with co-temporal events for
  modeling complex systems is a must.

\item Fast solvers for incremental causal reasoning can be built as a
  basis for dealing with dynamic controllable events where replanning
  is tightly integrated.

\item Planning should be tightly inter-leaved with execution to enable
  responsiveness to events which impact plan failure. This is also an
  important factor leading to platform adaptivity.

\item A model-based approach separating control formulation of the
  platform which describes operational characteristics from the
  domain. The controller can then be rigorously tested and certified
  even as the domain model evolves to fit diverse applications.

\end{itemize}

We have developed, tested and deployed the Teleo-Reactive EXecutive
(\texttt{T-REX}) an on-board adaptive control system that integrates
AI based planning and probabilistic state estimation in a hybrid
executive \cite{mcgann08a,mcgann08b,Py10}.  Probabilistic state
estimation integrates a number of science observations to produce a
likelihood that the vehicle sensors perceive a feature of
interest. Onboard planning and execution enables adaptation of
navigation and instrument control based on the probability of having
detected such a phenomenon. It further enables goal-directed
commanding within the context of projected mission state and allows
for replanning for off-nominal situations and opportunistic science
events. The framework in addition to being used on an AUV, is general
enough be used for controlling a personal robot \cite{mcgann2009} and
is deployed on a European planetary rover testbed
\cite{goac11}. Probabilistic state estimation is discussed elsewhere
\cite{mcgann08d} and out of scope for this chapter. Deliberation and
reaction are integrated systematically over different temporal and
functional scopes within a single agent and a single model that covers
the needs of high-level mission management, low-level navigation,
instrument control, and detection of unstructured and poorly
understood phenomena. \rx is deployed on MBARI's Dorado AUV shown in
Fig. \ref{fig:auv-fig}, which to the best of our knowledge is the only
operational marine platform anywhere being used for routine scientific
surveys with onboard plan synthesis.

\begin{figure}[t]
  \centering
  \vskip-5pt
  \includegraphics[scale=0.325]{figs/MBARI-AUV.jpg}
  \caption{\small The MBARI \emph{Dorado} AUV on its support vessel
    the R/V \emph{Zephyr}.}
  \label{fig:auv-fig}
  \vskip-0.3cm
\end{figure}

### 2012-01-12 concepts.tex
update

\section{Foundational Concepts}
\label{sec:concepts}

Our system, \texttt{T-REX} (or the Teleo-Reactive EXecutive) is an
adaptive, artificial intelligence based controller and provides
general framework for building reasoning systems for real-world
autonomous vehicles. At MBARI \texttt{T-REX} is used for AUV control;
another instantiation of the system is being used for control of a
terrestrial personal robot \cite{pr2}. The development of
\texttt{T-REX} has been targeted at surveying a number of
oceanographic features which are dynamic and unpredictable
spatio-temporally. Typically this requires our AUVs to balance the
goals of coverage spatially while opportunistically following features
of scientific interest and to do so while being operationally aware of
its own limitations in terms of resources (typically the battery state
of charge) and overall proximity to other observational assets for
obtaining scientific ground-truth. We are therefore interested in
representational frameworks that allow robots to pursue long-term
objectives while choosing short-term gain and can gracefully deal with
exogenous or endogenous change.

To enable this responsiveness to external observations, the agent has
to be able to synthesize plans insitu and to re-plan. We use a
temporal constraint-based planner EUROPA$_2$ with a demonstrated
legacy of having flown on NASA space missions
\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
innovations for AUV adaptation: the use of flexible plan
representations, compositional control with the use of partitioned
networks and off-line learning to inform insitu environmental state
estimation. In this section we detail some of the key concepts
relevant to flexible plan representation. Section \ref{sec:arch} has
details the \texttt{T-REX} partitioned architecture and estimation
driven vehicle adaptation.

% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
% for the Mars Exploration Rovers \cite{bresina05}.

EUROPA$_2$ uses a domain model written in a declarative language
(NDDL: New Domain Description Language), together with initial
conditions and goals also specified in NDDL, to construct a set of
temporal relations that must be true at the start time. These models
include assertions about the physics of the vehicle, i.e how it
responds to external stimulus and internally driven goals. By
propagating these relations forward using Simple Temporal Networks
\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
a set of conditions that should be true in the future, where some of
these conditions will correspond to actions the agent must take. The
planner can backtrack and try another path during search if a goal
cannot be reached while being capable of discarding unachievable
goals.

Traditionally robot execution has relied on dispatching commands at
precise times. Such linear sequences of precisely timed commands give
no ability to adjust execution on the basis of sensory
information. Although some commands can issue tests on sensor
readings, these tests have the objective of verifying whether expected
execution conditions are occuring. If not, the state of the system is
declared off-nominal and execution of the sequence of commands is
interrupted. More recently, executives have been proposed and
implemented that significantly broaden the way robots can be commanded
\cite{mus98,Alami:1998p820}. For example, the Remote Agent executive
interpreted a \textit{temporally flexible plan} which represents each
start time as a variable and contains an explicit network of bounded
delay constraints between such variables.

\begin{figure}[!t]
\centering
\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
\caption{\small Tokens with flexible temporal intervals and parametric
  constraints between tokens. This example shows the triggering of a
  water sampler based on a feature threshold while the vehicle
  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
  start \& end times.}
\label{fig:flex-timelines}
\vskip+0.1cm
\end{figure}

EUROPA$_2$ uses a \emph{state variable} representation to describe the
evolution of state over time. The instantiated history of such state
variable evolution over a temporal horizon we call \emph{timelines}
and which represent a single thread in the execution of a concurrent
system. At any given time each thread can execute a single procedure.
Thus each timeline consists of a sequence of procedures which
encapsulate and describe state evolution; we call these instantiated
atomic entities \emph{tokens}.  A token therefore describes a
procedure invocation, the state variables on which it can occur, the
parameter values of the procedure, and the time values defining the
interval. We allow encapsulation of uncertainty within these tokens
with a range of start and end times and parameters, all of which are
encoded as variables. A constraint solver in turn manipulates these
variables defined in a EUROPA$_2$ domain model. For example, consider
a scientific need to take a water samples 100 meters from a hotspot
while an AUV is performing a Yo-Yo in the water-column. Two samples
are needed if the feature's signal is above a threshold; one
otherwise. The token that is capturing the sensory threshold has a
parametric constraints to the token which fires the requisite water
sampler. In addition, the start time of the water sampling procedure
is highly dependant on the variability of sub-sea currents and actual
speed of the vehicle. Therefore a number of values are possible for
the start times and duration of the sampling all of which are valid
combinations for desired outcomes.

Unlike a traditional fixed time-tagged command sequences, such
flexible plans leave room for adaptation at execution time. When the
executive considers when to start a task, it propagates information
through the constraint network, computes a time bound for the
variable, selects an actual execution time within the bound, and
starts the task at that time. Temporally flexible plans therefore,
express a \textit{range of possible outcomes} of the robots
interaction with the environment within which the executive can elect
at run time the most appropriate one for the actual execution
conditions. The fact that constraints are explicitly represented
ensures that through constraint propagation the executive will respect
global limits expressed in the plan (e.g., don't start a task until a
certain condition has been satisfied but still satisfying some global
deadline). Such flexibility is critical when dealing with dynamic
ocean conditions where precise timing of a robotic action might be
indeterminate. Further, the advantage of flexibility can be contrasted
with the consequences of the intrinsic inflexibiliy of traditional
command sequences. Because they are inflexible, sequences must
necessarily be designed considering worst case scenarios.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} relations shown above.}
\label{fig:allen-algebra}
\vskip-0.3cm
\end{figure}

Each of the token variables, including the parameter variables, has a
domain of values assigned to it.  The variables may also participate
in constraints that specify which value combinations are valid.  To
allow the specification of multiple values, e.g, to express a range of
possible start times, variables are used to specify parameter, start
and end time values for a token.  As a result, a token $T$ is a tuple
$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
$s\leq e$) are variables encoded as flexible token start and end time
points. Consequently, partial plans represent a \textit{range of
  possible solutions} rather than an explicit definition of a single
trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
of a part of such plans with tokens representing constraints on
flexible temporal intervals. Systematic computation is reflected by
the use of the \texttt{Allen Algebra} \cite{allen84} temporal
relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
interval arithmetic in the EUROPA$_2$ domain model that allows
encoding of temporal constraint relationships between tokens on the
same or other concurrent timelines. Details of propagation algorithms
or of interval logic are beyond the scope of this paper.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
\caption{\small Timeline evolution can be described by simple finite
  state transitions where temporal relationships describe the arc
  transition and temporal extant of state are token
  durations. Constraints here are for \emph{meets} or its equivalent
  inverse \emph{met\_by}.}
\label{fig:FSM}
\vskip-0.3cm
\end{figure}

In describing the evolution of state via timelines, transitions within
a timeline can be described by simple finite state
machines. Fig. \ref{fig:FSM} for example describes a timeline
evolution for an abstract timeline entity describing navigation. Arc
transitions represent temporal constraints and token duration are
represented by temporal extant of state. The entire vehicle can then
be modeled as a number of state transitions with concurrency for each
of the modeled state variables defined a priori. However this simple
view is often complicated when temporal constraints between state
variables are articulated in our domain models. If there is a
criticism of this approach, modeling then, requires a proper balance
between abstraction, state variable description and concurrency
described by the inter-connectedness between such finite state
machines.

\begin{figure*}
\centering 
\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
\caption{\small An illustrative plan synthesis example with concurrent
  timelines. Each timeline is an instantiation of a subsystem tracked
  by the planner. Tokens describe the instantiated state of subsystem
  at a particular time instance and enforce temporal and parametric
  constraints. Abstract goals are decomoposed into successively less
  abstract tokens and instantiated in co-temporal timelines using
  \texttt{Allen Algebra} relations. All tokens represent flexible
  start/end times. \ref{fig:plan-evolve1} shows an initial state
  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
  \label{fig:Plan-evolve}
  \vskip-5pt
\end{figure*}

Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
timelines which track \texttt{Goal, Path, Navigation} and
\texttt{Command} state over time. Tokens on the goal timeline indicate
a survey within a bounded volume and a min/max depth envelope. As the
Volume Survey sub-goals on the \texttt{Path} timeline, the initial
token on that timeline gets \emph{squeezed} to have dependencies of
the goal token instantiated. Initially this dependency are the tokens
\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
coverage. The first of these in turn generates a sub-goal on the same
timeline to go to the next intermediate waypoint, which is illustrated
in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
figure) ultimately generate all the tokens to completion for this
timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
in plan generation and shows additional subgoals on the \texttt{Path,
  Navigation} and \texttt{Command} timelines. Each token can thus
trace its causality when instantiated in the plan. The domain model
provides source of these temporal and parametric dependencies in the
form of rules.

\begin{minipage}[c]{\textwidth}

% \framebox[\textwidth][t]{
Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
$\Rightarrow$ \\
met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
\ldots{} \\
ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
\end{minipage}

% \subsubsection{Partitioned Control}

% \subsubsection{State Estimation}

% Even as automated reasoning approaches have the ability to dynamically
% retarget the vehicle, estimating environmental signals of interest (or
% their proxies) is important to be able to enable opportunistic science
% in the water-column. In \texttt{T-REX} feature recognition revolves
% around a Hidden Markov Model (HMM) \cite{rabiner86} which is encoded
% directly within the unified representational and computational
% framework of a reactor. HMMs are useful since the stochastic nature of
% these models can correlate the type of features we want to detect with
% the sensor observations. Online targeted sensor data is classified by
% apriori generated cluster data which in turn determines the
% probability of having seen the feature of interest. Together with
% posterior probability, the HMM generates a probability of being within
% the target feature if it dominates the distribution. If other sampling
% conditions are satisfied (for instance sampling proximity), a
% constraint is activited in the model which can trigger a water sampler
% and also preserve the memory of sampling to alter a future transect
% resolution.

% The HMM is customized for the feature of interest and is built offline
% using clustering techniques from large data sets; we use Kohonen's
% Self Organizing Maps \cite{kohonen}. We use a semi-supervised learning
% method \cite{zhu05} to extract an environmental model that make uses
% of both labeled and unlabeled data for training. Online,

% The sensor data for classification depends on the
% feature of interest; for detecting INLs we use backscattering data
% from a HydroScat and for tracking riverine plumes to track fertilizer
% runoff, we use an ISUS Nitrate sensor.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ieee-ram09"
%%% End:

### 2012-01-12 intro.tex
update

@@
 % ocean.
+
+Autonomous platforms in the marine environment have had a substantial
+and immediate impact for both civil and military
+applications. Autonomous Underwater Vehicles (AUVs) in particular have
+altered the way oceanographers are able to reach beyond the surface
+and make observations which are not only at higher resolution but have
+altered both the discipline and impacted the engineering methodologies
+behind Sampling, Control and Robotics itself \comment{need higher
+  level citation here}. Yet challenges remain, principally in making
+such robotic devices more attune to the environment, to sample large
+swaths at the Meso-scale ($> 50$ km\textsuperscript{2}) and doing so
+systematically and adaptively. Further, over the last two decades
+substantial improvements in hardware including propulsion, battery
+technologies and hardware have not been kept abreast of in software
+especially in algorithmic methods in Sampling and Control and in
+systematic engineering and software engineering development.
+
+This is one important reason that periodic calls towards sustained
+exploration and sampling \cite{rudnick03} are issued by scientists for
+whom the scope and scale of observations have become increasingly
+tractable using autonomous platforms both below and on the
+surface. The complexity of oceanic processes and the chronic
+under-sampling of the coastal oceans especially of dynamic features
+spread over large spatial scales (from meters to kilometers in extent)
+with dynamic biological activity across the temporal spectrum (in the
+order of hours to weeks), has made it challenging for robotic
+platforms to autonomously track and sample. A prominent example of
+such a process is coastal algal blooms which are patchy and could
+cover large coastal zones. Persistent observation of such dynamic
+events dictates that our robotic assets track and sample such patches
+which can evolve rapidly due to inherent bio-geochemical activity,
+advection as well as diffusion with the water mass it resides in.
+
+At the Monterey Bay Aquarium Research Institute (MBARI), a unique
+inter-disciplinary collaboration between biologists, ecologists,
+geneticists and roboticists on the CANON (Controlled Agile Novel
+Observation Network) \cite{canon} has driven autonomous system
+exploration and control. Fundamental questions related to sampling
+strategies have driven the use cases for advanced concepts in
+autonomy, specifically to Artificial Intelligence (AI) Planning and
+Execution. Current work is also looking at how other fields in Machine
+Intelligence especially Machine Learning can be leveraged to impact
+the science of engineering systems in marine robotics and as a
+consequence oceanography itself.
+
+In this chapter, we propose an alternative view to decision-making
+using in-situ automated Planning \cite{ghallab04} and more important
+demonstrate that while earlier control paradigms primarily driven by
+\emph{reactive} approaches to control can be augmented to not only
+provide targeted observation and sampling, but also a more nuanced and
+balanced consideration of mission objectives, environmental conditions
+and available resources. Planning (or deliberation or projection in
+action space) is important to balance current needs of a robot with
+future desires or goals. And to do such planning in the context of
+time and available resources in-situ on the robot. Planning without
+time or action planning in of itself is not adequate; dealing with
+time ensures that the balance between now and the future is handled
+systematically. Yet, plans have to be tied to robotic action and in
+the real-world which is not static. Further, plans project future
+state which is not only likely to change, but is based on substantial
+uncertainty\footnote{It is important to remember that these principles
+  are broad much as Dwight Eisenhower is reputed to have said
+  ``\emph{In preparing for battle I have always found that plans are
+    useless, but planning is indispensable}'' and ``\emph{Failing to
+    plan is planning to fail}''}. Modeling of the marine environment
+to provide cues is hampered by poor understanding of the worlds ocean
+and poor predictive skill of ocean models and often little to no
+availability of a priori synoptic views of the survey area.
+
+AUV control architectures have primarily been driven by reactive
+Subsumption-based approaches \cite{brooks86}. They have been adequate
+to doing routine straight-line survey tracts which have played a
+valuable role in the acceptance of AUVs as a new tool for ocean
+science and DoD applications. However, scaling of problem domains in
+addition to dealing with event-response and discovery in the
+water-column challenges \emph{how} controllers are written in the
+first place, maintained and upgraded over the life-cycle of the
+platform. Dealing with off-nominal endogenous or exogenous conditions
+including sensor failures stretches how AUVs are currently being
+utilized; one consequence for instance, is any off-nominal condition
+results in a \emph{fail-safe} surface mode to radio for help. Further
+these approaches, words to the contrary are not only hard to maintain,
+are not 'adaptive' enough. Despite vigorous defense by Brooks early on
+\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
+require a model if not of the world then of the \emph{physics} of the
+vehicle when it interacts with the world.
+
+Persistent and early legitimate criticism of deliberation on the other
+hand also well articulated by Brooks
+\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
+was targeting the performance of in-situ deliberation. Terrestrial
+robotic platforms were well-known to frequently stop and ``plan''. In
+the marine robotics world, such a situation would simply not be
+permissible given the need to make continuous observations without
+stopping an under-articulated robotic platform such as an AUV. However
+algorithmic and implementational advances by the AI community over the
+years has produced a range of planners with demonstrated embedded
+capabilities \comment{cite}. Our work has been influenced by and in
+turn influenced a wide range of activities in the field of AI Planning
+and Plan Execution with demonstrated real-world and highly visible
+capabilities in the Space domain
+\cite{mus98,rajan00,ai-chang04,bresina05} for NASA. Lessons learned
+have been applied to the world of marine robotics which we bring
+together in this chapter. Important lessons that were derived from
+this rich legacy involved the following:
+
+\begin{itemize}
+
+\item Projecting plans (or using \emph{deliberation}) can and should
+  balance near-term objectives with end-term or even evolving goals.
+
+\item Dealing with time and resources is essential in real-world
+  planning.
+
+\item A rich representation that can deal with co-temporal events for
+  modeling complex systems is a must.
+
+\item Fast solvers for incremental causal reasoning can be built as a
+  basis for dealing with dynamic controllable events where replanning
+  is tightly integrated.
+
+\item Planning should be tightly inter-leaved with execution to enable
+  responsiveness to events which impact plan failure. This is also an
+  important factor leading to platform adaptivity.
+
+\item A model-based approach separating control formulation of the
+  platform which describes operational characteristics from the
+  domain. The controller can then be rigorously tested and certified
+  even as the domain model evolves to fit diverse applications.
+
+\end{itemize}
+
+We have developed, tested and deployed the Teleo-Reactive EXecutive
+(\texttt{T-REX}) an on-board adaptive control system that integrates
+AI based planning and probabilistic state estimation in a hybrid
+executive \cite{mcgann08a,mcgann08b,Py10}.  Probabilistic state
+estimation integrates a number of science observations to produce a
+likelihood that the vehicle sensors perceive a feature of
+interest. Onboard planning and execution enables adaptation of
+navigation and instrument control based on the probability of having
+detected such a phenomenon. It further enables goal-directed
+commanding within the context of projected mission state and allows
+for replanning for off-nominal situations and opportunistic science
+events. The framework in addition to being used on an AUV, is general
+enough be used for controlling a personal robot \cite{mcgann2009} and
+is deployed on a European planetary rover testbed
+\cite{goac11}. Probabilistic state estimation is discussed elsewhere
+\cite{mcgann08d} and out of scope for this chapter. Deliberation and
+reaction are integrated systematically over different temporal and
+functional scopes within a single agent and a single model that covers
+the needs of high-level mission management, low-level navigation,
+instrument control, and detection of unstructured and poorly
+understood phenomena. \rx is deployed on MBARI's Dorado AUV shown in
+Fig. \ref{fig:auv-fig}, which to the best of our knowledge is the only
+operational marine platform anywhere being used for routine scientific
+surveys with onboard plan synthesis.
+
+\begin{figure}[t]
+  \centering
+  \vskip-5pt
+  \includegraphics[scale=0.1]{figs/MBARI-AUV.jpg}
+  \caption{\small The MBARI \emph{Dorado} AUV on its support vessel
+    the R/V \emph{Zephyr}.}
+  \label{fig:auv-fig}
+  \vskip-0.3cm
+\end{figure}
+
+This chapter is organized as follows. We start with some foundational
+concepts in our Planning framework in Section
+\ref{sec:concepts}. \comment{add others} We conclude with Section
+\ref{sec:conclusion}.

### 2012-01-12 setobook.tex
Feww correction in order to ba able to compile
updates

@@
-\documentclass[12pt]{svmono}
+\documentclass[graybox,12pt]{svmult}
+
+% \documentclass[12pt]{svmono}
 
 \input{preamble}
+\setcounter{secnumdepth}{-1}
 
-\usepackage{graphicx}
+% \usepackage[agsm]{harvard}
 
-\usepackage[agsm]{harvard}
-
-\title{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
-\author{}
-\date{\small Monterey Bay Aquarium Research Institute \\
-7700 Sandholdt Rd.\\
-Moss Landing,  CA, 95039\\
-{\tt }}
+\title*{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
+\author{\normalsize Kanna~Rajan\inst{1} \and Fr\'ed\'eric Py\inst{1}
+  \and Javier Barreiro\inst{2}}
+\institute{Monterey Bay Aquarium Research Institute, California
+  \texttt{\{Kanna.Rajan,fpy\}@mbari.org}
+  \and
+  NASA Ames Research Center, Moffett Field, California
+  \texttt{javier.barreiro@nasa.gov}}
+% \date{\small Monterey Bay Aquarium Research Institute \\
+% 7700 Sandholdt Rd.\\
+% Moss Landing,  CA, 95039\\
+% {\tt }}
 
@@
 
-\begin{abstract}
-
-  We describe a general purpose Artificial Intelligence based control
-  architecture that implements the sense-plan-act paradigm. The
-  Teleo-Reactive EXecutive (T-REX) framework is deliberates about
-  future states, plans for actions and executes activities while
-  monitoring plans for anomalous conditions. It is unusual in being
-  operational and in routine use for operational oceanography,
-  providing a new tool for science to sample and observe the dynamic
-  coastal ocean. Plans are no longer scripted a priori but synthesized
+\abstract{We describe a general purpose Artificial Intelligence based
+  control architecture that incorporates in-situ decision making for
+  adaptation and control for autonomous underwater vehicles
+  (AUVs). The Teleo-Reactive EXecutive (\texttt{T-REX}) framework
+  deliberates about future states, plans for actions and executes
+  generated activities while monitoring plans for anomalous
+  conditions. Plans are no longer scripted a priori but synthesized
   onboard with high-level directives instead of low level
@@
   ability to deal with time and resources coupled with Machine
-  Learning techniques for event detection, allows balancing shorter
+  Learning techniques for event detection allows balancing shorter
   term benefits with longer term needs, an important need as AUV
   hardware becomes more robust allowing persistent ocean sampling and
-  observation.
-
-%  % Traditional approaches to commanding marine robots have relied on
-%  %  pre-scripted plans with a priori defined waypoints. These have
-%  %  served the oceanographic community well in enabling robots to
-%  %  cost-effectively perform surveys traditionally undertaken by
-%  %  ship-board observations. However, in dynamic coastal environments
-%  %  where is substantial interaction between atmospheric, oceanographic,
-%  %  estuarine/riverine and land-sea interaction processes requires a
-%  %  robot to be adaptive and receptive to opportunistic
-%  %  science. Sustained use of such autonomous platforms in military
-%  %  applications have also pushed.
-
-\end{abstract}
+  observation. \rx is in routine operational use at MBARI,
+  providing scientists a new tool to sample and observe the dynamic
+  coastal ocean.}
 
 \input{intro.tex}
+
+\input{concepts.tex}
 
@@
 
-\input{concepts.tex}
-
 \input{results.tex}
@@
 
-\input{conclde.tex}
+\input{conclude.tex}
 
@@
 
+\bibliographystyle{IEEEtran}
 \bibliography{references}

### 2012-01-12 trunk/arch.tex
- Added the cls file which was missing in svn - Initial draft with plan on the architecture part -- specifically a rough presentation of the synchronization flaws and their resolution process inside the deliberative reactor

@@
-\section{\texttt{T-REX} Architecture}
+\section{Planning and execution in the \texttt{T-REX} architecture}
 \label{sec:arch}
+
+\subsection{The T-REX architecture}
+\label{sec:arch:trex}
+
+{\em This section presents quickly T-REX architecture but move
+  quickly the focus on a single reactor and gross concepts on the
+  architecture:
+  \begin{itemize}
+  \item For t-trex the whole decision and control problem is reduced
+    around the state variables (or timeline construct). 
+  \item This reduce the execution tracking problem for each reactor on a state
+    identification problem (deduce internal state from the external
+    state), and the control problem to goal posting which will trigger
+    deliberation on the owner of the corresponding timeline. 
+  \item the architecture itself see each reactor as a black box and
+    therefore each reactor can implement its own mechanism in order to
+    resolve both synchronization and deliberation. Still we provide a
+    reactor based on the europa framework that leverage the automated
+    planning capabilities in order to do model based planning and
+    execution with a rich representation of resources and time.
+  \end{itemize}}
+
+\subsection{The Europa deliberative reactor}
+\label{sec:arch:europa}
+
+In T-REX architecture reactors have two core functions;
+synchronization where the main focus is to track and refine the
+evolution of the state of the world at any single tick, and
+deliberation which focus on providing a plan to the agent in order to
+complete the reactor {\em Internal} goals. These two processes are
+intertwined in the sense that one influences the outcome of the
+other.
+
+In  this section we present how both processes are implemented using
+the europa framework. We first present each of them independently by
+looking to their specific focus. We then develop how both processes
+are not only interleaved (as deliberation can take several tick while
+synchronization eneds to happen at every single tick) but also discuss
+on how each of these is impacting the other and how this intertwining 
+of the two result on an emergent behavior of the reactor by allowing
+the system to efficiently adapts its plan that can then be dispatched
+as objectives for the reactors it depends on in a seamless flow.
+
+\subsubsection{Synchronization identify internal state evolution}
+\label{sec:arch:synch}
+
+For the sake of simplicity we introduce first how a reactor can track
+and identify its state in the context where it does not have any
+compelling need to deliberate. Assume here that we have a reactor that
+have no future goal. {\em Need to develop the reason why it is a
+  necessity: primarily to provide its internal state to whoever
+  observe it but als simply to ensure that its current representation
+  of the world is up to date and still consistent}
+
+In the europa based reactor we introduce this requirement as a new
+type of flaw for the europa framework. This flaw enforces at the
+reactor to identify fully its internal state for the current
+tick. By using this new flaw we can describe the synchronization
+process as this general sequence :
+\begin{enumerate}
+\item Integrate the external state as provided by the owner of each
+  external timeline into the plan database.
+\item Propagate this information in the plan database following the
+  model $\mathcal{M}$ of this reactor.
+\item Resolve the current state value of each internal timelines.
+\end{enumerate}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
+  \caption{Ilustration of the synchronization in a reactor. The
+    reactor receive new observations when they are produced by the
+    owner(s) of its internal timlines. Meanwhile at every tick it
+    identifies the potential flaw of himself not having identify its
+    state for the current tick and will try to resolve it according to
+    its model $\mathcal{M}$} 
+  \label{fig:synch:flaw}
+\end{figure}
+
+The resolution of internal state flaw can be resolved using one of the
+following choices which are evaluated in the given order:
+\begin{enumerate}
+\item Extend the previous state value to end after this tick (ie
+  restrict its end time to $[\tau+1, \infty)$). 
+\item Start the next active token in the timeline and
+  attempt to start it now (ie restrict its start time to the single
+  value $\tau$).
+\item Create and insert a new token in this timeline that will start
+  at the current tick $\tau$ (attempt this for each possible token
+  type for this timeline if necessary). 
+\end{enumerate}
+
+All of these choices are evaluated sequentially until a consistent
+solution with no more flaw for this tick is identified. In order to do
+so we need to make the assumption that the current state value of an
+internal timeline do not depend on the future or more accurately
+that any choices made during this synchronization will not lead to a
+future inconsistency (meaning no possible solution) for a future
+synchronization. Such assumption deeply impact the set of possible
+domains one the reactor can support while remaining complete. Take for
+example the following model rule where {\tt Vehicle} is {\em Internal}
+and {\tt Command} is {\em External} to the reactor.
+\begin{verbatim}
+Vehicle::Dive {
+  contains(Command.Descend descend);
+  descend.start = start + 2;
+}
+\end{verbatim}
+
+Such model present the issue that the reactor needs at one point to
+take the decision to start the token {\tt Vehicle.Dive} and by doing
+so it assumes that in {\em exactly} 2 ticks in the future the {\tt
+  Command} timeline will change its state to {\tt Command.Descend}. If
+this observation does not occur at the given time this implies that
+the reactor estimation was wrong leading him to an inconsistent view
+of the world. Conversely the reactor can decide to not switch this
+timeline to the {\tt Vehicle.Dive} state and observe 2 ticks in the
+future that the {\tt Command.Descend} state value. In this case it
+would have missed the opportunity to set its {\tt Vehicle} state to
+the correct value which may similarly lead (depending on the rest of
+the model) to a similar inconsistency it won't be able to resolve (as
+TREX enforces that  the past is monotonic).
+
+While this limitation of the system is important to take into account,
+it is also important to note that the model snippet we gave here --
+while formally acceptable in the general planning problem -- is also
+illustrating a bad design for execution in a situated agent. Indeed,
+in this model we are tying the current decision to a future outcome
+that is {\em External} to the reactor and therefore not necessarily
+controllable by this reactor. Therefore we are entailing a part of the
+future to past decisions which can lead the system to a dead-end
+situation where the model cannot find a correct solution that both do
+not conflict with what we claimed in the past and still allow the
+planner to find its current {\em Internal} state.
+
+\subsubsection{Deliberation : planning for future state evolution}
+\label{sec:arch:plan}
+
+\subsubsection{Intertwining synchronization and execution }
+\label{sec:arch:intertwine}
 
@@
 % these principles aid in software engineering. Show T-REX block diagram.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-01-12 arch.tex
- rewrote the topic of the synchronization flaw figure

--- 2012-01-12 trunk/arch.tex
@@
   \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
-  \caption{Ilustration of the synchronization in a reactor. The
+  \caption{Ilustration of the synchronization flaws in a reactor. The
     reactor receive new observations when they are produced by the
-    owner(s) of its internal timlines. Meanwhile at every tick it
-    identifies the potential flaw of himself not having identify its
-    state for the current tick and will try to resolve it according to
-    its model $\mathcal{M}$} 
+    owner(s) of its internal timelines. The line after the last token
+    of each timeline represent the domain of possible values for the
+    end of this token. At every tick $\tau$ the reactor needs to
+    integrate the {\em External} state information it received and --
+    based on its model $\mathcal{M}$ -- resolve its {\em Internal}
+    state that will then be provided by the architecture to other
+    reactors using these state variables.}
   \label{fig:synch:flaw}

### 2012-01-12 trunk/concepts.tex
Added a paper reference for willow garage implementation on pr2

@@
-\section{Key Concepts}
+\section{Foundational Concepts}
 \label{sec:concepts}
 
-% This is the core of the chapter and talks about Planning and
-% Execution. Subsections with details on EUROPA as well as on T-REX
-% synchronization which will have to be different from the sutton
-% book. Most of our effort will be on this section with multiple
-% sub-sections.
+Our system, \texttt{T-REX} (or the Teleo-Reactive EXecutive) is an
+adaptive, artificial intelligence based controller and provides
+general framework for building reasoning systems for real-world
+autonomous vehicles. At MBARI \texttt{T-REX} is used for AUV control;
+another instantiation of the system is being used for control of a
+terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The development of
+\texttt{T-REX} has been targeted at surveying a number of
+oceanographic features which are dynamic and unpredictable
+spatio-temporally. Typically this requires our AUVs to balance the
+goals of coverage spatially while opportunistically following features
+of scientific interest and to do so while being operationally aware of
+its own limitations in terms of resources (typically the battery state
+of charge) and overall proximity to other observational assets for
+obtaining scientific ground-truth. We are therefore interested in
+representational frameworks that allow robots to pursue long-term
+objectives while choosing short-term gain and can gracefully deal with
+exogenous or endogenous change.
+
+To enable this responsiveness to external observations, the agent has
+to be able to synthesize plans insitu and to re-plan. We use a
+temporal constraint-based planner EUROPA$_2$ with a demonstrated
+legacy of having flown on NASA space missions
+\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
+innovations for AUV adaptation: the use of flexible plan
+representations, compositional control with the use of partitioned
+networks and off-line learning to inform insitu environmental state
+estimation. In this section we detail some of the key concepts
+relevant to flexible plan representation. Section \ref{sec:arch} has
+details the \texttt{T-REX} partitioned architecture and estimation
+driven vehicle adaptation.
+
+% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
+% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
+% for the Mars Exploration Rovers \cite{bresina05}.
+
+EUROPA$_2$ uses a domain model written in a declarative language
+(NDDL: New Domain Description Language), together with initial
+conditions and goals also specified in NDDL, to construct a set of
+temporal relations that must be true at the start time. These models
+include assertions about the physics of the vehicle, i.e how it
+responds to external stimulus and internally driven goals. By
+propagating these relations forward using Simple Temporal Networks
+\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
+a set of conditions that should be true in the future, where some of
+these conditions will correspond to actions the agent must take. The
+planner can backtrack and try another path during search if a goal
+cannot be reached while being capable of discarding unachievable
+goals.
+
+Traditionally robot execution has relied on dispatching commands at
+precise times. Such linear sequences of precisely timed commands give
+no ability to adjust execution on the basis of sensory
+information. Although some commands can issue tests on sensor
+readings, these tests have the objective of verifying whether expected
+execution conditions are occuring. If not, the state of the system is
+declared off-nominal and execution of the sequence of commands is
+interrupted. More recently, executives have been proposed and
+implemented that significantly broaden the way robots can be commanded
+\cite{mus98,Alami:1998p820}. For example, the Remote Agent executive
+interpreted a \textit{temporally flexible plan} which represents each
+start time as a variable and contains an explicit network of bounded
+delay constraints between such variables.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\caption{\small Tokens with flexible temporal intervals and parametric
+  constraints between tokens. This example shows the triggering of a
+  water sampler based on a feature threshold while the vehicle
+  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
+  start \& end times.}
+\label{fig:flex-timelines}
+\vskip+0.1cm
+\end{figure}
+
+EUROPA$_2$ uses a \emph{state variable} representation to describe the
+evolution of state over time. The instantiated history of such state
+variable evolution over a temporal horizon we call \emph{timelines}
+and which represent a single thread in the execution of a concurrent
+system. At any given time each thread can execute a single procedure.
+Thus each timeline consists of a sequence of procedures which
+encapsulate and describe state evolution; we call these instantiated
+atomic entities \emph{tokens}.  A token therefore describes a
+procedure invocation, the state variables on which it can occur, the
+parameter values of the procedure, and the time values defining the
+interval. We allow encapsulation of uncertainty within these tokens
+with a range of start and end times and parameters, all of which are
+encoded as variables. A constraint solver in turn manipulates these
+variables defined in a EUROPA$_2$ domain model. For example, consider
+a scientific need to take a water samples 100 meters from a hotspot
+while an AUV is performing a Yo-Yo in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time the most appropriate one for the actual execution
+conditions. The fact that constraints are explicitly represented
+ensures that through constraint propagation the executive will respect
+global limits expressed in the plan (e.g., don't start a task until a
+certain condition has been satisfied but still satisfying some global
+deadline). Such flexibility is critical when dealing with dynamic
+ocean conditions where precise timing of a robotic action might be
+indeterminate. Further, the advantage of flexibility can be contrasted
+with the consequences of the intrinsic inflexibiliy of traditional
+command sequences. Because they are inflexible, sequences must
+necessarily be designed considering worst case scenarios.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\caption{\small Temporal relations defined within the planner are
+  based on \texttt{Allen Algebra} relations shown above.}
+\label{fig:allen-algebra}
+\vskip-0.3cm
+\end{figure}
+
+Each of the token variables, including the parameter variables, has a
+domain of values assigned to it.  The variables may also participate
+in constraints that specify which value combinations are valid.  To
+allow the specification of multiple values, e.g, to express a range of
+possible start times, variables are used to specify parameter, start
+and end time values for a token.  As a result, a token $T$ is a tuple
+$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
+variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
+$s\leq e$) are variables encoded as flexible token start and end time
+points. Consequently, partial plans represent a \textit{range of
+  possible solutions} rather than an explicit definition of a single
+trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
+of a part of such plans with tokens representing constraints on
+flexible temporal intervals. Systematic computation is reflected by
+the use of the \texttt{Allen Algebra} \cite{allen84} temporal
+relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
+interval arithmetic in the EUROPA$_2$ domain model that allows
+encoding of temporal constraint relationships between tokens on the
+same or other concurrent timelines. Details of propagation algorithms
+or of interval logic are beyond the scope of this paper.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
+\caption{\small Timeline evolution can be described by simple finite
+  state transitions where temporal relationships describe the arc
+  transition and temporal extant of state are token
+  durations. Constraints here are for \emph{meets} or its equivalent
+  inverse \emph{met\_by}.}
+\label{fig:FSM}
+\vskip-0.3cm
+\end{figure}
+
+In describing the evolution of state via timelines, transitions within
+a timeline can be described by simple finite state
+machines. Fig. \ref{fig:FSM} for example describes a timeline
+evolution for an abstract timeline entity describing navigation. Arc
+transitions represent temporal constraints and token duration are
+represented by temporal extant of state. The entire vehicle can then
+be modeled as a number of state transitions with concurrency for each
+of the modeled state variables defined a priori. However this simple
+view is often complicated when temporal constraints between state
+variables are articulated in our domain models. If there is a
+criticism of this approach, modeling then, requires a proper balance
+between abstraction, state variable description and concurrency
+described by the inter-connectedness between such finite state
+machines.
+
+\begin{figure*}
+\centering 
+\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
+\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
+\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
+\caption{\small An illustrative plan synthesis example with concurrent
+  timelines. Each timeline is an instantiation of a subsystem tracked
+  by the planner. Tokens describe the instantiated state of subsystem
+  at a particular time instance and enforce temporal and parametric
+  constraints. Abstract goals are decomoposed into successively less
+  abstract tokens and instantiated in co-temporal timelines using
+  \texttt{Allen Algebra} relations. All tokens represent flexible
+  start/end times. \ref{fig:plan-evolve1} shows an initial state
+  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
+  \label{fig:Plan-evolve}
+  \vskip-5pt
+\end{figure*}
+
+Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
+timelines which track \texttt{Goal, Path, Navigation} and
+\texttt{Command} state over time. Tokens on the goal timeline indicate
+a survey within a bounded volume and a min/max depth envelope. As the
+Volume Survey sub-goals on the \texttt{Path} timeline, the initial
+token on that timeline gets \emph{squeezed} to have dependencies of
+the goal token instantiated. Initially this dependency are the tokens
+\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
+coverage. The first of these in turn generates a sub-goal on the same
+timeline to go to the next intermediate waypoint, which is illustrated
+in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
+figure) ultimately generate all the tokens to completion for this
+timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
+in plan generation and shows additional subgoals on the \texttt{Path,
+  Navigation} and \texttt{Command} timelines. Each token can thus
+trace its causality when instantiated in the plan. The domain model
+provides source of these temporal and parametric dependencies in the
+form of rules.
+
+\begin{minipage}[c]{\textwidth}
+
+% \framebox[\textwidth][t]{
+Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
+$\Rightarrow$ \\
+met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
+starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
+\ldots{} \\
+ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
+meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
+\end{minipage}
+
+% \subsubsection{Partitioned Control}
+
+% \subsubsection{State Estimation}
+
+% Even as automated reasoning approaches have the ability to dynamically
+% retarget the vehicle, estimating environmental signals of interest (or
+% their proxies) is important to be able to enable opportunistic science
+% in the water-column. In \texttt{T-REX} feature recognition revolves
+% around a Hidden Markov Model (HMM) \cite{rabiner86} which is encoded
+% directly within the unified representational and computational
+% framework of a reactor. HMMs are useful since the stochastic nature of
+% these models can correlate the type of features we want to detect with
+% the sensor observations. Online targeted sensor data is classified by
+% apriori generated cluster data which in turn determines the
+% probability of having seen the feature of interest. Together with
+% posterior probability, the HMM generates a probability of being within
+% the target feature if it dominates the distribution. If other sampling
+% conditions are satisfied (for instance sampling proximity), a
+% constraint is activited in the model which can trigger a water sampler
+% and also preserve the memory of sampling to alter a future transect
+% resolution.
+
+% The HMM is customized for the feature of interest and is built offline
+% using clustering techniques from large data sets; we use Kohonen's
+% Self Organizing Maps \cite{kohonen}. We use a semi-supervised learning
+% method \cite{zhu05} to extract an environmental model that make uses
+% of both labeled and unlabeled data for training. Online,
+
+% The sensor data for classification depends on the
+% feature of interest; for detecting INLs we use backscattering data
+% from a HydroScat and for tracking riverine plumes to track fertilizer
+% runoff, we use an ISUS Nitrate sensor.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "ieee-ram09"
+%%% End:

### 2012-01-13 trunk/setobook.tex
updates

@@
-\documentclass[12pt]{svmono}
+\documentclass[graybox,12pt]{svmult}
+
+% \documentclass[12pt]{svmono}
 
 \input{preamble}
+\setcounter{secnumdepth}{-1}
 
-\usepackage{graphicx}
+% \usepackage[agsm]{harvard}
 
-\usepackage[agsm]{harvard}
-
-\title{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
-\author{}
-\date{\small Monterey Bay Aquarium Research Institute \\
-7700 Sandholdt Rd.\\
-Moss Landing,  CA, 95039\\
-{\tt }}
+\title*{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
+\author{\normalsize Kanna~Rajan\inst{1} \and Fr\'ed\'eric Py\inst{1}
+  \and Javier Barreiro\inst{2}}
+\institute{Monterey Bay Aquarium Research Institute, California
+  \texttt{\{Kanna.Rajan,fpy\}@mbari.org}
+  \and
+  NASA Ames Research Center, Moffett Field, California
+  \texttt{javier.barreiro@nasa.gov}}
 
@@
 
-\begin{abstract}
-
-  We describe a general purpose Artificial Intelligence based control
-  architecture that incorporates in-situ decision making for
+\abstract{We describe a general purpose Artificial Intelligence based
+  control architecture that incorporates in-situ decision making for
   adaptation and control for autonomous underwater vehicles
@@
   hardware becomes more robust allowing persistent ocean sampling and
-  observation. \rx is in operational and routine use at MBARI,
+  observation. \rx is in routine operational use at MBARI,
   providing scientists a new tool to sample and observe the dynamic
-  coastal ocean.
-
-%  % Traditional approaches to commanding marine robots have relied on
-%  %  pre-scripted plans with a priori defined waypoints. These have
-%  %  served the oceanographic community well in enabling robots to
-%  %  cost-effectively perform surveys traditionally undertaken by
-%  %  ship-board observations. However, in dynamic coastal environments
-%  %  where is substantial interaction between atmospheric, oceanographic,
-%  %  estuarine/riverine and land-sea interaction processes requires a
-%  %  robot to be adaptive and receptive to opportunistic
-%  %  science. Sustained use of such autonomous platforms in military
-%  %  applications have also pushed.
-
-\end{abstract}
+  coastal ocean.}
 
 \input{intro.tex}
+
+\input{concepts.tex}
 

### 2012-01-16 europa.tex
outline for section on EUROPA

1. EUROPA

EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  Automated Planning can be reformulated as a constraint satisfaction problem (TODO: ref). EUROPA is an extension of that idea, however, unlike other approaches where the entire planning problem is translated into a CSP and then solved using traditional CSP methods, EUROPA uses a constraint reasoning engine as a fundamental building block, and enriches it with other architectural components that allow a more direct mapping for scheduling and planning problems, which in turn allows easier development of efficient solving methods.

TODO: motivate with a few examples of problems that can be solved in CSP, Scheduling and Planning.

We begin with a brief overview of Constraint Satisfaction Programming.

1.1 Constraint Satisfaction Programming
TODO: talk about variables, domains, constraints, arc consistency and possibly computational complexity.

1.2 EUROPA's Architecture

The following figure displays the main architectural components in EUROPA and their relationships:

(TODO: add figure)

- The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables. It also provides an efficient arc consistency mechanism (TODO: ref to AC-3). The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.

- The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does. Also manages tokens, which are a mechanism to group variables to represent temporally scoped state. Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can (TODO: provide snippet?)

- The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)

- The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)

- API Layer: makes all the services available so that EUROPA can be used to build applications.

- Extension Modules bundled with EUROPA: Temporal Network, NDDL, ANML.

1.3 Modeling

TODO: introduce model for Shopping or Rover example, link to architectural components.

1.4 Inference

TODO: explain how inference is used to detect violations and flaws.
- overview of the types of flaws and violations EUROPA detects.

1.5 Search

TODO: 
- explain how EUROPA's built-in solver uses flaws to search for a plan.
- explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.

1.6 Future Work?
explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).

### 2012-01-18 trunk/europa.tex
added some meat to CP and Search sections.

1. EUROPA

EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  Automated Planning can be reformulated as a constraint satisfaction problem (TODO: ref). EUROPA is an extension of that idea, however, unlike other approaches where the entire planning problem is translated into a CSP and then solved using traditional CSP methods, EUROPA uses a constraint reasoning engine as a fundamental building block, and enriches it with other architectural components that allow a more direct mapping for scheduling and planning problems, which in turn allows easier development of efficient solving methods.

TODO: motivate with a few examples of problems that can be solved in CSP, Scheduling and Planning.

We begin with a brief overview of Constraint Programming.

1.1 Constraint Programming

Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints. A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].

	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN. Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains. Given a defined conjunctive set of constraints on the variables: C={c1(x1,..,xN), ..., cK(x1,..,xN)}, the objective is to find one or more value assignments to V where all constraints are satisfied.

To solve a problem, CP uses logical inference to perform 2 operations:

	- Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2

	- Domain reduction: To infer a valid set of values for a variable. For example, for discrete variables, the constraints all_different(x1,x2,x3) and xi>=0, xi <=4. if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}

CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95]. Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.

TODO: talk about arc consistency algorithms, computational complexity and solvers.

1.2 EUROPA's Architecture

The following figure displays the main architectural components in EUROPA and their relationships:

(TODO: add figure)

- The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables. It also provides an efficient arc consistency mechanism (TODO: ref to AC-3). The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.

- The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does. Also manages tokens, which are a mechanism to group variables to represent temporally scoped state. Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can (TODO: provide snippet?)

- The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)

- The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)

- API Layer: makes all the services available so that EUROPA can be used to build applications.

- Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.

1.3 Modeling

TODO: introduce model for Shopping or Rover example, link to architectural components.
explain variable, object, predicate and action types and instances, constraints, goals and facts.

1.4 Inference

TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
- overview of the types of flaws and violations EUROPA detects.

1.5 Search

Now we have all the elements in place so that an automated problem solver can be created. Let's recap what those elements are:

- A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
- A problem instance (also called initial state) that consists of:
	- Variable and Object instances that exist for the entire planning horizon.
	- Temporally scoped predicate and action instances.
	- Temporally scoped goals.

All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.

EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals. The operations to refine the partial plan PP at any time are:
	1- Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
	2- Select one such flaw
	3- Select a resolver for the flaw
	4- Refine PP by applying the resolver
	5- If an inconsistency is found, try another resolver.
	6- If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.

This Plan Space Planning algorithm is translated into EUROPA's representation as follows:

The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances. Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:

	1. Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
	2. Open Condition: an open condition is an inactive token. Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
	3. Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects. Recall for example that a token may belong to objects (e.g. Timelines) which require a total order over their tokens. If any 2 tokens could possibly overlap (though not necessarily), then they pose a threat to each other in terms of achieving an extension of the current partial plan which is complete and consistent. Similarly, threats may arise where tokens share a common resource and their current state might yield extensions of the current partial plan which are inconsistent. Threats are resolved by imposing ordering constraints among tokens.

Open Conditions and Threats allow flaw detection and resolution at a higher-level of abstraction (i.e. in terms of objects and tokens) than that of simply binding variables as is common in Constraint Satisfaction Problems. This is advantageous when one applies heuristics for ordering choices since it provides a richer context in which to make decisions. Furthermore, it aids in reducing the amount of work done by a solver so that only the necessary refinements are made, otherwise leaving the partial plan with maximum flexibility. For example, one can omit unbound variables which are time-points of tokens since threats will force a solver to impose restrictions on these variables based on the semantics of the objects to which their tokens apply. Thus the planning process may yield partially-ordered plans for which all possible extensions are provably valid.

The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking. EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.

// Function to solve a partial plan by successive refinement
// until all flaws are resolved.
bool solve(PartialPlan plan)
{
	// Propagate the constraints to test for inconsistency. If found to be
	// inconsistent, then we can return false since this is a dead-end i.e. no refinements
	// to p can yield a consistent plan.

	if(isInconsistent(plan))
		return false;

	// Non-deterministically choose a flaw from the set of available flaws.
	Flaw flaw = chooseFlaw(plan);

	// If there are no flaws, then p is complete and we can terminate with success.
	if(flaw == NULL)
		return true;

	// Otherwise we formulate a decision point which is a branch in the search space. Each
	// choice is a particular refinement operation and the DecisionPoint collects all possible
	// refinement operations for the given flaw.
	DecisionPoint decision = makeDecisionPoint(flaw, plan);

	// Continue as long as we have something to try
	while(decision.hasNext()) {
		// A new partial plan is obtained by application of a refinement operator. Note that the ordering
    		// over refinement operators to select is a non-deterministic step.
		PartialPlan pp = decision.executeNext();

		// Recursive call to solve the new planning problem. If successful, then we are done.
		if(solve(pp))
			return true;
		else // Otherwise, retract the last refinement operation
			d.undo();
  	}

	// If we arrive here, then we have exhausted all options to resolve the flaw, including the case where
  	// no options were available initially. Thus the problem cannot be solved.
  	return false;
}

The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise. This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily. Dead-ends in the search are discovered through constraint propagation. Constraint propagation is a vehicle for evaluating the consistency of a partial plan and also for filtering infeasible values from consideration prior to commitment, allowing in some cases a strong look-ahead capability which is essential for tractable search. Consistency testing is initiated by the isConsistent procedure. The algorithm permits a heuristically controlled search by applying orderings for chooseFlaw and makeDecisionPoint. It results in a chronologically-backtracking, search.

It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?

TODO: 
- explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.

1.6 Future Work?
explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).

The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space

### 2012-01-18 setobook.tex
update

@@
 
-\title*{\textbf{\sc Deliberative Autonomy for Marine Robotics}}
+\titlerunning{Towards Deliberative Control of Autonomous Robots}
+\authorrunning{K.Rajan et. al.}
+
+\title*{\textbf{\sc Towards Deliberative Control of Autonomous Robots}}
 \author{\normalsize Kanna~Rajan\inst{1} \and Fr\'ed\'eric Py\inst{1}
@@
   \texttt{javier.barreiro@nasa.gov}}
-% \date{\small Monterey Bay Aquarium Research Institute \\
-% 7700 Sandholdt Rd.\\
-% Moss Landing,  CA, 95039\\
-% {\tt }}
 
@@
 
-The authors are funded by a block grant from the David and Lucile
-Packard Foundation to MBARI.
+Authors Rajan and Py are funded by a block grant from the David and
+Lucile Packard Foundation to MBARI.
 

### 2012-01-18 europa.tex
more detail on search.

@@
 
-We begin with a brief overview of Constraint Satisfaction Programming.
+We begin with a brief overview of Constraint Programming.
 
-1.1 Constraint Satisfaction Programming
-TODO: talk about variables, domains, constraints, arc consistency and possibly computational complexity.
+1.1 Constraint Programming
+
+Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
+A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].
+
+	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN.
+Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains.
+Given a defined conjunctive set of constraints on the variables: C={c1(x1,..,xN), ..., cK(x1,..,xN)}, the objective is to find one or more value assignments to V where all constraints are satisfied.
+
+To solve a problem, CP uses logical inference to perform 2 operations:
+
+	- Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
+
+	- Domain reduction: To infer a valid set of values for a variable.
+For example, for discrete variables, the constraints all_different(x1,x2,x3) and xi>=0, xi <=4.
+if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}
+
+CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
+Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other “primitive” CP constraint, such as <= or >=.
+
+TODO: talk about arc consistency algorithms, computational complexity and solvers.
 
@@
 
-- Extension Modules bundled with EUROPA: Temporal Network, NDDL, ANML.
+- Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
 
@@
 TODO: introduce model for Shopping or Rover example, link to architectural components.
+explain variable, object, predicate and action types and instances, constraints, goals and facts.
 
@@
 
-TODO: explain how inference is used to detect violations and flaws.
+TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
 - overview of the types of flaws and violations EUROPA detects.
@@
 
+Now we have all the elements in place so that an automated problem solver can be created. Let's recap what those elements are:
+
+- A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
+- A problem instance (also called initial state) that consists of:
+	- Variable and Object instances that exist for the entire planning horizon.
+	- Temporally scoped predicate and action instances.
+	- Temporally scoped goals.
+
+All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.
+
+Given that temporal intervals may extend indefinitely in both positive and negative directions, for problems with a temporal dimension it is also common to specify a Planning Horizon when invoking a solver.
+The Planning Horizon is the temporal interval that a problem solver has to worry about, any decisions that fall completely outside of the Planning Horizon can be ignored.
+This is very useful for recurrent activities, consider for instance a model where a medical checkup has to occur every 3 months, the user can model a recurring activity and then use the Planning Horizon to ensure that a finite number of medical checkup activities are generated as part of a plan.
+This is also useful to ignore activities that may happen too far in the future or in the past to be relevant for a particular plan, this way the user can ensure that the solver is making as few commitments as necessary.
+
+EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals.
+The operations to refine the partial plan PP at any time are:
+	1- Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
+	2- Select one such flaw
+	3- Select a resolver for the flaw
+	4- Refine PP by applying the resolver
+	5- If an inconsistency is found, try another resolver.
+	6- If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
+
+This Plan Space Planning algorithm is translated into EUROPA's representation as follows:
+
+The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances.
+Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:
+
+	1. Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
+	2.
+Open Condition: an open condition is an inactive token.
+Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
+Open Conditions can be resolved in three ways:
+		- Merging: A free activity is merged with a matching activity already in the plan.
+An activity a is said to match an activity a?
+if a and a?
+unify and the temporal constraints involving a are satisﬁed by a?.
+Thus,  a and a?
+can be considered the same activity and we do not need to introduce a in the plan.
+Consequently, the compatibilities associated with a are not ﬁred, because they have been already triggered when a?
+was introduced in the plan.
+		- Activation We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciﬁc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
+This results in both an ordering ﬂaw, corresponding to the just activated activity, and a number of open condition ﬂaws, corresponding to the new subgoal activities.
+		- Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
+
+	3.
+Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
+Recall for example that a token may belong to objects (e.g.
+Timelines) which require a total order over their tokens.
+If any 2 tokens could possibly overlap (though not necessarily), then they pose a threat to each other in terms of achieving an extension of the current partial plan which is complete and consistent.
+Similarly, threats may arise where tokens share a common resource and their current state might yield extensions of the current partial plan which are inconsistent.
+Threats are resolved by imposing ordering constraints among tokens.
+
+Open Conditions and Threats allow flaw detection and resolution at a higher-level of abstraction (i.e.
+in terms of objects and tokens) than that of simply binding variables as is common in Constraint Satisfaction Problems.
+This is advantageous when one applies heuristics for ordering choices since it provides a richer context in which to make decisions.
+Furthermore, it aids in reducing the amount of work done by a solver so that only the necessary refinements are made, otherwise leaving the partial plan with maximum flexibility.
+For example, one can omit unbound variables which are time-points of tokens since threats will force a solver to impose restrictions on these variables based on the semantics of the objects to which their tokens apply.
+Thus the planning process may yield partially-ordered plans for which all possible extensions are provably valid.
+
+The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
+EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
+
+// Function to solve a partial plan by successive refinement
+// until all flaws are resolved.
+bool solve(PartialPlan plan)
+{
+	// Propagate the constraints to test for inconsistency. If found to be
+	// inconsistent, then we can return false since this is a dead-end i.e. no refinements
+	// to p can yield a consistent plan.
+
+	if(isInconsistent(plan))
+		return false;
+
+	// Non-deterministically choose a flaw from the set of available flaws.
+	Flaw flaw = chooseFlaw(plan);
+
+	// If there are no flaws, then p is complete and we can terminate with success.
+	if(flaw == NULL)
+		return true;
+
+	// Otherwise we formulate a decision point which is a branch in the search space. Each
+	// choice is a particular refinement operation and the DecisionPoint collects all possible
+	// refinement operations for the given flaw.
+	DecisionPoint decision = makeDecisionPoint(flaw, plan);
+
+	// Continue as long as we have something to try
+	while(decision.hasNext()) {
+		// A new partial plan is obtained by application of a refinement operator. Note that the ordering
+    		// over refinement operators to select is a non-deterministic step.
+		PartialPlan pp = decision.executeNext();
+
+		// Recursive call to solve the new planning problem. If successful, then we are done.
+		if(solve(pp))
+			return true;
+		else // Otherwise, retract the last refinement operation
+			d.undo();
+  	}
+
+	// If we arrive here, then we have exhausted all options to resolve the flaw, including the case where
+  	// no options were available initially. Thus the problem cannot be solved.
+  	return false;
+}
+
+The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
+This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.
+Dead-ends in the search are discovered through constraint propagation.
+Constraint propagation is a vehicle for evaluating the consistency of a partial plan and also for filtering infeasible values from consideration prior to commitment, allowing in some cases a strong look-ahead capability which is essential for tractable search.
+Consistency testing is initiated by the isConsistent procedure.
+The algorithm permits a heuristically controlled search by applying orderings for chooseFlaw and makeDecisionPoint.
+It results in a chronologically-backtracking, search.
+
+It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?
+
 TODO: 
-- explain how EUROPA's built-in solver uses flaws to search for a plan.
 - explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.
-
-1.6 Future Work?
-explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).

### 2012-01-18 concepts.tex
updates

@@
 another instantiation of the system is being used for control of a
-terrestrial personal robot \cite{pr2}. The development of
+terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The development of
 \texttt{T-REX} has been targeted at surveying a number of
@@
 legacy of having flown on NASA space missions
-\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
+\cite{jonsson00,bresina05}. Our autonomy architecture brings three key
 innovations for AUV adaptation: the use of flexible plan
@@
 driven vehicle adaptation.
-
-% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
-% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
-% for the Mars Exploration Rovers \cite{bresina05}.
 
@@
 implemented that significantly broaden the way robots can be commanded
-\cite{mus98,Alami:1998p820}. For example, the Remote Agent executive
+\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
 interpreted a \textit{temporally flexible plan} which represents each

### 2012-01-19 trunk/related.tex
updates

\section{Background and Related Work}
\label{sec:related}

% Refer to past work in AUV autonomy as well as robotic autonomy in
% general. This is general purpose and should be a good literature survey
% for robotics and AI.

A distinct problem in Marine Robotics has been the use of the
overloaded term 'autonomy'. Not only does the notion transcend
different disciplines within engineering in this domain (e.g. the
Control Engineering community sees it distinctively from those in AI)
but users of marine platforms in oceanography as well as in maritime
defense conflate the methodology of use (tethered vs. untethered) with
their control. In this chapter, our notion of 'autonomy' deals
explicitly not only with the notion of dealing with the \emph{sense,
  plan and act} paradigm, but also the need to perform
\emph{computational search} between different action outcomes, an idea
central to the field of AI. Consequently, our literature survey here
is targeted towards a more focused and (to this chapter) relevant part
of the field of computation.

With few exceptions, most AUV control systems have been variations of
reactive approaches.  Typically, waypoint-based pre-defined mission
designs are uploaded to the AUV; specialized code fragments called
\emph{behaviors} are designed for the specific mission and a choice of
behaviors for the mission are used on the computational stack
\cite{bellingham94}. Swapping in and out of these behaviors using
conditionals or cleverer and quantitative approaches such as
\cite{Benjamin:2004} forms the basis for adaptation and safety in the
vehicle. Where such swapping does not aid adaptation, individual
behaviors are tweaked to generate some measure of responsiveness to
the surrounding environment \cite{yanwu08} to generate interesting
observations for oceanographers, yet technically naive in terms of
scalability and systematicity. \cite{Benjamin2006} is clear about the
extant of adaptation in stating:

{\footnotesize
  \begin{quote}
\small \emph{The primary difficulty often associated with behavior-based control
concerns action selection - namely how to ensure the chosen action
really is in the best overall interest of the robot or vehicle.}
\end{quote}
}

Such techniques have proved their mettle in the first round of use of
AUVs; they've provided operators a simple way to handle control
complexity of the vehicle while returned data at far higher resolution
than would have been possible using traditional ship-based methods and
at substantially reduced amortized costs of platform operation and
ownership over multiple years. AUV operators and clients for their
data products have reasons to be well satisfied with results thus
far. However, as hardware and more cost-effective sensors become more
robust to the environment and as mission durations increase (as
demonstrated by glider experiments such as \cite{rucool11}), sustained
presence in the ocean requires the ability to deal with off-nominal
conditions, that balances the needs of high-level mission management,
low-level navigation, instrument control and detection of unstructured
and poorly understood phenomena. Earlier reactive methods are unable
to perform such a balancing act without skewering the foundations of
the software development methodology leading to code bloat or worse.
\cite{carreras06} provides a reasonably good overview of AUV control
architectures that essentially is based on reactive control methods.

\begin{wrapfigure}{r}{0.2\textwidth}
 \centering
 \includegraphics[scale=0.3]{figs/planner-arch.pdf}
 \caption{\small A general architectural block diagram for an
   Artificial Intelligence based Planner.}
 \label{fig:planner}
\end{wrapfigure}

A word about generalized planning. Traditionally planning has been
considered computationally intensive (the general planning problem can
be worse than NP-complete \cite{ghallab04}). A major reason for this
belief had to do with the role of the planner (which was assumed to be
generative) and its place in the architecture (infrequently called to
re-plan the \emph{entire} mission in off-nominal conditions) within a
\emph{sense-plan-act} paradigm. This can limit the reactivity of a
robot especially when the environment could change at a rate faster
than the planner can plan.

The architecture of a typical AI planner is along the lines of
Fig. \ref{fig:planner}. Given a \emph{domain model} which encodes the
platform constraints in a higher-level language, an \emph{initial
  state}, a high-level objective or end \emph{goal}, the \emph{plan
  database} (or \texttt{plandb}) is where all assertions are tracked
and maintained. Theorem proving in the form of satisfaction of axioms
embodied in the domain model occurs within the \texttt{plandb}. It is
the job of the \emph{search engine} to provide the inferential
mechanism for state space exploration potentially aided by heuristics
encoded for \emph{search control}. In some complex domains
\cite{mus98} \emph{external modules} can provide additional domain
expertise which can augment the domain model. The end result is plan
formulation using a multitude of approaches either using forward or
backward search or using more generalized \emph{partial order} methods
\cite{ghallab04}.

Deliberative techniques for robotic control on the other hand
contrast in providing a distinct set of advantages:

\begin{enumerate}

\item action selection is driven axiomatically based on a systematic
  assessment of a range of causal conditions. An action is selected
  for instantiation only when there is causal support in the form of a
  constraint in a deterministic model of the vehicle's operation.

\item the system is goal directed which forms an objective towards
  which the system is expecting to converge providing a foci in action
  selection.

\item scaling to different (computational) problem size is dependent
  on incremental model building rather than being concerned about
  nuanced interactions between behaviors.

\item systematic software engineering methodologies such as spiral
  development \cite{boehm86} can further be used for scaling up the
  task as demonstrated by \cite{DS1report}.

\item when interactions between actions and/or behavior fragments do
  (and must) occur there are explicit constraints encoded in the model
  that must be computationally satisfied contrasting with carefully
  crafted strategies to instantiate behaviors on a stack. Such
  explicit recording of constraints in the model in turn aids
  long-term software maintenance and sustaining engineering.

\end{enumerate}

There are costs associated with such inference based control are also
not inconsequential. Foremost among them is the steep learning curve
that comes with the design of the domain model. With limited support
tools necessary for knowledge capture and design of the constraints
often in stylized higher-level languages (see \cite{NDDL} for an
example), it requires the modeler to think in alternative ways to do
problem solving. More importantly it is often necessary for the
modeler to be exposed to the internal representation of the planner
and how it performs search. THe level of expertise is often well above
what would be expected of a typical well-rounded Computer Science
graduate student.

However this is mild criticism given the end benefits and the
scientific and engineering goals associated with adaptive and
persistent observation for marine robots whether for civil or military
applications. Given the general problem-solving nature of such
deliberative systems, there is also a substantial overall benefit for
the field of robotics.

\subsection{Evolution of AI-based Robotic Planning: A Necessary Digression}

Planning and plan execution are not new to the overall field of
robotics. Early motivation of building computational mechanisms for
decision making were intended to be deployed on robotic devices by
very definition. The seminal volume \cite{computersthought} very
clearly articulates how physical manifestation of robots could be
controlled for task planning and execution. The very first planner
\cite{green69} was an exercise in computational theorem proving with
early state space planners popularized by \texttt{STRIPS}
\cite{strips71} dominating the academic landscape. An early
demonstration of planning as a situated agent was the highly
influential \texttt{Shakey} the robot experiment
\cite{shakey84}. Subsequent work driven largely by defense funding in
the United States sequestered the discipline in academic labs
principally applied for algorithmic development and if embedded then
in testing within the confines of modest indoor settings. The leap to
applying them for real world problem solving is relatively new and
driven largely by NASA applications pioneered by \cite{mus94,mus98,
  jonsson00, rajan00, chien05, bresina05}.

The dominant approach for building agent control systems utilize a
$3$-layered architecture (see \cite{gat98} for a well thought out
rationale), notable examples of which include IPEM
\cite{AmbrosIngerson88}, ROGUE \cite{Haigh98}, the LAAS Architecture
\cite{alami:1998p820}, the Remote Agent Experiment \texttt{RAX}
\cite{mus98} and the Autonomous Spacecraft Experiment \texttt{ASE}
\cite{chien99} (see \cite{Knight01} for a survey). Early efforts were
cognizant of if absent with, dealing with time and resources for such
experiments.

Representationally, the first paper on managing time systematically
within AI Planning frameworks was \cite{dean87}; subsequent efforts by
Boddy and others \cite{Dean88,Boddy93} refined the notion of using a
temporal database and the conceptualization of temporal
intervals. Using these base-level concepts Muscettola and others
\cite{mus92, ghallab94, laborie95, cesta96} developed the notion of
plan/schedule envelopes using the notion of state variable
instantiated timelines. Simultaneously work by Dechter et.al
\cite{dechter91} resulted in efficient temporal constraint propagation
which systematically defined the notion of Simple Temporal Networks
(STNs) and path and arc consistency algorithms. This work neatly put
together earlier efforts by Mackworth and Freuder \cite{mackworth77,
  mack85} and the Constraints community on propagation algorithms. In
parallel work in the UK as a result of US DARPA funding, resulted in
the design of \texttt{O-Plan} \cite{currie91}, which leveraged the
notion of constraints and metric time, but within a Hierarchical Task
Network (HTN) representation. All these efforts were soon after Vere’s
paper on planning and time \cite{vere83} which had a profound effect
on the AI Planning community. In particular the work of Muscettola
\cite{mus94} was embraced by NASA for telescope scheduling. Around the
same time period, researchers at LAAS came up with an architecture
\cite{alami:1998p820} which encapsulated the \texttt{IxTeT} planner
\cite{ghallab94}. It featured advanced concepts for a temporal planner
with the notion of chronicles, plan recognition for partial plan
fragments and early use of systematic resource
constraints. Development on \texttt{IxTeT} has continued albeit at a
slower pace with the contribution of various PhD students’ thesis.

\subsection{The Remote Agent Experiment and beyond}

While \texttt{HSTS} \cite{mus94} was being implemented as a
ground-based planning tool for decision support for US based space
observations (EUVE and Cassini) \cite{mus95}, the opportunity to fly
onboard NASA’s New Millennium Deep Space $1$ came about. The design of
the Remote Agent Experiment \texttt{RAX} \cite{pell97, bernard98,
  pell98, mus98, DS1report, rajan00, jonsson00} was a direct off-shoot
of this effort where \texttt{HSTS} (written in LISP) was flown on
board and successfully commanded the spacecraft for a week in May
1999. There were a number of significant software engineering lessons
learned with the HSTS technology as integrated for the RAX:

\begin{enumerate}

\item Constraint-based representations were not only sufficient for
  plan synthesis, but valuable during debugging and development as a
  means of building a viable domain model.

\item Domain models when separated from the search engine as
  articulated by the model-based approach \cite{williams96a} ensured
  that sufficient effort would be targeted on human validation of the
  model building phase while ensuring that search engine stability
  across applications and domains resulted in lower cost to deploy.

\item Flexible temporal representation generating a range of plans as
  against a single plan, were robust for plan execution.

\item If planning and execution were disconnected (as in
  \texttt{RAX}), dispatchabilty \cite{mus98a} and controllability
  \cite{morris00} issues within temporal plans need be addressed. In
  \texttt{RAX}, a post-processing step was added to counter these
  effects. It was clear that execution needed access to the planners
  database and be able to propagate execution time constraints prior
  to command dispatching.

\end{enumerate}

It was with the demonstration of \texttt{RAX} $65$ Million miles in
outer space, that temporal reasoning using methods came to the
forefront of AI planning and for situated agents. Lessons learned from
\texttt{RAX} led to the development of \texttt{IDEA}
\cite{mus02,mus04} with the central theme of using planners
collectively for problem solving arose. Another critical theme was
iteratively and incremental plan repair of an anytime plan
\cite{Zaimag96} as proposed by \texttt{CASPER}
\cite{chien00}.

This last lesson in particular had a lasting impact with the
observation that planning and execution are strongly
inter-related. This core concept was behind the next generation of the
Remote Agent, called \texttt{IDEA} (Intelligent Deployable Execution
Agents) where planning and execution were \emph{intertwined} within a
single domain model that spanned the most abstract (for high-level
planning) to the least (for diagnosis). \texttt{IDEA} \cite{mus02,
  mus04} agents were expected to interact to achieve plan formulation;
however there was no systematic framework for formally governing these
interactions. \texttt{IDEA} was also computationally heavy and
required substantial effort to customize for an application domain. It
did move to a retooled version of \texttt{HSTS}, now called EUROPA
\cite{frank2003, barreiro09} and ported to the C++ language.

While \texttt{IDEA} was being deployed as a coordinating executive for
earth analog rover deployments \cite{wetter05} a separate development
was undertaken to deploy the EUROPA planner for NASA’s Mars
Exploration Rovers (MER) mission. The \texttt{MAPGEN}
(Mixed-initiative Activity Plan GENerator) planner \cite{bresina03,
  aichang04, bresina05, bresina05a} as a decision support tool in the
mixed-initiative mode; \texttt{MAPGEN} continues to be used to this
day to command the twin rovers on the Red
Planet. % While EUROPA2 was not
% designed as an embedded planner, NASA undertook a bottom-up
% re-implementation of the planner and substantially increased its
% performance. EUROPA2 was then released as open source to the AI
% community at large [Europa].

The Autonomous Sciencecraft Experiment (\texttt{ASE}) \cite{chien99,
  chien03} was another autonomy demonstration in space, this time for
a more observable vehicle, the EO-1 in Earth orbit. It encapsulated a
classical 3-layered architecture like \texttt{RAX}. The fundamental
functional difference was that goals could not only be sent by a
ground segment, but also by an onboard science driver which could
encapsulate a range of Machine Learned feature detectors which could
trigger the planner. However ASE was not an integrated system; the
\texttt{CASPER} planner \cite{chien00} would generate plans which were
executed by a separate commercially available rule-based system
\cite{icl}. \texttt{CASPER} touts a \emph{continuous} paradigm for
planning; however this incremental modification simply takes the
overall mission plan and selectively plans near-term at a more
detailed level. The system also uses iterative-repair as a way to fix
plans; however this too is derived from a model of the mission and
spacecraft constraints which allow simple reconfiguration modes
towards replanning. Robust software engineering is also not an
important aspect targeted by \texttt{ASE} given the disparate models
between the planner and the rule-based executive. With such a system,
real-time updates from the environment cannot be accommodated since
the planner is a monolithic computational environment. The rate of
updates from the science drivers is roughly in the order of an earth
orbit on the EO-1 vehicle. As a result, the ASE is only an incremental
update on RAX and is not suitable for domains where complex automated
reasoning and asynchronous environmental observations need to be
factored in deliberation.

\texttt{TCA} \cite{simmons94} is another framework for robot control
and has similar motivations to \texttt{RAX} and
\texttt{ASE}. Reactivity and deliberation are also key to \texttt{TCA}
and a systematic approach to development led to its design. However it
suffers from three key weaknesses. First it has a weaker
representational framework of Task Trees which require representation
of tasks within cleanly formulated hierarchies. Constraints between
branches within a task hierarchy are allowed; timing between leaves of
different trees is however not possible. Second, its temporal
framework does not deal with flexibility; time points are fixed and
while partial orders are possible, it does deal with execution time
uncertainty. Third, while \texttt{TCA} shares the notion of using
domain expertise of different planners, it does so thru a centralized
message passing mechanism. While this allows disparate code-bases to
talk to each other thru IPC (a inter-process communication or message
passing mechanism), failure in the centralized controller is
tantamount to a system crash.

The \texttt{ReSSAC} project at ONERA \cite{teichteil07} has similar
motivations in controlling aerial UAV platforms for autonomous Search
and Rescue. Like \texttt{TCA}, the project uses a supervisor to
control planning and at the same time decouples the deliberation and
reactive components from the lower level functional layer. This
deliberate separation is partly because of the use of optimizing MDP’s
(Markov Decision Processes \cite{mdp93}) for planning which uses
probabilistic state transitions to determine the next course of action
based on perceived state. To overcome the typical problem of state
space explosion with MDPs, a local heuristic is used in order to
generate reachable goal states incrementally. This work however does
not deal with metric time with a modestly simple model.

\texttt{CIRCA} \cite{musliner95} is another effort to bring decision
making for real-world problems with the augmented need to have
\emph{verifiable} controllers synthesized in-situ to ensure that
within explored states, there are no erroneous transitions to
dangerous states by asynchronously generating Test-Action-Pairs
(TAPs).  These are annotated production rules consisting of a set of
tests (or preconditions) and a set of actions to take if all the tests
return true. TAPs are synthesized and scheduled by \texttt{CIRCA} and
provide a viable way to deal with a dynamically changing
world. \texttt{CIRCA} however differentiates between reasoning about
time and reasoning in real-time with the implication that reasoning
necessarily requires substantial computation for state space
exploration which negates the real-time (and thus real-world)
impact.

A key and early concern that dominated planning was that of
scalability.  The planning cycle in many approaches was monolithic
often making fast reaction times impractical when necessary for
embedded robotic agents. Many of these systems (which were
$3$-layered) also utilized very different techniques for specifying
each layer in the architecture resulting in duplication of effort and
a diffusion of knowledge.  The work we highlight in this chapter
builds on the approach used by \texttt{IDEA} \cite{mus02, mus04} in
utilizing a collection of controllers, each interleaving planning and
execution within a common framework. \texttt{IDEA} however, provided
no support for conflict resolution between controllers, nor does it
provide an efficient algorithm for integrating current state within a
controller, relying instead on a possibly exponential planning
algorithm. Efficient synchronization of state in a partitioned
structure is fundamental to making the approach effective in practice.

Our system, \texttt{T-REX}, the focus of this chapter, is an agent
control structure formally defined as a composition of coordinated
control loops with a recurring sense-plan-act (SPA)
cycle. \texttt{T-REX} was designed after carefully evaluating lessons
learned from the IDEA and owes substantially to the notion of
partitioned problem solving from that effort. However this
partitioning is systematic and methodologically principled where we
manage the information flow within the partitioned structure to ensure
consistency in order to direct the flow of goals and observations in a
timely manner. The resulting control structure improves scalability
since many details of each controller can be encapsulated within a
single control loop.  Furthermore, partitioning increases robustness
since controller failure can be localized to enable graceful system
degradation, making this an effective \emph{divide-and-conquer}
approach to the overall control problem.

A recent variation of the key idea of such controller composition
derived from \texttt{T-REX} is \texttt{ROAR} \cite{degroote11} where
lower-level functional modules that control specific sensors are
organized within a graph hierarchy. One reason they cite such a
structure is essential is in dealing robustly to failure; an
off-nominal condition will allow rapid graph traversal to identify
alternative ways in which sensing tasks can be executed and therefore
aiding in responsiveness. However, it is unclear whether to date this
system actually deliberates and whether it is actually instantiated on
a real platform for control.

% ``The whole partition must be accordingly reorganized: this kind of
% construction does not scale well over a large variety of missions,
% missing the objective of a versatile architecture for robots.''

### 2012-01-19 trunk/arch.tex
Added a cal description of  both synchronization and deliberation solvers in the intro of the europa reactor

@@
 intertwined in the sense that one influences the outcome of the
-other.
+other. They are represented as two europa solvers modifying the same
+plan database and for which the execution is managed by the T-REX 
+framework rythemed by the time advance during the execuion of the
+system:
+\begin{enumerate}
+\item The synchronization solver is a specialized eurpa solver that
+  will integrate in the plan new information about the evolution of
+  the External state varaible and ensure that the reactor propagate
+  these in order yto both identify its current stetate and inform
+  other reactors of any state change on its internal timelines. This
+  solver is summoned at the beginnning of every single tick.
+\item The deliberation solver is managing the deliberation process of
+  the reactor in order to produce a new plan or alter its current plan
+  as new goals are given to the reactor or the synchronization solver
+  identifying a conflict between the current state of the world and
+  the expectations of the previous plan. This process can  span other
+  multiple tick and therrfore can be interrupted at any single tick in
+  order for the synchronization solver to do its task.
+\end{enumerate}
 

### 2012-01-20 trunk/europa.tex
update
added some latex markup.

@@
-1. EUROPA
+\section{EUROPA}
+\label{sec:europa}
 
@@
 
-1.1 Constraint Programming
+\subsection{Constraint Programming}
+\label{sec:europa:cp}
 
@@
 CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
+Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other “primitive” CP constraint, such as <= or >=.
 
@@
 
-1.2 EUROPA's Architecture
+\subsection{EUROPA's Architecture}
+\label{sec:europa:arch}
 
@@
 
-1.3 Modeling
+\subsection{Modeling}
+\label{sec:europa:modeling}
 
@@
 
-1.4 Inference
+\subsection{Inference}
+\label{sec:europa:inference}
 
@@
 
-1.5 Search
+\subsection{Search}
+\label{sec:europa:search}
 
@@
 All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.
+
+Given that temporal intervals may extend indefinitely in both positive and negative directions, for problems with a temporal dimension it is also common to specify a Planning Horizon when invoking a solver.
+The Planning Horizon is the temporal interval that a problem solver has to worry about, any decisions that fall completely outside of the Planning Horizon can be ignored.
+This is very useful for recurrent activities, consider for instance a model where a medical checkup has to occur every 3 months, the user can model a recurring activity and then use the Planning Horizon to ensure that a finite number of medical checkup activities are generated as part of a plan.
+This is also useful to ignore activities that may happen too far in the future or in the past to be relevant for a particular plan, this way the user can ensure that the solver is making as few commitments as necessary.
 
@@
 Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
+Open Conditions can be resolved in three ways:
+		- Merging: A free activity is merged with a matching activity already in the plan.
+An activity a is said to match an activity a?
+if a and a?
+unify and the temporal constraints involving a are satisﬁed by a?.
+Thus,  a and a?
+can be considered the same activity and we do not need to introduce a in the plan.
+Consequently, the compatibilities associated with a are not ﬁred, because they have been already triggered when a?
+was introduced in the plan.
+		- Activation We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciﬁc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
+This results in both an ordering ﬂaw, corresponding to the just activated activity, and a number of open condition ﬂaws, corresponding to the new subgoal activities.
+		- Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
+
 	3.
@@
 EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
+
+\begin{verbatim}
 
@@
 
+\end{verbatim}
+
 The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
@@
 - explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.
-
-1.6 Future Work?
-explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).
-
-The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space

### 2012-01-20 trunk/europa.tex
added more latex markup to europa.tex and included europa.tex in setobook.tex

@@
 
-	- Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
+\begin{enumerate}
+	\item Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
 
-	- Domain reduction: To infer a valid set of values for a variable.
-For example, for discrete variables, the constraints all_different(x1,x2,x3) and xi>=0, xi <=4.
+	\item Domain reduction: To infer a valid set of values for a variable.
+For example, for discrete variables, the constraints allDifferent(x1,x2,x3) and xi>=0, xi <=4.
 if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}
+
+\end{enumerate}
 
@@
 
-- The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables.
+\begin{enumerate}
+
+	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables.
 It also provides an efficient arc consistency mechanism (TODO: ref to AC-3).
@@
 
-- The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does.
+	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does.
 Also manages tokens, which are a mechanism to group variables to represent temporally scoped state.
@@
 
-- The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
+	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
 
-- The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
+	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
 
-- API Layer: makes all the services available so that EUROPA can be used to build applications.
+	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
 
-- Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+
+\end{enumerate}
 
@@
 
-- A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
-- A problem instance (also called initial state) that consists of:
-	- Variable and Object instances that exist for the entire planning horizon.
-	- Temporally scoped predicate and action instances.
-	- Temporally scoped goals.
+\begin{enumerate}
+
+\item A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
+\item A problem instance (also called initial state) that consists of:
+	\item Variable and Object instances that exist for the entire planning horizon.
+	\item Temporally scoped predicate and action instances.
+	\item Temporally scoped goals.
+
+\end{enumerate}
 
@@
 The operations to refine the partial plan PP at any time are:
-	1- Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
-	2- Select one such flaw
-	3- Select a resolver for the flaw
-	4- Refine PP by applying the resolver
-	5- If an inconsistency is found, try another resolver.
-	6- If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
+\begin{enumerate}
+	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
+	\item Select one such flaw
+	\item Select a resolver for the flaw
+	\item Refine PP by applying the resolver
+	\item If an inconsistency is found, try another resolver.
+	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
+\end{enumerate}
 
@@
 
-	1. Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
-	2.
-Open Condition: an open condition is an inactive token.
+\begin{enumerate}
+	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
+
+	\item Open Condition: an open condition is an inactive token.
 Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
 Open Conditions can be resolved in three ways:
-		- Merging: A free activity is merged with a matching activity already in the plan.
+\begin{enumerate}
+		\item Merging: A free activity is merged with a matching activity already in the plan.
 An activity a is said to match an activity a?
@@
 was introduced in the plan.
-		- Activation We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciﬁc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
+		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciﬁc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
 This results in both an ordering ﬂaw, corresponding to the just activated activity, and a number of open condition ﬂaws, corresponding to the new subgoal activities.
-		- Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
+		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
+\end{enumerate}
 
-	3.
-Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
+	\item Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
 Recall for example that a token may belong to objects (e.g.
@@
 Threats are resolved by imposing ordering constraints among tokens.
+
+\end{enumerate}
 
@@
 
-// Function to solve a partial plan by successive refinement
-// until all flaws are resolved.
-bool solve(PartialPlan plan)
-{
-	// Propagate the constraints to test for inconsistency. If found to be
-	// inconsistent, then we can return false since this is a dead-end i.e. no refinements
-	// to p can yield a consistent plan.
-
-	if(isInconsistent(plan))
-		return false;
-
-	// Non-deterministically choose a flaw from the set of available flaws.
-	Flaw flaw = chooseFlaw(plan);
-
-	// If there are no flaws, then p is complete and we can terminate with success.
-	if(flaw == NULL)
-		return true;
-
-	// Otherwise we formulate a decision point which is a branch in the search space. Each
-	// choice is a particular refinement operation and the DecisionPoint collects all possible
-	// refinement operations for the given flaw.
-	DecisionPoint decision = makeDecisionPoint(flaw, plan);
-
-	// Continue as long as we have something to try
-	while(decision.hasNext()) {
-		// A new partial plan is obtained by application of a refinement operator. Note that the ordering
-    		// over refinement operators to select is a non-deterministic step.
-		PartialPlan pp = decision.executeNext();
-
-		// Recursive call to solve the new planning problem. If successful, then we are done.
-		if(solve(pp))
-			return true;
-		else // Otherwise, retract the last refinement operation
-			d.undo();
-  	}
-
-	// If we arrive here, then we have exhausted all options to resolve the flaw, including the case where
-  	// no options were available initially. Thus the problem cannot be solved.
-  	return false;
-}
+01.  bool solve(PartialPlan plan)
+02.   {
+03.      if (isInconsistent(plan))  return false;
+04.
+05.      Flaw flaw = chooseFlaw(plan);
+06.      if (flaw == NULL) return true;
+07.
+08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
+09.
+10.      while(decision.hasNext()) {
+11.          PartialPlan pp = decision.executeNext();
+12.
+13.          if (solve(pp)) return true;
+14.          else d.undo();
+25.      }
+16.
+17.      return false;
+18.  }
 
@@
 The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
+
+\begin{enumerate}
+	\item Line 3: Propagate the constraints to test for inconsistency. If found to be inconsistent, then we can return false since this is a dead-end i.e. no refinements to p can yield a consistent plan.
+	\item Line 5: Choose a flaw from the set of available flaws.
+	\item Line 6: If there are no flaws, then p is complete and we can terminate with success.
+	\item Line 8: Otherwise we formulate a decision point which is a branch in the search space.
+Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
+	\item Line 10: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
+	\item Line 11: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
+	\item Line 13: Recursive call to solve the new planning problem. If successful, then we are done.
+	\item Line 14: Otherwise, retract the last refinement operation and move on to try the next one.
+	\item Line 17:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially. Thus the problem cannot be solved.
+\end{enumerate}
+
 This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.
@@
 - explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-01-23 trunk/arch.tex
Short description of T-REX architecture … still not very clear

@@
 
-{\em This section presents quickly T-REX architecture but move
-  quickly the focus on a single reactor and gross concepts on the
-  architecture:
+% {\em\tiny This section introduces shortly T-REX architecture but move
+%   quickly the focus on a single reactor and gross concepts on the
+%   architecture:
+%   \begin{itemize}
+%   \item For t-trex the whole decision and control problem is reduced
+%     around the state variables (or timeline construct). 
+%   \item This reduce the execution tracking problem for each reactor on a state
+%     identification problem (deduce internal state from the external
+%     state), and the control problem to goal posting which will trigger
+%     deliberation on the owner of the corresponding timeline. 
+%   \item the architecture itself see each reactor as a black box and
+%     therefore each reactor can implement its own mechanism in order to
+%     resolve both synchronization and deliberation. Still we provide a
+%     reactor based on the europa framework that leverage the automated
+%     planning capabilities in order to do model based planning and
+%     execution with a rich representation of resources and time.
+%   \end{itemize}}
+
+The T-REX architecture was designed with the objective to bring 
+planning closer to the low-level control of the vehicle while
+maintaining its reactivity. Indeed, in most of the architecture where
+task planning is embedded, this task remains at a failry high level of
+abstraction as the complexity of this task would otherwise impede the
+system reactiveness. As a result the planning problem managed {\em in
+  situ} remains fairly detached from low-level control and most of the 
+reactive adaptation which in contrast is managed by specialized 
+component that relies on different representations for modeling 
+the local behavior they control or even do not have any explicit 
+formal model of this behavior. Taking cues from the IDEA 
+architecture \cite{mus02}, we designed a core framework that provides a
+formal basis \cite{Py:2010ti} on the way each of the sub-component --
+called reactors -- of our agent can interact by exchanging only
+information through {\em state variables} by ensuring that state 
+information can propagate through all the reactors ensuring a
+consistent view of the present state of the world at every single tick
+trough a bottom-up {\em synchronization} flow and collaboration for
+future state evolution by timely exchanges of goals on {\em state
+  variables}. This exchange does not impose to any reactor to be aware
+of the reactors it is connected to but only focus on the state
+variables it relies on (called {\em external}) and the state
+variables it manages and maintains (called {\em internal}).
+
+Take for example the T-REX agent instance we use in our AUV
+(Fig. \ref{fig:agent}). In this architecture all the reactors
+(symbolized by colored boxes) do not have explicit connection from one
+to another but instead rely on a publish/subscribe model on state
+variables. For Example the \textsf{Pilot} reactor needs the {\em
+  externally} managed state informations from \textsf{Sensor\_Data} 
+and \textsf{Nav\_cmd} states variables in order to maintain {\em
+  internally} the state variables \textsf{Waypoint} and
+\textsf{Nav\_policy}. This relation applies in two way :
 \begin{itemize}
-  \item For t-trex the whole decision and control problem is reduced
-    around the state variables (or timeline construct). 
-  \item This reduce the execution tracking problem for each reactor on a state
-    identification problem (deduce internal state from the external
-    state), and the control problem to goal posting which will trigger
-    deliberation on the owner of the corresponding timeline. 
-  \item the architecture itself see each reactor as a black box and
-    therefore each reactor can implement its own mechanism in order to
-    resolve both synchronization and deliberation. Still we provide a
-    reactor based on the europa framework that leverage the automated
-    planning capabilities in order to do model based planning and
-    execution with a rich representation of resources and time.
-  \end{itemize}}
+\item In order to identify its current {\em internal} state values
+  this reactor needs to know the current {\em external} state
+  information it relies on. In this case the current \textsf{Waypoint}
+  the vehicle is heading to can be extracted for example from the
+  current \textsf{Nav\_cmd} executed.
+\item The future objectives a reactor has on its future \textsf{Internal}
+  state will probably imply sub-objectives to be reported to the
+  owner of its {\em external} depending on their current values. Again
+  if this reactor wants to go ato a specific \textsf{Waypoint} --
+  depending on its current position provided by \textsf{Sensor\_Data}
+  it can identify a sequence of \textsf{Nav\_cmd} that should
+  eventually help him reach this location.
+\end{itemize}
+
+\begin{figure}[!htb]
+ \centering
+ \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
+ \caption{\small A \texttt{T-REX} agent is composed of multiple
+   reactors or control loops (rounded boxes) which are connected
+   through state variables provided by one reactor ({\color{blue}blue}
+   solid line) with multiple possible clients ({\color{green}green}
+   dashed lines)}
+  \label{fig:agent}
+\end{figure}
+
+Overal T-REX architecture helps to abstract to the reactor interaction
+to only state variables. Each reactor is practically ignoring where
+its {\em external} state information is managed as long as this
+information is available and properly maintained so they can properly 
+manage their own {\em internal} state information which in turn may be
+used by other reactors. Such formalism also constrain reactors
+interaction to state information exchanges which are two folds:
+\begin{itemize}
+\item {\em Observation on the current state values}: This information
+  propagates up in the reactor graph at the synchronization phase 
+  and occur at the beginning of every time steps (called
+  ticks). During this phase the agent provides to each reactor 
+  updates on their {\em external} state variables for this tick and
+  ask them to compute their {\em internal} state variables values for
+  this same tick. By doing so we propagate a consistent view of that
+  state of the world throughout all the reactors as time advance.
+\item {\em Goals request on future state}: This information propagates
+  down in the reactor graph. It reflects that, in order to satisfy own
+  of its {\em internal} objectives, a reactor may have a plan that
+  rely on future values of one of its {\em external} state. When such
+  a plan is identified we ensure that this part of the plan is given
+  as a goal to the owner of the given {\em external(s)} state variable(s). 
+\end{itemize}
+
+The overal execution cycle of the agent is emerging through these
+interaction and the processing that occurs in each reactors in order
+to process them. New observation occurring at every tick inform each
+reactor of the current state of the world and impose them to identify
+how it impact their state (and potentially what they planned for the
+future). Goals they receive will trigger in the reactor the need to
+identify a strategy (or plan) that will allow them to reach these new
+states knowing their current state deduced from synchronization.When
+such a plan has been identified, the {\em external}  part is then
+dispatched to other reactors that can do the same. This flow and focus
+around state and goals is perfectly fitting with automated planning
+and even more so to the Europa framework. As an illustration, in 
+ Fig. \ref{fig:agent}, only \textsf{Shore comm.} and \textsf{Vehicle}
+ reactors are specialized C++ code (acting as interface to external
+ programs). All the other reactors are based on the europa framework
+ differentiated only by the domain model they manipulate and som
+ heuristic configuration parameters. In the following sections we are
+ going to focus specifically ion this kind of reactor to illustrate
+ how  a formal planning framework can be used in our architecture to
+ both track state evolutions and plan for future objectives while
+ remaining efficient enough to do these tasks in a real-time control 
+loop.
 

### 2012-01-23 trunk/arch.tex
Added figures -- from sutton book chapter froor now but I can  reedit them in the future

@@
 plan database and for which the execution is managed by the T-REX 
-framework rythemed by the time advance during the execuion of the
-system:
+framework as time advance during the execution of the system:
 \begin{enumerate}
-\item The synchronization solver is a specialized eurpa solver that
+\item The synchronization solver is a specialized europa solver that
   will integrate in the plan new information about the evolution of
-  the External state varaible and ensure that the reactor propagate
-  these in order yto both identify its current stetate and inform
+  the External state variable and ensure that the reactor propagate
+  these in order to both identify its current state and inform
   other reactors of any state change on its internal timelines. This
-  solver is summoned at the beginnning of every single tick.
+  solver is summoned at the beginning of every single tick.
 \item The deliberation solver is managing the deliberation process of
@@
   the expectations of the previous plan. This process can  span other
-  multiple tick and therrfore can be interrupted at any single tick in
+  multiple tick and therefore can be interrupted at any single tick in
   order for the synchronization solver to do its task.
 \end{enumerate}
+
+While synchronization is assumed to be fast enough that it can be
+resolved in less than a tick. Deliberation on the other hand can take
+several ticks reflecting that automated planning can take time
+depending on how far in the future the planner is looking ahead and
+the complexity of the model. T-REX architecture, and more specifically
+the europa based reactor in our case, takes this into account by
+considering synchronization and deliberation as two concurrent tasks
+with their own timing requirements and priority. 
+While the two processes are separate, they share the same plan
+internal to a reactor and therefore at every synchronization cycle the
+planning process is informed of the new world state and its impact on
+the plan. Conversely, when the planning process eventually finds a
+solution, synchronization is informed about generating the planned
+state values on the {\em external} state variables. This {\em
+  external} plan defines the set of {\em goals} for this state
+variable managed by another {\em reactor}. Fig. \ref{fig:tick-exec}
+shows a conceptual view of these two tasks.
+
+\begin{figure}[!htbp]
+  \centering
+  \vskip-1pc
+  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
+  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
+      Deliberation} is interrupted by {\em synchronization} at the
+    beginning of every {\em tick} allowing integration of state
+    information.}
+  \label{fig:tick-exec}
+  \vskip-0.8pc
+\end{figure}
 
@@
 as objectives for the reactors it depends on in a seamless flow.
+
+{\em fpy: I still wonder whether I should start with planning assuming
+  first that the initial state is known and then develop
+  synchronization and how it ties state information in to the plan or
+  the other way around as it is now}
 

### 2012-01-24 arch.tex
Added som details about deliberation. Still need a lot of refinement.

@@
 intertwined in the sense that one influences the outcome of the
-other.
+other. They are represented as two europa solvers modifying the same
+plan database and for which the execution is managed by the T-REX 
+framework as time advance during the execution of the system:
+\begin{enumerate}
+\item The synchronization solver is a specialized europa solver that
+  will integrate in the plan new information about the evolution of
+  the External state variable and ensure that the reactor propagate
+  these in order to both identify its current state and inform
+  other reactors of any state change on its internal timelines. This
+  solver is summoned at the beginning of every single tick.
+\item The deliberation solver is managing the deliberation process of
+  the reactor in order to produce a new plan or alter its current plan
+  as new goals are given to the reactor or the synchronization solver
+  identifying a conflict between the current state of the world and
+  the expectations of the previous plan. This process can  span other
+  multiple tick and therefore can be interrupted at any single tick in
+  order for the synchronization solver to do its task.
+\end{enumerate}
+
+While synchronization is assumed to be fast enough that it can be
+resolved in less than a tick. Deliberation on the other hand can take
+several ticks reflecting that automated planning can take time
+depending on how far in the future the planner is looking ahead and
+the complexity of the model. T-REX architecture, and more specifically
+the europa based reactor in our case, takes this into account by
+considering synchronization and deliberation as two concurrent tasks
+with their own timing requirements and priority. 
+While the two processes are separate, they share the same plan
+internal to a reactor and therefore at every synchronization cycle the
+planning process is informed of the new world state and its impact on
+the plan. Conversely, when the planning process eventually finds a
+solution, synchronization is informed about generating the planned
+state values on the {\em external} state variables. This {\em
+  external} plan defines the set of {\em goals} for this state
+variable managed by another {\em reactor}. Fig. \ref{fig:tick-exec}
+shows a conceptual view of these two tasks.
+
+\begin{figure}[!htbp]
+  \centering
+  \vskip-1pc
+  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
+  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
+      Deliberation} is interrupted by {\em synchronization} at the
+    beginning of every {\em tick} allowing integration of state
+    information.}
+  \label{fig:tick-exec}
+  \vskip-0.8pc
+\end{figure}
 
@@
   \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
-  \caption{Ilustration of the synchronization flaws in a reactor. The
+  \caption{Illustration of the synchronization flaws in a reactor. The
     reactor receive new observations when they are produced by the
@@
 
+Synchronization remains here a deliberation process, and for this
+reason there's always the risk that the search of a solution is not a
+straightforward sequence of unit decision. As such process is required
+by the architecture at every ticks for every reactors presently
+running this need to be carefully designed. At the engine level we
+made careful choices in term of both the focus of the search -- limited
+only to tokens that can potentially overlap the current tick -- and
+its heuristic where the internal state flaw is considered last during
+the search. The later proved to limit the amount of backtrack in the
+search for a solution -- or at least how deep we need to backtrack in
+our decision tree -- similarly the sequence in which we evaluate the
+possible resolution for such a flaw -- presented previously -- was
+selected according to simple assumptions that takes benefit of a
+possibly already existing plan by assuming that if the outcome of
+synchronization appears consistent with our current plan then it
+probably reflects that this reactor plan is executed. Such assumption
+helps direct the internal state evaluation and, in nominal cases (where
+we assume that reactors are acting in concert), proved to be a sound
+solution.
+
+One final aspect to keep into account is that in T-REX point of view
+synchronization is the only critical task of a reactor. Indeed, such
+process is used in order for the agent to identify its
+current state and maintain it consistent throughout the agent
+life-time. A reactor not being able to find a consistent solution
+during synchronization -- meaning that after evaluating all the
+possible solutions provided by its model it did not found any plan
+that is not inconsistent -- not only expresses that this reactor is
+not able anymore to explain the state of the world but may jeopardize
+other reactors that relies on its {\em internal} state. For this
+reason a failure to do synchronization immediately result on the
+agent killing this reactor and notifying any reactor that depends on
+its {\em internal} state variables by posting the special observation
+\textsf{Failed}. By doing so we allow for graceful degradation where
+other reactors can readapt their plan as they receive this
+\textsf{Failed} observation.
+
 \subsubsection{Deliberation : planning for future state evolution}
 \label{sec:arch:plan}
+
+The planning process inside a reactor is directly based on the way
+europa work which was described in previous section. Still some simple
+alterations were made in  order to both include deliberation
+information one reactor provide to the architecture -- such as its
+deliberation latency and its planning look-ahead both expressed in 
+ticks -- and the fact that we potentially need to interleave this
+deliberation with not the synchronization process but also with 
+the deliberation of other reactors\footnote{Current implementation of
+  T-REX is running on a single process and it is the responsibility of
+  the agent itself to emulate the multi-threading of the reactors
+  deliberation and synchronization}. For this reason we 
+decompose the algorithm \ref{alg:europa:solve} inside the reactor by
+having the recursive calls decomposed in single steps. This step
+function will be called after synchronization and as long as the
+previous call in this tick did not return \textsf{true} -- which would
+express that the current solution is complete.
+{\em will need to change that in algorithm environment and refine it}
+
+\begin{verbatim}
+bool step(PartialPlan plan, DecisionStack stack) 
+   if( isInconsistent(plan) ) {
+     relaxPlan(plan, stack);
+     notify("No plan");
+     return true;
+   } 
+   Flaw flaw = chooseFlaw(flaw);
+   if( flaw==NULL )
+     return true;
+   DecisionPoint decision = makeDecisionPoint(flaw, plan);
+   while(1) {
+     while( decision.hasNext() ) {
+       PartialPlan pp = decision.executeNext();
+       if( isInconsistent(pp) ) {
+         decision.undo();
+       }  else {
+          plan = pp;
+          stack.push(decision);
+          return chooseFlaw(flaw)==NULL;
+       }
+     }
+     if( decisionStack.empty() ) {
+       notify("No plan");
+       return true;
+     } else {
+        repeat {
+          decision = decisionStack.pop();
+          decision.undo();
+        } until( decision.hasNext() );
+        return false;
+     }
+   } 
+}
+\end{verbatim}
+
+One critical aspect of this deliberation step is to ensure that at the
+exit of this call the $plan$ is not proven inconsistent. Indeed, it is
+always possible that the next call made by the agent is the
+synchronization which would immediately fail if the $plan$ is
+inconsistent before any solving attempt made for synchronization. 
+For this reason if the plan is proven inconsistent during a
+deliberation step we either backtrack in the decision stack until we
+find a node fo the tree that have remaining alternate solution or
+simply fully relax the plan in situation there's absolutely no other
+alternative  (by removing all the goals and keeping
+only recent observations produced). These constraints are yet again
+related to the fact that for the agent the only crucial functionality
+of a reactor is to resolve successfully its synchronization.
+
+The number of flaw between two steps can evolve for the following
+reasons that are exogenous to deliberation :
+\begin{enumerate}
+\item A new goal has been posted on one of the reactor timeline. This
+  is usually due to another reactor finding a plan and posting the
+  corresponding new objective to this reactor.
+\item The outcome of synchronization created new flaws that did not
+  need to be resolved during synchronization.
+\item A previously requested goal has be recalled by the initial
+  requester. This can happen when the reactor did identify that is
+  initial plan is no longer valid. For example as synchronization
+  proved this plan to be inconsistent.
+\end{enumerate}
+These 3 events will potentially produce new flaws to be resolved while
+any single deliberation step will usually reduce the number of flaws
+to be resolved.
+
+When the number of flaw is 0 A plan solution is found. This has two
+effects:
+\begin{enumerate}
+\item This reactor does not need to deliberate until the next synchronization.
+\item The {\em external} part of this plan can then be posted which
+  eventually (at the end of the current tick if these goals overlap
+  the owner planning window) will be dispatched as goals triggering
+  new deliberation on these reactors.
+\end{enumerate}
 

### 2012-01-24 europa.tex
Added som details about deliberation. Still need a lot of refinement.

@@
 Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains.
-Given a defined conjunctive set of constraints on the variables: C={c1(x1,..,xN), ..., cK(x1,..,xN)}, the objective is to find one or more value assignments to V where all constraints are satisfied.
+Given a defined conjunctive set of constraints on the variables: $C=\{c1(x1,..,xN), ..., cK(x1,..,xN)\}$, the objective is to find one or more value assignments to V where all constraints are satisfied.
 
@@
 
-- The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
+	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
 
-- The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
+	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
 
-- API Layer: makes all the services available so that EUROPA can be used to build applications.
+	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
 
-- Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
 
-1.3 Modeling
+\end{enumerate}
+
+\subsection{Modeling}
+\label{sec:europa:modeling}
 
@@
 
-// Function to solve a partial plan by successive refinement
-// until all flaws are resolved.
-bool solve(PartialPlan plan)
-{
-	// Propagate the constraints to test for inconsistency. If found to be
-	// inconsistent, then we can return false since this is a dead-end i.e. no refinements
-	// to p can yield a consistent plan.
+\begin{verbatim}
 
-	if(isInconsistent(plan))
-		return false;
+01.  bool solve(PartialPlan plan)
+02.   {
+03.      if (isInconsistent(plan))  return false;
+04.
+05.      Flaw flaw = chooseFlaw(plan);
+06.      if (flaw == NULL) return true;
+07.
+08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
+09.
+10.      while(decision.hasNext()) {
+11.          PartialPlan pp = decision.executeNext();
+12.
+13.          if (solve(pp)) return true;
+14.          else d.undo();
+25.      }
+16.
+17.      return false;
+18.  }
 
-	// Non-deterministically choose a flaw from the set of available flaws.
-	Flaw flaw = chooseFlaw(plan);
-
-	// If there are no flaws, then p is complete and we can terminate with success.
-	if(flaw == NULL)
-		return true;
-
-	// Otherwise we formulate a decision point which is a branch in the search space. Each
-	// choice is a particular refinement operation and the DecisionPoint collects all possible
-	// refinement operations for the given flaw.
-	DecisionPoint decision = makeDecisionPoint(flaw, plan);
-
-	// Continue as long as we have something to try
-	while(decision.hasNext()) {
-		// A new partial plan is obtained by application of a refinement operator. Note that the ordering
-    		// over refinement operators to select is a non-deterministic step.
-		PartialPlan pp = decision.executeNext();
-
-		// Recursive call to solve the new planning problem. If successful, then we are done.
-		if(solve(pp))
-			return true;
-		else // Otherwise, retract the last refinement operation
-			d.undo();
-  	}
-
-	// If we arrive here, then we have exhausted all options to resolve the flaw, including the case where
-  	// no options were available initially. Thus the problem cannot be solved.
-  	return false;
-}
+\end{verbatim}
 
 The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
+
+\begin{enumerate}
+	\item Line 3: Propagate the constraints to test for inconsistency. If found to be inconsistent, then we can return false since this is a dead-end i.e. no refinements to p can yield a consistent plan.
+	\item Line 5: Choose a flaw from the set of available flaws.
+	\item Line 6: If there are no flaws, then p is complete and we can terminate with success.
+	\item Line 8: Otherwise we formulate a decision point which is a branch in the search space.
+Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
+	\item Line 10: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
+	\item Line 11: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
+	\item Line 13: Recursive call to solve the new planning problem. If successful, then we are done.
+	\item Line 14: Otherwise, retract the last refinement operation and move on to try the next one.
+	\item Line 17:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially. Thus the problem cannot be solved.
+\end{enumerate}
+
 This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.

### 2012-01-24 results.tex
updates

@@
 % real. CANON will be the predominant driver of the work.
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.35\textwidth]{figs/gulper.jpg}
+\caption{\small{The Gulper water sampler \cite{Bird07} on MBARI's
+    upper water-column Dorado AUV.}}
+\label{fig:gulper}
+\end{figure}
+
+A significant driver of the applicability of our work was the
+implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
+$2$ litre syringe style containers $10$ of which are encased within
+the mid-body section of our Dorado vehicle as shown in Fig
+\ref{fig:gulper}. However, the original design and intended use of
+operation continued to dominated by a naive belief of collecting and
+returning water samples as traditional ship-borne methods often
+employ, namely using pre-scripted methods, albeit on a highly capable
+robot. With the installation of \rx onboard the Dorado with its
+capability to synthesize plans and replan on the fly, more imaginative
+methods of sample acquisition were targeted.
+
+Our early work focused on observing and sampling within Intermediate
+Nephaloid Layers (INLs) \footnote{INLs are fluid sheets of suspended
+  particulate matter that originate from the sea floor
+  \cite{mcphee-shaw2006}} \cite{ryan10}. The location and scale of
+INLs in the water column, which depend upon seasonal and episodic
+variability, are not predictable. In turn, they can be identified by
+their high turbidity and low chlorophyll fluorescence. Initial
+objectives focused on ways to enhance survey results by ensuring as
+large a coverage as possible so that INL fields were observed at
+tighter spaced transects and when no INL is detected, then transect
+spacing in the lawn mower can be incrementally increased. Follow on
+work directly targeted sample acquisition within INLs which are often
+sparse, patchy in large horizontal scales (in Kms) and small vertical
+scales (in meters). Fig. \ref{inl} shows one example from an AUV
+transect from 2005. The intent of both observation and sample
+acquisition was to interrupt the nominal plan-execution cycle and
+inject an action to alter and re-factor the plan dynamically; in one
+instance altering the navigation, the other in actuating a sampler.
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.5\textwidth]{figs/inl.jpg}
+\caption{\small An Intermediate Nepheloid Layer (INL) detected by an
+  AUV mission in the Monterey Bay in August 2005, showing the location
+  of the AUV transect in the bay (left) and optical backscatter image
+  at the 420 Nm frequency showing the INLs patchy nature
+  (right). \small{Courtesy: John Ryan/MBARI}}
+\label{fig:inl}
+\end{figure}
+
+The implied assumption of installing the samplers on an AUV was that
+robotic methods to actuate sample acquisition (with no human control),
+would be competitive with human knowledge of where and when to sample
+a feature given available shipboard data. To support this assumption,
+it required new ways to couple sample acquisition with AUV navigation
+and control. While some simple and ad-hoc approaches \cite{yanwu10,
+  yanwu11} have shown promise, these techniques were tailored to solve
+a specific problem with assumptions on feature structure.  Our
+approach on the other hand, stemmed from the need to build a coarse
+grained computational model of the INL feature where none existed and
+to use that model as a basis to close the \emph{sense-sample-acquire}
+loop within the AUV. \cite{fox2007} documents earlier work in using a
+well-known unsupervised method in classification the \emph{Kohonen
+  Network} or Self-Organizing Maps (SOM) \cite{kohonen} to
+\emph{reactively} acquire samples with plan alteration. In doing so we
+couple the immediacy of feature identification which is necessarily
+reactive, with the deliberation of an onboard temporal planner to
+actually trigger a sampler. This has two benefits. First, it goes to
+the core of having a deliberative approach to mission planning by
+balancing the immediate with that which is projected. Doing so allows
+a more systematic approach in dealing with spatial factors in sampling
+large volumes. Second, it allows other sampling parameters which are
+not directly impacted by perception to be brought to bear; in our
+case, scientists would like to introduce constraints (spatial
+separation, feature bias, distribution of samples when tracking
+multiple features). These are critical not only for science but for
+informed decision making which pure reactive approaches simply cannot
+undertake.
+
+While the early experiments in \cite{fox2007} coupled the classifier
+with the onboard planner, two further novel extensions to sampling
+were added. First, to enhance the classifier model by adding a
+temporal dimension with a Hidden Markov Model (HMM) \cite{rabiner86,
+  Rabiner89atutorial}.  An HMM is a sequential state transition model
+in which the current state is not directly visible, but an
+observation, dependent on the state, is.  Each state has a probability
+distribution over the possible observations.  Therefore the sequence
+of observations gives some information about the sequence of traversed
+states in the HMM. Unlike identification models based on the simple
+classification scheme, HMMs use the dynamics of transition between
+modeled states to exploit sequential information of the sensor
+readings providing more robust formulation for environmental state
+identification. The temporal probabilistic transition then acts as a
+filter on top of the clustering model since it is not only derived
+from it, but it informs the planner of a potential trigger state only
+when the models' transition occurs to an appropriate state of
+recognition of the INL feature \cite{mcgann08b}. Not only was
+deliberation aided by direct perception of the state (of the INL) but
+temporal transition functions added a further level of validation
+between expected states in the INL model. This coupling of Machine
+Learned model with a deliberative planner is novel in the ocean
+sciences and is a continuing direction of investigation \cite{kumar11,
+  sergio12}.
+
+The second innovation involved dynamic resource allocation associated
+with the water samplers \cite{olaya11}. Using a priori sample
+utilities coupled with temporal and spatial information maintained by
+the deliberating planner in-situ, the algorithm dynamically generates
+a policy on sample acquisition during the course of the mission. In so
+doing, utility functions can be used to balance sparcity or abundance
+of the feature encountered with the availability of samplers
+dynamically. Coupled with periodic sub-sampled science data returned
+by the vehicle when on the surface and goal-oriented commanding with
+\rx, such a capability allows judicious use of the limited samplers by
+a scientist from her desktop.
+
+\paragraph {Mixed-Initiative Frontal Tracking:} We exploit the
+in situ decision making capability of \rx in order to find and
+continuously map a thermal gradient. The strategy encapsulated in
+algorithms onboard is described by the following three phases:
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.8\textwidth]{figs/front-mi.jpg}
+\caption{\small{Survey trackline of AUV following a thermal
+    gradient. Image in the middle shows the temperature while the
+    bottom figure shows different phases in the algorithm.}}
+\label{fig:mi-front}
+\end{figure}
+
+\emph{Mapping}: The vehicle performs a yo-yo transect between two
+designated waypoints and sends temperature sampled every $5$ meters to
+shore via satellite. The temperature field is reconstructed from
+interpolation of the transmitted data, and the scientist identifies
+the target front. Using a graphical user interface, she is then able
+to generate a high-level command which represents the front as a
+temperature threshold ($t^{\circ}$) at a given depth ($d$). This goal is
+then appropriately packaged and sent to the AUV via satellite.
+
+\emph{Search}: After the mapping phase and on receipt of this message
+on the vehicle T-REX evaluates the goal, automatically synthesizes a
+set of waypoints by generating a plan and then dispatches the plan to
+execute by having the vehicle swim back and forth at depth $d$ all the
+while reactively evaluating when a temperature threshold $t^{\circ}$
+is crossed. In the process \rx checks both the depth and temperature
+sensed by the AUV; when the depth is around $d$ ($\pm 1$ meter to
+account for vehicle control uncertainty), it then compares the
+temperature read with $t^{\circ}$. When the comparison value changes \rx
+assumes that the front is identified and switches into the next phase.
+
+\emph{Sampling}: The vehicle undertakes a small transect around the
+identified front location while actuating water samplers, one at the
+center of the front with temperature threshold $t^{\circ}$, and one on
+either side of the front. When this transect is completed, \rx
+switches back to the Search phase to follow the frontal evolution.
+The survey is completed either when we reach the maximum mission
+duration or the vehicle returns to its initial waypoint during the
+search phase without crossing the threshold $t^{\circ}$. The latter is
+indicative of the frontal zone having left the survey domain.
+
+The novelty in this application was two-fold; in enabling the
+deliberation to decide when to decide on the phase changes and in
+showing viably how a human being coupled with a goal-oriented
+controller could do problem solving jointly. The approach, is general
+and can be adapted to any feature or scalar field.
+
+\paragraph {Lagrangian Surveys:} More recent experiments with \rx
+controlling the Dorado AUV have centered on the \texttt{CANON}
+(Controlled, Agile, and Novel Observing Network) program at MBARI. The
+fundamental science goals of the program are related to detecting and
+tracking of spatio-temporal coastal ocean features \cite{canon}. The
+dynamic features require a robotic platform to move with it, return
+water samples from across and within gradients and provide contextual
+observations to other drifting instruments. All of these observational
+strategies are tied to vehicle adaptation which is intrinsically tied
+to deliberation and difficult to accomplish otherwise.
+
+Various adaptation methodologies in \texttt{CANON} are still being
+investigated; however what has become a critical goal for the program
+has been to provide contextual environmental information about water
+patches. The form and content of such 'patches' varies; for instance
+initial attempts focused on Harmful Algal Blooms (HABs). They have
+large spatial ($>$ 50 km$^2$) and temporal (days to weeks) extent and
+are visible from space with visible coloring on the ocean surface. The
+direct impact of HABs is the introduction of toxins into the marine
+(and as a consequence human) food chain. In addition,
+\cite{anderson00} and \cite{hoagland06} estimate that the average
+economic impact resulting from HABs in the United States has been
+$\sim$\$75 million from 1987 to 2000. However, the processes behind
+bloom initiation, evolution and collapse is poorly understood in part
+due to the complex interactions between microbial communities and the
+surrounding environment. As a result, our capacity to assess the range
+of potential future scenarios that might result from ocean temperature
+changes, acidification, or nutrient shifts is highly limited.
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.5\textwidth]{figs/dta-app.pdf}
+\caption{\small{Overall setup of CANON lagrangian experiments with a
+    drifting instrument pack. Various combinations of computational
+    methods can be included in the Drifter Tracking App which is a
+    part of the Oceanographic Decision Support System (ODSS)
+    \cite{das11}.}}
+\label{fig:dta-setup}
+\end{figure}
+
+To augment our understanding of such microbial communities, the Dorado
+with \rx has been utilized in a range of activities in a
+quasi-lagrangian mode. As the water mass is being tracked with a
+drifting instrument, the vehicle is expected to move around it. While
+a number of different patterns have been tested in simulation, the
+majority of at-sea deployments have used a square pattern ($3X3$
+km\textsuperscript{2} or a $1X1$ km\textsuperscript{2})\footnote{Often
+  post-hoc reconstruction of the data set for visualization drives how
+  much adaptation the vehicle can be allowed to undertake.}. Plan
+synthesis occurs frequently, but within the context of a drifting
+point generated by a drifter. Fig. \ref{fig:dta-setup} shows the
+general setup of our experiments which can be tailored with the use of
+various computational methods within the \texttt{DTA} app which is
+contained within the ODSS (see more below) \cite{das11}. GPS data from
+the free-floating drifter is sent via satellite and processed onshore
+to determine a projection (speed and bearing) of the drifter. This
+data is sent to the AUV; at the completion of the current survey, the
+vehicle receives this data as a high-level goal which is decomposed
+and synthesized as a survey plan. Currently, the vehicle completes the
+survey without any further dynamic updates from the drifter. In order
+to minimize the range of errors that the survey needs to encapsulate
+(data latency or surface or sub-surface currents for those drifters
+which have a drogue) the estimation is a projection in time and space
+of the drifters centroid sent with this goal from the shore-side app.
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.35\textwidth]{figs/june11-drifter-follow-180-181.pdf}
+\caption{\small{A June 2011 drifter cruise shown in the Earth
+    frame. Deviations in the AUV transect are based on the frame of
+    reference in addition to positional errors as shown in
+    \cite{das11b}.}}
+\label{fig:drifter-errors}
+\end{figure}
+
+Typically this has meant that the vehicle computes a set of waypoints
+for the survey plan, moves to the start waypoint and then executes the
+survey even as the centroid is moving. Positional error of the vehicle
+and uncertainty in the speed and bearing of the vehicle\footnote{In
+  our September 2010 experiment as the drifter moved over the Davidson
+  Seamount \cite{clague10} deviation in the California current
+  resulted in visible directional change that can be seen in
+  Fig. \ref{fig:sep10-transects} at the end of the mission.} exist;
+however as \cite{das11b} clearly articulates, the observations of the
+survey to retain the environmental context require a change in
+reference frames so that the patch-center tagged by a GPS-tracked
+drifter remains within the survey perimeter at all times.
+
+\begin{figure*}
+\centering
+\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.38\textwidth]{figs/sep10-canon-cruise.png}}
+\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.35\textwidth]{figs/september-iterations.pdf}}
+  \caption{\small{The offshore CANON ESP-drifter following cruise
+      \cite{das11b}. Fig. \ref{fig:sep10-ship} shows the ESP drifter
+      foreground with the R/V \emph{Western Flyer} with the image
+      taken from the R/V \emph{Zephyr}. Fig \ref{fig:sep10-transects}
+      shows the 86 km transect of the Dorado following the ESP drifter
+      about 100 Nm from California shores.}}
+\label{fig:gulper}
+\end{figure*}
+
+\begin{figure*}
+\centering 
+\subfloat[]{\label{fig:sep11-temp}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-temp.png}} 
+\subfloat[]{\label{fig:sep11-sal}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-salinity.png}}\\
+\subfloat[]{\label{fig:sep11-flh}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-flh.png}} 
+\subfloat[]{\label{fig:sep11-n2}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-n2.png}} 
+\caption{\small }
+  \label{fig:sept11-wirewalker}
+\end{figure*}

### 2012-01-25 trunk/arch.tex
added a little paragraph presenting T-REX as conceptualization of every reactor as deliberation processes

@@
 \end{itemize}
+
+{\em\color{red} Following is an attempt to present TREX as a big
+  deliberation machine}
+The overall execution cycle of the agent is also abstracted as some
+continuous planning/deliberation cycle between all the
+reactors. Indeed , at the agent interface level, each reactor provided
+only two abstract calls that conceptualize their execution:
+\begin{itemize}
+\item \texttt{synchronize} that takes the last {\em external}
+  observations as argument and will return the {\em internal}
+  observations for this reactor or a failure report.
+\item \texttt{step} that accept new goals requested (if any) on the
+  {\em internal} timeline of this reactor and will execute one step of
+  deliberation. The returned value indicates if a complete plan has
+  been found and in that case will also provide the part of this plan
+  applied to {\em external} timelines of this reactor.
+\end{itemize}
+Both of these two deliberative functions have different scope and
+temporal constraints. The \texttt{synchronize} call is considered as
+atomic and will focus only on state inference from the current tick by
+identifying the current value of the {\em internal} state variables
+from the latest updates on the {\em external} variables. This process
+will then propagate up in the reactor graph at any single tick to
+ensure that all reactors are aware of the current state of the
+world. \texttt{step} relates to a single decision step of the planning
+process which is focused on the future evolution of the state
+variables and it can take multiple steps spanning other multiple
+ticks for a reactor to produce a plan that is both complete and
+valid. when such plan is identified its {\em external} part is then
+dispatched to the corresponding reactors which in turn can start their
+deliberation \texttt{step}s in order to satisfy these new
+objectives. as a consequence physical execution will occur as these
+goals and planning propagate down in the reactor graph until they
+eventually become goals for the lowest level reactor which -- instead
+of deliberating -- may just transform this goal in a physical command
+execution. 
+{\em\color{red} end of the attempt; need to rework the paragraphs
+  below accordingly}
 
@@
 
+Synchronization remains here a deliberation process, and for this
+reason there's always the risk that the search of a solution is not a
+straightforward sequence of unit decision. As such process is required
+by the architecture at every ticks for every reactors presently
+running this need to be carefully designed. At the engine level we
+made careful choices in term of both the focus of the search -- limited
+only to tokens that can potentially overlap the current tick -- and
+its heuristic where the internal state flaw is considered last during
+the search. The later proved to limit the amount of backtrack in the
+search for a solution -- or at least how deep we need to backtrack in
+our decision tree -- similarly the sequence in which we evaluate the
+possible resolution for such a flaw -- presented previously -- was
+selected according to simple assumptions that takes benefit of a
+possibly already existing plan by assuming that if the outcome of
+synchronization appears consistent with our current plan then it
+probably reflects that this reactor plan is executed. Such assumption
+helps direct the internal state evaluation and, in nominal cases (where
+we assume that reactors are acting in concert), proved to be a sound
+solution.
+
+One final aspect to keep into account is that in T-REX point of view
+synchronization is the only critical task of a reactor. Indeed, such
+process is used in order for the agent to identify its
+current state and maintain it consistent throughout the agent
+life-time. A reactor not being able to find a consistent solution
+during synchronization -- meaning that after evaluating all the
+possible solutions provided by its model it did not found any plan
+that is not inconsistent -- not only expresses that this reactor is
+not able anymore to explain the state of the world but may jeopardize
+other reactors that relies on its {\em internal} state. For this
+reason a failure to do synchronization immediately result on the
+agent killing this reactor and notifying any reactor that depends on
+its {\em internal} state variables by posting the special observation
+\textsf{Failed}. By doing so we allow for graceful degradation where
+other reactors can readapt their plan as they receive this
+\textsf{Failed} observation.
+
 \subsubsection{Deliberation : planning for future state evolution}
 \label{sec:arch:plan}
+
+The planning process inside a reactor is directly based on the way
+europa work which was described in previous section. Still some simple
+alterations were made in  order to both include deliberation
+information one reactor provide to the architecture -- such as its
+deliberation latency and its planning look-ahead both expressed in 
+ticks -- and the fact that we potentially need to interleave this
+deliberation with not the synchronization process but also with 
+the deliberation of other reactors\footnote{Current implementation of
+  T-REX is running on a single process and it is the responsibility of
+  the agent itself to emulate the multi-threading of the reactors
+  deliberation and synchronization}. For this reason we 
+decompose the algorithm \ref{alg:europa:solve} inside the reactor by
+having the recursive calls decomposed in single steps. This step
+function will be called after synchronization and as long as the
+previous call in this tick did not return \textsf{true} -- which would
+express that the current solution is complete.
+{\em will need to change that in algorithm environment and refine it}
+
+\begin{verbatim}
+bool step(PartialPlan plan, DecisionStack stack) 
+   if( isInconsistent(plan) ) {
+     relaxPlan(plan, stack);
+     notify("No plan");
+     return true;
+   } 
+   Flaw flaw = chooseFlaw(flaw);
+   if( flaw==NULL )
+     return true;
+   DecisionPoint decision = makeDecisionPoint(flaw, plan);
+   while(1) {
+     while( decision.hasNext() ) {
+       PartialPlan pp = decision.executeNext();
+       if( isInconsistent(pp) ) {
+         decision.undo();
+       }  else {
+          plan = pp;
+          stack.push(decision);
+          return chooseFlaw(flaw)==NULL;
+       }
+     }
+     if( decisionStack.empty() ) {
+       notify("No plan");
+       return true;
+     } else {
+        repeat {
+          decision = decisionStack.pop();
+          decision.undo();
+        } until( decision.hasNext() );
+        return false;
+     }
+   } 
+}
+\end{verbatim}
+
+One critical aspect of this deliberation step is to ensure that at the
+exit of this call the $plan$ is not proven inconsistent. Indeed, it is
+always possible that the next call made by the agent is the
+synchronization which would immediately fail if the $plan$ is
+inconsistent before any solving attempt made for synchronization. 
+For this reason if the plan is proven inconsistent during a
+deliberation step we either backtrack in the decision stack until we
+find a node fo the tree that have remaining alternate solution or
+simply fully relax the plan in situation there's absolutely no other
+alternative  (by removing all the goals and keeping
+only recent observations produced). These constraints are yet again
+related to the fact that for the agent the only crucial functionality
+of a reactor is to resolve successfully its synchronization.
+
+The number of flaw between two steps can evolve for the following
+reasons that are exogenous to deliberation :
+\begin{enumerate}
+\item A new goal has been posted on one of the reactor timeline. This
+  is usually due to another reactor finding a plan and posting the
+  corresponding new objective to this reactor.
+\item The outcome of synchronization created new flaws that did not
+  need to be resolved during synchronization.
+\item A previously requested goal has be recalled by the initial
+  requester. This can happen when the reactor did identify that is
+  initial plan is no longer valid. For example as synchronization
+  proved this plan to be inconsistent.
+\end{enumerate}
+These 3 events will potentially produce new flaws to be resolved while
+any single deliberation step will usually reduce the number of flaws
+to be resolved.
+
+The scope of planning for the reactor plays also a role in the
+deliberation process as we limit deliberation to the sliding temporal
+window $[\tau+\lambda, \tau+\lambda+\pi]$ where:
+\begin{itemize}
+\item $\tau$ is the current tick
+\item $\lambda$ is the specified latency of this reactor. Which is an
+  indicator on the expected maximum time this reactor is expected
+  before being able to produce a plan.\footnote{Do note that this
+    parameter remains indicative and, while it is highly recommended to
+    select a sound value, a failure to produce a plan in this delay is
+    not considered as a critical failure of the reactor}
+\item $\pi$ is the planning look-ahead of the reactor. which indicates
+  how many ticks in the future this reactor is looking ahead while planning.
+\end{itemize}
+This scope allow to filter out any tokens that are either necessarily
+ending before $\tau$ or necessarily starting after. Focusing the
+[planning problem to only the tokens that are in a reasonable
+future. This window is taken into account by the agent which will
+notify the reactor of new {\em internal} goals only when they overlap
+this window but also at the deliberation solver in order to reduce the
+number of tokens to be evaluated and consequently reduce the cost of
+deliberation. As time advance flaws that where initially beyond this
+window will become active flaw that the reactor can then evaluate
+resulting on an apparent continuous planning of the reactor.
+
+When there's no more flaw is present in the current planning scope a
+plan  solution is found. This has two effects:
+\begin{enumerate}
+\item This reactor does not need to deliberate until the next synchronization.
+\item The {\em external} part of this plan can then be posted which
+  eventually (at the end of the current tick if these goals overlap
+  the owner planning window) will be dispatched as goals triggering
+  new deliberation on these reactors.
+\end{enumerate}
 

### 2012-01-25 results.tex
updates

@@
   at the 420 Nm frequency showing the INLs patchy nature
-  (right). \small{Courtesy: John Ryan/MBARI}}
+  (right). \emph{Image courtesy: John Ryan, MBARI}}
 \label{fig:inl}
@@
 various computational methods within the \texttt{DTA} app which is
-contained within the ODSS (see more below) \cite{das11}. GPS data from
-the free-floating drifter is sent via satellite and processed onshore
-to determine a projection (speed and bearing) of the drifter. This
-data is sent to the AUV; at the completion of the current survey, the
-vehicle receives this data as a high-level goal which is decomposed
-and synthesized as a survey plan. Currently, the vehicle completes the
-survey without any further dynamic updates from the drifter. In order
-to minimize the range of errors that the survey needs to encapsulate
-(data latency or surface or sub-surface currents for those drifters
-which have a drogue) the estimation is a projection in time and space
-of the drifters centroid sent with this goal from the shore-side app.
+contained within the Oceanographic Decision Support System (ODSS; see
+below) \cite{das11}. GPS data from the free-floating drifter is sent
+via satellite and processed onshore to determine a projection (speed
+and bearing) of the drifter. This data is sent to the AUV; at the
+completion of the current survey, the vehicle receives this data as a
+high-level goal which is decomposed and synthesized as a survey
+plan. Currently, the vehicle completes the survey without any further
+dynamic updates from the drifter. In order to minimize the range of
+errors that the survey needs to encapsulate (data latency or surface
+or sub-surface currents for those drifters which have a drogue) the
+estimation is a projection in time and space of the drifters centroid
+sent with this goal from the shore-side app.
 
@@
     frame. Deviations in the AUV transect are based on the frame of
-    reference in addition to positional errors as shown in
-    \cite{das11b}.}}
+    reference in addition to positional errors \cite{das11b}.}}
 \label{fig:drifter-errors}
@@
   Fig. \ref{fig:sep10-transects} at the end of the mission.} exist;
-however as \cite{das11b} clearly articulates, the observations of the
-survey to retain the environmental context require a change in
-reference frames so that the patch-center tagged by a GPS-tracked
-drifter remains within the survey perimeter at all times.
+however, the observations of the survey to retain the environmental
+context require a change in reference frames so that the patch-center
+tagged by a GPS-tracked drifter remains within the survey perimeter at
+all times \cite{das11b}.
 
@@
 
+These theoretic results were validated in a five-day long off-shore
+deployment carried out in September 2010, 160 kms off the California
+coast (Fig.~\ref{fig:sep10-ship}). A specialized drifter with a
+genomic sensor \cite{scholin09} hanging $20$ m below, performing
+in-situ identification of micro-organisms was the target. The
+experiment was supported by crews on two support vessels, the R/V
+\emph{Western Flyer} and the R/V \emph{Zephyr}. The \emph{Flyer}
+visited the drifter every four hours to carry out a series of
+ship-based sampling experiments and lab analysis on water samples to
+ground-truth the drifter sensor data. The \emph{Zephyr} was meanwhile
+focused on Lagrangian observation studies using the Dorado with the
+goal to monitor the nutrient budget at the perimeter of a $1$km $X
+1$km water patch around the advecting drifter while the AUV performed
+a transformed box pattern around the drifter. A number of logistical
+issues were kept in mind while designing and executing the
+experiment. Each iteration began with the latest drifter update
+(position and velocity) received from the drifter through a satellite
+link which was transmitted to the vehicle for in-situ adaptation. \rx
+used this input to compute the $5$ waypoints necessary for an
+iteration of the box pattern, with the AUV traveling at constant
+velocity in the earth frame. Waypoints were computed once at the
+beginning of the survey with the AUV surfacing once for every survey,
+with each survey lasting $\sim1$ to $1.5$ hours. A total of 60 surveys
+were attempted over the course of the $5$
+days. Fig. \ref{fig:sep10-transects} shows the overall transects over
+the course of the mission, with gaps in between days when the AUV was
+being recharged. Each deployment lasted $\sim12$ hours. The black dots
+in the figure show the beginning of each iteration when drifter
+updates were received by the AUV.
+
+\begin{figure}[htpb]
+\centering
+\includegraphics[width=0.7\textwidth]{figs/oct10-ballet.pdf}
+\caption{\small An October 2010 robot ``ballet'' where a smaller
+  slower AUV tracked a patch center, sending periodic updates to the
+  Dorado using a setup shown in Fig. \ref{fig:dta-setup}. The Google
+  Earth post-hoc image shows red track lines of the \rx commanded
+  Dorado with the yellow transects of the LRAUV. Yellow dots show the
+  track line of a simple GPS-enabled drifter used for seeding the
+  experiment over the Monterey Canyon with Santa Cruz, CA to the right
+  of the figure.}
+\label{fig:ballet}
+\end{figure}
+
+Successive lagrangian experiments also followed a more innovative and
+robotically choreographed ``ballet'' between two AUVs, the Dorado and
+the LRAUV \cite{bellingham10} in October 2010 as part of another
+series of \texttt{CANON} field experiments. The slower vehicle with a
+fluorometer tracked a targeted chlorophyll patch seeded by a
+GPS-enabled drifter, sending periodic updates of the patch centroid to
+shore via the ODSS following a setup similar to
+Fig.\ref{fig:dta-setup}, to the Dorado running \rx.
+
 \begin{figure*}
 \centering 
-\subfloat[]{\label{fig:sep11-temp}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-temp.png}} 
-\subfloat[]{\label{fig:sep11-sal}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-salinity.png}}\\
-\subfloat[]{\label{fig:sep11-flh}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-flh.png}} 
-\subfloat[]{\label{fig:sep11-n2}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-n2.png}} 
-\caption{\small }
+\subfloat[Temperature]{\label{fig:sep11-temp}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-temp.png}} 
+\subfloat[Salinity]{\label{fig:sep11-sal}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-salinity.png}}\\
+\subfloat[Fluorescence line height]{\label{fig:sep11-flh}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-flh.png}} 
+\subfloat[Nitrate]{\label{fig:sep11-n2}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-n2.png}} 
+\caption{\small Interpolated results of a September 2011 \rx mission
+  in the Monterey Bay with the profiling \texttt{Wirewalker}
+  \cite{rainville01,pinkel11} instrument. Plots show the instrument
+  was entrained between two fronts. \emph{Image courtesy: Rishi
+    Graham, MBARI}}
   \label{fig:sept11-wirewalker}
 \end{figure*}
+
+More interesting experiments are being carried out as a basis to track
+more dynamic features such as ocean fronts \cite{fronts11} and
+tracking more capable drifting platforms. In particular experiments
+with a profiling CTD instrument, the \texttt{Wirewalker}
+\cite{rainville01,pinkel11} are resulting in new modalities of making
+Lagrangian and Eulerian observations. Fig. \ref{sep11-wirewalker}
+shows one such experiment in September 2011 with a GPS-enabled
+\texttt{Wirewalker} was tracked by the Dorado running \rx. We are
+currently investigating spatio-temporal correlations of CTD data
+obtained by the \texttt{Wirewalker} and tying it to the observations
+made in the contextual environment by the Dorado's sensors.

### 2012-01-25 trunk/results.tex
updates

--- 2012-01-25 results.tex
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figs/inl.jpg}
+\includegraphics[scale=0.65]{figs/inl.jpg}
 \caption{\small An Intermediate Nepheloid Layer (INL) detected by an
@@
 \texttt{Wirewalker} was tracked by the Dorado running \rx. We are
-currently investigating spatio-temporal correlations of CTD data
-obtained by the \texttt{Wirewalker} and tying it to the observations
-made in the contextual environment by the Dorado's sensors.
+currently investigating spatio-temporal correlations \cite{gneiting07}
+of CTD data obtained by the \texttt{Wirewalker} and tying it to the
+observations made in the contextual environment by the Dorado's
+sensors.

### 2012-01-25 related.tex
updates

@@
-\section{Related Work}
+\section{Background and Related Work}
 \label{sec:related}
@@
 % for robotics and AI.
+
+A distinct problem in Marine Robotics has been the use of the
+overloaded term 'autonomy'. Not only does the notion transcend
+different disciplines within engineering in this domain (e.g. the
+Control Engineering community sees it distinctively from those in AI)
+but users of marine platforms in oceanography as well as in maritime
+defense conflate the methodology of use (tethered vs. untethered) with
+their control. In this chapter, our notion of 'autonomy' deals
+explicitly not only with the notion of dealing with the \emph{sense,
+  plan and act} paradigm, but also the need to perform
+\emph{computational search} between different action outcomes, an idea
+central to the field of AI. Consequently, our literature survey here
+is targeted towards a more focused and (to this chapter) relevant part
+of the field of computation.
+
+With few exceptions, most AUV control systems have been variations of
+reactive approaches.  Typically, waypoint-based pre-defined mission
+designs are uploaded to the AUV; specialized code fragments called
+\emph{behaviors} are designed for the specific mission and a choice of
+behaviors for the mission are used on the computational stack
+\cite{bellingham94}. Swapping in and out of these behaviors using
+conditionals or cleverer and quantitative approaches such as
+\cite{Benjamin:2004} forms the basis for adaptation and safety in the
+vehicle. Where such swapping does not aid adaptation, individual
+behaviors are tweaked to generate some measure of responsiveness to
+the surrounding environment \cite{yanwu08} to generate interesting
+observations for oceanographers, yet technically naive in terms of
+scalability and systematicity. \cite{Benjamin2006} is clear about the
+extant of adaptation in stating:
+
+{\footnotesize
+  \begin{quote}
+\small \emph{The primary difficulty often associated with behavior-based control
+concerns action selection - namely how to ensure the chosen action
+really is in the best overall interest of the robot or vehicle.}
+\end{quote}
+}
+
+Such techniques have proved their mettle in the first round of use of
+AUVs; they've provided operators a simple way to handle control
+complexity of the vehicle while returned data at far higher resolution
+than would have been possible using traditional ship-based methods and
+at substantially reduced amortized costs of platform operation and
+ownership over multiple years. AUV operators and clients for their
+data products have reasons to be well satisfied with results thus
+far. However, as hardware and more cost-effective sensors become more
+robust to the environment and as mission durations increase (as
+demonstrated by glider experiments such as \cite{rucool11}), sustained
+presence in the ocean requires the ability to deal with off-nominal
+conditions, that balances the needs of high-level mission management,
+low-level navigation, instrument control and detection of unstructured
+and poorly understood phenomena. Earlier reactive methods are unable
+to perform such a balancing act without skewering the foundations of
+the software development methodology leading to code bloat or worse.
+\cite{carreras06} provides a reasonably good overview of AUV control
+architectures that essentially is based on reactive control methods.
+
+\begin{wrapfigure}{r}{0.2\textwidth}
+ \centering
+ \includegraphics[scale=0.3]{figs/planner-arch.pdf}
+ \caption{\small A general architectural block diagram for an
+   Artificial Intelligence based Planner.}
+ \label{fig:planner}
+\end{wrapfigure}
+
+A word about generalized planning. Traditionally planning has been
+considered computationally intensive (the general planning problem can
+be worse than NP-complete \cite{ghallab04}). A major reason for this
+belief had to do with the role of the planner (which was assumed to be
+generative) and its place in the architecture (infrequently called to
+re-plan the \emph{entire} mission in off-nominal conditions) within a
+\emph{sense-plan-act} paradigm. This can limit the reactivity of a
+robot especially when the environment could change at a rate faster
+than the planner can plan.
+
+The architecture of a typical AI planner is along the lines of
+Fig. \ref{fig:planner}. Given a \emph{domain model} which encodes the
+platform constraints in a higher-level language, an \emph{initial
+  state}, a high-level objective or end \emph{goal}, the \emph{plan
+  database} (or \texttt{plandb}) is where all assertions are tracked
+and maintained. Theorem proving in the form of satisfaction of axioms
+embodied in the domain model occurs within the \texttt{plandb}. It is
+the job of the \emph{search engine} to provide the inferential
+mechanism for state space exploration potentially aided by heuristics
+encoded for \emph{search control}. In some complex domains
+\cite{mus98} \emph{external modules} can provide additional domain
+expertise which can augment the domain model. The end result is plan
+formulation using a multitude of approaches either using forward or
+backward search or using more generalized \emph{partial order} methods
+\cite{ghallab04}.
+
+Deliberative techniques for robotic control on the other hand
+contrast in providing a distinct set of advantages:
+
+\begin{enumerate}
+
+\item action selection is driven axiomatically based on a systematic
+  assessment of a range of causal conditions. An action is selected
+  for instantiation only when there is causal support in the form of a
+  constraint in a deterministic model of the vehicle's operation.
+
+\item the system is goal directed which forms an objective towards
+  which the system is expecting to converge providing a foci in action
+  selection.
+
+\item scaling to different (computational) problem size is dependent
+  on incremental model building rather than being concerned about
+  nuanced interactions between behaviors.
+
+\item systematic software engineering methodologies such as spiral
+  development \cite{boehm86} can further be used for scaling up the
+  task as demonstrated by \cite{DS1report}.
+
+\item when interactions between actions and/or behavior fragments do
+  (and must) occur there are explicit constraints encoded in the model
+  that must be computationally satisfied contrasting with carefully
+  crafted strategies to instantiate behaviors on a stack. Such
+  explicit recording of constraints in the model in turn aids
+  long-term software maintenance and sustaining engineering.
+
+\end{enumerate}
+
+There are costs associated with such inference based control are also
+not inconsequential. Foremost among them is the steep learning curve
+that comes with the design of the domain model. With limited support
+tools necessary for knowledge capture and design of the constraints
+often in stylized higher-level languages (see \cite{NDDL} for an
+example), it requires the modeler to think in alternative ways to do
+problem solving. More importantly it is often necessary for the
+modeler to be exposed to the internal representation of the planner
+and how it performs search. THe level of expertise is often well above
+what would be expected of a typical well-rounded Computer Science
+graduate student.
+
+However this is mild criticism given the end benefits and the
+scientific and engineering goals associated with adaptive and
+persistent observation for marine robots whether for civil or military
+applications. Given the general problem-solving nature of such
+deliberative systems, there is also a substantial overall benefit for
+the field of robotics.
+
+\subsection{Evolution of AI-based Robotic Planning: A Necessary Digression}
+
+Planning and plan execution are not new to the overall field of
+robotics. Early motivation of building computational mechanisms for
+decision making were intended to be deployed on robotic devices by
+very definition. The seminal volume \cite{computersthought} very
+clearly articulates how physical manifestation of robots could be
+controlled for task planning and execution. The very first planner
+\cite{green69} was an exercise in computational theorem proving with
+early state space planners popularized by \texttt{STRIPS}
+\cite{strips71} dominating the academic landscape. An early
+demonstration of planning as a situated agent was the highly
+influential \texttt{Shakey} the robot experiment
+\cite{shakey84}. Subsequent work driven largely by defense funding in
+the United States sequestered the discipline in academic labs
+principally applied for algorithmic development and if embedded then
+in testing within the confines of modest indoor settings. The leap to
+applying them for real world problem solving is relatively new and
+driven largely by NASA applications pioneered by \cite{mus94,mus98,
+  jonsson00, rajan00, chien05, bresina05}.
+
+The dominant approach for building agent control systems utilize a
+$3$-layered architecture (see \cite{gat98} for a well thought out
+rationale), notable examples of which include IPEM
+\cite{AmbrosIngerson88}, ROGUE \cite{Haigh98}, the LAAS Architecture
+\cite{alami:1998p820}, the Remote Agent Experiment \texttt{RAX}
+\cite{mus98} and the Autonomous Spacecraft Experiment \texttt{ASE}
+\cite{chien99} (see \cite{Knight01} for a survey). Early efforts were
+cognizant of if absent with, dealing with time and resources for such
+experiments.
+
+Representationally, the first paper on managing time systematically
+within AI Planning frameworks was \cite{dean87}; subsequent efforts by
+Boddy and others \cite{Dean88,Boddy93} refined the notion of using a
+temporal database and the conceptualization of temporal
+intervals. Using these base-level concepts Muscettola and others
+\cite{mus92, ghallab94, laborie95, cesta96} developed the notion of
+plan/schedule envelopes using the notion of state variable
+instantiated timelines. Simultaneously work by Dechter et.al
+\cite{dechter91} resulted in efficient temporal constraint propagation
+which systematically defined the notion of Simple Temporal Networks
+(STNs) and path and arc consistency algorithms. This work neatly put
+together earlier efforts by Mackworth and Freuder \cite{mackworth77,
+  mack85} and the Constraints community on propagation algorithms. In
+parallel work in the UK as a result of US DARPA funding, resulted in
+the design of \texttt{O-Plan} \cite{currie91}, which leveraged the
+notion of constraints and metric time, but within a Hierarchical Task
+Network (HTN) representation. All these efforts were soon after Vere’s
+paper on planning and time \cite{vere83} which had a profound effect
+on the AI Planning community. In particular the work of Muscettola
+\cite{mus94} was embraced by NASA for telescope scheduling. Around the
+same time period, researchers at LAAS came up with an architecture
+\cite{alami:1998p820} which encapsulated the \texttt{IxTeT} planner
+\cite{ghallab94}. It featured advanced concepts for a temporal planner
+with the notion of chronicles, plan recognition for partial plan
+fragments and early use of systematic resource
+constraints. Development on \texttt{IxTeT} has continued albeit at a
+slower pace with the contribution of various PhD students’ thesis.
+
+\subsection{The Remote Agent Experiment and beyond}
+
+While \texttt{HSTS} \cite{mus94} was being implemented as a
+ground-based planning tool for decision support for US based space
+observations (EUVE and Cassini) \cite{mus95}, the opportunity to fly
+onboard NASA’s New Millennium Deep Space $1$ came about. The design of
+the Remote Agent Experiment \texttt{RAX} \cite{pell97, bernard98,
+  pell98, mus98, DS1report, rajan00, jonsson00} was a direct off-shoot
+of this effort where \texttt{HSTS} (written in LISP) was flown on
+board and successfully commanded the spacecraft for a week in May
+1999. There were a number of significant software engineering lessons
+learned with the HSTS technology as integrated for the RAX:
+
+\begin{enumerate}
+
+\item Constraint-based representations were not only sufficient for
+  plan synthesis, but valuable during debugging and development as a
+  means of building a viable domain model.
+
+\item Domain models when separated from the search engine as
+  articulated by the model-based approach \cite{williams96a} ensured
+  that sufficient effort would be targeted on human validation of the
+  model building phase while ensuring that search engine stability
+  across applications and domains resulted in lower cost to deploy.
+
+\item Flexible temporal representation generating a range of plans as
+  against a single plan, were robust for plan execution.
+
+\item If planning and execution were disconnected (as in
+  \texttt{RAX}), dispatchabilty \cite{mus98a} and controllability
+  \cite{morris00} issues within temporal plans need be addressed. In
+  \texttt{RAX}, a post-processing step was added to counter these
+  effects. It was clear that execution needed access to the planners
+  database and be able to propagate execution time constraints prior
+  to command dispatching.
+
+\end{enumerate}
+
+It was with the demonstration of \texttt{RAX} $65$ Million miles in
+outer space, that temporal reasoning using methods came to the
+forefront of AI planning and for situated agents. Lessons learned from
+\texttt{RAX} led to the development of \texttt{IDEA}
+\cite{mus02,mus04} with the central theme of using planners
+collectively for problem solving arose. Another critical theme was
+iteratively and incremental plan repair of an anytime plan
+\cite{Zaimag96} as proposed by \texttt{CASPER}
+\cite{chien00}.
+
+This last lesson in particular had a lasting impact with the
+observation that planning and execution are strongly
+inter-related. This core concept was behind the next generation of the
+Remote Agent, called \texttt{IDEA} (Intelligent Deployable Execution
+Agents) where planning and execution were \emph{intertwined} within a
+single domain model that spanned the most abstract (for high-level
+planning) to the least (for diagnosis). \texttt{IDEA} \cite{mus02,
+  mus04} agents were expected to interact to achieve plan formulation;
+however there was no systematic framework for formally governing these
+interactions. \texttt{IDEA} was also computationally heavy and
+required substantial effort to customize for an application domain. It
+did move to a retooled version of \texttt{HSTS}, now called EUROPA
+\cite{frank2003, barreiro09} and ported to the C++ language.
+
+While \texttt{IDEA} was being deployed as a coordinating executive for
+earth analog rover deployments \cite{wetter05} a separate development
+was undertaken to deploy the EUROPA planner for NASA’s Mars
+Exploration Rovers (MER) mission. The \texttt{MAPGEN}
+(Mixed-initiative Activity Plan GENerator) planner \cite{bresina03,
+  aichang04, bresina05, bresina05a} as a decision support tool in the
+mixed-initiative mode; \texttt{MAPGEN} continues to be used to this
+day to command the twin rovers on the Red
+Planet. % While EUROPA2 was not
+% designed as an embedded planner, NASA undertook a bottom-up
+% re-implementation of the planner and substantially increased its
+% performance. EUROPA2 was then released as open source to the AI
+% community at large [Europa].
+
+The Autonomous Sciencecraft Experiment (\texttt{ASE}) \cite{chien99,
+  chien03} was another autonomy demonstration in space, this time for
+a more observable vehicle, the EO-1 in Earth orbit. It encapsulated a
+classical 3-layered architecture like \texttt{RAX}. The fundamental
+functional difference was that goals could not only be sent by a
+ground segment, but also by an onboard science driver which could
+encapsulate a range of Machine Learned feature detectors which could
+trigger the planner. However ASE was not an integrated system; the
+\texttt{CASPER} planner \cite{chien00} would generate plans which were
+executed by a separate commercially available rule-based system
+\cite{icl}. \texttt{CASPER} touts a \emph{continuous} paradigm for
+planning; however this incremental modification simply takes the
+overall mission plan and selectively plans near-term at a more
+detailed level. The system also uses iterative-repair as a way to fix
+plans; however this too is derived from a model of the mission and
+spacecraft constraints which allow simple reconfiguration modes
+towards replanning. Robust software engineering is also not an
+important aspect targeted by \texttt{ASE} given the disparate models
+between the planner and the rule-based executive. With such a system,
+real-time updates from the environment cannot be accommodated since
+the planner is a monolithic computational environment. The rate of
+updates from the science drivers is roughly in the order of an earth
+orbit on the EO-1 vehicle. As a result, the ASE is only an incremental
+update on RAX and is not suitable for domains where complex automated
+reasoning and asynchronous environmental observations need to be
+factored in deliberation.
+
+\texttt{TCA} \cite{simmons94} is another framework for robot control
+and has similar motivations to \texttt{RAX} and
+\texttt{ASE}. Reactivity and deliberation are also key to \texttt{TCA}
+and a systematic approach to development led to its design. However it
+suffers from three key weaknesses. First it has a weaker
+representational framework of Task Trees which require representation
+of tasks within cleanly formulated hierarchies. Constraints between
+branches within a task hierarchy are allowed; timing between leaves of
+different trees is however not possible. Second, its temporal
+framework does not deal with flexibility; time points are fixed and
+while partial orders are possible, it does deal with execution time
+uncertainty. Third, while \texttt{TCA} shares the notion of using
+domain expertise of different planners, it does so thru a centralized
+message passing mechanism. While this allows disparate code-bases to
+talk to each other thru IPC (a inter-process communication or message
+passing mechanism), failure in the centralized controller is
+tantamount to a system crash.
+
+The \texttt{ReSSAC} project at ONERA \cite{teichteil07} has similar
+motivations in controlling aerial UAV platforms for autonomous Search
+and Rescue. Like \texttt{TCA}, the project uses a supervisor to
+control planning and at the same time decouples the deliberation and
+reactive components from the lower level functional layer. This
+deliberate separation is partly because of the use of optimizing MDP’s
+(Markov Decision Processes \cite{mdp93}) for planning which uses
+probabilistic state transitions to determine the next course of action
+based on perceived state. To overcome the typical problem of state
+space explosion with MDPs, a local heuristic is used in order to
+generate reachable goal states incrementally. This work however does
+not deal with metric time with a modestly simple model.
+
+\texttt{CIRCA} \cite{musliner95} is another effort to bring decision
+making for real-world problems with the augmented need to have
+\emph{verifiable} controllers synthesized in-situ to ensure that
+within explored states, there are no erroneous transitions to
+dangerous states by asynchronously generating Test-Action-Pairs
+(TAPs).  These are annotated production rules consisting of a set of
+tests (or preconditions) and a set of actions to take if all the tests
+return true. TAPs are synthesized and scheduled by \texttt{CIRCA} and
+provide a viable way to deal with a dynamically changing
+world. \texttt{CIRCA} however differentiates between reasoning about
+time and reasoning in real-time with the implication that reasoning
+necessarily requires substantial computation for state space
+exploration which negates the real-time (and thus real-world)
+impact.
+
+A key and early concern that dominated planning was that of
+scalability.  The planning cycle in many approaches was monolithic
+often making fast reaction times impractical when necessary for
+embedded robotic agents. Many of these systems (which were
+$3$-layered) also utilized very different techniques for specifying
+each layer in the architecture resulting in duplication of effort and
+a diffusion of knowledge.  The work we highlight in this chapter
+builds on the approach used by \texttt{IDEA} \cite{mus02, mus04} in
+utilizing a collection of controllers, each interleaving planning and
+execution within a common framework. \texttt{IDEA} however, provided
+no support for conflict resolution between controllers, nor does it
+provide an efficient algorithm for integrating current state within a
+controller, relying instead on a possibly exponential planning
+algorithm. Efficient synchronization of state in a partitioned
+structure is fundamental to making the approach effective in practice.
+
+Our system, \texttt{T-REX}, the focus of this chapter, is an agent
+control structure formally defined as a composition of coordinated
+control loops with a recurring sense-plan-act (SPA)
+cycle. \texttt{T-REX} was designed after carefully evaluating lessons
+learned from \texttt{IDEA} and owes substantially to the notion of
+partitioned problem solving from that effort. However this
+partitioning is systematic and methodologically principled where we
+manage the information flow within the partitioned structure to ensure
+consistency in order to direct the flow of goals and observations in a
+timely manner. The resulting control structure improves scalability
+since many details of each controller can be encapsulated within a
+single control loop.  Furthermore, partitioning increases robustness
+since controller failure can be localized to enable graceful system
+degradation, making this an effective \emph{divide-and-conquer}
+approach to the overall control problem.
+
+A recent variation of the key idea of such controller composition
+derived from \texttt{T-REX} is \texttt{ROAR} \cite{degroote11} where
+lower-level functional modules that control specific sensors are
+organized within a graph hierarchy. One reason they cite such a
+structure is essential is in dealing robustly to failure; an
+off-nominal condition will allow rapid graph traversal to identify
+alternative ways in which sensing tasks can be executed and therefore
+aiding in responsiveness. However, it is unclear whether to date this
+system actually deliberates and whether it is actually instantiated on
+a real platform for control.
+
+% ``The whole partition must be accordingly reorganized: this kind of
+% construction does not scale well over a large variety of missions,
+% missing the objective of a versatile architecture for robots.''

### 2012-01-26 trunk/results.tex
updates

@@
 sensors.
+
+\paragraph {Mixed-Initiative Autonomy}
+
+\begin{figure*}
+\centering
+\subfloat[The ODSS Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+\subfloat[Current Implementation for \texttt{CANON}]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
+\caption{\small The Oceanographic Decision Support System (ODSS) is a
+  single portal for planning, visualization, situational awareness,
+  collaboration and archival of information before, during and after
+  field experiments. The ODSS has been used since mid-year 2010 for
+  \texttt{CANON} field experiments.}
+\label{fig:odss}
+\end{figure*}
+
+Large scale oceanographic field experiments akin to those targeted by
+\can are usually logistically constrained driven largely by ship
+schedules.  Once a feature set has been identified for scientific
+observation and science goals have been collated, the appropriate
+robotic assets need to be assigned and deployed for observation and
+sampling. This asset allocation in turn depends on parameters like
+sensor payload, the speed of the robot \situ in comparison to the
+spatio-temporal scales of the evolution of the phenomena being
+sampled, the number of assets in a class, the kinds of measurements
+needed (e.g synoptic or close-up), logistical considerations for
+deployments such as battery charge needed or distance/time to deploy,
+impact to long-range deployment plans among others. All drive the
+strategy to balance short and long-term scientific and engineering
+needs. Our \can deliberations during the 2010 field season for example,
+were driven by all of the above \cite{das11}. Often these strategies
+were rebalanced during each decision-making cycle, especially in light
+of opportunistic science goals. For example, gliders running at $0.5$
+knots were often retargeted when surface support vessels noted the
+apparent dispersal of the bloom to subsurface layers when HABs were a
+\texttt{CANON} target. However, their retargeting was carefully
+monitored to balance the long-term desire of obtaining time series
+within fixed tracks and with the need to ensure these assets were not
+exposed to strong currents. Powered AUVs were targeted to chase after
+advecting blooms, but the decision on their deployment locations and
+survey areas too close to the surf-zone were avoided. In
+patch-tracking experiments, the deployment of drifters close to shore
+and where depths of $20m$ or less were encountered, was avoided even
+when contrary to scientific intent. The need to obtain high-quality
+data for post-hoc analysis was also driven by the response time of
+data returned by experiment partners.
+
+\begin{figure*}
+\centering
+\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.3\textwidth]{figs/october-canon-meeting.png}}
+\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.33\textwidth]{figs/virtual-drifter-tool.png}}
+\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
+\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with the
+  ODSS in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
+  device.}
+xf\label{fig:odss2}
+\end{figure*}
+
+These distinctions make some vehicles more or less suitable to address
+certain science goals and suggest some degree of intelligent planning
+in the assignment strategy. Further, even if partial plans can be
+synthesized automatically, the need to capture qualitative 'intent'
+and desires in the final outcome of an asset allocation strategy
+drives us towards a human-in-the-loop approach for asset
+allocation. The vast uncertainty and complex interaction between
+unknown quantities in the ocean environment dictate that any
+comprehensive experimental methodology must rely in some measure on
+human reasoning and adaptability.  Furthermore, the multidisciplinary
+nature of ocean experiments suggests collaboration between multiple
+human actors. Human cognitive capability then to capture these
+uncertainties, evaluate options, factor intent and synthesize data
+needed for asset allocation becomes increasingly important. Therefore,
+mixed-initiative approaches are viable as a means to apportion problem
+solving and inference between human and computational approaches. A
+viable example was shown in the command and control of the twin Mars
+rovers in \cite{aichang04,bresina05,bresina05a}\footnote{The MAPGEN
+  system uses the same \eu planner used in \rx. It continues to be
+  used routinely to this day on the MER mission.}.
+
+Additionally, our work on ODSS is driven by a vision of how marine
+field robotics will likely be conducted in the future and the
+necessary computational infrastructure that will augment decision
+making, onshore as well as on-board. We believe ultimately ODSS will
+combine capabilities inferential and data gathering and analysis
+capabilities and more uniquely, provide an event response capability
+that could trigger the deployment of a robot and/or adapt its sampling
+strategy, all from a scientist's desktop on shore. Data returned from
+robotic assets at sea will be filtered for signals associated with
+feature(s) of interest along with information on its contextual
+environment. The association between the signal and a prospective
+event will be undertaken through tagging by experts and machine
+learning techniques used in recommender
+systems\cite{Adomavicius05}. Adaptive sampling strategies will then
+balance decision-making \situ with that done onshore with
+computational models and human agents, the subject of a new NSF
+project in which the authors are participating.
+
+Operationally, the ODSS has become a software system that the
+scientist consults during a \can field campaign. S/he consults
+up-to-date model outputs, synoptic views of ocean color from remote
+sensing data and weather conditions for the next $24$-hour planning
+cycle and then queries the system to compute candidate sampling
+locations. In the near future, the ODSS will consult its situational
+awareness database, incorporate model outputs projecting these outputs
+into the space-time region of interest to generates candidate
+hypotheses with associated utility metrics. With the scientist's
+guidance the automated planner would then determine the most viable
+robotic option(s) to be dispatched (underwater, surface or aerial),
+given available resources, distance of travel, required payload and
+the environment. For instance, unpowered autonomous gliders would be
+chosen for longer-duration missions with smaller payload requirements
+in areas of moderate to weak ocean currents. The automated planner
+would balance these criteria to compute a viable trajectory plan for
+the selected vehicle. Vehicles with more controllability would be
+given a specified area to survey and the path generated \situ. Goals
+are communicated to the robot after analysis on shore for selective
+targeting of features by the conglomerate of vehicles in the ocean
+being tracked by the ODSS. Thus far, the ODSS has provided a simple
+yet functional tool for scientists to use in the daily complex
+decision making and the following experiment and survey design
+needs. It has helped train scientists to think of expanding their
+repertoire of tools for the scientific process and in the process it
+enabled a mixed team of engineers and scientists to interact at the
+functional level of the systems' capabilities.

### 2012-01-27 intro.tex
updates

@@
 inter-disciplinary collaboration between biologists, ecologists,
-geneticists and roboticists on the CANON (Controlled Agile Novel
+geneticists and roboticists on the \can (Controlled Agile Novel
 Observation Network) \cite{canon} has driven autonomous system
@@
 capabilities in the Space domain
-\cite{mus98,rajan00,ai-chang04,bresina05} for NASA. Lessons learned
+\cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons learned
 have been applied to the world of marine robotics which we bring

### 2012-01-27 trunk/future.tex
update
updates

@@
 % multiple vehicles is an important result of this activity.
+
+% 1.6 Future Work?
+% explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).
+
+% The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space
+
+The Ocean Sciences are at a cusp, straddling between traditional
+ship-based exploration with newer observatory-based methods. In the
+United States the NSF funded Ocean Observatories Initiative \cite{ooi}
+and regional bodies such as \texttt{CeNCOOS} \cite{cencoos} are
+investing in new observational methods and technologies which propose
+to aid oceanographers by producing substantially higher-resolution
+data. These will use moorings, cabled observatories and glider fleets
+generating near real-time data from one or more locations as a
+continuous stream. The tack chosen by \can has been to complement such
+methods with the aim of developing methods for a \emph{portable
+  mobile} observatory which is dependent on autonomous platforms. In
+this context what we advocate in this chapter is an initial step in
+making robotic platforms more adaptive, scaling them towards fleets of
+vehicles (underwater or on the surface) and most importantly using
+them to solve the larger challenge of Sampling spatio-temporal fields.
+
+While our trajectory of research in the oceanographic domain follows
+that in spacecraft autonomy, from onboard \cite{mus98} to
+mixed-initiative methods \cite{bresina05}, the challenges are
+significantly different with the environment playing a far larger
+role. Moreover, prediction and projection of a future course of action
+for a robot given the level of uncertainty calls for representational
+and architectural methods which would enable adaptation at different
+levels of abstraction, from mission-planning to low level actuation.
+
+Our efforts were initially targeted to full robotic autonomy; while
+this is an important goal that we continue to strive towards, it is
+increasingly clear that human-in-the-loop methods also have a sizable
+role to play in ocean exploration and by extension to Marine
+Robotics. Not only is an embedded \rx an important element of the
+effort towards smarter vehicles, but we are working towards an
+implementation of \rx which will perform multi-vehicle plan synthesis
+for the \od.
+
+A key role that \rx has played and will continue to play is towards
+\emph{event response} scenarios. In Section \ref{sec:results} we show
+the role Machine Learning (ML) techniques have played thus far in
+in-situ plan adaptation. This interaction between Planning and ML we
+believe will play a significant role going forward in the general
+theme of exploration and discovery. Interpretation of incoming sensor
+data to deal with scientific surprise and readapt for opportunistic
+science will be critical to dealing with undersampling the
+ocean\footnote{Walter Munk has famously stated \emph{``Most of the
+    previous century could be called a “century of undersampling”''}
+  -- Testimony to the U.S. Commission On Ocean Policy, 18 April
+  2002}. Much of what we know of the ocean was derived by exploring in
+the blind. We anticipate longer duration robotic vehicles with enough
+onboard computational capacity to run deliberative agents such as \rx
+make water-column measurements and reliably track and characterize
+dynamic features. While our experiments have already shown such
+capabilities the exploration Vs. exploitation tradeoff is better
+determined when vehicles can make choices over longer time periods
+(weeks and months) to study the impacts of bio-geochemical interaction
+driven by longer duration physical forcing.
+
+Given our poor understanding of the coastal ocean in particular,
+mixed-initiative methods of command and control for Sampling with
+heterogenous assets is an important goal that is clearly at our
+doorstep. We believe deliberation is small yet critical part of that
+unfolding research problem for AI and Robotics.

### 2012-01-27 trunk/results.tex
updates

@@
 \paragraph {Lagrangian Surveys:} More recent experiments with \rx
-controlling the Dorado AUV have centered on the \texttt{CANON}
-(Controlled, Agile, and Novel Observing Network) program at MBARI. The
-fundamental science goals of the program are related to detecting and
-tracking of spatio-temporal coastal ocean features \cite{canon}. The
-dynamic features require a robotic platform to move with it, return
-water samples from across and within gradients and provide contextual
+controlling the Dorado AUV have centered on the \can (Controlled,
+Agile, and Novel Observing Network) program at MBARI. The fundamental
+science goals of the program are related to detecting and tracking of
+spatio-temporal coastal ocean features \cite{canon}. The dynamic
+features require a robotic platform to move with it, return water
+samples from across and within gradients and provide contextual
 observations to other drifting instruments. All of these observational
@@
 
-Various adaptation methodologies in \texttt{CANON} are still being
+Various adaptation methodologies in \can are still being
 investigated; however what has become a critical goal for the program
@@
 \includegraphics[width=0.5\textwidth]{figs/dta-app.pdf}
-\caption{\small{Overall setup of CANON lagrangian experiments with a
+\caption{\small{Overall setup of \can lagrangian experiments with a
     drifting instrument pack. Various combinations of computational
     methods can be included in the Drifter Tracking App which is a
-    part of the Oceanographic Decision Support System (ODSS)
+    part of the Oceanographic Decision Support System (\od)
     \cite{das11}.}}
@@
 various computational methods within the \texttt{DTA} app which is
-contained within the Oceanographic Decision Support System (ODSS; see
+contained within the Oceanographic Decision Support System (\od; see
 below) \cite{das11}. GPS data from the free-floating drifter is sent
@@
 \subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.35\textwidth]{figs/september-iterations.pdf}}
-  \caption{\small{The offshore CANON ESP-drifter following cruise
+  \caption{\small{The offshore \can ESP-drifter following cruise
       \cite{das11b}. Fig. \ref{fig:sep10-ship} shows the ESP drifter
@@
 the LRAUV \cite{bellingham10} in October 2010 as part of another
-series of \texttt{CANON} field experiments. The slower vehicle with a
+series of \can field experiments. The slower vehicle with a
 fluorometer tracked a targeted chlorophyll patch seeded by a
 GPS-enabled drifter, sending periodic updates of the patch centroid to
-shore via the ODSS following a setup similar to
+shore via the \od following a setup similar to
 Fig.\ref{fig:dta-setup}, to the Dorado running \rx.
@@
 \centering
-\subfloat[The ODSS Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
-\subfloat[Current Implementation for \texttt{CANON}]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
-\caption{\small The Oceanographic Decision Support System (ODSS) is a
+\subfloat[The \od Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+\subfloat[Current Implementation for \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
+\caption{\small The Oceanographic Decision Support System (\od) is a
   single portal for planning, visualization, situational awareness,
   collaboration and archival of information before, during and after
-  field experiments. The ODSS has been used since mid-year 2010 for
-  \texttt{CANON} field experiments.}
+  field experiments. The \od has been used since mid-year 2010 for
+  \can field experiments.}
 \label{fig:odss}
@@
 apparent dispersal of the bloom to subsurface layers when HABs were a
-\texttt{CANON} target. However, their retargeting was carefully
+\can target. However, their retargeting was carefully
 monitored to balance the long-term desire of obtaining time series
@@
 \caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with the
-  ODSS in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
+  \od in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
   device.}
@@
 
-Additionally, our work on ODSS is driven by a vision of how marine
+Additionally, our work on \od is driven by a vision of how marine
 field robotics will likely be conducted in the future and the
 necessary computational infrastructure that will augment decision
-making, onshore as well as on-board. We believe ultimately ODSS will
+making, onshore as well as on-board. We believe ultimately \od will
 combine capabilities inferential and data gathering and analysis
@@
 
-Operationally, the ODSS has become a software system that the
+Operationally, the \od has become a software system that the
 scientist consults during a \can field campaign. S/he consults
@@
 cycle and then queries the system to compute candidate sampling
-locations. In the near future, the ODSS will consult its situational
+locations. In the near future, the \od will consult its situational
 awareness database, incorporate model outputs projecting these outputs
@@
 targeting of features by the conglomerate of vehicles in the ocean
-being tracked by the ODSS. Thus far, the ODSS has provided a simple
+being tracked by the \od. Thus far, the \od has provided a simple
 yet functional tool for scientists to use in the daily complex

### 2012-01-27 conclude.tex
updates

@@
 % vehicle autonomy.
+
+The Ocean Sciences are at a cusp, straddling between traditional
+ship-based exploration with newer observatory-based methods. In the
+United States the NSF funded Ocean Observatories Initiative \cite{ooi}
+and regional bodies such as \texttt{CeNCOOS} \cite{cencoos} are
+investing in new observational methods and technologies which propose
+to aid oceanographers by producing substantially higher-resolution
+data. These will use moorings, cabled observatories and glider fleets
+generating near real-time data from one or more locations as a
+continuous stream. The tack chosen by \can has been to complement such
+methods with the aim of developing methods for a \emph{portable
+  mobile} observatory which is dependent on autonomous platforms. In
+this context what we advocate in this chapter is an initial step in
+making robotic platforms more adaptive, scaling them towards fleets of
+vehicles (underwater or on the surface) and most importantly using
+them to solve the larger challenge of Sampling spatio-temporal fields.
+
+While our trajectory of research in the oceanographic domain follows
+that in spacecraft autonomy, from onboard \cite{mus98} to
+mixed-initiative methods \cite{bresina05}, the challenges are
+significantly different with the environment playing a far larger
+role. Moreover, prediction and projection of a future course of action
+for a robot given the level of uncertainty calls for representational
+and architectural methods which would enable adaptation at different
+levels of abstraction, from mission-planning to low level actuation.
+
+Our efforts were initially targeted to full robotic autonomy; while
+this is an important goal that we continue to strive towards, it is
+increasingly clear that human-in-the-loop methods also have a sizable
+role to play in ocean exploration and by extension to Marine
+Robotics. Not only is an embedded \rx an important element of the
+effort towards smarter vehicles, but we are working towards an
+implementation of \rx which will perform multi-vehicle plan synthesis
+for the \od.
+
+A key role that \rx has played and will continue to play is towards
+\emph{event response} scenarios. In Section \ref{sec:results} we show
+the role Machine Learning (ML) techniques have played thus far in
+in-situ plan adaptation. This interaction between Planning and ML we
+believe will play a significant role going forward in the general
+theme of exploration and discovery. Interpretation of incoming sensor
+data to deal with scientific surprise and readapt for opportunistic
+science will be critical to dealing with undersampling the
+ocean\footnote{Walter Munk has famously stated \emph{``Most of the
+    previous century could be called a “century of undersampling”''}
+  -- Testimony to the U.S. Commission On Ocean Policy, 18 April
+  2002}. Much of what we know of the ocean was derived by exploring in
+the blind. We anticipate longer duration robotic vehicles with enough
+onboard computational capacity to run deliberative agents such as \rx
+make water-column measurements and reliably track and characterize
+dynamic features. While our experiments have already shown such
+capabilities the exploration Vs. exploitation tradeoff is better
+determined when vehicles can make choices over longer time periods
+(weeks and months) to study the impacts of bio-geochemical interaction
+driven by longer duration physical forcing.
+
+Given our poor understanding of the coastal ocean in particular,
+mixed-initiative methods of command and control for Sampling with
+heterogenous assets is an important goal that is clearly at our
+doorstep. We believe deliberation is small yet critical part of the
+solution to unfold open research problems for AI and Marine Robotics.

### 2012-01-27 results.tex
updates

@@
 \texttt{Wirewalker} was tracked by the Dorado running \rx. We are
-currently investigating spatio-temporal correlations of CTD data
-obtained by the \texttt{Wirewalker} and tying it to the observations
-made in the contextual environment by the Dorado's sensors.
+currently investigating spatio-temporal correlations \cite{gneiting07}
+of CTD data obtained by the \texttt{Wirewalker} and tying it to the
+observations made in the contextual environment by the Dorado's
+sensors.
+
+\paragraph {Mixed-Initiative Autonomy}
+
+\begin{figure*}
+\centering
+\subfloat[The \od Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+\subfloat[Current Implementation for \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
+\caption{\small The Oceanographic Decision Support System (\od) is a
+  single portal for planning, visualization, situational awareness,
+  collaboration and archival of information before, during and after
+  field experiments. The \od has been used since mid-year 2010 for
+  \can field experiments.}
+\label{fig:odss}
+\end{figure*}
+
+Large scale oceanographic field experiments akin to those targeted by
+\can are usually logistically constrained driven largely by ship
+schedules.  Once a feature set has been identified for scientific
+observation and science goals have been collated, the appropriate
+robotic assets need to be assigned and deployed for observation and
+sampling. This asset allocation in turn depends on parameters like
+sensor payload, the speed of the robot \situ in comparison to the
+spatio-temporal scales of the evolution of the phenomena being
+sampled, the number of assets in a class, the kinds of measurements
+needed (e.g synoptic or close-up), logistical considerations for
+deployments such as battery charge needed or distance/time to deploy,
+impact to long-range deployment plans among others. All drive the
+strategy to balance short and long-term scientific and engineering
+needs. Our \can deliberations during the 2010 field season for example,
+were driven by all of the above \cite{das11}. Often these strategies
+were rebalanced during each decision-making cycle, especially in light
+of opportunistic science goals. For example, gliders running at $0.5$
+knots were often retargeted when surface support vessels noted the
+apparent dispersal of the bloom to subsurface layers when HABs were a
+\can target. However, their retargeting was carefully
+monitored to balance the long-term desire of obtaining time series
+within fixed tracks and with the need to ensure these assets were not
+exposed to strong currents. Powered AUVs were targeted to chase after
+advecting blooms, but the decision on their deployment locations and
+survey areas too close to the surf-zone were avoided. In
+patch-tracking experiments, the deployment of drifters close to shore
+and where depths of $20m$ or less were encountered, was avoided even
+when contrary to scientific intent. The need to obtain high-quality
+data for post-hoc analysis was also driven by the response time of
+data returned by experiment partners.
+
+\begin{figure*}
+\centering
+\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.3\textwidth]{figs/october-canon-meeting.png}}
+\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.33\textwidth]{figs/virtual-drifter-tool.png}}
+\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
+\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
+  the \od in use. Fig. \ref{fig:odss-virtual} shows a drifter tracking
+  tool used during \can planning; it used a physical model of advected
+  surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
+  deployments. Fig. \ref{fig:odss-ipad} shows \od use on a portable
+  iPad device.}
+\label{fig:odss2}
+\end{figure*}
+
+These distinctions make some vehicles more or less suitable to address
+certain science goals and suggest some degree of intelligent planning
+in the assignment strategy. Further, even if partial plans can be
+synthesized automatically, the need to capture qualitative 'intent'
+and desires in the final outcome of an asset allocation strategy
+drives us towards a human-in-the-loop approach for asset
+allocation. The vast uncertainty and complex interaction between
+unknown quantities in the ocean environment dictate that any
+comprehensive experimental methodology must rely in some measure on
+human reasoning and adaptability.  Furthermore, the multidisciplinary
+nature of ocean experiments suggests collaboration between multiple
+human actors. Human cognitive capability then to capture these
+uncertainties, evaluate options, factor intent and synthesize data
+needed for asset allocation becomes increasingly important. Therefore,
+mixed-initiative approaches are viable as a means to apportion problem
+solving and inference between human and computational approaches. A
+viable example was shown in the command and control of the twin Mars
+rovers in \cite{aichang04,bresina05,bresina05a}\footnote{The MAPGEN
+  system uses the same \eu planner used in \rx. It continues to be
+  used routinely to this day on the MER mission.}.
+
+Additionally, our work on \od is driven by a vision of how marine
+field robotics will likely be conducted in the future and the
+necessary computational infrastructure that will augment decision
+making, onshore as well as on-board. We believe ultimately \od will
+combine capabilities inferential and data gathering and analysis
+capabilities and more uniquely, provide an event response capability
+that could trigger the deployment of a robot and/or adapt its sampling
+strategy, all from a scientist's desktop on shore. Data returned from
+robotic assets at sea will be filtered for signals associated with
+feature(s) of interest along with information on its contextual
+environment. The association between the signal and a prospective
+event will be undertaken through tagging by experts and machine
+learning techniques used in recommender
+systems\cite{Adomavicius05}. Adaptive sampling strategies will then
+balance decision-making \situ with that done onshore with
+computational models and human agents, the subject of a new NSF
+project in which the authors are participating.
+
+Operationally, the \od has become a software system that the
+scientist consults during a \can field campaign. S/he consults
+up-to-date model outputs, synoptic views of ocean color from remote
+sensing data and weather conditions for the next $24$-hour planning
+cycle and then queries the system to compute candidate sampling
+locations. In the near future, the \od will consult its situational
+awareness database, incorporate model outputs projecting these outputs
+into the space-time region of interest to generates candidate
+hypotheses with associated utility metrics. With the scientist's
+guidance the automated planner would then determine the most viable
+robotic option(s) to be dispatched (underwater, surface or aerial),
+given available resources, distance of travel, required payload and
+the environment. For instance, unpowered autonomous gliders would be
+chosen for longer-duration missions with smaller payload requirements
+in areas of moderate to weak ocean currents. The automated planner
+would balance these criteria to compute a viable trajectory plan for
+the selected vehicle. Vehicles with more controllability would be
+given a specified area to survey and the path generated \situ. Goals
+are communicated to the robot after analysis on shore for selective
+targeting of features by the conglomerate of vehicles in the ocean
+being tracked by the \od. Thus far, the \od has provided a simple
+yet functional tool for scientists to use in the daily complex
+decision making and the following experiment and survey design
+needs. It has helped train scientists to think of expanding their
+repertoire of tools for the scientific process and in the process it
+enabled a mixed team of engineers and scientists to interact at the
+functional level of the systems' capabilities.

### 2012-01-30 trunk/conclude.tex
updates

--- 2012-01-27 conclude.tex
@@
 data to deal with scientific surprise and readapt for opportunistic
-science will be critical to dealing with undersampling the
-ocean\footnote{Walter Munk has famously stated \emph{``Most of the
-    previous century could be called a “century of undersampling”''}
-  -- Testimony to the U.S. Commission On Ocean Policy, 18 April
+science will be critical to dealing with persistent problem of
+undersampling the ocean\footnote{Walter Munk of the Scripps Institute
+  of Oceanography has famously stated \emph{``Most of the previous
+    century could be called a “century of undersampling”''} --
+  Testimony to the U.S. Commission On Ocean Policy, 18 April
   2002}. Much of what we know of the ocean was derived by exploring in

### 2012-01-30 future.tex
updates

--- 2012-01-04 trunk/future.tex
@@
 % multiple vehicles is an important result of this activity.
+
+% 1.6 Future Work?
+% explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).
+
+% The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space
+
+Our experience of using deliberative techniques in a highly
+inter-disciplinary environment for targeted marine science
+applications has shown the need and applicability of such novel
+algorithms and methods. In addition to augmenting traditional AUV
+surveys more advanced Lagrangian observations, mixed-initiative
+methods, our colleagues at MBARI have understood the longer term
+implications of our work. Our current and consequently future efforts
+are therefore grounded in important science problems in an operational
+oceanographic setting.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[scale=0.45]{figs/autonomy-topics.pdf}
+  \caption{\small }
+  \label{fig:topics}
+\end{figure}
+
+The core of our problem domain for \can has been towards Sampling. In
+large part given the spatial and temporal scales of the
+bio-geochemical processes we are interested in in this project, it is
+imperative that a key direction is in formulating a multi-robot
+adaptive sampling problem. In this context while superficially \rx can
+be considered as a multi-agent system and could therefore imply a
+direct extension, sporadic communication with high-latencies in
+surface communication, not to mention low throughput acoustic
+communications imply that the extension of \rx for distributed
+applications is non-trivial. Further, in order to efficiently execute
+coordinated observation of ocean processes with multiple autonomous
+sensing platforms, significant methodological as well as technical
+challenges must be addressed. We are working towards a systematic
+approach of using \rx as a backend planner for \can field
+experiments. The expectation is that \od users will plan initially
+individual vehicle surveys which are variants of standard
+straight-line transects. Lessons learned with human planners in the
+loop will be used towards planning larger surveys, also with simple
+fixed transects but using vehicle capabilities in ontologies similar
+to \cite{patron09}. Apportioning the planning task between shore-side
+components with substantial contextual information and compute power
+with robots with in-situ reasoning capabilities a la \rx is a
+challenge and an open research problem. Fig. \ref{fig:topics} shows
+what we believe is the scope and influence of the different topic
+areas we are focusing on.
+
+Partitioned inference and decision-making within \rx also lends itself
+to graceful system degradation, making the overall approach effective
+towards increased robustness. There are two outstanding diagnosis and
+failure recovery challenges that we need to address; the first has to
+do with methodology related to model design of the instrument payload
+for a highly modular and configurable vehicle; this is local to a
+reactor. Doing so would enable (as a first step) using systematic
+Fault Diagnosis, Isolation and Recovery (FDIR) algorithms as
+demonstrated in the \texttt{RAX} experiment
+\cite{williams96,mus98,williams97} and more recently on an AUV by
+Dearden and others \cite{wang09,ernits10,dearden11}. The second
+challenge has to implementing a policy within the \rx framework that
+explicitly deals with recovery since controller failure can be
+distributed across reactors. Since reactors are hierarchical in their
+relationship to one another, their well defined dependencies allow
+them to be removed from an agent systematically. Reactors more
+abstract in scope can be removed before those which are less abstract
+(and closer to the hardware). In the event a Navigation reactor, for
+example, has a fault, it can be removed safely with lower level
+functionalities (such as those in the executive). When the Navigator
+is removed, it will recursively remove more abstract reactors (such as
+a Mission Manager) that depend on it. Such a safety feature then
+allows the system designers to ensure that early on they can invest
+more in ensuring that reactors lower in the hierarchy are robust and
+potentially formally verified. It also means that there is a safe and
+sure way to take control of the vehicle at any level within the
+reactors hierarchy; as long as low level reactors are still active one
+will always been able to send goals/commands to them and execute what
+are traditionally known as runout sequences (or clean up scripts).
+
+% multi-vehicle control
+% sampling
+% diagnosis and recovery
+feature recognition
+scalable architecture for missions

### 2012-01-30 trunk/europa.tex
checkpoint. more europa content.

@@
 
-TODO: motivate with a few examples of problems that can be solved in CSP, Scheduling and Planning.
+Let's look at some of the problems that can be tackled using EUROPA.
 
-We begin with a brief overview of Constraint Programming.
+Constraint Satisfaction: 
+In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.
+
+Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
+TODO: include figure.
+
+If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
+Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.
+
+then we have to satisfy the following constraints:
+Sum(Qrc) = 1 ForAll r (Only one Queen per row)
+Sum(Qrc) = 1 ForAll c (Only one Queen per column)
+
+Sum(Qr+i,c+i) = 1
+Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)
+
+We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.
+
+Below is a solution found by EUROPA using that formulation and a specialized search procedure.
+
+TODO: include figure
+
+Scheduling:
+In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints among them, as well as respects fixed limits on the availability of resources required to perform each activity.
+In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
+
+Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
+TODO: include figure
+
+Planning:
+In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
+
+Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.
+
+TODO: include figure
+
+As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both embed Constraint Satisfaction problems as temporal, resource and many other kinds of constraints must be respected in most real life problems.
+
+These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs.
+Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
+Next we will examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
 
@@
 
-Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
+Constraint Satisfaction Programming, also know as Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
 A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].
@@
 TODO: talk about arc consistency algorithms, computational complexity and solvers.
+
+\subsection{Constraint-Based Attribute and Interval Planning}
+\label{sec:europa:cp}
+
+There has been 
+State Variable representation
+
+CSP encodings
+
+Constraint Based and interval planning.
 
@@
 
+	TODO: talk about the token network, move stuff from the Concepts section if necessary
+
 	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
@@
 	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+		\begin{enumerate}
+			\item Temporal Network
+			\item Resources
+			\item NDDL
+			\item ANML (TODO:maybe it's better to leave out?)
+		\end{enumerate}
 
@@
 
-TODO: introduce model for Shopping or Rover example, link to architectural components.
+Shopping model:
+
+\begin{verbatim}
+
+// Locations (Home, SuperMarket, etc.)
+class Location {
+  string name;
+  Location(string _name){
+    name = _name;
+  }
+}
+
+// Products (Milk, Banana, etc.) and
+class Product {
+  string name;
+  Product(string _name) {
+    name = _name;
+  }
+}
+
+// ProductLocations (Banana can be found at SuperMarket, for example)
+class ProductLocation {
+  Location location;
+  Product product;
+
+  ProductLocation(Location _location, Product _product){
+    location = _location;
+    product = _product;
+  }
+}
+
+// Use built-in Timeline functionality to enforce that an agent:
+// a) Can't be at more than one place at a time.
+// b) Can't Go more than one place at a time.
+// c) Can't Go somewhere and be At somewhere at the same time.
+class AgentLocation extends Timeline{
+  predicate At {
+    Location loc;
+  }
+
+  predicate Going {
+    Location from;
+    Location to;
+  }
+}
+
+// In addition to having a location timeline, the agent can buy and own things.  
+// Note that the actions and the location predicates can be concurrent 
+// so they can't be on the same Timeline
+class Agent {
+  AgentLocation location;
+
+  Agent() {
+    location = new AgentLocation();
+  }
+
+  action Buy {
+    Product product;
+  }
+
+  action Go {
+    Location from;
+    Location to;
+  }
+
+  predicate Own {
+  	Product product;
+  }
+}
+
+// Define the rules for our actions:
+Agent::Go {
+  met_by(condition object.location.At origin);
+  eq(from, origin.loc);
+
+  equals(effect object.location.Going going);
+  eq(going.from, from);
+  eq(going.to, to);
+
+  meets(effect object.location.At destination);
+  eq(to, destination.loc);
+}
+
+Agent::Buy {
+  // A Buy takes 10 time units
+  eq(10, duration);
+
+  // initialized to all locations
+  ProductLocation possibleStores;
+
+  // limit possibleStores variables to ones that provide what we need to buy
+  eq(product, possibleStores.product);
+
+  // We must be At a location during our Buy, and that location must have the
+  // product we want available:
+  contained_by(condition object.location.At currLocation);
+  eq(currLocation.loc, possibleStores.location);
+
+  starts(effect Own purchase);
+  eq(purchase.product,product);
+}
+
+\end{verbatim}
+
+And a problem instance:
+
+\begin{verbatim}
+
+// Allocate instances
+Location Home = new Location("Home");
+Location SuperMarket = new Location("SuperMarket");
+Location HardwareStore = new Location("HardwareStore");
+
+Product Banana = new Product("Banana");
+Product Milk = new Product("Milk");
+Product Drill = new Product("Drill");
+
+ProductLocation bananaLocation = new ProductLocation(SuperMarket, Banana);
+ProductLocation milkLocation = new ProductLocation(SuperMarket, Milk);
+ProductLocation drillLocation = new ProductLocation(HardwareStore, Drill);
+
+Agent agent = new Agent();
+
+// Indicate that the database is closed - no new objects can be created
+close();
+
+// We start the day at Home:
+fact(agent.location.At atHomeForBreakfast);
+atHomeForBreakfast.loc.specify(Home);
+
+// Goals for all of the agent's needs, buy if needed
+goal(agent.Own gotMilk);
+gotMilk.product.specify(Milk);
+goal(agent.Own gotBanana);
+gotBanana.product.specify(Banana);
+goal(agent.Own gotDrill);
+gotDrill.product.specify(Drill);
+
+// Make sure agent is home for dinner
+goal(agent.location.At atHomeForDinner);
+atHomeForDinner.loc.specify(Home);
+
+// Agent has all day to satisfy goals:
+gotMilk after atHomeForBreakfast;
+gotMilk before atHomeForDinner;
+
+gotBanana after atHomeForBreakfast;
+gotBanana before atHomeForDinner;
+
+gotDrill after atHomeForBreakfast;
+gotDrill before atHomeForDinner;
+
+// Force things to happen within our planning horizon:
+int horizonStart = 0;
+int horizonEnd = 100;
+leq(horizonStart, atHomeForBreakfast.start);
+leq(atHomeForDinner.end, horizonEnd);
+
+\end{verbatim}
+
+TODO: introduce model for Shopping link to architectural components.
 explain variable, object, predicate and action types and instances, constraints, goals and facts.
@@
 TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
-- overview of the types of flaws and violations EUROPA detects.
+- talk about AC3 implementation and extension mechanisms
+- talk about europa's constraint library
+- talk about Temporal Network
+- talk about Resource Envelopes.
 
@@
 
+TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
+examples:
+- deal with earliest flaws first
+- deal with planning (open condition flaws) before scheduling (threats) ones
+
 It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?

### 2012-01-30 trunk/europa.tex
checkpoint. more europa content.

@@
 Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
-TODO: include figure.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
+\caption{\small N-Queens problem}
+\label{fig:nqueens-1}
+\vskip+0.1cm
+\end{figure}
 
@@
 
-TODO: include figure
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
+\caption{\small N-Queens solution}
+\label{fig:nqueens-2}
+\vskip+0.1cm
+\end{figure}
 
@@
 Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
-TODO: include figure
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
+\caption{\small RCPSP problem}
+\label{fig:rcpsp-1}
+\vskip+0.1cm
+\end{figure}
 
@@
 
-TODO: include figure
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
+\caption{\small Shopping Agent problem}
+\label{fig:shopping-1}
+\vskip+0.1cm
+\end{figure}
 
-As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both embed Constraint Satisfaction problems as temporal, resource and many other kinds of constraints must be respected in most real life problems.
+As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
 
@@
 
-(TODO: add figure)
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/EUROPA-Architecture.jpg}
+\caption{\small EUROPA's Architecture}
+\label{fig:europa-architecture}
+\vskip+0.1cm
+\end{figure}
 

### 2012-01-30 trunk/europa.tex
checkpoint. more europa content.

@@
 Scheduling:
-In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints among them, as well as respects fixed limits on the availability of resources required to perform each activity.
+In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints.
+The activity schedule must also respect fixed limits on the availability of resources required to perform each activity.
 In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
@@
 In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
+The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
 
@@
 
-As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
+As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
 
@@
 CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other “primitive” CP constraint, such as <= or >=.
+Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
+
+Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.
 
@@
 
-Shopping model:
+EUROPA can ingest descriptions of models, plans and goals.
+This means that a single planner can be applied to different systems and problems, simply by providing different models and goals.
+For such planners the expressiveness of the languages for models, goals and plans is of great importance.
+EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way.
+The main characteristics that make NDDL a powerful tool for describing problem domains are:
+
+\begin{enumerate}
+	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++.
+It also supports polymorphism for some of the planning components.
+Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
+	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
+	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
+\end{enumerate}
+
+To understand how the most important elements of NDDL work, let's take a look at how the Shopping Agent problem described above would be modeled:
 
@@
 
-And a problem instance:
+TODO: explain variable, object, predicate and action types, including constraints that can be part of them
+
+In this model we can see:
+
+\begin{enumerate}
+	\item Variable Types:
+	\item Object Types:
+	\item Predicate Types:
+	\item Action Types:
+\end{enumerate}
+
+Now let's look at how a specific problem instance would be specified in NDDL:
 
@@
 // Indicate that the database is closed - no new objects can be created
+// this allows EUROPA to perform more efficient inference
 close();
@@
 
-TODO: introduce model for Shopping link to architectural components.
-explain variable, object, predicate and action types and instances, constraints, goals and facts.
+TODO:  explain variable, object, predicate and constraint instances.
+explain goals and facts.
+
+\begin{enumerate}
+	\item Variable Instances:
+	\item Object Instances:
+	\item Facts:
+	\item Goals:
+	\item Constraint Instances:
+\end{enumerate}
 
@@
 - talk about europa's constraint library
-- talk about Temporal Network
-- talk about Resource Envelopes.
+
+The Temporal Network:
+The notion of time is central to temporal planning.
+EUROPA uses variables to explicitly represent timepoints for plan activities and states.
+Constraints among timepoints provide a natural way to express domain axioms.
+For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
+Dechter et al.
+(Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN).
+Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node.
+The diagram below illustrates a simple STN with just 2 variables and a single constraint.
+It also shows the resulting DG.
+
+TODO: add figure
+
+Dechter et al.
+(Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle.
+A negative cycle is a path from a node to itself that has a path length less than 0.
+If such a cycle exists, the network is inconsistent.
+It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search.
+Thus we have an efficient and complete algorithm for propagating an STN.
+These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.
+
+Resources:
+
+    Reservoir resources can have both 'consume' and 'produce' tokens.
+Note that these tokens do NOT have the usual start/end/duration token variables, since occur at a single instant in time.
+Instead, they have a single variable, 'time' used to represent that instant.
+    Reusable resources have 'uses' tokens that use a quantity for the duration of the token. These tokens do have the usual start/end/duration variables.
+    Unary resources are reusable resources with unit quantity and have 'use' tokens (again, with the usual start/end/duration variables).
+
+    TODO: talk about resource envelope computation by using MaxFlow formulation.
 
@@
 if a and a?
-unify and the temporal constraints involving a are satisﬁed by a?.
+unify and the temporal constraints involving a are satisÞed by a?.
 Thus,  a and a?
 can be considered the same activity and we do not need to introduce a in the plan.
-Consequently, the compatibilities associated with a are not ﬁred, because they have been already triggered when a?
+Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a?
 was introduced in the plan.
-		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciﬁc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
-This results in both an ordering ﬂaw, corresponding to the just activated activity, and a number of open condition ﬂaws, corresponding to the new subgoal activities.
+		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
+This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities.
 		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
@@
 
+N-Queens solver:
+\begin{verbatim}
+    public void solve(int maxIter)
+    {
+    	init();
+
+    	for (int i=0;psengine_.getViolation() > 0 && i < maxIter;i++) {
+    	    PSVariable queenToMove = getQueenWithMaxViolation();
+    		int curPos = queenToMove.getSingletonValue().asInt();
+
+    	    boolean moved = false;
+            SortedSet<Move> moves = getMoves(queenToMove,curPos);             
+            for (Move m : moves) {
+            	moved = makeMove(queenToMove,curPos,m,false);
+            	if (moved)
+            	    break;
+            }
+
+            if (!moved) 
+            	makeMove(queenToMove,curPos,moves.first(),true);
+
+            checkSolution(); // See if we have a new best solution            
+            notifyIterationCompleted(curIteration_++);
+
+            if (curIteration_-bestIter_ > 50) 
+            	restart();
+    	}
+    }
+\end{verbatim}
+
+RCPSP solver:
+\begin{verbatim}
+    public void solve(PSEngine psengine,
+    		          long timeout, // in msecs
+    		          int bound)
+    {
+       init(psengine,timeout,bound);
+
+       for (int i=0; true ; i++) {
+          flatten();
+          updateSolution(i);
+          updateCriticalPrecedences();
+
+          if ((nbStable_ > maxStable_) || (bestMakespan_ <= makespanBound_))
+             break;
+
+          if (timer_.getElapsed() > timeout) {
+              timedOut_ = true;
+              break;
+          }
+
+          relax();
+          curIteration_++;
+       }
+
+       restoreBestSolution();
+    }
+
+\end{verbatim}
+
 TODO: 

### 2012-01-30 future.tex
updates

@@
 
+Feature recognition for Sampling is a general research thrust; however
+it has specific relevance in the context of \rx and AUV control. As
+noted in Section \ref{sec:results} we have already demonstrated
+Machine Learning techniques for recognition of select features in the
+coastal ocean. A principal challenge that remains is that of obtaining
+expert labeled data for any supervised or semi-supervised learning
+methods.  While supervised techniques for learning classifiers such as
+decision trees (\cite{Quinlan93-dtrees}), instance-based
+classification (\cite{Aha-ibl-ml91}), Bayesian classifiers
+(\cite{Jensen2001-BNetworks}) and artificial neural networks
+(\cite{ANNsurvey-2000}) are popular, given the vast amount of data
+available and poor understanding of correlation between physical and
+biogeochemical variables in the coastal ocean, our work on INLs
+\cite{mcgann08d,ryan10} and reliance on a oceanographic expert has
+shown that scaling to different problems remains problematic. This is
+a prime motivation to move towards more semi-supervised methods such
+as \cite{kumar11,sergio12}. With the advent of the \od however,
+another direction our research is taking us is in Recommender systems
+\cite{Adomavicius05} as noted in Section \ref{sec:results}. We are
+working to design and deploy software infrastructure that will learn
+from user tagged data. To do so, we will use tagging techniques
+similar to what users of commercial image sharing sites like Flickr
+and Picassa do, to learn the context of the data (including remote
+sensing). In addition, the \od will integrate statistical Machine
+Learning techniques with incoming asynchronous data stream obtained by
+sensors and platforms. Identification models will be based on existing
+training sets for INLs and plumes; models for other features of
+scientific interests driven by \can requirements will be built in
+addition. Existing AUV archives for instance provide sufficient data
+for a learning base for a range of features of interest to \can; in
+addition we will build software that will learn to incorporate
+incoming data streams to provide an existence proof of a feature of
+interest.
+
 % multi-vehicle control
@@
 % diagnosis and recovery
-feature recognition
-scalable architecture for missions
+% feature recognition
+% scalable architecture for missions

### 2012-01-30 results.tex
updates

@@
 event will be undertaken through tagging by experts and machine
-learning techniques used in recommender
-systems\cite{Adomavicius05}. Adaptive sampling strategies will then
-balance decision-making \situ with that done onshore with
-computational models and human agents, the subject of a new NSF
-project in which the authors are participating.
+learning techniques used in recommender systems
+\cite{Adomavicius05}. Adaptive sampling strategies will then balance
+decision-making \situ with that done onshore with computational models
+and human agents, the subject of a new NSF project in which the
+authors are participating.
 

### 2012-01-30 trunk/related.tex
updates

@@
 
-\begin{wrapfigure}{r}{0.2\textwidth}
+\begin{figure}[htpb]
  \centering
@@
  \label{fig:planner}
-\end{wrapfigure}
+\end{figure}
 
@@
 cycle. \texttt{T-REX} was designed after carefully evaluating lessons
-learned from the IDEA and owes substantially to the notion of
+learned from \texttt{IDEA} and owes substantially to the notion of
 partitioned problem solving from that effort. However this

### 2012-01-30 trunk/intro.tex
updates

@@
 altered both the discipline and impacted the engineering methodologies
-behind Sampling, Control and Robotics itself \comment{need higher
-  level citation here}. Yet challenges remain, principally in making
-such robotic devices more attune to the environment, to sample large
-swaths at the Meso-scale ($> 50$ km\textsuperscript{2}) and doing so
-systematically and adaptively. Further, over the last two decades
-substantial improvements in hardware including propulsion, battery
-technologies and hardware have not been kept abreast of in software
-especially in algorithmic methods in Sampling and Control and in
-systematic engineering and software engineering development.
+behind Sampling, Control and Robotics itself
+\cite{Brierley08032002,ryan05,Thomas06,Yoerger01012007,Incze2009,Rigby10}. Yet
+challenges remain, principally in making such robotic devices more
+attune to the environment, to sample large swaths at the Meso-scale
+($> 50$ km\textsuperscript{2}) and doing so systematically and
+adaptively. Further, over the last two decades substantial
+improvements in hardware including propulsion, battery technologies
+and hardware have not been kept abreast of in software especially in
+algorithmic methods in Sampling and Control and in systematic
+engineering and software engineering development.
 
@@
 years has produced a range of planners with demonstrated embedded
-capabilities \comment{cite}. Our work has been influenced by and in
-turn influenced a wide range of activities in the field of AI Planning
-and Plan Execution with demonstrated real-world and highly visible
-capabilities in the Space domain \cite{mus98, rajan00, ai-chang04,
-  bresina05} for NASA. Lessons learned have been applied to the world
-of marine robotics which we bring together in this chapter. Important
-lessons that were derived from this rich legacy involved the following:
+capabilities
+\cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
+work has been influenced by and in turn influenced a wide range of
+activities in the field of AI Planning and Plan Execution with
+demonstrated real-world and highly visible capabilities in the Space
+domain \cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons
+learned have been applied to the world of marine robotics which we
+bring together in this chapter. Important lessons that were derived
+from this rich legacy involved the following:
 
@@
   \vskip-5pt
-  \includegraphics[scale=0.325]{figs/MBARI-AUV.jpg}
+  \includegraphics[scale=0.1]{figs/MBARI-AUV.jpg}
   \caption{\small The MBARI \emph{Dorado} AUV on its support vessel
@@
 \end{figure}
+
+This chapter is organized as follows. We start with some foundational
+concepts in our Planning framework in Section
+\ref{sec:concepts}. \comment{add others} We conclude with Section
+\ref{sec:conclusion}.

### 2012-01-31 trunk/arch.tex
- Updated europa solve algorithm so I can refer to it - updates in the t-rex section

@@
 
-{\em\color{red} Following is an attempt to present TREX as a big
-  deliberation machine}
+The choice of representation of state variables, observations and
+goals is directly derieved to the timleines and tokens from europa. 
+such representation is ich enough to allow the excahnge of state
+information in a formal and yet flexible way and its transaltion in
+planning frameworks that have an explicit representation of time 
+(such as but not limited to europa) is often trivial.
+
 The overall execution cycle of the agent is also abstracted as some
@@
 deliberation \texttt{step}s in order to satisfy these new
-objectives. as a consequence physical execution will occur as these
+objectives. As a consequence physical execution will occur as these
 goals and planning propagate down in the reactor graph until they
@@
 execution.
-{\em\color{red} end of the attempt; need to rework the paragraphs
-  below accordingly}
 
-The overal execution cycle of the agent is emerging through these
-interaction and the processing that occurs in each reactors in order
-to process them. New observation occurring at every tick inform each
-reactor of the current state of the world and impose them to identify
-how it impact their state (and potentially what they planned for the
-future). Goals they receive will trigger in the reactor the need to
-identify a strategy (or plan) that will allow them to reach these new
-states knowing their current state deduced from synchronization.When
-such a plan has been identified, the {\em external}  part is then
-dispatched to other reactors that can do the same. This flow and focus
-around state and goals is perfectly fitting with automated planning
-and even more so to the Europa framework. As an illustration, in 
- Fig. \ref{fig:agent}, only \textsf{Shore comm.} and \textsf{Vehicle}
- reactors are specialized C++ code (acting as interface to external
- programs). All the other reactors are based on the europa framework
- differentiated only by the domain model they manipulate and som
- heuristic configuration parameters. In the following sections we are
- going to focus specifically ion this kind of reactor to illustrate
- how  a formal planning framework can be used in our architecture to
- both track state evolutions and plan for future objectives while
- remaining efficient enough to do these tasks in a real-time control 
-loop.
+The way the agent interleave planning \texttt{step}s and
+synchronization is illustrated in Figure \ref{fig:tick-exec}.
+Synchronization happens at every single tick and will interrupt
+planning allowing this reactor to identify its current state that can
+propagate through the reactor graph in order to ensure t aconsitent
+view of the state of the world between all the reactors.  
+\begin{figure}[!htbp]
+  \centering
+  \vskip-1pc
+  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
+  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
+      Deliberation} is interrupted by {\em synchronization} at the
+    beginning of every {\em tick} allowing integration of state
+    information.}
+  \label{fig:tick-exec}
+  \vskip-0.8pc
+\end{figure}
+This view of the reactor has two parallel tasks aloow for long
+planning deliberation while ensuring that the state evolution
+propagates through the agent as time advance. These two tasks 
+intertwined together give the two natural flows (bottom-up for state
+estimation and top to bottom for future plan) that we expect to see in
+a control loop. As they are both conceptualized as inference processes
+it is natural to implement a reactor based on generic automated
+planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
+  comm.} and \textsf{Vehicle} acting as interfaces to the system,
+every other reactors (colored in blue) are different instance of our
+europa reactor differntiated only by their model and europa solver
+configurations.
 
@@
 
-In T-REX architecture reactors have two core functions;
-synchronization where the main focus is to track and refine the
-evolution of the state of the world at any single tick, and
-deliberation which focus on providing a plan to the agent in order to
-complete the reactor {\em Internal} goals. These two processes are
-intertwined in the sense that one influences the outcome of the
-other. They are represented as two europa solvers modifying the same
+The europa reactor is our implementation inside T-REX of a reactor
+that usess automated planning for its execution. This reactor is fully
+based on the europa framework and applies the core idea we of T-REX we
+described above to implement a reactor for which the sole purpse is to
+deliberate as time advance. The choice of europa as the planning
+framework was directed by the fact thta this framework provide  a rich
+and highly configurable framework to be able to infer around a rich
+temporal domain.
+
+This reactor is implemented following the idea that in T-REX, the
+whole execution cycle is two deliberation processes. Both of these
+processes are implemented as two europa solvers modifying the same
 plan database and for which the execution is managed by the T-REX 
@@
 \end{enumerate}
-
-While synchronization is assumed to be fast enough that it can be
-resolved in less than a tick. Deliberation on the other hand can take
-several ticks reflecting that automated planning can take time
-depending on how far in the future the planner is looking ahead and
-the complexity of the model. T-REX architecture, and more specifically
-the europa based reactor in our case, takes this into account by
-considering synchronization and deliberation as two concurrent tasks
-with their own timing requirements and priority. 
 While the two processes are separate, they share the same plan
@@
   external} plan defines the set of {\em goals} for this state
-variable managed by another {\em reactor}. Fig. \ref{fig:tick-exec}
-shows a conceptual view of these two tasks.
+variable managed by another {\em reactor}.
 
-\begin{figure}[!htbp]
-  \centering
-  \vskip-1pc
-  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
-  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
-      Deliberation} is interrupted by {\em synchronization} at the
-    beginning of every {\em tick} allowing integration of state
-    information.}
-  \label{fig:tick-exec}
-  \vskip-0.8pc
-\end{figure}
+An important challenge for embedding automayted planning i na situated
+agent as always been to go around the asumptions made by most classical
+planners. Specifically one assumption restrict the problem to ``offline
+planning''  defined in \cite{ghallab04} as follow:
+\begin{quotation}
+  The planner is not concerned with any change that may occur in
+  $\Sigma$ \footnote{In this book $\Sigma$ identifies the world modeled
+    by the plan domain} while it is planning; it plans for the given
+  initial and goal states regardless of the current dynamics, if any.
+\end{quotation}
+This assumption, while helpfull to reduce the scope of the planning
+problem, becomes problematic if one wants to embed a planner in a
+real-time system that evolves in a highly dynamic environment. In a
+typical robot this implies that the only point where the planner can
+be informed about the current world state is righe before the planning
+is started by giving to the planner the intial state and current
+objectives. During the planning, it is assumed that the agent can
+maintin the world as perceived by the planner as stable. In 
+\cite{lemai04, lemai-chenevier2004}, they do attempt to reduce the
+impact of planning by aloowing local plan repairs which attempt to
+insert repairing actions in the exisiting plan andf can be done fast
+enough. Still in ythe eventuality that such repair is not possible
+they add to stop their vehicle until the full (re-)planning was
+complete. {\em I may need more examples/refs here}
 
-In  this section we present how both processes are implemented using
-the europa framework. We first present each of them independently by
-looking to their specific focus. We then develop how both processes
-are not only interleaved (as deliberation can take several tick while
-synchronization eneds to happen at every single tick) but also discuss
-on how each of these is impacting the other and how this intertwining 
-of the two result on an emergent behavior of the reactor by allowing
-the system to efficiently adapts its plan that can then be dispatched
-as objectives for the reactors it depends on in a seamless flow.
-
-{\em fpy: I still wonder whether I should start with planning assuming
-  first that the initial state is known and then develop
-  synchronization and how it ties state information in to the plan or
-  the other way around as it is now}
+Looking back at Fig. \ref{fig:tick-exec}, one can already see that
+this asumtpion is problematic in our architecture. Indeed
+synchronization that tracks state evolution can occur several times 
+during the reactor deliberation process. By making synchronization and
+deliberation share the same plan we provide a solution that relax this
+assumption. In the following sections we present how synchronization
+and deliberation are implemented independently first to finally show
+how their interaction through the plan helps relax this assumption by
+making the planning fully integrated to the reactor execution cycle
+while potentially allowing for more informed planning.
 

### 2012-01-31 europa.tex
- Updated europa solve algorithm so I can refer to it - updates in the t-rex section

@@
 
-TODO: motivate with a few examples of problems that can be solved in CSP, Scheduling and Planning.
+Let's look at some of the problems that can be tackled using EUROPA.
 
-We begin with a brief overview of Constraint Programming.
+Constraint Satisfaction: 
+In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.
+
+Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
+\caption{\small N-Queens problem}
+\label{fig:nqueens-1}
+\vskip+0.1cm
+\end{figure}
+
+If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
+Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.
+
+then we have to satisfy the following constraints:
+Sum(Qrc) = 1 ForAll r (Only one Queen per row)
+Sum(Qrc) = 1 ForAll c (Only one Queen per column)
+
+Sum(Qr+i,c+i) = 1
+Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)
+
+We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.
+
+Below is a solution found by EUROPA using that formulation and a specialized search procedure.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
+\caption{\small N-Queens solution}
+\label{fig:nqueens-2}
+\vskip+0.1cm
+\end{figure}
+
+Scheduling:
+In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints.
+The activity schedule must also respect fixed limits on the availability of resources required to perform each activity.
+In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
+
+Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
+\caption{\small RCPSP problem}
+\label{fig:rcpsp-1}
+\vskip+0.1cm
+\end{figure}
+
+Planning:
+In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
+The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
+
+Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
+\caption{\small Shopping Agent problem}
+\label{fig:shopping-1}
+\vskip+0.1cm
+\end{figure}
+
+As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
+
+These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs.
+Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
+Next we will examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
 
@@
 CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other “primitive” CP constraint, such as <= or >=.
+Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
+
+Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.
 
 TODO: talk about arc consistency algorithms, computational complexity and solvers.
+
+\subsection{Constraint-Based Attribute and Interval Planning}
+\label{sec:europa:cp}
+
+There has been 
+State Variable representation
+
+CSP encodings
+
+Constraint Based and interval planning.
 
@@
 
-TODO: introduce model for Shopping or Rover example, link to architectural components.
-explain variable, object, predicate and action types and instances, constraints, goals and facts.
+EUROPA can ingest descriptions of models, plans and goals.
+This means that a single planner can be applied to different systems and problems, simply by providing different models and goals.
+For such planners the expressiveness of the languages for models, goals and plans is of great importance.
+EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way.
+The main characteristics that make NDDL a powerful tool for describing problem domains are:
+
+\begin{enumerate}
+	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++.
+It also supports polymorphism for some of the planning components.
+Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
+	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
+	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
+\end{enumerate}
+
+To understand how the most important elements of NDDL work, let's take a look at how the Shopping Agent problem described above would be modeled:
+
+\begin{verbatim}
+
+// Locations (Home, SuperMarket, etc.)
+class Location {
+  string name;
+  Location(string _name){
+    name = _name;
+  }
+}
+
+// Products (Milk, Banana, etc.) and
+class Product {
+  string name;
+  Product(string _name) {
+    name = _name;
+  }
+}
+
+// ProductLocations (Banana can be found at SuperMarket, for example)
+class ProductLocation {
+  Location location;
+  Product product;
+
+  ProductLocation(Location _location, Product _product){
+    location = _location;
+    product = _product;
+  }
+}
+
+// Use built-in Timeline functionality to enforce that an agent:
+// a) Can't be at more than one place at a time.
+// b) Can't Go more than one place at a time.
+// c) Can't Go somewhere and be At somewhere at the same time.
+class AgentLocation extends Timeline{
+  predicate At {
+    Location loc;
+  }
+
+  predicate Going {
+    Location from;
+    Location to;
+  }
+}
+
+// In addition to having a location timeline, the agent can buy and own things.  
+// Note that the actions and the location predicates can be concurrent 
+// so they can't be on the same Timeline
+class Agent {
+  AgentLocation location;
+
+  Agent() {
+    location = new AgentLocation();
+  }
+
+  action Buy {
+    Product product;
+  }
+
+  action Go {
+    Location from;
+    Location to;
+  }
+
+  predicate Own {
+  	Product product;
+  }
+}
+
+// Define the rules for our actions:
+Agent::Go {
+  met_by(condition object.location.At origin);
+  eq(from, origin.loc);
+
+  equals(effect object.location.Going going);
+  eq(going.from, from);
+  eq(going.to, to);
+
+  meets(effect object.location.At destination);
+  eq(to, destination.loc);
+}
+
+Agent::Buy {
+  // A Buy takes 10 time units
+  eq(10, duration);
+
+  // initialized to all locations
+  ProductLocation possibleStores;
+
+  // limit possibleStores variables to ones that provide what we need to buy
+  eq(product, possibleStores.product);
+
+  // We must be At a location during our Buy, and that location must have the
+  // product we want available:
+  contained_by(condition object.location.At currLocation);
+  eq(currLocation.loc, possibleStores.location);
+
+  starts(effect Own purchase);
+  eq(purchase.product,product);
+}
+
+\end{verbatim}
+
+TODO: explain variable, object, predicate and action types, including constraints that can be part of them
+
+In this model we can see:
+
+\begin{enumerate}
+	\item Variable Types:
+	\item Object Types:
+	\item Predicate Types:
+	\item Action Types:
+\end{enumerate}
+
+Now let's look at how a specific problem instance would be specified in NDDL:
+
+\begin{verbatim}
+
+// Allocate instances
+Location Home = new Location("Home");
+Location SuperMarket = new Location("SuperMarket");
+Location HardwareStore = new Location("HardwareStore");
+
+Product Banana = new Product("Banana");
+Product Milk = new Product("Milk");
+Product Drill = new Product("Drill");
+
+ProductLocation bananaLocation = new ProductLocation(SuperMarket, Banana);
+ProductLocation milkLocation = new ProductLocation(SuperMarket, Milk);
+ProductLocation drillLocation = new ProductLocation(HardwareStore, Drill);
+
+Agent agent = new Agent();
+
+// Indicate that the database is closed - no new objects can be created
+// this allows EUROPA to perform more efficient inference
+close();
+
+// We start the day at Home:
+fact(agent.location.At atHomeForBreakfast);
+atHomeForBreakfast.loc.specify(Home);
+
+// Goals for all of the agent's needs, buy if needed
+goal(agent.Own gotMilk);
+gotMilk.product.specify(Milk);
+goal(agent.Own gotBanana);
+gotBanana.product.specify(Banana);
+goal(agent.Own gotDrill);
+gotDrill.product.specify(Drill);
+
+// Make sure agent is home for dinner
+goal(agent.location.At atHomeForDinner);
+atHomeForDinner.loc.specify(Home);
+
+// Agent has all day to satisfy goals:
+gotMilk after atHomeForBreakfast;
+gotMilk before atHomeForDinner;
+
+gotBanana after atHomeForBreakfast;
+gotBanana before atHomeForDinner;
+
+gotDrill after atHomeForBreakfast;
+gotDrill before atHomeForDinner;
+
+// Force things to happen within our planning horizon:
+int horizonStart = 0;
+int horizonEnd = 100;
+leq(horizonStart, atHomeForBreakfast.start);
+leq(atHomeForDinner.end, horizonEnd);
+
+\end{verbatim}
+
+TODO:  explain variable, object, predicate and constraint instances.
+explain goals and facts.
+
+\begin{enumerate}
+	\item Variable Instances:
+	\item Object Instances:
+	\item Facts:
+	\item Goals:
+	\item Constraint Instances:
+\end{enumerate}
 
@@
 TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
-- overview of the types of flaws and violations EUROPA detects.
+- talk about AC3 implementation and extension mechanisms
+- talk about europa's constraint library
+
+The Temporal Network:
+The notion of time is central to temporal planning.
+EUROPA uses variables to explicitly represent timepoints for plan activities and states.
+Constraints among timepoints provide a natural way to express domain axioms.
+For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
+Dechter et al.
+(Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN).
+Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node.
+The diagram below illustrates a simple STN with just 2 variables and a single constraint.
+It also shows the resulting DG.
+
+TODO: add figure
+
+Dechter et al.
+(Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle.
+A negative cycle is a path from a node to itself that has a path length less than 0.
+If such a cycle exists, the network is inconsistent.
+It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search.
+Thus we have an efficient and complete algorithm for propagating an STN.
+These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.
+
+Resources:
+
+    Reservoir resources can have both 'consume' and 'produce' tokens.
+Note that these tokens do NOT have the usual start/end/duration token variables, since occur at a single instant in time.
+Instead, they have a single variable, 'time' used to represent that instant.
+    Reusable resources have 'uses' tokens that use a quantity for the duration of the token. These tokens do have the usual start/end/duration variables.
+    Unary resources are reusable resources with unit quantity and have 'use' tokens (again, with the usual start/end/duration variables).
+
+    TODO: talk about resource envelope computation by using MaxFlow formulation.
 
@@
 The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
+EUROPA's built-in solver implements a chronological backtracking algorithm \ref{alg:europa:solve}
+ % summarized in the figure below.
 
-\begin{verbatim}
+\begin{algorithm}[H]
+\KwIn{PartialPlan $plan$}
+\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
+\Begin{
+  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
+  \BlankLine
+  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
+  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
+  \BlankLine
+  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
+  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
+    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
+    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
+    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
+   }
+   \Return \texttt{false} \; \label{li:noplan}
+ }
+\caption{$\mathrm{bool} ~ solve(plan)$}
+\label{alg:europa:solve}
+\end{algorithm}
 
-01.  bool solve(PartialPlan plan)
-02.   {
-03.      if (isInconsistent(plan))  return false;
-04.
-05.      Flaw flaw = chooseFlaw(plan);
-06.      if (flaw == NULL) return true;
-07.
-08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
-09.
-10.      while(decision.hasNext()) {
-11.          PartialPlan pp = decision.executeNext();
-12.
-13.          if (solve(pp)) return true;
-14.          else d.undo();
-25.      }
-16.
-17.      return false;
-18.  }
+% \begin{verbatim}
 
-\end{verbatim}
+% 01.  bool solve(PartialPlan plan)
+% 02.   {
+% 03.      if (isInconsistent(plan))  return false;
+% 04.
+% 05.      Flaw flaw = chooseFlaw(plan);
+% 06.      if (flaw == NULL) return true;
+% 07.
+% 08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
+% 09.
+% 10.      while(decision.hasNext()) {
+% 11.          PartialPlan pp = decision.executeNext();
+% 12.
+% 13.          if (solve(pp)) return true;
+% 14.          else d.undo();
+% 25.      }
+% 16.
+% 17.      return false;
+% 18.  }
+
+% \end{verbatim}
 
@@
 \begin{enumerate}
-	\item Line 3: Propagate the constraints to test for inconsistency. If found to be inconsistent, then we can return false since this is a dead-end i.e. no refinements to p can yield a consistent plan.
-	\item Line 5: Choose a flaw from the set of available flaws.
-	\item Line 6: If there are no flaws, then p is complete and we can terminate with success.
-	\item Line 8: Otherwise we formulate a decision point which is a branch in the search space.
+	\item Line \ref{li:propagate}: Propagate the constraints to test for inconsistency.
+If found to be inconsistent, then we can return false since this is a dead-end i.e.
+no refinements to p can yield a consistent plan.
+	\item Line \ref{li:selectflaw}: Choose a flaw from the set of available flaws.
+	\item Line \ref{li:noflaw}: If there are no flaws, then p is complete and we can terminate with success.
+	\item Line \ref{li:decisionpt}: Otherwise we formulate a decision point which is a branch in the search space.
 Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
-	\item Line 10: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
-	\item Line 11: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
-	\item Line 13: Recursive call to solve the new planning problem. If successful, then we are done.
-	\item Line 14: Otherwise, retract the last refinement operation and move on to try the next one.
-	\item Line 17:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially. Thus the problem cannot be solved.
+	\item Line \ref{li:decisionloop}: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
+	\item Line \ref{li:decide}: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
+	\item Line \ref{li:recurse}: Recursive call to solve the new planning problem. If successful, then we are done.
+	\item Line \ref{li:backtrack}: Otherwise, retract the last refinement operation and move on to try the next one.
+	\item Line \ref{li:noplan}:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially.
+Thus the problem cannot be solved.
 \end{enumerate}
@@
 
+TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
+examples:
+- deal with earliest flaws first
+- deal with planning (open condition flaws) before scheduling (threats) ones
+
 It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?
+
+N-Queens solver:
+\begin{verbatim}
+    public void solve(int maxIter)
+    {
+    	init();
+
+    	for (int i=0;psengine_.getViolation() > 0 && i < maxIter;i++) {
+    	    PSVariable queenToMove = getQueenWithMaxViolation();
+    		int curPos = queenToMove.getSingletonValue().asInt();
+
+    	    boolean moved = false;
+            SortedSet<Move> moves = getMoves(queenToMove,curPos);             
+            for (Move m : moves) {
+            	moved = makeMove(queenToMove,curPos,m,false);
+            	if (moved)
+            	    break;
+            }
+
+            if (!moved) 
+            	makeMove(queenToMove,curPos,moves.first(),true);
+
+            checkSolution(); // See if we have a new best solution            
+            notifyIterationCompleted(curIteration_++);
+
+            if (curIteration_-bestIter_ > 50) 
+            	restart();
+    	}
+    }
+\end{verbatim}
+
+RCPSP solver:
+\begin{verbatim}
+    public void solve(PSEngine psengine,
+    		          long timeout, // in msecs
+    		          int bound)
+    {
+       init(psengine,timeout,bound);
+
+       for (int i=0; true ; i++) {
+          flatten();
+          updateSolution(i);
+          updateCriticalPrecedences();
+
+          if ((nbStable_ > maxStable_) || (bestMakespan_ <= makespanBound_))
+             break;
+
+          if (timer_.getElapsed() > timeout) {
+              timedOut_ = true;
+              break;
+          }
+
+          relax();
+          curIteration_++;
+       }
+
+       restoreBestSolution();
+    }
+
+\end{verbatim}
 

### 2012-01-31 arch.tex
updated deliberation section of trex to refer to solve algorithm

@@
 
-The overal execution cycle of the agent is emerging through these
-interaction and the processing that occurs in each reactors in order
-to process them. New observation occurring at every tick inform each
-reactor of the current state of the world and impose them to identify
-how it impact their state (and potentially what they planned for the
-future). Goals they receive will trigger in the reactor the need to
-identify a strategy (or plan) that will allow them to reach these new
-states knowing their current state deduced from synchronization.When
-such a plan has been identified, the {\em external}  part is then
-dispatched to other reactors that can do the same. This flow and focus
-around state and goals is perfectly fitting with automated planning
-and even more so to the Europa framework. As an illustration, in 
- Fig. \ref{fig:agent}, only \textsf{Shore comm.} and \textsf{Vehicle}
- reactors are specialized C++ code (acting as interface to external
- programs). All the other reactors are based on the europa framework
- differentiated only by the domain model they manipulate and som
- heuristic configuration parameters. In the following sections we are
- going to focus specifically ion this kind of reactor to illustrate
- how  a formal planning framework can be used in our architecture to
- both track state evolutions and plan for future objectives while
- remaining efficient enough to do these tasks in a real-time control 
-loop.
+The choice of representation of state variables, observations and
+goals is directly derieved to the timleines and tokens from europa. 
+such representation is ich enough to allow the excahnge of state
+information in a formal and yet flexible way and its transaltion in
+planning frameworks that have an explicit representation of time 
+(such as but not limited to europa) is often trivial.
+
+The overall execution cycle of the agent is also abstracted as some
+continuous planning/deliberation cycle between all the
+reactors. Indeed , at the agent interface level, each reactor provided
+only two abstract calls that conceptualize their execution:
+\begin{itemize}
+\item \texttt{synchronize} that takes the last {\em external}
+  observations as argument and will return the {\em internal}
+  observations for this reactor or a failure report.
+\item \texttt{step} that accept new goals requested (if any) on the
+  {\em internal} timeline of this reactor and will execute one step of
+  deliberation. The returned value indicates if a complete plan has
+  been found and in that case will also provide the part of this plan
+  applied to {\em external} timelines of this reactor.
+\end{itemize}
+Both of these two deliberative functions have different scope and
+temporal constraints. The \texttt{synchronize} call is considered as
+atomic and will focus only on state inference from the current tick by
+identifying the current value of the {\em internal} state variables
+from the latest updates on the {\em external} variables. This process
+will then propagate up in the reactor graph at any single tick to
+ensure that all reactors are aware of the current state of the
+world. \texttt{step} relates to a single decision step of the planning
+process which is focused on the future evolution of the state
+variables and it can take multiple steps spanning other multiple
+ticks for a reactor to produce a plan that is both complete and
+valid. when such plan is identified its {\em external} part is then
+dispatched to the corresponding reactors which in turn can start their
+deliberation \texttt{step}s in order to satisfy these new
+objectives. As a consequence physical execution will occur as these
+goals and planning propagate down in the reactor graph until they
+eventually become goals for the lowest level reactor which -- instead
+of deliberating -- may just transform this goal in a physical command
+execution.
+
+The way the agent interleave planning \texttt{step}s and
+synchronization is illustrated in Figure \ref{fig:tick-exec}.
+Synchronization happens at every single tick and will interrupt
+planning allowing this reactor to identify its current state that can
+propagate through the reactor graph in order to ensure t aconsitent
+view of the state of the world between all the reactors.  
+\begin{figure}[!htbp]
+  \centering
+  \vskip-1pc
+  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
+  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
+      Deliberation} is interrupted by {\em synchronization} at the
+    beginning of every {\em tick} allowing integration of state
+    information.}
+  \label{fig:tick-exec}
+  \vskip-0.8pc
+\end{figure}
+This view of the reactor has two parallel tasks aloow for long
+planning deliberation while ensuring that the state evolution
+propagates through the agent as time advance. These two tasks 
+intertwined together give the two natural flows (bottom-up for state
+estimation and top to bottom for future plan) that we expect to see in
+a control loop. As they are both conceptualized as inference processes
+it is natural to implement a reactor based on generic automated
+planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
+  comm.} and \textsf{Vehicle} acting as interfaces to the system,
+every other reactors (colored in blue) are different instance of our
+europa reactor differntiated only by their model and europa solver
+configurations.
 
@@
   the agent itself to emulate the multi-threading of the reactors
-  deliberation and synchronization}. For this reason we 
-decompose the algorithm \ref{alg:europa:solve} inside the reactor by
-having the recursive calls decomposed in single steps. This step
-function will be called after synchronization and as long as the
-previous call in this tick did not return \textsf{true} -- which would
-express that the current solution is complete.
-{\em will need to change that in algorithm environment and refine it}
-
-\begin{verbatim}
-bool step(PartialPlan plan, DecisionStack stack) 
-   if( isInconsistent(plan) ) {
-     relaxPlan(plan, stack);
-     notify("No plan");
-     return true;
-   } 
-   Flaw flaw = chooseFlaw(flaw);
-   if( flaw==NULL )
-     return true;
-   DecisionPoint decision = makeDecisionPoint(flaw, plan);
-   while(1) {
-     while( decision.hasNext() ) {
-       PartialPlan pp = decision.executeNext();
-       if( isInconsistent(pp) ) {
-         decision.undo();
-       }  else {
-          plan = pp;
-          stack.push(decision);
-          return chooseFlaw(flaw)==NULL;
-       }
-     }
-     if( decisionStack.empty() ) {
-       notify("No plan");
-       return true;
-     } else {
-        repeat {
-          decision = decisionStack.pop();
-          decision.undo();
-        } until( decision.hasNext() );
-        return false;
-     }
-   } 
-}
-\end{verbatim}
+  deliberation and synchronization}. For this reason while the overal
+planning of a europa reactor strictly relies on the europa framework
+plnanning solver, its slices the execution of Algorithm
+\ref{alg:europa:solve} in atomic steps. The interruption is done
+around the recursive call (line \ref{li:recurse}) while always
+ensuring that the current partial plan mainitained is never poroven
+inconsitent at any single step. This can be managed by allowing the
+call to backtrack until a DecisionPoInt that is not exhausted (ie
+$decision.hasNext()$ is \texttt{true}) is find in the call stack.
 
@@
 
-When the number of flaw is 0 A plan solution is found. This has two
-effects:
+The scope of planning for the reactor plays also a role in the
+deliberation process as we limit deliberation to the sliding temporal
+window $[\tau+\lambda, \tau+\lambda+\pi]$ where:
+\begin{itemize}
+\item $\tau$ is the current tick
+\item $\lambda$ is the specified latency of this reactor. Which is an
+  indicator on the expected maximum time this reactor is expected
+  before being able to produce a plan.\footnote{Do note that this
+    parameter remains indicative and, while it is highly recommended to
+    select a sound value, a failure to produce a plan in this delay is
+    not considered as a critical failure of the reactor}
+\item $\pi$ is the planning look-ahead of the reactor. which indicates
+  how many ticks in the future this reactor is looking ahead while planning.
+\end{itemize}
+This scope allow to filter out any tokens that are either necessarily
+ending before $\tau$ or necessarily starting after. Focusing the
+[planning problem to only the tokens that are in a reasonable
+future. This window is taken into account by the agent which will
+notify the reactor of new {\em internal} goals only when they overlap
+this window but also at the deliberation solver in order to reduce the
+number of tokens to be evaluated and consequently reduce the cost of
+deliberation. As time advance flaws that where initially beyond this
+window will become active flaw that the reactor can then evaluate
+resulting on an apparent continuous planning of the reactor.
+
+When there's no more flaw is present in the current planning scope a
+plan  solution is found. This has two effects:
 \begin{enumerate}

### 2012-02-01 europa.tex
checkpoint. more europa content.

@@
 
+\subsection{EUROPA's Plan Representation}
+\label{sec:europa:pr}
+
+Variables
+
+Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+
+Objects
+
+The things we wish to describe and refer to in a domain are considered Objects.
+As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
+In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
+Objects have state and behavior.
+For example, a camera can be:
+
+    off,
+    ready, or
+    taking a picture.
+
+An attitude controller can be:
+
+    pointing at a position, or
+    turning from one position to another.
+
+An object is an instance of a class.
+In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
+In order to describe such state and behavior we build on the formalism of first order logic.
+Temporally Qualified Predicates
+
+A predicate defines a relation between objects and properties.
+In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
+For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
+Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
+Similarly, p is a variable whose values are the set of possible positions.
+In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
+
+In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
+Predicates that are always true can be thought to hold from the beginning to the end of time.
+However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
+So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
+In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
+In EUROPA, all predicate instances are Tokens.
+Tokens
+
+A Token is an instance of a predicate and is defined over a temporal extent. Every token has five built-in variables:
+
+    start: The beginning of the temporal extent over which the predicate is defined.
+    end: The end of the temporal extent over which the predicate is defined.
+    duration: The constraints start + duration = end is enforced automatically.
+    object: made.
+The set of objects to which a token might apply.
+In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
+its state or behavior) in time.
+However, in a partial plan, the commitment to a specific object may not yet have been made.
+    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
+The state variable captures the token's current state and its reachable states through further restriction.
+See the Token State Model discussion below for details.
+
+Objects (continued)
+
+As we have noted, Tokens describe some aspect of an Object in time.
+Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
+Within this general framework, we note a few particulars:
+
+    Static Facts: Classes without predicates lead to Objects without Tokens.
+This arises where an objects state or behavior is independent of time.
+For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
+    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
+Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
+Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
+See below for further discussion.
+    Resources: Metric resources, e.g.
+the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
+produce, consume, use, change.
+These changes are captured as tokens.
+Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
+Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
+
+Dynamic Objects
+
+TODO (there was a stub in the old doxygen docs for this)
+Timelines
+
+It may be sufficient to maintain only a partial-order among tokens.
+However, it is often the case that tokens represent states and actions of a single object in the system.
+Such tokens are typically mutually-exclusive.
+EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
+
+For example, consider the tire-world domain.
+The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
+Clearly, the same tire cannot be in both places at once.
+A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
+The tire is an object in the tire-world represented as a timeline.
+It has predicates associated with it which can describe its states and actions over time.
+The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
+an instance of a Timeline).
+While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
+A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
+This can be important in supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3 contiguous tokens.
+The end and start time-points are related by an equality constraint.
+A Moving predicate has been introduced to cover the transition from one location to another.
+It takes 2 location arguments.
+Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
+To illustrate this, sample times are included.
+Assume a total time range of interest between 0 and 1000.
+In addition, tokens on Timelines have a minimum duration of 1.
+As a result the values shown are the most flexible possible values for the start and end of each token.
+There are a number of advantages of allowing this flexibility.
+First, the basic structure of the plan can be developed without over-committing to specific times.
+If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
+Such an approach permits a least-commitment approach to planning.
+Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
+For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
+In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
+In such domains, flexibility in planning aids robustness in execution.
+Rules
+
+In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
+Rules govern the internal and external relationships of a token.
+For example, consider a parameterized predicate describing a transition from one location to another.
+Let the parameters be from and to.
+The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
+A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
+This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
+It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
+This is an example of a rule governing an external relationship among tokens.
+It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
+The token on which the rule applies is referred to as the master.
+Each Located token required by the master is referred to as a slave.
+All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
+Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
+Token State Model
+
+The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
+Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
+There are 2 possibilities to consider:
+
+    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
+    The token is merged with a matching token already in the plan.
+Once a slave is merged, the requirement it represents are considered satisfied.
+The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
+Merging a token requires finding a target active token that is compatible with the inactive token.
+For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
+The effect of merging is illustrated in Figure 3.
+
+Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
+
+On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
+There is a nuance to the state model for tokens which relate to the mode of its creation.
+If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
+A valid plan can include rejected tokens.
+This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
+This state is not reachable for slaves since that would imply selective adherence to the domain model.
+Figure 4 presents the state transition diagram for token states relevant for planning.
+The transitions are operations on a partial plan which can decide an outcome for an inactive token.
+As operations which may arise in search, they must be reversible during backtracking.
+The cancellation operations in each case are also shown.
+
+Figure 4: Token States and Transitions for Planning
+
+The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
+The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
+The INACTIVE state is captured by the variable being unbound.
+If a value is removed from the domain, that state will not be reachable.
+For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
+Summary
+
+This page has presented the main elements of the EUROPA planning approach. They are:
+
+    Temporally scoped predicates (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
+    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
+    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
+It provides semantics of mutual exclusion sequencing in time.
+Other core abstractions are available in EUROPA for handling metric resources.
+    Domain rules to describe internal and external relationships on and between tokens respectively.
+Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
+The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
+effects which may not occur until some temporal distance after the end of the action).
+    Token States which are the basis of planning operations on a partial plan.
+
 \subsection{EUROPA's Architecture}
@@
 \centering
-\includegraphics[scale=0.35]{figs/EUROPA-Architecture.jpg}
+\includegraphics[scale=0.25]{figs/EUROPA-Architecture.jpg}
 \caption{\small EUROPA's Architecture}
@@
 Also manages tokens, which are a mechanism to group variables to represent temporally scoped state.
-Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can (TODO: provide snippet?)
+Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can.
 
-	TODO: talk about the token network, move stuff from the Concepts section if necessary
-
-	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
+	\item The Rules Engine (RE): manages token relationships in a planning model
 
@@
 	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
+\caption{\small EUROPA's Client API}
+\label{fig:europa-api}
+\vskip+0.1cm
+\end{figure}
 
@@
 \begin{enumerate}
-	\item Variable Types:
+	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
 	\item Object Types:
@@
 The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm \ref{alg:europa:solve}
- % summarized in the figure below.
+EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
 
-\begin{algorithm}[H]
-\KwIn{PartialPlan $plan$}
-\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
-\Begin{
-  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
-  \BlankLine
-  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
-  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
-  \BlankLine
-  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
-  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
-    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
-    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
-    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
-   }
-   \Return \texttt{false} \; \label{li:noplan}
- }
-\caption{$\mathrm{bool} ~ solve(plan)$}
-\label{alg:europa:solve}
-\end{algorithm}
+\begin{verbatim}
 
-% \begin{verbatim}
+01.  bool solve(PartialPlan plan)
+02.   {
+03.      if (isInconsistent(plan))  return false;
+04.
+05.      Flaw flaw = chooseFlaw(plan);
+06.      if (flaw == NULL) return true;
+07.
+08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
+09.
+10.      while(decision.hasNext()) {
+11.          PartialPlan pp = decision.executeNext();
+12.
+13.          if (solve(pp)) return true;
+14.          else d.undo();
+25.      }
+16.
+17.      return false;
+18.  }
 
-% 01.  bool solve(PartialPlan plan)
-% 02.   {
-% 03.      if (isInconsistent(plan))  return false;
-% 04.
-% 05.      Flaw flaw = chooseFlaw(plan);
-% 06.      if (flaw == NULL) return true;
-% 07.
-% 08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
-% 09.
-% 10.      while(decision.hasNext()) {
-% 11.          PartialPlan pp = decision.executeNext();
-% 12.
-% 13.          if (solve(pp)) return true;
-% 14.          else d.undo();
-% 25.      }
-% 16.
-% 17.      return false;
-% 18.  }
-
-% \end{verbatim}
+\end{verbatim}
 
@@
 \begin{enumerate}
-	\item Line \ref{li:propagate}: Propagate the constraints to test for inconsistency.
-If found to be inconsistent, then we can return false since this is a dead-end i.e.
-no refinements to p can yield a consistent plan.
-	\item Line \ref{li:selectflaw}: Choose a flaw from the set of available flaws.
-	\item Line \ref{li:noflaw}: If there are no flaws, then p is complete and we can terminate with success.
-	\item Line \ref{li:decisionpt}: Otherwise we formulate a decision point which is a branch in the search space.
+	\item Line 3: Propagate the constraints to test for inconsistency. If found to be inconsistent, then we can return false since this is a dead-end i.e. no refinements to p can yield a consistent plan.
+	\item Line 5: Choose a flaw from the set of available flaws.
+	\item Line 6: If there are no flaws, then p is complete and we can terminate with success.
+	\item Line 8: Otherwise we formulate a decision point which is a branch in the search space.
 Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
-	\item Line \ref{li:decisionloop}: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
-	\item Line \ref{li:decide}: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
-	\item Line \ref{li:recurse}: Recursive call to solve the new planning problem. If successful, then we are done.
-	\item Line \ref{li:backtrack}: Otherwise, retract the last refinement operation and move on to try the next one.
-	\item Line \ref{li:noplan}:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially.
-Thus the problem cannot be solved.
+	\item Line 10: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
+	\item Line 11: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
+	\item Line 13: Recursive call to solve the new planning problem. If successful, then we are done.
+	\item Line 14: Otherwise, retract the last refinement operation and move on to try the next one.
+	\item Line 17:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially. Thus the problem cannot be solved.
 \end{enumerate}

### 2012-02-01 arch.tex
some initials text on last subsection

@@
 
-\subsubsection{Intertwining synchronization and execution }
+\subsubsection{Intertwining synchronization and execution}
 \label{sec:arch:intertwine}
+
+A core aspect of this reactor is that it takes advantage of the
+specific sequencing enforced by the T-REX architecture along with the
+fact that only one $plan$ structure is maintined in the reactor and
+shared between synchronization and planning to allow these two processes
+to not only be interleaved but impact each other through the way they
+manipulate this plan.
 

### 2012-02-01 trunk/europa.tex
checkpoint. more europa content.

@@
 
+\subsection{EUROPA's Plan Representation}
+\label{sec:europa:pr}
+
+Variables
+
+Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+
+Objects
+
+The things we wish to describe and refer to in a domain are considered Objects.
+As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
+In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
+Objects have state and behavior.
+For example, a camera can be:
+
+    off,
+    ready, or
+    taking a picture.
+
+An attitude controller can be:
+
+    pointing at a position, or
+    turning from one position to another.
+
+An object is an instance of a class.
+In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
+In order to describe such state and behavior we build on the formalism of first order logic.
+
+Temporally Qualified Predicates and Actions
+
+A predicate defines a relation between objects and properties.
+In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
+For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
+Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
+Similarly, p is a variable whose values are the set of possible positions.
+In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
+
+In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
+Predicates that are always true can be thought to hold from the beginning to the end of time.
+However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
+So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
+In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
+In EUROPA, all predicate instances are Tokens.
+
+Tokens
+
+A Token is an instance of a predicate or an action and is defined over a temporal extent. Every token has five built-in variables:
+
+    start: The beginning of the temporal extent over which the predicate is defined.
+    end: The end of the temporal extent over which the predicate is defined.
+    duration: The constraints start + duration = end is enforced automatically.
+    object: made.
+The set of objects to which a token might apply.
+In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
+its state or behavior) in time.
+However, in a partial plan, the commitment to a specific object may not yet have been made.
+    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
+The state variable captures the token's current state and its reachable states through further restriction.
+See the Token State Model discussion below for details.
+
+Objects (continued)
+
+As we have noted, Tokens describe some aspect of an Object in time.
+Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
+Within this general framework, we note a few particulars:
+
+    Static Facts: Classes without predicates lead to Objects without Tokens.
+This arises where an objects state or behavior is independent of time.
+For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
+    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
+Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
+Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
+See below for further discussion.
+    Resources: Metric resources, e.g.
+the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
+produce, consume, use, change.
+These changes are captured as tokens.
+Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
+Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
+
+Dynamic Objects
+
+TODO (there was a stub in the old doxygen docs for this)
+Timelines
+
+It may be sufficient to maintain only a partial-order among tokens.
+However, it is often the case that tokens represent states and actions of a single object in the system.
+Such tokens are typically mutually-exclusive.
+EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
+
+For example, consider the tire-world domain.
+The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
+Clearly, the same tire cannot be in both places at once.
+A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
+The tire is an object in the tire-world represented as a timeline.
+It has predicates associated with it which can describe its states and actions over time.
+The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
+an instance of a Timeline).
+While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
+A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
+This can be important in supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3 contiguous tokens.
+The end and start time-points are related by an equality constraint.
+A Moving predicate has been introduced to cover the transition from one location to another.
+It takes 2 location arguments.
+Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
+To illustrate this, sample times are included.
+Assume a total time range of interest between 0 and 1000.
+In addition, tokens on Timelines have a minimum duration of 1.
+As a result the values shown are the most flexible possible values for the start and end of each token.
+There are a number of advantages of allowing this flexibility.
+First, the basic structure of the plan can be developed without over-committing to specific times.
+If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
+Such an approach permits a least-commitment approach to planning.
+Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
+For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
+In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
+In such domains, flexibility in planning aids robustness in execution.
+Rules
+
+In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
+Rules govern the internal and external relationships of a token.
+For example, consider a parameterized predicate describing a transition from one location to another.
+Let the parameters be from and to.
+The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
+A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
+This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
+It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
+This is an example of a rule governing an external relationship among tokens.
+It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
+The token on which the rule applies is referred to as the master.
+Each Located token required by the master is referred to as a slave.
+All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
+Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
+Token State Model
+
+The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
+Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
+There are 2 possibilities to consider:
+
+    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
+    The token is merged with a matching token already in the plan.
+Once a slave is merged, the requirement it represents are considered satisfied.
+The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
+Merging a token requires finding a target active token that is compatible with the inactive token.
+For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
+The effect of merging is illustrated in Figure 3.
+
+Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
+
+On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
+There is a nuance to the state model for tokens which relate to the mode of its creation.
+If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
+A valid plan can include rejected tokens.
+This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
+This state is not reachable for slaves since that would imply selective adherence to the domain model.
+Figure 4 presents the state transition diagram for token states relevant for planning.
+The transitions are operations on a partial plan which can decide an outcome for an inactive token.
+As operations which may arise in search, they must be reversible during backtracking.
+The cancellation operations in each case are also shown.
+
+Figure 4: Token States and Transitions for Planning
+
+The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
+The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
+The INACTIVE state is captured by the variable being unbound.
+If a value is removed from the domain, that state will not be reachable.
+For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
+Summary
+
+This page has presented the main elements of the EUROPA planning approach. They are:
+
+    Temporally scoped predicates and actions (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
+    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
+    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
+It provides semantics of mutual exclusion sequencing in time.
+Other core abstractions are available in EUROPA for handling metric resources.
+    Domain rules to describe internal and external relationships on and between tokens respectively.
+Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
+The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
+effects which may not occur until some temporal distance after the end of the action).
+    Token States which are the basis of planning operations on a partial plan.
+
 \subsection{EUROPA's Architecture}
@@
 For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\caption{\small Temporal relations defined within the planner are
+  based on \texttt{Allen Algebra} relations shown above.}
+\label{fig:allen-algebra}
+\vskip-0.3cm
+\end{figure}
+
 Dechter et al.

### 2012-02-02 trunk/europa.tex
checkpoint. more europa content.

@@
 
+Planning, for our purposes, can be thought of as determining all the small tasks that must be carried out in order to accomplish a goal.
+Let's say your goal is to buy a gallon of milk.
+It may sound like a simple task, but if you break it down, there are many small tasks involved: obtain keys, obtain wallet, start car, drive to store, find and obtain milk, purchase milk, etc.
+Planning also takes into account rules, called constraints, which control when certain tasks can or cannot happen.
+Two (of many) constraints in this example are that you must obtain your keys and wallet before driving to the store and you must obtain the milk before purchasing it.
+
+Here is what a simple plan for buying milk at the store might look like:
+
 EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  
-Automated Planning can be reformulated as a constraint satisfaction problem (TODO: ref).
-EUROPA is an extension of that idea, however, unlike other approaches where the entire planning problem is translated into a CSP and then solved using traditional CSP methods, EUROPA uses a constraint reasoning engine as a fundamental building block, and enriches it with other architectural components that allow a more direct mapping for scheduling and planning problems, which in turn allows easier development of efficient solving methods.
+TODO: Talk about NASA missions where EUROPA has been used.
 
@@
 Planning:
-In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
+In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products (milk, drill, etc) that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
 The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
@@
 Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
-Next we will examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
+Below we examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
 
@@
 
-There has been 
-State Variable representation
+The most common planning formulations use a propositional representation, where the state of the world is represented by a set of propositions (statements that can be true or false), and operators change the truth values of these propositions (TODO ref?).
+Although these formulations are powerful and have allowed researchers to develop numerous contributions in automated planning, there are many classes of problem domains that are difficult it is difficult to represent using them.
 
-CSP encodings
+In particular, it is hard to represent time, resources, mutual exclusion and concurrency using propositions (TODO: ref?).
+It is straightforward to represent and reason about all of those elements using a CSP representation, as a result there has been some work on doing automated planning while taking advantage of CSP  (TODO: ref).
+Traditionally, the entire planning problem is translated into a CSP and then solved using traditional CSP methods, this leads to a formulation where action choices and relationships are represented as variables and/or constraints.
+However, this has at least two major drawbacks:
+\begin{enumerate}
+	\item Given that action choices and relationships (rules) are expresed through variables, the domain descriptions that result from this approach are not intuitive and therefore difficult to understand and debug
+	\item If the structure of a planning model (actions, conditions, effects, dependencies at the action level, etc) is not explicitly maintained by the CSP planner, the search algorithms are deprived of critical information to make better decisions.
+If that structure is maintained somehow (for instance, by internally marking variables that represent action choice and relationships between actions), it is still hard to write search algorithms as any planning-specific insight has to be translated into the CSP representation of variables and constraints  (TODO: example?)
+\end{enumerate}
 
-Constraint Based and interval planning.
+Constraint-Based Attribute and Interval planning (CAIP) (TODO:ref) is intended to close that gap: on the one hand, it takes advantage of a CSP representation, but it also uses attributes and intervals to maintain an explicit representation of the elements relevant to planning, so that it is easier to write algorithms that search for an reason abut plans.
+The main elements of CAIP are:
+
+\begin{enumerate}
+	\item Intervals:
+	\item Attributes:
+	\item Domain Constraints an Configuration rules:
+\end{enumerate}
+
+TODO: Explain how planning problem instances and their solutions are represented.
 
@@
 \label{sec:europa:pr}
+
+EUROPA is an implementation of the CAIP framework explained in the previous section
 
@@
 
-Dynamic Objects
-
-TODO (there was a stub in the old doxygen docs for this)
 Timelines
@@
 
-TODO: explain variable, object, predicate and action types, including constraints that can be part of them
-
 In this model we can see:
 
+TODO: explain variable, object, predicate and action types, including constraints that can be part of them
 \begin{enumerate}

### 2012-02-02 trunk/europa.tex
Put back algorithm in LaTeX for as I refer to it

@@
 The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
+EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in Algorithm \ref{alg:europa:solve}.
+ % summarized in the figure below.
 
-\begin{verbatim}
-
-01.  bool solve(PartialPlan plan)
-02.   {
-03.      if (isInconsistent(plan))  return false;
-04.
-05.      Flaw flaw = chooseFlaw(plan);
-06.      if (flaw == NULL) return true;
-07.
-08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
-09.
-10.      while(decision.hasNext()) {
-11.          PartialPlan pp = decision.executeNext();
-12.
-13.          if (solve(pp)) return true;
-14.          else d.undo();
-25.      }
-16.
-17.      return false;
-18.  }
-
-\end{verbatim}
+\begin{algorithm}[H]
+\KwIn{PartialPlan $plan$}
+\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
+\Begin{
+  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
+  \BlankLine
+  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
+  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
+  \BlankLine
+  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
+  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
+    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
+    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
+    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
+   }
+   \Return \texttt{false} \; \label{li:noplan}
+ }
+\caption{$\mathrm{bool} ~ solve(plan)$}
+\label{alg:europa:solve}
+\end{algorithm}
 

### 2012-02-02 arch.tex
Further development in intertwined synch & planning

--- 2012-01-31 arch.tex
@@
 
-\subsubsection{Intertwining synchronization and execution }
+\subsubsection{Intertwining synchronization and execution}
 \label{sec:arch:intertwine}
+
+A core aspect of this reactor is that it takes advantage of the
+specific sequencing enforced by the T-REX architecture along with the
+fact that only one $plan$ structure is maintained in the reactor and
+shared between synchronization and planning to allow these two processes
+to not only be interleaved but impact each other through the way they
+manipulate this plan.
+
+The most obvious case is how synchronization allow propagate of {\em
+  external} observations as a new tick occur in the plan. In or
+presentation of synchronization, we isolated the problem by stating
+that we will assume at this point that the reactorhas no future goal
+or more accurateely no plan to enforce at this stage. By doing so we
+were able to develop the way we augment yhe europpa solver to be able
+to propagete {\em external} observations in order to identify the
+reactor {\em internal} state for the same tick. By doing so we were
+able to focus on synchronization as a state evluation process that can
+be resolved within the europa framework with few extensions. simliarly
+we presented deliberation at this fairly high level not necessarily
+discussing how the disruptive nature of synchronization in this
+process could impact the way this planning process will evolve.
+
+We first anlyse the case during which synchronization occurs after the
+planning has vcompleted (ie the last seep of deliberation resulted into
+a plan that is considered as complete for the current look-ahead). As
+synchronization starts with  the plan produced this will gave a frame
+for the search process. Therefore this deliberation will not only be a
+model-based estimation uniquely based on the {\em external}
+observations but will also include the reactor intentions as described
+by the plan produced.
 

### 2012-02-02 trunk/arch.tex
A lot of text added to the closing section of TREX : need to be massaged but I think the ideas are already in there

@@
 
-\subsubsection{Intertwining synchronization and execution }
+\subsubsection{Intertwining synchronization and execution}
 \label{sec:arch:intertwine}
+
+A core aspect of this reactor is that it takes advantage of the
+specific sequencing enforced by the T-REX architecture along with the
+fact that only one $plan$ structure is maintained in the reactor and
+shared between synchronization and planning to allow these two processes
+to not only be interleaved but impact each other through the way they
+manipulate this plan.
+
+The most obvious case is how synchronization allow propagate of {\em
+  external} observations as a new tick occur in the plan. In or
+presentation of synchronization, we isolated the problem by stating
+that we will assume at this point that the reactorhas no future goal
+or more accurateely no plan to enforce at this stage. By doing so we
+were able to develop the way we augment yhe europpa solver to be able
+to propagete {\em external} observations in order to identify the
+reactor {\em internal} state for the same tick. By doing so we were
+able to focus on synchronization as a state evluation process that can
+be resolved within the europa framework with few extensions. simliarly
+we presented deliberation at this fairly high level not necessarily
+discussing how the disruptive nature of synchronization in this
+process could impact the way this planning process will evolve.
+
+We first anlyse the case during which synchronization occurs after the
+planning has vcompleted (ie the last seep of deliberation resulted into
+a plan that is considered as complete for the current look-ahead). As
+synchronization starts with  the plan produced this will gave a frame
+for the search process. Therefore this deliberation will not only be a
+model-based estimation uniquely based on the {\em external}
+observations but will also include the reactor intentions as described
+by the plan produced. Consider for example the Shopping Agent model
+described during the europa presentation and that the
+location of the \texttt{agent} is an {\em external} timeline. For
+example as, the reactor add the goal to have \texttt{Own} milk and the
+current {\em external} states indicated that it was \texttt{At(home)} it
+produced the plan in Fig. \ref{fig:shop:exec0}. The  {\em external}
+  part of the plan -- namely the tokens \texttt{Going(home,
+    superMarket)} and \texttt{At(SuperMarket)} -- and integrated as
+  goals for the reactor managing this specific state varaiable {\em
+    internally}.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}
+  \caption{Shopping example : plan produced before synchronization. 
+    Tokens in red indicate observations, 
+    tokens in blue the goal that this plan attempt to solve. Arrows
+    indicate temporal constraints.}
+  \label{fig:shop:exec0}
+\end{figure}
+
+Now, consider that at the next tick $\tau+1$ the reactor managing the
+agent location was able to change the location state varaibale to 
+the \texttt{Going}. T-REX informs our reactor of this new observation
+which is added in the plan structure as a fact -- starting exactly at
+$\tau+1$ and lastign for an unlknown duration which is at least 1
+tick. As this new observation is integrated in the plan through
+synchronization the solver can identify the similarity between this
+new observation oand the next token in the existing plan. Which mean
+that instead of making a blind exhaustive search of all the possible
+implication the synchronization process starts its execution with the
+assumption that these new observations are consitent with the current
+plan maintained. As a result at the end of synchronization we will
+have the plan depicted in Fig. \ref{fig:shop:exec1}. Which just
+propagated the information provided by the new observation  in the
+exisiting plan (restricting in turn the start time of 
+\texttt{Going} and by propagtion of the constraints \texttt{Go} to  be
+{\em exactly} $\tau+1$). While focusing uniquely on the execution
+frontier ($\tau+1$ in this figure), the europa solver took benefit of
+the exisiting plan to evaluate a solution that asumes that the
+observation is consistent with what was decided  inside this
+reactor.
+
+This choice is directed not only by some natural heurisitc
+choices made in the solver but even more enforced in general by the
+heuristic order of the decision point reolution choices for our {\em
+  internla} state flaw presented in section \ref{sec:arch:sync}
+{\em\color{red} TODO: need to rework the presentation of thee choices
+  so I can refer to them more directly -- algorithm or pseudo theorem
+  should do the trick}. Should such a flaw be revealed during
+synchronization -- meaning that the current state of an {\em internal}
+state varaibale is not fully grounded -- the sequencing of the choices
+will atempt first a conservative choice in term of maintaining
+previous state, then an optimisitic choice in term of the
+execution/advance in our current plan to finally attempt other choices
+that were not predicted by the model. The two first possible choices
+are strongly directed by the exisiting plan which help direct the
+search within this frame. This often allow to have a synchronization
+that is much more directed and i, in nominal situation such as the one
+we showed resolve the deliberation in few steps (often without any
+backtrack in the search).
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
+  \caption{Shopping example : Result of synchronization after the {\em
+      Going} observation was received by the reactor with thew plan
+    showed in Fig. \ref{fig:shop:exec0}.}
+  \label{fig:shop:exec1}
+\end{figure}
+
+The synchronization process is also in this process supporting fully
+the tracking of the plan execution as it specifies tokens parameters
+such as the time points and allow this to propagate through the plan
+in order to idenitfy if the plan remains valid -- as in our example --
+or that it cannot be executed as it is. The later  is often identified
+by the synchronization failing to find a consistent solution. for
+example dshould the \texttt{Going} observation we receive be to the 
+\texttt{hardwareStore}, this would break the current plan of our
+reactor as we cannot go simultaneously at 2 different places as
+illustrated in Fig. \ref{fig:shop:relax}-1. as synchronization is not
+possible in this context our strategy is to fully relax all decisions
+made by the previous plan steps by just keeping n the plan all the
+past observations in both {\em external} and {\em internal} state
+variables and the exsiting {\em internal} goals received by this
+reactor (\texttt{Own(Milk)} in our example) resulting on the plrtial
+plan sghowned in Fig. \ref{fig:shop:relax}-2. In this process the
+reactor inform the agent that all {\em external} goals it requested
+for the future are note valid anymore which often result in the ownere
+of htese timelines to not have to maintain this goal anynmore.
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
+  \caption{Illustration of a conflict suring synchronization and the resulting relaxed plan for recovery}
+  \label{fig:shop:relax}
+\end{figure}
+
+As the plan is not anymore in the way the synchronization can resume 
+and find a solution as we initial lly described. The newly pending goals will enforce  at
+the next \texttt{step} to resumed deliberation on this new fully
+relaxed partial plan. By the simple use of synchronization the reactor 
+was not only able to identify that its plan was not executable anymore 
+but recover from it while offering a new incomplete partial plan to be
+resolved during the next deliberation \texttt{steps}. This on its own
+ogfer the core functionality one should except to close the loop
+between planning and execution in the sense that synchronization plays
+the role of what is typically named as the {\em Executive} in other
+architectures {\em\color{red} references to diverse archis, 3-tiered,
+  laas, claraty, B Wiliams, and many more}.
+
+In our case though the interaction between planning and execution
+does not stop at this level. A core point of the overal schedulling of
+both {\em synchronization} and {\em planning} steps resides on our
+architecture enforsincing that the synchroniuzation process can occur
+in between any step of the plan. This lead us to the fact that not
+only synchronization is imnfluenced by the outcome of planning but
+this influence goes also in the alternate direction. Specifically when
+synchronuiation occuers before we found a complete plan, it will
+inject as we described new facts (the {\em external} observations and
+resulting {\em internal} state) in the plan database. This
+informatiuon will be present when the next planning \texttt{step} will
+resume. Therefore they are not intergal part of the plan resolution
+(generating potentially new flaws for the planner or contributing to
+resolve previously exisiting flaws). The planning process is therefore
+perturbed internallyand fully informed on how the world evolves as it
+is planning.  This aspect trueluy contrast to other approaches we have
+idenitifed in the literature which avoid to perturb the planner during
+it search as it is not compatible with the ``off-line planning'' assumption.
+
+By our design choices -- both at the architecture level and the integration of
+the europa framework for embedded deliberation -- we were able to
+implement an architecture that avoid this limitation while remaining in
+a formal frame that still allow for deliberation and model-based
+agent control. We believe that this tighter integration of planning
+and execution allow the system to ba more accute of its environament
+alowing better informaed decision and tighter reaction. This is
+possible while still integrating information such as a long latency
+for the reactors as long as one can ensure that synchronization of al
+the reactors and deliberation steps can be done in a time that is
+resonably small in comparison of the agent tick duration.
 

### 2012-02-03 concepts.tex
updates

@@
 to be able to synthesize plans insitu and to re-plan. We use a
-temporal constraint-based planner EUROPA$_2$ with a demonstrated
-legacy of having flown on NASA space missions
-\cite{jonsson00,bresina05}. Our autonomy architecture brings three key
-innovations for AUV adaptation: the use of flexible plan
-representations, compositional control with the use of partitioned
-networks and off-line learning to inform insitu environmental state
-estimation. In this section we detail some of the key concepts
-relevant to flexible plan representation. Section \ref{sec:arch} has
-details the \texttt{T-REX} partitioned architecture and estimation
-driven vehicle adaptation.
+temporal constraint-based planner \eu with a demonstrated legacy of
+having flown on NASA space missions \cite{jonsson00,bresina05,
+  barreiro09}. Our autonomy architecture brings three key innovations
+for AUV adaptation: the use of flexible plan representations,
+compositional control with the use of partitioned networks and
+off-line learning to inform insitu environmental state estimation. In
+this section we detail some of the key concepts relevant to flexible
+plan representation. Section \ref{sec:arch} has details the
+\texttt{T-REX} partitioned architecture and estimation driven vehicle
+adaptation.
 
-EUROPA$_2$ uses a domain model written in a declarative language
-(NDDL: New Domain Description Language), together with initial
-conditions and goals also specified in NDDL, to construct a set of
-temporal relations that must be true at the start time. These models
-include assertions about the physics of the vehicle, i.e how it
-responds to external stimulus and internally driven goals. By
-propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
-a set of conditions that should be true in the future, where some of
-these conditions will correspond to actions the agent must take. The
-planner can backtrack and try another path during search if a goal
-cannot be reached while being capable of discarding unachievable
-goals.
+\eu uses a domain model written in a declarative language (NDDL: New
+Domain Description Language), together with initial conditions and
+goals also specified in NDDL, to construct a set of temporal relations
+that must be true at the start time. These models include assertions
+about the physics of the vehicle, i.e how it responds to external
+stimulus and internally driven goals. By propagating these relations
+forward using Simple Temporal Networks \cite{dechter91} and applying
+goal constraints, \eu can select a set of conditions that should be
+true in the future, where some of these conditions will correspond to
+actions the agent must take. The planner can backtrack and try another
+path during search if a goal cannot be reached while being capable of
+discarding unachievable goals.
 
@@
 
-EUROPA$_2$ uses a \emph{state variable} representation to describe the
+\eu uses a \emph{state variable} representation to describe the
 evolution of state over time. The instantiated history of such state
@@
 encoded as variables. A constraint solver in turn manipulates these
-variables defined in a EUROPA$_2$ domain model. For example, consider
+variables defined in a \eu domain model. For example, consider
 a scientific need to take a water samples 100 meters from a hotspot
@@
 relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
-interval arithmetic in the EUROPA$_2$ domain model that allows
+interval arithmetic in the \eu domain model that allows
 encoding of temporal constraint relationships between tokens on the
@@
 \begin{minipage}[c]{\textwidth}
-
+\vspace{+0.5cm}
 % \framebox[\textwidth][t]{

### 2012-02-03 trunk/europa-basics.tex
updates

--- 2012-02-02 trunk/europa.tex
@@
-\section{EUROPA}
+\section{Fundamentals of  Automated Planning}
 \label{sec:europa}
 
-Planning, for our purposes, can be thought of as determining all the small tasks that must be carried out in order to accomplish a goal.
-Let's say your goal is to buy a gallon of milk.
-It may sound like a simple task, but if you break it down, there are many small tasks involved: obtain keys, obtain wallet, start car, drive to store, find and obtain milk, purchase milk, etc.
-Planning also takes into account rules, called constraints, which control when certain tasks can or cannot happen.
-Two (of many) constraints in this example are that you must obtain your keys and wallet before driving to the store and you must obtain the milk before purchasing it.
+{\footnotesize
+  \begin{quote}
+[Planning] \emph{is an abstract, explicit deliberation process that chooses and
+organizes actions by anticipating their expected outcomes. This
+deliberation aims at achieving as best as possible some prestated
+objectives. Automated planning is an area of Artificial Intelligence
+(AI) that studies this deliberation process computationally.} -- from
+\textbf{Automated Planning Theory and Practice} by Ghallab, Nau and
+Traverso \cite{ghallab04} 
+\end{quote}
+}
 
-Here is what a simple plan for buying milk at the store might look like:
+% Planning, for our purposes, can be thought of as determining all the
+% small tasks that must be carried out in order to accomplish a goal.
+To articulate the fundamentals of automated planning briefly and use
+that to motivate the mechanisms we use in our specific form of the
+technique we start with a simple example.
 
-EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  
-TODO: Talk about NASA missions where EUROPA has been used.
+\begin{quotation}
 
-Let's look at some of the problems that can be tackled using EUROPA.
+  In the near future, a personal robot sets out to buy a gallon of milk
+  This involves a number of tasks: obtain keys, obtain wallet,
+  start car, drive to store, find and obtain milk, purchase milk, etc.
+  The embedded planner has to have a ``model'' of the world in which it
+  lives and has to use the task primitives in this model to structure
+  the actions so it achieves its goal. Constraints control when certain
+  tasks can or cannot occur. For example the robot must obtain the keys
+  and wallet \emph{before} driving to the store and pick up the milk
+  \emph{before} purchasing it.
 
-Constraint Satisfaction: 
-In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.
+\end{quotation}
 
-Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
+For such a robot the milk buying plan at the store might look like
+\comment{Figure needed}.
 
-\begin{figure}
-\centering
+% \eu is a general purpose AI planning toolkit developed at NASA Ames.  
+% TODO: Talk about NASA missions where \eu has been used.
+
+\eu is a versatile planning framework which can be used in a range of
+problem-solving methods.
+
+\paragraph \texttt{Constraint Satisfaction}: A canonical problem in
+dealing with constraints is the $N$-Queens problem in which chess queens
+must be placed on an  $N$x$N$ chessboard so no queens attack the other. Fig.
+\ref{fig:nqueens-1} shows an example of a random positioning of queens
+on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
+are highlighted in red.
+
+\begin{figure} \centering
   \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
-\caption{\small N-Queens problem}
+  \caption{\small N-Queens problem. Queens in violation of the
+    non-attack constraint are highlighted in red.}
 \label{fig:nqueens-1}
@@
 
-If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
-Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.
+If we define $N$x$N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in [1,N]$,
+Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen, $0$
+otherwise. Then the following constraints need to be satisfied:
 
-then we have to satisfy the following constraints:
-Sum(Qrc) = 1 ForAll r (Only one Queen per row)
-Sum(Qrc) = 1 ForAll c (Only one Queen per column)
+\begin{equation}
+ Sum(Q_{rc})= \left\{
+\begin{array}{l l}
+  1 & \forall r \quad \mbox{(only one Queen per row)}\\
+  1 & \forall c \quad \mbox{(only one Queen per column)}\\ 
+\end{array} \right.
+Sum(Q_{r+i,c+i}) = 1\\
+Sum(Q_{r-1,c+i}) = 1 \quad \mbox{(only one Queen on each diagonal)}\\
+\end {equation}
 
-Sum(Qr+i,c+i) = 1
-Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)
+\comment{fix \newline and numbering problem above}
 
-We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.
-
-Below is a solution found by EUROPA using that formulation and a specialized search procedure.
+The problem can be solved by finding assignments for all variables
+Q$_{rc}$ that satisfy the above constraints. Fig. \ref{fig:nqueens-2} is
+a solution found by \eu using that formulation and a specialized search
+procedure.
 
@@
 \includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small N-Queens solution}
+\caption{\small N-Queens solution generated by \eu}
 \label{fig:nqueens-2}
@@
 
-Scheduling:
-In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints.
-The activity schedule must also respect fixed limits on the availability of resources required to perform each activity.
-In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
-
-Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
+\paragraph \texttt{Scheduling}: In the Resource Constrained Project
+Scheduling Problem (RCPSP), a project consisting of a set of activities
+must be scheduled in a way that satisfies minimum and/or maximum
+temporal separation constraints. The activity schedule must also respect
+fixed limits on the availability of resources required to perform each
+activity. In addition to satisfying temporal and resource constraints,
+it is common for the user to want to minimize makespan so that the
+entire project is finished as early as possible. Fig. \ref{fig:rcpsp-1}
+shows an example of a a solution provided by \eu for an RCPSP instance
+with 10 activities, 5 resources and 30 temporal constraints.
 
@@
 
-Planning:
-In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products (milk, drill, etc) that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
-The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
+\paragraph \texttt{Planning}: In the Shopping Agent Problem
+\cite{russelnorvig} an agent needs to purchase a set of products (milk,
+drill, etc) that are available at specific locations (supermarket,
+hardware store, etc), the agent is subject to temporal (must complete
+tasks by specific deadlines) and resource (fuel, carrying capacity, etc)
+constraints. The agent must figure out what actions need to be performed
+to find and acquire the required items, as well as when to perform each
+of those actions.
 
-Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.
+Below is a solution produced by \eu for a simple problem instance where
+the shopping agent needs to buy Bananas, Milk and a Drill.
 
@@
 
-As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
+As we can see, a planning problem (what actions must be performed to
+achieve a goal) may embed a scheduling problem (in what order should
+those actions be performed), and both planning and scheduling embed
+Constraint Satisfaction problems since temporal, resource and many other
+kinds of constraints must be respected in most real life problems.
 
-These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs.
-Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
-Below we examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
+These relationships between Planning, Scheduling and Constraint
+Satisfaction have been examined in detail in the literature (TODOL ref
+Frank et al on Planning vs. Scheduling), and they lead to \eu's use of
+constraint reasoning as its innermost building block. Below we examine
+the main ideas from Constraint Satisfaction Programming, and from
+Constraint Based Planning and Scheduling that constitute \eu's
+theoretical underpinnings, then we dive into how \eu implements those
+ideas.
 
@@
 TODO: Explain how planning problem instances and their solutions are represented.
-
-\subsection{EUROPA's Plan Representation}
-\label{sec:europa:pr}
-
-EUROPA is an implementation of the CAIP framework explained in the previous section
-
-Variables
-
-Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
-
-Objects
-
-The things we wish to describe and refer to in a domain are considered Objects.
-As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
-In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
-Objects have state and behavior.
-For example, a camera can be:
-
-    off,
-    ready, or
-    taking a picture.
-
-An attitude controller can be:
-
-    pointing at a position, or
-    turning from one position to another.
-
-An object is an instance of a class.
-In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
-In order to describe such state and behavior we build on the formalism of first order logic.
-
-Temporally Qualified Predicates and Actions
-
-A predicate defines a relation between objects and properties.
-In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
-For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
-Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
-Similarly, p is a variable whose values are the set of possible positions.
-In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
-
-In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
-Predicates that are always true can be thought to hold from the beginning to the end of time.
-However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
-So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
-In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
-In EUROPA, all predicate instances are Tokens.
-
-Tokens
-
-A Token is an instance of a predicate or an action and is defined over a temporal extent. Every token has five built-in variables:
-
-    start: The beginning of the temporal extent over which the predicate is defined.
-    end: The end of the temporal extent over which the predicate is defined.
-    duration: The constraints start + duration = end is enforced automatically.
-    object: made.
-The set of objects to which a token might apply.
-In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
-its state or behavior) in time.
-However, in a partial plan, the commitment to a specific object may not yet have been made.
-    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
-The state variable captures the token's current state and its reachable states through further restriction.
-See the Token State Model discussion below for details.
-
-Objects (continued)
-
-As we have noted, Tokens describe some aspect of an Object in time.
-Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
-Within this general framework, we note a few particulars:
-
-    Static Facts: Classes without predicates lead to Objects without Tokens.
-This arises where an objects state or behavior is independent of time.
-For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
-    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
-Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
-Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
-See below for further discussion.
-    Resources: Metric resources, e.g.
-the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
-produce, consume, use, change.
-These changes are captured as tokens.
-Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
-Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
-
-Timelines
-
-It may be sufficient to maintain only a partial-order among tokens.
-However, it is often the case that tokens represent states and actions of a single object in the system.
-Such tokens are typically mutually-exclusive.
-EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
-
-For example, consider the tire-world domain.
-The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
-Clearly, the same tire cannot be in both places at once.
-A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
-The tire is an object in the tire-world represented as a timeline.
-It has predicates associated with it which can describe its states and actions over time.
-The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
-an instance of a Timeline).
-While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
-A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
-This can be important in supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3 contiguous tokens.
-The end and start time-points are related by an equality constraint.
-A Moving predicate has been introduced to cover the transition from one location to another.
-It takes 2 location arguments.
-Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
-To illustrate this, sample times are included.
-Assume a total time range of interest between 0 and 1000.
-In addition, tokens on Timelines have a minimum duration of 1.
-As a result the values shown are the most flexible possible values for the start and end of each token.
-There are a number of advantages of allowing this flexibility.
-First, the basic structure of the plan can be developed without over-committing to specific times.
-If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
-Such an approach permits a least-commitment approach to planning.
-Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
-For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
-In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
-In such domains, flexibility in planning aids robustness in execution.
-Rules
-
-In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
-Rules govern the internal and external relationships of a token.
-For example, consider a parameterized predicate describing a transition from one location to another.
-Let the parameters be from and to.
-The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
-A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
-This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
-It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
-This is an example of a rule governing an external relationship among tokens.
-It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
-The token on which the rule applies is referred to as the master.
-Each Located token required by the master is referred to as a slave.
-All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
-Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
-Token State Model
-
-The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
-Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
-There are 2 possibilities to consider:
-
-    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
-    The token is merged with a matching token already in the plan.
-Once a slave is merged, the requirement it represents are considered satisfied.
-The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
-Merging a token requires finding a target active token that is compatible with the inactive token.
-For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
-The effect of merging is illustrated in Figure 3.
-
-Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
-
-On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
-There is a nuance to the state model for tokens which relate to the mode of its creation.
-If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
-A valid plan can include rejected tokens.
-This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
-This state is not reachable for slaves since that would imply selective adherence to the domain model.
-Figure 4 presents the state transition diagram for token states relevant for planning.
-The transitions are operations on a partial plan which can decide an outcome for an inactive token.
-As operations which may arise in search, they must be reversible during backtracking.
-The cancellation operations in each case are also shown.
-
-Figure 4: Token States and Transitions for Planning
-
-The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
-The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
-The INACTIVE state is captured by the variable being unbound.
-If a value is removed from the domain, that state will not be reachable.
-For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
-Summary
-
-This page has presented the main elements of the EUROPA planning approach. They are:
-
-    Temporally scoped predicates and actions (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
-    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
-    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
-It provides semantics of mutual exclusion sequencing in time.
-Other core abstractions are available in EUROPA for handling metric resources.
-    Domain rules to describe internal and external relationships on and between tokens respectively.
-Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
-The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
-effects which may not occur until some temporal distance after the end of the action).
-    Token States which are the basis of planning operations on a partial plan.
-
-\subsection{EUROPA's Architecture}
-\label{sec:europa:arch}
-
-The following figure displays the main architectural components in EUROPA and their relationships:
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.25]{figs/EUROPA-Architecture.jpg}
-\caption{\small EUROPA's Architecture}
-\label{fig:europa-architecture}
-\vskip+0.1cm
-\end{figure}
-
-\begin{enumerate}
-
-	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables.
-It also provides an efficient arc consistency mechanism (TODO: ref to AC-3).
-The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.
-
-	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does.
-Also manages tokens, which are a mechanism to group variables to represent temporally scoped state.
-Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can.
-
-	\item The Rules Engine (RE): manages token relationships in a planning model
-
-	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
-
-	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
-\caption{\small EUROPA's Client API}
-\label{fig:europa-api}
-\vskip+0.1cm
-\end{figure}
-
-	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
-		\begin{enumerate}
-			\item Temporal Network
-			\item Resources
-			\item NDDL
-			\item ANML (TODO:maybe it's better to leave out?)
-		\end{enumerate}
-
-\end{enumerate}
-
-\subsection{Modeling}
-\label{sec:europa:modeling}
-
-EUROPA can ingest descriptions of models, plans and goals.
-This means that a single planner can be applied to different systems and problems, simply by providing different models and goals.
-For such planners the expressiveness of the languages for models, goals and plans is of great importance.
-EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way.
-The main characteristics that make NDDL a powerful tool for describing problem domains are:
-
-\begin{enumerate}
-	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++.
-It also supports polymorphism for some of the planning components.
-Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
-	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
-	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
-\end{enumerate}
-
-To understand how the most important elements of NDDL work, let's take a look at how the Shopping Agent problem described above would be modeled:
-
-\begin{verbatim}
-
-// Locations (Home, SuperMarket, etc.)
-class Location {
-  string name;
-  Location(string _name){
-    name = _name;
-  }
-}
-
-// Products (Milk, Banana, etc.) and
-class Product {
-  string name;
-  Product(string _name) {
-    name = _name;
-  }
-}
-
-// ProductLocations (Banana can be found at SuperMarket, for example)
-class ProductLocation {
-  Location location;
-  Product product;
-
-  ProductLocation(Location _location, Product _product){
-    location = _location;
-    product = _product;
-  }
-}
-
-// Use built-in Timeline functionality to enforce that an agent:
-// a) Can't be at more than one place at a time.
-// b) Can't Go more than one place at a time.
-// c) Can't Go somewhere and be At somewhere at the same time.
-class AgentLocation extends Timeline{
-  predicate At {
-    Location loc;
-  }
-
-  predicate Going {
-    Location from;
-    Location to;
-  }
-}
-
-// In addition to having a location timeline, the agent can buy and own things.  
-// Note that the actions and the location predicates can be concurrent 
-// so they can't be on the same Timeline
-class Agent {
-  AgentLocation location;
-
-  Agent() {
-    location = new AgentLocation();
-  }
-
-  action Buy {
-    Product product;
-  }
-
-  action Go {
-    Location from;
-    Location to;
-  }
-
-  predicate Own {
-  	Product product;
-  }
-}
-
-// Define the rules for our actions:
-Agent::Go {
-  met_by(condition object.location.At origin);
-  eq(from, origin.loc);
-
-  equals(effect object.location.Going going);
-  eq(going.from, from);
-  eq(going.to, to);
-
-  meets(effect object.location.At destination);
-  eq(to, destination.loc);
-}
-
-Agent::Buy {
-  // A Buy takes 10 time units
-  eq(10, duration);
-
-  // initialized to all locations
-  ProductLocation possibleStores;
-
-  // limit possibleStores variables to ones that provide what we need to buy
-  eq(product, possibleStores.product);
-
-  // We must be At a location during our Buy, and that location must have the
-  // product we want available:
-  contained_by(condition object.location.At currLocation);
-  eq(currLocation.loc, possibleStores.location);
-
-  starts(effect Own purchase);
-  eq(purchase.product,product);
-}
-
-\end{verbatim}
-
-In this model we can see:
-
-TODO: explain variable, object, predicate and action types, including constraints that can be part of them
-\begin{enumerate}
-	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
-	\item Object Types:
-	\item Predicate Types:
-	\item Action Types:
-\end{enumerate}
-
-Now let's look at how a specific problem instance would be specified in NDDL:
-
-\begin{verbatim}
-
-// Allocate instances
-Location Home = new Location("Home");
-Location SuperMarket = new Location("SuperMarket");
-Location HardwareStore = new Location("HardwareStore");
-
-Product Banana = new Product("Banana");
-Product Milk = new Product("Milk");
-Product Drill = new Product("Drill");
-
-ProductLocation bananaLocation = new ProductLocation(SuperMarket, Banana);
-ProductLocation milkLocation = new ProductLocation(SuperMarket, Milk);
-ProductLocation drillLocation = new ProductLocation(HardwareStore, Drill);
-
-Agent agent = new Agent();
-
-// Indicate that the database is closed - no new objects can be created
-// this allows EUROPA to perform more efficient inference
-close();
-
-// We start the day at Home:
-fact(agent.location.At atHomeForBreakfast);
-atHomeForBreakfast.loc.specify(Home);
-
-// Goals for all of the agent's needs, buy if needed
-goal(agent.Own gotMilk);
-gotMilk.product.specify(Milk);
-goal(agent.Own gotBanana);
-gotBanana.product.specify(Banana);
-goal(agent.Own gotDrill);
-gotDrill.product.specify(Drill);
-
-// Make sure agent is home for dinner
-goal(agent.location.At atHomeForDinner);
-atHomeForDinner.loc.specify(Home);
-
-// Agent has all day to satisfy goals:
-gotMilk after atHomeForBreakfast;
-gotMilk before atHomeForDinner;
-
-gotBanana after atHomeForBreakfast;
-gotBanana before atHomeForDinner;
-
-gotDrill after atHomeForBreakfast;
-gotDrill before atHomeForDinner;
-
-// Force things to happen within our planning horizon:
-int horizonStart = 0;
-int horizonEnd = 100;
-leq(horizonStart, atHomeForBreakfast.start);
-leq(atHomeForDinner.end, horizonEnd);
-
-\end{verbatim}
-
-TODO:  explain variable, object, predicate and constraint instances.
-explain goals and facts.
-
-\begin{enumerate}
-	\item Variable Instances:
-	\item Object Instances:
-	\item Facts:
-	\item Goals:
-	\item Constraint Instances:
-\end{enumerate}
-
-\subsection{Inference}
-\label{sec:europa:inference}
-
-TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
-- talk about AC3 implementation and extension mechanisms
-- talk about europa's constraint library
-
-The Temporal Network:
-The notion of time is central to temporal planning.
-EUROPA uses variables to explicitly represent timepoints for plan activities and states.
-Constraints among timepoints provide a natural way to express domain axioms.
-For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
-\label{fig:allen-algebra}
-\vskip-0.3cm
-\end{figure}
-
-Dechter et al.
-(Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN).
-Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node.
-The diagram below illustrates a simple STN with just 2 variables and a single constraint.
-It also shows the resulting DG.
-
-TODO: add figure
-
-Dechter et al.
-(Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle.
-A negative cycle is a path from a node to itself that has a path length less than 0.
-If such a cycle exists, the network is inconsistent.
-It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search.
-Thus we have an efficient and complete algorithm for propagating an STN.
-These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.
-
-Resources:
-
-    Reservoir resources can have both 'consume' and 'produce' tokens.
-Note that these tokens do NOT have the usual start/end/duration token variables, since occur at a single instant in time.
-Instead, they have a single variable, 'time' used to represent that instant.
-    Reusable resources have 'uses' tokens that use a quantity for the duration of the token. These tokens do have the usual start/end/duration variables.
-    Unary resources are reusable resources with unit quantity and have 'use' tokens (again, with the usual start/end/duration variables).
-
-    TODO: talk about resource envelope computation by using MaxFlow formulation.
-
-\subsection{Search}
-\label{sec:europa:search}
-
-Now we have all the elements in place so that an automated problem solver can be created. Let's recap what those elements are:
-
-\begin{enumerate}
-
-\item A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
-\item A problem instance (also called initial state) that consists of:
-	\item Variable and Object instances that exist for the entire planning horizon.
-	\item Temporally scoped predicate and action instances.
-	\item Temporally scoped goals.
-
-\end{enumerate}
-
-All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.
-
-Given that temporal intervals may extend indefinitely in both positive and negative directions, for problems with a temporal dimension it is also common to specify a Planning Horizon when invoking a solver.
-The Planning Horizon is the temporal interval that a problem solver has to worry about, any decisions that fall completely outside of the Planning Horizon can be ignored.
-This is very useful for recurrent activities, consider for instance a model where a medical checkup has to occur every 3 months, the user can model a recurring activity and then use the Planning Horizon to ensure that a finite number of medical checkup activities are generated as part of a plan.
-This is also useful to ignore activities that may happen too far in the future or in the past to be relevant for a particular plan, this way the user can ensure that the solver is making as few commitments as necessary.
-
-EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals.
-The operations to refine the partial plan PP at any time are:
-\begin{enumerate}
-	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
-	\item Select one such flaw
-	\item Select a resolver for the flaw
-	\item Refine PP by applying the resolver
-	\item If an inconsistency is found, try another resolver.
-	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
-\end{enumerate}
-
-This Plan Space Planning algorithm is translated into EUROPA's representation as follows:
-
-The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances.
-Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:
-
-\begin{enumerate}
-	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
-
-	\item Open Condition: an open condition is an inactive token.
-Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
-Open Conditions can be resolved in three ways:
-\begin{enumerate}
-		\item Merging: A free activity is merged with a matching activity already in the plan.
-An activity a is said to match an activity a?
-if a and a?
-unify and the temporal constraints involving a are satisÞed by a?.
-Thus,  a and a?
-can be considered the same activity and we do not need to introduce a in the plan.
-Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a?
-was introduced in the plan.
-		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
-This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities.
-		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
-\end{enumerate}
-
-	\item Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
-Recall for example that a token may belong to objects (e.g.
-Timelines) which require a total order over their tokens.
-If any 2 tokens could possibly overlap (though not necessarily), then they pose a threat to each other in terms of achieving an extension of the current partial plan which is complete and consistent.
-Similarly, threats may arise where tokens share a common resource and their current state might yield extensions of the current partial plan which are inconsistent.
-Threats are resolved by imposing ordering constraints among tokens.
-
-\end{enumerate}
-
-Open Conditions and Threats allow flaw detection and resolution at a higher-level of abstraction (i.e.
-in terms of objects and tokens) than that of simply binding variables as is common in Constraint Satisfaction Problems.
-This is advantageous when one applies heuristics for ordering choices since it provides a richer context in which to make decisions.
-Furthermore, it aids in reducing the amount of work done by a solver so that only the necessary refinements are made, otherwise leaving the partial plan with maximum flexibility.
-For example, one can omit unbound variables which are time-points of tokens since threats will force a solver to impose restrictions on these variables based on the semantics of the objects to which their tokens apply.
-Thus the planning process may yield partially-ordered plans for which all possible extensions are provably valid.
-
-The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in the figure below.
-
-\begin{verbatim}
-
-01.  bool solve(PartialPlan plan)
-02.   {
-03.      if (isInconsistent(plan))  return false;
-04.
-05.      Flaw flaw = chooseFlaw(plan);
-06.      if (flaw == NULL) return true;
-07.
-08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
-09.
-10.      while(decision.hasNext()) {
-11.          PartialPlan pp = decision.executeNext();
-12.
-13.          if (solve(pp)) return true;
-14.          else d.undo();
-25.      }
-16.
-17.      return false;
-18.  }
-
-\end{verbatim}
-
-The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
-
-\begin{enumerate}
-	\item Line 3: Propagate the constraints to test for inconsistency. If found to be inconsistent, then we can return false since this is a dead-end i.e. no refinements to p can yield a consistent plan.
-	\item Line 5: Choose a flaw from the set of available flaws.
-	\item Line 6: If there are no flaws, then p is complete and we can terminate with success.
-	\item Line 8: Otherwise we formulate a decision point which is a branch in the search space.
-Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
-	\item Line 10: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
-	\item Line 11: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
-	\item Line 13: Recursive call to solve the new planning problem. If successful, then we are done.
-	\item Line 14: Otherwise, retract the last refinement operation and move on to try the next one.
-	\item Line 17:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially. Thus the problem cannot be solved.
-\end{enumerate}
-
-This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.
-Dead-ends in the search are discovered through constraint propagation.
-Constraint propagation is a vehicle for evaluating the consistency of a partial plan and also for filtering infeasible values from consideration prior to commitment, allowing in some cases a strong look-ahead capability which is essential for tractable search.
-Consistency testing is initiated by the isConsistent procedure.
-The algorithm permits a heuristically controlled search by applying orderings for chooseFlaw and makeDecisionPoint.
-It results in a chronologically-backtracking, search.
-
-TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
-examples:
-- deal with earliest flaws first
-- deal with planning (open condition flaws) before scheduling (threats) ones
-
-It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?
-
-N-Queens solver:
-\begin{verbatim}
-    public void solve(int maxIter)
-    {
-    	init();
-
-    	for (int i=0;psengine_.getViolation() > 0 && i < maxIter;i++) {
-    	    PSVariable queenToMove = getQueenWithMaxViolation();
-    		int curPos = queenToMove.getSingletonValue().asInt();
-
-    	    boolean moved = false;
-            SortedSet<Move> moves = getMoves(queenToMove,curPos);             
-            for (Move m : moves) {
-            	moved = makeMove(queenToMove,curPos,m,false);
-            	if (moved)
-            	    break;
-            }
-
-            if (!moved) 
-            	makeMove(queenToMove,curPos,moves.first(),true);
-
-            checkSolution(); // See if we have a new best solution            
-            notifyIterationCompleted(curIteration_++);
-
-            if (curIteration_-bestIter_ > 50) 
-            	restart();
-    	}
-    }
-\end{verbatim}
-
-RCPSP solver:
-\begin{verbatim}
-    public void solve(PSEngine psengine,
-    		          long timeout, // in msecs
-    		          int bound)
-    {
-       init(psengine,timeout,bound);
-
-       for (int i=0; true ; i++) {
-          flatten();
-          updateSolution(i);
-          updateCriticalPrecedences();
-
-          if ((nbStable_ > maxStable_) || (bestMakespan_ <= makespanBound_))
-             break;
-
-          if (timer_.getElapsed() > timeout) {
-              timedOut_ = true;
-              break;
-          }
-
-          relax();
-          curIteration_++;
-       }
-
-       restoreBestSolution();
-    }
-
-\end{verbatim}
-
-TODO: 
-- explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "setobook"
-%%% End:

### 2012-02-03 trunk/europa.tex
updates

@@
-\section{EUROPA}
-\label{sec:europa}
-
-Planning, for our purposes, can be thought of as determining all the small tasks that must be carried out in order to accomplish a goal.
-Let's say your goal is to buy a gallon of milk.
-It may sound like a simple task, but if you break it down, there are many small tasks involved: obtain keys, obtain wallet, start car, drive to store, find and obtain milk, purchase milk, etc.
-Planning also takes into account rules, called constraints, which control when certain tasks can or cannot happen.
-Two (of many) constraints in this example are that you must obtain your keys and wallet before driving to the store and you must obtain the milk before purchasing it.
-
-Here is what a simple plan for buying milk at the store might look like:
-
-EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  
-TODO: Talk about NASA missions where EUROPA has been used.
-
-Let's look at some of the problems that can be tackled using EUROPA.
-
-Constraint Satisfaction: 
-In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.
-
-Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
-\caption{\small N-Queens problem}
-\label{fig:nqueens-1}
-\vskip+0.1cm
-\end{figure}
-
-If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
-Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.
-
-then we have to satisfy the following constraints:
-Sum(Qrc) = 1 ForAll r (Only one Queen per row)
-Sum(Qrc) = 1 ForAll c (Only one Queen per column)
-
-Sum(Qr+i,c+i) = 1
-Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)
-
-We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.
-
-Below is a solution found by EUROPA using that formulation and a specialized search procedure.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small N-Queens solution}
-\label{fig:nqueens-2}
-\vskip+0.1cm
-\end{figure}
-
-Scheduling:
-In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints.
-The activity schedule must also respect fixed limits on the availability of resources required to perform each activity.
-In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
-
-Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small RCPSP problem}
-\label{fig:rcpsp-1}
-\vskip+0.1cm
-\end{figure}
-
-Planning:
-In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products (milk, drill, etc) that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
-The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
-
-Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
-\caption{\small Shopping Agent problem}
-\label{fig:shopping-1}
-\vskip+0.1cm
-\end{figure}
-
-As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
-
-These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs.
-Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
-Below we examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
-
-\subsection{Constraint Programming}
-\label{sec:europa:cp}
-
-Constraint Satisfaction Programming, also know as Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
-A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].
-
-	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN.
-Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains.
-Given a defined conjunctive set of constraints on the variables: $C=\{c1(x1,..,xN), ..., cK(x1,..,xN)\}$, the objective is to find one or more value assignments to V where all constraints are satisfied.
-
-To solve a problem, CP uses logical inference to perform 2 operations:
-
-\begin{enumerate}
-	\item Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
-
-	\item Domain reduction: To infer a valid set of values for a variable.
-For example, for discrete variables, the constraints allDifferent(x1,x2,x3) and xi>=0, xi <=4.
-if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}
-
-\end{enumerate}
-
-CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
-
-Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.
-
-TODO: talk about arc consistency algorithms, computational complexity and solvers.
-
-\subsection{Constraint-Based Attribute and Interval Planning}
-\label{sec:europa:cp}
-
-The most common planning formulations use a propositional representation, where the state of the world is represented by a set of propositions (statements that can be true or false), and operators change the truth values of these propositions (TODO ref?).
-Although these formulations are powerful and have allowed researchers to develop numerous contributions in automated planning, there are many classes of problem domains that are difficult it is difficult to represent using them.
-
-In particular, it is hard to represent time, resources, mutual exclusion and concurrency using propositions (TODO: ref?).
-It is straightforward to represent and reason about all of those elements using a CSP representation, as a result there has been some work on doing automated planning while taking advantage of CSP  (TODO: ref).
-Traditionally, the entire planning problem is translated into a CSP and then solved using traditional CSP methods, this leads to a formulation where action choices and relationships are represented as variables and/or constraints.
-However, this has at least two major drawbacks:
-\begin{enumerate}
-	\item Given that action choices and relationships (rules) are expresed through variables, the domain descriptions that result from this approach are not intuitive and therefore difficult to understand and debug
-	\item If the structure of a planning model (actions, conditions, effects, dependencies at the action level, etc) is not explicitly maintained by the CSP planner, the search algorithms are deprived of critical information to make better decisions.
-If that structure is maintained somehow (for instance, by internally marking variables that represent action choice and relationships between actions), it is still hard to write search algorithms as any planning-specific insight has to be translated into the CSP representation of variables and constraints  (TODO: example?)
-\end{enumerate}
-
-Constraint-Based Attribute and Interval planning (CAIP) (TODO:ref) is intended to close that gap: on the one hand, it takes advantage of a CSP representation, but it also uses attributes and intervals to maintain an explicit representation of the elements relevant to planning, so that it is easier to write algorithms that search for an reason abut plans.
-The main elements of CAIP are:
-
-\begin{enumerate}
-	\item Intervals:
-	\item Attributes:
-	\item Domain Constraints an Configuration rules:
-\end{enumerate}
-
-TODO: Explain how planning problem instances and their solutions are represented.
-
 \subsection{EUROPA's Plan Representation}

### 2012-02-03 trunk/setobook.tex
updates

@@
 
+\input{related.tex}
+
 \input{concepts.tex}
 
-\input{related.tex}
+\input{europa-basics.tex}
+
+\input{europa.tex}
 

### 2012-02-03 europa-basics.tex
updates

--- 2012-02-03 trunk/europa-basics.tex
@@
 
-\comment{fix \newline and numbering problem above}
+\comment{fix newline and numbering problem above}
 
@@
 \paragraph \texttt{Scheduling}: In the Resource Constrained Project
-Scheduling Problem (RCPSP), a project consisting of a set of activities
-must be scheduled in a way that satisfies minimum and/or maximum
-temporal separation constraints. The activity schedule must also respect
-fixed limits on the availability of resources required to perform each
-activity. In addition to satisfying temporal and resource constraints,
-it is common for the user to want to minimize makespan so that the
-entire project is finished as early as possible. Fig. \ref{fig:rcpsp-1}
-shows an example of a a solution provided by \eu for an RCPSP instance
-with 10 activities, 5 resources and 30 temporal constraints.
+Scheduling Problem (RCPSP) \comment{citations needed}, a project
+consisting of a set of activities must be scheduled in a way that
+satisfies minimum and/or maximum temporal separation constraints. The
+activity schedule must also respect fixed limits on the availability of
+resources required to perform each activity. In addition to satisfying
+temporal and resource constraints, it is common for the user to want to
+minimize makespan so that the entire project is finished as early as
+possible. Fig. \ref{fig:rcpsp-1} shows an example of a a solution
+provided by \eu for an RCPSP instance with 10 activities, 5 resources
+and 30 temporal constraints.
 
@@
 \includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small RCPSP problem}
+\caption{\small A \eu solution to an RCPSP \comment{citation} problem.}
 \label{fig:rcpsp-1}
@@
 tasks by specific deadlines) and resource (fuel, carrying capacity, etc)
-constraints. The agent must figure out what actions need to be performed
-to find and acquire the required items, as well as when to perform each
-of those actions.
-
-Below is a solution produced by \eu for a simple problem instance where
-the shopping agent needs to buy Bananas, Milk and a Drill.
+constraints. The agent needs to figure out what actions need to be
+performed to find and acquire the required items, as well as when to
+perform each of those actions. Fig. \ref{fig:shopping-1} shows a
+solution produced by \eu for such a problem instance.
 
@@
 \includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
-\caption{\small Shopping Agent problem}
+\caption{\small A \eu solution to a shopping agent problem domain where
+  the agent needs to buy Bananas, Milk and a Drill.}
 \label{fig:shopping-1}
@@
 
-As we can see, a planning problem (what actions must be performed to
-achieve a goal) may embed a scheduling problem (in what order should
-those actions be performed), and both planning and scheduling embed
-Constraint Satisfaction problems since temporal, resource and many other
-kinds of constraints must be respected in most real life problems.
-
-These relationships between Planning, Scheduling and Constraint
-Satisfaction have been examined in detail in the literature (TODOL ref
-Frank et al on Planning vs. Scheduling), and they lead to \eu's use of
-constraint reasoning as its innermost building block. Below we examine
-the main ideas from Constraint Satisfaction Programming, and from
-Constraint Based Planning and Scheduling that constitute \eu's
-theoretical underpinnings, then we dive into how \eu implements those
-ideas.
+As the examples above show, a planning problem (actions to achieve a
+goal) may embed a scheduling problem (what resources are necessary to
+achieve stated goals) and both planning and scheduling could embed a
+Constraint Satisfaction problem. % since temporal, resource and  other
+% kinds of constraints must be respected in most real life problems.
+The relationships between Planning, Scheduling and Constraint
+Satisfaction have been examined \cite{smith00} and have lead to use of
+constraint reasoning in \eu as its innermost building block. % Below we
+% examine the main ideas from Constraint Satisfaction Programming, and
+% from Constraint Based Planning and Scheduling that constitute
+% \texttt{EUROPA}$_2$'s theoretical underpinnings.
 
@@
 
-Constraint Satisfaction Programming, also know as Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
-A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].
+Constraint Satisfaction Programming (also known as Constraint
+Programming (CP)) is a discipline that provides a generic framework for
+representing, solving and making logical inference on constraints. A
+complete treatment of this discipline is given in [MAR98] and [APT03]
+and very concise introductions are provided in [BAR99] and
+[LUS01].\comment{fix citations}
 
-	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN.
-Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains.
-Given a defined conjunctive set of constraints on the variables: $C=\{c1(x1,..,xN), ..., cK(x1,..,xN)\}$, the objective is to find one or more value assignments to V where all constraints are satisfied.
+A constraint programming problem consists of a set of variables $V=
+{x_1,..,x_N}$, where each variable takes values from a domain
+$d_1,..,d_N$; in this chapter we will deal only with discrete and
+finite domains. Given a defined conjunctive set of constraints on the
+variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the objective
+is to find one or more value assignments to $V$ where all constraints
+are satisfied.
 
-To solve a problem, CP uses logical inference to perform 2 operations:
+To solve a problem, CP techniques use logical inference to perform two
+important operations:
 
 \begin{enumerate}
-	\item Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
 
-	\item Domain reduction: To infer a valid set of values for a variable.
-For example, for discrete variables, the constraints allDifferent(x1,x2,x3) and xi>=0, xi <=4.
-if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}
+\item \textbf{Bounds propagation}: To infer upper and lower variable
+  bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \&
+  $x_i \geq 0$, we can infer $[0,2]$ bounds for $x_1$ and $x_2$
+
+\item \textbf{Domain reduction}: To infer a valid set of values for a variable.
+  For example, for  constraints allDifferent($x_1,x_2,x_3$), $x_i \geq
+  0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can infer that the
+  valid domain for $x_3$ is reduced  to $\{2,4\}$
 
@@
 
-CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
+CP is normally implemented as part of a programming language and
+constraints are normally represented as objects [PUG95]
+\comment{citation}. Any constraint can be introduced and, as long as the
+bounds propagation and domain reduction protocols specified by the host
+CP system are enforced, it will be indistinguishable from any other
+``primitive'' CP constraint, such as $\leq$ or $\geq$. In theory,
+solving a CSP problem is NP-Hard \cite{ghallab04}, but often very
+efficient in practice using a number of algorithms and techniques that
+are well understood.
 
-Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.
-
-TODO: talk about arc consistency algorithms, computational complexity and solvers.
+\comment{TODO: talk about arc consistency algorithms, computational
+  complexity and solvers.}
 
@@
 
-The most common planning formulations use a propositional representation, where the state of the world is represented by a set of propositions (statements that can be true or false), and operators change the truth values of these propositions (TODO ref?).
-Although these formulations are powerful and have allowed researchers to develop numerous contributions in automated planning, there are many classes of problem domains that are difficult it is difficult to represent using them.
+The most common planning formulations use a propositional
+representation, where the state of the world is represented by a set
+of propositions (statements that can be true or false), and operators
+change the truth values of these propositions \cite{gen87}. Although
+these formulations are powerful and have allowed researchers to
+develop numerous contributions in automated planning, there are many
+classes of problem domains that are difficult to represent using this
+formalism. In particular, it is hard to represent time, resources,
+mutual exclusion and concurrency using propositions \comment{citation
+  needed}. While CSP representations have traditionally had an edge in
+formulating and representing planning problems, 
+% It is straightforward to represent and reason about all of
+% those elements using a CSP representation, as a result there has been
+% some work on doing automated planning while taking advantage of CSP
+% (TODO: ref). Traditionally, the entire planning problem is translated
+% into a CSP and then solved using traditional CSP methods, this
+leading to a formulation where action choices and relationships are
+represented as variables and/or constraints, there are at least two
+major drawbacks:
 
-In particular, it is hard to represent time, resources, mutual exclusion and concurrency using propositions (TODO: ref?).
-It is straightforward to represent and reason about all of those elements using a CSP representation, as a result there has been some work on doing automated planning while taking advantage of CSP  (TODO: ref).
-Traditionally, the entire planning problem is translated into a CSP and then solved using traditional CSP methods, this leads to a formulation where action choices and relationships are represented as variables and/or constraints.
-However, this has at least two major drawbacks:
 \begin{enumerate}
-	\item Given that action choices and relationships (rules) are expresed through variables, the domain descriptions that result from this approach are not intuitive and therefore difficult to understand and debug
-	\item If the structure of a planning model (actions, conditions, effects, dependencies at the action level, etc) is not explicitly maintained by the CSP planner, the search algorithms are deprived of critical information to make better decisions.
-If that structure is maintained somehow (for instance, by internally marking variables that represent action choice and relationships between actions), it is still hard to write search algorithms as any planning-specific insight has to be translated into the CSP representation of variables and constraints  (TODO: example?)
+
+\item Given that action choices and relationships (rules) are expressed
+  through variables, the domain descriptions that result from this
+  approach are not intuitive and therefore difficult to understand and
+  debug
+
+\item If the structure of a planning model (actions, conditions,
+  effects, dependencies at the action level) is not explicitly
+  maintained by the CSP planner, the search algorithms are deprived of
+  critical information to make better decisions. If that structure is
+  maintained (for instance, by internally marking variables that
+  represent action choice and relationships between actions), it is
+  still hard to write search algorithms as any planning-specific
+  insight has to be translated into the CSP representation of
+  variables and constraints \comment{For example?}
+
 \end{enumerate}
 
-Constraint-Based Attribute and Interval planning (CAIP) (TODO:ref) is intended to close that gap: on the one hand, it takes advantage of a CSP representation, but it also uses attributes and intervals to maintain an explicit representation of the elements relevant to planning, so that it is easier to write algorithms that search for an reason abut plans.
-The main elements of CAIP are:
+Constraint-based Planning and specifically Constraint-Based Attribute
+and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
+close that gap. On one hand, it takes advantage of a CSP
+representation; one the other it also uses attributes and intervals to
+maintain an explicit representation of the elements relevant to
+planning. This makes it easier to write algorithms that search for an
+reason abut plans. The main elements of CAIP are:
 
 \begin{enumerate}
-	\item Intervals:
+	\item Intervals: \comment{needs to be defined}
 	\item Attributes:
@@
 
-TODO: Explain how planning problem instances and their solutions are represented.
+\comment{Explain how planning problem instances and their solutions are represented}

### 2012-02-03 related.tex
updates

@@
 
-\begin{wrapfigure}{r}{0.2\textwidth}
+\begin{figure}[htpb]
  \centering
  \includegraphics[scale=0.3]{figs/planner-arch.pdf}
- \caption{\small A general architectural block diagram for an
-   Artificial Intelligence based Planner.}
+ \caption{\small A general architectural block diagram for an AI based
+   Planner.}
  \label{fig:planner}
-\end{wrapfigure}
+\end{figure}
 
@@
 $3$-layered architecture (see \cite{gat98} for a well thought out
-rationale), notable examples of which include IPEM
-\cite{AmbrosIngerson88}, ROGUE \cite{Haigh98}, the LAAS Architecture
-\cite{alami:1998p820}, the Remote Agent Experiment \texttt{RAX}
-\cite{mus98} and the Autonomous Spacecraft Experiment \texttt{ASE}
-\cite{chien99} (see \cite{Knight01} for a survey). Early efforts were
-cognizant of if absent with, dealing with time and resources for such
-experiments.
+rationale), notable examples of which include \texttt{IPEM}
+\cite{AmbrosIngerson88}, \texttt{ROGUE} \cite{Haigh98}, the LAAS
+Architecture \cite{alami:1998p820}, the Remote Agent Experiment
+\texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
+\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). Early
+efforts were cognizant of if absent with, dealing with time and
+resources for such experiments.
 
@@
 iteratively and incremental plan repair of an anytime plan
-\cite{Zaimag96} as proposed by \texttt{CASPER}
-\cite{chien00}.
+\cite{Zaimag96} as proposed by \texttt{CASPER} \cite{chien00}.
 

### 2012-02-04 trunk/europa.tex
updates

@@
-\subsection{EUROPA's Plan Representation}
-\label{sec:europa:pr}
-
-EUROPA is an implementation of the CAIP framework explained in the previous section
-
-Variables
-
-Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
-
-Objects
-
-The things we wish to describe and refer to in a domain are considered Objects.
-As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
-In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
-Objects have state and behavior.
-For example, a camera can be:
-
-    off,
-    ready, or
-    taking a picture.
-
-An attitude controller can be:
-
-    pointing at a position, or
-    turning from one position to another.
-
-An object is an instance of a class.
-In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
-In order to describe such state and behavior we build on the formalism of first order logic.
-
-Temporally Qualified Predicates and Actions
-
-A predicate defines a relation between objects and properties.
-In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
-For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
-Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
-Similarly, p is a variable whose values are the set of possible positions.
-In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
-
-In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
-Predicates that are always true can be thought to hold from the beginning to the end of time.
-However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
-So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
-In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
-In EUROPA, all predicate instances are Tokens.
-
-Tokens
-
-A Token is an instance of a predicate or an action and is defined over a temporal extent. Every token has five built-in variables:
-
-    start: The beginning of the temporal extent over which the predicate is defined.
-    end: The end of the temporal extent over which the predicate is defined.
-    duration: The constraints start + duration = end is enforced automatically.
-    object: made.
-The set of objects to which a token might apply.
-In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
-its state or behavior) in time.
-However, in a partial plan, the commitment to a specific object may not yet have been made.
-    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
-The state variable captures the token's current state and its reachable states through further restriction.
-See the Token State Model discussion below for details.
-
-Objects (continued)
-
-As we have noted, Tokens describe some aspect of an Object in time.
-Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
-Within this general framework, we note a few particulars:
-
-    Static Facts: Classes without predicates lead to Objects without Tokens.
-This arises where an objects state or behavior is independent of time.
-For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
-    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
-Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
-Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
-See below for further discussion.
-    Resources: Metric resources, e.g.
-the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
-produce, consume, use, change.
-These changes are captured as tokens.
-Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
-Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
-
-Timelines
-
-It may be sufficient to maintain only a partial-order among tokens.
-However, it is often the case that tokens represent states and actions of a single object in the system.
-Such tokens are typically mutually-exclusive.
-EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
-
-For example, consider the tire-world domain.
-The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
-Clearly, the same tire cannot be in both places at once.
-A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
-The tire is an object in the tire-world represented as a timeline.
-It has predicates associated with it which can describe its states and actions over time.
-The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
-an instance of a Timeline).
-While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
-A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
-This can be important in supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3 contiguous tokens.
-The end and start time-points are related by an equality constraint.
-A Moving predicate has been introduced to cover the transition from one location to another.
-It takes 2 location arguments.
-Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
-To illustrate this, sample times are included.
-Assume a total time range of interest between 0 and 1000.
-In addition, tokens on Timelines have a minimum duration of 1.
-As a result the values shown are the most flexible possible values for the start and end of each token.
-There are a number of advantages of allowing this flexibility.
-First, the basic structure of the plan can be developed without over-committing to specific times.
-If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
-Such an approach permits a least-commitment approach to planning.
-Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
-For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
-In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
-In such domains, flexibility in planning aids robustness in execution.
-Rules
-
-In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
-Rules govern the internal and external relationships of a token.
-For example, consider a parameterized predicate describing a transition from one location to another.
-Let the parameters be from and to.
-The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
-A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
-This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
-It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
-This is an example of a rule governing an external relationship among tokens.
-It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
-The token on which the rule applies is referred to as the master.
-Each Located token required by the master is referred to as a slave.
-All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
-Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
-Token State Model
-
-The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
-Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
-There are 2 possibilities to consider:
-
-    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
-    The token is merged with a matching token already in the plan.
-Once a slave is merged, the requirement it represents are considered satisfied.
-The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
-Merging a token requires finding a target active token that is compatible with the inactive token.
-For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
-The effect of merging is illustrated in Figure 3.
-
-Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
-
-On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
-There is a nuance to the state model for tokens which relate to the mode of its creation.
-If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
-A valid plan can include rejected tokens.
-This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
-This state is not reachable for slaves since that would imply selective adherence to the domain model.
-Figure 4 presents the state transition diagram for token states relevant for planning.
-The transitions are operations on a partial plan which can decide an outcome for an inactive token.
-As operations which may arise in search, they must be reversible during backtracking.
-The cancellation operations in each case are also shown.
-
-Figure 4: Token States and Transitions for Planning
-
-The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
-The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
-The INACTIVE state is captured by the variable being unbound.
-If a value is removed from the domain, that state will not be reachable.
-For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
-Summary
-
-This page has presented the main elements of the EUROPA planning approach. They are:
-
-    Temporally scoped predicates and actions (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
-    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
-    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
-It provides semantics of mutual exclusion sequencing in time.
-Other core abstractions are available in EUROPA for handling metric resources.
-    Domain rules to describe internal and external relationships on and between tokens respectively.
-Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
-The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
-effects which may not occur until some temporal distance after the end of the action).
-    Token States which are the basis of planning operations on a partial plan.
-
-\subsection{EUROPA's Architecture}
+\subsection{\eu Architecture}
 \label{sec:europa:arch}
@@
 \centering
-\includegraphics[scale=0.25]{figs/EUROPA-Architecture.jpg}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
 \caption{\small EUROPA's Architecture}

### 2012-02-04 europa-basics.tex
updates

@@
 \comment{Explain how planning problem instances and their solutions are represented}
+
+\subsection{EUROPA's Plan Representation}
+\label{sec:europa:pr}
+
+EUROPA is an implementation of the CAIP framework explained in the previous section
+
+Variables
+
+Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+
+Objects
+
+The things we wish to describe and refer to in a domain are considered Objects.
+As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
+In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
+Objects have state and behavior.
+For example, a camera can be:
+
+    off,
+    ready, or
+    taking a picture.
+
+An attitude controller can be:
+
+    pointing at a position, or
+    turning from one position to another.
+
+An object is an instance of a class.
+In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
+In order to describe such state and behavior we build on the formalism of first order logic.
+
+Temporally Qualified Predicates and Actions
+
+A predicate defines a relation between objects and properties.
+In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
+For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
+Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
+Similarly, p is a variable whose values are the set of possible positions.
+In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
+
+In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
+Predicates that are always true can be thought to hold from the beginning to the end of time.
+However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
+So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
+In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
+In EUROPA, all predicate instances are Tokens.
+
+Tokens
+
+A Token is an instance of a predicate or an action and is defined over a temporal extent. Every token has five built-in variables:
+
+    start: The beginning of the temporal extent over which the predicate is defined.
+    end: The end of the temporal extent over which the predicate is defined.
+    duration: The constraints start + duration = end is enforced automatically.
+    object: made.
+The set of objects to which a token might apply.
+In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
+its state or behavior) in time.
+However, in a partial plan, the commitment to a specific object may not yet have been made.
+    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
+The state variable captures the token's current state and its reachable states through further restriction.
+See the Token State Model discussion below for details.
+
+Objects (continued)
+
+As we have noted, Tokens describe some aspect of an Object in time.
+Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
+Within this general framework, we note a few particulars:
+
+    Static Facts: Classes without predicates lead to Objects without Tokens.
+This arises where an objects state or behavior is independent of time.
+For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
+    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
+Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
+Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
+See below for further discussion.
+    Resources: Metric resources, e.g.
+the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
+produce, consume, use, change.
+These changes are captured as tokens.
+Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
+Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
+
+Timelines
+
+It may be sufficient to maintain only a partial-order among tokens.
+However, it is often the case that tokens represent states and actions of a single object in the system.
+Such tokens are typically mutually-exclusive.
+EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
+
+For example, consider the tire-world domain.
+The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
+Clearly, the same tire cannot be in both places at once.
+A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
+The tire is an object in the tire-world represented as a timeline.
+It has predicates associated with it which can describe its states and actions over time.
+The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
+an instance of a Timeline).
+While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
+A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
+This can be important in supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3 contiguous tokens.
+The end and start time-points are related by an equality constraint.
+A Moving predicate has been introduced to cover the transition from one location to another.
+It takes 2 location arguments.
+Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
+To illustrate this, sample times are included.
+Assume a total time range of interest between 0 and 1000.
+In addition, tokens on Timelines have a minimum duration of 1.
+As a result the values shown are the most flexible possible values for the start and end of each token.
+There are a number of advantages of allowing this flexibility.
+First, the basic structure of the plan can be developed without over-committing to specific times.
+If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
+Such an approach permits a least-commitment approach to planning.
+Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
+For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
+In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
+In such domains, flexibility in planning aids robustness in execution.
+Rules
+
+In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
+Rules govern the internal and external relationships of a token.
+For example, consider a parameterized predicate describing a transition from one location to another.
+Let the parameters be from and to.
+The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
+A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
+This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
+It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
+This is an example of a rule governing an external relationship among tokens.
+It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
+The token on which the rule applies is referred to as the master.
+Each Located token required by the master is referred to as a slave.
+All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
+Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
+Token State Model
+
+The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
+Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
+There are 2 possibilities to consider:
+
+    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
+    The token is merged with a matching token already in the plan.
+Once a slave is merged, the requirement it represents are considered satisfied.
+The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
+Merging a token requires finding a target active token that is compatible with the inactive token.
+For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
+The effect of merging is illustrated in Figure 3.
+
+Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
+
+On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
+There is a nuance to the state model for tokens which relate to the mode of its creation.
+If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
+A valid plan can include rejected tokens.
+This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
+This state is not reachable for slaves since that would imply selective adherence to the domain model.
+Figure 4 presents the state transition diagram for token states relevant for planning.
+The transitions are operations on a partial plan which can decide an outcome for an inactive token.
+As operations which may arise in search, they must be reversible during backtracking.
+The cancellation operations in each case are also shown.
+
+Figure 4: Token States and Transitions for Planning
+
+The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
+The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
+The INACTIVE state is captured by the variable being unbound.
+If a value is removed from the domain, that state will not be reachable.
+For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
+Summary
+
+This page has presented the main elements of the EUROPA planning approach. They are:
+
+    Temporally scoped predicates and actions (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
+    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
+    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
+It provides semantics of mutual exclusion sequencing in time.
+Other core abstractions are available in EUROPA for handling metric resources.
+    Domain rules to describe internal and external relationships on and between tokens respectively.
+Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
+The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
+effects which may not occur until some temporal distance after the end of the action).
+    Token States which are the basis of planning operations on a partial plan.

### 2012-02-04 trunk/europa-basics.tex
updates

@@
 
-TODO: Explain how planning problem instances and their solutions are represented.
+\comment{Explain how planning problem instances and their solutions are represented}
+
+\subsection{\eu Plan Representation}
+\label{sec:europa:pr}
+
+\eu is an implementation of the CAIP framework.
+
+Variables
+
+Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+
+Objects
+
+The things we wish to describe and refer to in a domain are considered
+Objects. As in the case with object-oriented analysis and design, one
+can seek out the nouns in any domain description to find likely
+objects. In a Mars rover example, we might consider the satellite, the
+camera, and the attitude controller (which orients the robot for
+taking pictures) to be objects. Objects have state and behavior. For
+example, a camera can be:
+
+    off,
+    ready, or
+    taking a picture.
+
+An attitude controller can be:
+
+    pointing at a position, or
+    turning from one position to another.
+
+    An object is an instance of a class. In EUROPA we model using the
+    abstraction of a class to speak about all instances having certain
+    properties of state and behavior. In order to describe such state
+    and behavior we build on the formalism of first order logic.
+
+Temporally Qualified Predicates and Actions
+
+A predicate defines a relation between objects and properties. In
+EUROPA, we define such relations between variables whose domains are
+sets of objects and sets of properties. For example, we might use a
+predicate Pointing(a,p) to indicate that a rover's attitude controller
+a is pointing at position p. Note that a is a variable which may have
+a number of possible values in a problem with multiple
+satellites. Similarly, p is a variable whose values are the set of
+possible positions. In a grounded plan, single values will be
+specified for each variable as we saw in the case of a solved
+Constraint Satisfaction Problem.
+
+In general, is is not sufficient to state that a predicate is true
+without giving it some temporal extent over which it holds. Predicates
+that are always true can be thought to hold from the beginning to the
+end of time. However, in practice, the temporal extent of interest
+must be defined with timepoints to represent its start and end. So we
+might want to write Pointing(a,s,e,p) to indicate that the attitude
+controller a is pointing at position p from time s to time e. In fact,
+this pattern of using such predicates to describe both state and
+behavior of objects is so prevalent in EUROPA that we have introduced
+a special construct called a Token which has the built in variables to
+indicate the object to which the statement principally applies and the
+timepoints over which it holds. In EUROPA, all predicate instances are
+Tokens.
+
+Tokens
+
+A Token is an instance of a predicate or an action and is defined over
+a temporal extent. Every token has five built-in variables:
+
+start: The beginning of the temporal extent over which the predicate
+is defined.  end: The end of the temporal extent over which the
+predicate is defined.  duration: The constraints start + duration =
+end is enforced automatically.  object: made. The set of objects to
+which a token might apply. In a grounded plan each Token applies to a
+specific Object, reflecting the intuition that we are using Tokens to
+describe some aspect of an Object (i.e. its state or behavior) in
+time. However, in a partial plan, the commitment to a specific object
+may not yet have been made.
+
+state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
+variable captures the token's current state and its reachable states
+through further restriction. See the Token State Model discussion
+below for details.
+
+Objects (continued)
+
+As we have noted, Tokens describe some aspect of an Object in
+time. Objects thus may have many Tokens in a plan in order to describe
+their state and behavior throughout all points in time of the
+plan. Within this general framework, we note a few particulars:
+
+Static Facts: Classes without predicates lead to Objects without
+Tokens. This arises where an objects state or behavior is independent
+of time. For example, a domain may have a set of locations and/or
+paths for which there is nothing more to say than that they exist.
+
+Timelines: Often objects in a domain must be described by exactly one
+token for every given timepoint in the plan. Such objects are so
+common that we provide a special construct in EUROPA to extend these
+semantics to derived classes. Any instances of a class derived from a
+Timeline will induce ordering requirements among its tokens in order
+to ensure no temporal overlap may occur among them. See below for
+further discussion.
+
+Resources: Metric resources, e.g. the energy of a battery or the
+capacity of a cargo hold, are objects with an explicit quantitative
+state in time and with a circumscribed range of changes that can occur
+to impact that state i.e. produce, consume, use, change. These changes
+are captured as tokens. Resources are such a common requirement for
+EUROPA users that special constructs are also provided for
+them. Instances of classes derived from a Resource will induce
+ordering requirements on their Tokens in order to ensure that the
+level of the resource remains within specified limits.
+
+Timelines
+
+It may be sufficient to maintain only a partial-order among
+tokens. However, it is often the case that tokens represent states and
+actions of a single object in the system. Such tokens are typically
+mutually-exclusive. EUROPA uses a Timeline structure developed in (??)
+to concisely capture system components whose behavior is described
+over time in this manner.
+
+For example, consider the tire-world domain. The tire was located in
+the trunk with the predicate tireLocated(Trunk) and located on the
+ground with the predicate tireLocated(Ground). Clearly, the same tire
+cannot be in both places at once. A Timeline provides a simple method
+for aggregating the statements about a tire such that they are
+mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the
+whereabouts of a tire. The tire is an object in the tire-world
+represented as a timeline. It has predicates associated with it which
+can describe its states and actions over time. The predicate names
+need not contain the tire prefix; this is now implicit since the
+tokens are assigned to a specific tire instance (i.e. an instance of a
+Timeline). While Timelines are a useful element of the EUROPA planning
+paradigm, they are not essential. A more general notion of a system
+Object can be used which does not impose restrictions of
+mutual-exclusion or non-zero duration. This can be important in
+supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3
+contiguous tokens. The end and start time-points are related by an
+equality constraint. A Moving predicate has been introduced to cover
+the transition from one location to another. It takes 2 location
+arguments. Notice that precedence relationships exist between tokens
+such that they cannot overlap but the start and end times may remain
+flexible. To illustrate this, sample times are included. Assume a
+total time range of interest between 0 and 1000. In addition, tokens
+on Timelines have a minimum duration of 1. As a result the values
+shown are the most flexible possible values for the start and end of
+each token. There are a number of advantages of allowing this
+flexibility. First, the basic structure of the plan can be developed
+without over-committing to specific times. If it is not necessary for
+the tire to be on the ground at time-step 4, then a planner should not
+be forced to specify it. Such an approach permits a least-commitment
+approach to planning. Second, in many domains it is simply impossible
+to know in planning exactly how long an activity might take. For
+example, a common activity of driving a car from one point to another
+on a road can take varying amounts of time depending on traffic and
+traffic lights. In such circumstances it is more practical to express
+upper and lower bounds on durations which naturally lead to intervals
+for start and end times. In such domains, flexibility in planning aids
+robustness in execution.  Rules
+
+In order for a plan to be valid, it must comply with all rules and
+regulations pertinent to the application domain in question. Rules
+govern the internal and external relationships of a token. For
+example, consider a parameterized predicate describing a transition
+from one location to another. Let the parameters be from and to. The
+parameters are instantiated on a token as variables whose domain of
+values is the set of all locations in a given problem. A rule
+governing an internal relationship among token variables might
+stipulate that a transition must involve a change in location. This
+can be easily expressed as a constraint on the definition of a
+predicate of the form: from != to. It is reasonable to further
+stipulate that one must be Located somewhere before a transition can
+occur, and one must end up Located somewhere when completed. This is
+an example of a rule governing an external relationship among
+tokens. It specifies a requirement that tokens of the predicate
+Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving
+Figure 2 illustrates the entities and relations involved in specifying
+such a rule on a Moving predicate. The token on which the rule applies
+is referred to as the master. Each Located token required by the
+master is referred to as a slave. All variables are indicated by name
+and their domains are expressed as intervals in the case of temporal
+variables and as enumerations for the remainder. Application of a rule
+on a token can thus cause slave tokens, variables, and constraints to
+be introduced.  Token State Model
+
+The capability of a domain rule to cause a slave token to be created
+is a key vehicle through which planning occurs. Semantically, this
+rule imposes a requirement for supporting tokens to be in the plan in
+order for the master to be valid. There are 2 possibilities to
+consider:
+
+    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
+    The token is merged with a matching token already in the plan.
+Once a slave is merged, the requirement it represents are considered satisfied.
+The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
+Merging a token requires finding a target active token that is compatible with the inactive token.
+For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
+The effect of merging is illustrated in Figure 3.
+
+    Figure 3: Merging an inactive token on an active token. Domain
+    restrictions occur in highlighted variables of active token.
+
+    On creation of a token, where the commitment has not been made yet
+    to activate or merge the token, the token is said to be
+    inactive. There is a nuance to the state model for tokens which
+    relate to the mode of its creation. If the token is allocated
+    explicitly by an external actor, rather than internally through
+    rule firing, it may include the state rejected indicating the
+    planner is permitted to reject the token. A valid plan can include
+    rejected tokens. This typically arises where the token represents
+    a goal that is preferable to achieve but not mandatory. This state
+    is not reachable for slaves since that would imply selective
+    adherence to the domain model. Figure 4 presents the state
+    transition diagram for token states relevant for planning. The
+    transitions are operations on a partial plan which can decide an
+    outcome for an inactive token. As operations which may arise in
+    search, they must be reversible during backtracking. The
+    cancellation operations in each case are also shown.
+
+Figure 4: Token States and Transitions for Planning
+
+The state of a token is embodied with a 5th and final built-in
+variable referred to as the state variable. The planner states are
+values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
+INACTIVE state is captured by the variable being unbound. If a value
+is removed from the domain, that state will not be reachable. For
+example, when a slave is created via a rule, the REJECTED value is
+removed from its state variable.  Summary
+
+This page has presented the main elements of the EUROPA planning approach. They are:
+
+Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+states and actions in time. Note that tokens do not discriminate
+between state and action.
+
+Constraints to describe relationships among tokens. This provides an
+expressive method of describing interactions among tokens in the
+context of Temporal Planning.
+
+Timelines as a concise abstraction to express the evolution of state
+and behavior for a system component. It provides semantics of mutual
+exclusion sequencing in time. Other core abstractions are available in
+EUROPA for handling metric resources.
+
+Domain rules to describe internal and external relationships on and
+between tokens respectively. Rules are applied to active tokens
+(master tokens) referred to as masters and typically produce inactive
+tokens (slaves) which are a key vehicle through which the planning
+process evolves. The rule structure in EUROPA differs from the more
+restrictive commitment to preconditions and effects in classical
+planning which prohibits durative actions and disembodied effects
+(i.e. effects which may not occur until some temporal distance after
+the end of the action).
+
+Token States which are the basis of planning operations on a partial
+plan.

### 2012-02-06 trunk/europa-basics.tex
updates

@@
 Constraint Satisfaction Programming (also known as Constraint
-Programming (CP)) is a discipline that provides a generic framework for
-representing, solving and making logical inference on constraints. A
-complete treatment of this discipline is given in [MAR98] and [APT03]
-and very concise introductions are provided in [BAR99] and
-[LUS01].\comment{fix citations}
+Programming (CP)) is a discipline that provides a generic framework
+for representing, solving and making logical inference on
+constraints. A complete treatment of this discipline is given in
+\cite{marriott98,apt03} and very concise introductions are provided in
+\cite{bartak99,lustig01}.
 
@@
 CP is normally implemented as part of a programming language and
-constraints are normally represented as objects [PUG95]
+constraints are normally represented as objects \cite{puget95}
 \comment{citation}. Any constraint can be introduced and, as long as the
@@
 
-    An object is an instance of a class. In EUROPA we model using the
+    An object is an instance of a class. In \eu we model using the
     abstraction of a class to speak about all instances having certain
@@
 A predicate defines a relation between objects and properties. In
-EUROPA, we define such relations between variables whose domains are
+\eu, we define such relations between variables whose domains are
 sets of objects and sets of properties. For example, we might use a
@@
 this pattern of using such predicates to describe both state and
-behavior of objects is so prevalent in EUROPA that we have introduced
+behavior of objects is so prevalent in \eu that we have introduced
 a special construct called a Token which has the built in variables to
 indicate the object to which the statement principally applies and the
-timepoints over which it holds. In EUROPA, all predicate instances are
+timepoints over which it holds. In \eu, all predicate instances are
 Tokens.
@@
 token for every given timepoint in the plan. Such objects are so
-common that we provide a special construct in EUROPA to extend these
+common that we provide a special construct in \eu to extend these
 semantics to derived classes. Any instances of a class derived from a
@@
 are captured as tokens. Resources are such a common requirement for
-EUROPA users that special constructs are also provided for
+\eu users that special constructs are also provided for
 them. Instances of classes derived from a Resource will induce
@@
 actions of a single object in the system. Such tokens are typically
-mutually-exclusive. EUROPA uses a Timeline structure developed in (??)
+mutually-exclusive. \eu uses a Timeline structure developed in (??)
 to concisely capture system components whose behavior is described
@@
 tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the EUROPA planning
+Timeline). While Timelines are a useful element of the \eu planning
 paradigm, they are not essential. A more general notion of a system
@@
 
-    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
-    The token is merged with a matching token already in the plan.
-Once a slave is merged, the requirement it represents are considered satisfied.
-The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
-Merging a token requires finding a target active token that is compatible with the inactive token.
-For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
-The effect of merging is illustrated in Figure 3.
+The slave is inserted as an active token in the plan. As such, rules
+may be activated on the slave, and it may consume resources.
+
+The token is merged with a matching token already in the plan. Once a
+slave is merged, the requirement it represents are considered
+satisfied. The process of merging passes on all restrictions imposed
+on the slave to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. For an active token and an inactive token to be
+compatible requires that they are instances of the same predicate and
+that no intersections between corresponding variables are empty. The
+effect of merging is illustrated in Figure 3.
 
@@
 
-This page has presented the main elements of the EUROPA planning approach. They are:
+This page has presented the main elements of the \eu planning
+approach. They are:
 
@@
 exclusion sequencing in time. Other core abstractions are available in
-EUROPA for handling metric resources.
+\eu for handling metric resources.
 
@@
 tokens (slaves) which are a key vehicle through which the planning
-process evolves. The rule structure in EUROPA differs from the more
+process evolves. The rule structure in \eu differs from the more
 restrictive commitment to preconditions and effects in classical

### 2012-02-06 trunk/europa-basics.tex
updates

@@
 
-A predicate defines a relation between objects and properties. In
-\eu, we define such relations between variables whose domains are
-sets of objects and sets of properties. For example, we might use a
-predicate Pointing(a,p) to indicate that a rover's attitude controller
-a is pointing at position p. Note that a is a variable which may have
-a number of possible values in a problem with multiple
-satellites. Similarly, p is a variable whose values are the set of
+A predicate defines a relation between objects and properties. In \eu,
+we define such relations between variables whose domains are sets of
+objects and sets of properties. For example, we might use a predicate
+\texttt{Pointing($a,p$)} to indicate that a rover's attitude controller
+$a$ is pointing at position $p$. Note that $a$ is a variable which may
+have a number of possible values in a problem with multiple
+satellites. Similarly, $p$ is a variable whose values are the set of
 possible positions. In a grounded plan, single values will be
@@
 
-In general, is is not sufficient to state that a predicate is true
+In general, it is not sufficient to state that a predicate is true
 without giving it some temporal extent over which it holds. Predicates
@@
 must be defined with timepoints to represent its start and end. So we
-might want to write Pointing(a,s,e,p) to indicate that the attitude
-controller a is pointing at position p from time s to time e. In fact,
-this pattern of using such predicates to describe both state and
-behavior of objects is so prevalent in \eu that we have introduced
-a special construct called a Token which has the built in variables to
-indicate the object to which the statement principally applies and the
-timepoints over which it holds. In \eu, all predicate instances are
-Tokens.
+might want to write \texttt{Pointing($a,s,e,p$)} to indicate that the
+attitude controller a is pointing at position $p$ from time $s$ to
+time $e$. In fact, this pattern of using such predicates to describe
+both state and behavior of objects is so prevalent in \eu that we have
+introduced a special construct called a Token which has the built in
+variables to indicate the object to which the statement principally
+applies and the timepoints over which it holds. In \eu, all predicate
+instances are Tokens.
 

### 2012-02-06 europa-basics.tex
updates
CSP  background section looking close to complete now.

@@
 
-\eu is a versatile planning framework which can be used in a range of
-problem-solving methods.
+\eu is a versatile planning framework, let's start by looking at some examples of the main problem types that it can tackle.
 
-\paragraph \texttt{Constraint Satisfaction}: A canonical problem in
+\paragraph \textit{Constraint Satisfaction}: A canonical problem in
 dealing with constraints is the $N$-Queens problem in which chess queens
@@
 
-\paragraph \texttt{Scheduling}: In the Resource Constrained Project
+\paragraph \textit{Scheduling}: In the Resource Constrained Project
 Scheduling Problem (RCPSP) \comment{citations needed}, a project
@@
 
-\paragraph \texttt{Planning}: In the Shopping Agent Problem
+\paragraph \textit{Planning}: In the Shopping Agent Problem
 \cite{russelnorvig} an agent needs to purchase a set of products (milk,
@@
 
-As the examples above show, a planning problem (actions to achieve a
-goal) may embed a scheduling problem (what resources are necessary to
-achieve stated goals) and both planning and scheduling could embed a
-Constraint Satisfaction problem. % since temporal, resource and  other
+As the examples above show, a Planning problem (actions to achieve a goal) may embed a Scheduling problem (what resources are necessary to achieve stated goals) and both Planning and Scheduling may embed a Constraint Satisfaction problem.
+% since temporal, resource and  other
 % kinds of constraints must be respected in most real life problems.
@@
 Satisfaction have been examined \cite{smith00} and have lead to use of
-constraint reasoning in \eu as its innermost building block. % Below we
-% examine the main ideas from Constraint Satisfaction Programming, and
-% from Constraint Based Planning and Scheduling that constitute
-% \texttt{EUROPA}$_2$'s theoretical underpinnings.
+constraint reasoning in \eu as its innermost building block.
 
@@
 
-To solve a problem, CP techniques use logical inference to perform two
-important operations:
+To solve a problem, CP techniques use logical inference to perform Constraint Propagation, which consists of two
+operations:
 
@@
 
-CP is normally implemented as part of a programming language and
-constraints are normally represented as objects [PUG95]
-\comment{citation}. Any constraint can be introduced and, as long as the
-bounds propagation and domain reduction protocols specified by the host
-CP system are enforced, it will be indistinguishable from any other
-``primitive'' CP constraint, such as $\leq$ or $\geq$. In theory,
-solving a CSP problem is NP-Hard \cite{ghallab04}, but often very
-efficient in practice using a number of algorithms and techniques that
-are well understood.
+A set of constraints and variables that describe a problem domain are typically represented as a Constraint Network, where the variables are nodes , and constraints are arcs between the variables.
+Having this network representation in mind, Arc Consistency for a a pair of variables x,y is achieved if for every value in x's domain, there is a value in the y's domain that is consistent with all the constraints that connect x and y.
+A naive approach to ensuring arc consistency consists of cycling through all the variable pairs and performing constraint propagation until there are no variable domain changes, the AC-3 algorithm \comment {citation [MAC77]} improves over the naive approach by ignoring constraints whose variables were not affected since the last iteration.
+AC-3 is very popular in CP implementations.
 
-\comment{TODO: talk about arc consistency algorithms, computational
-  complexity and solvers.}
+Arc Consistency and other, more sophisticated consistency algorithms \comment {citation CP Handbook} can be used to efficiently prove that a CP problem is unsatisfiable, however, they generally can't prove that a CP problem is satisfiable and provide specific variable assignments that constitute solutions.
+For that, search algorithms like global backtracking or local search are used \comment {citation CP Handbook}, which use consistency as an efficient way to prune the search space.
+
+In theory, solving a CP problem is NP-Hard \cite{ghallab04}, but often very efficient in practice using a number of consistency and search algorithms that are well understood.
+
+CP is usually implemented as part of a programming language and constraints are usually represented as objects \cite{puget95}.
+Any constraint can be introduced and, as long as the constraint propagation protocols specified by the hostCP system are enforced, it will be indistinguishable from any other ``primitive'' CP constraint, such as $\leq$ or $\geq$.
 
@@
 
-\subsection{EUROPA's Plan Representation}
+\subsection{\eu Plan Representation}
 \label{sec:europa:pr}
 
-EUROPA is an implementation of the CAIP framework explained in the previous section
+\eu is an implementation of the CAIP framework.
 
-Variables
+\textit{Variables} Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
 
-Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
-
-Objects
-
-The things we wish to describe and refer to in a domain are considered Objects.
-As in the case with object-oriented analysis and design, one can seek out the nouns in any domain description to find likely objects.
-In a Mars rover example, we might consider the satellite, the camera, and the attitude controller (which orients the robot for taking pictures) to be objects.
-Objects have state and behavior.
-For example, a camera can be:
+\textit{Objects} The things we wish to describe and refer to in a domain are considered
+Objects. As in the case with object-oriented analysis and design, one
+can seek out the nouns in any domain description to find likely
+objects. In a Mars rover example, we might consider the satellite, the
+camera, and the attitude controller (which orients the robot for
+taking pictures) to be objects. Objects have state and behavior. For
+example, a camera can be:
 
@@
 
-An object is an instance of a class.
-In EUROPA we model using the abstraction of a class to speak about all instances having certain properties of state and behavior.
-In order to describe such state and behavior we build on the formalism of first order logic.
+    An object is an instance of a class. In \eu we model using the
+    abstraction of a class to speak about all instances having certain
+    properties of state and behavior. In order to describe such state
+    and behavior we build on the formalism of first order logic.
 
-Temporally Qualified Predicates and Actions
+\textit{Tokens}
 
-A predicate defines a relation between objects and properties.
-In EUROPA, we define such relations between variables whose domains are sets of objects and sets of properties.
-For example, we might use a predicate Pointing(a,p) to indicate that a rover's attitude controller a is pointing at position p.
-Note that a is a variable which may have a number of possible values in a problem with multiple satellites.
-Similarly, p is a variable whose values are the set of possible positions.
-In a grounded plan, single values will be specified for each variable as we saw in the case of a solved Constraint Satisfaction Problem.
+A predicate defines a relation between objects and properties. In \eu,
+we define such relations between variables whose domains are sets of
+objects and sets of properties. For example, we might use a predicate
+\texttt{Pointing($a,p$)} to indicate that a rover's attitude controller
+$a$ is pointing at position $p$. Note that $a$ is a variable which may
+have a number of possible values in a problem with multiple
+satellites. Similarly, $p$ is a variable whose values are the set of
+possible positions. In a grounded plan, single values will be
+specified for each variable as we saw in the case of a solved
+Constraint Satisfaction Problem.
 
-In general, is is not sufficient to state that a predicate is true without giving it some temporal extent over which it holds.
-Predicates that are always true can be thought to hold from the beginning to the end of time.
-However, in practice, the temporal extent of interest must be defined with timepoints to represent its start and end.
-So we might want to write Pointing(a,s,e,p) to indicate that the attitude controller a is pointing at position p from time s to time e.
-In fact, this pattern of using such predicates to describe both state and behavior of objects is so prevalent in EUROPA that we have introduced a special construct called a Token which has the built in variables to indicate the object to which the statement principally applies and the timepoints over which it holds.
-In EUROPA, all predicate instances are Tokens.
+In general, it is not sufficient to state that a predicate is true
+without giving it some temporal extent over which it holds. Predicates
+that are always true can be thought to hold from the beginning to the
+end of time. However, in practice, the temporal extent of interest
+must be defined with timepoints to represent its start and end. So we
+might want to write \texttt{Pointing($a,s,e,p$)} to indicate that the
+attitude controller a is pointing at position $p$ from time $s$ to
+time $e$. In fact, this pattern of using such predicates to describe
+both state and behavior of objects is so prevalent in \eu that we have
+introduced a special construct called a Token which has the built in
+variables to indicate the object to which the statement principally
+applies and the timepoints over which it holds. In \eu, all predicate
+instances are Tokens.
 
-Tokens
+TODO: talk about actions
 
-A Token is an instance of a predicate or an action and is defined over a temporal extent. Every token has five built-in variables:
+A Token is an instance of a predicate or an action and is defined over
+a temporal extent. Every token has five built-in variables:
 
-    start: The beginning of the temporal extent over which the predicate is defined.
-    end: The end of the temporal extent over which the predicate is defined.
-    duration: The constraints start + duration = end is enforced automatically.
-    object: made.
-The set of objects to which a token might apply.
-In a grounded plan each Token applies to a specific Object, reflecting the intuition that we are using Tokens to describe some aspect of an Object (i.e.
-its state or behavior) in time.
-However, in a partial plan, the commitment to a specific object may not yet have been made.
-    state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED.
-The state variable captures the token's current state and its reachable states through further restriction.
-See the Token State Model discussion below for details.
+start: The beginning of the temporal extent over which the predicate
+is defined.  end: The end of the temporal extent over which the
+predicate is defined.  duration: The constraints start + duration =
+end is enforced automatically.  object: made. The set of objects to
+which a token might apply. In a grounded plan each Token applies to a
+specific Object, reflecting the intuition that we are using Tokens to
+describe some aspect of an Object (i.e. its state or behavior) in
+time. However, in a partial plan, the commitment to a specific object
+may not yet have been made.
 
-Objects (continued)
+state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
+variable captures the token's current state and its reachable states
+through further restriction. See the Token State Model discussion
+below for details.
 
-As we have noted, Tokens describe some aspect of an Object in time.
-Objects thus may have many Tokens in a plan in order to describe their state and behavior throughout all points in time of the plan.
-Within this general framework, we note a few particulars:
+\textit{Objects (continued)}
 
-    Static Facts: Classes without predicates lead to Objects without Tokens.
-This arises where an objects state or behavior is independent of time.
-For example, a domain may have a set of locations and/or paths for which there is nothing more to say than that they exist.
-    Timelines: Often objects in a domain must be described by exactly one token for every given timepoint in the plan.
-Such objects are so common that we provide a special construct in EUROPA to extend these semantics to derived classes.
-Any instances of a class derived from a Timeline will induce ordering requirements among its tokens in order to ensure no temporal overlap may occur among them.
-See below for further discussion.
-    Resources: Metric resources, e.g.
-the energy of a battery or the capacity of a cargo hold, are objects with an explicit quantitative state in time and with a circumscribed range of changes that can occur to impact that state i.e.
-produce, consume, use, change.
-These changes are captured as tokens.
-Resources are such a common requirement for EUROPA users that special constructs are also provided for them.
-Instances of classes derived from a Resource will induce ordering requirements on their Tokens in order to ensure that the level of the resource remains within specified limits.
+As we have noted, Tokens describe some aspect of an Object in time. Objects thus may have many Tokens in a plan in order to describe
+their state and behavior throughout all points in time of the plan. Within this general framework, we note a few particulars:
+
+Static Facts: Classes without predicates lead to Objects without
+Tokens. This arises where an objects state or behavior is independent
+of time. For example, a domain may have a set of locations and/or
+paths for which there is nothing more to say than that they exist.
+
+Timelines: Often objects in a domain must be described by exactly one
+token for every given timepoint in the plan. Such objects are so
+common that we provide a special construct in \eu to extend these
+semantics to derived classes. Any instances of a class derived from a
+Timeline will induce ordering requirements among its tokens in order
+to ensure no temporal overlap may occur among them. See below for
+further discussion.
+
+Resources: Metric resources, e.g. the energy of a battery or the
+capacity of a cargo hold, are objects with an explicit quantitative
+state in time and with a circumscribed range of changes that can occur
+to impact that state i.e. produce, consume, use, change. These changes
+are captured as tokens. Resources are such a common requirement for
+\eu users that special constructs are also provided for
+them. Instances of classes derived from a Resource will induce
+ordering requirements on their Tokens in order to ensure that the
+level of the resource remains within specified limits.
 
@@
 
-It may be sufficient to maintain only a partial-order among tokens.
-However, it is often the case that tokens represent states and actions of a single object in the system.
-Such tokens are typically mutually-exclusive.
-EUROPA uses a Timeline structure developed in (??) to concisely capture system components whose behavior is described over time in this manner.
+It may be sufficient to maintain only a partial-order among
+tokens. However, it is often the case that tokens represent states and
+actions of a single object in the system. Such tokens are typically
+mutually-exclusive. \eu uses a Timeline structure developed in (??)
+to concisely capture system components whose behavior is described
+over time in this manner.
 
-For example, consider the tire-world domain.
-The tire was located in the trunk with the predicate tireLocated(Trunk) and located on the ground with the predicate tireLocated(Ground).
-Clearly, the same tire cannot be in both places at once.
-A Timeline provides a simple method for aggregating the statements about a tire such that they are mutually exclusive.
+For example, consider the tire-world domain. The tire was located in
+the trunk with the predicate tireLocated(Trunk) and located on the
+ground with the predicate tireLocated(Ground). Clearly, the same tire
+cannot be in both places at once. A Timeline provides a simple method
+for aggregating the statements about a tire such that they are
+mutually exclusive.
 
@@
 
-Figure 1 illustrates how a Timeline can be used to describe the whereabouts of a tire.
-The tire is an object in the tire-world represented as a timeline.
-It has predicates associated with it which can describe its states and actions over time.
-The predicate names need not contain the tire prefix; this is now implicit since the tokens are assigned to a specific tire instance (i.e.
-an instance of a Timeline).
-While Timelines are a useful element of the EUROPA planning paradigm, they are not essential.
-A more general notion of a system Object can be used which does not impose restrictions of mutual-exclusion or non-zero duration.
-This can be important in supporting partial-order planning.
+Figure 1 illustrates how a Timeline can be used to describe the
+whereabouts of a tire. The tire is an object in the tire-world
+represented as a timeline. It has predicates associated with it which
+can describe its states and actions over time. The predicate names
+need not contain the tire prefix; this is now implicit since the
+tokens are assigned to a specific tire instance (i.e. an instance of a
+Timeline). While Timelines are a useful element of the \eu planning
+paradigm, they are not essential. A more general notion of a system
+Object can be used which does not impose restrictions of
+mutual-exclusion or non-zero duration. This can be important in
+supporting partial-order planning.
 
-For this example, the state of the tire is specified using 3 contiguous tokens.
-The end and start time-points are related by an equality constraint.
-A Moving predicate has been introduced to cover the transition from one location to another.
-It takes 2 location arguments.
-Notice that precedence relationships exist between tokens such that they cannot overlap but the start and end times may remain flexible.
-To illustrate this, sample times are included.
-Assume a total time range of interest between 0 and 1000.
-In addition, tokens on Timelines have a minimum duration of 1.
-As a result the values shown are the most flexible possible values for the start and end of each token.
-There are a number of advantages of allowing this flexibility.
-First, the basic structure of the plan can be developed without over-committing to specific times.
-If it is not necessary for the tire to be on the ground at time-step 4, then a planner should not be forced to specify it.
-Such an approach permits a least-commitment approach to planning.
-Second, in many domains it is simply impossible to know in planning exactly how long an activity might take.
-For example, a common activity of driving a car from one point to another on a road can take varying amounts of time depending on traffic and traffic lights.
-In such circumstances it is more practical to express upper and lower bounds on durations which naturally lead to intervals for start and end times.
-In such domains, flexibility in planning aids robustness in execution.
-Rules
+For this example, the state of the tire is specified using 3
+contiguous tokens. The end and start time-points are related by an
+equality constraint. A Moving predicate has been introduced to cover
+the transition from one location to another. It takes 2 location
+arguments. Notice that precedence relationships exist between tokens
+such that they cannot overlap but the start and end times may remain
+flexible. To illustrate this, sample times are included. Assume a
+total time range of interest between 0 and 1000. In addition, tokens
+on Timelines have a minimum duration of 1. As a result the values
+shown are the most flexible possible values for the start and end of
+each token. There are a number of advantages of allowing this
+flexibility. First, the basic structure of the plan can be developed
+without over-committing to specific times. If it is not necessary for
+the tire to be on the ground at time-step 4, then a planner should not
+be forced to specify it. Such an approach permits a least-commitment
+approach to planning. Second, in many domains it is simply impossible
+to know in planning exactly how long an activity might take. For
+example, a common activity of driving a car from one point to another
+on a road can take varying amounts of time depending on traffic and
+traffic lights. In such circumstances it is more practical to express
+upper and lower bounds on durations which naturally lead to intervals
+for start and end times. In such domains, flexibility in planning aids
+robustness in execution.  Rules
 
-In order for a plan to be valid, it must comply with all rules and regulations pertinent to the application domain in question.
-Rules govern the internal and external relationships of a token.
-For example, consider a parameterized predicate describing a transition from one location to another.
-Let the parameters be from and to.
-The parameters are instantiated on a token as variables whose domain of values is the set of all locations in a given problem.
-A rule governing an internal relationship among token variables might stipulate that a transition must involve a change in location.
-This can be easily expressed as a constraint on the definition of a predicate of the form: from != to.
-It is reasonable to further stipulate that one must be Located somewhere before a transition can occur, and one must end up Located somewhere when completed.
-This is an example of a rule governing an external relationship among tokens.
-It specifies a requirement that tokens of the predicate Located precede and succeed tokens of the predicate Moving.
+In order for a plan to be valid, it must comply with all rules and
+regulations pertinent to the application domain in question. Rules
+govern the internal and external relationships of a token. For
+example, consider a parameterized predicate describing a transition
+from one location to another. Let the parameters be from and to. The
+parameters are instantiated on a token as variables whose domain of
+values is the set of all locations in a given problem. A rule
+governing an internal relationship among token variables might
+stipulate that a transition must involve a change in location. This
+can be easily expressed as a constraint on the definition of a
+predicate of the form: from != to. It is reasonable to further
+stipulate that one must be Located somewhere before a transition can
+occur, and one must end up Located somewhere when completed. This is
+an example of a rule governing an external relationship among
+tokens. It specifies a requirement that tokens of the predicate
+Located precede and succeed tokens of the predicate Moving.
 
-Figure 2: Internal and external relationships from rules on Moving Figure 2 illustrates the entities and relations involved in specifying such a rule on a Moving predicate.
-The token on which the rule applies is referred to as the master.
-Each Located token required by the master is referred to as a slave.
-All variables are indicated by name and their domains are expressed as intervals in the case of temporal variables and as enumerations for the remainder.
-Application of a rule on a token can thus cause slave tokens, variables, and constraints to be introduced.
-Token State Model
+Figure 2: Internal and external relationships from rules on Moving
+Figure 2 illustrates the entities and relations involved in specifying
+such a rule on a Moving predicate. The token on which the rule applies
+is referred to as the master. Each Located token required by the
+master is referred to as a slave. All variables are indicated by name
+and their domains are expressed as intervals in the case of temporal
+variables and as enumerations for the remainder. Application of a rule
+on a token can thus cause slave tokens, variables, and constraints to
+be introduced.  Token State Model
 
-The capability of a domain rule to cause a slave token to be created is a key vehicle through which planning occurs.
-Semantically, this rule imposes a requirement for supporting tokens to be in the plan in order for the master to be valid.
-There are 2 possibilities to consider:
+The capability of a domain rule to cause a slave token to be created
+is a key vehicle through which planning occurs. Semantically, this
+rule imposes a requirement for supporting tokens to be in the plan in
+order for the master to be valid. There are 2 possibilities to
+consider:
 
-    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
-    The token is merged with a matching token already in the plan.
-Once a slave is merged, the requirement it represents are considered satisfied.
-The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged.
-Merging a token requires finding a target active token that is compatible with the inactive token.
-For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty.
-The effect of merging is illustrated in Figure 3.
+The slave is inserted as an active token in the plan. As such, rules
+may be activated on the slave, and it may consume resources.
 
-Figure 3: Merging an inactive token on an active token. Domain restrictions occur in highlighted variables of active token.
+The token is merged with a matching token already in the plan. Once a
+slave is merged, the requirement it represents are considered
+satisfied. The process of merging passes on all restrictions imposed
+on the slave to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. For an active token and an inactive token to be
+compatible requires that they are instances of the same predicate and
+that no intersections between corresponding variables are empty. The
+effect of merging is illustrated in Figure 3.
 
-On creation of a token, where the commitment has not been made yet to activate or merge the token, the token is said to be inactive.
-There is a nuance to the state model for tokens which relate to the mode of its creation.
-If the token is allocated explicitly by an external actor, rather than internally through rule firing, it may include the state rejected indicating the planner is permitted to reject the token.
-A valid plan can include rejected tokens.
-This typically arises where the token represents a goal that is preferable to achieve but not mandatory.
-This state is not reachable for slaves since that would imply selective adherence to the domain model.
-Figure 4 presents the state transition diagram for token states relevant for planning.
-The transitions are operations on a partial plan which can decide an outcome for an inactive token.
-As operations which may arise in search, they must be reversible during backtracking.
-The cancellation operations in each case are also shown.
+    Figure 3: Merging an inactive token on an active token. Domain
+    restrictions occur in highlighted variables of active token.
+
+    On creation of a token, where the commitment has not been made yet
+    to activate or merge the token, the token is said to be
+    inactive. There is a nuance to the state model for tokens which
+    relate to the mode of its creation. If the token is allocated
+    explicitly by an external actor, rather than internally through
+    rule firing, it may include the state rejected indicating the
+    planner is permitted to reject the token. A valid plan can include
+    rejected tokens. This typically arises where the token represents
+    a goal that is preferable to achieve but not mandatory. This state
+    is not reachable for slaves since that would imply selective
+    adherence to the domain model. Figure 4 presents the state
+    transition diagram for token states relevant for planning. The
+    transitions are operations on a partial plan which can decide an
+    outcome for an inactive token. As operations which may arise in
+    search, they must be reversible during backtracking. The
+    cancellation operations in each case are also shown.
 
@@
 
-The state of a token is embodied with a 5th and final built-in variable referred to as the state variable.
-The planner states are values in the domain of this variable {MERGED, ACTIVE, REJECTED}.
-The INACTIVE state is captured by the variable being unbound.
-If a value is removed from the domain, that state will not be reachable.
-For example, when a slave is created via a rule, the REJECTED value is removed from its state variable.
-Summary
+The state of a token is embodied with a 5th and final built-in
+variable referred to as the state variable. The planner states are
+values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
+INACTIVE state is captured by the variable being unbound. If a value
+is removed from the domain, that state will not be reachable. For
+example, when a slave is created via a rule, the REJECTED value is
+removed from its state variable.
 
-This page has presented the main elements of the EUROPA planning approach. They are:
+\textit{Summary}
 
-    Temporally scoped predicates and actions (a.k.a. Tokens) to represent states and actions in time. Note that tokens do not discriminate between state and action.
-    Constraints to describe relationships among tokens. This provides an expressive method of describing interactions among tokens in the context of Temporal Planning.
-    Timelines as a concise abstraction to express the evolution of state and behavior for a system component.
-It provides semantics of mutual exclusion sequencing in time.
-Other core abstractions are available in EUROPA for handling metric resources.
-    Domain rules to describe internal and external relationships on and between tokens respectively.
-Rules are applied to active tokens (master tokens) referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning process evolves.
-The rule structure in EUROPA differs from the more restrictive commitment to preconditions and effects in classical planning which prohibits durative actions and disembodied effects (i.e.
-effects which may not occur until some temporal distance after the end of the action).
-    Token States which are the basis of planning operations on a partial plan.
+This has presented the main elements of the EUROPA planning approach. They are:
+
+\begin{enumerate}
+
+\item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+states and actions in time. Note that tokens do not discriminate
+between state and action.
+
+\item Token States which are the basis of planning operations on a partial
+plan.
+
+\item Constraints to describe relationships among tokens. This provides an
+expressive method of describing interactions among tokens in the
+context of Temporal Planning.
+
+\item Timelines as a concise abstraction to express the evolution of state
+and behavior for a system component. It provides semantics of mutual
+exclusion sequencing in time. Other core abstractions are available in
+\eu for handling metric resources.
+
+\item Domain rules to describe internal and external relationships on and
+between tokens respectively. Rules are applied to active tokens
+(master tokens) referred to as masters and typically produce inactive
+tokens (slaves) which are a key vehicle through which the planning
+process evolves. The rule structure in \eu differs from the more
+restrictive commitment to preconditions and effects in classical
+planning which prohibits durative actions and disembodied effects
+(i.e. effects which may not occur until some temporal distance after
+the end of the action).
+
+\end{enumerate}

### 2012-02-06 europa-basics.tex
updates

@@
 
-\eu is a versatile planning framework, let's start by looking at some examples of the main problem types that it can tackle.
+At the core of the \rx framework, is the deliberation engine, \eu,
+which has a rich legacy from NASA missions. \eu is a versatile
+Constraint-based temporal planner which continues to be deployed on a
+diverse set of applications. We motivate this section with some
+problem domains this planner can handle and then delve in substantial
+detail.
+
+% \eu is a versatile planning framework.  let's start by looking at some examples of the main problem types that it can tackle.
 
@@
 
-As the examples above show, a Planning problem (actions to achieve a goal) may embed a Scheduling problem (what resources are necessary to achieve stated goals) and both Planning and Scheduling may embed a Constraint Satisfaction problem.
-% since temporal, resource and  other
+As the examples above show, a Planning problem (actions to achieve a
+goal) may embed a Scheduling problem (what resources are necessary to
+achieve stated goals) and both Planning and Scheduling may embed a
+Constraint Satisfaction problem. % since temporal, resource and other
 % kinds of constraints must be respected in most real life problems.
@@
 finite domains. Given a defined conjunctive set of constraints on the
-variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the objective
-is to find one or more value assignments to $V$ where all constraints
-are satisfied.
-
-To solve a problem, CP techniques use logical inference to perform Constraint Propagation, which consists of two
-operations:
+variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the
+objective is to find one or more value assignments to $V$ where all
+constraints are satisfied. To solve a problem, CP techniques use
+logical inference to perform Constraint Propagation, which consists of
+two operations:
 
@@
 
-A set of constraints and variables that describe a problem domain are typically represented as a Constraint Network, where the variables are nodes , and constraints are arcs between the variables.
-Having this network representation in mind, Arc Consistency for a a pair of variables x,y is achieved if for every value in x's domain, there is a value in the y's domain that is consistent with all the constraints that connect x and y.
-A naive approach to ensuring arc consistency consists of cycling through all the variable pairs and performing constraint propagation until there are no variable domain changes, the AC-3 algorithm \comment {citation [MAC77]} improves over the naive approach by ignoring constraints whose variables were not affected since the last iteration.
-AC-3 is very popular in CP implementations.
+A set of constraints and variables that describe a problem domain are
+typically represented as a Constraint Network, where the variables are
+nodes and constraints are arcs between the variables. Having this
+network representation in mind, Arc Consistency \comment{needs a
+  definition} for a pair of variables $x,y$ is achieved if for every
+value in $x$'s domain, there is a value in $y$'s domain that is
+consistent with all the constraints that connect $x$ and $y$.  A naive
+approach to ensuring arc consistency consists of cycling through all
+the variable pairs and performing constraint propagation until there
+are no variable domain changes. The AC-3 algorithm \cite{mackworth77},
+a popular implementation, improves over the naive approach by ignoring
+constraints whose variables were not affected since the last
+iteration.
 
-Arc Consistency and other, more sophisticated consistency algorithms \comment {citation CP Handbook} can be used to efficiently prove that a CP problem is unsatisfiable, however, they generally can't prove that a CP problem is satisfiable and provide specific variable assignments that constitute solutions.
-For that, search algorithms like global backtracking or local search are used \comment {citation CP Handbook}, which use consistency as an efficient way to prune the search space.
+Arc Consistency and other sophisticated consistency algorithms can be
+used to efficiently prove that a CP problem is unsatisfiable. However,
+they generally cannot prove that a CP problem is satisfiable and
+provide specific variable assignments that constitute solutions. For
+that, search algorithms like global backtracking \comment{citation
+  needed} or local search are used; these use consistency as an
+efficient way to prune the search space \comment {citation CP
+  Handbook}. In theory, solving a CP problem is NP-Hard
+\cite{ghallab04}, but often very efficient in practice using a number
+of consistency and search algorithms that are well understood.
 
-In theory, solving a CP problem is NP-Hard \cite{ghallab04}, but often very efficient in practice using a number of consistency and search algorithms that are well understood.
-
-CP is usually implemented as part of a programming language and constraints are usually represented as objects \cite{puget95}.
-Any constraint can be introduced and, as long as the constraint propagation protocols specified by the hostCP system are enforced, it will be indistinguishable from any other ``primitive'' CP constraint, such as $\leq$ or $\geq$.
+CP is usually implemented as part of a programming language and
+constraints are usually represented as objects \cite{puget95}. Any
+constraint can be introduced and, as long as the constraint
+propagation protocols specified by the hostCP system are enforced, it
+will be indistinguishable from any other ``primitive'' CP constraint,
+such as $\leq$ or $\geq$.
 
@@
 close that gap. On one hand, it takes advantage of a CSP
-representation; one the other it also uses attributes and intervals to
+representation; on the other it uses attributes and intervals to
 maintain an explicit representation of the elements relevant to
@@
 
-\textit{Variables} Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+\textit{Variables} Values that need to be represented to describe the
+problem domain and over which we may want to specify constraints.
 
-\textit{Objects} The things we wish to describe and refer to in a domain are considered
-Objects. As in the case with object-oriented analysis and design, one
-can seek out the nouns in any domain description to find likely
-objects. In a Mars rover example, we might consider the satellite, the
-camera, and the attitude controller (which orients the robot for
-taking pictures) to be objects. Objects have state and behavior. For
-example, a camera can be:
+\textit{Objects} The things we wish to describe and refer to in a
+domain are considered Objects. As in the case with object-oriented
+analysis and design, one can seek out the nouns in any domain
+description to find likely objects. In a Mars rover example, we might
+consider the satellite, the camera, and the attitude controller (which
+orients the robot for taking pictures) to be objects. Objects have
+state and behavior. For example, a camera can be:
 
@@
 
-TODO: talk about actions
+\comment{talk about actions}
 
@@
 
-As we have noted, Tokens describe some aspect of an Object in time. Objects thus may have many Tokens in a plan in order to describe
-their state and behavior throughout all points in time of the plan. Within this general framework, we note a few particulars:
+As we have noted, Tokens describe some aspect of an Object in
+time. Objects thus may have many Tokens in a plan in order to describe
+their state and behavior throughout all points in time of the
+plan. Within this general framework, we note a few particulars:
 
@@
 
-For example, consider the tire-world domain. The tire was located in
-the trunk with the predicate tireLocated(Trunk) and located on the
-ground with the predicate tireLocated(Ground). Clearly, the same tire
-cannot be in both places at once. A Timeline provides a simple method
-for aggregating the statements about a tire such that they are
-mutually exclusive.
+\comment{This is the first instance that this domain is being
+  introduced. Either we need more explanation or it ought to be
+  removed.} For example, consider the tire-world domain. The tire was
+located in the trunk with the predicate tireLocated(Trunk) and located
+on the ground with the predicate tireLocated(Ground). Clearly, the
+same tire cannot be in both places at once. A Timeline provides a
+simple method for aggregating the statements about a tire such that
+they are mutually exclusive.
 
@@
 
-    On creation of a token, where the commitment has not been made yet
-    to activate or merge the token, the token is said to be
-    inactive. There is a nuance to the state model for tokens which
-    relate to the mode of its creation. If the token is allocated
-    explicitly by an external actor, rather than internally through
-    rule firing, it may include the state rejected indicating the
-    planner is permitted to reject the token. A valid plan can include
-    rejected tokens. This typically arises where the token represents
-    a goal that is preferable to achieve but not mandatory. This state
-    is not reachable for slaves since that would imply selective
-    adherence to the domain model. Figure 4 presents the state
-    transition diagram for token states relevant for planning. The
+On creation of a token, where the commitment has not been made yet to
+activate or merge the token, the token is said to be inactive. There
+is a nuance to the state model for tokens which relate to the mode of
+its creation. If the token is allocated explicitly by an external
+actor, rather than internally through rule firing, it may include the
+state rejected indicating the planner is permitted to reject the
+token. A valid plan can include rejected tokens. This typically arises
+where the token represents a goal that is preferable to achieve but
+not mandatory. This state is not reachable for slaves since that would
+imply selective adherence to the domain model. Figure 4 presents the
+state transition diagram for token states relevant for planning. The
 transitions are operations on a partial plan which can decide an
 outcome for an inactive token. As operations which may arise in
-    search, they must be reversible during backtracking. The
-    cancellation operations in each case are also shown.
+search, they must be reversible during backtracking. The cancellation
+operations in each case are also shown.
 

### 2012-02-06 europa-basics.tex
cpl of changes to CP section.

@@
 nodes and constraints are arcs between the variables. Having this
-network representation in mind, Arc Consistency \comment{needs a
-  definition} for a pair of variables $x,y$ is achieved if for every
+network representation in mind, a pair of variables $x,y$ is said to be arc consistent if for every
 value in $x$'s domain, there is a value in $y$'s domain that is
@@
 
-Arc Consistency and other sophisticated consistency algorithms can be
+Arc Consistency and other, more stringent consistency definitions can be
 used to efficiently prove that a CP problem is unsatisfiable. However,
@@
 constraints are usually represented as objects \cite{puget95}. Any
-constraint can be introduced and, as long as the constraint
-propagation protocols specified by the hostCP system are enforced, it
-will be indistinguishable from any other ``primitive'' CP constraint,
+constraint that the user is able to reason about can be introduced into the system and, as long as the constraint
+propagation protocols specified by the host CP system are enforced, it
+will be indistinguishable from any other "primitive'' CP constraint,
 such as $\leq$ or $\geq$.

### 2012-02-06 trunk/europa-basics.tex
re-working EUROPA's plan representation. checkpoint before lunch.

@@
 
-\eu is a versatile planning framework which can be used in a range of
-problem-solving methods.
+At the core of the \rx framework, is the deliberation engine, \eu,
+which has a rich legacy from NASA missions. \eu is a versatile
+Constraint-based temporal planner which continues to be deployed on a
+diverse set of applications. We motivate this section with some
+problem domains this planner can handle and then delve in substantial
+detail.
 
-\paragraph \texttt{Constraint Satisfaction}: A canonical problem in
+% \eu is a versatile planning framework.  let's start by looking at some examples of the main problem types that it can tackle.
+
+\paragraph \textit{Constraint Satisfaction}: A canonical problem in
 dealing with constraints is the $N$-Queens problem in which chess queens
@@
 
-As the examples above show, a planning problem (actions to achieve a
-goal) may embed a scheduling problem (what resources are necessary to
-achieve stated goals) and both planning and scheduling could embed a
+As the examples above show, a Planning problem (actions to achieve a
+goal) may embed a Scheduling problem (what resources are necessary to
+achieve stated goals) and both Planning and Scheduling may embed a
 Constraint Satisfaction problem. % since temporal, resource and other
@@
 finite domains. Given a defined conjunctive set of constraints on the
-variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the objective
-is to find one or more value assignments to $V$ where all constraints
-are satisfied.
-
-To solve a problem, CP techniques use logical inference to perform two
-important operations:
+variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the
+objective is to find one or more value assignments to $V$ where all
+constraints are satisfied. To solve a problem, CP techniques use
+logical inference to perform Constraint Propagation, which consists of
+two operations:
 
@@
 
-CP is normally implemented as part of a programming language and
-constraints are normally represented as objects \cite{puget95}
-\comment{citation}. Any constraint can be introduced and, as long as the
-bounds propagation and domain reduction protocols specified by the host
-CP system are enforced, it will be indistinguishable from any other
-``primitive'' CP constraint, such as $\leq$ or $\geq$. In theory,
-solving a CSP problem is NP-Hard \cite{ghallab04}, but often very
-efficient in practice using a number of algorithms and techniques that
-are well understood.
+A set of constraints and variables that describe a problem domain are
+typically represented as a Constraint Network, where the variables are
+nodes and constraints are arcs between the variables. Having this
+network representation in mind, a pair of variables $x,y$ is said to be arc consistent if for every
+value in $x$'s domain, there is a value in $y$'s domain that is
+consistent with all the constraints that connect $x$ and $y$.  A naive
+approach to ensuring arc consistency consists of cycling through all
+the variable pairs and performing constraint propagation until there
+are no variable domain changes. The AC-3 algorithm \cite{mackworth77},
+a popular implementation, improves over the naive approach by ignoring
+constraints whose variables were not affected since the last
+iteration.
 
-\comment{TODO: talk about arc consistency algorithms, computational
-  complexity and solvers.}
+Arc Consistency and other, more stringent consistency definitions can be
+used to efficiently prove that a CP problem is unsatisfiable. However,
+they generally cannot prove that a CP problem is satisfiable and
+provide specific variable assignments that constitute solutions. For
+that, search algorithms like global backtracking \comment{citation
+  needed} or local search are used; these use consistency as an
+efficient way to prune the search space \comment {citation CP
+  Handbook}. In theory, solving a CP problem is NP-Hard
+\cite{ghallab04}, but often very efficient in practice using a number
+of consistency and search algorithms that are well understood.
+
+CP is usually implemented as part of a programming language and
+constraints are usually represented as objects \cite{puget95}. Any
+constraint that the user is able to reason about can be introduced into the system and, as long as the constraint
+propagation protocols specified by the host CP system are enforced, it
+will be indistinguishable from any other "primitive'' CP constraint,
+such as $\leq$ or $\geq$.
 
@@
 \eu is an implementation of the CAIP framework. 
+\comment{once the CAIP section is finished, this section will need an iteration to provide a few pointers back to it}
 
-Variables
+\textit{Variables} Values that need to be represented to describe the
+problem domain and over which we may want to specify constraints.
+In the Shopping Agent example, the times at which the agent needs to leave or be back, or executes a purchase, are all instances where variables representing time would be used.
 
-Values that need to be represented to describe the problem domain and over which we may want to specify constraints.
+\textit{Objects} The things we wish to describe and refer to in a
+domain are considered Objects. As in the case with object-oriented
+analysis and design, one can seek out the nouns in any domain
+description to find likely objects.
+In the Shopping Agent example, we night consider the Agent itself, Products and Locations all to be objects.
+Objects have state and behavior.
+For example, a Shopping Agent can have:
+\begin{enumerate}
+	\item State: it's location, a bag that contains the products it has already purchased (the bag can in turn be another object), etc.
+	\item Behavior: go to a location to look for a product, perform a purchase, return home, etc. 
+\end{enumerate}
 
-Objects
+As in object-oriented design, objects that have similar state and behavior can be described generically in terms of object types or classes.
+\eu allows the definition of object classes in the same way that it is done in popular object-oriented programming languages.
+However, to describe state and behavior for the purposes of planning we need a different mechanism, for that we build on the formalism of first order logic as explained below.
 
-The things we wish to describe and refer to in a domain are considered
-Objects. As in the case with object-oriented analysis and design, one
-can seek out the nouns in any domain description to find likely
-objects. In a Mars rover example, we might consider the satellite, the
-camera, and the attitude controller (which orients the robot for
-taking pictures) to be objects. Objects have state and behavior. For
-example, a camera can be:
+\textit{Tokens}
 
-    off,
-    ready, or
-    taking a picture.
+In first order logic, a predicate defines a relation between objects and properties \comment{citation?}. 
+In \eu, we define such relations between variables whose domains are sets of
+objects and sets of properties to describe state and behavior. For example, we might use a predicate
+\texttt{At($a$,$l$)} to indicate that agent
+$a$ is at location $l$, or a predicate \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to buy product $p$. Note that $a$ is a variable which may
+have a number of possible values in a problem with multiple
+agents. Similarly, $l$ and $p$ are variables whose values are the set of
+possible positions and products respectively.
+\eu can be used to create partial plans, where the domains of variables $a$ and $l$ can have more than one possible value in them, or grounded plans, where single values will be
+specified for each variable as we saw in the case of the N-queen's problem.
 
-An attitude controller can be:
-
-    pointing at a position, or
-    turning from one position to another.
-
-    An object is an instance of a class. In \eu we model using the
-    abstraction of a class to speak about all instances having certain
-    properties of state and behavior. In order to describe such state
-    and behavior we build on the formalism of first order logic.
-
-Temporally Qualified Predicates and Actions
-
-A predicate defines a relation between objects and properties. In \eu,
-we define such relations between variables whose domains are sets of
-objects and sets of properties. For example, we might use a predicate
-\texttt{Pointing($a,p$)} to indicate that a rover's attitude controller
-$a$ is pointing at position $p$. Note that $a$ is a variable which may
-have a number of possible values in a problem with multiple
-satellites. Similarly, $p$ is a variable whose values are the set of
-possible positions. In a grounded plan, single values will be
-specified for each variable as we saw in the case of a solved
-Constraint Satisfaction Problem.
-
-In general, it is not sufficient to state that a predicate is true
-without giving it some temporal extent over which it holds. Predicates
+In general, to come up with an executable plan it is not sufficient to state predicates that describe required state or behavior
+without also specifying some temporal extent over which each of those predicates hold. Predicates
 that are always true can be thought to hold from the beginning to the
@@
 must be defined with timepoints to represent its start and end. So we
-might want to write \texttt{Pointing($a,s,e,p$)} to indicate that the
-attitude controller a is pointing at position $p$ from time $s$ to
+might want to write \texttt{At($a$,$s$,$e$,$l$)} to indicate that the
+agent $a$ is at location $p$ from time $s$ to
 time $e$. In fact, this pattern of using such predicates to describe
@@
 applies and the timepoints over which it holds. In \eu, all predicate
-instances are Tokens.
+instances are Tokens. Also, in the same way that objects are described generically by classes, in \eu Tokens can be described generically by Token Types.
 
-Tokens
-
-A Token is an instance of a predicate or an action and is defined over
+A Token is an instance of a predicate that represents and object's state or behavior and is defined over
 a temporal extent. Every token has five built-in variables:
 
-start: The beginning of the temporal extent over which the predicate
-is defined.  end: The end of the temporal extent over which the
-predicate is defined.  duration: The constraints start + duration =
-end is enforced automatically.  object: made. The set of objects to
+\begin{enumerate}
+    \item \textit {start}: The beginning of the temporal extent over which the predicate
+is defined.  
+    \item \textit {end}: The end of the temporal extent over which the
+predicate is defined.  
+    \item \textit {duration}: The constraint \textit{start} + \textit{duration} =
+\textit{end} is enforced automatically.  
+    \item \textit{object}: The set of objects to
 which a token might apply. In a grounded plan each Token applies to a
@@
 may not yet have been made.
+    \item \textit{state}: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
+variable captures the token's current state and its reachable states
+through further restriction.  This is \eu's mechanism to support CAIP's planning approach, see the Token State Model discussion
+below for details.
+\end{enumerate}
 
-state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
-variable captures the token's current state and its reachable states
-through further restriction. See the Token State Model discussion
-below for details.
+\comment {resume cleanup here}
 
-Objects (continued)
+\textit{Objects (continued)}
 
@@
 example, when a slave is created via a rule, the REJECTED value is
-removed from its state variable.  Summary
+removed from its state variable.
 
-This page has presented the main elements of the \eu planning
-approach. They are:
+\textit{Summary}
 
-Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+This has presented the main elements of the EUROPA planning approach. They are:
+
+\begin{enumerate}
+
+\item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
 states and actions in time. Note that tokens do not discriminate
@@
 
-Constraints to describe relationships among tokens. This provides an
+\item Token States which are the basis of planning operations on a partial
+plan.
+
+\item Constraints to describe relationships among tokens. This provides an
 expressive method of describing interactions among tokens in the
@@
 
-Timelines as a concise abstraction to express the evolution of state
+\item Timelines as a concise abstraction to express the evolution of state
 and behavior for a system component. It provides semantics of mutual
@@
 
-Domain rules to describe internal and external relationships on and
+\item Domain rules to describe internal and external relationships on and
 between tokens respectively. Rules are applied to active tokens
@@
 
-Token States which are the basis of planning operations on a partial
-plan.
+\end{enumerate}

### 2012-02-06 trunk/concepts.tex
updates

@@
 
-Our system, \texttt{T-REX} (or the Teleo-Reactive EXecutive) is an
-adaptive, artificial intelligence based controller and provides
-general framework for building reasoning systems for real-world
-autonomous vehicles. At MBARI \texttt{T-REX} is used for AUV control;
-another instantiation of the system is being used for control of a
-terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The development of
-\texttt{T-REX} has been targeted at surveying a number of
-oceanographic features which are dynamic and unpredictable
-spatio-temporally. Typically this requires our AUVs to balance the
-goals of coverage spatially while opportunistically following features
-of scientific interest and to do so while being operationally aware of
-its own limitations in terms of resources (typically the battery state
-of charge) and overall proximity to other observational assets for
-obtaining scientific ground-truth. We are therefore interested in
-representational frameworks that allow robots to pursue long-term
-objectives while choosing short-term gain and can gracefully deal with
-exogenous or endogenous change.
+\texttt{T-REX} is an adaptive, artificial intelligence based
+controller and provides general framework for building reasoning
+systems for real-world autonomous vehicles. At MBARI \texttt{T-REX} is
+used for AUV control; another instantiation of the system is being
+used for control of a terrestrial personal robot \cite{pr2,
+  Meeussen:2010dn}. The development of \texttt{T-REX} has been
+targeted at surveying a number of oceanographic features which are
+dynamic and unpredictable spatio-temporally. Typically this requires
+our AUVs to balance the goals of coverage spatially while
+opportunistically following features of scientific interest and to do
+so while being operationally aware of its own limitations in terms of
+resources (typically the battery state of charge) and overall
+proximity to other observational assets for obtaining scientific
+ground-truth. We are therefore interested in representational
+frameworks that allow robots to pursue long-term objectives while
+choosing short-term gain and can gracefully deal with exogenous or
+endogenous change.
 
@@
 to be able to synthesize plans insitu and to re-plan. We use a
-temporal constraint-based planner EUROPA$_2$ with a demonstrated
-legacy of having flown on NASA space missions
-\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
-innovations for AUV adaptation: the use of flexible plan
-representations, compositional control with the use of partitioned
-networks and off-line learning to inform insitu environmental state
-estimation. In this section we detail some of the key concepts
-relevant to flexible plan representation. Section \ref{sec:arch} has
-details the \texttt{T-REX} partitioned architecture and estimation
-driven vehicle adaptation.
+temporal constraint-based planner \eu with a demonstrated legacy of
+having flown on NASA space missions \cite{jonsson00,bresina05,
+  barreiro09}. Our autonomy architecture brings three key innovations
+for AUV adaptation: the use of flexible plan representations,
+compositional control with the use of partitioned networks and
+off-line learning to inform insitu environmental state estimation. In
+this section we detail some of the key concepts relevant to flexible
+plan representation. Section \ref{sec:arch} has details the
+\rx partitioned architecture and estimation driven vehicle
+adaptation.
 
-% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
-% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
-% for the Mars Exploration Rovers \cite{bresina05}.
+\subsection{The basics}
+\label{sec:basics}
 
-EUROPA$_2$ uses a domain model written in a declarative language
+\eu uses a \emph{domain model} written in a declarative language
 (NDDL: New Domain Description Language), together with initial
@@
 propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
-a set of conditions that should be true in the future, where some of
-these conditions will correspond to actions the agent must take. The
-planner can backtrack and try another path during search if a goal
-cannot be reached while being capable of discarding unachievable
-goals.
+\cite{dechter91} and applying goal constraints, \eu can select a set
+of conditions that should be true in the future, where some of these
+conditions will correspond to actions the agent must take. The planner
+can backtrack and try another path during search if a goal cannot be
+reached while being capable of discarding unachievable goals.
 
@@
 \end{figure}
-
-EUROPA$_2$ uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a EUROPA$_2$ domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
 
@@
 
+\eu uses a \emph{state variable} representation to describe the
+evolution of state over time. The instantiated history of such state
+variable evolution over a temporal horizon we call \emph{timelines}
+and which represent a single thread in the execution of a concurrent
+system. At any given time each thread can execute a single procedure.
+Thus each timeline consists of a sequence of procedures which
+encapsulate and describe state evolution; we call these instantiated
+atomic entities \emph{tokens}.  A token therefore describes a
+procedure invocation, the state variables on which it can occur, the
+parameter values of the procedure, and the time values defining the
+interval. We allow encapsulation of uncertainty within these tokens
+with a range of start and end times and parameters, all of which are
+encoded as variables. A constraint solver in turn manipulates these
+variables defined in a \eu domain model. For example, consider
+a scientific need to take a water samples 100 meters from a hotspot
+while an AUV is performing a Yo-Yo in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
 \begin{figure}
@@
 \end{figure}
+
+\subsubsection{\eu Plan Representation}
+\label{sec:europa:pr}
+
+This section describes the fundamental entities within \eu which are
+typical to other CAIP-based formalisms.
+
+\begin{description}
+
+\item[\textbf{Variables}] Values that need to be represented to
+  describe the problem domain and over which we may want to specify
+  constraints. In the Shopping Agent example, the times at which the
+  agent needs to leave or be back, or executes a purchase, are all
+  instances where variables representing time would be used.
+
+\item[\textbf{Objects}] The items we wish to describe and refer to in
+  a domain are considered Objects. As in the case with object-oriented
+  analysis and design, one can seek out the nouns in any domain
+  description to find likely objects. In the Shopping Agent example,
+  we night consider the Agent itself, Products and Locations all to be
+  objects. Objects have state and behavior. For example, a Shopping
+  Agent can have:
+
+  \begin{itemize}
+
+  \item State: it's location, a bag that contains the products it has
+    already purchased (the bag can in turn be another object), etc.
+
+  \item Behavior: go to a location to look for a product, perform a
+    purchase, return home, etc.
+
+  \end{itemize}
+
+  Similarly objects that have similar state and behavior can be
+  described generically in terms of object types or classes. \eu
+  allows the definition of object classes in the same way that it is
+  done in popular object-oriented programming languages. However, to
+  describe state and behavior for the purposes of planning we need to
+  build on the formalism of first order logic as explained below.
+
+\item[\textbf{Tokens}] In first order logic, a predicate defines a
+  relation between objects and properties \comment{citation?}.  In
+  \eu, we define such relations between variables whose domains are
+  sets of objects and sets of properties to describe state and
+  behavior. For example, we might use a predicate \texttt{At($a$,$l$)}
+  to indicate that agent $a$ is at location $l$, or a predicate
+  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
+  buy product $p$. Note that $a$ is a variable which may have a number
+  of possible values in a problem with multiple agents. Similarly, $l$
+  and $p$ are variables whose values are the set of possible positions
+  and products respectively. \eu can be used to create partial plans,
+  where the domains of variables $a$ and $l$ can have more than one
+  possible value in them, or grounded plans, where single values will
+  be specified for each variable as we saw in the case of the
+  N-queen's problem.
+
+  In general, to come up with an executable plan it is not sufficient
+  to state predicates that describe required state or behavior without
+  also specifying some temporal extent over which each of those
+  predicates hold. Predicates that are always true can be thought to
+  hold from the beginning to the end of time. However, in practice,
+  the temporal extent of interest must be defined with timepoints to
+  represent its start and end. So we might want to write
+  \texttt{At($a$,$s$,$e$,$l$)} to indicate that the agent $a$ is at
+  location $p$ from time $s$ to time $e$. In fact, this pattern of
+  using such predicates to describe both state and behavior of objects
+  is so prevalent in \eu that we have introduced a special construct
+  called a Token which has the built in variables to indicate the
+  object to which the statement principally applies and the timepoints
+  over which it holds. A Token is an instance of a predicate that
+  represents and object's state or behavior and is defined over a
+  temporal extent. In \eu, all predicate instances are Tokens. Also,
+  in the same way that objects are described generically by classes,
+  in \eu Tokens can be described generically by Token Types. Every
+  token has five built-in variables:
+
+  \begin{enumerate}
+
+  \item \textit {start}: The beginning of the temporal extent over
+    which the predicate is defined.
+
+  \item \textit {end}: The end of the temporal extent over which the
+    predicate is defined.
+
+  \item \textit {duration}: The constraint \textit{start} $+$
+    \textit{duration} = \textit{end} is enforced automatically.
+
+  \item \textit{object}: The set of objects to which a token might
+    apply. In a grounded plan each Token applies to a specific Object,
+    reflecting the intuition that we are using Tokens to describe some
+    aspect of an Object (i.e. its state or behavior) in time. However,
+    in a partial plan, the commitment to a specific object may not yet
+    have been made.
+
+  \item \textit{state}: Tokens can be \texttt{ACTIVE, INACTIVE,
+      MERGED}, or \texttt{REJECTED}. The state variable captures the
+    token's current state and its reachable states through further
+    restriction.  This is \eus mechanism to support the CAIP approach.
+
+\end{enumerate}
+
+\comment {resume cleanup here}
+
+% \textit{Objects (continued)}
+
+% As we have noted, Tokens describe some aspect of an Object in
+% time. Objects thus may have many Tokens in a plan in order to describe
+% their state and behavior throughout all points in time of the
+% plan. Within this general framework, we note a few particulars:
+
+% Static Facts: Classes without predicates lead to Objects without
+% Tokens. This arises where an objects state or behavior is independent
+% of time. For example, a domain may have a set of locations and/or
+% paths for which there is nothing more to say than that they exist.
+
+% Timelines: Often objects in a domain must be described by exactly one
+% token for every given timepoint in the plan. Such objects are so
+% common that we provide a special construct in \eu to extend these
+% semantics to derived classes. Any instances of a class derived from a
+% Timeline will induce ordering requirements among its tokens in order
+% to ensure no temporal overlap may occur among them. See below for
+% further discussion.
+
+% Resources: Metric resources, e.g. the energy of a battery or the
+% capacity of a cargo hold, are objects with an explicit quantitative
+% state in time and with a circumscribed range of changes that can occur
+% to impact that state i.e. produce, consume, use, change. These changes
+% are captured as tokens. Resources are such a common requirement for
+% \eu users that special constructs are also provided for
+% them. Instances of classes derived from a Resource will induce
+% ordering requirements on their Tokens in order to ensure that the
+% level of the resource remains within specified limits.
+
+\item[\textbf{Timelines}] It may be sufficient to maintain only a
+  partial-order among tokens. However, it is often the case that
+  tokens represent states and actions of a single object in the
+  system. Such tokens are typically mutually-exclusive. \eu uses a
+  Timeline structure \cite{mus94} to concisely capture system
+  components whose behavior is described over time in this manner.
+
+\end{description}
+
+For example, consider the tire-world domain. The tire was located in
+the trunk with the predicate tireLocated(Trunk) and located on the
+ground with the predicate tireLocated(Ground). Clearly, the same tire
+cannot be in both places at once. A Timeline provides a simple method
+for aggregating the statements about a tire such that they are
+mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the
+whereabouts of a tire. The tire is an object in the tire-world
+represented as a timeline. It has predicates associated with it which
+can describe its states and actions over time. The predicate names
+need not contain the tire prefix; this is now implicit since the
+tokens are assigned to a specific tire instance (i.e. an instance of a
+Timeline). While Timelines are a useful element of the \eu planning
+paradigm, they are not essential. A more general notion of a system
+Object can be used which does not impose restrictions of
+mutual-exclusion or non-zero duration. This can be important in
+supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3
+contiguous tokens. The end and start time-points are related by an
+equality constraint. A Moving predicate has been introduced to cover
+the transition from one location to another. It takes 2 location
+arguments. Notice that precedence relationships exist between tokens
+such that they cannot overlap but the start and end times may remain
+flexible. To illustrate this, sample times are included. Assume a
+total time range of interest between 0 and 1000. In addition, tokens
+on Timelines have a minimum duration of 1. As a result the values
+shown are the most flexible possible values for the start and end of
+each token. There are a number of advantages of allowing this
+flexibility. First, the basic structure of the plan can be developed
+without over-committing to specific times. If it is not necessary for
+the tire to be on the ground at time-step 4, then a planner should not
+be forced to specify it. Such an approach permits a least-commitment
+approach to planning. Second, in many domains it is simply impossible
+to know in planning exactly how long an activity might take. For
+example, a common activity of driving a car from one point to another
+on a road can take varying amounts of time depending on traffic and
+traffic lights. In such circumstances it is more practical to express
+upper and lower bounds on durations which naturally lead to intervals
+for start and end times. In such domains, flexibility in planning aids
+robustness in execution.  Rules
+
+In order for a plan to be valid, it must comply with all rules and
+regulations pertinent to the application domain in question. Rules
+govern the internal and external relationships of a token. For
+example, consider a parameterized predicate describing a transition
+from one location to another. Let the parameters be from and to. The
+parameters are instantiated on a token as variables whose domain of
+values is the set of all locations in a given problem. A rule
+governing an internal relationship among token variables might
+stipulate that a transition must involve a change in location. This
+can be easily expressed as a constraint on the definition of a
+predicate of the form: from != to. It is reasonable to further
+stipulate that one must be Located somewhere before a transition can
+occur, and one must end up Located somewhere when completed. This is
+an example of a rule governing an external relationship among
+tokens. It specifies a requirement that tokens of the predicate
+Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving
+Figure 2 illustrates the entities and relations involved in specifying
+such a rule on a Moving predicate. The token on which the rule applies
+is referred to as the master. Each Located token required by the
+master is referred to as a slave. All variables are indicated by name
+and their domains are expressed as intervals in the case of temporal
+variables and as enumerations for the remainder. Application of a rule
+on a token can thus cause slave tokens, variables, and constraints to
+be introduced.  Token State Model
+
+The capability of a domain rule to cause a slave token to be created
+is a key vehicle through which planning occurs. Semantically, this
+rule imposes a requirement for supporting tokens to be in the plan in
+order for the master to be valid. There are 2 possibilities to
+consider:
+
+The slave is inserted as an active token in the plan. As such, rules
+may be activated on the slave, and it may consume resources.
+
+The token is merged with a matching token already in the plan. Once a
+slave is merged, the requirement it represents are considered
+satisfied. The process of merging passes on all restrictions imposed
+on the slave to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. For an active token and an inactive token to be
+compatible requires that they are instances of the same predicate and
+that no intersections between corresponding variables are empty. The
+effect of merging is illustrated in Figure 3.
+
+Figure 3: Merging an inactive token on an active token. Domain
+restrictions occur in highlighted variables of active token.
+
+On creation of a token, where the commitment has not been made yet to
+activate or merge the token, the token is said to be inactive. There
+is a nuance to the state model for tokens which relate to the mode of
+its creation. If the token is allocated explicitly by an external
+actor, rather than internally through rule firing, it may include the
+state rejected indicating the planner is permitted to reject the
+token. A valid plan can include rejected tokens. This typically arises
+where the token represents a goal that is preferable to achieve but
+not mandatory. This state is not reachable for slaves since that would
+imply selective adherence to the domain model. Figure 4 presents the
+state transition diagram for token states relevant for planning. The
+transitions are operations on a partial plan which can decide an
+outcome for an inactive token. As operations which may arise in
+search, they must be reversible during backtracking. The cancellation
+operations in each case are also shown.
 
@@
 \end{minipage}
+
+%*********
+
+Timelines
+
+It may be sufficient to maintain only a partial-order among
+tokens. However, it is often the case that tokens represent states and
+actions of a single object in the system. Such tokens are typically
+mutually-exclusive. \eu uses a Timeline structure developed in (??)
+to concisely capture system components whose behavior is described
+over time in this manner.
+
+\comment{This is the first instance that this domain is being
+  introduced. Either we need more explanation or it ought to be
+  removed.} For example, consider the tire-world domain. The tire was
+located in the trunk with the predicate tireLocated(Trunk) and located
+on the ground with the predicate tireLocated(Ground). Clearly, the
+same tire cannot be in both places at once. A Timeline provides a
+simple method for aggregating the statements about a tire such that
+they are mutually exclusive.
+
+Figure 1: A Timeline for a Tire
+
+Figure 1 illustrates how a Timeline can be used to describe the
+whereabouts of a tire. The tire is an object in the tire-world
+represented as a timeline. It has predicates associated with it which
+can describe its states and actions over time. The predicate names
+need not contain the tire prefix; this is now implicit since the
+tokens are assigned to a specific tire instance (i.e. an instance of a
+Timeline). While Timelines are a useful element of the \eu planning
+paradigm, they are not essential. A more general notion of a system
+Object can be used which does not impose restrictions of
+mutual-exclusion or non-zero duration. This can be important in
+supporting partial-order planning.
+
+For this example, the state of the tire is specified using 3
+contiguous tokens. The end and start time-points are related by an
+equality constraint. A Moving predicate has been introduced to cover
+the transition from one location to another. It takes 2 location
+arguments. Notice that precedence relationships exist between tokens
+such that they cannot overlap but the start and end times may remain
+flexible. To illustrate this, sample times are included. Assume a
+total time range of interest between 0 and 1000. In addition, tokens
+on Timelines have a minimum duration of 1. As a result the values
+shown are the most flexible possible values for the start and end of
+each token. There are a number of advantages of allowing this
+flexibility. First, the basic structure of the plan can be developed
+without over-committing to specific times. If it is not necessary for
+the tire to be on the ground at time-step 4, then a planner should not
+be forced to specify it. Such an approach permits a least-commitment
+approach to planning. Second, in many domains it is simply impossible
+to know in planning exactly how long an activity might take. For
+example, a common activity of driving a car from one point to another
+on a road can take varying amounts of time depending on traffic and
+traffic lights. In such circumstances it is more practical to express
+upper and lower bounds on durations which naturally lead to intervals
+for start and end times. In such domains, flexibility in planning aids
+robustness in execution.  Rules
+
+In order for a plan to be valid, it must comply with all rules and
+regulations pertinent to the application domain in question. Rules
+govern the internal and external relationships of a token. For
+example, consider a parameterized predicate describing a transition
+from one location to another. Let the parameters be from and to. The
+parameters are instantiated on a token as variables whose domain of
+values is the set of all locations in a given problem. A rule
+governing an internal relationship among token variables might
+stipulate that a transition must involve a change in location. This
+can be easily expressed as a constraint on the definition of a
+predicate of the form: from != to. It is reasonable to further
+stipulate that one must be Located somewhere before a transition can
+occur, and one must end up Located somewhere when completed. This is
+an example of a rule governing an external relationship among
+tokens. It specifies a requirement that tokens of the predicate
+Located precede and succeed tokens of the predicate Moving.
+
+Figure 2: Internal and external relationships from rules on Moving
+Figure 2 illustrates the entities and relations involved in specifying
+such a rule on a Moving predicate. The token on which the rule applies
+is referred to as the master. Each Located token required by the
+master is referred to as a slave. All variables are indicated by name
+and their domains are expressed as intervals in the case of temporal
+variables and as enumerations for the remainder. Application of a rule
+on a token can thus cause slave tokens, variables, and constraints to
+be introduced.  Token State Model
+
+The capability of a domain rule to cause a slave token to be created
+is a key vehicle through which planning occurs. Semantically, this
+rule imposes a requirement for supporting tokens to be in the plan in
+order for the master to be valid. There are 2 possibilities to
+consider:
+
+The slave is inserted as an active token in the plan. As such, rules
+may be activated on the slave, and it may consume resources.
+
+The token is merged with a matching token already in the plan. Once a
+slave is merged, the requirement it represents are considered
+satisfied. The process of merging passes on all restrictions imposed
+on the slave to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. For an active token and an inactive token to be
+compatible requires that they are instances of the same predicate and
+that no intersections between corresponding variables are empty. The
+effect of merging is illustrated in Figure 3.
+
+Figure 3: Merging an inactive token on an active token. Domain
+restrictions occur in highlighted variables of active token.
+
+On creation of a token, where the commitment has not been made yet to
+activate or merge the token, the token is said to be inactive. There
+is a nuance to the state model for tokens which relate to the mode of
+its creation. If the token is allocated explicitly by an external
+actor, rather than internally through rule firing, it may include the
+state rejected indicating the planner is permitted to reject the
+token. A valid plan can include rejected tokens. This typically arises
+where the token represents a goal that is preferable to achieve but
+not mandatory. This state is not reachable for slaves since that would
+imply selective adherence to the domain model. Figure 4 presents the
+state transition diagram for token states relevant for planning. The
+transitions are operations on a partial plan which can decide an
+outcome for an inactive token. As operations which may arise in
+search, they must be reversible during backtracking. The cancellation
+operations in each case are also shown.
+
+Figure 4: Token States and Transitions for Planning
+
+The state of a token is embodied with a 5th and final built-in
+variable referred to as the state variable. The planner states are
+values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
+INACTIVE state is captured by the variable being unbound. If a value
+is removed from the domain, that state will not be reachable. For
+example, when a slave is created via a rule, the REJECTED value is
+removed from its state variable.
+
+% \textit{Summary}
+
+% This has presented the main elements of the EUROPA planning approach. They are:
+
+% \begin{enumerate}
+
+% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+% states and actions in time. Note that tokens do not discriminate
+% between state and action.
+
+% \item Token States which are the basis of planning operations on a partial
+% plan.
+
+% \item Constraints to describe relationships among tokens. This provides an
+% expressive method of describing interactions among tokens in the
+% context of Temporal Planning.
+
+% \item Timelines as a concise abstraction to express the evolution of state
+% and behavior for a system component. It provides semantics of mutual
+% exclusion sequencing in time. Other core abstractions are available in
+% \eu for handling metric resources.
+
+% \item Domain rules to describe internal and external relationships on and
+% between tokens respectively. Rules are applied to active tokens
+% (master tokens) referred to as masters and typically produce inactive
+% tokens (slaves) which are a key vehicle through which the planning
+% process evolves. The rule structure in \eu differs from the more
+% restrictive commitment to preconditions and effects in classical
+% planning which prohibits durative actions and disembodied effects
+% (i.e. effects which may not occur until some temporal distance after
+% the end of the action).
+
+% \end{enumerate}
 

### 2012-02-06 trunk/europa-basics.tex
updates

@@
 
-% \eu is a general purpose AI planning toolkit developed at NASA Ames.  
-% TODO: Talk about NASA missions where \eu has been used.
-
 At the core of the \rx framework, is the deliberation engine, \eu,
@@
 detail.
-
-% \eu is a versatile planning framework.  let's start by looking at some examples of the main problem types that it can tackle.
 
@@
 nodes and constraints are arcs between the variables. Having this
-network representation in mind, a pair of variables $x,y$ is said to be arc consistent if for every
-value in $x$'s domain, there is a value in $y$'s domain that is
-consistent with all the constraints that connect $x$ and $y$.  A naive
-approach to ensuring arc consistency consists of cycling through all
-the variable pairs and performing constraint propagation until there
-are no variable domain changes. The AC-3 algorithm \cite{mackworth77},
-a popular implementation, improves over the naive approach by ignoring
-constraints whose variables were not affected since the last
-iteration.
+network representation in mind, a pair of variables $x,y$ is said to
+be arc consistent if for every value in $x$'s domain, there is a value
+in $y$'s domain that is consistent with all the constraints that
+connect $x$ and $y$.  A naive approach to ensuring arc consistency
+consists of cycling through all the variable pairs and performing
+constraint propagation until there are no variable domain changes. The
+AC-3 algorithm \cite{mackworth77}, a popular implementation, improves
+over the naive approach by ignoring constraints whose variables were
+not affected since the last iteration.
 
-Arc Consistency and other, more stringent consistency definitions can be
-used to efficiently prove that a CP problem is unsatisfiable. However,
-they generally cannot prove that a CP problem is satisfiable and
-provide specific variable assignments that constitute solutions. For
-that, search algorithms like global backtracking \comment{citation
-  needed} or local search are used; these use consistency as an
-efficient way to prune the search space \comment {citation CP
-  Handbook}. In theory, solving a CP problem is NP-Hard
+Arc Consistency and other, more stringent consistency definitions can
+be used to efficiently prove that a CP problem is
+unsatisfiable. However, they generally cannot prove that a CP problem
+is satisfiable and provide specific variable assignments that
+constitute solutions. For that, search algorithms like global
+backtracking \comment{citation needed} or local search are used; these
+use consistency as an efficient way to prune the search space
+\cite{cp06}. In theory, solving a CP problem is NP-Hard
 \cite{ghallab04}, but often very efficient in practice using a number
@@
 constraints are usually represented as objects \cite{puget95}. Any
-constraint that the user is able to reason about can be introduced into the system and, as long as the constraint
-propagation protocols specified by the host CP system are enforced, it
-will be indistinguishable from any other "primitive'' CP constraint,
-such as $\leq$ or $\geq$.
+constraint that the user is able to reason about can be introduced
+into the system and, as long as the constraint propagation protocols
+specified by the host CP system are enforced, it will be
+indistinguishable from any other "primitive'' CP constraint, such as
+$\leq$ or $\geq$.
 
@@
 maintain an explicit representation of the elements relevant to
-planning. This makes it easier to write algorithms that search for an
-reason abut plans. The main elements of CAIP are:
+planning. \eu is an implementation of the CAIP framework.  This makes
+it easier to write algorithms that search for an reason abut
+plans. The main elements of CAIP are:
 
@@
 \comment{Explain how planning problem instances and their solutions are represented}
-
-\subsection{\eu Plan Representation}
-\label{sec:europa:pr}
-
-\eu is an implementation of the CAIP framework. 
-\comment{once the CAIP section is finished, this section will need an iteration to provide a few pointers back to it}
-
-\textit{Variables} Values that need to be represented to describe the
-problem domain and over which we may want to specify constraints.
-In the Shopping Agent example, the times at which the agent needs to leave or be back, or executes a purchase, are all instances where variables representing time would be used.
-
-\textit{Objects} The things we wish to describe and refer to in a
-domain are considered Objects. As in the case with object-oriented
-analysis and design, one can seek out the nouns in any domain
-description to find likely objects.
-In the Shopping Agent example, we night consider the Agent itself, Products and Locations all to be objects.
-Objects have state and behavior.
-For example, a Shopping Agent can have:
-\begin{enumerate}
-	\item State: it's location, a bag that contains the products it has already purchased (the bag can in turn be another object), etc.
-	\item Behavior: go to a location to look for a product, perform a purchase, return home, etc. 
-\end{enumerate}
-
-As in object-oriented design, objects that have similar state and behavior can be described generically in terms of object types or classes.
-\eu allows the definition of object classes in the same way that it is done in popular object-oriented programming languages.
-However, to describe state and behavior for the purposes of planning we need a different mechanism, for that we build on the formalism of first order logic as explained below.
-
-\textit{Tokens}
-
-In first order logic, a predicate defines a relation between objects and properties \comment{citation?}. 
-In \eu, we define such relations between variables whose domains are sets of
-objects and sets of properties to describe state and behavior. For example, we might use a predicate
-\texttt{At($a$,$l$)} to indicate that agent
-$a$ is at location $l$, or a predicate \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to buy product $p$. Note that $a$ is a variable which may
-have a number of possible values in a problem with multiple
-agents. Similarly, $l$ and $p$ are variables whose values are the set of
-possible positions and products respectively.
-\eu can be used to create partial plans, where the domains of variables $a$ and $l$ can have more than one possible value in them, or grounded plans, where single values will be
-specified for each variable as we saw in the case of the N-queen's problem.
-
-In general, to come up with an executable plan it is not sufficient to state predicates that describe required state or behavior
-without also specifying some temporal extent over which each of those predicates hold. Predicates
-that are always true can be thought to hold from the beginning to the
-end of time. However, in practice, the temporal extent of interest
-must be defined with timepoints to represent its start and end. So we
-might want to write \texttt{At($a$,$s$,$e$,$l$)} to indicate that the
-agent $a$ is at location $p$ from time $s$ to
-time $e$. In fact, this pattern of using such predicates to describe
-both state and behavior of objects is so prevalent in \eu that we have
-introduced a special construct called a Token which has the built in
-variables to indicate the object to which the statement principally
-applies and the timepoints over which it holds. In \eu, all predicate
-instances are Tokens. Also, in the same way that objects are described generically by classes, in \eu Tokens can be described generically by Token Types.
-
-A Token is an instance of a predicate that represents and object's state or behavior and is defined over
-a temporal extent. Every token has five built-in variables:
-
-\begin{enumerate}
-    \item \textit {start}: The beginning of the temporal extent over which the predicate
-is defined.  
-    \item \textit {end}: The end of the temporal extent over which the
-predicate is defined.  
-    \item \textit {duration}: The constraint \textit{start} + \textit{duration} =
-\textit{end} is enforced automatically.  
-    \item \textit{object}: The set of objects to
-which a token might apply. In a grounded plan each Token applies to a
-specific Object, reflecting the intuition that we are using Tokens to
-describe some aspect of an Object (i.e. its state or behavior) in
-time. However, in a partial plan, the commitment to a specific object
-may not yet have been made.
-    \item \textit{state}: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
-variable captures the token's current state and its reachable states
-through further restriction.  This is \eu's mechanism to support CAIP's planning approach, see the Token State Model discussion
-below for details.
-\end{enumerate}
-
-\comment {resume cleanup here}
-
-\textit{Objects (continued)}
-
-As we have noted, Tokens describe some aspect of an Object in
-time. Objects thus may have many Tokens in a plan in order to describe
-their state and behavior throughout all points in time of the
-plan. Within this general framework, we note a few particulars:
-
-Static Facts: Classes without predicates lead to Objects without
-Tokens. This arises where an objects state or behavior is independent
-of time. For example, a domain may have a set of locations and/or
-paths for which there is nothing more to say than that they exist.
-
-Timelines: Often objects in a domain must be described by exactly one
-token for every given timepoint in the plan. Such objects are so
-common that we provide a special construct in \eu to extend these
-semantics to derived classes. Any instances of a class derived from a
-Timeline will induce ordering requirements among its tokens in order
-to ensure no temporal overlap may occur among them. See below for
-further discussion.
-
-Resources: Metric resources, e.g. the energy of a battery or the
-capacity of a cargo hold, are objects with an explicit quantitative
-state in time and with a circumscribed range of changes that can occur
-to impact that state i.e. produce, consume, use, change. These changes
-are captured as tokens. Resources are such a common requirement for
-\eu users that special constructs are also provided for
-them. Instances of classes derived from a Resource will induce
-ordering requirements on their Tokens in order to ensure that the
-level of the resource remains within specified limits.
-
-Timelines
-
-It may be sufficient to maintain only a partial-order among
-tokens. However, it is often the case that tokens represent states and
-actions of a single object in the system. Such tokens are typically
-mutually-exclusive. \eu uses a Timeline structure developed in (??)
-to concisely capture system components whose behavior is described
-over time in this manner.
-
-\comment{This is the first instance that this domain is being
-  introduced. Either we need more explanation or it ought to be
-  removed.} For example, consider the tire-world domain. The tire was
-located in the trunk with the predicate tireLocated(Trunk) and located
-on the ground with the predicate tireLocated(Ground). Clearly, the
-same tire cannot be in both places at once. A Timeline provides a
-simple method for aggregating the statements about a tire such that
-they are mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the
-whereabouts of a tire. The tire is an object in the tire-world
-represented as a timeline. It has predicates associated with it which
-can describe its states and actions over time. The predicate names
-need not contain the tire prefix; this is now implicit since the
-tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the \eu planning
-paradigm, they are not essential. A more general notion of a system
-Object can be used which does not impose restrictions of
-mutual-exclusion or non-zero duration. This can be important in
-supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3
-contiguous tokens. The end and start time-points are related by an
-equality constraint. A Moving predicate has been introduced to cover
-the transition from one location to another. It takes 2 location
-arguments. Notice that precedence relationships exist between tokens
-such that they cannot overlap but the start and end times may remain
-flexible. To illustrate this, sample times are included. Assume a
-total time range of interest between 0 and 1000. In addition, tokens
-on Timelines have a minimum duration of 1. As a result the values
-shown are the most flexible possible values for the start and end of
-each token. There are a number of advantages of allowing this
-flexibility. First, the basic structure of the plan can be developed
-without over-committing to specific times. If it is not necessary for
-the tire to be on the ground at time-step 4, then a planner should not
-be forced to specify it. Such an approach permits a least-commitment
-approach to planning. Second, in many domains it is simply impossible
-to know in planning exactly how long an activity might take. For
-example, a common activity of driving a car from one point to another
-on a road can take varying amounts of time depending on traffic and
-traffic lights. In such circumstances it is more practical to express
-upper and lower bounds on durations which naturally lead to intervals
-for start and end times. In such domains, flexibility in planning aids
-robustness in execution.  Rules
-
-In order for a plan to be valid, it must comply with all rules and
-regulations pertinent to the application domain in question. Rules
-govern the internal and external relationships of a token. For
-example, consider a parameterized predicate describing a transition
-from one location to another. Let the parameters be from and to. The
-parameters are instantiated on a token as variables whose domain of
-values is the set of all locations in a given problem. A rule
-governing an internal relationship among token variables might
-stipulate that a transition must involve a change in location. This
-can be easily expressed as a constraint on the definition of a
-predicate of the form: from != to. It is reasonable to further
-stipulate that one must be Located somewhere before a transition can
-occur, and one must end up Located somewhere when completed. This is
-an example of a rule governing an external relationship among
-tokens. It specifies a requirement that tokens of the predicate
-Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving
-Figure 2 illustrates the entities and relations involved in specifying
-such a rule on a Moving predicate. The token on which the rule applies
-is referred to as the master. Each Located token required by the
-master is referred to as a slave. All variables are indicated by name
-and their domains are expressed as intervals in the case of temporal
-variables and as enumerations for the remainder. Application of a rule
-on a token can thus cause slave tokens, variables, and constraints to
-be introduced.  Token State Model
-
-The capability of a domain rule to cause a slave token to be created
-is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting tokens to be in the plan in
-order for the master to be valid. There are 2 possibilities to
-consider:
-
-The slave is inserted as an active token in the plan. As such, rules
-may be activated on the slave, and it may consume resources.
-
-The token is merged with a matching token already in the plan. Once a
-slave is merged, the requirement it represents are considered
-satisfied. The process of merging passes on all restrictions imposed
-on the slave to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. For an active token and an inactive token to be
-compatible requires that they are instances of the same predicate and
-that no intersections between corresponding variables are empty. The
-effect of merging is illustrated in Figure 3.
-
-Figure 3: Merging an inactive token on an active token. Domain
-restrictions occur in highlighted variables of active token.
-
-On creation of a token, where the commitment has not been made yet to
-activate or merge the token, the token is said to be inactive. There
-is a nuance to the state model for tokens which relate to the mode of
-its creation. If the token is allocated explicitly by an external
-actor, rather than internally through rule firing, it may include the
-state rejected indicating the planner is permitted to reject the
-token. A valid plan can include rejected tokens. This typically arises
-where the token represents a goal that is preferable to achieve but
-not mandatory. This state is not reachable for slaves since that would
-imply selective adherence to the domain model. Figure 4 presents the
-state transition diagram for token states relevant for planning. The
-transitions are operations on a partial plan which can decide an
-outcome for an inactive token. As operations which may arise in
-search, they must be reversible during backtracking. The cancellation
-operations in each case are also shown.
-
-Figure 4: Token States and Transitions for Planning
-
-The state of a token is embodied with a 5th and final built-in
-variable referred to as the state variable. The planner states are
-values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
-INACTIVE state is captured by the variable being unbound. If a value
-is removed from the domain, that state will not be reachable. For
-example, when a slave is created via a rule, the REJECTED value is
-removed from its state variable.
-
-\textit{Summary}
-
-This has presented the main elements of the EUROPA planning approach. They are:
-
-\begin{enumerate}
-
-\item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
-states and actions in time. Note that tokens do not discriminate
-between state and action.
-
-\item Token States which are the basis of planning operations on a partial
-plan.
-
-\item Constraints to describe relationships among tokens. This provides an
-expressive method of describing interactions among tokens in the
-context of Temporal Planning.
-
-\item Timelines as a concise abstraction to express the evolution of state
-and behavior for a system component. It provides semantics of mutual
-exclusion sequencing in time. Other core abstractions are available in
-\eu for handling metric resources.
-
-\item Domain rules to describe internal and external relationships on and
-between tokens respectively. Rules are applied to active tokens
-(master tokens) referred to as masters and typically produce inactive
-tokens (slaves) which are a key vehicle through which the planning
-process evolves. The rule structure in \eu differs from the more
-restrictive commitment to preconditions and effects in classical
-planning which prohibits durative actions and disembodied effects
-(i.e. effects which may not occur until some temporal distance after
-the end of the action).
-
-\end{enumerate}

### 2012-02-06 trunk/europa.tex
updates

@@
 \label{sec:europa:arch}
-
-The following figure displays the main architectural components in EUROPA and their relationships:
 
@@
 \includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
-\caption{\small EUROPA's Architecture}
+\caption{\small The \eu Architecture}
 \label{fig:europa-architecture}
@@
 
-\begin{enumerate}
+Fig. \ref{fig:europa-architecture} shows the main architectural
+components in \eu and their relationships.
 
-	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables.
-It also provides an efficient arc consistency mechanism (TODO: ref to AC-3).
-The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.
+The Constraint Reasoning Engine (CRE) manages variables, the domains
+from which they can take values, and constraints that define
+relationships among variables. It also provides an efficient arc
+consistency mechanism \cite{mackworth77}. The CRE is designed so that
+specialized reasoning algorithms for specific constraints can be easily
+and efficiently plugged in. The Plan Database (PDB) manages object types
+and objects, which are a mechanism to group variables to more naturally
+model the real world in much the same way that object oriented
+programming does. Also manages tokens, which are a mechanism to group
+variables to represent temporally scoped state. Objects and tokens can
+be used to model planning domains in a much more natural and extensible
+way that a pure CSP approach can. The Rules Engine (RE) manages token
+relationships in a planning model. The language modules (NDDL, ANML, and
+other higher level modeling languages) are used to create domain models
+and problem instances \comment{point advantages over pure programming
+  API approaches, provide snippets}. Finally the API Layer (Fig.
+\ref{fig:europa-api}) makes all the services available so that \eu can
+be used to build applications.
 
-	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does.
-Also manages tokens, which are a mechanism to group variables to represent temporally scoped state.
-Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can.
-
-	\item The Rules Engine (RE): manages token relationships in a planning model
-
-	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
-
-	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
+\comment{Is Fig. \ref{fig:europa-api} necessary?}
 
@@
 
-	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
-		\begin{enumerate}
-			\item Temporal Network
-			\item Resources
-			\item NDDL
-			\item ANML (TODO:maybe it's better to leave out?)
-		\end{enumerate}
-
-\end{enumerate}
+	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+	% 	\begin{enumerate}
+	% 		\item Temporal Network
+	% 		\item Resources
+	% 		\item NDDL
+	% 		\item ANML (TODO:maybe it's better to leave out?)
+	% 	\end{enumerate}
 
@@
 
-EUROPA can ingest descriptions of models, plans and goals.
-This means that a single planner can be applied to different systems and problems, simply by providing different models and goals.
-For such planners the expressiveness of the languages for models, goals and plans is of great importance.
-EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way.
-The main characteristics that make NDDL a powerful tool for describing problem domains are:
+\eu can ingest descriptions of models, plans and goals and thereby can
+be applied to different systems and problems, simply by providing
+different models and goals. Consequently the expressiveness of the
+languages for models, goals and plans is of great importance. For \eu
+NDDL (New Domain Description Language [pronounced ``noodle'']
+\comment{cite}) allows the user to specify the different components of a
+problem domain in a concise way. The main characteristics that make NDDL
+a powerful tool for describing problem domains are:
 
 \begin{enumerate}
-	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++.
-It also supports polymorphism for some of the planning components.
-Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
-	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
-	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
+
+\item Object Oriented: NDDL supports classes and
+  inheritance in similar fashion to popular Object-Oriented languages
+  like Java and C++. It also supports polymorphism for some of the
+  planning components. Using object classes and instances is a
+  time-tested approach to naturally describe problem domains \comment{OO
+    citation}.
+
+\item It offers declarative constructs to define constraints
+  and causality dependencies between actions and effects.
+
+\item It offers procedural constructs to populate \eus plan database
+  with specific problem instances expressed in terms of objects,
+  variables, constraints, facts and goals.
+
 \end{enumerate}
 
-To understand how the most important elements of NDDL work, let's take a look at how the Shopping Agent problem described above would be modeled:
+To understand how the most important elements of NDDL work, let's take a
+look at how the Shopping Agent problem described above would be modeled:
 
@@
 }
-
 \end{verbatim}
@@
 
-TODO: explain variable, object, predicate and action types, including constraints that can be part of them
+\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
 \begin{enumerate}
@@
 
-Now let's look at how a specific problem instance would be specified in NDDL:
+ shows how a specific problem instance would be
+specified in NDDL.
 
@@
 
+\label{eu:model}
 \end{verbatim}
 
-TODO:  explain variable, object, predicate and constraint instances.
-explain goals and facts.
+\comment{explain variable, object, predicate and constraint instances.
+explain goals and facts.}
 
@@
 
-TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
+\comment{explain how inference is used to detect perform bounds
+  propagation, domain reduction and to detect constraint violations.} 
 - talk about AC3 implementation and extension mechanisms
@@
 
-The Temporal Network:
-The notion of time is central to temporal planning.
-EUROPA uses variables to explicitly represent timepoints for plan activities and states.
-Constraints among timepoints provide a natural way to express domain axioms.
-For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
+The Temporal Network: The notion of time is central to temporal
+planning. \eu uses variables to explicitly represent timepoints for plan
+activities and states. Constraints among timepoints provide a natural
+way to express domain axioms. For example, in order to state that
+activity A must occur before activity B we can say that the end
+timepoint of A is $leq$ the start timepoint of B.
 
@@
 
-Dechter et al.
-(Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN).
-Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node.
-The diagram below illustrates a simple STN with just 2 variables and a single constraint.
-It also shows the resulting DG.
+\cite{dechter91} proposed that constraints among timepoints can be
+grouped together to form a Simple Temporal Network (STN). Such a network
+can be transformed into a Distance Graph (DG) where the outward arc from
+a node represents the maximum distance from the source node to the
+target node. They also showed that shortest path algorithms could be
+used to propagate values in the network and discover a negative cycle. A
+negative cycle is a path from a node to itself that has a path length
+less than 0. If such a cycle exists, the network is inconsistent. It was
+further shown that a single-source shortest path algorithm was
+sufficient to detect a negative cycle and provide sufficient propagation
+to yield a backtrack-free search. Thus we have an efficient and complete
+algorithm for propagating an STN. These results build on the already
+established notion of a CSP and are naturally incorporated into the
+general representation and propagation scheme used in \eu.
+The diagram below illustrates a simple STN with just 2
+variables and a single constraint. It also shows the resulting DG.
 
-TODO: add figure
-
-Dechter et al.
-(Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle.
-A negative cycle is a path from a node to itself that has a path length less than 0.
-If such a cycle exists, the network is inconsistent.
-It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search.
-Thus we have an efficient and complete algorithm for propagating an STN.
-These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.
+\comment{ add figure}
 
@@
 
-    Reservoir resources can have both 'consume' and 'produce' tokens.
-Note that these tokens do NOT have the usual start/end/duration token variables, since occur at a single instant in time.
-Instead, they have a single variable, 'time' used to represent that instant.
-    Reusable resources have 'uses' tokens that use a quantity for the duration of the token. These tokens do have the usual start/end/duration variables.
-    Unary resources are reusable resources with unit quantity and have 'use' tokens (again, with the usual start/end/duration variables).
+Reservoir resources can have both 'consume' and 'produce' tokens. Note
+that these tokens do NOT have the usual start/end/duration token
+variables, since occur at a single instant in time. Instead, they have a
+single variable, 'time' used to represent that instant.
 
-    TODO: talk about resource envelope computation by using MaxFlow formulation.
+Reusable resources have 'uses' tokens that use a quantity for the
+duration of the token. These tokens do have the usual start/end/duration
+variables.
+
+Unary resources are reusable resources with unit quantity and have 'use'
+tokens (again, with the usual start/end/duration variables).
+
+\comment{talk about resource envelope computation by using MaxFlow formulation.}
 
@@
 
-Now we have all the elements in place so that an automated problem solver can be created. Let's recap what those elements are:
+Now we have all the elements in place so that an automated problem
+solver can be created. Let's recap what those elements are:
 
@@
 
-\item A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
+\item A domain model that describes the variable, object, predicate and
+  action types that are relevant for the problem.
+
 \item A problem instance (also called initial state) that consists of:
@@
 
-All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.
+All this information is kept in \eus Plan Database so that inference and
+search mechanisms can be used to look for a problem solution.
 
-Given that temporal intervals may extend indefinitely in both positive and negative directions, for problems with a temporal dimension it is also common to specify a Planning Horizon when invoking a solver.
-The Planning Horizon is the temporal interval that a problem solver has to worry about, any decisions that fall completely outside of the Planning Horizon can be ignored.
-This is very useful for recurrent activities, consider for instance a model where a medical checkup has to occur every 3 months, the user can model a recurring activity and then use the Planning Horizon to ensure that a finite number of medical checkup activities are generated as part of a plan.
-This is also useful to ignore activities that may happen too far in the future or in the past to be relevant for a particular plan, this way the user can ensure that the solver is making as few commitments as necessary.
+Given that temporal intervals may extend indefinitely in both positive
+and negative directions, for problems with a temporal dimension it is
+also common to specify a Planning Horizon when invoking a solver. The
+Planning Horizon is the temporal interval that a problem solver has to
+worry about, any decisions that fall completely outside of the Planning
+Horizon can be ignored. This is very useful for recurrent activities,
+consider for instance a model where a medical checkup has to occur every
+3 months, the user can model a recurring activity and then use the
+Planning Horizon to ensure that a finite number of medical checkup
+activities are generated as part of a plan. This is also useful to
+ignore activities that may happen too far in the future or in the past
+to be relevant for a particular plan, this way the user can ensure that
+the solver is making as few commitments as necessary.
 
-EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals.
-The operations to refine the partial plan PP at any time are:
+\eu provides a built-in solver that performs Plan Space Planning
+\cite{ghallab04} in this approach, the initial state is considered a
+partial plan that needs to be refined toward a solution plan that
+achieves the goals. The operations to refine the partial plan PP at any
+time are:
+
 \begin{enumerate}
@@
 
-This Plan Space Planning algorithm is translated into EUROPA's representation as follows:
+This Plan Space Planning algorithm is translated into \eus
+representation as follows:
 
-The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances.
-Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:
+The initial partial plan is the state of \eus Plan Database after the
+initial state has been instantiated, this results in a set of variable,
+object and token instances. Inference then takes place to detect flaws
+in the partial plan, there are 3 kinds of flaws that can be detected:
 
 \begin{enumerate}
-	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
 
-	\item Open Condition: an open condition is an inactive token.
-Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
-Open Conditions can be resolved in three ways:
+\item Unbound Variable: a variable in the partial plan whose domain is
+  not a singleton. Unbound Variables are resolved by specification of a
+  value from the domain of the variable.
+
+\item Open Condition: an open condition is an inactive token. Inactive
+  tokens can be generated by explicitly posted goals, or when a token is
+  activated, rules may fire that create inactive slave tokens (TODO:
+  point to more detailed decription above in Rules Engine section). Open
+  Conditions can be resolved in three ways:
+
   \begin{enumerate}
-		\item Merging: A free activity is merged with a matching activity already in the plan.
-An activity a is said to match an activity a?
-if a and a?
-unify and the temporal constraints involving a are satisÞed by a?.
-Thus,  a and a?
-can be considered the same activity and we do not need to introduce a in the plan.
-Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a?
-was introduced in the plan.
-		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
-This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities.
-		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
-\end{enumerate}
 
-	\item Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
-Recall for example that a token may belong to objects (e.g.
-Timelines) which require a total order over their tokens.
-If any 2 tokens could possibly overlap (though not necessarily), then they pose a threat to each other in terms of achieving an extension of the current partial plan which is complete and consistent.
-Similarly, threats may arise where tokens share a common resource and their current state might yield extensions of the current partial plan which are inconsistent.
-Threats are resolved by imposing ordering constraints among tokens.
+  \item Merging: A free activity is merged with a matching activity
+    already in the plan. An activity a is said to match an activity a?
+    if a and a? unify and the temporal constraints involving a are
+    satisÞed by a?. Thus, a and a? can be considered the same activity
+    and we do not need to introduce a in the plan. Consequently, the
+    compatibilities associated with a are not Þred, because they have
+    been already triggered when a? was introduced in the plan.
+
+  \item Activation: We introduce a new activity a in the current plan
+    associating it with the proper timeline, but without choosing a
+    speciÞc time slot for it.The compatibilities associated with a are
+    applied and the subgoal activities resulting from those
+    compatibilities are introduced as free activities. This results in
+    both an ordering flaw, corresponding to the just activated activity,
+    and a number of open condition flaws, corresponding to the new
+    subgoal activities.
+
+  \item Rejection: Some goals may be optional, if that is the case for a
+    particular order condition, its corresponding flaw can be resolved
+    by discarding the inactive token.
 
@@
 
-Open Conditions and Threats allow flaw detection and resolution at a higher-level of abstraction (i.e.
-in terms of objects and tokens) than that of simply binding variables as is common in Constraint Satisfaction Problems.
-This is advantageous when one applies heuristics for ordering choices since it provides a richer context in which to make decisions.
-Furthermore, it aids in reducing the amount of work done by a solver so that only the necessary refinements are made, otherwise leaving the partial plan with maximum flexibility.
-For example, one can omit unbound variables which are time-points of tokens since threats will force a solver to impose restrictions on these variables based on the semantics of the objects to which their tokens apply.
-Thus the planning process may yield partially-ordered plans for which all possible extensions are provably valid.
+\item Threat: once a token has been placed in the partial plan it may
+  impact other tokens indirectly through possible overlapping
+  requirements on objects. Recall for example that a token may belong to
+  objects (e.g. Timelines) which require a total order over their
+  tokens. If any 2 tokens could possibly overlap (though not
+  necessarily), then they pose a threat to each other in terms of
+  achieving an extension of the current partial plan which is complete
+  and consistent. Similarly, threats may arise where tokens share a
+  common resource and their current state might yield extensions of the
+  current partial plan which are inconsistent. Threats are resolved by
+  imposing ordering constraints among tokens.
 
-The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in Algorithm \ref{alg:europa:solve}.
- % summarized in the figure below.
+\end{enumerate}
+
+Open Conditions and Threats allow flaw detection and resolution at a
+higher-level of abstraction (i.e. in terms of objects and tokens) than
+that of simply binding variables as is common in Constraint Satisfaction
+Problems. This is advantageous when one applies heuristics for ordering
+choices since it provides a richer context in which to make decisions.
+Furthermore, it aids in reducing the amount of work done by a solver so
+that only the necessary refinements are made, otherwise leaving the
+partial plan with maximum flexibility. For example, one can omit unbound
+variables which are time-points of tokens since threats will force a
+solver to impose restrictions on these variables based on the semantics
+of the objects to which their tokens apply. Thus the planning process
+may yield partially-ordered plans for which all possible extensions are
+provably valid.
+
+The Plan Space Planning algorithm described above can be implemented in
+many different ways depending on the approach chosen for flaw and
+resolver selection and for backtracking. \eus built-in solver implements
+a chronological backtracking algorithm that is summarized in Algorithm
+\ref{alg:europa:solve}.
 
@@
 
-The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
+The algorithm takes as input a partial plan p and returns true if a
+complete and consistent refinement of p could be found (or if p is
+initially complete and consistent), and false otherwise.
 
@@
 
-This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.
+This algorithm provides for a sound and complete search, assuming that
+no flaws or available refinement operators are pruned unnecessarily.
 Dead-ends in the search are discovered through constraint propagation.
-Constraint propagation is a vehicle for evaluating the consistency of a partial plan and also for filtering infeasible values from consideration prior to commitment, allowing in some cases a strong look-ahead capability which is essential for tractable search.
-Consistency testing is initiated by the isConsistent procedure.
-The algorithm permits a heuristically controlled search by applying orderings for chooseFlaw and makeDecisionPoint.
-It results in a chronologically-backtracking, search.
+Constraint propagation is a vehicle for evaluating the consistency of a
+partial plan and also for filtering infeasible values from consideration
+prior to commitment, allowing in some cases a strong look-ahead
+capability which is essential for tractable search. Consistency testing
+is initiated by the isConsistent procedure. The algorithm permits a
+heuristically controlled search by applying orderings for chooseFlaw and
+makeDecisionPoint. It results in a chronologically-backtracking, search.
 
-TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
+\comment{expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint}
 examples:
@@
 
-It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?
+It should be emphasized that while this algorithm is commonly employed,
+it is only one of many that could be implemented. TODO: give examples of
+local search solvers for scheduling and CSP?
 
@@
 
-TODO: 
-- explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.
+\comment{explain how other solvers can be built, maybe use NQueens or
+  RCPSP example to illustrate. Or maybe build a little specialized
+  solver for shopping or rover examples.}
 

### 2012-02-06 trunk/arch.tex
Update on the introduction of T-REX

@@
 the local behavior they control or even do not have any explicit 
-formal model of this behavior. Taking cues from the IDEA 
-architecture \cite{mus02}, we designed a core framework that provides a
-formal basis \cite{Py:2010ti} on the way each of the sub-component --
-called reactors -- of our agent can interact by exchanging only
-information through {\em state variables} by ensuring that state 
-information can propagate through all the reactors ensuring a
-consistent view of the present state of the world at every single tick
-trough a bottom-up {\em synchronization} flow and collaboration for
-future state evolution by timely exchanges of goals on {\em state
-  variables}. This exchange does not impose to any reactor to be aware
-of the reactors it is connected to but only focus on the state
-variables it relies on (called {\em external}) and the state
-variables it manages and maintains (called {\em internal}).
+formal model of this behavior. The two angles our architecture attck
+this issue are the following:
+\begin{itemize}
+\item Distribute the global planning problem into multiple decision
+  loops, called {\em reactors}. Each reactor have its own scope both 
+  functionally (one for example focusing on the conversion of high 
+  level waypoints into low-level commands) and temporally as 
+  each reactor specifies to the agent their planning look-ahead and
+  their expected latency before being able to produce plan.
+\item Enforce a stronger interaction between the planning process 
+  and the integration of the world evolution. The objective being to
+  make sure that the plannign task is done while continuously informed 
+  on new world state change. By doing so, we expect to allow the planner
+  tohave better informed decision despite the fact that its latency is
+  larger that the world change rate.
+\end{itemize}
+
+For the first point T-REX provides a formal framework described in
+details in \cite{Py:2010ti} and is not the focus of this paper. The
+second have deep implication on how on should integrate a planner in
+this framework in general and was applied to the europa framework
+specifically. We'll develop later how this requirement implies to see
+in-situ planning as two parallels processes that are tightly linked
+through the same plan structure. But before that we need to introduce
+at least the high leve lconcepts and ideas behind our architecture as
+a whole.
+
+T-REX architecture is structured around the notion of 
+{\em state variables} as the unique basis of interactionalong
+with a well defined ownership  model for each of these. 
+In this framework each reactor is seen as a
+black box providing its own {\em internal} state variables for which
+its is reponsible to maintain its value at every single instant --
+represented by a {\em tick} with a fixed duration. In order to
+identify its state a reactor can subscribe to {\em external} state
+variables for  which it will received new updates when the owner of
+these will produce its {\em internal} state. This {\em
+  synchronization} process progate through the reactor dependency 
+graph on a bottom up flow at every {\em tick} ensuring that all the
+reactors have a consistent view of the world. Conversely each reactor
+can request a future change on one of its {\em external} state
+varaiable which will be transfromed as an {\em internal} goal to the
+owner. This goal is expected to porduce a deliberation phase on the
+receiver which can in turn produce a plan impacting its {\em external}
+state variables that will propagate down following the same
+mechanism.
+
+% Taking cues from the IDEA 
+% architecture \cite{mus02}, we designed a core framework that provides a
+% formal basis \cite{Py:2010ti} on the way each of the sub-component --
+% called reactors -- of our agent can interact by exchanging only
+% information through {\em state variables} by ensuring that state 
+% information can propagate through all the reactors ensuring a
+% consistent view of the present state of the world at every single tick
+% trough a bottom-up {\em synchronization} flow and collaboration for
+% future state evolution by timely exchanges of goals on {\em state
+%   variables}. This exchange does not impose to any reactor to be aware
+% of the reactors it is connected to but only focus on the state
+% variables it relies on (called {\em external}) and the state
+% variables it manages and maintains (called {\em internal}).
 

### 2012-02-06 concepts.tex
updates

@@
 plan representation. Section \ref{sec:arch} has details the
-\texttt{T-REX} partitioned architecture and estimation driven vehicle
+\rx partitioned architecture and estimation driven vehicle
 adaptation.
 
-\eu uses a domain model written in a declarative language (NDDL: New
-Domain Description Language), together with initial conditions and
-goals also specified in NDDL, to construct a set of temporal relations
-that must be true at the start time. These models include assertions
-about the physics of the vehicle, i.e how it responds to external
-stimulus and internally driven goals. By propagating these relations
-forward using Simple Temporal Networks \cite{dechter91} and applying
-goal constraints, \eu can select a set of conditions that should be
-true in the future, where some of these conditions will correspond to
-actions the agent must take. The planner can backtrack and try another
-path during search if a goal cannot be reached while being capable of
-discarding unachievable goals.
+\subsection{The basics}
+\label{sec:basics}
+
+\eu uses a \emph{domain model} written in a declarative language
+(NDDL: New Domain Description Language), together with initial
+conditions and goals also specified in NDDL, to construct a set of
+temporal relations that must be true at the start time. These models
+include assertions about the physics of the vehicle, i.e how it
+responds to external stimulus and internally driven goals. By
+propagating these relations forward using Simple Temporal Networks
+\cite{dechter91} and applying goal constraints, \eu can select a set
+of conditions that should be true in the future, where some of these
+conditions will correspond to actions the agent must take. The planner
+can backtrack and try another path during search if a goal cannot be
+reached while being capable of discarding unachievable goals.
 
@@
 
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time the most appropriate one for the actual execution
+conditions. The fact that constraints are explicitly represented
+ensures that through constraint propagation the executive will respect
+global limits expressed in the plan (e.g., don't start a task until a
+certain condition has been satisfied but still satisfying some global
+deadline). Such flexibility is critical when dealing with dynamic
+ocean conditions where precise timing of a robotic action might be
+indeterminate. Further, the advantage of flexibility can be contrasted
+with the consequences of the intrinsic inflexibiliy of traditional
+command sequences. Because they are inflexible, sequences must
+necessarily be designed considering worst case scenarios.
+
 \eu uses a \emph{state variable} representation to describe the
@@
 
-Unlike a traditional fixed time-tagged command sequences, such
-flexible plans leave room for adaptation at execution time. When the
-executive considers when to start a task, it propagates information
-through the constraint network, computes a time bound for the
-variable, selects an actual execution time within the bound, and
-starts the task at that time. Temporally flexible plans therefore,
-express a \textit{range of possible outcomes} of the robots
-interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
-
 \begin{figure}
@@
 
+% \textit{Summary}
+
+% This has presented the main elements of the EUROPA planning approach. They are:
+
+% \begin{enumerate}
+
+% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+% states and actions in time. Note that tokens do not discriminate
+% between state and action.
+
+% \item Token States which are the basis of planning operations on a partial
+% plan.
+
+% \item Constraints to describe relationships among tokens. This provides an
+% expressive method of describing interactions among tokens in the
+% context of Temporal Planning.
+
+% \item Timelines as a concise abstraction to express the evolution of state
+% and behavior for a system component. It provides semantics of mutual
+% exclusion sequencing in time. Other core abstractions are available in
+% \eu for handling metric resources.
+
+% \item Domain rules to describe internal and external relationships on and
+% between tokens respectively. Rules are applied to active tokens
+% (master tokens) referred to as masters and typically produce inactive
+% tokens (slaves) which are a key vehicle through which the planning
+% process evolves. The rule structure in \eu differs from the more
+% restrictive commitment to preconditions and effects in classical
+% planning which prohibits durative actions and disembodied effects
+% (i.e. effects which may not occur until some temporal distance after
+% the end of the action).
+
+% \end{enumerate}
+
 % \subsubsection{Partitioned Control}

### 2012-02-07 europa.tex
updates

--- 2012-01-31 europa.tex
@@
-\section{EUROPA}
-\label{sec:europa}
-
-EUROPA is a general purpose AI planning toolkit developed at NASA Ames.
-Automated Planning can be reformulated as a constraint satisfaction problem (TODO: ref).
-EUROPA is an extension of that idea, however, unlike other approaches where the entire planning problem is translated into a CSP and then solved using traditional CSP methods, EUROPA uses a constraint reasoning engine as a fundamental building block, and enriches it with other architectural components that allow a more direct mapping for scheduling and planning problems, which in turn allows easier development of efficient solving methods.
-
-Let's look at some of the problems that can be tackled using EUROPA.
-
-Constraint Satisfaction: 
-In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.
-
-Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.
+\subsection{\eu Architecture}
+\label{sec:europa:arch}
 
@@
 \centering
-\includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
-\caption{\small N-Queens problem}
-\label{fig:nqueens-1}
-\vskip+0.1cm
-\end{figure}
-
-If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
-Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.
-
-then we have to satisfy the following constraints:
-Sum(Qrc) = 1 ForAll r (Only one Queen per row)
-Sum(Qrc) = 1 ForAll c (Only one Queen per column)
-
-Sum(Qr+i,c+i) = 1
-Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)
-
-We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.
-
-Below is a solution found by EUROPA using that formulation and a specialized search procedure.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small N-Queens solution}
-\label{fig:nqueens-2}
-\vskip+0.1cm
-\end{figure}
-
-Scheduling:
-In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints.
-The activity schedule must also respect fixed limits on the availability of resources required to perform each activity.
-In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.
-
-Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small RCPSP problem}
-\label{fig:rcpsp-1}
-\vskip+0.1cm
-\end{figure}
-
-Planning:
-In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints.
-The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.
-
-Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
-\caption{\small Shopping Agent problem}
-\label{fig:shopping-1}
-\vskip+0.1cm
-\end{figure}
-
-As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.
-
-These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs.
-Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block.
-Next we will examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.
-
-\subsection{Constraint Programming}
-\label{sec:europa:cp}
-
-Constraint Satisfaction Programming, also know as Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints.
-A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].
-
-	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN.
-Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains.
-Given a defined conjunctive set of constraints on the variables: $C=\{c1(x1,..,xN), ..., cK(x1,..,xN)\}$, the objective is to find one or more value assignments to V where all constraints are satisfied.
-
-To solve a problem, CP uses logical inference to perform 2 operations:
-
-\begin{enumerate}
-	\item Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2
-
-	\item Domain reduction: To infer a valid set of values for a variable.
-For example, for discrete variables, the constraints allDifferent(x1,x2,x3) and xi>=0, xi <=4.
-if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}
-
-\end{enumerate}
-
-CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95].
-Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.
-
-Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.
-
-TODO: talk about arc consistency algorithms, computational complexity and solvers.
-
-\subsection{Constraint-Based Attribute and Interval Planning}
-\label{sec:europa:cp}
-
-There has been 
-State Variable representation
-
-CSP encodings
-
-Constraint Based and interval planning.
-
-\subsection{EUROPA's Architecture}
-\label{sec:europa:arch}
-
-The following figure displays the main architectural components in EUROPA and their relationships:
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/EUROPA-Architecture.jpg}
-\caption{\small EUROPA's Architecture}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
+\caption{\small The \eu Architecture}
 \label{fig:europa-architecture}
@@
 
-\begin{enumerate}
+Fig. \ref{fig:europa-architecture} shows the main architectural
+components in \eu and their relationships.
 
-	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables.
-It also provides an efficient arc consistency mechanism (TODO: ref to AC-3).
-The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.
+The Constraint Reasoning Engine (CRE) manages variables, the domains
+from which they can take values, and constraints that define
+relationships among variables. It also provides an efficient arc
+consistency mechanism \cite{mackworth77}. The CRE is designed so that
+specialized reasoning algorithms for specific constraints can be easily
+and efficiently plugged in. The Plan Database (PDB) manages object types
+and objects, which are a mechanism to group variables to more naturally
+model the real world in much the same way that object oriented
+programming does. Also manages tokens, which are a mechanism to group
+variables to represent temporally scoped state. Objects and tokens can
+be used to model planning domains in a much more natural and extensible
+way that a pure CSP approach can. The Rules Engine (RE) manages token
+relationships in a planning model. The language modules (NDDL, ANML, and
+other higher level modeling languages) are used to create domain models
+and problem instances \comment{point advantages over pure programming
+  API approaches, provide snippets}. Finally the API Layer (Fig.
+\ref{fig:europa-api}) makes all the services available so that \eu can
+be used to build applications.
 
-	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does.
-Also manages tokens, which are a mechanism to group variables to represent temporally scoped state.
-Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can (TODO: provide snippet?)
+\comment{Is Fig. \ref{fig:europa-api} necessary?}
 
-	TODO: talk about the token network, move stuff from the Concepts section if necessary
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
+\caption{\small EUROPA's Client API}
+\label{fig:europa-api}
+\vskip+0.1cm
+\end{figure}
 
-	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)
-
-	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)
-
-	\item API Layer: makes all the services available so that EUROPA can be used to build applications.
-
-	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
-		\begin{enumerate}
-			\item Temporal Network
-			\item Resources
-			\item NDDL
-			\item ANML (TODO:maybe it's better to leave out?)
-		\end{enumerate}
-
-\end{enumerate}
+	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+	% 	\begin{enumerate}
+	% 		\item Temporal Network
+	% 		\item Resources
+	% 		\item NDDL
+	% 		\item ANML (TODO:maybe it's better to leave out?)
+	% 	\end{enumerate}
 
@@
 }
-
 \end{verbatim}
-
-TODO: explain variable, object, predicate and action types, including constraints that can be part of them
 
@@
 
+\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
 \begin{enumerate}
-	\item Variable Types:
+	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
 	\item Object Types:
@@
 
-The Temporal Network:
-The notion of time is central to temporal planning.
-EUROPA uses variables to explicitly represent timepoints for plan activities and states.
-Constraints among timepoints provide a natural way to express domain axioms.
-For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.
-Dechter et al.
-(Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN).
-Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node.
-The diagram below illustrates a simple STN with just 2 variables and a single constraint.
-It also shows the resulting DG.
+The Temporal Network: The notion of time is central to temporal
+planning. \eu uses variables to explicitly represent timepoints for plan
+activities and states. Constraints among timepoints provide a natural
+way to express domain axioms. For example, in order to state that
+activity A must occur before activity B we can say that the end
+timepoint of A is $leq$ the start timepoint of B.
 
-TODO: add figure
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\caption{\small Temporal relations defined within the planner are
+  based on \texttt{Allen Algebra} relations shown above.}
+\label{fig:allen-algebra}
+\vskip-0.3cm
+\end{figure}
 
-Dechter et al.
-(Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle.
-A negative cycle is a path from a node to itself that has a path length less than 0.
-If such a cycle exists, the network is inconsistent.
-It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search.
-Thus we have an efficient and complete algorithm for propagating an STN.
-These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.
+\cite{dechter91} proposed that constraints among timepoints can be
+grouped together to form a Simple Temporal Network (STN). Such a network
+can be transformed into a Distance Graph (DG) where the outward arc from
+a node represents the maximum distance from the source node to the
+target node. They also showed that shortest path algorithms could be
+used to propagate values in the network and discover a negative cycle. A
+negative cycle is a path from a node to itself that has a path length
+less than 0. If such a cycle exists, the network is inconsistent. It was
+further shown that a single-source shortest path algorithm was
+sufficient to detect a negative cycle and provide sufficient propagation
+to yield a backtrack-free search. Thus we have an efficient and complete
+algorithm for propagating an STN. These results build on the already
+established notion of a CSP and are naturally incorporated into the
+general representation and propagation scheme used in \eu.
+The diagram below illustrates a simple STN with just 2
+variables and a single constraint. It also shows the resulting DG.
+
+\comment{ add figure}
 
@@
 
-This Plan Space Planning algorithm is translated into EUROPA's representation as follows:
+This Plan Space Planning algorithm is translated into \eus
+representation as follows:
 
-The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances.
-Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:
+The initial partial plan is the state of \eus Plan Database after the
+initial state has been instantiated, this results in a set of variable,
+object and token instances. Inference then takes place to detect flaws
+in the partial plan, there are 3 kinds of flaws that can be detected:
 
 \begin{enumerate}
-	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.
 
-	\item Open Condition: an open condition is an inactive token.
-Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).
-Open Conditions can be resolved in three ways:
+\item Unbound Variable: a variable in the partial plan whose domain is
+  not a singleton. Unbound Variables are resolved by specification of a
+  value from the domain of the variable.
+
+\item Open Condition: an open condition is an inactive token. Inactive
+  tokens can be generated by explicitly posted goals, or when a token is
+  activated, rules may fire that create inactive slave tokens
+  \comment{point to more detailed decription above in Rules Engine
+    section}.
+  Open
+  Conditions can be resolved in three ways:
+
   \begin{enumerate}
-		\item Merging: A free activity is merged with a matching activity already in the plan.
-An activity a is said to match an activity a?
-if a and a?
-unify and the temporal constraints involving a are satisÞed by a?.
-Thus,  a and a?
-can be considered the same activity and we do not need to introduce a in the plan.
-Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a?
-was introduced in the plan.
-		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities.
-This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities.
-		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
-\end{enumerate}
 
-	\item Threat: once a token has been placed in the partial plan it may impact other tokens indirectly through possible overlapping requirements on objects.
-Recall for example that a token may belong to objects (e.g.
-Timelines) which require a total order over their tokens.
-If any 2 tokens could possibly overlap (though not necessarily), then they pose a threat to each other in terms of achieving an extension of the current partial plan which is complete and consistent.
-Similarly, threats may arise where tokens share a common resource and their current state might yield extensions of the current partial plan which are inconsistent.
-Threats are resolved by imposing ordering constraints among tokens.
+  \item Merging: A free activity is merged with a matching activity
+    already in the plan. An activity a is said to match an activity a?
+    if a and a? unify and the temporal constraints involving a are
+    satisÞed by a?. Thus, a and a? can be considered the same activity
+    and we do not need to introduce a in the plan. Consequently, the
+    compatibilities associated with a are not Þred, because they have
+    been already triggered when a? was introduced in the plan.
+
+  \item Activation: We introduce a new activity a in the current plan
+    associating it with the proper timeline, but without choosing a
+    speciÞc time slot for it.The compatibilities associated with a are
+    applied and the subgoal activities resulting from those
+    compatibilities are introduced as free activities. This results in
+    both an ordering flaw, corresponding to the just activated activity,
+    and a number of open condition flaws, corresponding to the new
+    subgoal activities.
+
+  \item Rejection: Some goals may be optional, if that is the case for a
+    particular order condition, its corresponding flaw can be resolved
+    by discarding the inactive token.
 
@@
 
-Open Conditions and Threats allow flaw detection and resolution at a higher-level of abstraction (i.e.
-in terms of objects and tokens) than that of simply binding variables as is common in Constraint Satisfaction Problems.
-This is advantageous when one applies heuristics for ordering choices since it provides a richer context in which to make decisions.
-Furthermore, it aids in reducing the amount of work done by a solver so that only the necessary refinements are made, otherwise leaving the partial plan with maximum flexibility.
-For example, one can omit unbound variables which are time-points of tokens since threats will force a solver to impose restrictions on these variables based on the semantics of the objects to which their tokens apply.
-Thus the planning process may yield partially-ordered plans for which all possible extensions are provably valid.
+\item Threat: once a token has been placed in the partial plan it may
+  impact other tokens indirectly through possible overlapping
+  requirements on objects. Recall for example that a token may belong to
+  objects (e.g. Timelines) which require a total order over their
+  tokens. If any 2 tokens could possibly overlap (though not
+  necessarily), then they pose a threat to each other in terms of
+  achieving an extension of the current partial plan which is complete
+  and consistent. Similarly, threats may arise where tokens share a
+  common resource and their current state might yield extensions of the
+  current partial plan which are inconsistent. Threats are resolved by
+  imposing ordering constraints among tokens.
 
-The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking.
-EUROPA's built-in solver implements a chronological backtracking algorithm \ref{alg:europa:solve}
- % summarized in the figure below.
+\end{enumerate}
+
+Open Conditions and Threats allow flaw detection and resolution at a
+higher-level of abstraction (i.e. in terms of objects and tokens) than
+that of simply binding variables as is common in Constraint Satisfaction
+Problems. This is advantageous when one applies heuristics for ordering
+choices since it provides a richer context in which to make decisions.
+Furthermore, it aids in reducing the amount of work done by a solver so
+that only the necessary refinements are made, otherwise leaving the
+partial plan with maximum flexibility. For example, one can omit unbound
+variables which are time-points of tokens since threats will force a
+solver to impose restrictions on these variables based on the semantics
+of the objects to which their tokens apply. Thus the planning process
+may yield partially-ordered plans for which all possible extensions are
+provably valid.
+
+The Plan Space Planning algorithm described above can be implemented in
+many different ways depending on the approach chosen for flaw and
+resolver selection and for backtracking. \eus built-in solver implements
+a chronological backtracking algorithm that is summarized in Algorithm
+\ref{alg:europa:solve}.
 
@@
 
-% \begin{verbatim}
-
-% 01.  bool solve(PartialPlan plan)
-% 02.   {
-% 03.      if (isInconsistent(plan))  return false;
-% 04.
-% 05.      Flaw flaw = chooseFlaw(plan);
-% 06.      if (flaw == NULL) return true;
-% 07.
-% 08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
-% 09.
-% 10.      while(decision.hasNext()) {
-% 11.          PartialPlan pp = decision.executeNext();
-% 12.
-% 13.          if (solve(pp)) return true;
-% 14.          else d.undo();
-% 25.      }
-% 16.
-% 17.      return false;
-% 18.  }
-
-% \end{verbatim}
-
-The algorithm takes as input a partial plan p and returns true if a complete and consistent refinement of p could be found (or if p is initially complete and consistent), and false otherwise.
+The algorithm takes as input a partial plan p and returns true if a
+complete and consistent refinement of p could be found (or if p is
+initially complete and consistent), and false otherwise.
 
@@
 
-This algorithm provides for a sound and complete search, assuming that no flaws or available refinement operators are pruned unnecessarily.
+This algorithm provides for a sound and complete search, assuming that
+no flaws or available refinement operators are pruned unnecessarily.
 Dead-ends in the search are discovered through constraint propagation.
-Constraint propagation is a vehicle for evaluating the consistency of a partial plan and also for filtering infeasible values from consideration prior to commitment, allowing in some cases a strong look-ahead capability which is essential for tractable search.
-Consistency testing is initiated by the isConsistent procedure.
-The algorithm permits a heuristically controlled search by applying orderings for chooseFlaw and makeDecisionPoint.
-It results in a chronologically-backtracking, search.
+Constraint propagation is a vehicle for evaluating the consistency of a
+partial plan and also for filtering infeasible values from consideration
+prior to commitment, allowing in some cases a strong look-ahead
+capability which is essential for tractable search. Consistency testing
+is initiated by the isConsistent procedure. The algorithm permits a
+heuristically controlled search by applying orderings for chooseFlaw and
+makeDecisionPoint. It results in a chronologically-backtracking, search.
 
-TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
+\comment{expand a little more on how heuristics can be implemented by
+  applying ordering/filtering for chooseFlaw and makeDecisionPoint}
 examples:
@@
 
-It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?
+It should be emphasized that while this algorithm is commonly employed,
+it is only one of many that could be implemented. \comment{give examples
+  of local search solvers for scheduling and CSP?}
 

### 2012-02-07 trunk/intro.tex
updates

@@
 AI based planning and probabilistic state estimation in a hybrid
-executive \cite{mcgann08a,mcgann08b,Py10}.  Probabilistic state
+executive \cite{mcgann08a,mcgann08b,py10}.  Probabilistic state
 estimation integrates a number of science observations to produce a

### 2012-02-07 arch.tex
updates

@@
 the local behavior they control or even do not have any explicit 
-formal model of this behavior. Taking cues from the IDEA 
-architecture \cite{mus02}, we designed a core framework that provides a
-formal basis \cite{Py:2010ti} on the way each of the sub-component --
-called reactors -- of our agent can interact by exchanging only
-information through {\em state variables} by ensuring that state 
-information can propagate through all the reactors ensuring a
-consistent view of the present state of the world at every single tick
-trough a bottom-up {\em synchronization} flow and collaboration for
-future state evolution by timely exchanges of goals on {\em state
-  variables}. This exchange does not impose to any reactor to be aware
-of the reactors it is connected to but only focus on the state
-variables it relies on (called {\em external}) and the state
-variables it manages and maintains (called {\em internal}).
+formal model of this behavior. The two angles our architecture attck
+this issue are the following:
+\begin{itemize}
+\item Distribute the global planning problem into multiple decision
+  loops, called {\em reactors}. Each reactor have its own scope both 
+  functionally (one for example focusing on the conversion of high 
+  level waypoints into low-level commands) and temporally as 
+  each reactor specifies to the agent their planning look-ahead and
+  their expected latency before being able to produce plan.
+\item Enforce a stronger interaction between the planning process 
+  and the integration of the world evolution. The objective being to
+  make sure that the plannign task is done while continuously informed 
+  on new world state change. By doing so, we expect to allow the planner
+  tohave better informed decision despite the fact that its latency is
+  larger that the world change rate.
+\end{itemize}
+
+For the first point T-REX provides a formal framework described in
+details in \cite{py10} and is not the focus of this paper. The
+second have deep implication on how on should integrate a planner in
+this framework in general and was applied to the europa framework
+specifically. We'll develop later how this requirement implies to see
+in-situ planning as two parallels processes that are tightly linked
+through the same plan structure. But before that we need to introduce
+at least the high leve lconcepts and ideas behind our architecture as
+a whole.
+
+T-REX architecture is structured around the notion of 
+{\em state variables} as the unique basis of interactionalong
+with a well defined ownership  model for each of these. 
+In this framework each reactor is seen as a
+black box providing its own {\em internal} state variables for which
+its is reponsible to maintain its value at every single instant --
+represented by a {\em tick} with a fixed duration. In order to
+identify its state a reactor can subscribe to {\em external} state
+variables for  which it will received new updates when the owner of
+these will produce its {\em internal} state. This {\em
+  synchronization} process progate through the reactor dependency 
+graph on a bottom up flow at every {\em tick} ensuring that all the
+reactors have a consistent view of the world. Conversely each reactor
+can request a future change on one of its {\em external} state
+varaiable which will be transfromed as an {\em internal} goal to the
+owner. This goal is expected to porduce a deliberation phase on the
+receiver which can in turn produce a plan impacting its {\em external}
+state variables that will propagate down following the same
+mechanism.
+
+% Taking cues from the IDEA 
+% architecture \cite{mus02}, we designed a core framework that provides a
+% formal basis \cite{Py:2010ti} on the way each of the sub-component --
+% called reactors -- of our agent can interact by exchanging only
+% information through {\em state variables} by ensuring that state 
+% information can propagate through all the reactors ensuring a
+% consistent view of the present state of the world at every single tick
+% trough a bottom-up {\em synchronization} flow and collaboration for
+% future state evolution by timely exchanges of goals on {\em state
+%   variables}. This exchange does not impose to any reactor to be aware
+% of the reactors it is connected to but only focus on the state
+% variables it relies on (called {\em external}) and the state
+% variables it manages and maintains (called {\em internal}).
 
@@
 observations but will also include the reactor intentions as described
-by the plan produced.
+by the plan produced. Consider for example the Shopping Agent model
+described during the europa presentation and that the
+location of the \texttt{agent} is an {\em external} timeline. For
+example as, the reactor add the goal to have \texttt{Own} milk and the
+current {\em external} states indicated that it was \texttt{At(home)} it
+produced the plan in Fig. \ref{fig:shop:exec0}. The  {\em external}
+  part of the plan -- namely the tokens \texttt{Going(home,
+    superMarket)} and \texttt{At(SuperMarket)} -- and integrated as
+  goals for the reactor managing this specific state varaiable {\em
+    internally}.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}
+  \caption{Shopping example : plan produced before synchronization. 
+    Tokens in red indicate observations, 
+    tokens in blue the goal that this plan attempt to solve. Arrows
+    indicate temporal constraints.}
+  \label{fig:shop:exec0}
+\end{figure}
+
+Now, consider that at the next tick $\tau+1$ the reactor managing the
+agent location was able to change the location state varaibale to 
+the \texttt{Going}. T-REX informs our reactor of this new observation
+which is added in the plan structure as a fact -- starting exactly at
+$\tau+1$ and lastign for an unlknown duration which is at least 1
+tick. As this new observation is integrated in the plan through
+synchronization the solver can identify the similarity between this
+new observation oand the next token in the existing plan. Which mean
+that instead of making a blind exhaustive search of all the possible
+implication the synchronization process starts its execution with the
+assumption that these new observations are consitent with the current
+plan maintained. As a result at the end of synchronization we will
+have the plan depicted in Fig. \ref{fig:shop:exec1}. Which just
+propagated the information provided by the new observation  in the
+exisiting plan (restricting in turn the start time of 
+\texttt{Going} and by propagtion of the constraints \texttt{Go} to  be
+{\em exactly} $\tau+1$). While focusing uniquely on the execution
+frontier ($\tau+1$ in this figure), the europa solver took benefit of
+the exisiting plan to evaluate a solution that asumes that the
+observation is consistent with what was decided  inside this
+reactor.
+
+This choice is directed not only by some natural heurisitc
+choices made in the solver but even more enforced in general by the
+heuristic order of the decision point reolution choices for our {\em
+  internla} state flaw presented in section \ref{sec:arch:sync}
+{\em\color{red} TODO: need to rework the presentation of thee choices
+  so I can refer to them more directly -- algorithm or pseudo theorem
+  should do the trick}. Should such a flaw be revealed during
+synchronization -- meaning that the current state of an {\em internal}
+state varaibale is not fully grounded -- the sequencing of the choices
+will atempt first a conservative choice in term of maintaining
+previous state, then an optimisitic choice in term of the
+execution/advance in our current plan to finally attempt other choices
+that were not predicted by the model. The two first possible choices
+are strongly directed by the exisiting plan which help direct the
+search within this frame. This often allow to have a synchronization
+that is much more directed and i, in nominal situation such as the one
+we showed resolve the deliberation in few steps (often without any
+backtrack in the search).
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
+  \caption{Shopping example : Result of synchronization after the {\em
+      Going} observation was received by the reactor with thew plan
+    showed in Fig. \ref{fig:shop:exec0}.}
+  \label{fig:shop:exec1}
+\end{figure}
+
+The synchronization process is also in this process supporting fully
+the tracking of the plan execution as it specifies tokens parameters
+such as the time points and allow this to propagate through the plan
+in order to idenitfy if the plan remains valid -- as in our example --
+or that it cannot be executed as it is. The later  is often identified
+by the synchronization failing to find a consistent solution. for
+example dshould the \texttt{Going} observation we receive be to the 
+\texttt{hardwareStore}, this would break the current plan of our
+reactor as we cannot go simultaneously at 2 different places as
+illustrated in Fig. \ref{fig:shop:relax}-1. as synchronization is not
+possible in this context our strategy is to fully relax all decisions
+made by the previous plan steps by just keeping n the plan all the
+past observations in both {\em external} and {\em internal} state
+variables and the exsiting {\em internal} goals received by this
+reactor (\texttt{Own(Milk)} in our example) resulting on the plrtial
+plan sghowned in Fig. \ref{fig:shop:relax}-2. In this process the
+reactor inform the agent that all {\em external} goals it requested
+for the future are note valid anymore which often result in the ownere
+of htese timelines to not have to maintain this goal anynmore.
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
+  \caption{Illustration of a conflict suring synchronization and the resulting relaxed plan for recovery}
+  \label{fig:shop:relax}
+\end{figure}
+
+As the plan is not anymore in the way the synchronization can resume 
+and find a solution as we initial lly described. The newly pending goals will enforce  at
+the next \texttt{step} to resumed deliberation on this new fully
+relaxed partial plan. By the simple use of synchronization the reactor 
+was not only able to identify that its plan was not executable anymore 
+but recover from it while offering a new incomplete partial plan to be
+resolved during the next deliberation \texttt{steps}. This on its own
+ogfer the core functionality one should except to close the loop
+between planning and execution in the sense that synchronization plays
+the role of what is typically named as the {\em Executive} in other
+architectures {\em\color{red} references to diverse archis, 3-tiered,
+  laas, claraty, B Wiliams, and many more}.
+
+In our case though the interaction between planning and execution
+does not stop at this level. A core point of the overal schedulling of
+both {\em synchronization} and {\em planning} steps resides on our
+architecture enforsincing that the synchroniuzation process can occur
+in between any step of the plan. This lead us to the fact that not
+only synchronization is imnfluenced by the outcome of planning but
+this influence goes also in the alternate direction. Specifically when
+synchronuiation occuers before we found a complete plan, it will
+inject as we described new facts (the {\em external} observations and
+resulting {\em internal} state) in the plan database. This
+informatiuon will be present when the next planning \texttt{step} will
+resume. Therefore they are not intergal part of the plan resolution
+(generating potentially new flaws for the planner or contributing to
+resolve previously exisiting flaws). The planning process is therefore
+perturbed internallyand fully informed on how the world evolves as it
+is planning.  This aspect trueluy contrast to other approaches we have
+idenitifed in the literature which avoid to perturb the planner during
+it search as it is not compatible with the ``off-line planning'' assumption.
+
+By our design choices -- both at the architecture level and the integration of
+the europa framework for embedded deliberation -- we were able to
+implement an architecture that avoid this limitation while remaining in
+a formal frame that still allow for deliberation and model-based
+agent control. We believe that this tighter integration of planning
+and execution allow the system to ba more accute of its environament
+alowing better informaed decision and tighter reaction. This is
+possible while still integrating information such as a long latency
+for the reactors as long as one can ensure that synchronization of al
+the reactors and deliberation steps can be done in a time that is
+resonably small in comparison of the agent tick duration.
 

### 2012-02-07 trunk/arch.tex
updates

@@
 
-\subsection{The T-REX architecture}
+\subsection{The \rx architecture}
 \label{sec:arch:trex}
@@
 
-The T-REX architecture was designed with the objective to bring 
-planning closer to the low-level control of the vehicle while
-maintaining its reactivity. Indeed, in most of the architecture where
-task planning is embedded, this task remains at a failry high level of
-abstraction as the complexity of this task would otherwise impede the
-system reactiveness. As a result the planning problem managed {\em in
-  situ} remains fairly detached from low-level control and most of the 
+The \rx architecture was designed with the objective to bring planning
+closer to the low-level control of the vehicle while maintaining its
+reactivity. Indeed, in most of the architecture where task planning is
+embedded, this task remains at a failry high level of abstraction as
+the complexity of this task would otherwise impede the system
+reactiveness. As a result the planning problem managed {\em in situ}
+remains fairly detached from low-level control and most of the
 reactive adaptation which in contrast is managed by specialized
-component that relies on different representations for modeling 
-the local behavior they control or even do not have any explicit 
-formal model of this behavior. The two angles our architecture attck
-this issue are the following:
+component that relies on different representations for modeling the
+local behavior they control or even do not have any explicit formal
+model of this behavior. The two angles our architecture attck this
+issue are the following:
+
 \begin{itemize}
+
 \item Distribute the global planning problem into multiple decision
@@
   their expected latency before being able to produce plan.
+
 \item Enforce a stronger interaction between the planning process 
@@
   larger that the world change rate.
+
 \end{itemize}
 
-For the first point T-REX provides a formal framework described in
-details in \cite{Py:2010ti} and is not the focus of this paper. The
-second have deep implication on how on should integrate a planner in
-this framework in general and was applied to the europa framework
-specifically. We'll develop later how this requirement implies to see
-in-situ planning as two parallels processes that are tightly linked
-through the same plan structure. But before that we need to introduce
-at least the high leve lconcepts and ideas behind our architecture as
-a whole.
+For the first point \rx provides a formal framework described in
+details in \cite{py10} and \cite{rajan12} and is not the focus of this
+paper. The second have deep implication on how on should integrate a
+planner in this framework in general and was applied to the europa
+framework specifically. We'll develop later how this requirement
+implies to see in-situ planning as two parallels processes that are
+tightly linked through the same plan structure. But before that we
+need to introduce at least the high leve lconcepts and ideas behind
+our architecture as a whole.
 
-T-REX architecture is structured around the notion of 
-{\em state variables} as the unique basis of interactionalong
-with a well defined ownership  model for each of these. 
-In this framework each reactor is seen as a
-black box providing its own {\em internal} state variables for which
-its is reponsible to maintain its value at every single instant --
-represented by a {\em tick} with a fixed duration. In order to
-identify its state a reactor can subscribe to {\em external} state
-variables for  which it will received new updates when the owner of
-these will produce its {\em internal} state. This {\em
-  synchronization} process progate through the reactor dependency 
+\rx architecture is structured around the notion of {\em state
+  variables} as the unique basis of interactionalong with a well
+defined ownership model for each of these.  In this framework each
+reactor is seen as a black box providing its own {\em internal} state
+variables for which its is reponsible to maintain its value at every
+single instant -- represented by a {\em tick} with a fixed
+duration. In order to identify its state a reactor can subscribe to
+{\em external} state variables for which it will received new updates
+when the owner of these will produce its {\em internal} state. This
+{\em synchronization} process progate through the reactor dependency
 graph on a bottom up flow at every {\em tick} ensuring that all the
@@
 can request a future change on one of its {\em external} state
-varaiable which will be transfromed as an {\em internal} goal to the
+variable which will be transfromed as an {\em internal} goal to the
 owner. This goal is expected to porduce a deliberation phase on the
 receiver which can in turn produce a plan impacting its {\em external}
-state variables that will propagate down following the same
-mechanism.
+state variables that will propagate down following the same mechanism.
 
@@
 
-Take for example the T-REX agent instance we use in our AUV
+Take for example the \rx agent instance we use in our AUV
 (Fig. \ref{fig:agent}). In this architecture all the reactors
@@
 \textsf{Nav\_policy}. This relation applies in two way :
+
 \begin{itemize}
+
 \item In order to identify its current {\em internal} state values
@@
   current \textsf{Nav\_cmd} executed.
+
 \item The future objectives a reactor has on its future \textsf{Internal}
@@
   eventually help him reach this location.
+
 \end{itemize}
@@
 
-Overal T-REX architecture helps to abstract to the reactor interaction
-to only state variables. Each reactor is practically ignoring where
-its {\em external} state information is managed as long as this
-information is available and properly maintained so they can properly 
-manage their own {\em internal} state information which in turn may be
-used by other reactors. Such formalism also constrain reactors
-interaction to state information exchanges which are two folds:
+Overall the \rx architecture helps to abstract to the reactor
+interaction to only state variables. Each reactor is practically
+ignoring where its {\em external} state information is managed as long
+as this information is available and properly maintained so they can
+properly manage their own {\em internal} state information which in
+turn may be used by other reactors. Such formalism also constrain
+reactors interaction to state information exchanges which are two
+folds:
+
 \begin{itemize}
+
 \item {\em Observation on the current state values}: This information
@@
   state of the world throughout all the reactors as time advance.
+
 \item {\em Goals request on future state}: This information propagates
@@
   as a goal to the owner of the given {\em external(s)} state variable(s).
+
 \end{itemize}
@@
 information in a formal and yet flexible way and its transaltion in
-planning frameworks that have an explicit representation of time 
-(such as but not limited to europa) is often trivial.
+planning frameworks that have an explicit representation of time (such
+as but not limited to europa) is often trivial.
 
@@
 only two abstract calls that conceptualize their execution:
+
 \begin{itemize}
+
 \item \texttt{synchronize} that takes the last {\em external}
@@
   observations for this reactor or a failure report.
+
 \item \texttt{step} that accept new goals requested (if any) on the
@@
   applied to {\em external} timelines of this reactor.
+
 \end{itemize}
+
 Both of these two deliberative functions have different scope and
@@
 process which is focused on the future evolution of the state
-variables and it can take multiple steps spanning other multiple
-ticks for a reactor to produce a plan that is both complete and
-valid. when such plan is identified its {\em external} part is then
-dispatched to the corresponding reactors which in turn can start their
-deliberation \texttt{step}s in order to satisfy these new
-objectives. As a consequence physical execution will occur as these
-goals and planning propagate down in the reactor graph until they
-eventually become goals for the lowest level reactor which -- instead
-of deliberating -- may just transform this goal in a physical command
-execution.
+variables and it can take multiple steps spanning other multiple ticks
+for a reactor to produce a plan that is both complete and valid. when
+such plan is identified its {\em external} part is then dispatched to
+the corresponding reactors which in turn can start their deliberation
+\texttt{step}s in order to satisfy these new objectives. As a
+consequence physical execution will occur as these goals and planning
+propagate down in the reactor graph until they eventually become goals
+for the lowest level reactor which -- instead of deliberating -- may
+just transform this goal in a physical command execution.
 
@@
 view of the state of the world between all the reactors.
+
 \begin{figure}[!htbp]
@@
 \end{figure}
+
 This view of the reactor has two parallel tasks aloow for long
@@
 
-The europa reactor is our implementation inside T-REX of a reactor
-that usess automated planning for its execution. This reactor is fully
-based on the europa framework and applies the core idea we of T-REX we
+The \eu reactor is our implementation inside \rx of a reactor that
+usess automated planning for its execution. This reactor is fully
+based on the europa framework and applies the core idea we of \rx we
 described above to implement a reactor for which the sole purpse is to
@@
 
-This reactor is implemented following the idea that in T-REX, the
+This reactor is implemented following the idea that in \rx, the
 whole execution cycle is two deliberation processes. Both of these
 processes are implemented as two europa solvers modifying the same
-plan database and for which the execution is managed by the T-REX 
+plan database and for which the execution is managed by the \rx 
 framework as time advance during the execution of the system:
+
 \begin{enumerate}
+
 \item The synchronization solver is a specialized europa solver that
@@
   solver is summoned at the beginning of every single tick.
+
 \item The deliberation solver is managing the deliberation process of
@@
   order for the synchronization solver to do its task.
+
 \end{enumerate}
+
 While the two processes are separate, they share the same plan
@@
 planning''  defined in \cite{ghallab04} as follow:
+
 \begin{quotation}
@@
 \end{quotation}
+
 This assumption, while helpfull to reduce the scope of the planning
@@
 type of flaw for the europa framework. This flaw enforces at the
-reactor to identify fully its internal state for the current
-tick. By using this new flaw we can describe the synchronization
-process as this general sequence :
+reactor to identify fully its internal state for the current tick. By
+using this new flaw we can describe the synchronization process as
+this general sequence :
+
 \begin{enumerate}
+
 \item Integrate the external state as provided by the owner of each
   external timeline into the plan database.
+
 \item Propagate this information in the plan database following the
   model $\mathcal{M}$ of this reactor.
+
 \item Resolve the current state value of each internal timelines.
+
 \end{enumerate}
@@
 following choices which are evaluated in the given order:
+
 \begin{enumerate}
+
 \item Extend the previous state value to end after this tick (ie
   restrict its end time to $[\tau+1, \infty)$).
+
 \item Start the next active token in the timeline and
@@
   value $\tau$).
+
 \item Create and insert a new token in this timeline that will start
@@
   type for this timeline if necessary).
+
 \end{enumerate}
@@
 so we need to make the assumption that the current state value of an
-internal timeline do not depend on the future or more accurately
-that any choices made during this synchronization will not lead to a
-future inconsistency (meaning no possible solution) for a future
+internal timeline do not depend on the future or more accurately that
+any choices made during this synchronization will not lead to a future
+inconsistency (meaning no possible solution) for a future
 synchronization. Such assumption deeply impact the set of possible
@@
 and {\tt Command} is {\em External} to the reactor.
+
 \begin{verbatim}
@@
 
-Such model present the issue that the reactor needs at one point to
+Such model presents the issue that the reactor needs at one point to
 take the decision to start the token {\tt Vehicle.Dive} and by doing
@@
 running this need to be carefully designed. At the engine level we
-made careful choices in term of both the focus of the search -- limited
-only to tokens that can potentially overlap the current tick -- and
-its heuristic where the internal state flaw is considered last during
-the search. The later proved to limit the amount of backtrack in the
-search for a solution -- or at least how deep we need to backtrack in
-our decision tree -- similarly the sequence in which we evaluate the
-possible resolution for such a flaw -- presented previously -- was
-selected according to simple assumptions that takes benefit of a
-possibly already existing plan by assuming that if the outcome of
-synchronization appears consistent with our current plan then it
-probably reflects that this reactor plan is executed. Such assumption
-helps direct the internal state evaluation and, in nominal cases (where
-we assume that reactors are acting in concert), proved to be a sound
-solution.
+made careful choices in term of both the focus of the search --
+limited only to tokens that can potentially overlap the current tick
+-- and its heuristic where the internal state flaw is considered last
+during the search. The later proved to limit the amount of backtrack
+in the search for a solution -- or at least how deep we need to
+backtrack in our decision tree -- similarly the sequence in which we
+evaluate the possible resolution for such a flaw -- presented
+previously -- was selected according to simple assumptions that takes
+benefit of a possibly already existing plan by assuming that if the
+outcome of synchronization appears consistent with our current plan
+then it probably reflects that this reactor plan is executed. Such
+assumption helps direct the internal state evaluation and, in nominal
+cases (where we assume that reactors are acting in concert), proved to
+be a sound solution.
 
-One final aspect to keep into account is that in T-REX point of view
+One final aspect to keep into account is that in \rx point of view
 synchronization is the only critical task of a reactor. Indeed, such
-process is used in order for the agent to identify its
-current state and maintain it consistent throughout the agent
-life-time. A reactor not being able to find a consistent solution
-during synchronization -- meaning that after evaluating all the
-possible solutions provided by its model it did not found any plan
-that is not inconsistent -- not only expresses that this reactor is
-not able anymore to explain the state of the world but may jeopardize
-other reactors that relies on its {\em internal} state. For this
-reason a failure to do synchronization immediately result on the
-agent killing this reactor and notifying any reactor that depends on
-its {\em internal} state variables by posting the special observation
-\textsf{Failed}. By doing so we allow for graceful degradation where
-other reactors can readapt their plan as they receive this
-\textsf{Failed} observation.
+process is used in order for the agent to identify its current state
+and maintain it consistent throughout the agent life-time. A reactor
+not being able to find a consistent solution during synchronization --
+meaning that after evaluating all the possible solutions provided by
+its model it did not found any plan that is not inconsistent -- not
+only expresses that this reactor is not able anymore to explain the
+state of the world but may jeopardize other reactors that relies on
+its {\em internal} state. For this reason a failure to do
+synchronization immediately result on the agent killing this reactor
+and notifying any reactor that depends on its {\em internal} state
+variables by posting the special observation \textsf{Failed}. By doing
+so we allow for graceful degradation where other reactors can readapt
+their plan as they receive this \textsf{Failed} observation.
 
@@
 ticks -- and the fact that we potentially need to interleave this
-deliberation with not the synchronization process but also with 
-the deliberation of other reactors\footnote{Current implementation of
-  T-REX is running on a single process and it is the responsibility of
-  the agent itself to emulate the multi-threading of the reactors
+deliberation with not the synchronization process but also with the
+deliberation of other reactors\footnote{Current implementation of \rx
+  is running on a single process and it is the responsibility of the
+  agent itself to emulate the multi-threading of the reactors
   deliberation and synchronization}. For this reason while the overal
@@
 around the recursive call (line \ref{li:recurse}) while always
-ensuring that the current partial plan mainitained is never poroven
-inconsitent at any single step. This can be managed by allowing the
+ensuring that the current partial plan mainitained is never proven
+inconsistant at any single step. This can be managed by allowing the
 call to backtrack until a DecisionPoInt that is not exhausted (ie
@@
 synchronization which would immediately fail if the $plan$ is
-inconsistent before any solving attempt made for synchronization. 
-For this reason if the plan is proven inconsistent during a
-deliberation step we either backtrack in the decision stack until we
-find a node fo the tree that have remaining alternate solution or
-simply fully relax the plan in situation there's absolutely no other
-alternative  (by removing all the goals and keeping
-only recent observations produced). These constraints are yet again
-related to the fact that for the agent the only crucial functionality
-of a reactor is to resolve successfully its synchronization.
+inconsistent before any solving attempt made for synchronization.  For
+this reason if the plan is proven inconsistent during a deliberation
+step we either backtrack in the decision stack until we find a node fo
+the tree that have remaining alternate solution or simply fully relax
+the plan in situation there's absolutely no other alternative (by
+removing all the goals and keeping only recent observations
+produced). These constraints are yet again related to the fact that
+for the agent the only crucial functionality of a reactor is to
+resolve successfully its synchronization.
 
-The number of flaw between two steps can evolve for the following
+The number of flaws between two steps can evolve for the following
 reasons that are exogenous to deliberation :
+
 \begin{enumerate}
+
 \item A new goal has been posted on one of the reactor timeline. This
@@
   corresponding new objective to this reactor.
+
 \item The outcome of synchronization created new flaws that did not
   need to be resolved during synchronization.
+
 \item A previously requested goal has be recalled by the initial
@@
   proved this plan to be inconsistent.
+
 \end{enumerate}
+
 These 3 events will potentially produce new flaws to be resolved while
@@
 window $[\tau+\lambda, \tau+\lambda+\pi]$ where:
+
 \begin{itemize}
+
 \item $\tau$ is the current tick
+
 \item $\lambda$ is the specified latency of this reactor. Which is an
@@
   before being able to produce a plan.\footnote{Do note that this
-    parameter remains indicative and, while it is highly recommended to
-    select a sound value, a failure to produce a plan in this delay is
-    not considered as a critical failure of the reactor}
+    parameter remains indicative and, while it is highly recommended
+    to select a sound value, a failure to produce a plan in this delay
+    is not considered as a critical failure of the reactor}
+
 \item $\pi$ is the planning look-ahead of the reactor. which indicates
   how many ticks in the future this reactor is looking ahead while planning.
+
 \end{itemize}
+
 This scope allow to filter out any tokens that are either necessarily
@@
 
-When there's no more flaw is present in the current planning scope a
+When there are no more flaw is present in the current planning scope a
 plan solution is found. This has two effects:
+
 \begin{enumerate}
+
 \item This reactor does not need to deliberate until the next synchronization.
+
 \item The {\em external} part of this plan can then be posted which
@@
   new deliberation on these reactors.
+
 \end{enumerate}
@@
 A core aspect of this reactor is that it takes advantage of the
-specific sequencing enforced by the T-REX architecture along with the
+specific sequencing enforced by the \rx architecture along with the
 fact that only one $plan$ structure is maintained in the reactor and
-shared between synchronization and planning to allow these two processes
-to not only be interleaved but impact each other through the way they
-manipulate this plan.
+shared between synchronization and planning to allow these two
+processes to not only be interleaved but impact each other through the
+way they manipulate this plan.
 
@@
 We first anlyse the case during which synchronization occurs after the
-planning has vcompleted (ie the last seep of deliberation resulted into
-a plan that is considered as complete for the current look-ahead). As
-synchronization starts with  the plan produced this will gave a frame
-for the search process. Therefore this deliberation will not only be a
-model-based estimation uniquely based on the {\em external}
-observations but will also include the reactor intentions as described
-by the plan produced. Consider for example the Shopping Agent model
-described during the europa presentation and that the
+planning has vcompleted (ie the last seep of deliberation resulted
+into a plan that is considered as complete for the current
+look-ahead). As synchronization starts with the plan produced this
+will gave a frame for the search process. Therefore this deliberation
+will not only be a model-based estimation uniquely based on the {\em
+  external} observations but will also include the reactor intentions
+as described by the plan produced. Consider for example the Shopping
+Agent model described during the europa presentation and that the
 location of the \texttt{agent} is an {\em external} timeline. For
 example as, the reactor add the goal to have \texttt{Own} milk and the
-current {\em external} states indicated that it was \texttt{At(home)} it
-produced the plan in Fig. \ref{fig:shop:exec0}. The  {\em external}
+current {\em external} states indicated that it was \texttt{At(home)}
+it produced the plan in Fig. \ref{fig:shop:exec0}. The {\em external}
 part of the plan -- namely the tokens \texttt{Going(home,
@@
 Now, consider that at the next tick $\tau+1$ the reactor managing the
-agent location was able to change the location state varaibale to 
-the \texttt{Going}. T-REX informs our reactor of this new observation
-which is added in the plan structure as a fact -- starting exactly at
+agent location was able to change the location state varaibale to the
+\texttt{Going}. \rx informs our reactor of this new observation which
+is added in the plan structure as a fact -- starting exactly at
 $\tau+1$ and lastign for an unlknown duration which is at least 1
@@
 propagated the information provided by the new observation in the
-exisiting plan (restricting in turn the start time of 
-\texttt{Going} and by propagtion of the constraints \texttt{Go} to  be
-{\em exactly} $\tau+1$). While focusing uniquely on the execution
-frontier ($\tau+1$ in this figure), the europa solver took benefit of
-the exisiting plan to evaluate a solution that asumes that the
-observation is consistent with what was decided  inside this
-reactor.
+exisiting plan (restricting in turn the start time of \texttt{Going}
+and by propagtion of the constraints \texttt{Go} to be {\em exactly}
+$\tau+1$). While focusing uniquely on the execution frontier ($\tau+1$
+in this figure), the europa solver took benefit of the exisiting plan
+to evaluate a solution that asumes that the observation is consistent
+with what was decided inside this reactor.
 
-This choice is directed not only by some natural heurisitc
-choices made in the solver but even more enforced in general by the
-heuristic order of the decision point reolution choices for our {\em
-  internla} state flaw presented in section \ref{sec:arch:sync}
-{\em\color{red} TODO: need to rework the presentation of thee choices
-  so I can refer to them more directly -- algorithm or pseudo theorem
-  should do the trick}. Should such a flaw be revealed during
-synchronization -- meaning that the current state of an {\em internal}
-state varaibale is not fully grounded -- the sequencing of the choices
-will atempt first a conservative choice in term of maintaining
-previous state, then an optimisitic choice in term of the
-execution/advance in our current plan to finally attempt other choices
-that were not predicted by the model. The two first possible choices
-are strongly directed by the exisiting plan which help direct the
-search within this frame. This often allow to have a synchronization
-that is much more directed and i, in nominal situation such as the one
-we showed resolve the deliberation in few steps (often without any
-backtrack in the search).
+This choice is directed not only by some natural heurisitc choices
+made in the solver but even more enforced in general by the heuristic
+order of the decision point reolution choices for our {\em internla}
+state flaw presented in section \ref{sec:arch:sync} {\em\color{red}
+  TODO: need to rework the presentation of thee choices so I can refer
+  to them more directly -- algorithm or pseudo theorem should do the
+  trick}. Should such a flaw be revealed during synchronization --
+meaning that the current state of an {\em internal} state varaibale is
+not fully grounded -- the sequencing of the choices will atempt first
+a conservative choice in term of maintaining previous state, then an
+optimisitic choice in term of the execution/advance in our current
+plan to finally attempt other choices that were not predicted by the
+model. The two first possible choices are strongly directed by the
+exisiting plan which help direct the search within this frame. This
+often allow to have a synchronization that is much more directed and
+i, in nominal situation such as the one we showed resolve the
+deliberation in few steps (often without any backtrack in the search).
 
@@
   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
-  \caption{Illustration of a conflict suring synchronization and the resulting relaxed plan for recovery}
+  \caption{Illustration of a conflict suring synchronization and the
+    resulting relaxed plan for recovery} 
   \label{fig:shop:relax}
@@
 As the plan is not anymore in the way the synchronization can resume
-and find a solution as we initial lly described. The newly pending goals will enforce  at
-the next \texttt{step} to resumed deliberation on this new fully
-relaxed partial plan. By the simple use of synchronization the reactor 
-was not only able to identify that its plan was not executable anymore 
-but recover from it while offering a new incomplete partial plan to be
-resolved during the next deliberation \texttt{steps}. This on its own
-ogfer the core functionality one should except to close the loop
-between planning and execution in the sense that synchronization plays
-the role of what is typically named as the {\em Executive} in other
-architectures {\em\color{red} references to diverse archis, 3-tiered,
-  laas, claraty, B Wiliams, and many more}.
+and find a solution as we initial lly described. The newly pending
+goals will enforce at the next \texttt{step} to resumed deliberation
+on this new fully relaxed partial plan. By the simple use of
+synchronization the reactor was not only able to identify that its
+plan was not executable anymore but recover from it while offering a
+new incomplete partial plan to be resolved during the next
+deliberation \texttt{steps}. This on its own ogfer the core
+functionality one should except to close the loop between planning and
+execution in the sense that synchronization plays the role of what is
+typically named as the {\em Executive} in other architectures
+{\em\color{red} references to diverse archis, 3-tiered, laas, claraty,
+  B Wiliams, and many more}.
 
-In our case though the interaction between planning and execution
-does not stop at this level. A core point of the overal schedulling of
-both {\em synchronization} and {\em planning} steps resides on our
+In our case though the interaction between planning and execution does
+not stop at this level. A core point of the overal schedulling of both
+{\em synchronization} and {\em planning} steps resides on our
 architecture enforsincing that the synchroniuzation process can occur
@@
 is planning.  This aspect trueluy contrast to other approaches we have
-idenitifed in the literature which avoid to perturb the planner during
-it search as it is not compatible with the ``off-line planning'' assumption.
+identified in the literature which avoid to perturb the planner during
+it search as it is not compatible with the ``off-line planning''
+assumption.
 
-By our design choices -- both at the architecture level and the integration of
-the europa framework for embedded deliberation -- we were able to
-implement an architecture that avoid this limitation while remaining in
-a formal frame that still allow for deliberation and model-based
-agent control. We believe that this tighter integration of planning
-and execution allow the system to ba more accute of its environament
-alowing better informaed decision and tighter reaction. This is
-possible while still integrating information such as a long latency
-for the reactors as long as one can ensure that synchronization of al
-the reactors and deliberation steps can be done in a time that is
-resonably small in comparison of the agent tick duration.
+By our design choices -- both at the architecture level and the
+integration of the europa framework for embedded deliberation -- we
+were able to implement an architecture that avoid this limitation
+while remaining in a formal frame that still allow for deliberation
+and model-based agent control. We believe that this tighter
+integration of planning and execution allow the system to ba more
+acute of its environment allowing better informaed decision and
+tighter reaction. This is possible while still integrating information
+such as a long latency for the reactors as long as one can ensure that
+synchronization of al the reactors and deliberation steps can be done
+in a time that is resonably small in comparison of the agent tick
+duration.
 

### 2012-02-07 arch.tex
Reintroduced reference to IDEA:   - this time in europa reactor section   - I need to read in depth AAAI fall symposium 2006 as they may share a lot of our ideas presented here

@@
 
-The T-REX architecture was designed with the objective to bring 
-planning closer to the low-level control of the vehicle while
-maintaining its reactivity. Indeed, in most of the architecture where
-task planning is embedded, this task remains at a failry high level of
-abstraction as the complexity of this task would otherwise impede the
-system reactiveness. As a result the planning problem managed {\em in
-  situ} remains fairly detached from low-level control and most of the 
+The \rx architecture was designed with the objective to bring planning
+closer to the low-level control of the vehicle while maintaining its
+reactivity. Indeed, in most of the architecture where task planning is
+embedded, this task remains at a failry high level of abstraction as
+the complexity of this task would otherwise impede the system
+reactiveness. As a result the planning problem managed {\em in situ}
+remains fairly detached from low-level control and most of the
 reactive adaptation which in contrast is managed by specialized
-component that relies on different representations for modeling 
-the local behavior they control or even do not have any explicit 
-formal model of this behavior. The two angles our architecture attck
-this issue are the following:
+component that relies on different representations for modeling the
+local behavior they control or even do not have any explicit formal
+model of this behavior. The two angles our architecture attack this issue 
+are the following:
+
 \begin{itemize}
+
 \item Distribute the global planning problem into multiple decision
@@
   larger that the world change rate.
+
 \end{itemize}
 
-For the first point T-REX provides a formal framework described in
-details in \cite{py10} and is not the focus of this paper. The
-second have deep implication on how on should integrate a planner in
-this framework in general and was applied to the europa framework
-specifically. We'll develop later how this requirement implies to see
-in-situ planning as two parallels processes that are tightly linked
-through the same plan structure. But before that we need to introduce
-at least the high leve lconcepts and ideas behind our architecture as
-a whole.
+For the first point \rx provides a formal framework described in
+details in \cite{py10} and \cite{rajan12} and is not the focus of this
+paper. The second have deep implication on how on should integrate a
+planner in this framework in general and was applied to the europa
+framework specifically. We'll develop later how this requirement
+implies to see in-situ planning as two parallels processes that are
+tightly linked through the same plan structure. But before that we
+need to introduce at least the high leve lconcepts and ideas behind
+our architecture as a whole.
 
-T-REX architecture is structured around the notion of 
-{\em state variables} as the unique basis of interactionalong
-with a well defined ownership  model for each of these. 
-In this framework each reactor is seen as a
-black box providing its own {\em internal} state variables for which
-its is reponsible to maintain its value at every single instant --
-represented by a {\em tick} with a fixed duration. In order to
-identify its state a reactor can subscribe to {\em external} state
-variables for  which it will received new updates when the owner of
-these will produce its {\em internal} state. This {\em
-  synchronization} process progate through the reactor dependency 
+\rx architecture is structured around the notion of {\em state
+  variables} as the unique basis of interactionalong with a well
+defined ownership model for each of these.  In this framework each
+reactor is seen as a black box providing its own {\em internal} state
+variables for which its is reponsible to maintain its value at every
+single instant -- represented by a {\em tick} with a fixed
+duration. In order to identify its state a reactor can subscribe to
+{\em external} state variables for which it will received new updates
+when the owner of these will produce its {\em internal} state. This
+{\em synchronization} process progate through the reactor dependency
 graph on a bottom up flow at every {\em tick} ensuring that all the
@@
 The choice of representation of state variables, observations and
-goals is directly derieved to the timleines and tokens from europa. 
-such representation is ich enough to allow the excahnge of state
+goals is directly derieved to the timelines and tokens from europa. 
+such representation is rich enough to allow the exchange of state
 information in a formal and yet flexible way and its transaltion in
-planning frameworks that have an explicit representation of time 
-(such as but not limited to europa) is often trivial.
+planning frameworks that have an explicit representation of time (such
+as but not limited to europa) is often trivial.
 
-The overall execution cycle of the agent is also abstracted as some
+The overall execution cycle of the agent is also abstracted as a
 continuous planning/deliberation cycle between all the
-reactors. Indeed , at the agent interface level, each reactor provided
+reactors. At the agent interface level, each reactor provided
 only two abstract calls that conceptualize their execution:
+
 \begin{itemize}
+
 \item \texttt{synchronize} that takes the last {\em external}
@@
 process which is focused on the future evolution of the state
-variables and it can take multiple steps spanning other multiple
-ticks for a reactor to produce a plan that is both complete and
-valid. when such plan is identified its {\em external} part is then
-dispatched to the corresponding reactors which in turn can start their
-deliberation \texttt{step}s in order to satisfy these new
-objectives. As a consequence physical execution will occur as these
-goals and planning propagate down in the reactor graph until they
-eventually become goals for the lowest level reactor which -- instead
-of deliberating -- may just transform this goal in a physical command
-execution.
+variables and it can take multiple steps spanning other multiple ticks
+for a reactor to produce a plan that is both complete and valid. when
+such plan is identified its {\em external} part is then dispatched to
+the corresponding reactors which in turn can start their deliberation
+\texttt{step}s in order to satisfy these new objectives. As a
+consequence physical execution will occur as these goals and planning
+propagate down in the reactor graph until they eventually become goals
+for the lowest level reactor which -- instead of deliberating -- may
+just transform this goal in a physical command execution.
 
 The way the agent interleave planning \texttt{step}s and
-synchronization is illustrated in Figure \ref{fig:tick-exec}.
-Synchronization happens at every single tick and will interrupt
-planning allowing this reactor to identify its current state that can
-propagate through the reactor graph in order to ensure t aconsitent
-view of the state of the world between all the reactors.  
+synchronization for a single reactor is illustrated in Figure 
+\ref{fig:tick-exec}. Synchronization happens at every single 
+tick and will interrupt planning allowing this reactor to identify
+ its current state that can propagate through the reactor graph
+ in order to ensure t aconsitent view of the state of the world 
+between all the reactors.
+
 \begin{figure}[!htbp]
@@
 \end{figure}
-This view of the reactor has two parallel tasks aloow for long
+
+This view of the reactor has two parallel tasks allow for long
 planning deliberation while ensuring that the state evolution
@@
 
-The europa reactor is our implementation inside T-REX of a reactor
-that usess automated planning for its execution. This reactor is fully
-based on the europa framework and applies the core idea we of T-REX we
-described above to implement a reactor for which the sole purpse is to
-deliberate as time advance. The choice of europa as the planning
-framework was directed by the fact thta this framework provide  a rich
-and highly configurable framework to be able to infer around a rich
-temporal domain.
+The \eu reactor is our implementation inside \rx of a reactor that
+usess automated planning for its execution. This reactor is fully
+based on the europa framework and applies the core idea of \rx we
+described above to implement a reactor for which the sole purpose is to
+deliberate as time advance. The design of this reactor takes cues from 
+the IDEA architecture presented on \cite{mus02, mus06} in the sense taht both
+planning and execution are based on a single unified model. The choice 
+of \eu as the planning framework was directed by the fact that this 
+framework provide a rich and highly configurable toolset to be able to 
+infer around a rich temporal domain.
 
-This reactor is implemented following the idea that in T-REX, the
+This reactor is implemented following the idea that in \rx, the
 whole execution cycle is two deliberation processes. Both of these
 processes are implemented as two europa solvers modifying the same
-plan database and for which the execution is managed by the T-REX 
+plan database and for which the execution is managed by the \rx 
 framework as time advance during the execution of the system:
+
 \begin{enumerate}
+
 \item The synchronization solver is a specialized europa solver that
@@
 typical robot this implies that the only point where the planner can
-be informed about the current world state is righe before the planning
+be informed about the current world state is right before the planning
 is started by giving to the planner the intial state and current
@@
 \cite{lemai04, lemai-chenevier2004}, they do attempt to reduce the
-impact of planning by aloowing local plan repairs which attempt to
+impact of planning by allowing local plan repairs which attempt to
 insert repairing actions in the exisiting plan andf can be done fast

### 2012-02-08 trunk/concepts.tex
next iteration of EUROPA's plan representation, still need to cleanup Token State model and Rules.

@@
   instances where variables representing time would be used. 
+  As explained in the CP section, a variable can take values from a domain. In \eu, domains are defined over a specific data type, the primitive data types supported are:
+  \begin{enumerate}
+    \item \textit{String}: sequences of characters, like "Red", "Spirit", etc 
+    \item \textit{Boolean}: {True,False}
+    \item \textit{Numeric}: integers and floats
+    \item \textit{Object}: reference to an object instance
+  \end {enumerate}
+  Two types of variable Domains can be defined over a data type: 
+  \begin{enumerate}
+    \item \textit{Enumerated}: Finite sets over any data type, specified explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
+    \item \textit{Interval}: Only for Numeric data types, they are specified by a [LowerBound,UpperBound] pair, for instance: [1,10], [5.0,100.0], etc.
+  \end {enumerate}
+  \eu's ability to represent and reason over numeric intervals is a very useful characteristic that allow users to create flexible plans. \comment{say more about the benefits of flexibility?}
 
@@
     token's current state and its reachable states through further
-    restriction.  This is \eus mechanism to support the CAIP approach.
+    restriction.  This is \eu's mechanism to support the CAIP approach.
 
+  \end{enumerate}
+
+\item[\textbf{Built-in Object Types}] In a domain model, there may be some classes that don't have any time-dependent state or behavior, and therefore those classes will not have any Token types associated with them, for instance in the Shopping Agent model, the set locations and products are static for the agent's purposes so we may only want to say what they are but not have any state or behavior associated with them.
+
+However, the most common case in any non-trivial domain model is that Objects will have many Tokens  in order to describe their state and behavior throughout a plan.
+There are a couple of Object Types that are so commonly used in domain descriptions, that \eu provides a built-in implementation for them:
+
+\begin{enumerate}
+    \item \textit{Timelines}: Often objects in a domain must be described by exactly one
+    token for every given timepoint in the plan. \eu provides a built-in Timeline class, any instances of a class derived from a
+    Timeline will induce ordering requirements among its tokens in order
+     to ensure no temporal overlap may occur among them  \cite{mus94}. 
+     \comment{add small example?}
+
+    \item \textit{Resources}: Metric resources, e.g. the energy of a battery or the
+    capacity of a cargo hold, are objects with an explicit quantitative
+    state in time and with a circumscribed range of changes that can occur
+     to impact that state i.e. produce, consume, use, change. 
+     Resources are such a common requirement for \eu users that built-in object types (with their corresponding token types to denote production, consumption, etc) are provided for them.
+Instances of classes derived from a Resource will induce
+     ordering requirements on their Tokens in order to ensure that the
+     level of the resource remains within specified limits.
+     \comment{add small example?}
 \end{enumerate}
@@
 
-% \textit{Objects (continued)}
-
-% As we have noted, Tokens describe some aspect of an Object in
-% time. Objects thus may have many Tokens in a plan in order to describe
-% their state and behavior throughout all points in time of the
-% plan. Within this general framework, we note a few particulars:
-
-% Static Facts: Classes without predicates lead to Objects without
-% Tokens. This arises where an objects state or behavior is independent
-% of time. For example, a domain may have a set of locations and/or
-% paths for which there is nothing more to say than that they exist.
-
-% Timelines: Often objects in a domain must be described by exactly one
-% token for every given timepoint in the plan. Such objects are so
-% common that we provide a special construct in \eu to extend these
-% semantics to derived classes. Any instances of a class derived from a
-% Timeline will induce ordering requirements among its tokens in order
-% to ensure no temporal overlap may occur among them. See below for
-% further discussion.
-
-% Resources: Metric resources, e.g. the energy of a battery or the
-% capacity of a cargo hold, are objects with an explicit quantitative
-% state in time and with a circumscribed range of changes that can occur
-% to impact that state i.e. produce, consume, use, change. These changes
-% are captured as tokens. Resources are such a common requirement for
-% \eu users that special constructs are also provided for
-% them. Instances of classes derived from a Resource will induce
-% ordering requirements on their Tokens in order to ensure that the
-% level of the resource remains within specified limits.
-
-\item[\textbf{Timelines}] It may be sufficient to maintain only a
-  partial-order among tokens. However, it is often the case that
-  tokens represent states and actions of a single object in the
-  system. Such tokens are typically mutually-exclusive. \eu uses a
-  Timeline structure \cite{mus94} to concisely capture system
-  components whose behavior is described over time in this manner.
-
-\end{description}
-
-For example, consider the tire-world domain. The tire was located in
-the trunk with the predicate tireLocated(Trunk) and located on the
-ground with the predicate tireLocated(Ground). Clearly, the same tire
-cannot be in both places at once. A Timeline provides a simple method
-for aggregating the statements about a tire such that they are
-mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the
-whereabouts of a tire. The tire is an object in the tire-world
-represented as a timeline. It has predicates associated with it which
-can describe its states and actions over time. The predicate names
-need not contain the tire prefix; this is now implicit since the
-tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the \eu planning
-paradigm, they are not essential. A more general notion of a system
-Object can be used which does not impose restrictions of
-mutual-exclusion or non-zero duration. This can be important in
-supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3
-contiguous tokens. The end and start time-points are related by an
-equality constraint. A Moving predicate has been introduced to cover
-the transition from one location to another. It takes 2 location
-arguments. Notice that precedence relationships exist between tokens
-such that they cannot overlap but the start and end times may remain
-flexible. To illustrate this, sample times are included. Assume a
-total time range of interest between 0 and 1000. In addition, tokens
-on Timelines have a minimum duration of 1. As a result the values
-shown are the most flexible possible values for the start and end of
-each token. There are a number of advantages of allowing this
-flexibility. First, the basic structure of the plan can be developed
-without over-committing to specific times. If it is not necessary for
-the tire to be on the ground at time-step 4, then a planner should not
-be forced to specify it. Such an approach permits a least-commitment
-approach to planning. Second, in many domains it is simply impossible
-to know in planning exactly how long an activity might take. For
-example, a common activity of driving a car from one point to another
-on a road can take varying amounts of time depending on traffic and
-traffic lights. In such circumstances it is more practical to express
-upper and lower bounds on durations which naturally lead to intervals
-for start and end times. In such domains, flexibility in planning aids
-robustness in execution.  Rules
-
-In order for a plan to be valid, it must comply with all rules and
-regulations pertinent to the application domain in question. Rules
-govern the internal and external relationships of a token. For
-example, consider a parameterized predicate describing a transition
-from one location to another. Let the parameters be from and to. The
-parameters are instantiated on a token as variables whose domain of
-values is the set of all locations in a given problem. A rule
-governing an internal relationship among token variables might
-stipulate that a transition must involve a change in location. This
-can be easily expressed as a constraint on the definition of a
-predicate of the form: from != to. It is reasonable to further
-stipulate that one must be Located somewhere before a transition can
-occur, and one must end up Located somewhere when completed. This is
-an example of a rule governing an external relationship among
-tokens. It specifies a requirement that tokens of the predicate
-Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving
-Figure 2 illustrates the entities and relations involved in specifying
-such a rule on a Moving predicate. The token on which the rule applies
-is referred to as the master. Each Located token required by the
-master is referred to as a slave. All variables are indicated by name
-and their domains are expressed as intervals in the case of temporal
-variables and as enumerations for the remainder. Application of a rule
-on a token can thus cause slave tokens, variables, and constraints to
-be introduced.  Token State Model
-
-The capability of a domain rule to cause a slave token to be created
+\item[\textbf{Token State Model}] The capability of a domain rule to cause a slave token to be created
 is a key vehicle through which planning occurs. Semantically, this
@@
 
-Each of the token variables, including the parameter variables, has a
-domain of values assigned to it.  The variables may also participate
-in constraints that specify which value combinations are valid.  To
-allow the specification of multiple values, e.g, to express a range of
-possible start times, variables are used to specify parameter, start
-and end time values for a token.  As a result, a token $T$ is a tuple
-$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
-variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
-$s\leq e$) are variables encoded as flexible token start and end time
-points. Consequently, partial plans represent a \textit{range of
-  possible solutions} rather than an explicit definition of a single
-trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
-of a part of such plans with tokens representing constraints on
-flexible temporal intervals. Systematic computation is reflected by
-the use of the \texttt{Allen Algebra} \cite{allen84} temporal
-relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
-interval arithmetic in the \eu domain model that allows
-encoding of temporal constraint relationships between tokens on the
-same or other concurrent timelines. Details of propagation algorithms
-or of interval logic are beyond the scope of this paper.
+\item[\textbf{Rules}] In order for a plan to be valid, it must comply with all rules and
+regulations pertinent to the application domain in question. Rules
+govern the internal and external relationships of a token. For
+example, consider a parameterized predicate describing a transition
+from one location to another. Let the parameters be from and to. The
+parameters are instantiated on a token as variables whose domain of
+values is the set of all locations in a given problem. A rule
+governing an internal relationship among token variables might
+stipulate that a transition must involve a change in location. This
+can be easily expressed as a constraint on the definition of a
+predicate of the form: from != to. It is reasonable to further
+stipulate that one must be Located somewhere before a transition can
+occur, and one must end up Located somewhere when completed. This is
+an example of a rule governing an external relationship among
+tokens. It specifies a requirement that tokens of the predicate
+Located precede and succeed tokens of the predicate Moving.
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
-\caption{\small Timeline evolution can be described by simple finite
-  state transitions where temporal relationships describe the arc
-  transition and temporal extant of state are token
-  durations. Constraints here are for \emph{meets} or its equivalent
-  inverse \emph{met\_by}.}
-\label{fig:FSM}
-\vskip-0.3cm
-\end{figure}
+Figure 2: Internal and external relationships from rules on Moving
+Figure 2 illustrates the entities and relations involved in specifying
+such a rule on a Moving predicate. The token on which the rule applies
+is referred to as the master. Each Located token required by the
+master is referred to as a slave. All variables are indicated by name
+and their domains are expressed as intervals in the case of temporal
+variables and as enumerations for the remainder. Application of a rule
+on a token can thus cause slave tokens, variables, and constraints to
+be introduced.
 
-In describing the evolution of state via timelines, transitions within
-a timeline can be described by simple finite state
-machines. Fig. \ref{fig:FSM} for example describes a timeline
-evolution for an abstract timeline entity describing navigation. Arc
-transitions represent temporal constraints and token duration are
-represented by temporal extant of state. The entire vehicle can then
-be modeled as a number of state transitions with concurrency for each
-of the modeled state variables defined a priori. However this simple
-view is often complicated when temporal constraints between state
-variables are articulated in our domain models. If there is a
-criticism of this approach, modeling then, requires a proper balance
-between abstraction, state variable description and concurrency
-described by the inter-connectedness between such finite state
-machines.
+\end{description}
+
+\comment{instead of summary, tie every concept to the shopping example?}
+
+This section has presented the main elements of \eu's planning representation:
+
+\begin{enumerate}
+    \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+    states and actions in time. Note that tokens do not discriminate between state and action.
+
+    \item Token States which are the basis of planning operations on a partial plan.
+
+    \item Constraints to describe relationships among tokens. This provides an
+    expressive method of describing interactions among tokens in the context of Temporal Planning.
+
+    \item Timelines as a concise abstraction to express the evolution of state
+    and behavior for a system component. It provides semantics of mutual
+    exclusion sequencing in time. Other core abstractions are available in
+    \eu for handling metric resources.
+
+    \item Domain rules to describe internal and external relationships on and
+     between tokens respectively. Rules are applied to active tokens (master tokens) 
+     referred to as masters and typically produce inactive
+     tokens (slaves) which are a key vehicle through which the planning
+     process evolves. The rule structure in \eu differs from the more
+     restrictive commitment to preconditions and effects in classical
+     planning which prohibits durative actions and disembodied effects
+     (i.e. effects which may not occur until some temporal distance after
+     the end of the action).
+\end{enumerate}
+
+%Each of the token variables, including the parameter variables, has a
+%domain of values assigned to it.  The variables may also participate
+%in constraints that specify which value combinations are valid.  To
+%allow the specification of multiple values, e.g, to express a range of
+%possible start times, variables are used to specify parameter, start
+%and end time values for a token.  As a result, a token $T$ is a tuple
+%$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
+%variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
+%$s\leq e$) are variables encoded as flexible token start and end time
+%points. Consequently, partial plans represent a \textit{range of
+%  possible solutions} rather than an explicit definition of a single
+%trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
+%of a part of such plans with tokens representing constraints on
+%flexible temporal intervals. Systematic computation is reflected by
+%the use of the \texttt{Allen Algebra} \cite{allen84} temporal
+%relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
+%interval arithmetic in the \eu domain model that allows
+%encoding of temporal constraint relationships between tokens on the
+%same or other concurrent timelines. Details of propagation algorithms
+%or of interval logic are beyond the scope of this paper.
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
+%\caption{\small Timeline evolution can be described by simple finite
+%  state transitions where temporal relationships describe the arc
+%  transition and temporal extant of state are token
+%  durations. Constraints here are for \emph{meets} or its equivalent
+%  inverse \emph{met\_by}.}
+%\label{fig:FSM}
+%\vskip-0.3cm
+%\end{figure}
+%
+%In describing the evolution of state via timelines, transitions within
+%a timeline can be described by simple finite state
+%machines. Fig. \ref{fig:FSM} for example describes a timeline
+%evolution for an abstract timeline entity describing navigation. Arc
+%transitions represent temporal constraints and token duration are
+%represented by temporal extant of state. The entire vehicle can then
+%be modeled as a number of state transitions with concurrency for each
+%of the modeled state variables defined a priori. However this simple
+%view is often complicated when temporal constraints between state
+%variables are articulated in our domain models. If there is a
+%criticism of this approach, modeling then, requires a proper balance
+%between abstraction, state variable description and concurrency
+%described by the inter-connectedness between such finite state
+%machines.
+
+%For example, consider the tire-world domain. The tire was located in
+%the trunk with the predicate tireLocated(Trunk) and located on the
+%ground with the predicate tireLocated(Ground). Clearly, the same tire
+%cannot be in both places at once. A Timeline provides a simple method
+%for aggregating the statements about a tire such that they are
+%mutually exclusive.
+%
+%Figure 1: A Timeline for a Tire
+%
+%Figure 1 illustrates how a Timeline can be used to describe the
+%whereabouts of a tire. The tire is an object in the tire-world
+%represented as a timeline. It has predicates associated with it which
+%can describe its states and actions over time. The predicate names
+%need not contain the tire prefix; this is now implicit since the
+%tokens are assigned to a specific tire instance (i.e. an instance of a
+%Timeline). While Timelines are a useful element of the \eu planning
+%paradigm, they are not essential. A more general notion of a system
+%Object can be used which does not impose restrictions of
+%mutual-exclusion or non-zero duration. This can be important in
+%supporting partial-order planning.
+%
+%For this example, the state of the tire is specified using 3
+%contiguous tokens. The end and start time-points are related by an
+%equality constraint. A Moving predicate has been introduced to cover
+%the transition from one location to another. It takes 2 location
+%arguments. Notice that precedence relationships exist between tokens
+%such that they cannot overlap but the start and end times may remain
+%flexible. To illustrate this, sample times are included. Assume a
+%total time range of interest between 0 and 1000. In addition, tokens
+%on Timelines have a minimum duration of 1. As a result the values
+%shown are the most flexible possible values for the start and end of
+%each token. There are a number of advantages of allowing this
+%flexibility. First, the basic structure of the plan can be developed
+%without over-committing to specific times. If it is not necessary for
+%the tire to be on the ground at time-step 4, then a planner should not
+%be forced to specify it. Such an approach permits a least-commitment
+%approach to planning. Second, in many domains it is simply impossible
+%to know in planning exactly how long an activity might take. For
+%example, a common activity of driving a car from one point to another
+%on a road can take varying amounts of time depending on traffic and
+%traffic lights. In such circumstances it is more practical to express
+%upper and lower bounds on durations which naturally lead to intervals
+%for start and end times. In such domains, flexibility in planning aids
+%robustness in execution.
 
@@
 
-%*********
-
-Timelines
-
-It may be sufficient to maintain only a partial-order among
-tokens. However, it is often the case that tokens represent states and
-actions of a single object in the system. Such tokens are typically
-mutually-exclusive. \eu uses a Timeline structure developed in (??)
-to concisely capture system components whose behavior is described
-over time in this manner.
-
-\comment{This is the first instance that this domain is being
-  introduced. Either we need more explanation or it ought to be
-  removed.} For example, consider the tire-world domain. The tire was
-located in the trunk with the predicate tireLocated(Trunk) and located
-on the ground with the predicate tireLocated(Ground). Clearly, the
-same tire cannot be in both places at once. A Timeline provides a
-simple method for aggregating the statements about a tire such that
-they are mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the
-whereabouts of a tire. The tire is an object in the tire-world
-represented as a timeline. It has predicates associated with it which
-can describe its states and actions over time. The predicate names
-need not contain the tire prefix; this is now implicit since the
-tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the \eu planning
-paradigm, they are not essential. A more general notion of a system
-Object can be used which does not impose restrictions of
-mutual-exclusion or non-zero duration. This can be important in
-supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3
-contiguous tokens. The end and start time-points are related by an
-equality constraint. A Moving predicate has been introduced to cover
-the transition from one location to another. It takes 2 location
-arguments. Notice that precedence relationships exist between tokens
-such that they cannot overlap but the start and end times may remain
-flexible. To illustrate this, sample times are included. Assume a
-total time range of interest between 0 and 1000. In addition, tokens
-on Timelines have a minimum duration of 1. As a result the values
-shown are the most flexible possible values for the start and end of
-each token. There are a number of advantages of allowing this
-flexibility. First, the basic structure of the plan can be developed
-without over-committing to specific times. If it is not necessary for
-the tire to be on the ground at time-step 4, then a planner should not
-be forced to specify it. Such an approach permits a least-commitment
-approach to planning. Second, in many domains it is simply impossible
-to know in planning exactly how long an activity might take. For
-example, a common activity of driving a car from one point to another
-on a road can take varying amounts of time depending on traffic and
-traffic lights. In such circumstances it is more practical to express
-upper and lower bounds on durations which naturally lead to intervals
-for start and end times. In such domains, flexibility in planning aids
-robustness in execution.  Rules
-
-In order for a plan to be valid, it must comply with all rules and
-regulations pertinent to the application domain in question. Rules
-govern the internal and external relationships of a token. For
-example, consider a parameterized predicate describing a transition
-from one location to another. Let the parameters be from and to. The
-parameters are instantiated on a token as variables whose domain of
-values is the set of all locations in a given problem. A rule
-governing an internal relationship among token variables might
-stipulate that a transition must involve a change in location. This
-can be easily expressed as a constraint on the definition of a
-predicate of the form: from != to. It is reasonable to further
-stipulate that one must be Located somewhere before a transition can
-occur, and one must end up Located somewhere when completed. This is
-an example of a rule governing an external relationship among
-tokens. It specifies a requirement that tokens of the predicate
-Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving
-Figure 2 illustrates the entities and relations involved in specifying
-such a rule on a Moving predicate. The token on which the rule applies
-is referred to as the master. Each Located token required by the
-master is referred to as a slave. All variables are indicated by name
-and their domains are expressed as intervals in the case of temporal
-variables and as enumerations for the remainder. Application of a rule
-on a token can thus cause slave tokens, variables, and constraints to
-be introduced.  Token State Model
-
-The capability of a domain rule to cause a slave token to be created
-is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting tokens to be in the plan in
-order for the master to be valid. There are 2 possibilities to
-consider:
-
-The slave is inserted as an active token in the plan. As such, rules
-may be activated on the slave, and it may consume resources.
-
-The token is merged with a matching token already in the plan. Once a
-slave is merged, the requirement it represents are considered
-satisfied. The process of merging passes on all restrictions imposed
-on the slave to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. For an active token and an inactive token to be
-compatible requires that they are instances of the same predicate and
-that no intersections between corresponding variables are empty. The
-effect of merging is illustrated in Figure 3.
-
-Figure 3: Merging an inactive token on an active token. Domain
-restrictions occur in highlighted variables of active token.
-
-On creation of a token, where the commitment has not been made yet to
-activate or merge the token, the token is said to be inactive. There
-is a nuance to the state model for tokens which relate to the mode of
-its creation. If the token is allocated explicitly by an external
-actor, rather than internally through rule firing, it may include the
-state rejected indicating the planner is permitted to reject the
-token. A valid plan can include rejected tokens. This typically arises
-where the token represents a goal that is preferable to achieve but
-not mandatory. This state is not reachable for slaves since that would
-imply selective adherence to the domain model. Figure 4 presents the
-state transition diagram for token states relevant for planning. The
-transitions are operations on a partial plan which can decide an
-outcome for an inactive token. As operations which may arise in
-search, they must be reversible during backtracking. The cancellation
-operations in each case are also shown.
-
-Figure 4: Token States and Transitions for Planning
-
-The state of a token is embodied with a 5th and final built-in
-variable referred to as the state variable. The planner states are
-values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
-INACTIVE state is captured by the variable being unbound. If a value
-is removed from the domain, that state will not be reachable. For
-example, when a slave is created via a rule, the REJECTED value is
-removed from its state variable.
-
-% \textit{Summary}
-
-% This has presented the main elements of the EUROPA planning approach. They are:
-
-% \begin{enumerate}
-
-% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
-% states and actions in time. Note that tokens do not discriminate
-% between state and action.
-
-% \item Token States which are the basis of planning operations on a partial
-% plan.
-
-% \item Constraints to describe relationships among tokens. This provides an
-% expressive method of describing interactions among tokens in the
-% context of Temporal Planning.
-
-% \item Timelines as a concise abstraction to express the evolution of state
-% and behavior for a system component. It provides semantics of mutual
-% exclusion sequencing in time. Other core abstractions are available in
-% \eu for handling metric resources.
-
-% \item Domain rules to describe internal and external relationships on and
-% between tokens respectively. Rules are applied to active tokens
-% (master tokens) referred to as masters and typically produce inactive
-% tokens (slaves) which are a key vehicle through which the planning
-% process evolves. The rule structure in \eu differs from the more
-% restrictive commitment to preconditions and effects in classical
-% planning which prohibits durative actions and disembodied effects
-% (i.e. effects which may not occur until some temporal distance after
-% the end of the action).
-
 % \end{enumerate}

### 2012-02-08 arch.tex
Adapted illustration of model ties to the future to the Shopping example

@@
 domains one the reactor can support while remaining complete. Take for
-example the following model rule where {\tt Vehicle} is {\em Internal}
-and {\tt Command} is {\em External} to the reactor.
+example this rule from our \texttt{Shopping} model :
+\begin{verbatim}
+ 1 Agent::Go {
+ 2   met_by(condition object.location.At origin);
+ 3   eq(from, origin.loc);
+ 4
+ 5   equals(effect object.location.Going going);
+ 6   eq(going.from, from);
+ 7   eq(going.to, to);
+ 8   
+ 9   meets(effect object.location.At destination);
+10   eq(to, destination.loc);
+11 }
+\end{verbatim}
+While this model is perfectly standard and acceptable in the overall
+case. It can become problematic while tied to the
+execution. Specificially it puts a strong tie between the {\em fact} that we
+are \texttt{Going} to a location and the {\em expected} future outcome  of
+ending at this location. This ties is even stronger due to the
+constraint at line 5. Knwoing this aspect assume that the
+\texttt{Agent.location} is {\em external} to the reactor which receive
+the observation \texttt{Going(Home,SuperMarket)} which synchronization
+resolved by producing the observation \texttt{Go(Home, SuperMarket)}
+in the Agent timeline. From this point this means that the only
+possible outcome on \texttt{Agent.location} is to be
+\texttt{At(SuperMarket)} in the directly foreseeable future. In the
+real world the fact that we are \texttt{Going} to a certain location
+does not necessarly imply that we will succeed on ending up there. An
+extreme case could be that due to some exogenous event, the road to
+the \texttt{SuperMarket} is closed which could result on an alternate
+observation \texttt{At(7th st. \& Foo Ave.)}. This observation
+contradicts thre existing plan the reactor maintin and it is important
+here to relaize the temp[oral direction of the causal link that is
+triggereing this inconsitency. Indeed at the time we will observe this
+contradicting \texttt{At}, the rule that generated this conlict is
+connected to the past observations \texttt{Going} and \texttt{Go}
+which cannot be fully retracted without potential global impact in the
+rest of the exisiting reactors -- implying  a huge cost in
+backprogation that would be problematic.
 
+More generally tying a token
+to future outcomes in a model that is meant to be confronted to rela
+executionis something that is problematic. It may be valid ifg you can
+gurantee that this will always happen (for example the reactor
+managing \texttt{Agent.location} timeline guarantees that the outcome
+of the \texttt{Going} will {\em always} result on ending at the traget
+location) but should be manipulated while keeping in mind that
+predicting the future is not absolute. In this case one could alter
+this part of the model as follow in order to give the reactor more
+flexibility:
 \begin{verbatim}
-Vehicle::Dive {
-  contains(Command.Descend descend);
-  descend.start = start + 2;
-}
+ 1 Agent::Go {
+ 2   met_by(condition object.location.At origin);
+ 3   eq(from, origin.loc);
+ 4
+ 5   contains(effect object.location.Going going);
+ 6   eq(going.from, from);
+ 7   eq(going.to, to);
+ 8   
+ 9   meets(effect object.location.At destination);
+10   eq(to, destination.loc);
+11   going meets destination;
+12 }
 \end{verbatim}
-
-Such model presents the issue that the reactor needs at one point to
-take the decision to start the token {\tt Vehicle.Dive} and by doing
-so it assumes that in {\em exactly} 2 ticks in the future the {\tt
-  Command} timeline will change its state to {\tt Command.Descend}. If
-this observation does not occur at the given time this implies that
-the reactor estimation was wrong leading him to an inconsistent view
-of the world. Conversely the reactor can decide to not switch this
-timeline to the {\tt Vehicle.Dive} state and observe 2 ticks in the
-future that the {\tt Command.Descend} state value. In this case it
-would have missed the opportunity to set its {\tt Vehicle} state to
-the correct value which may similarly lead (depending on the rest of
-the model) to a similar inconsistency it won't be able to resolve (as
-TREX enforces that the past is monotonic).
-
-While this limitation of the system is important to take into account,
-it is also important to note that the model snippet we gave here --
-while formally acceptable in the general planning problem -- is also
-illustrating a bad design for execution in a situated agent. Indeed,
-in this model we are tying the current decision to a future outcome
-that is {\em External} to the reactor and therefore not necessarily
-controllable by this reactor. Therefore we are entailing a part of the
-future to past decisions which can lead the system to a dead-end
-situation where the model cannot find a correct solution that both do
-not conflict with what we claimed in the past and still allow the
-planner to find its current {\em Internal} state.
+The alteration made to relax the constraint on linbe 5 from a strict 
+\texttt{equals} to contains and then add the new constraint at line 11
+that ensure that the last \texttt{Going} done during this \texttt{Go}
+ends up in the location the agent wnats to be. This firts level of
+relaxation will therefore allow the agent to usies multiple
+\texttt{Going} should the first one fail to reach our objective.  The
+probalem can even be relaxed further if needed but this relaxation
+gicves already more felxibility for the reactor to recover from local
+failures.
 

### 2012-02-09 trunk/results.tex
updates

@@
 sparse, patchy in large horizontal scales (in Kms) and small vertical
-scales (in meters). Fig. \ref{inl} shows one example from an AUV
+scales (in meters). Fig. \ref{fig:inl} shows one example from an AUV
 transect from 2005. The intent of both observation and sample
@@
 \subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
-\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with the
-  \od in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
-  device.}
-xf\label{fig:odss2}
+\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
+  the \od in use. Fig. \ref{fig:odss-virtual} shows a drifter tracking
+  tool used during \can planning; it used a physical model of advected
+  surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
+  deployments. Fig. \ref{fig:odss-ipad} shows \od use on a portable
+  iPad device.}
+\label{fig:odss2}
 \end{figure*}

### 2012-02-09 concepts.tex
cleaned up section on TokenStates, moving to section on Rules now.

@@
 
-  \item \textit {duration}: The constraint \textit{start} $+$
+  \item \textit {duration}: The duration of the temporal extent over which the predicate is defined. The constraint \textit{start} $+$
     \textit{duration} = \textit{end} is enforced automatically.
@@
     token's current state and its reachable states through further
-    restriction.  This is \eus mechanism to support the CAIP approach.
+    restriction.
+This is required to support CAIP's approach to planning where goals and subgoals can be satisfied by either adding new intervals (Tokens) to the plan, or by merging with existing intervals.
+Token State lifecycle is examined in detailed in a separate section below.
 
@@
 
+\item[\textbf{Built-in Object Types}] In a domain model, there may be some classes that don't have any time-dependent state or behavior, and therefore those classes will not have any Token types associated with them, for instance in the Shopping Agent model, the set locations and products are static for the agent's purposes so we may only want to say what they are but not have any state or behavior associated with them.
+
+However, the most common case in any non-trivial domain model is that Objects will have many associated Tokens in order to describe their state and behavior throughout a plan.
+There are a couple of Object Types that are so commonly used in domain descriptions, that \eu provides a built-in implementation for them:
+
+\begin{enumerate}
+    \item \textit{Timelines}: Often objects in a domain must be described by exactly one
+    token for every given timepoint in the plan. Any instances of a class derived from \eu's built-in
+    Timeline class will induce ordering requirements among its tokens
+     to ensure no temporal overlap may occur among them  \cite{mus94}. 
+     \comment{add small example?}
+
+    \item \textit{Resources}: Metric resources, e.g. the energy of a battery or the
+    capacity of a cargo hold, are objects with an explicit quantitative
+    state in time and with a circumscribed range of changes that can occur
+     to impact that state i.e. produce, consume, use, change. 
+     Resources are such a common requirement for \eu users that built-in object types (with their corresponding token types to denote production, consumption, etc) are provided for them.
+Instances of classes derived from a Resource will induce
+     ordering requirements on their Tokens in order to ensure that the
+     level of the resource remains within specified limits.
+     \comment{add small example?}
+\end{enumerate}
+
+\item[\textbf{Token State Model}] In \eu's representation, goals are temporally scoped states that need to be achieved, or actions that must be performed.
+As a result, \eu internally represents goals as Tokens.
+Goals can be posted explicitly, for instance, when a user asks the Shopping Agent to own a specific product by a specific time, or implicitly when subgoals are created as a result of domain compatibilities (described in the CAIP framework above), for instance, when the Shopping Agent must get to a Location where a Product is available before it is able to purchase it, in this case the original goal of purchasing the product will spawn a subgoal of being at a specific location.
+
+Looking at a how a simple Shopping Agent goal would be addresed by \eu helps illustrate how Token States support the planning process.
+
+Let's assume we start with a partial plan that places the Shopping agent at home at time 0 and with no Products in its possession, then we post a goal for the Shopping Agent to own a Drill at time 10.
+
+\comment{Add figure showing an AgentLocation timeline with an At(Home) token with start=0 and end open, an empty AgentBag timeline, and a Own(Drill) token with start=10 floating below both timelines}
+
+Notice that the \texttt{Own(Drill)} token that represents the goal is not immediately placed on the \texttt{AgentBag} timeline since \eu must determine whether it can achieve that goal and how.
+When the goal is posted and before any decisions are made on it, the  \texttt{Own(Drill)} token is created in an INACTIVE state, then \eu looks at the current partial plan and considers 2 alternatives:
+
+\begin{enumerate}
+    \item If there are any tokens in the plan that are compatible with the new goal, \eu may attempt to satisfy the goal by merging it with any of them (if there is more than one it can try them one at a time), in this case, the state of the token would change to MERGED.
+    \item Instead of merging with existing tokens in the plan, \eu may decide to insert the new token into the plan, in that case its state would change to ACTIVE.
+\end{enumerate}
+
+When a token is merged, all restrictions imposed
+on the merged token are transferred to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. Two tokens are merge-compatible if they are instances of the same predicate and
+that no intersections between corresponding variables are empty. The
+effect of merging is illustrated in Figure 3.
+
+\comment{Add figure showing the effect of merging}
+
+When a token is activated all compatibilities associated with its token type are evaluated, which may lead to new subgoals being generated.
+In our example, there should be a compatibility stating that the Shopping Agent must be at the same location as the product it is purchasing.
+In that case, activating the \texttt{Own(Drill)} token will cause a new  \texttt{At(Drill.location)} to be created in an INACTIVE state, which in turn will have to be dealt with.
+
+\comment{Add figure showing the effect of activating the token by putting it on the AgentBag timeline, and a new At(Drill.location) token floating below}
+
+If neither activating nor merging works, a token may become REJECTED, this is possible only for goals posted by the end user and specified as optional.
+Subgoals generated as a result of token activation cannot be rejected, if a subgoal cannot be satisfied through activation or merging then it will cause the original activation decision that spawned it to be retracted (\eu's search mechanisms are explained in detail in a separate section later on).
+
+As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED or REJECTED, this lifecycle is illustrated in the figure below.
+
+\comment{Add figure showing possible state transitions}
+
 \comment {resume cleanup here}
 
-% \textit{Objects (continued)}
-
-% As we have noted, Tokens describe some aspect of an Object in
-% time. Objects thus may have many Tokens in a plan in order to describe
-% their state and behavior throughout all points in time of the
-% plan. Within this general framework, we note a few particulars:
-
-% Static Facts: Classes without predicates lead to Objects without
-% Tokens. This arises where an objects state or behavior is independent
-% of time. For example, a domain may have a set of locations and/or
-% paths for which there is nothing more to say than that they exist.
-
-% Timelines: Often objects in a domain must be described by exactly one
-% token for every given timepoint in the plan. Such objects are so
-% common that we provide a special construct in \eu to extend these
-% semantics to derived classes. Any instances of a class derived from a
-% Timeline will induce ordering requirements among its tokens in order
-% to ensure no temporal overlap may occur among them. See below for
-% further discussion.
-
-% Resources: Metric resources, e.g. the energy of a battery or the
-% capacity of a cargo hold, are objects with an explicit quantitative
-% state in time and with a circumscribed range of changes that can occur
-% to impact that state i.e. produce, consume, use, change. These changes
-% are captured as tokens. Resources are such a common requirement for
-% \eu users that special constructs are also provided for
-% them. Instances of classes derived from a Resource will induce
-% ordering requirements on their Tokens in order to ensure that the
-% level of the resource remains within specified limits.
-
-\item[\textbf{Timelines}] It may be sufficient to maintain only a
-  partial-order among tokens. However, it is often the case that
-  tokens represent states and actions of a single object in the
-  system. Such tokens are typically mutually-exclusive. \eu uses a
-  Timeline structure \cite{mus94} to concisely capture system
-  components whose behavior is described over time in this manner.
-
-\end{description}
-
-For example, consider the tire-world domain. The tire was located in
-the trunk with the predicate tireLocated(Trunk) and located on the
-ground with the predicate tireLocated(Ground). Clearly, the same tire
-cannot be in both places at once. A Timeline provides a simple method
-for aggregating the statements about a tire such that they are
-mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the
-whereabouts of a tire. The tire is an object in the tire-world
-represented as a timeline. It has predicates associated with it which
-can describe its states and actions over time. The predicate names
-need not contain the tire prefix; this is now implicit since the
-tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the \eu planning
-paradigm, they are not essential. A more general notion of a system
-Object can be used which does not impose restrictions of
-mutual-exclusion or non-zero duration. This can be important in
-supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3
-contiguous tokens. The end and start time-points are related by an
-equality constraint. A Moving predicate has been introduced to cover
-the transition from one location to another. It takes 2 location
-arguments. Notice that precedence relationships exist between tokens
-such that they cannot overlap but the start and end times may remain
-flexible. To illustrate this, sample times are included. Assume a
-total time range of interest between 0 and 1000. In addition, tokens
-on Timelines have a minimum duration of 1. As a result the values
-shown are the most flexible possible values for the start and end of
-each token. There are a number of advantages of allowing this
-flexibility. First, the basic structure of the plan can be developed
-without over-committing to specific times. If it is not necessary for
-the tire to be on the ground at time-step 4, then a planner should not
-be forced to specify it. Such an approach permits a least-commitment
-approach to planning. Second, in many domains it is simply impossible
-to know in planning exactly how long an activity might take. For
-example, a common activity of driving a car from one point to another
-on a road can take varying amounts of time depending on traffic and
-traffic lights. In such circumstances it is more practical to express
-upper and lower bounds on durations which naturally lead to intervals
-for start and end times. In such domains, flexibility in planning aids
-robustness in execution.  Rules
-
-In order for a plan to be valid, it must comply with all rules and
+\item[\textbf{Rules}] In order for a plan to be valid, it must comply with all rules and
 regulations pertinent to the application domain in question. Rules
@@
 on a token can thus cause slave tokens, variables, and constraints to
-be introduced.  Token State Model
+be introduced.
 
@@
 rule imposes a requirement for supporting tokens to be in the plan in
-order for the master to be valid. There are 2 possibilities to
-consider:
+order for the master to be valid.
 
-The slave is inserted as an active token in the plan. As such, rules
-may be activated on the slave, and it may consume resources.
+\end{description}
 
-The token is merged with a matching token already in the plan. Once a
-slave is merged, the requirement it represents are considered
-satisfied. The process of merging passes on all restrictions imposed
-on the slave to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. For an active token and an inactive token to be
-compatible requires that they are instances of the same predicate and
-that no intersections between corresponding variables are empty. The
-effect of merging is illustrated in Figure 3.
+\comment{instead of summary, tie every concept to the shopping example?}
 
-Figure 3: Merging an inactive token on an active token. Domain
-restrictions occur in highlighted variables of active token.
+This section has presented the main elements of \eu's planning representation:
 
-On creation of a token, where the commitment has not been made yet to
-activate or merge the token, the token is said to be inactive. There
-is a nuance to the state model for tokens which relate to the mode of
-its creation. If the token is allocated explicitly by an external
-actor, rather than internally through rule firing, it may include the
-state rejected indicating the planner is permitted to reject the
-token. A valid plan can include rejected tokens. This typically arises
-where the token represents a goal that is preferable to achieve but
-not mandatory. This state is not reachable for slaves since that would
-imply selective adherence to the domain model. Figure 4 presents the
-state transition diagram for token states relevant for planning. The
-transitions are operations on a partial plan which can decide an
-outcome for an inactive token. As operations which may arise in
-search, they must be reversible during backtracking. The cancellation
-operations in each case are also shown.
+\begin{enumerate}
+    \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
+    states and actions in time. Note that tokens do not discriminate between state and action.
 
-Each of the token variables, including the parameter variables, has a
-domain of values assigned to it.  The variables may also participate
-in constraints that specify which value combinations are valid.  To
-allow the specification of multiple values, e.g, to express a range of
-possible start times, variables are used to specify parameter, start
-and end time values for a token.  As a result, a token $T$ is a tuple
-$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
-variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
-$s\leq e$) are variables encoded as flexible token start and end time
-points. Consequently, partial plans represent a \textit{range of
-  possible solutions} rather than an explicit definition of a single
-trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
-of a part of such plans with tokens representing constraints on
-flexible temporal intervals. Systematic computation is reflected by
-the use of the \texttt{Allen Algebra} \cite{allen84} temporal
-relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
-interval arithmetic in the \eu domain model that allows
-encoding of temporal constraint relationships between tokens on the
-same or other concurrent timelines. Details of propagation algorithms
-or of interval logic are beyond the scope of this paper.
+    \item Token States which are the basis of planning operations on a partial plan.
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
-\caption{\small Timeline evolution can be described by simple finite
-  state transitions where temporal relationships describe the arc
-  transition and temporal extant of state are token
-  durations. Constraints here are for \emph{meets} or its equivalent
-  inverse \emph{met\_by}.}
-\label{fig:FSM}
-\vskip-0.3cm
-\end{figure}
+    \item Constraints to describe relationships among tokens. This provides an
+    expressive method of describing interactions among tokens in the context of Temporal Planning.
 
-In describing the evolution of state via timelines, transitions within
-a timeline can be described by simple finite state
-machines. Fig. \ref{fig:FSM} for example describes a timeline
-evolution for an abstract timeline entity describing navigation. Arc
-transitions represent temporal constraints and token duration are
-represented by temporal extant of state. The entire vehicle can then
-be modeled as a number of state transitions with concurrency for each
-of the modeled state variables defined a priori. However this simple
-view is often complicated when temporal constraints between state
-variables are articulated in our domain models. If there is a
-criticism of this approach, modeling then, requires a proper balance
-between abstraction, state variable description and concurrency
-described by the inter-connectedness between such finite state
-machines.
+    \item Timelines as a concise abstraction to express the evolution of state
+    and behavior for a system component. It provides semantics of mutual
+    exclusion sequencing in time. Other core abstractions are available in
+    \eu for handling metric resources.
+
+    \item Domain rules to describe internal and external relationships on and
+     between tokens respectively. Rules are applied to active tokens (master tokens) 
+     referred to as masters and typically produce inactive
+     tokens (slaves) which are a key vehicle through which the planning
+     process evolves. The rule structure in \eu differs from the more
+     restrictive commitment to preconditions and effects in classical
+     planning which prohibits durative actions and disembodied effects
+     (i.e. effects which may not occur until some temporal distance after
+     the end of the action).
+\end{enumerate}
+
+%Each of the token variables, including the parameter variables, has a
+%domain of values assigned to it.  The variables may also participate
+%in constraints that specify which value combinations are valid.  To
+%allow the specification of multiple values, e.g, to express a range of
+%possible start times, variables are used to specify parameter, start
+%and end time values for a token.  As a result, a token $T$ is a tuple
+%$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
+%variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
+%$s\leq e$) are variables encoded as flexible token start and end time
+%points. Consequently, partial plans represent a \textit{range of
+%  possible solutions} rather than an explicit definition of a single
+%trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
+%of a part of such plans with tokens representing constraints on
+%flexible temporal intervals. Systematic computation is reflected by
+%the use of the \texttt{Allen Algebra} \cite{allen84} temporal
+%relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
+%interval arithmetic in the \eu domain model that allows
+%encoding of temporal constraint relationships between tokens on the
+%same or other concurrent timelines. Details of propagation algorithms
+%or of interval logic are beyond the scope of this paper.
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
+%\caption{\small Timeline evolution can be described by simple finite
+%  state transitions where temporal relationships describe the arc
+%  transition and temporal extant of state are token
+%  durations. Constraints here are for \emph{meets} or its equivalent
+%  inverse \emph{met\_by}.}
+%\label{fig:FSM}
+%\vskip-0.3cm
+%\end{figure}
+%
+%In describing the evolution of state via timelines, transitions within
+%a timeline can be described by simple finite state
+%machines. Fig. \ref{fig:FSM} for example describes a timeline
+%evolution for an abstract timeline entity describing navigation. Arc
+%transitions represent temporal constraints and token duration are
+%represented by temporal extant of state. The entire vehicle can then
+%be modeled as a number of state transitions with concurrency for each
+%of the modeled state variables defined a priori. However this simple
+%view is often complicated when temporal constraints between state
+%variables are articulated in our domain models. If there is a
+%criticism of this approach, modeling then, requires a proper balance
+%between abstraction, state variable description and concurrency
+%described by the inter-connectedness between such finite state
+%machines.
+
+%For example, consider the tire-world domain. The tire was located in
+%the trunk with the predicate tireLocated(Trunk) and located on the
+%ground with the predicate tireLocated(Ground). Clearly, the same tire
+%cannot be in both places at once. A Timeline provides a simple method
+%for aggregating the statements about a tire such that they are
+%mutually exclusive.
+%
+%Figure 1: A Timeline for a Tire
+%
+%Figure 1 illustrates how a Timeline can be used to describe the
+%whereabouts of a tire. The tire is an object in the tire-world
+%represented as a timeline. It has predicates associated with it which
+%can describe its states and actions over time. The predicate names
+%need not contain the tire prefix; this is now implicit since the
+%tokens are assigned to a specific tire instance (i.e. an instance of a
+%Timeline). While Timelines are a useful element of the \eu planning
+%paradigm, they are not essential. A more general notion of a system
+%Object can be used which does not impose restrictions of
+%mutual-exclusion or non-zero duration. This can be important in
+%supporting partial-order planning.
+%
+%For this example, the state of the tire is specified using 3
+%contiguous tokens. The end and start time-points are related by an
+%equality constraint. A Moving predicate has been introduced to cover
+%the transition from one location to another. It takes 2 location
+%arguments. Notice that precedence relationships exist between tokens
+%such that they cannot overlap but the start and end times may remain
+%flexible. To illustrate this, sample times are included. Assume a
+%total time range of interest between 0 and 1000. In addition, tokens
+%on Timelines have a minimum duration of 1. As a result the values
+%shown are the most flexible possible values for the start and end of
+%each token. There are a number of advantages of allowing this
+%flexibility. First, the basic structure of the plan can be developed
+%without over-committing to specific times. If it is not necessary for
+%the tire to be on the ground at time-step 4, then a planner should not
+%be forced to specify it. Such an approach permits a least-commitment
+%approach to planning. Second, in many domains it is simply impossible
+%to know in planning exactly how long an activity might take. For
+%example, a common activity of driving a car from one point to another
+%on a road can take varying amounts of time depending on traffic and
+%traffic lights. In such circumstances it is more practical to express
+%upper and lower bounds on durations which naturally lead to intervals
+%for start and end times. In such domains, flexibility in planning aids
+%robustness in execution.
 
@@
 
-% \textit{Summary}
-
-% This has presented the main elements of the EUROPA planning approach. They are:
-
-% \begin{enumerate}
-
-% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
-% states and actions in time. Note that tokens do not discriminate
-% between state and action.
-
-% \item Token States which are the basis of planning operations on a partial
-% plan.
-
-% \item Constraints to describe relationships among tokens. This provides an
-% expressive method of describing interactions among tokens in the
-% context of Temporal Planning.
-
-% \item Timelines as a concise abstraction to express the evolution of state
-% and behavior for a system component. It provides semantics of mutual
-% exclusion sequencing in time. Other core abstractions are available in
-% \eu for handling metric resources.
-
-% \item Domain rules to describe internal and external relationships on and
-% between tokens respectively. Rules are applied to active tokens
-% (master tokens) referred to as masters and typically produce inactive
-% tokens (slaves) which are a key vehicle through which the planning
-% process evolves. The rule structure in \eu differs from the more
-% restrictive commitment to preconditions and effects in classical
-% planning which prohibits durative actions and disembodied effects
-% (i.e. effects which may not occur until some temporal distance after
-% the end of the action).
-
 % \end{enumerate}

### 2012-02-09 trunk/related.tex
updates

@@
  \includegraphics[scale=0.3]{figs/planner-arch.pdf}
- \caption{\small A general architectural block diagram for an
-   Artificial Intelligence based Planner.}
+ \caption{\small A general architectural block diagram for an AI based
+   Planner.}
  \label{fig:planner}
@@
 \texttt{RAX} led to the development of \texttt{IDEA}
-\cite{mus02,mus04} with the central theme of using planners
-collectively for problem solving arose. Another critical theme was
-iteratively and incremental plan repair of an anytime plan
-\cite{Zaimag96} as proposed by \texttt{CASPER}
-\cite{chien00}.
+\cite{mus02,mus04,Dias:2003ua,mus06} with the central theme of using
+planners collectively for problem solving arose. Another critical
+theme was iteratively and incremental plan repair of an anytime plan
+\cite{Zaimag96} as proposed by \texttt{CASPER} \cite{chien00}.
 
@@
 control structure formally defined as a composition of coordinated
-control loops with a recurring sense-plan-act (SPA)
-cycle. \texttt{T-REX} was designed after carefully evaluating lessons
-learned from \texttt{IDEA} and owes substantially to the notion of
-partitioned problem solving from that effort. However this
-partitioning is systematic and methodologically principled where we
-manage the information flow within the partitioned structure to ensure
-consistency in order to direct the flow of goals and observations in a
-timely manner. The resulting control structure improves scalability
-since many details of each controller can be encapsulated within a
-single control loop.  Furthermore, partitioning increases robustness
-since controller failure can be localized to enable graceful system
-degradation, making this an effective \emph{divide-and-conquer}
-approach to the overall control problem.
+control loops with a recurring sense-plan-act (SPA) cycle with
+deliberation at its core. \texttt{T-REX} was designed after carefully
+evaluating lessons learned from \texttt{IDEA} and owes substantially
+to the notion of partitioned problem solving from that effort. However
+this partitioning is systematic and methodologically principled where
+we manage the information flow within the partitioned structure to
+ensure consistency in order to direct the flow of goals and
+observations in a timely manner. The resulting control structure
+improves scalability since many details of each controller can be
+encapsulated within a single control loop.  Furthermore, partitioning
+increases robustness since controller failure can be localized to
+enable graceful system degradation, making this an effective
+\emph{divide-and-conquer} approach to the overall control problem.
 

### 2012-02-09 arch.tex
updates

@@
 
-The \rx architecture was designed with the objective to bring planning
-closer to the low-level control of the vehicle while maintaining its
-reactivity. Indeed, in most of the architecture where task planning is
-embedded, this task remains at a failry high level of abstraction as
-the complexity of this task would otherwise impede the system
-reactiveness. As a result the planning problem managed {\em in situ}
-remains fairly detached from low-level control and most of the
-reactive adaptation which in contrast is managed by specialized
-component that relies on different representations for modeling the
-local behavior they control or even do not have any explicit formal
-model of this behavior. The two angles our architecture attack this issue 
-are the following:
+A prime motivation behind the \rx architecture was to design a
+controller which could bring automated planning closer to low-level
+control of the vehicle while maintaining its reactivity. Traditionally
+planning has been deemed computationally expensive with planner
+performance restricting robot reactivity
+\cite{ghallab04,Dias:2003ua}. In control architectures where task
+planning is embedded, planning consequently remains at an abstract
+level with the assumption that it would otherwise impede system
+reactiveness and that the environment can change at a faster rate than
+the planner can plan for it. In such a situation, the agent may thrash
+if the internal state of the plan gets out of synch with the actual
+state of the world. As a result the planning problem managed {\em in
+  situ} remains fairly detached from low-level control. Most
+adaptations at the lower-level reactive layers in contrast are managed
+by specialized components (for example GeNoM \comment{} and CLARATy
+\comment{}) that rely on different representations for modeling local
+control behavior or worse do not have any explicit formal agent
+behavior model. The consequent different techniques for specifying
+each layer in the architecture results in duplication of effort and a
+diffusion of knowledge leading to significant design and integration
+issues \cite{ds1report}.  \texttt{IDEA} mitigated most of these issues
+by tightly integrating planning and execution in a single
+representational and computational framework with a unified
+declarative model. Our approach with \texttt{T-REX} builds on this
+design methodology albeit in a substantially systematic manner
+
+The \rx architecture contributes to agent architectures over and
+beyond \texttt{IDEA} are in the following ways:
 
@@
 
-\item Distribute the global planning problem into multiple decision
-  loops, called {\em reactors}. Each reactor have its own scope both 
-  functionally (one for example focusing on the conversion of high 
-  level waypoints into low-level commands) and temporally as 
-  each reactor specifies to the agent their planning look-ahead and
-  their expected latency before being able to produce plan.
+\item \emph{Partitioning} the global planning problem into multiple
+  decision loops, called {\em reactors}. Each reactor has its own
+  scope both functionally (one for example focusing on the conversion
+  of high level waypoints into low-level commands) and temporally as
+  each reactor declares its planning look-ahead and expected latency
+  before being able to produce plan to the agent depending on where it
+  is located in a reactor dependency graph such as shown in
+  Fig. \ref{fig:agent}.
 
-\item Enforce a stronger interaction between the planning process 
-  and the integration of the world evolution. The objective being to
-  make sure that the plannign task is done while continuously informed 
-  on new world state change. By doing so, we expect to allow the planner
-  tohave better informed decision despite the fact that its latency is
-  larger that the world change rate.
+\item \emph{Coupling} a tighter integration between the planning
+  process and world evolution to ensure that planning occurs while
+  being continuously informed of state change. By doing so, the
+  planner is better informed in its deliberation despite a larger
+  latency than changing world change state.
 
@@
 
-For the first point \rx provides a formal framework described in
-details in \cite{py10} and \cite{rajan12} and is not the focus of this
-paper. The second have deep implication on how on should integrate a
-planner in this framework in general and was applied to the europa
-framework specifically. We'll develop later how this requirement
-implies to see in-situ planning as two parallels processes that are
-tightly linked through the same plan structure. But before that we
-need to introduce at least the high leve lconcepts and ideas behind
-our architecture as a whole.
+\rx provides a formal framework described which details partitioning
+in \cite{py10} and \cite{rajan12} and is not the focus of this
+chapter. Coupled interaction between planning and the world on the
+other hand, has a significant implication on how planning can be
+integrated in such frameworks; here we show using the \eu planner can
+be coupled to our framework. First we link concepts in \eu planning
+with the \rx framework.
 
-\rx architecture is structured around the notion of {\em state
-  variables} as the unique basis of interactionalong with a well
-defined ownership model for each of these.  In this framework each
-reactor is seen as a black box providing its own {\em internal} state
-variables for which its is reponsible to maintain its value at every
-single instant -- represented by a {\em tick} with a fixed
-duration. In order to identify its state a reactor can subscribe to
-{\em external} state variables for which it will received new updates
-when the owner of these will produce its {\em internal} state. This
-{\em synchronization} process progate through the reactor dependency
-graph on a bottom up flow at every {\em tick} ensuring that all the
+% on should integrate a
+% planner in this framework in general and was applied to the europa
+% framework specifically. We'll develop later how this requirement
+% implies to see in-situ planning as two parallels processes that are
+% tightly linked through the same plan structure. But before that we
+% need to introduce at least the high leve lconcepts and ideas behind
+% our architecture as a whole.
+
+The \rx architecture is structured around the notion of \eu {\em state
+  variables} as the fundamental basis of interaction. Additionally it
+comes with a well-defined ownership model associated with each state
+variable. In this framework each reactor is seen as a 'black box'
+providing its own {\em internal} state variables which it is
+responsible for in order to maintain its value at every instant of
+time. Each such instance is represented by a {\em tick} with a fixed
+duration; in our implementation of \rx on the Dorado, we use a tick of
+$1$ second. In order to identify its state a reactor can subscribe to
+{\em external} state variables owned by other reactors, for which it
+will receive new updates when the owner reactor generates its {\em
+  internal} state.
+
+This {\em synchronization} process propagates up through the reactor
+dependency graph bottom up at every {\em tick} ensuring that all the
 reactors have a consistent view of the world. Conversely each reactor
 can request a future change on one of its {\em external} state
-variable which will be transfromed as an {\em internal} goal to the
-owner. This goal is expected to porduce a deliberation phase on the
-receiver which can in turn produce a plan impacting its {\em external}
-state variables that will propagate down following the same mechanism.
+variables which will be transfromed as an {\em internal} goal to the
+owner. Such \emph{goal dispatching} is expected to produce a
+deliberation phase for the receiving reactor which can cyclically and
+in turn, produce a plan impacting its {\em external} state variables
+that will propagate down following the same mechanism.
 
@@
 
-Take for example the \rx agent instance we use in our AUV
-(Fig. \ref{fig:agent}). In this architecture all the reactors
-(symbolized by colored boxes) do not have explicit connection from one
-to another but instead rely on a publish/subscribe model on state
-variables. For Example the \textsf{Pilot} reactor needs the {\em
-  externally} managed state informations from \textsf{Sensor\_Data}
-and \textsf{Nav\_cmd} states variables in order to maintain {\em
-  internally} the state variables \textsf{Waypoint} and
-\textsf{Nav\_policy}. This relation applies in two way :
+Consider for example the \rx agent instance we use on our AUV as shown
+in Fig. \ref{fig:agent}. In this instance all the reactors (symbolized
+by colored boxes) do not have an explicit connection from one to
+another but instead rely on a publish/subscribe model on shared state
+variables. For instance the \textsf{Pilot} reactor needs the {\em
+  externally} managed state information from \textsf{Sensor\_Data} and
+\textsf{Nav\_cmd} state variables in order to maintain the internal
+state variables \textsf{Waypoint} and \textsf{Nav\_policy}. This
+relation applies in two ways:
 
@@
 The way the agent interleave planning \texttt{step}s and
-synchronization for a single reactor is illustrated in Figure 
-\ref{fig:tick-exec}. Synchronization happens at every single 
-tick and will interrupt planning allowing this reactor to identify
- its current state that can propagate through the reactor graph
- in order to ensure t aconsitent view of the state of the world 
-between all the reactors.
+synchronization for a single reactor is illustrated in
+Fig.~\ref{fig:tick-exec}. Synchronization happens at every single tick
+and will interrupt planning allowing this reactor to identify its
+current state that can propagate through the reactor graph in order to
+ensure t aconsitent view of the state of the world between all the
+reactors.
 
@@
 based on the europa framework and applies the core idea of \rx we
-described above to implement a reactor for which the sole purpose is to
-deliberate as time advance. The design of this reactor takes cues from 
-the IDEA architecture presented on \cite{mus02, mus06} in the sense taht both
-planning and execution are based on a single unified model. The choice 
-of \eu as the planning framework was directed by the fact that this 
-framework provide a rich and highly configurable toolset to be able to 
-infer around a rich temporal domain.
+described above to implement a reactor for which the sole purpose is
+to deliberate as time advance. The design of this reactor takes cues
+from the IDEA architecture presented on \cite{mus02, mus06} in the
+sense that both planning and execution are based on a single unified
+model. The choice of \eu as the planning framework was directed by the
+fact that this framework provide a rich and highly configurable
+toolset to be able to infer around a rich temporal domain.
 
@@
 $\tau+1$). While focusing uniquely on the execution frontier ($\tau+1$
-in this figure), the europa solver took benefit of the exisiting plan
-to evaluate a solution that asumes that the observation is consistent
-with what was decided inside this reactor.
+in Fig. \ref{fig:shop:exec1}), the europa solver took benefit of the
+exisiting plan to evaluate a solution that asumes that the observation
+is consistent with what was decided inside this reactor.
 

### 2012-02-09 arch.tex
updates

@@
-\section{Planning and execution in the \texttt{T-REX} architecture}
+\section{Planning and execution in the \rx architecture}
 \label{sec:arch}
-
-\subsection{The \rx architecture}
-\label{sec:arch:trex}
 
@@
 %   \end{itemize}}
+
+\subsection{Introductory concepts}
+\label{sec:arch:intro}
 
@@
 diffusion of knowledge leading to significant design and integration
-issues \cite{ds1report}.  \texttt{IDEA} mitigated most of these issues
+issues \cite{DS1report}.  \texttt{IDEA} mitigated most of these issues
 by tightly integrating planning and execution in a single
@@
 be coupled to our framework. First we link concepts in \eu planning
-with the \rx framework.
+with the \rx framework. Note that in the following treatise, we call
+the formal semantic framework of \rx as the 'agent'. Reactors
+constitute an entity within this agent and are standalone blocks of
+inference.
 
@@
 % our architecture as a whole.
+
+\subsection{Architectural concepts in \rx}
+\label{sec:arch:trex}
 
@@
   externally} managed state information from \textsf{Sensor\_Data} and
-\textsf{Nav\_cmd} state variables in order to maintain the internal
-state variables \textsf{Waypoint} and \textsf{Nav\_policy}. This
-relation applies in two ways:
+\textsf{Nav\_cmd} state variables in order to maintain the
+\emph{internal} state variables \textsf{Waypoint} and
+\textsf{Nav\_policy}. This relation applies in two ways:
 
@@
 
-\item In order to identify its current {\em internal} state values
-  this reactor needs to know the current {\em external} state
-  information it relies on. In this case the current \textsf{Waypoint}
-  the vehicle is heading to can be extracted for example from the
-  current \textsf{Nav\_cmd} executed.
+\item In order to identify its current {\em internal} state values the
+  \textsf{Pilot} \comment{double check this mod} needs to know the
+  current {\em external} state information it relies on. In this case
+  the current \textsf{Waypoint} the vehicle is heading to can be
+  extracted for example from the current \textsf{Nav\_cmd} executed.
 
-\item The future objectives a reactor has on its future \textsf{Internal}
-  state will probably imply sub-objectives to be reported to the
-  owner of its {\em external} depending on their current values. Again
-  if this reactor wants to go ato a specific \textsf{Waypoint} --
-  depending on its current position provided by \textsf{Sensor\_Data}
-  it can identify a sequence of \textsf{Nav\_cmd} that should
-  eventually help him reach this location.
+\item The future objectives a reactor has on \textsf{Internal} state
+  will likely imply sub-objectives to be reported to the owner of its
+  {\em external} state variables depending on their current
+  values. Should the \textsf{Pilot} want to visit a specific
+  \textsf{Waypoint} -- given its current position as provided by
+  \textsf{Sensor\_Data}, it can identify a sequence of
+  \textsf{Nav\_cmd} states that should eventually help it reach this
+  location. \comment{this explanation could do with a figure which I
+    believe we have used previously}
 
@@
  \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
- \caption{\small A \texttt{T-REX} agent is composed of multiple
-   reactors or control loops (rounded boxes) which are connected
-   through state variables provided by one reactor ({\color{blue}blue}
-   solid line) with multiple possible clients ({\color{green}green}
-   dashed lines)}
+ \caption{\small A \rx agent is composed of multiple reactors or
+   control loops (rounded boxes) which are connected through state
+   variables provided by one reactor ({\color{blue}blue} solid line)
+   with multiple possible clients ({\color{green}green} dashed lines)}
   \label{fig:agent}
@@
 
-Overall the \rx architecture helps to abstract to the reactor
-interaction to only state variables. Each reactor is practically
-ignoring where its {\em external} state information is managed as long
-as this information is available and properly maintained so they can
-properly manage their own {\em internal} state information which in
-turn may be used by other reactors. Such formalism also constrain
-reactors interaction to state information exchanges which are two
-folds:
+Given the above example, the \rx architecture therefore helps abstract
+reactor interaction by containing them to state variables. Each
+reactor is therefore agnostic where its {\em external} state
+information is managed as long as this information is available and
+properly maintained for managing {\em internal} state
+information. This {\em internal} state in turn may be used by other
+reactors. Such a formalism also constrains reactor interaction to
+state information exchange which comes in two forms:
 
@@
 \item {\em Observation on the current state values}: This information
-  propagates up in the reactor graph at the synchronization phase 
-  and occur at the beginning of every time steps (called
-  ticks). During this phase the agent provides to each reactor 
-  updates on their {\em external} state variables for this tick and
-  ask them to compute their {\em internal} state variables values for
-  this same tick. By doing so we propagate a consistent view of that
-  state of the world throughout all the reactors as time advance.
+  propagates up in the reactor dependency graph at the synchronization
+  phase and occurs at the beginning of every tick. During this phase
+  the agent provides to each reactor updates on their {\em external}
+  state variables for this tick and ask them to compute their {\em
+    internal} state variables values for this same tick. By doing so
+  we propagate a consistent view of that state of the world throughout
+  all the reactors as time advance.
 

### 2012-02-09 trunk/arch.tex
updates

@@
 
-The \rx architecture was designed with the objective to bring planning
-closer to the low-level control of the vehicle while maintaining its
-reactivity. Indeed, in most of the architecture where task planning is
-embedded, this task remains at a failry high level of abstraction as
-the complexity of this task would otherwise impede the system
-reactiveness. As a result the planning problem managed {\em in situ}
-remains fairly detached from low-level control and most of the
-reactive adaptation which in contrast is managed by specialized
-component that relies on different representations for modeling the
-local behavior they control or even do not have any explicit formal
-model of this behavior. The two angles our architecture attck this
-issue are the following:
+\subsection{Introductory concepts}
+\label{sec:arch:intro}
+
+A prime motivation behind the \rx architecture was to design a
+controller which could bring automated planning closer to low-level
+control of the vehicle while maintaining its reactivity. Traditionally
+planning has been deemed computationally expensive with planner
+performance restricting robot reactivity
+\cite{ghallab04,Dias:2003ua}. In control architectures where task
+planning is embedded, planning consequently remains at an abstract
+level with the assumption that it would otherwise impede system
+reactiveness and that the environment can change at a faster rate than
+the planner can plan for it. In such a situation, the agent may thrash
+if the internal state of the plan gets out of synch with the actual
+state of the world. As a result the planning problem managed {\em in
+  situ} remains fairly detached from low-level control. Most
+adaptations at the lower-level reactive layers in contrast are managed
+by specialized components (for example GeNoM \comment{} and CLARATy
+\comment{}) that rely on different representations for modeling local
+control behavior or worse do not have any explicit formal agent
+behavior model. The consequent different techniques for specifying
+each layer in the architecture results in duplication of effort and a
+diffusion of knowledge leading to significant design and integration
+issues \cite{DS1report}.  \texttt{IDEA} mitigated most of these issues
+by tightly integrating planning and execution in a single
+representational and computational framework with a unified
+declarative model. Our approach with \texttt{T-REX} builds on this
+design methodology albeit in a substantially systematic manner
+
+The \rx architecture contributes to agent architectures over and
+beyond \texttt{IDEA} are in the following ways:
 
@@
 
-For the first point \rx provides a formal framework described in
-details in \cite{py10} and \cite{rajan12} and is not the focus of this
-paper. The second have deep implication on how on should integrate a
-planner in this framework in general and was applied to the europa
-framework specifically. We'll develop later how this requirement
-implies to see in-situ planning as two parallels processes that are
-tightly linked through the same plan structure. But before that we
-need to introduce at least the high leve lconcepts and ideas behind
-our architecture as a whole.
+\rx provides a formal framework which details partitioning in
+\cite{py10} and \cite{rajan12} and is not the focus of this
+chapter. Coupled interaction between planning and the world on the
+other hand, has a significant implication on how planning can be
+integrated in such frameworks; here we show using the \eu planner can
+be coupled to our framework. First we link concepts in \eu planning
+with the \rx framework. Note that in the following treatise, we call
+the formal semantic framework of \rx as the 'agent'. Reactors
+constitute an entity within this agent and are standalone blocks of
+inference.
 
-\rx architecture is structured around the notion of {\em state
-  variables} as the unique basis of interactionalong with a well
-defined ownership model for each of these.  In this framework each
-reactor is seen as a black box providing its own {\em internal} state
-variables for which its is reponsible to maintain its value at every
-single instant -- represented by a {\em tick} with a fixed
-duration. In order to identify its state a reactor can subscribe to
-{\em external} state variables for which it will received new updates
-when the owner of these will produce its {\em internal} state. This
-{\em synchronization} process progate through the reactor dependency
-graph on a bottom up flow at every {\em tick} ensuring that all the
+% on should integrate a
+% planner in this framework in general and was applied to the europa
+% framework specifically. We'll develop later how this requirement
+% implies to see in-situ planning as two parallels processes that are
+% tightly linked through the same plan structure. But before that we
+% need to introduce at least the high leve lconcepts and ideas behind
+% our architecture as a whole.
+
+\subsection{Architectural concepts in \rx}
+\label{sec:arch:trex}
+
+The \rx architecture is structured around the notion of \eu {\em state
+  variables} as the fundamental basis of interaction. Additionally it
+comes with a well-defined ownership model associated with each state
+variable. In this framework each reactor is seen as a 'black box'
+providing its own {\em internal} state variables which it is
+responsible for in order to maintain its value at every instant of
+time. Each such instance is represented by a {\em tick} with a fixed
+duration; in our implementation of \rx on the Dorado, we use a tick of
+$1$ second. In order to identify its state a reactor can subscribe to
+{\em external} state variables owned by other reactors, for which it
+will receive new updates when the owner reactor generates its {\em
+  internal} state.
+
+This {\em synchronization} process propagates up through the reactor
+dependency graph bottom up at every {\em tick} ensuring that all the
 reactors have a consistent view of the world. Conversely each reactor
 can request a future change on one of its {\em external} state
-variable which will be transfromed as an {\em internal} goal to the
-owner. This goal is expected to porduce a deliberation phase on the
-receiver which can in turn produce a plan impacting its {\em external}
-state variables that will propagate down following the same mechanism.
+variables which will be transfromed as an {\em internal} goal to the
+owner. Such \emph{goal dispatching} is expected to produce a
+deliberation phase for the receiving reactor which can cyclically and
+in turn, produce a plan impacting its {\em external} state variables
+that will propagate down following the same mechanism.
 
@@
 
-Take for example the \rx agent instance we use in our AUV
-(Fig. \ref{fig:agent}). In this architecture all the reactors
-(symbolized by colored boxes) do not have explicit connection from one
-to another but instead rely on a publish/subscribe model on state
-variables. For Example the \textsf{Pilot} reactor needs the {\em
-  externally} managed state informations from \textsf{Sensor\_Data}
-and \textsf{Nav\_cmd} states variables in order to maintain {\em
-  internally} the state variables \textsf{Waypoint} and
-\textsf{Nav\_policy}. This relation applies in two way :
+\begin{figure}[!htb]
+ \centering
+ \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
+ \caption{\small A \rx agent is composed of multiple reactors or
+   control loops (rounded boxes) which are connected through state
+   variables provided by one reactor ({\color{blue}blue} solid line)
+   with multiple possible clients ({\color{green}green} dashed lines)}
+  \label{fig:agent}
+\end{figure}
+
+Consider for example the \rx agent instance we use on our AUV as shown
+in Fig. \ref{fig:agent}. In this instance all the reactors (symbolized
+by colored boxes) do not have an explicit connection from one to
+another but instead rely on a publish/subscribe model on shared state
+variables. For instance the \textsf{Pilot} reactor needs the {\em
+  externally} managed state information from \textsf{Sensor\_Data} and
+\textsf{Nav\_cmd} state variables in order to maintain the
+\emph{internal} state variables \textsf{Waypoint} and
+\textsf{Nav\_policy}. This relation applies in two ways:
 
@@
 
-\begin{figure}[!htb]
- \centering
- \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
- \caption{\small A \texttt{T-REX} agent is composed of multiple
-   reactors or control loops (rounded boxes) which are connected
-   through state variables provided by one reactor ({\color{blue}blue}
-   solid line) with multiple possible clients ({\color{green}green}
-   dashed lines)}
-  \label{fig:agent}
-\end{figure}
-
-Overall the \rx architecture helps to abstract to the reactor
-interaction to only state variables. Each reactor is practically
-ignoring where its {\em external} state information is managed as long
-as this information is available and properly maintained so they can
-properly manage their own {\em internal} state information which in
-turn may be used by other reactors. Such formalism also constrain
-reactors interaction to state information exchanges which are two
-folds:
+Given the above example, the \rx architecture therefore helps abstract
+reactor interaction by containing them to state variables. Each
+reactor is therefore agnostic where its {\em external} state
+information is managed as long as this information is available and
+properly maintained for managing {\em internal} state
+information. This {\em internal} state in turn may be used by other
+reactors. Such a formalism also constrains reactor interaction to
+state information exchange which comes in two forms:
 
@@
 \item {\em Observation on the current state values}: This information
-  propagates up in the reactor graph at the synchronization phase 
-  and occur at the beginning of every time steps (called
-  ticks). During this phase the agent provides to each reactor 
-  updates on their {\em external} state variables for this tick and
-  ask them to compute their {\em internal} state variables values for
-  this same tick. By doing so we propagate a consistent view of that
-  state of the world throughout all the reactors as time advance.
+  propagates \emph{up} in the reactor dependency graph at the
+  synchronization phase and occurs at the beginning of every
+  tick. During this phase the agent provides updates to each reactor
+  on their {\em external} state variables for this tick so as they
+  could compute their {\em internal} state variable values for this
+  same tick. By doing so we propagate a consistent view of that state
+  of the world throughout all the reactors as time
+  advances. \comment{could do with a figure here}
 
 \item {\em Goals request on future state}: This information propagates
-  down in the reactor graph. It reflects that, in order to satisfy own
-  of its {\em internal} objectives, a reactor may have a plan that
-  rely on future values of one of its {\em external} state. When such
-  a plan is identified we ensure that this part of the plan is given
-  as a goal to the owner of the given {\em external(s)} state variable(s).
+  \emph{down} in the reactor dependency graph. In order to satisfy own
+  {\em internal} objectives, a reactor may have a plan that relies on
+  future values of one of its {\em external} state. When such a plan
+  is identified the agent ensures that this part of the plan is given
+  as a goal to the owner of the given {\em external} state
+  variable(s). \comment{could do with a figure here}
 
@@
 The choice of representation of state variables, observations and
-goals is directly derieved to the timleines and tokens from europa.
-such representation is ich enough to allow the excahnge of state
+goals is directly derieved to the timelines and tokens from europa. 
+such representation is rich enough to allow the exchange of state
 information in a formal and yet flexible way and its transaltion in
@@
 
-The overall execution cycle of the agent is also abstracted as some
+The overall execution cycle of the agent is also abstracted as a
 continuous planning/deliberation cycle between all the
-reactors. Indeed , at the agent interface level, each reactor provided
+reactors. At the agent interface level, each reactor provided
 only two abstract calls that conceptualize their execution:
@@
 The way the agent interleave planning \texttt{step}s and
-synchronization is illustrated in Figure \ref{fig:tick-exec}.
-Synchronization happens at every single tick and will interrupt
-planning allowing this reactor to identify its current state that can
-propagate through the reactor graph in order to ensure t aconsitent
-view of the state of the world between all the reactors.
+synchronization for a single reactor is illustrated in
+Fig.~\ref{fig:tick-exec}. Synchronization happens at every single tick
+and will interrupt planning allowing this reactor to identify its
+current state that can propagate through the reactor graph in order to
+ensure t aconsitent view of the state of the world between all the
+reactors.
 
@@
 
-This view of the reactor has two parallel tasks aloow for long
+This view of the reactor has two parallel tasks allow for long
 planning deliberation while ensuring that the state evolution
@@
 usess automated planning for its execution. This reactor is fully
-based on the europa framework and applies the core idea we of \rx we
-described above to implement a reactor for which the sole purpse is to
-deliberate as time advance. The choice of europa as the planning
-framework was directed by the fact thta this framework provide a rich
-and highly configurable framework to be able to infer around a rich
-temporal domain.
+based on the europa framework and applies the core idea of \rx we
+described above to implement a reactor for which the sole purpose is
+to deliberate as time advance. The design of this reactor takes cues
+from the IDEA architecture presented on \cite{mus02, mus06} in the
+sense that both planning and execution are based on a single unified
+model. The choice of \eu as the planning framework was directed by the
+fact that this framework provide a rich and highly configurable
+toolset to be able to infer around a rich temporal domain.
 

### 2012-02-10 concepts.tex
finished cleanup of "EUROPA plan representation section". Still need figures and a few minor notes remain.

@@
     \item \textit{Numeric}: integers and floats
-    \item \textit{Object}: reference to an object instance
+    \item \textit{Object}: reference to an object instance (see below for a description of objects in \eu)
   \end {enumerate}
@@
 
+\item[\textbf{Constraints}] valid plans in most problem domains have to satisfy a number of restrictions, for instance, a Shopping Agent may only be able to perform one task at a time, may have limited time or an energy or fuel budget to perform a task, and so on.
+In \eu, these restrictions are represented by Constraints, which can be defined over any combination of Variables in a domain.
+\eu provides a Constraint Library that already implements many useful constraints (temporal, resource, relational, etc), it is also possible to add new constraints if required by a particular domain.
+
 \item[\textbf{Objects}] The items we wish to describe and refer to in
@@
   description to find likely objects. In the Shopping Agent example,
-  we night consider the Agent itself, Products and Locations all to be
+  we might consider the Agent itself, Products and Locations all to be
   objects. Objects have state and behavior. For example, a Shopping
@@
 
-  \item State: it's location, a bag that contains the products it has
+  \item State: its location, a bag that contains the products it has
     already purchased (the bag can in turn be another object), etc.
@@
 
-  Similarly objects that have similar state and behavior can be
+  Objects that have similar state and behavior can be
   described generically in terms of object types or classes. \eu
-  allows the definition of object classes in the same way that it is
-  done in popular object-oriented programming languages. However, to
-  describe state and behavior for the purposes of planning we need to
+  allows the definition of classes in the same way that it is
+  done in popular object-oriented programming languages, where a class 
+  can have attributes (represented by Variables), and classes can be arranged in a single-parent hierarchy through inheritance. 
+  However, one key aspect needs to be introduced to class definition in \eu that is not found 
+  in object-oriented programming:  to describe state and behavior for the purposes of planning we need to
   build on the formalism of first order logic as explained below.
@@
 
-\comment {resume cleanup here}
-
 \item[\textbf{Rules}] In order for a plan to be valid, it must comply with all rules and
@@
 govern the internal and external relationships of a token. For
-example, consider a parameterized predicate describing a transition
-from one location to another. Let the parameters be from and to. The
+example, consider a parameterized action \texttt{Go(origin,destination)} which describes
+a Shopping Agent traveling from one location ( \textit{origin}) to another ( \textit{destination}). The
 parameters are instantiated on a token as variables whose domain of
@@
 can be easily expressed as a constraint on the definition of a
-predicate of the form: from != to. It is reasonable to further
-stipulate that one must be Located somewhere before a transition can
-occur, and one must end up Located somewhere when completed. This is
+predicate of the form: \textit{origin != destination}. It makes sense to further
+stipulate that the agent must be located at  \textit{origin} before travel can happen
+and the agent must be located at  \textit{destination} when completed. This is
 an example of a rule governing an external relationship among
 tokens. It specifies a requirement that tokens of the predicate
-Located precede and succeed tokens of the predicate Moving.
+\texttt{At} that represent the location state for the Shopping Agent 
+precede and succeed tokens of the predicate \texttt{Go} that represents traveling.
 
-Figure 2: Internal and external relationships from rules on Moving
-Figure 2 illustrates the entities and relations involved in specifying
-such a rule on a Moving predicate. The token on which the rule applies
-is referred to as the master. Each Located token required by the
+\comment{ Add figure showing internal and external relationships from rules on action Go}
+
+Figure xx illustrates the entities and relations involved in specifying
+such a rule on a \texttt{Go} action. The token on which the rule applies
+is referred to (\texttt{Go}) as the master. Each  \texttt{At} token required by the
 master is referred to as a slave. All variables are indicated by name
@@
 is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting tokens to be in the plan in
+rule imposes a requirement for supporting (slave) tokens to be in the plan in
 order for the master to be valid.
+
+\comment {talk about condition/effect annotations?}
 
@@
 
-\comment{instead of summary, tie every concept to the shopping example?}
-
-This section has presented the main elements of \eu's planning representation:
+We have just covered the main elements of \eu's planning representation:
 
 \begin{enumerate}
-    \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
-    states and actions in time. Note that tokens do not discriminate between state and action.
-
-    \item Token States which are the basis of planning operations on a partial plan.
-
-    \item Constraints to describe relationships among tokens. This provides an
-    expressive method of describing interactions among tokens in the context of Temporal Planning.
-
-    \item Timelines as a concise abstraction to express the evolution of state
-    and behavior for a system component. It provides semantics of mutual
-    exclusion sequencing in time. Other core abstractions are available in
-    \eu for handling metric resources.
-
-    \item Domain rules to describe internal and external relationships on and
+   \item \textbf{Domains, Variables and Objects} to represent the entities (Objects) and attributes (Variables) needed to represent a problem domain	
+    \item \textbf{Tokens} Temporally scoped predicates to represent states and actions in time. 
+    \item \textbf{Token States} which are necessary for the implementation of planning operations on a partial plan.
+    \item \textbf{Constraints} to describe relationships among attributes of objects and tokens.
+\eu has a built-in library of constraints that covers time, resources and others that are common in real life problem domains.
+    \item \textbf{Domain Rules} to describe internal and external relationships on and
      between tokens respectively. Rules are applied to active tokens (master tokens) 
-     referred to as masters and typically produce inactive
-     tokens (slaves) which are a key vehicle through which the planning
-     process evolves. The rule structure in \eu differs from the more
-     restrictive commitment to preconditions and effects in classical
-     planning which prohibits durative actions and disembodied effects
-     (i.e. effects which may not occur until some temporal distance after
+     referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning
+     process evolves. \comment{citation, or perhaps remove this statement?:}The rule structure in \eu differs from the more restrictive commitment to preconditions and effects in classical
+     planning which prohibits durative actions and disembodied effects (i.e. effects which may not occur until some temporal distance after
      the end of the action).    
+    \item \textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
 \end{enumerate}
 
-%Each of the token variables, including the parameter variables, has a
-%domain of values assigned to it.  The variables may also participate
-%in constraints that specify which value combinations are valid.  To
-%allow the specification of multiple values, e.g, to express a range of
-%possible start times, variables are used to specify parameter, start
-%and end time values for a token.  As a result, a token $T$ is a tuple
-%$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
-%variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
-%$s\leq e$) are variables encoded as flexible token start and end time
-%points. Consequently, partial plans represent a \textit{range of
-%  possible solutions} rather than an explicit definition of a single
-%trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
-%of a part of such plans with tokens representing constraints on
-%flexible temporal intervals. Systematic computation is reflected by
-%the use of the \texttt{Allen Algebra} \cite{allen84} temporal
-%relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
-%interval arithmetic in the \eu domain model that allows
-%encoding of temporal constraint relationships between tokens on the
-%same or other concurrent timelines. Details of propagation algorithms
-%or of interval logic are beyond the scope of this paper.
-%
-%\begin{figure}
+%\begin{figure*}
 %\centering 
-%\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
-%\caption{\small Timeline evolution can be described by simple finite
-%  state transitions where temporal relationships describe the arc
-%  transition and temporal extant of state are token
-%  durations. Constraints here are for \emph{meets} or its equivalent
-%  inverse \emph{met\_by}.}
-%\label{fig:FSM}
-%\vskip-0.3cm
-%\end{figure}
-%
-%In describing the evolution of state via timelines, transitions within
-%a timeline can be described by simple finite state
-%machines. Fig. \ref{fig:FSM} for example describes a timeline
-%evolution for an abstract timeline entity describing navigation. Arc
-%transitions represent temporal constraints and token duration are
-%represented by temporal extant of state. The entire vehicle can then
-%be modeled as a number of state transitions with concurrency for each
-%of the modeled state variables defined a priori. However this simple
-%view is often complicated when temporal constraints between state
-%variables are articulated in our domain models. If there is a
-%criticism of this approach, modeling then, requires a proper balance
-%between abstraction, state variable description and concurrency
-%described by the inter-connectedness between such finite state
-%machines.
+%\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
+%\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
+%\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
+%\caption{\small An illustrative plan synthesis example with concurrent
+%  timelines. Each timeline is an instantiation of a subsystem tracked
+%  by the planner. Tokens describe the instantiated state of subsystem
+%  at a particular time instance and enforce temporal and parametric
+%  constraints. Abstract goals are decomoposed into successively less
+%  abstract tokens and instantiated in co-temporal timelines using
+%  \texttt{Allen Algebra} relations. All tokens represent flexible
+%  start/end times. \ref{fig:plan-evolve1} shows an initial state
+%  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
+%  \label{fig:Plan-evolve}
+%  \vskip-5pt
+%\end{figure*}
 
-%For example, consider the tire-world domain. The tire was located in
-%the trunk with the predicate tireLocated(Trunk) and located on the
-%ground with the predicate tireLocated(Ground). Clearly, the same tire
-%cannot be in both places at once. A Timeline provides a simple method
-%for aggregating the statements about a tire such that they are
-%mutually exclusive.
-%
-%Figure 1: A Timeline for a Tire
-%
-%Figure 1 illustrates how a Timeline can be used to describe the
-%whereabouts of a tire. The tire is an object in the tire-world
-%represented as a timeline. It has predicates associated with it which
-%can describe its states and actions over time. The predicate names
-%need not contain the tire prefix; this is now implicit since the
-%tokens are assigned to a specific tire instance (i.e. an instance of a
-%Timeline). While Timelines are a useful element of the \eu planning
-%paradigm, they are not essential. A more general notion of a system
-%Object can be used which does not impose restrictions of
-%mutual-exclusion or non-zero duration. This can be important in
-%supporting partial-order planning.
-%
-%For this example, the state of the tire is specified using 3
-%contiguous tokens. The end and start time-points are related by an
-%equality constraint. A Moving predicate has been introduced to cover
-%the transition from one location to another. It takes 2 location
-%arguments. Notice that precedence relationships exist between tokens
-%such that they cannot overlap but the start and end times may remain
-%flexible. To illustrate this, sample times are included. Assume a
-%total time range of interest between 0 and 1000. In addition, tokens
-%on Timelines have a minimum duration of 1. As a result the values
-%shown are the most flexible possible values for the start and end of
-%each token. There are a number of advantages of allowing this
-%flexibility. First, the basic structure of the plan can be developed
-%without over-committing to specific times. If it is not necessary for
-%the tire to be on the ground at time-step 4, then a planner should not
-%be forced to specify it. Such an approach permits a least-commitment
-%approach to planning. Second, in many domains it is simply impossible
-%to know in planning exactly how long an activity might take. For
-%example, a common activity of driving a car from one point to another
-%on a road can take varying amounts of time depending on traffic and
-%traffic lights. In such circumstances it is more practical to express
-%upper and lower bounds on durations which naturally lead to intervals
-%for start and end times. In such domains, flexibility in planning aids
-%robustness in execution.
+%Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
+%timelines which track \texttt{Goal, Path, Navigation} and
+%\texttt{Command} state over time. Tokens on the goal timeline indicate
+%a survey within a bounded volume and a min/max depth envelope. As the
+%Volume Survey sub-goals on the \texttt{Path} timeline, the initial
+%token on that timeline gets \emph{squeezed} to have dependencies of
+%the goal token instantiated. Initially this dependency are the tokens
+%\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
+%coverage. The first of these in turn generates a sub-goal on the same
+%timeline to go to the next intermediate waypoint, which is illustrated
+%in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
+%figure) ultimately generate all the tokens to completion for this
+%timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
+%in plan generation and shows additional subgoals on the \texttt{Path,
+%  Navigation} and \texttt{Command} timelines. Each token can thus
+%trace its causality when instantiated in the plan. The domain model
+%provides source of these temporal and parametric dependencies in the
+%form of rules.
 
-\begin{figure*}
-\centering 
-\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
-\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
-\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
-\caption{\small An illustrative plan synthesis example with concurrent
-  timelines. Each timeline is an instantiation of a subsystem tracked
-  by the planner. Tokens describe the instantiated state of subsystem
-  at a particular time instance and enforce temporal and parametric
-  constraints. Abstract goals are decomoposed into successively less
-  abstract tokens and instantiated in co-temporal timelines using
-  \texttt{Allen Algebra} relations. All tokens represent flexible
-  start/end times. \ref{fig:plan-evolve1} shows an initial state
-  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
-  \label{fig:Plan-evolve}
-  \vskip-5pt
-\end{figure*}
-
-Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
-timelines which track \texttt{Goal, Path, Navigation} and
-\texttt{Command} state over time. Tokens on the goal timeline indicate
-a survey within a bounded volume and a min/max depth envelope. As the
-Volume Survey sub-goals on the \texttt{Path} timeline, the initial
-token on that timeline gets \emph{squeezed} to have dependencies of
-the goal token instantiated. Initially this dependency are the tokens
-\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
-coverage. The first of these in turn generates a sub-goal on the same
-timeline to go to the next intermediate waypoint, which is illustrated
-in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
-figure) ultimately generate all the tokens to completion for this
-timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
-in plan generation and shows additional subgoals on the \texttt{Path,
-  Navigation} and \texttt{Command} timelines. Each token can thus
-trace its causality when instantiated in the plan. The domain model
-provides source of these temporal and parametric dependencies in the
-form of rules.
-
-\begin{minipage}[c]{\textwidth}
-\vspace{+0.5cm}
+%\begin{minipage}[c]{\textwidth}
+%\vspace{+0.5cm}
 % \framebox[\textwidth][t]{
-Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
-$\Rightarrow$ \\
-met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
-starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
-\ldots{} \\
-ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
-meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
-\end{minipage}
+%Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
+%$\Rightarrow$ \\
+%met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
+%starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
+%\ldots{} \\
+%ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
+%meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
+%\end{minipage}
 

### 2012-02-10 trunk/europa.tex
checkpoint. cleaning up europa's architecture section before diving into modeling.

@@
 
-The Constraint Reasoning Engine (CRE) manages variables, the domains
-from which they can take values, and constraints that define
+The \textbf{Constraint Reasoning Engine} (CRE) manages domain types, and constraints that define
 relationships among variables. It also provides an efficient arc
@@
 specialized reasoning algorithms for specific constraints can be easily
-and efficiently plugged in. The Plan Database (PDB) manages object types
-and objects, which are a mechanism to group variables to more naturally
-model the real world in much the same way that object oriented
-programming does. Also manages tokens, which are a mechanism to group
-variables to represent temporally scoped state. Objects and tokens can
-be used to model planning domains in a much more natural and extensible
-way that a pure CSP approach can. The Rules Engine (RE) manages token
-relationships in a planning model. The language modules (NDDL, ANML, and
+and efficiently plugged in.
+
+The \textbf{Plan Database} (PDB) manages object types
+and objects,
+
+The \textbf{Rules Engine} (RE) manages token
+relationships in a planning model.
+
+The {Language Modules} (NDDL, ANML, and
 other higher level modeling languages) are used to create domain models
 and problem instances \comment{point advantages over pure programming
-  API approaches, provide snippets}. Finally the API Layer (Fig.
-\ref{fig:europa-api}) makes all the services available so that \eu can
+  API approaches, provide snippets}.
+
+  Finally,  the \textbf{API Layer}  makes all the services available so that \eu can
 be used to build applications.
 
-\comment{Is Fig. \ref{fig:europa-api} necessary?}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
-\caption{\small EUROPA's Client API}
-\label{fig:europa-api}
-\vskip+0.1cm
-\end{figure}
-
-	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
-	% 	\begin{enumerate}
-	% 		\item Temporal Network
-	% 		\item Resources
-	% 		\item NDDL
-	% 		\item ANML (TODO:maybe it's better to leave out?)
-	% 	\end{enumerate}
+Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+\begin{enumerate}
+	\item Temporal Network
+	\item Resources
+	\\item NDDL. ANML (TODO:maybe it's better to leave out?)
+\end{enumerate}
 
@@
   tokens can be generated by explicitly posted goals, or when a token is
-  activated, rules may fire that create inactive slave tokens (TODO:
-  point to more detailed decription above in Rules Engine section). Open
+  activated, rules may fire that create inactive slave tokens
+  \comment{point to more detailed decription above in Rules Engine
+    section}.
+  Open
   Conditions can be resolved in three ways:
@@
 
-\comment{expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint}
+\comment{expand a little more on how heuristics can be implemented by
+  applying ordering/filtering for chooseFlaw and makeDecisionPoint}
 examples:
@@
 It should be emphasized that while this algorithm is commonly employed,
-it is only one of many that could be implemented. TODO: give examples of
-local search solvers for scheduling and CSP?
+it is only one of many that could be implemented. \comment{give examples
+  of local search solvers for scheduling and CSP?}
 

### 2012-02-11 trunk/europa.tex
more europa architecture cleanup.

@@
 
-The \textbf{Constraint Reasoning Engine} (CRE) manages domain types, and constraints that define
-relationships among variables. It also provides an efficient arc
+The \textbf{Constraint Reasoning Engine} (CRE) manages domain types, variables and constraints that define
+relationships among them. It also provides an efficient arc
 consistency mechanism \cite{mackworth77}. The CRE is designed so that
@@
 
-The \textbf{Plan Database} (PDB) manages object types
-and objects,
+The \textbf{Plan Database} (PDB) manages object and token types.
+It also maintains the state of partial plans in terms of object, variable and token instances.
+This constitutes the data foundation on which end user applications and automated planners are built.
 
-The \textbf{Rules Engine} (RE) manages token
-relationships in a planning model.
+The \textbf{Rules Engine} (RE) modularly adds the capability to support domain rules that describe dependencies between tokens,
 
-The {Language Modules} (NDDL, ANML, and
-other higher level modeling languages) are used to create domain models
-and problem instances \comment{point advantages over pure programming
-  API approaches, provide snippets}.
+The \textbf{Solver Module} provides the framework for flaw detection and resolution that is needed to implement search algorithms for planning.
 
-  Finally,  the \textbf{API Layer}  makes all the services available so that \eu can
-be used to build applications.
+The \textbf {Language Modules}: Describing problem domains and problem instances in terms of objects, variables and domain rules in a programmatic way is cumbersome to the extent of being impractical, that motivated the creation of modeling languages that support those definitions in a much more natural and compact way.
+NDDL is the main modeling language for \eu and will be described in detail in the Modeling section below.
+ANML is a language under development that provides a more intuitive syntax \comment{provide ANML citation}.
 
-Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
+The \textbf{API Layer}  provides access to services from all the different modules in a programmatic way so that \eu can be easily embedded into specialized applications.
+
+In an effort to mimimize the amount of work that a user needs to perform to solve specific problems, \eu  also provides extension modules that have been found to be useful in many domains: 
 \begin{enumerate}
-	\item Temporal Network
-	\item Resources
-	\\item NDDL. ANML (TODO:maybe it's better to leave out?)
+	\item \textit{Temporal Network}: Extensions to the CRE module to reason about temporal constraints 
+	\item \textit{Resources}: Extensions to the CRE, PDB and Solver modules to provide out-of-the-box reasoning and search for metric resources.
+	\item \textit{Chronological Backtracking Solver}: A general purpose built-in solver to provide the ability to plan out-of-the-box, problems that require challenging search may require specialized solvers to be built as explained in the Search section below.
 \end{enumerate}

### 2012-02-11 arch.tex
updates

@@
 
-\rx provides a formal framework described which details partitioning
-in \cite{py10} and \cite{rajan12} and is not the focus of this
+\rx provides a formal framework which details partitioning in
+\cite{py10} and \cite{rajan12} and is not the focus of this
 chapter. Coupled interaction between planning and the world on the
@@
 
+\begin{figure}[!htb]
+ \centering
+ \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
+ \caption{\small A \rx agent is composed of multiple reactors or
+   control loops (rounded boxes) which are connected through state
+   variables provided by one reactor ({\color{blue}blue} solid line)
+   with multiple possible clients ({\color{green}green} dashed lines)}
+  \label{fig:agent}
+\end{figure}
+
 Consider for example the \rx agent instance we use on our AUV as shown
@@
 
-\begin{figure}[!htb]
- \centering
- \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
- \caption{\small A \rx agent is composed of multiple reactors or
-   control loops (rounded boxes) which are connected through state
-   variables provided by one reactor ({\color{blue}blue} solid line)
-   with multiple possible clients ({\color{green}green} dashed lines)}
-  \label{fig:agent}
-\end{figure}
-
 Given the above example, the \rx architecture therefore helps abstract
@@
 
-\item {\em Observation on the current state values}: This information
-  propagates up in the reactor dependency graph at the synchronization
-  phase and occurs at the beginning of every tick. During this phase
-  the agent provides to each reactor updates on their {\em external}
-  state variables for this tick and ask them to compute their {\em
-    internal} state variables values for this same tick. By doing so
-  we propagate a consistent view of that state of the world throughout
-  all the reactors as time advance.
+\item {\em Observation on current state values}: This information
+  propagates \emph{up} in the reactor dependency graph at the
+  synchronization phase and occurs at the beginning of every
+  tick. During this phase the agent provides updates to each reactor
+  on their {\em external} state variables for this tick so as they
+  could compute their {\em internal} state variable values for this
+  same tick. By doing so we propagate a consistent view of that state
+  of the world throughout all the reactors as time
+  advances. \comment{could do with a figure here}
 
-\item {\em Goals request on future state}: This information propagates
-  down in the reactor graph. It reflects that, in order to satisfy own
-  of its {\em internal} objectives, a reactor may have a plan that
-  rely on future values of one of its {\em external} state. When such
-  a plan is identified we ensure that this part of the plan is given
-  as a goal to the owner of the given {\em external(s)} state variable(s).
+\item {\em Goal request on future state}: This information propagates
+  \emph{down} in the reactor dependency graph. In order to satisfy own
+  {\em internal} objectives, a reactor may have a plan that relies on
+  future values of one of its {\em external} state. When such a plan
+  is identified the agent ensures that this part of the plan is given
+  as a goal to the owner of the given {\em external} state
+  variable(s). \comment{could do with a figure here}
 
@@
 The choice of representation of state variables, observations and
-goals is directly derieved to the timelines and tokens from europa. 
-such representation is rich enough to allow the exchange of state
-information in a formal and yet flexible way and its transaltion in
-planning frameworks that have an explicit representation of time (such
-as but not limited to europa) is often trivial.
+goals is directly derived from \eu timelines and tokens.  While such a
+rich representation allows exchange of state information in a formal
+yet flexible manner, its translation into planning frameworks that
+have an explicit representation of time (such as but not limited to
+\eu) is often trivial.
 
-The overall execution cycle of the agent is also abstracted as a
-continuous planning/deliberation cycle between all the
-reactors. At the agent interface level, each reactor provided
-only two abstract calls that conceptualize their execution:
+\subsection{The Execution Cycle}
+\label{sec:arch:exec}
+
+The overall execution cycle of the agent is also abstracted out as a
+continuous planning/deliberation cycle between all reactors. At the
+agent interface level, each reactor provides only two abstract calls
+that conceptualize execution:
 
@@
 
-\item \texttt{synchronize} that takes the last {\em external}
-  observations as argument and will return the {\em internal}
-  observations for this reactor or a failure report.
+\item \texttt{synchronize} takes the last {\em external} observations
+  as an argument and returns the {\em internal} observations for a
+  reactor or a failure report.
 
-\item \texttt{step} that accept new goals requested (if any) on the
-  {\em internal} timeline of this reactor and will execute one step of
-  deliberation. The returned value indicates if a complete plan has
-  been found and in that case will also provide the part of this plan
-  applied to {\em external} timelines of this reactor.
+\item \texttt{step} that accepts new goals requested (if any) on the
+  {\em internal} timeline of a reactor and will execute one step of
+  deliberation \comment{Might be good to articulate what ``one'' step
+    looks like}. The returned value indicates if a complete plan has
+  been found. If so this part \comment{what does 'this part' mean? Can
+    we sketch a figure or make it more formal?} of the plan is
+  provided to {\em external} timelines of this reactor.
 
@@
 
-Both of these two deliberative functions have different scope and
-temporal constraints. The \texttt{synchronize} call is considered as
-atomic and will focus only on state inference from the current tick by
-identifying the current value of the {\em internal} state variables
-from the latest updates on the {\em external} variables. This process
-will then propagate up in the reactor graph at any single tick to
-ensure that all reactors are aware of the current state of the
-world. \texttt{step} relates to a single decision step of the planning
-process which is focused on the future evolution of the state
-variables and it can take multiple steps spanning other multiple ticks
-for a reactor to produce a plan that is both complete and valid. when
-such plan is identified its {\em external} part is then dispatched to
-the corresponding reactors which in turn can start their deliberation
-\texttt{step}s in order to satisfy these new objectives. As a
-consequence physical execution will occur as these goals and planning
-propagate down in the reactor graph until they eventually become goals
-for the lowest level reactor which -- instead of deliberating -- may
-just transform this goal in a physical command execution.
-
-The way the agent interleave planning \texttt{step}s and
-synchronization for a single reactor is illustrated in
-Fig.~\ref{fig:tick-exec}. Synchronization happens at every single tick
-and will interrupt planning allowing this reactor to identify its
-current state that can propagate through the reactor graph in order to
-ensure t aconsitent view of the state of the world between all the
-reactors.
+Both these deliberative functions have different scope and temporal
+constraints. The \texttt{synchronize} call is considered as atomic and
+will focus only on state inference for the current tick by identifying
+the current value of {\em internal} state variables from the latest
+updates on the {\em external} state variables. This process propagates
+up in the reactor dependency graph at any single tick \comment{at
+  every single tick??? maybe} to ensure that all reactors are aware of
+the current state of the world. \texttt{step} relates to a single
+decision step of the planning process which is focused on the future
+evolution of state variables; it can take multiple steps spanning
+multiple ticks for a reactor to produce a plan that is both complete
+and valid. When such plan is identified \comment{how? Is there a
+  terminating condition?} its {\em external} timeline(s) are then
+dispatched to the corresponding reactors which own these times and
+which in turn can start their own deliberation \texttt{step}s to
+satisfy these new objectives. Consequently execution occurs as these
+goals and plans propagate down in the reactor hierarchy until they
+eventually become goals for the lowest level reactor which -- instead
+of deliberating -- may just transform this goal into a command for
+execution in hardware.
 
@@
 
-This view of the reactor has two parallel tasks allow for long
-planning deliberation while ensuring that the state evolution
-propagates through the agent as time advance. These two tasks 
-intertwined together give the two natural flows (bottom-up for state
-estimation and top to bottom for future plan) that we expect to see in
-a control loop. As they are both conceptualized as inference processes
-it is natural to implement a reactor based on generic automated
-planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
-  comm.} and \textsf{Vehicle} acting as interfaces to the system,
-every other reactors (colored in blue) are different instance of our
-europa reactor differntiated only by their model and europa solver
-configurations.
+The manner in which the agent interleaves planning \texttt{step}s and
+synchronization for a single reactor is illustrated in
+Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick
+interrupting planning; this allows a reactor to identify its current
+state that can propagate through the reactor hierarchy in order to
+ensure a consistent view of the state of the world between all the
+reactors.
+% This view of a reactor has two parallel tasks allowing for long
+% planning deliberation while ensuring that the state evolution
+% propagates through the agent as time advance.
+These intertwined tasks together provide two natural information flow
+methods (bottom-up for state estimation and top-down for plan
+projection) that we expect to see in a control loop. As they are both
+inference-based processes it is natural to implement a reactor based
+on generic automated planning. For example in Fig. \ref{fig:agent},
+apart from \textsf{Shore comm.} and \textsf{Vehicle} acting as
+interfaces to the system, every other reactor (in blue) are different
+instances of a \eu reactor differentiated only by their model and \eu
+solver configurations.
 
-\subsection{The Europa deliberative reactor}
+\subsection{The  \eu Deliberative Reactor}
 \label{sec:arch:europa}

### 2012-02-13 trunk/europa-basics.tex
updates

@@
 \paragraph \textit{Scheduling}: In the Resource Constrained Project
-Scheduling Problem (RCPSP) \comment{citations needed}, a project
-consisting of a set of activities must be scheduled in a way that
-satisfies minimum and/or maximum temporal separation constraints. The
-activity schedule must also respect fixed limits on the availability of
-resources required to perform each activity. In addition to satisfying
-temporal and resource constraints, it is common for the user to want to
+Scheduling Problem (RCPSP) \cite{Bruckera99}, a project consisting of
+a set of activities must be scheduled in a way that satisfies minimum
+and/or maximum temporal separation constraints. The activity schedule
+must also respect fixed limits on the availability of resources
+required to perform each activity. In addition to satisfying temporal
+and resource constraints, it is common for the user to want to
 minimize makespan so that the entire project is finished as early as

### 2012-02-13 trunk/arch.tex
updates

@@
 
-\rx provides a formal framework which details partitioning in
-\cite{py10} and \cite{rajan12} and is not the focus of this
-chapter. Coupled interaction between planning and the world on the
-other hand, has a significant implication on how planning can be
+Partitioned inference is a major contribution of \rx. However that is
+not the focus of this chapter; details are in \cite{py10} and
+\cite{rajan12}. Coupled interaction between planning and the world on
+the other hand, has a significant implication on how planning can be
 integrated in such frameworks; here we show using the \eu planner can
@@
 Given the above example, the \rx architecture therefore helps abstract
-reactor interaction by containing them to state variables. Each
+reactor interaction by constraining them to state variables. Each
 reactor is therefore agnostic where its {\em external} state
@@
 
-\item {\em Observation on the current state values}: This information
+\item {\em Observation on current state values}: This information
   propagates \emph{up} in the reactor dependency graph at the
@@
 
-\item {\em Goals request on future state}: This information propagates
+\item {\em Goal request on future state}: This information propagates
   \emph{down} in the reactor dependency graph. In order to satisfy own
@@
 
-Both of these two deliberative functions have different scope and
-temporal constraints. The \texttt{synchronize} call is considered as
-atomic and will focus only on state inference from the current tick by
-identifying the current value of the {\em internal} state variables
-from the latest updates on the {\em external} variables. This process
-will then propagate up in the reactor graph at any single tick to
-ensure that all reactors are aware of the current state of the
-world. \texttt{step} relates to a single decision step of the planning
-process which is focused on the future evolution of the state
-variables and it can take multiple steps spanning other multiple ticks
-for a reactor to produce a plan that is both complete and valid. when
-such plan is identified its {\em external} part is then dispatched to
-the corresponding reactors which in turn can start their deliberation
-\texttt{step}s in order to satisfy these new objectives. As a
-consequence physical execution will occur as these goals and planning
-propagate down in the reactor graph until they eventually become goals
-for the lowest level reactor which -- instead of deliberating -- may
-just transform this goal in a physical command execution.
-
-The way the agent interleave planning \texttt{step}s and
-synchronization for a single reactor is illustrated in
-Fig.~\ref{fig:tick-exec}. Synchronization happens at every single tick
-and will interrupt planning allowing this reactor to identify its
-current state that can propagate through the reactor graph in order to
-ensure t aconsitent view of the state of the world between all the
-reactors.
+Both these deliberative functions have different scope and temporal
+constraints. The \texttt{synchronize} call is considered as atomic and
+will focus only on state inference for the current tick by identifying
+the current value of {\em internal} state variables from the latest
+updates on the {\em external} state variables. This process propagates
+up in the reactor dependency graph at every tick to ensure that all
+reactors are aware of the current state of the world. \texttt{step}
+relates to a single decision step of the planning process which is
+focused on the future evolution of state variables; it can take
+multiple steps spanning multiple ticks for a reactor to produce a plan
+that is both complete and valid. When such plan is identified
+\comment{how? Is there a terminating condition?} its {\em external}
+timeline(s) are then dispatched to the corresponding reactors which
+own these times and which in turn can start their own deliberation
+\texttt{step}s to satisfy these new objectives. Consequently execution
+occurs as these goals and plans propagate down in the reactor
+hierarchy until they eventually become goals for the lowest level
+reactor which -- instead of deliberating -- may just transform this
+goal into a command for execution in hardware.
 
@@
   \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
-  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
-      Deliberation} is interrupted by {\em synchronization} at the
-    beginning of every {\em tick} allowing integration of state
-    information.}
+  \caption{\small \rx reactor execution cycle: {\em Deliberation} is
+    interrupted by {\em synchronization} at the beginning of every
+    {\em tick} allowing integration of state information.}
   \label{fig:tick-exec}
@@
 
-This view of the reactor has two parallel tasks allow for long
-planning deliberation while ensuring that the state evolution
-propagates through the agent as time advance. These two tasks 
-intertwined together give the two natural flows (bottom-up for state
-estimation and top to bottom for future plan) that we expect to see in
-a control loop. As they are both conceptualized as inference processes
-it is natural to implement a reactor based on generic automated
-planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
-  comm.} and \textsf{Vehicle} acting as interfaces to the system,
-every other reactors (colored in blue) are different instance of our
-europa reactor differntiated only by their model and europa solver
+The manner in which the agent interleaves planning \texttt{step}s and
+synchronization for a single reactor is illustrated in
+Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick while
+interrupting planning; this allows a reactor to identify its current
+state that can propagate through the reactor hierarchy in order to
+ensure a consistent view of the state of the world between all the
+reactors.
+% This view of a reactor has two parallel tasks allowing for long
+% planning deliberation while ensuring that the state evolution
+% propagates through the agent as time advance.
+These intertwined tasks together provide two natural information flows
+(bottom-up for state estimation and top-down for plan projection) that
+we expect to see in a control loop. As they are both inference-based
+processes it is natural to implement a reactor based on generic
+automated planning. For example in Fig. \ref{fig:agent}, apart from
+\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces to the
+system, every other reactor (in blue) are different instances of a \eu
+reactor differentiated only by their model and \eu solver
 configurations.
 
-\subsection{The Europa deliberative reactor}
+\subsection{The  \eu Deliberative Reactor}
 \label{sec:arch:europa}
 
-The \eu reactor is our implementation inside \rx of a reactor that
-usess automated planning for its execution. This reactor is fully
-based on the europa framework and applies the core idea of \rx we
-described above to implement a reactor for which the sole purpose is
-to deliberate as time advance. The design of this reactor takes cues
-from the IDEA architecture presented on \cite{mus02, mus06} in the
-sense that both planning and execution are based on a single unified
-model. The choice of \eu as the planning framework was directed by the
-fact that this framework provide a rich and highly configurable
-toolset to be able to infer around a rich temporal domain.
-
-This reactor is implemented following the idea that in \rx, the
-whole execution cycle is two deliberation processes. Both of these
-processes are implemented as two europa solvers modifying the same
-plan database and for which the execution is managed by the \rx 
-framework as time advance during the execution of the system:
+Within \rx, an instance of an \eu reactor comes with the sole purpose
+to deliberate. A typical design of such a reactor takes its cues from
+\texttt{IDEA} \cite{mus02, mus06} where both planning and execution
+are based on a single unified model within a rich representation that
+provides support for temporal inference. In \rx, the execution cycle
+consists of two deliberation processes which are implemented as two
+\eu solvers modifying the \emph{same} plan database. %  and for which the
+% execution is managed by the \rx framework as time advance during the
+% execution of the system:
 
@@
 
-\item The synchronization solver is a specialized europa solver that
-  will integrate in the plan new information about the evolution of
-  the External state variable and ensure that the reactor propagate
-  these in order to both identify its current state and inform
-  other reactors of any state change on its internal timelines. This
-  solver is summoned at the beginning of every single tick.
+\item The \emph{synchronization solver} is a specialized \eu solver
+  that integrates new information in the plan, the evolution of
+  \emph{external} state variables and ensures that the reactor
+  propagates these to identify the reactors current state as well as
+  to inform other reactors of any state change on its \emph{internal}
+  timelines. This solver is summoned at the beginning of every single
+  tick.
 
-\item The deliberation solver is managing the deliberation process of
-  the reactor in order to produce a new plan or alter its current plan
-  as new goals are given to the reactor or the synchronization solver
-  identifying a conflict between the current state of the world and
-  the expectations of the previous plan. This process can  span other
-  multiple tick and therefore can be interrupted at any single tick in
-  order for the synchronization solver to do its task.
+\item The \emph{deliberation solver} manages the deliberation process
+  of the reactor either to produce a new plan or alter its current
+  plan as new goals are provided or when the synchronization solver
+  identifies a conflict between current state and expectations from a
+  previously generated plan. This process can span multiple ticks and
+  can therefore be interrupted by the synchronization solver.
 
@@
 While the two processes are separate, they share the same plan
-internal to a reactor and therefore at every synchronization cycle the
+internal to a reactor; consequently at every synchronization cycle the
 planning process is informed of the new world state and its impact on
-the plan. Conversely, when the planning process eventually finds a
+the plan. Conversely, when the deliberation process eventually finds a
 solution, synchronization is informed about generating the planned
-state values on the {\em external} state variables. This {\em
-  external} plan defines the set of {\em goals} for this state
-variable managed by another {\em reactor}.
+state values \comment{should this read as ``...is informed about
+  generated state values..'' instead?} on the {\em external} state
+variables. This {\em external} plan defines the set of {\em goals} for
+this state variable managed by another {\em reactor}.
 
-An important challenge for embedding automayted planning i na situated
-agent as always been to go around the asumptions made by most classical
-planners. Specifically one assumption restrict the problem to ``offline
-planning''  defined in \cite{ghallab04} as follow:
+Synchronization offers an important challenge for embedding planning
+in a situated agent, typically %  as always been to go around the asumptions made by
+% most classical planners. Specifically one
+an assumption to restrict the problem to ``offline planning'' defined
+in \cite{ghallab04} as follows:
 
@@
   The planner is not concerned with any change that may occur in
-  $\Sigma$ \footnote{In this book $\Sigma$ identifies the world modeled
-    by the plan domain} while it is planning; it plans for the given
-  initial and goal states regardless of the current dynamics, if any.
+  $\Sigma$ \footnote{the world modeled by the plan domain} while it is
+  planning; it plans for the given initial and goal states regardless
+  of the current dynamics, if any. 
 \end{quotation}
 
-This assumption, while helpfull to reduce the scope of the planning
-problem, becomes problematic if one wants to embed a planner in a
-real-time system that evolves in a highly dynamic environment. In a
-typical robot this implies that the only point where the planner can
-be informed about the current world state is right before the planning
-is started by giving to the planner the intial state and current
-objectives. During the planning, it is assumed that the agent can
-maintin the world as perceived by the planner as stable. In
-\cite{lemai04, lemai-chenevier2004}, they do attempt to reduce the
-impact of planning by allowing local plan repairs which attempt to
-insert repairing actions in the exisiting plan andf can be done fast
-enough. Still in ythe eventuality that such repair is not possible
-they add to stop their vehicle until the full (re-)planning was
-complete. {\em I may need more examples/refs here}
+This assumption, while helpful to reduce the scope of the planning
+problem, becomes problematic for planning within a highly dynamic
+environment. In a typical robot this implies that the only point where
+the planner can be informed about the current world state is prior to
+the commencement of a planning cycle by giving to the planner the
+initial state and current objectives. During the planning, it is
+assumed that the agent can maintin the world as perceived by the
+planner as stable. \cite{lemai04, lemai-chenevier2004} attempts to
+reduce the impact of planning by allowing local plan repair in the
+existing plan. In the eventuality that such repair is not feasible
+they require that their robot comes to a full stop until (re-)planning
+is complete. \comment{I may need more examples/refs here}. As Fig.
+\ref{fig:tick-exec} shows, this assumption is problematic for our
+architecture. Synchronization that tracks state evolution can occur
+several times during the reactor deliberation process. By making
+synchronization and deliberation share the same plan we provide a
+solution that relaxes this static world assumption.
 
-Looking back at Fig. \ref{fig:tick-exec}, one can already see that
-this asumtpion is problematic in our architecture. Indeed
-synchronization that tracks state evolution can occur several times
-during the reactor deliberation process. By making synchronization and
-deliberation share the same plan we provide a solution that relax this
-assumption. In the following sections we present how synchronization
-and deliberation are implemented independently first to finally show
-how their interaction through the plan helps relax this assumption by
-making the planning fully integrated to the reactor execution cycle
-while potentially allowing for more informed planning.
+In the following sections we show how synchronization and deliberation
+are implemented independently to show how their interaction through
+the plan helps relax this assumption by making planning fully
+integrated in the reactor execution cycle while potentially allowing
+for more informed planning.
 

### 2012-02-13 arch.tex
updates

@@
 updates on the {\em external} state variables. This process propagates
-up in the reactor dependency graph at any single tick \comment{at
-  every single tick??? maybe} to ensure that all reactors are aware of
-the current state of the world. \texttt{step} relates to a single
-decision step of the planning process which is focused on the future
-evolution of state variables; it can take multiple steps spanning
-multiple ticks for a reactor to produce a plan that is both complete
-and valid. When such plan is identified \comment{how? Is there a
-  terminating condition?} its {\em external} timeline(s) are then
-dispatched to the corresponding reactors which own these times and
-which in turn can start their own deliberation \texttt{step}s to
-satisfy these new objectives. Consequently execution occurs as these
-goals and plans propagate down in the reactor hierarchy until they
-eventually become goals for the lowest level reactor which -- instead
-of deliberating -- may just transform this goal into a command for
-execution in hardware.
+up in the reactor dependency graph at every tick to ensure that all
+reactors are aware of the current state of the world. \texttt{step}
+relates to a single decision step of the planning process which is
+focused on the future evolution of state variables; it can take
+multiple steps spanning multiple ticks for a reactor to produce a plan
+that is both complete and valid. When such plan is identified
+\comment{how? Is there a terminating condition?} its {\em external}
+timeline(s) are then dispatched to the corresponding reactors which
+own these times and which in turn can start their own deliberation
+\texttt{step}s to satisfy these new objectives. Consequently execution
+occurs as these goals and plans propagate down in the reactor
+hierarchy until they eventually become goals for the lowest level
+reactor which -- instead of deliberating -- may just transform this
+goal into a command for execution in hardware.
 
@@
 synchronization for a single reactor is illustrated in
-Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick
+Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick while
 interrupting planning; this allows a reactor to identify its current
@@
 % propagates through the agent as time advance.
-These intertwined tasks together provide two natural information flow
-methods (bottom-up for state estimation and top-down for plan
-projection) that we expect to see in a control loop. As they are both
-inference-based processes it is natural to implement a reactor based
-on generic automated planning. For example in Fig. \ref{fig:agent},
-apart from \textsf{Shore comm.} and \textsf{Vehicle} acting as
-interfaces to the system, every other reactor (in blue) are different
-instances of a \eu reactor differentiated only by their model and \eu
-solver configurations.
+These intertwined tasks together provide two natural information flows
+(bottom-up for state estimation and top-down for plan projection) that
+we expect to see in a control loop. As they are both inference-based
+processes it is natural to implement a reactor based on generic
+automated planning. For example in Fig. \ref{fig:agent}, apart from
+\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces to the
+system, every other reactor (in blue) are different instances of a \eu
+reactor differentiated only by their model and \eu solver
+configurations.
 
@@
 
-The \eu reactor is our implementation inside \rx of a reactor that
-usess automated planning for its execution. This reactor is fully
-based on the europa framework and applies the core idea of \rx we
-described above to implement a reactor for which the sole purpose is
-to deliberate as time advance. The design of this reactor takes cues
-from the IDEA architecture presented on \cite{mus02, mus06} in the
-sense that both planning and execution are based on a single unified
-model. The choice of \eu as the planning framework was directed by the
-fact that this framework provide a rich and highly configurable
-toolset to be able to infer around a rich temporal domain.
-
-This reactor is implemented following the idea that in \rx, the
-whole execution cycle is two deliberation processes. Both of these
-processes are implemented as two europa solvers modifying the same
-plan database and for which the execution is managed by the \rx 
-framework as time advance during the execution of the system:
+Within \rx, an instance of an \eu reactor comes with the sole purpose
+to deliberate. A typical design of such a reactor takes its cues from
+\texttt{IDEA} \cite{mus02, mus06} where both planning and execution
+are based on a single unified model within a rich representation that
+provides support for temporal inference. In \rx, the execution cycle
+consists of two deliberation processes which are implemented as two
+\eu solvers modifying the \emph{same} plan database. %  and for which the
+% execution is managed by the \rx framework as time advance during the
+% execution of the system:
 
@@
 
-For the sake of simplicity we introduce first how a reactor can track
-and identify its state in the context where it does not have any
-compelling need to deliberate. Assume here that we have a reactor that
-have no future goal. {\em Need to develop the reason why it is a
-  necessity: primarily to provide its internal state to whoever
-  observe it but als simply to ensure that its current representation
-  of the world is up to date and still consistent}
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
+  \caption{\small Illustration of the synchronization flaws in a reactor. The
+    reactor receive new observations when they are produced by the
+    owner(s) of its internal timelines. The line after the last token
+    of each timeline represent the domain of possible values for the
+    end of this token. At every tick $\tau$ the reactor needs to
+    integrate the {\em External} state information it received and --
+    based on its model $\mathcal{M}$ -- resolve its {\em Internal}
+    state that will then be provided by the architecture to other
+    reactors using these state variables.}
+  \label{fig:synch:flaw}
+\end{figure}
 
-In the europa based reactor we introduce this requirement as a new
-type of flaw for the europa framework. This flaw enforces at the
-reactor to identify fully its internal state for the current tick. By
-using this new flaw we can describe the synchronization process as
-this general sequence :
+We first show how a reactor can track and identify its state in the
+context where it does not have any compelling need to deliberate. We
+assume that we have a reactor that has no future goal. \comment{Need
+  to develop the reason why it is a necessity: primarily to provide
+  its internal state to whoever observe it but also simply to ensure
+  that its current representation of the world is up to date and still
+  consistent}
+
+We introduce this requirement as a new type of flaw (Section
+\ref{sec:europa:arch}) for the \eu framework. This flaw forces the
+reactor to identify its internal state for the current tick. By using
+this flaw one can describe the synchronization process as:
 
@@
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
-  \caption{Illustration of the synchronization flaws in a reactor. The
-    reactor receive new observations when they are produced by the
-    owner(s) of its internal timelines. The line after the last token
-    of each timeline represent the domain of possible values for the
-    end of this token. At every tick $\tau$ the reactor needs to
-    integrate the {\em External} state information it received and --
-    based on its model $\mathcal{M}$ -- resolve its {\em Internal}
-    state that will then be provided by the architecture to other
-    reactors using these state variables.}
-  \label{fig:synch:flaw}
-\end{figure}
-
-The resolution of internal state flaw can be resolved using one of the
-following choices which are evaluated in the given order:
+An internal state flaw can be resolved using one of the following
+choices which are evaluated in order:
 
@@
 
-\item Extend the previous state value to end after this tick (ie
-  restrict its end time to $[\tau+1, \infty)$).
+\item Extend the previous state value to end after this tick (\ie
+  restrict its end time to $[\tau+1, \infty)$). \comment{first use of
+    $\tau$, so needs to be defined earlier}
 
-\item Start the next active token in the timeline and
-  attempt to start it now (ie restrict its start time to the single
-  value $\tau$).
+\item Start the next active token in the timeline 
+  by restricting its start time to the single value $\tau$
 
@@
 All of these choices are evaluated sequentially until a consistent
-solution with no more flaw for this tick is identified. In order to do
-so we need to make the assumption that the current state value of an
-internal timeline do not depend on the future or more accurately that
-any choices made during this synchronization will not lead to a future
-inconsistency (meaning no possible solution) for a future
-synchronization. Such assumption deeply impact the set of possible
-domains one the reactor can support while remaining complete. Take for
+solution with no flaws at this tick, is identified. In order to do so
+we need to make the assumption that the current state value of an
+internal timeline does not depend on the future or more accurately
+that any choices made during this synchronization will not lead to a
+future inconsistency. Such an assumption impacts the set of possible
+domains the reactor can support while remaining complete. Take for
 example this rule from our \texttt{Shopping} model :
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 While this model is perfectly standard and acceptable in the overall
-case. It can become problematic while tied to the
-execution. Specificially it puts a strong tie between the {\em fact} that we
-are \texttt{Going} to a location and the {\em expected} future outcome  of
-ending at this location. This ties is even stronger due to the
-constraint at line 5. Knwoing this aspect assume that the
-\texttt{Agent.location} is {\em external} to the reactor which receive
-the observation \texttt{Going(Home,SuperMarket)} which synchronization
-resolved by producing the observation \texttt{Go(Home, SuperMarket)}
-in the Agent timeline. From this point this means that the only
-possible outcome on \texttt{Agent.location} is to be
-\texttt{At(SuperMarket)} in the directly foreseeable future. In the
-real world the fact that we are \texttt{Going} to a certain location
-does not necessarly imply that we will succeed on ending up there. An
-extreme case could be that due to some exogenous event, the road to
-the \texttt{SuperMarket} is closed which could result on an alternate
-observation \texttt{At(7th st. \& Foo Ave.)}. This observation
-contradicts thre existing plan the reactor maintin and it is important
-here to relaize the temp[oral direction of the causal link that is
-triggereing this inconsitency. Indeed at the time we will observe this
-contradicting \texttt{At}, the rule that generated this conlict is
-connected to the past observations \texttt{Going} and \texttt{Go}
-which cannot be fully retracted without potential global impact in the
-rest of the exisiting reactors -- implying  a huge cost in
-backprogation that would be problematic.
+case. It can become problematic when tied to execution. Specifically it
+puts a strong tie between the {\em fact} that we are \texttt{Going} to a
+location and the {\em expected} future outcome of ending at this
+location. This ties is even stronger due to the constraint at line 5.
+Knowing this aspect assume that the \texttt{Agent.location} is {\em
+  external} to the reactor which receive the observation
+\texttt{Going(Home,SuperMarket)} which synchronization resolved by
+producing the observation \texttt{Go(Home, SuperMarket)} in the Agent
+timeline. From this point this means that the only possible outcome on
+\texttt{Agent.location} is to be \texttt{At(SuperMarket)} in the
+directly foreseeable future. In the real world the fact that we are
+\texttt{Going} to a certain location does not necessarly imply that we
+will succeed on ending up there. An extreme case could be that due to
+some exogenous event, the road to the \texttt{SuperMarket} is closed
+which could result on an alternate observation \texttt{At(7th st. \& Foo
+  Ave.)}. This observation contradicts thre existing plan the reactor
+maintin and it is important here to relaize the temp[oral direction of
+the causal link that is triggereing this inconsitency. Indeed at the
+time we will observe this contradicting \texttt{At}, the rule that
+generated this conlict is connected to the past observations
+\texttt{Going} and \texttt{Go} which cannot be fully retracted without
+potential global impact in the rest of the exisiting reactors --
+implying a huge cost in backprogation that would be problematic.
 
-More generally tying a token
-to future outcomes in a model that is meant to be confronted to rela
+More generally tying a token to future outcomes in a model that is meant
+to be confronted to rela 
 executionis something that is problematic. It may be valid ifg you can
@@
 flexibility:
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 The alteration made to relax the constraint on linbe 5 from a strict 

### 2012-02-14 trunk/europa.tex
checkpoint. fleshing out modeling section.

@@
 
-\item It offers procedural constructs to populate \eus plan database
+\item It offers procedural constructs to populate \eu's plan database
   with specific problem instances expressed in terms of objects,
@@
 To understand how the most important elements of NDDL work, let's take a
-look at how the Shopping Agent problem described above would be modeled:
+look at how the Shopping Agent problem could be modeled:
 
@@
 }
+\end{verbatim}
+
+In the model above we can see how the main elements or a problem domain can be mapped to \eu's representation:
+
+\begin{enumerate}
+	\item \textbf{Object Types}: NDDL supports an object-oriented approach, where common data and behavior can be abstracted as a class.
+In this case, we have the \textit{Product} class represent the products that the Shopping Agent needs, \textit{Location} represents the places where products can be found, \textit{ProductLocation} represents the association arising when a product can be purchased at a location.
+The Shopping Agent itself is represented by an \textit{Agent} class, which relies on the \textit{AgentLocation} class to keep track of where the agent is at any point in time.
+NDDL supports a single rooted class hierarchy so that state and behavior can be extended through inheritance.
+We took advantage of that by making \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline} class which doesn't allow predicates to overlap in time for the same object instance, in this way we can ensure that and agent can only be at one place at a time.
+	\item \textbf{Predicate Types}: Some classes like \textit{Product}, \textit{Location} and \textit{ProductLocation} are completely static and so don't have any predicate or action types associated with them.
+The \textit{AgentLocation} class is purely used to keep track of state and so it has \texttt{At} and \texttt{Going} predicate types associated with it.
+	\item \textbf{Action Types}: The \textit{Agent} class is the only one that has behavior associated with it, consequently, it defines action types to represent when the agent is taking action to \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a \textit{Product}.
+We'll see how the actual behavior for those actions is specified below.
+	\item \textbf{Variable Types and Instances} : All the object attributes are represented as variables in this model, for instance, it uses integer variables to represent the built-in temporal variables for each token (activity duration, start and end times), string variables to represent Location and Product names, and object reference variables to represent the structural relationships between different objects (for instance, each Agent has variable to keep track of its AgentLocation).
+\end{enumerate}
+
+\comment{explain rule definition below}
+
+\begin{verbatim}
 
@@
 
-In this model we can see:
-
-\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
-\begin{enumerate}
-	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
-	\item Object Types:
-	\item Predicate Types:
-	\item Action Types:
-\end{enumerate}
-
- shows how a specific problem instance would be
-specified in NDDL.
+Now, having defined the Shopping agent domain that way, here is how a particular problem instance could be specified:
 
@@
 \label{eu:model}
+%\caption{Specifying a problem instance for the Shopping Agent in NDDL}
 \end{verbatim}

### 2012-02-14 trunk/arch.tex
Some updates and correction

@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation \comment{Might be good to articulate what ``one'' step
-    looks like}. The returned value indicates if a complete plan has
-  been found. If so this part \comment{what does 'this part' mean? Can
-    we sketch a figure or make it more formal?} of the plan is
-  provided to {\em external} timelines of this reactor.
-
+  deliberation. In the context of Europa for example pone step of
+  deliberation correspod to resolve one sibngle flaw while ensuring
+  that the resulting plan is not proven inconsistent. This implies
+  that it also includes the backtracking of the search in such case. 
+  The returned value by this step function is a boolean which
+  indicates whether the reactor needs some extra step to deliberate or
+  if this last step resulted on a complete plan for the reactor
+  look-ahead. On the later, the part of the plan that concern this
+  reactor {\em external} timelines is provided in order to be
+  dispatched as goals to whichever reactor that is reponsible for
+  these. \comment{By extending I may have saud things I developped
+    below ... will have  to check}
 \end{itemize}
@@
 that is both complete and valid. When such plan is identified
-\comment{how? Is there a terminating condition?} its {\em external}
+\comment{how? Is there a terminating condition? No more flaws but I
+  washoping it would have been defined in europa section} its {\em external}
 timeline(s) are then dispatched to the corresponding reactors which
@@
 planning process is informed of the new world state and its impact on
-the plan. Conversely, when the deliberation process eventually finds a
+the plan. Conversely, when synchronization occurs after deliberation
+steps, its starting point is the last partial plan resulting from
+deliberation. This gives an initial frame to the problemm that helps
+-- at least initial -- to target the synchronization problem by
+ensuring that the new {\em external} state updates are compatible with
+this plan. If so, the synchronization propagates these events in the
+plan and can uses it as a help on identifying its {\em internal}
+timelines based on this plan. If they are not compatibles, then
+synchronization identifies a plan inconsistency  reflecting that the
+current plan is not valid anymore. \comment{fpy: I am here and will
+  develop further}
+
+\comment{following is not what I intended to express whatsoever
+  ... synchroniuzation od not deal with dispatching but execution
+  tracking ... it is though helped by the structure of the exisiting
+  plan and often identifies execution failures}
+ Conversely, when the deliberation process eventually finds a
 solution, synchronization is informed about generating the planned
@@
 synchronization and deliberation share the same plan we provide a
-solution that relaxes this static world assumption.
+solution that relaxes this static world assumption. \comment{Need to
+  talk about IDEA and the fact that they had similar aspect even though
+  they were never very clear nor well articulated in their papers}
 
@@
 compelling need to deliberate. Assume here that we have a reactor that
-have no future goal. {\em Need to develop the reason why it is a
-  necessity: primarily to provide its internal state to whoever
-  observe it but als simply to ensure that its current representation
-  of the world is up to date and still consistent}
+have no future goal. Even in this context, it is a core requirement in
+T-REX (see \cite{py10}) to have each reactor to produce at every
+single tick their {\em internal} state or at least, for the ones not owning
+{\em internal} timelines, make sure that {\em external} observations
+are taken into account for every tick. This requirement is enforced to
+ensure that all the reactors of an agent share a consistent view of
+the world as time advance.
 
-In the europa based reactor we introduce this requirement as a new
-type of flaw for the europa framework. This flaw enforces at the
+Taking into account {\em external} state updates is manage d by
+introducing these observation in the plan structure as factsand ensure
+that they remain compatible sewith the current plan mainainted by the
+reactor. The more delicate part of synchronization is to decide of the 
+{\em internla} state values for the current tick. In the Shopping
+example for example we have the following code snippet whihch helps
+identify when we can \texttt{Buy} a product:
+\begin{verbatim}
+Agent::Buy {
+  eq(duration, 10);
+  // [...]
+  contained_by(condition object.location.At currLocation);
+  // [...]
+}
+\end{verbatim}
+If \texttt{object.location} is {\em external} and the \texttt{Agent}
+is {\em internal}. I may encounter the situation where I am at the
+location where I can buy one (possibly among others) product. Already
+there's here  a decision problem related to the state value of the
+\texttt{Agent} timeline where I can either {\em Buy} a specific
+product now, wait a little longer before doing so or just simply not
+buy any product at this location.
+
+In the europa based reactor we introduce this decision reuqirement 
+as a new type of flaw for the europa framework. This flaw enforces at the
 reactor to identify fully its internal state for the current tick. By
@@
   \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
-  \caption{Illustration of the synchronization flaws in a reactor. The
+  \caption{\small Illustration of the synchronization flaws in a reactor. The
     reactor receive new observations when they are produced by the
@@
 
-The resolution of internal state flaw can be resolved using one of the
-following choices which are evaluated in the given order:
+We first show how a reactor can track and identify its state in the
+context where it does not have any compelling need to deliberate. We
+assume that we have a reactor that has no future goal. \comment{Need
+  to develop the reason why it is a necessity: primarily to provide
+  its internal state to whoever observe it but also simply to ensure
+  that its current representation of the world is up to date and still
+  consistent}
+
+We introduce this requirement as a new type of flaw (Section
+\ref{sec:europa:arch}) for the \eu framework. This flaw forces the
+reactor to identify its internal state for the current tick. By using
+this flaw one can describe the synchronization process as:
 
@@
 
-\item Extend the previous state value to end after this tick (ie
-  restrict its end time to $[\tau+1, \infty)$).
+\item Integrate the external state as provided by the owner of each
+  external timeline into the plan database.
 
-\item Start the next active token in the timeline and
-  attempt to start it now (ie restrict its start time to the single
-  value $\tau$).
+\item Propagate this information in the plan database following the
+  model $\mathcal{M}$ of this reactor.
+
+\item Resolve the current state value of each internal timelines.
+
+\end{enumerate}
+
+An internal state flaw can be resolved using one of the following
+choices which are evaluated in order:
+
+\begin{enumerate}
+
+\item Extend the previous state value to end after this tick (\ie
+  restrict its end time to $[\tau+1, \infty)$). \comment{first use of
+    $\tau$, so needs to be defined earlier}
+
+\item Start the next active token in the timeline 
+  by restricting its start time to the single value $\tau$
 

### 2012-02-14 trunk/europa.tex
fleshed out modeling section.

@@
 \eu can ingest descriptions of models, plans and goals and thereby can
-be applied to different systems and problems, simply by providing
+be applied to different domains and problems, simply by providing
 different models and goals. Consequently the expressiveness of the
-languages for models, goals and plans is of great importance. For \eu
+language for models, goals and plans is of great importance. For \eu,
 NDDL (New Domain Description Language [pronounced ``noodle'']
-\comment{cite}) allows the user to specify the different components of a
+allows the user to specify the different components of a
 problem domain in a concise way. The main characteristics that make NDDL
@@
 We took advantage of that by making \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline} class which doesn't allow predicates to overlap in time for the same object instance, in this way we can ensure that and agent can only be at one place at a time.
+	\item \textbf{Variable Types and Instances} : All object attributes are represented as variables in this model, for instance, it uses integer variables to represent the built-in temporal parameters for each token (activity duration, start and end times), string variables to represent  \textit{Location} and  \textit{Product} names, and object reference variables to represent the structural relationships between different objects (for instance, each Agent has variable to keep track of its  \textit{AgentLocation}).
 	\item \textbf{Predicate Types}: Some classes like \textit{Product}, \textit{Location} and \textit{ProductLocation} are completely static and so don't have any predicate or action types associated with them.
 The \textit{AgentLocation} class is purely used to keep track of state and so it has \texttt{At} and \texttt{Going} predicate types associated with it.
+The \textit{Agent} class keeps track of the products it purchases through the \texttt{Own} predicate, having a separate \textit{AgentBag} class with its own predicates would have been an equally valid modeling choice.
 	\item \textbf{Action Types}: The \textit{Agent} class is the only one that has behavior associated with it, consequently, it defines action types to represent when the agent is taking action to \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a \textit{Product}.
 We'll see how the actual behavior for those actions is specified below.
-	\item \textbf{Variable Types and Instances} : All the object attributes are represented as variables in this model, for instance, it uses integer variables to represent the built-in temporal variables for each token (activity duration, start and end times), string variables to represent Location and Product names, and object reference variables to represent the structural relationships between different objects (for instance, each Agent has variable to keep track of its AgentLocation).
 \end{enumerate}
-
-\comment{explain rule definition below}
 
@@
 
-Now, having defined the Shopping agent domain that way, here is how a particular problem instance could be specified:
+Domain rules can be defined in NDDL in the context of a Predicate or an Action. In the example above we can see the main elements of a rule definition:
+
+\begin{enumerate}
+  \item \textbf{Variables}: A Token Type (predicate or action) is part of a class, therefore class attributes (see references to \texttt{object.location} above), Token Type parameters (see references to \texttt{[from, to]}  in \texttt{Agent::Go}, or to \texttt{[duration, product]} in \texttt{Agent::Buy}  and locally declared variables (see \texttt{possibleStores} in \texttt{Agent::Buy}) can all be used in the definition of a Rule.
+  \item \textbf{Constraints}: restrictions on the values that variables can take inside a Rule are expressed through constraints, in the example above only the \texttt{eq} constraint is used, but in general, any constraint available in \eu's library can be used.
+ \item \textbf{Subgoals}: In \texttt{Agent::Go}, subgoals are used to specify that the Shopping Agent's location must transition from origin to destination in a temporally consistent way.
+The \texttt{[meets, equals, metby]} operators are used to specify subgoals (Allen operators are explained in more detail below), along with the temporal relationship each token representing a subgoal has with the the token representing the \texttt{Agent::Go} action.
+Also, \texttt{[condition, effect]} annotations can be used when defining a subgoal inside an action, \eu doesn't immediately associate any semantics with them, but they can be extremely useful for search modules when creating plans.
+\end{enumerate}
+
+Now, having that we have a specification for the Shopping Agent domain, here is how a particular problem instance could be specified:
 
@@
 
-\comment{explain variable, object, predicate and constraint instances.
-explain goals and facts.}
-
+A problem instance consists mainly of 
 \begin{enumerate}
-	\item Variable Instances:
-	\item Object Instances:
-	\item Facts:
-	\item Goals:
-	\item Constraint Instances:
+  \item \textbf{Variables} and \textbf{Objects}: In this case we defined temporal variables that are used to specify the planning horizon (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are used to specify the locations, products, product-loction associations and the Shopping Agent that we want to plan for.
+  \item \textbf{Facts}: These are used to specify the current state of the world and don't need any support.
+Facts are represented as tokens in \eu's Plan Database.
+In this case we use facts to specify that the Shipping Agent is initially at home.
+  \item \textbf{Goals}:  Goals are represented as tokens in \eu's Plan Database, differently from Facts, Goals trigger inference in \eu which determines whether flaws (unresolved subgoals, overloaded resources, etc., flaws will be covered in detail in the search section below) need to be resolved, if that is the case, search modules can be invoked which will look for a plan where all the flaws are resolved.
+  \item \textbf{Constraints}: restrictions on and between actions and goals are specified through constraints, in this case we use temporal constraints to specify when the facts hold and when the goals should be accomplished.
+As always, all the constraints in \eu's library (built-in, or user defined), can be used to specify these restrictions.
 \end{enumerate}
@@
 
-Reservoir resources can have both 'consume' and 'produce' tokens. Note
+\begin{enumerate}
+  \item \textbf{Reservoir} resources can have both 'consume' and 'produce' tokens. Note
 that these tokens do NOT have the usual start/end/duration token
@@
 
-Reusable resources have 'uses' tokens that use a quantity for the
+  \item \textbf{Reusable} resources have 'uses' tokens that use a quantity for the
 duration of the token. These tokens do have the usual start/end/duration
@@
 
-Unary resources are reusable resources with unit quantity and have 'use'
+  \item \textbf {Unary} resources are reusable resources with unit quantity and have 'use'
 tokens (again, with the usual start/end/duration variables).
+\end{enumerate}
 
@@
 
-This Plan Space Planning algorithm is translated into \eus
+This Plan Space Planning algorithm is translated into \eu's
 representation as follows:
@@
 heuristically controlled search by applying orderings for chooseFlaw and
-makeDecisionPoint. It results in a chronologically-backtracking, search.
+makeDecisionPoint. It results in a chronologically-backtracking search.
 

### 2012-02-14 arch.tex
Updated execution plan figures :   - replaced \tau+1 by \tau' to avoid some confusion   - corrected the conflicting case as it was not a conflict previously Few typos and correction in  arch.tex

@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation \comment{Might be good to articulate what ``one'' step
-    looks like}. The returned value indicates if a complete plan has
-  been found. If so this part \comment{what does 'this part' mean? Can
-    we sketch a figure or make it more formal?} of the plan is
-  provided to {\em external} timelines of this reactor.
-
+  deliberation. In the context of \eu for example pone step of
+  deliberation correspod to resolve one sibngle flaw while ensuring
+  that the resulting plan is not proven inconsistent. This implies
+  that it also includes the backtracking of the search in such case. 
+  The returned value by this step function is a boolean which
+  indicates whether the reactor needs some extra step to deliberate or
+  if this last step resulted on a complete plan for the reactor
+  look-ahead. On the later, the part of the plan that concern this
+  reactor {\em external} timelines is provided in order to be
+  dispatched as goals to whichever reactor that is reponsible for
+  these. \comment{By extending I may have saud things I developped
+    below ... will have  to check}
 \end{itemize}
@@
 that is both complete and valid. When such plan is identified
-\comment{how? Is there a terminating condition?} its {\em external}
+\comment{how? Is there a terminating condition? No more flaws but I
+  washoping it would have been defined in \eu section} its {\em external}
 timeline(s) are then dispatched to the corresponding reactors which
@@
 \label{sec:arch:synch}
+
+For the sake of simplicity we introduce first how a reactor can track
+and identify its state in the context where it does not have any
+compelling need to deliberate. Assume here that we have a reactor that
+have no future goal. Even in this context, it is a core requirement in
+T-REX (see \cite{py10}) to have each reactor to produce at every
+single tick their {\em internal} state or at least, for the ones not owning
+{\em internal} timelines, make sure that {\em external} observations
+are taken into account for every tick. This requirement is enforced to
+ensure that all the reactors of an agent share a consistent view of
+the world as time advance.
+
+Taking into account {\em external} state updates is manage d by
+introducing these observation in the plan structure as factsand ensure
+that they remain compatible sewith the current plan mainainted by the
+reactor. The more delicate part of synchronization is to decide of the 
+{\em internla} state values for the current tick. In the Shopping
+example for example we have the following code snippet whihch helps
+identify when we can \texttt{Buy} a product:
+\begin{verbatim}
+Agent::Buy {
+  eq(duration, 10);
+  // [...]
+  contained_by(condition object.location.At currLocation);
+  // [...]
+}
+\end{verbatim}
+If \texttt{object.location} is {\em external} and the \texttt{Agent}
+is {\em internal}. I may encounter the situation where I am at the
+location where I can buy one (possibly among others) product. Already
+there's here  a decision problem related to the state value of the
+\texttt{Agent} timeline where I can either {\em Buy} a specific
+product now, wait a little longer before doing so or just simply not
+buy any product at this location.
+
+In the \eu based reactor we introduce this decision reuqirement 
+as a new type of flaw for the \eu framework. This flaw enforces at the
+reactor to identify fully its internal state for the current tick. By
+using this new flaw we can describe the synchronization process as
+this general sequence :
+
+\begin{enumerate}
+
+\item Integrate the external state as provided by the owner of each
+  external timeline into the plan database.
+
+\item Propagate this information in the plan database following the
+  model $\mathcal{M}$ of this reactor.
+
+\item Resolve the current state value of each internal timelines.
+
+\end{enumerate}
 
@@
 The planning process inside a reactor is directly based on the way
-europa work which was described in previous section. Still some simple
+\eu work which was described in previous section. Still some simple
 alterations were made in order to both include deliberation
@@
   deliberation and synchronization}. For this reason while the overal
-planning of a europa reactor strictly relies on the europa framework
+planning of a \eu reactor strictly relies on the \eu framework
 plnanning solver, its slices the execution of Algorithm
@@
 able to focus on synchronization as a state evluation process that can
-be resolved within the europa framework with few extensions. simliarly
+be resolved within the \eu framework with few extensions. simliarly
 we presented deliberation at this fairly high level not necessarily
@@
 as described by the plan produced. Consider for example the Shopping
-Agent model described during the europa presentation and that the
+Agent model described during the \eu presentation and that the
 location of the \texttt{agent} is an {\em external} timeline. For
@@
 
-Now, consider that at the next tick $\tau+1$ the reactor managing the
+Now, consider that at the next tick $\tau' = \tau+1$ the reactor managing the
 agent location was able to change the location state varaibale to the
@@
 is added in the plan structure as a fact -- starting exactly at
-$\tau+1$ and lastign for an unlknown duration which is at least 1
+$\tau'$ and lastign for an unlknown duration which is at least 1
 tick. As this new observation is integrated in the plan through
@@
 exisiting plan (restricting in turn the start time of \texttt{Going}
-and by propagtion of the constraints \texttt{Go} to be {\em exactly}
-$\tau+1$). While focusing uniquely on the execution frontier ($\tau+1$
-in Fig. \ref{fig:shop:exec1}), the europa solver took benefit of the
+and by propagation of the constraints \texttt{Go} to be {\em exactly}
+$\tau'$). While focusing uniquely on the execution frontier ($\tau'$
+in Fig. \ref{fig:shop:exec1}), the \eu solver took benefit of the
 exisiting plan to evaluate a solution that asumes that the observation
@@
 exisiting plan which help direct the search within this frame. This
-often allow to have a synchronization that is much more directed and
-i, in nominal situation such as the one we showed resolve the
-deliberation in few steps (often without any backtrack in the search).
+often allow to have a synchronization that is much more directed and, 
+in nominal cases such as the one we showed resolve the deliberation 
+in few steps (often without any backtrack in the search).
 
@@
   \caption{Shopping example : Result of synchronization after the {\em
-      Going} observation was received by the reactor with thew plan
+      Going} observation was received by the reactor with the plan
     showed in Fig. \ref{fig:shop:exec0}.}
@@
   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
-  \caption{Illustration of a conflict suring synchronization and the
+  \caption{Illustration of a conflict during synchronization and the
     resulting relaxed plan for recovery} 
@@
 By our design choices -- both at the architecture level and the
-integration of the europa framework for embedded deliberation -- we
+integration of the \eu framework for embedded deliberation -- we
 were able to implement an architecture that avoid this limitation

### 2012-02-14 trunk/arch.tex
Added a figure showing state  flow between 2 reactors

@@
 \item In order to identify its current {\em internal} state values the
-  \textsf{Pilot} \comment{double check this mod} needs to know the
-  current {\em external} state information it relies on. In this case
-  the current \textsf{Waypoint} the vehicle is heading to can be
-  extracted for example from the current \textsf{Nav\_cmd} executed.
+  \textsf{Pilot} needs to know the current {\em external} state 
+  information it relies on. This information is providede in turn by
+  the reactors that provided these timelines as {\em internal}. In
+  this specific case, the current \texttt{Waypoint} where the vehicle
+  is heading can be evaluated based on the \texttt{Nav\_cmd} executed
+  along with usefull \texttt{Sensor\_Data}, both being provided by the
+  \texttt{Vehicle} reactor.
 
@@
 \end{itemize}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.45\columnwidth]{figs/TREX_flow}
+  \vskip-1cm
+  \caption{Illustration of state information flow between the
+    \texttt{Pilot} and the \texttt{Vehicle} reactors.
+    Observations transit in a bottom-up during synchronization 
+    and apply to the execution frontier $\tau$. 
+    Goal information flow from the top to the bottom and concern the
+    desired future values of state variables.}
+  \label{fig:trex:flow}
+\end{figure}
 
@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation. In the context of Europa for example pone step of
+  deliberation. In the context of \eu for example pone step of
   deliberation correspod to resolve one sibngle flaw while ensuring
@@
 \comment{how? Is there a terminating condition? No more flaws but I
-  washoping it would have been defined in europa section} its {\em external}
+  washoping it would have been defined in \eu section} its {\em external}
 timeline(s) are then dispatched to the corresponding reactors which
@@
 
-In the europa based reactor we introduce this decision reuqirement 
-as a new type of flaw for the europa framework. This flaw enforces at the
+In the \eu based reactor we introduce this decision reuqirement 
+as a new type of flaw for the \eu framework. This flaw enforces at the
 reactor to identify fully its internal state for the current tick. By

### 2012-02-14 arch.tex
updates

@@
   each reactor declares its planning look-ahead and expected latency
-  before being able to produce plan to the agent depending on where it
-  is located in a reactor dependency graph such as shown in
+  \comment{$\pi \&  \lambda$ both defined for the first time} before
+  being able to produce plan to the agent depending on where it is
+  located in a reactor dependency graph such as shown in
   Fig. \ref{fig:agent}.
@@
 \end{itemize}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.45\columnwidth]{figs/TREX_flow}
+  \vskip-1cm
+  \caption{\small Illustration of state information flow between the
+    \texttt{Pilot} and the \texttt{Vehicle} reactors.  Observations
+    transit in a bottom-up during synchronization and apply to the
+    execution frontier $\tau$.  Goal information flow from the top to
+    the bottom and concern the desired future values of state
+    variables.\comment{ Fig should be cited in the text}}
+  \label{fig:trex:flow}
+\end{figure}
 
@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation. In the context of \eu for example pone step of
-  deliberation correspod to resolve one sibngle flaw while ensuring
-  that the resulting plan is not proven inconsistent. This implies
-  that it also includes the backtracking of the search in such case. 
-  The returned value by this step function is a boolean which
-  indicates whether the reactor needs some extra step to deliberate or
-  if this last step resulted on a complete plan for the reactor
-  look-ahead. On the later, the part of the plan that concern this
-  reactor {\em external} timelines is provided in order to be
-  dispatched as goals to whichever reactor that is reponsible for
-  these. \comment{By extending I may have saud things I developped
-    below ... will have  to check}
+  deliberation. In the context of \eu for example one deliberation
+  step corresponds to a single flaw resolution (Section
+  \ref{sec:europa:arch}) while ensuring that the resulting plan is not
+  proven inconsistent. This implies that it also includes the
+  backtracking of the search in such a case.  The returned value by
+  the \texttt{step} function is a boolean which indicates whether the
+  reactor needs an extra step to deliberate or if this last step
+  resulted in a complete plan within the reactor's look-ahead. On the
+  later, the part of the plan that concern this reactor {\em external}
+  timelines is provided in order to be dispatched as goals to
+  whichever reactor that is reponsible for these. \comment{By
+    extending I may have saud things I developed below ... will have
+    to check}
 \end{itemize}
@@
 
-For the sake of simplicity we introduce first how a reactor can track
-and identify its state in the context where it does not have any
-compelling need to deliberate. Assume here that we have a reactor that
-have no future goal. Even in this context, it is a core requirement in
-T-REX (see \cite{py10}) to have each reactor to produce at every
-single tick their {\em internal} state or at least, for the ones not owning
-{\em internal} timelines, make sure that {\em external} observations
-are taken into account for every tick. This requirement is enforced to
-ensure that all the reactors of an agent share a consistent view of
-the world as time advance.
+We first show how a reactor can track and identify state when it does
+not have any compelling need to deliberate with no future goals. Even
+in this condition, \rx requires each reactor to produce at every tick
+its {\em internal} state. For those reactors not owning {\em internal}
+timelines, \rx makes sure that {\em external} observations are taken
+into account for every tick as shown in \cite{py10}. Such a
+requirement is enforced to ensure that all reactors share a consistent
+view of the world with advancing time.
 
-Taking into account {\em external} state updates is manage d by
-introducing these observation in the plan structure as factsand ensure
-that they remain compatible sewith the current plan mainainted by the
-reactor. The more delicate part of synchronization is to decide of the 
-{\em internla} state values for the current tick. In the Shopping
-example for example we have the following code snippet whihch helps
-identify when we can \texttt{Buy} a product:
+{\em External} state updates are managed by introducing these
+observation in the plan structure as facts and ensure that they remain
+compatible with the current plan mainainted by the reactor. The
+challenging part of synchronization is to decide the {\em internal}
+state values for the current tick. In the Shopping example for example
+we have the following code snippet which helps identify when we can
+\texttt{Buy} a product:
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 If \texttt{object.location} is {\em external} and the \texttt{Agent}
-is {\em internal}. I may encounter the situation where I am at the
-location where I can buy one (possibly among others) product. Already
-there's here  a decision problem related to the state value of the
-\texttt{Agent} timeline where I can either {\em Buy} a specific
-product now, wait a little longer before doing so or just simply not
-buy any product at this location.
+is {\em internal}, a robot may encounter the situation where it is at
+the location where it can buy one (possibly among others)
+product. This results om a decision problem related to the state value
+of the \texttt{Agent} timeline where the robot can either {\em Buy} a
+specific product now, wait a little longer before doing so or just
+simply not buy any product at this location.
 
-In the \eu based reactor we introduce this decision reuqirement 
-as a new type of flaw for the \eu framework. This flaw enforces at the
-reactor to identify fully its internal state for the current tick. By
-using this new flaw we can describe the synchronization process as
-this general sequence :
+In the \eu based reactor this decision requirement is a new type of
+flaw which forces full identification of internal state of the reactor
+for the current tick. Using this new flaw we can describe the
+synchronization process as:
 
@@
 
-We introduce this requirement as a new type of flaw (Section
-\ref{sec:europa:arch}) for the \eu framework. This flaw forces the
-reactor to identify its internal state for the current tick. By using
-this flaw one can describe the synchronization process as:
+We introduce this requirement as a new type of flaw for the \eu
+framework. This flaw forces the reactor to identify its internal state
+for the current tick. By using this flaw one can describe the
+synchronization process as:
 
@@
 
-While this model is perfectly standard and acceptable in the overall
-case. It can become problematic when tied to execution. Specifically it
-puts a strong tie between the {\em fact} that we are \texttt{Going} to a
-location and the {\em expected} future outcome of ending at this
-location. This ties is even stronger due to the constraint at line 5.
-Knowing this aspect assume that the \texttt{Agent.location} is {\em
-  external} to the reactor which receive the observation
-\texttt{Going(Home,SuperMarket)} which synchronization resolved by
-producing the observation \texttt{Go(Home, SuperMarket)} in the Agent
-timeline. From this point this means that the only possible outcome on
-\texttt{Agent.location} is to be \texttt{At(SuperMarket)} in the
-directly foreseeable future. In the real world the fact that we are
-\texttt{Going} to a certain location does not necessarly imply that we
-will succeed on ending up there. An extreme case could be that due to
-some exogenous event, the road to the \texttt{SuperMarket} is closed
-which could result on an alternate observation \texttt{At(7th st. \& Foo
-  Ave.)}. This observation contradicts thre existing plan the reactor
-maintin and it is important here to relaize the temp[oral direction of
-the causal link that is triggereing this inconsitency. Indeed at the
-time we will observe this contradicting \texttt{At}, the rule that
-generated this conlict is connected to the past observations
-\texttt{Going} and \texttt{Go} which cannot be fully retracted without
-potential global impact in the rest of the exisiting reactors --
-implying a huge cost in backprogation that would be problematic.
+While this model is perfectly acceptable for planning it becomes
+problematic when execution is taken into account especially as it puts
+a strong emphasis between an evolving plan and an unknown future
+(\texttt{Going} to a location and the {\em expected} future outcome of
+ending at this location, ties it strongly to the constraint in line
+$5$ above).  So for example, assume that the \texttt{Agent.location}
+is {\em external} to the reactor which receives the observation
+\texttt{Going(Home,SuperMarket)} which synchronization has resolved by
+producing the observation \texttt{Go(Home, SuperMarket)} in the
+\texttt{Agent} timeline. This would imply that the only possible
+outcome on \texttt{Agent.location} is to be \texttt{At(SuperMarket)}
+in the directly foreseeable future. In the real world the fact that we
+are \texttt{Going} to a certain location does not necessarly imply
+that we will succeed on ending up there. In one extreme case due to an
+exogenous event, the road to the \texttt{SuperMarket} is closed which
+could result in an alternate observation that the robot is
+\texttt{At(7th st. \& Main)}. This observation contradicts the
+existing plan the reactor is maintaining. Further, it is important to
+understand the temporal direction of the \emph{causal link} that is
+triggering the inconsistency. At the time the robot observes this
+contradicting the \texttt{At} predicate, the rule that generated this
+conflict is connected to past observations \texttt{Going} and
+\texttt{Go} which cannot be easily retracted without a potential
+global impact to the plan implying a huge cost in back-propagation
+that would be problematic.
 
-More generally tying a token to future outcomes in a model that is meant
-to be confronted to rela 
-executionis something that is problematic. It may be valid ifg you can
-gurantee that this will always happen (for example the reactor
-managing \texttt{Agent.location} timeline guarantees that the outcome
-of the \texttt{Going} will {\em always} result on ending at the traget
-location) but should be manipulated while keeping in mind that
-predicting the future is not absolute. In this case one could alter
-this part of the model as follow in order to give the reactor more
-flexibility:
+More generally tying a token to future outcomes in a model that is
+meant to be confronted by execution is problematic. It might be valid
+if one can guarantee that an outcome is predisposed to occur (for
+example the reactor managing \texttt{Agent.location} timeline
+guarantees that the outcome of the \texttt{Going} will {\em always}
+result on ending at the traget location). However, given the very
+nature of the planning problem and the ability to search outcomes and
+replan, the model should be written so that a future projection can be
+manipulated. %  while keeping in mind that predicting the future is not
+% absolute. 
+In this case one could alter this part of the model in order to give
+the reactor more flexibility as follows:
 
@@
 
-The alteration made to relax the constraint on linbe 5 from a strict 
-\texttt{equals} to contains and then add the new constraint at line 11
-that ensure that the last \texttt{Going} done during this \texttt{Go}
-ends up in the location the agent wnats to be. This firts level of
-relaxation will therefore allow the agent to usies multiple
-\texttt{Going} should the first one fail to reach our objective.  The
-probalem can even be relaxed further if needed but this relaxation
-gicves already more felxibility for the reactor to recover from local
-failures.
+We do so by relaxing the constraint in line $5$ from a strict
+\texttt{equals} to a \texttt{contains} and to add the new constraint
+at line $11$ which ensures that the last \texttt{Going} ends up in the
+location the agent actually wants to be at. This first level of
+relaxation will therefore allow the agent to use multiple
+\texttt{Going} tokens should the first one fail to reach the
+objective.  % The problem can even be relaxed further if needed but
+% this relaxation gives already more felxibility for the reactor to
+% recover from local failures.
+As the simple example above shows, synchronization involves a
+deliberation process with the risk that the search for a solution is
+not a straightforward sequence of unit decisions. Since such a process
+is required by the architecture at \emph{every} tick for \emph{every}
+reactor the model needs to be carefully designed.
 
-Synchronization remains here a deliberation process, and for this
-reason there's always the risk that the search of a solution is not a
-straightforward sequence of unit decision. As such process is required
-by the architecture at every ticks for every reactors presently
-running this need to be carefully designed. At the engine level we
-made careful choices in term of both the focus of the search --
-limited only to tokens that can potentially overlap the current tick
--- and its heuristic where the internal state flaw is considered last
-during the search. The later proved to limit the amount of backtrack
-in the search for a solution -- or at least how deep we need to
-backtrack in our decision tree -- similarly the sequence in which we
+At the engine level careful choices were made in both the focus of the
+search -- limited only to tokens that can potentially overlap the
+current tick -- and heuristics where the internal state flaw is
+considered last during the search process. The later limits the amount
+of backtracking for a solution -- or at least how deep one needs to
+backtrack in a decision tree. Similarly the sequence in which we
 evaluate the possible resolution for such a flaw -- presented
-previously -- was selected according to simple assumptions that takes
-benefit of a possibly already existing plan by assuming that if the
-outcome of synchronization appears consistent with our current plan
-then it probably reflects that this reactor plan is executed. Such
-assumption helps direct the internal state evaluation and, in nominal
-cases (where we assume that reactors are acting in concert), proved to
-be a sound solution.
+previously -- was selected according to some simple assumptions that % takes
+% benefit of a possibly assume the existing plan by assuming that 
+if the outcome of synchronization appears consistent with our current
+plan then it probably reflects that the plan has executed
+\comment{that the past is monotonic?}. Such an assumption helps direct
+the internal state evaluation and in nominal cases, proves to be a
+sound solution.
 
-One final aspect to keep into account is that in \rx point of view
-synchronization is the only critical task of a reactor. Indeed, such
+One final aspect to take into account is that from a \rx point of
+view, synchronization is the only critical task of a reactor. Such a
 process is used in order for the agent to identify its current state
-and maintain it consistent throughout the agent life-time. A reactor
-not being able to find a consistent solution during synchronization --
-meaning that after evaluating all the possible solutions provided by
-its model it did not found any plan that is not inconsistent -- not
-only expresses that this reactor is not able anymore to explain the
-state of the world but may jeopardize other reactors that relies on
-its {\em internal} state. For this reason a failure to do
-synchronization immediately result on the agent killing this reactor
-and notifying any reactor that depends on its {\em internal} state
-variables by posting the special observation \textsf{Failed}. By doing
-so we allow for graceful degradation where other reactors can readapt
-their plan as they receive this \textsf{Failed} observation.
+and maintain it consistently throughout the agent's lifetime. For a
+reactor not being able to find a consistent solution during
+synchronization -- implying that after evaluating all possible
+solutions provided by a model it was unable to find a consistent plan
+-- not only indicates that a reactor is unable to explain the state of
+the world, but more importantly it may jeopardize other reactors that
+rely on its {\em internal} state. For this reason a failure to
+synchronize immediately results in the agent killing a reactor and
+notifying any reactor that depends on its {\em internal} state
+variables; it does so by posting the special observation
+\textsf{Failed}. And by doing so we allow for graceful degradation
+where other reactors can readapt their plans as they receive this
+observation.
 

### 2012-02-14 trunk/arch.tex
Some tweaking in the figures

@@
   \centering
- \includegraphics[scale=0.45]{figs/AUV-agent.pdf}
- \caption{\small A \rx agent is composed of multiple reactors or
+  \subfloat[\small A \rx agent is composed of multiple reactors or
     control loops (rounded boxes) which are connected through state
     variables provided by one reactor ({\color{blue}blue} solid line)
-   with multiple possible clients ({\color{green}green} dashed lines)}
+    with multiple possible clients ({\color{green}green} dashed
+    lines)]{\includegraphics[scale=0.45]{figs/AUV-agent.pdf}
   \label{fig:agent}
+  }
+  \qquad
+  \subfloat[\small State information flow between the
+    \texttt{Pilot} and the \texttt{Vehicle} reactors.
+    Observations transit as a bottom-up flow 
+    and apply to the execution frontier $\tau$. 
+    Goal information flow from the top to the bottom and relates to 
+    desired future values of state variables.]{
+      \includegraphics[width=0.4\columnwidth]{figs/TREX_flow}
+      %\vskip-1cm
+      \label{fig:trex:flow}
+    }
+    \caption{\rx connection of reactors through state variables based on the
+    AUV agent we use at sea.}
 \end{figure}
@@
   \textsf{Nav\_cmd} states that should eventually help it reach this
-  location. \comment{this explanation could do with a figure which I
-    believe we have used previously}
+  location. %\comment{this explanation could do with a figure which I
+    %believe we have used previously}
 
 \end{itemize}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.45\columnwidth]{figs/TREX_flow}
-  \vskip-1cm
-  \caption{Illustration of state information flow between the
-    \texttt{Pilot} and the \texttt{Vehicle} reactors.
-    Observations transit in a bottom-up during synchronization 
-    and apply to the execution frontier $\tau$. 
-    Goal information flow from the top to the bottom and concern the
-    desired future values of state variables.}
-  \label{fig:trex:flow}
-\end{figure}
 
@@
 reactors. Such a formalism also constrains reactor interaction to
-state information exchange which comes in two forms:
+state information exchange which comes in two forms as illustrated in
+Figure \ref{fig:trex:flow}:
 
@@
   of the world throughout all the reactors as time
-  advances. \comment{could do with a figure here}
+  advances. %\comment{could do with a figure here}
 
@@
   as a goal to the owner of the given {\em external} state
-  variable(s). \comment{could do with a figure here}
+  variable(s). %\comment{could do with a figure here}
 

### 2012-02-14 trunk/arch.tex
updates

@@
   The planner is not concerned with any change that may occur in
-  $\Sigma$ \footnote{the world modeled by the plan domain} while it is
+  $\Sigma$ \footnote{The world modeled by the plan domain.} while it is
   planning; it plans for the given initial and goal states regardless
@@
 
-\subsubsection{Deliberation : planning for future state evolution}
+\subsubsection{Deliberation: Planning for future state evolution}
 \label{sec:arch:plan}
 
-The planning process inside a reactor is directly based on the way
-\eu work which was described in previous section. Still some simple
-alterations were made in order to both include deliberation
-information one reactor provide to the architecture -- such as its
-deliberation latency and its planning look-ahead both expressed in
-ticks -- and the fact that we potentially need to interleave this
-deliberation with not the synchronization process but also with the
-deliberation of other reactors\footnote{Current implementation of \rx
-  is running on a single process and it is the responsibility of the
-  agent itself to emulate the multi-threading of the reactors
-  deliberation and synchronization}. For this reason while the overal
-planning of a \eu reactor strictly relies on the \eu framework
-plnanning solver, its slices the execution of Algorithm
-\ref{alg:europa:solve} in atomic steps. The interruption is done
-around the recursive call (line \ref{li:recurse}) while always
-ensuring that the current partial plan mainitained is never proven
-inconsistant at any single step. This can be managed by allowing the
-call to backtrack until a DecisionPoInt that is not exhausted (ie
-$decision.hasNext()$ is \texttt{true}) is find in the call stack.
+Deliberation inside a reactor is based on \eu capabilities described
+in Section \ref{sec:europa:arch}. However deliberation within \rx
+comes with some augmentation -- deliberation information one reactor
+provide to the architecture such as its deliberation latency and its
+plan look-ahead both expressed in ticks -- and the fact that we
+potentially need to interleave deliberation with not just the
+synchronization process but also with deliberation of other
+reactors\footnote{The current implementation of \rx is running on a
+  single process and it is the responsibility of the agent itself to
+  emulate reactor multi-threading for deliberation and
+  synchronization.}. For this reason while the overall planning of a
+deliberative reactor strictly relies on the \eu framework planning
+solver, it slices the execution of Algorithm \ref{alg:europa:solve} in
+atomic steps. The interruption is done around the recursive call (line
+\ref{li:recurse}) while ensuring that the current partial plan is
+never proven inconsistant at any single step. This can be managed by
+allowing the call to backtrack until a \textsf{DecisionPoInt} that is
+not exhausted (\ie $decision.hasNext()$ is \texttt{true}) is found in
+the call stack.
 
-One critical aspect of this deliberation step is to ensure that at the
-exit of this call the $plan$ is not proven inconsistent. Indeed, it is
-always possible that the next call made by the agent is the
-synchronization which would immediately fail if the $plan$ is
-inconsistent before any solving attempt made for synchronization.  For
-this reason if the plan is proven inconsistent during a deliberation
-step we either backtrack in the decision stack until we find a node fo
-the tree that have remaining alternate solution or simply fully relax
-the plan in situation there's absolutely no other alternative (by
-removing all the goals and keeping only recent observations
-produced). These constraints are yet again related to the fact that
-for the agent the only crucial functionality of a reactor is to
-resolve successfully its synchronization.
+A critical aspect of this deliberation step is to ensure that at the
+exit of this call the $plan$ is not inconsistent. % Indeed, it is always
+% possible that the next call made by the agent is for synchronization
+% which could fail if the $plan$ is inconsistent before any attempt is
+% made for synchronization.
+If the partial plan is found to be inconsistent during a deliberation
+step we either backtrack in the decision stack until a node of the
+tree is found with an alternate solution or the plan is relaxed since
+there is no other alternative solution. We do this by removing all
+goals and keeping only recent observations. % These constraints are yet
+% again related to the fact that for the agent the only crucial
+% functionality of a reactor is to resolve successfully its
+% synchronization.
 
-The number of flaws between two steps can evolve for the following
-reasons that are exogenous to deliberation :
+The number of flaws between two steps \comment{between deliberation
+  and synch.?} can evolve for the following reasons that are exogenous
+to deliberation:
 
@@
 
-\item A new goal has been posted on one of the reactor timeline. This
-  is usually due to another reactor finding a plan and posting the
-  corresponding new objective to this reactor.
+\item A new goal has been posted on a reactor timeline. This is
+  usually due to another reactor finding a partial plan and posting
+  the corresponding new objective.
 
@@
 \item A previously requested goal has be recalled by the initial
-  requester. This can happen when the reactor did identify that is
-  initial plan is no longer valid. For example as synchronization
-  proved this plan to be inconsistent.
+  requester. This can occur when the reactor's initial plan is no
+  longer valid, for instance with synchronization resulting in
+  inconsistency.
 
@@
 
-These 3 events will potentially produce new flaws to be resolved while
-any single deliberation step will usually reduce the number of flaws
-to be resolved.
-
-The scope of planning for the reactor plays also a role in the
-deliberation process as we limit deliberation to the sliding temporal
-window $[\tau+\lambda, \tau+\lambda+\pi]$ where:
+\comment{this might be a subsubsection before someplace since it
+  introduces the concept of latency/lookahead.}
+These events could potentially produce new flaws to be resolved while
+any single deliberation step will likely reduce the number of flaws
+for resolution. The scope of planning for a reactor also plays a role
+in the deliberation process as we limit deliberation to the sliding
+temporal window $[\tau+\lambda, \tau+\lambda+\pi]$ \cite{py10} where:
 
@@
 
-\item $\tau$ is the current tick
+\item $\tau$ is the current tick indicative of the execution frontier
 
-\item $\lambda$ is the specified latency of this reactor. Which is an
-  indicator on the expected maximum time this reactor is expected
-  before being able to produce a plan.\footnote{Do note that this
-    parameter remains indicative and, while it is highly recommended
-    to select a sound value, a failure to produce a plan in this delay
-    is not considered as a critical failure of the reactor}
+\item $\lambda$ is the specified latency of this reactor, which is an
+  indicator of the maximum expected time for a reactor to produce a
+  plan.\footnote{Note that while it is highly recommended to select a
+    sound value, a failure to produce a plan within a chosen value for
+    this parameter is not considered as a critical failure of the
+    reactor.}
 
 \item $\pi$ is the planning look-ahead of the reactor. which indicates
-  how many ticks in the future this reactor is looking ahead while planning.
+  how many ticks in the future this reactor is looking ahead during
+  deliberation.
 
@@
 
-This scope allow to filter out any tokens that are either necessarily
-ending before $\tau$ or necessarily starting after. Focusing the
-[planning problem to only the tokens that are in a reasonable
-future. This window is taken into account by the agent which will
-notify the reactor of new {\em internal} goals only when they overlap
-this window but also at the deliberation solver in order to reduce the
-number of tokens to be evaluated and consequently reduce the cost of
-deliberation. As time advance flaws that where initially beyond this
-window will become active flaw that the reactor can then evaluate
-resulting on an apparent continuous planning of the reactor.
-
-When there are no more flaw is present in the current planning scope a
-plan solution is found. This has two effects:
+Such a scope allows filtering of tokens that are either necessarily
+ending before $\tau$ or necessarily starting after, focusing the
+planning problem on only those tokens that are within a ``reasonable''
+future. This window is taken into account by the \rx agent which will
+notify a reactor of new {\em internal} goals not only when they
+overlap this reactor specific window but also for the deliberation
+solver in order to reduce the number of tokens to be evaluated to
+reduce deliberation cost and performance. As time advances flaws that
+were initially beyond this temporal horizon will become active which a
+reactor can evaluate. This results in an apparent \emph{continuous
+  planning} of the reactor. When there are no more flaws present in
+the current planning scope a partial plan solution is found. This has
+two effects:
 
@@
 
-\item This reactor does not need to deliberate until the next synchronization.
+\item This reactor does not need to deliberate until the next
+  synchronization.
 
-\item The {\em external} part of this plan can then be posted which
-  eventually (at the end of the current tick if these goals overlap
-  the owner planning window) will be dispatched as goals triggering
-  new deliberation on these reactors.
+\item The token on the {\em external} timelines of this partial plan
+  can then be posted which eventually (at the end of the current tick
+  if these goals overlap the owner reactor's planning window) will be
+  dispatched as goals, in turn triggering further deliberation.
 

### 2012-02-15 trunk/arch.tex
Refined  some text. Ensured that some definitions (such as tau , lambda and pi) are at least mentioned early Reworked the synchronization with a focus on the current state flaw definition and proposed resolution

@@
 adaptations at the lower-level reactive layers in contrast are managed
-by specialized components (for example GeNoM \comment{} and CLARATy
-\comment{}) that rely on different representations for modeling local
+by specialized components (for example LAAS architecture
+\cite{alami:1998p820,Ingrand07} and CLARATy \cite{Nesnas:2003do}
+) that rely on different representations for modeling local
 control behavior or worse do not have any explicit formal agent
@@
   of high level waypoints into low-level commands) and temporally as
-  each reactor declares its planning look-ahead and expected latency
-  \comment{$\pi \&  \lambda$ both defined for the first time} before
+  each reactor declares its planning look-ahead $\pi$  and expected
+  latency $\lambda$ 
+  % \comment{$\pi \&  \lambda$ both defined for the first time} 
+  before
   being able to produce plan to the agent depending on where it is
@@
 
-This {\em synchronization} process propagates up through the reactor
-dependency graph bottom up at every {\em tick} ensuring that all the
-reactors have a consistent view of the world. Conversely each reactor
-can request a future change on one of its {\em external} state
-variables which will be transfromed as an {\em internal} goal to the
-owner. Such \emph{goal dispatching} is expected to produce a
+This {\em synchronization} process propagates the current state
+information at the execution frontier $\tau$ throughout all the 
+reactors. The sequencing of the reactors synchronization is identified
+by traversing the reactor graph from the least dependents reactors --
+the ones with no {\em external} timelines -- to the most dependent
+ones which typicallu do not provide {\em internal} timelines.
+Conversely each reactor can request a future change on one of 
+its {\em external} state variables which will be transfromed as 
+an {\em internal} goal to the owner. 
+Such \emph{goal dispatching} is expected to produce a
 deliberation phase for the receiving reactor which can cyclically and
-in turn, produce a plan impacting its {\em external} state variables
-that will propagate down following the same mechanism.
+in turn, produce a plan impacting its {\em external} state variables' 
+future that will propagate down following the same mechanism.
 
@@
 
-We first show how a reactor can track and identify state when it does
-not have any compelling need to deliberate with no future goals. Even
+We first show how a reactor can track and identify its state in the
+context where it does not have any compelling need to deliberate. We
+assume that we have a reactor that has no future goal.Even
 in this condition, \rx requires each reactor to produce at every tick
@@
 
-We first show how a reactor can track and identify its state in the
-context where it does not have any compelling need to deliberate. We
-assume that we have a reactor that has no future goal. \comment{Need
-  to develop the reason why it is a necessity: primarily to provide
-  its internal state to whoever observe it but also simply to ensure
-  that its current representation of the world is up to date and still
-  consistent}
+We introduce this requirement as a new type of flaw for the \eu
+framework defined as follow: 
+\begin{definition}
+  \label{def:csf}
+  A {\em current state flaw} indicate that one of the reactor {\em
+    internal} timelines does not provide a fully defined state value
+  for the execution frontier $\tau$. Such flaw can be due to one 
+  of the 2 following reasons:
+  \begin{itemize}
+  \item The absence of a token in this timeline which overlaps the
+    instant $\tau$. 
+  \item An open decision between terminating the last past token for
+    this timeline or pushing its end in the future.
+  \end{itemize}
+\end{definition}
 
-We introduce this requirement as a new type of flaw for the \eu
-framework. This flaw forces the reactor to identify its internal state
+The core idea bewteen this flaw is to ensure that at any single tick
+the {\em synchronization} solver will identifies all the {\em
+  internal} timelines for which the current state value is yet to be
+identified. We handle their resolution based on following
+proposition. 
+\begin{proposition}
+  \label{prop:csf:resolve}
+  A {\em current state flaw} can  be resolved using one of the
+  following choices which are eveluated accroding to this order:
+\begin{enumerate}
+
+\item Extend the previous state value to end after the execution 
+  frontier  (\ie restrict its end time to $[\tau+1, \infty)$). %\comment{first use of
+%    $\tau$, so needs to be defined earlier}
+
+\item Start the next active token in the timeline 
+  by restricting its start time to the single value $\tau$
+
+\item Create and insert a new token in this timeline that will start
+  at the current tick $\tau$ (attempt this for each possible token
+  type for this timeline if necessary).
+
+\end{enumerate}
+\end{proposition}
+Intuitively the first choice enforces an intertail value asumption
+where the reactor will tend to maintain its current as long as the
+model and external observations allow it. The second choice 
+favorise the adavnce of the reactor current plan if any; this direct
+the {\em synchronization} decision toward assuming that its current
+plan is properly executed. Finally if none of these two choices did
+provide a valid solution -- meaning that both lead to an inconsittant
+plan -- the reactor tries all the possible state values remaining for
+this timeline until it finds one that allows it to identify its
+state as required by \rx architecture.
+
+This flaw forces the reactor to identify its internal state
 for the current tick. By using this flaw one can describe the
@@
 
-\item Resolve the current state value of each internal timelines.
-
+\item Resolve the current state value of each internal timelines based
+  on Proposition \ref{prop:csf:resolve}.
 \end{enumerate}
 
-An internal state flaw can be resolved using one of the following
-choices which are evaluated in order:
-
-\begin{enumerate}
-
-\item Extend the previous state value to end after this tick (\ie
-  restrict its end time to $[\tau+1, \infty)$). \comment{first use of
-    $\tau$, so needs to be defined earlier}
-
-\item Start the next active token in the timeline 
-  by restricting its start time to the single value $\tau$
-
-\item Create and insert a new token in this timeline that will start
-  at the current tick $\tau$ (attempt this for each possible token
-  type for this timeline if necessary).
-
-\end{enumerate}
-
-All of these choices are evaluated sequentially until a consistent
-solution with no flaws at this tick, is identified. In order to do so
-we need to make the assumption that the current state value of an
-internal timeline does not depend on the future or more accurately
+The steps 2 \& 3 are handled by the \eu solver from algorithm
+\ref{alg:europa:solve}. We only enforce that the {\em current state
+  flaws} are selected last by the algorithm in order to follow the
+sequencing given above. As {\em synchronization} focusses on the
+execution frontier wee also filter out all flaws that are outside of
+the current tick scope. This mean that if this flaw relates to a token
+that is {\em necessarily} ending before $\tau$ or starting after 
+$\tau+1$ then the solver will ignore it. This helps to reduce the
+scope of the synchronization to exclusively the current state of the
+reactor and reduce dramatically the search space for the {\em
+  synchronization} solver. In order to do so we need to make the
+assumption that the current state value of an
+{\em internal} timeline does not depend on the future or more accurately
 that any choices made during this synchronization will not lead to a

### 2012-02-15 arch.tex
updates

@@
   of high level waypoints into low-level commands) and temporally as
+
   each reactor declares its planning look-ahead and expected latency
-  \comment{$\pi \&  \lambda$ both defined for the first time} before
-  being able to produce plan to the agent depending on where it is
-  located in a reactor dependency graph such as shown in
+  \comment{$\pi \& \lambda$ both defined here for the first time}
+  before being able to produce plan to the agent depending on where it
+  is located in a reactor dependency graph such as shown in
   Fig. \ref{fig:agent}.
@@
   \textsf{Nav\_cmd} states that should eventually help it reach this
-  location. \comment{this explanation could do with a figure which I
-    believe we have used previously}
+  location. %\comment{this explanation could do with a figure which I
+    %believe we have used previously}
 
 \end{itemize}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.45\columnwidth]{figs/TREX_flow}
-  \vskip-1cm
-  \caption{\small Illustration of state information flow between the
-    \texttt{Pilot} and the \texttt{Vehicle} reactors.  Observations
-    transit in a bottom-up during synchronization and apply to the
-    execution frontier $\tau$.  Goal information flow from the top to
-    the bottom and concern the desired future values of state
-    variables.\comment{ Fig should be cited in the text}}
-  \label{fig:trex:flow}
-\end{figure}
 
@@
 solution that relaxes this static world assumption. \comment{Need to
-  talk about IDEA and the fact that they had similar aspect even though
-  they were never very clear nor well articulated in their papers}
+  talk about IDEA and the fact that they had similar aspect even
+  though they were never very clear nor well articulated in their
+  papers}
 
@@
 
-We first show how a reactor can track and identify state when it does
-not have any compelling need to deliberate with no future goals. Even
-in this condition, \rx requires each reactor to produce at every tick
-its {\em internal} state. For those reactors not owning {\em internal}
+We first show how a reactor can track and identify its state in the
+context where it does not have any compelling need to deliberate. We
+assume that we have a reactor that has no future goal. Even in this
+condition, \rx requires each reactor to produce at every tick its {\em
+  internal} state. For those reactors not owning {\em internal}
 timelines, \rx makes sure that {\em external} observations are taken
@@
 
-We first show how a reactor can track and identify its state in the
-context where it does not have any compelling need to deliberate. We
-assume that we have a reactor that has no future goal. \comment{Need
-  to develop the reason why it is a necessity: primarily to provide
-  its internal state to whoever observe it but also simply to ensure
-  that its current representation of the world is up to date and still
-  consistent}
+We introduce this requirement as a new type of flaw for the \eu
+framework defined as follows:
 
-We introduce this requirement as a new type of flaw for the \eu
-framework. This flaw forces the reactor to identify its internal state
+\begin{definition}
+  \label{def:csf}
+  A {\em current state flaw} indicates that a reactor {\em internal}
+  timeline does not provide a fully defined state value for the
+  execution frontier $\tau$. Such a flaw can be result of either of
+  the following reasons:
+
+  \begin{itemize}
+  \item The absence of a token in this timeline which overlaps the
+    instant $\tau$.
+  \item An open decision between terminating the last token for this
+    timeline or pushing its end in the future.
+  \end{itemize}
+
+\end{definition}
+
+The core idea between this flaw is to ensure that at any single tick
+the {\em synchronization} solver will identify all the {\em internal}
+timelines for which the current state value is yet to be
+identified. We handle their resolution based on the following
+proposition.
+
+\begin{proposition}
+  \label{prop:csf:resolve}
+  A {\em current state flaw} can be resolved using one of the
+  following choices which are evaluated in the following order:
+
+  \begin{enumerate}
+
+  \item Extend the previous state value to end after the execution 
+    frontier  (\ie restrict its end time to $[\tau+1, \infty)$). %\comment{first use of
+    % $\tau$, so needs to be defined earlier}
+
+  \item Start the next active token in the timeline by restricting its
+    start time to the single value $\tau$
+
+  \item Create and insert a new token in this timeline that will start
+    at the current tick $\tau$ (attempt this for each possible token
+    type for this timeline if necessary).
+
+\end{enumerate}
+\end{proposition}
+
+Intuitively the first choice enforces an intertail value assumption
+where the reactor will tend to maintain its current as long as the
+model and external observations allow it. The second choice favorise
+the adavnce of the reactor current plan if any; this direct the {\em
+  synchronization} decision toward assuming that its current plan is
+properly executed. Finally if none of these two choices did provide a
+valid solution -- meaning that both lead to an inconsittant plan --
+the reactor tries all the possible state values remaining for this
+timeline until it finds one that allows it to identify its state as
+required by \rx architecture.
+
+This flaw forces the reactor to identify its internal state
 for the current tick. By using this flaw one can describe the
@@
 
-\item Resolve the current state value of each internal timelines.
-
+\item Resolve the current state value of each internal timelines based
+  on Proposition \ref{prop:csf:resolve}.
 \end{enumerate}
 
-An internal state flaw can be resolved using one of the following
-choices which are evaluated in order:
-
-\begin{enumerate}
-
-\item Extend the previous state value to end after this tick (\ie
-  restrict its end time to $[\tau+1, \infty)$). \comment{first use of
-    $\tau$, so needs to be defined earlier}
-
-\item Start the next active token in the timeline 
-  by restricting its start time to the single value $\tau$
-
-\item Create and insert a new token in this timeline that will start
-  at the current tick $\tau$ (attempt this for each possible token
-  type for this timeline if necessary).
-
-\end{enumerate}
-
-All of these choices are evaluated sequentially until a consistent
-solution with no flaws at this tick, is identified. In order to do so
-we need to make the assumption that the current state value of an
-internal timeline does not depend on the future or more accurately
-that any choices made during this synchronization will not lead to a
-future inconsistency. Such an assumption impacts the set of possible
-domains the reactor can support while remaining complete. Take for
-example this rule from our \texttt{Shopping} model :
+Steps $2$ \& $3$ are handled by the \eu solver from algorithm
+\ref{alg:europa:solve}. We only enforce that the {\em current state
+  flaws} are selected last by the algorithm in order to follow the
+sequencing given above. As {\em synchronization} focusses on the
+execution frontier wee also filter out all flaws that are outside of
+the current tick scope. This mean that if this flaw relates to a token
+that is {\em necessarily} ending before $\tau$ or starting after
+$\tau+1$ then the solver will ignore it. This helps to reduce the
+scope of the synchronization to exclusively the current state of the
+reactor and reduce dramatically the search space for the {\em
+  synchronization} solver. In order to do so we need to make the
+assumption that the current state value of an {\em internal} timeline
+does not depend on the future or more accurately that any choices made
+during this synchronization will not lead to a future
+inconsistency. Such an assumption impacts the set of possible domains
+the reactor can support while remaining complete. Take for example
+this rule from our \texttt{Shopping} model:
 
@@
 While this model is perfectly acceptable for planning it becomes
-problematic when execution is taken into account especially as it puts
-a strong emphasis between an evolving plan and an unknown future
+problematic when execution is taken into account especially since it
+puts a strong emphasis between an evolving plan and an unknown future
 (\texttt{Going} to a location and the {\em expected} future outcome of
@@
 
+% \begin{minipage}[t]{\textwidth}
+% \begin{figure}
 \begin{verbatim}
@@
 \end{verbatim}
+% \label{lab:model:shopping}
+% \caption{\small Model fragment to enforce a \texttt{Going} end condition}
+% \end{figure}
+% \end{minipage}
 
@@
 One final aspect to take into account is that from a \rx point of
-view, synchronization is the only critical task of a reactor. Such a
-process is used in order for the agent to identify its current state
-and maintain it consistently throughout the agent's lifetime. For a
-reactor not being able to find a consistent solution during
-synchronization -- implying that after evaluating all possible
+view, \emph{synchronization is the only critical task of a
+  reactor}. Such a process is used in order for the agent to identify
+its current state and maintain it consistently throughout the agent's
+lifetime. For a reactor not being able to find a consistent solution
+during synchronization -- implying that after evaluating all possible
 solutions provided by a model it was unable to find a consistent plan
@@
 
-\subsubsection{Deliberation : planning for future state evolution}
+\subsubsection{Deliberation: Planning for future state evolution}
 \label{sec:arch:plan}
 
-The planning process inside a reactor is directly based on the way
-\eu work which was described in previous section. Still some simple
-alterations were made in order to both include deliberation
-information one reactor provide to the architecture -- such as its
-deliberation latency and its planning look-ahead both expressed in
-ticks -- and the fact that we potentially need to interleave this
-deliberation with not the synchronization process but also with the
-deliberation of other reactors\footnote{Current implementation of \rx
-  is running on a single process and it is the responsibility of the
-  agent itself to emulate the multi-threading of the reactors
-  deliberation and synchronization}. For this reason while the overal
-planning of a \eu reactor strictly relies on the \eu framework
-plnanning solver, its slices the execution of Algorithm
-\ref{alg:europa:solve} in atomic steps. The interruption is done
-around the recursive call (line \ref{li:recurse}) while always
-ensuring that the current partial plan mainitained is never proven
-inconsistant at any single step. This can be managed by allowing the
-call to backtrack until a DecisionPoInt that is not exhausted (ie
-$decision.hasNext()$ is \texttt{true}) is find in the call stack.
+Deliberation inside a reactor is based on \eu capabilities described
+in Section \ref{sec:europa:arch}. However deliberation within \rx
+comes with some augmentation -- deliberation information one reactor
+provide to the architecture such as its deliberation latency and its
+plan look-ahead both expressed in ticks -- and the fact that we
+potentially need to interleave deliberation with not just the
+synchronization process but also with deliberation of other
+reactors\footnote{The current implementation of \rx is running on a
+  single process and it is the responsibility of the agent itself to
+  emulate reactor multi-threading for deliberation and
+  synchronization.}. For this reason while the overall planning of a
+deliberative reactor strictly relies on the \eu framework planning
+solver, it slices the execution of Algorithm \ref{alg:europa:solve} in
+atomic steps. The interruption is done around the recursive call (line
+\ref{li:recurse}) while ensuring that the current partial plan is
+never proven inconsistant at any single step. This can be managed by
+allowing the call to backtrack until a \textsf{DecisionPoInt} that is
+not exhausted (\ie $decision.hasNext()$ is \texttt{true}) is found in
+the call stack.
 
-One critical aspect of this deliberation step is to ensure that at the
-exit of this call the $plan$ is not proven inconsistent. Indeed, it is
-always possible that the next call made by the agent is the
-synchronization which would immediately fail if the $plan$ is
-inconsistent before any solving attempt made for synchronization.  For
-this reason if the plan is proven inconsistent during a deliberation
-step we either backtrack in the decision stack until we find a node fo
-the tree that have remaining alternate solution or simply fully relax
-the plan in situation there's absolutely no other alternative (by
-removing all the goals and keeping only recent observations
-produced). These constraints are yet again related to the fact that
-for the agent the only crucial functionality of a reactor is to
-resolve successfully its synchronization.
+A critical aspect of this deliberation step is to ensure that at the
+exit of this call the $plan$ is not inconsistent. % Indeed, it is always
+% possible that the next call made by the agent is for synchronization
+% which could fail if the $plan$ is inconsistent before any attempt is
+% made for synchronization.
+If the partial plan is found to be inconsistent during a deliberation
+step we either backtrack in the decision stack until a node of the
+tree is found with an alternate solution or the plan is relaxed since
+there is no other alternative solution. We do this by removing all
+goals and keeping only recent observations. % These constraints are yet
+% again related to the fact that for the agent the only crucial
+% functionality of a reactor is to resolve successfully its
+% synchronization.
 
-The number of flaws between two steps can evolve for the following
-reasons that are exogenous to deliberation :
+The number of flaws between two steps \comment{between synch. and
+  execution?} can evolve for the following reasons that are exogenous
+to deliberation:
 
@@
 
-This scope allow to filter out any tokens that are either necessarily
-ending before $\tau$ or necessarily starting after. Focusing the
-[planning problem to only the tokens that are in a reasonable
-future. This window is taken into account by the agent which will
-notify the reactor of new {\em internal} goals only when they overlap
-this window but also at the deliberation solver in order to reduce the
-number of tokens to be evaluated and consequently reduce the cost of
-deliberation. As time advance flaws that where initially beyond this
-window will become active flaw that the reactor can then evaluate
-resulting on an apparent continuous planning of the reactor.
-
-When there are no more flaw is present in the current planning scope a
-plan solution is found. This has two effects:
+\comment{Insert a figure here [AAMAS?] showing $\pi \lambda$} Such a
+scope allows filtering of tokens that are either necessarily ending
+before $\tau$ or necessarily starting after, focusing the planning
+problem on only those tokens that are within a ``reasonable''
+future. This window is taken into account by the \rx agent which will
+notify a reactor of new {\em internal} goals not only when they
+overlap this reactor specific window but also for the deliberation
+solver in order to reduce the number of tokens to be evaluated to
+reduce deliberation cost and performance. As time advances flaws that
+were initially beyond this temporal horizon will become active which a
+reactor can evaluate. This results in an apparent \emph{continuous
+  planning} of the reactor. When there are no more flaws present in
+the current planning scope a partial plan solution is found. This has
+two effects:
 
@@
 
-A core aspect of this reactor is that it takes advantage of the
-specific sequencing enforced by the \rx architecture along with the
-fact that only one $plan$ structure is maintained in the reactor and
-shared between synchronization and planning to allow these two
-processes to not only be interleaved but impact each other through the
-way they manipulate this plan.
+One key aspect of the \rx framework is intertwined plan execution. As
+seen above, it is synchronization which drives planning, so the
+co-mingling between these two processes is critical to maintaining
+state within a robot immersed in a dynamic environment. Also important
+is that \rx maintains 
+% A core aspect of this reactor is that it takes advantage of the
+% specific sequencing enforced by the \rx architecture along with the
+% fact that only 
+one $plan$ structure in the reactor and shared between synchronization
+and planning. This allows these two processes to not only be
+interleaved but how they can impact each other via plan manipulation.
 
-The most obvious case is how synchronization allow propagate of {\em
-  external} observations as a new tick occur in the plan. In or
-presentation of synchronization, we isolated the problem by stating
-that we will assume at this point that the reactorhas no future goal
-or more accurateely no plan to enforce at this stage. By doing so we
-were able to develop the way we augment yhe europpa solver to be able
-to propagete {\em external} observations in order to identify the
-reactor {\em internal} state for the same tick. By doing so we were
-able to focus on synchronization as a state evluation process that can
-be resolved within the \eu framework with few extensions. simliarly
-we presented deliberation at this fairly high level not necessarily
-discussing how the disruptive nature of synchronization in this
-process could impact the way this planning process will evolve.
+The most obvious case is how synchronization allows propagate of {\em
+  external} observations when a new tick occurs. In our presentation
+of synchronization in Section \ref{sec:arch:synch}, we isolated the
+problem by assuming that the reactor has no future goal or more
+accurateely no plan to enforce.  By doing so we were able to develop
+the way we augment the \eu solver to be able to propagete {\em
+  external} observations in order to identify the reactor {\em
+  internal} state for the same tick. By doing so we were able to focus
+on synchronization as a state evaluation process that can be resolved
+within the \eu framework with few extensions. Similarly we presented
+deliberation as an abstraction while not detailing how the disruptive
+nature of synchronization could impact the way planning could evolve.
 
-We first anlyse the case during which synchronization occurs after the
-planning has vcompleted (ie the last seep of deliberation resulted
-into a plan that is considered as complete for the current
-look-ahead). As synchronization starts with the plan produced this
-will gave a frame for the search process. Therefore this deliberation
-will not only be a model-based estimation uniquely based on the {\em
-  external} observations but will also include the reactor intentions
-as described by the plan produced. Consider for example the Shopping
-Agent model described during the \eu presentation and that the
-location of the \texttt{agent} is an {\em external} timeline. For
-example as, the reactor add the goal to have \texttt{Own} milk and the
-current {\em external} states indicated that it was \texttt{At(home)}
-it produced the plan in Fig. \ref{fig:shop:exec0}. The {\em external}
-part of the plan -- namely the tokens \texttt{Going(home,
-  superMarket)} and \texttt{At(SuperMarket)} -- and integrated as
-goals for the reactor managing this specific state varaiable {\em
-  internally}.
+We first analyze the case during which synchronization occurs after
+planning has completed (\ie the last step of deliberation results in a
+plan that is considered complete for the current look-ahead). As
+synchronization starts with a plan % produced this will gave a frame for
+% the search process. Therefore 
+this deliberation will not only be a model-derived estimation uniquely
+based on {\em external} observations but will also include reactor
+intentions specified in the available partial plan. Consider for
+example the Shopping Agent model described where the \texttt{Agent} is
+an {\em external} timeline. As the reactor adds the goal to
+\texttt{Own} milk and the current {\em external} timeline indicates
+that the robot was \texttt{At(home)}, it produces a plan as shown in
+Fig. \ref{fig:shop:exec0}. The {\em external} part of the plan --
+namely the tokens \texttt{Going(home, superMarket)} and
+\texttt{At(SuperMarket)} -- are installed as goals for the reactor
+managing this state variable {\em internally}.
 
@@
   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}
-  \caption{Shopping example : plan produced before synchronization. 
-    Tokens in red indicate observations, 
-    tokens in blue the goal that this plan attempt to solve. Arrows
+  \caption{\small Shopping example: A plan produced before
+    synchronization.  Tokens in red indicate observations, tokens in
+    blue the goal that this plan attempts to solve for. Arrows
     indicate temporal constraints.}
@@
 
-Now, consider that at the next tick $\tau' = \tau+1$ the reactor managing the
-agent location was able to change the location state varaibale to the
+Consider that at the next tick $\tau' = \tau+1$ the reactor managing
+agent location is able to change the location state variable to
 \texttt{Going}. \rx informs our reactor of this new observation which
 is added in the plan structure as a fact -- starting exactly at
-$\tau'$ and lastign for an unlknown duration which is at least 1
+$\tau'$ and lasting for an unknown duration which is at least 1
 tick. As this new observation is integrated in the plan through
-synchronization the solver can identify the similarity between this
-new observation oand the next token in the existing plan. Which mean
-that instead of making a blind exhaustive search of all the possible
-implication the synchronization process starts its execution with the
-assumption that these new observations are consitent with the current
-plan maintained. As a result at the end of synchronization we will
-have the plan depicted in Fig. \ref{fig:shop:exec1}. Which just
-propagated the information provided by the new observation in the
-exisiting plan (restricting in turn the start time of \texttt{Going}
-and by propagation of the constraints \texttt{Go} to be {\em exactly}
-$\tau'$). While focusing uniquely on the execution frontier ($\tau'$
-in Fig. \ref{fig:shop:exec1}), the \eu solver took benefit of the
+synchronization, the solver can identify the similarity between this
+new observation and the next token in the existing plan. This implies
+that instead of initiating an exhaustive search of all possible
+implications, the synchronization process starts its execution with
+the assumption that these new observations are consitent with the
+currently maintained plan. As a consequence, at the end of
+synchronization we will likely have the plan depicted in
+Fig. \ref{fig:shop:exec1}. Which just propagated the information
+provided by the new observation in the exisiting plan (restricting in
+turn the start time of \texttt{Going} and by propagation of the
+constraints \texttt{Go} to be {\em exactly} $\tau'$). While focusing
+uniquely on the execution frontier ($\tau'$ in
+Fig. \ref{fig:shop:exec1}), the \eu solver took benefit of the
 exisiting plan to evaluate a solution that asumes that the observation
 is consistent with what was decided inside this reactor.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
+  \caption{Shopping example : Result of synchronization after the {\em
+      Going} observation was received by the reactor with the plan
+    showed in Fig. \ref{fig:shop:exec0}.}
+  \label{fig:shop:exec1}
+\end{figure}
 
@@
 in few steps (often without any backtrack in the search).
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
-  \caption{Shopping example : Result of synchronization after the {\em
-      Going} observation was received by the reactor with the plan
-    showed in Fig. \ref{fig:shop:exec0}.}
-  \label{fig:shop:exec1}
-\end{figure}
 

### 2012-02-15 trunk/arch.tex
- Further refinements on synchronization presentation by statin many of the assumption we do in our reactor to help for quick decision - Reorganized/rewrote the end of synchronization in order to smoothly transition to deliberation/planning section

@@
   of high level waypoints into low-level commands) and temporally as
+
   each reactor declares its planning look-ahead $\pi$ and expected
   latency $\lambda$
-  % \comment{$\pi \&  \lambda$ both defined for the first time} 
-  before
-  being able to produce plan to the agent depending on where it is
-  located in a reactor dependency graph such as shown in
+  % \comment{$\pi \& \lambda$ both defined here for the first time}
+  before being able to produce plan to the agent depending on where it
+  is located in a reactor dependency graph such as shown in
   Fig. \ref{fig:agent}.
@@
 context where it does not have any compelling need to deliberate. We
-assume that we have a reactor that has no future goal.Even
-in this condition, \rx requires each reactor to produce at every tick
-its {\em internal} state. For those reactors not owning {\em internal}
+assume that we have a reactor that has no future goal. Even in this
+condition, \rx requires each reactor to produce at every tick its {\em
+  internal} state. For those reactors not owning {\em internal}
 timelines, \rx makes sure that {\em external} observations are taken
@@
 framework defined as follow:
+
 \begin{definition}
   \label{def:csf}
-  A {\em current state flaw} indicate that one of the reactor {\em
-    internal} timelines does not provide a fully defined state value
-  for the execution frontier $\tau$. Such flaw can be due to one 
-  of the 2 following reasons:
+  A {\em current state flaw} indicates that a reactor {\em internal}
+  timeline does not provide a fully defined state value for the
+  execution frontier $\tau$. Such a flaw can be result of either of
+  the following reasons:
+
   \begin{itemize}
@@
     instant $\tau$.
-  \item An open decision between terminating the last past token for
-    this timeline or pushing its end in the future.
+  \item An open decision between terminating the last token for this
+    timeline or pushing its end in the future.
   \end{itemize}
+
 \end{definition}
 
-The core idea bewteen this flaw is to ensure that at any single tick
-the {\em synchronization} solver will identifies all the {\em
-  internal} timelines for which the current state value is yet to be
-identified. We handle their resolution based on following
+The core idea between this flaw is to ensure that at any single tick
+the {\em synchronization} solver will identify all the {\em internal}
+timelines for which the current state value is yet to be
+identified. We handle their resolution based on the following
 proposition.
+
 \begin{proposition}
@@
   A {\em current state flaw} can be resolved using one of the
-  following choices which are eveluated accroding to this order:
+  following choices which are evaluated in the following order:
+
   \begin{enumerate}
@@
 
-\item Start the next active token in the timeline 
-  by restricting its start time to the single value $\tau$
+  \item Start the next active token in the timeline by restricting its
+    start time to the single value $\tau$
 
@@
 \end{proposition}
-Intuitively the first choice enforces an intertail value asumption
+Intuitively the first choice enforces an intertial value assumption
 where the reactor will tend to maintain its current as long as the
-model and external observations allow it. The second choice 
-favorise the adavnce of the reactor current plan if any; this direct
-the {\em synchronization} decision toward assuming that its current
-plan is properly executed. Finally if none of these two choices did
-provide a valid solution -- meaning that both lead to an inconsittant
-plan -- the reactor tries all the possible state values remaining for
-this timeline until it finds one that allows it to identify its
-state as required by \rx architecture.
+model and external observations allow it. The second choice favorise
+the advance of the reactor current plan if any; this direct the {\em
+  synchronization} decision toward assuming that its current plan is
+properly executed. Finally if none of these two choices did provide a
+valid solution -- meaning that both lead to an inconsistant plan --
+the reactor tries all the possible state values remaining for this
+timeline until it finds one that allows it to identify its state as
+required by \rx architecture.
 
@@
 \begin{enumerate}
-
 \item Integrate the external state as provided by the owner of each
   external timeline into the plan database.
-
 \item Propagate this information in the plan database following the
   model $\mathcal{M}$ of this reactor.
-
 \item Resolve the current state value of each internal timelines based
@@
 
-The steps 2 \& 3 are handled by the \eu solver from algorithm
+%%%% Extra paragraph:  targeted to a more AI centric audience
+% The step 1 translates new {\em external} observations received from
+% \rx into europa tokens with a start time restricted to $\tau$ or, for
+% the timelines that did not receive a new observation,  restrict the
+% previous token end time to the interval $[\tau+1, \infty)$. All of
+% these tokens are produced as facts and consequently cannot be 
+% excluded from the plan. They are also maintained by the reactors as
+% observation tokens which will allow to avoid to mistakenly remove 
+% them when one of the two reactor solvers attempt to relax former
+% decisions made.\comment{This relaxation is developed on the last 
+%   subsection further in depth}
+%%%%
+
+Steps $2$ \& $3$ are handled by the \eu solver from algorithm
 \ref{alg:europa:solve}. We only enforce that the {\em current state
@@
 reactor and reduce dramatically the search space for the {\em
-  synchronization} solver. In order to do so we need to make the
-assumption that the current state value of an
-{\em internal} timeline does not depend on the future or more accurately
-that any choices made during this synchronization will not lead to a
-future inconsistency. Such an assumption impacts the set of possible 
-domains the reactor can support while remaining complete. Take for 
-example this rule from our \texttt{Shopping} model :
+  synchronization} solver. In order to do so we need to assume 
+that the current state value of an {\em internal} timeline does not 
+depend on the future or more accurately that any choices made during 
+this {\em synchronization} will not lead to a future inconsistency. 
+Such an assumption impacts the set of possible domains the reactor 
+can support while remaining complete. Take for  example this rule from 
+our \texttt{Shopping} model :
 
@@
 objective.  % The problem can even be relaxed further if needed but
-% this relaxation gives already more felxibility for the reactor to
+% this relaxation gives already more flexibility for the reactor to
 % recover from local failures.
@@
 
-At the engine level careful choices were made in both the focus of the
-search -- limited only to tokens that can potentially overlap the
-current tick -- and heuristics where the internal state flaw is
-considered last during the search process. The later limits the amount
-of backtracking for a solution -- or at least how deep one needs to
-backtrack in a decision tree. Similarly the sequence in which we
-evaluate the possible resolution for such a flaw -- presented
-previously -- was selected according to some simple assumptions that % takes
-% benefit of a possibly assume the existing plan by assuming that 
-if the outcome of synchronization appears consistent with our current
-plan then it probably reflects that the plan has executed
-\comment{that the past is monotonic?}. Such an assumption helps direct
-the internal state evaluation and in nominal cases, proves to be a
-sound solution.
-
-One final aspect to take into account is that from a \rx point of
+One critical aspect to take into account is that from a \rx point of
 view, synchronization is the only critical task of a reactor. Such a
@@
 observation.
+
+Finally, at the engine level careful choices were made in 
+both the focus of the search -- limited only to tokens that 
+can potentially overlap the current tick -- and heuristics where 
+the {\em current state flaw} is considered last during the search
+process. The later limits the amount of backtracking for a solution 
+-- or at least how deep one needs to backtrack in a decision tree. 
+Similarly the sequence in which we evaluate the possible resolution 
+for such a flaw -- presented in Proposition \ref{prop:csf:resolve} -- 
+was selected according to  some simple assumptions that any external 
+observation that is compatible with an existing token in the current
+plan reflects the execution of the plan itself. In other words this
+means that our solver always prefer to use tokens already existing in
+the plan in order to identify its current internal state. By doing so
+we directs the synchronization search by exploiting the structure of
+the plan produced during previous deliberation steps. This implies
+that a deliberation steps have been done and did impact the plan
+structure and calls for a description on how our \eu reactor resolve
+such steps which we are now going to describe further.
 
@@
 
-The number of flaws between two steps \comment{between deliberation
-  and synch.?} can evolve for the following reasons that are exogenous
+The number of flaws between two steps \comment{between synch. and
+  execution?} can evolve for the following reasons that are exogenous
 to deliberation:
@@
 
-Such a scope allows filtering of tokens that are either necessarily
-ending before $\tau$ or necessarily starting after, focusing the
-planning problem on only those tokens that are within a ``reasonable''
+\comment{Insert a figure here [AAMAS?] showing $\pi \lambda$} Such a
+scope allows filtering of tokens that are either necessarily ending
+before $\tau$ or necessarily starting after, focusing the planning
+problem on only those tokens that are within a ``reasonable''
 future. This window is taken into account by the \rx agent which will
@@
 
-A core aspect of this reactor is that it takes advantage of the
-specific sequencing enforced by the \rx architecture along with the
-fact that only one $plan$ structure is maintained in the reactor and
-shared between synchronization and planning to allow these two
-processes to not only be interleaved but impact each other through the
-way they manipulate this plan.
+One key aspect of the \rx framework is intertwined plan execution. As
+seen above, it is synchronization which drives planning, so the
+co-mingling between these two processes is critical to maintaining
+state within a robot immersed in a dynamic environment. Also important
+is that \rx maintains 
+% A core aspect of this reactor is that it takes advantage of the
+% specific sequencing enforced by the \rx architecture along with the
+% fact that only 
+one $plan$ structure in the reactor and shared between synchronization
+and planning. This allows these two processes to not only be
+interleaved but how they can impact each other via plan manipulation.
 
 The most obvious case is how synchronization allow propagate of {\em
-  external} observations as a new tick occur in the plan. In or
-presentation of synchronization, we isolated the problem by stating
-that we will assume at this point that the reactorhas no future goal
-or more accurateely no plan to enforce at this stage. By doing so we
-were able to develop the way we augment yhe europpa solver to be able
-to propagete {\em external} observations in order to identify the
-reactor {\em internal} state for the same tick. By doing so we were
-able to focus on synchronization as a state evluation process that can
-be resolved within the \eu framework with few extensions. simliarly
+  external} observations as a new tick occurs. In our presentation of 
+{\em synchronization}, we isolated the problem by stating
+that we will assume at this point that the reactor has no future goal
+or more accurately no plan to enforce. By doing so we
+were able to develop the way we augment the \eu solver to be able
+to propagate {\em external} observations in order to identify the
+reactor {\em internal} state for the same tick. This allowed us 
+to focus on synchronization as a state evaluation process that can
+be resolved within the \eu framework with few extensions. Similarly
 we presented deliberation at this fairly high level not necessarily
 discussing how the disruptive nature of synchronization in this
-process could impact the way this planning process will evolve.
+process could impact the way this planning could evolve.
 
-We first anlyse the case during which synchronization occurs after the
-planning has vcompleted (ie the last seep of deliberation resulted
-into a plan that is considered as complete for the current
-look-ahead). As synchronization starts with the plan produced this
-will gave a frame for the search process. Therefore this deliberation
-will not only be a model-based estimation uniquely based on the {\em
-  external} observations but will also include the reactor intentions
-as described by the plan produced. Consider for example the Shopping
-Agent model described during the \eu presentation and that the
-location of the \texttt{agent} is an {\em external} timeline. For
-example as, the reactor add the goal to have \texttt{Own} milk and the
-current {\em external} states indicated that it was \texttt{At(home)}
-it produced the plan in Fig. \ref{fig:shop:exec0}. The {\em external}
-part of the plan -- namely the tokens \texttt{Going(home,
-  superMarket)} and \texttt{At(SuperMarket)} -- and integrated as
-goals for the reactor managing this specific state varaiable {\em
-  internally}.
+We first analyze the case during which synchronization occurs after
+planning has completed (\ie the last step of deliberation results in a
+plan that is considered complete for the current look-ahead). As
+synchronization starts with a plan % produced this will gave a frame for
+% the search process. Therefore 
+this deliberation will not only be a model-derived estimation uniquely
+based on {\em external} observations but will also include reactor
+intentions specified in the available partial plan. Consider for
+example the Shopping Agent model described where the \texttt{Agent} is
+an {\em external} timeline. As the reactor adds the goal to
+\texttt{Own} milk and the current {\em external} timeline indicates
+that the robot was \texttt{At(home)}, it produces a plan as shown in
+Fig. \ref{fig:shop:exec0}. The {\em external} part of the plan --
+namely the tokens \texttt{Going(home, superMarket)} and
+\texttt{At(SuperMarket)} -- are installed as goals for the reactor
+managing this state variable {\em internally}.
 

### 2012-02-15 trunk/arch.tex
updates

@@
   \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}
-  \caption{\small Shopping example: A plan produced before
-    synchronization.  Tokens in red indicate observations, tokens in
-    blue the goal that this plan attempts to solve for. Arrows
-    indicate temporal constraints.}
-  \label{fig:shop:exec0}
+  \subfloat[\small A plan produced before synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.45\columnwidth]{figs/shoping_exec_t0}}
+  \subfloat[\small Result of synchronization after the {\em Going}
+  observation was received by the reactor with the plan showed in
+    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=0.48\columnwidth]{figs/shoping_exec_t1}}
+    \caption{\small Shopping example: Tokens in red indicate
+      observations, tokens in blue the goal that this plan attempts to
+      solve for. Arrows indicate temporal constraints. }
+  \label{fig:shop:exec}
 \end{figure}
@@
 synchronization, the solver can identify the similarity between this
-new observation and the next token in the existing plan. This implies
-that instead of initiating an exhaustive search of all possible
-implications, the synchronization process starts its execution with
-the assumption that these new observations are consitent with the
-currently maintained plan. As a consequence, at the end of
-synchronization we will likely have the plan depicted in
-Fig. \ref{fig:shop:exec1}. Which just propagated the information
-provided by the new observation in the exisiting plan (restricting in
-turn the start time of \texttt{Going} and by propagation of the
-constraints \texttt{Go} to be {\em exactly} $\tau'$). While focusing
-uniquely on the execution frontier ($\tau'$ in
-Fig. \ref{fig:shop:exec1}), the \eu solver took benefit of the
-exisiting plan to evaluate a solution that asumes that the observation
-is consistent with what was decided inside this reactor.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
-  \caption{Shopping example : Result of synchronization after the {\em
-      Going} observation was received by the reactor with the plan
-    showed in Fig. \ref{fig:shop:exec0}.}
-  \label{fig:shop:exec1}
-\end{figure}
-
-This choice is directed not only by some natural heurisitc choices
-made in the solver but even more enforced in general by the heuristic
-order of the decision point reolution choices for our {\em internla}
-state flaw presented in section \ref{sec:arch:sync} {\em\color{red}
-  TODO: need to rework the presentation of thee choices so I can refer
-  to them more directly -- algorithm or pseudo theorem should do the
-  trick}. Should such a flaw be revealed during synchronization --
-meaning that the current state of an {\em internal} state varaibale is
-not fully grounded -- the sequencing of the choices will atempt first
-a conservative choice in term of maintaining previous state, then an
-optimisitic choice in term of the execution/advance in our current
-plan to finally attempt other choices that were not predicted by the
-model. The two first possible choices are strongly directed by the
-exisiting plan which help direct the search within this frame. This
-often allow to have a synchronization that is much more directed and, 
-in nominal cases such as the one we showed resolve the deliberation 
-in few steps (often without any backtrack in the search).
-
-The synchronization process is also in this process supporting fully
-the tracking of the plan execution as it specifies tokens parameters
-such as the time points and allow this to propagate through the plan
-in order to idenitfy if the plan remains valid -- as in our example --
-or that it cannot be executed as it is. The later is often identified
-by the synchronization failing to find a consistent solution. for
-example dshould the \texttt{Going} observation we receive be to the
+observation and the next token in the existing plan. Instead of
+initiating an exhaustive search of all possible implications, the
+synchronization process starts its execution with the assumption that
+these new observations are consistent with the currently maintained
+plan. As a consequence, at the end of synchronization the plan
+depicted in Fig. \ref{fig:shop:exec1} is likely. This is a result of
+propagation of information provided by the new observation in the
+existing plan while restricting the start time of \texttt{Going} and
+propagating temporal constraints, making \texttt{Go} to start {\em
+  exactly} at $\tau'$. In doing so, the \eu solver's focus on the
+execution frontier ($\tau'$ in Fig. \ref{fig:shop:exec1}) takes
+advantage of the existing plan to evaluate a solution that assumes
+that the observation is consistent with what was decided by the
+reactor.
+
+% \begin{figure}[!htb]
+%   \centering
+%   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
+%   \caption{Shopping example : Result of synchronization after the {\em
+%       Going} observation was received by the reactor with the plan
+%     showed in Fig. \ref{fig:shop:exec0}.}
+%   \label{fig:shop:exec1}
+% \end{figure}
+
+This choice is directed not only by heuristic choices made in the
+solver but also enforced by the heuristic order of the decision point
+resolution choices for the {\em internal} state flaw presented in
+Section \ref{sec:arch:sync} \fcomment{TODO: need to rework the
+  presentation of thee choices so I can refer to them more directly --
+  algorithm or pseudo theorem should do the trick}. Should such a flaw
+be revealed during synchronization -- such that the current state of
+an {\em internal} state variable is not fully grounded -- the
+sequencing of the choices will attempt first a conservative choice in
+maintaining previous state and only then try a more optimistic
+approach in terms of the \comment{I don't follow this last bit. Is
+  this token addition that you're talking about?} execution/advance in
+our current plan to finally attempt other choices that were not
+predicted by the model. The two first possible choices are strongly
+directed by the existing plan which helps direct the search within
+this frame. This often allow to have a synchronization that is much
+more directed and, in nominal cases such as the one we showed resolve
+the deliberation in few steps (often without any backtrack in the
+search) \comment{All this is confusing; probably reword}.
+
+The synchronization process also supports the tracking of plan
+execution. It does so by specifying token parameters such as time
+points that are propagated through the plan in order to identify if
+the plan remains valid -- as in our example -- or that it cannot be
+executed as it is. The later is often identified by the
+synchronization failing to find a consistent solution. for example
+dshould the \texttt{Going} observation we receive be to the
 \texttt{hardwareStore}, this would break the current plan of our

### 2012-02-15 trunk/arch.tex
updates

@@
 
-The synchronization process also supports the tracking of plan
-execution. It does so by specifying token parameters such as time
-points that are propagated through the plan in order to identify if
-the plan remains valid -- as in our example -- or that it cannot be
-executed as it is. The later is often identified by the
-synchronization failing to find a consistent solution. for example
-dshould the \texttt{Going} observation we receive be to the
-\texttt{hardwareStore}, this would break the current plan of our
-reactor as we cannot go simultaneously at 2 different places as
-illustrated in Fig. \ref{fig:shop:relax}-1. as synchronization is not
-possible in this context our strategy is to fully relax all decisions
-made by the previous plan steps by just keeping n the plan all the
-past observations in both {\em external} and {\em internal} state
-variables and the exsiting {\em internal} goals received by this
-reactor (\texttt{Own(Milk)} in our example) resulting on the plrtial
-plan sghowned in Fig. \ref{fig:shop:relax}-2. In this process the
-reactor inform the agent that all {\em external} goals it requested
-for the future are note valid anymore which often result in the ownere
-of htese timelines to not have to maintain this goal anynmore.
+Synchronization also supports the tracking of plan execution. It does
+so by specifying token parameters such as time points that are
+propagated through the plan in order to identify if the plan remains
+valid -- as in our example -- or that it cannot be executed as it
+is. The later is often identified by the synchronization failing to
+find a consistent solution. For example should the \texttt{Going}
+observation received be to the \texttt{hardwareStore}, this would
+break the current partial plan in the reactor as the robot cannot
+possible be going simultaneously to $2$ different locations; see
+Fig. \ref{fig:shop:relax1}. As synchronization is not possible in such
+a situation the strategy followed is to relax all decisions made by
+previous plan steps. This is done by keeping all past observations in
+both {\em external} and {\em internal} state variables as well as the
+existing {\em internal} goals (\texttt{Own(Milk)} in this example)
+received by this reactor. The resulting partial plan is shown in
+Fig. \ref{fig:shop:relax2}. In this process the reactor informs the
+agent that all {\em external} goals it requested for the future are no
+longer valid which results in turn, in timeline owners not having to
+maintain such goals.
 
@@
   \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
-  \caption{Illustration of a conflict during synchronization and the
-    resulting relaxed plan for recovery} 
-  \label{fig:shop:relax}
+  \subfloat[Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
+  \subfloat[Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \caption{\small Illustration of a conflict during synchronization
+    and the resulting relaxed plan for recovery.}
 \end{figure}
 
-As the plan is not anymore in the way the synchronization can resume
-and find a solution as we initial lly described. The newly pending
-goals will enforce at the next \texttt{step} to resumed deliberation
-on this new fully relaxed partial plan. By the simple use of
-synchronization the reactor was not only able to identify that its
-plan was not executable anymore but recover from it while offering a
-new incomplete partial plan to be resolved during the next
-deliberation \texttt{steps}. This on its own ogfer the core
-functionality one should except to close the loop between planning and
-execution in the sense that synchronization plays the role of what is
-typically named as the {\em Executive} in other architectures
-{\em\color{red} references to diverse archis, 3-tiered, laas, claraty,
-  B Wiliams, and many more}.
-
-In our case though the interaction between planning and execution does
-not stop at this level. A core point of the overal schedulling of both
-{\em synchronization} and {\em planning} steps resides on our
-architecture enforsincing that the synchroniuzation process can occur
-in between any step of the plan. This lead us to the fact that not
-only synchronization is imnfluenced by the outcome of planning but
-this influence goes also in the alternate direction. Specifically when
-synchronuiation occuers before we found a complete plan, it will
-inject as we described new facts (the {\em external} observations and
-resulting {\em internal} state) in the plan database. This
-informatiuon will be present when the next planning \texttt{step} will
-resume. Therefore they are not intergal part of the plan resolution
-(generating potentially new flaws for the planner or contributing to
-resolve previously exisiting flaws). The planning process is therefore
-perturbed internallyand fully informed on how the world evolves as it
-is planning.  This aspect trueluy contrast to other approaches we have
-identified in the literature which avoid to perturb the planner during
-it search as it is not compatible with the ``off-line planning''
-assumption.
+As the plan is no longer conflicted, synchronization can resume and
+find a solution described above. Newly pending goals would be enforced
+at the next \texttt{step} to resume deliberation on this relaxed
+partial plan. By using synchronization the reactor was not only able
+to identify that its plan was not consistent but able to recover from
+it by offering a new incomplete partial plan to be resolved in the
+next deliberation \texttt{steps}. % This on its own offer the core
+% functionality one should except to close the loop between planning and
+% execution in the sense that 
+Synchronization plays the role of what is typically termed as the {\em
+  Executive} in other architectures \cite{gat98, alami:1998p820,
+  mus98, williams03, Nesnas:2003do}.
+
+In \rx however the interaction between planning and execution does not
+stop at this level. The overall scheduling of both {\em
+  synchronization} and {\em planning} is such that the synchronization
+can occur in between any planning step. Not only is synchronization
+influenced by the outcome, but it also influences
+planning. Specifically when synchronization occurs before a complete
+plan is found, it will inject new facts ({\em external} observations
+and resulting {\em internal} state) in the plan database available for
+inference when the next planning \texttt{step} resumes. Therefore they
+\comment{what is the ``they''?} are not integral part of the plan
+resolution (generating potentially new flaws for the planner or
+contributing to resolve previously existing flaws). The planning
+process is therefore perturbed internally and fully informed by how
+the world is evolving even as it is planning.  This specific aspect
+contrasts with other approaches we have which avoid perturbing the
+planner during its search as it is not compatible with the ``off-line
+planning'' assumption.
 
 By our design choices -- both at the architecture level and the
-integration of the \eu framework for embedded deliberation -- we
-were able to implement an architecture that avoid this limitation
-while remaining in a formal frame that still allow for deliberation
-and model-based agent control. We believe that this tighter
-integration of planning and execution allow the system to ba more
-acute of its environment allowing better informaed decision and
-tighter reaction. This is possible while still integrating information
-such as a long latency for the reactors as long as one can ensure that
-synchronization of al the reactors and deliberation steps can be done
-in a time that is resonably small in comparison of the agent tick
-duration.
+integration of the \eu framework for embedded deliberation -- we are
+able to implement an architecture that avoids this critical off-line
+planning limitation while remaining within a formal framework that
+allows for deliberation and model-based agent control. We believe that
+this tight integration of planning and execution allows the system to
+be more acute vis-a-vis the environment allowing informed
+decision-making and a tighter reaction loop. This is possible while
+integrating information despite a long latency for the reactors, as
+long as one can ensure that synchronization of all reactors and
+deliberation steps can be done within a time that is resonably small
+in comparison to the agent tick duration.
 

### 2012-02-15 trunk/arch.tex
- Reworked paragraph in the last section in resp. to Kanna's comments - corrected one bibtex entry that refused to be embedded parsed

@@
 solver but also enforced by the heuristic order of the decision point
-resolution choices for the {\em internal} state flaw presented in
-Section \ref{sec:arch:sync} \fcomment{TODO: need to rework the
-  presentation of thee choices so I can refer to them more directly --
-  algorithm or pseudo theorem should do the trick}. Should such a flaw
-be revealed during synchronization -- such that the current state of
-an {\em internal} state variable is not fully grounded -- the
-sequencing of the choices will attempt first a conservative choice in
-maintaining previous state and only then try a more optimistic
-approach in terms of the \comment{I don't follow this last bit. Is
-  this token addition that you're talking about?} execution/advance in
-our current plan to finally attempt other choices that were not
-predicted by the model. The two first possible choices are strongly
-directed by the existing plan which helps direct the search within
-this frame. This often allow to have a synchronization that is much
-more directed and, in nominal cases such as the one we showed resolve
-the deliberation in few steps (often without any backtrack in the
-search) \comment{All this is confusing; probably reword}.
+resolution choices for the {\em current  state flaw} presented in
+Proposition \ref{prop:csf:resolve}. Indeed, as we stated then the two
+first choices tend to privilege a conservative view in respect to the
+existing plan. It attempts first to extend the previous observation we
+produced for this {\em internal} timeline in order to avoid to commit
+to early in the plan advance. If the formaer failed it will then
+attempt to start the next exisiting token in the plan which alow the
+reactor to advance its plan execution. Both of these two initial
+choices atetempt to resolve the plan with the least amount of
+perturbation of the plan. They just constrain further the start and
+end time of the exisiting tokens which were provided by the previous 
+deliberations steps. This means that in a nominal situation where the
+plan this reactor produced is properly executed, the synchronization
+remains limited to restrict time points in the plan and progate their
+impact. Most of the heavy dutty work of creating new tokens in the
+plan and schedule them -- both being much more difficult problems 
+to resolve -- have been managed during deliberation steps. In that
+sense we see that most of the time synchronization is limited to very
+few node in \eu search space with often no backtrack required. The
+synchronization may become slightly nore costfull when we need to
+identify a new token to be injected by the plan (the 3rd choice in
+Proposition \ref{prop:csf:resolve}) as it needs to explore on tis own
+all the possible token types this timeline accept as value. This is
+the reason this choice is evaluated last in our flaw resolution in
+order to avoid such complex action during synchronization unless 
+necessary. In that sense,  ansd due to carefull choice on our
+synchrponization heurisitic we see that the plan produced during \rx
+deliberation steps helps sthe synchronization process to be more
+efficient.
+
+% Should such a flaw be revealed 
+% during synchronization -- such that the current state of
+% an {\em internal} state variable is not fully grounded -- the
+% sequencing of the choices will attempt first a conservative choice in
+% maintaining previous state and only then try a more optimistic
+% approach in terms of the \comment{I don't follow this last bit. Is
+%   this token addition that you're talking about?} execution/advance in
+% our current plan to finally attempt other choices that were not
+% predicted by the model. The two first possible choices are strongly
+% directed by the existing plan which helps direct the search within
+% this frame. This often allow to have a synchronization that is much
+% more directed and, in nominal cases such as the one we showed resolve
+% the deliberation in few steps (often without any backtrack in the
+% search) \comment{All this is confusing; probably reword}.
 
@@
 and resulting {\em internal} state) in the plan database available for
-inference when the next planning \texttt{step} resumes. Therefore they
-\comment{what is the ``they''?} are not integral part of the plan
-resolution (generating potentially new flaws for the planner or
-contributing to resolve previously existing flaws). The planning
-process is therefore perturbed internally and fully informed by how
-the world is evolving even as it is planning.  This specific aspect
-contrasts with other approaches we have which avoid perturbing the
-planner during its search as it is not compatible with the ``off-line
-planning'' assumption.
+inference when the next planning \texttt{step} resumes. On this next
+step the \eu solver used for planning will have these new facts
+included in the partial plan which can introduce new flaws in the plan
+while potetnially having reolved previously exisiting flaw. The
+deliberation process is therefore perturbed externally by the {\em
+  synchronization} and consequently fully informed of the world state
+evolution as it is planning.  This specific aspect contrasts with
+other more classical approaches we have seen, which avoid 
+perturbing the planner during its search as it is not compatible 
+with the ``off-line planning'' assumption.
 

### 2012-02-15 arch.tex
Added an illustration based on Fig. 14 to explain how the existing plan supports synchronization search

@@
 
-  each reactor declares its planning look-ahead and expected latency
-  \comment{$\pi \& \lambda$ both defined here for the first time}
+  each reactor declares its planning look-ahead $\pi$ and expected
+  latency $\lambda$
+  % \comment{$\pi \& \lambda$ both defined here for the first time}
   before being able to produce plan to the agent depending on where it
@@
 We introduce this requirement as a new type of flaw for the \eu
-framework defined as follows:
+framework defined as follow:
 
@@
 \end{proposition}
-
-Intuitively the first choice enforces an intertail value assumption
+Intuitively the first choice enforces an intertial value assumption
 where the reactor will tend to maintain its current as long as the
 model and external observations allow it. The second choice favorise
-the adavnce of the reactor current plan if any; this direct the {\em
+the advance of the reactor current plan if any; this direct the {\em
   synchronization} decision toward assuming that its current plan is
 properly executed. Finally if none of these two choices did provide a
-valid solution -- meaning that both lead to an inconsittant plan --
+valid solution -- meaning that both lead to an inconsistant plan --
 the reactor tries all the possible state values remaining for this
@@
 \end{enumerate}
+
+%%%% Extra paragraph:  targeted to a more AI centric audience
+% The step 1 translates new {\em external} observations received from
+% \rx into europa tokens with a start time restricted to $\tau$ or, for
+% the timelines that did not receive a new observation,  restrict the
+% previous token end time to the interval $[\tau+1, \infty)$. All of
+% these tokens are produced as facts and consequently cannot be 
+% excluded from the plan. They are also maintained by the reactors as
+% observation tokens which will allow to avoid to mistakenly remove 
+% them when one of the two reactor solvers attempt to relax former
+% decisions made.\comment{This relaxation is developed on the last 
+%   subsection further in depth}
+%%%%
 
@@
 reactor and reduce dramatically the search space for the {\em
-  synchronization} solver. In order to do so we need to make the
-assumption that the current state value of an {\em internal} timeline
-does not depend on the future or more accurately that any choices made
-during this synchronization will not lead to a future
-inconsistency. Such an assumption impacts the set of possible domains
-the reactor can support while remaining complete. Take for example
-this rule from our \texttt{Shopping} model:
+  synchronization} solver. In order to do so we need to assume 
+that the current state value of an {\em internal} timeline does not 
+depend on the future or more accurately that any choices made during 
+this {\em synchronization} will not lead to a future inconsistency. 
+Such an assumption impacts the set of possible domains the reactor 
+can support while remaining complete. Take for  example this rule from 
+our \texttt{Shopping} model :
 
@@
 
-At the engine level careful choices were made in both the focus of the
-search -- limited only to tokens that can potentially overlap the
-current tick -- and heuristics where the internal state flaw is
-considered last during the search process. The later limits the amount
-of backtracking for a solution -- or at least how deep one needs to
-backtrack in a decision tree. Similarly the sequence in which we
-evaluate the possible resolution for such a flaw -- presented
-previously -- was selected according to some simple assumptions that % takes
-% benefit of a possibly assume the existing plan by assuming that 
-if the outcome of synchronization appears consistent with our current
-plan then it probably reflects that the plan has executed
-\comment{that the past is monotonic?}. Such an assumption helps direct
-the internal state evaluation and in nominal cases, proves to be a
-sound solution.
-
-One final aspect to take into account is that from a \rx point of
-view, \emph{synchronization is the only critical task of a
-  reactor}. Such a process is used in order for the agent to identify
-its current state and maintain it consistently throughout the agent's
-lifetime. For a reactor not being able to find a consistent solution
-during synchronization -- implying that after evaluating all possible
+One critical aspect to take into account is that from a \rx point of
+view, synchronization is the only critical task of a reactor. Such a
+process is used in order for the agent to identify its current state
+and maintain it consistently throughout the agent's lifetime. For a
+reactor not being able to find a consistent solution during
+synchronization -- implying that after evaluating all possible
 solutions provided by a model it was unable to find a consistent plan
@@
 
-The most obvious case is how synchronization allows propagate of {\em
-  external} observations when a new tick occurs. In our presentation
-of synchronization in Section \ref{sec:arch:synch}, we isolated the
-problem by assuming that the reactor has no future goal or more
-accurateely no plan to enforce.  By doing so we were able to develop
-the way we augment the \eu solver to be able to propagete {\em
-  external} observations in order to identify the reactor {\em
-  internal} state for the same tick. By doing so we were able to focus
-on synchronization as a state evaluation process that can be resolved
-within the \eu framework with few extensions. Similarly we presented
-deliberation as an abstraction while not detailing how the disruptive
-nature of synchronization could impact the way planning could evolve.
+The most obvious case is how synchronization allow propagate of {\em
+  external} observations as a new tick occurs. In our presentation of 
+{\em synchronization}, we isolated the problem by stating
+that we will assume at this point that the reactor has no future goal
+or more accurately no plan to enforce. By doing so we
+were able to develop the way we augment the \eu solver to be able
+to propagate {\em external} observations in order to identify the
+reactor {\em internal} state for the same tick. This allowed us 
+to focus on synchronization as a state evaluation process that can
+be resolved within the \eu framework with few extensions. Similarly
+we presented deliberation at this fairly high level not necessarily
+discussing how the disruptive nature of synchronization in this
+process could impact the way this planning could evolve.
 
@@
 synchronization, the solver can identify the similarity between this
-new observation and the next token in the existing plan. This implies
-that instead of initiating an exhaustive search of all possible
-implications, the synchronization process starts its execution with
-the assumption that these new observations are consitent with the
-currently maintained plan. As a consequence, at the end of
-synchronization we will likely have the plan depicted in
-Fig. \ref{fig:shop:exec1}. Which just propagated the information
-provided by the new observation in the exisiting plan (restricting in
-turn the start time of \texttt{Going} and by propagation of the
-constraints \texttt{Go} to be {\em exactly} $\tau'$). While focusing
-uniquely on the execution frontier ($\tau'$ in
-Fig. \ref{fig:shop:exec1}), the \eu solver took benefit of the
-exisiting plan to evaluate a solution that asumes that the observation
-is consistent with what was decided inside this reactor.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
-  \caption{Shopping example : Result of synchronization after the {\em
-      Going} observation was received by the reactor with the plan
-    showed in Fig. \ref{fig:shop:exec0}.}
-  \label{fig:shop:exec1}
-\end{figure}
-
-This choice is directed not only by some natural heurisitc choices
-made in the solver but even more enforced in general by the heuristic
-order of the decision point reolution choices for our {\em internla}
-state flaw presented in section \ref{sec:arch:sync} {\em\color{red}
-  TODO: need to rework the presentation of thee choices so I can refer
-  to them more directly -- algorithm or pseudo theorem should do the
-  trick}. Should such a flaw be revealed during synchronization --
-meaning that the current state of an {\em internal} state varaibale is
-not fully grounded -- the sequencing of the choices will atempt first
-a conservative choice in term of maintaining previous state, then an
-optimisitic choice in term of the execution/advance in our current
-plan to finally attempt other choices that were not predicted by the
-model. The two first possible choices are strongly directed by the
-exisiting plan which help direct the search within this frame. This
-often allow to have a synchronization that is much more directed and, 
-in nominal cases such as the one we showed resolve the deliberation 
-in few steps (often without any backtrack in the search).
-
-The synchronization process is also in this process supporting fully
-the tracking of the plan execution as it specifies tokens parameters
-such as the time points and allow this to propagate through the plan
-in order to idenitfy if the plan remains valid -- as in our example --
-or that it cannot be executed as it is. The later is often identified
-by the synchronization failing to find a consistent solution. for
-example dshould the \texttt{Going} observation we receive be to the
-\texttt{hardwareStore}, this would break the current plan of our
-reactor as we cannot go simultaneously at 2 different places as
-illustrated in Fig. \ref{fig:shop:relax}-1. as synchronization is not
-possible in this context our strategy is to fully relax all decisions
-made by the previous plan steps by just keeping n the plan all the
-past observations in both {\em external} and {\em internal} state
-variables and the exsiting {\em internal} goals received by this
-reactor (\texttt{Own(Milk)} in our example) resulting on the plrtial
-plan sghowned in Fig. \ref{fig:shop:relax}-2. In this process the
-reactor inform the agent that all {\em external} goals it requested
-for the future are note valid anymore which often result in the ownere
-of htese timelines to not have to maintain this goal anynmore.
+observation and the next token in the existing plan. Instead of
+initiating an exhaustive search of all possible implications, the
+synchronization process starts its execution with the assumption that
+these new observations are consistent with the currently maintained
+plan. As a consequence, at the end of synchronization the plan
+depicted in Fig. \ref{fig:shop:exec1} is likely. This is a result of
+propagation of information provided by the new observation in the
+existing plan while restricting the start time of \texttt{Going} and
+propagating temporal constraints, making \texttt{Go} to start {\em
+  exactly} at $\tau'$. In doing so, the \eu solver's focus on the
+execution frontier ($\tau'$ in Fig. \ref{fig:shop:exec1}) takes
+advantage of the existing plan to evaluate a solution that assumes
+that the observation is consistent with what was decided by the
+reactor.
+
+% \begin{figure}[!htb]
+%   \centering
+%   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
+%   \caption{Shopping example : Result of synchronization after the {\em
+%       Going} observation was received by the reactor with the plan
+%     showed in Fig. \ref{fig:shop:exec0}.}
+%   \label{fig:shop:exec1}
+% \end{figure}
+
+This choice is directed not only by heuristic choices made in the
+solver but also enforced by the heuristic order of the decision point
+resolution choices for the {\em current  state flaw} presented in
+Proposition \ref{prop:csf:resolve}. Indeed, as we stated then the two
+first choices tend to privilege a conservative view in respect to the
+existing plan. It attempts first to extend the previous observation we
+produced for this {\em internal} timeline in order to avoid to commit
+to early in the plan advance. If the formaer failed it will then
+attempt to start the next exisiting token in the plan which alow the
+reactor to advance its plan execution. Both of these two initial
+choices atetempt to resolve the plan with the least amount of
+perturbation of the plan. They just constrain further the start and
+end time of the exisiting tokens which were provided by the previous 
+deliberations steps. This means that in a nominal situation where the
+plan this reactor produced is properly executed, the synchronization
+remains limited to restrict time points in the plan and progate their
+impact. Most of the heavy dutty work of creating new tokens in the
+plan and schedule them -- both being much more difficult problems 
+to resolve -- have been managed during deliberation steps. In that
+sense we see that most of the time synchronization is limited to very
+few node in \eu search space with often no backtrack required. 
+Fig. \ref{fig:shop:exec} shows how the existing plan can support the
+synchronization process. In that case, the observation \texttt{Going}
+occuring at $\tau'$ was fully compatible with the one already present
+in our plan, by updating the plan and progating the equality
+constraints, the resolution of the {\em current state flaw} for the
+{\em agent} timeline became obvious as the token \texttt{Go} from 
+the plan presented itself as a natural solution.
+
+The synchronization may become slightly nore costfull when we need to
+identify a new token to be injected by the plan (the 3rd choice in
+Proposition \ref{prop:csf:resolve}) as it needs to explore on tis own
+all the possible token types this timeline accept as value. This is
+the reason this choice is evaluated last in our flaw resolution in
+order to avoid such complex action during synchronization unless 
+necessary. In that sense,  and thanks to the carefull choice on our
+synchronization heurisitic we see that the plan produced during \rx
+deliberation steps helps sthe synchronization process to be more
+efficient.
+
+% Should such a flaw be revealed 
+% during synchronization -- such that the current state of
+% an {\em internal} state variable is not fully grounded -- the
+% sequencing of the choices will attempt first a conservative choice in
+% maintaining previous state and only then try a more optimistic
+% approach in terms of the \comment{I don't follow this last bit. Is
+%   this token addition that you're talking about?} execution/advance in
+% our current plan to finally attempt other choices that were not
+% predicted by the model. The two first possible choices are strongly
+% directed by the existing plan which helps direct the search within
+% this frame. This often allow to have a synchronization that is much
+% more directed and, in nominal cases such as the one we showed resolve
+% the deliberation in few steps (often without any backtrack in the
+% search) \comment{All this is confusing; probably reword}.
+
+Synchronization also supports the tracking of plan execution. It does
+so by specifying token parameters such as time points that are
+propagated through the plan in order to identify if the plan remains
+valid -- as in our example -- or that it cannot be executed as it
+is. The later is often identified by the synchronization failing to
+find a consistent solution. For example should the \texttt{Going}
+observation received be to the \texttt{hardwareStore}, this would
+break the current partial plan in the reactor as the robot cannot
+possible be going simultaneously to $2$ different locations; see
+Fig. \ref{fig:shop:relax1}. As synchronization is not possible in such
+a situation the strategy followed is to relax all decisions made by
+previous plan steps. This is done by keeping all past observations in
+both {\em external} and {\em internal} state variables as well as the
+existing {\em internal} goals (\texttt{Own(Milk)} in this example)
+received by this reactor. The resulting partial plan is shown in
+Fig. \ref{fig:shop:relax2}. In this process the reactor informs the
+agent that all {\em external} goals it requested for the future are no
+longer valid which results in turn, in timeline owners not having to
+maintain such goals.
 
@@
   \centering
-  \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax}  
-  \caption{Illustration of a conflict during synchronization and the
-    resulting relaxed plan for recovery} 
-  \label{fig:shop:relax}
+  \subfloat[Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
+  \subfloat[Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \caption{\small Illustration of a conflict during synchronization
+    and the resulting relaxed plan for recovery.}
 \end{figure}
 
-As the plan is not anymore in the way the synchronization can resume
-and find a solution as we initial lly described. The newly pending
-goals will enforce at the next \texttt{step} to resumed deliberation
-on this new fully relaxed partial plan. By the simple use of
-synchronization the reactor was not only able to identify that its
-plan was not executable anymore but recover from it while offering a
-new incomplete partial plan to be resolved during the next
-deliberation \texttt{steps}. This on its own ogfer the core
-functionality one should except to close the loop between planning and
-execution in the sense that synchronization plays the role of what is
-typically named as the {\em Executive} in other architectures
-{\em\color{red} references to diverse archis, 3-tiered, laas, claraty,
-  B Wiliams, and many more}.
-
-In our case though the interaction between planning and execution does
-not stop at this level. A core point of the overal schedulling of both
-{\em synchronization} and {\em planning} steps resides on our
-architecture enforsincing that the synchroniuzation process can occur
-in between any step of the plan. This lead us to the fact that not
-only synchronization is imnfluenced by the outcome of planning but
-this influence goes also in the alternate direction. Specifically when
-synchronuiation occuers before we found a complete plan, it will
-inject as we described new facts (the {\em external} observations and
-resulting {\em internal} state) in the plan database. This
-informatiuon will be present when the next planning \texttt{step} will
-resume. Therefore they are not intergal part of the plan resolution
-(generating potentially new flaws for the planner or contributing to
-resolve previously exisiting flaws). The planning process is therefore
-perturbed internallyand fully informed on how the world evolves as it
-is planning.  This aspect trueluy contrast to other approaches we have
-identified in the literature which avoid to perturb the planner during
-it search as it is not compatible with the ``off-line planning''
-assumption.
+As the plan is no longer conflicted, synchronization can resume and
+find a solution described above. Newly pending goals would be enforced
+at the next \texttt{step} to resume deliberation on this relaxed
+partial plan. By using synchronization the reactor was not only able
+to identify that its plan was not consistent but able to recover from
+it by offering a new incomplete partial plan to be resolved in the
+next deliberation \texttt{steps}. % This on its own offer the core
+% functionality one should except to close the loop between planning and
+% execution in the sense that 
+Synchronization plays the role of what is typically termed as the {\em
+  Executive} in other architectures \cite{gat98, alami:1998p820,
+  mus98, williams03, Nesnas:2003do}.
+
+In \rx however the interaction between planning and execution does not
+stop at this level. The overall scheduling of both {\em
+  synchronization} and {\em planning} is such that the synchronization
+can occur in between any planning step. Not only is synchronization
+influenced by the outcome, but it also influences
+planning. Specifically when synchronization occurs before a complete
+plan is found, it will inject new facts ({\em external} observations
+and resulting {\em internal} state) in the plan database available for
+inference when the next planning \texttt{step} resumes. On this next
+step the \eu solver used for planning will have these new facts
+included in the partial plan which can introduce new flaws in the plan
+while potetnially having reolved previously exisiting flaw. The
+deliberation process is therefore perturbed externally by the {\em
+  synchronization} and consequently fully informed of the world state
+evolution as it is planning.  This specific aspect contrasts with
+other more classical approaches we have seen, which avoid 
+perturbing the planner during its search as it is not compatible 
+with the ``off-line planning'' assumption.
 
 By our design choices -- both at the architecture level and the
-integration of the \eu framework for embedded deliberation -- we
-were able to implement an architecture that avoid this limitation
-while remaining in a formal frame that still allow for deliberation
-and model-based agent control. We believe that this tighter
-integration of planning and execution allow the system to ba more
-acute of its environment allowing better informaed decision and
-tighter reaction. This is possible while still integrating information
-such as a long latency for the reactors as long as one can ensure that
-synchronization of al the reactors and deliberation steps can be done
-in a time that is resonably small in comparison of the agent tick
-duration.
+integration of the \eu framework for embedded deliberation -- we are
+able to implement an architecture that avoids this critical off-line
+planning limitation while remaining within a formal framework that
+allows for deliberation and model-based agent control. We believe that
+this tight integration of planning and execution allows the system to
+be more acute vis-a-vis the environment allowing informed
+decision-making and a tighter reaction loop. This is possible while
+integrating information despite a long latency for the reactors, as
+long as one can ensure that synchronization of all reactors and
+deliberation steps can be done within a time that is resonably small
+in comparison to the agent tick duration.
 

### 2012-02-15 trunk/europa-basics.tex
rearranging text from concepts into europa-basics

@@
+\section{Foundational Concepts}
+\label{sec:concepts}
+
+\rx is an adaptive, artificial intelligence based controller and
+provides general framework for building reasoning systems for
+real-world autonomous vehicles. At MBARI \rx is used for AUV control;
+another instantiation of the system is being used for control of a
+terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
+development of \rx has been targeted at surveying a number of
+oceanographic features which are dynamic and unpredictable
+spatio-temporally. Typically this requires our AUVs to balance the
+goals of coverage spatially while opportunistically following features
+of scientific interest and to do so while being operationally aware of
+its own limitations in terms of resources (typically the battery state
+of charge) and overall proximity to other observational assets for
+obtaining scientific ground-truth. We are therefore interested in
+representational frameworks that allow robots to pursue long-term
+objectives while choosing short-term gain and can gracefully deal with
+exogenous or endogenous change.
+
+To enable this responsiveness to external observations, the agent has
+to be able to synthesize plans insitu and to re-plan. We use a
+temporal constraint-based planner \eu with a demonstrated legacy of
+having flown on NASA space missions \cite{jonsson00,bresina05,
+  barreiro09}. Our autonomy architecture brings three key innovations
+for AUV adaptation: the use of flexible plan representations,
+compositional control with the use of partitioned networks and
+off-line learning to inform insitu environmental state estimation. In
+this section we detail some of the key concepts relevant to flexible
+plan representation. Section \ref{sec:arch} has details the
+\rx partitioned architecture and estimation driven vehicle
+adaptation.
+
+\subsection{The basics}
+\label{sec:basics}
+
+\eu uses a \emph{domain model} written in a declarative language
+(NDDL: New Domain Description Language), together with initial
+conditions and goals also specified in NDDL, to construct a set of
+temporal relations that must be true at the start time. These models
+include assertions about the physics of the vehicle, i.e how it
+responds to external stimulus and internally driven goals. By
+propagating these relations forward using Simple Temporal Networks
+\cite{dechter91} and applying goal constraints, \eu can select a set
+of conditions that should be true in the future, where some of these
+conditions will correspond to actions the agent must take. The planner
+can backtrack and try another path during search if a goal cannot be
+reached while being capable of discarding unachievable goals.
+
+Traditionally robot execution has relied on dispatching commands at
+precise times. Such linear sequences of precisely timed commands give
+no ability to adjust execution on the basis of sensory
+information. Although some commands can issue tests on sensor
+readings, these tests have the objective of verifying whether expected
+execution conditions are occuring. If not, the state of the system is
+declared off-nominal and execution of the sequence of commands is
+interrupted. More recently, executives have been proposed and
+implemented that significantly broaden the way robots can be commanded
+\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
+interpreted a \textit{temporally flexible plan} which represents each
+start time as a variable and contains an explicit network of bounded
+delay constraints between such variables.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\caption{\small Tokens with flexible temporal intervals and parametric
+  constraints between tokens. This example shows the triggering of a
+  water sampler based on a feature threshold while the vehicle
+  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
+  start \& end times.}
+\label{fig:flex-timelines}
+\vskip+0.1cm
+\end{figure}
+
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time the most appropriate one for the actual execution
+conditions. The fact that constraints are explicitly represented
+ensures that through constraint propagation the executive will respect
+global limits expressed in the plan (e.g., don't start a task until a
+certain condition has been satisfied but still satisfying some global
+deadline). Such flexibility is critical when dealing with dynamic
+ocean conditions where precise timing of a robotic action might be
+indeterminate. Further, the advantage of flexibility can be contrasted
+with the consequences of the intrinsic inflexibiliy of traditional
+command sequences. Because they are inflexible, sequences must
+necessarily be designed considering worst case scenarios.
+
+\eu uses a \emph{state variable} representation to describe the
+evolution of state over time. The instantiated history of such state
+variable evolution over a temporal horizon we call \emph{timelines}
+and which represent a single thread in the execution of a concurrent
+system. At any given time each thread can execute a single procedure.
+Thus each timeline consists of a sequence of procedures which
+encapsulate and describe state evolution; we call these instantiated
+atomic entities \emph{tokens}.  A token therefore describes a
+procedure invocation, the state variables on which it can occur, the
+parameter values of the procedure, and the time values defining the
+interval. We allow encapsulation of uncertainty within these tokens
+with a range of start and end times and parameters, all of which are
+encoded as variables. A constraint solver in turn manipulates these
+variables defined in a \eu domain model. For example, consider
+a scientific need to take a water samples 100 meters from a hotspot
+while an AUV is performing a Yo-Yo in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
+% \begin{figure}
+% \centering
+% \includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+% \caption{\small Temporal relations defined within the planner are
+%   based on \texttt{Allen Algebra} relations shown above.}
+% \label{fig:allen-algebra}
+% \vskip-0.3cm
+% \end{figure}
+
 \section{Fundamentals of  Automated Planning}
@@
 Constraint Satisfaction Programming (also known as Constraint
-Programming (CP)) is a discipline that provides a generic framework
+Programming (\textsf{CP})) is a discipline that provides a generic framework
 for representing, solving and making logical inference on
@@
 objective is to find one or more value assignments to $V$ where all
-constraints are satisfied. To solve a problem, CP techniques use
+constraints are satisfied. To solve a problem, \textsf{CP} techniques use
 logical inference to perform Constraint Propagation, which consists of
@@
 over the naive approach by ignoring constraints whose variables were
-not affected since the last iteration.
+not affected since the last iteration. \comment{A figure of a
+  constraint network preferably of the shopping example would be
+  useful here.}
 
 Arc Consistency and other, more stringent consistency definitions can
-be used to efficiently prove that a CP problem is
-unsatisfiable. However, they generally cannot prove that a CP problem
-is satisfiable and provide specific variable assignments that
+be used to efficiently prove that a \textsf{CP} problem is
+unsatisfiable. However, they generally cannot prove that a \textsf{CP}
+problem is satisfiable and provide specific variable assignments that
 constitute solutions. For that, search algorithms like global
@@
 use consistency as an efficient way to prune the search space
-\cite{cp06}. In theory, solving a CP problem is NP-Hard
+\cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
 \cite{ghallab04}, but often very efficient in practice using a number
@@
 
-CP is usually implemented as part of a programming language and
-constraints are usually represented as objects \cite{puget95}. Any
+\textsf{CP} is usually implemented as part of a programming language
+and constraints are usually represented as objects \cite{puget95}. Any
 constraint that the user is able to reason about can be introduced
 into the system and, as long as the constraint propagation protocols
-specified by the host CP system are enforced, it will be
-indistinguishable from any other "primitive'' CP constraint, such as
-$\leq$ or $\geq$.
+specified by the host \textsf{CP} system are enforced, it will be
+indistinguishable from any other "primitive'' \textsf{CP} constraint,
+such as $\leq$ or $\geq$.
 

### 2012-02-15 arch.tex
rearranging text from concepts into europa-basics

@@
   of high level waypoints into low-level commands) and temporally as
-
   each reactor declares its planning look-ahead $\pi$ and expected
-  latency $\lambda$
-  % \comment{$\pi \& \lambda$ both defined here for the first time}
-  before being able to produce plan to the agent depending on where it
-  is located in a reactor dependency graph such as shown in
-  Fig. \ref{fig:agent}.
+  latency $\lambda$ before being able to produce plan to the agent
+  depending on where it is located in a reactor dependency graph such
+  as shown in Fig. \ref{fig:agent}.
 
@@
 ones which typicallu do not provide {\em internal} timelines.
-Conversely each reactor can request a future change on one of 
-its {\em external} state variables which will be transfromed as 
-an {\em internal} goal to the owner. 
-Such \emph{goal dispatching} is expected to produce a
-deliberation phase for the receiving reactor which can cyclically and
-in turn, produce a plan impacting its {\em external} state variables' 
-future that will propagate down following the same mechanism.
+Conversely each reactor can request a future change on one of its {\em
+  external} state variables which will be transfromed as an {\em
+  internal} goal to the owner.  Such \emph{goal dispatching} is
+expected to produce a deliberation phase for the receiving reactor
+which can cyclically and in turn, produce a plan impacting its {\em
+  external} state variables' future that will propagate down following
+the same mechanism.
 

### 2012-02-15 concepts.tex
rearranging text from concepts into europa-basics

@@
-\section{Foundational Concepts}
-\label{sec:concepts}
-
-\texttt{T-REX} is an adaptive, artificial intelligence based
-controller and provides general framework for building reasoning
-systems for real-world autonomous vehicles. At MBARI \texttt{T-REX} is
-used for AUV control; another instantiation of the system is being
-used for control of a terrestrial personal robot \cite{pr2,
-  Meeussen:2010dn}. The development of \texttt{T-REX} has been
-targeted at surveying a number of oceanographic features which are
-dynamic and unpredictable spatio-temporally. Typically this requires
-our AUVs to balance the goals of coverage spatially while
-opportunistically following features of scientific interest and to do
-so while being operationally aware of its own limitations in terms of
-resources (typically the battery state of charge) and overall
-proximity to other observational assets for obtaining scientific
-ground-truth. We are therefore interested in representational
-frameworks that allow robots to pursue long-term objectives while
-choosing short-term gain and can gracefully deal with exogenous or
-endogenous change.
-
-To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use a
-temporal constraint-based planner \eu with a demonstrated legacy of
-having flown on NASA space missions \cite{jonsson00,bresina05,
-  barreiro09}. Our autonomy architecture brings three key innovations
-for AUV adaptation: the use of flexible plan representations,
-compositional control with the use of partitioned networks and
-off-line learning to inform insitu environmental state estimation. In
-this section we detail some of the key concepts relevant to flexible
-plan representation. Section \ref{sec:arch} has details the
-\rx partitioned architecture and estimation driven vehicle
-adaptation.
-
-\subsection{The basics}
-\label{sec:basics}
-
-\eu uses a \emph{domain model} written in a declarative language
-(NDDL: New Domain Description Language), together with initial
-conditions and goals also specified in NDDL, to construct a set of
-temporal relations that must be true at the start time. These models
-include assertions about the physics of the vehicle, i.e how it
-responds to external stimulus and internally driven goals. By
-propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, \eu can select a set
-of conditions that should be true in the future, where some of these
-conditions will correspond to actions the agent must take. The planner
-can backtrack and try another path during search if a goal cannot be
-reached while being capable of discarding unachievable goals.
-
-Traditionally robot execution has relied on dispatching commands at
-precise times. Such linear sequences of precisely timed commands give
-no ability to adjust execution on the basis of sensory
-information. Although some commands can issue tests on sensor
-readings, these tests have the objective of verifying whether expected
-execution conditions are occuring. If not, the state of the system is
-declared off-nominal and execution of the sequence of commands is
-interrupted. More recently, executives have been proposed and
-implemented that significantly broaden the way robots can be commanded
-\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
-interpreted a \textit{temporally flexible plan} which represents each
-start time as a variable and contains an explicit network of bounded
-delay constraints between such variables.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
-\caption{\small Tokens with flexible temporal intervals and parametric
-  constraints between tokens. This example shows the triggering of a
-  water sampler based on a feature threshold while the vehicle
-  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
-  start \& end times.}
-\label{fig:flex-timelines}
-\vskip+0.1cm
-\end{figure}
-
-Unlike a traditional fixed time-tagged command sequences, such
-flexible plans leave room for adaptation at execution time. When the
-executive considers when to start a task, it propagates information
-through the constraint network, computes a time bound for the
-variable, selects an actual execution time within the bound, and
-starts the task at that time. Temporally flexible plans therefore,
-express a \textit{range of possible outcomes} of the robots
-interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
-
-\eu uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a \eu domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
-\label{fig:allen-algebra}
-\vskip-0.3cm
-\end{figure}
-
 \subsubsection{\eu Plan Representation}
@@
   agent needs to leave or be back, or executes a purchase, are all
-  instances where variables representing time would be used. 
-  As explained in the CP section, a variable can take values from a domain. In \eu, domains are defined over a specific data type, the primitive data types supported are:
+  instances where variables representing time would be used.  As
+  explained in the \textsf{CP} section, a variable can take values
+  from a domain. In \eu, domains are defined over a specific data
+  type, the primitive data types supported are:
+
   \begin{enumerate}
-    \item \textit{String}: sequences of characters, like "Red", "Spirit", etc 
+  \item \textit{String}: sequences of characters, like "Red",
+    "Spirit", etc
+
   \item \textit{Boolean}: {True,False}
+
   \item \textit{Numeric}: integers and floats
-    \item \textit{Object}: reference to an object instance (see below for a description of objects in \eu)
+
+  \item \textit{Object}: reference to an object instance (see below
+    for a description of objects in \eu)
+
 \end {enumerate}
+
 Two types of variable Domains can be defined over a data type:
+
   \begin{enumerate}
-    \item \textit{Enumerated}: Finite sets over any data type, specified explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
-    \item \textit{Interval}: Only for Numeric data types, they are specified by a [LowerBound,UpperBound] pair, for instance: [1,10], [5.0,100.0], etc.
+  \item \textit{Enumerated}: Finite sets over any data type, specified
+    explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
+  \item \textit{Interval}: Only for Numeric data types, they are
+    specified by a [LowerBound,UpperBound] pair, for instance: [1,10],
+    [5.0,100.0], etc.
   \end {enumerate}
-  \eu's ability to represent and reason over numeric intervals is a very useful characteristic that allow users to create flexible plans. \comment{say more about the benefits of flexibility?}
 
-\item[\textbf{Constraints}] valid plans in most problem domains have to satisfy a number of restrictions, for instance, a Shopping Agent may only be able to perform one task at a time, may have limited time or an energy or fuel budget to perform a task, and so on.
-In \eu, these restrictions are represented by Constraints, which can be defined over any combination of Variables in a domain.
-\eu provides a Constraint Library that already implements many useful constraints (temporal, resource, relational, etc), it is also possible to add new constraints if required by a particular domain.
+  \eu's ability to represent and reason over numeric intervals is a
+  very useful characteristic that allow users to create flexible
+  plans. \comment{say more about the benefits of flexibility?}
+
+\item[\textbf{Constraints}] valid plans in most problem domains have
+  to satisfy a number of restrictions, for instance, a Shopping Agent
+  may only be able to perform one task at a time, may have limited
+  time or an energy or fuel budget to perform a task, and so on. In
+  \eu, these restrictions are represented by Constraints, which can be
+  defined over any combination of Variables in a domain. \eu provides
+  a Constraint Library that already implements many useful constraints
+  (temporal, resource, relational, etc), it is also possible to add
+  new constraints if required by a particular domain.
 
@@
 
-  Objects that have similar state and behavior can be
-  described generically in terms of object types or classes. \eu
-  allows the definition of classes in the same way that it is
-  done in popular object-oriented programming languages, where a class 
-  can have attributes (represented by Variables), and classes can be arranged in a single-parent hierarchy through inheritance. 
-  However, one key aspect needs to be introduced to class definition in \eu that is not found 
-  in object-oriented programming:  to describe state and behavior for the purposes of planning we need to
-  build on the formalism of first order logic as explained below.
+  Objects that have similar state and behavior can be described
+  generically in terms of object types or classes. \eu allows the
+  definition of classes in the same way that it is done in popular
+  object-oriented programming languages, where a class can have
+  attributes (represented by Variables), and classes can be arranged
+  in a single-parent hierarchy through inheritance.  However, one key
+  aspect needs to be introduced to class definition in \eu that is not
+  found in object-oriented programming: to describe state and behavior
+  for the purposes of planning we need to build on the formalism of
+  first order logic as explained below.
 
@@
 
-  \item \textit {duration}: The duration of the temporal extent over which the predicate is defined. The constraint \textit{start} $+$
+  \item \textit {duration}: The duration of the temporal extent over
+    which the predicate is defined. The constraint \textit{start} $+$
     \textit{duration} = \textit{end} is enforced automatically.
@@
     token's current state and its reachable states through further
-    restriction.
-This is required to support CAIP's approach to planning where goals and subgoals can be satisfied by either adding new intervals (Tokens) to the plan, or by merging with existing intervals.
-Token State lifecycle is examined in detailed in a separate section below.
+    restriction.  This is required to support CAIP's approach to
+    planning where goals and subgoals can be satisfied by either
+    adding new intervals (Tokens) to the plan, or by merging with
+    existing intervals. Token State lifecycle is examined in detailed
+    in a separate section below.
 
@@
 
-\item[\textbf{Built-in Object Types}] In a domain model, there may be some classes that don't have any time-dependent state or behavior, and therefore those classes will not have any Token types associated with them, for instance in the Shopping Agent model, the set locations and products are static for the agent's purposes so we may only want to say what they are but not have any state or behavior associated with them.
+\item[\textbf{Built-in Object Types}] In a domain model, there may be
+  some classes that don't have any time-dependent state or behavior,
+  and therefore those classes will not have any Token types associated
+  with them, for instance in the Shopping Agent model, the set
+  locations and products are static for the agent's purposes so we may
+  only want to say what they are but not have any state or behavior
+  associated with them.
 
-However, the most common case in any non-trivial domain model is that Objects will have many associated Tokens in order to describe their state and behavior throughout a plan.
-There are a couple of Object Types that are so commonly used in domain descriptions, that \eu provides a built-in implementation for them:
+  However, the most common case in any non-trivial domain model is
+  that Objects will have many associated Tokens in order to describe
+  their state and behavior throughout a plan. There are a couple of
+  Object Types that are so commonly used in domain descriptions, that
+  \eu provides a built-in implementation for them:
 
 \begin{enumerate}
-    \item \textit{Timelines}: Often objects in a domain must be described by exactly one
-    token for every given timepoint in the plan. Any instances of a class derived from \eu's built-in
-    Timeline class will induce ordering requirements among its tokens
-     to ensure no temporal overlap may occur among them  \cite{mus94}. 
-     \comment{add small example?}
 
-    \item \textit{Resources}: Metric resources, e.g. the energy of a battery or the
-    capacity of a cargo hold, are objects with an explicit quantitative
-    state in time and with a circumscribed range of changes that can occur
-     to impact that state i.e. produce, consume, use, change. 
-     Resources are such a common requirement for \eu users that built-in object types (with their corresponding token types to denote production, consumption, etc) are provided for them.
-Instances of classes derived from a Resource will induce
+\item \textit{Timelines}: Often objects in a domain must be described
+  by exactly one token for every given timepoint in the plan. Any
+  instances of a class derived from \eu's built-in Timeline class will
+  induce ordering requirements among its tokens to ensure no temporal
+  overlap may occur among them \cite{mus94}.  \comment{add small
+    example?}
+
+\item \textit{Resources}: Metric resources, e.g. the energy of a
+  battery or the capacity of a cargo hold, are objects with an
+  explicit quantitative state in time and with a circumscribed range
+  of changes that can occur to impact that state i.e. produce,
+  consume, use, change.  Resources are such a common requirement for
+  \eu users that built-in object types (with their corresponding token
+  types to denote production, consumption, etc) are provided for
+  them. Instances of classes derived from a Resource will induce
   ordering requirements on their Tokens in order to ensure that the
-     level of the resource remains within specified limits.
-     \comment{add small example?}
+  level of the resource remains within specified limits.  \comment{add
+    small example?}
 \end{enumerate}
 
-\item[\textbf{Token State Model}] In \eu's representation, goals are temporally scoped states that need to be achieved, or actions that must be performed.
-As a result, \eu internally represents goals as Tokens.
-Goals can be posted explicitly, for instance, when a user asks the Shopping Agent to own a specific product by a specific time, or implicitly when subgoals are created as a result of domain compatibilities (described in the CAIP framework above), for instance, when the Shopping Agent must get to a Location where a Product is available before it is able to purchase it, in this case the original goal of purchasing the product will spawn a subgoal of being at a specific location.
+\item[\textbf{Token State Model}] In \eu's representation, goals are
+  temporally scoped states that need to be achieved, or actions that
+  must be performed. As a result, \eu internally represents goals as
+  Tokens.  Goals can be posted explicitly, for instance, when a user
+  asks the Shopping Agent to own a specific product by a specific
+  time, or implicitly when subgoals are created as a result of domain
+  compatibilities (described in the CAIP framework above), for
+  instance, when the Shopping Agent must get to a Location where a
+  Product is available before it is able to purchase it, in this case
+  the original goal of purchasing the product will spawn a subgoal of
+  being at a specific location.
 
-Looking at a how a simple Shopping Agent goal would be addresed by \eu helps illustrate how Token States support the planning process.
+  Looking at a how a simple Shopping Agent goal would be addresed by
+  \eu helps illustrate how Token States support the planning process.
 
-Let's assume we start with a partial plan that places the Shopping agent at home at time 0 and with no Products in its possession, then we post a goal for the Shopping Agent to own a Drill at time 10.
+  Let's assume we start with a partial plan that places the Shopping
+  agent at home at time 0 and with no Products in its possession, then
+  we post a goal for the Shopping Agent to own a Drill at time 10.
 
-\comment{Add figure showing an AgentLocation timeline with an At(Home) token with start=0 and end open, an empty AgentBag timeline, and a Own(Drill) token with start=10 floating below both timelines}
+  \comment{Add figure showing an AgentLocation timeline with an
+    At(Home) token with start=0 and end open, an empty AgentBag
+    timeline, and a Own(Drill) token with start=10 floating below both
+    timelines}
 
-Notice that the \texttt{Own(Drill)} token that represents the goal is not immediately placed on the \texttt{AgentBag} timeline since \eu must determine whether it can achieve that goal and how.
-When the goal is posted and before any decisions are made on it, the  \texttt{Own(Drill)} token is created in an INACTIVE state, then \eu looks at the current partial plan and considers 2 alternatives:
+Notice that the \texttt{Own(Drill)} token that represents the goal is
+not immediately placed on the \texttt{AgentBag} timeline since \eu
+must determine whether it can achieve that goal and how. When the goal
+is posted and before any decisions are made on it, the
+\texttt{Own(Drill)} token is created in an INACTIVE state, then \eu
+looks at the current partial plan and considers 2 alternatives:
 
 \begin{enumerate}
-    \item If there are any tokens in the plan that are compatible with the new goal, \eu may attempt to satisfy the goal by merging it with any of them (if there is more than one it can try them one at a time), in this case, the state of the token would change to MERGED.
-    \item Instead of merging with existing tokens in the plan, \eu may decide to insert the new token into the plan, in that case its state would change to ACTIVE.
+
+\item If there are any tokens in the plan that are compatible with the
+  new goal, \eu may attempt to satisfy the goal by merging it with any
+  of them (if there is more than one it can try them one at a time),
+  in this case, the state of the token would change to MERGED.
+
+\item Instead of merging with existing tokens in the plan, \eu may
+  decide to insert the new token into the plan, in that case its state
+  would change to ACTIVE.
+
 \end{enumerate}
 
-When a token is merged, all restrictions imposed
-on the merged token are transferred to the active token upon which it is merged. Merging a
+When a token is merged, all restrictions imposed on the merged token
+are transferred to the active token upon which it is merged. Merging a
 token requires finding a target active token that is compatible with
-the inactive token. Two tokens are merge-compatible if they are instances of the same predicate and
-that no intersections between corresponding variables are empty. The
-effect of merging is illustrated in Figure 3.
+the inactive token. Two tokens are merge-compatible if they are
+instances of the same predicate and that no intersections between
+corresponding variables are empty. The effect of merging is
+illustrated in Figure 3.
 
@@
 
-When a token is activated all compatibilities associated with its token type are evaluated, which may lead to new subgoals being generated.
-In our example, there should be a compatibility stating that the Shopping Agent must be at the same location as the product it is purchasing.
-In that case, activating the \texttt{Own(Drill)} token will cause a new  \texttt{At(Drill.location)} to be created in an INACTIVE state, which in turn will have to be dealt with.
+When a token is activated all compatibilities associated with its
+token type are evaluated, which may lead to new subgoals being
+generated. In our example, there should be a compatibility stating
+that the Shopping Agent must be at the same location as the product it
+is purchasing. In that case, activating the \texttt{Own(Drill)} token
+will cause a new \texttt{At(Drill.location)} to be created in an
+INACTIVE state, which in turn will have to be dealt with.
 
-\comment{Add figure showing the effect of activating the token by putting it on the AgentBag timeline, and a new At(Drill.location) token floating below}
+\comment{Add figure showing the effect of activating the token by
+  putting it on the AgentBag timeline, and a new At(Drill.location)
+  token floating below}
 
-If neither activating nor merging works, a token may become REJECTED, this is possible only for goals posted by the end user and specified as optional.
-Subgoals generated as a result of token activation cannot be rejected, if a subgoal cannot be satisfied through activation or merging then it will cause the original activation decision that spawned it to be retracted (\eu's search mechanisms are explained in detail in a separate section later on).
+If neither activating nor merging works, a token may become REJECTED,
+this is possible only for goals posted by the end user and specified
+as optional. Subgoals generated as a result of token activation cannot
+be rejected, if a subgoal cannot be satisfied through activation or
+merging then it will cause the original activation decision that
+spawned it to be retracted (\eu's search mechanisms are explained in
+detail in a separate section later on).
 
-As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED or REJECTED, this lifecycle is illustrated in the figure below.
+As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED
+or REJECTED, this lifecycle is illustrated in the figure below.
 
@@
 
-\item[\textbf{Rules}] In order for a plan to be valid, it must comply with all rules and
-regulations pertinent to the application domain in question. Rules
-govern the internal and external relationships of a token. For
-example, consider a parameterized action \texttt{Go(origin,destination)} which describes
-a Shopping Agent traveling from one location ( \textit{origin}) to another ( \textit{destination}). The
-parameters are instantiated on a token as variables whose domain of
-values is the set of all locations in a given problem. A rule
-governing an internal relationship among token variables might
-stipulate that a transition must involve a change in location. This
-can be easily expressed as a constraint on the definition of a
-predicate of the form: \textit{origin != destination}. It makes sense to further
-stipulate that the agent must be located at  \textit{origin} before travel can happen
-and the agent must be located at  \textit{destination} when completed. This is
-an example of a rule governing an external relationship among
+\item[\textbf{Rules}] In order for a plan to be valid, it must comply
+  with all rules and regulations pertinent to the application domain
+  in question. Rules govern the internal and external relationships of
+  a token. For example, consider a parameterized action
+  \texttt{Go(origin,destination)} which describes a Shopping Agent
+  traveling from one location ( \textit{origin}) to another (
+  \textit{destination}). The parameters are instantiated on a token as
+  variables whose domain of values is the set of all locations in a
+  given problem. A rule governing an internal relationship among token
+  variables might stipulate that a transition must involve a change in
+  location. This can be easily expressed as a constraint on the
+  definition of a predicate of the form: \textit{origin !=
+    destination}. It makes sense to further stipulate that the agent
+  must be located at \textit{origin} before travel can happen and the
+  agent must be located at \textit{destination} when completed. This
+  is an example of a rule governing an external relationship among
   tokens. It specifies a requirement that tokens of the predicate
   \texttt{At} that represent the location state for the Shopping Agent
-precede and succeed tokens of the predicate \texttt{Go} that represents traveling.
+  precede and succeed tokens of the predicate \texttt{Go} that
+  represents traveling.
 
-\comment{ Add figure showing internal and external relationships from rules on action Go}
+  \comment{ Add figure showing internal and external relationships
+    from rules on action Go}
 
-Figure xx illustrates the entities and relations involved in specifying
-such a rule on a \texttt{Go} action. The token on which the rule applies
-is referred to (\texttt{Go}) as the master. Each  \texttt{At} token required by the
-master is referred to as a slave. All variables are indicated by name
-and their domains are expressed as intervals in the case of temporal
-variables and as enumerations for the remainder. Application of a rule
-on a token can thus cause slave tokens, variables, and constraints to
-be introduced.
+Figure xx illustrates the entities and relations involved in
+specifying such a rule on a \texttt{Go} action. The token on which the
+rule applies is referred to (\texttt{Go}) as the master. Each
+\texttt{At} token required by the master is referred to as a
+slave. All variables are indicated by name and their domains are
+expressed as intervals in the case of temporal variables and as
+enumerations for the remainder. Application of a rule on a token can
+thus cause slave tokens, variables, and constraints to be introduced.
 
@@
 is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting (slave) tokens to be in the plan in
-order for the master to be valid.
+rule imposes a requirement for supporting (slave) tokens to be in the
+plan in order for the master to be valid.
 
@@
 
-We have just covered the main elements of \eu's planning representation:
+We have just covered the main elements of \eu's planning
+representation:
 
 \begin{enumerate}
-   \item \textbf{Domains, Variables and Objects} to represent the entities (Objects) and attributes (Variables) needed to represent a problem domain	
-    \item \textbf{Tokens} Temporally scoped predicates to represent states and actions in time. 
-    \item \textbf{Token States} which are necessary for the implementation of planning operations on a partial plan.
-    \item \textbf{Constraints} to describe relationships among attributes of objects and tokens.
-\eu has a built-in library of constraints that covers time, resources and others that are common in real life problem domains.
-    \item \textbf{Domain Rules} to describe internal and external relationships on and
-     between tokens respectively. Rules are applied to active tokens (master tokens) 
-     referred to as masters and typically produce inactive tokens (slaves) which are a key vehicle through which the planning
-     process evolves. \comment{citation, or perhaps remove this statement?:}The rule structure in \eu differs from the more restrictive commitment to preconditions and effects in classical
-     planning which prohibits durative actions and disembodied effects (i.e. effects which may not occur until some temporal distance after
-     the end of the action).    
-    \item \textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
+\item \textbf{Domains, Variables and Objects} to represent the
+  entities (Objects) and attributes (Variables) needed to represent a
+  problem domain
+
+\item \textbf{Tokens} Temporally scoped predicates to represent states
+  and actions in time.
+
+\item \textbf{Token States} which are necessary for the implementation
+  of planning operations on a partial plan.
+
+\item \textbf{Constraints} to describe relationships among attributes
+  of objects and tokens. \eu has a built-in library of constraints
+  that covers time, resources and others that are common in real life
+  problem domains.
+
+\item \textbf{Domain Rules} to describe internal and external
+  relationships on and between tokens respectively. Rules are applied
+  to active tokens (master tokens) referred to as masters and
+  typically produce inactive tokens (slaves) which are a key vehicle
+  through which the planning process evolves. \comment{citation, or
+    perhaps remove this statement?:}The rule structure in \eu differs
+  from the more restrictive commitment to preconditions and effects in
+  classical planning which prohibits durative actions and disembodied
+  effects (i.e. effects which may not occur until some temporal
+  distance after the end of the action).
+
+\item \textbf{Timelines and Resources} are built in implementations of
+  Object types that appear frequently in real life domains.
+
 \end{enumerate}

### 2012-02-15 arch.tex
added a new figure to illustrate planning window of a reactor and how it relates to reactor latency and look-ahead

@@
 \comment{this might be a subsubsection before someplace since it
-  introduces the concept of latency/lookahead.}
+  introduces the concept of latency/lookahead. TODO: yep it needs to
+  move up to the architecture section.}
 These events could potentially produce new flaws to be resolved while
@@
 
-\comment{Insert a figure here [AAMAS?] showing $\pi \lambda$} Such a
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.65\columnwidth]{figs/plan_window}
+  \caption{The planning window of the deliberation solver and its
+    relation with the reactor latency and look-ahead. The dark grey tokens
+    with dashed borders are excluded from deliberation as they do not
+    overlap the interval $[\tau, \tau+\lambda+\pi]$}
+  \label{fig:plan:window}
+\end{figure}
+
+As Fig. \ref{fig:plan:window} illustrates,  such a
 scope allows filtering of tokens that are either necessarily ending

### 2012-02-16 trunk/europa-basics.tex
cleaned up constraint-bsed planning section.

@@
 achieve stated goals) and both Planning and Scheduling may embed a
-Constraint Satisfaction problem. % since temporal, resource and other
-% kinds of constraints must be respected in most real life problems.
+Constraint Satisfaction problem. 
 The relationships between Planning, Scheduling and Constraint
@@
 
-\subsection{Constraint-Based Attribute and Interval Planning}
+\subsection{Constraint-Based Planning}
 \label{sec:europa:cp}
@@
 mutual exclusion and concurrency using propositions \comment{citation
-  needed}. While CSP representations have traditionally had an edge in
-formulating and representing planning problems, 
-% It is straightforward to represent and reason about all of
-% those elements using a CSP representation, as a result there has been
-% some work on doing automated planning while taking advantage of CSP
-% (TODO: ref). Traditionally, the entire planning problem is translated
-% into a CSP and then solved using traditional CSP methods, this
-leading to a formulation where action choices and relationships are
-represented as variables and/or constraints, there are at least two
-major drawbacks:
+  needed}. 
+Constraint Programming is a framework that lends itself very well to represent and reason about all of those elements \comment{Ixtet citation?},
+consequently, there has been a good amount of work building automated planners that take advantage of CP techniques.
+The most common approach in constraint-based planning has been to encode the entire planning problem as a CP problem, 
+where action choices and relationships are represented as variables and/or constraints \comment {citations 158,519,539,556 from malik's book}.
+The advantage of using a pure CP formulation is that standard CP solvers can be then used to solve the planning problem, however, this approach also has a couple of significant drawbacks:
 
@@
 
-Constraint-based Planning and specifically Constraint-Based Attribute
-and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
-close that gap. On one hand, it takes advantage of a CSP
-representation; on the other it uses attributes and intervals to
-maintain an explicit representation of the elements relevant to
-planning. \eu is an implementation of the CAIP framework.  This makes
-it easier to write algorithms that search for an reason abut
-plans. The main elements of CAIP are:
+These drawbacks have been addressed by a small number of planning frameworks like Descartes \comment{citation 289,290 from malik's book}, IxTet \comment{citation  223 from Malik's book}, and CAIP (Constraint-Based Interval and Attribute Planning)\cite{mus94,frank2003}.
+These frameworks take advantage of a CSP representation to deal with time and resources, at the same time maintaining an explicit representation of the elements relevant to planning (action choices, causal relationships, etc).
+This combination results in a planning approach that can deal with the type of constraints found in real-life problems, while allowing for intuitive problem descriptions and a natural way to express search algorithms and heuristics.
+\eu is an implementation of the CAIP framework.
+In CAIP, planning problems are specified in terms of:
 
 \begin{enumerate}
-	\item Intervals: \comment{needs to be defined}
-	\item Attributes:
-	\item Domain Constraints an Configuration rules:
+	\item \textbf{Intervals}: entities that have a temporal scope (that is, a beginning and an  end in time) and a set of attributes.
+Intervals are used to represent actions and states in the planning domain.
+	\item \textbf{Variables and Constraints}: Variables represent Interval attributes and Constraints enable a direct representation of temporal, resource and any other the restrictions that a valid plan must comply with.
+	\item \textbf{Domain Configuration Rules}: this is an explicit representation of the conceptual relationships between actions and states in the problem domain, for instance, for a Shopping Agent to be able to purchase a product, it must be at a location where the product is available.
 \end{enumerate}
 
-\comment{Explain how planning problem instances and their solutions are represented}
+To generate plans, the CAIP framework uses a plan-space search approach \comment{cite malik's chapter 5}, where a partial plan is evolved until all open goals are achieved, and all inconsistencies are removed, the specific search algorithm used by \eu will be covered in detail below.

### 2012-02-16 concepts.tex
rearranging text from concepts into europa-basics
smoother transition into "europa's plan representation" section

@@
 
-This section describes the fundamental entities within \eu which are
-typical to other CAIP-based formalisms.
+\eu follows the representation outlined by the CAIP framework, it also adds a few elements to it that are necessary for an efficient implementation of constraint-based planning.
+Below is a brief summary of the main elements, followed by a detailed description for each one:
+
+\begin{enumerate}
+\item \textbf{Tokens} Temporally scoped entities that correspond to Intervals in the CAIP framework, are used to represent  state and actions in time.
+A\textbf{Token State model} is defined to support an efficient implementation of plan-space search algorithms.
+
+\item \textbf{Domains, Variables and Constraint} are used to represent restrictions in the problem domain in terms of a CP formulation.
+
+\item \textbf{Domain Rules} are used to describe internal and external relationships on and between tokens, as presecribed by the CAIP framework.
+
+\item \textbf{Objects} are used to represent problem domains in a more modular and scaleable way.
+\textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
+
+\end{enumerate}
 
@@
 
-We have just covered the main elements of \eu's planning
-representation:
-
-\begin{enumerate}
-\item \textbf{Domains, Variables and Objects} to represent the
-  entities (Objects) and attributes (Variables) needed to represent a
-  problem domain
-
-\item \textbf{Tokens} Temporally scoped predicates to represent states
-  and actions in time.
-
-\item \textbf{Token States} which are necessary for the implementation
-  of planning operations on a partial plan.
-
-\item \textbf{Constraints} to describe relationships among attributes
-  of objects and tokens. \eu has a built-in library of constraints
-  that covers time, resources and others that are common in real life
-  problem domains.
-
-\item \textbf{Domain Rules} to describe internal and external
-  relationships on and between tokens respectively. Rules are applied
-  to active tokens (master tokens) referred to as masters and
-  typically produce inactive tokens (slaves) which are a key vehicle
-  through which the planning process evolves. \comment{citation, or
-    perhaps remove this statement?:}The rule structure in \eu differs
-  from the more restrictive commitment to preconditions and effects in
-  classical planning which prohibits durative actions and disembodied
-  effects (i.e. effects which may not occur until some temporal
-  distance after the end of the action).
-
-\item \textbf{Timelines and Resources} are built in implementations of
-  Object types that appear frequently in real life domains.
-
-\end{enumerate}
-
 %\begin{figure*}

### 2012-02-16 trunk/europa.tex
smoother transition into the "europa architecture" section.

@@
 \label{sec:europa:arch}
+
+\eu is implemented as a library that encapsulates all the elements described above and makes them easily accessible through modeling and API layers.
+A user will typically use the modeling facilities to inject a problem domain description, and the API to extract and manipulate plans, and to obtain information about plan flaws and constraint violations at any stage of the planning process.
 

### 2012-02-16 arch.tex
Added/refined material around the notion of planning and dispatch window for a reactor

@@
 
-This {\em synchronization} process propagates the current state
-information at the execution frontier $\tau$ throughout all the
-reactors. The sequencing of the reactors synchronization is identified
+As the execution frontier -- called $\tau$ -- advance, the \rx agent
+propagates the current state information accross all the reactors states
+variables through the {\em synchronization} process.
+ The sequencing of the reactors synchronization is identified
 by traversing the reactor graph from the least dependents reactors --
@@
 ones which typicallu do not provide {\em internal} timelines.
+
 Conversely each reactor can request a future change on one of its {\em
-  external} state variables which will be transfromed as an {\em
+  external} state variables which will be transformed as an {\em
   internal} goal to the owner.  Such \emph{goal dispatching} is
@@
 the same mechanism.
+
+In order to avoid to be overflood by goals that may threats the
+ability of a reactor to process these goals and integrate them in tis
+plan. In order to do so each reactor specifies the two following
+values that apply for the full scope of the agent lifetime:
+\begin{itemize}
+\item $\lambda$ is the specified latency of this reactor, which is an
+  indicator of the maximum expected time for a reactor to produce a
+  plan.\footnote{Note that while it is highly recommended to select a
+    sound value, a failure to produce a plan within a chosen value for
+    this parameter is not considered as a critical failure of the
+    reactor.}
+\item $\pi$ is the planning look-ahead of the reactor. which indicates
+  how many ticks in the future this reactor is looking ahead during
+  deliberation. 
+\end{itemize}
+
+Using this information along with the dependency relation betwen
+reactor \rx defines the notion of execution latency of a state
+variable as follow:
+\begin{definition}
+  The {\em execution latency}  of a {\em state variable} $s$ which is
+    internal to the reactor $r$ is the maximum expected delay before 
+    a goal on this state varabale can be included in its plan and
+    executed by the  reactor $r$.
+
+    It is defined recursrevely as follow: 
+    \begin{equation*}
+      \lambda_{exec}(s) = \lambda(r) + \max_{e \in exts(r)}
+      \lambda_{exec}(e) 
+  \end{equation*}
+    Where  $exts(r)$ is the set of the state varaibles which are
+    {\em externals} to $r$. 
+\end{definition}
+
+The concept of execution latency is critical at the agent level
+and can also impact the planning/deliberation process on the {\em
+  external} state variables of a reactor. For simplicity sake in the
+future sections more focused in the \eu reactor, we abstract it out 
+as being equivalent to the reactor latency (meaning that we 
+assume that all {\em external} timelines of this reactor have
+$\lambda_{exec} = 0$). At the \rx agent level though this
+ execution latency is used in turn to define the {\em dispatch
+window} of the same state variable:
+\comment{Needed to say that early on but the flow of this paragraph 
+is not very tight}
+\begin{definition}
+  \label{def:dispatch}
+  For every timelines $s$ {\em internals} to the reactor $r$, the {\em 
+    dispatch window} of this timeline at the execution frontier $\tau$
+  is the time interval:
+  \begin{equation*}
+    \Pi(s) = [\tau + \lambda_{exec}(s), \tau+\lambda_{exec}(s)+\pi(r)]
+  \end{equation*}
+\end{definition}
+
+This window is used by \rx in order to identify when to dispatch a
+request to a {\em reactor} as a new goal. This happens only if the
+request start time -- represented as the interval of possible start
+instants -- intersects this window. Requests further in the future 
+are buffered by the agent until this window advanced enough to 
+overlap their start time. This gives a simple mechanism in order to
+request a change from the {\em external} state variable of a a reactor 
+while avoiding to dispatch too much goals at once to the reactor that
+declare this same state variable as {\em internal}. The choice of both
+the latency and look-ahead of a reactor can be delicate as one is
+dependent on the other. Latency also highly dependent on exogenous
+paramaters such as the computing power of the host system and the
+number of reactors needing to deliberate at once. As of today these
+should be evaluated carefully by the system designers and while
+having an important impact in the efficacy of the agent to find and
+execute a global strategy they are just indicative and a failure from
+a reactor to produce a plan in its latency is not considered as a
+critical failure from \rx point of view.
+
+Requests from a reactor can  be {\em recalled} by this reactor when it
+they are not part of its plan anymore. For example if this reactor
+need to replan due to an unexpected event, it will recall ol the
+requests it did propduce that have not been observed yet. These
+recalls are notified to the reactor that received this request as a
+goal in order to remove it from its current objectives.
 
@@
 
-The number of flaws between two steps \comment{between synch. and
-  execution?} can evolve for the following reasons that are exogenous
-to deliberation:
-
+{\em Synchronization} and the dispatching of new {\em goals} from
+\rx can occur between two steps and alter the plan. These alterations
+exogenous to the deliberation process are dynamically impacting the set
+of flaws that need to be resolved. Obviously new flaws are created
+when a new goal is received from \rx and conversely often the recall
+of an objective will remove the corresponding flaws\footnote{although
+  if the flaws of this gaol were previously resolved, the removal of
+  this goal may create new flaws}. The impact of {\em synchronization}
+can be more subtle as its alteration of the plan at the execution
+frontier is much more drastic. We can still give a non exhaustive list
+of possible alterations of the flaws directly related to {\em
+  synchronization}:
 \begin{enumerate}
-
-\item A new goal has been posted on a reactor timeline. This is
-  usually due to another reactor finding a partial plan and posting
-  the corresponding new objective.
-
-\item The outcome of synchronization created new flaws that did not
-  need to be resolved during synchronization.
-
-\item A previously requested goal has be recalled by the initial
-  requester. This can occur when the reactor's initial plan is no
-  longer valid, for instance with synchronization resulting in
-  inconsistency.
-
+\item Previous flaws directly related to the execution frontier have
+  been resolved by the {\em synchronization} solver.
+\item The {\em synchronization} solver may have created new flaws that
+  were outside of its scope but are now part of the {\em deliberation}
+  solver scope.
+\item Last observations proved the current plan to be invalid
+  requiring the {\em synchronization} solver to relax the existing
+  plan in order to identify its current {\em external} state.
 \end{enumerate}
-
-\comment{this might be a subsubsection before someplace since it
-  introduces the concept of latency/lookahead. TODO: yep it needs to
-  move up to the architecture section.}
-These events could potentially produce new flaws to be resolved while
-any single deliberation step will likely reduce the number of flaws
-for resolution. The scope of planning for a reactor also plays a role
-in the deliberation process as we limit deliberation to the sliding
-temporal window $[\tau+\lambda, \tau+\lambda+\pi]$ \cite{py10} where:
-
-\begin{itemize}
-
-\item $\tau$ is the current tick indicative of the execution frontier
-
-\item $\lambda$ is the specified latency of this reactor, which is an
-  indicator of the maximum expected time for a reactor to produce a
-  plan.\footnote{Note that while it is highly recommended to select a
-    sound value, a failure to produce a plan within a chosen value for
-    this parameter is not considered as a critical failure of the
-    reactor.}
-
-\item $\pi$ is the planning look-ahead of the reactor. which indicates
-  how many ticks in the future this reactor is looking ahead during
-  deliberation.
-
-\end{itemize}
+Finally as {\em synchronization} occurs when the {\em
+  execution frontier} advances, it implies that the reactor planning
+scope is further in the future. This need to be reflected by the {\em
+  deliberation} solver which need a new deliberation scope one tick
+further in the future.
 
@@
   \includegraphics[width=0.65\columnwidth]{figs/plan_window}
-  \caption{The planning window of the deliberation solver and its
-    relation with the reactor latency and look-ahead. The dark grey tokens
-    with dashed borders are excluded from deliberation as they do not
-    overlap the interval $[\tau, \tau+\lambda+\pi]$}
+  \caption{\small The planning window of the deliberation solver and its
+    relation with the {\em dispatch window}. The dark grey
+    tokens   with dashed borders are excluded from deliberation as they do not
+    overlap the interval $[\tau, \tau+\lambda+\pi]$. All the other
+    tokens (in {\color{green} green}) -- along with the goals received from the
+    agent (in {\color{blue} blue}) which can overlap this window are
+    evaluated.}
   \label{fig:plan:window}
@@
 problem on only those tokens that are within a ``reasonable''
-future. This window is taken into account by the \rx agent which will
-notify a reactor of new {\em internal} goals not only when they
-overlap this reactor specific window but also for the deliberation
-solver in order to reduce the number of tokens to be evaluated to
+future. This filtering window complements the \rx {\em dispatch
+  window} from definition \ref{def:dispatch} which already filters
+out at the agent level the goals that are before the reactor execution
+latency or too far in the future. The planning window inside of the
+reactor refines further this filtering as tokens can be excluded
+dynamically from deliberation as their insertion in the plan and
+scheduling with other tokens generated can reduced their temporal
+scope outside of the planning scope. This helps the deliberation 
+solver reduce the number of tokens to be evaluated and therefore
 reduce deliberation cost and performance. As time advances flaws that

### 2012-02-16 trunk/concepts.tex
merged minor mods to europa secs

--- 2012-01-12 trunk/concepts.tex
@@
-\section{Foundational Concepts}
-\label{sec:concepts}
+\subsubsection{\eu Plan Representation}
+\label{sec:europa:pr}
 
-Our system, \texttt{T-REX} (or the Teleo-Reactive EXecutive) is an
-adaptive, artificial intelligence based controller and provides
-general framework for building reasoning systems for real-world
-autonomous vehicles. At MBARI \texttt{T-REX} is used for AUV control;
-another instantiation of the system is being used for control of a
-terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The development of
-\texttt{T-REX} has been targeted at surveying a number of
-oceanographic features which are dynamic and unpredictable
-spatio-temporally. Typically this requires our AUVs to balance the
-goals of coverage spatially while opportunistically following features
-of scientific interest and to do so while being operationally aware of
-its own limitations in terms of resources (typically the battery state
-of charge) and overall proximity to other observational assets for
-obtaining scientific ground-truth. We are therefore interested in
-representational frameworks that allow robots to pursue long-term
-objectives while choosing short-term gain and can gracefully deal with
-exogenous or endogenous change.
+\eu follows the representation outlined by the CAIP framework, it also adds a few elements to it that are necessary for an efficient implementation of constraint-based planning.
+Below is a brief summary of the main elements, followed by a detailed description for each one:
 
-To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use a
-temporal constraint-based planner EUROPA$_2$ with a demonstrated
-legacy of having flown on NASA space missions
-\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
-innovations for AUV adaptation: the use of flexible plan
-representations, compositional control with the use of partitioned
-networks and off-line learning to inform insitu environmental state
-estimation. In this section we detail some of the key concepts
-relevant to flexible plan representation. Section \ref{sec:arch} has
-details the \texttt{T-REX} partitioned architecture and estimation
-driven vehicle adaptation.
+\begin{enumerate}
+\item \textbf{Tokens} Temporally scoped entities that correspond to Intervals in the CAIP framework, are used to represent  state and actions in time.
+A\textbf{Token State model} is defined to support an efficient implementation of plan-space search algorithms.
 
-% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
-% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
-% for the Mars Exploration Rovers \cite{bresina05}.
+\item \textbf{Domains, Variables and Constraint} are used to represent restrictions in the problem domain in terms of a CP formulation.
 
-EUROPA$_2$ uses a domain model written in a declarative language
-(NDDL: New Domain Description Language), together with initial
-conditions and goals also specified in NDDL, to construct a set of
-temporal relations that must be true at the start time. These models
-include assertions about the physics of the vehicle, i.e how it
-responds to external stimulus and internally driven goals. By
-propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
-a set of conditions that should be true in the future, where some of
-these conditions will correspond to actions the agent must take. The
-planner can backtrack and try another path during search if a goal
-cannot be reached while being capable of discarding unachievable
-goals.
+\item \textbf{Domain Rules} are used to describe internal and external relationships on and between tokens, as presecribed by the CAIP framework.
 
-Traditionally robot execution has relied on dispatching commands at
-precise times. Such linear sequences of precisely timed commands give
-no ability to adjust execution on the basis of sensory
-information. Although some commands can issue tests on sensor
-readings, these tests have the objective of verifying whether expected
-execution conditions are occuring. If not, the state of the system is
-declared off-nominal and execution of the sequence of commands is
-interrupted. More recently, executives have been proposed and
-implemented that significantly broaden the way robots can be commanded
-\cite{mus98,Alami:1998p820}. For example, the Remote Agent executive
-interpreted a \textit{temporally flexible plan} which represents each
-start time as a variable and contains an explicit network of bounded
-delay constraints between such variables.
+\item \textbf{Objects} are used to represent problem domains in a more modular and scaleable way.
+\textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
 
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
-\caption{\small Tokens with flexible temporal intervals and parametric
-  constraints between tokens. This example shows the triggering of a
-  water sampler based on a feature threshold while the vehicle
-  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
-  start \& end times.}
-\label{fig:flex-timelines}
-\vskip+0.1cm
-\end{figure}
+\end{enumerate}
 
-EUROPA$_2$ uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a EUROPA$_2$ domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
+\begin{description}
 
-Unlike a traditional fixed time-tagged command sequences, such
-flexible plans leave room for adaptation at execution time. When the
-executive considers when to start a task, it propagates information
-through the constraint network, computes a time bound for the
-variable, selects an actual execution time within the bound, and
-starts the task at that time. Temporally flexible plans therefore,
-express a \textit{range of possible outcomes} of the robots
-interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
+\item[\textbf{Variables}] Values that need to be represented to
+  describe the problem domain and over which we may want to specify
+  constraints. In the Shopping Agent example, the times at which the
+  agent needs to leave or be back, or executes a purchase, are all
+  instances where variables representing time would be used.  As
+  explained in the \textsf{CP} section, a variable can take values
+  from a domain. In \eu, domains are defined over a specific data
+  type, the primitive data types supported are:
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
-\label{fig:allen-algebra}
-\vskip-0.3cm
-\end{figure}
+  \begin{enumerate}
+  \item \textit{String}: sequences of characters, like "Red",
+    "Spirit", etc
 
-Each of the token variables, including the parameter variables, has a
-domain of values assigned to it.  The variables may also participate
-in constraints that specify which value combinations are valid.  To
-allow the specification of multiple values, e.g, to express a range of
-possible start times, variables are used to specify parameter, start
-and end time values for a token.  As a result, a token $T$ is a tuple
-$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
-variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
-$s\leq e$) are variables encoded as flexible token start and end time
-points. Consequently, partial plans represent a \textit{range of
-  possible solutions} rather than an explicit definition of a single
-trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
-of a part of such plans with tokens representing constraints on
-flexible temporal intervals. Systematic computation is reflected by
-the use of the \texttt{Allen Algebra} \cite{allen84} temporal
-relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
-interval arithmetic in the EUROPA$_2$ domain model that allows
-encoding of temporal constraint relationships between tokens on the
-same or other concurrent timelines. Details of propagation algorithms
-or of interval logic are beyond the scope of this paper.
+  \item \textit{Boolean}: {True,False}
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
-\caption{\small Timeline evolution can be described by simple finite
-  state transitions where temporal relationships describe the arc
-  transition and temporal extant of state are token
-  durations. Constraints here are for \emph{meets} or its equivalent
-  inverse \emph{met\_by}.}
-\label{fig:FSM}
-\vskip-0.3cm
-\end{figure}
+  \item \textit{Numeric}: integers and floats
 
-In describing the evolution of state via timelines, transitions within
-a timeline can be described by simple finite state
-machines. Fig. \ref{fig:FSM} for example describes a timeline
-evolution for an abstract timeline entity describing navigation. Arc
-transitions represent temporal constraints and token duration are
-represented by temporal extant of state. The entire vehicle can then
-be modeled as a number of state transitions with concurrency for each
-of the modeled state variables defined a priori. However this simple
-view is often complicated when temporal constraints between state
-variables are articulated in our domain models. If there is a
-criticism of this approach, modeling then, requires a proper balance
-between abstraction, state variable description and concurrency
-described by the inter-connectedness between such finite state
-machines.
+  \item \textit{Object}: reference to an object instance (see below
+    for a description of objects in \eu)
 
-\begin{figure*}
-\centering 
-\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
-\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
-\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
-\caption{\small An illustrative plan synthesis example with concurrent
-  timelines. Each timeline is an instantiation of a subsystem tracked
-  by the planner. Tokens describe the instantiated state of subsystem
-  at a particular time instance and enforce temporal and parametric
-  constraints. Abstract goals are decomoposed into successively less
-  abstract tokens and instantiated in co-temporal timelines using
-  \texttt{Allen Algebra} relations. All tokens represent flexible
-  start/end times. \ref{fig:plan-evolve1} shows an initial state
-  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
-  \label{fig:Plan-evolve}
-  \vskip-5pt
-\end{figure*}
+\end {enumerate}
 
-Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
-timelines which track \texttt{Goal, Path, Navigation} and
-\texttt{Command} state over time. Tokens on the goal timeline indicate
-a survey within a bounded volume and a min/max depth envelope. As the
-Volume Survey sub-goals on the \texttt{Path} timeline, the initial
-token on that timeline gets \emph{squeezed} to have dependencies of
-the goal token instantiated. Initially this dependency are the tokens
-\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
-coverage. The first of these in turn generates a sub-goal on the same
-timeline to go to the next intermediate waypoint, which is illustrated
-in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
-figure) ultimately generate all the tokens to completion for this
-timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
-in plan generation and shows additional subgoals on the \texttt{Path,
-  Navigation} and \texttt{Command} timelines. Each token can thus
-trace its causality when instantiated in the plan. The domain model
-provides source of these temporal and parametric dependencies in the
-form of rules.
+Two types of variable Domains can be defined over a data type:
 
-\begin{minipage}[c]{\textwidth}
+  \begin{enumerate}
+  \item \textit{Enumerated}: Finite sets over any data type, specified
+    explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
+  \item \textit{Interval}: Only for Numeric data types, they are
+    specified by a [LowerBound,UpperBound] pair, for instance: [1,10],
+    [5.0,100.0], etc.
+  \end {enumerate}
 
+  \eu's ability to represent and reason over numeric intervals is a
+  very useful characteristic that allow users to create flexible
+  plans. \comment{say more about the benefits of flexibility?}
+
+\item[\textbf{Constraints}] valid plans in most problem domains have
+  to satisfy a number of restrictions, for instance, a Shopping Agent
+  may only be able to perform one task at a time, may have limited
+  time or an energy or fuel budget to perform a task, and so on. In
+  \eu, these restrictions are represented by Constraints, which can be
+  defined over any combination of Variables in a domain. \eu provides
+  a Constraint Library that already implements many useful constraints
+  (temporal, resource, relational, etc), it is also possible to add
+  new constraints if required by a particular domain.
+
+\item[\textbf{Objects}] The items we wish to describe and refer to in
+  a domain are considered Objects. As in the case with object-oriented
+  analysis and design, one can seek out the nouns in any domain
+  description to find likely objects. In the Shopping Agent example,
+  we might consider the Agent itself, Products and Locations all to be
+  objects. Objects have state and behavior. For example, a Shopping
+  Agent can have:
+
+  \begin{itemize}
+
+  \item State: its location, a bag that contains the products it has
+    already purchased (the bag can in turn be another object), etc.
+
+  \item Behavior: go to a location to look for a product, perform a
+    purchase, return home, etc.
+
+  \end{itemize}
+
+  Objects that have similar state and behavior can be described
+  generically in terms of object types or classes. \eu allows the
+  definition of classes in the same way that it is done in popular
+  object-oriented programming languages, where a class can have
+  attributes (represented by Variables), and classes can be arranged
+  in a single-parent hierarchy through inheritance.  However, one key
+  aspect needs to be introduced to class definition in \eu that is not
+  found in object-oriented programming: to describe state and behavior
+  for the purposes of planning we need to build on the formalism of
+  first order logic as explained below.
+
+\item[\textbf{Tokens}] In first order logic, a predicate defines a
+  relation between objects and properties \comment{citation?}.  In
+  \eu, we define such relations between variables whose domains are
+  sets of objects and sets of properties to describe state and
+  behavior. For example, we might use a predicate \texttt{At($a$,$l$)}
+  to indicate that agent $a$ is at location $l$, or a predicate
+  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
+  buy product $p$. Note that $a$ is a variable which may have a number
+  of possible values in a problem with multiple agents. Similarly, $l$
+  and $p$ are variables whose values are the set of possible positions
+  and products respectively. \eu can be used to create partial plans,
+  where the domains of variables $a$ and $l$ can have more than one
+  possible value in them, or grounded plans, where single values will
+  be specified for each variable as we saw in the case of the
+  N-queen's problem.
+
+  In general, to come up with an executable plan it is not sufficient
+  to state predicates that describe required state or behavior without
+  also specifying some temporal extent over which each of those
+  predicates hold. Predicates that are always true can be thought to
+  hold from the beginning to the end of time. However, in practice,
+  the temporal extent of interest must be defined with timepoints to
+  represent its start and end. So we might want to write
+  \texttt{At($a$,$s$,$e$,$l$)} to indicate that the agent $a$ is at
+  location $p$ from time $s$ to time $e$. In fact, this pattern of
+  using such predicates to describe both state and behavior of objects
+  is so prevalent in \eu that we have introduced a special construct
+  called a Token which has the built in variables to indicate the
+  object to which the statement principally applies and the timepoints
+  over which it holds. A Token is an instance of a predicate that
+  represents and object's state or behavior and is defined over a
+  temporal extent. In \eu, all predicate instances are Tokens. Also,
+  in the same way that objects are described generically by classes,
+  in \eu Tokens can be described generically by Token Types. Every
+  token has five built-in variables:
+
+  \begin{enumerate}
+
+  \item \textit {start}: The beginning of the temporal extent over
+    which the predicate is defined.
+
+  \item \textit {end}: The end of the temporal extent over which the
+    predicate is defined.
+
+  \item \textit {duration}: The duration of the temporal extent over
+    which the predicate is defined. The constraint \textit{start} $+$
+    \textit{duration} = \textit{end} is enforced automatically.
+
+  \item \textit{object}: The set of objects to which a token might
+    apply. In a grounded plan each Token applies to a specific Object,
+    reflecting the intuition that we are using Tokens to describe some
+    aspect of an Object (i.e. its state or behavior) in time. However,
+    in a partial plan, the commitment to a specific object may not yet
+    have been made.
+
+  \item \textit{state}: Tokens can be \texttt{ACTIVE, INACTIVE,
+      MERGED}, or \texttt{REJECTED}. The state variable captures the
+    token's current state and its reachable states through further
+    restriction.  This is required to support CAIP's approach to
+    planning where goals and subgoals can be satisfied by either
+    adding new intervals (Tokens) to the plan, or by merging with
+    existing intervals. Token State lifecycle is examined in detailed
+    in a separate section below.
+
+  \end{enumerate}
+
+\item[\textbf{Built-in Object Types}] In a domain model, there may be
+  some classes that don't have any time-dependent state or behavior,
+  and therefore those classes will not have any Token types associated
+  with them, for instance in the Shopping Agent model, the set
+  locations and products are static for the agent's purposes so we may
+  only want to say what they are but not have any state or behavior
+  associated with them.
+
+  However, the most common case in any non-trivial domain model is
+  that Objects will have many associated Tokens in order to describe
+  their state and behavior throughout a plan. There are a couple of
+  Object Types that are so commonly used in domain descriptions, that
+  \eu provides a built-in implementation for them:
+
+\begin{enumerate}
+
+\item \textit{Timelines}: Often objects in a domain must be described
+  by exactly one token for every given timepoint in the plan. Any
+  instances of a class derived from \eu's built-in Timeline class will
+  induce ordering requirements among its tokens to ensure no temporal
+  overlap may occur among them \cite{mus94}.  \comment{add small
+    example?}
+
+\item \textit{Resources}: Metric resources, e.g. the energy of a
+  battery or the capacity of a cargo hold, are objects with an
+  explicit quantitative state in time and with a circumscribed range
+  of changes that can occur to impact that state i.e. produce,
+  consume, use, change.  Resources are such a common requirement for
+  \eu users that built-in object types (with their corresponding token
+  types to denote production, consumption, etc) are provided for
+  them. Instances of classes derived from a Resource will induce
+  ordering requirements on their Tokens in order to ensure that the
+  level of the resource remains within specified limits.  \comment{add
+    small example?}
+\end{enumerate}
+
+\item[\textbf{Token State Model}] In \eu's representation, goals are
+  temporally scoped states that need to be achieved, or actions that
+  must be performed. As a result, \eu internally represents goals as
+  Tokens.  Goals can be posted explicitly, for instance, when a user
+  asks the Shopping Agent to own a specific product by a specific
+  time, or implicitly when subgoals are created as a result of domain
+  compatibilities (described in the CAIP framework above), for
+  instance, when the Shopping Agent must get to a Location where a
+  Product is available before it is able to purchase it, in this case
+  the original goal of purchasing the product will spawn a subgoal of
+  being at a specific location.
+
+  Looking at a how a simple Shopping Agent goal would be addresed by
+  \eu helps illustrate how Token States support the planning process.
+
+  Let's assume we start with a partial plan that places the Shopping
+  agent at home at time 0 and with no Products in its possession, then
+  we post a goal for the Shopping Agent to own a Drill at time 10.
+
+  \comment{Add figure showing an AgentLocation timeline with an
+    At(Home) token with start=0 and end open, an empty AgentBag
+    timeline, and a Own(Drill) token with start=10 floating below both
+    timelines}
+
+Notice that the \texttt{Own(Drill)} token that represents the goal is
+not immediately placed on the \texttt{AgentBag} timeline since \eu
+must determine whether it can achieve that goal and how. When the goal
+is posted and before any decisions are made on it, the
+\texttt{Own(Drill)} token is created in an INACTIVE state, then \eu
+looks at the current partial plan and considers 2 alternatives:
+
+\begin{enumerate}
+
+\item If there are any tokens in the plan that are compatible with the
+  new goal, \eu may attempt to satisfy the goal by merging it with any
+  of them (if there is more than one it can try them one at a time),
+  in this case, the state of the token would change to MERGED.
+
+\item Instead of merging with existing tokens in the plan, \eu may
+  decide to insert the new token into the plan, in that case its state
+  would change to ACTIVE.
+
+\end{enumerate}
+
+When a token is merged, all restrictions imposed on the merged token
+are transferred to the active token upon which it is merged. Merging a
+token requires finding a target active token that is compatible with
+the inactive token. Two tokens are merge-compatible if they are
+instances of the same predicate and that no intersections between
+corresponding variables are empty. The effect of merging is
+illustrated in Figure 3.
+
+\comment{Add figure showing the effect of merging; all figs on this
+  section are conjoined. JB to sketch it out.}
+
+When a token is activated all compatibilities associated with its
+token type are evaluated, which may lead to new subgoals being
+generated. In our example, there should be a compatibility stating
+that the Shopping Agent must be at the same location as the product it
+is purchasing. In that case, activating the \texttt{Own(Drill)} token
+will cause a new \texttt{At(Drill.location)} to be created in an
+INACTIVE state, which in turn will have to be dealt with.
+
+\comment{Add figure showing the effect of activating the token by
+  putting it on the AgentBag timeline, and a new At(Drill.location)
+  token floating below}
+
+If neither activating nor merging works, a token may become REJECTED,
+this is possible only for goals posted by the end user and specified
+as optional. Subgoals generated as a result of token activation cannot
+be rejected, if a subgoal cannot be satisfied through activation or
+merging then it will cause the original activation decision that
+spawned it to be retracted (\eu's search mechanisms are explained in
+detail in a separate section later on).
+
+As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED
+or REJECTED, this lifecycle is illustrated in the figure below.
+
+\comment{Add figure showing possible state transitions. Need a FSM.}
+
+\item[\textbf{Rules}] In order for a plan to be valid, it must comply
+  with all rules and regulations pertinent to the application domain
+  in question. Rules govern the internal and external relationships of
+  a token. For example, consider a parameterized action
+  \texttt{Go(origin,destination)} which describes a Shopping Agent
+  traveling from one location ( \textit{origin}) to another (
+  \textit{destination}). The parameters are instantiated on a token as
+  variables whose domain of values is the set of all locations in a
+  given problem. A rule governing an internal relationship among token
+  variables might stipulate that a transition must involve a change in
+  location. This can be easily expressed as a constraint on the
+  definition of a predicate of the form: \textit{origin !=
+    destination}. It makes sense to further stipulate that the agent
+  must be located at \textit{origin} before travel can happen and the
+  agent must be located at \textit{destination} when completed. This
+  is an example of a rule governing an external relationship among
+  tokens. It specifies a requirement that tokens of the predicate
+  \texttt{At} that represent the location state for the Shopping Agent
+  precede and succeed tokens of the predicate \texttt{Go} that
+  represents traveling.
+
+  \comment{ Add figure showing internal and external relationships
+    from rules on action Go. Need a fig from JB}
+
+Figure xx illustrates the entities and relations involved in
+specifying such a rule on a \texttt{Go} action. The token on which the
+rule applies is referred to (\texttt{Go}) as the master. Each
+\texttt{At} token required by the master is referred to as a
+slave. All variables are indicated by name and their domains are
+expressed as intervals in the case of temporal variables and as
+enumerations for the remainder. Application of a rule on a token can
+thus cause slave tokens, variables, and constraints to be introduced.
+
+The capability of a domain rule to cause a slave token to be created
+is a key vehicle through which planning occurs. Semantically, this
+rule imposes a requirement for supporting (slave) tokens to be in the
+plan in order for the master to be valid.
+
+\comment {talk about condition/effect annotations?}
+
+\end{description}
+
+%\begin{figure*}
+%\centering 
+%\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
+%\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
+%\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
+%\caption{\small An illustrative plan synthesis example with concurrent
+%  timelines. Each timeline is an instantiation of a subsystem tracked
+%  by the planner. Tokens describe the instantiated state of subsystem
+%  at a particular time instance and enforce temporal and parametric
+%  constraints. Abstract goals are decomoposed into successively less
+%  abstract tokens and instantiated in co-temporal timelines using
+%  \texttt{Allen Algebra} relations. All tokens represent flexible
+%  start/end times. \ref{fig:plan-evolve1} shows an initial state
+%  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
+%  \label{fig:Plan-evolve}
+%  \vskip-5pt
+%\end{figure*}
+
+%Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
+%timelines which track \texttt{Goal, Path, Navigation} and
+%\texttt{Command} state over time. Tokens on the goal timeline indicate
+%a survey within a bounded volume and a min/max depth envelope. As the
+%Volume Survey sub-goals on the \texttt{Path} timeline, the initial
+%token on that timeline gets \emph{squeezed} to have dependencies of
+%the goal token instantiated. Initially this dependency are the tokens
+%\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
+%coverage. The first of these in turn generates a sub-goal on the same
+%timeline to go to the next intermediate waypoint, which is illustrated
+%in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
+%figure) ultimately generate all the tokens to completion for this
+%timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
+%in plan generation and shows additional subgoals on the \texttt{Path,
+%  Navigation} and \texttt{Command} timelines. Each token can thus
+%trace its causality when instantiated in the plan. The domain model
+%provides source of these temporal and parametric dependencies in the
+%form of rules.
+
+%\begin{minipage}[c]{\textwidth}
+%\vspace{+0.5cm}
 % \framebox[\textwidth][t]{
-Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
-$\Rightarrow$ \\
-met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
-starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
-\ldots{} \\
-ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
-meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
-\end{minipage}
+%Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
+%$\Rightarrow$ \\
+%met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
+%starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
+%\ldots{} \\
+%ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
+%meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
+%\end{minipage}
+
+% \end{enumerate}
 

### 2012-02-16 trunk/europa-basics.tex
merged minor mods to europa secs

@@
 \includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small A \eu solution to an RCPSP \comment{citation} problem.}
+\caption{\small A \eu solution to an RCPSP \cite{Bruckera99} problem.}
 \label{fig:rcpsp-1}
@@
   constraint network preferably of the shopping example would be
-  useful here.}
+  useful here. JB to generate a napkin fig}
 
@@
 formalism. In particular, it is hard to represent time, resources,
-mutual exclusion and concurrency using propositions \comment{citation
-  needed}. 
-Constraint Programming is a framework that lends itself very well to represent and reason about all of those elements \comment{Ixtet citation?},
-consequently, there has been a good amount of work building automated planners that take advantage of CP techniques.
-The most common approach in constraint-based planning has been to encode the entire planning problem as a CP problem, 
-where action choices and relationships are represented as variables and/or constraints \comment {citations 158,519,539,556 from malik's book}.
-The advantage of using a pure CP formulation is that standard CP solvers can be then used to solve the planning problem, however, this approach also has a couple of significant drawbacks:
+
+% mutual exclusion and concurrency using propositions
+% \cite{ghallab04}. While CSP representations have traditionally had an
+% edge in formulating and representing planning problems,
+% % It is straightforward to represent and reason about all of
+% % those elements using a CSP representation, as a result there has been
+% % some work on doing automated planning while taking advantage of CSP
+% % (TODO: ref). Traditionally, the entire planning problem is translated
+% % into a CSP and then solved using traditional CSP methods, this
+% leading to a formulation where action choices and relationships are
+% represented as variables and/or constraints, there are at least two
+% major drawbacks:
+
+mutual exclusion and concurrency using propositions
+\comment{ghallab04}.  Constraint Programming is a framework that lends
+itself very well to represent and reason about all of those elements
+\comment{Ixtet citation?}, consequently, there has been a good amount
+of work building automated planners that take advantage of CP
+techniques.  The most common approach in constraint-based planning has
+been to encode the entire planning problem as a CP problem, where
+action choices and relationships are represented as variables and/or
+constraints \comment {citations 158,519,539,556 from malik's book}.
+The advantage of using a pure CP formulation is that standard CP
+solvers can be then used to solve the planning problem, however, this
+approach also has a couple of significant drawbacks:
 

### 2012-02-16 arch.tex
"Final draft" on arch.tex:   Addressed Kanna's comments and/or rewrote some sections   The overall flow is here along with the core concepts that needed to be introduced

@@
   timelines is provided in order to be dispatched as goals to
-  whichever reactor that is reponsible for these. \comment{By
-    extending I may have saud things I developed below ... will have
-    to check}
+  whichever reactor that is reponsible for these. % \comment{By
+    % extending I may have saud things I developed below ... will have
+    % to check}
 \end{itemize}
@@
 multiple steps spanning multiple ticks for a reactor to produce a plan
-that is both complete and valid. When such plan is identified
-\comment{how? Is there a terminating condition? No more flaws but I
-  washoping it would have been defined in \eu section} its {\em external}
+that is both complete and valid. How the reactor identifies that its
+plan is complete and valid is its responsibility and depend on the way
+such reactor resolve its plan during each {\em steps}. The general
+idea is that the reactor was able to find a valid strategy tin order
+to satisfy as much of its {\em internal} goals as possible. In the
+context of \eu, which we are going to develop, this means that the
+solver do not identify any more flaw to be resolved in the plan.
+
+ When such plan is identified, its {\em external}
 timeline(s) are then dispatched to the corresponding reactors which
@@
 steps, its starting point is the last partial plan resulting from
-deliberation. This gives an initial frame to the problemm that helps
--- at least initial -- to target the synchronization problem by
+deliberation. This gives an initial frame to the problem that helps
+-- at least initially -- to target the synchronization problem by
 ensuring that the new {\em external} state updates are compatible with
@@
 timelines based on this plan. If they are not compatibles, then
-synchronization identifies a plan inconsistency  reflecting that the
-current plan is not valid anymore. \comment{fpy: I am here and will
-  develop further}
-
-\comment{following is not what I intended to express whatsoever
-  ... synchroniuzation od not deal with dispatching but execution
-  tracking ... it is though helped by the structure of the exisiting
-  plan and often identifies execution failures}
- Conversely, when the deliberation process eventually finds a
-solution, synchronization is informed about generating the planned
-state values \comment{should this read as ``...is informed about
-  generated state values..'' instead?} on the {\em external} state
-variables. This {\em external} plan defines the set of {\em goals} for
-this state variable managed by another {\em reactor}.
+{\em synchronization} identifies a plan inconsistency  reflecting that the
+current plan is not valid anymore. This results on the retraction of
+decisions made during {\em deliberation} steps in order to allow 
+{\em synchronization} to identify its {\em internal}  state. Such
+relaxation can be viewed as a backtracking on the plan search
+triggered exogenously by the world evolution. It is the way our
+reactor will identify that its current plan failed to execute and can
+trigger further deliberations steps in order to find an alternate 
+strategy.
+
+Finally, when the deliberation process did identify a complete
+solution plan for its current {\em internal} goals. The part of its
+plan that applies to its {\em external} timelines is sent as a 
+request to \rx that will handle their dispatching to the corresponding 
+reactors. If things go smoothly the {\em synchronization} process will
+eventually receive {\em external } observations that confirm the
+execution of this plan and will allow the plan execution to progress
+inside the reactor (by having the part of the plan observed restricted
+to the start time of these observations).
 
@@
 they require that their robot comes to a full stop until (re-)planning
-is complete. \comment{I may need more examples/refs here}. As Fig.
-\ref{fig:tick-exec} shows, this assumption is problematic for our
+is complete.% \comment{I may need more examples/refs here}. 
+As Fig. \ref{fig:tick-exec} shows, this assumption is problematic for our
 architecture. Synchronization that tracks state evolution can occur
@@
 synchronization and deliberation share the same plan we provide a
-solution that relaxes this static world assumption. \comment{Need to
-  talk about IDEA and the fact that they had similar aspect even
-  though they were never very clear nor well articulated in their
-  papers}
+solution that relaxes this static world assumption. According to 
+\cite{Dias:2003ua, mus06}, IDEA was sharing the same core idea 
+where all the solvers of an agent shared the same plan as a basis of
+communication and specifically their {\em Reactive planner} --roughly
+similar to our {\em synchronization} solver -- was often the only one 
+targeting the execution frontier. Although, none of the papers we
+found were detailing how this was handled in the
+architecture. Moreover, in \cite{Dias:2003ua} we can read:
+\begin{quotation}
+  [\textellipsis] as long as the planning horizon over which the
+  deliberative planner is working never intersects the current
+  execution time, deliberative planning can operate in parallel with
+  reactive execution and does not affect the reactivity of the agent. 
+\end{quotation}
+\comment{fpy: This is my interpretation of this sentence at least .... feel
+  free to comment in case you disagree}
+Which suggest a non overlap between {\em reactive planner} and {\em
+  deliberative planners} horizons. In \rx, and more specifically in
+our \eu reactor the planning horizon of {\em synchronization} -- 
+limited to the current tick -- is a subset of the planning horizon 
+of the deliberation steps -- starting at the current tick and with an
+extend based on the latency $\lambda$ and look-ahead $\pi$ of 
+the reactor. The way we avoid to create conflict between the two is by
+a careful control of the scheduling of these two process by the \rx 
+agent.
 
@@
 of an objective will remove the corresponding flaws\footnote{although
-  if the flaws of this gaol were previously resolved, the removal of
+  if the flaws of this goal were previously resolved, the removal of
   this goal may create new flaws}. The impact of {\em synchronization}
@@
   requiring the {\em synchronization} solver to relax the existing
-  plan in order to identify its current {\em external} state.
+  plan in order to identify its current {\em external} state. (This
+  case is developed further on the next section)
 \end{enumerate}

### 2012-02-16 trunk/europa-basics.tex
citations added to europa text

@@
 
-{\footnotesize
+{\scriptsize
   \begin{quote}
@@
 constitute solutions. For that, search algorithms like global
-backtracking \comment{citation needed} or local search are used; these
-use consistency as an efficient way to prune the search space
+backtracking \cite{hooker05} or local search are used; these use
+consistency as an efficient way to prune the search space
 \cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
@@
 formalism. In particular, it is hard to represent time, resources,
-
 % mutual exclusion and concurrency using propositions
@@
 % major drawbacks:
-
-mutual exclusion and concurrency using propositions
-\comment{ghallab04}.  Constraint Programming is a framework that lends
-itself very well to represent and reason about all of those elements
-\comment{Ixtet citation?}, consequently, there has been a good amount
-of work building automated planners that take advantage of CP
-techniques.  The most common approach in constraint-based planning has
-been to encode the entire planning problem as a CP problem, where
-action choices and relationships are represented as variables and/or
-constraints \comment {citations 158,519,539,556 from malik's book}.
-The advantage of using a pure CP formulation is that standard CP
-solvers can be then used to solve the planning problem, however, this
-approach also has a couple of significant drawbacks:
+mutual exclusion and concurrency using propositions \cite{ghallab04}.
+Constraint Programming is a framework that lends itself very well to
+represent and reason about all of those elements \cite{ghallab94},
+consequently, there has been a good amount of work building automated
+planners that take advantage of CP techniques.  The most common
+approach in constraint-based planning has been to encode the entire
+planning problem as a CP problem, where action choices and
+relationships are represented as variables and/or constraints
+\cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
+pure CP formulation is that standard CP solvers can be then used to
+solve the planning problem, however, this approach also has a couple
+of significant drawbacks:
 
@@
 
-These drawbacks have been addressed by a small number of planning frameworks like Descartes \comment{citation 289,290 from malik's book}, IxTet \comment{citation  223 from Malik's book}, and CAIP (Constraint-Based Interval and Attribute Planning)\cite{mus94,frank2003}.
-These frameworks take advantage of a CSP representation to deal with time and resources, at the same time maintaining an explicit representation of the elements relevant to planning (action choices, causal relationships, etc).
-This combination results in a planning approach that can deal with the type of constraints found in real-life problems, while allowing for intuitive problem descriptions and a natural way to express search algorithms and heuristics.
-\eu is an implementation of the CAIP framework.
-In CAIP, planning problems are specified in terms of:
+These drawbacks have been addressed by a small number of planning
+frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
+\texttt{IxTet} \cite{ghallab87,lemai04}, and CAIP (Constraint-Based
+Interval and Attribute Planning)\cite{mus94,frank2003}.  These
+frameworks take advantage of a CSP representation to deal with time
+and resources, at the same time maintaining an explicit representation
+of the elements relevant to planning (action choices, causal
+relationships, etc). This combination results in a planning approach
+that can deal with the type of constraints found in real-life
+problems, while allowing for intuitive problem descriptions and a
+natural way to express search algorithms and heuristics. \eu is an
+implementation of the CAIP framework. In CAIP, planning problems are
+specified in terms of:
 
 \begin{enumerate}
-	\item \textbf{Intervals}: entities that have a temporal scope (that is, a beginning and an  end in time) and a set of attributes.
-Intervals are used to represent actions and states in the planning domain.
-	\item \textbf{Variables and Constraints}: Variables represent Interval attributes and Constraints enable a direct representation of temporal, resource and any other the restrictions that a valid plan must comply with.
-	\item \textbf{Domain Configuration Rules}: this is an explicit representation of the conceptual relationships between actions and states in the problem domain, for instance, for a Shopping Agent to be able to purchase a product, it must be at a location where the product is available.
+
+\item \textbf{Intervals}: entities that have a temporal scope (that
+  is, a beginning and an end in time) and a set of
+  attributes. Intervals are used to represent actions and states in
+  the planning domain.
+
+\item \textbf{Variables and Constraints}: Variables represent Interval
+  attributes and Constraints enable a direct representation of
+  temporal, resource and any other the restrictions that a valid plan
+  must comply with.
+
+\item \textbf{Domain Configuration Rules}: this is an explicit
+  representation of the conceptual relationships between actions and
+  states in the problem domain, for instance, for a Shopping Agent to
+  be able to purchase a product, it must be at a location where the
+  product is available.
 \end{enumerate}
 
-To generate plans, the CAIP framework uses a plan-space search approach \comment{cite malik's chapter 5}, where a partial plan is evolved until all open goals are achieved, and all inconsistencies are removed, the specific search algorithm used by \eu will be covered in detail below.
+To generate plans, the CAIP framework uses a plan-space search
+approach \cite{ghallab04}, where a partial plan is evolved until all
+open goals are achieved, and all inconsistencies are removed, the
+specific search algorithm used by \eu will be covered in detail below.

### 2012-02-16 europa.tex
citations added to europa text

@@
 
-The Constraint Reasoning Engine (CRE) manages variables, the domains
-from which they can take values, and constraints that define
-relationships among variables. It also provides an efficient arc
+The \textbf{Constraint Reasoning Engine} (CRE) manages domain types, variables and constraints that define
+relationships among them. It also provides an efficient arc
 consistency mechanism \cite{mackworth77}. The CRE is designed so that
 specialized reasoning algorithms for specific constraints can be easily
-and efficiently plugged in. The Plan Database (PDB) manages object types
-and objects, which are a mechanism to group variables to more naturally
-model the real world in much the same way that object oriented
-programming does. Also manages tokens, which are a mechanism to group
-variables to represent temporally scoped state. Objects and tokens can
-be used to model planning domains in a much more natural and extensible
-way that a pure CSP approach can. The Rules Engine (RE) manages token
-relationships in a planning model. The language modules (NDDL, ANML, and
-other higher level modeling languages) are used to create domain models
-and problem instances \comment{point advantages over pure programming
-  API approaches, provide snippets}. Finally the API Layer (Fig.
-\ref{fig:europa-api}) makes all the services available so that \eu can
-be used to build applications.
+and efficiently plugged in.
 
-\comment{Is Fig. \ref{fig:europa-api} necessary?}
+The \textbf{Plan Database} (PDB) manages object and token types.
+It also maintains the state of partial plans in terms of object, variable and token instances.
+This constitutes the data foundation on which end user applications and automated planners are built.
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
-\caption{\small EUROPA's Client API}
-\label{fig:europa-api}
-\vskip+0.1cm
-\end{figure}
+The \textbf{Rules Engine} (RE) modularly adds the capability to support domain rules that describe dependencies between tokens,
 
-	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
-	% 	\begin{enumerate}
-	% 		\item Temporal Network
-	% 		\item Resources
-	% 		\item NDDL
-	% 		\item ANML (TODO:maybe it's better to leave out?)
-	% 	\end{enumerate}
+The \textbf{Solver Module} provides the framework for flaw detection and resolution that is needed to implement search algorithms for planning.
+
+The \textbf {Language Modules}: Describing problem domains and problem instances in terms of objects, variables and domain rules in a programmatic way is cumbersome to the extent of being impractical, that motivated the creation of modeling languages that support those definitions in a much more natural and compact way.
+NDDL is the main modeling language for \eu and will be described in detail in the Modeling section below.
+ANML is a language under development that provides a more intuitive syntax \comment{provide ANML citation}.
+
+The \textbf{API Layer}  provides access to services from all the different modules in a programmatic way so that \eu can be easily embedded into specialized applications.
+
+In an effort to mimimize the amount of work that a user needs to perform to solve specific problems, \eu  also provides extension modules that have been found to be useful in many domains: 
+\begin{enumerate}
+	\item \textit{Temporal Network}: Extensions to the CRE module to reason about temporal constraints 
+	\item \textit{Resources}: Extensions to the CRE, PDB and Solver modules to provide out-of-the-box reasoning and search for metric resources.
+	\item \textit{Chronological Backtracking Solver}: A general purpose built-in solver to provide the ability to plan out-of-the-box, problems that require challenging search may require specialized solvers to be built as explained in the Search section below.
+\end{enumerate}
 
@@
 }
+\end{verbatim}
+
+In the model above we can see how the main elements or a problem domain can be mapped to \eu's representation:
+
+\begin{enumerate}
+	\item \textbf{Object Types}: NDDL supports an object-oriented approach, where common data and behavior can be abstracted as a class.
+In this case, we have the \textit{Product} class represent the products that the Shopping Agent needs, \textit{Location} represents the places where products can be found, \textit{ProductLocation} represents the association arising when a product can be purchased at a location.
+The Shopping Agent itself is represented by an \textit{Agent} class, which relies on the \textit{AgentLocation} class to keep track of where the agent is at any point in time.
+NDDL supports a single rooted class hierarchy so that state and behavior can be extended through inheritance.
+We took advantage of that by making \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline} class which doesn't allow predicates to overlap in time for the same object instance, in this way we can ensure that and agent can only be at one place at a time.
+	\item \textbf{Variable Types and Instances} : All object attributes are represented as variables in this model, for instance, it uses integer variables to represent the built-in temporal parameters for each token (activity duration, start and end times), string variables to represent  \textit{Location} and  \textit{Product} names, and object reference variables to represent the structural relationships between different objects (for instance, each Agent has variable to keep track of its  \textit{AgentLocation}).
+	\item \textbf{Predicate Types}: Some classes like \textit{Product}, \textit{Location} and \textit{ProductLocation} are completely static and so don't have any predicate or action types associated with them.
+The \textit{AgentLocation} class is purely used to keep track of state and so it has \texttt{At} and \texttt{Going} predicate types associated with it.
+The \textit{Agent} class keeps track of the products it purchases through the \texttt{Own} predicate, having a separate \textit{AgentBag} class with its own predicates would have been an equally valid modeling choice.
+	\item \textbf{Action Types}: The \textit{Agent} class is the only one that has behavior associated with it, consequently, it defines action types to represent when the agent is taking action to \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a \textit{Product}.
+We'll see how the actual behavior for those actions is specified below.
+\end{enumerate}
+
+\begin{verbatim}
 
@@
 
-In this model we can see:
+Domain rules can be defined in NDDL in the context of a Predicate or an Action. In the example above we can see the main elements of a rule definition:
 
-\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
 \begin{enumerate}
-	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
-	\item Object Types:
-	\item Predicate Types:
-	\item Action Types:
+  \item \textbf{Variables}: A Token Type (predicate or action) is part of a class, therefore class attributes (see references to \texttt{object.location} above), Token Type parameters (see references to \texttt{[from, to]}  in \texttt{Agent::Go}, or to \texttt{[duration, product]} in \texttt{Agent::Buy}  and locally declared variables (see \texttt{possibleStores} in \texttt{Agent::Buy}) can all be used in the definition of a Rule.
+  \item \textbf{Constraints}: restrictions on the values that variables can take inside a Rule are expressed through constraints, in the example above only the \texttt{eq} constraint is used, but in general, any constraint available in \eu's library can be used.
+ \item \textbf{Subgoals}: In \texttt{Agent::Go}, subgoals are used to specify that the Shopping Agent's location must transition from origin to destination in a temporally consistent way.
+The \texttt{[meets, equals, metby]} operators are used to specify subgoals (Allen operators are explained in more detail below), along with the temporal relationship each token representing a subgoal has with the the token representing the \texttt{Agent::Go} action.
+Also, \texttt{[condition, effect]} annotations can be used when defining a subgoal inside an action, \eu doesn't immediately associate any semantics with them, but they can be extremely useful for search modules when creating plans.
 \end{enumerate}
 
- shows how a specific problem instance would be
-specified in NDDL.
+Now, having that we have a specification for the Shopping Agent domain, here is how a particular problem instance could be specified:
 
@@
 \label{eu:model}
+%\caption{Specifying a problem instance for the Shopping Agent in NDDL}
 \end{verbatim}
 
-\comment{explain variable, object, predicate and constraint instances.
-explain goals and facts.}
-
+A problem instance consists mainly of 
 \begin{enumerate}
-	\item Variable Instances:
-	\item Object Instances:
-	\item Facts:
-	\item Goals:
-	\item Constraint Instances:
+  \item \textbf{Variables} and \textbf{Objects}: In this case we defined temporal variables that are used to specify the planning horizon (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are used to specify the locations, products, product-loction associations and the Shopping Agent that we want to plan for.
+  \item \textbf{Facts}: These are used to specify the current state of the world and don't need any support.
+Facts are represented as tokens in \eu's Plan Database.
+In this case we use facts to specify that the Shipping Agent is initially at home.
+  \item \textbf{Goals}:  Goals are represented as tokens in \eu's Plan Database, differently from Facts, Goals trigger inference in \eu which determines whether flaws (unresolved subgoals, overloaded resources, etc., flaws will be covered in detail in the search section below) need to be resolved, if that is the case, search modules can be invoked which will look for a plan where all the flaws are resolved.
+  \item \textbf{Constraints}: restrictions on and between actions and goals are specified through constraints, in this case we use temporal constraints to specify when the facts hold and when the goals should be accomplished.
+As always, all the constraints in \eu's library (built-in, or user defined), can be used to specify these restrictions.
 \end{enumerate}
@@
 
-\comment{ add figure}
+\begin{figure}[!htb]
+  \centering
+  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
+  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
+    \caption{\small An illustration of a STN and its corresponding
+      distance graph.}
+  \label{fig:stn}
+\end{figure}
+
+\comment{ add figure. Use Europa wiki figure; or use conor figure from slides}
 

### 2012-02-16 trunk/europa.tex
merged minor mods to europa secs
citations added to europa text

@@
 
-\eu is implemented as a library that encapsulates all the elements described above and makes them easily accessible through modeling and API layers.
-A user will typically use the modeling facilities to inject a problem domain description, and the API to extract and manipulate plans, and to obtain information about plan flaws and constraint violations at any stage of the planning process.
+\eu is implemented as a library that encapsulates all the elements
+described above and makes them easily accessible through modeling and
+API layers.  A user will typically use the modeling facilities to
+inject a problem domain description, and the API to extract and
+manipulate plans, and to obtain information about plan flaws and
+constraint violations at any stage of the planning process.
 
@@
 
-The \textbf{Constraint Reasoning Engine} (CRE) manages domain types, variables and constraints that define
-relationships among them. It also provides an efficient arc
-consistency mechanism \cite{mackworth77}. The CRE is designed so that
-specialized reasoning algorithms for specific constraints can be easily
-and efficiently plugged in.
+The \textbf{Constraint Reasoning Engine} (CRE) manages domain types,
+variables and constraints that define relationships among them. It
+also provides an efficient arc consistency mechanism
+\cite{mackworth77}. The CRE is designed so that specialized reasoning
+algorithms for specific constraints can be easily and efficiently
+plugged in.
 
-The \textbf{Plan Database} (PDB) manages object and token types.
-It also maintains the state of partial plans in terms of object, variable and token instances.
-This constitutes the data foundation on which end user applications and automated planners are built.
+The \textbf{Plan Database} (PDB) manages object and token types. It
+also maintains the state of partial plans in terms of object, variable
+and token instances. This constitutes the data foundation on which end
+user applications and automated planners are built.
 
-The \textbf{Rules Engine} (RE) modularly adds the capability to support domain rules that describe dependencies between tokens,
+The \textbf{Rules Engine} (RE) modularly adds the capability to
+support domain rules that describe dependencies between tokens,
 
-The \textbf{Solver Module} provides the framework for flaw detection and resolution that is needed to implement search algorithms for planning.
+The \textbf{Solver Module} provides the framework for flaw detection
+and resolution that is needed to implement search algorithms for
+planning.
 
-The \textbf {Language Modules}: Describing problem domains and problem instances in terms of objects, variables and domain rules in a programmatic way is cumbersome to the extent of being impractical, that motivated the creation of modeling languages that support those definitions in a much more natural and compact way.
-NDDL is the main modeling language for \eu and will be described in detail in the Modeling section below.
-ANML is a language under development that provides a more intuitive syntax \comment{provide ANML citation}.
+The \textbf {Language Modules}: Describing problem domains and problem
+instances in terms of objects, variables and domain rules in a
+programmatic way is cumbersome to the extent of being impractical,
+that motivated the creation of modeling languages that support those
+definitions in a much more natural and compact way. NDDL is the main
+modeling language for \eu and will be described in detail in the
+Modeling section below. ANML is a language under development that
+provides a more intuitive syntax \cite{smith08}.
 
-The \textbf{API Layer}  provides access to services from all the different modules in a programmatic way so that \eu can be easily embedded into specialized applications.
+The \textbf{API Layer} provides access to services from all the
+different modules in a programmatic way so that \eu can be easily
+embedded into specialized applications.
 
-In an effort to mimimize the amount of work that a user needs to perform to solve specific problems, \eu  also provides extension modules that have been found to be useful in many domains: 
+In an effort to mimimize the amount of work that a user needs to
+perform to solve specific problems, \eu also provides extension
+modules that have been found to be useful in many domains:
+
 \begin{enumerate}
-	\item \textit{Temporal Network}: Extensions to the CRE module to reason about temporal constraints 
-	\item \textit{Resources}: Extensions to the CRE, PDB and Solver modules to provide out-of-the-box reasoning and search for metric resources.
-	\item \textit{Chronological Backtracking Solver}: A general purpose built-in solver to provide the ability to plan out-of-the-box, problems that require challenging search may require specialized solvers to be built as explained in the Search section below.
+
+\item \textit{Temporal Network}: Extensions to the CRE module to
+  reason about temporal constraints
+
+\item \textit{Resources}: Extensions to the CRE, PDB and Solver
+  modules to provide out-of-the-box reasoning and search for metric
+  resources.
+
+\item \textit{Chronological Backtracking Solver}: A general purpose
+  built-in solver to provide the ability to plan out-of-the-box,
+  problems that require challenging search may require specialized
+  solvers to be built as explained in the Search section below.
 \end{enumerate}
@@
 language for models, goals and plans is of great importance. For \eu,
-NDDL (New Domain Description Language [pronounced ``noodle'']
-allows the user to specify the different components of a
-problem domain in a concise way. The main characteristics that make NDDL
-a powerful tool for describing problem domains are:
+NDDL (New Domain Description Language [pronounced ``noodle''] allows
+the user to specify the different components of a problem domain in a
+concise way. The main characteristics that make NDDL a powerful tool
+for describing problem domains are:
 
@@
 
-\item Object Oriented: NDDL supports classes and
-  inheritance in similar fashion to popular Object-Oriented languages
-  like Java and C++. It also supports polymorphism for some of the
-  planning components. Using object classes and instances is a
-  time-tested approach to naturally describe problem domains \comment{OO
+\item Object Oriented: NDDL supports classes and inheritance in
+  similar fashion to popular Object-Oriented languages like Java and
+  C++. It also supports polymorphism for some of the planning
+  components. Using object classes and instances is a time-tested
+  approach to naturally describe problem domains \comment{OO
     citation}.
 
-\item It offers declarative constructs to define constraints
-  and causality dependencies between actions and effects.
+\item It offers declarative constructs to define constraints and
+  causality dependencies between actions and effects.
 
@@
 
-To understand how the most important elements of NDDL work, let's take a
-look at how the Shopping Agent problem could be modeled:
+To understand how the most important elements of NDDL work, we show
+how the Shopping Agent problem could be modeled:
 
@@
 
-In the model above we can see how the main elements or a problem domain can be mapped to \eu's representation:
+In the model above we can see how the main elements or a problem
+domain can be mapped to \eu's representation:
 
 \begin{enumerate}
-	\item \textbf{Object Types}: NDDL supports an object-oriented approach, where common data and behavior can be abstracted as a class.
-In this case, we have the \textit{Product} class represent the products that the Shopping Agent needs, \textit{Location} represents the places where products can be found, \textit{ProductLocation} represents the association arising when a product can be purchased at a location.
-The Shopping Agent itself is represented by an \textit{Agent} class, which relies on the \textit{AgentLocation} class to keep track of where the agent is at any point in time.
-NDDL supports a single rooted class hierarchy so that state and behavior can be extended through inheritance.
-We took advantage of that by making \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline} class which doesn't allow predicates to overlap in time for the same object instance, in this way we can ensure that and agent can only be at one place at a time.
-	\item \textbf{Variable Types and Instances} : All object attributes are represented as variables in this model, for instance, it uses integer variables to represent the built-in temporal parameters for each token (activity duration, start and end times), string variables to represent  \textit{Location} and  \textit{Product} names, and object reference variables to represent the structural relationships between different objects (for instance, each Agent has variable to keep track of its  \textit{AgentLocation}).
-	\item \textbf{Predicate Types}: Some classes like \textit{Product}, \textit{Location} and \textit{ProductLocation} are completely static and so don't have any predicate or action types associated with them.
-The \textit{AgentLocation} class is purely used to keep track of state and so it has \texttt{At} and \texttt{Going} predicate types associated with it.
-The \textit{Agent} class keeps track of the products it purchases through the \texttt{Own} predicate, having a separate \textit{AgentBag} class with its own predicates would have been an equally valid modeling choice.
-	\item \textbf{Action Types}: The \textit{Agent} class is the only one that has behavior associated with it, consequently, it defines action types to represent when the agent is taking action to \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a \textit{Product}.
-We'll see how the actual behavior for those actions is specified below.
+
+\item \textbf{Object Types}: NDDL supports an object-oriented
+  approach, where common data and behavior can be abstracted as a
+  class. In this case, we have the \textit{Product} class represent
+  the products that the Shopping Agent needs, \textit{Location}
+  represents the places where products can be found,
+  \textit{ProductLocation} represents the association arising when a
+  product can be purchased at a location. The Shopping Agent itself is
+  represented by an \textit{Agent} class, which relies on the
+  \textit{AgentLocation} class to keep track of where the agent is at
+  any point in time. NDDL supports a single rooted class hierarchy so
+  that state and behavior can be extended through inheritance. We took
+  advantage of that by making \textit{AgentLocation} inherit from
+  \eu's built-in \textit{Timeline} class which doesn't allow
+  predicates to overlap in time for the same object instance, in this
+  way we can ensure that and agent can only be at one place at a time.
+
+\item \textbf{Variable Types and Instances} : All object attributes
+  are represented as variables in this model, for instance, it uses
+  integer variables to represent the built-in temporal parameters for
+  each token (activity duration, start and end times), string
+  variables to represent \textit{Location} and \textit{Product} names,
+  and object reference variables to represent the structural
+  relationships between different objects (for instance, each Agent
+  has variable to keep track of its \textit{AgentLocation}).
+
+\item \textbf{Predicate Types}: Some classes like \textit{Product},
+  \textit{Location} and \textit{ProductLocation} are completely static
+  and so don't have any predicate or action types associated with
+  them. The \textit{AgentLocation} class is purely used to keep track
+  of state and so it has \texttt{At} and \texttt{Going} predicate
+  types associated with it. The \textit{Agent} class keeps track of
+  the products it purchases through the \texttt{Own} predicate, having
+  a separate \textit{AgentBag} class with its own predicates would
+  have been an equally valid modeling choice.
+
+\item \textbf{Action Types}: The \textit{Agent} class is the only one
+  that has behavior associated with it, consequently, it defines
+  action types to represent when the agent is taking action to
+  \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
+  \textit{Product}. We'll see how the actual behavior for those
+  actions is specified below.
 \end{enumerate}
@@
 
-Domain rules can be defined in NDDL in the context of a Predicate or an Action. In the example above we can see the main elements of a rule definition:
+Domain rules can be defined in NDDL in the context of a Predicate or
+an Action. In the example above we can see the main elements of a rule
+definition:
 
 \begin{enumerate}
-  \item \textbf{Variables}: A Token Type (predicate or action) is part of a class, therefore class attributes (see references to \texttt{object.location} above), Token Type parameters (see references to \texttt{[from, to]}  in \texttt{Agent::Go}, or to \texttt{[duration, product]} in \texttt{Agent::Buy}  and locally declared variables (see \texttt{possibleStores} in \texttt{Agent::Buy}) can all be used in the definition of a Rule.
-  \item \textbf{Constraints}: restrictions on the values that variables can take inside a Rule are expressed through constraints, in the example above only the \texttt{eq} constraint is used, but in general, any constraint available in \eu's library can be used.
- \item \textbf{Subgoals}: In \texttt{Agent::Go}, subgoals are used to specify that the Shopping Agent's location must transition from origin to destination in a temporally consistent way.
-The \texttt{[meets, equals, metby]} operators are used to specify subgoals (Allen operators are explained in more detail below), along with the temporal relationship each token representing a subgoal has with the the token representing the \texttt{Agent::Go} action.
-Also, \texttt{[condition, effect]} annotations can be used when defining a subgoal inside an action, \eu doesn't immediately associate any semantics with them, but they can be extremely useful for search modules when creating plans.
+\item \textbf{Variables}: A Token Type (predicate or action) is part
+  of a class, therefore class attributes (see references to
+  \texttt{object.location} above), Token Type parameters (see
+  references to \texttt{[from, to]} in \texttt{Agent::Go}, or to
+  \texttt{[duration, product]} in \texttt{Agent::Buy} and locally
+  declared variables (see \texttt{possibleStores} in
+  \texttt{Agent::Buy}) can all be used in the definition of a Rule.
+
+\item \textbf{Constraints}: restrictions on the values that variables
+  can take inside a Rule are expressed through constraints, in the
+  example above only the \texttt{eq} constraint is used, but in
+  general, any constraint available in \eu's library can be used.
+
+  \item \textbf{Subgoals}: In \texttt{Agent::Go}, subgoals are used to
+    specify that the Shopping Agent's location must transition from
+    origin to destination in a temporally consistent way. The
+    \texttt{[meets, equals, metby]} operators are used to specify
+    subgoals (Allen operators are explained in more detail below),
+    along with the temporal relationship each token representing a
+    subgoal has with the the token representing the \texttt{Agent::Go}
+    action. Also, \texttt{[condition, effect]} annotations can be used
+    when defining a subgoal inside an action, \eu doesn't immediately
+    associate any semantics with them, but they can be extremely
+    useful for search modules when creating plans.
 \end{enumerate}
 
-Now, having that we have a specification for the Shopping Agent domain, here is how a particular problem instance could be specified:
+Now, having that we have a specification for the Shopping Agent
+domain, here is how a particular problem instance could be specified:
 
@@
 A problem instance consists mainly of
+
 \begin{enumerate}
-  \item \textbf{Variables} and \textbf{Objects}: In this case we defined temporal variables that are used to specify the planning horizon (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are used to specify the locations, products, product-loction associations and the Shopping Agent that we want to plan for.
-  \item \textbf{Facts}: These are used to specify the current state of the world and don't need any support.
-Facts are represented as tokens in \eu's Plan Database.
-In this case we use facts to specify that the Shipping Agent is initially at home.
-  \item \textbf{Goals}:  Goals are represented as tokens in \eu's Plan Database, differently from Facts, Goals trigger inference in \eu which determines whether flaws (unresolved subgoals, overloaded resources, etc., flaws will be covered in detail in the search section below) need to be resolved, if that is the case, search modules can be invoked which will look for a plan where all the flaws are resolved.
-  \item \textbf{Constraints}: restrictions on and between actions and goals are specified through constraints, in this case we use temporal constraints to specify when the facts hold and when the goals should be accomplished.
-As always, all the constraints in \eu's library (built-in, or user defined), can be used to specify these restrictions.
+
+\item \textbf{Variables} and \textbf{Objects}: In this case we defined
+  temporal variables that are used to specify the planning horizon
+  (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are
+  used to specify the locations, products, product-location
+  associations and the Shopping Agent that we want to plan for.
+
+\item \textbf{Facts}: These are used to specify the current state of
+  the world and don't need any support. Facts are represented as
+  tokens in \eu's Plan Database. In this case we use facts to specify
+  that the Shipping Agent is initially at home.
+
+\item \textbf{Goals}: Goals are represented as tokens in \eu's Plan
+  Database, differently from Facts, Goals trigger inference in \eu
+  which determines whether flaws (unresolved subgoals, overloaded
+  resources, etc., flaws will be covered in detail in the search
+  section below) need to be resolved, if that is the case, search
+  modules can be invoked which will look for a plan where all the
+  flaws are resolved.
+
+\item \textbf{Constraints}: restrictions on and between actions and
+  goals are specified through constraints, in this case we use
+  temporal constraints to specify when the facts hold and when the
+  goals should be accomplished. As always, all the constraints in
+  \eu's library (built-in, or user defined), can be used to specify
+  these restrictions.
 \end{enumerate}
@@
 The Temporal Network: The notion of time is central to temporal
-planning. \eu uses variables to explicitly represent timepoints for plan
-activities and states. Constraints among timepoints provide a natural
-way to express domain axioms. For example, in order to state that
-activity A must occur before activity B we can say that the end
+planning. \eu uses variables to explicitly represent timepoints for
+plan activities and states. Constraints among timepoints provide a
+natural way to express domain axioms. For example, in order to state
+that activity A must occur before activity B we can say that the end
 timepoint of A is $leq$ the start timepoint of B.
@@
 \cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network (STN). Such a network
-can be transformed into a Distance Graph (DG) where the outward arc from
-a node represents the maximum distance from the source node to the
-target node. They also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative cycle. A
-negative cycle is a path from a node to itself that has a path length
-less than 0. If such a cycle exists, the network is inconsistent. It was
-further shown that a single-source shortest path algorithm was
-sufficient to detect a negative cycle and provide sufficient propagation
-to yield a backtrack-free search. Thus we have an efficient and complete
-algorithm for propagating an STN. These results build on the already
-established notion of a CSP and are naturally incorporated into the
-general representation and propagation scheme used in \eu.
-The diagram below illustrates a simple STN with just 2
-variables and a single constraint. It also shows the resulting DG.
+grouped together to form a Simple Temporal Network (STN). Such a
+network can be transformed into a Distance Graph (DG) where the
+outward arc from a node represents the maximum distance from the
+source node to the target node. They also showed that shortest path
+algorithms could be used to propagate values in the network and
+discover a negative cycle. A negative cycle is a path from a node to
+itself that has a path length less than 0. If such a cycle exists, the
+network is inconsistent. It was further shown that a single-source
+shortest path algorithm was sufficient to detect a negative cycle and
+provide sufficient propagation to yield a backtrack-free search. Thus
+we have an efficient and complete algorithm for propagating an
+STN. These results build on the already established notion of a CSP
+and are naturally incorporated into the general representation and
+propagation scheme used in \eu.  The diagram below illustrates a
+simple STN with just 2 variables and a single constraint. It also
+shows the resulting DG.
 
-\comment{ add figure}
+\begin{figure}[!htb]
+  \centering
+  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
+  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
+    \caption{\small An illustration of a STN and its corresponding
+      distance graph.}
+  \label{fig:stn}
+\end{figure}
+
+% \comment{ add figure. Use Europa wiki figure; or use conor figure from slides}
 
@@
 \begin{enumerate}
-  \item \textbf{Reservoir} resources can have both 'consume' and 'produce' tokens. Note
-that these tokens do NOT have the usual start/end/duration token
-variables, since occur at a single instant in time. Instead, they have a
-single variable, 'time' used to represent that instant.
 
-  \item \textbf{Reusable} resources have 'uses' tokens that use a quantity for the
-duration of the token. These tokens do have the usual start/end/duration
-variables.
+\item \textbf{Reservoir} resources can have both 'consume' and
+  'produce' tokens. Note that these tokens do NOT have the usual
+  start/end/duration token variables, since occur at a single instant
+  in time. Instead, they have a single variable, 'time' used to
+  represent that instant.
 
-  \item \textbf {Unary} resources are reusable resources with unit quantity and have 'use'
-tokens (again, with the usual start/end/duration variables).
+\item \textbf{Reusable} resources have 'uses' tokens that use a
+  quantity for the duration of the token. These tokens do have the
+  usual start/end/duration variables.
+
+\item \textbf {Unary} resources are reusable resources with unit
+  quantity and have 'use' tokens (again, with the usual
+  start/end/duration variables).
 \end{enumerate}
@@
 
-\item A domain model that describes the variable, object, predicate and
-  action types that are relevant for the problem.
+\item A domain model that describes the variable, object, predicate
+  and action types that are relevant for the problem.
 
 \item A problem instance (also called initial state) that consists of:
-	\item Variable and Object instances that exist for the entire planning horizon.
+\item Variable and Object instances that exist for the entire planning
+  horizon.
 \item Temporally scoped predicate and action instances.
@@
 
-All this information is kept in \eus Plan Database so that inference and
+This information is kept in \eus Plan Database so that inference and
 search mechanisms can be used to look for a problem solution.
@@
 Planning Horizon is the temporal interval that a problem solver has to
-worry about, any decisions that fall completely outside of the Planning
-Horizon can be ignored. This is very useful for recurrent activities,
-consider for instance a model where a medical checkup has to occur every
-3 months, the user can model a recurring activity and then use the
-Planning Horizon to ensure that a finite number of medical checkup
-activities are generated as part of a plan. This is also useful to
-ignore activities that may happen too far in the future or in the past
-to be relevant for a particular plan, this way the user can ensure that
-the solver is making as few commitments as necessary.
+worry about, any decisions that fall completely outside of the
+Planning Horizon can be ignored. This is very useful for recurrent
+activities, consider for instance a model where a medical checkup has
+to occur every 3 months, the user can model a recurring activity and
+then use the Planning Horizon to ensure that a finite number of
+medical checkup activities are generated as part of a plan. This is
+also useful to ignore activities that may happen too far in the future
+or in the past to be relevant for a particular plan, this way the user
+can ensure that the solver is making as few commitments as necessary.
 
@@
 partial plan that needs to be refined toward a solution plan that
-achieves the goals. The operations to refine the partial plan PP at any
-time are:
+achieves the goals. The operations to refine the partial plan PP at
+any time are:
 
 \begin{enumerate}
-	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
+\item Find the flaws of PP, that is the conditions that prevent it
+  from being a solution plan (flaw types are explained below).
 \item Select one such flaw
@@
 \item If an inconsistency is found, try another resolver.
-	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
+\item If all possible resolvers for a particular flaw fail, return
+  failure, otherwise continue until resolving all the flaws.
 \end{enumerate}
@@
 The initial partial plan is the state of \eus Plan Database after the
-initial state has been instantiated, this results in a set of variable,
-object and token instances. Inference then takes place to detect flaws
-in the partial plan, there are 3 kinds of flaws that can be detected:
+initial state has been instantiated, this results in a set of
+variable, object and token instances. Inference then takes place to
+detect flaws in the partial plan, there are 3 kinds of flaws that can
+be detected:
 
@@
 \item Unbound Variable: a variable in the partial plan whose domain is
-  not a singleton. Unbound Variables are resolved by specification of a
-  value from the domain of the variable.
+  not a singleton. Unbound Variables are resolved by specification of
+  a value from the domain of the variable.
 
 \item Open Condition: an open condition is an inactive token. Inactive
-  tokens can be generated by explicitly posted goals, or when a token is
-  activated, rules may fire that create inactive slave tokens
+  tokens can be generated by explicitly posted goals, or when a token
+  is activated, rules may fire that create inactive slave tokens
   \comment{point to more detailed decription above in Rules Engine
-    section}.
-  Open
-  Conditions can be resolved in three ways:
+    section}.  Open Conditions can be resolved in three ways:
 
@@
   impact other tokens indirectly through possible overlapping
-  requirements on objects. Recall for example that a token may belong to
-  objects (e.g. Timelines) which require a total order over their
+  requirements on objects. Recall for example that a token may belong
+  to objects (e.g. Timelines) which require a total order over their
   tokens. If any 2 tokens could possibly overlap (though not
@@
   and consistent. Similarly, threats may arise where tokens share a
-  common resource and their current state might yield extensions of the
-  current partial plan which are inconsistent. Threats are resolved by
-  imposing ordering constraints among tokens.
+  common resource and their current state might yield extensions of
+  the current partial plan which are inconsistent. Threats are
+  resolved by imposing ordering constraints among tokens.
 
@@
 higher-level of abstraction (i.e. in terms of objects and tokens) than
-that of simply binding variables as is common in Constraint Satisfaction
-Problems. This is advantageous when one applies heuristics for ordering
-choices since it provides a richer context in which to make decisions.
-Furthermore, it aids in reducing the amount of work done by a solver so
-that only the necessary refinements are made, otherwise leaving the
-partial plan with maximum flexibility. For example, one can omit unbound
-variables which are time-points of tokens since threats will force a
-solver to impose restrictions on these variables based on the semantics
-of the objects to which their tokens apply. Thus the planning process
-may yield partially-ordered plans for which all possible extensions are
-provably valid.
+that of simply binding variables as is common in Constraint
+Satisfaction Problems. This is advantageous when one applies
+heuristics for ordering choices since it provides a richer context in
+which to make decisions.  Furthermore, it aids in reducing the amount
+of work done by a solver so that only the necessary refinements are
+made, otherwise leaving the partial plan with maximum flexibility. For
+example, one can omit unbound variables which are time-points of
+tokens since threats will force a solver to impose restrictions on
+these variables based on the semantics of the objects to which their
+tokens apply. Thus the planning process may yield partially-ordered
+plans for which all possible extensions are provably valid.
 
-The Plan Space Planning algorithm described above can be implemented in
-many different ways depending on the approach chosen for flaw and
-resolver selection and for backtracking. \eus built-in solver implements
-a chronological backtracking algorithm that is summarized in Algorithm
-\ref{alg:europa:solve}.
+The Plan Space Planning algorithm described above can be implemented
+in many different ways depending on the approach chosen for flaw and
+resolver selection and for backtracking. \eus built-in solver
+implements a chronological backtracking algorithm that is summarized
+in Algorithm \ref{alg:europa:solve}.
 
@@
 \begin{enumerate}
-	\item Line \ref{li:propagate}: Propagate the constraints to test for inconsistency.
-If found to be inconsistent, then we can return false since this is a dead-end i.e.
-no refinements to p can yield a consistent plan.
-	\item Line \ref{li:selectflaw}: Choose a flaw from the set of available flaws.
-	\item Line \ref{li:noflaw}: If there are no flaws, then p is complete and we can terminate with success.
-	\item Line \ref{li:decisionpt}: Otherwise we formulate a decision point which is a branch in the search space.
-Each choice is a particular refinement operation and the DecisionPoint collects all possible refinement operations for the given flaw.
-	\item Line \ref{li:decisionloop}: We keep trying until the chosen flaw is resolved, or until we run out of resolvers to try.
-	\item Line \ref{li:decide}: A new partial plan is obtained by application of a refinement operator. Note that the ordering over refinement operators to select is a non-deterministic step.
-	\item Line \ref{li:recurse}: Recursive call to solve the new planning problem. If successful, then we are done.
-	\item Line \ref{li:backtrack}: Otherwise, retract the last refinement operation and move on to try the next one.
-	\item Line \ref{li:noplan}:  If we arrive here, then we have exhausted all options to resolve the flaw, including the case where no options were available initially.
-Thus the problem cannot be solved.
+
+\item Line \ref{li:propagate}: Propagate the constraints to test for
+  inconsistency. If found to be inconsistent, then we can return false
+  since this is a dead-end i.e. no refinements to p can yield a
+  consistent plan.
+
+\item Line \ref{li:selectflaw}: Choose a flaw from the set of
+  available flaws.
+
+\item Line \ref{li:noflaw}: If there are no flaws, then p is complete
+  and we can terminate with success.
+
+\item Line \ref{li:decisionpt}: Otherwise we formulate a decision
+  point which is a branch in the search space. Each choice is a
+  particular refinement operation and the DecisionPoint collects all
+  possible refinement operations for the given flaw.
+
+\item Line \ref{li:decisionloop}: We keep trying until the chosen flaw
+  is resolved, or until we run out of resolvers to try.
+
+\item Line \ref{li:decide}: A new partial plan is obtained by
+  application of a refinement operator. Note that the ordering over
+  refinement operators to select is a non-deterministic step.
+
+\item Line \ref{li:recurse}: Recursive call to solve the new planning
+  problem. If successful, then we are done.
+
+\item Line \ref{li:backtrack}: Otherwise, retract the last refinement
+  operation and move on to try the next one.
+
+\item Line \ref{li:noplan}: If we arrive here, then we have exhausted
+  all options to resolve the flaw, including the case where no options
+  were available initially. Thus the problem cannot be solved.
 \end{enumerate}
@@
 Dead-ends in the search are discovered through constraint propagation.
-Constraint propagation is a vehicle for evaluating the consistency of a
-partial plan and also for filtering infeasible values from consideration
-prior to commitment, allowing in some cases a strong look-ahead
-capability which is essential for tractable search. Consistency testing
-is initiated by the isConsistent procedure. The algorithm permits a
-heuristically controlled search by applying orderings for chooseFlaw and
-makeDecisionPoint. It results in a chronologically-backtracking search.
+Constraint propagation is a vehicle for evaluating the consistency of
+a partial plan and also for filtering infeasible values from
+consideration prior to commitment, allowing in some cases a strong
+look-ahead capability which is essential for tractable
+search. Consistency testing is initiated by the isConsistent
+procedure. The algorithm permits a heuristically controlled search by
+applying orderings for chooseFlaw and makeDecisionPoint. It results in
+a chronologically-backtracking search.
 
@@
 
-It should be emphasized that while this algorithm is commonly employed,
-it is only one of many that could be implemented. \comment{give examples
-  of local search solvers for scheduling and CSP?}
+It should be emphasized that while this algorithm is commonly
+employed, it is only one of many that could be
+implemented. \comment{give examples of local search solvers for
+  scheduling and CSP?}
 
@@
 
-RCPSP solver:
+RCPSP solver \comment{Do we need this example?}:
 \begin{verbatim}

### 2012-02-16 europa.tex
- added figs/shopping_europa.pdf that show the plan for buying Milk with a more europa centric point of view [europa.tex] changed the Agent::Buy rule to be constant with my figures

@@
 
-  starts(effect Own purchase);
+  meets(effect Own purchase);
   eq(purchase.product,product);
@@
 \cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network (STN). Such a network
-can be transformed into a Distance Graph (DG) where the outward arc from
-a node represents the maximum distance from the source node to the
-target node. They also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative cycle. A
-negative cycle is a path from a node to itself that has a path length
-less than 0. If such a cycle exists, the network is inconsistent. It was
-further shown that a single-source shortest path algorithm was
-sufficient to detect a negative cycle and provide sufficient propagation
-to yield a backtrack-free search. Thus we have an efficient and complete
-algorithm for propagating an STN. These results build on the already
-established notion of a CSP and are naturally incorporated into the
-general representation and propagation scheme used in \eu.
-The diagram below illustrates a simple STN with just 2
-variables and a single constraint. It also shows the resulting DG.
+grouped together to form a Simple Temporal Network (STN). Such a
+network can be transformed into a Distance Graph (DG) where the
+outward arc from a node represents the maximum distance from the
+source node to the target node. They also showed that shortest path
+algorithms could be used to propagate values in the network and
+discover a negative cycle. A negative cycle is a path from a node to
+itself that has a path length less than 0. If such a cycle exists, the
+network is inconsistent. It was further shown that a single-source
+shortest path algorithm was sufficient to detect a negative cycle and
+provide sufficient propagation to yield a backtrack-free search. Thus
+we have an efficient and complete algorithm for propagating an
+STN. These results build on the already established notion of a CSP
+and are naturally incorporated into the general representation and
+propagation scheme used in \eu.  The diagram below illustrates a
+simple STN with just 2 variables and a single constraint. It also
+shows the resulting DG.
 
@@
 
-\comment{ add figure. Use Europa wiki figure; or use conor figure from slides}
+% \comment{ add figure. Use Europa wiki figure; or use conor figure from slides}
 

### 2012-02-16 related.tex
updates
updates to europa text, figs and citations

@@
 \subsection{The Remote Agent Experiment and beyond}
+\label{sec:rabeyond}
 
@@
 \texttt{RAX} led to the development of \texttt{IDEA}
-\cite{mus02,mus04} with the central theme of using planners
-collectively for problem solving arose. Another critical theme was
-iteratively and incremental plan repair of an anytime plan
+\cite{mus02,mus04,Dias:2003ua,mus06} with the central theme of using
+planners collectively for problem solving arose. Another critical
+theme was iteratively and incremental plan repair of an anytime plan
 \cite{Zaimag96} as proposed by \texttt{CASPER} \cite{chien00}.

### 2012-02-16 trunk/europa-basics.tex
updates to europa text, figs and citations

@@
 
-\subsection{The basics}
+\paragraph {A brief note on \texttt{EUROPA} history and notation} A
+brief word about the history of planning paradigm and its
+implementation in this chapter. The earliest origin of what we will
+call \eu is from the \texttt{HSTS} planner \cite{mus94} which was used
+for a number of demonstrations for space telescopes as noted in
+Section \ref{sec:rabeyond}. \texttt{HSTS} was the
+\texttt{LISP}-based\footnote{This was the first (and only to our
+  knowledge) software ever to been written in \texttt{LISP} to be
+  flown in space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
+the planner was designed at NASA Ames which went on to be used for the
+\texttt{MAPGEN} system \cite{bresina05}. A vastly re-factored, higher
+performance open source version \cite{europapso} was named \eu and is
+the subject of this chapter. Note however, in using the name \eu, we
+refer to the \emph{fundamental ideas as well as the planning paradigm}
+even if there is a specific implementation online.
+
+\subsection{The Basics}
 \label{sec:basics}
@@
 
+\begin{figure} \centering
+  \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
+  \caption{\small A potential plan for a robot shopping for groceries.}
+\label{fig:shop:shopping}
+\vskip+0.1cm
+\end{figure}
+
 % Planning, for our purposes, can be thought of as determining all the
@@
 For such a robot the milk buying plan at the store might look like
-\comment{Figure needed}.
+that shown in Fig. \ref{fig:shop:shopping}.
 

### 2012-02-17 trunk/europa.tex
in europa.tex: fleshed out inference section, finished search section.

@@
 described above and makes them easily accessible through modeling and
-API layers.  A user will typically use the modeling facilities to
+Application Programming Interface (API) layers.  A user will typically use the modeling facilities to
 inject a problem domain description, and the API to extract and
@@
 \cite{mackworth77}. The CRE is designed so that specialized reasoning
-algorithms for specific constraints can be easily and efficiently
+algorithms for new constraints can be easily and efficiently
 plugged in.
@@
 
-The \textbf{Rules Engine} (RE) modularly adds the capability to
-support domain rules that describe dependencies between tokens,
+The \textbf{Rules Engine} (RE) modularly adds the semantics of domain rules that describe dependencies between tokens.
 
@@
 programmatic way is cumbersome to the extent of being impractical,
-that motivated the creation of modeling languages that support those
+this motivated the creation of modeling languages that support those
 definitions in a much more natural and compact way. NDDL is the main
@@
 In an effort to mimimize the amount of work that a user needs to
-perform to solve specific problems, \eu also provides extension
+perform to solve specific problems, \eu also bundles extension
 modules that have been found to be useful in many domains:
@@
 
-Now, having that we have a specification for the Shopping Agent
-domain, here is how a particular problem instance could be specified:
+Having that specification for the Shopping Agent
+domain, here is how a particular problem instance could be stated:
 
@@
 
-A problem instance consists mainly of
+A problem instance consists of:
 
@@
 \item \textbf{Goals}: Goals are represented as tokens in \eu's Plan
-  Database, differently from Facts, Goals trigger inference in \eu
+  Database. Differently from Facts, Goals trigger inference in \eu
   which determines whether flaws (unresolved subgoals, overloaded
-  resources, etc., flaws will be covered in detail in the search
-  section below) need to be resolved, if that is the case, search
-  modules can be invoked which will look for a plan where all the
-  flaws are resolved.
+  resources, etc.) need to be resolved, if that is the case, search
+  modules can be invoked which will look for a plan to resolve those
+  flaws.
 
@@
 
-\comment{explain how inference is used to detect perform bounds
-  propagation, domain reduction and to detect constraint violations.} 
-- talk about AC3 implementation and extension mechanisms
-- talk about europa's constraint library
+In \eu,  all the Variables and Constraints from a problem domain domain description are kept inside the Constraint Reasoning Engine (CRE).
+The CRE provides a framework where any number of constraint types can collaborate to perform inference and prune variable domains using Bounds Propagation and Domain Reduction techniques.
+\eu uses a Propagator pattern, in which a Propagator object keeps track of a set of constraint instances of the same type and is responsible for ensuring that whenever any variable changes or constraints are introduced or dropped, all related variables are updated through inference.
 
-The Temporal Network: The notion of time is central to temporal
-planning. \eu uses variables to explicitly represent timepoints for
+In a planning problem, variable or constraint changes can come directly from an end user (by introducing new activities or goals and relating them to the rest of the plan through constraints, by changing the start time of an activity, etc), or can come from an automated planner (by expanding subgoals, introducing precedence constraints to resolve mutual exclusion conflicts, etc).
+When these changes happen, each Propagator is only able to ensure consistency for the constraints that are associated with it.
+To maintain global consistency, the CRE implements a variant of the AC-3 algorithm \cite{mackworth77} where it asks Propagators to perform inference until the Constraint Network returns to an stable state.
+
+\comment {add europa's AC-3 algorithm here}
+
+A \eu user can introduce new constraint types by implementing specialized Constraint and Propagator classes in C++, however, \eu provides a built-in with over 25 arithmetic, temporal, set and resource constraint types so that a large number of problem domains can be modeled without having to introduce new constraint types.
+\comment{insert table from http://code.google.com/p/europa-pso/wiki/ConstraintLibraryReference ?
+}
+Temporal and resource constraints are specially important to model real-life planning domains, so it is worth going into some more detail about how \eu's built-in support works for those constraint types.
+
+\textit{Temporal Constraints:} \eu uses variables to explicitly represent timepoints for
 plan activities and states. Constraints among timepoints provide a
@@
 outward arc from a node represents the maximum distance from the
-source node to the target node. They also showed that shortest path
-algorithms could be used to propagate values in the network and
-discover a negative cycle. A negative cycle is a path from a node to
-itself that has a path length less than 0. If such a cycle exists, the
-network is inconsistent. It was further shown that a single-source
-shortest path algorithm was sufficient to detect a negative cycle and
-provide sufficient propagation to yield a backtrack-free search. Thus
-we have an efficient and complete algorithm for propagating an
-STN. These results build on the already established notion of a CSP
-and are naturally incorporated into the general representation and
-propagation scheme used in \eu.  The diagram below illustrates a
-simple STN with just 2 variables and a single constraint. It also
+source node to the target node. 
+The diagram below illustrates a simple STN with just 2 variables and a single constraint. It also
 shows the resulting DG.
@@
 
-% \comment{ add figure. Use Europa wiki figure; or use conor figure from slides}
+\cite{dechter91} also showed that shortest path
+algorithms could be used to propagate values in the network and
+discover a negative cycle. A negative cycle is a path from a node to
+itself that has a path length less than 0. If such a cycle exists, the
+network is inconsistent. It was further shown that a single-source
+shortest path algorithm was sufficient to detect a negative cycle and
+provide sufficient propagation to yield a backtrack-free search. Thus
+we have an efficient and complete algorithm for propagating an
+STN. \eu's Propagator for temporal constraints is a straight implementation of this algorithm.
 
-Resources:
+\textit{Resource Constraints:} Most real life planning problems have to deal with 
+the fact that plans made up of actions that require resources to be performed typically
+have only a limited amount of those resources available (think of fuel, storage capacity, etc).
+\eu provides built-in support for resources whose availability can be measured with real numbers, 
+in terms of \eu's  representation, resource object types and their related constraints are provided as follows:
 
@@
 
-\item \textbf{Reservoir} resources can have both 'consume' and
-  'produce' tokens. Note that these tokens do NOT have the usual
-  start/end/duration token variables, since occur at a single instant
-  in time. Instead, they have a single variable, 'time' used to
-  represent that instant.
+\item \textbf{Reservoir} resources can have both \texttt{consume(resource,time,quantity)} and
+  \texttt{produce(resource,time,quantity)} constraints. Production and consumption is instantaneous (it happens at a single point in time)
 
-\item \textbf{Reusable} resources have 'uses' tokens that use a
-  quantity for the duration of the token. These tokens do have the
-  usual start/end/duration variables.
+\item \textbf{Reusable} resources have \texttt{uses(resource,start,duration,quantity)} constraints that use a
+  quantity of the resource for a specific duration, then release it.
 
-\item \textbf {Unary} resources are reusable resources with unit
-  quantity and have 'use' tokens (again, with the usual
-  start/end/duration variables).
+\item \textbf {Unary} resources are Reusable resources with capacity=1, which allows \eu to perform 
+some run time savings compared to the generic Reusable class.
 \end{enumerate}
 
-\comment{talk about resource envelope computation by using MaxFlow formulation.}
+\eu offers configuration options so that in addition to the usual constraint violations when a resource is completely depleted, plan flaws can be generated whenever a resource level 
+is outside a specific upper or lower limit, or when instantaneous production or consumption exceeds a specific threshold.
+
+Recall that \eu supports a flexible temporal representation, where time is represented by intervals, instead of single values.
+In this context, determining whether a resource is the source of a plan flaw or a constraint violation in general is more difficult than just computing a profile
+of the resource and checking against some limits. 
+In the context of time flexibility, there are many resource profiles that are possible, depending on when exactly each production/consumption event happens out of all the possible values that are in the domain of the variables that represent the time for the event.
+\cite{mus02} proposed an algorithm where the variables that represent the time for production/consumption events are placed in a temporal network, similar to the network used by the Temporal Propagator above, then the constraints on the resource can be seen as quantities flowing through that network.
+Computing the maximum flow will yield the maximum resource usage, and thus a lower bound on the resource level, and computing a minimum flow will similarly yield an upper bound.
+Since maximum flow on a temporal network can be computed in polynomial time, \eu takes advantage of this algorithm to efficiently compute upper and lower bound envelopes on the resource level.
+
+\comment{provide snapshot of resource envelopes generated by europa}
+
+Resource envelopes can be very useful when trying to figure how brittle a plan in the face of changing activity durations or start times.
+They can also be useful when searching for plans, although they are relatively expensive to compute so their ability to prune the search space must be weighted against the ability to search faster \comment{reference Barreiro,Morris, Bresina paper on grounded inference}.
 
@@
 \item A problem instance (also called initial state) that consists of:
+    \begin{enumerate}
         \item Variable and Object instances that exist for the entire planning
@@
         \item Temporally scoped goals.
+\end{enumerate}
 
@@
 
-\comment{expand a little more on how heuristics can be implemented by
-  applying ordering/filtering for chooseFlaw and makeDecisionPoint}
-examples:
-- deal with earliest flaws first
-- deal with planning (open condition flaws) before scheduling (threats) ones
+Heuristics can be implemented by applying ordering/filtering policies for
+\texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance, by dealing with earliest flaws first, 
+or by making planning decisions (open condition flaws) before scheduling ones (threats), etc.
 
-It should be emphasized that while this algorithm is commonly
-employed, it is only one of many that could be
-implemented. \comment{give examples of local search solvers for
-  scheduling and CSP?}
+It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be
+implemented. As a simple example, consider the following solver for the N-Queens problem, where there is an integer variable 
+Vi to represent the queen in column i, so that for instance, V1=3 means that the variable on column 1 is placed on row 3. 
+non-attack constraints are posted as arithmetic constraints among the Vi variables to ensure that all the rows and diagonals contain only one queen.
 
-N-Queens solver:
 \begin{verbatim}
-    public void solve(int maxIter)
+    public void solve(ConstraintEngine ce, int maxIter)
     {
-    	init();
+    	curIter=0;
 
-    	for (int i=0;psengine_.getViolation() > 0 && i < maxIter;i++) {
-    	    PSVariable queenToMove = getQueenWithMaxViolation();
-    		int curPos = queenToMove.getSingletonValue().asInt();
+    	while ((ce.getViolations() > 0) && (curIter < maxIter)) {
+	    Variable queenToMove = getQueenWithMaxViolation(ce);
+    	     int curPos = queenToMove.getCurrentValue();
 
+            SortedSet<int> moves = getMoves(ce,queenToMove,curPos);             
   	   boolean moved = false;
-            SortedSet<Move> moves = getMoves(queenToMove,curPos);             
-            for (Move m : moves) {
-            	moved = makeMove(queenToMove,curPos,m,false);
+            for (int newPos : moves) {
+            	moved = tryMove(queenToMove,curPos,newPos);
             	if (moved)
@@
             if (!moved) 
-            	makeMove(queenToMove,curPos,moves.first(),true);
+            	forceMove(queenToMove,curPos,moves.first());
 
             checkSolution(); // See if we have a new best solution            
-            notifyIterationCompleted(curIteration_++);
-
-            if (curIteration_-bestIter_ > 50) 
-            	restart();
+            curIteration_++;
     	}
@@
 
-RCPSP solver \comment{Do we need this example?}:
-\begin{verbatim}
-    public void solve(PSEngine psengine,
-    		          long timeout, // in msecs
-    		          int bound)
-    {
-       init(psengine,timeout,bound);
+\eu's Constraint Reasoning Engine can be used to compute the non-attack constraint violations (\texttt{ce.getViolations()} is a straight call on the \eu API), 
+to easily pick the queen that has he most violations by implementing \texttt{getQueenWithMaxViolation(ce)} and to look ahead and see what 
+moves are possible for a queen that would remove its violations by implementing \texttt{getMoves(ce,queenToMove,curPos)}.
 
-       for (int i=0; true ; i++) {
-          flatten();
-          updateSolution(i);
-          updateCriticalPrecedences();
-
-          if ((nbStable_ > maxStable_) || (bestMakespan_ <= makespanBound_))
-             break;
-
-          if (timer_.getElapsed() > timeout) {
-              timedOut_ = true;
-              break;
-          }
-
-          relax();
-          curIteration_++;
-       }
-
-       restoreBestSolution();
-    }
-
-\end{verbatim}
-
-\comment{explain how other solvers can be built, maybe use NQueens or
-  RCPSP example to illustrate. Or maybe build a little specialized
-  solver for shopping or rover examples.}
+In similar fashion, the constraint violation and flaw reporting mechanisms that \eu provides can be used as building blocks that specialized or general purpose search algorithms can use to construct plans.
 

### 2012-02-17 trunk/europa.tex
europ+concepts touch ups with Javier, fig for arch and updates to arch

@@
 
-\comment{provide snapshot of resource envelopes generated by europa}
+\comment{figure :provide snapshot of resource envelopes generated by europa}
 
@@
   is activated, rules may fire that create inactive slave tokens
-  \comment{point to more detailed decription above in Rules Engine
+  \comment{reference: point to more detailed decription above in Rules Engine
     section}.  Open Conditions can be resolved in three ways:

### 2012-02-17 arch.tex
europ+concepts touch ups with Javier, fig for arch and updates to arch

@@
 controller which could bring automated planning closer to low-level
-control of the vehicle while maintaining its reactivity. Traditionally
-planning has been deemed computationally expensive with planner
-performance restricting robot reactivity
+control of the vehicle while maintaining its reactivity.
+
+Traditionally planning has been deemed computationally expensive with
+planner performance restricting robot reactivity
 \cite{ghallab04,Dias:2003ua}. In control architectures where task
@@
 reactiveness and that the environment can change at a faster rate than
-the planner can plan for it. In such a situation, the agent may thrash
-if the internal state of the plan gets out of synch with the actual
-state of the world. As a result the planning problem managed {\em in
-  situ} remains fairly detached from low-level control. Most
-adaptations at the lower-level reactive layers in contrast are managed
-by specialized components (for example LAAS architecture
-\cite{alami:1998p820,Ingrand07} and CLARATy \cite{Nesnas:2003do}
-) that rely on different representations for modeling local
-control behavior or worse do not have any explicit formal agent
-behavior model. The consequent different techniques for specifying
-each layer in the architecture results in duplication of effort and a
-diffusion of knowledge leading to significant design and integration
-issues \cite{DS1report}.  \texttt{IDEA} mitigated most of these issues
-by tightly integrating planning and execution in a single
-representational and computational framework with a unified
-declarative model. Our approach with \texttt{T-REX} builds on this
-design methodology albeit in a substantially systematic manner
-
-The \rx architecture contributes to agent architectures over and
-beyond \texttt{IDEA} are in the following ways:
+the planner can plan for. In such a situation, the agent may thrash if
+the internal state of the plan gets out of synch with the actual state
+of the world. As a result the planning problem managed {\em in situ}
+has traditionally remained fairly detached from low-level
+control. Most adaptations at the lower-level reactive layers in
+contrast are managed by specialized components (for example
+\texttt{LAAS architecture} \cite{alami:1998p820,Ingrand07} and
+\texttt{CLARATy} \cite{Nesnas:2003do}) that rely on different
+representations for modeling local control behavior or worse do not
+have any explicit formal agent behavior model. The consequent
+different techniques for specifying each layer in the architecture
+results in duplication of effort and a diffusion of knowledge leading
+to significant design and integration issues \cite{DS1report}.
+\texttt{IDEA} mitigated most of these issues by tightly integrating
+planning and execution in a single representational and computational
+framework with a unified declarative model. Our approach with \rx
+builds on this design methodology albeit in a substantially systematic
+manner. The \rx architecture contributes to agent architectures over
+and beyond \texttt{IDEA} are in the following ways:
 
@@
   depending on where it is located in a reactor dependency graph such
-  as shown in Fig. \ref{fig:agent}.
+  as shown in Fig. \ref{fig:agent}. A reactor closer to the hardware
+  in the hierarchy, is expected to deliberate less and be highly
+  responsive to observations from the world. This implies lower values
+  of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex} below).
 
 \item \emph{Coupling} a tighter integration between the planning
-  process and world evolution to ensure that planning occurs while
-  being continuously informed of state change. By doing so, the
-  planner is better informed in its deliberation despite a larger
-  latency than changing world change state.
+  process and world evolution to ensure that planning occurs
+  \emph{during continuous} state update. By doing so, the planner is
+  better informed in its deliberation despite a larger latency than
+  the changing world.
 
@@
 
-Partitioned inference is a major contribution of \rx. However that is
-not the focus of this chapter; details are in \cite{py10} and
-\cite{rajan12}. Coupled interaction between planning and the world on
-the other hand, has a significant implication on how planning can be
-integrated in such frameworks; here we show using the \eu planner can
-be coupled to our framework. First we link concepts in \eu planning
-with the \rx framework. Note that in the following treatise, we call
-the formal semantic framework of \rx as the 'agent'. Reactors
-constitute an entity within this agent and are standalone blocks of
-inference.
+Systematic partitioned inference is a major contribution of
+\rx. However that is not the focus of this chapter with details
+covered in \cite{py10, rajan12}. Coupled interaction between planning
+and the world on the other hand, has a significant implication on how
+planning can be integrated in such frameworks. In this chapter we show
+how using the \eu planner can be coupled to our framework by first
+linking concepts in temporal planning with the \rx framework. Note
+that in the following treatise, we call the formal semantic framework
+of \rx as the 'agent'. Reactors constitute an entity within this agent
+and are standalone blocks of inference. The scope, detail and number
+of reactors are application specific and outside the scope of this
+chapter.
 
@@
   variables} as the fundamental basis of interaction. Additionally it
-comes with a well-defined ownership model associated with each state
-variable. In this framework each reactor is seen as a 'black box'
-providing its own {\em internal} state variables which it is
+comes with a well-defined ownership model associated with each such
+state variable. In this framework each reactor is seen as a 'black
+box' providing its own {\em internal} state variable which it is
 responsible for in order to maintain its value at every instant of
 time. Each such instance is represented by a {\em tick} with a fixed
-duration; in our implementation of \rx on the Dorado, we use a tick of
-$1$ second. In order to identify its state a reactor can subscribe to
-{\em external} state variables owned by other reactors, for which it
-will receive new updates when the owner reactor generates its {\em
-  internal} state.
-
-As the execution frontier -- called $\tau$ -- advance, the \rx agent
-propagates the current state information accross all the reactors states
-variables through the {\em synchronization} process.
- The sequencing of the reactors synchronization is identified
-by traversing the reactor graph from the least dependents reactors --
-the ones with no {\em external} timelines -- to the most dependent
-ones which typicallu do not provide {\em internal} timelines.
+duration; in our implementation of \rx on the Dorado AUV, we use a
+tick of $1$ second. In order to identify its state a reactor can
+subscribe to {\em external} state variables owned by other reactors,
+for which it will receive new updates when the owner reactor generates
+its {\em internal} state.
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
+  \caption{\small An example of a \rx reactor hierarchy which is used
+    to determine a synchronization order for propagation of values in
+    \emph{external} state variables.}
+  \label{fig:reactor-DAG}
+\end{figure}
+
+As the execution frontier, $\tau$ advances, the \rx agent propagates
+the current state information across all reactor state variables
+through the {\em synchronization} process.  The sequencing of the
+reactors synchronization is identified by traversing a reactor
+dependency graph from the least dependent -- ones with no {\em
+  external} timelines -- to the most, \ie which typically do not have
+{\em internal} timelines. Fig. \ref{fig:reactor-DAG} shows an example
+of such a dependency graph implemented as a directed acyclic graph.
 
@@
   external} state variables which will be transformed as an {\em
-  internal} goal to the owner.  Such \emph{goal dispatching} is
-expected to produce a deliberation phase for the receiving reactor
-which can cyclically and in turn, produce a plan impacting its {\em
-  external} state variables' future that will propagate down following
-the same mechanism.
-
-In order to avoid to be overflood by goals that may threats the
-ability of a reactor to process these goals and integrate them in tis
-plan. In order to do so each reactor specifies the two following
-values that apply for the full scope of the agent lifetime:
+  internal} goal to the owner reactor(s).  Such \emph{goal
+  dispatching} is expected to produce a deliberation phase for the
+receiving reactor which can cyclically in turn, produce a plan
+impacting its {\em external} state variables' future that will
+propagate down following the same mechanism.
+
+In such a process, one challenge is to avoid being overwhelmed by
+goals that may threaten the ability of a reactor to process and
+integrate in its evolving partial plan. For that reason, each reactor
+specifies the following two parameters that applied towards the scope
+of the agent lifetime:
+
 \begin{itemize}
+
 \item $\lambda$ is the specified latency of this reactor, which is an
-  indicator of the maximum expected time for a reactor to produce a
-  plan.\footnote{Note that while it is highly recommended to select a
-    sound value, a failure to produce a plan within a chosen value for
-    this parameter is not considered as a critical failure of the
-    reactor.}
-\item $\pi$ is the planning look-ahead of the reactor. which indicates
-  how many ticks in the future this reactor is looking ahead during
-  deliberation. 
+  indicator of the \emph{maximum expected time} for a reactor to
+  produce a plan.\footnote{Note that while it is highly recommended to
+    select a sound value, a failure to produce a plan within a chosen
+    value for this parameter is not considered as a critical failure
+    of the reactor.}
+
+\item $\pi$ is the planning \emph{look-ahead} of the reactor. which is
+  indicative of the number of ticks in the future this reactor is
+  looking ahead during deliberation.
+
 \end{itemize}
 
-Using this information along with the dependency relation betwen
-reactor \rx defines the notion of execution latency of a state
+Using this information along with the dependency relation between
+reactors, \rx defines the notion of execution latency of a state
 variable as follow:
+
 \begin{definition}
   The {\em execution latency} of a {\em state variable} $s$ which is
-    internal to the reactor $r$ is the maximum expected delay before 
-    a goal on this state varabale can be included in its plan and
-    executed by the  reactor $r$.
-
-    It is defined recursrevely as follow: 
+  internal to the reactor $r$ is the maximum expected delay before a
+  goal on this state variable can be included in its plan and executed
+  by the reactor $r$. It is defined recursively as: 
   \begin{equation*}
@@
   \end{equation*}
-    Where  $exts(r)$ is the set of the state varaibles which are
+  Where  $exts(r)$ is the set of the state variables which are
   {\em externals} to $r$. 
@@
 
-The concept of execution latency is critical at the agent level
-and can also impact the planning/deliberation process on the {\em
-  external} state variables of a reactor. For simplicity sake in the
-future sections more focused in the \eu reactor, we abstract it out 
-as being equivalent to the reactor latency (meaning that we 
-assume that all {\em external} timelines of this reactor have
-$\lambda_{exec} = 0$). At the \rx agent level though this
- execution latency is used in turn to define the {\em dispatch
-window} of the same state variable:
-\comment{Needed to say that early on but the flow of this paragraph 
-is not very tight}
+For intertwined plan execution, the concept of execution latency is
+critical at the agent level since it can impact the
+planning/deliberation process on {\em external} state variables of a
+reactor. For simplicity we abstract it out as being equivalent to the
+reactor latency implying that we assume that all {\em external}
+timelines of this reactor have $\lambda_{exec} = 0$. At the \rx agent
+level this execution latency is used to define the {\em dispatch
+  window} of the same state variable.% : \comment{Needed to say that
+  % early on but the flow of this paragraph is not very tight}
+\kcomment{Looks fine to me unless you want to clarify further.}
+
 \begin{definition}
   \label{def:dispatch}
-  For every timelines $s$ {\em internals} to the reactor $r$, the {\em 
+  For every timeline $s$ {\em internal} to a reactor $r$, the {\em
     dispatch window} of this timeline at the execution frontier $\tau$
@@
 
-This window is used by \rx in order to identify when to dispatch a
-request to a {\em reactor} as a new goal. This happens only if the
-request start time -- represented as the interval of possible start
-instants -- intersects this window. Requests further in the future 
-are buffered by the agent until this window advanced enough to 
-overlap their start time. This gives a simple mechanism in order to
-request a change from the {\em external} state variable of a a reactor 
-while avoiding to dispatch too much goals at once to the reactor that
-declare this same state variable as {\em internal}. The choice of both
-the latency and look-ahead of a reactor can be delicate as one is
-dependent on the other. Latency also highly dependent on exogenous
-paramaters such as the computing power of the host system and the
-number of reactors needing to deliberate at once. As of today these
-should be evaluated carefully by the system designers and while
-having an important impact in the efficacy of the agent to find and
-execute a global strategy they are just indicative and a failure from
-a reactor to produce a plan in its latency is not considered as a
-critical failure from \rx point of view.
-
-Requests from a reactor can  be {\em recalled} by this reactor when it
-they are not part of its plan anymore. For example if this reactor
-need to replan due to an unexpected event, it will recall ol the
-requests it did propduce that have not been observed yet. These
-recalls are notified to the reactor that received this request as a
-goal in order to remove it from its current objectives.
+This window is used by \rx to identify when to dispatch a request to a
+{\em reactor} as a new goal. This happens only if the request start
+time -- represented as the interval of possible start times --
+intersects this window. Requests further in the future are buffered by
+the agent until this time window advances for an overlap. This
+provides an elegant mechanism to request a change from the {\em
+  external} state variable of a reactor while avoiding flooding the
+owner reactor of this state variable.
+
+The choice of both the latency and look-ahead of a reactor can be
+delicate as one is dependent on the other. Needless to say, latency is
+highly dependent on exogenous paramaters such as the host computers
+computing power and the number of reactors needing to deliberate at
+once, among others. As the design stands, these variables should be
+evaluated carefully by the system designer balancing agent efficacy.
+
+% efficacy of the agent to find and execute a
+% global strategy they are just indicative and a failure from a reactor
+% to produce a plan in its latency is not considered as a critical
+% failure from \rx point of view.
+
+Requests from a reactor can be {\em recalled} when it is determined
+that they are not part of its projected plan anymore. For example if a
+reactor needs to replan due to an unexpected event, it will recall all
+requests it produced that have not been observed \kcomment{replace
+  ``observed'' with ``not incorporated by the owner reactor''?}
+yet. This recall is sent to the reactor that received the goal(s) to
+remove it/them from its current set of objectives.
 
@@
   \qquad
-  \subfloat[\small State information flow between the
-    \texttt{Pilot} and the \texttt{Vehicle} reactors.
-    Observations transit as a bottom-up flow 
-    and apply to the execution frontier $\tau$. 
-    Goal information flow from the top to the bottom and relates to 
-    desired future values of state variables.]{
+  \subfloat[\small State information flow between the \texttt{Pilot}
+  and the \texttt{Vehicle} reactors in an \rx instance.  Observations
+  transit as a bottom-up flow and apply to the execution frontier
+  $\tau$.  Goal information flow from the top to the bottom and
+  relates to desired future values of state variables.]{
     \includegraphics[width=0.4\columnwidth]{figs/TREX_flow}
-      %\vskip-1cm
     \label{fig:trex:flow}
   }
-    \caption{\rx connection of reactors through state variables based on the
-    AUV agent we use at sea.}
+\caption{\small \rx reactors and shared state variables.}
 \end{figure}
@@
 in Fig. \ref{fig:agent}. In this instance all the reactors (symbolized
-by colored boxes) do not have an explicit connection from one to
-another but instead rely on a publish/subscribe model on shared state
-variables. For instance the \textsf{Pilot} reactor needs the {\em
-  externally} managed state information from \textsf{Sensor\_Data} and
-\textsf{Nav\_cmd} state variables in order to maintain the
-\emph{internal} state variables \textsf{Waypoint} and
-\textsf{Nav\_policy}. This relation applies in two ways:
+by colored boxes) % do not have an explicit connection from one to
+% another but instead
+rely on a publish/subscribe model on shared state variables. For
+instance the \textsf{Pilot} reactor needs the {\em externally} managed
+state information from \textsf{Sensor\_Data} and \textsf{Nav\_cmd}
+state variables in order to maintain the \emph{internal} state
+variables \textsf{Waypoint} and \textsf{Nav\_policy}. This relation
+applies in two ways:
 
@@
   \textsf{Pilot} needs to know the current {\em external} state
-  information it relies on. This information is providede in turn by
-  the reactors that provided these timelines as {\em internal}. In
-  this specific case, the current \texttt{Waypoint} where the vehicle
-  is heading can be evaluated based on the \texttt{Nav\_cmd} executed
-  along with usefull \texttt{Sensor\_Data}, both being provided by the
+  information it relies on. This information is provided in turn by
+  the reactors that define these timelines as {\em internal}. In this
+  specific case, the current \texttt{Waypoint} where the vehicle is
+  heading can be evaluated based on the \texttt{Nav\_cmd} executed
+  along with useful \texttt{Sensor\_Data}, both being provided by the
   \texttt{Vehicle} reactor.
@@
 
-Given the above example, the \rx architecture therefore helps abstract
-reactor interaction by constraining them to state variables. Each
-reactor is therefore agnostic where its {\em external} state
-information is managed as long as this information is available and
-properly maintained for managing {\em internal} state
-information. This {\em internal} state in turn may be used by other
-reactors. Such a formalism also constrains reactor interaction to
-state information exchange which comes in two forms as illustrated in
-Figure \ref{fig:trex:flow}:
+Given the above example, \rx abstracts reactor interaction by
+constraining them to state variables. Each reactor is therefore
+agnostic where its {\em external} state information is managed as long
+as this information is available and properly maintained for managing
+{\em internal} state information. This {\em internal} state in turn
+may be used by other reactors. Such a formalism also constrains
+reactor interaction to state information exchange which comes in two
+forms as illustrated in Fig. \ref{fig:trex:flow}:
 
@@
   tick. During this phase the agent provides updates to each reactor
-  on their {\em external} state variables for this tick so as they
-  could compute their {\em internal} state variable values for this
-  same tick. By doing so we propagate a consistent view of that state
-  of the world throughout all the reactors as time
-  advances. %\comment{could do with a figure here}
+  on their {\em external} state variables for this tick so they can
+  compute their {\em internal} state variable values for this same
+  tick. By doing so we propagate a consistent view of that state of
+  the world throughout all the reactors as time
+  advances.
 
@@
   future values of one of its {\em external} state. When such a plan
-  is identified the agent ensures that this part of the plan is given
-  as a goal to the owner of the given {\em external} state
-  variable(s). %\comment{could do with a figure here}
+  is identified the agent ensures that this part of the plan is
+  provided as a goal to the owner of the given {\em external} state
+  variable(s).
 

### 2012-02-17 concepts.tex
europ+concepts touch ups with Javier, fig for arch and updates to arch

@@
   very useful characteristic that allow users to create flexible
-  plans. \comment{say more about the benefits of flexibility?}
+  plans.
 
@@
 \item[\textbf{Tokens}] In first order logic, a predicate defines a
-  relation between objects and properties \comment{citation?}.  In
+  relation between objects and properties \comment{use Genesereth citation?}.  In
   \eu, we define such relations between variables whose domains are
@@
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \comment{add small
+  overlap may occur among them \cite{mus94}.  \comment{Maybe: add small
     example?}
@@
   ordering requirements on their Tokens in order to ensure that the
-  level of the resource remains within specified limits.  \comment{add
-    small example?}
+  level of the resource remains within specified limits.
+  \comment{Maybe: add small example?}
 \end{enumerate}
@@
 
-  \comment{Add figure showing an AgentLocation timeline with an
+  \comment{Use Fig 4 and edit:: Add figure showing an AgentLocation timeline with an
     At(Home) token with start=0 and end open, an empty AgentBag
@@
 
-\comment{Add figure showing the effect of merging}
+\comment{Use previous fig and modify as appropriately:: Add figure
+  showing the effect of merging; all figs on this section are
+  conjoined. JB to sketch it out.}
 
@@
 
-\comment{Add figure showing the effect of activating the token by
-  putting it on the AgentBag timeline, and a new At(Drill.location)
-  token floating below}
+\comment{Use previous fig and modify as appropriately: Add figure
+  showing the effect of activating the token by putting it on the
+  AgentBag timeline, and a new At(Drill.location) token floating
+  below}
 
@@
 
-\comment{Add figure showing possible state transitions}
+\comment{Maybe: Add figure showing possible state transitions. Need a FSM.}
 
@@
 
-  \comment{ Add figure showing internal and external relationships
-    from rules on action Go}
+  \comment{Use previous fig and modify as appropriately (used as
+    above): Add figure showing internal and external relationships
+    from rules on action Go. Need a fig from JB}
 
@@
 plan in order for the master to be valid.
-
-\comment {talk about condition/effect annotations?}
 

### 2012-02-17 trunk/setobook.tex
moved concepts section after europa section
updates
europ+concepts touch ups with Javier, fig for arch and updates to arch

@@
 \documentclass[graybox,12pt]{svmult}
-
-% \documentclass[12pt]{svmono}
 
@@
 
-% \usepackage[agsm]{harvard}
-
-\titlerunning{Towards Deliberative Control of Autonomous Robots}
-\authorrunning{K.Rajan et. al.}
+% \titlerunning{Towards Deliberative Control of Autonomous Robots}
+% \authorrunning{K.Rajan et. al.}
 
@@
 \institute{Monterey Bay Aquarium Research Institute, California
-  \texttt{\{Kanna.Rajan,fpy\}@mbari.org}
+  \emph{\{Kanna.Rajan,fpy\}@mbari.org}
   \and
   NASA Ames Research Center, Moffett Field, California
-  \texttt{javier.barreiro@nasa.gov}}
+  \emph{javier.barreiro@nasa.gov}}
 
@@
   hardware becomes more robust allowing persistent ocean sampling and
-  observation. \rx is in routine operational use at MBARI,
-  providing scientists a new tool to sample and observe the dynamic
-  coastal ocean.}
+  observation. \rx is in routine operational use at MBARI, providing
+  scientists a new tool to sample and observe the dynamic coastal
+  ocean.}
 
@@
 
+\input{europa-basics.tex}
+
 \input{concepts.tex}
-
-\input{europa-basics.tex}
 
@@
 Authors Rajan and Py are funded by a block grant from the David and
-Lucile Packard Foundation to MBARI.
+Lucile Packard Foundation to MBARI. Rajan and Py would like to thank
+their collaborators at MBARI, J. Ryan, J. Harvey, R. Vrijenhoek,
+J. Bellingham. They also thank colleagues at NASA Ames who made \eu
+available as open source; specifically M. del Alto, D. Korsmeyer and
+S. Zornetzer. We are indebted to N. Muscettola for his ideas and early
+work at NASA Ames which forms the basis of this work.
 

### 2012-02-18 arch.tex
updates to arch with few comments

@@
 
-\item \texttt{synchronize} takes the last {\em external} observations
-  as an argument and returns the {\em internal} observations for a
-  reactor or a failure report.
+\item \texttt{synchronize} takes the last set of {\em external}
+  observations as an argument and returns the {\em internal}
+  observations for a reactor or a failure report.
 
@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation. In the context of \eu for example one deliberation
+  deliberation. In the context of \eu for example, one deliberation
   step corresponds to a single flaw resolution (Section
   \ref{sec:europa:arch}) while ensuring that the resulting plan is not
-  proven inconsistent. This implies that it also includes the
-  backtracking of the search in such a case.  The returned value by
-  the \texttt{step} function is a boolean which indicates whether the
-  reactor needs an extra step to deliberate or if this last step
-  resulted in a complete plan within the reactor's look-ahead. On the
-  later, the part of the plan that concern this reactor {\em external}
-  timelines is provided in order to be dispatched as goals to
-  whichever reactor that is reponsible for these. % \comment{By
+  inconsistent. Backtracking is included in the step should it be such
+  a case.  The returned value by the \texttt{step} function is a
+  boolean which indicates whether the reactor needs an extra step to
+  deliberate or if this last step resulted in a complete plan within
+  the reactor's look-ahead. The part of the plan that covers the
+  reactor's {\em external} timelines is provided to be dispatched as a
+  goal.
+% to whichever reactor that is responsible for these. % \comment{By
     % extending I may have saud things I developed below ... will have
     % to check}
+
 \end{itemize}
 
-Both these deliberative functions have different scope and temporal
-constraints. The \texttt{synchronize} call is considered as atomic and
-will focus only on state inference for the current tick by identifying
-the current value of {\em internal} state variables from the latest
-updates on the {\em external} state variables. This process propagates
-up in the reactor dependency graph at every tick to ensure that all
-reactors are aware of the current state of the world. \texttt{step}
-relates to a single decision step of the planning process which is
-focused on the future evolution of state variables; it can take
-multiple steps spanning multiple ticks for a reactor to produce a plan
-that is both complete and valid. How the reactor identifies that its
-plan is complete and valid is its responsibility and depend on the way
-such reactor resolve its plan during each {\em steps}. The general
-idea is that the reactor was able to find a valid strategy tin order
-to satisfy as much of its {\em internal} goals as possible. In the
-context of \eu, which we are going to develop, this means that the
-solver do not identify any more flaw to be resolved in the plan.
-
- When such plan is identified, its {\em external}
-timeline(s) are then dispatched to the corresponding reactors which
-own these times and which in turn can start their own deliberation
-\texttt{step}s to satisfy these new objectives. Consequently execution
-occurs as these goals and plans propagate down in the reactor
-hierarchy until they eventually become goals for the lowest level
-reactor which -- instead of deliberating -- may just transform this
-goal into a command for execution in hardware.
+Both these deliberative \kcomment{``deliberative??''; something else
+  or remove altogether perhaps?} functions have different scope and
+temporal constraints. The \texttt{synchronize} call is considered as
+atomic and will focus only on state inference for the current tick by
+identifying the current value of {\em internal} state variables from
+the latest updates on {\em external} state variables. This process
+propagates up in the reactor dependency graph at every tick to ensure
+that all reactors are aware of the current state of the
+world. \texttt{step} relates to a single decision step of the planning
+process focused on the future evolution of state variables; it can
+take multiple steps spanning multiple ticks for a reactor to produce a
+plan that is both complete and valid. How the reactor identifies that
+its plan is complete and valid is its responsibility and depends on
+the way each reactor resolves its plan at each {\em step}. The general
+idea is that the reactor is able to find a valid strategy in order to
+satisfy as many of its {\em internal} goals as possible. For \eu, this
+means that the solver is unable to identify any further flaws to be
+resolved in the plan.
+
+When such plan is identified, its {\em external} timeline(s) are then
+dispatched to the corresponding reactors which own these times and
+which in turn can start their own deliberation \texttt{step}s to
+satisfy these new objectives. Consequently execution occurs as these
+goals and plans propagate down in the reactor hierarchy until they
+eventually become goals for the lowest level reactor which -- instead
+of deliberating -- may just transform this goal into a command for
+execution in hardware.
 
@@
 automated planning. For example in Fig. \ref{fig:agent}, apart from
-\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces to the
-system, every other reactor (in blue) are different instances of a \eu
-reactor differentiated only by their model and \eu solver
-configurations.
+\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces, every
+other reactor (in blue) is a different instance of an \eu reactor
+differentiated only by their model and \eu solver configurations.
 
@@
 consists of two deliberation processes which are implemented as two
-\eu solvers modifying the \emph{same} plan database. %  and for which the
-% execution is managed by the \rx framework as time advance during the
-% execution of the system:
+separate \eu solvers modifying the \emph{same} plan database.
 
@@
 \item The \emph{synchronization solver} is a specialized \eu solver
-  that integrates new information in the plan, the evolution of
-  \emph{external} state variables and ensures that the reactor
-  propagates these to identify the reactors current state as well as
-  to inform other reactors of any state change on its \emph{internal}
-  timelines. This solver is summoned at the beginning of every single
-  tick.
+  that integrates new information in the plan. This signifies the
+  evolution of \emph{external} state variables and ensures that the
+  reactor propagates these to identify current reactor state as well
+  as to inform other reactors of any state change on its
+  \emph{internal} timelines. This solver is summoned at the beginning
+  of every single tick.
 
 \item The \emph{deliberation solver} manages the deliberation process
-  of the reactor either to produce a new plan or alter its current
-  plan as new goals are provided or when the synchronization solver
-  identifies a conflict between current state and expectations from a
-  previously generated plan. This process can span multiple ticks and
-  can therefore be interrupted by the synchronization solver.
+  of the reactor. It does so either to produce a new plan, to alter
+  its current plan as new goals are provided or when the
+  synchronization solver identifies a conflict between current state
+  and expectations from a previously generated plan. This process can
+  span multiple ticks and can therefore be interrupted by the
+  synchronization solver.
 
@@
 the plan. Conversely, when synchronization occurs after deliberation
-steps, its starting point is the last partial plan resulting from
-deliberation. This gives an initial frame to the problem that helps
--- at least initially -- to target the synchronization problem by
-ensuring that the new {\em external} state updates are compatible with
-this plan. If so, the synchronization propagates these events in the
-plan and can uses it as a help on identifying its {\em internal}
-timelines based on this plan. If they are not compatibles, then
-{\em synchronization} identifies a plan inconsistency  reflecting that the
-current plan is not valid anymore. This results on the retraction of
-decisions made during {\em deliberation} steps in order to allow 
-{\em synchronization} to identify its {\em internal}  state. Such
-relaxation can be viewed as a backtracking on the plan search
-triggered exogenously by the world evolution. It is the way our
-reactor will identify that its current plan failed to execute and can
-trigger further deliberations steps in order to find an alternate 
+stops, its starting point is the last partial plan resulting from
+deliberation. This gives an initial frame to the problem that helps to
+target the synchronization problem by ensuring that the new {\em
+  external} state updates are compatible with this plan. If so,
+synchronization propagates these events in the plan and can uses it as
+an aid in identifying its {\em internal} timelines \kcomment{``use it
+  as an aid to identify'' what though?} based on this plan.
+
+If they are not compatible, then {\em synchronization} identifies a
+plan inconsistency. This results in the retraction of decisions made
+during the \kcomment{previous?} {\em deliberation} step. % in order to
+% allow {\em synchronization} to identify its {\em internal} state.
+Such a relaxation can be viewed as a backtracking % on the plan search
+% triggered exogenously by the world evolution. It is the way our
+allowing the reactor to identify that its current plan has failed to
+execute triggering further deliberations steps to find an alternate
 strategy.
 
-Finally, when the deliberation process did identify a complete
-solution plan for its current {\em internal} goals. The part of its
-plan that applies to its {\em external} timelines is sent as a 
-request to \rx that will handle their dispatching to the corresponding 
-reactors. If things go smoothly the {\em synchronization} process will
-eventually receive {\em external } observations that confirm the
-execution of this plan and will allow the plan execution to progress
-inside the reactor (by having the part of the plan observed restricted
-to the start time of these observations).
+Finally, when the deliberation process has identified a complete
+solution for its current {\em internal} goals, a part of its plan that
+applies to its {\em external} timelines is sent as a request to \rx to
+handle their dispatch to corresponding reactors.  Synchronization
+eventually receives {\em external } observations that confirms plan
+execution, in turn recording execution progress by having the part of
+the plan observed restricted to observation start times.
 
 Synchronization offers an important challenge for embedding planning
-in a situated agent, typically %  as always been to go around the asumptions made by
-% most classical planners. Specifically one
-an assumption to restrict the problem to ``offline planning'' defined
-in \cite{ghallab04} as follows:
-
-\begin{quotation}
+in a situated agent, typically an assumption to restrict the problem
+to ``offline planning'' defined in \cite{ghallab04}:
+
+{\scriptsize
+  \begin{quote}
     The planner is not concerned with any change that may occur in
-  $\Sigma$ \footnote{The world modeled by the plan domain.} while it is
-  planning; it plans for the given initial and goal states regardless
-  of the current dynamics, if any. 
-\end{quotation}
-
-This assumption, while helpful to reduce the scope of the planning
-problem, becomes problematic for planning within a highly dynamic
-environment. In a typical robot this implies that the only point where
-the planner can be informed about the current world state is prior to
-the commencement of a planning cycle by giving to the planner the
-initial state and current objectives. During the planning, it is
-assumed that the agent can maintin the world as perceived by the
-planner as stable. \cite{lemai04, lemai-chenevier2004} attempts to
-reduce the impact of planning by allowing local plan repair in the
-existing plan. In the eventuality that such repair is not feasible
-they require that their robot comes to a full stop until (re-)planning
-is complete.% \comment{I may need more examples/refs here}. 
-As Fig. \ref{fig:tick-exec} shows, this assumption is problematic for our
-architecture. Synchronization that tracks state evolution can occur
-several times during the reactor deliberation process. By making
-synchronization and deliberation share the same plan we provide a
-solution that relaxes this static world assumption. According to 
-\cite{Dias:2003ua, mus06}, IDEA was sharing the same core idea 
-where all the solvers of an agent shared the same plan as a basis of
-communication and specifically their {\em Reactive planner} --roughly
-similar to our {\em synchronization} solver -- was often the only one 
-targeting the execution frontier. Although, none of the papers we
-found were detailing how this was handled in the
-architecture. Moreover, in \cite{Dias:2003ua} we can read:
-\begin{quotation}
+    $\Sigma$ \footnote{The world modeled by the plan domain.} while it
+    is planning; it plans for the given initial and goal states
+    regardless of the current dynamics, if any.
+\end{quote}}
+
+This assumption, while helpful to reduce the scope and complexity of
+the planning problem, becomes problematic for planning within a highly
+dynamic environment. In a typical robot this implies that the only
+point where the planner can be informed about the current world state
+is prior to the commencement of a planning cycle, by providing the
+planner an initial state and current objectives. During the planning,
+it is assumed that the agent can maintain the world as perceived by
+the planner. \cite{lemai04, lemai-chenevier2004} attempts to reduce
+the impact of planning by allowing local plan repair in the existing
+plan. In the eventuality that such repair is not feasible they require
+that their robot comes to a stop until (re-)planning is complete.
+
+This assumption is problematic for situated agents in general and as
+Fig. \ref{fig:tick-exec} shows, for \rx. Synchronization that tracks
+state evolution can occur several times during reactor
+deliberation. By making synchronization and deliberation share the
+same plan we provide a solution that relaxes this static world
+assumption. \texttt{IDEA} uses the same principle with all solvers in
+an agent sharing the same plan for communicating environmental change
+\cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
+-- similar to our {\em synchronization} solver -- was often the solver
+targeting the change at the execution frontier. 
+% Although, none of the papers we found were detailing how this was
+% handled in the architecture.
+\cite{Dias:2003ua} states:
+
+{\scriptsize
+  \begin{quote}
   [\textellipsis] as long as the planning horizon over which the
@@
   reactive execution and does not affect the reactivity of the agent. 
-\end{quotation}
-\comment{fpy: This is my interpretation of this sentence at least .... feel
-  free to comment in case you disagree}
-Which suggest a non overlap between {\em reactive planner} and {\em
-  deliberative planners} horizons. In \rx, and more specifically in
-our \eu reactor the planning horizon of {\em synchronization} -- 
-limited to the current tick -- is a subset of the planning horizon 
-of the deliberation steps -- starting at the current tick and with an
-extend based on the latency $\lambda$ and look-ahead $\pi$ of 
-the reactor. The way we avoid to create conflict between the two is by
-a careful control of the scheduling of these two process by the \rx 
-agent.
+\end{quote}}
+
+This suggests a deliberate non overlap between {\em reactive planner}
+and {\em deliberative planners} horizons. In \rx, and more
+specifically in our \eu reactor the planning horizon of {\em
+  synchronization} -- limited to the current tick -- is a subset of
+the planning horizon of the deliberation steps -- starting at the
+current tick and with an extent based on latency $\lambda$ and
+look-ahead $\pi$ of the reactor. In this manner we avoid creating a
+conflict between the deliberation and synchronization by controlling
+the scheduling of these two process by the \rx agent.
 
@@
 are implemented independently to show how their interaction through
-the plan helps relax this assumption by making planning fully
-integrated in the reactor execution cycle while potentially allowing
-for more informed planning.
+the plan helps relax this assumption. It does so by integrating
+planning in the reactor execution cycle even while allowing for more
+informed deliberation.
 

### 2012-02-19 arch.tex
updates to arch with embedded comments for FP

@@
 
-\subsubsection{Synchronization identify internal state evolution}
+\subsubsection{Synchronization and Identifying Internal State
+  Evolution}
 \label{sec:arch:synch}
@@
 We first show how a reactor can track and identify its state in the
-context where it does not have any compelling need to deliberate. We
-assume that we have a reactor that has no future goal. Even in this
-condition, \rx requires each reactor to produce at every tick its {\em
-  internal} state. For those reactors not owning {\em internal}
-timelines, \rx makes sure that {\em external} observations are taken
-into account for every tick as shown in \cite{py10}. Such a
-requirement is enforced to ensure that all reactors share a consistent
-view of the world with advancing time.
+context where it does not have any compelling need to deliberate. In
+doing so we assume that we have a reactor that has no future
+goal. Under such a condition, \rx requires each reactor to produce at
+every tick its {\em internal} state. For those reactors not owning
+{\em internal} timelines, \rx makes sure that {\em external}
+observations are taken into account for every tick as shown in
+\cite{py10}. Such a requirement is enforced to ensure that all
+reactors share a consistent view of the world with advancing time.
 
 {\em External} state updates are managed by introducing these
-observation in the plan structure as facts and ensure that they remain
-compatible with the current plan mainainted by the reactor. The
+observations in the plan structure as facts and ensuring that they
+remain compatible with the current plan mainainted by the reactor. The
 challenging part of synchronization is to decide the {\em internal}
-state values for the current tick. In the Shopping example for example
+state values for the current tick. In the Shopping example for example,
 we have the following code snippet which helps identify when we can
@@
 
-If \texttt{object.location} is {\em external} and the \texttt{Agent}
-is {\em internal}, a robot may encounter the situation where it is at
-the location where it can buy one (possibly among others)
-product. This results om a decision problem related to the state value
-of the \texttt{Agent} timeline where the robot can either {\em Buy} a
-specific product now, wait a little longer before doing so or just
-simply not buy any product at this location.
-
-In the \eu based reactor this decision requirement is a new type of
-flaw which forces full identification of internal state of the reactor
-for the current tick. Using this new flaw we can describe the
-synchronization process as:
-
-\begin{enumerate}
-
-\item Integrate the external state as provided by the owner of each
-  external timeline into the plan database.
-
-\item Propagate this information in the plan database following the
-  model $\mathcal{M}$ of this reactor.
-
-\item Resolve the current state value of each internal timelines.
-
-\end{enumerate}
+If \texttt{object.location} is {\em external} and \texttt{Agent} is an
+{\em internal} timeline, a robot may encounter the situation where it
+is at the location where it can buy one (possibly among others)
+product. This results in a decision-making problem related to the
+state value of the \texttt{Agent} timeline; the robot can either {\em
+  Buy} a specific product now, wait a little longer before doing so or
+just simply not buy any product at this location.
+
+In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
+new type of flaw which forces identification \kcomment{identification
+  of what? Probably need a little more explanation here.} of internal
+state of the reactor for the current tick. This flaw can be defined
+as:
+
+\begin{definition}
+  \label{def:csf}
+  A {\em current state flaw} indicates that a reactor's {\em internal}
+  timeline does not provide a fully defined state value for the
+  execution frontier $\tau$. Such a flaw can be the result of either:
+
+  \begin{itemize}
+  \item The absence of a token in this timeline which overlaps the
+    execution frontier $\tau$.
+  \item An open decision between terminating the last token for this
+    timeline or extending its end in the future.
+  \end{itemize}
+
+\end{definition}
 
@@
   \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
-  \caption{\small Illustration of the synchronization flaws in a reactor. The
-    reactor receive new observations when they are produced by the
-    owner(s) of its internal timelines. The line after the last token
-    of each timeline represent the domain of possible values for the
-    end of this token. At every tick $\tau$ the reactor needs to
-    integrate the {\em External} state information it received and --
-    based on its model $\mathcal{M}$ -- resolve its {\em Internal}
-    state that will then be provided by the architecture to other
-    reactors using these state variables.}
+  \caption{\small Illustration of synchronization flaws in a
+    reactor. The reactor receives new observations when they are
+    produced by the owner(s) of its internal timelines. The line
+    \kcomment{what line???} after the last token of each timeline
+    represent the domain of possible values for the end of this
+    token. At every tick $\tau$ the reactor needs to integrate the
+    {\em External} state information it received and -- based on its
+    model $\mathcal{M}$ -- resolve its {\em Internal} state that will
+    then be provided by the agent to other reactors using these state
+    variables.}
   \label{fig:synch:flaw}
@@
 
-We introduce this requirement as a new type of flaw for the \eu
-framework defined as follow:
-
-\begin{definition}
-  \label{def:csf}
-  A {\em current state flaw} indicates that a reactor {\em internal}
-  timeline does not provide a fully defined state value for the
-  execution frontier $\tau$. Such a flaw can be result of either of
-  the following reasons:
-
-  \begin{itemize}
-  \item The absence of a token in this timeline which overlaps the
-    instant $\tau$.
-  \item An open decision between terminating the last token for this
-    timeline or pushing its end in the future.
-  \end{itemize}
-
-\end{definition}
-
-The core idea between this flaw is to ensure that at any single tick
-the {\em synchronization} solver will identify all the {\em internal}
+The key idea is to ensure that at any single tick the {\em
+  synchronization} solver will identify all the {\em internal}
 timelines for which the current state value is yet to be
 identified. We handle their resolution based on the following
-proposition.
+proposition:
 
@@
   A {\em current state flaw} can be resolved using one of the
-  following choices which are evaluated in the following order:
+  following choices which are evaluated in order:
 
@@
   \item Extend the previous state value to end after the execution
-    frontier  (\ie restrict its end time to $[\tau+1, \infty)$). %\comment{first use of
-    % $\tau$, so needs to be defined earlier}
+    frontier (\ie restrict its end time to $[\tau+1, \infty)$).
 
@@
   \item Create and insert a new token in this timeline that will start
-    at the current tick $\tau$ (attempt this for each possible token
-    type for this timeline if necessary).
+    at the current tick $\tau$ and attempt this for each possible
+    token type for this timeline if necessary.
 
@@
 \end{proposition}
-Intuitively the first choice enforces an intertial value assumption
-where the reactor will tend to maintain its current as long as the
-model and external observations allow it. The second choice favorise
-the advance of the reactor current plan if any; this direct the {\em
-  synchronization} decision toward assuming that its current plan is
-properly executed. Finally if none of these two choices did provide a
-valid solution -- meaning that both lead to an inconsistant plan --
-the reactor tries all the possible state values remaining for this
-timeline until it finds one that allows it to identify its state as
-required by \rx architecture.
-
-This flaw forces the reactor to identify its internal state
-for the current tick. By using this flaw one can describe the
-synchronization process as:
+
+Intuitively the first choice enforces an \emph{inertial value
+  assumption} where the reactor will tend to maintain its current
+state as long as the model and external observations allow for it. The
+second choice favors the advance of the reactor's current plan if any;
+this directs the {\em synchronization} decision toward assuming that
+its current plan is properly executed. Finally if none of these two
+choices provides a valid solution -- \ie lead to an inconsistant state
+-- the reactor tries all the possible state values remaining for the
+timeline until it finds one that brings plan consistency.
+
+% allows it to identify its state as required by \rx architecture.
+
+% The flaw forces the reactor to identify its internal state for the
+% current tick. By using this flaw one can describe the synchronization
+% process as:
+
+The synchronization process in turn can be described as:
 
 \begin{enumerate}
+
 \item Integrate the external state as provided by the owner of each
   external timeline into the plan database.
-\item Propagate this information in the plan database following the
-  model $\mathcal{M}$ of this reactor.
+
+\item Propagate this information in the plan database using model
+  $\mathcal{M}$ of this reactor.
+
 \item Resolve the current state value of each internal timelines based
   on Proposition \ref{prop:csf:resolve}.
+
 \end{enumerate}
+
+% \begin{enumerate}
+% \item Integrate the external state as provided by the owner of each
+%   external timeline into the plan database.
+% \item Propagate this information in the plan database following the
+%   model $\mathcal{M}$ of this reactor.
+% \end{enumerate}
 
@@
 
-Steps $2$ \& $3$ are handled by the \eu solver from algorithm
-\ref{alg:europa:solve}. We only enforce that the {\em current state
-  flaws} are selected last by the algorithm in order to follow the
-sequencing given above. As {\em synchronization} focusses on the
-execution frontier wee also filter out all flaws that are outside of
-the current tick scope. This mean that if this flaw relates to a token
-that is {\em necessarily} ending before $\tau$ or starting after
-$\tau+1$ then the solver will ignore it. This helps to reduce the
-scope of the synchronization to exclusively the current state of the
-reactor and reduce dramatically the search space for the {\em
-  synchronization} solver. In order to do so we need to assume 
-that the current state value of an {\em internal} timeline does not 
-depend on the future or more accurately that any choices made during 
-this {\em synchronization} will not lead to a future inconsistency. 
-Such an assumption impacts the set of possible domains the reactor 
-can support while remaining complete. Take for  example this rule from 
-our \texttt{Shopping} model :
+Steps $2$ \& $3$ above are handled by the \eu solver from algorithm
+\ref{alg:europa:solve} which ensures that the {\em current state
+  flaws} are selected last so as to follow the sequence given
+above.
+
+As {\em synchronization} focuses on the execution frontier it also
+filters out all flaws that are outside of the current tick's scope.
+If the flaw correlates to a token that is {\em necessarily} ending
+before $\tau$ or starting after $\tau+1$ then the solver ignores
+it. Such a strategy helps to reduce the scope of synchronization
+exclusively to the current state of the reactor and in turn reduces
+the search space for the {\em synchronization} solver.  To do so we
+need to assume that the current state value of an {\em internal}
+timeline does not depend on the future or more accurately that any
+choice made during this {\em synchronization} will not lead to a
+future inconsistency.  Such an assumption impacts the set of possible
+domains the reactor can support while remaining complete. Take for
+example this rule from our \texttt{Shopping} model:
 
@@
 problematic when execution is taken into account especially since it
-puts a strong emphasis between an evolving plan and an unknown future
-(\texttt{Going} to a location and the {\em expected} future outcome of
+puts a strong emphasis between an evolving plan and an unknown future;
+\texttt{Going} to a location and the {\em expected} future outcome of
 ending at this location, ties it strongly to the constraint in line
-$5$ above).  So for example, assume that the \texttt{Agent.location}
-is {\em external} to the reactor which receives the observation
+$5$ above. For instance, assume that the \texttt{Agent.location} is
+{\em external} to the reactor which receives the observation
 \texttt{Going(Home,SuperMarket)} which synchronization has resolved by
@@
 \texttt{Agent} timeline. This would imply that the only possible
-outcome on \texttt{Agent.location} is to be \texttt{At(SuperMarket)}
-in the directly foreseeable future. In the real world the fact that we
-are \texttt{Going} to a certain location does not necessarly imply
-that we will succeed on ending up there. In one extreme case due to an
-exogenous event, the road to the \texttt{SuperMarket} is closed which
-could result in an alternate observation that the robot is
-\texttt{At(7th st. \& Main)}. This observation contradicts the
-existing plan the reactor is maintaining. Further, it is important to
+outcome on the \texttt{Agent.location} timeline, is to be
+\texttt{At(SuperMarket)} in the directly foreseeable future.
+
+In the real world the fact that the robot is \texttt{Going} to a
+certain location does not necessarly imply that it will succeed in
+ending up there. Due to an exogenous event, for example, the road to
+the \texttt{SuperMarket} could be closed, which could result in an
+alternate observation that the robot is \texttt{At(7th st. \& Main)}
+contradicting the existing plan. Further, it is important to
 understand the temporal direction of the \emph{causal link} that is
-triggering the inconsistency. At the time the robot observes this
-contradicting the \texttt{At} predicate, the rule that generated this
-conflict is connected to past observations \texttt{Going} and
-\texttt{Go} which cannot be easily retracted without a potential
-global impact to the plan implying a huge cost in back-propagation
-that would be problematic.
+triggering the inconsistency. At the time the robot observes the
+contradiction of the \texttt{At} predicate, the rule that generated
+this conflict is connected to past observations \texttt{Going} and
+\texttt{Go}. These cannot be retracted easily without a potential
+impact to the plan implying in terms of a cost in back-propagation.
 
@@
 if one can guarantee that an outcome is predisposed to occur (for
-example the reactor managing \texttt{Agent.location} timeline
+example the reactor managing the \texttt{Agent.location} timeline
 guarantees that the outcome of the \texttt{Going} will {\em always}
@@
 nature of the planning problem and the ability to search outcomes and
-replan, the model should be written so that a future projection can be
-manipulated. %  while keeping in mind that predicting the future is not
-% absolute. 
-In this case one could alter this part of the model in order to give
-the reactor more flexibility as follows:
+to replan, the model should be written so that a future projection can
+be manipulated.
+
+One could therefore alter this part of the model to give the reactor
+more flexibility as follows:
 
@@
 at line $11$ which ensures that the last \texttt{Going} ends up in the
-location the agent actually wants to be at. This first level of
-relaxation will therefore allow the agent to use multiple
+location the agent actually wants the robot to be at. This first level
+of relaxation will therefore allow the agent to use multiple
 \texttt{Going} tokens should the first one fail to reach the
-objective.  % The problem can even be relaxed further if needed but
-% this relaxation gives already more flexibility for the reactor to
-% recover from local failures.
+objective.
+
 As the simple example above shows, synchronization involves a
 deliberation process with the risk that the search for a solution is
-not a straightforward sequence of unit decisions. Since such a process
-is required by the architecture at \emph{every} tick for \emph{every}
-reactor the model needs to be carefully designed.
+not a straightforward sequence of unit decisions \kcomment{has a unit
+  decision been explained in Europa?}. Since such a process is
+required by the architecture at \emph{every} tick for \emph{every}
+reactor the model needs to be designed carefully.
 
@@
 rely on its {\em internal} state. For this reason a failure to
-synchronize immediately results in the agent killing a reactor and
+synchronize immediately results in the \rx agent killing a reactor and
 notifying any reactor that depends on its {\em internal} state
 variables; it does so by posting the special observation
-\textsf{Failed}. And by doing so we allow for graceful degradation
-where other reactors can readapt their plans as they receive this
-observation.
-
-Finally, at the engine level careful choices were made in 
-both the focus of the search -- limited only to tokens that 
-can potentially overlap the current tick -- and heuristics where 
-the {\em current state flaw} is considered last during the search
-process. The later limits the amount of backtracking for a solution 
--- or at least how deep one needs to backtrack in a decision tree. 
-Similarly the sequence in which we evaluate the possible resolution 
-for such a flaw -- presented in Proposition \ref{prop:csf:resolve} -- 
-was selected according to  some simple assumptions that any external 
-observation that is compatible with an existing token in the current
-plan reflects the execution of the plan itself. In other words this
-means that our solver always prefer to use tokens already existing in
-the plan in order to identify its current internal state. By doing so
-we directs the synchronization search by exploiting the structure of
-the plan produced during previous deliberation steps. This implies
-that a deliberation steps have been done and did impact the plan
-structure and calls for a description on how our \eu reactor resolve
-such steps which we are now going to describe further.
+\textsf{Failed}. By so doing we allow for graceful degradation where
+other reactors can readapt their plans on receipt of this observation.
+
+Finally, at the engine level careful choices were made in both the
+focus of the search -- limited only to tokens that can potentially
+overlap the current tick -- and heuristics where the {\em current
+  state flaw} is considered last during the search process. The later
+limits the amount of backtracking for a solution -- or at least how
+deep one needs to backtrack in a decision tree.  Similarly the
+sequence in which we evaluate the possible resolution for such a flaw
+-- presented in Proposition \ref{prop:csf:resolve} -- was selected
+according to some simple assumptions, that any external observation
+that is compatible with an existing token in the current plan reflects
+the execution of the plan itself. In other words this means that our
+solver always prefers to use tokens \emph{already existing} in the
+plan as a way to identify current internal state. By doing so
+synchronization exploits the structure of the plan produced during
+previous deliberation steps.
+
+% This implies that a deliberation steps
+% have been done and did impact the plan structure and calls for a
+% description on how our \eu reactor resolve such steps which we are now
+% going to describe further.
 

### 2012-02-19 arch.tex
further updates to arch

@@
 
-\subsubsection{Synchronization and Identifying Internal State
-  Evolution}
+\subsubsection{Synchronization and Identifying Internal State}
 \label{sec:arch:synch}
@@
 Deliberation inside a reactor is based on \eu capabilities described
-in Section \ref{sec:europa:arch}. However deliberation within \rx
-comes with some augmentation -- deliberation information one reactor
-provide to the architecture such as its deliberation latency and its
-plan look-ahead both expressed in ticks -- and the fact that we
-potentially need to interleave deliberation with not just the
-synchronization process but also with deliberation of other
-reactors\footnote{The current implementation of \rx is running on a
-  single process and it is the responsibility of the agent itself to
-  emulate reactor multi-threading for deliberation and
-  synchronization.}. For this reason while the overall planning of a
-deliberative reactor strictly relies on the \eu framework planning
-solver, it slices the execution of Algorithm \ref{alg:europa:solve} in
-atomic steps. The interruption is done around the recursive call (line
-\ref{li:recurse}) while ensuring that the current partial plan is
-never proven inconsistant at any single step. This can be managed by
-allowing the call to backtrack until a \textsf{DecisionPoInt} that is
-not exhausted (\ie $decision.hasNext()$ is \texttt{true}) is found in
-the call stack.
+in Section \ref{sec:europa:arch}. However this deliberation comes with
+some augmentation namely the reactor's deliberation latency and its
+plan look-ahead both expressed in ticks and that the agent needs to
+interleave deliberation with not just the synchronization but also
+with deliberation of other reactors\footnote{The current
+  implementation of \rx is running on a single process and it is the
+  responsibility of the agent itself to emulate reactor
+  multi-threading for deliberation and synchronization.}.  For this
+reason while the reactor relies on the \eu solver, it slices the
+execution of Algorithm \ref{alg:europa:solve} into atomic steps. This
+interruption occurs around the recursive call line \ref{li:recurse}
+while ensuring that the current partial plan is not inconsistant at
+any single step. This is managed by allowing the call to backtrack
+until a \textsf{DecisionPoInt} that is not exhausted (\ie
+$decision.hasNext()$ is \texttt{true}) is found in the call stack.
 
@@
 there is no other alternative solution. We do this by removing all
-goals and keeping only recent observations. % These constraints are yet
-% again related to the fact that for the agent the only crucial
-% functionality of a reactor is to resolve successfully its
-% synchronization.
-
-{\em Synchronization} and the dispatching of new {\em goals} from
-\rx can occur between two steps and alter the plan. These alterations
-exogenous to the deliberation process are dynamically impacting the set
-of flaws that need to be resolved. Obviously new flaws are created
-when a new goal is received from \rx and conversely often the recall
-of an objective will remove the corresponding flaws\footnote{although
-  if the flaws of this goal were previously resolved, the removal of
-  this goal may create new flaws}. The impact of {\em synchronization}
-can be more subtle as its alteration of the plan at the execution
-frontier is much more drastic. We can still give a non exhaustive list
-of possible alterations of the flaws directly related to {\em
-  synchronization}:
+goals while keeping only recent observations.
+
+{\em Synchronization} and the dispatching of new {\em goals} from \rx
+can occur between two steps and alter the plan. These alterations
+exogenous to the deliberation process dynamically impact the set of
+flaws that need to be resolved. New flaws are created when a new goal
+is received from \rx and conversely the recall of an objective could
+remove the corresponding flaws\footnote{However if the flaws related
+  to this goal were resolved, the removal of this goal may likely
+  create new flaws.}. The impact of {\em synchronization} can be more
+subtle as its alteration of the plan at the execution frontier is more
+drastic. We can still give a non exhaustive list of possible
+alterations of the flaws directly related to {\em synchronization}:
+
 \begin{enumerate}
+
 \item Previous flaws directly related to the execution frontier have
   been resolved by the {\em synchronization} solver.
+
 \item The {\em synchronization} solver may have created new flaws that
-  were outside of its scope but are now part of the {\em deliberation}
-  solver scope.
-\item Last observations proved the current plan to be invalid
+  were outside of its scope but are now part of the scope of the {\em
+    deliberation} solver.
+
+\item The last observation has proved that the current plan is invalid
   requiring the {\em synchronization} solver to relax the existing
-  plan in order to identify its current {\em external} state. (This
-  case is developed further on the next section)
+  plan to identify its current {\em external} state. (This case is
+  developed further on the next section).
+
 \end{enumerate}
-Finally as {\em synchronization} occurs when the {\em
-  execution frontier} advances, it implies that the reactor planning
-scope is further in the future. This need to be reflected by the {\em
-  deliberation} solver which need a new deliberation scope one tick
-further in the future.
 
@@
   \includegraphics[width=0.65\columnwidth]{figs/plan_window}
-  \caption{\small The planning window of the deliberation solver and its
-    relation with the {\em dispatch window}. The dark grey
-    tokens   with dashed borders are excluded from deliberation as they do not
+  \caption{\small The planning window of the deliberation solver and
+    its relation to the {\em dispatch window}. The dark grey tokens
+    with dashed borders are excluded from deliberation as they do not
     overlap the interval $[\tau, \tau+\lambda+\pi]$. All the other
-    tokens (in {\color{green} green}) -- along with the goals received from the
-    agent (in {\color{blue} blue}) which can overlap this window are
+    tokens ({\color{green} green}) along with the goals received from
+    the agent ({\color{blue} blue}) which can overlap this window, are
     evaluated.}
@@
 
-As Fig. \ref{fig:plan:window} illustrates,  such a
-scope allows filtering of tokens that are either necessarily ending
+When the {\em execution frontier} advances, the reactor planning scope
+moves further into the future. This is reflected by the {\em
+  deliberation} solver which uses a new deliberation scope one tick
+further in the future. As Fig. \ref{fig:plan:window} illustrates, such
+a scope allows filtering of tokens that are either necessarily ending
 before $\tau$ or necessarily starting after, focusing the planning
-problem on only those tokens that are within a ``reasonable''
-future. This filtering window complements the \rx {\em dispatch
-  window} from definition \ref{def:dispatch} which already filters
-out at the agent level the goals that are before the reactor execution
-latency or too far in the future. The planning window inside of the
-reactor refines further this filtering as tokens can be excluded
-dynamically from deliberation as their insertion in the plan and
-scheduling with other tokens generated can reduced their temporal
-scope outside of the planning scope. This helps the deliberation 
-solver reduce the number of tokens to be evaluated and therefore
-reduce deliberation cost and performance. As time advances flaws that
-were initially beyond this temporal horizon will become active which a
-reactor can evaluate. This results in an apparent \emph{continuous
-  planning} of the reactor. When there are no more flaws present in
-the current planning scope a partial plan solution is found. This has
-two effects:
+problem on only those tokens that are within a ``reasonable'' future.
+This filtering window complements the \rx {\em dispatch
+  window} from Definition \ref{def:dispatch} which filters out at the
+agent level those goals that are before the reactor execution latency
+or too far into the future. The planning window inside of a reactor
+refines this filtering further, as tokens can be excluded dynamically
+from deliberation as their insertion in the plan and scheduling with
+other tokens generated, can reduce their temporal scope outside of the
+planning scope \kcomment{ this last sentence should be reworked}. This
+helps the deliberation solver reduce the number of tokens to be
+evaluated reducing deliberation cost and increasing performance. As
+time advances, flaws that were initially beyond this temporal horizon
+will become active which a reactor can evaluate. This results in an
+apparent \emph{continuous planning} of the reactor. When there are no
+more flaws present in the current planning scope a partial plan
+solution is found. This has two effects:
 

### 2012-02-19 arch.tex
further updates to arch

@@
 
-\subsubsection{Intertwining synchronization and execution}
+\subsubsection{Interleaving Synchronization and Execution}
 \label{sec:arch:intertwine}
 
-One key aspect of the \rx framework is intertwined plan execution. As
-seen above, it is synchronization which drives planning, so the
+A key aspect of the \rx framework is interleaved plan execution. As
+noted above, it is synchronization which drives planning, so the
 co-mingling between these two processes is critical to maintaining
-state within a robot immersed in a dynamic environment. Also important
-is that \rx maintains 
+state within a robot especially when it is immersed in a dynamic
+environment. Further, it is important that \rx maintain
 % A core aspect of this reactor is that it takes advantage of the
@@
 % fact that only 
-one $plan$ structure in the reactor and shared between synchronization
-and planning. This allows these two processes to not only be
-interleaved but how they can impact each other via plan manipulation.
-
-The most obvious case is how synchronization allow propagate of {\em
-  external} observations as a new tick occurs. In our presentation of 
-{\em synchronization}, we isolated the problem by stating
-that we will assume at this point that the reactor has no future goal
-or more accurately no plan to enforce. By doing so we
-were able to develop the way we augment the \eu solver to be able
-to propagate {\em external} observations in order to identify the
-reactor {\em internal} state for the same tick. This allowed us 
-to focus on synchronization as a state evaluation process that can
-be resolved within the \eu framework with few extensions. Similarly
-we presented deliberation at this fairly high level not necessarily
-discussing how the disruptive nature of synchronization in this
-process could impact the way this planning could evolve.
-
-We first analyze the case during which synchronization occurs after
-planning has completed (\ie the last step of deliberation results in a
-plan that is considered complete for the current look-ahead). As
-synchronization starts with a plan % produced this will gave a frame for
-% the search process. Therefore 
-this deliberation will not only be a model-derived estimation uniquely
-based on {\em external} observations but will also include reactor
-intentions specified in the available partial plan. Consider for
-example the Shopping Agent model described where the \texttt{Agent} is
-an {\em external} timeline. As the reactor adds the goal to
+one $plan$ structure shared between synchronization and planning. This
+allows these two processes to not only be interleaved for agent
+responsiveness, but ensures a viable mechanism to communicate.
+
+% but also impact
+% each other via plan manipulation.
+
+As noted earlier, synchronization propagates {\em
+  external} observations when a new tick occurs. Earlier, we isolated
+the case when a reactor has no future goal or no plan to enforce. By
+doing so we were able to develop the way we augment the \eu solver to
+propagate {\em external} observations to identify {\em
+  internal} state. This allowed us to focus on synchronization as a
+state evaluation process that can be resolved within the \eu framework
+with few extensions. Similarly we presented deliberation as an
+abstraction without discussing the disruptive nature of
+synchronization and how it impacts the planning process. We first
+analyze the case when synchronization occurs after planning has
+completed (\ie the last step of deliberation results in a plan that is
+considered complete for the current look-ahead). As synchronization
+starts with a plan, this deliberation will not only be a model-derived
+estimation uniquely based on {\em external} observations but will also
+include reactor intentions specified in the available partial plan.
+
+Consider for example the Shopping Agent model where the \texttt{Agent}
+is an {\em external} timeline. As the reactor adds a goal to
 \texttt{Own} milk and the current {\em external} timeline indicates
@@
 \texttt{At(SuperMarket)} -- are installed as goals for the reactor
-managing this state variable {\em internally}.
+managing this state variable internally.
 
@@
 agent location is able to change the location state variable to
-\texttt{Going}. \rx informs our reactor of this new observation which
-is added in the plan structure as a fact -- starting exactly at
-$\tau'$ and lasting for an unknown duration which is at least 1
-tick. As this new observation is integrated in the plan through
-synchronization, the solver can identify the similarity between this
-observation and the next token in the existing plan. Instead of
-initiating an exhaustive search of all possible implications, the
-synchronization process starts its execution with the assumption that
-these new observations are consistent with the currently maintained
-plan. As a consequence, at the end of synchronization the plan
-depicted in Fig. \ref{fig:shop:exec1} is likely. This is a result of
-propagation of information provided by the new observation in the
-existing plan while restricting the start time of \texttt{Going} and
-propagating temporal constraints, making \texttt{Go} to start {\em
-  exactly} at $\tau'$. In doing so, the \eu solver's focus on the
-execution frontier ($\tau'$ in Fig. \ref{fig:shop:exec1}) takes
-advantage of the existing plan to evaluate a solution that assumes
-that the observation is consistent with what was decided by the
-reactor.
+\texttt{Going}. \rx informs the reactor of this observation which is
+added to the partial plan as a fact -- starting exactly at $\tau'$ and
+lasting for an unknown duration which is at least one tick. As this
+new observation is integrated in the plan through synchronization, the
+solver can identify the similarity between this observation and the
+next token in the existing plan. Instead of initiating an exhaustive
+search of all possible implications, the synchronization process
+starts its execution with the assumption that these new observations
+are consistent with the currently maintained plan. As a consequence,
+at the end of synchronization the plan depicted in Fig.
+\ref{fig:shop:exec1} is likely. This is a result of propagation of
+information provided by the new observation in the existing plan while
+restricting the start time of \texttt{Going} and propagating temporal
+constraints, making \texttt{Go} to start {\em exactly} at $\tau'$. In
+doing so, the \eu solver's focus on the execution frontier ($\tau'$ in
+Fig. \ref{fig:shop:exec1}) takes advantage of the existing plan to
+evaluate a solution that assumes that the observation is consistent
+with what was decided by the reactor.
 
@@
 solver but also enforced by the heuristic order of the decision point
-resolution choices for the {\em current  state flaw} presented in
-Proposition \ref{prop:csf:resolve}. Indeed, as we stated then the two
-first choices tend to privilege a conservative view in respect to the
-existing plan. It attempts first to extend the previous observation we
-produced for this {\em internal} timeline in order to avoid to commit
-to early in the plan advance. If the formaer failed it will then
-attempt to start the next exisiting token in the plan which alow the
-reactor to advance its plan execution. Both of these two initial
-choices atetempt to resolve the plan with the least amount of
-perturbation of the plan. They just constrain further the start and
-end time of the exisiting tokens which were provided by the previous 
-deliberations steps. This means that in a nominal situation where the
-plan this reactor produced is properly executed, the synchronization
-remains limited to restrict time points in the plan and progate their
-impact. Most of the heavy dutty work of creating new tokens in the
-plan and schedule them -- both being much more difficult problems 
-to resolve -- have been managed during deliberation steps. In that
-sense we see that most of the time synchronization is limited to very
-few node in \eu search space with often no backtrack required. 
-Fig. \ref{fig:shop:exec} shows how the existing plan can support the
-synchronization process. In that case, the observation \texttt{Going}
-occuring at $\tau'$ was fully compatible with the one already present
-in our plan, by updating the plan and progating the equality
-constraints, the resolution of the {\em current state flaw} for the
-{\em agent} timeline became obvious as the token \texttt{Go} from 
-the plan presented itself as a natural solution.
-
-The synchronization may become slightly nore costfull when we need to
-identify a new token to be injected by the plan (the 3rd choice in
-Proposition \ref{prop:csf:resolve}) as it needs to explore on tis own
-all the possible token types this timeline accept as value. This is
-the reason this choice is evaluated last in our flaw resolution in
-order to avoid such complex action during synchronization unless 
-necessary. In that sense,  and thanks to the carefull choice on our
-synchronization heurisitic we see that the plan produced during \rx
-deliberation steps helps sthe synchronization process to be more
-efficient.
+resolution choices for the {\em current state flaw} shown in
+Proposition \ref{prop:csf:resolve}. As noted, the first two choices
+are biased towards a conservative view with respect to the existing
+plan. The heuristics first attempt to extend the previous observation
+for this {\em internal} timeline to avoid early commitment. If this
+fails \rx attempts to start the next existing token in the plan to
+advance plan execution. Both choices attempt to resolve state with the
+least amount of plan perturbation by constraining further the start
+and end time of existing tokens generated by previous deliberations
+steps. By doing so, in a nominal situation, synchronization remains
+limited to restrict time points in the plan and propagates their
+impact. Most of the computational effort of creating new tokens in the
+plan and scheduling them onto timelines are managed during
+deliberation steps. Consequently, most of the time synchronization is
+limited to very few nodes in the \eu search space often requiring no
+backtracking. Fig. \ref{fig:shop:exec} shows how the existing plan can
+support the synchronization process. The observation \texttt{Going}
+occuring at $\tau'$ is fully compatible with the token already present
+in our plan. By updating the plan and progating equality constraints,
+the resolution of the {\em current state flaw} for the {\em Agent}
+timeline becomes obvious as a solution, as the token \texttt{Go} from
+the plan presents itself.
+
+Synchronization may become computationally more expensive when we need
+to identify a new token to be injected into the plan (the third choice
+in Proposition \ref{prop:csf:resolve}), as it needs to explore the
+outcome of inserting all possible token types the timeline can accept.
+This is the reason this choice is evaluated last in our flaw
+resolution strategy.
+
+% In that sense, and thanks to the carefull choice on our
+% synchronization heurisitic we see that the plan produced during \rx
+% deliberation steps helps sthe synchronization process to be more
+% efficient.
 
@@
 so by specifying token parameters such as time points that are
-propagated through the plan in order to identify if the plan remains
-valid -- as in our example -- or that it cannot be executed as it
-is. The later is often identified by the synchronization failing to
-find a consistent solution. For example should the \texttt{Going}
+propagated throughout the plan to identify plan validity.
+%  -- as in our example -- or that it cannot be executed as it is. 
+When invalid, identification occurs by synchronization failing to find
+a consistent solution. For example should the \texttt{Going}
 observation received be to the \texttt{hardwareStore}, this would
-break the current partial plan in the reactor as the robot cannot
-possible be going simultaneously to $2$ different locations; see
-Fig. \ref{fig:shop:relax1}. As synchronization is not possible in such
-a situation the strategy followed is to relax all decisions made by
+break the partial plan in the reactor as the robot cannot possible be
+going simultaneously to two different locations; see Fig.
+\ref{fig:shop:relax1}. As synchronization is not possible in such a
+situation the strategy followed is to relax all decisions made by
 previous plan steps. This is done by keeping all past observations in
@@
 existing {\em internal} goals (\texttt{Own(Milk)} in this example)
-received by this reactor. The resulting partial plan is shown in
-Fig. \ref{fig:shop:relax2}. In this process the reactor informs the
-agent that all {\em external} goals it requested for the future are no
-longer valid which results in turn, in timeline owners not having to
-maintain such goals.
+received by the reactor. The resulting partial plan is shown in Fig.
+\ref{fig:shop:relax2}. In this process the reactor informs the agent
+that all {\em external} goals it requested for the future are no
+longer valid, allowing timeline owners to divest maintaining such
+goals.
 
@@
 
-As the plan is no longer conflicted, synchronization can resume and
-find a solution described above. Newly pending goals would be enforced
-at the next \texttt{step} to resume deliberation on this relaxed
-partial plan. By using synchronization the reactor was not only able
-to identify that its plan was not consistent but able to recover from
+When the plan is no longer conflicted, synchronization can resume and
+find a solution as described above. Newly pending goals would be
+enforced at the next \texttt{step} to resume deliberation on this
+relaxed partial plan. By using synchronization the reactor is not only
+able to identify that the plan is consistent but able to recover from
 it by offering a new incomplete partial plan to be resolved in the
@@
 % execution in the sense that 
-Synchronization plays the role of what is typically termed as the {\em
-  Executive} in other architectures \cite{gat98, alami:1998p820,
-  mus98, williams03, Nesnas:2003do}.
+Synchronization therefore, plays the role of what is typically termed
+as the {\em Executive} in other architectures \cite{gat98,
+  alami:1998p820, mus98, williams03, Nesnas:2003do}.
 
@@
 can occur in between any planning step. Not only is synchronization
-influenced by the outcome, but it also influences
-planning. Specifically when synchronization occurs before a complete
-plan is found, it will inject new facts ({\em external} observations
-and resulting {\em internal} state) in the plan database available for
+influenced by the outcome, but it also influences planning.
+Specifically when synchronization occurs before a complete plan is
+found, it will inject new facts ({\em external} observations and
+resulting {\em internal} state) in the plan database available for
 inference when the next planning \texttt{step} resumes. On this next
@@
 included in the partial plan which can introduce new flaws in the plan
-while potetnially having reolved previously exisiting flaw. The
-deliberation process is therefore perturbed externally by the {\em
+while potentially having resolved previously existing flaws. The
+deliberation process is therefore perturbed externally by {\em
   synchronization} and consequently fully informed of the world state
-evolution as it is planning.  This specific aspect contrasts with
-other more classical approaches we have seen, which avoid 
-perturbing the planner during its search as it is not compatible 
-with the ``off-line planning'' assumption.
-
-By our design choices -- both at the architecture level and the
-integration of the \eu framework for embedded deliberation -- we are
-able to implement an architecture that avoids this critical off-line
-planning limitation while remaining within a formal framework that
-allows for deliberation and model-based agent control. We believe that
-this tight integration of planning and execution allows the system to
-be more acute vis-a-vis the environment allowing informed
-decision-making and a tighter reaction loop. This is possible while
-integrating information despite a long latency for the reactors, as
-long as one can ensure that synchronization of all reactors and
-deliberation steps can be done within a time that is resonably small
-in comparison to the agent tick duration.
+evolution as it is planning. This contrasts with other classical
+approaches, which avoid perturbing the planner during its search as it
+is not compatible with the ``off-line planning'' assumption, noted in
+Section \ref{sec:arch:europa}.
+
+By design choices made in the \rx framework -- both at the
+architectural level and the integration of the \eu framework for
+embedded deliberation -- we are able to implement a system that avoids
+this critical off-line planning limitation while enforcing a formal
+framework that allows for deliberation and model-based agent control.
+This tight integration of planning and execution allows the system to
+be more responsive to environmental change allowing informed
+decision-making within a tighter reaction loop and despite reactor
+latency.
+
+% As long as one can ensure that synchronization of all
+% reactors and deliberation steps can be done within a time that is
+% resonably small in comparison to the agent tick duration.
 

### 2012-02-19 trunk/setobook.tex
further updates to arch

@@
 
-Authors Rajan and Py are funded by a block grant from the David and
-Lucile Packard Foundation to MBARI. Rajan and Py would like to thank
-their collaborators at MBARI, J. Ryan, J. Harvey, R. Vrijenhoek,
-J. Bellingham. They also thank colleagues at NASA Ames who made \eu
-available as open source; specifically M. del Alto, D. Korsmeyer and
-S. Zornetzer. We are indebted to N. Muscettola for his ideas and early
-work at NASA Ames which forms the basis of this work.
+Rajan and Py are funded by a block grant from the David and Lucile
+Packard Foundation to MBARI. They thank their collaborators J. Ryan,
+J. Harvey, R. Vrijenhoek and J. Bellingham at MBARI. In addition they
+thank colleagues M. del Alto, D. Korsmeyer and S. Zornetzer at NASA
+Ames who made \eu available as open source. The authors are indebted
+to N. Muscettola for his ideas and early work at NASA Ames which forms
+the basis of this work.
 

### 2012-02-20 constraints-table.tex
added EUROPA constraints in the form of a table

\begin{table*}[ht]
  \centering
  \renewcommand{\arraystretch}{1.2}%{0.99}
  \begin{tabular}{|l|l|p{8cm}|p{3cm}|}
    \hline
    \textbf{Constraint}& \textbf{Syntax}& \textbf{Description}& \textbf{Variable Types}\\
    \hline
    distanceSquares& distanceSquares(a,b,c)& c = sqrt(a+b) if a and b are singleton& Numeric intervals\\
    \hline
    calcDistance&calcDistance(a,b,c,d,e)& a is Euclidean distance between points (b,c) and (d,e)& Numeric intervals\\
    \hline
    sin& sin(a,b)& a = sin(b)& Numeric intervals\\
    \hline
    allDiff& allDiff(a,b,\ldots)& Restrict all domains so the intersection of any pair of domains is empty& Comparable\\
    \hline
    EqualMaximum& EqualMax(a,b,\ldots)& a = max(b,\ldots)& Numeric\\
    \hline
    EqualMinimum& EqualMin(a,b,\ldots)& a = min(b,\ldots)& Numeric\\
    \hline
    CountZeros& CountZeros(a,b,\ldots)& a is the count of the rest that can be zero& Numeric\\
    \hline
    CountNonZeros& CountNonZeros(a,b,\ldots)& a is the count of the rest that can be non-zero& Numeric\\
    \hline
    diffSquare& diffSquare(a,b,\ldots)& c = (a - b)$^2$ if a and b are singleton& Numeric intervals\\
    \hline
    card& card(a,b,\ldots)& a must be greater than or equal the count of the other variables that are true& Numeric\\
    \hline
  \end{tabular}
  \caption{\small Calculation Constraints: These constraints enforce that one variable (usually a) is constrained by a calculation done on the remaining variables.}
  \label{tab:calconst}
\end{table*}

%%%%%%%%%%%%%

\begin{table*}[ht]
  \centering
  \renewcommand{\arraystretch}{1.0}%{0.99}
  \begin{tabular}{|l|l|p{8cm}|p{3cm}|}
    \hline
    \textbf{Constraint}& \textbf{Syntax}& \textbf{Description}& \textbf{Variable Types}\\
    \hline
    subsetOf& subsetOf(a,b)& a is a subset of b& Comparable\\
    \hline
    memberImply& memberImply(a,b,c,d)& If a is a subset of b, then require that c is a subset of d& a and b comparable, c and d comparable\\
    \hline
    Lock& Lock(a,b)& Restrict a's domain to be contained in b's domain& Comparable\\
    \hline
  \end{tabular}
  \caption{\small Set Constraints: These constraints are likely to be used for set comparisons etc. Note, however, that many of the other constraints could be used for sets (\ie Comparable variables) just as these constraints could be applied to numeric domains as well.}
  \label{tab:setconst}
\end{table*}

%%%%%%%%%%%%%

\begin{table*}[ht]
  \centering
  \renewcommand{\arraystretch}{1.0}%{0.99}
  \begin{tabular}{|l|l|p{5cm}|p{3.5cm}|}
    \hline
    \textbf{Constraint}& \textbf{Syntax}& \textbf{Description}& \textbf{Variable Types}\\
    \hline
    commonAncestor& commonAncestor(a,b,c)& a and b must be contained by the same object in c (or by c itself)& a, b and c are all objects\\
    \hline
    hasAncestor& hasAncestor(a,b)& a must be contained by some object in b& a and b are objects\\
    \hline
  \end{tabular}
  \caption{\small Object Hierarchy Constraints: These constraints are imposed on objects and tokens. These constraints are used to assert which object one or more tokens is contained by. Most often, the commonAncestor constraint is used to subgoal across timelines which must share a contained object in common.}
  \label{tab:objhierarchy}
\end{table*}

%%%%%%%%%%%%%

\begin{table*}[ht]
  \centering
  \renewcommand{\arraystretch}{1.0}%{0.99}
  \begin{tabular}{|l|l|p{10cm}|p{3cm}|}
    \hline
    \textbf{Constraint}& \textbf{Syntax}& \textbf{Description}& \textbf{Variable Types}\\
    \hline
    UNARY& NA& Restrict a variable's domain; given a variable and a domain, intersect them. Note that this constraints is only available internally and not exposed in NDDL& Anything\\
    \hline
    or& or(a,b\ldots)& At least one of the variables must be true& Numeric\\
    \hline
    absVal& absVal(a,b)& a.lb $\geq$ 0, a.ub = max(abs(b.lb)), b.lb $\geq$ -a.lb, b.ub $\leq$ a.ub& Numeric\\
    \hline
  \end{tabular}
 \caption{\small Miscellaneous Constraints}
  \label{tab:misc}
\end{table*}

### 2012-02-20 europa-basics.tex
added EUROPA constraints in the form of a table

@@
+\section{Foundational Concepts}
+\label{sec:concepts}
+
+\rx is an adaptive, artificial intelligence based controller and
+provides general framework for building reasoning systems for
+real-world autonomous vehicles. At MBARI \rx is used for AUV control;
+another instantiation of the system is being used for control of a
+terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
+development of \rx has been targeted at surveying a number of
+oceanographic features which are dynamic and unpredictable
+spatio-temporally. Typically this requires our AUVs to balance the
+goals of coverage spatially while opportunistically following features
+of scientific interest and to do so while being operationally aware of
+its own limitations in terms of resources (typically the battery state
+of charge) and overall proximity to other observational assets for
+obtaining scientific ground-truth. We are therefore interested in
+representational frameworks that allow robots to pursue long-term
+objectives while choosing short-term gain and can gracefully deal with
+exogenous or endogenous change.
+
+To enable this responsiveness to external observations, the agent has
+to be able to synthesize plans insitu and to re-plan. We use a
+temporal constraint-based planner \eu with a demonstrated legacy of
+having flown on NASA space missions \cite{jonsson00,bresina05,
+  barreiro09}. Our autonomy architecture brings three key innovations
+for AUV adaptation: the use of flexible plan representations,
+compositional control with the use of partitioned networks and
+off-line learning to inform insitu environmental state estimation. In
+this section we detail some of the key concepts relevant to flexible
+plan representation. Section \ref{sec:arch} has details the
+\rx partitioned architecture and estimation driven vehicle
+adaptation.
+
+\paragraph {A brief note on \texttt{EUROPA} history and notation} The
+earliest origin of what we will call \eu in this chapter, is from the
+\texttt{HSTS} planner \cite{mus94} which was used for a number of
+demonstrations for space telescopes as noted in Section
+\ref{sec:rabeyond}. \texttt{HSTS} was the
+\texttt{LISP}-based\footnote{This was the first (and only to our
+  knowledge) software ever to been written in \texttt{LISP} to be
+  flown in space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
+the planner was implemented at NASA Ames which went on to be used for
+the \texttt{MAPGEN} \cite{bresina05} and continues to be used to date,
+in the mission-critical uplink process for the command/control of the
+Mars Exploration Rovers (MER) mission. A vastly re-factored, higher
+performance open source version \cite{europapso} was named \eu and is
+the subject of this chapter. Note however, in using the name \eu, we
+refer to the \emph{fundamental ideas as well as the planning paradigm}
+even if there exists a specific implementation.
+
+\subsection{The Basics}
+\label{sec:basics}
+
+\eu uses a \emph{domain model} written in a declarative language
+(NDDL: New Domain Description Language), together with initial
+conditions and goals also specified in NDDL, to construct a set of
+temporal relations that must be true at the start time. These models
+include assertions about the physics of the vehicle, i.e how it
+responds to external stimulus and internally driven goals. By
+propagating these relations forward using Simple Temporal Networks
+\cite{dechter91} and applying goal constraints, \eu can select a set
+of conditions that should be true in the future, where some of these
+conditions will correspond to actions the agent must take. The planner
+can backtrack and try another path during search if a goal cannot be
+reached while being capable of discarding unachievable goals.
+
+Traditionally robot execution has relied on dispatching commands at
+precise times. Such linear sequences of precisely timed commands give
+no ability to adjust execution on the basis of sensory
+information. Although some commands can issue tests on sensor
+readings, these tests have the objective of verifying whether expected
+execution conditions are occuring. If not, the state of the system is
+declared off-nominal and execution of the sequence of commands is
+interrupted. More recently, executives have been proposed and
+implemented that significantly broaden the way robots can be commanded
+\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
+interpreted a \textit{temporally flexible plan} which represents each
+start time as a variable and contains an explicit network of bounded
+delay constraints between such variables.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\caption{\small Tokens with flexible temporal intervals and parametric
+  constraints between tokens. This example shows the triggering of a
+  water sampler based on a feature threshold while the vehicle
+  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
+  start \& end times.}
+\label{fig:flex-timelines}
+\vskip+0.1cm
+\end{figure}
+
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time the most appropriate one for the actual execution
+conditions. The fact that constraints are explicitly represented
+ensures that through constraint propagation the executive will respect
+global limits expressed in the plan (e.g., don't start a task until a
+certain condition has been satisfied but still satisfying some global
+deadline). Such flexibility is critical when dealing with dynamic
+ocean conditions where precise timing of a robotic action might be
+indeterminate. Further, the advantage of flexibility can be contrasted
+with the consequences of the intrinsic inflexibiliy of traditional
+command sequences. Because they are inflexible, sequences must
+necessarily be designed considering worst case scenarios.
+
+\eu uses a \emph{state variable} representation to describe the
+evolution of state over time. The instantiated history of such state
+variable evolution over a temporal horizon we call \emph{timelines}
+and which represent a single thread in the execution of a concurrent
+system. At any given time each thread can execute a single procedure.
+Thus each timeline consists of a sequence of procedures which
+encapsulate and describe state evolution; we call these instantiated
+atomic entities \emph{tokens}.  A token therefore describes a
+procedure invocation, the state variables on which it can occur, the
+parameter values of the procedure, and the time values defining the
+interval. We allow encapsulation of uncertainty within these tokens
+with a range of start and end times and parameters, all of which are
+encoded as variables. A constraint solver in turn manipulates these
+variables defined in a \eu domain model. For example, consider
+a scientific need to take a water samples 100 meters from a hotspot
+while an AUV is performing a Yo-Yo in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
+% \begin{figure}
+% \centering
+% \includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+% \caption{\small Temporal relations defined within the planner are
+%   based on \texttt{Allen Algebra} relations shown above.}
+% \label{fig:allen-algebra}
+% \vskip-0.3cm
+% \end{figure}
+
 \section{Fundamentals of  Automated Planning}
@@
 }
+
+\begin{figure} \centering
+  \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
+  \caption{\small A potential plan for a robot shopping for groceries.}
+\label{fig:shop:shopping}
+\vskip+0.1cm
+\end{figure}
 
@@
 For such a robot the milk buying plan at the store might look like
-\comment{Figure needed}.
-
-% \eu is a general purpose AI planning toolkit developed at NASA Ames.  
-% TODO: Talk about NASA missions where \eu has been used.
+that shown in Fig. \ref{fig:shop:shopping}.
 
@@
 nodes and constraints are arcs between the variables. Having this
-network representation in mind, a pair of variables $x,y$ is said to be arc consistent if for every
-value in $x$'s domain, there is a value in $y$'s domain that is
-consistent with all the constraints that connect $x$ and $y$.  A naive
-approach to ensuring arc consistency consists of cycling through all
-the variable pairs and performing constraint propagation until there
-are no variable domain changes. The AC-3 algorithm \cite{mackworth77},
-a popular implementation, improves over the naive approach by ignoring
-constraints whose variables were not affected since the last
-iteration.
+network representation in mind, a pair of variables $x,y$ is said to
+be arc consistent if for every value in $x$'s domain, there is a value
+in $y$'s domain that is consistent with all the constraints that
+connect $x$ and $y$.  A naive approach to ensuring arc consistency
+consists of cycling through all the variable pairs and performing
+constraint propagation until there are no variable domain changes. The
+AC-3 algorithm \cite{mackworth77}, a popular implementation, improves
+over the naive approach by ignoring constraints whose variables were
+not affected since the last iteration. \comment{A figure of a
+  constraint network preferably of the shopping example would be
+  useful here. JB to generate a napkin fig}
 
-Arc Consistency and other, more stringent consistency definitions can be
-used to efficiently prove that a CP problem is unsatisfiable. However,
-they generally cannot prove that a CP problem is satisfiable and
-provide specific variable assignments that constitute solutions. For
-that, search algorithms like global backtracking \comment{citation
-  needed} or local search are used; these use consistency as an
-efficient way to prune the search space \comment {citation CP
-  Handbook}. In theory, solving a CP problem is NP-Hard
+Arc Consistency and other, more stringent consistency definitions can
+be used to efficiently prove that a \textsf{CP} problem is
+unsatisfiable. However, they generally cannot prove that a \textsf{CP}
+problem is satisfiable and provide specific variable assignments that
+constitute solutions. For that, search algorithms like global
+backtracking \cite{hooker05} or local search are used; these use
+consistency as an efficient way to prune the search space
+\cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
 \cite{ghallab04}, but often very efficient in practice using a number
@@
 
-CP is usually implemented as part of a programming language and
-constraints are usually represented as objects \cite{puget95}. Any
-constraint that the user is able to reason about can be introduced into the system and, as long as the constraint
-propagation protocols specified by the host CP system are enforced, it
-will be indistinguishable from any other "primitive'' CP constraint,
+\textsf{CP} is usually implemented as part of a programming language
+and constraints are usually represented as objects \cite{puget95}. Any
+constraint that the user is able to reason about can be introduced
+into the system and, as long as the constraint propagation protocols
+specified by the host \textsf{CP} system are enforced, it will be
+indistinguishable from any other "primitive'' \textsf{CP} constraint,
 such as $\leq$ or $\geq$.
 
-\subsection{Constraint-Based Attribute and Interval Planning}
+\subsection{Constraint-Based Planning}
 \label{sec:europa:cp}
@@
 formalism. In particular, it is hard to represent time, resources,
-mutual exclusion and concurrency using propositions \comment{citation
-  needed}. While CSP representations have traditionally had an edge in
-formulating and representing planning problems, 
-% It is straightforward to represent and reason about all of
-% those elements using a CSP representation, as a result there has been
-% some work on doing automated planning while taking advantage of CSP
-% (TODO: ref). Traditionally, the entire planning problem is translated
-% into a CSP and then solved using traditional CSP methods, this
-leading to a formulation where action choices and relationships are
-represented as variables and/or constraints, there are at least two
-major drawbacks:
+% mutual exclusion and concurrency using propositions
+% \cite{ghallab04}. While CSP representations have traditionally had an
+% edge in formulating and representing planning problems,
+% % It is straightforward to represent and reason about all of
+% % those elements using a CSP representation, as a result there has been
+% % some work on doing automated planning while taking advantage of CSP
+% % (TODO: ref). Traditionally, the entire planning problem is translated
+% % into a CSP and then solved using traditional CSP methods, this
+% leading to a formulation where action choices and relationships are
+% represented as variables and/or constraints, there are at least two
+% major drawbacks:
+mutual exclusion and concurrency using propositions \cite{ghallab04}.
+Constraint Programming is a framework that lends itself very well to
+represent and reason about all of those elements \cite{ghallab94},
+consequently, there has been a good amount of work building automated
+planners that take advantage of CP techniques.  The most common
+approach in constraint-based planning has been to encode the entire
+planning problem as a CP problem, where action choices and
+relationships are represented as variables and/or constraints
+\cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
+pure CP formulation is that standard CP solvers can be then used to
+solve the planning problem, however, this approach also has a couple
+of significant drawbacks:
 
@@
 
-Constraint-based Planning and specifically Constraint-Based Attribute
-and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
-close that gap. On one hand, it takes advantage of a CSP
-representation; on the other it uses attributes and intervals to
-maintain an explicit representation of the elements relevant to
-planning. This makes it easier to write algorithms that search for an
-reason abut plans. The main elements of CAIP are:
-
-\begin{enumerate}
-	\item Intervals: \comment{needs to be defined}
-	\item Attributes:
-	\item Domain Constraints an Configuration rules:
-\end{enumerate}
-
-\comment{Explain how planning problem instances and their solutions are represented}
-
-\subsection{\eu Plan Representation}
-\label{sec:europa:pr}
-
-\eu is an implementation of the CAIP framework.
-
-\textit{Variables} Values that need to be represented to describe the
-problem domain and over which we may want to specify constraints.
-
-\textit{Objects} The things we wish to describe and refer to in a
-domain are considered Objects. As in the case with object-oriented
-analysis and design, one can seek out the nouns in any domain
-description to find likely objects. In a Mars rover example, we might
-consider the satellite, the camera, and the attitude controller (which
-orients the robot for taking pictures) to be objects. Objects have
-state and behavior. For example, a camera can be:
-
-    off,
-    ready, or
-    taking a picture.
-
-An attitude controller can be:
-
-    pointing at a position, or
-    turning from one position to another.
-
-    An object is an instance of a class. In \eu we model using the
-    abstraction of a class to speak about all instances having certain
-    properties of state and behavior. In order to describe such state
-    and behavior we build on the formalism of first order logic.
-
-\textit{Tokens}
-
-A predicate defines a relation between objects and properties. In \eu,
-we define such relations between variables whose domains are sets of
-objects and sets of properties. For example, we might use a predicate
-\texttt{Pointing($a,p$)} to indicate that a rover's attitude controller
-$a$ is pointing at position $p$. Note that $a$ is a variable which may
-have a number of possible values in a problem with multiple
-satellites. Similarly, $p$ is a variable whose values are the set of
-possible positions. In a grounded plan, single values will be
-specified for each variable as we saw in the case of a solved
-Constraint Satisfaction Problem.
-
-In general, it is not sufficient to state that a predicate is true
-without giving it some temporal extent over which it holds. Predicates
-that are always true can be thought to hold from the beginning to the
-end of time. However, in practice, the temporal extent of interest
-must be defined with timepoints to represent its start and end. So we
-might want to write \texttt{Pointing($a,s,e,p$)} to indicate that the
-attitude controller a is pointing at position $p$ from time $s$ to
-time $e$. In fact, this pattern of using such predicates to describe
-both state and behavior of objects is so prevalent in \eu that we have
-introduced a special construct called a Token which has the built in
-variables to indicate the object to which the statement principally
-applies and the timepoints over which it holds. In \eu, all predicate
-instances are Tokens.
-
-\comment{talk about actions}
-
-A Token is an instance of a predicate or an action and is defined over
-a temporal extent. Every token has five built-in variables:
-
-start: The beginning of the temporal extent over which the predicate
-is defined.  end: The end of the temporal extent over which the
-predicate is defined.  duration: The constraints start + duration =
-end is enforced automatically.  object: made. The set of objects to
-which a token might apply. In a grounded plan each Token applies to a
-specific Object, reflecting the intuition that we are using Tokens to
-describe some aspect of an Object (i.e. its state or behavior) in
-time. However, in a partial plan, the commitment to a specific object
-may not yet have been made.
-
-state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
-variable captures the token's current state and its reachable states
-through further restriction. See the Token State Model discussion
-below for details.
-
-\textit{Objects (continued)}
-
-As we have noted, Tokens describe some aspect of an Object in
-time. Objects thus may have many Tokens in a plan in order to describe
-their state and behavior throughout all points in time of the
-plan. Within this general framework, we note a few particulars:
-
-Static Facts: Classes without predicates lead to Objects without
-Tokens. This arises where an objects state or behavior is independent
-of time. For example, a domain may have a set of locations and/or
-paths for which there is nothing more to say than that they exist.
-
-Timelines: Often objects in a domain must be described by exactly one
-token for every given timepoint in the plan. Such objects are so
-common that we provide a special construct in \eu to extend these
-semantics to derived classes. Any instances of a class derived from a
-Timeline will induce ordering requirements among its tokens in order
-to ensure no temporal overlap may occur among them. See below for
-further discussion.
-
-Resources: Metric resources, e.g. the energy of a battery or the
-capacity of a cargo hold, are objects with an explicit quantitative
-state in time and with a circumscribed range of changes that can occur
-to impact that state i.e. produce, consume, use, change. These changes
-are captured as tokens. Resources are such a common requirement for
-\eu users that special constructs are also provided for
-them. Instances of classes derived from a Resource will induce
-ordering requirements on their Tokens in order to ensure that the
-level of the resource remains within specified limits.
-
-Timelines
-
-It may be sufficient to maintain only a partial-order among
-tokens. However, it is often the case that tokens represent states and
-actions of a single object in the system. Such tokens are typically
-mutually-exclusive. \eu uses a Timeline structure developed in (??)
-to concisely capture system components whose behavior is described
-over time in this manner.
-
-\comment{This is the first instance that this domain is being
-  introduced. Either we need more explanation or it ought to be
-  removed.} For example, consider the tire-world domain. The tire was
-located in the trunk with the predicate tireLocated(Trunk) and located
-on the ground with the predicate tireLocated(Ground). Clearly, the
-same tire cannot be in both places at once. A Timeline provides a
-simple method for aggregating the statements about a tire such that
-they are mutually exclusive.
-
-Figure 1: A Timeline for a Tire
-
-Figure 1 illustrates how a Timeline can be used to describe the
-whereabouts of a tire. The tire is an object in the tire-world
-represented as a timeline. It has predicates associated with it which
-can describe its states and actions over time. The predicate names
-need not contain the tire prefix; this is now implicit since the
-tokens are assigned to a specific tire instance (i.e. an instance of a
-Timeline). While Timelines are a useful element of the \eu planning
-paradigm, they are not essential. A more general notion of a system
-Object can be used which does not impose restrictions of
-mutual-exclusion or non-zero duration. This can be important in
-supporting partial-order planning.
-
-For this example, the state of the tire is specified using 3
-contiguous tokens. The end and start time-points are related by an
-equality constraint. A Moving predicate has been introduced to cover
-the transition from one location to another. It takes 2 location
-arguments. Notice that precedence relationships exist between tokens
-such that they cannot overlap but the start and end times may remain
-flexible. To illustrate this, sample times are included. Assume a
-total time range of interest between 0 and 1000. In addition, tokens
-on Timelines have a minimum duration of 1. As a result the values
-shown are the most flexible possible values for the start and end of
-each token. There are a number of advantages of allowing this
-flexibility. First, the basic structure of the plan can be developed
-without over-committing to specific times. If it is not necessary for
-the tire to be on the ground at time-step 4, then a planner should not
-be forced to specify it. Such an approach permits a least-commitment
-approach to planning. Second, in many domains it is simply impossible
-to know in planning exactly how long an activity might take. For
-example, a common activity of driving a car from one point to another
-on a road can take varying amounts of time depending on traffic and
-traffic lights. In such circumstances it is more practical to express
-upper and lower bounds on durations which naturally lead to intervals
-for start and end times. In such domains, flexibility in planning aids
-robustness in execution.  Rules
-
-In order for a plan to be valid, it must comply with all rules and
-regulations pertinent to the application domain in question. Rules
-govern the internal and external relationships of a token. For
-example, consider a parameterized predicate describing a transition
-from one location to another. Let the parameters be from and to. The
-parameters are instantiated on a token as variables whose domain of
-values is the set of all locations in a given problem. A rule
-governing an internal relationship among token variables might
-stipulate that a transition must involve a change in location. This
-can be easily expressed as a constraint on the definition of a
-predicate of the form: from != to. It is reasonable to further
-stipulate that one must be Located somewhere before a transition can
-occur, and one must end up Located somewhere when completed. This is
-an example of a rule governing an external relationship among
-tokens. It specifies a requirement that tokens of the predicate
-Located precede and succeed tokens of the predicate Moving.
-
-Figure 2: Internal and external relationships from rules on Moving
-Figure 2 illustrates the entities and relations involved in specifying
-such a rule on a Moving predicate. The token on which the rule applies
-is referred to as the master. Each Located token required by the
-master is referred to as a slave. All variables are indicated by name
-and their domains are expressed as intervals in the case of temporal
-variables and as enumerations for the remainder. Application of a rule
-on a token can thus cause slave tokens, variables, and constraints to
-be introduced.  Token State Model
-
-The capability of a domain rule to cause a slave token to be created
-is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting tokens to be in the plan in
-order for the master to be valid. There are 2 possibilities to
-consider:
-
-The slave is inserted as an active token in the plan. As such, rules
-may be activated on the slave, and it may consume resources.
-
-The token is merged with a matching token already in the plan. Once a
-slave is merged, the requirement it represents are considered
-satisfied. The process of merging passes on all restrictions imposed
-on the slave to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. For an active token and an inactive token to be
-compatible requires that they are instances of the same predicate and
-that no intersections between corresponding variables are empty. The
-effect of merging is illustrated in Figure 3.
-
-Figure 3: Merging an inactive token on an active token. Domain
-restrictions occur in highlighted variables of active token.
-
-On creation of a token, where the commitment has not been made yet to
-activate or merge the token, the token is said to be inactive. There
-is a nuance to the state model for tokens which relate to the mode of
-its creation. If the token is allocated explicitly by an external
-actor, rather than internally through rule firing, it may include the
-state rejected indicating the planner is permitted to reject the
-token. A valid plan can include rejected tokens. This typically arises
-where the token represents a goal that is preferable to achieve but
-not mandatory. This state is not reachable for slaves since that would
-imply selective adherence to the domain model. Figure 4 presents the
-state transition diagram for token states relevant for planning. The
-transitions are operations on a partial plan which can decide an
-outcome for an inactive token. As operations which may arise in
-search, they must be reversible during backtracking. The cancellation
-operations in each case are also shown.
-
-Figure 4: Token States and Transitions for Planning
-
-The state of a token is embodied with a 5th and final built-in
-variable referred to as the state variable. The planner states are
-values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
-INACTIVE state is captured by the variable being unbound. If a value
-is removed from the domain, that state will not be reachable. For
-example, when a slave is created via a rule, the REJECTED value is
-removed from its state variable.
-
-\textit{Summary}
-
-This has presented the main elements of the EUROPA planning approach. They are:
+These drawbacks have been addressed by a small number of planning
+frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
+\texttt{IxTet} \cite{ghallab87,lemai04}, and CAIP (Constraint-Based
+Interval and Attribute Planning)\cite{mus94,frank2003}.  These
+frameworks take advantage of a CSP representation to deal with time
+and resources, at the same time maintaining an explicit representation
+of the elements relevant to planning (action choices, causal
+relationships, etc). This combination results in a planning approach
+that can deal with the type of constraints found in real-life
+problems, while allowing for intuitive problem descriptions and a
+natural way to express search algorithms and heuristics. \eu is an
+implementation of the CAIP framework. In CAIP, planning problems are
+specified in terms of:
 
@@
 
-\item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
-states and actions in time. Note that tokens do not discriminate
-between state and action.
+\item \textbf{Intervals}: entities that have a temporal scope (that
+  is, a beginning and an end in time) and a set of
+  attributes. Intervals are used to represent actions and states in
+  the planning domain.
 
-\item Token States which are the basis of planning operations on a partial
-plan.
+\item \textbf{Variables and Constraints}: Variables represent Interval
+  attributes and Constraints enable a direct representation of
+  temporal, resource and any other the restrictions that a valid plan
+  must comply with.
 
-\item Constraints to describe relationships among tokens. This provides an
-expressive method of describing interactions among tokens in the
-context of Temporal Planning.
+\item \textbf{Domain Configuration Rules}: this is an explicit
+  representation of the conceptual relationships between actions and
+  states in the problem domain, for instance, for a Shopping Agent to
+  be able to purchase a product, it must be at a location where the
+  product is available.
+\end{enumerate}
 
-\item Timelines as a concise abstraction to express the evolution of state
-and behavior for a system component. It provides semantics of mutual
-exclusion sequencing in time. Other core abstractions are available in
-\eu for handling metric resources.
+To generate plans, the CAIP framework uses a plan-space search
+approach \cite{ghallab04}, where a partial plan is evolved until all
+open goals are achieved, and all inconsistencies are removed, the
+specific search algorithm used by \eu will be covered in detail below.
 
-\item Domain rules to describe internal and external relationships on and
-between tokens respectively. Rules are applied to active tokens
-(master tokens) referred to as masters and typically produce inactive
-tokens (slaves) which are a key vehicle through which the planning
-process evolves. The rule structure in \eu differs from the more
-restrictive commitment to preconditions and effects in classical
-planning which prohibits durative actions and disembodied effects
-(i.e. effects which may not occur until some temporal distance after
-the end of the action).
-
-\end{enumerate}
+\input{constraints-table}

### 2012-02-20 arch.tex
addressed kcomments the best I could

@@
   % early on but the flow of this paragraph is not very tight}
-\kcomment{Looks fine to me unless you want to clarify further.}
+%\kcomment{Looks fine to me unless you want to clarify further.}
 
@@
 reactor needs to replan due to an unexpected event, it will recall all
-requests it produced that have not been observed \kcomment{replace
-  ``observed'' with ``not incorporated by the owner reactor''?}
-yet. This recall is sent to the reactor that received the goal(s) to
+requests it produced that have not been executed yet.
+% \kcomment{replace
+%   ``observed'' with ``not incorporated by the owner reactor''?}
+% yet
+ This recall is sent to the reactor that received the goal(s) to
 remove it/them from its current set of objectives.
@@
 
-Both these deliberative \kcomment{``deliberative??''; something else
-  or remove altogether perhaps?} functions have different scope and
-temporal constraints. The \texttt{synchronize} call is considered as
-atomic and will focus only on state inference for the current tick by
+Both these 
+% deliberative \kcomment{``deliberative??''; something else
+%   or remove altogether perhaps?} 
+functions have different scope and temporal constraints. 
+The \texttt{synchronize} call is considered as atomic and will 
+focus only on state inference for the current tick by
 identifying the current value of {\em internal} state variables from
@@
 synchronization propagates these events in the plan and can uses it as
-an aid in identifying its {\em internal} timelines \kcomment{``use it
-  as an aid to identify'' what though?} based on this plan.
+an aid in resolving its {\em current state flaw} by using the plan
+information on the exepected {\em internal} state evolution nearby the
+execution frontier. 
+% \kcomment{``use it
+%   as an aid to identify'' what though?}
 
@@
 plan inconsistency. This results in the retraction of decisions made
-during the \kcomment{previous?} {\em deliberation} step. % in order to
+during previous {\em deliberation} steps. % in order to
 % allow {\em synchronization} to identify its {\em internal} state.
@@
 In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
-new type of flaw which forces identification \kcomment{identification
-  of what? Probably need a little more explanation here.} of internal
-state of the reactor for the current tick. This flaw can be defined
-as:
+new type of flaw which requires to identify the value of all the {\em
+  internal} state variables for the current tick.
+% forces identification \kcomment{identification
+%   of what? Probably need a little more explanation here.} of internal
+% state of the reactor for the current tick. 
+This flaw can be defined as:
 
@@
 deliberation process with the risk that the search for a solution is
-not a straightforward sequence of unit decisions \kcomment{has a unit
-  decision been explained in Europa?}. Since such a process is
-required by the architecture at \emph{every} tick for \emph{every}
-reactor the model needs to be designed carefully.
+not a straightforward sequence of staright-forward decisions where 
+the resolution can be easily identified by the solver as any other
+option would leafd immediately to an inonsistency in the plan.
+% \kcomment{has a unit decision been explained in Europa?}. 
+Since such a process is required by the architecture at \emph{every}
+tick for \emph{every} reactor the model needs to be designed carefully.
 
@@
 agent level those goals that are before the reactor execution latency
-or too far into the future. The planning window inside of a reactor
-refines this filtering further, as tokens can be excluded dynamically
-from deliberation as their insertion in the plan and scheduling with
-other tokens generated, can reduce their temporal scope outside of the
-planning scope \kcomment{ this last sentence should be reworked}. This
-helps the deliberation solver reduce the number of tokens to be
-evaluated reducing deliberation cost and increasing performance. As
-time advances, flaws that were initially beyond this temporal horizon
-will become active which a reactor can evaluate. This results in an
-apparent \emph{continuous planning} of the reactor. When there are no
+or too far into the future. The planning window reflect this filtering
+of the agent by applying to tokens that are generated internally by the
+reactor during its deliberation or the {\em goals} for which he
+planner decided to postpone after the planning window -- probably
+due to other tasks that needed completion beforehand.
+% The planning window inside of a reactor
+% refines this filtering further, as tokens can be excluded dynamically
+% from deliberation as their insertion in the plan and scheduling with
+% other tokens generated, can reduce their temporal scope outside of the
+% planning scope \kcomment{ this last sentence should be reworked}. 
+This allow the planner to focus on its planning horizon and report
+decision on flawss thgat are beyond this scope on later ticks. As, 
+time advance, these future flaws will eventuallty be part of the
+planning window of the reactor and will be then resolved along with
+potential new goals received from the agent.
+% This helps the deliberation solver reduce the number of tokens to be
+% evaluated reducing deliberation cost and increasing performance. As
+% time advances, flaws that were initially beyond this temporal horizon
+% will become active which a reactor can evaluate.
+ This results in an apparent \emph{continuous planning} of the reactor. When there are no
 more flaws present in the current planning scope a partial plan

### 2012-02-20 trunk/arch.tex
typos

@@
   of high level waypoints into low-level commands) and temporally as
-
   each reactor declares its planning look-ahead $\pi$ and expected
-  latency $\lambda$
-  % \comment{$\pi \& \lambda$ both defined here for the first time}
-  before being able to produce plan to the agent depending on where it
-  is located in a reactor dependency graph such as shown in
-  Fig. \ref{fig:agent}.
+  latency $\lambda$ before being able to produce plan to the agent
+  depending on where it is located in a reactor dependency graph such
+  as shown in Fig. \ref{fig:agent}. A reactor closer to the hardware
+  in the hierarchy, is expected to deliberate less and be highly
+  responsive to observations from the world. This implies lower values
+  of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex} below).
 
 \item \emph{Coupling} a tighter integration between the planning
-  process and world evolution to ensure that planning occurs while
-  being continuously informed of state change. By doing so, the
-  planner is better informed in its deliberation despite a larger
-  latency than changing world change state.
+  process and world evolution to ensure that planning occurs
+  \emph{during continuous} state update. By doing so, the planner is
+  better informed in its deliberation despite a larger latency than
+  the changing world.
 
@@
   variables} as the fundamental basis of interaction. Additionally it
-comes with a well-defined ownership model associated with each state
-variable. In this framework each reactor is seen as a 'black box'
-providing its own {\em internal} state variables which it is
+comes with a well-defined ownership model associated with each such
+state variable. In this framework each reactor is seen as a 'black
+box' providing its own {\em internal} state variable which it is
 responsible for in order to maintain its value at every instant of
 time. Each such instance is represented by a {\em tick} with a fixed
-duration; in our implementation of \rx on the Dorado, we use a tick of
-$1$ second. In order to identify its state a reactor can subscribe to
-{\em external} state variables owned by other reactors, for which it
-will receive new updates when the owner reactor generates its {\em
-  internal} state.
-
-This {\em synchronization} process propagates the current state
-information at the execution frontier $\tau$ throughout all the 
-reactors. The sequencing of the reactors synchronization is identified
-by traversing the reactor graph from the least dependents reactors --
-the ones with no {\em external} timelines -- to the most dependent
-ones which typicallu do not provide {\em internal} timelines.
-Conversely each reactor can request a future change on one of 
-its {\em external} state variables which will be transfromed as 
-an {\em internal} goal to the owner. 
-Such \emph{goal dispatching} is expected to produce a
-deliberation phase for the receiving reactor which can cyclically and
-in turn, produce a plan impacting its {\em external} state variables' 
-future that will propagate down following the same mechanism.
+duration; in our implementation of \rx on the Dorado AUV, we use a
+tick of $1$ second. In order to identify its state a reactor can
+subscribe to {\em external} state variables owned by other reactors,
+for which it will receive new updates when the owner reactor generates
+its {\em internal} state.
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
+  \caption{\small An example of a \rx reactor hierarchy which is used
+    to determine a synchronization order for propagation of values in
+    \emph{external} state variables.}
+  \label{fig:reactor-DAG}
+\end{figure}
+
+As the execution frontier, $\tau$ advances, the \rx agent propagates
+the current state information across all reactor state variables
+through the {\em synchronization} process.  The sequencing of the
+reactors synchronization is identified by traversing a reactor
+dependency graph from the least dependent -- ones with no {\em
+  external} timelines -- to the most, \ie which typically do not have
+{\em internal} timelines. Fig. \ref{fig:reactor-DAG} shows an example
+of such a dependency graph implemented as a directed acyclic graph.
+
+Conversely each reactor can request a future change on one of its {\em
+  external} state variables which will be transformed as an {\em
+  internal} goal to the owner reactor(s).  Such \emph{goal
+  dispatching} is expected to produce a deliberation phase for the
+receiving reactor which can cyclically in turn, produce a plan
+impacting its {\em external} state variables' future that will
+propagate down following the same mechanism.
+
+In such a process, one challenge is to avoid being overwhelmed by
+goals that may threaten the ability of a reactor to process and
+integrate in its evolving partial plan. For that reason, each reactor
+specifies the following two parameters that applied towards the scope
+of the agent lifetime:
+
+\begin{itemize}
+
+\item $\lambda$ is the specified latency of this reactor, which is an
+  indicator of the \emph{maximum expected time} for a reactor to
+  produce a plan.\footnote{Note that while it is highly recommended to
+    select a sound value, a failure to produce a plan within a chosen
+    value for this parameter is not considered as a critical failure
+    of the reactor.}
+
+\item $\pi$ is the planning \emph{look-ahead} of the reactor. which is
+  indicative of the number of ticks in the future this reactor is
+  looking ahead during deliberation.
+
+\end{itemize}
+
+Using this information along with the dependency relation between
+reactors, \rx defines the notion of execution latency of a state
+variable as follow:
+
+\begin{definition}
+  The {\em execution latency} of a {\em state variable} $s$ which is
+  internal to the reactor $r$ is the maximum expected delay before a
+  goal on this state variable can be included in its plan and executed
+  by the reactor $r$. It is defined recursively as: 
+  \begin{equation*}
+    \lambda_{exec}(s) = \lambda(r) + \max_{e \in exts(r)}
+    \lambda_{exec}(e) 
+  \end{equation*}
+  Where  $exts(r)$ is the set of the state variables which are
+  {\em externals} to $r$. 
+\end{definition}
+
+For intertwined plan execution, the concept of execution latency is
+critical at the agent level since it can impact the
+planning/deliberation process on {\em external} state variables of a
+reactor. For simplicity we abstract it out as being equivalent to the
+reactor latency implying that we assume that all {\em external}
+timelines of this reactor have $\lambda_{exec} = 0$. At the \rx agent
+level this execution latency is used to define the {\em dispatch
+  window} of the same state variable.% : \comment{Needed to say that
+  % early on but the flow of this paragraph is not very tight}
+%\kcomment{Looks fine to me unless you want to clarify further.}
+
+\begin{definition}
+  \label{def:dispatch}
+  For every timeline $s$ {\em internal} to a reactor $r$, the {\em
+    dispatch window} of this timeline at the execution frontier $\tau$
+  is the time interval:
+  \begin{equation*}
+    \Pi(s) = [\tau + \lambda_{exec}(s), \tau+\lambda_{exec}(s)+\pi(r)]
+  \end{equation*}
+\end{definition}
+
+This window is used by \rx to identify when to dispatch a request to a
+{\em reactor} as a new goal. This happens only if the request start
+time -- represented as the interval of possible start times --
+intersects this window. Requests further in the future are buffered by
+the agent until this time window advances for an overlap. This
+provides an elegant mechanism to request a change from the {\em
+  external} state variable of a reactor while avoiding flooding the
+owner reactor of this state variable.
+
+The choice of both the latency and look-ahead of a reactor can be
+delicate as one is dependent on the other. Needless to say, latency is
+highly dependent on exogenous paramaters such as the host computers
+computing power and the number of reactors needing to deliberate at
+once, among others. As the design stands, these variables should be
+evaluated carefully by the system designer balancing agent efficacy.
+
+% efficacy of the agent to find and execute a
+% global strategy they are just indicative and a failure from a reactor
+% to produce a plan in its latency is not considered as a critical
+% failure from \rx point of view.
+
+Requests from a reactor can be {\em recalled} when it is determined
+that they are not part of its projected plan anymore. For example if a
+reactor needs to replan due to an unexpected event, it will recall all
+requests it produced that have not been executed yet.
+% \kcomment{replace
+%   ``observed'' with ``not incorporated by the owner reactor''?}
+% yet
+ This recall is sent to the reactor that received the goal(s) to
+remove it/them from its current set of objectives.
 
@@
   {\em internal} timeline of a reactor and will execute one step of
-  deliberation. In the context of \eu for example one deliberation
+  deliberation. In the context of \eu for example, one deliberation
   step corresponds to a single flaw resolution (Section
   \ref{sec:europa:arch}) while ensuring that the resulting plan is not
-  proven inconsistent. This implies that it also includes the
-  backtracking of the search in such a case.  The returned value by
-  the \texttt{step} function is a boolean which indicates whether the
-  reactor needs an extra step to deliberate or if this last step
-  resulted in a complete plan within the reactor's look-ahead. On the
-  later, the part of the plan that concern this reactor {\em external}
-  timelines is provided in order to be dispatched as goals to
-  whichever reactor that is reponsible for these. \comment{By
-    extending I may have saud things I developed below ... will have
-    to check}
+  inconsistent. Backtracking is included in the step should it be such
+  a case.  The returned value by the \texttt{step} function is a
+  boolean which indicates whether the reactor needs an extra step to
+  deliberate or if this last step resulted in a complete plan within
+  the reactor's look-ahead. The part of the plan that covers the
+  reactor's {\em external} timelines is provided to be dispatched as a
+  goal.
+% to whichever reactor that is responsible for these. % \comment{By
+    % extending I may have saud things I developed below ... will have
+    % to check}
+
 \end{itemize}
 
-Both these deliberative functions have different scope and temporal
-constraints. The \texttt{synchronize} call is considered as atomic and
-will focus only on state inference for the current tick by identifying
-the current value of {\em internal} state variables from the latest
-updates on the {\em external} state variables. This process propagates
-up in the reactor dependency graph at every tick to ensure that all
-reactors are aware of the current state of the world. \texttt{step}
-relates to a single decision step of the planning process which is
-focused on the future evolution of state variables; it can take
-multiple steps spanning multiple ticks for a reactor to produce a plan
-that is both complete and valid. When such plan is identified
-\comment{how? Is there a terminating condition? No more flaws but I
-  washoping it would have been defined in \eu section} its {\em external}
-timeline(s) are then dispatched to the corresponding reactors which
-own these times and which in turn can start their own deliberation
-\texttt{step}s to satisfy these new objectives. Consequently execution
-occurs as these goals and plans propagate down in the reactor
-hierarchy until they eventually become goals for the lowest level
-reactor which -- instead of deliberating -- may just transform this
-goal into a command for execution in hardware.
+Both these 
+% deliberative \kcomment{``deliberative??''; something else
+%   or remove altogether perhaps?} 
+functions have different scope and temporal constraints. 
+The \texttt{synchronize} call is considered as atomic and will 
+focus only on state inference for the current tick by
+identifying the current value of {\em internal} state variables from
+the latest updates on {\em external} state variables. This process
+propagates up in the reactor dependency graph at every tick to ensure
+that all reactors are aware of the current state of the
+world. \texttt{step} relates to a single decision step of the planning
+process focused on the future evolution of state variables; it can
+take multiple steps spanning multiple ticks for a reactor to produce a
+plan that is both complete and valid. How the reactor identifies that
+its plan is complete and valid is its responsibility and depends on
+the way each reactor resolves its plan at each {\em step}. The general
+idea is that the reactor is able to find a valid strategy in order to
+satisfy as many of its {\em internal} goals as possible. For \eu, this
+means that the solver is unable to identify any further flaws to be
+resolved in the plan.
+
+When such plan is identified, its {\em external} timeline(s) are then
+dispatched to the corresponding reactors which own these times and
+which in turn can start their own deliberation \texttt{step}s to
+satisfy these new objectives. Consequently execution occurs as these
+goals and plans propagate down in the reactor hierarchy until they
+eventually become goals for the lowest level reactor which -- instead
+of deliberating -- may just transform this goal into a command for
+execution in hardware.
 
@@
 the plan. Conversely, when synchronization occurs after deliberation
-steps, its starting point is the last partial plan resulting from
-deliberation. This gives an initial frame to the problemm that helps
--- at least initial -- to target the synchronization problem by
-ensuring that the new {\em external} state updates are compatible with
-this plan. If so, the synchronization propagates these events in the
-plan and can uses it as a help on identifying its {\em internal}
-timelines based on this plan. If they are not compatibles, then
-synchronization identifies a plan inconsistency  reflecting that the
-current plan is not valid anymore. \comment{fpy: I am here and will
-  develop further}
-
-\comment{following is not what I intended to express whatsoever
-  ... synchroniuzation od not deal with dispatching but execution
-  tracking ... it is though helped by the structure of the exisiting
-  plan and often identifies execution failures}
- Conversely, when the deliberation process eventually finds a
-solution, synchronization is informed about generating the planned
-state values \comment{should this read as ``...is informed about
-  generated state values..'' instead?} on the {\em external} state
-variables. This {\em external} plan defines the set of {\em goals} for
-this state variable managed by another {\em reactor}.
+stops, its starting point is the last partial plan resulting from
+deliberation. This gives an initial frame to the problem that helps to
+target the synchronization problem by ensuring that the new {\em
+  external} state updates are compatible with this plan. If so,
+synchronization propagates these events in the plan and can uses it as
+an aid in resolving its {\em current state flaw} by using the plan
+information on the exepected {\em internal} state evolution nearby the
+execution frontier. 
+% \kcomment{``use it
+%   as an aid to identify'' what though?}
+
+If they are not compatible, then {\em synchronization} identifies a
+plan inconsistency. This results in the retraction of decisions made
+during previous {\em deliberation} steps. % in order to
+% allow {\em synchronization} to identify its {\em internal} state.
+Such a relaxation can be viewed as a backtracking % on the plan search
+% triggered exogenously by the world evolution. It is the way our
+allowing the reactor to identify that its current plan has failed to
+execute triggering further deliberations steps to find an alternate
+strategy.
+
+Finally, when the deliberation process has identified a complete
+solution for its current {\em internal} goals, a part of its plan that
+applies to its {\em external} timelines is sent as a request to \rx to
+handle their dispatch to corresponding reactors.  Synchronization
+eventually receives {\em external } observations that confirms plan
+execution, in turn recording execution progress by having the part of
+the plan observed restricted to observation start times.
 
 Synchronization offers an important challenge for embedding planning
-in a situated agent, typically %  as always been to go around the asumptions made by
-% most classical planners. Specifically one
-an assumption to restrict the problem to ``offline planning'' defined
-in \cite{ghallab04} as follows:
-
-\begin{quotation}
+in a situated agent, typically an assumption to restrict the problem
+to ``offline planning'' defined in \cite{ghallab04}:
+
+{\scriptsize
+  \begin{quote}
     The planner is not concerned with any change that may occur in
-  $\Sigma$ \footnote{The world modeled by the plan domain.} while it is
-  planning; it plans for the given initial and goal states regardless
-  of the current dynamics, if any. 
-\end{quotation}
-
-This assumption, while helpful to reduce the scope of the planning
-problem, becomes problematic for planning within a highly dynamic
-environment. In a typical robot this implies that the only point where
-the planner can be informed about the current world state is prior to
-the commencement of a planning cycle by giving to the planner the
-initial state and current objectives. During the planning, it is
-assumed that the agent can maintin the world as perceived by the
-planner as stable. \cite{lemai04, lemai-chenevier2004} attempts to
-reduce the impact of planning by allowing local plan repair in the
-existing plan. In the eventuality that such repair is not feasible
-they require that their robot comes to a full stop until (re-)planning
-is complete. \comment{I may need more examples/refs here}. As Fig.
-\ref{fig:tick-exec} shows, this assumption is problematic for our
-architecture. Synchronization that tracks state evolution can occur
-several times during the reactor deliberation process. By making
-synchronization and deliberation share the same plan we provide a
-solution that relaxes this static world assumption. \comment{Need to
-  talk about IDEA and the fact that they had similar aspect even
-  though they were never very clear nor well articulated in their
-  papers}
+    $\Sigma$ \footnote{The world modeled by the plan domain.} while it
+    is planning; it plans for the given initial and goal states
+    regardless of the current dynamics, if any.
+\end{quote}}
+
+This assumption, while helpful to reduce the scope and complexity of
+the planning problem, becomes problematic for planning within a highly
+dynamic environment. In a typical robot this implies that the only
+point where the planner can be informed about the current world state
+is prior to the commencement of a planning cycle, by providing the
+planner an initial state and current objectives. During the planning,
+it is assumed that the agent can maintain the world as perceived by
+the planner. \cite{lemai04, lemai-chenevier2004} attempts to reduce
+the impact of planning by allowing local plan repair in the existing
+plan. In the eventuality that such repair is not feasible they require
+that their robot comes to a stop until (re-)planning is complete.
+
+This assumption is problematic for situated agents in general and as
+Fig. \ref{fig:tick-exec} shows, for \rx. Synchronization that tracks
+state evolution can occur several times during reactor
+deliberation. By making synchronization and deliberation share the
+same plan we provide a solution that relaxes this static world
+assumption. \texttt{IDEA} uses the same principle with all solvers in
+an agent sharing the same plan for communicating environmental change
+\cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
+-- similar to our {\em synchronization} solver -- was often the solver
+targeting the change at the execution frontier. 
+% Although, none of the papers we found were detailing how this was
+% handled in the architecture.
+\cite{Dias:2003ua} states:
+
+{\scriptsize
+  \begin{quote}
+  [\textellipsis] as long as the planning horizon over which the
+  deliberative planner is working never intersects the current
+  execution time, deliberative planning can operate in parallel with
+  reactive execution and does not affect the reactivity of the agent. 
+\end{quote}}
+
+This suggests a deliberate non overlap between {\em reactive planner}
+and {\em deliberative planners} horizons. In \rx, and more
+specifically in our \eu reactor the planning horizon of {\em
+  synchronization} -- limited to the current tick -- is a subset of
+the planning horizon of the deliberation steps -- starting at the
+current tick and with an extent based on latency $\lambda$ and
+look-ahead $\pi$ of the reactor. In this manner we avoid creating a
+conflict between the deliberation and synchronization by controlling
+the scheduling of these two process by the \rx agent.
 
@@
 are implemented independently to show how their interaction through
-the plan helps relax this assumption by making planning fully
-integrated in the reactor execution cycle while potentially allowing
-for more informed planning.
-
-\subsubsection{Synchronization identify internal state evolution}
+the plan helps relax this assumption. It does so by integrating
+planning in the reactor execution cycle even while allowing for more
+informed deliberation.
+
+\subsubsection{Synchronization and Identifying Internal State}
 \label{sec:arch:synch}
@@
 We first show how a reactor can track and identify its state in the
-context where it does not have any compelling need to deliberate. We
-assume that we have a reactor that has no future goal. Even in this
-condition, \rx requires each reactor to produce at every tick its {\em
-  internal} state. For those reactors not owning {\em internal}
-timelines, \rx makes sure that {\em external} observations are taken
-into account for every tick as shown in \cite{py10}. Such a
-requirement is enforced to ensure that all reactors share a consistent
-view of the world with advancing time.
+context where it does not have any compelling need to deliberate. In
+doing so we assume that we have a reactor that has no future
+goal. Under such a condition, \rx requires each reactor to produce at
+every tick its {\em internal} state. For those reactors not owning
+{\em internal} timelines, \rx makes sure that {\em external}
+observations are taken into account for every tick as shown in
+\cite{py10}. Such a requirement is enforced to ensure that all
+reactors share a consistent view of the world with advancing time.
 
 {\em External} state updates are managed by introducing these
-observation in the plan structure as facts and ensure that they remain
-compatible with the current plan mainainted by the reactor. The
+observations in the plan structure as facts and ensuring that they
+remain compatible with the current plan maintained by the reactor. The
 challenging part of synchronization is to decide the {\em internal}
-state values for the current tick. In the Shopping example for example
+state values for the current tick. In the Shopping example for example,
 we have the following code snippet which helps identify when we can
@@
 
-If \texttt{object.location} is {\em external} and the \texttt{Agent}
-is {\em internal}, a robot may encounter the situation where it is at
-the location where it can buy one (possibly among others)
-product. This results om a decision problem related to the state value
-of the \texttt{Agent} timeline where the robot can either {\em Buy} a
-specific product now, wait a little longer before doing so or just
-simply not buy any product at this location.
-
-In the \eu based reactor this decision requirement is a new type of
-flaw which forces full identification of internal state of the reactor
-for the current tick. Using this new flaw we can describe the
-synchronization process as:
-
-\begin{enumerate}
-
-\item Integrate the external state as provided by the owner of each
-  external timeline into the plan database.
-
-\item Propagate this information in the plan database following the
-  model $\mathcal{M}$ of this reactor.
-
-\item Resolve the current state value of each internal timelines.
-
-\end{enumerate}
+If \texttt{object.location} is {\em external} and \texttt{Agent} is an
+{\em internal} timeline, a robot may encounter the situation where it
+is at the location where it can buy one (possibly among others)
+product. This results in a decision-making problem related to the
+state value of the \texttt{Agent} timeline; the robot can either {\em
+  Buy} a specific product now, wait a little longer before doing so or
+just simply not buy any product at this location.
+
+In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
+new type of flaw which requires to identify the value of all the {\em
+  internal} state variables for the current tick.
+% forces identification \kcomment{identification
+%   of what? Probably need a little more explanation here.} of internal
+% state of the reactor for the current tick. 
+This flaw can be defined as:
+
+\begin{definition}
+  \label{def:csf}
+  A {\em current state flaw} indicates that a reactor's {\em internal}
+  timeline does not provide a fully defined state value for the
+  execution frontier $\tau$. Such a flaw can be the result of either:
+
+  \begin{itemize}
+  \item The absence of a token in this timeline which overlaps the
+    execution frontier $\tau$.
+  \item An open decision between terminating the last token for this
+    timeline or extending its end in the future.
+  \end{itemize}
+
+\end{definition}
 
@@
 at line $11$ which ensures that the last \texttt{Going} ends up in the
-location the agent actually wants to be at. This first level of
-relaxation will therefore allow the agent to use multiple
+location the agent actually wants the robot to be at. This first level
+of relaxation will therefore allow the agent to use multiple
 \texttt{Going} tokens should the first one fail to reach the
-objective.  % The problem can even be relaxed further if needed but
-% this relaxation gives already more flexibility for the reactor to
-% recover from local failures.
+objective.
+
 As the simple example above shows, synchronization involves a
 deliberation process with the risk that the search for a solution is
-not a straightforward sequence of unit decisions. Since such a process
-is required by the architecture at \emph{every} tick for \emph{every}
-reactor the model needs to be carefully designed.
+not a straightforward sequence of staright-forward decisions where 
+the resolution can be easily identified by the solver as any other
+option would leafd immediately to an inonsistency in the plan.
+% \kcomment{has a unit decision been explained in Europa?}. 
+Since such a process is required by the architecture at \emph{every}
+tick for \emph{every} reactor the model needs to be designed carefully.
 
@@
 there is no other alternative solution. We do this by removing all
-goals and keeping only recent observations. % These constraints are yet
-% again related to the fact that for the agent the only crucial
-% functionality of a reactor is to resolve successfully its
-% synchronization.
-
-The number of flaws between two steps \comment{between synch. and
-  execution?} can evolve for the following reasons that are exogenous
-to deliberation:
+goals while keeping only recent observations.
+
+{\em Synchronization} and the dispatching of new {\em goals} from \rx
+can occur between two steps and alter the plan. These alterations
+exogenous to the deliberation process dynamically impact the set of
+flaws that need to be resolved. New flaws are created when a new goal
+is received from \rx and conversely the recall of an objective could
+remove the corresponding flaws\footnote{However if the flaws related
+  to this goal were resolved, the removal of this goal may likely
+  create new flaws.}. The impact of {\em synchronization} can be more
+subtle as its alteration of the plan at the execution frontier is more
+drastic. We can still give a non exhaustive list of possible
+alterations of the flaws directly related to {\em synchronization}:
 
@@
 
-\item A new goal has been posted on a reactor timeline. This is
-  usually due to another reactor finding a partial plan and posting
-  the corresponding new objective.
-
-\item The outcome of synchronization created new flaws that did not
-  need to be resolved during synchronization.
-
-\item A previously requested goal has be recalled by the initial
-  requester. This can occur when the reactor's initial plan is no
-  longer valid, for instance with synchronization resulting in
-  inconsistency.
+\item Previous flaws directly related to the execution frontier have
+  been resolved by the {\em synchronization} solver.
+
+\item The {\em synchronization} solver may have created new flaws that
+  were outside of its scope but are now part of the scope of the {\em
+    deliberation} solver.
+
+\item The last observation has proved that the current plan is invalid
+  requiring the {\em synchronization} solver to relax the existing
+  plan to identify its current {\em external} state. (This case is
+  developed further on the next section).
 
@@
 
-\comment{this might be a subsubsection before someplace since it
-  introduces the concept of latency/lookahead.}
-These events could potentially produce new flaws to be resolved while
-any single deliberation step will likely reduce the number of flaws
-for resolution. The scope of planning for a reactor also plays a role
-in the deliberation process as we limit deliberation to the sliding
-temporal window $[\tau+\lambda, \tau+\lambda+\pi]$ \cite{py10} where:
-
-\begin{itemize}
-
-\item $\tau$ is the current tick indicative of the execution frontier
-
-\item $\lambda$ is the specified latency of this reactor, which is an
-  indicator of the maximum expected time for a reactor to produce a
-  plan.\footnote{Note that while it is highly recommended to select a
-    sound value, a failure to produce a plan within a chosen value for
-    this parameter is not considered as a critical failure of the
-    reactor.}
-
-\item $\pi$ is the planning look-ahead of the reactor. which indicates
-  how many ticks in the future this reactor is looking ahead during
-  deliberation.
-
-\end{itemize}
-
-\comment{Insert a figure here [AAMAS?] showing $\pi \lambda$} Such a
-scope allows filtering of tokens that are either necessarily ending
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.65\columnwidth]{figs/plan_window}
+  \caption{\small The planning window of the deliberation solver and
+    its relation to the {\em dispatch window}. The dark grey tokens
+    with dashed borders are excluded from deliberation as they do not
+    overlap the interval $[\tau, \tau+\lambda+\pi]$. All the other
+    tokens ({\color{green} green}) along with the goals received from
+    the agent ({\color{blue} blue}) which can overlap this window, are
+    evaluated.}
+  \label{fig:plan:window}
+\end{figure}
+
+When the {\em execution frontier} advances, the reactor planning scope
+moves further into the future. This is reflected by the {\em
+  deliberation} solver which uses a new deliberation scope one tick
+further in the future. As Fig. \ref{fig:plan:window} illustrates, such
+a scope allows filtering of tokens that are either necessarily ending
 before $\tau$ or necessarily starting after, focusing the planning
-problem on only those tokens that are within a ``reasonable''
-future. This window is taken into account by the \rx agent which will
-notify a reactor of new {\em internal} goals not only when they
-overlap this reactor specific window but also for the deliberation
-solver in order to reduce the number of tokens to be evaluated to
-reduce deliberation cost and performance. As time advances flaws that
-were initially beyond this temporal horizon will become active which a
-reactor can evaluate. This results in an apparent \emph{continuous
-  planning} of the reactor. When there are no more flaws present in
-the current planning scope a partial plan solution is found. This has
-two effects:
+problem on only those tokens that are within a ``reasonable'' future.
+This filtering window complements the \rx {\em dispatch
+  window} from Definition \ref{def:dispatch} which filters out at the
+agent level those goals that are before the reactor execution latency
+or too far into the future. The planning window reflect this filtering
+of the agent by applying to tokens that are generated internally by the
+reactor during its deliberation or the {\em goals} for which he
+planner decided to postpone after the planning window -- probably
+due to other tasks that needed completion beforehand.
+% The planning window inside of a reactor
+% refines this filtering further, as tokens can be excluded dynamically
+% from deliberation as their insertion in the plan and scheduling with
+% other tokens generated, can reduce their temporal scope outside of the
+% planning scope \kcomment{ this last sentence should be reworked}. 
+This allow the planner to focus on its planning horizon and report
+decision on flawss thgat are beyond this scope on later ticks. As, 
+time advance, these future flaws will eventuallty be part of the
+planning window of the reactor and will be then resolved along with
+potential new goals received from the agent.
+% This helps the deliberation solver reduce the number of tokens to be
+% evaluated reducing deliberation cost and increasing performance. As
+% time advances, flaws that were initially beyond this temporal horizon
+% will become active which a reactor can evaluate.
+ This results in an apparent \emph{continuous planning} of the reactor. When there are no
+more flaws present in the current planning scope a partial plan
+solution is found. This has two effects:
 
@@
 solver but also enforced by the heuristic order of the decision point
-resolution choices for the {\em current  state flaw} presented in
-Proposition \ref{prop:csf:resolve}. Indeed, as we stated then the two
-first choices tend to privilege a conservative view in respect to the
-existing plan. It attempts first to extend the previous observation we
-produced for this {\em internal} timeline in order to avoid to commit
-to early in the plan advance. If the formaer failed it will then
-attempt to start the next exisiting token in the plan which alow the
-reactor to advance its plan execution. Both of these two initial
-choices atetempt to resolve the plan with the least amount of
-perturbation of the plan. They just constrain further the start and
-end time of the exisiting tokens which were provided by the previous 
-deliberations steps. This means that in a nominal situation where the
-plan this reactor produced is properly executed, the synchronization
-remains limited to restrict time points in the plan and progate their
-impact. Most of the heavy dutty work of creating new tokens in the
-plan and schedule them -- both being much more difficult problems 
-to resolve -- have been managed during deliberation steps. In that
-sense we see that most of the time synchronization is limited to very
-few node in \eu search space with often no backtrack required. The
-synchronization may become slightly nore costfull when we need to
-identify a new token to be injected by the plan (the 3rd choice in
-Proposition \ref{prop:csf:resolve}) as it needs to explore on tis own
-all the possible token types this timeline accept as value. This is
-the reason this choice is evaluated last in our flaw resolution in
-order to avoid such complex action during synchronization unless 
-necessary. In that sense,  ansd due to carefull choice on our
-synchrponization heurisitic we see that the plan produced during \rx
-deliberation steps helps sthe synchronization process to be more
-efficient.
+resolution choices for the {\em current state flaw} shown in
+Proposition \ref{prop:csf:resolve}. As noted, the first two choices
+are biased towards a conservative view with respect to the existing
+plan. The heuristics first attempt to extend the previous observation
+for this {\em internal} timeline to avoid early commitment. If this
+fails \rx attempts to start the next existing token in the plan to
+advance plan execution. Both choices attempt to resolve state with the
+least amount of plan perturbation by constraining further the start
+and end time of existing tokens generated by previous deliberations
+steps. By doing so, in a nominal situation, synchronization remains
+limited to restrict time points in the plan and propagates their
+impact. Most of the computational effort of creating new tokens in the
+plan and scheduling them onto timelines are managed during
+deliberation steps. Consequently, most of the time synchronization is
+limited to very few nodes in the \eu search space often requiring no
+backtracking. Fig. \ref{fig:shop:exec} shows how the existing plan can
+support the synchronization process. The observation \texttt{Going}
+occuring at $\tau'$ is fully compatible with the token already present
+in our plan. By updating the plan and progating equality constraints,
+the resolution of the {\em current state flaw} for the {\em Agent}
+timeline becomes obvious as a solution, as the token \texttt{Go} from
+the plan presents itself.
+
+Synchronization may become computationally more expensive when we need
+to identify a new token to be injected into the plan (the third choice
+in Proposition \ref{prop:csf:resolve}), as it needs to explore the
+outcome of inserting all possible token types the timeline can accept.
+This is the reason this choice is evaluated last in our flaw
+resolution strategy.
+
+% In that sense, and thanks to the carefull choice on our
+% synchronization heurisitic we see that the plan produced during \rx
+% deliberation steps helps sthe synchronization process to be more
+% efficient.
 

### 2012-02-21 arch.tex
updated constraint table in correct location within europa.tex and corrected arch.tex typos

--- 2012-02-19 arch.tex
@@
 \label{sec:arch}
-
-% {\em\tiny This section introduces shortly T-REX architecture but move
-%   quickly the focus on a single reactor and gross concepts on the
-%   architecture:
-%   \begin{itemize}
-%   \item For t-trex the whole decision and control problem is reduced
-%     around the state variables (or timeline construct). 
-%   \item This reduce the execution tracking problem for each reactor on a state
-%     identification problem (deduce internal state from the external
-%     state), and the control problem to goal posting which will trigger
-%     deliberation on the owner of the corresponding timeline. 
-%   \item the architecture itself see each reactor as a black box and
-%     therefore each reactor can implement its own mechanism in order to
-%     resolve both synchronization and deliberation. Still we provide a
-%     reactor based on the europa framework that leverage the automated
-%     planning capabilities in order to do model based planning and
-%     execution with a rich representation of resources and time.
-%   \end{itemize}}
 
@@
 level this execution latency is used to define the {\em dispatch
-  window} of the same state variable.% : \comment{Needed to say that
-  % early on but the flow of this paragraph is not very tight}
-\kcomment{Looks fine to me unless you want to clarify further.}
+  window} of the same state variable.
 
@@
 reactor needs to replan due to an unexpected event, it will recall all
-requests it produced that have not been observed \kcomment{replace
-  ``observed'' with ``not incorporated by the owner reactor''?}
-yet. This recall is sent to the reactor that received the goal(s) to
-remove it/them from its current set of objectives.
+requests produced that have yet to be executed.  This recall is also
+sent to the reactor(s) that received the goal(s) to remove it/them
+from its current set of objectives.
 
@@
   goal.
-% to whichever reactor that is responsible for these. % \comment{By
-    % extending I may have saud things I developed below ... will have
-    % to check}
 
@@
 
-Both these deliberative \kcomment{``deliberative??''; something else
-  or remove altogether perhaps?} functions have different scope and
-temporal constraints. The \texttt{synchronize} call is considered as
-atomic and will focus only on state inference for the current tick by
-identifying the current value of {\em internal} state variables from
-the latest updates on {\em external} state variables. This process
-propagates up in the reactor dependency graph at every tick to ensure
-that all reactors are aware of the current state of the
-world. \texttt{step} relates to a single decision step of the planning
-process focused on the future evolution of state variables; it can
-take multiple steps spanning multiple ticks for a reactor to produce a
-plan that is both complete and valid. How the reactor identifies that
-its plan is complete and valid is its responsibility and depends on
-the way each reactor resolves its plan at each {\em step}. The general
-idea is that the reactor is able to find a valid strategy in order to
-satisfy as many of its {\em internal} goals as possible. For \eu, this
-means that the solver is unable to identify any further flaws to be
-resolved in the plan.
+Both these functions have different scope and temporal constraints.
+The \texttt{synchronize} call is considered as atomic and will focus
+only on state inference for the current tick by identifying the
+current value of {\em internal} state variables from the latest
+updates on {\em external} state variables. This process propagates up
+in the reactor dependency graph at every tick to ensure that all
+reactors are aware of the current state of the world. \texttt{step}
+relates to a single decision step of the planning process focused on
+the future evolution of state variables; it can take multiple steps
+spanning multiple ticks for a reactor to produce a plan that is both
+complete and valid. How the reactor identifies that its plan is
+complete and valid is its responsibility and depends on the way each
+reactor resolves its plan at each {\em step}. The general idea is that
+the reactor is able to find a valid strategy in order to satisfy as
+many of its {\em internal} goals as possible. For \eu, this means that
+the solver is unable to identify any further flaws to be resolved in
+the plan.
 
@@
   external} state updates are compatible with this plan. If so,
-synchronization propagates these events in the plan and can uses it as
-an aid in identifying its {\em internal} timelines \kcomment{``use it
-  as an aid to identify'' what though?} based on this plan.
+synchronization propagates these events in the plan and can use it as
+an aid in resolving its {\em current state flaw}. It does so by using
+the plan information on the exepected {\em internal} state evolution
+near the execution frontier.
 
@@
 plan inconsistency. This results in the retraction of decisions made
-during the \kcomment{previous?} {\em deliberation} step. % in order to
-% allow {\em synchronization} to identify its {\em internal} state.
-Such a relaxation can be viewed as a backtracking % on the plan search
-% triggered exogenously by the world evolution. It is the way our
-allowing the reactor to identify that its current plan has failed to
-execute triggering further deliberations steps to find an alternate
-strategy.
+during previous {\em deliberation} steps.  Such a relaxation can be
+viewed as a backtracking allowing the reactor to identify that its
+current plan has failed to execute triggering further deliberations
+steps to find an alternate strategy.
 
@@
 observations in the plan structure as facts and ensuring that they
-remain compatible with the current plan mainainted by the reactor. The
+remain compatible with the current plan maintained by the reactor. The
 challenging part of synchronization is to decide the {\em internal}
@@
 In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
-new type of flaw which forces identification \kcomment{identification
-  of what? Probably need a little more explanation here.} of internal
-state of the reactor for the current tick. This flaw can be defined
-as:
+new type of flaw which requires identification of all values of {\em
+  internal} state variables for the current tick. This flaw can be
+defined as:
 
@@
     reactor. The reactor receives new observations when they are
-    produced by the owner(s) of its internal timelines. The line
-    \kcomment{what line???} after the last token of each timeline
-    represent the domain of possible values for the end of this
-    token. At every tick $\tau$ the reactor needs to integrate the
-    {\em External} state information it received and -- based on its
-    model $\mathcal{M}$ -- resolve its {\em Internal} state that will
-    then be provided by the agent to other reactors using these state
+    produced by the owner(s) of its internal timelines. The open
+    ``holes'' after the last token of each timeline represents the
+    domain of all possible values for the end of this token. At every
+    tick $\tau$ the reactor needs to integrate the {\em External}
+    state information it received and -- based on its model
+    $\mathcal{M}$ -- resolve its {\em Internal} state that will then
+    be provided by the agent to other reactors using these state
     variables.}
@@
 
-As the simple example above shows, synchronization involves a
-deliberation process with the risk that the search for a solution is
-not a straightforward sequence of unit decisions \kcomment{has a unit
-  decision been explained in Europa?}. Since such a process is
-required by the architecture at \emph{every} tick for \emph{every}
-reactor the model needs to be designed carefully.
+As the example above shows, synchronization involves a deliberation
+process. The risk is such that the search for a solution is not a
+simple sequence of decisions where the resolution can be easily
+identified by the solver as any other option would lead immediately to
+n plan inconsistency. Since such a process is required by the \rx at
+\emph{every} tick for \emph{every} reactor the model needs to be
+designed carefully.
 
@@
 problem on only those tokens that are within a ``reasonable'' future.
-This filtering window complements the \rx {\em dispatch
-  window} from Definition \ref{def:dispatch} which filters out at the
-agent level those goals that are before the reactor execution latency
-or too far into the future. The planning window inside of a reactor
-refines this filtering further, as tokens can be excluded dynamically
-from deliberation as their insertion in the plan and scheduling with
-other tokens generated, can reduce their temporal scope outside of the
-planning scope \kcomment{ this last sentence should be reworked}. This
-helps the deliberation solver reduce the number of tokens to be
-evaluated reducing deliberation cost and increasing performance. As
-time advances, flaws that were initially beyond this temporal horizon
-will become active which a reactor can evaluate. This results in an
-apparent \emph{continuous planning} of the reactor. When there are no
-more flaws present in the current planning scope a partial plan
-solution is found. This has two effects:
+This filtering window complements the \rx {\em dispatch window} from
+Definition \ref{def:dispatch} which filters out at the agent level
+those goals that are before the reactor execution latency or too far
+into the future.  The planning window refines this filtering further,
+as tokens can be excluded dynamically from deliberation as their
+insertion in the plan and scheduling with other generated tokens can
+alter their temporal scope to be outside of the planning window.  This
+allows the planner to focus on its planning horizon and report
+decision on flaws that are beyond this scope. As, time advances, these
+potential future flaws will eventually be part of the planning window
+of the reactor and will be then resolved along with potentially new
+goals.  This in turn, results in an apparent \emph{continuous
+  planning} of the reactor. When there are no more flaws present in
+the current planning scope a partial plan solution is found. This has
+two effects:
 

### 2012-02-21 europa.tex
updated constraint table in correct location within europa.tex and corrected arch.tex typos

@@
 
-\comment{explain how inference is used to detect perform bounds
-  propagation, domain reduction and to detect constraint violations.} 
-- talk about AC3 implementation and extension mechanisms
-- talk about europa's constraint library
+In \eu,  all the Variables and Constraints from a problem domain domain description are kept inside the Constraint Reasoning Engine (CRE).
+The CRE provides a framework where any number of constraint types can collaborate to perform inference and prune variable domains using Bounds Propagation and Domain Reduction techniques.
+\eu uses a Propagator pattern, in which a Propagator object keeps track of a set of constraint instances of the same type and is responsible for ensuring that whenever any variable changes or constraints are introduced or dropped, all related variables are updated through inference.
 
-The Temporal Network: The notion of time is central to temporal
-planning. \eu uses variables to explicitly represent timepoints for
+In a planning problem, variable or constraint changes can come
+directly from an end user (by introducing new activities or goals and
+relating them to the rest of the plan through constraints, by changing
+the start time of an activity, etc), or can come from an automated
+planner (by expanding subgoals, introducing precedence constraints to
+resolve mutual exclusion conflicts, etc). When these changes happen,
+each Propagator is only able to ensure consistency for the constraints
+that are associated with it.  To maintain global consistency, the CRE
+implements a variant of the \texttt{AC-3} algorithm \cite{mackworth77}
+where it asks Propagators to perform inference until the Constraint
+Network returns to an stable state.
+
+\comment {add europa's AC-3 algorithm here}
+
+\input{constraints-table}
+
+A \eu user can introduce new constraint types by implementing
+specialized Constraint and Propagator classes in C++, however, \eu
+provides a built-in with over 25 arithmetic, temporal, set and
+resource constraint types so that a large number of problem domains
+can be modeled without having to introduce new constraint
+types. Tables \ref{{tab:calconst, tab:setconst,
+    tab:objhierarchy,tab:misc} show the constraint types.
+
+Temporal and resource constraints are specially important to model
+real-life planning domains, so it is worth going into some more detail
+about how \eu's built-in support works for those constraint types.
+
+\textit{Temporal Constraints:} \eu uses variables to explicitly represent timepoints for
 plan activities and states. Constraints among timepoints provide a
@@
 
-\item \textbf{Reservoir} resources can have both 'consume' and
-  'produce' tokens. Note that these tokens do NOT have the usual
-  start/end/duration token variables, since occur at a single instant
-  in time. Instead, they have a single variable, 'time' used to
-  represent that instant.
+\item \textbf{Reservoir} resources can have both \texttt{consume(resource,time,quantity)} and
+  \texttt{produce(resource,time,quantity)} constraints. Production and consumption is instantaneous (it happens at a single point in time)
 
-\item \textbf{Reusable} resources have 'uses' tokens that use a
-  quantity for the duration of the token. These tokens do have the
-  usual start/end/duration variables.
+\item \textbf{Reusable} resources have \texttt{uses(resource,start,duration,quantity)} constraints that use a
+  quantity of the resource for a specific duration, then release it.
 
-\item \textbf {Unary} resources are reusable resources with unit
-  quantity and have 'use' tokens (again, with the usual
-  start/end/duration variables).
+\item \textbf {Unary} resources are Reusable resources with capacity=1, which allows \eu to perform 
+some run time savings compared to the generic Reusable class.
 \end{enumerate}
 
-\comment{talk about resource envelope computation by using MaxFlow formulation.}
+\eu offers configuration options so that in addition to the usual constraint violations when a resource is completely depleted, plan flaws can be generated whenever a resource level 
+is outside a specific upper or lower limit, or when instantaneous production or consumption exceeds a specific threshold.
+
+Recall that \eu supports a flexible temporal representation, where time is represented by intervals, instead of single values.
+In this context, determining whether a resource is the source of a plan flaw or a constraint violation in general is more difficult than just computing a profile
+of the resource and checking against some limits. 
+In the context of time flexibility, there are many resource profiles that are possible, depending on when exactly each production/consumption event happens out of all the possible values that are in the domain of the variables that represent the time for the event.
+\cite{mus02} proposed an algorithm where the variables that represent the time for production/consumption events are placed in a temporal network, similar to the network used by the Temporal Propagator above, then the constraints on the resource can be seen as quantities flowing through that network.
+Computing the maximum flow will yield the maximum resource usage, and thus a lower bound on the resource level, and computing a minimum flow will similarly yield an upper bound.
+Since maximum flow on a temporal network can be computed in polynomial time, \eu takes advantage of this algorithm to efficiently compute upper and lower bound envelopes on the resource level.
+
+\comment{figure :provide snapshot of resource envelopes generated by europa}
+
+Resource envelopes can be very useful when trying to figure how brittle a plan in the face of changing activity durations or start times.
+They can also be useful when searching for plans, although they are relatively expensive to compute so their ability to prune the search space must be weighted against the ability to search faster \comment{reference Barreiro,Morris, Bresina paper on grounded inference}.
 

### 2012-02-21 trunk/europa.tex
updated constraint table in correct location within europa.tex and corrected arch.tex typos

@@
 
-In a planning problem, variable or constraint changes can come directly from an end user (by introducing new activities or goals and relating them to the rest of the plan through constraints, by changing the start time of an activity, etc), or can come from an automated planner (by expanding subgoals, introducing precedence constraints to resolve mutual exclusion conflicts, etc).
-When these changes happen, each Propagator is only able to ensure consistency for the constraints that are associated with it.
-To maintain global consistency, the CRE implements a variant of the AC-3 algorithm \cite{mackworth77} where it asks Propagators to perform inference until the Constraint Network returns to an stable state.
+In a planning problem, variable or constraint changes can come
+directly from an end user (by introducing new activities or goals and
+relating them to the rest of the plan through constraints, by changing
+the start time of an activity, etc), or can come from an automated
+planner (by expanding subgoals, introducing precedence constraints to
+resolve mutual exclusion conflicts, etc). When these changes happen,
+each Propagator is only able to ensure consistency for the constraints
+that are associated with it.  To maintain global consistency, the CRE
+implements a variant of the \texttt{AC-3} algorithm \cite{mackworth77}
+where it asks Propagators to perform inference until the Constraint
+Network returns to an stable state.
 
@@
 
-A \eu user can introduce new constraint types by implementing specialized Constraint and Propagator classes in C++, however, \eu provides a built-in with over 25 arithmetic, temporal, set and resource constraint types so that a large number of problem domains can be modeled without having to introduce new constraint types.
-\comment{insert table from http://code.google.com/p/europa-pso/wiki/ConstraintLibraryReference ?
-}
-Temporal and resource constraints are specially important to model real-life planning domains, so it is worth going into some more detail about how \eu's built-in support works for those constraint types.
+\input{constraints-table}
+
+A \eu user can introduce new constraint types by implementing
+specialized Constraint and Propagator classes in C++, however, \eu
+provides a built-in with over 25 arithmetic, temporal, set and
+resource constraint types so that a large number of problem domains
+can be modeled without having to introduce new constraint
+types. Tables \ref{tab:calconst}, \ref{tab:setconst},
+\ref{tab:objhierarchy} and \ref{tab:misc} show various constraint
+types.
+
+Temporal and resource constraints are specially important to model
+real-life planning domains, so it is worth going into some more detail
+about how \eu's built-in support works for those constraint types.
 
@@
 
-Recall that \eu supports a flexible temporal representation, where time is represented by intervals, instead of single values.
-In this context, determining whether a resource is the source of a plan flaw or a constraint violation in general is more difficult than just computing a profile
-of the resource and checking against some limits. 
-In the context of time flexibility, there are many resource profiles that are possible, depending on when exactly each production/consumption event happens out of all the possible values that are in the domain of the variables that represent the time for the event.
-\cite{mus02} proposed an algorithm where the variables that represent the time for production/consumption events are placed in a temporal network, similar to the network used by the Temporal Propagator above, then the constraints on the resource can be seen as quantities flowing through that network.
-Computing the maximum flow will yield the maximum resource usage, and thus a lower bound on the resource level, and computing a minimum flow will similarly yield an upper bound.
-Since maximum flow on a temporal network can be computed in polynomial time, \eu takes advantage of this algorithm to efficiently compute upper and lower bound envelopes on the resource level.
+Recall that \eu supports a flexible temporal representation, where
+time is represented by intervals, instead of single values. In this
+context, determining whether a resource is the source of a plan flaw
+or a constraint violation in general is more difficult than just
+computing a profile of the resource and checking against some limits.
+In the context of time flexibility, there are many resource profiles
+that are possible, depending on when exactly each
+production/consumption event happens out of all the possible values
+that are in the domain of the variables that represent the time for
+the event.  \cite{mus02} proposed an algorithm where the variables
+that represent the time for production/consumption events are placed
+in a temporal network, similar to the network used by the Temporal
+Propagator above, then the constraints on the resource can be seen as
+quantities flowing through that network. Computing the maximum flow
+will yield the maximum resource usage, and thus a lower bound on the
+resource level, and computing a minimum flow will similarly yield an
+upper bound. Since maximum flow on a temporal network can be computed
+in polynomial time, \eu takes advantage of this algorithm to
+efficiently compute upper and lower bound envelopes on the resource
+level.
 
@@
 
-Resource envelopes can be very useful when trying to figure how brittle a plan in the face of changing activity durations or start times.
-They can also be useful when searching for plans, although they are relatively expensive to compute so their ability to prune the search space must be weighted against the ability to search faster \comment{reference Barreiro,Morris, Bresina paper on grounded inference}.
+Resource envelopes can be very useful when trying to figure how
+brittle a plan in the face of changing activity durations or start
+times. They can also be useful when searching for plans, although they
+are relatively expensive to compute so their ability to prune the
+search space must be weighted against the ability to search faster
+\cite{6007772}.
 

### 2012-02-21 concepts.tex
post-telecon updates to title europa sections

@@
 
-\eu follows the representation outlined by the CAIP framework, it also adds a few elements to it that are necessary for an efficient implementation of constraint-based planning.
-Below is a brief summary of the main elements, followed by a detailed description for each one:
+\eu follows the representation outlined by the CAIP framework, it also
+adds a few elements to it that are necessary for an efficient
+implementation of constraint-based planning. Below is a brief summary
+of the main elements, followed by a detailed description for each one:
 
 \begin{enumerate}
-\item \textbf{Tokens} Temporally scoped entities that correspond to Intervals in the CAIP framework, are used to represent  state and actions in time.
-A\textbf{Token State model} is defined to support an efficient implementation of plan-space search algorithms.
 
-\item \textbf{Domains, Variables and Constraint} are used to represent restrictions in the problem domain in terms of a CP formulation.
+\item \textbf{Tokens} Temporally scoped entities that correspond to
+  Intervals in the CAIP framework, are used to represent state and
+  actions in time. A\textbf{Token State model} is defined to support
+  an efficient implementation of plan-space search algorithms.
 
-\item \textbf{Domain Rules} are used to describe internal and external relationships on and between tokens, as presecribed by the CAIP framework.
+\item \textbf{Domains, Variables and Constraint} are used to represent
+  restrictions in the problem domain in terms of a CP formulation.
 
-\item \textbf{Objects} are used to represent problem domains in a more modular and scaleable way.
-\textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
+\item \textbf{Domain Rules} are used to describe internal and external
+  relationships on and between tokens, as presecribed by the CAIP
+  framework.
+
+\item \textbf{Objects} are used to represent problem domains in a more
+  modular and scaleable way. \textbf{Timelines and Resources} are
+  built in implementations of Object types that appear frequently in
+  real life domains.
 
@@
   \begin{enumerate}
+
   \item \textit{String}: sequences of characters, like "Red",
@@
   \begin{enumerate}
+
   \item \textit{Enumerated}: Finite sets over any data type, specified
     explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
+
   \item \textit{Interval}: Only for Numeric data types, they are
@@
     [5.0,100.0], etc.
+
   \end {enumerate}
@@
 \item[\textbf{Tokens}] In first order logic, a predicate defines a
-  relation between objects and properties \comment{use Genesereth citation?}.  In
-  \eu, we define such relations between variables whose domains are
-  sets of objects and sets of properties to describe state and
-  behavior. For example, we might use a predicate \texttt{At($a$,$l$)}
-  to indicate that agent $a$ is at location $l$, or a predicate
+  relation between objects and properties \cite{gen87}.  In \eu, we
+  define such relations between variables whose domains are sets of
+  objects and sets of properties to describe state and behavior. For
+  example, we might use a predicate \texttt{At($a$,$l$)} to indicate
+  that agent $a$ is at location $l$, or a predicate
   \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
@@
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \comment{Maybe: add small
-    example?}
+  overlap may occur among them \cite{mus94}.  \comment{Maybe: add
+    small example?}
 
@@
 
-  \comment{Use Fig 4 and edit:: Add figure showing an AgentLocation timeline with an
-    At(Home) token with start=0 and end open, an empty AgentBag
-    timeline, and a Own(Drill) token with start=10 floating below both
-    timelines}
+  \comment{Use Fig 4 and edit:: Add figure showing an AgentLocation
+    timeline with an At(Home) token with start=0 and end open, an
+    empty AgentBag timeline, and a Own(Drill) token with start=10
+    floating below both timelines}
 
-Notice that the \texttt{Own(Drill)} token that represents the goal is
+  Note that the \texttt{Own(Drill)} token that represents the goal is
   not immediately placed on the \texttt{AgentBag} timeline since \eu
-must determine whether it can achieve that goal and how. When the goal
-is posted and before any decisions are made on it, the
+  must determine whether it can achieve that goal and how. When the
+  goal is posted and before any decisions are made on it, the
   \texttt{Own(Drill)} token is created in an INACTIVE state, then \eu
@@
   of them (if there is more than one it can try them one at a time),
-  in this case, the state of the token would change to MERGED.
+  in this case, the state of the token would change to
+  \texttt{MERGED}.
 
@@
 
-As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED
-or REJECTED, this lifecycle is illustrated in the figure below.
+As we can see, a token's state can go from \texttt{INACTIVE} to
+\texttt{ACTIVE, MERGED} or \texttt{REJECTED}, this lifecycle is
+illustrated in the figure below.
 
@@
   Figure xx illustrates the entities and relations involved in
-specifying such a rule on a \texttt{Go} action. The token on which the
-rule applies is referred to (\texttt{Go}) as the master. Each
+  specifying such a rule on a \texttt{Go} action. The token on which
+  the rule applies is referred to (\texttt{Go}) as the master. Each
   \texttt{At} token required by the master is referred to as a
@@
   enumerations for the remainder. Application of a rule on a token can
-thus cause slave tokens, variables, and constraints to be introduced.
+  thus cause slave tokens, variables, and constraints to be
+  introduced.
 
@@
   is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting (slave) tokens to be in the
-plan in order for the master to be valid.
+  rule imposes a requirement for supporting (slave) tokens to be in
+  the plan in order for the master to be valid.
 

### 2012-02-21 setobook.tex
added more latex markup to europa.tex and included europa.tex in setobook.tex
citations added to europa text
post-telecon updates to title europa sections

@@
 
-% \usepackage[agsm]{harvard}
+% \titlerunning{Towards Deliberative Control of Autonomous Robots}
+% \authorrunning{K.Rajan et. al.}
 
-\titlerunning{Towards Deliberative Control of Autonomous Robots}
-\authorrunning{K.Rajan et. al.}
-
-\title*{\textbf{\sc Towards Deliberative Control of Autonomous Robots}}
+\title*{\textbf{\sc Towards Deliberative Control in Marine Robotics}}
 \author{\normalsize Kanna~Rajan\inst{1} \and Fr\'ed\'eric Py\inst{1}
@@
 
+\input{related.tex}
+
+\input{europa-basics.tex}
+
 \input{concepts.tex}
 
-\input{related.tex}
+\input{europa.tex}
 
@@
 
-Authors Rajan and Py are funded by a block grant from the David and
-Lucile Packard Foundation to MBARI.
+Rajan and Py are funded by a block grant from the David and Lucile
+Packard Foundation to MBARI. They thank their collaborators J. Ryan,
+J. Harvey, R. Vrijenhoek and J. Bellingham at MBARI. In addition they
+thank colleagues M. del Alto, D. Korsmeyer and S. Zornetzer at NASA
+Ames who made \eu available as open source. The authors are indebted
+to N. Muscettola for his ideas and early work at NASA Ames which forms
+the basis of this work.
 

### 2012-02-21 trunk/europa-basics.tex
post-telecon updates to title europa sections

--- 2012-02-16 trunk/europa-basics.tex
@@
 
-\subsection{The basics}
+\paragraph {A brief note on \eu history and notation} The earliest
+origin of what we will call \eu in this chapter, is from the
+\texttt{HSTS} planner \cite{mus94} which was used for a number of
+demonstrations for space telescopes as noted in Section
+\ref{sec:rabeyond}. \texttt{HSTS} was the
+\texttt{LISP}-based\footnote{This was the first (and only to our
+  knowledge) software ever to been written in \texttt{LISP} to be
+  flown in space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
+the planner was implemented at NASA Ames which went on to be used for
+the \texttt{MAPGEN} \cite{bresina05} and continues to be used to date,
+in the mission-critical uplink process for the command/control of the
+Mars Exploration Rovers (MER) mission. A vastly re-factored, higher
+performance open source version \cite{europapso} was named \eut and is
+the subject of this chapter. Note however, in using the name \eu, we
+refer to the \emph{fundamental ideas as well as the planning paradigm}
+even if there exists a specific implementation.
+
+\subsection{The Basics}
 \label{sec:basics}
@@
 
-\begin{quotation}
-
-  In the near future, a personal robot sets out to buy a gallon of milk
-  This involves a number of tasks: obtain keys, obtain wallet,
-  start car, drive to store, find and obtain milk, purchase milk, etc.
-  The embedded planner has to have a ``model'' of the world in which it
-  lives and has to use the task primitives in this model to structure
-  the actions so it achieves its goal. Constraints control when certain
-  tasks can or cannot occur. For example the robot must obtain the keys
-  and wallet \emph{before} driving to the store and pick up the milk
-  \emph{before} purchasing it.
-
-\end{quotation}
+{\small
+  \begin{quote}
+    In the near future, a personal robot sets out to buy a gallon of
+    milk This involves a number of tasks: obtain keys, obtain wallet,
+    start car, drive to store, find and obtain milk, purchase milk,
+    etc.  The embedded planner has to have a ``model'' of the world in
+    which it lives and has to use the task primitives in this model to
+    structure the actions so it achieves its goal. Constraints control
+    when certain tasks can or cannot occur. For example the robot must
+    obtain the keys and wallet \emph{before} driving to the store and
+    pick up the milk \emph{before} purchasing it.
+\end{quote}
 
 For such a robot the milk buying plan at the store might look like
-\comment{Figure needed}.
+that shown in Fig. \ref{fig:shop:shopping}.
 

### 2012-02-21 trunk/europa.tex
post-telecon updates to title europa sections

@@
 described above and makes them easily accessible through modeling and
-Application Programming Interface (API) layers.  A user will typically use the modeling facilities to
-inject a problem domain description, and the API to extract and
-manipulate plans, and to obtain information about plan flaws and
-constraint violations at any stage of the planning process.
+Application Programming Interface (API) layers.  A user will typically
+use the modeling facilities to inject a problem domain description,
+and the API to extract and manipulate plans, and to obtain information
+about plan flaws and constraint violations at any stage of the
+planning process.
 
@@
 \cite{mackworth77}. The CRE is designed so that specialized reasoning
-algorithms for new constraints can be easily and efficiently
-plugged in.
+algorithms for new constraints can be easily and efficiently plugged
+in.
 
@@
 
-The \textbf{Rules Engine} (RE) modularly adds the semantics of domain rules that describe dependencies between tokens.
+The \textbf{Rules Engine} (RE) modularly adds the semantics of domain
+rules that describe dependencies between tokens.
 
@@
 this motivated the creation of modeling languages that support those
-definitions in a much more natural and compact way. NDDL is the main
-modeling language for \eu and will be described in detail in the
-Modeling section below. ANML is a language under development that
+definitions in a much more natural and compact way. \texttt{NDDL} (New
+Domain Description Language [pronounced ``noodle'']) \cite{NDDL} is
+the main modeling language for \eu and will be described in detail in
+the Modeling section below. ANML is a language under development that
 provides a more intuitive syntax \cite{smith08}.
@@
 language for models, goals and plans is of great importance. For \eu,
-NDDL (New Domain Description Language [pronounced ``noodle''] allows
-the user to specify the different components of a problem domain in a
-concise way. The main characteristics that make NDDL a powerful tool
-for describing problem domains are:
+\texttt{NDDL} allows the user to specify the different components of a
+problem domain in a concise way. The main characteristics that make
+\texttt{NDDL} a powerful tool for describing problem domains are:
 
@@
 
-\item Object Oriented: NDDL supports classes and inheritance in
-  similar fashion to popular Object-Oriented languages like Java and
-  C++. It also supports polymorphism for some of the planning
+\item Object Oriented: \texttt{NDDL} supports classes and inheritance
+  in similar fashion to popular Object-Oriented languages like Java
+  and C++. It also supports polymorphism for some of the planning
   components. Using object classes and instances is a time-tested
-  approach to naturally describe problem domains \comment{OO
-    citation}.
+  approach to naturally describe problem domains \cite{rumbaugh}.
 
@@
 
-To understand how the most important elements of NDDL work, we show
-how the Shopping Agent problem could be modeled:
+To understand how the most important elements of \texttt{NDDL} work,
+we show how the Shopping Agent problem could be modeled:
 
@@
 
-\item \textbf{Object Types}: NDDL supports an object-oriented
+\item \textbf{Object Types}: \texttt{NDDL} supports an object-oriented
   approach, where common data and behavior can be abstracted as a
@@
   \textit{AgentLocation} class to keep track of where the agent is at
-  any point in time. NDDL supports a single rooted class hierarchy so
-  that state and behavior can be extended through inheritance. We took
-  advantage of that by making \textit{AgentLocation} inherit from
-  \eu's built-in \textit{Timeline} class which doesn't allow
-  predicates to overlap in time for the same object instance, in this
-  way we can ensure that and agent can only be at one place at a time.
+  any point in time. \texttt{NDDL} supports a single rooted class
+  hierarchy so that state and behavior can be extended through
+  inheritance. We took advantage of that by making
+  \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline}
+  class which doesn't allow predicates to overlap in time for the same
+  object instance, in this way we can ensure that and agent can only
+  be at one place at a time.
 
@@
 
-Domain rules can be defined in NDDL in the context of a Predicate or
-an Action. In the example above we can see the main elements of a rule
-definition:
+Domain rules can be defined in \texttt{NDDL} in the context of a
+Predicate or an Action. In the example above we can see the main
+elements of a rule definition:
 
 \begin{enumerate}
+
 \item \textbf{Variables}: A Token Type (predicate or action) is part
@@
   \texttt{[meets, equals, metby]} operators are used to specify
-    subgoals (Allen operators are explained in more detail below),
-    along with the temporal relationship each token representing a
-    subgoal has with the the token representing the \texttt{Agent::Go}
-    action. Also, \texttt{[condition, effect]} annotations can be used
-    when defining a subgoal inside an action, \eu doesn't immediately
-    associate any semantics with them, but they can be extremely
-    useful for search modules when creating plans.
+  subgoals (Allen operators are explained in more detail below), along
+  with the temporal relationship each token representing a subgoal has
+  with the the token representing the \texttt{Agent::Go} action. Also,
+  \texttt{[condition, effect]} annotations can be used when defining a
+  subgoal inside an action, \eu doesn't immediately associate any
+  semantics with them, but they can be extremely useful for search
+  modules when creating plans.
+
 \end{enumerate}
 
-Having that specification for the Shopping Agent
-domain, here is how a particular problem instance could be stated:
+Having that specification for the Shopping Agent domain, here is how a
+particular problem instance could be stated:
 
@@
 
-In \eu,  all the Variables and Constraints from a problem domain domain description are kept inside the Constraint Reasoning Engine (CRE).
-The CRE provides a framework where any number of constraint types can collaborate to perform inference and prune variable domains using Bounds Propagation and Domain Reduction techniques.
-\eu uses a Propagator pattern, in which a Propagator object keeps track of a set of constraint instances of the same type and is responsible for ensuring that whenever any variable changes or constraints are introduced or dropped, all related variables are updated through inference.
+In \eu, all the Variables and Constraints from a problem domain,
+domain descriptions are kept inside the Constraint Reasoning Engine
+(CRE).  The CRE provides a framework where any number of constraint
+types can collaborate to perform inference and prune variable domains
+using Bounds Propagation and Domain Reduction techniques. \eu uses a
+Propagator pattern, in which a Propagator object keeps track of a set
+of constraint instances of the same type and is responsible for
+ensuring that whenever any variable changes or constraints are
+introduced or dropped, all related variables are updated through
+inference.
 
@@
 
-\comment {add europa's AC-3 algorithm here}
+\kcomment{add europa's AC-3 algorithm here}
 
@@
 
-\textit{Temporal Constraints:} \eu uses variables to explicitly represent timepoints for
-plan activities and states. Constraints among timepoints provide a
-natural way to express domain axioms. For example, in order to state
-that activity A must occur before activity B we can say that the end
-timepoint of A is $leq$ the start timepoint of B.
+\textit{Temporal Constraints:} \eu uses variables to explicitly
+represent timepoints for plan activities and states. Constraints among
+timepoints provide a natural way to express domain axioms. For
+example, in order to state that activity $A$ must occur before
+activity $B$ we can say that the end timepoint of $A$ is $leq$ the
+start timepoint of $B$.
 
@@
 outward arc from a node represents the maximum distance from the
-source node to the target node. 
-The diagram below illustrates a simple STN with just 2 variables and a single constraint. It also
+source node to the target node.  The diagram below illustrates a
+simple STN with just 2 variables and a single constraint. It also
 shows the resulting DG.
@@
 
-\cite{dechter91} also showed that shortest path
-algorithms could be used to propagate values in the network and
-discover a negative cycle. A negative cycle is a path from a node to
-itself that has a path length less than 0. If such a cycle exists, the
-network is inconsistent. It was further shown that a single-source
-shortest path algorithm was sufficient to detect a negative cycle and
-provide sufficient propagation to yield a backtrack-free search. Thus
-we have an efficient and complete algorithm for propagating an
-STN. \eu's Propagator for temporal constraints is a straight implementation of this algorithm.
+\cite{dechter91} also showed that shortest path algorithms could be
+used to propagate values in the network and discover a negative
+cycle. A negative cycle is a path from a node to itself that has a
+path length less than 0. If such a cycle exists, the network is
+inconsistent. It was further shown that a single-source shortest path
+algorithm was sufficient to detect a negative cycle and provide
+sufficient propagation to yield a backtrack-free search. Thus we have
+an efficient and complete algorithm for propagating an STN. \eu's
+Propagator for temporal constraints is a straight implementation of
+this algorithm.
 
-\textit{Resource Constraints:} Most real life planning problems have to deal with 
-the fact that plans made up of actions that require resources to be performed typically
-have only a limited amount of those resources available (think of fuel, storage capacity, etc).
-\eu provides built-in support for resources whose availability can be measured with real numbers, 
-in terms of \eu's  representation, resource object types and their related constraints are provided as follows:
+\textit{Resource Constraints:} Most real life planning problems have
+to deal with the fact that plans made up of actions that require
+resources to be performed typically have only a limited amount of
+those resources available (think of fuel, storage capacity, etc).  \eu
+provides built-in support for resources whose availability can be
+measured with real numbers, in terms of \eu's representation, resource
+object types and their related constraints are provided as follows:
 
@@
 
-\item \textbf{Reservoir} resources can have both \texttt{consume(resource,time,quantity)} and
-  \texttt{produce(resource,time,quantity)} constraints. Production and consumption is instantaneous (it happens at a single point in time)
+\item \textbf{Reservoir} resources can have both
+  \texttt{consume(resource,time,quantity)} and
+  \texttt{produce(resource,time,quantity)} constraints. Production and
+  consumption is instantaneous (it happens at a single point in time)
 
-\item \textbf{Reusable} resources have \texttt{uses(resource,start,duration,quantity)} constraints that use a
-  quantity of the resource for a specific duration, then release it.
+\item \textbf{Reusable} resources have
+  \texttt{uses(resource,start,duration,quantity)} constraints that use
+  a quantity of the resource for a specific duration, then release it.
 
-\item \textbf {Unary} resources are Reusable resources with capacity=1, which allows \eu to perform 
-some run time savings compared to the generic Reusable class.
+\item \textbf {Unary} resources are Reusable resources with
+  capacity=1, which allows \eu to perform some run time savings
+  compared to the generic Reusable class.
+
 \end{enumerate}
 
-\eu offers configuration options so that in addition to the usual constraint violations when a resource is completely depleted, plan flaws can be generated whenever a resource level 
-is outside a specific upper or lower limit, or when instantaneous production or consumption exceeds a specific threshold.
+\eu offers configuration options so that in addition to the usual
+constraint violations when a resource is completely depleted, plan
+flaws can be generated whenever a resource level is outside a specific
+upper or lower limit, or when instantaneous production or consumption
+exceeds a specific threshold.
 
@@
 
-Now we have all the elements in place so that an automated problem
-solver can be created. Let's recap what those elements are:
+Given the above elements, an automated problem solver can be
+created using:
 
@@
 \item A problem instance (also called initial state) that consists of:
+
     \begin{enumerate}
-        \item Variable and Object instances that exist for the entire planning
-  horizon.
+
+    \item Variable and Object instances that exist for the entire
+      planning horizon.
+
     \item Temporally scoped predicate and action instances.
+
     \item Temporally scoped goals.
+
 \end{enumerate}
@@
 \begin{enumerate}
+
 \item Find the flaws of PP, that is the conditions that prevent it
   from being a solution plan (flaw types are explained below).
+
 \item Select one such flaw
+
 \item Select a resolver for the flaw
+
 \item Refine PP by applying the resolver
+
 \item If an inconsistency is found, try another resolver.
+
 \item If all possible resolvers for a particular flaw fail, return
   failure, otherwise continue until resolving all the flaws.
+
 \end{enumerate}
 
-This Plan Space Planning algorithm is translated into \eu's
+This Plan Space Planning algorithm is translated into \eus
 representation as follows:
@@
   is activated, rules may fire that create inactive slave tokens
-  \comment{reference: point to more detailed decription above in Rules Engine
-    section}.  Open Conditions can be resolved in three ways:
+  \comment{reference: point to more detailed decription above in Rules
+    Engine section}.  Open Conditions can be resolved in three ways:
 
@@
     compatibilities are introduced as free activities. This results in
-    both an ordering flaw, corresponding to the just activated activity,
-    and a number of open condition flaws, corresponding to the new
-    subgoal activities.
+    both an ordering flaw, corresponding to the just activated
+    activity, and a number of open condition flaws, corresponding to
+    the new subgoal activities.
 
-  \item Rejection: Some goals may be optional, if that is the case for a
-    particular order condition, its corresponding flaw can be resolved
-    by discarding the inactive token.
+  \item Rejection: Some goals may be optional, if that is the case for
+    a particular order condition, its corresponding flaw can be
+    resolved by discarding the inactive token.
 
@@
   were available initially. Thus the problem cannot be solved.
+
 \end{enumerate}
@@
 
-Heuristics can be implemented by applying ordering/filtering policies for
-\texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance, by dealing with earliest flaws first, 
-or by making planning decisions (open condition flaws) before scheduling ones (threats), etc.
+Heuristics can be implemented by applying ordering/filtering policies
+for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance,
+by dealing with earliest flaws first, or by making planning decisions
+(open condition flaws) before scheduling ones (threats), etc.
 
-It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be
-implemented. As a simple example, consider the following solver for the N-Queens problem, where there is an integer variable 
-Vi to represent the queen in column i, so that for instance, V1=3 means that the variable on column 1 is placed on row 3. 
-non-attack constraints are posted as arithmetic constraints among the Vi variables to ensure that all the rows and diagonals contain only one queen.
+It should be emphasized that while this algorithm is commonly
+employed, it is only one of many that could be implemented. As a
+simple example, consider the following solver for the N-Queens
+problem, where there is an integer variable Vi to represent the queen
+in column i, so that for instance, V1=3 means that the variable on
+column 1 is placed on row 3.  non-attack constraints are posted as
+arithmetic constraints among the Vi variables to ensure that all the
+rows and diagonals contain only one queen.
 
@@
 
-\eu's Constraint Reasoning Engine can be used to compute the non-attack constraint violations (\texttt{ce.getViolations()} is a straight call on the \eu API), 
-to easily pick the queen that has he most violations by implementing \texttt{getQueenWithMaxViolation(ce)} and to look ahead and see what 
-moves are possible for a queen that would remove its violations by implementing \texttt{getMoves(ce,queenToMove,curPos)}.
+\eus Constraint Reasoning Engine can be used to compute the non-attack
+constraint violations (\texttt{ce.getViolations()} is a straight call
+on the \eu API), to easily pick the queen that has he most violations
+by implementing \texttt{getQueenWithMaxViolation(ce)} and to look
+ahead and see what moves are possible for a queen that would remove
+its violations by implementing
+\texttt{getMoves(ce,queenToMove,curPos)}.
 
-In similar fashion, the constraint violation and flaw reporting mechanisms that \eu provides can be used as building blocks that specialized or general purpose search algorithms can use to construct plans.
+In similar fashion, the constraint violation and flaw reporting
+mechanisms that \eu provides can be used as building blocks that
+specialized or general purpose search algorithms can use to construct
+plans.
 

### 2012-02-21 arch.tex
merged duplicate picture from kana with auv-agent

@@
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
-  \caption{\small An example of a \rx reactor hierarchy which is used
-    to determine a synchronization order for propagation of values in
-    \emph{external} state variables.}
-  \label{fig:reactor-DAG}
-\end{figure}
+% \begin{figure}[!htbp]
+%   \centering
+%   \includegraphics[scale=0.35]{figs/DAG-hierarchy}
+%   \caption{\small An example of a \rx reactor hierarchy which is used
+%     to determine a synchronization order for propagation of values in
+%     \emph{external} state variables.}
+%   \label{fig:reactor-DAG}
+% \end{figure}
+\fcomment{I removed this figure as the next one is mucjh more
+  detailed and vehiculate the same idea}
 
@@
 dependency graph from the least dependent -- ones with no {\em
-  external} timelines -- to the most, \ie which typically do not have
-{\em internal} timelines. Fig. \ref{fig:reactor-DAG} shows an example
+  external} state variables -- to the most, \ie which typically do not have
+{\em internal} state variables. Fig. \ref{fig:agent} shows an example
 of such a dependency graph implemented as a directed acyclic graph.

### 2012-02-21 concepts.tex
updates to intro etc

@@
-\subsubsection{\eu Plan Representation}
+\section{\eu Basics}
+\label{sec:basics}
+
+\paragraph {A brief note on \eu history and notation} The earliest
+origin of what we will call \eu in this chapter, is from the
+\texttt{HSTS} planner \cite{mus94} which was used for a number of
+demonstrations for space telescopes as noted in Section
+\ref{sec:rabeyond}. \texttt{HSTS} was also the
+\texttt{LISP}-based\footnote{This was the first (and only to our
+  knowledge) software ever to been written in \texttt{LISP} to be
+  flown in space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
+the planner was implemented at which went on to be used for the
+\texttt{MAPGEN} \cite{bresina05} system and continues to be used to
+date, in the mission-critical uplink process for the command/control
+of the Mars Exploration Rovers (\texttt{MER}) mission. A vastly
+re-factored, higher performance open source version \cite{europapso}
+was named \eut and is the subject of this chapter and used onboard
+MBARI's Dorado platform. Note however, in using the name \eu, we refer
+to the \emph{fundamental ideas as well as the planning paradigm} even
+if there exists a specific implementation.
+
+\subsection{Introduction}
+\label{sec:euintro}
+
+\eu uses a \emph{domain model} written in a declarative language,
+together with initial conditions and goals also in that specified
+language, to construct a set of temporal relations that must be true
+at the start time. These models include assertions about the physics
+of the vehicle, i.e how it responds to external stimulus and
+internally driven goals. By propagating these relations forward using
+Simple Temporal Networks \cite{dechter91} and applying goal
+constraints, \eu can select a set of conditions that should be true in
+the future, where some of these conditions will correspond to actions
+the agent must take. The planner can backtrack and try another path
+during search if a goal cannot be reached while being capable of
+discarding unachievable goals.
+
+Traditionally robot execution has relied on dispatching commands at
+precise times. Such linear sequences of precisely timed commands give
+no ability to adjust execution on the basis of sensory
+information. Although some commands can issue tests on sensor
+readings, these tests have the objective of verifying whether expected
+execution conditions are occuring. If not, the state of the system is
+declared off-nominal and execution of the sequence of commands is
+interrupted. More recently, executives have been proposed and
+implemented that significantly broaden the way robots can be commanded
+\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
+interpreted a \textit{temporally flexible plan} which represents each
+start time as a variable and contains an explicit network of bounded
+delay constraints between such variables.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\caption{\small Tokens with flexible temporal intervals and parametric
+  constraints between tokens. This example shows the triggering of a
+  water sampler based on a feature threshold while the vehicle
+  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
+  start \& end times.}
+\label{fig:flex-timelines}
+\vskip+0.1cm
+\end{figure}
+
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time the most appropriate one for the actual execution
+conditions. The fact that constraints are explicitly represented
+ensures that through constraint propagation the executive will respect
+global limits expressed in the plan (e.g., don't start a task until a
+certain condition has been satisfied but still satisfying some global
+deadline). Such flexibility is critical when dealing with dynamic
+ocean conditions where precise timing of a robotic action might be
+indeterminate. Further, the advantage of flexibility can be contrasted
+with the consequences of the intrinsic inflexibiliy of traditional
+command sequences. Because they are inflexible, sequences must
+necessarily be designed considering worst case scenarios.
+
+\eu uses a \emph{state variable} representation to describe the
+evolution of state over time. The instantiated history of such state
+variable evolution over a temporal horizon we call \emph{timelines}
+and which represent a single thread in the execution of a concurrent
+system. At any given time each thread can execute a single procedure.
+Thus each timeline consists of a sequence of procedures which
+encapsulate and describe state evolution; we call these instantiated
+atomic entities \emph{tokens}.  A token therefore describes a
+procedure invocation, the state variables on which it can occur, the
+parameter values of the procedure, and the time values defining the
+interval. We allow encapsulation of uncertainty within these tokens
+with a range of start and end times and parameters, all of which are
+encoded as variables. A constraint solver in turn manipulates these
+variables defined in a \eu domain model. For example, consider
+a scientific need to take a water samples 100 meters from a hotspot
+while an AUV is performing a Yo-Yo in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
+\subsection{\eu Plan Representation}
 \label{sec:europa:pr}
@@
 
-  \eu's ability to represent and reason over numeric intervals is a
+  \eus ability to represent and reason over numeric intervals is a
   very useful characteristic that allow users to create flexible
@@
   by exactly one token for every given timepoint in the plan. Any
-  instances of a class derived from \eu's built-in Timeline class will
+  instances of a class derived from \eus built-in Timeline class will
   induce ordering requirements among its tokens to ensure no temporal
@@
 
-\item[\textbf{Token State Model}] In \eu's representation, goals are
+\item[\textbf{Token State Model}] In \eus representation, goals are
   temporally scoped states that need to be achieved, or actions that
@@
 merging then it will cause the original activation decision that
-spawned it to be retracted (\eu's search mechanisms are explained in
+spawned it to be retracted (\eus search mechanisms are explained in
 detail in a separate section later on).

### 2012-02-21 intro.tex
updates to intro etc

@@
 years has produced a range of planners with demonstrated embedded
-capabilities \comment{cite}. Our work has been influenced by and in
-turn influenced a wide range of activities in the field of AI Planning
-and Plan Execution with demonstrated real-world and highly visible
-capabilities in the Space domain
-\cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons learned
-have been applied to the world of marine robotics which we bring
-together in this chapter. Important lessons that were derived from
-this rich legacy involved the following:
+capabilities
+\cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
+work has been influenced by and in turn influenced a wide range of
+activities in the field of AI Planning and Plan Execution with
+demonstrated real-world and highly visible capabilities in the Space
+domain \cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons
+learned have been applied to the world of marine robotics which we
+bring together in this chapter. Important lessons that were derived
+from this rich legacy involved the following:
 
@@
 \begin{figure}[t]
-  \centering
-  \vskip-5pt
+  \centering \vskip-5pt
   \includegraphics[scale=0.1]{figs/MBARI-AUV.jpg}
@@
 This chapter is organized as follows. We start with some foundational
-concepts in our Planning framework in Section
-\ref{sec:concepts}. \comment{add others} We conclude with Section
-\ref{sec:conclusion}.
+concepts in our Planning framework in Section~\ref{sec:concepts} which
+includes a brief overview of AI Planning in
+Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
+deeper into the \eu framework itself, the basis of all deliberation
+within \rx. A key portion is Section~\ref{sec:europa:pr} which details
+the Internal plan representation essential to planning within a
+situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
+Inference (Section~\ref{sec:europa:inference}) and Search
+(Section~\ref{sec:europa:search}) further detail the important details
+associated with \eu for the reader. With these foundational elements
+out of the way, we transition to \rx in Section~\ref{sec:arch}, bring
+out architectural details in Section~\ref{sec:arch:trex}, deal with
+the execution cycle in Section~\ref{sec:arch:exec}, highlight how \rx
+deliberates in Section~\ref{sec:arch:europa} finally delving deeper
+into the key concept of synchronization in
+Section~\ref{sec:arch:synch}. Further details on deliberation within
+\rx (Section~\ref{sec:arch:plan}) and how planning and execution are
+interleaved are shown in Section~\ref{sec:arch:intertwine}. We offer
+results from our experiments at sea in Section~\ref{sec:results},
+offer some insights into the near-term future goals of our research in
+Section~\ref{sec:future} and conclude with
+Section~\ref{sec:conclusion}.

### 2012-02-21 trunk/europa-basics.tex
[preamble] removed dimensions forcing : this reset paper layout to its default [europa-basics] chenged N-Queens equation to a eqnarray for better layout
updates to intro etc

--- 2012-02-16 trunk/europa-basics.tex
@@
 of scientific interest and to do so while being operationally aware of
-its own limitations in terms of resources (typically the battery state
-of charge) and overall proximity to other observational assets for
-obtaining scientific ground-truth. We are therefore interested in
+its own limitations in terms of resources (\eg the battery state of
+charge) and overall proximity to other observational assets for
+obtaining scientific ground-truth. We are interested in
 representational frameworks that allow robots to pursue long-term
@@
 To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use a
+to be able to synthesize plans insitu and to re-plan. We use the
 temporal constraint-based planner \eu with a demonstrated legacy of
-having flown on NASA space missions \cite{jonsson00,bresina05,
+having flown on NASA space missions \cite{mus98,jonsson00,bresina05,
   barreiro09}. Our autonomy architecture brings three key innovations
 for AUV adaptation: the use of flexible plan representations,
-compositional control with the use of partitioned networks and
-off-line learning to inform insitu environmental state estimation. In
-this section we detail some of the key concepts relevant to flexible
-plan representation. Section \ref{sec:arch} has details the
-\rx partitioned architecture and estimation driven vehicle
-adaptation.
+compositional control with the use of partitioned networks and a
+demonstrated capability at sea in novel ocean observation methods.
 
-\subsection{The basics}
-\label{sec:basics}
-
-\eu uses a \emph{domain model} written in a declarative language
-(NDDL: New Domain Description Language), together with initial
-conditions and goals also specified in NDDL, to construct a set of
-temporal relations that must be true at the start time. These models
-include assertions about the physics of the vehicle, i.e how it
-responds to external stimulus and internally driven goals. By
-propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, \eu can select a set
-of conditions that should be true in the future, where some of these
-conditions will correspond to actions the agent must take. The planner
-can backtrack and try another path during search if a goal cannot be
-reached while being capable of discarding unachievable goals.
-
-Traditionally robot execution has relied on dispatching commands at
-precise times. Such linear sequences of precisely timed commands give
-no ability to adjust execution on the basis of sensory
-information. Although some commands can issue tests on sensor
-readings, these tests have the objective of verifying whether expected
-execution conditions are occuring. If not, the state of the system is
-declared off-nominal and execution of the sequence of commands is
-interrupted. More recently, executives have been proposed and
-implemented that significantly broaden the way robots can be commanded
-\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
-interpreted a \textit{temporally flexible plan} which represents each
-start time as a variable and contains an explicit network of bounded
-delay constraints between such variables.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
-\caption{\small Tokens with flexible temporal intervals and parametric
-  constraints between tokens. This example shows the triggering of a
-  water sampler based on a feature threshold while the vehicle
-  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
-  start \& end times.}
-\label{fig:flex-timelines}
-\vskip+0.1cm
-\end{figure}
-
-Unlike a traditional fixed time-tagged command sequences, such
-flexible plans leave room for adaptation at execution time. When the
-executive considers when to start a task, it propagates information
-through the constraint network, computes a time bound for the
-variable, selects an actual execution time within the bound, and
-starts the task at that time. Temporally flexible plans therefore,
-express a \textit{range of possible outcomes} of the robots
-interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
-
-\eu uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a \eu domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
+% . In
+% this section we detail some of the key concepts relevant to flexible
+% plan representation. Section \ref{sec:arch} has details the \rx
+% partitioned architecture and estimation driven vehicle adaptation.
 
@@
 
-\section{Fundamentals of  Automated Planning}
-\label{sec:europa}
+\subsection{Fundamentals of  Automated Planning}
+\label{sec:planningfound}
 
@@
 
-\begin{equation}
- Sum(Q_{rc})= \left\{
+\begin{eqnarray*}
+ Sum(Q_{rc}) & = & \left\{
    \begin{array}{l l}
@@
      1 & \forall c \quad \mbox{(only one Queen per column)}\\ 
-\end{array} \right.
-Sum(Q_{r+i,c+i}) = 1\\
-Sum(Q_{r-1,c+i}) = 1 \quad \mbox{(only one Queen on each diagonal)}\\
-\end {equation}
+   \end{array} \right. \\
+Sum(Q_{r+i,c+i}) & = & 1\\
+Sum(Q_{r-i,c+i}) & = & 1 \qquad \mbox{(only one Queen on each diagonal)}\\
+\end{eqnarray*}
 
-\comment{fix newline and numbering problem above}
+%\comment{fix newline and numbering problem above}
 
@@
 
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/constraint-net.pdf}
+\caption{\small Illustration of a constraint network as represented as
+a graph.}
+\label{fig:constraintnet}
+\end{figure}
+
 A set of constraints and variables that describe a problem domain are
@@
 constraint propagation until there are no variable domain changes. The
-AC-3 algorithm \cite{mackworth77}, a popular implementation, improves
-over the naive approach by ignoring constraints whose variables were
-not affected since the last iteration. \comment{A figure of a
-  constraint network preferably of the shopping example would be
-  useful here. JB to generate a napkin fig}
+\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation,
+improves over the naive approach by ignoring constraints whose
+variables were not affected since the last
+iteration. Fig. \ref{fig:constraintnet} shows an example of a
+constraint network with three variables, their domains and the
+constraint relations between them.
 
@@
 specific search algorithm used by \eu will be covered in detail below.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-21 trunk/future.tex
updates to intro etc

--- 2012-01-04 trunk/future.tex
@@
 
-% Where do we see all this going esp. in the context of embedded AI Planning
-% for marine robots? This can be (and should be)
-% speculative. Shore-side/onboard autonomy coordination and control of
-% multiple vehicles is an important result of this activity.
+Our experience of using deliberative techniques in a highly
+inter-disciplinary environment for targeted marine science
+applications has shown the need and applicability of such novel
+algorithms and methods. In addition to augmenting traditional AUV
+surveys more advanced Lagrangian observations, mixed-initiative
+methods, our colleagues at MBARI have understood the longer term
+implications of our work. Our current and consequently future efforts
+are therefore grounded in important science problems in an operational
+oceanographic setting.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[scale=0.45]{figs/autonomy-topics.pdf}
+  \caption{\small }
+  \label{fig:topics}
+\end{figure}
+
+The core of our problem domain for \can has been towards Sampling. In
+large part given the spatial and temporal scales of the
+bio-geochemical processes we are interested in in this project, it is
+imperative that a key direction is in formulating a multi-robot
+adaptive sampling problem. In this context while superficially \rx can
+be considered as a multi-agent system and could therefore imply a
+direct extension, sporadic communication with high-latencies in
+surface communication, not to mention low throughput acoustic
+communications imply that the extension of \rx for distributed
+applications is non-trivial. Further, in order to efficiently execute
+coordinated observation of ocean processes with multiple autonomous
+sensing platforms, significant methodological as well as technical
+challenges must be addressed. We are working towards a systematic
+approach of using \rx as a backend planner for \can field
+experiments. The expectation is that \od users will plan initially
+individual vehicle surveys which are variants of standard
+straight-line transects. Lessons learned with human planners in the
+loop will be used towards planning larger surveys, also with simple
+fixed transects but using vehicle capabilities in ontologies similar
+to \cite{patron09}. Apportioning the planning task between shore-side
+components with substantial contextual information and compute power
+with robots with in-situ reasoning capabilities a la \rx is a
+challenge and an open research problem. Fig. \ref{fig:topics} shows
+what we believe is the scope and influence of the different topic
+areas we are focusing on.
+
+Partitioned inference and decision-making within \rx also lends itself
+to graceful system degradation, making the overall approach effective
+towards increased robustness. There are two outstanding diagnosis and
+failure recovery challenges that we need to address; the first has to
+do with methodology related to model design of the instrument payload
+for a highly modular and configurable vehicle; this is local to a
+reactor. Doing so would enable (as a first step) using systematic
+Fault Diagnosis, Isolation and Recovery (FDIR) algorithms as
+demonstrated in the \texttt{RAX} experiment
+\cite{williams96,mus98,williams97} and more recently on an AUV by
+Dearden and others \cite{wang09,ernits10,dearden11}. The second
+challenge has to implementing a policy within the \rx framework that
+explicitly deals with recovery since controller failure can be
+distributed across reactors. Since reactors are hierarchical in their
+relationship to one another, their well defined dependencies allow
+them to be removed from an agent systematically. Reactors more
+abstract in scope can be removed before those which are less abstract
+(and closer to the hardware). In the event a Navigation reactor, for
+example, has a fault, it can be removed safely with lower level
+functionalities (such as those in the executive). When the Navigator
+is removed, it will recursively remove more abstract reactors (such as
+a Mission Manager) that depend on it. Such a safety feature then
+allows the system designers to ensure that early on they can invest
+more in ensuring that reactors lower in the hierarchy are robust and
+potentially formally verified. It also means that there is a safe and
+sure way to take control of the vehicle at any level within the
+reactors hierarchy; as long as low level reactors are still active one
+will always been able to send goals/commands to them and execute what
+are traditionally known as runout sequences (or clean up scripts).
+
+Feature recognition for Sampling is a general research thrust; however
+it has specific relevance in the context of \rx and AUV control. As
+noted in Section \ref{sec:results} we have already demonstrated
+Machine Learning techniques for recognition of select features in the
+coastal ocean. A principal challenge that remains is that of obtaining
+expert labeled data for any supervised or semi-supervised learning
+methods.  While supervised techniques for learning classifiers such as
+decision trees (\cite{Quinlan93-dtrees}), instance-based
+classification (\cite{Aha-ibl-ml91}), Bayesian classifiers
+(\cite{Jensen2001-BNetworks}) and artificial neural networks
+(\cite{ANNsurvey-2000}) are popular, given the vast amount of data
+available and poor understanding of correlation between physical and
+biogeochemical variables in the coastal ocean, our work on INLs
+\cite{mcgann08d,ryan10} and reliance on a oceanographic expert has
+shown that scaling to different problems remains problematic. This is
+a prime motivation to move towards more semi-supervised methods such
+as \cite{kumar11,sergio12}. With the advent of the \od however,
+another direction our research is taking us is in Recommender systems
+\cite{Adomavicius05} as noted in Section \ref{sec:results}. We are
+working to design and deploy software infrastructure that will learn
+from user tagged data. To do so, we will use tagging techniques
+similar to what users of commercial image sharing sites like Flickr
+and Picassa do, to learn the context of the data (including remote
+sensing). In addition, the \od will integrate statistical Machine
+Learning techniques with incoming asynchronous data stream obtained by
+sensors and platforms. Identification models will be based on existing
+training sets for INLs and plumes; models for other features of
+scientific interests driven by \can requirements will be built in
+addition. Existing AUV archives for instance provide sufficient data
+for a learning base for a range of features of interest to \can; in
+addition we will build software that will learn to incorporate
+incoming data streams to provide an existence proof of a feature of
+interest.
+
+% multi-vehicle control
+% sampling
+% diagnosis and recovery
+% feature recognition
+% scalable architecture for missions

### 2012-02-21 trunk/intro.tex
Activated section numbering

@@
 This chapter is organized as follows. We start with some foundational
-concepts in our Planning framework in Section
-\ref{sec:concepts}. \comment{add others} We conclude with Section
-\ref{sec:conclusion}.
+concepts in our Planning framework in Section~\ref{sec:concepts} which
+includes a brief overview of AI Planning in
+Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
+deeper into the \eu framework itself, the basis of all deliberation
+within \rx. A key portion is Section~\ref{sec:europa:pr} which details
+the Internal plan representation essential to planning within a
+situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
+Inference (Section~\ref{sec:europa:inference}) and Search
+(Section~\ref{sec:europa:search}) further detail the important details
+associated with \eu for the reader. With these foundational elements
+out of the way, we transition to \rx in Section~\ref{sec:arch}, bring
+out architectural details in Section~\ref{sec:arch:trex}, deal with
+the execution cycle in Section~\ref{sec:arch:exec}, highlight how \rx
+deliberates in Section~\ref{sec:arch:europa} finally delving deeper
+into the key concept of synchronization in
+Section~\ref{sec:arch:synch}. Further details on deliberation within
+\rx (Section~\ref{sec:arch:plan}) and how planning and execution are
+interleaved are shown in Section~\ref{sec:arch:intertwine}. We offer
+results from our experiments at sea in Section~\ref{sec:results},
+offer some insights into the near-term future goals of our research in
+Section~\ref{sec:future} and conclude with
+Section~\ref{sec:conclusion}.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-21 conclude.tex
Activated section numbering

@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2012-02-21 trunk/europa-basics.tex
europa-basics cleanup.

@@
 
-% . In
-% this section we detail some of the key concepts relevant to flexible
-% plan representation. Section \ref{sec:arch} has details the \rx
-% partitioned architecture and estimation driven vehicle adaptation.
+The next few sections will cover \eu in a bottom-up fashion, going from the 
+generic theoretical frameworks that underlie \eu, to the specific representation, 
+modeling and algorithms implemented by it. First, a quick introduction to 
+Constraint Programming and Constraint-Based Planning is provided. 
+Next, the specific representation that \eu uses to perform Constraint-Based Planning is presented.
+After that comes an overview of \eu's architecture and how it can be embedded in a planning application
+such as \rx. Finally,  \eu's modeling, inference and search mechanisms are covered in 
+detail so that the reader can understand how \rx can perform deliberative planning and react to
+execution events by embedding and customizing \eu.
 
-% \begin{figure}
-% \centering
-% \includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-% \caption{\small Temporal relations defined within the planner are
-%   based on \texttt{Allen Algebra} relations shown above.}
-% \label{fig:allen-algebra}
-% \vskip-0.3cm
-% \end{figure}
-
-\subsection{Fundamentals of  Automated Planning}
+\subsection{Automated Planning and \eu}
 \label{sec:planningfound}
@@
 
-% Planning, for our purposes, can be thought of as determining all the
-% small tasks that must be carried out in order to accomplish a goal.
 To articulate the fundamentals of automated planning briefly and use
@@
 Sum(Q_{r+i,c+i}) & = & 1\\
-Sum(Q_{r-i,c+i}) & = & 1 \qquad \mbox{(only one Queen on each diagonal)}\\
+Sum(Q_{r-i,c-i}) & = & 1 \qquad \mbox{(only one Queen on each diagonal)}\\
 \end{eqnarray*}
-
-%\comment{fix newline and numbering problem above}
 
@@
 Q$_{rc}$ that satisfy the above constraints. Fig. \ref{fig:nqueens-2} is
-a solution found by \eu using that formulation and a specialized search
+a solution found by \eu using a compact version of that formulation and a specialized search
 procedure.
@@
 Constraint Satisfaction Programming (also known as Constraint
-Programming (\textsf{CP})) is a discipline that provides a generic framework
+Programming or \textsf{CP}) is a discipline that provides a generic framework
 for representing, solving and making logical inference on
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.35]{figs/constraint-net.pdf}
-\caption{\small Illustration of a constraint network as represented as
-a graph.}
-\label{fig:constraintnet}
-\end{figure}
-
 A set of constraints and variables that describe a problem domain are
@@
 constraint propagation until there are no variable domain changes. The
-\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation,
-improves over the naive approach by ignoring constraints whose
-variables were not affected since the last
-iteration. Fig. \ref{fig:constraintnet} shows an example of a
-constraint network with three variables, their domains and the
-constraint relations between them.
+\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation, improves
+over the naive approach by ignoring constraints whose variables were
+not affected since the last iteration. Fig.  \ref{fig:constraintnet} shows a set of constraints on 3 variables
+($x$, $y$ and $z$) represented as a constraint network, and their domains before arc consistency is enforced.
+After the AC-3 algorithm is executed, the domains for the 3 variables
+would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/constraint-net.pdf}
+\caption{\small Variables and Constraints represented as a Constraint Network}
+\label{fig:constraintnet}
+\end{figure}
 
@@
 formalism. In particular, it is hard to represent time, resources,
-% mutual exclusion and concurrency using propositions
-% \cite{ghallab04}. While CSP representations have traditionally had an
-% edge in formulating and representing planning problems,
-% % It is straightforward to represent and reason about all of
-% % those elements using a CSP representation, as a result there has been
-% % some work on doing automated planning while taking advantage of CSP
-% % (TODO: ref). Traditionally, the entire planning problem is translated
-% % into a CSP and then solved using traditional CSP methods, this
-% leading to a formulation where action choices and relationships are
-% represented as variables and/or constraints, there are at least two
-% major drawbacks:
 mutual exclusion and concurrency using propositions \cite{ghallab04}.
@@
   insight has to be translated into the CSP representation of
-  variables and constraints \comment{For example?}
+  variables and constraints.
 
@@
 frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
-\texttt{IxTet} \cite{ghallab87,lemai04}, and CAIP (Constraint-Based
+\texttt{IxTet} \cite{ghallab87,lemai04}, and \texttt{CAIP} (Constraint-Based
 Interval and Attribute Planning)\cite{mus94,frank2003}.  These
@@
 natural way to express search algorithms and heuristics. \eu is an
-implementation of the CAIP framework. In CAIP, planning problems are
+implementation of the \texttt{CAIP} framework. In \texttt{CAIP}, planning problems are
 specified in terms of:
@@
 
-To generate plans, the CAIP framework uses a plan-space search
+To generate plans, the \texttt{CAIP} framework uses a plan-space search
 approach \cite{ghallab04}, where a partial plan is evolved until all
-open goals are achieved, and all inconsistencies are removed, the
+open goals are achieved, and all inconsistencies are removed. The
 specific search algorithm used by \eu will be covered in detail below.

### 2012-02-21 europa.tex
converted queen algorithm into an algorithm enviroment

@@
 can be modeled without having to introduce new constraint
-types. Tables \ref{{tab:calconst, tab:setconst,
-    tab:objhierarchy,tab:misc} show the constraint types.
+types. Tables \ref{tab:calconst}, \ref{tab:setconst},
+\ref{tab:objhierarchy} and \ref{tab:misc} show various constraint
+types.
 
@@
 
-\item \textbf{Reservoir} resources can have both \texttt{consume(resource,time,quantity)} and
-  \texttt{produce(resource,time,quantity)} constraints. Production and consumption is instantaneous (it happens at a single point in time)
+\item \textbf{Reservoir} resources can have both
+  \texttt{consume(resource,time,quantity)} and
+  \texttt{produce(resource,time,quantity)} constraints. Production and
+  consumption is instantaneous (it happens at a single point in time)
 
-\item \textbf{Reusable} resources have \texttt{uses(resource,start,duration,quantity)} constraints that use a
-  quantity of the resource for a specific duration, then release it.
+\item \textbf{Reusable} resources have
+  \texttt{uses(resource,start,duration,quantity)} constraints that use
+  a quantity of the resource for a specific duration, then release it.
 
-\item \textbf {Unary} resources are Reusable resources with capacity=1, which allows \eu to perform 
-some run time savings compared to the generic Reusable class.
+\item \textbf {Unary} resources are Reusable resources with
+  capacity=1, which allows \eu to perform some run time savings
+  compared to the generic Reusable class.
+
 \end{enumerate}
 
-\eu offers configuration options so that in addition to the usual constraint violations when a resource is completely depleted, plan flaws can be generated whenever a resource level 
-is outside a specific upper or lower limit, or when instantaneous production or consumption exceeds a specific threshold.
+\eu offers configuration options so that in addition to the usual
+constraint violations when a resource is completely depleted, plan
+flaws can be generated whenever a resource level is outside a specific
+upper or lower limit, or when instantaneous production or consumption
+exceeds a specific threshold.
 
-Recall that \eu supports a flexible temporal representation, where time is represented by intervals, instead of single values.
-In this context, determining whether a resource is the source of a plan flaw or a constraint violation in general is more difficult than just computing a profile
-of the resource and checking against some limits. 
-In the context of time flexibility, there are many resource profiles that are possible, depending on when exactly each production/consumption event happens out of all the possible values that are in the domain of the variables that represent the time for the event.
-\cite{mus02} proposed an algorithm where the variables that represent the time for production/consumption events are placed in a temporal network, similar to the network used by the Temporal Propagator above, then the constraints on the resource can be seen as quantities flowing through that network.
-Computing the maximum flow will yield the maximum resource usage, and thus a lower bound on the resource level, and computing a minimum flow will similarly yield an upper bound.
-Since maximum flow on a temporal network can be computed in polynomial time, \eu takes advantage of this algorithm to efficiently compute upper and lower bound envelopes on the resource level.
+Recall that \eu supports a flexible temporal representation, where
+time is represented by intervals, instead of single values. In this
+context, determining whether a resource is the source of a plan flaw
+or a constraint violation in general is more difficult than just
+computing a profile of the resource and checking against some limits.
+In the context of time flexibility, there are many resource profiles
+that are possible, depending on when exactly each
+production/consumption event happens out of all the possible values
+that are in the domain of the variables that represent the time for
+the event.  \cite{mus02} proposed an algorithm where the variables
+that represent the time for production/consumption events are placed
+in a temporal network, similar to the network used by the Temporal
+Propagator above, then the constraints on the resource can be seen as
+quantities flowing through that network. Computing the maximum flow
+will yield the maximum resource usage, and thus a lower bound on the
+resource level, and computing a minimum flow will similarly yield an
+upper bound. Since maximum flow on a temporal network can be computed
+in polynomial time, \eu takes advantage of this algorithm to
+efficiently compute upper and lower bound envelopes on the resource
+level.
 
@@
 
-Heuristics can be implemented by applying ordering/filtering policies for
-\texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance, by dealing with earliest flaws first, 
-or by making planning decisions (open condition flaws) before scheduling ones (threats), etc.
+Heuristics can be implemented by applying ordering/filtering policies
+for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance,
+by dealing with earliest flaws first, or by making planning decisions
+(open condition flaws) before scheduling ones (threats), etc.
 
-It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be
-implemented. As a simple example, consider the following solver for the N-Queens problem, where there is an integer variable 
-Vi to represent the queen in column i, so that for instance, V1=3 means that the variable on column 1 is placed on row 3. 
-non-attack constraints are posted as arithmetic constraints among the Vi variables to ensure that all the rows and diagonals contain only one queen.
+It should be emphasized that while this algorithm is commonly
+employed, it is only one of many that could be implemented. As a
+simple example, consider the following solver for the N-Queens
+problem, where there is an integer variable Vi to represent the queen
+in column i, so that for instance, V1=3 means that the variable on
+column 1 is placed on row 3.  non-attack constraints are posted as
+arithmetic constraints among the Vi variables to ensure that all the
+rows and diagonals contain only one queen.
 
-\begin{verbatim}
-    public void solve(ConstraintEngine ce, int maxIter)
-    {
-    	curIter=0;
-
-    	while ((ce.getViolations() > 0) && (curIter < maxIter)) {
-	    Variable queenToMove = getQueenWithMaxViolation(ce);
-    	     int curPos = queenToMove.getCurrentValue();
-
-            SortedSet<int> moves = getMoves(ce,queenToMove,curPos);             
-  	   boolean moved = false;
-            for (int newPos : moves) {
-            	moved = tryMove(queenToMove,curPos,newPos);
-            	if (moved)
-            	    break;
+\begin{algorithm}[H]
+  \KwIn{ConstraintEngine $ce$}
+  \KwIn{int $maxIter$}
+  \Begin{
+    $curIter \leftarrow 0$ \;
+    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
+        maxIter \right)$ }{
+      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
+      \;
+      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
+      curPos)$ \;
+      bool $moved \leftarrow $ \texttt{false} \;
+      \ForEach{int $newPos$ of $moves$}{
+        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
+        \lIf{$moved$}{ \textbf{break} \; }
       }
-
-            if (!moved) 
-            	forceMove(queenToMove,curPos,moves.first());
-
-            checkSolution(); // See if we have a new best solution            
-            curIteration_++;
+      \lIf{$! moved$}{
+        $forceMove(queenToMove, curPos, moves.first())$ \;
+      }
+      $checkSolution()$ \tcp*{See if we have a new best solution}
+      $curIter++$ \;
     }
   }
-\end{verbatim}
+  \caption{$solveNQueens(ce, maxIter)$}
+\end{algorithm}
 
-\eu's Constraint Reasoning Engine can be used to compute the non-attack constraint violations (\texttt{ce.getViolations()} is a straight call on the \eu API), 
-to easily pick the queen that has he most violations by implementing \texttt{getQueenWithMaxViolation(ce)} and to look ahead and see what 
-moves are possible for a queen that would remove its violations by implementing \texttt{getMoves(ce,queenToMove,curPos)}.
+% \begin{verbatim}
+%     public void solve(ConstraintEngine ce, int maxIter)
+%     {
+%     	curIter=0;
 
-In similar fashion, the constraint violation and flaw reporting mechanisms that \eu provides can be used as building blocks that specialized or general purpose search algorithms can use to construct plans.
+%     	while ((ce.getViolations() > 0) && (curIter < maxIter)) {
+% 	    Variable queenToMove = getQueenWithMaxViolation(ce);
+%     	     int curPos = queenToMove.getCurrentValue();
+
+%             SortedSet<int> moves = getMoves(ce,queenToMove,curPos);             
+%   	   boolean moved = false;
+%             for (int newPos : moves) {
+%             	moved = tryMove(queenToMove,curPos,newPos);
+%             	if (moved)
+%             	    break;
+%             }
+
+%             if (!moved) 
+%             	forceMove(queenToMove,curPos,moves.first());
+
+%             checkSolution(); // See if we have a new best solution            
+%             curIteration_++;
+%     	}
+%     }
+% \end{verbatim}
+
+\eus Constraint Reasoning Engine can be used to compute the non-attack
+constraint violations (\texttt{ce.getViolations()} is a straight call
+on the \eu API), to easily pick the queen that has he most violations
+by implementing \texttt{getQueenWithMaxViolation(ce)} and to look
+ahead and see what moves are possible for a queen that would remove
+its violations by implementing
+\texttt{getMoves(ce,queenToMove,curPos)}.
+
+In similar fashion, the constraint violation and flaw reporting
+mechanisms that \eu provides can be used as building blocks that
+specialized or general purpose search algorithms can use to construct
+plans.
 

### 2012-02-21 europa-basics.tex
updated constraint table in correct location within europa.tex and corrected arch.tex typos
updates to intro etc

@@
 To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use a
+to be able to synthesize plans insitu and to re-plan. We use the
 temporal constraint-based planner \eu with a demonstrated legacy of
-having flown on NASA space missions \cite{jonsson00,bresina05,
+having flown on NASA space missions \cite{mus98,jonsson00,bresina05,
   barreiro09}. Our autonomy architecture brings three key innovations
 for AUV adaptation: the use of flexible plan representations,
-compositional control with the use of partitioned networks and
-off-line learning to inform insitu environmental state estimation. In
-this section we detail some of the key concepts relevant to flexible
-plan representation. Section \ref{sec:arch} has details the
-\rx partitioned architecture and estimation driven vehicle
-adaptation.
+compositional control with the use of partitioned networks and a
+demonstrated capability at sea in novel ocean observation methods.
 
-\paragraph {A brief note on \texttt{EUROPA} history and notation} The
-earliest origin of what we will call \eu in this chapter, is from the
-\texttt{HSTS} planner \cite{mus94} which was used for a number of
-demonstrations for space telescopes as noted in Section
-\ref{sec:rabeyond}. \texttt{HSTS} was the
-\texttt{LISP}-based\footnote{This was the first (and only to our
-  knowledge) software ever to been written in \texttt{LISP} to be
-  flown in space.} planner that was flown as part of the Remote Agent
-\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
-the planner was implemented at NASA Ames which went on to be used for
-the \texttt{MAPGEN} \cite{bresina05} and continues to be used to date,
-in the mission-critical uplink process for the command/control of the
-Mars Exploration Rovers (MER) mission. A vastly re-factored, higher
-performance open source version \cite{europapso} was named \eu and is
-the subject of this chapter. Note however, in using the name \eu, we
-refer to the \emph{fundamental ideas as well as the planning paradigm}
-even if there exists a specific implementation.
+The next few sections will cover \eu in a bottom-up fashion, going
+from the generic theoretical frameworks that underlie \eu, to the
+specific representation, modeling and algorithms implemented by
+it. First, a quick introduction to Constraint Programming and
+Constraint-Based Planning is provided.  Next, the specific
+representation that \eu uses to perform Constraint-Based Planning is
+presented.  After that comes an overview of \eu's architecture and how
+it can be embedded in a planning application such as \rx. Finally,
+\eu's modeling, inference and search mechanisms are covered in detail
+so that the reader can understand how \rx can perform deliberative
+planning and react to execution events by embedding and customizing
+\eu.
 
-\subsection{The Basics}
-\label{sec:basics}
-
-\eu uses a \emph{domain model} written in a declarative language
-(NDDL: New Domain Description Language), together with initial
-conditions and goals also specified in NDDL, to construct a set of
-temporal relations that must be true at the start time. These models
-include assertions about the physics of the vehicle, i.e how it
-responds to external stimulus and internally driven goals. By
-propagating these relations forward using Simple Temporal Networks
-\cite{dechter91} and applying goal constraints, \eu can select a set
-of conditions that should be true in the future, where some of these
-conditions will correspond to actions the agent must take. The planner
-can backtrack and try another path during search if a goal cannot be
-reached while being capable of discarding unachievable goals.
-
-Traditionally robot execution has relied on dispatching commands at
-precise times. Such linear sequences of precisely timed commands give
-no ability to adjust execution on the basis of sensory
-information. Although some commands can issue tests on sensor
-readings, these tests have the objective of verifying whether expected
-execution conditions are occuring. If not, the state of the system is
-declared off-nominal and execution of the sequence of commands is
-interrupted. More recently, executives have been proposed and
-implemented that significantly broaden the way robots can be commanded
-\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
-interpreted a \textit{temporally flexible plan} which represents each
-start time as a variable and contains an explicit network of bounded
-delay constraints between such variables.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
-\caption{\small Tokens with flexible temporal intervals and parametric
-  constraints between tokens. This example shows the triggering of a
-  water sampler based on a feature threshold while the vehicle
-  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
-  start \& end times.}
-\label{fig:flex-timelines}
-\vskip+0.1cm
-\end{figure}
-
-Unlike a traditional fixed time-tagged command sequences, such
-flexible plans leave room for adaptation at execution time. When the
-executive considers when to start a task, it propagates information
-through the constraint network, computes a time bound for the
-variable, selects an actual execution time within the bound, and
-starts the task at that time. Temporally flexible plans therefore,
-express a \textit{range of possible outcomes} of the robots
-interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
-
-\eu uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a \eu domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
-
-% \begin{figure}
-% \centering
-% \includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-% \caption{\small Temporal relations defined within the planner are
-%   based on \texttt{Allen Algebra} relations shown above.}
-% \label{fig:allen-algebra}
-% \vskip-0.3cm
-% \end{figure}
-
-\section{Fundamentals of  Automated Planning}
-\label{sec:europa}
+\subsection{Automated Planning and \eu}
+\label{sec:planningfound}
 
@@
 
-\begin{quotation}
-
-  In the near future, a personal robot sets out to buy a gallon of milk
-  This involves a number of tasks: obtain keys, obtain wallet,
-  start car, drive to store, find and obtain milk, purchase milk, etc.
-  The embedded planner has to have a ``model'' of the world in which it
-  lives and has to use the task primitives in this model to structure
-  the actions so it achieves its goal. Constraints control when certain
-  tasks can or cannot occur. For example the robot must obtain the keys
-  and wallet \emph{before} driving to the store and pick up the milk
-  \emph{before} purchasing it.
-
-\end{quotation}
+{\small
+  \begin{quote}
+    In the near future, a personal robot sets out to buy a gallon of
+    milk This involves a number of tasks: obtain keys, obtain wallet,
+    start car, drive to store, find and obtain milk, purchase milk,
+    etc.  The embedded planner has to have a ``model'' of the world in
+    which it lives and has to use the task primitives in this model to
+    structure the actions so it achieves its goal. Constraints control
+    when certain tasks can or cannot occur. For example the robot must
+    obtain the keys and wallet \emph{before} driving to the store and
+    pick up the milk \emph{before} purchasing it.
+\end{quote}
 
@@
      1 & \forall c \quad \mbox{(only one Queen per column)}\\ 
-\end{array} \right.
-Sum(Q_{r+i,c+i}) = 1\\
-Sum(Q_{r-1,c+i}) = 1 \quad \mbox{(only one Queen on each diagonal)}\\
-\end {equation}
-
-\comment{fix newline and numbering problem above}
+   \end{array} \right. \\
+Sum(Q_{r+i,c+i}) & = & 1\\
+Sum(Q_{r-i,c-i}) & = & 1 \qquad \mbox{(only one Queen on each diagonal)}\\
+\end{eqnarray*}
 
@@
 constraint propagation until there are no variable domain changes. The
-AC-3 algorithm \cite{mackworth77}, a popular implementation, improves
+\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation, improves
 over the naive approach by ignoring constraints whose variables were
-not affected since the last iteration. \comment{A figure of a
-  constraint network preferably of the shopping example would be
-  useful here. JB to generate a napkin fig}
+not affected since the last iteration. Fig.  \ref{fig:constraintnet} shows a set of constraints on 3 variables
+($x$, $y$ and $z$) represented as a constraint network, and their domains before arc consistency is enforced.
+After the AC-3 algorithm is executed, the domains for the 3 variables
+would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.35]{figs/constraint-net.pdf}
+\caption{\small Variables and Constraints represented as a Constraint Network}
+\label{fig:constraintnet}
+\end{figure}
 
@@
 
-To generate plans, the CAIP framework uses a plan-space search
+To generate plans, the \texttt{CAIP} framework uses a plan-space search
 approach \cite{ghallab04}, where a partial plan is evolved until all
-open goals are achieved, and all inconsistencies are removed, the
+open goals are achieved, and all inconsistencies are removed. The
 specific search algorithm used by \eu will be covered in detail below.
 
-\input{constraints-table}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-22 trunk/concepts.tex
updates to europa-basics and concepts

@@
-\subsubsection{\eu Plan Representation}
+\section{Fundamentals of \eu}
+\label{sec:basics}
+
+\paragraph {A brief note on \eu history and notation} The origin of
+what we will call \eu in this chapter, is derived from the
+\texttt{HSTS} planner \cite{mus94} used on a number of ground-based
+demonstrations for space telescope scheduling as noted in Section
+\ref{sec:rabeyond}. \texttt{HSTS} was also the
+\texttt{LISP}-based\footnote{This was the first (and to our knowledge
+  only) software ever to been written in \texttt{LISP} to be flown in
+  space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX} deployment, a C++
+version of the planner was implemented which went on to be used for
+the \texttt{MAPGEN} \cite{bresina05} system and continues to be used
+to date, in the mission-critical uplink process for the
+command/control of the Mars Exploration Rovers (\texttt{MER})
+mission. A vastly re-factored, higher performance open source version
+\cite{europapso} was named \eut and is the subject of this chapter and
+used by \rx.  Note however, in using the name \eu, we refer to the
+\emph{fundamental ideas as well as the planning paradigm} even if
+there exists a specific implementation.
+
+\subsection{Introduction}
+\label{sec:euintro}
+
+\eu uses a \emph{domain model} written in a declarative language,
+together with initial conditions and goals (also in that specified
+language), to construct a set of temporal relations that must be true
+at start time. These models include assertions about the physics of
+the vehicle, i.e how it responds to external stimulus and internally
+driven goals. By propagating these relations forward using Simple
+Temporal Networks \cite{dechter91} and applying goal constraints, \eu
+can select a set of conditions that should be true in the future,
+where some of these conditions will correspond to actions the agent
+must take. The planner can backtrack and try another path during the
+search process if a goal cannot be reached while being capable of
+discarding unachievable goals.
+
+Traditionally robot execution has relied on dispatching commands at
+precise times. Such linear sequences of precisely timed commands give
+no ability to adjust execution on the basis of sensory
+information. Although some commands can issue tests on sensor
+readings, these tests have the objective of verifying whether expected
+execution conditions are occuring. If not, the state of the system is
+declared off-nominal and execution of the sequence of commands is
+interrupted. More recently, executives have been proposed and
+implemented that significantly broaden the way robots can be commanded
+\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
+interpreted a \textit{temporally flexible plan} which represents each
+start time as a variable and contains an explicit network of bounded
+delay constraints between such
+variables. Fig. \ref{fig:flex-timelines} shows an example of a
+flexible partial plan.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\caption{\small Tokens with flexible temporal intervals and parametric
+  constraints between tokens. This example shows the triggering of a
+  water sampler based on a feature threshold while the vehicle
+  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
+  start \& end times.}
+\label{fig:flex-timelines}
+\end{figure}
+
+Unlike a traditional fixed time-tagged command sequences, such
+flexible plans leave room for adaptation at execution time. When the
+executive considers when to start a task, it propagates information
+through the constraint network, computes a time bound for the
+variable, selects an actual execution time within the bound, and
+starts the task at that time. Temporally flexible plans therefore,
+express a \textit{range of possible outcomes} of the robots
+interaction with the environment within which the executive can elect
+at run time, the most appropriate for actual execution. The fact that
+constraints are explicitly represented ensures that through constraint
+propagation the executive will respect global limits expressed in the
+plan (e.g., don't start a task until a certain condition has been
+satisfied but still satisfying some overall deadline). Such
+flexibility is critical when dealing with dynamic and uncertain
+conditions such as in the ocean, where precise timing of a robotic
+action might be indeterminate. Further, the advantage of flexibility
+can be contrasted with the consequences of the intrinsic inflexibility
+of traditional command sequences. Because they are inflexible,
+sequences must necessarily be designed considering worst case
+scenarios.
+
+\eu uses a \emph{state variable} representation
+\cite{mus94,jonsson00,mcgann09} to describe the evolution of state
+over time. The instantiated history of such state variable evolution
+over a temporal horizon we call \emph{timelines} and which represent a
+single thread in the execution of a concurrent system. At any given
+time each thread can execute a single procedure.  Thus each timeline
+consists of a sequence of procedures which encapsulate and describe
+state evolution; we call these instantiated atomic entities
+\emph{tokens}.  A token therefore describes a procedure invocation,
+the state variables on which it can occur, the parameter values of the
+procedure, and the time values defining the interval. We allow
+encapsulation of uncertainty within these tokens with a range of start
+and end times and parameters, all of which are encoded as variables. A
+constraint solver in turn manipulates these variables defined in a \eu
+domain model. For example, consider a scientific need to take water
+samples 100 meters from an environmental hotspot while an AUV is
+performing a Yo-Yo sawtooth pattern in the water-column. Two samples
+are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
+
+\subsection{\eu Plan Representation}
 \label{sec:europa:pr}
 
-\eu follows the representation outlined by the CAIP framework, it also adds a few elements to it that are necessary for an efficient implementation of constraint-based planning.
-Below is a brief summary of the main elements, followed by a detailed description for each one:
+\eu follows the representation outlined by the \texttt{CAIP} framework
+augmenting with elements that are necessary for an efficient
+implementation of constraint-based planning. Below we describe briefly
+the main elements, followed by a detailed description for each of
+these elements and how together they constitute a rich
+representational paradigm for inference in Sections
+\ref{sec:europa:modeling}, \ref{sec:europa:inference} and
+\ref{sec:europa:search}.
 
 \begin{enumerate}
-\item \textbf{Tokens} Temporally scoped entities that correspond to Intervals in the CAIP framework, are used to represent  state and actions in time.
-A\textbf{Token State model} is defined to support an efficient implementation of plan-space search algorithms.
 
-\item \textbf{Domains, Variables and Constraint} are used to represent restrictions in the problem domain in terms of a CP formulation.
+\item \textbf{Tokens} Temporally scoped entities that correspond to
+  Intervals in the \texttt{CAIP} framework, are used to represent
+  state and actions in time. A\textbf{Token State model} is defined to
+  support an efficient implementation of plan-space search algorithms.
 
-\item \textbf{Domain Rules} are used to describe internal and external relationships on and between tokens, as presecribed by the CAIP framework.
+\item \textbf{Domains, Variables and Constraint} are used to represent
+  restrictions in the problem domain in terms of a \texttt{CP}
+  formulation.
 
-\item \textbf{Objects} are used to represent problem domains in a more modular and scaleable way.
-\textbf{Timelines and Resources} are built in implementations of Object types that appear frequently in real life domains.
+\item \textbf{Domain Rules} are used to describe internal and external
+  relationships on and between tokens, as presecribed by
+  \texttt{CAIP}.
+
+\item \textbf{Objects} are used to represent problem domains in a more
+  modular and scaleable way. \textbf{Timelines and Resources} are
+  built in implementations of Object types that appear frequently in
+  real-world domains.
 
 \end{enumerate}
+
+At the atomic level, the entities in the \eu plan representation are:
 
@@
   agent needs to leave or be back, or executes a purchase, are all
-  instances where variables representing time would be used.  As
-  explained in the \textsf{CP} section, a variable can take values
-  from a domain. In \eu, domains are defined over a specific data
-  type, the primitive data types supported are:
-
-  \begin{enumerate}
-  \item \textit{String}: sequences of characters, like "Red",
-    "Spirit", etc
-
-  \item \textit{Boolean}: {True,False}
-
-  \item \textit{Numeric}: integers and floats
-
-  \item \textit{Object}: reference to an object instance (see below
-    for a description of objects in \eu)
-
-\end {enumerate}
-
-Two types of variable Domains can be defined over a data type:
-
-  \begin{enumerate}
-  \item \textit{Enumerated}: Finite sets over any data type, specified
-    explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
-  \item \textit{Interval}: Only for Numeric data types, they are
-    specified by a [LowerBound,UpperBound] pair, for instance: [1,10],
-    [5.0,100.0], etc.
-  \end {enumerate}
-
-  \eu's ability to represent and reason over numeric intervals is a
-  very useful characteristic that allow users to create flexible
-  plans. \comment{say more about the benefits of flexibility?}
+  instances where variables representing time is used.  As noted in
+  Section \ref{sec:europa:cp}, a variable can take values from a
+  domain. In \eu, domains are defined over a specific data type, the
+  primitive data types supported are, \textit{String},
+  \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
+  below). Further, two types of variable Domains can be defined over a
+  data type, \textit{Enumerated} and \textit{Interval} (used only for
+  numeric data types, they are specified by a [LowerBound,UpperBound]
+  pair, for instance: [1,10], [5.0,100.0]). It is this latter ability
+  to represent and reason over numeric intervals that forms the basis
+  of representing flexible plans.
 
@@
   may only be able to perform one task at a time, may have limited
-  time or an energy or fuel budget to perform a task, and so on. In
-  \eu, these restrictions are represented by Constraints, which can be
-  defined over any combination of Variables in a domain. \eu provides
-  a Constraint Library that already implements many useful constraints
-  (temporal, resource, relational, etc), it is also possible to add
-  new constraints if required by a particular domain.
+  time or an energy or fuel budget to perform a task. In \eu, these
+  restrictions are represented by constraints, which can be defined
+  over any combination of variables in a domain. \eu provides a
+  \emph{Constraint Library} that implements a number of useful
+  constraints (temporal, resource, relational), while allowing the
+  possibility to add new constraints if required by a particular
+  domain.
 
 \item[\textbf{Objects}] The items we wish to describe and refer to in
-  a domain are considered Objects. As in the case with object-oriented
+  a domain are considered objects. As in the case with object-oriented
   analysis and design, one can seek out the nouns in any domain
   description to find likely objects. In the Shopping Agent example,
-  we might consider the Agent itself, Products and Locations all to be
-  objects. Objects have state and behavior. For example, a Shopping
-  Agent can have:
-
-  \begin{itemize}
-
-  \item State: its location, a bag that contains the products it has
-    already purchased (the bag can in turn be another object), etc.
-
-  \item Behavior: go to a location to look for a product, perform a
+  we might consider the Agent, Products and Locations all to be
+  objects. Objects have state and behavior, for example, a Shopping
+  Agent can have, its location, a bag that contains the products it
+  has already purchased (the bag can in turn be another object) and
+  behavior as in going to a location to look for a product, perform a
   purchase, return home, etc.
-
-  \end{itemize}
 
@@
   generically in terms of object types or classes. \eu allows the
-  definition of classes in the same way that it is done in popular
-  object-oriented programming languages, where a class can have
-  attributes (represented by Variables), and classes can be arranged
-  in a single-parent hierarchy through inheritance.  However, one key
-  aspect needs to be introduced to class definition in \eu that is not
+  definition of classes similar to object-oriented programming
+  languages, where a class can have attributes (represented by
+  variables), and classes can be arranged in a single-parent hierarchy
+  through inheritance.  However, one key aspect in \eu that is not
   found in object-oriented programming: to describe state and behavior
   for the purposes of planning we need to build on the formalism of
-  first order logic as explained below.
+  first order logic.
 
 \item[\textbf{Tokens}] In first order logic, a predicate defines a
-  relation between objects and properties \comment{citation?}.  In
-  \eu, we define such relations between variables whose domains are
-  sets of objects and sets of properties to describe state and
-  behavior. For example, we might use a predicate \texttt{At($a$,$l$)}
-  to indicate that agent $a$ is at location $l$, or a predicate
-  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
+  relation between objects and properties \cite{gen87}.  In \eu, we
+  define such relations between variables whose domains are sets of
+  objects and sets of properties to describe state and behavior. For
+  example, one might use a predicate \texttt{At($a$,$l$)} to indicate
+  that agent $a$ is at location $l$, or a predicate
+  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking action to
   buy product $p$. Note that $a$ is a variable which may have a number
@@
   where the domains of variables $a$ and $l$ can have more than one
-  possible value in them, or grounded plans, where single values will
-  be specified for each variable as we saw in the case of the
-  N-queen's problem.
+  possible value. Or they can be grounded plans, where single values
+  will be specified for each variable as showing in the N-queen's
+  problem.
 
-  In general, to come up with an executable plan it is not sufficient
-  to state predicates that describe required state or behavior without
-  also specifying some temporal extent over which each of those
-  predicates hold. Predicates that are always true can be thought to
-  hold from the beginning to the end of time. However, in practice,
-  the temporal extent of interest must be defined with timepoints to
-  represent its start and end. So we might want to write
-  \texttt{At($a$,$s$,$e$,$l$)} to indicate that the agent $a$ is at
-  location $p$ from time $s$ to time $e$. In fact, this pattern of
-  using such predicates to describe both state and behavior of objects
-  is so prevalent in \eu that we have introduced a special construct
-  called a Token which has the built in variables to indicate the
-  object to which the statement principally applies and the timepoints
-  over which it holds. A Token is an instance of a predicate that
-  represents and object's state or behavior and is defined over a
-  temporal extent. In \eu, all predicate instances are Tokens. Also,
+  In general, for an executable plan it is insufficient to state
+  predicates that describe required state or behavior without also
+  specifying some temporal extent over which each of those predicates
+  hold. Predicates that are always true can be thought to hold from
+  the beginning to the end of the planning horizon. However, in
+  practice, the temporal extent of interest must be defined with
+  timepoints \cite{Dean88} to represent its start and end.  In such
+  representation, one writes \texttt{At($a$,$s$,$e$,$l$)} to indicate
+  that agent $a$ is at location $l$ from time $s$ to time $e$. In
+  fact, this pattern of using such predicates to describe both state
+  and behavior of objects is prevalent enough that \eu the token
+  representation. Token have built in variables to indicate the object
+  to which the statement principally applies as well as the timepoints
+  over which it holds. A token is an instance of a predicate that
+  represents an object's state or behavior and is defined over a
+  temporal extent. In \eu, all predicate instances are tokens. Also,
   in the same way that objects are described generically by classes,
-  in \eu Tokens can be described generically by Token Types. Every
+  in \eu tokens can be described generically by token types. Every
   token has five built-in variables:
@@
   \item \textit {start}: The beginning of the temporal extent over
-    which the predicate is defined.
+    which the token predicate is defined.
 
@@
 
-  \item \textit {duration}: The duration of the temporal extent over
-    which the predicate is defined. The constraint \textit{start} $+$
-    \textit{duration} = \textit{end} is enforced automatically.
+  \item \textit {duration}: The duration of the temporal extent. The
+    constraint \textit{start} $+$ \textit{duration} = \textit{end} is
+    enforced automatically.
 
   \item \textit{object}: The set of objects to which a token might
-    apply. In a grounded plan each Token applies to a specific Object,
-    reflecting the intuition that we are using Tokens to describe some
-    aspect of an Object (i.e. its state or behavior) in time. However,
-    in a partial plan, the commitment to a specific object may not yet
-    have been made.
+    apply. In a grounded plan representation each token applies to a
+    specific object, reflecting the intuition that tokens are used to
+    describe some aspect of an object (\ie state or behavior) in
+    time. However, in a partial plan, it is possible that the
+    commitment to a specific object may not yet have been made.
 
@@
     token's current state and its reachable states through further
-    restriction.  This is required to support CAIP's approach to
-    planning where goals and subgoals can be satisfied by either
-    adding new intervals (Tokens) to the plan, or by merging with
-    existing intervals. Token State lifecycle is examined in detailed
-    in a separate section below.
+    restriction.  This is required to support \texttt{CAIP}'s approach
+    to planning where goals and subgoals can be satisfied by either
+    adding new intervals (tokens) to the plan, or by merging with
+    existing intervals. % Token state lifecycle is examined in detailed
+    % in a separate section below.
 
@@
   some classes that don't have any time-dependent state or behavior,
-  and therefore those classes will not have any Token types associated
-  with them, for instance in the Shopping Agent model, the set
-  locations and products are static for the agent's purposes so we may
-  only want to say what they are but not have any state or behavior
-  associated with them.
-
-  However, the most common case in any non-trivial domain model is
-  that Objects will have many associated Tokens in order to describe
-  their state and behavior throughout a plan. There are a couple of
-  Object Types that are so commonly used in domain descriptions, that
-  \eu provides a built-in implementation for them:
+  and therefore those classes will not have any token types associated
+  with them. For instance in the Shopping Agent model, the set
+  locations and products are static for the agent's purposes so one
+  may only want to declare them but not associate any state or
+  behavior with them.  However, a common case in a non-trivial domain
+  model is that objects will have many associated tokens in order to
+  describe their state and behavior throughout a plan. There are some
+  object types that are commonly used in domain descriptions, that \eu
+  provides a built-in implementation for them:
 
@@
   by exactly one token for every given timepoint in the plan. Any
-  instances of a class derived from \eu's built-in Timeline class will
+  instances of a class derived from \eus built-in timeline class will
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \comment{add small
-    example?}
+  overlap may occur among them \cite{mus94}.  \kcomment{add a figure}
 
@@
   of changes that can occur to impact that state i.e. produce,
-  consume, use, change.  Resources are such a common requirement for
-  \eu users that built-in object types (with their corresponding token
-  types to denote production, consumption, etc) are provided for
-  them. Instances of classes derived from a Resource will induce
-  ordering requirements on their Tokens in order to ensure that the
-  level of the resource remains within specified limits.  \comment{add
-    small example?}
+  consume, use, change.  Resources are a common requirement for \eu
+  users that built-in object types (with their corresponding token
+  types to denote production, consumption, etc) are provided.
+  Instances of classes derived from a resource will induce ordering
+  requirements on their tokens in order to ensure that the level of
+  the resource remains within specified limits.  \kcomment{This
+    definitely needs a visual example and some text esp. since it
+    impacts plan structure via scheduling}
+
 \end{enumerate}
 
-\item[\textbf{Token State Model}] In \eu's representation, goals are
+\item[\textbf{Token State Model}] In \eus representation, goals are
   temporally scoped states that need to be achieved, or actions that
   must be performed. As a result, \eu internally represents goals as
-  Tokens.  Goals can be posted explicitly, for instance, when a user
+  tokens.  Goals can be posted explicitly, for example, when a user
   asks the Shopping Agent to own a specific product by a specific
   time, or implicitly when subgoals are created as a result of domain
-  compatibilities (described in the CAIP framework above), for
-  instance, when the Shopping Agent must get to a Location where a
-  Product is available before it is able to purchase it, in this case
-  the original goal of purchasing the product will spawn a subgoal of
-  being at a specific location.
+  compatibilities \kcomment{Define a compatibility}. For instance,
+  when the Shopping Agent must get to a location where a product is
+  available before it is able to purchase it: in this case the
+  original goal of purchasing the product spawns a subgoal of being at
+  a specific location without which the purchase will not occur as
+  defined in the domain model.  Looking at a how a simple Shopping
+  Agent goal would be addresed by \eu helps illustrate how Token
+  States support the planning process.  Let's assume the planner
+  starts with a partial plan that places the Shopping agent at home at
+  time $0$ with no products in possession.  If a goal is posted for
+  the Agent to own a drill by time 10. \kcomment{Use Fig 4 and edit::
+    Add figure showing an AgentLocation timeline with an At(Home)
+    token with start=0 and end open, an empty AgentBag timeline, and a
+    Own(Drill) token with start=10 floating below both timelines}
 
-  Looking at a how a simple Shopping Agent goal would be addresed by
-  \eu helps illustrate how Token States support the planning process.
-
-  Let's assume we start with a partial plan that places the Shopping
-  agent at home at time 0 and with no Products in its possession, then
-  we post a goal for the Shopping Agent to own a Drill at time 10.
-
-  \comment{Add figure showing an AgentLocation timeline with an
-    At(Home) token with start=0 and end open, an empty AgentBag
-    timeline, and a Own(Drill) token with start=10 floating below both
-    timelines}
-
-Notice that the \texttt{Own(Drill)} token that represents the goal is
+  Note that the \texttt{Own(Drill)} token that represents the goal is
   not immediately placed on the \texttt{AgentBag} timeline since \eu
-must determine whether it can achieve that goal and how. When the goal
-is posted and before any decisions are made on it, the
-\texttt{Own(Drill)} token is created in an INACTIVE state, then \eu
-looks at the current partial plan and considers 2 alternatives:
+  must determine whether it can achieve that goal and how. When the
+  goal is posted and before any decisions are made, the
+  \texttt{Own(Drill)} token is created in an \texttt{INACTIVE}
+  state. \eu then considers the partial plan and evaluates two
+  possible alternatives:
 
@@
 \item If there are any tokens in the plan that are compatible with the
-  new goal, \eu may attempt to satisfy the goal by merging it with any
-  of them (if there is more than one it can try them one at a time),
-  in this case, the state of the token would change to MERGED.
+  new goal, \eu may attempt to satisfy the goal by \emph{merging} it
+  with any of them (if there is more than one it can try them one at a
+  time or in some heuristic order), in this case, the state of the
+  token would change to \texttt{MERGED}.
 
 \item Instead of merging with existing tokens in the plan, \eu may
-  decide to insert the new token into the plan, in that case its state
-  would change to ACTIVE.
+  decide to \emph{insert} a new token into the plan, in which case the
+  state would change to \texttt{ACTIVE}.
 
@@
 are transferred to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. Two tokens are merge-compatible if they are
-instances of the same predicate and that no intersections between
-corresponding variables are empty. The effect of merging is
-illustrated in Figure 3.
+token requires finding a target active token that is compatible. Two
+tokens are merge-compatible if they are instances of the same
+predicate and that no intersections between corresponding variables
+are empty. The effect of merging is illustrated in Figure 3
+\kcomment{Which figure?}.\comment{Use previous fig and modify as
+  appropriately:: Add figure showing the effect of merging; all figs
+  on this section are conjoined. JB to sketch it out.}
 
-\comment{Add figure showing the effect of merging; all figs on this
-  section are conjoined. JB to sketch it out.}
+When a token is activated all compatibilities \kcomment{this should
+  have been defined earlier} associated with its type are evaluated,
+which may lead to new subgoals being generated. In our example, there
+could be a compatibility stating that the Shopping Agent must be at
+the same location as the product it is purchasing. In such a case,
+activating the \texttt{Own(Drill)} token will cause a new
+\texttt{At(Drill.location)} to be created in an \texttt{INACTIVE}
+state, which in turn could spawn its own subgoals.
 
-When a token is activated all compatibilities associated with its
-token type are evaluated, which may lead to new subgoals being
-generated. In our example, there should be a compatibility stating
-that the Shopping Agent must be at the same location as the product it
-is purchasing. In that case, activating the \texttt{Own(Drill)} token
-will cause a new \texttt{At(Drill.location)} to be created in an
-INACTIVE state, which in turn will have to be dealt with.
+\comment{Use previous fig and modify as appropriately: Add figure
+  showing the effect of activating the token by putting it on the
+  AgentBag timeline, and a new At(Drill.location) token floating
+  below}
 
-\comment{Add figure showing the effect of activating the token by
-  putting it on the AgentBag timeline, and a new At(Drill.location)
-  token floating below}
+If neither activating nor merging work, a token may be
+\texttt{REJECTED}, this is possible only for goals posted by the end
+user and specified as optional. Subgoals generated as a result of
+token activation cannot be rejected; if a subgoal cannot be satisfied
+through activation or merging then it will cause the original
+activation decision that spawned it to be retracted.
 
-If neither activating nor merging works, a token may become REJECTED,
-this is possible only for goals posted by the end user and specified
-as optional. Subgoals generated as a result of token activation cannot
-be rejected, if a subgoal cannot be satisfied through activation or
-merging then it will cause the original activation decision that
-spawned it to be retracted (\eu's search mechanisms are explained in
-detail in a separate section later on).
+As can be seen from the above example, a token's state can go from
+\texttt{INACTIVE} to \texttt{ACTIVE, MERGED} or \texttt{REJECTED}, as
+shown in figure below \kcomment{Which figure?}.
 
-As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED
-or REJECTED, this lifecycle is illustrated in the figure below.
-
-\comment{Add figure showing possible state transitions. Need a FSM.}
+\comment{Maybe: Add figure showing possible state transitions. Need a FSM.}
 
 \item[\textbf{Rules}] In order for a plan to be valid, it must comply
-  with all rules and regulations pertinent to the application domain
-  in question. Rules govern the internal and external relationships of
-  a token. For example, consider a parameterized action
+  with all rules pertinent to the relevant application domain. Rules
+  govern the internal and external relationships of a token. For
+  example, consider a parameterized action
   \texttt{Go(origin,destination)} which describes a Shopping Agent
-  traveling from one location ( \textit{origin}) to another (
-  \textit{destination}). The parameters are instantiated on a token as
-  variables whose domain of values is the set of all locations in a
-  given problem. A rule governing an internal relationship among token
-  variables might stipulate that a transition must involve a change in
-  location. This can be easily expressed as a constraint on the
-  definition of a predicate of the form: \textit{origin !=
-    destination}. It makes sense to further stipulate that the agent
-  must be located at \textit{origin} before travel can happen and the
-  agent must be located at \textit{destination} when completed. This
-  is an example of a rule governing an external relationship among
-  tokens. It specifies a requirement that tokens of the predicate
-  \texttt{At} that represent the location state for the Shopping Agent
-  precede and succeed tokens of the predicate \texttt{Go} that
-  represents traveling.
+  traveling from one location (\textit{origin}) to another
+  (\textit{destination}). The parameters are instantiated on a token
+  as variables whose domain of values is the set of all possible
+  locations in a given problem. A rule governing an internal
+  relationship among token variables might stipulate that a transition
+  must involve a change in location. This can be expressed as a
+  constraint on the definition of a predicate of the form:
+  \textit{origin != destination}. A further stipulation could be that
+  the agent must be located at \textit{origin} before travel can occur
+  and the agent must be located at \textit{destination} when travel is
+  completed. This is an example of a rule governing an external
+  relationship among tokens. It specifies a requirement that tokens of
+  the predicate \texttt{At} that represent the location state for the
+  Shopping Agent precede and succeed tokens of the predicate
+  \texttt{Go} that represents traveling.
 
-  \comment{ Add figure showing internal and external relationships
+  \comment{Use previous fig and modify as appropriately (used as
+    above): Add figure showing internal and external relationships
     from rules on action Go. Need a fig from JB}
 
-Figure xx illustrates the entities and relations involved in
-specifying such a rule on a \texttt{Go} action. The token on which the
-rule applies is referred to (\texttt{Go}) as the master. Each
-\texttt{At} token required by the master is referred to as a
-slave. All variables are indicated by name and their domains are
-expressed as intervals in the case of temporal variables and as
-enumerations for the remainder. Application of a rule on a token can
-thus cause slave tokens, variables, and constraints to be introduced.
-
-The capability of a domain rule to cause a slave token to be created
-is a key vehicle through which planning occurs. Semantically, this
-rule imposes a requirement for supporting (slave) tokens to be in the
+  Figure xx \kcomment{which figure?} illustrates the entities and
+  relations involved in specifying such a rule on a \texttt{Go}
+  action. The token on which the rule applies is referred to
+  (\texttt{Go}) as the \emph{master}. Each \texttt{At} token required
+  by the master is referred to as a \emph{slave}. All variables are
+  indicated by name and their domains are expressed as intervals in
+  the case of temporal variables and as enumerations for the
+  remainder. Application of a rule on a token can thus cause slave
+  tokens, variables and constraints to be introduced.  The capability
+  of a domain rule to cause a slave token to be created is a key
+  vehicle through which planning occurs. Semantically, this rule
+  imposes a requirement for supporting (slave) tokens to be in the
   plan in order for the master to be valid.
 
-\comment {talk about condition/effect annotations?}
-
 \end{description}
-
-%\begin{figure*}
-%\centering 
-%\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
-%\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
-%\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
-%\caption{\small An illustrative plan synthesis example with concurrent
-%  timelines. Each timeline is an instantiation of a subsystem tracked
-%  by the planner. Tokens describe the instantiated state of subsystem
-%  at a particular time instance and enforce temporal and parametric
-%  constraints. Abstract goals are decomoposed into successively less
-%  abstract tokens and instantiated in co-temporal timelines using
-%  \texttt{Allen Algebra} relations. All tokens represent flexible
-%  start/end times. \ref{fig:plan-evolve1} shows an initial state
-%  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
-%  \label{fig:Plan-evolve}
-%  \vskip-5pt
-%\end{figure*}
-
-%Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
-%timelines which track \texttt{Goal, Path, Navigation} and
-%\texttt{Command} state over time. Tokens on the goal timeline indicate
-%a survey within a bounded volume and a min/max depth envelope. As the
-%Volume Survey sub-goals on the \texttt{Path} timeline, the initial
-%token on that timeline gets \emph{squeezed} to have dependencies of
-%the goal token instantiated. Initially this dependency are the tokens
-%\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
-%coverage. The first of these in turn generates a sub-goal on the same
-%timeline to go to the next intermediate waypoint, which is illustrated
-%in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
-%figure) ultimately generate all the tokens to completion for this
-%timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
-%in plan generation and shows additional subgoals on the \texttt{Path,
-%  Navigation} and \texttt{Command} timelines. Each token can thus
-%trace its causality when instantiated in the plan. The domain model
-%provides source of these temporal and parametric dependencies in the
-%form of rules.
-
-%\begin{minipage}[c]{\textwidth}
-%\vspace{+0.5cm}
-% \framebox[\textwidth][t]{
-%Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
-%$\Rightarrow$ \\
-%met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
-%starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
-%\ldots{} \\
-%ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
-%meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
-%\end{minipage}
-
-% \end{enumerate}
-
-% \subsubsection{Partitioned Control}
-
-% \subsubsection{State Estimation}
-
-% Even as automated reasoning approaches have the ability to dynamically
-% retarget the vehicle, estimating environmental signals of interest (or
-% their proxies) is important to be able to enable opportunistic science
-% in the water-column. In \texttt{T-REX} feature recognition revolves
-% around a Hidden Markov Model (HMM) \cite{rabiner86} which is encoded
-% directly within the unified representational and computational
-% framework of a reactor. HMMs are useful since the stochastic nature of
-% these models can correlate the type of features we want to detect with
-% the sensor observations. Online targeted sensor data is classified by
-% apriori generated cluster data which in turn determines the
-% probability of having seen the feature of interest. Together with
-% posterior probability, the HMM generates a probability of being within
-% the target feature if it dominates the distribution. If other sampling
-% conditions are satisfied (for instance sampling proximity), a
-% constraint is activited in the model which can trigger a water sampler
-% and also preserve the memory of sampling to alter a future transect
-% resolution.
-
-% The HMM is customized for the feature of interest and is built offline
-% using clustering techniques from large data sets; we use Kohonen's
-% Self Organizing Maps \cite{kohonen}. We use a semi-supervised learning
-% method \cite{zhu05} to extract an environmental model that make uses
-% of both labeled and unlabeled data for training. Online,
-
-% The sensor data for classification depends on the
-% feature of interest; for detecting INLs we use backscattering data
-% from a HydroScat and for tracking riverine plumes to track fertilizer
-% runoff, we use an ISUS Nitrate sensor.
 

### 2012-02-22 europa-basics.tex
updates to europa-basics and concepts

@@
 development of \rx has been targeted at surveying a number of
-oceanographic features which are dynamic and unpredictable
-spatio-temporally. Typically this requires our AUVs to balance the
-goals of coverage spatially while opportunistically following features
-of scientific interest and to do so while being operationally aware of
-its own limitations in terms of resources (\eg the battery state of
-charge) and overall proximity to other observational assets for
-obtaining scientific ground-truth. We are interested in
-representational frameworks that allow robots to pursue long-term
-objectives while choosing short-term gain and can gracefully deal with
-exogenous or endogenous change.
+oceanographic features which are dynamic and spatio-temporally
+unpredictable. Typically this requires our AUVs to balance the goals
+of coverage spatially while opportunistically following features of
+scientific interest and to do so while being aware of its own
+limitations in terms of resources (\eg the battery state of charge)
+and overall proximity to other observational assets for obtaining
+scientific ground-truth. We are interested in representational
+frameworks that allow robots to pursue long-term objectives while
+choosing short-term gain and can gracefully deal with exogenous or
+endogenous change.
 
 To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use the
-temporal constraint-based planner \eu with a demonstrated legacy of
-having flown on NASA space missions \cite{mus98,jonsson00,bresina05,
-  barreiro09}. Our autonomy architecture brings three key innovations
-for AUV adaptation: the use of flexible plan representations,
-compositional control with the use of partitioned networks and a
-demonstrated capability at sea in novel ocean observation methods.
+to be able to synthesize plans insitu and to re-plan without human
+intervention. To do so we use a temporal constraint-based planner with
+a demonstrated legacy of having flown on NASA space missions. Our
+autonomy architecture brings three key innovations for AUV control:
+in-situ plan synthesis, the use of flexible plan representations with
+a sound theoretical basis and systematic compositional control with
+the use of partitioned networks doing so with a demonstrated
+capability at sea in novel ocean observation methods.
 
 The next few sections will cover \eu in a bottom-up fashion, going
-from the generic theoretical frameworks that underlie \eu, to the
-specific representation, modeling and algorithms implemented by
-it. First, a quick introduction to Constraint Programming and
-Constraint-Based Planning is provided.  Next, the specific
-representation that \eu uses to perform Constraint-Based Planning is
-presented.  After that comes an overview of \eu's architecture and how
-it can be embedded in a planning application such as \rx. Finally,
-\eu's modeling, inference and search mechanisms are covered in detail
-so that the reader can understand how \rx can perform deliberative
-planning and react to execution events by embedding and customizing
-\eu.
+from the generic theoretical framework that underlies \eu, to the
+specific representation, modeling and algorithms implemented by it.  A
+brief introduction to Constraint Programming and Constraint-Based
+Planning is provided followed by the specific representation that \eu
+uses to undertake Constraint-Based Planning.  Subsequently, we provide
+an overview of \eus architecture and how it can be embedded in a
+planning application such as \rx. Finally, \eus modeling, inference
+and search mechanisms are covered in detail so that the reader can
+understand how \rx can perform deliberative planning and react to
+execution events by embedding and customizing \eu.
 
@@
 To articulate the fundamentals of automated planning briefly and use
-that to motivate the mechanisms we use in our specific form of the
-technique we start with a simple example.
+that to motivate the mechanisms we use in our specific form of
+planning we start with a simple example.
 
@@
 At the core of the \rx framework, is the deliberation engine, \eu,
-which has a rich legacy from NASA missions. \eu is a versatile
-Constraint-based temporal planner which continues to be deployed on a
-diverse set of applications. We motivate this section with some
-problem domains this planner can handle and then delve in substantial
-detail.
+which has a rich legacy from NASA missions \cite{mus98,rajan00,
+  jonsson00,aichang04, bresina05}. \eu is a versatile Constraint-based
+Temporal Planner which continues to be deployed on a diverse set of
+applications including recently, planning for the International Space
+Station \cite{barreiro09}. We motivate this section with some problem
+domains this planner can handle and then proceed to go into detail.
 
@@
 dealing with constraints is the $N$-Queens problem in which chess queens
-must be placed on an  $N$x$N$ chessboard so no queens attack the other. Fig.
-\ref{fig:nqueens-1} shows an example of a random positioning of queens
-on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
-are highlighted in red.
-
-\begin{figure} \centering
-  \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
-  \caption{\small N-Queens problem. Queens in violation of the
-    non-attack constraint are highlighted in red.}
-\label{fig:nqueens-1}
-\vskip+0.1cm
-\end{figure}
-
+must be placed on an  $N$x$N$ chessboard so no queens attack the other. % Fig.
+% \ref{fig:nqueens-1} shows an example of a random positioning of queens
+% on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
+% are highlighted in red.
+% \begin{figure} \centering
+%   \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
+%   \caption{\small N-Queens problem. Queens in violation of the
+%     non-attack constraint are highlighted in red.}
+% \label{fig:nqueens-1}
+% \vskip+0.1cm
+% \end{figure}
 If we define $N$x$N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in [1,N]$,
-Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen, $0$
-otherwise. Then the following constraints need to be satisfied:
+Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen,
+$0$ otherwise. Then the following constraints need to be satisfied:
 
@@
 The problem can be solved by finding assignments for all variables
-Q$_{rc}$ that satisfy the above constraints. Fig. \ref{fig:nqueens-2} is
-a solution found by \eu using a compact version of that formulation and a specialized search
-procedure.
+Q$_{rc}$ that satisfy the above constraints. Fig. \ref{fig:nqueens-2}
+is a solution found by \eu using a compact version of that formulation
+and a specialized search procedure.
 
@@
 \paragraph \textit{Scheduling}: In the Resource Constrained Project
-Scheduling Problem (RCPSP) \cite{Bruckera99}, a project consisting of
-a set of activities must be scheduled in a way that satisfies minimum
-and/or maximum temporal separation constraints. The activity schedule
-must also respect fixed limits on the availability of resources
-required to perform each activity. In addition to satisfying temporal
-and resource constraints, it is common for the user to want to
-minimize makespan so that the entire project is finished as early as
-possible. Fig. \ref{fig:rcpsp-1} shows an example of a a solution
-provided by \eu for an RCPSP instance with 10 activities, 5 resources
-and 30 temporal constraints.
+Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, a project
+consisting of a set of activities must be scheduled in a way that
+satisfies minimum and/or maximum temporal separation constraints. The
+activity schedule must also respect fixed limits on the availability
+of resources required to perform each activity. In addition to
+satisfying temporal and resource constraints, it is common for the
+user to want to minimize makespan \cite{ghallab04} so that the entire
+project is finished as early as possible. Fig. \ref{fig:rcpsp-1} shows
+an example of a a solution provided by \eu for an \texttt{RCPSP}
+instance with 10 activities, 5 resources and 30 temporal constraints.
 
@@
 \includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small A \eu solution to an RCPSP \cite{Bruckera99} problem.}
+\caption{\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
+  problem.}
 \label{fig:rcpsp-1}
@@
 \paragraph \textit{Planning}: In the Shopping Agent Problem
-\cite{russelnorvig} an agent needs to purchase a set of products (milk,
-drill, etc) that are available at specific locations (supermarket,
-hardware store, etc), the agent is subject to temporal (must complete
-tasks by specific deadlines) and resource (fuel, carrying capacity, etc)
-constraints. The agent needs to figure out what actions need to be
-performed to find and acquire the required items, as well as when to
-perform each of those actions. Fig. \ref{fig:shopping-1} shows a
-solution produced by \eu for such a problem instance.
+\cite{russelnorvig} an agent needs to purchase a set of products
+(milk, drill, etc) that are available at specific locations
+(supermarket, hardware store, etc) and the agent is subject to
+temporal (must complete tasks by specific deadlines) and resource
+(fuel, carrying capacity, etc) constraints. The agent needs to figure
+out what actions need to be performed to find and acquire the required
+items, as well as when to perform each of those
+actions. Fig. \ref{fig:shopping-1} shows a solution produced by \eu
+for such a problem instance.
 
@@
 achieve stated goals) and both Planning and Scheduling may embed a
-Constraint Satisfaction problem. 
-The relationships between Planning, Scheduling and Constraint
-Satisfaction have been examined \cite{smith00} and have lead to use of
-constraint reasoning in \eu as its innermost building block.
+Constraint Satisfaction problem.  The relationships between Planning,
+Scheduling and Constraint Satisfaction have been examined
+\cite{smith00} and have lead to use of constraint reasoning in \eu as
+a core building block.
 
@@
 Constraint Satisfaction Programming (also known as Constraint
-Programming or \textsf{CP}) is a discipline that provides a generic framework
-for representing, solving and making logical inference on
+Programming or \textsf{CP}) is a discipline that provides a generic
+framework for representing, solving and making logical inference on
 constraints. A complete treatment of this discipline is given in
-\cite{marriott98,apt03} and very concise introductions are provided in
+\cite{marriott98,apt03} with concise introductions provided in
 \cite{bartak99,lustig01}.
@@
 objective is to find one or more value assignments to $V$ where all
-constraints are satisfied. To solve a problem, \textsf{CP} techniques use
-logical inference to perform Constraint Propagation, which consists of
-two operations:
+constraints are satisfied. To solve a problem, \textsf{CP} techniques
+use logical inference to perform Constraint Propagation, which
+consists of two operations:
 
@@
 
-\item \textbf{Bounds propagation}: To infer upper and lower variable
-  bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \&
-  $x_i \geq 0$, we can infer $[0,2]$ bounds for $x_1$ and $x_2$
+\item \textbf{Bounds propagation} infers upper and lower variable
+  bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \& $x_i
+  \geq 0$, we can infer $[0,2]$ bounds for $x_1$ and $x_2$.
 
-\item \textbf{Domain reduction}: To infer a valid set of values for a variable.
-  For example, for  constraints allDifferent($x_1,x_2,x_3$), $x_i \geq
-  0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can infer that the
-  valid domain for $x_3$ is reduced  to $\{2,4\}$
+\item \textbf{Domain reduction} infers a valid set of values for a
+  variable.  For example, for constraints allDifferent($x_1,x_2,x_3$),
+  $x_i \geq 0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can
+  infer that the valid and reduced domain for $x_3$ is $\{2,4\}$.
 
@@
 A set of constraints and variables that describe a problem domain are
-typically represented as a Constraint Network, where the variables are
-nodes and constraints are arcs between the variables. Having this
-network representation in mind, a pair of variables $x,y$ is said to
-be arc consistent if for every value in $x$'s domain, there is a value
-in $y$'s domain that is consistent with all the constraints that
-connect $x$ and $y$.  A naive approach to ensuring arc consistency
-consists of cycling through all the variable pairs and performing
-constraint propagation until there are no variable domain changes. The
-\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation, improves
-over the naive approach by ignoring constraints whose variables were
-not affected since the last iteration. Fig.  \ref{fig:constraintnet} shows a set of constraints on 3 variables
-($x$, $y$ and $z$) represented as a constraint network, and their domains before arc consistency is enforced.
-After the AC-3 algorithm is executed, the domains for the 3 variables
-would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.
+typically represented as a \emph{Constraint Network}, where the
+variables are nodes and constraints are arcs between the variables. A
+pair of variables $x \& y$ is said to be \emph{arc consistent} if for
+every value in $x$'s domain, there is a value in $y$'s domain that is
+consistent with all the constraints that connect $x$ and $y$.  A naive
+approach to ensuring arc consistency consists of cycling through all
+the variable pairs and performing constraint propagation until there
+are no variable domain changes. The \texttt{AC-3} algorithm
+\cite{mackworth77}, a popular implementation, improves over the naive
+approach by ignoring constraints whose variables were not affected
+since the last iteration. Fig. \ref{fig:constraintnet} shows a set of
+constraints on $3$ variables ($x$, $y$ and $z$) represented as a
+constraint network, and their domains before arc consistency is
+enforced.  After the \texttt{AC-3} algorithm is executed, the domains
+for the variables would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.
 
@@
 
-Arc Consistency and other, more stringent consistency definitions can
+Arc Consistency and other more stringent consistency definitions can
 be used to efficiently prove that a \textsf{CP} problem is
@@
 problem is satisfiable and provide specific variable assignments that
-constitute solutions. For that, search algorithms like global
+constitute solutions. To do so, search algorithms like global
 backtracking \cite{hooker05} or local search are used; these use
@@
 \cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
-\cite{ghallab04}, but often very efficient in practice using a number
-of consistency and search algorithms that are well understood.
+\cite{ghallab04}, but often in practice a number of consistency and
+search algorithms can be used.
 
@@
 mutual exclusion and concurrency using propositions \cite{ghallab04}.
-Constraint Programming is a framework that lends itself very well to
-represent and reason about all of those elements \cite{ghallab94},
-consequently, there has been a good amount of work building automated
-planners that take advantage of CP techniques.  The most common
-approach in constraint-based planning has been to encode the entire
-planning problem as a CP problem, where action choices and
-relationships are represented as variables and/or constraints
+Constraint Programming is a framework that lends itself well to
+represent and reason about all of those elements \cite{ghallab94}.
+Consequently, there has been substantial effort in building automated
+planners that take advantage of \textsf{CP} techniques.  The most
+common approach in constraint-based planning has been to encode the
+entire planning problem as a \textsf{CP} problem, where action choices
+and relationships are represented as variables and/or constraints
 \cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
-pure CP formulation is that standard CP solvers can be then used to
-solve the planning problem, however, this approach also has a couple
-of significant drawbacks:
+pure \textsf{CP} formulation is that standard solvers can be then used
+to solve the planning problem. However, this approach also has a
+couple of significant drawbacks:
 
@@
 
-\item Given that action choices and relationships (rules) are expressed
-  through variables, the domain descriptions that result from this
-  approach are not intuitive and therefore difficult to understand and
-  debug
+\item Given that action choices and relationships (rules) are
+  expressed through variables, the domain descriptions that result
+  from this approach are not intuitive and therefore difficult to
+  understand and debug.
 
@@
 
-These drawbacks have been addressed by a small number of planning
-frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
-\texttt{IxTet} \cite{ghallab87,lemai04}, and \texttt{CAIP} (Constraint-Based
-Interval and Attribute Planning)\cite{mus94,frank2003}.  These
-frameworks take advantage of a CSP representation to deal with time
-and resources, at the same time maintaining an explicit representation
-of the elements relevant to planning (action choices, causal
-relationships, etc). This combination results in a planning approach
-that can deal with the type of constraints found in real-life
-problems, while allowing for intuitive problem descriptions and a
-natural way to express search algorithms and heuristics. \eu is an
-implementation of the \texttt{CAIP} framework. In \texttt{CAIP}, planning problems are
-specified in terms of:
+These drawbacks have been addressed by a number of planning frameworks
+like \texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
+\cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
+and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
+advantage of a CSP representation to deal with time and resources, at
+the same time maintaining an explicit representation of the elements
+relevant to planning (action choices, causal relationships, etc). This
+combination results in a planning approach that can deal with the type
+of constraints found in real-world problems while allowing for
+intuitive problem descriptions and a natural way to express search
+algorithms and heuristics. \eu is one such implementation of the
+\texttt{CAIP} framework, in which planning problems are specified in
+terms of:
 
@@
 \item \textbf{Variables and Constraints}: Variables represent Interval
-  attributes and Constraints enable a direct representation of
-  temporal, resource and any other the restrictions that a valid plan
-  must comply with.
+  attributes and constraints enable a direct representation of
+  temporal, resource and any other restrictions that a valid plan must
+  comply with.
 
-\item \textbf{Domain Configuration Rules}: this is an explicit
+\item \textbf{Domain Configuration Rules}: This is an explicit
   representation of the conceptual relationships between actions and
@@
   product is available.
+
 \end{enumerate}

### 2012-02-22 intro.tex
updates to europa-basics and concepts

@@
 within \rx. A key portion is Section~\ref{sec:europa:pr} which details
-the Internal plan representation essential to planning within a
+the internal plan representation essential to planning within a
 situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
@@
 (Section~\ref{sec:europa:search}) further detail the important details
-associated with \eu for the reader. With these foundational elements
-out of the way, we transition to \rx in Section~\ref{sec:arch}, bring
-out architectural details in Section~\ref{sec:arch:trex}, deal with
-the execution cycle in Section~\ref{sec:arch:exec}, highlight how \rx
-deliberates in Section~\ref{sec:arch:europa} finally delving deeper
-into the key concept of synchronization in
-Section~\ref{sec:arch:synch}. Further details on deliberation within
-\rx (Section~\ref{sec:arch:plan}) and how planning and execution are
-interleaved are shown in Section~\ref{sec:arch:intertwine}. We offer
-results from our experiments at sea in Section~\ref{sec:results},
-offer some insights into the near-term future goals of our research in
+associated with \eu for the and relevant to how \rx works to the
+reader. With these foundational elements out of the way, we transition
+to \rx itself, in Section~\ref{sec:arch}, bring out architectural
+details in Section~\ref{sec:arch:trex}, deal with the execution cycle
+in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
+Section~\ref{sec:arch:europa} finally delving deeper into the key
+concept of synchronization in Section~\ref{sec:arch:synch}. Further
+details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
+how planning and execution are interleaved are shown in
+Section~\ref{sec:arch:intertwine}. We offer results from our
+experiments at sea in Section~\ref{sec:results}, offer some insights
+into the near-term future goals of our research in
 Section~\ref{sec:future} and conclude with
 Section~\ref{sec:conclusion}.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-22 trunk/concepts.tex
added figures missing in concepts.tex

@@
     adding new intervals (tokens) to the plan, or by merging with
-    existing intervals. % Token state lifecycle is examined in detailed
-    % in a separate section below.
+    existing intervals.
 
@@
   describe their state and behavior throughout a plan. There are some
-  object types that are commonly used in domain descriptions, that \eu
+  object types that are used so often in domain descriptions that \eu
   provides a built-in implementation for them:
@@
 
-\item[\textbf{Token State Model}] In \eus representation, goals are
+\item[\textbf{Token State Model}] In \eu's representation, goals are
   temporally scoped states that need to be achieved, or actions that
@@
   time, or implicitly when subgoals are created as a result of domain
-  compatibilities \kcomment{Define a compatibility}. For instance,
-  when the Shopping Agent must get to a location where a product is
+  rules. For instance, when the Shopping Agent must get to a location where a product is
   available before it is able to purchase it: in this case the
@@
   a specific location without which the purchase will not occur as
-  defined in the domain model.  Looking at a how a simple Shopping
+  defined in the domain model.  Looking at how a simple Shopping
   Agent goal would be addresed by \eu helps illustrate how Token
@@
   starts with a partial plan that places the Shopping agent at home at
-  time $0$ with no products in possession.  If a goal is posted for
-  the Agent to own a drill by time 10. \kcomment{Use Fig 4 and edit::
-    Add figure showing an AgentLocation timeline with an At(Home)
-    token with start=0 and end open, an empty AgentBag timeline, and a
-    Own(Drill) token with start=10 floating below both timelines}
+  time $0$ with no products in possession. Let's also assume that goals are posted for
+  the Agent to own a drill by time 10 and to be home by time 20 as depicted in Fig. \ref{fig:europapr1}
 
-  Note that the \texttt{Own(Drill)} token that represents the goal is
-  not immediately placed on the \texttt{AgentBag} timeline since \eu
-  must determine whether it can achieve that goal and how. When the
-  goal is posted and before any decisions are made, the
-  \texttt{Own(Drill)} token is created in an \texttt{INACTIVE}
+\begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-1.jpg}
+  \caption{\small Timelines representing initial state and goals for a Shopping Agent.}
+\label{fig:europapr1}
+\vskip+0.1cm
+\end{figure}
+
+  Note that the tokens \texttt{Own(Drill)}  and \texttt{At(Home)} that represent the goals are
+  not immediately placed on the \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu
+  must first determine whether it can achieve those goal and how. When the
+  goals are posted and before any decisions are made, the
+  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an \texttt{INACTIVE}
   state. \eu then considers the partial plan and evaluates two
@@
 tokens are merge-compatible if they are instances of the same
-predicate and that no intersections between corresponding variables
-are empty. The effect of merging is illustrated in Figure 3
-\kcomment{Which figure?}.\comment{Use previous fig and modify as
-  appropriately:: Add figure showing the effect of merging; all figs
-  on this section are conjoined. JB to sketch it out.}
+predicate and there are no empty intersections between corresponding token variables. 
+The effect of merging is illustrated in Fig. \ref{fig:europapr:merge}
 
-When a token is activated all compatibilities \kcomment{this should
-  have been defined earlier} associated with its type are evaluated,
+\begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-merge.jpg}
+  \caption{\small Timeline states after satisfying goal to be home by time 20 by merging with initial state}
+\label{fig:europapr:merge}
+\vskip+0.1cm
+\end{figure}
+
+The goal to be home by time 20 is satisfied by merging with the initial state that said that the Agent is home at time 0.
+The merge is possible because the tokens are of the same type \texttt{At}, and intersecting the domains for all of their variables
+doesn't yield an empty domain (which represents an inconsistency), in this case the token variables are the location, 
+which is \texttt{Home} in both cases, and the temporal variables (start, end, duration), which can be intersected without causing 
+an inconsistency. After the token representing the goal is merged onto the token representing the initial state, the resulting token
+represents not only the fact that the Agent must be \texttt{At(Home)} by time 0 but also \texttt{At(Home)} at time 20, this is reflected in 
+the new constraints on the token's start and end variables as depicted in Fig. \ref{fig:europaprmerge}.
+
+When a token is activated, all domain rules associated with its type are evaluated,
 which may lead to new subgoals being generated. In our example, there
-could be a compatibility stating that the Shopping Agent must be at
-the same location as the product it is purchasing. In such a case,
+could be a domain rule stating that if a  Shopping Agent doesn't already own a product,
+then it must \texttt{Buy} it before it can \texttt{Own} it. In such a case,
 activating the \texttt{Own(Drill)} token will cause a new
-\texttt{At(Drill.location)} to be created in an \texttt{INACTIVE}
-state, which in turn could spawn its own subgoals.
+\texttt{Buy(Drill)} to be created in an \texttt{INACTIVE}
+state, which in turn could spawn its own subgoals, like being at the location where the product
+can be purchased before being able to buy it, and so on, until all domain rules are satisfied.
 
-\comment{Use previous fig and modify as appropriately: Add figure
-  showing the effect of activating the token by putting it on the
-  AgentBag timeline, and a new At(Drill.location) token floating
-  below}
+\begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-activate.jpg}
+  \caption{\small Timeline states after satisfying goal to own drill by time 10 by activating corresponding token}
+\label{fig:europapr:activate}
+\vskip+0.1cm
+\end{figure}
 
-If neither activating nor merging work, a token may be
-\texttt{REJECTED}, this is possible only for goals posted by the end
+It is easy to see in Fig. \ref{fig:europapr:activate} that the decision to merge the goal to be \texttt{At(Home)} by time 20 
+with the initial state was not a good one since the Agent will have to get away from \texttt{Home} to be able to 
+buy the drill which is inconsistent with the constraints that resulted after the merge. As a result that decision will have 
+to be retracted at some point and the alternative to activate or merge with a different token will have to be considered.
+
+If, after looking at all possibilities, neither activating nor merging work, a token may be
+\texttt{REJECTED}. This is possible only for goals posted by the end
 user and specified as optional. Subgoals generated as a result of
-token activation cannot be rejected; if a subgoal cannot be satisfied
+token activation cannot be rejected. If a subgoal cannot be satisfied
 through activation or merging then it will cause the original
@@
 
-As can be seen from the above example, a token's state can go from
+As can be seen from the above example, during the search for a plan, a token's state can go from
 \texttt{INACTIVE} to \texttt{ACTIVE, MERGED} or \texttt{REJECTED}, as
-shown in figure below \kcomment{Which figure?}.
+shown in Fig. \ref{fig:europapr:states}.
 
-\comment{Maybe: Add figure showing possible state transitions. Need a FSM.}
+\begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-states.jpg}
+  \caption{\small Possible transitions for a Token State}
+\label{fig:europapr:states}
+\vskip+0.1cm
+\end{figure}
 
@@
 
-  \comment{Use previous fig and modify as appropriately (used as
-    above): Add figure showing internal and external relationships
-    from rules on action Go. Need a fig from JB}
+  \begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-rules.jpg}
+  \caption{\small Master and Slave tokens generated as a result of a Domain Rule}
+\label{fig:europapr:rules}
+\vskip+0.1cm
+\end{figure}
 
-  Figure xx \kcomment{which figure?} illustrates the entities and
+  Fig. \ref{fig:europapr:rules} illustrates the entities and
   relations involved in specifying such a rule on a \texttt{Go}
-  action. The token on which the rule applies is referred to
-  (\texttt{Go}) as the \emph{master}. Each \texttt{At} token required
-  by the master is referred to as a \emph{slave}. All variables are
-  indicated by name and their domains are expressed as intervals in
-  the case of temporal variables and as enumerations for the
-  remainder. Application of a rule on a token can thus cause slave
+  action. The token on which the rule applies  (\texttt{Go}) is referred to
+ as the \emph{master}. Each \texttt{At} and \texttt{Going} token required
+  by the master is referred to as a \emph{slave}. 
+  Application of a rule on a token can thus cause slave
   tokens, variables and constraints to be introduced.  The capability
   of a domain rule to cause a slave token to be created is a key
-  vehicle through which planning occurs. Semantically, this rule
+  vehicle through which planning occurs. Semantically, a rule
   imposes a requirement for supporting (slave) tokens to be in the
-  plan in order for the master to be valid.
+  plan in order for the master token to be valid.
 

### 2012-02-22 trunk/europa-basics.tex
added figures missing in concepts.tex

@@
 development of \rx has been targeted at surveying a number of
-oceanographic features which are dynamic and unpredictable
-spatio-temporally. Typically this requires our AUVs to balance the
-goals of coverage spatially while opportunistically following features
-of scientific interest and to do so while being operationally aware of
-its own limitations in terms of resources (\eg the battery state of
-charge) and overall proximity to other observational assets for
-obtaining scientific ground-truth. We are interested in
-representational frameworks that allow robots to pursue long-term
-objectives while choosing short-term gain and can gracefully deal with
-exogenous or endogenous change.
+oceanographic features which are dynamic and spatio-temporally
+unpredictable. Typically this requires our AUVs to balance the goals
+of coverage spatially while opportunistically following features of
+scientific interest and to do so while being aware of its own
+limitations in terms of resources (\eg the battery state of charge)
+and overall proximity to other observational assets for obtaining
+scientific ground-truth. We are interested in representational
+frameworks that allow robots to pursue long-term objectives while
+choosing short-term gain and can gracefully deal with exogenous or
+endogenous change.
 
 To enable this responsiveness to external observations, the agent has
-to be able to synthesize plans insitu and to re-plan. We use the
-temporal constraint-based planner \eu with a demonstrated legacy of
-having flown on NASA space missions \cite{mus98,jonsson00,bresina05,
-  barreiro09}. Our autonomy architecture brings three key innovations
-for AUV adaptation: the use of flexible plan representations,
-compositional control with the use of partitioned networks and a
-demonstrated capability at sea in novel ocean observation methods.
+to be able to synthesize plans insitu and to re-plan without human
+intervention. To do so we use a temporal constraint-based planner with
+a demonstrated legacy of having flown on NASA space missions. Our
+autonomy architecture brings three key innovations for AUV control:
+in-situ plan synthesis, the use of flexible plan representations with
+a sound theoretical basis and systematic compositional control with
+the use of partitioned networks doing so with a demonstrated
+capability at sea in novel ocean observation methods.
 
-The next few sections will cover \eu in a bottom-up fashion, going from the 
-generic theoretical frameworks that underlie \eu, to the specific representation, 
-modeling and algorithms implemented by it. First, a quick introduction to 
-Constraint Programming and Constraint-Based Planning is provided. 
-Next, the specific representation that \eu uses to perform Constraint-Based Planning is presented.
-After that comes an overview of \eu's architecture and how it can be embedded in a planning application
-such as \rx. Finally,  \eu's modeling, inference and search mechanisms are covered in 
-detail so that the reader can understand how \rx can perform deliberative planning and react to
+The next few sections will cover \eu in a bottom-up fashion, going
+from the generic theoretical framework that underlies \eu, to the
+specific representation, modeling and algorithms implemented by it.  A
+brief introduction to Constraint Programming and Constraint-Based
+Planning is provided followed by the specific representation that \eu
+uses to undertake Constraint-Based Planning.  Subsequently, we provide
+an overview of \eus architecture and how it can be embedded in a
+planning application such as \rx. Finally, \eus modeling, inference
+and search mechanisms are covered in detail so that the reader can
+understand how \rx can perform deliberative planning and react to
 execution events by embedding and customizing \eu.
@@
 
-\paragraph \textit{Planning}: In the Shopping Agent Problem
-\cite{russelnorvig} an agent needs to purchase a set of products (milk,
-drill, etc) that are available at specific locations (supermarket,
-hardware store, etc), the agent is subject to temporal (must complete
-tasks by specific deadlines) and resource (fuel, carrying capacity, etc)
-constraints. The agent needs to figure out what actions need to be
-performed to find and acquire the required items, as well as when to
-perform each of those actions. Fig. \ref{fig:shopping-1} shows a
-solution produced by \eu for such a problem instance.
+\paragraph \textit{Planning}: The Shopping Agent Problem
+\cite{russelnorvig} is a variation of the problem with which we started this chapter.
+In this problem, an agent needs to purchase a set of products
+(milk, drill, etc) that are available at specific locations
+(supermarket, hardware store, etc) and the agent is subject to
+temporal (must complete tasks by specific deadlines) and resource
+(fuel, carrying capacity, etc) constraints. The agent needs to figure
+out what actions need to be performed to find and acquire the required
+items, as well as when to perform each of those
+actions. Fig. \ref{fig:shopping-1} shows a solution produced by \eu
+for such a problem instance.
 
@@
   effects, dependencies at the action level) is not explicitly
-  maintained by the CSP planner, the search algorithms are deprived of
+  maintained by the \textsf{CP} planner, the search algorithms are deprived of
   critical information to make better decisions. If that structure is
@@
   still hard to write search algorithms as any planning-specific
-  insight has to be translated into the CSP representation of
+  insight has to be translated into the \textsf{CP} representation of
   variables and constraints.
@@
 
-These drawbacks have been addressed by a small number of planning
-frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
-\texttt{IxTet} \cite{ghallab87,lemai04}, and \texttt{CAIP} (Constraint-Based
-Interval and Attribute Planning)\cite{mus94,frank2003}.  These
-frameworks take advantage of a CSP representation to deal with time
-and resources, at the same time maintaining an explicit representation
-of the elements relevant to planning (action choices, causal
-relationships, etc). This combination results in a planning approach
-that can deal with the type of constraints found in real-life
-problems, while allowing for intuitive problem descriptions and a
-natural way to express search algorithms and heuristics. \eu is an
-implementation of the \texttt{CAIP} framework. In \texttt{CAIP}, planning problems are
-specified in terms of:
+These drawbacks have been addressed by a number of planning frameworks
+like \texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
+\cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
+and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
+advantage of a \textsf{CP} representation to deal with time and resources, at
+the same time maintaining an explicit representation of the elements
+relevant to planning (action choices, causal relationships, etc). This
+combination results in a planning approach that can deal with the type
+of constraints found in real-world problems while allowing for
+intuitive problem descriptions and a natural way to express search
+algorithms and heuristics. \eu is one such implementation of the
+\texttt{CAIP} framework, in which planning problems are specified in
+terms of:
 

### 2012-02-22 trunk/europa.tex
added resource envelopes figure.

@@
 \label{eu:model}
-%\caption{Specifying a problem instance for the Shopping Agent in NDDL}
 \end{verbatim}
@@
 
-\comment{figure :provide snapshot of resource envelopes generated by europa}
+\begin{figure}
+\centering
+\includegraphics[scale=0.35]{figs/europa-resource-envelopes.png}
+\caption{\small Resource Envelopes computed by  \eu}
+\label{fig:resenvelopes}
+\vskip+0.1cm
+\end{figure}
 
@@
   tokens can be generated by explicitly posted goals, or when a token
-  is activated, rules may fire that create inactive slave tokens
-  \comment{reference: point to more detailed decription above in Rules
-    Engine section}.  Open Conditions can be resolved in three ways:
+  is activated, rules may fire that create inactive slave tokens as described in section
+  \comment{point to EUROPA Plan Representation section}.  Open Conditions can be resolved in three ways:
 
@@
 
-\begin{verbatim}
-    public void solve(ConstraintEngine ce, int maxIter)
-    {
-    	curIter=0;
-
-    	while ((ce.getViolations() > 0) && (curIter < maxIter)) {
-	    Variable queenToMove = getQueenWithMaxViolation(ce);
-    	     int curPos = queenToMove.getCurrentValue();
-
-            SortedSet<int> moves = getMoves(ce,queenToMove,curPos);             
-  	   boolean moved = false;
-            for (int newPos : moves) {
-            	moved = tryMove(queenToMove,curPos,newPos);
-            	if (moved)
-            	    break;
+\begin{algorithm}[H]
+  \KwIn{ConstraintEngine $ce$}
+  \KwIn{int $maxIter$}
+  \Begin{
+    int $curIter \leftarrow 0$ \;
+    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
+        maxIter \right)$ }{
+      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
+      \;
+      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
+      curPos)$ \;
+      bool $moved \leftarrow $ \texttt{false} \;
+      \ForEach{int $newPos$ of $moves$}{
+        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
+        \lIf{$moved$}{ \textbf{break} \; }
       }
-
-            if (!moved) 
-            	forceMove(queenToMove,curPos,moves.first());
-
-            checkSolution(); // See if we have a new best solution            
-            curIteration_++;
+      \lIf{$! moved$}{
+        $forceMove(queenToMove, curPos, moves.first())$ \;
+      }
+      $checkSolution()$ \tcp*{See if we have a new best solution}
+      $curIter$\texttt{++} \;
     }
   }
-\end{verbatim}
+  \caption{$solveNQueens(ce, maxIter)$}
+\end{algorithm}
 

### 2012-02-22 trunk/europa.tex
skipped adding europa's AC-3 algorithm since it doesn't seem to be possible to add it at a level of detail consistent with the rest of the section. adding standard AC-3 algorithm doesn't add value.

@@
 
-\kcomment{add europa's AC-3 algorithm here}
+%\kcomment{add europa's AC-3 algorithm here}
 
@@
 specialized Constraint and Propagator classes in C++, however, \eu
-provides a built-in with over 25 arithmetic, temporal, set and
+provides a built-in Constraint Library with over 25 arithmetic, temporal, set and
 resource constraint types so that a large number of problem domains

### 2012-02-22 arch.tex
updates to intro etc
Corrections from William comments and refined shopping pictures to clarify internal/external timelines

@@
-\section{Planning and execution in the \rx architecture}
+\section{Planning and execution with \rx}
 \label{sec:arch}
@@
 % \end{figure}
-\fcomment{I removed this figure as the next one is mucjh more
-  detailed and vehiculate the same idea}
+% \fcomment{I removed this figure as the next one is mucjh more
+%   detailed and vehiculate the same idea}
 
@@
 variables; it does so by posting the special observation
-\textsf{Failed}. By so doing we allow for graceful degradation where
+\textsf{Failed}. This special state is a special predicate which is
+acceptable for all the {\em internal} and {\em external} timelines of
+a \eu reactor.  By so doing we allow for graceful degradation where
 other reactors can readapt their plans on receipt of this observation.
@@
 
-\subsubsection{Deliberation: Planning for future state evolution}
+\subsubsection{Deliberation: Planning for  Evolving State}
 \label{sec:arch:plan}
@@
 
-Consider for example the Shopping Agent model where the \texttt{Agent}
-is an {\em external} timeline. As the reactor adds a goal to
+Consider for example the Shopping Agent model where the timeline 
+\texttt{Agent.position} is {\em external} while \texttt{Agent} is
+implemented a an {\em internal} timeline. As the reactor adds a goal to
 \texttt{Own} milk and the current {\em external} timeline indicates
@@
   \centering
-  \subfloat[\small A plan produced before synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.45\columnwidth]{figs/shoping_exec_t0}}
+  \subfloat[\small A plan produced before
+  synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}} \\
   \subfloat[\small Result of synchronization after the {\em Going}
   observation was received by the reactor with the plan showed in
-    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=0.48\columnwidth]{figs/shoping_exec_t1}}
+    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
     \caption{\small Shopping example: Tokens in red indicate

### 2012-02-22 intro.tex
removed references to probabilistic state estimation as neither t-rex or europa are formally handling that.  We may have demonstrated that a HMM can be modeled and executed within nddl but still neither of the 2 frameworks where aware of it.

@@
 geneticists and roboticists on the \can (Controlled Agile Novel
-Observation Network) \cite{canon} has driven autonomous system
+Observation Network) program \cite{canon} has driven autonomous system
 exploration and control. Fundamental questions related to sampling
@@
 are not 'adaptive' enough. Despite vigorous defense by Brooks early on
-\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
-require a model if not of the world then of the \emph{physics} of the
-vehicle when it interacts with the world.
+\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep},
+autonomous systems require a model if not of the world then of the 
+\emph{physics} of the vehicle when it interacts with the world.
 
@@
 was targeting the performance of in-situ deliberation. Terrestrial
-robotic platforms were well-known to frequently stop and ``plan''. In
+robotic platforms were well-known to frequently stop and ``plan''; 
+assuming in the meantime that the world will remain static. In
 the marine robotics world, such a situation would simply not be
@@
 (\texttt{T-REX}) an on-board adaptive control system that integrates
-AI based planning and probabilistic state estimation in a hybrid
-executive \cite{mcgann08a,mcgann08b,py10}.  Probabilistic state
-estimation integrates a number of science observations to produce a
-likelihood that the vehicle sensors perceive a feature of
-interest. Onboard planning and execution enables adaptation of
-navigation and instrument control based on the probability of having
-detected such a phenomenon. It further enables goal-directed
+AI based planning and %\sout{probabilistic} 
+state estimation in a hybrid executive\cite{mcgann08a,mcgann08b,py10}. 
+%  \sout{Probabilistic state estimation integrates a number of science observations to produce a
+% likelihood that the vehicle sensors perceive a feature of
+% interest}.\fcomment{\rx {\em does not} integrate probabilistic state
+%   estimation! ... The HMM estimation was coded on top of \eu model but
+%   neither \eu or \rx had explicit management of it} 
+State estimation allow the system to keep track of the world evolution
+as time advances based on a formal model distributed through the architecture.
+Onboard planning and execution enables adaptation of
+navigation and instrument control based on the state identified during
+state estimation.
+% \sout{the probability of having detected such a phenomenon} 
+It further enables goal-directed
 commanding within the context of projected mission state and allows
@@
 is deployed on a European planetary rover testbed
-\cite{goac11}. Probabilistic state estimation is discussed elsewhere
-\cite{mcgann08d} and out of scope for this chapter. Deliberation and
-reaction are integrated systematically over different temporal and
+\cite{goac11}. While it is possible to integrate probabilistic
+estimation on top of this framework \cite{mcgann08d}, such aspect is
+out of scope of this chapter.
+% Probabilistic state estimation through modeling within 
+% this framework can be done is discussed elsewhere
+% \cite{mcgann08d} and out of scope for this chapter. 
+Deliberation and reaction are integrated systematically over different temporal and
 functional scopes within a single agent and a single model that covers

### 2012-02-22 trunk/intro.tex
Extra updates

--- 2012-02-07 trunk/intro.tex
@@
 is deployed on a European planetary rover testbed
-\cite{goac11}. Probabilistic state estimation is discussed elsewhere
-\cite{mcgann08d} and out of scope for this chapter. Deliberation and
-reaction are integrated systematically over different temporal and
+\cite{goac11}. While it is possible to integrate probabilistic
+estimation on top of this framework \cite{mcgann08d}, such
+representation is not explicitly handled by the framework and is
+out of scope of this chapter.
+% Probabilistic state estimation through modeling within 
+% this framework can be done is discussed elsewhere
+% \cite{mcgann08d} and out of scope for this chapter. 
+Deliberation and reaction are integrated systematically over different temporal and
 functional scopes within a single agent and a single model that covers
@@
 This chapter is organized as follows. We start with some foundational
-concepts in our Planning framework in Section
-\ref{sec:concepts}. \comment{add others} We conclude with Section
-\ref{sec:conclusion}.
+concepts in our Planning framework in Section~\ref{sec:concepts} which
+includes a brief overview of AI Planning in
+Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
+deeper into the \eu framework itself, the basis of all deliberation
+within \rx. A key portion is Section~\ref{sec:europa:pr} which details
+the internal plan representation essential to planning within a
+situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
+Inference (Section~\ref{sec:europa:inference}) and Search
+(Section~\ref{sec:europa:search}) further detail the important details
+associated with \eu for the and relevant to how \rx works to the
+reader. With these foundational elements out of the way, we transition
+to \rx itself, in Section~\ref{sec:arch}, bring out architectural
+details in Section~\ref{sec:arch:trex}, deal with the execution cycle
+in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
+Section~\ref{sec:arch:europa} finally delving deeper into the key
+concept of synchronization in Section~\ref{sec:arch:synch}. Further
+details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
+how planning and execution are interleaved are shown in
+Section~\ref{sec:arch:intertwine}. We offer results from our
+experiments at sea in Section~\ref{sec:results}, offer some insights
+into the near-term future goals of our research in
+Section~\ref{sec:future} and conclude with
+Section~\ref{sec:conclusion}.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-23 trunk/arch.tex
Rewrote some parts of related.tex with comments

@@
-\section{Planning and execution in the \rx architecture}
+\section{Planning and execution with \rx}
 \label{sec:arch}
-
-% {\em\tiny This section introduces shortly T-REX architecture but move
-%   quickly the focus on a single reactor and gross concepts on the
-%   architecture:
-%   \begin{itemize}
-%   \item For t-trex the whole decision and control problem is reduced
-%     around the state variables (or timeline construct). 
-%   \item This reduce the execution tracking problem for each reactor on a state
-%     identification problem (deduce internal state from the external
-%     state), and the control problem to goal posting which will trigger
-%     deliberation on the owner of the corresponding timeline. 
-%   \item the architecture itself see each reactor as a black box and
-%     therefore each reactor can implement its own mechanism in order to
-%     resolve both synchronization and deliberation. Still we provide a
-%     reactor based on the europa framework that leverage the automated
-%     planning capabilities in order to do model based planning and
-%     execution with a rich representation of resources and time.
-%   \end{itemize}}
 
@@
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
-  \caption{\small An example of a \rx reactor hierarchy which is used
-    to determine a synchronization order for propagation of values in
-    \emph{external} state variables.}
-  \label{fig:reactor-DAG}
-\end{figure}
+% \begin{figure}[!htbp]
+%   \centering
+%   \includegraphics[scale=0.35]{figs/DAG-hierarchy}
+%   \caption{\small An example of a \rx reactor hierarchy which is used
+%     to determine a synchronization order for propagation of values in
+%     \emph{external} state variables.}
+%   \label{fig:reactor-DAG}
+% \end{figure}
+% \fcomment{I removed this figure as the next one is mucjh more
+%   detailed and vehiculate the same idea}
 
@@
 level this execution latency is used to define the {\em dispatch
-  window} of the same state variable.% : \comment{Needed to say that
-  % early on but the flow of this paragraph is not very tight}
-%\kcomment{Looks fine to me unless you want to clarify further.}
+  window} of the same state variable.
 
@@
 reactor needs to replan due to an unexpected event, it will recall all
-requests it produced that have not been executed yet.
-% \kcomment{replace
-%   ``observed'' with ``not incorporated by the owner reactor''?}
-% yet
- This recall is sent to the reactor that received the goal(s) to
-remove it/them from its current set of objectives.
+requests produced that have yet to be executed.  This recall is also
+sent to the reactor(s) that received the goal(s) to remove it/them
+from its current set of objectives.
 
@@
 
-Both these 
-% deliberative \kcomment{``deliberative??''; something else
-%   or remove altogether perhaps?} 
-functions have different scope and temporal constraints. 
-The \texttt{synchronize} call is considered as atomic and will 
-focus only on state inference for the current tick by
-identifying the current value of {\em internal} state variables from
-the latest updates on {\em external} state variables. This process
-propagates up in the reactor dependency graph at every tick to ensure
-that all reactors are aware of the current state of the
-world. \texttt{step} relates to a single decision step of the planning
-process focused on the future evolution of state variables; it can
-take multiple steps spanning multiple ticks for a reactor to produce a
-plan that is both complete and valid. How the reactor identifies that
-its plan is complete and valid is its responsibility and depends on
-the way each reactor resolves its plan at each {\em step}. The general
-idea is that the reactor is able to find a valid strategy in order to
-satisfy as many of its {\em internal} goals as possible. For \eu, this
-means that the solver is unable to identify any further flaws to be
-resolved in the plan.
+Both these functions have different scope and temporal constraints.
+The \texttt{synchronize} call is considered as atomic and will focus
+only on state inference for the current tick by identifying the
+current value of {\em internal} state variables from the latest
+updates on {\em external} state variables. This process propagates up
+in the reactor dependency graph at every tick to ensure that all
+reactors are aware of the current state of the world. \texttt{step}
+relates to a single decision step of the planning process focused on
+the future evolution of state variables; it can take multiple steps
+spanning multiple ticks for a reactor to produce a plan that is both
+complete and valid. How the reactor identifies that its plan is
+complete and valid is its responsibility and depends on the way each
+reactor resolves its plan at each {\em step}. The general idea is that
+the reactor is able to find a valid strategy in order to satisfy as
+many of its {\em internal} goals as possible. For \eu, this means that
+the solver is unable to identify any further flaws to be resolved in
+the plan.
 
@@
   external} state updates are compatible with this plan. If so,
-synchronization propagates these events in the plan and can uses it as
-an aid in resolving its {\em current state flaw} by using the plan
-information on the exepected {\em internal} state evolution nearby the
-execution frontier. 
-% \kcomment{``use it
-%   as an aid to identify'' what though?}
+synchronization propagates these events in the plan and can use it as
+an aid in resolving its {\em current state flaw}. It does so by using
+the plan information on the exepected {\em internal} state evolution
+near the execution frontier.
 
@@
 plan inconsistency. This results in the retraction of decisions made
-during previous {\em deliberation} steps. % in order to
-% allow {\em synchronization} to identify its {\em internal} state.
-Such a relaxation can be viewed as a backtracking % on the plan search
-% triggered exogenously by the world evolution. It is the way our
-allowing the reactor to identify that its current plan has failed to
-execute triggering further deliberations steps to find an alternate
-strategy.
+during previous {\em deliberation} steps.  Such a relaxation can be
+viewed as a backtracking allowing the reactor to identify that its
+current plan has failed to execute triggering further deliberations
+steps to find an alternate strategy.
 
@@
 In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
-new type of flaw which requires to identify the value of all the {\em
-  internal} state variables for the current tick.
-% forces identification \kcomment{identification
-%   of what? Probably need a little more explanation here.} of internal
-% state of the reactor for the current tick. 
-This flaw can be defined as:
+new type of flaw which requires identification of all values of {\em
+  internal} state variables for the current tick. This flaw can be
+defined as:
 
@@
 
-As the simple example above shows, synchronization involves a
-deliberation process with the risk that the search for a solution is
-not a straightforward sequence of staright-forward decisions where 
-the resolution can be easily identified by the solver as any other
-option would leafd immediately to an inonsistency in the plan.
-% \kcomment{has a unit decision been explained in Europa?}. 
-Since such a process is required by the architecture at \emph{every}
-tick for \emph{every} reactor the model needs to be designed carefully.
+As the example above shows, synchronization involves a deliberation
+process. The risk is such that the search for a solution is not a
+simple sequence of decisions where the resolution can be easily
+identified by the solver as any other option would lead immediately to
+n plan inconsistency. Since such a process is required by the \rx at
+\emph{every} tick for \emph{every} reactor the model needs to be
+designed carefully.
 
@@
 problem on only those tokens that are within a ``reasonable'' future.
-This filtering window complements the \rx {\em dispatch
-  window} from Definition \ref{def:dispatch} which filters out at the
-agent level those goals that are before the reactor execution latency
-or too far into the future. The planning window reflect this filtering
-of the agent by applying to tokens that are generated internally by the
-reactor during its deliberation or the {\em goals} for which he
-planner decided to postpone after the planning window -- probably
-due to other tasks that needed completion beforehand.
-% The planning window inside of a reactor
-% refines this filtering further, as tokens can be excluded dynamically
-% from deliberation as their insertion in the plan and scheduling with
-% other tokens generated, can reduce their temporal scope outside of the
-% planning scope \kcomment{ this last sentence should be reworked}. 
-This allow the planner to focus on its planning horizon and report
-decision on flawss thgat are beyond this scope on later ticks. As, 
-time advance, these future flaws will eventuallty be part of the
-planning window of the reactor and will be then resolved along with
-potential new goals received from the agent.
-% This helps the deliberation solver reduce the number of tokens to be
-% evaluated reducing deliberation cost and increasing performance. As
-% time advances, flaws that were initially beyond this temporal horizon
-% will become active which a reactor can evaluate.
- This results in an apparent \emph{continuous planning} of the reactor. When there are no
-more flaws present in the current planning scope a partial plan
-solution is found. This has two effects:
+This filtering window complements the \rx {\em dispatch window} from
+Definition \ref{def:dispatch} which filters out at the agent level
+those goals that are before the reactor execution latency or too far
+into the future.  The planning window refines this filtering further,
+as tokens can be excluded dynamically from deliberation as their
+insertion in the plan and scheduling with other generated tokens can
+alter their temporal scope to be outside of the planning window.  This
+allows the planner to focus on its planning horizon and report
+decision on flaws that are beyond this scope. As, time advances, these
+potential future flaws will eventually be part of the planning window
+of the reactor and will be then resolved along with potentially new
+goals.  This in turn, results in an apparent \emph{continuous
+  planning} of the reactor. When there are no more flaws present in
+the current planning scope a partial plan solution is found. This has
+two effects:
 
@@
 managing this state variable internally.
-
+x1
 \begin{figure}[!htb]
   \centering
-  \subfloat[\small A plan produced before synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.45\columnwidth]{figs/shoping_exec_t0}}
+  \subfloat[\small A plan produced before
+  synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}} \\
   \subfloat[\small Result of synchronization after the {\em Going}
   observation was received by the reactor with the plan showed in
-    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=0.48\columnwidth]{figs/shoping_exec_t1}}
+    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
     \caption{\small Shopping example: Tokens in red indicate

### 2012-02-23 trunk/related.tex
Rewrote some parts of related.tex with comments

@@
 
-There are costs associated with such inference based control are also
+There are costs associated with such inference based control which are 
 not inconsequential. Foremost among them is the steep learning curve
@@
 modeler to be exposed to the internal representation of the planner
-and how it performs search. THe level of expertise is often well above
+and how it performs search. The level of expertise is often well above
 what would be expected of a typical well-rounded Computer Science
@@
 the field of robotics. 
+\fcomment{Though it woulfd  be usefull to emphasize this ...}
+The cost of modeling put aduring the intial design, for example,
+should be balanced against the cost of redesigning
+a new mission script eate every new deployment in more ``classical''
+script-based controllers.
 
@@
 clearly articulates how physical manifestation of robots could be
-controlled for task planning and execution. The very first planner
+controlled by task planning and execution. The very first planner
 \cite{green69} was an exercise in computational theorem proving with
@@
 efforts were cognizant of if absent with, dealing with time and
-resources for such experiments.
+resources for such experiments.\fcomment{This sentence does not make
+  sence to me}
 
@@
 control structure formally defined as a composition of coordinated
-control loops with a recurring sense-plan-act (SPA) cycle with
-deliberation at its core. \texttt{T-REX} was designed after carefully
+control loops where sensing, planning and acting are resulting from 
+concurrent deliberating tasks within a formal framework. \fcomment{Got
+  rid of sense-plan-act reference here. \rx plan/sense while sensing/planning.}
+\texttt{T-REX} was designed after carefully
 evaluating lessons learned from \texttt{IDEA} and owes substantially
@@
 % missing the objective of a versatile architecture for robots.''
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-23 europa-basics.tex
Made sure that paragraph tittles are correctly displayed in italic

@@
 
-\paragraph \textit{Constraint Satisfaction}: A canonical problem in
+\paragraph{} {\em Constraint Satisfaction}: A canonical problem in
 dealing with constraints is the $N$-Queens problem in which chess queens
@@
 
-\paragraph \textit{Scheduling}: In the Resource Constrained Project
+\paragraph{} \textit{Scheduling}: In the Resource Constrained Project
 Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, a project
@@
 
-\paragraph \textit{Planning}: In the Shopping Agent Problem
-\cite{russelnorvig} an agent needs to purchase a set of products
+\paragraph{} \textit{Planning}: The Shopping Agent Problem
+\cite{russelnorvig} is a variation of the problem with which we started this chapter.
+In this problem, an agent needs to purchase a set of products
 (milk, drill, etc) that are available at specific locations
@@
 and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
-advantage of a CSP representation to deal with time and resources, at
+advantage of a \textsf{CP} representation to deal with time and resources, at
 the same time maintaining an explicit representation of the elements

### 2012-02-23 related.tex
heading changes based on Minh and discussion; also noted FP comments

@@
 
-There are costs associated with such inference based control are also
+There are costs associated with such inference based control which are
 not inconsequential. Foremost among them is the steep learning curve
@@
 modeler to be exposed to the internal representation of the planner
-and how it performs search. THe level of expertise is often well above
+and how it performs search. The level of expertise is often well above
 what would be expected of a typical well-rounded Computer Science
-graduate student.
+graduate student. Yet in our experience, the cost of model design in
+such systems, is well balanced against the cost of re-engineering new
+mission scripts for new deployment in more ``classical'' script-based
+controllers.
 
@@
 \texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
-\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). Early
-efforts were cognizant of if absent with, dealing with time and
-resources for such experiments.
+\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). % Early
+% efforts were cognizant of if absent with, dealing with time and
+% resources for such experiments.
 
@@
 
-Our system, \texttt{T-REX}, the focus of this chapter, is an agent
-control structure formally defined as a composition of coordinated
-control loops with a recurring sense-plan-act (SPA) cycle with
-deliberation at its core. \texttt{T-REX} was designed after carefully
-evaluating lessons learned from \texttt{IDEA} and owes substantially
-to the notion of partitioned problem solving from that effort. However
-this partitioning is systematic and methodologically principled where
-we manage the information flow within the partitioned structure to
-ensure consistency in order to direct the flow of goals and
+Our system, \rx, the focus of this chapter, is an agent control
+structure formally defined as a composition of coordinated control
+loops where sensing, planning and acting are resulting from concurrent
+deliberating tasks within a formal framework.  \rx was designed after
+carefully evaluating lessons learned from \texttt{IDEA} to which it
+owes substantially the notion of partitioned problem solving. However
+partitioning in \rx is systematic and methodologically principled
+where we manage the information flow within the partitioned structure
+to ensure consistency in order to direct the flow of goals and
 observations in a timely manner. The resulting control structure
@@
 A recent variation of the key idea of such controller composition
-derived from \texttt{T-REX} is \texttt{ROAR} \cite{degroote11} where
-lower-level functional modules that control specific sensors are
-organized within a graph hierarchy. One reason they cite such a
-structure is essential is in dealing robustly to failure; an
-off-nominal condition will allow rapid graph traversal to identify
-alternative ways in which sensing tasks can be executed and therefore
-aiding in responsiveness. However, it is unclear whether to date this
-system actually deliberates and whether it is actually instantiated on
-a real platform for control.
+derived from \rx is \texttt{ROAR} \cite{degroote11} where lower-level
+functional modules that control specific sensors are organized within
+a graph hierarchy. One reason they cite such a structure is essential
+is in dealing robustly to failure; an off-nominal condition will allow
+rapid graph traversal to identify alternative ways in which sensing
+tasks can be executed and therefore aiding in responsiveness. However,
+it is unclear whether to date this system actually deliberates and
+whether it is actually instantiated on a real platform for control.
 

### 2012-02-23 trunk/europa-basics.tex
heading changes based on Minh and discussion; also noted FP comments

@@
 
-\subsection{Automated Planning and \eu}
+\subsection{Automated Planning}
 \label{sec:planningfound}
@@
 
-\paragraph \textit{Planning}: The Shopping Agent Problem
-\cite{russelnorvig} is a variation of the problem with which we started this chapter.
-In this problem, an agent needs to purchase a set of products
-(milk, drill, etc) that are available at specific locations
-(supermarket, hardware store, etc) and the agent is subject to
+\paragraph{} \textit{Planning}: The Shopping Agent Problem
+\cite{russelnorvig} is a variation of the problem with which we
+started this chapter.  In this problem, an agent needs to purchase a
+set of products (milk, drill, etc) that are available at specific
+locations (supermarket, hardware store, etc). The agent is subject to
 temporal (must complete tasks by specific deadlines) and resource
-(fuel, carrying capacity, etc) constraints. The agent needs to figure
-out what actions need to be performed to find and acquire the required
-items, as well as when to perform each of those
-actions. Fig. \ref{fig:shopping-1} shows a solution produced by \eu
-for such a problem instance.
+(fuel, carrying capacity, etc) constraints and it needs to synthesize
+actions that need to be performed to find and acquire the required
+items, as well as when to perform each of those actions.
+Fig. \ref{fig:shopping-1} shows a solution produced by \eu for such a
+problem instance.
 

### 2012-02-23 intro.tex
heading changes based on Minh and discussion; also noted FP comments

@@
 We have developed, tested and deployed the Teleo-Reactive EXecutive
-(\texttt{T-REX}) an on-board adaptive control system that integrates
-AI based planning and %\sout{probabilistic} 
-state estimation in a hybrid executive\cite{mcgann08a,mcgann08b,py10}. 
-%  \sout{Probabilistic state estimation integrates a number of science observations to produce a
-% likelihood that the vehicle sensors perceive a feature of
-% interest}.\fcomment{\rx {\em does not} integrate probabilistic state
-%   estimation! ... The HMM estimation was coded on top of \eu model but
-%   neither \eu or \rx had explicit management of it} 
-State estimation allow the system to keep track of the world evolution
-as time advances based on a formal model distributed through the architecture.
-Onboard planning and execution enables adaptation of
-navigation and instrument control based on the state identified during
-state estimation.
-% \sout{the probability of having detected such a phenomenon} 
-It further enables goal-directed
-commanding within the context of projected mission state and allows
-for replanning for off-nominal situations and opportunistic science
-events. The framework in addition to being used on an AUV, is general
-enough be used for controlling a personal robot \cite{mcgann2009} and
-is deployed on a European planetary rover testbed
-\cite{goac11}. While it is possible to integrate probabilistic
-estimation on top of this framework \cite{mcgann08d}, such aspect is
-out of scope of this chapter.
-% Probabilistic state estimation through modeling within 
-% this framework can be done is discussed elsewhere
-% \cite{mcgann08d} and out of scope for this chapter. 
-Deliberation and reaction are integrated systematically over different temporal and
+(\rx) an on-board adaptive control system that integrates AI based
+planning and state estimation in a hybrid
+executive\cite{mcgann08a,mcgann08b,py10}.  State estimation allows the
+system to keep track of world evolution as time advances based on a
+formal model distributed through the architecture.  Onboard planning
+and execution enables adaptation of navigation and instrument control
+based on the state identified during state estimation.  It further
+enables goal-directed commanding within the context of projected
+mission state and allows for replanning for off-nominal situations and
+opportunistic science events. The framework in addition to being used
+on an AUV, is general enough be used for controlling a personal robot
+\cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed on a European
+planetary rover testbed \cite{goac11}. While it is possible to
+integrate probabilistic estimation on top of this framework
+\cite{mcgann08d}, such representation is not explicitly handled by the
+framework and is out of scope of this chapter.  Deliberation and
+reaction are integrated systematically over different temporal and
 functional scopes within a single agent and a single model that covers
@@
 instrument control, and detection of unstructured and poorly
-understood phenomena. \rx is deployed on MBARI's Dorado AUV shown in
-Fig. \ref{fig:auv-fig}, which to the best of our knowledge is the only
-operational marine platform anywhere being used for routine scientific
-surveys with onboard plan synthesis.
+understood phenomena. \rx is deployed on MBARI's \emph{Dorado} AUV
+shown in Fig. \ref{fig:auv-fig}, which to the best of our knowledge is
+the only operational marine platform anywhere being used for routine
+scientific surveys with onboard plan synthesis.
 

### 2012-02-23 concepts.tex
minor typos and jpg to PDF converstion

@@
-\section{\eu Basics}
+\section{The \eu Planner }
 \label{sec:basics}
 
-\paragraph {A brief note on \eu history and notation} The earliest
-origin of what we will call \eu in this chapter, is from the
-\texttt{HSTS} planner \cite{mus94} which was used for a number of
-demonstrations for space telescopes as noted in Section
+\paragraph {A brief note on \eu history and notation} The origin of
+what we will call \eu in this chapter, is derived from the
+\texttt{HSTS} planner \cite{mus94} used on a number of ground-based
+demonstrations for space telescope scheduling as noted in Section
 \ref{sec:rabeyond}. \texttt{HSTS} was also the
-\texttt{LISP}-based\footnote{This was the first (and only to our
-  knowledge) software ever to been written in \texttt{LISP} to be
-  flown in space.} planner that was flown as part of the Remote Agent
-\texttt{RAX} experiment. Concurrent to \texttt{RAX}, a C++ version of
-the planner was implemented at which went on to be used for the
-\texttt{MAPGEN} \cite{bresina05} system and continues to be used to
-date, in the mission-critical uplink process for the command/control
-of the Mars Exploration Rovers (\texttt{MER}) mission. A vastly
-re-factored, higher performance open source version \cite{europapso}
-was named \eut and is the subject of this chapter and used onboard
-MBARI's Dorado platform. Note however, in using the name \eu, we refer
-to the \emph{fundamental ideas as well as the planning paradigm} even
-if there exists a specific implementation.
+\texttt{LISP}-based\footnote{This was the first (and to our knowledge
+  only) software ever to been written in \texttt{LISP} to be flown in
+  space.} planner that was flown as part of the Remote Agent
+\texttt{RAX} experiment. Concurrent to \texttt{RAX} deployment, a C++
+version of the planner was implemented which went on to be used for
+the \texttt{MAPGEN} \cite{bresina05} system and continues to be used
+to date, in the mission-critical uplink process for the
+command/control of the Mars Exploration Rovers (\texttt{MER})
+mission. A vastly re-factored, higher performance open source version
+\cite{europapso} was named \eut and is the subject of this chapter and
+used by \rxe.  Note however, in using the name \eu, we refer to the
+\emph{fundamental ideas as well as the planning paradigm} even if
+there exists a specific implementation.
 
@@
 \eu uses a \emph{domain model} written in a declarative language,
-together with initial conditions and goals also in that specified
-language, to construct a set of temporal relations that must be true
-at the start time. These models include assertions about the physics
-of the vehicle, i.e how it responds to external stimulus and
-internally driven goals. By propagating these relations forward using
-Simple Temporal Networks \cite{dechter91} and applying goal
-constraints, \eu can select a set of conditions that should be true in
-the future, where some of these conditions will correspond to actions
-the agent must take. The planner can backtrack and try another path
-during search if a goal cannot be reached while being capable of
+together with initial conditions and goals (also in that specified
+language), to construct a set of temporal relations that must be true
+at start time. These models include assertions about the physics of
+the vehicle, i.e how it responds to external stimulus and internally
+driven goals. By propagating these relations forward using Simple
+Temporal Networks \cite{dechter91} and applying goal constraints, \eu
+can select a set of conditions that should be true in the future,
+where some of these conditions will correspond to actions the agent
+must take. The planner can backtrack and try another path during the
+search process if a goal cannot be reached while being capable of
 discarding unachievable goals.
@@
 start time as a variable and contains an explicit network of bounded
-delay constraints between such variables.
+delay constraints between such
+variables. Fig. \ref{fig:flex-timelines} shows an example of a
+flexible partial plan.
 
-\begin{figure}[!t]
+\begin{figure}[!htb]
 \centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\includegraphics[scale=0.3]{figs/flexible-timelines.pdf}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
 \label{fig:flex-timelines}
-\vskip+0.1cm
 \end{figure}
@@
 interaction with the environment within which the executive can elect
-at run time the most appropriate one for the actual execution
-conditions. The fact that constraints are explicitly represented
-ensures that through constraint propagation the executive will respect
-global limits expressed in the plan (e.g., don't start a task until a
-certain condition has been satisfied but still satisfying some global
-deadline). Such flexibility is critical when dealing with dynamic
-ocean conditions where precise timing of a robotic action might be
-indeterminate. Further, the advantage of flexibility can be contrasted
-with the consequences of the intrinsic inflexibiliy of traditional
-command sequences. Because they are inflexible, sequences must
-necessarily be designed considering worst case scenarios.
+at run time, the most appropriate for actual execution. The fact that
+constraints are explicitly represented ensures that through constraint
+propagation the executive will respect global limits expressed in the
+plan (e.g., don't start a task until a certain condition has been
+satisfied but still satisfying some overall deadline). Such
+flexibility is critical when dealing with dynamic and uncertain
+conditions such as in the ocean, where precise timing of a robotic
+action might be indeterminate. Further, the advantage of flexibility
+can be contrasted with the consequences of the intrinsic inflexibility
+of traditional command sequences. Because they are inflexible,
+sequences must necessarily be designed considering worst case
+scenarios.
 
-\eu uses a \emph{state variable} representation to describe the
-evolution of state over time. The instantiated history of such state
-variable evolution over a temporal horizon we call \emph{timelines}
-and which represent a single thread in the execution of a concurrent
-system. At any given time each thread can execute a single procedure.
-Thus each timeline consists of a sequence of procedures which
-encapsulate and describe state evolution; we call these instantiated
-atomic entities \emph{tokens}.  A token therefore describes a
-procedure invocation, the state variables on which it can occur, the
-parameter values of the procedure, and the time values defining the
-interval. We allow encapsulation of uncertainty within these tokens
-with a range of start and end times and parameters, all of which are
-encoded as variables. A constraint solver in turn manipulates these
-variables defined in a \eu domain model. For example, consider
-a scientific need to take a water samples 100 meters from a hotspot
-while an AUV is performing a Yo-Yo in the water-column. Two samples
+\eu uses a \emph{state variable} representation
+\cite{mus94,jonsson00,mcgann09} to describe the evolution of state
+over time. The instantiated history of such state variable evolution
+over a temporal horizon we call \emph{timelines} and which represent a
+single thread in the execution of a concurrent system. At any given
+time each thread can execute a single procedure.  Thus each timeline
+consists of a sequence of procedures which encapsulate and describe
+state evolution; we call these instantiated atomic entities
+\emph{tokens}.  A token therefore describes a procedure invocation,
+the state variables on which it can occur, the parameter values of the
+procedure, and the time values defining the interval. We allow
+encapsulation of uncertainty within these tokens with a range of start
+and end times and parameters, all of which are encoded as variables. A
+constraint solver in turn manipulates these variables defined in a \eu
+domain model. For example, consider a scientific need to take water
+samples 100 meters from an environmental hotspot while an AUV is
+performing a Yo-Yo sawtooth pattern in the water-column. Two samples
 are needed if the feature's signal is above a threshold; one
@@
 
-\eu follows the representation outlined by the CAIP framework, it also
-adds a few elements to it that are necessary for an efficient
-implementation of constraint-based planning. Below is a brief summary
-of the main elements, followed by a detailed description for each one:
+\eu follows the representation outlined by the \texttt{CAIP} framework
+augmenting with elements that are necessary for an efficient
+implementation of constraint-based planning. Below we describe briefly
+the main elements, followed by a detailed description for each of
+these elements and how together they constitute a rich
+representational paradigm for inference in Sections
+\ref{sec:europa:modeling}, \ref{sec:europa:inference} and
+\ref{sec:europa:search}.
 
@@
 \item \textbf{Tokens} Temporally scoped entities that correspond to
-  Intervals in the CAIP framework, are used to represent state and
-  actions in time. A\textbf{Token State model} is defined to support
-  an efficient implementation of plan-space search algorithms.
+  Intervals in the \texttt{CAIP} framework, are used to represent
+  state and actions in time. A\textbf{Token State model} is defined to
+  support an efficient implementation of plan-space search algorithms.
 
 \item \textbf{Domains, Variables and Constraint} are used to represent
-  restrictions in the problem domain in terms of a CP formulation.
+  restrictions in the problem domain in terms of a \texttt{CP}
+  formulation.
 
 \item \textbf{Domain Rules} are used to describe internal and external
-  relationships on and between tokens, as presecribed by the CAIP
-  framework.
+  relationships on and between tokens, as presecribed by
+  \texttt{CAIP}.
 
@@
   built in implementations of Object types that appear frequently in
-  real life domains.
+  real-world domains.
 
 \end{enumerate}
+
+At the atomic level, the entities in the \eu plan representation are:
 
@@
   agent needs to leave or be back, or executes a purchase, are all
-  instances where variables representing time would be used.  As
-  explained in the \textsf{CP} section, a variable can take values
-  from a domain. In \eu, domains are defined over a specific data
-  type, the primitive data types supported are:
-
-  \begin{enumerate}
-
-  \item \textit{String}: sequences of characters, like "Red",
-    "Spirit", etc
-
-  \item \textit{Boolean}: {True,False}
-
-  \item \textit{Numeric}: integers and floats
-
-  \item \textit{Object}: reference to an object instance (see below
-    for a description of objects in \eu)
-
-\end {enumerate}
-
-Two types of variable Domains can be defined over a data type:
-
-  \begin{enumerate}
-
-  \item \textit{Enumerated}: Finite sets over any data type, specified
-    explicitly, for instance: ["Red","Yellow","Green"], [1,18,32], etc
-
-  \item \textit{Interval}: Only for Numeric data types, they are
-    specified by a [LowerBound,UpperBound] pair, for instance: [1,10],
-    [5.0,100.0], etc.
-
-  \end {enumerate}
-
-  \eus ability to represent and reason over numeric intervals is a
-  very useful characteristic that allow users to create flexible
-  plans.
+  instances where variables representing time is used.  As noted in
+  Section \ref{sec:europa:cp}, a variable can take values from a
+  domain. In \eu, domains are defined over a specific data type, the
+  primitive data types supported are, \textit{String},
+  \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
+  below). Further, two types of variable Domains can be defined over a
+  data type, \textit{Enumerated} and \textit{Interval} (used only for
+  numeric data types, they are specified by a [LowerBound,UpperBound]
+  pair, for instance: [1,10], [5.0,100.0]). It is this latter ability
+  to represent and reason over numeric intervals that forms the basis
+  of representing flexible plans.
 
@@
   generically in terms of object types or classes. \eu allows the
-  definition of classes in the same way that it is done in popular
-  object-oriented programming languages, where a class can have
-  attributes (represented by Variables), and classes can be arranged
-  in a single-parent hierarchy through inheritance.  However, one key
-  aspect needs to be introduced to class definition in \eu that is not
+  definition of classes similar to object-oriented programming
+  languages, where a class can have attributes (represented by
+  variables), and classes can be arranged in a single-parent hierarchy
+  through inheritance.  However, one key aspect in \eu that is not
   found in object-oriented programming: to describe state and behavior
   for the purposes of planning we need to build on the formalism of
-  first order logic as explained below.
+  first order logic.
 
@@
   objects and sets of properties to describe state and behavior. For
-  example, we might use a predicate \texttt{At($a$,$l$)} to indicate
+  example, one might use a predicate \texttt{At($a$,$l$)} to indicate
   that agent $a$ is at location $l$, or a predicate
-  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
+  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking action to
   buy product $p$. Note that $a$ is a variable which may have a number
@@
     token's current state and its reachable states through further
-    restriction.  This is required to support CAIP's approach to
-    planning where goals and subgoals can be satisfied by either
-    adding new intervals (Tokens) to the plan, or by merging with
-    existing intervals. Token State lifecycle is examined in detailed
-    in a separate section below.
+    restriction.  This is required to support \texttt{CAIP}'s approach
+    to planning where goals and subgoals can be satisfied by either
+    adding new intervals (tokens) to the plan, or by merging with
+    existing intervals.
 
@@
   some classes that don't have any time-dependent state or behavior,
-  and therefore those classes will not have any Token types associated
-  with them, for instance in the Shopping Agent model, the set
-  locations and products are static for the agent's purposes so we may
-  only want to say what they are but not have any state or behavior
-  associated with them.
-
-  However, the most common case in any non-trivial domain model is
-  that Objects will have many associated Tokens in order to describe
-  their state and behavior throughout a plan. There are a couple of
-  Object Types that are so commonly used in domain descriptions, that
-  \eu provides a built-in implementation for them:
+  and therefore those classes will not have any token types associated
+  with them. For instance in the Shopping Agent model, the set
+  locations and products are static for the agent's purposes so one
+  may only want to declare them but not associate any state or
+  behavior with them.  However, a common case in a non-trivial domain
+  model is that objects will have many associated tokens in order to
+  describe their state and behavior throughout a plan. There are some
+  object types that are used so often in domain descriptions that \eu
+  provides a built-in implementation for them:
 
@@
   by exactly one token for every given timepoint in the plan. Any
-  instances of a class derived from \eus built-in Timeline class will
+  instances of a class derived from \eus built-in timeline class will
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \comment{Maybe: add
-    small example?}
+  overlap may occur among them \cite{mus94}.  \kcomment{add a figure}
 
@@
   of changes that can occur to impact that state i.e. produce,
-  consume, use, change.  Resources are such a common requirement for
-  \eu users that built-in object types (with their corresponding token
-  types to denote production, consumption, etc) are provided for
-  them. Instances of classes derived from a Resource will induce
-  ordering requirements on their Tokens in order to ensure that the
-  level of the resource remains within specified limits.
-  \comment{Maybe: add small example?}
+  consume, use, change.  Resources are a common requirement for \eu
+  users that built-in object types (with their corresponding token
+  types to denote production, consumption, etc) are provided.
+  Instances of classes derived from a resource will induce ordering
+  requirements on their tokens in order to ensure that the level of
+  the resource remains within specified limits.  \kcomment{This
+    definitely needs a visual example and some text esp. since it
+    impacts plan structure via scheduling}
+
 \end{enumerate}
 
-\item[\textbf{Token State Model}] In \eus representation, goals are
+\item[\textbf{Token State Model}] In \eu's representation, goals are
   temporally scoped states that need to be achieved, or actions that
   must be performed. As a result, \eu internally represents goals as
-  Tokens.  Goals can be posted explicitly, for instance, when a user
+  tokens.  Goals can be posted explicitly, for example, when a user
   asks the Shopping Agent to own a specific product by a specific
   time, or implicitly when subgoals are created as a result of domain
-  compatibilities (described in the CAIP framework above), for
-  instance, when the Shopping Agent must get to a Location where a
-  Product is available before it is able to purchase it, in this case
-  the original goal of purchasing the product will spawn a subgoal of
-  being at a specific location.
+  rules. For instance, when the Shopping Agent must get to a location
+  where a product is available before it is able to purchase it: in
+  this case the original goal of purchasing the product spawns a
+  subgoal of being at a specific location without which the purchase
+  will not occur as defined in the domain model.  Looking at how a
+  simple Shopping Agent goal would be addresed by \eu helps illustrate
+  how Token States support the planning process.  Let's assume the
+  planner starts with a partial plan that places the Shopping agent at
+  home at time $0$ with no products in possession. Let's also assume
+  that goals are posted for the Agent to own a drill by time $10$ and
+  to be home by time $20$ as shown in Fig. \ref{fig:europapr1}
 
-  Looking at a how a simple Shopping Agent goal would be addresed by
-  \eu helps illustrate how Token States support the planning process.
+  \begin{figure} \centering
+    \includegraphics[scale=0.4]{figs/europa-pr-1.pdf}
+    \caption{\small Timelines representing initial state and goals for a Shopping Agent.}
+    \label{fig:europapr1}
+  \end{figure}
 
-  Let's assume we start with a partial plan that places the Shopping
-  agent at home at time 0 and with no Products in its possession, then
-  we post a goal for the Shopping Agent to own a Drill at time 10.
-
-  \comment{Use Fig 4 and edit:: Add figure showing an AgentLocation
-    timeline with an At(Home) token with start=0 and end open, an
-    empty AgentBag timeline, and a Own(Drill) token with start=10
-    floating below both timelines}
-
-  Note that the \texttt{Own(Drill)} token that represents the goal is
-  not immediately placed on the \texttt{AgentBag} timeline since \eu
-  must determine whether it can achieve that goal and how. When the
-  goal is posted and before any decisions are made on it, the
-  \texttt{Own(Drill)} token is created in an INACTIVE state, then \eu
-  looks at the current partial plan and considers 2 alternatives:
+  Note that the tokens \texttt{Own(Drill)} and \texttt{At(Home)} that
+  represent the goals are not immediately placed on the
+  \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu must
+  first determine whether it can achieve those goal and how. When the
+  goals are posted and before any decisions are made, the
+  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an
+  \texttt{INACTIVE} state. \eu then considers the partial plan and
+  evaluates two possible alternatives:
 
@@
 \item If there are any tokens in the plan that are compatible with the
-  new goal, \eu may attempt to satisfy the goal by merging it with any
-  of them (if there is more than one it can try them one at a time),
-  in this case, the state of the token would change to
-  \texttt{MERGED}.
+  new goal, \eu may attempt to satisfy the goal by \emph{merging} it
+  with any of them (if there is more than one it can try them one at a
+  time or in some heuristic order), in this case, the state of the
+  token would change to \texttt{MERGED}.
 
 \item Instead of merging with existing tokens in the plan, \eu may
-  decide to insert the new token into the plan, in that case its state
-  would change to ACTIVE.
+  decide to \emph{insert} a new token into the plan, in which case the
+  state would change to \texttt{ACTIVE}.
 
@@
 are transferred to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible with
-the inactive token. Two tokens are merge-compatible if they are
-instances of the same predicate and that no intersections between
-corresponding variables are empty. The effect of merging is
-illustrated in Figure 3.
+token requires finding a target active token that is compatible. Two
+tokens are merge-compatible if they are instances of the same
+predicate and there are no empty intersections between corresponding
+token variables.  The effect of merging is illustrated in
+Fig. \ref{fig:europapr:merge}
 
-\comment{Use previous fig and modify as appropriately:: Add figure
-  showing the effect of merging; all figs on this section are
-  conjoined. JB to sketch it out.}
+\begin{figure} 
+  \centering
+  \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
+  \caption{\small State of timelines after satisfying a goal to be
+    home by time $20$ on merging with initial state.}
+  \label{fig:europapr:merge}
+\end{figure}
 
-When a token is activated all compatibilities associated with its
-token type are evaluated, which may lead to new subgoals being
-generated. In our example, there should be a compatibility stating
-that the Shopping Agent must be at the same location as the product it
-is purchasing. In that case, activating the \texttt{Own(Drill)} token
-will cause a new \texttt{At(Drill.location)} to be created in an
-INACTIVE state, which in turn will have to be dealt with.
+The goal to be home by time $20$ is satisfied by merging with the
+initial state that said that the Agent is home at time $0$.  The merge
+is possible because the tokens are of the same type \texttt{At}, and
+intersecting the domains for all of their variables doesn't yield an
+empty domain (which represents an inconsistency), in this case the
+token variables are the location, which is \texttt{Home} in both
+cases, and the temporal variables (start, end, duration), which can be
+intersected without causing an inconsistency. After the token
+representing the goal is merged onto the token representing the
+initial state, the resulting token represents not only the fact that
+the Agent must be \texttt{At(Home)} by time 0 but also
+\texttt{At(Home)} at time $20$, this is reflected in the new
+constraints on the token's start and end variables as depicted in
+Fig. \ref{fig:europapr:merge}.
 
-\comment{Use previous fig and modify as appropriately: Add figure
-  showing the effect of activating the token by putting it on the
-  AgentBag timeline, and a new At(Drill.location) token floating
-  below}
+When a token is activated, all domain rules associated with its type
+are evaluated, which may lead to new subgoals being generated. In our
+example, there could be a domain rule stating that if a Shopping Agent
+doesn't already own a product, then it must \texttt{Buy} it before it
+can \texttt{Own} it. In such a case, activating the
+\texttt{Own(Drill)} token will cause a new \texttt{Buy(Drill)} to be
+created in an \texttt{INACTIVE} state, which in turn could spawn its
+own subgoals, like being at the location where the product can be
+purchased before being able to buy it, and so on, until all domain
+rules are satisfied.
 
-If neither activating nor merging works, a token may become REJECTED,
-this is possible only for goals posted by the end user and specified
-as optional. Subgoals generated as a result of token activation cannot
-be rejected, if a subgoal cannot be satisfied through activation or
-merging then it will cause the original activation decision that
-spawned it to be retracted (\eus search mechanisms are explained in
-detail in a separate section later on).
+\begin{figure} \centering
+  \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
+  \caption{\small State of timelines after satisfying a goal to own a
+    drill by time $10$ by activating corresponding token.}
+\label{fig:europapr:activate}
+\end{figure}
 
-As we can see, a token's state can go from \texttt{INACTIVE} to
-\texttt{ACTIVE, MERGED} or \texttt{REJECTED}, this lifecycle is
-illustrated in the figure below.
+As Fig. \ref{fig:europapr:activate} shows the decision to merge the
+goal to be \texttt{At(Home)} by time $20$ with the initial state was
+not appropriate since the Agent will have to get away from
+\texttt{Home} to be able to buy the drill which is inconsistent with
+the constraints that resulted after the merge. As a result that
+decision will have to be retracted at some point and the alternative
+to activate or merge with a different token will have to be
+considered.
 
-\comment{Maybe: Add figure showing possible state transitions. Need a FSM.}
+If, after looking at all possibilities, neither activating nor merging
+work, a token may be \texttt{REJECTED}. This is possible only for
+goals posted by the end user and specified as optional. Subgoals
+generated as a result of token activation cannot be rejected. If a
+subgoal cannot be satisfied through activation or merging then it will
+cause the original activation decision that spawned it to be
+retracted.
+
+As can be seen from the above example, during the search for a plan, a
+token's state can go from \texttt{INACTIVE} to \texttt{ACTIVE, MERGED}
+or \texttt{REJECTED}, as shown in Fig. \ref{fig:europapr:states}.
+
+\begin{figure} \centering
+  \includegraphics[scale=0.45]{figs/europa-pr-states.jpg}
+  \caption{\small Possible transitions for a Token State}
+\label{fig:europapr:states}
+\end{figure}
 
 \item[\textbf{Rules}] In order for a plan to be valid, it must comply
-  with all rules and regulations pertinent to the application domain
-  in question. Rules govern the internal and external relationships of
-  a token. For example, consider a parameterized action
+  with all rules pertinent to the relevant application domain. Rules
+  govern the internal and external relationships of a token. For
+  example, consider a parameterized action
   \texttt{Go(origin,destination)} which describes a Shopping Agent
-  traveling from one location ( \textit{origin}) to another (
-  \textit{destination}). The parameters are instantiated on a token as
-  variables whose domain of values is the set of all locations in a
-  given problem. A rule governing an internal relationship among token
-  variables might stipulate that a transition must involve a change in
-  location. This can be easily expressed as a constraint on the
-  definition of a predicate of the form: \textit{origin !=
-    destination}. It makes sense to further stipulate that the agent
-  must be located at \textit{origin} before travel can happen and the
-  agent must be located at \textit{destination} when completed. This
-  is an example of a rule governing an external relationship among
-  tokens. It specifies a requirement that tokens of the predicate
-  \texttt{At} that represent the location state for the Shopping Agent
-  precede and succeed tokens of the predicate \texttt{Go} that
-  represents traveling.
+  traveling from one location (\textit{origin}) to another
+  (\textit{destination}). The parameters are instantiated on a token
+  as variables whose domain of values is the set of all possible
+  locations in a given problem. A rule governing an internal
+  relationship among token variables might stipulate that a transition
+  must involve a change in location. This can be expressed as a
+  constraint on the definition of a predicate of the form:
+  \textit{origin != destination}. A further stipulation could be that
+  the agent must be located at \textit{origin} before travel can occur
+  and the agent must be located at \textit{destination} when travel is
+  completed. This is an example of a rule governing an external
+  relationship among tokens. It specifies a requirement that tokens of
+  the predicate \texttt{At} that represent the location state for the
+  Shopping Agent precede and succeed tokens of the predicate
+  \texttt{Go} that represents traveling.
 
-  \comment{Use previous fig and modify as appropriately (used as
-    above): Add figure showing internal and external relationships
-    from rules on action Go. Need a fig from JB}
+  \begin{figure} 
+    \centering
+    \includegraphics[scale=0.4]{figs/europa-pr-rules.pdf}
+    \caption{\small Master and Slave tokens generated as a result of a domain rule.}
+    \label{fig:europapr:rules}
+  \end{figure}
 
-  Figure xx illustrates the entities and relations involved in
-  specifying such a rule on a \texttt{Go} action. The token on which
-  the rule applies is referred to (\texttt{Go}) as the master. Each
-  \texttt{At} token required by the master is referred to as a
-  slave. All variables are indicated by name and their domains are
-  expressed as intervals in the case of temporal variables and as
-  enumerations for the remainder. Application of a rule on a token can
-  thus cause slave tokens, variables, and constraints to be
-  introduced.
-
-  The capability of a domain rule to cause a slave token to be created
-  is a key vehicle through which planning occurs. Semantically, this
-  rule imposes a requirement for supporting (slave) tokens to be in
-  the plan in order for the master to be valid.
+  Fig. \ref{fig:europapr:rules} illustrates the entities and relations
+  involved in specifying such a rule on a \texttt{Go} action. The
+  token on which the rule applies (\texttt{Go}) is referred to as the
+  \emph{master}. Each \texttt{At} and \texttt{Going} token required by
+  the master is referred to as a \emph{slave}.  Application of a rule
+  on a token can thus cause slave tokens, variables and constraints to
+  be introduced.  The capability of a domain rule to cause a slave
+  token to be created is a key vehicle through which planning
+  occurs. Semantically, a rule imposes a requirement for supporting
+  (slave) tokens to be in the plan in order for the master token to be
+  valid.
 
 \end{description}
-
-%\begin{figure*}
-%\centering 
-%\subfloat[]{\label{fig:plan-evolve1}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-1.pdf}} 
-%\subfloat[]{\label{fig:plan-evolve2}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-2.pdf}} 
-%\subfloat[]{\label{fig:plan-evolve3}\includegraphics[width=0.3\textwidth]{figs/Plan-evolve-3.pdf}} 
-%\caption{\small An illustrative plan synthesis example with concurrent
-%  timelines. Each timeline is an instantiation of a subsystem tracked
-%  by the planner. Tokens describe the instantiated state of subsystem
-%  at a particular time instance and enforce temporal and parametric
-%  constraints. Abstract goals are decomoposed into successively less
-%  abstract tokens and instantiated in co-temporal timelines using
-%  \texttt{Allen Algebra} relations. All tokens represent flexible
-%  start/end times. \ref{fig:plan-evolve1} shows an initial state
-%  evolving into \ref{fig:plan-evolve2} and \ref{fig:plan-evolve3}.}
-%  \label{fig:Plan-evolve}
-%  \vskip-5pt
-%\end{figure*}
-
-%Fig. \ref{fig:Plan-evolve} shows an illustrative example. We show four
-%timelines which track \texttt{Goal, Path, Navigation} and
-%\texttt{Command} state over time. Tokens on the goal timeline indicate
-%a survey within a bounded volume and a min/max depth envelope. As the
-%Volume Survey sub-goals on the \texttt{Path} timeline, the initial
-%token on that timeline gets \emph{squeezed} to have dependencies of
-%the goal token instantiated. Initially this dependency are the tokens
-%\texttt{Go(x1,y1)} and \texttt{Go(xn,yn)} indicative of the area of
-%coverage. The first of these in turn generates a sub-goal on the same
-%timeline to go to the next intermediate waypoint, which is illustrated
-%in Fig. \ref{fig:plan-evolve2}. Subsequent sub-goals (not shown in the
-%figure) ultimately generate all the tokens to completion for this
-%timeline. Fig. \ref {fig:plan-evolve3} shows a snapshot further along
-%in plan generation and shows additional subgoals on the \texttt{Path,
-%  Navigation} and \texttt{Command} timelines. Each token can thus
-%trace its causality when instantiated in the plan. The domain model
-%provides source of these temporal and parametric dependencies in the
-%form of rules.
-
-%\begin{minipage}[c]{\textwidth}
-%\vspace{+0.5cm}
-% \framebox[\textwidth][t]{
-%Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
-%$\Rightarrow$ \\
-%met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
-%starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
-%\ldots{} \\
-%ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
-%meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
-%\end{minipage}
-
-% \end{enumerate}
-
-% \subsubsection{Partitioned Control}
-
-% \subsubsection{State Estimation}
-
-% Even as automated reasoning approaches have the ability to dynamically
-% retarget the vehicle, estimating environmental signals of interest (or
-% their proxies) is important to be able to enable opportunistic science
-% in the water-column. In \texttt{T-REX} feature recognition revolves
-% around a Hidden Markov Model (HMM) \cite{rabiner86} which is encoded
-% directly within the unified representational and computational
-% framework of a reactor. HMMs are useful since the stochastic nature of
-% these models can correlate the type of features we want to detect with
-% the sensor observations. Online targeted sensor data is classified by
-% apriori generated cluster data which in turn determines the
-% probability of having seen the feature of interest. Together with
-% posterior probability, the HMM generates a probability of being within
-% the target feature if it dominates the distribution. If other sampling
-% conditions are satisfied (for instance sampling proximity), a
-% constraint is activited in the model which can trigger a water sampler
-% and also preserve the memory of sampling to alter a future transect
-% resolution.
-
-% The HMM is customized for the feature of interest and is built offline
-% using clustering techniques from large data sets; we use Kohonen's
-% Self Organizing Maps \cite{kohonen}. We use a semi-supervised learning
-% method \cite{zhu05} to extract an environmental model that make uses
-% of both labeled and unlabeled data for training. Online,
-
-% The sensor data for classification depends on the
-% feature of interest; for detecting INLs we use backscattering data
-% from a HydroScat and for tracking riverine plumes to track fertilizer
-% runoff, we use an ISUS Nitrate sensor.
 

### 2012-02-23 europa-basics.tex
minor typos and jpg to PDF converstion

@@
 The next few sections will cover \eu in a bottom-up fashion, going
-from the generic theoretical framework that underlies \eu, to the
+from the generic theoretical framework that underlies \eue, to the
 specific representation, modeling and algorithms implemented by it.  A
@@
 an overview of \eus architecture and how it can be embedded in a
-planning application such as \rx. Finally, \eus modeling, inference
+planning application such as \rxe. Finally, \eus modeling, inference
 and search mechanisms are covered in detail so that the reader can
@@
 \label{fig:shop:shopping}
-\vskip+0.1cm
 \end{figure}
@@
 that to motivate the mechanisms we use in our specific form of
-planning we start with a simple example.
+planning we start with an example.
 
@@
 
-At the core of the \rx framework, is the deliberation engine, \eu,
+At the core of the \rx framework, is the deliberation engine, \eue,
 which has a rich legacy from NASA missions \cite{mus98,rajan00,
@@
 \label{fig:nqueens-2}
-\vskip+0.1cm
 \end{figure}
@@
 \centering
-\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
+\includegraphics[scale=0.25]{figs/Example-UBO0.jpg}
 \caption{\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
@@
 \label{fig:rcpsp-1}
-\vskip+0.1cm
 \end{figure}
@@
 \paragraph{} \textit{Planning}: The Shopping Agent Problem
-\cite{russelnorvig} is a variation of the problem with which we started this chapter.
-In this problem, an agent needs to purchase a set of products
-(milk, drill, etc) that are available at specific locations
-(supermarket, hardware store, etc) and the agent is subject to
+\cite{russelnorvig} is a variation of the problem with which we
+started this chapter.  In this problem, an agent needs to purchase a
+set of products (milk, drill, etc) that are available at specific
+locations (supermarket, hardware store, etc). The agent is subject to
 temporal (must complete tasks by specific deadlines) and resource
-(fuel, carrying capacity, etc) constraints. The agent needs to figure
-out what actions need to be performed to find and acquire the required
-items, as well as when to perform each of those
-actions. Fig. \ref{fig:shopping-1} shows a solution produced by \eu
-for such a problem instance.
+(fuel, carrying capacity, etc) constraints and it needs to synthesize
+actions that need to be performed to find and acquire the required
+items, as well as when to perform each of those actions.
+Fig. \ref{fig:shopping-1} shows a solution produced by \eu for such a
+problem instance.
 
@@
 \centering
-\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
+\includegraphics[scale=0.3]{figs/Example-Shopping0.jpg}
 \caption{\small A \eu solution to a shopping agent problem domain where
@@
 \label{fig:shopping-1}
-\vskip+0.1cm
 \end{figure}
@@
 \centering
-\includegraphics[scale=0.35]{figs/constraint-net.pdf}
+\includegraphics[scale=0.3]{figs/constraint-net.pdf}
 \caption{\small Variables and Constraints represented as a Constraint Network}

### 2012-02-23 intro.tex
minor typos and jpg to PDF converstion

@@
 deeper into the \eu framework itself, the basis of all deliberation
-within \rx. A key portion is Section~\ref{sec:europa:pr} which details
-the internal plan representation essential to planning within a
-situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
-Inference (Section~\ref{sec:europa:inference}) and Search
+within \rxe. A key portion is Section~\ref{sec:europa:pr} which
+details the internal plan representation essential to planning within
+a situated agent like \rxe. Modeling
+(Section~\ref{sec:europa:modeling}), Inference
+(Section~\ref{sec:europa:inference}) and Search
 (Section~\ref{sec:europa:search}) further detail the important details

### 2012-02-23 related.tex
minor typos and jpg to PDF converstion

@@
 
-Our system, \rx, the focus of this chapter, is an agent control
+Our system, \rxe, the focus of this chapter, is an agent control
 structure formally defined as a composition of coordinated control

### 2012-02-23 results.tex
minor typos and jpg to PDF converstion

@@
 by the vehicle when on the surface and goal-oriented commanding with
-\rx, such a capability allows judicious use of the limited samplers by
-a scientist from her desktop.
+\rxe, such a capability allows judicious use of the limited samplers
+by a scientist from her desktop.
 
@@
 shore via the \od following a setup similar to
-Fig.\ref{fig:dta-setup}, to the Dorado running \rx.
+Fig.\ref{fig:dta-setup}, to the Dorado running \rxe.
 
@@
 shows one such experiment in September 2011 with a GPS-enabled
-\texttt{Wirewalker} was tracked by the Dorado running \rx. We are
+\texttt{Wirewalker} was tracked by the Dorado running \rxe. We are
 currently investigating spatio-temporal correlations \cite{gneiting07}
@@
 rovers in \cite{aichang04,bresina05,bresina05a}\footnote{The MAPGEN
-  system uses the same \eu planner used in \rx. It continues to be
+  system uses the same \eu planner used in \rxe. It continues to be
   used routinely to this day on the MER mission.}.

### 2012-02-23 trunk/arch.tex
minor typos and jpg to PDF converstion

@@
-\section{Planning and execution with \rx}
+\section{Planning and Execution with \rx}
 \label{sec:arch}
 
-\subsection{Introductory concepts}
+\subsection{Introduction}
 \label{sec:arch:intro}
@@
 Systematic partitioned inference is a major contribution of
-\rx. However that is not the focus of this chapter with details
+\rxe. However that is not the focus of this chapter with details
 covered in \cite{py10, rajan12}. Coupled interaction between planning
@@
 
-\subsection{The  \eu Deliberative Reactor}
+\subsection{The  Deliberative Reactor}
 \label{sec:arch:europa}
 
-Within \rx, an instance of an \eu reactor comes with the sole purpose
+Within \rxe, an instance of an \eu reactor comes with the sole purpose
 to deliberate. A typical design of such a reactor takes its cues from
@@
 are based on a single unified model within a rich representation that
-provides support for temporal inference. In \rx, the execution cycle
+provides support for temporal inference. In \rxe, the execution cycle
 consists of two deliberation processes which are implemented as two
@@
 This assumption is problematic for situated agents in general and as
-Fig. \ref{fig:tick-exec} shows, for \rx. Synchronization that tracks
+Fig. \ref{fig:tick-exec} shows, for \rxe. Synchronization that tracks
 state evolution can occur several times during reactor
@@
 This suggests a deliberate non overlap between {\em reactive planner}
-and {\em deliberative planners} horizons. In \rx, and more
+and {\em deliberative planners} horizons. In \rxe, and more
 specifically in our \eu reactor the planning horizon of {\em
@@
 
-\subsubsection{Deliberation: Planning for  Evolving State}
+\subsubsection{Deliberation: Planning for  State Evolution}
 \label{sec:arch:plan}

### 2012-02-23 trunk/europa.tex
heading changes based on Minh and discussion; also noted FP comments
minor typos and jpg to PDF converstion

@@
-\subsection{\eu Architecture}
+\subsection{Architecture}
 \label{sec:europa:arch}
@@
 \centering
-\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
-\caption{\small The \eu Architecture}
+\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
+\caption{\small The \eu software architecture.}
 \label{fig:europa-architecture}
-\vskip+0.1cm
 \end{figure}
@@
 
-\subsection{Modeling}
+\subsection{Modeling in \eu}
 \label{sec:europa:modeling}
@@
 different models and goals. Consequently the expressiveness of the
-language for models, goals and plans is of great importance. For \eu,
+language for models, goals and plans is of great importance. For \eue,
 \texttt{NDDL} allows the user to specify the different components of a
@@
 
-\subsection{Inference}
+\subsection{Automated Inference}
 \label{sec:europa:inference}
 
-In \eu, all the Variables and Constraints from a problem domain,
+In \eue, all the Variables and Constraints from a problem domain,
 domain descriptions are kept inside the Constraint Reasoning Engine
@@
 \centering
-\includegraphics[scale=0.35]{figs/europa-resource-envelopes.png}
+\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
 \caption{\small Resource Envelopes computed by  \eu}
@@
 
-\subsection{Search}
+\subsection{Search in \eu}
 \label{sec:europa:search}
@@
   tokens can be generated by explicitly posted goals, or when a token
-  is activated, rules may fire that create inactive slave tokens as described in section
-  \comment{point to EUROPA Plan Representation section}.  Open Conditions can be resolved in three ways:
+  is activated, rules may fire that create inactive slave tokens as
+  described in section \comment{point to EUROPA Plan Representation
+    section}.  Open Conditions can be resolved in three ways:
 

### 2012-02-23 trunk/setobook.tex
- added table of content  I was unable to embed it as pdf bookmarks due to apparent incompatibility of hyperref with our maketitle in svmult document class Insteed I just generated it before the document
minor typos and jpg to PDF converstion

@@
 \input{preamble}
-\setcounter{secnumdepth}{-1}
+\setcounter{tocdepth}{3}
 
@@
 
-\title*{\textbf{\sc Towards Deliberative Control of Autonomous Robots}}
+\title*{\textbf{\sc Towards Deliberative Control in Marine Robotics}}
 \author{\normalsize Kanna~Rajan\inst{1} \and Fr\'ed\'eric Py\inst{1}
@@
 \institute{Monterey Bay Aquarium Research Institute, California
-  \emph{\{Kanna.Rajan,fpy\}@mbari.org}
+  \emph{\{Kanna.Rajan, fpy\}@mbari.org}
   \and
@@
 \begin{document}
+
+\tableofcontents{}
 
@@
 Rajan and Py are funded by a block grant from the David and Lucile
-Packard Foundation to MBARI. They thank their collaborators J. Ryan,
-J. Harvey, R. Vrijenhoek and J. Bellingham at MBARI. In addition they
-thank colleagues M. del Alto, D. Korsmeyer and S. Zornetzer at NASA
-Ames who made \eu available as open source. The authors are indebted
-to N. Muscettola for his ideas and early work at NASA Ames which forms
-the basis of this work.
+Packard Foundation to MBARI. They thank their collaborators John Ryan,
+Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. In
+addition they thank colleagues Martha del Alto, David Korsmeyer and
+Steven Zornetzer at NASA Ames who made \eu available as open
+source. The authors are indebted to Nicola Muscettola for his ideas
+and early efforts in Constraint-based Temporal Planning at NASA Ames
+which forms the basis of this work.
 

### 2012-02-23 europa.tex
display updates
updates on europa.tex for Javier

@@
-\subsection{\eu Architecture}
+\subsection{Architecture}
 \label{sec:europa:arch}
 
-\eu is implemented as a library that encapsulates all the elements
-described above and makes them easily accessible through modeling and
-Application Programming Interface (API) layers.  A user will typically
-use the modeling facilities to inject a problem domain description,
-and the API to extract and manipulate plans, and to obtain information
-about plan flaws and constraint violations at any stage of the
-planning process.
+\eu is implemented as a library that encapsulates all elements
+described above making them accessible via modeling and Application
+Programming Interface (API) layers.  A user will typically use the
+modeling facilities to inject a problem domain description and the
+API to extract and manipulate plans and to obtain information about
+plan flaws and constraint violations at any stage of the planning
+process.
 
@@
 Fig. \ref{fig:europa-architecture} shows the main architectural
-components in \eu and their relationships.
+components in \eu and their relationships. The \textbf{Constraint
+  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
+constraints that define relationships among them. It also provides an
+efficient arc consistency mechanism \cite{mackworth77}. The
+\texttt{CRE} is designed so that specialized reasoning algorithms for
+new constraints can be easily and efficiently plugged in. The
+\textbf{Plan Database} (\texttt{PDB}) manages object and token
+types. It also maintains the state of partial plans in terms of
+object, variable and token instances. This constitutes the data
+foundation on which end user applications and automated planners
+within \eu are built. The \textbf{Rules Engine} (\texttt{RE})
+modularly adds the semantics of domain rules that describe
+dependencies between tokens. The \textbf{Solver Module} provides the
+framework for flaw detection and resolution that is needed to
+implement search algorithms for planning. The \textbf{API Layer}
+provides access to services from all the different modules in a
+programmatic way so that \eu can be easily embedded into specialized
+applications.
 
-The \textbf{Constraint Reasoning Engine} (CRE) manages domain types,
-variables and constraints that define relationships among them. It
-also provides an efficient arc consistency mechanism
-\cite{mackworth77}. The CRE is designed so that specialized reasoning
-algorithms for new constraints can be easily and efficiently plugged
-in.
-
-The \textbf{Plan Database} (PDB) manages object and token types. It
-also maintains the state of partial plans in terms of object, variable
-and token instances. This constitutes the data foundation on which end
-user applications and automated planners are built.
-
-The \textbf{Rules Engine} (RE) modularly adds the semantics of domain
-rules that describe dependencies between tokens.
-
-The \textbf{Solver Module} provides the framework for flaw detection
-and resolution that is needed to implement search algorithms for
-planning.
-
-The \textbf {Language Modules}: Describing problem domains and problem
-instances in terms of objects, variables and domain rules in a
-programmatic way is cumbersome to the extent of being impractical,
-this motivated the creation of modeling languages that support those
-definitions in a much more natural and compact way. \texttt{NDDL} (New
-Domain Description Language [pronounced ``noodle'']) \cite{NDDL} is
-the main modeling language for \eu and will be described in detail in
-the Modeling section below. ANML is a language under development that
+Describing problem domains and problem instances in terms of objects,
+variables and domain rules in a programmatic way is cumbersome to the
+extent of being impractical; this motivated the creation of modeling
+languages that support those definitions in a much more natural and
+compact way. \texttt{NDDL} (New Domain Description Language
+[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
+\eu and will be described in detail in the Modeling section
+below. Another, \texttt{ANML} is a language under development that
 provides a more intuitive syntax \cite{smith08}.
-
-The \textbf{API Layer} provides access to services from all the
-different modules in a programmatic way so that \eu can be easily
-embedded into specialized applications.
 
@@
 
-\item \textit{Temporal Network}: Extensions to the CRE module to
+\item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
   reason about temporal constraints
 
-\item \textit{Resources}: Extensions to the CRE, PDB and Solver
-  modules to provide out-of-the-box reasoning and search for metric
+\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
+  and Solver modules to provide reasoning and search for metric
   resources.
@@
 \item \textit{Chronological Backtracking Solver}: A general purpose
-  built-in solver to provide the ability to plan out-of-the-box,
-  problems that require challenging search may require specialized
-  solvers to be built as explained in the Search section below.
+  built-in solver to provide the ability to plan out-of-the-box.
+  Problems that require challenging search may require specialized
+  solvers to be built as explained in Section \ref{sec:europa:search}.
+
 \end{enumerate}
 
-\subsection{Modeling}
+\subsection{Modeling in \eu}
 \label{sec:europa:modeling}
@@
 different models and goals. Consequently the expressiveness of the
-language for models, goals and plans is of great importance. For \eu,
+language for models, goals and plans is of great importance. For \eue,
 \texttt{NDDL} allows the user to specify the different components of a
-problem domain in a concise way. The main characteristics that make
-\texttt{NDDL} a powerful tool for describing problem domains are:
-
-\begin{enumerate}
-
-\item Object Oriented: \texttt{NDDL} supports classes and inheritance
-  in similar fashion to popular Object-Oriented languages like Java
-  and C++. It also supports polymorphism for some of the planning
-  components. Using object classes and instances is a time-tested
-  approach to naturally describe problem domains \cite{rumbaugh}.
-
-\item It offers declarative constructs to define constraints and
-  causality dependencies between actions and effects.
-
-\item It offers procedural constructs to populate \eu's plan database
-  with specific problem instances expressed in terms of objects,
-  variables, constraints, facts and goals.
-
-\end{enumerate}
-
-To understand how the most important elements of \texttt{NDDL} work,
-we show how the Shopping Agent problem could be modeled:
+problem domain in a precise and concise manner. Among the main
+features that make \texttt{NDDL} a powerful tool for describing
+problem domains are that it is object oriented. \texttt{NDDL} supports
+classes and inheritance in similar fashion to popular Object-Oriented
+languages like Java and C++. It also supports polymorphism for some of
+the planning components. Using object classes and instances is a
+time-tested approach to naturally describe problem domains
+\cite{rumbaugh}. It offers declarative constructs to define
+constraints and causality dependencies between actions and effects and
+finally it offers procedural constructs to populate \eus plan
+database with specific problem instances expressed in terms of
+objects, variables, constraints, facts and goals. To articulate
+important elements of \texttt{NDDL}, we show how the Shopping Agent
+problem could be modeled:
 
@@
 
-In the model above we can see how the main elements or a problem
-domain can be mapped to \eu's representation:
+In the model above one can see the main elements of a problem domain
+can be mapped to \eus representation:
 
-\begin{enumerate}
+\begin{description}
 
-\item \textbf{Object Types}: \texttt{NDDL} supports an object-oriented
+\item[\textbf{Object Types}]: \texttt{NDDL} supports an object-oriented
   approach, where common data and behavior can be abstracted as a
@@
   hierarchy so that state and behavior can be extended through
-  inheritance. We took advantage of that by making
-  \textit{AgentLocation} inherit from \eu's built-in \textit{Timeline}
+  inheritance. We take advantage of that feature by making
+  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
   class which doesn't allow predicates to overlap in time for the same
-  object instance, in this way we can ensure that and agent can only
-  be at one place at a time.
+  object instance. This feature is used for instance, to ensure that
+  an agent can only be at one place at a time in our example.
 
-\item \textbf{Variable Types and Instances} : All object attributes
-  are represented as variables in this model, for instance, it uses
+\item[\textbf{Variable Types and Instances}]: All object attributes
+  are represented as variables in this model. For instance, it uses
   integer variables to represent the built-in temporal parameters for
@@
   relationships between different objects (for instance, each Agent
-  has variable to keep track of its \textit{AgentLocation}).
+  maintains a variable to keep track of its \textit{AgentLocation}).
 
-\item \textbf{Predicate Types}: Some classes like \textit{Product},
-  \textit{Location} and \textit{ProductLocation} are completely static
-  and so don't have any predicate or action types associated with
-  them. The \textit{AgentLocation} class is purely used to keep track
-  of state and so it has \texttt{At} and \texttt{Going} predicate
-  types associated with it. The \textit{Agent} class keeps track of
-  the products it purchases through the \texttt{Own} predicate, having
-  a separate \textit{AgentBag} class with its own predicates would
-  have been an equally valid modeling choice.
+\item[\textbf{Predicate Types}]: Some classes like \textit{Product},
+  \textit{Location} and \textit{ProductLocation} are static and so
+  don't have any predicate or action types associated with them. The
+  \textit{AgentLocation} class is used to keep track of state and so
+  has \texttt{At} and \texttt{Going} predicate types associated with
+  it. The \textit{Agent} class keeps track of the products it
+  purchases through the \texttt{Own} predicate; having a separate
+  \textit{AgentBag} class with its own predicates would have been an
+  equally valid modeling choice.
 
-\item \textbf{Action Types}: The \textit{Agent} class is the only one
-  that has behavior associated with it, consequently, it defines
-  action types to represent when the agent is taking action to
+\item[\textbf{Action Types}]: In our example, \textit{Agent} is solely
+  the class that has behavior associated with it. Consequently, it
+  defines action types to represent when the agent is taking action to
   \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
-  \textit{Product}. We'll see how the actual behavior for those
-  actions is specified below.
-\end{enumerate}
+  \textit{Product}. % We'll see how the actual behavior for those
+  % actions is specified below.
+
+\end{description}
 
@@
 Domain rules can be defined in \texttt{NDDL} in the context of a
-Predicate or an Action. In the example above we can see the main
+predicate or an action. In the example above we can see the main
 elements of a rule definition:
 
-\begin{enumerate}
+\begin{description}
 
-\item \textbf{Variables}: A Token Type (predicate or action) is part
-  of a class, therefore class attributes (see references to
-  \texttt{object.location} above), Token Type parameters (see
-  references to \texttt{[from, to]} in \texttt{Agent::Go}, or to
-  \texttt{[duration, product]} in \texttt{Agent::Buy} and locally
-  declared variables (see \texttt{possibleStores} in
-  \texttt{Agent::Buy}) can all be used in the definition of a Rule.
+\item[\textbf{Variables}]: A token type (predicate or action) is part
+  of a class, therefore class attributes (see \texttt{object.location}
+  above), token type parameters (see \texttt{[from, to]} in
+  \texttt{Agent::Go}, or \texttt{[duration, product]} in
+  \texttt{Agent::Buy}) and locally declared variables (see
+  \texttt{possibleStores} in \texttt{Agent::Buy}) can all be used in
+  the definition of a rule.
 
-\item \textbf{Constraints}: restrictions on the values that variables
-  can take inside a Rule are expressed through constraints, in the
-  example above only the \texttt{eq} constraint is used, but in
-  general, any constraint available in \eu's library can be used.
+\item[\textbf{Constraints}]: Restrictions on values that variables can
+  take within a rule are expressed through constraints. In the example
+  above the \texttt{eq} constraint is used; in general, any constraint
+  available in \eus library can be used.
 
-\item \textbf{Subgoals}: In \texttt{Agent::Go}, subgoals are used to
+\item[\textbf{Subgoals}]: In \texttt{Agent::Go}, subgoals are used to
   specify that the Shopping Agent's location must transition from
   origin to destination in a temporally consistent way. The
-  \texttt{[meets, equals, metby]} operators are used to specify
-  subgoals (Allen operators are explained in more detail below), along
-  with the temporal relationship each token representing a subgoal has
-  with the the token representing the \texttt{Agent::Go} action. Also,
+  \texttt{[meets, equals, metby]} operators (see Section
+  \ref{sec:europa:inference}) are used to specify subgoals, along with
+  the temporal relationship each token representing a subgoal has with
+  the the token representing the \texttt{Agent::Go} action. Also,
   \texttt{[condition, effect]} annotations can be used when defining a
-  subgoal inside an action, \eu doesn't immediately associate any
-  semantics with them, but they can be extremely useful for search
+  subgoal inside an action. Currently \eu does not associate semantics
+  with these annotation; but they can be extremely useful for search
   modules when creating plans.
 
-\end{enumerate}
+\end{description}
 
-Having that specification for the Shopping Agent domain, here is how a
-particular problem instance could be stated:
+With that specification for the Shopping Agent domain, a particular
+problem instance could be stated as follows:
 
@@
 \label{eu:model}
-%\caption{Specifying a problem instance for the Shopping Agent in NDDL}
 \end{verbatim}
 
-A problem instance consists of:
+This problem instance consists of \textbf{Variables} and
+\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
+\textbf{Constraints}. In our example we define temporal variables that
+are used to specify the planning horizon (\texttt{horizonStart} and
+\texttt{horizonEnd}) and objects are used to specify locations,
+products, product-location associations and the Shopping Agent that we
+want to plan for. Facts are used to specify the current state of the
+world and do not need any support and are represented as tokens in
+\eus plan database. In this case we use facts to specify that the
+Shipping Agent is initially at home. Goals are also represented as
+tokens in \eus plan database. Different from facts, goals trigger
+inference in \eu which determines whether flaws (unresolved subgoals,
+overloaded resources, etc.) need to be resolved; if so, search modules
+can be invoked to resolve those flaws as part of the deliberation
+process. Finally, constraints are restrictions on and between actions
+and goals. In the example above, we use temporal constraints to
+specify when facts hold and when the goals should be
+accomplished. % As
+% always, all constraints in \eus library could be used.
 
-\begin{enumerate}
-
-\item \textbf{Variables} and \textbf{Objects}: In this case we defined
-  temporal variables that are used to specify the planning horizon
-  (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are
-  used to specify the locations, products, product-location
-  associations and the Shopping Agent that we want to plan for.
-
-\item \textbf{Facts}: These are used to specify the current state of
-  the world and don't need any support. Facts are represented as
-  tokens in \eu's Plan Database. In this case we use facts to specify
-  that the Shipping Agent is initially at home.
-
-\item \textbf{Goals}: Goals are represented as tokens in \eu's Plan
-  Database. Differently from Facts, Goals trigger inference in \eu
-  which determines whether flaws (unresolved subgoals, overloaded
-  resources, etc.) need to be resolved, if that is the case, search
-  modules can be invoked which will look for a plan to resolve those
-  flaws.
-
-\item \textbf{Constraints}: restrictions on and between actions and
-  goals are specified through constraints, in this case we use
-  temporal constraints to specify when the facts hold and when the
-  goals should be accomplished. As always, all the constraints in
-  \eu's library (built-in, or user defined), can be used to specify
-  these restrictions.
-\end{enumerate}
-
-\subsection{Inference}
+\subsection{Automated Inference}
 \label{sec:europa:inference}
 
-In \eu, all the Variables and Constraints from a problem domain,
-domain descriptions are kept inside the Constraint Reasoning Engine
-(CRE).  The CRE provides a framework where any number of constraint
-types can collaborate to perform inference and prune variable domains
-using Bounds Propagation and Domain Reduction techniques. \eu uses a
-Propagator pattern, in which a Propagator object keeps track of a set
-of constraint instances of the same type and is responsible for
-ensuring that whenever any variable changes or constraints are
-introduced or dropped, all related variables are updated through
-inference.
+In \eue, all variables and constraints from a problem domain
+description are stored inside the \texttt{CRE}.  The \texttt{CRE}
+provides a framework where any number of constraint types can
+collaborate to perform inference and prune variable domains using
+Bounds Propagation and Domain Reduction techniques \kcomment{citation
+  needed}. \eu uses a Propagator pattern which keeps track of a set of
+constraint instances of the same type and is responsible for ensuring
+that whenever any variable is modified or constraints are introduced
+or dropped, all related variables are updated through inference.
 
@@
 each Propagator is only able to ensure consistency for the constraints
-that are associated with it.  To maintain global consistency, the CRE
-implements a variant of the \texttt{AC-3} algorithm \cite{mackworth77}
-where it asks Propagators to perform inference until the Constraint
-Network returns to an stable state.
-
-\kcomment{add europa's AC-3 algorithm here}
+that are associated with it.  To maintain global consistency, the
+\texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
+\cite{mackworth77} where it asks Propagators to perform inference
+until the Constraint Network returns to an stable state.
 
@@
 real-life planning domains, so it is worth going into some more detail
-about how \eu's built-in support works for those constraint types.
+about how \eus built-in support works for those constraint types.
 
@@
 
+reflected by the use of the \texttt{Allen Algebra} \cite{allen84}
+temporal relations shown in Fig. \ref{fig:allen-algebra}. Interval
+arithmetic in the EUROPA$_2$ domain model using \texttt{Allen Algebra}
+allows encoding of temporal constraint relationships between tokens on
+concurrent timelines. See \cite{Jonsson00} for additional details
+including propagation algorithms and interval logic.
+
 \begin{figure}[!htb]
@@
 sufficient propagation to yield a backtrack-free search. Thus we have
-an efficient and complete algorithm for propagating an STN. \eu's
+an efficient and complete algorithm for propagating an STN. \eus
 Propagator for temporal constraints is a straight implementation of
@@
 provides built-in support for resources whose availability can be
-measured with real numbers, in terms of \eu's representation, resource
+measured with real numbers, in terms of \eus representation, resource
 object types and their related constraints are provided as follows:
@@
 
-\comment{figure :provide snapshot of resource envelopes generated by europa}
+\begin{figure}
+\centering
+\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
+\caption{\small Resource Envelopes computed by  \eu}
+\label{fig:resenvelopes}
+\vskip+0.1cm
+\end{figure}
 
@@
   tokens can be generated by explicitly posted goals, or when a token
-  is activated, rules may fire that create inactive slave tokens
-  \comment{reference: point to more detailed decription above in Rules
-    Engine section}.  Open Conditions can be resolved in three ways:
+  is activated, rules may fire that create inactive slave tokens as
+  described in section \comment{point to EUROPA Plan Representation
+    section}.  Open Conditions can be resolved in three ways:
 
@@
   \Begin{
-    $curIter \leftarrow 0$ \;
+    int $curIter \leftarrow 0$ \;
     \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
@@
       $checkSolution()$ \tcp*{See if we have a new best solution}
-      $curIter++$ \;
+      $curIter$\texttt{++} \;
     }
@@
 \end{algorithm}
-
-% \begin{verbatim}
-%     public void solve(ConstraintEngine ce, int maxIter)
-%     {
-%     	curIter=0;
-
-%     	while ((ce.getViolations() > 0) && (curIter < maxIter)) {
-% 	    Variable queenToMove = getQueenWithMaxViolation(ce);
-%     	     int curPos = queenToMove.getCurrentValue();
-
-%             SortedSet<int> moves = getMoves(ce,queenToMove,curPos);             
-%   	   boolean moved = false;
-%             for (int newPos : moves) {
-%             	moved = tryMove(queenToMove,curPos,newPos);
-%             	if (moved)
-%             	    break;
-%             }
-
-%             if (!moved) 
-%             	forceMove(queenToMove,curPos,moves.first());
-
-%             checkSolution(); // See if we have a new best solution            
-%             curIteration_++;
-%     	}
-%     }
-% \end{verbatim}
 

### 2012-02-23 trunk/europa.tex
Corrected missed references

@@
 
-\eu is implemented as a library that encapsulates all the elements
-described above and makes them easily accessible through modeling and
-Application Programming Interface (API) layers.  A user will typically
-use the modeling facilities to inject a problem domain description,
-and the API to extract and manipulate plans, and to obtain information
-about plan flaws and constraint violations at any stage of the
-planning process.
+\eu is implemented as a library that encapsulates all elements
+described above making them accessible via modeling and Application
+Programming Interface (API) layers.  A user will typically use the
+modeling facilities to inject a problem domain description and the
+API to extract and manipulate plans and to obtain information about
+plan flaws and constraint violations at any stage of the planning
+process.
 
@@
 \item \textit{Chronological Backtracking Solver}: A general purpose
-  built-in solver to provide the ability to plan out-of-the-box,
-  problems that require challenging search may require specialized
-  solvers to be built as explained in the Search section below.
+  built-in solver to provide the ability to plan out-of-the-box.
+  Problems that require challenging search may require specialized
+  solvers to be built as explained in Section \ref{sec:europa:search}.
+
 \end{enumerate}
@@
 \texttt{NDDL} allows the user to specify the different components of a
-problem domain in a concise way. The main characteristics that make
-\texttt{NDDL} a powerful tool for describing problem domains are:
-
-\begin{enumerate}
-
-\item Object Oriented: \texttt{NDDL} supports classes and inheritance
-  in similar fashion to popular Object-Oriented languages like Java
-  and C++. It also supports polymorphism for some of the planning
-  components. Using object classes and instances is a time-tested
-  approach to naturally describe problem domains \cite{rumbaugh}.
-
-\item It offers declarative constructs to define constraints and
-  causality dependencies between actions and effects.
-
-\item It offers procedural constructs to populate \eu's plan database
-  with specific problem instances expressed in terms of objects,
-  variables, constraints, facts and goals.
-
-\end{enumerate}
-
-To understand how the most important elements of \texttt{NDDL} work,
-we show how the Shopping Agent problem could be modeled:
+problem domain in a precise and concise manner. Among the main
+features that make \texttt{NDDL} a powerful tool for describing
+problem domains are that it is object oriented. \texttt{NDDL} supports
+classes and inheritance in similar fashion to popular Object-Oriented
+languages like Java and C++. It also supports polymorphism for some of
+the planning components. Using object classes and instances is a
+time-tested approach to naturally describe problem domains
+\cite{rumbaugh}. It offers declarative constructs to define
+constraints and causality dependencies between actions and effects and
+finally it offers procedural constructs to populate \eus plan
+database with specific problem instances expressed in terms of
+objects, variables, constraints, facts and goals. To articulate
+important elements of \texttt{NDDL}, we show how the Shopping Agent
+problem could be modeled:
 
@@
 
-A problem instance consists of:
-
-\begin{enumerate}
-
-\item \textbf{Variables} and \textbf{Objects}: In this case we defined
-  temporal variables that are used to specify the planning horizon
-  (see \texttt{horizonStart} and \texttt{horizonEnd}), and objects are
-  used to specify the locations, products, product-location
-  associations and the Shopping Agent that we want to plan for.
-
-\item \textbf{Facts}: These are used to specify the current state of
-  the world and don't need any support. Facts are represented as
-  tokens in \eu's Plan Database. In this case we use facts to specify
-  that the Shipping Agent is initially at home.
-
-\item \textbf{Goals}: Goals are represented as tokens in \eu's Plan
-  Database. Differently from Facts, Goals trigger inference in \eu
-  which determines whether flaws (unresolved subgoals, overloaded
-  resources, etc.) need to be resolved, if that is the case, search
-  modules can be invoked which will look for a plan to resolve those
-  flaws.
-
-\item \textbf{Constraints}: restrictions on and between actions and
-  goals are specified through constraints, in this case we use
-  temporal constraints to specify when the facts hold and when the
-  goals should be accomplished. As always, all the constraints in
-  \eu's library (built-in, or user defined), can be used to specify
-  these restrictions.
-\end{enumerate}
+This problem instance consists of \textbf{Variables} and
+\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
+\textbf{Constraints}. In our example we define temporal variables that
+are used to specify the planning horizon (\texttt{horizonStart} and
+\texttt{horizonEnd}) and objects are used to specify locations,
+products, product-location associations and the Shopping Agent that we
+want to plan for. Facts are used to specify the current state of the
+world and do not need any support and are represented as tokens in
+\eus plan database. In this case we use facts to specify that the
+Shipping Agent is initially at home. Goals are also represented as
+tokens in \eus plan database. Different from facts, goals trigger
+inference in \eu which determines whether flaws (unresolved subgoals,
+overloaded resources, etc.) need to be resolved; if so, search modules
+can be invoked to resolve those flaws as part of the deliberation
+process. Finally, constraints are restrictions on and between actions
+and goals. In the example above, we use temporal constraints to
+specify when facts hold and when the goals should be
+accomplished. % As
+% always, all constraints in \eus library could be used.
 
@@
 
-In \eue, all the Variables and Constraints from a problem domain,
-domain descriptions are kept inside the Constraint Reasoning Engine
-(CRE).  The CRE provides a framework where any number of constraint
-types can collaborate to perform inference and prune variable domains
-using Bounds Propagation and Domain Reduction techniques. \eu uses a
-Propagator pattern, in which a Propagator object keeps track of a set
-of constraint instances of the same type and is responsible for
-ensuring that whenever any variable changes or constraints are
-introduced or dropped, all related variables are updated through
-inference.
+In \eue, all variables and constraints from a problem domain
+description are stored inside the \texttt{CRE}.  The \texttt{CRE}
+provides a framework where any number of constraint types can
+collaborate to perform inference and prune variable domains using
+Bounds Propagation and Domain Reduction techniques \kcomment{citation
+  needed}. \eu uses a Propagator pattern which keeps track of a set of
+constraint instances of the same type and is responsible for ensuring
+that whenever any variable is modified or constraints are introduced
+or dropped, all related variables are updated through inference.
 
@@
 each Propagator is only able to ensure consistency for the constraints
-that are associated with it.  To maintain global consistency, the CRE
-implements a variant of the \texttt{AC-3} algorithm \cite{mackworth77}
-where it asks Propagators to perform inference until the Constraint
-Network returns to an stable state.
-
-%\kcomment{add europa's AC-3 algorithm here}
+that are associated with it.  To maintain global consistency, the
+\texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
+\cite{mackworth77} where it asks Propagators to perform inference
+until the Constraint Network returns to an stable state.
 
@@
 
+reflected by the use of the \texttt{Allen Algebra} \cite{allen84}
+temporal relations shown in Fig. \ref{fig:allen-algebra}. Interval
+arithmetic in the EUROPA$_2$ domain model using \texttt{Allen Algebra}
+allows encoding of temporal constraint relationships between tokens on
+concurrent timelines. See \cite{jonsson00} for additional details
+including propagation algorithms and interval logic.
+
 \begin{figure}[!htb]

### 2012-02-23 trunk/europa.tex
updates on europa.tex for Javier

@@
 constraints and causality dependencies between actions and effects and
-finally it offers procedural constructs to populate \eus plan
-database with specific problem instances expressed in terms of
-objects, variables, constraints, facts and goals. To articulate
-important elements of \texttt{NDDL}, we show how the Shopping Agent
-problem could be modeled:
+finally it offers procedural constructs to populate \eus \texttt{PDB}
+with specific problem instances expressed in terms of objects,
+variables, constraints, facts and goals. To articulate important
+elements of \texttt{NDDL}, we show how the Shopping Agent problem
+could be modeled:
 
@@
 want to plan for. Facts are used to specify the current state of the
-world and do not need any support and are represented as tokens in
-\eus plan database. In this case we use facts to specify that the
-Shipping Agent is initially at home. Goals are also represented as
-tokens in \eus plan database. Different from facts, goals trigger
-inference in \eu which determines whether flaws (unresolved subgoals,
-overloaded resources, etc.) need to be resolved; if so, search modules
-can be invoked to resolve those flaws as part of the deliberation
+world and do not need any support and are represented as tokens in the
+\texttt{PDB}. In this case we use facts to specify that the Shipping
+Agent is initially at home. Goals are also represented as tokens in
+the \texttt{PDB}. Different from facts, goals trigger inference in \eu
+which determines whether flaws (unresolved subgoals, overloaded
+resources, etc.) need to be resolved; if so, search modules can be
+invoked to resolve those flaws as part of the deliberation
 process. Finally, constraints are restrictions on and between actions
@@
 % always, all constraints in \eus library could be used.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\caption{\small Temporal relations defined within the planner are
+  based on \texttt{Allen Algebra} relations shown above.}
+\label{fig:allen-algebra}
+\end{figure}
+
+In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
+temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
+encoding of temporal constraint relationships between tokens on
+concurrent timelines. \kcomment{This para could do with a little
+  embellishment especially since it is the only material that actually
+  talks about the kinds of constraints and their origin.}
 
@@
 constraint instances of the same type and is responsible for ensuring
-that whenever any variable is modified or constraints are introduced
-or dropped, all related variables are updated through inference.
+variables are inferentially updated with change (modifications or
+constraint addition/deletion).
+
+% that whenever any variable is modified or constraints are introduced
+% or dropped, all related variables are updated through inference.
 
@@
 relating them to the rest of the plan through constraints, by changing
-the start time of an activity, etc), or can come from an automated
-planner (by expanding subgoals, introducing precedence constraints to
-resolve mutual exclusion conflicts, etc). When these changes happen,
-each Propagator is only able to ensure consistency for the constraints
-that are associated with it.  To maintain global consistency, the
+the start time of an activity, etc), or from an automated planner (by
+expanding subgoals, introducing precedence constraints to resolve
+mutual exclusion conflicts, etc). When these changes occur, each
+propagator is only able to ensure consistency for the constraints that
+are associated with it.  To maintain global consistency, the
 \texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
-\cite{mackworth77} where it asks Propagators to perform inference
-until the Constraint Network returns to an stable state.
+\cite{mackworth77} where propagators are triggered to infer to ensure
+the constraint network returns to a quiescent state.
 
-\input{constraints-table}
+\subsubsection{Constraint Formulation}
+\label{sec:europa:constraints}
 
-A \eu user can introduce new constraint types by implementing
-specialized Constraint and Propagator classes in C++, however, \eu
-provides a built-in Constraint Library with over 25 arithmetic, temporal, set and
-resource constraint types so that a large number of problem domains
-can be modeled without having to introduce new constraint
+An \eu user can introduce new constraint types by implementing
+specialized constraint and propagator classes; however, \eu provides a
+built-in constraint library with a range of arithmetic, temporal, set
+and resource constraint types so that a large number of problem
+domains can be modeled without having to introduce new constraint
 types. Tables \ref{tab:calconst}, \ref{tab:setconst},
@@
 
-Temporal and resource constraints are specially important to model
-real-life planning domains, so it is worth going into some more detail
-about how \eus built-in support works for those constraint types.
+\input{constraints-table}
 
-\textit{Temporal Constraints:} \eu uses variables to explicitly
-represent timepoints for plan activities and states. Constraints among
-timepoints provide a natural way to express domain axioms. For
-example, in order to state that activity $A$ must occur before
-activity $B$ we can say that the end timepoint of $A$ is $leq$ the
-start timepoint of $B$.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
-\label{fig:allen-algebra}
-\vskip-0.3cm
-\end{figure}
-
+Temporal and resource constraints are especially important to model
+real-world planning domains. % so it is worth going into some more
+% detail about how \eus built-in support works for those constraint
+% types.
+\eu uses variables to explicitly represent timepoints for plan
+activities and states. Constraints among timepoints provide a natural
+way to express domain axioms. For example, in order to state that
+activity $A$ must occur before activity $B$ we can say that the end
+timepoint of $A$ is $leq$ the start timepoint of $B$.
 \cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network (STN). Such a
-network can be transformed into a Distance Graph (DG) where the
-outward arc from a node represents the maximum distance from the
-source node to the target node.  The diagram below illustrates a
-simple STN with just 2 variables and a single constraint. It also
-shows the resulting DG.
-
-reflected by the use of the \texttt{Allen Algebra} \cite{allen84}
-temporal relations shown in Fig. \ref{fig:allen-algebra}. Interval
-arithmetic in the EUROPA$_2$ domain model using \texttt{Allen Algebra}
-allows encoding of temporal constraint relationships between tokens on
-concurrent timelines. See \cite{jonsson00} for additional details
-including propagation algorithms and interval logic.
+grouped together to form a Simple Temporal Network
+(\texttt{STN}). Such a network can be transformed into a Distance
+Graph (\texttt{DG}) where the outward arc from a node represents the
+maximum distance from the source node to the target node.
+Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
+a single constraint along with the resulting \texttt{DG}.
+\cite{dechter91} also showed that shortest path algorithms could be
+used to propagate values in the network and discover a negative cycle:
+namely a path from a node to itself that has a path length less than
+$0$. If such a cycle exists, the network is inconsistent. It was
+further shown that a single-source shortest path algorithm was
+sufficient to detect a negative cycle and provide sufficient
+propagation to yield a backtrack-free search. \eus propagator for
+temporal constraints is a straight implementation of this algorithm.
 
@@
 
-\cite{dechter91} also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative
-cycle. A negative cycle is a path from a node to itself that has a
-path length less than 0. If such a cycle exists, the network is
-inconsistent. It was further shown that a single-source shortest path
-algorithm was sufficient to detect a negative cycle and provide
-sufficient propagation to yield a backtrack-free search. Thus we have
-an efficient and complete algorithm for propagating an STN. \eus
-Propagator for temporal constraints is a straight implementation of
-this algorithm.
-
-\textit{Resource Constraints:} Most real life planning problems have
-to deal with the fact that plans made up of actions that require
-resources to be performed typically have only a limited amount of
-those resources available (think of fuel, storage capacity, etc).  \eu
-provides built-in support for resources whose availability can be
-measured with real numbers, in terms of \eus representation, resource
-object types and their related constraints are provided as follows:
-
-\begin{enumerate}
-
-\item \textbf{Reservoir} resources can have both
+Most real-world planning problems need to deal with plans made up of
+actions that typically depend upon limited resources (\eg fuel,
+storage capacity, battery state of charge, data buffers).  \eu
+provides built-in support for resource computation. In \eus
+representation resource object types come in three forms with related
+constraints. \textbf{Reservoir} resources can have both
 \texttt{consume(resource,time,quantity)} and
 \texttt{produce(resource,time,quantity)} constraints. Production and
-  consumption is instantaneous (it happens at a single point in time)
+consumption is instantaneous. \textbf{Reusable} resources have
+\texttt{uses(resource,start,duration,quantity)} constraints that use a
+quantity of the resource for a specific duration, which are then
+released. \textbf {Unary} resources are reusable with capacity$ = 1$,
+which allows \eu to perform run time savings in comparison to the
+generic reusable class. Configuration options are offered such that in
+addition to the usual constraint violations when a resource is
+completely depleted, plan flaws can be generated whenever a resource
+level is outside a specific upper or lower bound, or when
+instantaneous production or consumption exceeds a specific threshold.
 
-\item \textbf{Reusable} resources have
-  \texttt{uses(resource,start,duration,quantity)} constraints that use
-  a quantity of the resource for a specific duration, then release it.
-
-\item \textbf {Unary} resources are Reusable resources with
-  capacity=1, which allows \eu to perform some run time savings
-  compared to the generic Reusable class.
-
-\end{enumerate}
-
-\eu offers configuration options so that in addition to the usual
-constraint violations when a resource is completely depleted, plan
-flaws can be generated whenever a resource level is outside a specific
-upper or lower limit, or when instantaneous production or consumption
-exceeds a specific threshold.
-
-Recall that \eu supports a flexible temporal representation, where
-time is represented by intervals, instead of single values. In this
-context, determining whether a resource is the source of a plan flaw
-or a constraint violation in general is more difficult than just
+Since \eu supports a flexible temporal representation, where time is
+represented by intervals, instead of single values, determining
+whether a resource is the source of a plan flaw or a constraint
+violation in general is substantially more difficult than just
 computing a profile of the resource and checking against some limits.
-In the context of time flexibility, there are many resource profiles
-that are possible, depending on when exactly each
-production/consumption event happens out of all the possible values
-that are in the domain of the variables that represent the time for
-the event.  \cite{mus02} proposed an algorithm where the variables
-that represent the time for production/consumption events are placed
-in a temporal network, similar to the network used by the Temporal
-Propagator above, then the constraints on the resource can be seen as
+With temporal flexibility, a number of resource profiles could be
+possible depending on precisely when each production/consumption event
+occurs out of a range of possible values that are in the domain of the
+variables that represent the temporal extant for the event.
+\cite{Muscettola04,Muscettola06} proposed an algorithm where the
+variables that represent the time for production/consumption events
+are placed in a temporal network, similar to the network used by the
+temporal propagator; constraints on the resource can then be seen as
 quantities flowing through that network. Computing the maximum flow
-will yield the maximum resource usage, and thus a lower bound on the
-resource level, and computing a minimum flow will similarly yield an
-upper bound. Since maximum flow on a temporal network can be computed
-in polynomial time, \eu takes advantage of this algorithm to
-efficiently compute upper and lower bound envelopes on the resource
-level.
+then, yields the maximum resource usage and thus a lower bound on the
+resource level; computing a minimum flow will similarly yield an upper
+bound. Since maximum flow on a temporal network can be computed in
+polynomial time, \eu takes advantage of this algorithm to efficiently
+compute upper and lower bound envelopes on the resource
+level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
+computation.
 
@@
 \includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
-\caption{\small Resource Envelopes computed by  \eu}
+\caption{\small An example of resource envelopes computed by \eu.}
 \label{fig:resenvelopes}
-\vskip+0.1cm
 \end{figure}
 
-Resource envelopes can be very useful when trying to figure how
-brittle a plan in the face of changing activity durations or start
-times. They can also be useful when searching for plans, although they
-are relatively expensive to compute so their ability to prune the
-search space must be weighted against the ability to search faster
+Resource envelopes can be useful when trying to evaluate plan
+brittleness in the face of changing activity durations or start
+times. They can also be useful when searching for plans; however they
+are relatively expensive to compute. Therefore their ability to prune
+the search space must be weighted against the ability to search faster
 \cite{6007772}.
@@
 
-Given the above elements, an automated problem solver can be
-created using:
+Given the above, an automated problem solver can be created using, a
+\emph{domain model} that describes the variable, object, predicate and
+action types that are relevant for the problem, a \emph{problem
+  instance} or initial state that consists of variable and object
+instances that exist for the entire planning horizon, \emph{temporally
+  scoped predicate and action} instances and \emph{temporally scoped
+  goals}. This information is retained in \eus \texttt{PDB} so that
+inference and search mechanisms can be used to look for a problem
+solution.
 
-\begin{enumerate}
+Since temporal intervals may extend indefinitely in both positive and
+negative directions, for problems with a temporal dimension it is also
+common to specify a \emph{planning horizon} when invoking a solver;
+any decisions that fall completely outside of the horizon can be
+ignored. This is particularly useful for recurrent activities;
+consider for instance a model where a medical checkup has to occur
+every $3$ months. The user can model a recurring activity and then use
+the horizon to ensure that a finite number of medical checkup
+activities are generated as part of a plan. This is also useful to
+ignore activities that may occur far in the future or in the past to
+be relevant for a particular plan, ensuring that the solver is making
+as few commitments as necessary and making planning tractable.
 
-\item A domain model that describes the variable, object, predicate
-  and action types that are relevant for the problem.
-
-\item A problem instance (also called initial state) that consists of:
-
-    \begin{enumerate}
-
-    \item Variable and Object instances that exist for the entire
-      planning horizon.
-
-    \item Temporally scoped predicate and action instances.
-
-    \item Temporally scoped goals.
-
-\end{enumerate}
-
-\end{enumerate}
-
-This information is kept in \eus Plan Database so that inference and
-search mechanisms can be used to look for a problem solution.
-
-Given that temporal intervals may extend indefinitely in both positive
-and negative directions, for problems with a temporal dimension it is
-also common to specify a Planning Horizon when invoking a solver. The
-Planning Horizon is the temporal interval that a problem solver has to
-worry about, any decisions that fall completely outside of the
-Planning Horizon can be ignored. This is very useful for recurrent
-activities, consider for instance a model where a medical checkup has
-to occur every 3 months, the user can model a recurring activity and
-then use the Planning Horizon to ensure that a finite number of
-medical checkup activities are generated as part of a plan. This is
-also useful to ignore activities that may happen too far in the future
-or in the past to be relevant for a particular plan, this way the user
-can ensure that the solver is making as few commitments as necessary.
-
-\eu provides a built-in solver that performs Plan Space Planning
-\cite{ghallab04} in this approach, the initial state is considered a
-partial plan that needs to be refined toward a solution plan that
-achieves the goals. The operations to refine the partial plan PP at
+\eu provides a built-in solver that performs \emph{Plan Space
+  Planning} \cite{ghallab04}; the initial state is considered a
+partial plan that needs to be refined toward a solution that achieves
+stated goals. The operations to refine the partial plan \texttt{PP} at
 any time are:
@@
 
-\item Find the flaws of PP, that is the conditions that prevent it
-  from being a solution plan (flaw types are explained below).
+\item Find flaws of \texttt{PP}, \ie conditions that prevent it from
+  being a solution plan.
 
-\item Select one such flaw
+\item Select one such flaw.
 
-\item Select a resolver for the flaw
+\item Select a resolver for the flaw.
 
-\item Refine PP by applying the resolver
+\item Refine \texttt{PP} by applying the resolver.
 
@@
 \item If all possible resolvers for a particular flaw fail, return
-  failure, otherwise continue until resolving all the flaws.
+  failure, otherwise continue until resolving all flaws.
 
@@
 
-This Plan Space Planning algorithm is translated into \eus
-representation as follows:
+% This Plan Space Planning algorithm is translated into \eus
+% representation as follows:
 
 The initial partial plan is the state of \eus Plan Database after the
-initial state has been instantiated, this results in a set of
+initial state has been instantiated. This results in a set of
 variable, object and token instances. Inference then takes place to
-detect flaws in the partial plan, there are 3 kinds of flaws that can
-be detected:
+detect flaws in the partial plan; the three kinds of flaws that can be
+detected are:
 
-\begin{enumerate}
+\begin{description}
 
-\item Unbound Variable: a variable in the partial plan whose domain is
-  not a singleton. Unbound Variables are resolved by specification of
-  a value from the domain of the variable.
+\item[\textbf{Unbound Variable}]: a variable in the partial plan whose
+  domain is not a singleton. Unbound Variables are resolved by value
+  selection from their domain.
 
-\item Open Condition: an open condition is an inactive token. Inactive
-  tokens can be generated by explicitly posted goals, or when a token
-  is activated, rules may fire that create inactive slave tokens as
-  described in section \comment{point to EUROPA Plan Representation
-    section}.  Open Conditions can be resolved in three ways:
+\item[\textbf{Open Condition}]: an open condition is an
+  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
+  by explicitly posted goals, or when a token is activated, rules may
+  activate the creation of \texttt{INACTIVE} slave tokens as described
+  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
+  three ways:
 
-  \begin{enumerate}
+  \begin{description}
 
-  \item Merging: A free activity is merged with a matching activity
-    already in the plan. An activity a is said to match an activity a?
-    if a and a? unify and the temporal constraints involving a are
-    satisÞed by a?. Thus, a and a? can be considered the same activity
-    and we do not need to introduce a in the plan. Consequently, the
-    compatibilities associated with a are not Þred, because they have
-    been already triggered when a? was introduced in the plan.
+  \item[\textbf{Merging}]: A free activity is merged with a matching
+    activity already in the plan. An activity $a$ is said to match an
+    activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig} and the
+    temporal constraints involving $a$ are satisfied by $a?$. Thus,
+    $a$ and $a?$ can be considered the same activity without the need
+    to introduce $a$ in the plan. Consequently, the compatibilities
+    \kcomment{compatibilities have not been defined} associated with
+    $a$ are not Þred \kcomment{???}, because they have been already
+    triggered when $a?$ was introduced in the plan.
 
-  \item Activation: We introduce a new activity a in the current plan
-    associating it with the proper timeline, but without choosing a
-    speciÞc time slot for it.The compatibilities associated with a are
-    applied and the subgoal activities resulting from those
-    compatibilities are introduced as free activities. This results in
-    both an ordering flaw, corresponding to the just activated
+  \item[\textbf{Activation}]: A new activity $a$ is introduced in the
+    current plan associating it with the correct timeline, without
+    choosing a specific time slot. The compatibilities \kcomment{as
+      above} associated with $a$ are applied and the subgoal
+    activities resulting from those compatibilities \kcomment{as
+      above} are introduced as free activities \kcomment{what is a
+      free activity?}. This results in both an ordering flaw
+    \kcomment{undefined}, corresponding to the just activated
     activity, and a number of open condition flaws, corresponding to
@@
 
-  \item Rejection: Some goals may be optional, if that is the case for
-    a particular order condition, its corresponding flaw can be
-    resolved by discarding the inactive token.
+  \item[\textbf{Rejection}]: Some goals may be optional; if so its
+    corresponding flaw can be resolved by discarding the
+    \texttt{INACTIVE} token.
 
-\end{enumerate}
+  \end{description}
 
-\item Threat: once a token has been placed in the partial plan it may
-  impact other tokens indirectly through possible overlapping
-  requirements on objects. Recall for example that a token may belong
-  to objects (e.g. Timelines) which require a total order over their
-  tokens. If any 2 tokens could possibly overlap (though not
-  necessarily), then they pose a threat to each other in terms of
+\item[\textbf{Threat}]: Once a token has been placed in the partial
+  plan it may impact other tokens indirectly through possible
+  overlapping requirements on objects. Recall for example that a token
+  may belong to objects (\eg timelines) which require a total order
+  over their tokens. If any two tokens could possibly overlap (though
+  not necessarily), then they pose a threat to each other in terms of
   achieving an extension of the current partial plan which is complete
@@
 
-\end{enumerate}
+\end{description}
 

### 2012-02-23 trunk/arch.tex
Few additions in regard to Javier comments

@@
   process and world evolution to ensure that planning occurs
-  \emph{during continuous} state update. By doing so, the planner is
+  \emph{during continuous} state update in the {\em same} polan
+  database used for planning. By doing so, the planner is
   better informed in its deliberation despite a larger latency than
   the changing world. 
-
 \end{itemize}
@@
 
+It is important to note that in T-REX architecture the only critical
+task of a reactor is to sucessfully identify its state through
+synchronization as the execution frontier advances. Indeed, a reactor
+that fails to integrate its {\em external} state updates and/or
+identify its {\em internal} state at any tick express that it is not
+able anymore to explain the state of the world. In the agent point of
+view that means that any state information from this reactor is not to
+be trucsted anymore and will result on this reactor exclusion from the
+execution cycle. As the agent destroy the  faulty reactor is
+destraoyed the agent produced on all of its former {\em internal}
+timelines the special observation \texttt{Failed} to all the reactors
+which did declare these state variables as {\em external} informing
+them through this that this sepcific state varaiable is currently
+unavailable. On the other hand all goals exchanged between reactors
+are considered as rejectable by the reactor that manages them. As a
+result the falure to find a plwn for these goals, while to be avoided,
+is not considered as a critical failure of the reactor and will only
+result on these goals being ignored until new goals are received.
+
+Take for example our AUV agent from Fig. \ref{fig:agent},
+and assume that the \texttt{Sampling} reactor failed to synchronize;
+in  such case this reactor is destroyed and both
+\texttt{Sampling\_policy} and \texttt{Relevant\_data} state variables
+produced the \texttt{Failed} observation. Therefore, the
+\texttt{Survey controller} will received this observation on its {\em
+  external} \texttt{Sampling\_policy} which indicates that this state
+varaiable is not managed anymore. At this point depending on this
+reactor implementationm we have two possible outcomes:
+\begin{enumerate}
+\item \texttt{Survey  controller} model allows him to integrate this
+  \texttt{Failed} observation. For example it can integarte that from
+  now there's now way to take water samples. It is still able to
+  navigate as \texttt{Pilot} is still available which allow him to
+  continue this aspect of its current plan.
+\item \texttt{Survey controller} models is not able to handle this
+  unexpected failure resulting in turn on its failure to
+  synchronize. This result on this reactor being also destroyed and
+  the \texttt{Failed} observation is produced on the \texttt{survey}
+  timeline.
+\end{enumerate}
+Even on the worst case scenario of the later case, we
+note that a reactor failure only propagates up in the reactor graph. 
+In our example, even if all the reactors that depends on
+\texttt{Sampling} are entrained by its failure, we still keep the
+three reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
+  operation} allowing to  maintain the vehicle navigation control
+(\texttt{Pilot}) along with high level safety operation 
+(\texttt{Ship operation}). It illustrates how \rx can handle graceful
+system degradation despite the failure of one or more reactors. It
+also emphasize that it is critical to ensure that the low level
+reactors are properly designed in order to be as robust as 
+possible.
+
 \subsection{The  Deliberative Reactor}
@@
   external timeline into the plan database. 
+  \begin{itemize}
+  \item New observations are integrated by creating the
+    corresponding token as a fact in the plan database
+  \item If no observations have been received at this tick the last
+    observation token restrict its \texttt{end} domaine to the
+    interval $[\tau+1, +\infty)$
+  \end{itemize}
 
@@
 
-A key aspect of the \rx framework is interleaved plan execution. As
-noted above, it is synchronization which drives planning, so the
+A key aspect of the \rx framework is interleaved plan and
+execution. Both {\em synchronization} -- keeping track of execution --
+and {\em deliberation} -- dealing with reactor planning -- of one
+reactor are sharing the same {\em single} plan database which 
+they modify alternatively as the \rx agent schedule their execution.
+
+As noted above, it is synchronization which drives planning, so the
 co-mingling between these two processes is critical to maintaining

### 2012-02-23 trunk/europa.tex
updates on europa.tex for Javier

@@
 
+\subsubsection{Flaw Resolution}
+\label{sec:europa:flaws}
+
 \eu provides a built-in solver that performs \emph{Plan Space
@@
 
-Open Conditions and Threats allow flaw detection and resolution at a
-higher-level of abstraction (i.e. in terms of objects and tokens) than
-that of simply binding variables as is common in Constraint
-Satisfaction Problems. This is advantageous when one applies
-heuristics for ordering choices since it provides a richer context in
-which to make decisions.  Furthermore, it aids in reducing the amount
-of work done by a solver so that only the necessary refinements are
-made, otherwise leaving the partial plan with maximum flexibility. For
-example, one can omit unbound variables which are time-points of
-tokens since threats will force a solver to impose restrictions on
-these variables based on the semantics of the objects to which their
-tokens apply. Thus the planning process may yield partially-ordered
-plans for which all possible extensions are provably valid.
+Open conditions and threats allow flaw detection and resolution at a
+higher-level of abstraction (\ie in terms of objects and tokens) than
+that of binding variables as is common in Constraint Satisfaction
+Problems. This is advantageous when a solver applies heuristics for
+ordering choices since it provides a richer context in which to make
+decisions.  Furthermore, it aids in reducing the amount of work done
+by a solver so that only the necessary refinements are made leaving a
+partial plan with sufficient flexibility. For example, one can bypass
+committing unbound timepoint variables since threats will force a
+solver to impose ordering restrictions on these variables.  Thus the
+planning process may yield partially-ordered plans for which all
+possible extensions are provably valid.
 
 The Plan Space Planning algorithm described above can be implemented
-in many different ways depending on the approach chosen for flaw and
+in multiple ways depending on the approach chosen for flaw and
 resolver selection and for backtracking. \eus built-in solver
@@
 
-The algorithm takes as input a partial plan p and returns true if a
-complete and consistent refinement of p could be found (or if p is
-initially complete and consistent), and false otherwise.
+The algorithm takes as input a partial plan $p$ and returns true if a
+complete and consistent refinement of $p$ could be found (or if $p$ is
+initially complete and consistent) and false otherwise. Briefly the
+algorithm is described as:
 
-\begin{enumerate}
+\begin{description}
 
-\item Line \ref{li:propagate}: Propagate the constraints to test for
-  inconsistency. If found to be inconsistent, then we can return false
-  since this is a dead-end i.e. no refinements to p can yield a
-  consistent plan.
+\item[Line \textbf{\ref{li:propagate}}]: Propagate the constraints to
+  test for inconsistency. If inconsistent, then return false since no
+  refinements to $p$ can yield a consistent plan.
 
-\item Line \ref{li:selectflaw}: Choose a flaw from the set of
-  available flaws.
+\item[Line \textbf{\ref{li:selectflaw}}]: Choose a flaw from the set
+  of available flaws.
 
-\item Line \ref{li:noflaw}: If there are no flaws, then p is complete
-  and we can terminate with success.
+\item[Line \textbf{\ref{li:noflaw}}]: If there are no flaws, then $p$
+  is complete and terminate with success.
 
-\item Line \ref{li:decisionpt}: Otherwise we formulate a decision
-  point which is a branch in the search space. Each choice is a
-  particular refinement operation and the DecisionPoint collects all
+\item[Line \textbf{\ref{li:decisionpt}}]: Formulate a decision point
+  which is a branch in the search space. Each choice is a particular
+  refinement operation and the \texttt{DecisionPoint} collects all
   possible refinement operations for the given flaw.
 
-\item Line \ref{li:decisionloop}: We keep trying until the chosen flaw
-  is resolved, or until we run out of resolvers to try.
+\item[Line \textbf{\ref{li:decisionloop}}]: Keep trying until the
+  chosen flaw is resolved, or until there are no further resolvers to
+  try.
 
-\item Line \ref{li:decide}: A new partial plan is obtained by
-  application of a refinement operator. Note that the ordering over
+\item[Line \textbf{\ref{li:decide}}]: A new partial plan is obtained
+  by application of a refinement operator. Note that the ordering over
   refinement operators to select is a non-deterministic step.
 
-\item Line \ref{li:recurse}: Recursive call to solve the new planning
-  problem. If successful, then we are done.
+\item[Line \textbf{\ref{li:recurse}}]: Recursive call to solve the
+  new planning problem. If successful, then success.
 
-\item Line \ref{li:backtrack}: Otherwise, retract the last refinement
-  operation and move on to try the next one.
+\item[Line \textbf{\ref{li:backtrack}}]: Otherwise, retract the last
+  refinement operation and move on to try the next one.
 
-\item Line \ref{li:noplan}: If we arrive here, then we have exhausted
-  all options to resolve the flaw, including the case where no options
-  were available initially. Thus the problem cannot be solved.
+\item[Line \textbf{\ref{li:noplan}}]: Arriving here implies all
+  options to resolve the flaw have been exhausted, including the case
+  where no options were available initially. Thus the problem is
+  insoluble.
 
-\end{enumerate}
+\end{description}
 
-This algorithm provides for a sound and complete search, assuming that
-no flaws or available refinement operators are pruned unnecessarily.
-Dead-ends in the search are discovered through constraint propagation.
-Constraint propagation is a vehicle for evaluating the consistency of
-a partial plan and also for filtering infeasible values from
-consideration prior to commitment, allowing in some cases a strong
-look-ahead capability which is essential for tractable
-search. Consistency testing is initiated by the isConsistent
+Algorithm \ref{alg:europa:solve} provides for a sound and complete
+search, assuming that no flaws or available refinement operators are
+pruned unnecessarily.  Dead-ends in the search are discovered through
+constraint propagation, a vehicle for evaluating the consistency of a
+partial plan and also for filtering infeasible values from
+consideration prior to commitment. % This allows in some cases a strong
+% look-ahead capability which is essential for tractable
+% search.
+Consistency testing is initiated by the \texttt{isInconsistent}
 procedure. The algorithm permits a heuristically controlled search by
-applying orderings for chooseFlaw and makeDecisionPoint. It results in
-a chronologically-backtracking search.
+applying orderings for \texttt{chooseFlaw} and
+\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
+search. Heuristics can be implemented by applying ordering/filtering
+policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
+instance, by dealing with earliest flaws first, or by making planning
+decisions (open condition flaws) before scheduling decision (threats).
 
-Heuristics can be implemented by applying ordering/filtering policies
-for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for instance,
-by dealing with earliest flaws first, or by making planning decisions
-(open condition flaws) before scheduling ones (threats), etc.
-
-It should be emphasized that while this algorithm is commonly
-employed, it is only one of many that could be implemented. As a
-simple example, consider the following solver for the N-Queens
-problem, where there is an integer variable Vi to represent the queen
-in column i, so that for instance, V1=3 means that the variable on
-column 1 is placed on row 3.  non-attack constraints are posted as
-arithmetic constraints among the Vi variables to ensure that all the
-rows and diagonals contain only one queen.
+It should be emphasized that while this algorithm is commonly employed
+with \eu implementations, it is only one of many that could be
+implemented. As a simple example, consider the following solver for
+the $N$-Queens problem, where an integer variable V$_i$ represents the
+queen in column $i$; V$_1$ $= 3$ implies that the variable on column
+$1$ is placed on row $3$.  Non-attack constraints are posted as
+arithmetic constraints among the V$_i$ variables to ensure that all
+the rows and diagonals contain only one queen. \eus Constraint
+Reasoning Engine or \texttt{CRE} can then be used to compute the
+non-attack constraint violations, with \texttt{ce.getViolations()} a
+call on the API, to pick the queen that has he most violations.  By
+implementing \texttt{getQueenWithMaxViolation()} and looking ahead to
+see what moves are possible for a queen that would remove its
+violations with the implementing of \texttt{getMoves()}, this solver
+provides a viable planner implementation for a range of possible
+initial states.
 
@@
 
-\eus Constraint Reasoning Engine can be used to compute the non-attack
-constraint violations (\texttt{ce.getViolations()} is a straight call
-on the \eu API), to easily pick the queen that has he most violations
-by implementing \texttt{getQueenWithMaxViolation(ce)} and to look
-ahead and see what moves are possible for a queen that would remove
-its violations by implementing
-\texttt{getMoves(ce,queenToMove,curPos)}.
-
 In similar fashion, the constraint violation and flaw reporting

### 2012-02-23 europa-basics.tex
updates on europa.tex for Javier

@@
 dealing with constraints is the $N$-Queens problem in which chess queens
-must be placed on an  $N$x$N$ chessboard so no queens attack the other. % Fig.
+must be placed on an  $N \times N$ chessboard so no queens attack the other. % Fig.
 % \ref{fig:nqueens-1} shows an example of a random positioning of queens
@@
 \includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small N-Queens solution generated by \eu}
+\caption{\small $N$-Queens solution generated by \eu}
 \label{fig:nqueens-2}

### 2012-02-23 trunk/arch.tex
Typos in arch.tex

@@
 
-It is important to note that in T-REX architecture the only critical
-task of a reactor is to sucessfully identify its state through
+It is important to note that, in T-REX architecture, the only critical
+task of a reactor is to successfully identify its state through
 synchronization as the execution frontier advances. Indeed, a reactor
@@
 view that means that any state information from this reactor is not to
-be trucsted anymore and will result on this reactor exclusion from the
-execution cycle. As the agent destroy the  faulty reactor is
-destraoyed the agent produced on all of its former {\em internal}
-timelines the special observation \texttt{Failed} to all the reactors
-which did declare these state variables as {\em external} informing
-them through this that this sepcific state varaiable is currently
+be trusted anymore and results on this reactor exclusion from the
+execution cycle and consequent destruction. As the agent destroy 
+the  faulty reactor is destroyed the agent produced on all of 
+its former {\em internal} timelines the special observation 
+\texttt{Failed} to all the reactors which did declare these state 
+variables as {\em external} informing
+them through this that this specific state variable is currently
 unavailable. On the other hand all goals exchanged between reactors
 are considered as rejectable by the reactor that manages them. As a
-result the falure to find a plwn for these goals, while to be avoided,
+result the failure to find a plan for these goals, while to be avoided,
 is not considered as a critical failure of the reactor and will only
@@
   external} \texttt{Sampling\_policy} which indicates that this state
-varaiable is not managed anymore. At this point depending on this
-reactor implementationm we have two possible outcomes:
+variable is not managed anymore. At this point depending on this
+reactor implementation we have two possible outcomes:
 \begin{enumerate}
 \item \texttt{Survey  controller} model allows him to integrate this
-  \texttt{Failed} observation. For example it can integarte that from
+  \texttt{Failed} observation. For example it can integrate that from
   now there's now way to take water samples. It is still able to
@@
 \end{enumerate}
-Even on the worst case scenario of the later case, we
+Even on the worst case scenario presented on this second case, we
 note that a reactor failure only propagates up in the reactor graph. 
@@
 an aid in resolving its {\em current state flaw}. It does so by using
-the plan information on the exepected {\em internal} state evolution
+the plan information on the expected {\em internal} state evolution
 near the execution frontier.
@@
 In the real world the fact that the robot is \texttt{Going} to a
-certain location does not necessarly imply that it will succeed in
+certain location does not necessarily imply that it will succeed in
 ending up there. Due to an exogenous event, for example, the road to
@@
 support the synchronization process. The observation \texttt{Going}
-occuring at $\tau'$ is fully compatible with the token already present
-in our plan. By updating the plan and progating equality constraints,
+occurring at $\tau'$ is fully compatible with the token already present
+in our plan. By updating the plan and propagating equality constraints,
 the resolution of the {\em current state flaw} for the {\em Agent}
@@
 
-% In that sense, and thanks to the carefull choice on our
-% synchronization heurisitic we see that the plan produced during \rx
-% deliberation steps helps sthe synchronization process to be more
+% In that sense, and thanks to the careful choice on our
+% synchronization heuristic we see that the plan produced during \rx
+% deliberation steps helps the synchronization process to be more
 % efficient.
@@
 % reactors and deliberation steps can be done within a time that is
-% resonably small in comparison to the agent tick duration.
+% reasonably small in comparison to the agent tick duration.
 

### 2012-02-23 arch.tex
heading changes based on Minh and discussion; also noted FP comments
Extra typos an rewording

@@
 
-\subsection{The  \eu Deliberative Reactor}
+It is important to note that, in T-REX architecture, the only critical
+task of a reactor is to successfully identify its state through
+synchronization as the execution frontier advances. Indeed, a reactor
+that fails to integrate its {\em external} state updates and/or
+identify its {\em internal} state at any tick express that it is not
+able anymore to explain the state of the world. In the agent point of
+view that means that any state information from this reactor is not to
+be trusted anymore and results on this reactor exclusion from the
+execution cycle and consequent destruction. As the agent destroy 
+the  faulty reactor is destroyed the agent produced on all of 
+its former {\em internal} timelines the special observation 
+\texttt{Failed} to all the reactors which did declare these state 
+variables as {\em external} informing
+them through this that this specific state variable is currently
+unavailable. On the other hand all goals exchanged between reactors
+are considered as rejectable by the reactor that manages them. As a
+result the failure to find a plan for these goals, while to be avoided,
+is not considered as a critical failure of the reactor and will only
+result on these goals being ignored until new goals are received.
+
+Take for example our AUV agent from Fig. \ref{fig:agent},
+and assume that the \texttt{Sampling} reactor failed to synchronize;
+in  such case this reactor is destroyed and both
+\texttt{Sampling\_policy} and \texttt{Relevant\_data} state variables
+produced the \texttt{Failed} observation. Therefore, the
+\texttt{Survey controller} will received this observation on its {\em
+  external} \texttt{Sampling\_policy} which indicates that this state
+variable is not managed anymore. At this point depending on this
+reactor implementation we have two possible outcomes:
+\begin{enumerate}
+\item \texttt{Survey  controller} model allows him to integrate this
+  \texttt{Failed} observation. For example it can identify that, from
+  now on, there's no way to take water samples. It still can 
+  navigate as \texttt{Pilot} is available which allow him to continue
+  this aspect of its current plan.
+\item \texttt{Survey controller} models is not able to handle this
+  unexpected failure resulting in turn on its failure to
+  synchronize. This result on this reactor being also destroyed and
+  the \texttt{Failed} observation is produced on the \texttt{survey}
+  timeline.
+\end{enumerate}
+Even on the worst case scenario presented on this second case, we
+note that a reactor failure only propagates up in the reactor graph. 
+In our example, even if all the reactors that depends on
+\texttt{Sampling} are entrained by its failure, we still keep the
+three reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
+  operation} allowing to  maintain the vehicle navigation control
+(\texttt{Pilot}) along with high level safety operation 
+(\texttt{Ship operation}). It illustrates how \rx can handle graceful
+system degradation despite the failure of one or more reactors. It
+also emphasize that it is critical to ensure that the low level
+reactors are properly designed in order to be as robust as 
+possible.
+
+\subsection{The  Deliberative Reactor}
 \label{sec:arch:europa}
 
-Within \rx, an instance of an \eu reactor comes with the sole purpose
+Within \rxe, an instance of an \eu reactor comes with the sole purpose
 to deliberate. A typical design of such a reactor takes its cues from
@@
 managing this state variable internally.
-
+x1
 \begin{figure}[!htb]

### 2012-02-24 trunk/europa.tex
updates on europa.tex for Javier

@@
 \texttt{consume(resource,time,quantity)} and
-\texttt{produce(resource,time,quantity)} constraints. Production and
-consumption is instantaneous. \textbf{Reusable} resources have
-\texttt{uses(resource,start,duration,quantity)} constraints that use a
-quantity of the resource for a specific duration, which are then
-released. \textbf {Unary} resources are reusable with capacity$ = 1$,
-which allows \eu to perform run time savings in comparison to the
-generic reusable class. Configuration options are offered such that in
-addition to the usual constraint violations when a resource is
-completely depleted, plan flaws can be generated whenever a resource
-level is outside a specific upper or lower bound, or when
-instantaneous production or consumption exceeds a specific threshold.
+\texttt{produce(resource,time,quantity)} constraints; production and
+consumption is instantaneous. \textbf{Reusable} resources on the other
+hand, have constraints that use a quantity of the resource for a
+specific duration (\texttt{uses(resource,start,duration,quantity)}),
+which are then subsequently released. \textbf {Unary} resources are
+reusable with capacity$ = 1$, which allows \eu to perform run time
+savings in comparison to the generic reusable class. Configuration
+options are offered such that in addition to the usual constraint
+violations when a resource is completely depleted, plan flaws can be
+generated whenever a resource level is outside a specific upper or
+lower bound, or when instantaneous production or consumption exceeds a
+specific threshold.
 

### 2012-02-24 concepts.tex
updates on europa.tex for Javier

@@
   Shopping Agent precede and succeed tokens of the predicate
-  \texttt{Go} that represents traveling.
+  \texttt{Go} that represents traveling. Fig. \ref{fig:europapr:rules}
+  illustrates the entities and relations involved in specifying such a
+  rule on a \texttt{Go} action. The token on which the rule applies
+  (\texttt{Go}) is referred to as the \emph{master}. Each \texttt{At}
+  and \texttt{Going} token required by the master is referred to as a
+  \emph{slave}.  Application of a rule on a token can thus cause slave
+  tokens, variables and constraints to be introduced.  The capability
+  of a domain rule to cause a slave token to be created is a key
+  vehicle through which planning occurs. Semantically, a rule imposes
+  a requirement for supporting (slave) tokens to be in the plan in
+  order for the master token to be valid.
 
@@
 
-  Fig. \ref{fig:europapr:rules} illustrates the entities and relations
-  involved in specifying such a rule on a \texttt{Go} action. The
-  token on which the rule applies (\texttt{Go}) is referred to as the
-  \emph{master}. Each \texttt{At} and \texttt{Going} token required by
-  the master is referred to as a \emph{slave}.  Application of a rule
-  on a token can thus cause slave tokens, variables and constraints to
-  be introduced.  The capability of a domain rule to cause a slave
-  token to be created is a key vehicle through which planning
-  occurs. Semantically, a rule imposes a requirement for supporting
-  (slave) tokens to be in the plan in order for the master token to be
-  valid.
-
 \end{description}

### 2012-02-24 constraints-table.tex
updates on europa.tex for Javier

@@
   \centering
-  \renewcommand{\arraystretch}{1.2}%{0.99}
-  \begin{tabular}{|l|l|p{8cm}|p{3cm}|}
+  \renewcommand{\arraystretch}{1.0}%{0.99}
+  \begin{tabular}{|l|l|p{7cm}|p{3cm}|}
     \hline
@@
   \end{tabular}
-  \caption{\small Calculation Constraints: These constraints enforce that one variable (usually a) is constrained by a calculation done on the remaining variables.}
+  \caption{\small Calculation Constraints: These constraints enforce that one variable (usually $a$) is constrained by a calculation done on the remaining variables.}
   \label{tab:calconst}
@@
   \renewcommand{\arraystretch}{1.0}%{0.99}
-  \begin{tabular}{|l|l|p{8cm}|p{3cm}|}
+  \begin{tabular}{|l|l|p{7.5cm}|p{3cm}|}
     \hline
@@
   \renewcommand{\arraystretch}{1.0}%{0.99}
-  \begin{tabular}{|l|l|p{5cm}|p{3.5cm}|}
+  \begin{tabular}{|l|l|p{7cm}|p{3cm}|}
     \hline
@@
   \renewcommand{\arraystretch}{1.0}%{0.99}
-  \begin{tabular}{|l|l|p{10cm}|p{3cm}|}
+  \begin{tabular}{|l|l|p{9.5cm}|p{3cm}|}
     \hline

### 2012-02-24 europa-basics.tex
updates on europa.tex for Javier

@@
 % \end{figure}
-If we define $N$x$N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in [1,N]$,
-Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen,
-$0$ otherwise. Then the following constraints need to be satisfied:
+If we define $N \times N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in
+[1,N]$, Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a
+Queen, $0$ otherwise. Then the following constraints need to be
+satisfied:
 
@@
 variables are nodes and constraints are arcs between the variables. A
-pair of variables $x \& y$ is said to be \emph{arc consistent} if for
-every value in $x$'s domain, there is a value in $y$'s domain that is
-consistent with all the constraints that connect $x$ and $y$.  A naive
-approach to ensuring arc consistency consists of cycling through all
-the variable pairs and performing constraint propagation until there
-are no variable domain changes. The \texttt{AC-3} algorithm
+pair of variables $x$ and $y$ is said to be \emph{arc consistent} if
+for every value in $x$'s domain, there is a value in $y$'s domain that
+is consistent with all the constraints that connect $x$ and $y$.  A
+naive approach to ensuring arc consistency consists of cycling through
+all the variable pairs and performing constraint propagation until
+there are no variable domain changes. The \texttt{AC-3} algorithm
 \cite{mackworth77}, a popular implementation, improves over the naive

### 2012-02-24 trunk/results.tex
[arch] moved synch failure example earlier in the text [result] reword paragraph about gulper in order to enlarge the motivation from simply Gulper to "sensor based" missions

@@
 A significant driver of the applicability of our work was the
+growing need for AUVs mission to be driven by scientific sensor 
+data.Thid trend is illustrated by the 
 implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
@@
 \cite{rainville01,pinkel11} are resulting in new modalities of making
-Lagrangian and Eulerian observations. Fig. \ref{sep11-wirewalker}
+Lagrangian and Eulerian observations. Fig. \ref{fig:sept11-wirewalker}
 shows one such experiment in September 2011 with a GPS-enabled
-\texttt{Wirewalker} was tracked by the Dorado running \rx. We are
+\texttt{Wirewalker} was tracked by the Dorado running \rxe. We are
 currently investigating spatio-temporal correlations \cite{gneiting07}
@@
 functional level of the systems' capabilities.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-24 trunk/arch.tex
updates to arch and results mostly cosmetic

@@
 
-\begin{figure}[!htbp]
-  \centering
-  \vskip-1pc
-  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
-  \caption{\small \rx reactor execution cycle: {\em Deliberation} is
-    interrupted by {\em synchronization} at the beginning of every
-    {\em tick} allowing integration of state information.}
-  \label{fig:tick-exec}
-  \vskip-0.8pc
-\end{figure}
-
-The manner in which the agent interleaves planning \texttt{step}s and
-synchronization for a single reactor is illustrated in
-Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick while
-interrupting planning; this allows a reactor to identify its current
-state that can propagate through the reactor hierarchy in order to
-ensure a consistent view of the state of the world between all the
-reactors.
-% This view of a reactor has two parallel tasks allowing for long
-% planning deliberation while ensuring that the state evolution
-% propagates through the agent as time advance.
-These intertwined tasks together provide two natural information flows
-(bottom-up for state estimation and top-down for plan projection) that
-we expect to see in a control loop. As they are both inference-based
-processes it is natural to implement a reactor based on generic
-automated planning. For example in Fig. \ref{fig:agent}, apart from
-\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces, every
-other reactor (in blue) is a different instance of an \eu reactor
-differentiated only by their model and \eu solver configurations.
-
 It is important to note that, in T-REX architecture, the only critical
@@
 \item \texttt{Survey  controller} model allows him to integrate this
-  \texttt{Failed} observation. For example it can integrate that from
-  now there's now way to take water samples. It is still able to
-  navigate as \texttt{Pilot} is still available which allow him to
-  continue this aspect of its current plan.
+  \texttt{Failed} observation. For example it can identify that, from
+  now on, there's no way to take water samples. It still can 
+  navigate as \texttt{Pilot} is available which allow him to continue
+  this aspect of its current plan.
 \item \texttt{Survey controller} models is not able to handle this
@@
 possible.
+
+\begin{figure}[!htbp]
+  \centering
+  \vskip-1pc
+  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
+  \caption{\small \rx reactor execution cycle: {\em Deliberation} is
+    interrupted by {\em synchronization} at the beginning of every
+    {\em tick} allowing integration of state information.}
+  \label{fig:tick-exec}
+  \vskip-0.8pc
+\end{figure}
+
+The manner in which the agent interleaves planning \texttt{step}s and
+synchronization for a single reactor is illustrated in
+Fig.~\ref{fig:tick-exec}. Synchronization occurs at every tick while
+interrupting planning; this allows a reactor to identify its current
+state that can propagate through the reactor hierarchy in order to
+ensure a consistent view of the state of the world between all the
+reactors.
+% This view of a reactor has two parallel tasks allowing for long
+% planning deliberation while ensuring that the state evolution
+% propagates through the agent as time advance.
+These intertwined tasks together provide two natural information flows
+(bottom-up for state estimation and top-down for plan projection) that
+we expect to see in a control loop. As they are both inference-based
+processes it is natural to implement a reactor based on generic
+automated planning. For example in Fig. \ref{fig:agent}, apart from
+\textsf{Shore comm.} and \textsf{Vehicle} acting as interfaces, every
+other reactor (in blue) is a different instance of an \eu reactor
+differentiated only by their model and \eu solver configurations.
 
@@
 managing this state variable internally.
-x1
+
 \begin{figure}[!htb]
@@
     \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
-    \caption{\small Shopping example: Tokens in red indicate
-      observations, tokens in blue the goal that this plan attempts to
-      solve for. Arrows indicate temporal constraints. }
+    \caption{\small The Shopping example: Tokens in {\color{red}red}
+      indicate observations, tokens in {\color{blue}blue} the goal
+      that this plan attempts to solve for. Arrows indicate temporal
+      constraints. }
   \label{fig:shop:exec}

### 2012-02-24 trunk/results.tex
updates to arch and results mostly cosmetic

@@
   Dorado using a setup shown in Fig. \ref{fig:dta-setup}. The Google
-  Earth post-hoc image shows red track lines of the \rx commanded
-  Dorado with the yellow transects of the LRAUV. Yellow dots show the
-  track line of a simple GPS-enabled drifter used for seeding the
-  experiment over the Monterey Canyon with Santa Cruz, CA to the right
-  of the figure.}
+  Earth post-hoc image shows {\color{red}red} track lines of the \rx
+  commanded Dorado with the {\color{yellow}yellow} transects of the
+  LRAUV. {\color{yellow}Yellow} dots show the track line of a
+  GPS-enabled drifter used for seeding the experiment over the
+  Monterey Canyon with Santa Cruz, California to the right of the
+  figure.}
 \label{fig:ballet}
@@
   \centering
-\subfloat[The \od Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
-\subfloat[Current Implementation for \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
+  \subfloat[A conceptual view of the \od]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+  \subfloat[Current Implementation for
+  \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
 \caption{\small The Oceanographic Decision Support System (\od) is a
@@
 \subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
-\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
-  the \od in use. Fig. \ref{fig:odss-virtual} shows a drifter tracking
-  tool used during \can planning; it used a physical model of advected
+\caption{\small \ref{fig:odss-mtg} shows a 2010 \can meeting with the
+  \od in use. \ref{fig:odss-virtual} shows a drifter tracking tool
+  used during \can planning; it used a physical model of advected
   surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
-  deployments. Fig. \ref{fig:odss-ipad} shows \od use on a portable
-  iPad device.}
+  deployments. \ref{fig:odss-ipad} shows \od use on a portable iPad
+  device.}
 \label{fig:odss2}

### 2012-02-24 trunk/concepts.tex
heading changes based on Minh and discussion; also noted FP comments
addressed Kanna's comments.

@@
-\section{Fundamentals of \eu}
+\section{The \eu Planner }
 \label{sec:basics}
@@
 \cite{europapso} was named \eut and is the subject of this chapter and
-used by \rx.  Note however, in using the name \eu, we refer to the
+used by \rxe.  Note however, in using the name \eu, we refer to the
 \emph{fundamental ideas as well as the planning paradigm} even if
@@
 \centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\includegraphics[scale=0.3]{figs/flexible-timelines.pdf}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \kcomment{add a figure}
+  overlap may occur among them \cite{mus94}. Fig. \ref{fig:europapr1} below and several others
+  after that illustrate how timelines are used to represent state in a plan.
 
@@
   of changes that can occur to impact that state i.e. produce,
-  consume, use, change.  Resources are a common requirement for \eu
+  consume, use, change.  Resources are such a common requirement for \eu
   users that built-in object types (with their corresponding token
@@
   requirements on their tokens in order to ensure that the level of
-  the resource remains within specified limits.  \kcomment{This
-    definitely needs a visual example and some text esp. since it
-    impacts plan structure via scheduling}
+  the resource remains within specified limits.
 
@@
   time, or implicitly when subgoals are created as a result of domain
-  rules. For instance, when the Shopping Agent must get to a location where a product is
-  available before it is able to purchase it: in this case the
-  original goal of purchasing the product spawns a subgoal of being at
-  a specific location without which the purchase will not occur as
-  defined in the domain model.  Looking at how a simple Shopping
-  Agent goal would be addresed by \eu helps illustrate how Token
-  States support the planning process.  Let's assume the planner
-  starts with a partial plan that places the Shopping agent at home at
-  time $0$ with no products in possession. Let's also assume that goals are posted for
-  the Agent to own a drill by time 10 and to be home by time 20 as depicted in Fig. \ref{fig:europapr1}
+  rules. For instance, when the Shopping Agent must get to a location
+  where a product is available before it is able to purchase it: in
+  this case the original goal of purchasing the product spawns a
+  subgoal of being at a specific location without which the purchase
+  will not occur as defined in the domain model.  Looking at how a
+  simple Shopping Agent goal would be addresed by \eu helps illustrate
+  how Token States support the planning process.  Let's assume the
+  planner starts with a partial plan that places the Shopping agent at
+  home at time $0$ with no products in possession. Let's also assume
+  that goals are posted for the Agent to own a drill by time $10$ and
+  to be home by time $20$ as shown in Fig. \ref{fig:europapr1}
 
   \begin{figure} \centering
-  \includegraphics[scale=0.45]{figs/europa-pr-1.jpg}
+    \includegraphics[scale=0.4]{figs/europa-pr-1.pdf}
     \caption{\small Timelines representing initial state and goals for a Shopping Agent.}
     \label{fig:europapr1}
-\vskip+0.1cm
   \end{figure}
 
-  Note that the tokens \texttt{Own(Drill)}  and \texttt{At(Home)} that represent the goals are
-  not immediately placed on the \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu
-  must first determine whether it can achieve those goal and how. When the
+  Note that the tokens \texttt{Own(Drill)} and \texttt{At(Home)} that
+  represent the goals are not immediately placed on the
+  \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu must
+  first determine whether it can achieve those goal and how. When the
   goals are posted and before any decisions are made, the
-  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an \texttt{INACTIVE}
-  state. \eu then considers the partial plan and evaluates two
-  possible alternatives:
+  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an
+  \texttt{INACTIVE} state. \eu then considers the partial plan and
+  evaluates two possible alternatives:
 
@@
 tokens are merge-compatible if they are instances of the same
-predicate and there are no empty intersections between corresponding token variables. 
-The effect of merging is illustrated in Fig. \ref{fig:europapr:merge}
+predicate and there are no empty intersections between corresponding
+token variables.  The effect of merging is illustrated in
+Fig. \ref{fig:europapr:merge}
+
+\begin{figure} 
+  \centering
+  \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
+  \caption{\small State of timelines after satisfying a goal to be
+    home by time $20$ on merging with initial state.}
+  \label{fig:europapr:merge}
+\end{figure}
+
+The goal to be home by time $20$ is satisfied by merging with the
+initial state that said that the Agent is home at time $0$.  The merge
+is possible because the tokens are of the same type \texttt{At}, and
+intersecting the domains for all of their variables doesn't yield an
+empty domain (which represents an inconsistency), in this case the
+token variables are the location, which is \texttt{Home} in both
+cases, and the temporal variables (start, end, duration), which can be
+intersected without causing an inconsistency. After the token
+representing the goal is merged onto the token representing the
+initial state, the resulting token represents not only the fact that
+the Agent must be \texttt{At(Home)} by time 0 but also
+\texttt{At(Home)} at time $20$, this is reflected in the new
+constraints on the token's start and end variables as depicted in
+Fig. \ref{fig:europapr:merge}.
+
+When a token is activated, all domain rules associated with its type
+are evaluated, which may lead to new subgoals being generated. In our
+example, there could be a domain rule stating that if a Shopping Agent
+doesn't already own a product, then it must \texttt{Buy} it before it
+can \texttt{Own} it. In such a case, activating the
+\texttt{Own(Drill)} token will cause a new \texttt{Buy(Drill)} to be
+created in an \texttt{INACTIVE} state, which in turn could spawn its
+own subgoals, like being at the location where the product can be
+purchased before being able to buy it, and so on, until all domain
+rules are satisfied.
 
 \begin{figure} \centering
-  \includegraphics[scale=0.45]{figs/europa-pr-merge.jpg}
-  \caption{\small Timeline states after satisfying goal to be home by time 20 by merging with initial state}
-\label{fig:europapr:merge}
-\vskip+0.1cm
+  \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
+  \caption{\small State of timelines after satisfying a goal to own a
+    drill by time $10$ by activating corresponding token.}
+\label{fig:europapr:activate}
 \end{figure}
 
-The goal to be home by time 20 is satisfied by merging with the initial state that said that the Agent is home at time 0.
-The merge is possible because the tokens are of the same type \texttt{At}, and intersecting the domains for all of their variables
-doesn't yield an empty domain (which represents an inconsistency), in this case the token variables are the location, 
-which is \texttt{Home} in both cases, and the temporal variables (start, end, duration), which can be intersected without causing 
-an inconsistency. After the token representing the goal is merged onto the token representing the initial state, the resulting token
-represents not only the fact that the Agent must be \texttt{At(Home)} by time 0 but also \texttt{At(Home)} at time 20, this is reflected in 
-the new constraints on the token's start and end variables as depicted in Fig. \ref{fig:europaprmerge}.
+As Fig. \ref{fig:europapr:activate} shows the decision to merge the
+goal to be \texttt{At(Home)} by time $20$ with the initial state was
+not appropriate since the Agent will have to get away from
+\texttt{Home} to be able to buy the drill which is inconsistent with
+the constraints that resulted after the merge. As a result that
+decision will have to be retracted at some point and the alternative
+to activate or merge with a different token will have to be
+considered.
 
-When a token is activated, all domain rules associated with its type are evaluated,
-which may lead to new subgoals being generated. In our example, there
-could be a domain rule stating that if a  Shopping Agent doesn't already own a product,
-then it must \texttt{Buy} it before it can \texttt{Own} it. In such a case,
-activating the \texttt{Own(Drill)} token will cause a new
-\texttt{Buy(Drill)} to be created in an \texttt{INACTIVE}
-state, which in turn could spawn its own subgoals, like being at the location where the product
-can be purchased before being able to buy it, and so on, until all domain rules are satisfied.
+If, after looking at all possibilities, neither activating nor merging
+work, a token may be \texttt{REJECTED}. This is possible only for
+goals posted by the end user and specified as optional. Subgoals
+generated as a result of token activation cannot be rejected. If a
+subgoal cannot be satisfied through activation or merging then it will
+cause the original activation decision that spawned it to be
+retracted.
 
-\begin{figure} \centering
-  \includegraphics[scale=0.45]{figs/europa-pr-activate.jpg}
-  \caption{\small Timeline states after satisfying goal to own drill by time 10 by activating corresponding token}
-\label{fig:europapr:activate}
-\vskip+0.1cm
-\end{figure}
-
-It is easy to see in Fig. \ref{fig:europapr:activate} that the decision to merge the goal to be \texttt{At(Home)} by time 20 
-with the initial state was not a good one since the Agent will have to get away from \texttt{Home} to be able to 
-buy the drill which is inconsistent with the constraints that resulted after the merge. As a result that decision will have 
-to be retracted at some point and the alternative to activate or merge with a different token will have to be considered.
-
-If, after looking at all possibilities, neither activating nor merging work, a token may be
-\texttt{REJECTED}. This is possible only for goals posted by the end
-user and specified as optional. Subgoals generated as a result of
-token activation cannot be rejected. If a subgoal cannot be satisfied
-through activation or merging then it will cause the original
-activation decision that spawned it to be retracted.
-
-As can be seen from the above example, during the search for a plan, a token's state can go from
-\texttt{INACTIVE} to \texttt{ACTIVE, MERGED} or \texttt{REJECTED}, as
-shown in Fig. \ref{fig:europapr:states}.
+As can be seen from the above example, during the search for a plan, a
+token's state can go from \texttt{INACTIVE} to \texttt{ACTIVE, MERGED}
+or \texttt{REJECTED}, as shown in Fig. \ref{fig:europapr:states}.
 
@@
 \label{fig:europapr:states}
-\vskip+0.1cm
 \end{figure}
@@
   Shopping Agent precede and succeed tokens of the predicate
-  \texttt{Go} that represents traveling.
-
-  \begin{figure} \centering
-  \includegraphics[scale=0.45]{figs/europa-pr-rules.jpg}
-  \caption{\small Master and Slave tokens generated as a result of a Domain Rule}
-\label{fig:europapr:rules}
-\vskip+0.1cm
-\end{figure}
-
-  Fig. \ref{fig:europapr:rules} illustrates the entities and
-  relations involved in specifying such a rule on a \texttt{Go}
-  action. The token on which the rule applies  (\texttt{Go}) is referred to
- as the \emph{master}. Each \texttt{At} and \texttt{Going} token required
-  by the master is referred to as a \emph{slave}. 
-  Application of a rule on a token can thus cause slave
+  \texttt{Go} that represents traveling. Fig. \ref{fig:europapr:rules}
+  illustrates the entities and relations involved in specifying such a
+  rule on a \texttt{Go} action. The token on which the rule applies
+  (\texttt{Go}) is referred to as the \emph{master}. Each \texttt{At}
+  and \texttt{Going} token required by the master is referred to as a
+  \emph{slave}.  Application of a rule on a token can thus cause slave
   tokens, variables and constraints to be introduced.  The capability
   of a domain rule to cause a slave token to be created is a key
-  vehicle through which planning occurs. Semantically, a rule
-  imposes a requirement for supporting (slave) tokens to be in the
-  plan in order for the master token to be valid.
+  vehicle through which planning occurs. Semantically, a rule imposes
+  a requirement for supporting (slave) tokens to be in the plan in
+  order for the master token to be valid.
+
+  \begin{figure} 
+    \centering
+    \includegraphics[scale=0.4]{figs/europa-pr-rules.pdf}
+    \caption{\small Master and Slave tokens generated as a result of a domain rule.}
+    \label{fig:europapr:rules}
+  \end{figure}
 

### 2012-02-24 europa.tex
addressed Kanna's comments.

@@
 % always, all constraints in \eus library could be used.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\caption{\small Temporal relations defined within the planner are
+  based on \texttt{Allen Algebra} relations shown above.}
+\label{fig:allen-algebra}
+\end{figure}
+
+In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
+temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
+encoding of temporal constraint relationships between tokens. 
+%\kcomment{This para could do with a little
+  embellishment especially since it is the only material that actually
+  talks about the kinds of constraints and their origin.JB: I don't really know what else to say given the table above.}
 
@@
 collaborate to perform inference and prune variable domains using
-Bounds Propagation and Domain Reduction techniques \kcomment{citation
-  needed}. \eu uses a Propagator pattern which keeps track of a set of
+Bounds Propagation and Domain Reduction techniques \cite{marriott98}. 
+\eu uses a Propagator pattern which keeps track of a set of
 constraint instances of the same type and is responsible for ensuring
-that whenever any variable is modified or constraints are introduced
-or dropped, all related variables are updated through inference.
+variables are inferentially updated upon change (domain modifications or
+constraint addition/deletion).
+
+% that whenever any variable is modified or constraints are introduced
+% or dropped, all related variables are updated through inference.
 
@@
 
-Temporal and resource constraints are specially important to model
-real-life planning domains, so it is worth going into some more detail
-about how \eus built-in support works for those constraint types.
+\input{constraints-table}
 
-\textit{Temporal Constraints:} \eu uses variables to explicitly
-represent timepoints for plan activities and states. Constraints among
-timepoints provide a natural way to express domain axioms. For
-example, in order to state that activity $A$ must occur before
-activity $B$ we can say that the end timepoint of $A$ is $leq$ the
-start timepoint of $B$.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
-\label{fig:allen-algebra}
-\vskip-0.3cm
-\end{figure}
-
+Temporal and resource constraints are especially important to model
+real-world planning domains. % so it is worth going into some more
+% detail about how \eus built-in support works for those constraint
+% types.
+\eu uses variables to explicitly represent timepoints for plan
+activities and states. Constraints among timepoints provide a natural
+way to express domain axioms. For example, in order to state that
+activity $A$ must occur before activity $B$ we can say that the end
+timepoint of $A$ is $leq$ the start timepoint of $B$.
 \cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network (STN). Such a
-network can be transformed into a Distance Graph (DG) where the
-outward arc from a node represents the maximum distance from the
-source node to the target node.  The diagram below illustrates a
-simple STN with just 2 variables and a single constraint. It also
-shows the resulting DG.
-
-reflected by the use of the \texttt{Allen Algebra} \cite{allen84}
-temporal relations shown in Fig. \ref{fig:allen-algebra}. Interval
-arithmetic in the EUROPA$_2$ domain model using \texttt{Allen Algebra}
-allows encoding of temporal constraint relationships between tokens on
-concurrent timelines. See \cite{Jonsson00} for additional details
-including propagation algorithms and interval logic.
+grouped together to form a Simple Temporal Network
+(\texttt{STN}). Such a network can be transformed into a Distance
+Graph (\texttt{DG}) where the outward arc from a node represents the
+maximum distance from the source node to the target node.
+Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
+a single constraint along with the resulting \texttt{DG}.
+\cite{dechter91} also showed that shortest path algorithms could be
+used to propagate values in the network and discover a negative cycle:
+namely a path from a node to itself that has a path length less than
+$0$. If such a cycle exists, the network is inconsistent. It was
+further shown that a single-source shortest path algorithm was
+sufficient to detect a negative cycle and provide sufficient
+propagation to yield a backtrack-free search. \eus propagator for
+temporal constraints is a straight implementation of this algorithm.
 
@@
 
-\cite{dechter91} also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative
-cycle. A negative cycle is a path from a node to itself that has a
-path length less than 0. If such a cycle exists, the network is
-inconsistent. It was further shown that a single-source shortest path
-algorithm was sufficient to detect a negative cycle and provide
-sufficient propagation to yield a backtrack-free search. Thus we have
-an efficient and complete algorithm for propagating an STN. \eus
-Propagator for temporal constraints is a straight implementation of
-this algorithm.
+Most real-world planning problems need to deal with plans made up of
+actions that typically depend upon limited resources (\eg fuel,
+storage capacity, battery state of charge, data buffers).  \eu
+provides built-in support for resource computation. In \eus
+representation resource object types come in three forms with related
+constraints. \textbf{Reservoir} resources can have both
+\texttt{consume(resource,time,quantity)} and
+\texttt{produce(resource,time,quantity)} constraints; production and
+consumption is instantaneous. \textbf{Reusable} resources on the other
+hand, have constraints that use a quantity of the resource for a
+specific duration (\texttt{uses(resource,start,duration,quantity)}),
+which are then subsequently released. \textbf {Unary} resources are
+reusable with capacity$ = 1$, which allows \eu to perform run time
+savings in comparison to the generic reusable class. Configuration
+options are offered such that in addition to the usual constraint
+violations when a resource is completely depleted, plan flaws can be
+generated whenever a resource level is outside a specific upper or
+lower bound, or when instantaneous production or consumption exceeds a
+specific threshold.
 
-\textit{Resource Constraints:} Most real life planning problems have
-to deal with the fact that plans made up of actions that require
-resources to be performed typically have only a limited amount of
-those resources available (think of fuel, storage capacity, etc).  \eu
-provides built-in support for resources whose availability can be
-measured with real numbers, in terms of \eus representation, resource
-object types and their related constraints are provided as follows:
-
-\begin{enumerate}
-
-\item \textbf{Reservoir} resources can have both
-  \texttt{consume(resource,time,quantity)} and
-  \texttt{produce(resource,time,quantity)} constraints. Production and
-  consumption is instantaneous (it happens at a single point in time)
-
-\item \textbf{Reusable} resources have
-  \texttt{uses(resource,start,duration,quantity)} constraints that use
-  a quantity of the resource for a specific duration, then release it.
-
-\item \textbf {Unary} resources are Reusable resources with
-  capacity=1, which allows \eu to perform some run time savings
-  compared to the generic Reusable class.
-
-\end{enumerate}
-
-\eu offers configuration options so that in addition to the usual
-constraint violations when a resource is completely depleted, plan
-flaws can be generated whenever a resource level is outside a specific
-upper or lower limit, or when instantaneous production or consumption
-exceeds a specific threshold.
-
-Recall that \eu supports a flexible temporal representation, where
-time is represented by intervals, instead of single values. In this
-context, determining whether a resource is the source of a plan flaw
-or a constraint violation in general is more difficult than just
+Since \eu supports a flexible temporal representation, where time is
+represented by intervals, instead of single values, determining
+whether a resource is the source of a plan flaw or a constraint
+violation in general is substantially more difficult than just
 computing a profile of the resource and checking against some limits.
-In the context of time flexibility, there are many resource profiles
-that are possible, depending on when exactly each
-production/consumption event happens out of all the possible values
-that are in the domain of the variables that represent the time for
-the event.  \cite{mus02} proposed an algorithm where the variables
-that represent the time for production/consumption events are placed
-in a temporal network, similar to the network used by the Temporal
-Propagator above, then the constraints on the resource can be seen as
+With temporal flexibility, a number of resource profiles could be
+possible depending on precisely when each production/consumption event
+occurs out of a range of possible values that are in the domain of the
+variables that represent the temporal extant for the event.
+\cite{Muscettola04,Muscettola06} proposed an algorithm where the
+variables that represent the time for production/consumption events
+are placed in a temporal network, similar to the network used by the
+temporal propagator; constraints on the resource can then be seen as
 quantities flowing through that network. Computing the maximum flow
-will yield the maximum resource usage, and thus a lower bound on the
-resource level, and computing a minimum flow will similarly yield an
-upper bound. Since maximum flow on a temporal network can be computed
-in polynomial time, \eu takes advantage of this algorithm to
-efficiently compute upper and lower bound envelopes on the resource
-level.
+then, yields the maximum resource usage and thus a lower bound on the
+resource level; computing a minimum flow will similarly yield an upper
+bound. Since maximum flow on a temporal network can be computed in
+polynomial time, \eu takes advantage of this algorithm to efficiently
+compute upper and lower bound envelopes on the resource
+level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
+computation.
 
@@
 
-Given the above elements, an automated problem solver can be
-created using:
+Given the above, an automated problem solver can be created using, a
+\emph{domain model} that describes the variable, object, predicate and
+action types that are relevant for the problem, a \emph{problem
+  instance} or initial state that consists of variable and object
+instances that exist for the entire planning horizon, \emph{temporally
+  scoped predicate and action} instances and \emph{temporally scoped
+  goals}. This information is retained in \eus \texttt{PDB} so that
+inference and search mechanisms can be used to look for a problem
+solution.
 
-\begin{enumerate}
+Since temporal intervals may extend indefinitely in both positive and
+negative directions, for problems with a temporal dimension it is also
+common to specify a \emph{planning horizon} when invoking a solver;
+any decisions that fall completely outside of the horizon can be
+ignored. This is particularly useful for recurrent activities;
+consider for instance a model where a medical checkup has to occur
+every $3$ months. The user can model a recurring activity and then use
+the horizon to ensure that a finite number of medical checkup
+activities are generated as part of a plan. This is also useful to
+ignore activities that may occur far in the future or in the past to
+be relevant for a particular plan, ensuring that the solver is making
+as few commitments as necessary and making planning tractable.
 
-\item A domain model that describes the variable, object, predicate
-  and action types that are relevant for the problem.
+\subsubsection{Flaw Resolution}
+\label{sec:europa:flaws}
 
-\item A problem instance (also called initial state) that consists of:
-
-    \begin{enumerate}
-
-    \item Variable and Object instances that exist for the entire
-      planning horizon.
-
-    \item Temporally scoped predicate and action instances.
-
-    \item Temporally scoped goals.
-
-\end{enumerate}
-
-\end{enumerate}
-
-This information is kept in \eus Plan Database so that inference and
-search mechanisms can be used to look for a problem solution.
-
-Given that temporal intervals may extend indefinitely in both positive
-and negative directions, for problems with a temporal dimension it is
-also common to specify a Planning Horizon when invoking a solver. The
-Planning Horizon is the temporal interval that a problem solver has to
-worry about, any decisions that fall completely outside of the
-Planning Horizon can be ignored. This is very useful for recurrent
-activities, consider for instance a model where a medical checkup has
-to occur every 3 months, the user can model a recurring activity and
-then use the Planning Horizon to ensure that a finite number of
-medical checkup activities are generated as part of a plan. This is
-also useful to ignore activities that may happen too far in the future
-or in the past to be relevant for a particular plan, this way the user
-can ensure that the solver is making as few commitments as necessary.
-
-\eu provides a built-in solver that performs Plan Space Planning
-\cite{ghallab04} in this approach, the initial state is considered a
-partial plan that needs to be refined toward a solution plan that
-achieves the goals. The operations to refine the partial plan PP at
+\eu provides a built-in solver that performs \emph{Plan Space
+  Planning} \cite{ghallab04}; the initial state is considered a
+partial plan that needs to be refined toward a solution that achieves
+stated goals. The operations to refine the partial plan \texttt{PP} at
 any time are:
@@
 
-This Plan Space Planning algorithm is translated into \eus
-representation as follows:
+% This Plan Space Planning algorithm is translated into \eus
+% representation as follows:
 
 The initial partial plan is the state of \eus Plan Database after the
-initial state has been instantiated, this results in a set of
+initial state has been instantiated. This results in a set of
 variable, object and token instances. Inference then takes place to
-detect flaws in the partial plan, there are 3 kinds of flaws that can
-be detected:
+detect flaws in the partial plan; the three kinds of flaws that can be
+detected are:
 
-\begin{enumerate}
+\begin{description}
 
-\item Unbound Variable: a variable in the partial plan whose domain is
-  not a singleton. Unbound Variables are resolved by specification of
-  a value from the domain of the variable.
+\item[\textbf{Unbound Variable}]: a variable in the partial plan whose
+  domain is not a singleton. Unbound Variables are resolved by value
+  selection from their domain.
 
-\item Open Condition: an open condition is an inactive token. Inactive
-  tokens can be generated by explicitly posted goals, or when a token
-  is activated, rules may fire that create inactive slave tokens as
-  described in section \comment{point to EUROPA Plan Representation
-    section}.  Open Conditions can be resolved in three ways:
+\item[\textbf{Open Condition}]: an open condition is an
+  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
+  by explicitly posted goals, or when a token is activated, rules may
+  activate the creation of \texttt{INACTIVE} slave tokens as described
+  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
+  three ways:
 
-  \begin{enumerate}
+  \begin{description}
 
-  \item Merging: A free activity is merged with a matching activity
-    already in the plan. An activity a is said to match an activity a?
-    if a and a? unify and the temporal constraints involving a are
-    satisÞed by a?. Thus, a and a? can be considered the same activity
-    and we do not need to introduce a in the plan. Consequently, the
-    compatibilities associated with a are not Þred, because they have
-    been already triggered when a? was introduced in the plan.
+  \item[\textbf{Merging}]: A free activity is merged with a matching
+    activity already in the plan. An activity $a$ is said to match an
+    activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig} and the
+    temporal constraints involving $a$ are satisfied by $a?$. Thus,
+    $a$ and $a?$ can be considered the same activity without the need
+    to introduce $a$ in the plan. Consequently, the domain rules associated with
+    $a$ are not introduced into the planning process, because they have been already
+    triggered when $a?$ was introduced in the plan.
 
-  \item Activation: We introduce a new activity a in the current plan
-    associating it with the proper timeline, but without choosing a
-    speciÞc time slot for it.The compatibilities associated with a are
-    applied and the subgoal activities resulting from those
-    compatibilities are introduced as free activities. This results in
-    both an ordering flaw, corresponding to the just activated
-    activity, and a number of open condition flaws, corresponding to
+  \item[\textbf{Activation}]: A new activity $a$ is introduced in the
+    current plan associating it with the correct timeline, without
+    choosing a specific time slot (if its timing causes a problem 
+    it'll trigger a Threat flaw, explained below). 
+    The domain rules associated with $a$ are applied and the subgoal
+    activities resulting from those domain rules are introduced as inactive tokens. 
+    This results in a number of open condition flaws, corresponding to
     the new subgoal activities.
 
-  \item Rejection: Some goals may be optional, if that is the case for
-    a particular order condition, its corresponding flaw can be
-    resolved by discarding the inactive token.
+  \item[\textbf{Rejection}]: Some goals may be optional; if so its
+    corresponding flaw may be resolved by discarding the
+    \texttt{INACTIVE} token.
 
-\end{enumerate}
+  \end{description}
 
-\item Threat: once a token has been placed in the partial plan it may
-  impact other tokens indirectly through possible overlapping
-  requirements on objects. Recall for example that a token may belong
-  to objects (e.g. Timelines) which require a total order over their
-  tokens. If any 2 tokens could possibly overlap (though not
-  necessarily), then they pose a threat to each other in terms of
+\item[\textbf{Threat}]: Once a token has been placed in the partial
+  plan it may impact other tokens indirectly through possible
+  overlapping requirements on objects. Recall for example that a token
+  may belong to objects (\eg timelines) which require a total order
+  over their tokens. If any two tokens could possibly overlap (though
+  not necessarily), then they pose a threat to each other in terms of
   achieving an extension of the current partial plan which is complete
@@
 
-\end{enumerate}
+\end{description}
 
-Open Conditions and Threats allow flaw detection and resolution at a
-higher-level of abstraction (i.e. in terms of objects and tokens) than
-that of simply binding variables as is common in Constraint
-Satisfaction Problems. This is advantageous when one applies
-heuristics for ordering choices since it provides a richer context in
-which to make decisions.  Furthermore, it aids in reducing the amount
-of work done by a solver so that only the necessary refinements are
-made, otherwise leaving the partial plan with maximum flexibility. For
-example, one can omit unbound variables which are time-points of
-tokens since threats will force a solver to impose restrictions on
-these variables based on the semantics of the objects to which their
-tokens apply. Thus the planning process may yield partially-ordered
-plans for which all possible extensions are provably valid.
+Open conditions and threats allow flaw detection and resolution at a
+higher-level of abstraction (\ie in terms of objects and tokens) than
+that of binding variables as is common in Constraint Satisfaction
+Problems. This is advantageous when a solver applies heuristics for
+ordering choices since it provides a richer context in which to make
+decisions.  Furthermore, it aids in reducing the amount of work done
+by a solver so that only the necessary refinements are made leaving a
+partial plan with sufficient flexibility. For example, one can bypass
+committing unbound timepoint variables since threats will force a
+solver to impose ordering restrictions on these variables.  Thus the
+planning process may yield partially-ordered plans for which all
+possible extensions are provably valid.
 
 The Plan Space Planning algorithm described above can be implemented
-in many different ways depending on the approach chosen for flaw and
+in multiple ways depending on the approach chosen for flaw and
 resolver selection and for backtracking. \eus built-in solver

### 2012-02-24 trunk/concepts.tex
updates to arch and results mostly cosmetic

@@
 
-  \begin{figure} \centering
-    \includegraphics[scale=0.4]{figs/europa-pr-1.pdf}
-    \caption{\small Timelines representing initial state and goals for a Shopping Agent.}
-    \label{fig:europapr1}
+  \begin{figure}[t]
+    \centering \subfloat[\small Timelines representing initial state
+    and goals for a Shopping
+    Agent.]{\label{fig:europapr1}\includegraphics[scale=0.35]{figs/europa-pr-1.pdf}}\qquad
+    \subfloat[\small State of timelines after satisfying a goal to be
+    home by time $20$ on merging with initial
+    state.]{\label{fig:europapr:merge}\includegraphics[scale=0.35]{figs/europa-pr-merge.pdf}}\qquad
+    \subfloat[\small State of timelines after satisfying a goal to own
+    a drill by time $10$ by activating corresponding
+    token.]{\label{fig:europapr:activate}\includegraphics[scale=0.35]{figs/europa-pr-activate.pdf}}
+    \caption{\small Evolution of a timeline from the Shopping Agent example}
   \end{figure}
@@
 
-\begin{figure} 
-  \centering
-  \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
-  \caption{\small State of timelines after satisfying a goal to be
-    home by time $20$ on merging with initial state.}
-  \label{fig:europapr:merge}
-\end{figure}
+% \begin{figure} 
+%   \centering
+%   \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
+%   \caption{\small State of timelines after satisfying a goal to be
+%     home by time $20$ on merging with initial state.}
+%   \label{fig:europapr:merge}
+% \end{figure}
 
@@
 
-\begin{figure} \centering
-  \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
-  \caption{\small State of timelines after satisfying a goal to own a
-    drill by time $10$ by activating corresponding token.}
-\label{fig:europapr:activate}
-\end{figure}
+% \begin{figure} \centering
+%   \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
+%   \caption{\small State of timelines after satisfying a goal to own a
+%     drill by time $10$ by activating corresponding token.}
+% \label{fig:europapr:activate}
+% \end{figure}
 

### 2012-02-24 trunk/europa.tex
updates to arch and results mostly cosmetic

@@
 
-\item[\textbf{Object Types}]: \texttt{NDDL} supports an object-oriented
+\item[\textbf{Object Types}:] \texttt{NDDL} supports an object-oriented
   approach, where common data and behavior can be abstracted as a
@@
 
-\item[\textbf{Variable Types and Instances}]: All object attributes
+\item[\textbf{Variable Types and Instances}:] All object attributes
   are represented as variables in this model. For instance, it uses
@@
 
-\item[\textbf{Predicate Types}]: Some classes like \textit{Product},
+\item[\textbf{Predicate Types}:] Some classes like \textit{Product},
   \textit{Location} and \textit{ProductLocation} are static and so
@@
 
-\item[\textbf{Action Types}]: In our example, \textit{Agent} is solely
+\item[\textbf{Action Types}:] In our example, \textit{Agent} is solely
   the class that has behavior associated with it. Consequently, it
@@
 
-\item[\textbf{Variables}]: A token type (predicate or action) is part
+\item[\textbf{Variables}:] A token type (predicate or action) is part
   of a class, therefore class attributes (see \texttt{object.location}
@@
 
-\item[\textbf{Constraints}]: Restrictions on values that variables can
+\item[\textbf{Constraints}:] Restrictions on values that variables can
   take within a rule are expressed through constraints. In the example
@@
 
-\item[\textbf{Subgoals}]: In \texttt{Agent::Go}, subgoals are used to
+\item[\textbf{Subgoals}:] In \texttt{Agent::Go}, subgoals are used to
   specify that the Shopping Agent's location must transition from
@@
 \caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
+  based on \texttt{Allen Algebra} \cite{allen84} relations shown
+  above.}
 \label{fig:allen-algebra}
@@
 temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
-encoding of temporal constraint relationships between tokens on
-concurrent timelines. \kcomment{This para could do with a little
-  embellishment especially since it is the only material that actually
-  talks about the kinds of constraints and their origin.}
+encoding of temporal constraint relationships between tokens.
+\kcomment{This para could do with a little embellishment especially
+  since it is the only material that actually talks about the kinds of
+  constraints and their origin.JB: I don't really know what else to
+  say given the table above.}
 
@@
 collaborate to perform inference and prune variable domains using
-Bounds Propagation and Domain Reduction techniques \kcomment{citation
-  needed}. \eu uses a Propagator pattern which keeps track of a set of
+Bounds Propagation and Domain Reduction techniques \cite{marriott98}. 
+\eu uses a Propagator pattern which keeps track of a set of
 constraint instances of the same type and is responsible for ensuring
-variables are inferentially updated with change (modifications or
+variables are inferentially updated upon change (domain modifications or
 constraint addition/deletion).
@@
 
-\item[\textbf{Unbound Variable}]: a variable in the partial plan whose
+\item[\textbf{Unbound Variable}:] a variable in the partial plan whose
   domain is not a singleton. Unbound Variables are resolved by value
@@
 
-\item[\textbf{Open Condition}]: an open condition is an
+\item[\textbf{Open Condition}:] an open condition is an
   \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
@@
 
-  \item[\textbf{Merging}]: A free activity is merged with a matching
-    activity already in the plan. An activity $a$ is said to match an
-    activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig} and the
-    temporal constraints involving $a$ are satisfied by $a?$. Thus,
-    $a$ and $a?$ can be considered the same activity without the need
-    to introduce $a$ in the plan. Consequently, the compatibilities
-    \kcomment{compatibilities have not been defined} associated with
-    $a$ are not Þred \kcomment{???}, because they have been already
-    triggered when $a?$ was introduced in the plan.
+  \item[\textbf{Merging}:] A free activity \kcomment{this could be
+      defined, since no def. for 'free' is provided} is merged with a
+    matching activity already in the plan. An activity $a$ is said to
+    match an activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
+    and the temporal constraints involving $a$ are satisfied by
+    $a?$. Thus, $a$ and $a?$ can be considered the same activity
+    without the need to introduce $a$ in the plan. Consequently, the
+    domain rules associated with $a$ are not introduced into the
+    planning process, since they are already triggered when $a?$ was
+    introduced in the plan.
 
-  \item[\textbf{Activation}]: A new activity $a$ is introduced in the
-    current plan associating it with the correct timeline, without
-    choosing a specific time slot. The compatibilities \kcomment{as
-      above} associated with $a$ are applied and the subgoal
-    activities resulting from those compatibilities \kcomment{as
-      above} are introduced as free activities \kcomment{what is a
-      free activity?}. This results in both an ordering flaw
-    \kcomment{undefined}, corresponding to the just activated
-    activity, and a number of open condition flaws, corresponding to
-    the new subgoal activities.
+  \item[\textbf{Activation}:] A new activity $a$ is introduced in the
+    current plan associating it with the appropriate timeline, without
+    choosing a specific time slot; if its timing causes a problem a
+    Threat flaw will be triggered (see below).  The domain rules
+    associated with $a$ are applied and the subgoal activities
+    resulting from those domain rules are introduced as inactive
+    tokens.  This results in a number of open condition flaws,
+    corresponding to the new subgoal activities.
 
-  \item[\textbf{Rejection}]: Some goals may be optional; if so its
-    corresponding flaw can be resolved by discarding the
+  \item[\textbf{Rejection}:] Some goals may be optional; if so its
+    corresponding flaw may be resolved by discarding the
     \texttt{INACTIVE} token.
@@
 
-\item[\textbf{Threat}]: Once a token has been placed in the partial
+\item[\textbf{Threat}:] Once a token has been placed in the partial
   plan it may impact other tokens indirectly through possible
@@
 
-\item[Line \textbf{\ref{li:propagate}}]: Propagate the constraints to
+\item[Line \textbf{\ref{li:propagate}}:] Propagate the constraints to
   test for inconsistency. If inconsistent, then return false since no
@@
 
-\item[Line \textbf{\ref{li:selectflaw}}]: Choose a flaw from the set
+\item[Line \textbf{\ref{li:selectflaw}}:] Choose a flaw from the set
   of available flaws.
 
-\item[Line \textbf{\ref{li:noflaw}}]: If there are no flaws, then $p$
+\item[Line \textbf{\ref{li:noflaw}}:] If there are no flaws, then $p$
   is complete and terminate with success.
 
-\item[Line \textbf{\ref{li:decisionpt}}]: Formulate a decision point
+\item[Line \textbf{\ref{li:decisionpt}}:] Formulate a decision point
   which is a branch in the search space. Each choice is a particular
@@
 
-\item[Line \textbf{\ref{li:decisionloop}}]: Keep trying until the
+\item[Line \textbf{\ref{li:decisionloop}}:] Keep trying until the
   chosen flaw is resolved, or until there are no further resolvers to
@@
 
-\item[Line \textbf{\ref{li:decide}}]: A new partial plan is obtained
+\item[Line \textbf{\ref{li:decide}}:] A new partial plan is obtained
   by application of a refinement operator. Note that the ordering over
@@
 
-\item[Line \textbf{\ref{li:recurse}}]: Recursive call to solve the
+\item[Line \textbf{\ref{li:recurse}}:] Recursive call to solve the
   new planning problem. If successful, then success.
 
-\item[Line \textbf{\ref{li:backtrack}}]: Otherwise, retract the last
+\item[Line \textbf{\ref{li:backtrack}}:] Otherwise, retract the last
   refinement operation and move on to try the next one.
 
-\item[Line \textbf{\ref{li:noplan}}]: Arriving here implies all
+\item[Line \textbf{\ref{li:noplan}}:] Arriving here implies all
   options to resolve the flaw have been exhausted, including the case

### 2012-02-24 trunk/results.tex
updates to arch and results mostly cosmetic

@@
 
-\begin{figure}[htpb]
+\begin{figure}[b]
 \centering
@@
 
-\begin{figure}[htpb]
+\begin{figure}[t]
 \centering
@@
   \centering
-  \subfloat[A conceptual view of the \od]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
-  \subfloat[Current Implementation for
+  \subfloat[A conceptual view of the \od and how it fits into \can
+  field experiments]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+  \subfloat[Current implementation of \od for
   \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
@@
   surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
-  deployments. \ref{fig:odss-ipad} shows \od use on a portable iPad
-  device.}
+  deployments. \ref{fig:odss-ipad} shows \od use on an iPad.}
 \label{fig:odss2}

### 2012-02-24 europa-basics.tex
updates to arch and results mostly cosmetic

@@
 
-\begin{figure} \centering
+\begin{figure}[b]
+\centering
   \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
@@
 \centering
-\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small $N$-Queens solution generated by \eu}
+\includegraphics[scale=0.3]{figs/Example-NQueens1.jpg}
+\caption{\small An $N$-Queens solution generated by \eu}
 \label{fig:nqueens-2}
@@
   effects, dependencies at the action level) is not explicitly
-  maintained by the \textsf{CP} planner, the search algorithms are deprived of
-  critical information to make better decisions. If that structure is
-  maintained (for instance, by internally marking variables that
-  represent action choice and relationships between actions), it is
-  still hard to write search algorithms as any planning-specific
-  insight has to be translated into the \textsf{CP} representation of
-  variables and constraints.
+  maintained by the \textsf{CP} planner, the search algorithms are
+  deprived of critical information to make better decisions. If that
+  structure is maintained (for instance, by internally marking
+  variables that represent action choice and relationships between
+  actions), it is still hard to write search algorithms as any
+  planning-specific insight has to be translated into the \textsf{CP}
+  representation of variables and constraints.
 
@@
 and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
-advantage of a \textsf{CP} representation to deal with time and resources, at
-the same time maintaining an explicit representation of the elements
-relevant to planning (action choices, causal relationships, etc). This
-combination results in a planning approach that can deal with the type
-of constraints found in real-world problems while allowing for
-intuitive problem descriptions and a natural way to express search
-algorithms and heuristics. \eu is one such implementation of the
-\texttt{CAIP} framework, in which planning problems are specified in
-terms of:
+advantage of a \textsf{CP} representation to deal with time and
+resources, at the same time maintaining an explicit representation of
+the elements relevant to planning (action choices, causal
+relationships, etc). This combination results in a planning approach
+that can deal with the type of constraints found in real-world
+problems while allowing for intuitive problem descriptions and a
+natural way to express search algorithms and heuristics. \eu is one
+such implementation of the \texttt{CAIP} framework, in which planning
+problems are specified in terms of:
 
-\begin{enumerate}
+\begin{description}
 
-\item \textbf{Intervals}: entities that have a temporal scope (that
+\item[\textbf{Intervals}:] entities that have a temporal scope (that
   is, a beginning and an end in time) and a set of
@@
 
-\item \textbf{Variables and Constraints}: Variables represent Interval
+\item[\textbf{Variables and Constraints}:] Variables represent Interval
   attributes and constraints enable a direct representation of
@@
 
-\item \textbf{Domain Configuration Rules}: This is an explicit
+\item[\textbf{Domain Configuration Rules}:] This is an explicit
   representation of the conceptual relationships between actions and

### 2012-02-24 trunk/future.tex
Added few paragraphs in future work more related to T-REX refinements

@@
   \includegraphics[scale=0.45]{figs/autonomy-topics.pdf}
-  \caption{\small }
+  \caption{\small Overlapping topics in our research in deliberation
+    and autonomy with sampling the dominant theme.}
   \label{fig:topics}
@@
 
+Many aspect remain to be fully explored in how to improve \rx and its
+integration with \eu. We present here  few of these that we consider
+as important in order to improve \rx abilities and reduce the
+engineering required to circonvey them.
+
+One of them is related to the dispatchability of
+the {\em external} part of the plan \cite{mus98a}. While we were able
+to limit its impact so far by adding extra controls inside the \eu
+model we know that it can be impoved. For example by analyzing the
+plan structure we dshould be able to distinguish part of the plan that
+contribute to a  reactor goal -- which often needs to be dispatched
+as early as possible -- to other tokens describing the world evolution
+for which it can be preferrable to dispatch as late as needed. Being
+able to distinguish these two kinds of tokens should largely improve
+the quality of the overal agent behavior while avoiding to add extra
+complexity on every reactor models.
+
+At the architecture level we would like to take full benefits of the
+reactor graph structure maintained by the agent and how to take full
+benefits. One aspect, we need to improve further is the ability to
+alter this graph dynamically by creating new reactors or new timeline
+connections. Offering such abilty would allow to provide better
+reovery mechanism after one reactor has failed to sunchronize by the 
+agent. For example, one alternate reactor could take ownership of the
+timelines left invalid by the destruction of the faulty reactor. This
+would allow to dimplement in atrasnparent way components
+redudancy. Still, even though our implementation already allow such
+mechanism we do not provide yet this at \rx API level as we need to
+ensure that this kind of dynamic alteration of the graph can be done
+wothout breaking our current assumptions both at \rx level and \eu
+level.
+
+More specifically to \eu level we would like to improve \rx
+integration of \eu ability to deliberate about resources. More
+specifically we need to ba able to update resource level as 
+modeled by \eu based on \rx observation.  As of today the 
+support of resource is limited inside the \eu solver and there's no
+direct way for \rx to update them dynamically based on recent
+observations. This result on a potential discrepeancy between measured
+given by, for example, the battery as modeled in \eu and the current
+battery level as read in a \rx timeline. Finding an automated way to
+link these two constructs would allow the plan to dynamically readapt
+when these two values differs -- for example if the vehicle looses a
+battery bank during a mission.
+
 % multi-vehicle control
@@
 % scalable architecture for missions
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "setobook"
+%%% End:

### 2012-02-24 trunk/results.tex
Added few paragraphs in future work more related to T-REX refinements

@@
 The novelty in this application was two-fold; in enabling the
-deliberation to decide when to decide on the phase changes and in
+deliberation to decide when to trigger the phase changes and in
 showing viably how a human being coupled with a goal-oriented
@@
 drifting instrument, the vehicle is expected to move around it. While
-a number of different patterns have been tested in simulation, the
+our model suppirt multiples patterns(square, lawn-mower,
+transect(s)\dots) with any possible dimension factor, the
 majority of at-sea deployments have used a square pattern ($3X3$

### 2012-02-24 arch.tex
[arch] moved synch failure example earlier in the text [result] reword paragraph about gulper in order to enlarge the motivation from simply Gulper to "sensor based" missions
Made failure handling as a subsubsection

@@
 
-\begin{figure}[!htb]
+\begin{figure}[t]
   \centering
@@
 
+\label{sec:delib-exec-as}
 \begin{figure}[!htbp]
@@
 
+\subsubsection{Reactors failure handling in \rx}
+\label{sec:rx-reactor-failure}
 It is important to note that, in T-REX architecture, the only critical

### 2012-02-24 arch.tex
added figures to illustrate reactor failure to synchronize

@@
 \texttt{Sampling\_policy} and \texttt{Relevant\_data} state variables
-produced the \texttt{Failed} observation. Therefore, the
+produced the \texttt{Failed} observation as illustrated in in
+Fig. \ref{fig:fail0}. Therefore, the
 \texttt{Survey controller} will received this observation on its {\em
@@
   navigate as \texttt{Pilot} is available which allow him to continue
-  this aspect of its current plan.
+  this aspect of its current plan. In such case the agent structure
+  remains as in Fig. \ref{fig:fail0} with no further impact.
 \item \texttt{Survey controller} models is not able to handle this
@@
   the \texttt{Failed} observation is produced on the \texttt{survey}
-  timeline.
+  timeline (Fig. \ref{fig:fail1}).
 \end{enumerate}
+\begin{figure}[!htb]
+  \centering
+  \subfloat[\small When \texttt{Sampling} reactor fails to
+  synchronize its removed from the graph and all its {\em internal}
+  timelines produces the \texttt{Failed} observation]{\includegraphics[width=0.4\columnwidth]{figs/AUV-fail0.pdf}
+  \label{fig:fail0}
+  }
+  \qquad
+  \subfloat[\small In the worst case this \texttt{Failed} may imply
+  other reactors hat depends on \texttt{Sampling} to  fail to
+  synchronize implying a bottom-up destruction of faulty reactors.]{
+    \includegraphics[width=0.4\columnwidth]{figs/AUV-fail1}
+    \label{fig:fail1}
+  }
+  \caption{Impact of the \texttt{Sampling} reactor failure in the AUV
+    agent from Fig. \ref{fig:agent}}
+\end{figure}
+
 Even on the worst case scenario presented on this second case, we

### 2012-02-24 results.tex
Corrected missed references
corrected a typo

@@
 A significant driver of the applicability of our work was the
+growing need for AUVs mission to be driven by scientific sensor 
+data.This trend is illustrated by the 
 implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
@@
 \cite{rainville01,pinkel11} are resulting in new modalities of making
-Lagrangian and Eulerian observations. Fig. \ref{sep11-wirewalker}
+Lagrangian and Eulerian observations. Fig. \ref{fig:sept11-wirewalker}
 shows one such experiment in September 2011 with a GPS-enabled
@@
   \centering
-\subfloat[The \od Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
-\subfloat[Current Implementation for \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
+  \subfloat[A conceptual view of the \od and how it fits into \can
+  field experiments]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+  \subfloat[Current implementation of \od for
+  \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
 \caption{\small The Oceanographic Decision Support System (\od) is a
@@
 \subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
-\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
-  the \od in use. Fig. \ref{fig:odss-virtual} shows a drifter tracking
-  tool used during \can planning; it used a physical model of advected
+\caption{\small \ref{fig:odss-mtg} shows a 2010 \can meeting with the
+  \od in use. \ref{fig:odss-virtual} shows a drifter tracking tool
+  used during \can planning; it used a physical model of advected
   surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
-  deployments. Fig. \ref{fig:odss-ipad} shows \od use on a portable
-  iPad device.}
+  deployments. \ref{fig:odss-ipad} shows \od use on an iPad.}
 \label{fig:odss2}

### 2012-02-24 europa.tex
addressed pending items from our last conversation. a number of fixes after a final review.

@@
 object, variable and token instances. This constitutes the data
-foundation on which end user applications and automated planners
-within \eu are built. The \textbf{Rules Engine} (\texttt{RE})
+foundation on which end user applications and automated problem solvers
+ are built. The \textbf{Rules Engine} (\texttt{RE})
 modularly adds the semantics of domain rules that describe
@@
 with specific problem instances expressed in terms of objects,
-variables, constraints, facts and goals. To articulate important
+variables, constraints, facts and goals. To articulate the most important
 elements of \texttt{NDDL}, we show how the Shopping Agent problem
@@
 
-In the model above one can see the main elements of a problem domain
+In the model above one can see how the main elements of a problem domain
 can be mapped to \eus representation:
@@
 
-\item[\textbf{Variable Types and Instances}]: All object attributes
+\item[\textbf{Variable Types and Instances}:] All object attributes
   are represented as variables in this model. For instance, it uses
   integer variables to represent the built-in temporal parameters for
-  each token (activity duration, start and end times), string
+  each token (duration, start and end times), string
   variables to represent \textit{Location} and \textit{Product} names,
@@
   \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
-  \textit{Product}. % We'll see how the actual behavior for those
-  % actions is specified below.
+  \textit{Product}.
 
@@
   \texttt{[condition, effect]} annotations can be used when defining a
-  subgoal inside an action. Currently \eu does not associate semantics
-  with these annotation; but they can be extremely useful for search
+  subgoal inside an action. \eu does not intrinsicaly associate semantics
+  with these annotations; but they can be extremely useful for search
   modules when creating plans.
@@
 
-\label{eu:model}
 \end{verbatim}
@@
 want to plan for. Facts are used to specify the current state of the
-world and do not need any support and are represented as tokens in the
+world, do not need any support and are represented as tokens in the
 \texttt{PDB}. In this case we use facts to specify that the Shipping
@@
 invoked to resolve those flaws as part of the deliberation
-process. Finally, constraints are restrictions on and between actions
+process. Finally, constraints are restrictions on and between facts 
 and goals. In the example above, we use temporal constraints to
 specify when facts hold and when the goals should be
-accomplished. % As
-% always, all constraints in \eus library could be used.
+accomplished. In the model above, \texttt{before} and \texttt{after} 
+are \texttt{Allen Algebra} \cite{allen84} operators, which \eu supports to specify
+temporal relations between tokens (specifically between their start and end timepoints)
+as summarized in Fig. \ref{fig:allen-algebra}
 
@@
 \caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} relations shown above.}
+  based on \texttt{Allen Algebra} \cite{allen84} relations shown
+  above.}
 \label{fig:allen-algebra}
 \end{figure}
-
-In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
-temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
-encoding of temporal constraint relationships between tokens. 
-%\kcomment{This para could do with a little
-  embellishment especially since it is the only material that actually
-  talks about the kinds of constraints and their origin.JB: I don't really know what else to say given the table above.}
 
@@
 constraint instances of the same type and is responsible for ensuring
-variables are inferentially updated upon change (domain modifications or
-constraint addition/deletion).
-
-% that whenever any variable is modified or constraints are introduced
-% or dropped, all related variables are updated through inference.
+variables are updated through inference whenever a relevant change occurs
+ (variable domain modifications or constraint addition/deletion).
 
@@
 relating them to the rest of the plan through constraints, by changing
-the start time of an activity, etc), or from an automated planner (by
+the start or end time of an activity, etc), or from an automated planner (by
 expanding subgoals, introducing precedence constraints to resolve
@@
 \texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
-\cite{mackworth77} where propagators are triggered to infer to ensure
+\cite{mackworth77} where propagators are invoked repeatedly until
 the constraint network returns to a quiescent state.
@@
 types. Tables \ref{tab:calconst}, \ref{tab:setconst},
-\ref{tab:objhierarchy} and \ref{tab:misc} show various constraint
-types.
+\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
+types included in \eus Constraint Library.
 
@@
 Temporal and resource constraints are especially important to model
-real-world planning domains. % so it is worth going into some more
-% detail about how \eus built-in support works for those constraint
-% types.
+real-world planning domains.
+
 \eu uses variables to explicitly represent timepoints for plan
-activities and states. Constraints among timepoints provide a natural
+actions and states. Constraints among timepoints provide a natural
 way to express domain axioms. For example, in order to state that
@@
 which are then subsequently released. \textbf {Unary} resources are
-reusable with capacity$ = 1$, which allows \eu to perform run time
+reusables with capacity$ = 1$, which allows \eu to perform run time
 savings in comparison to the generic reusable class. Configuration
@@
 occurs out of a range of possible values that are in the domain of the
-variables that represent the temporal extant for the event.
+variables that represent the temporal extent for the event.
 \cite{Muscettola04,Muscettola06} proposed an algorithm where the
@@
 the search space must be weighted against the ability to search faster
-\cite{6007772}.
+\cite{6007772}, \comment{cite "Stable Grounded Inference in Flexible Resource Scheduling" by Morris, Bresina, Barreiro}.
 
@@
 
-Given the above, an automated problem solver can be created using, a
+Given the above, an automated problem solver can be created using a
 \emph{domain model} that describes the variable, object, predicate and
@@
 any decisions that fall completely outside of the horizon can be
-ignored. This is particularly useful for recurrent activities;
+ignored. This is useful to ignore activities that may occur far in the future 
+or in the past to be relevant for a particular plan, ensuring that the solver is making
+as few commitments as necessary and reducing the complexity of plan generation.
+The \emph{planning horizon} is also useful when dealing with recurrent activities;
 consider for instance a model where a medical checkup has to occur
@@
 the horizon to ensure that a finite number of medical checkup
-activities are generated as part of a plan. This is also useful to
-ignore activities that may occur far in the future or in the past to
-be relevant for a particular plan, ensuring that the solver is making
-as few commitments as necessary and making planning tractable.
-
+activities are generated as part of a plan. 
 \subsubsection{Flaw Resolution}
@@
 
-% This Plan Space Planning algorithm is translated into \eus
-% representation as follows:
-
 The initial partial plan is the state of \eus Plan Database after the
@@
 
-\item[\textbf{Open Condition}]: an open condition is an
+\item[\textbf{Open Condition}:] an open condition is an
   \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
   by explicitly posted goals, or when a token is activated, rules may
-  activate the creation of \texttt{INACTIVE} slave tokens as described
+  cause the creation of \texttt{INACTIVE} slave tokens as described
   in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
@@
 
-  \item[\textbf{Merging}]: A free activity is merged with a matching
-    activity already in the plan. An activity $a$ is said to match an
-    activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig} and the
-    temporal constraints involving $a$ are satisfied by $a?$. Thus,
-    $a$ and $a?$ can be considered the same activity without the need
-    to introduce $a$ in the plan. Consequently, the domain rules associated with
-    $a$ are not introduced into the planning process, because they have been already
-    triggered when $a?$ was introduced in the plan.
+  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
+    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
+    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
+    and the temporal constraints involving $a$ are satisfied by
+    $a?$. Thus, $a$ and $a?$ can be considered the same token
+    without the need to introduce $a$ in the plan. Consequently, the
+    domain rules associated with $a$ are not introduced into the
+    planning process, since they are already triggered when $a?$ was
+    introduced in the plan.
 
-  \item[\textbf{Activation}]: A new activity $a$ is introduced in the
-    current plan associating it with the correct timeline, without
-    choosing a specific time slot (if its timing causes a problem 
-    it'll trigger a Threat flaw, explained below). 
-    The domain rules associated with $a$ are applied and the subgoal
-    activities resulting from those domain rules are introduced as inactive tokens. 
-    This results in a number of open condition flaws, corresponding to
-    the new subgoal activities.
+  \item[\textbf{Activation}:] A token $a$ is introduced in the
+    current plan associating it with the appropriate timeline, without
+    choosing a specific time slot; if its timing causes a problem a
+    Threat flaw will be triggered (see below).  The domain rules
+    associated with $a$ are applied and the subgoal activities
+    resulting from those domain rules are introduced as \texttt{INACTIVE}
+    tokens.  This results in a number of open condition flaws,
+    corresponding to the new subgoal activities.
 
-  \item[\textbf{Rejection}]: Some goals may be optional; if so its
+  \item[\textbf{Rejection}:] Some goals may be optional; if so its
     corresponding flaw may be resolved by discarding the
@@
 partial plan and also for filtering infeasible values from
-consideration prior to commitment. % This allows in some cases a strong
-% look-ahead capability which is essential for tractable
-% search.
-Consistency testing is initiated by the \texttt{isInconsistent}
-procedure. The algorithm permits a heuristically controlled search by
+consideration prior to commitment. Consistency testing is initiated by the \texttt{isInconsistent}
+procedure.
+
+The algorithm permits a heuristically controlled search by
 applying orderings for \texttt{chooseFlaw} and
@@
 It should be emphasized that while this algorithm is commonly employed
-with \eu implementations, it is only one of many that could be
+by applications that embed  \eu, it is only one of many that could be
 implemented. As a simple example, consider the following solver for
@@
 arithmetic constraints among the V$_i$ variables to ensure that all
-the rows and diagonals contain only one queen. \eus Constraint
-Reasoning Engine or \texttt{CRE} can then be used to compute the
-non-attack constraint violations, with \texttt{ce.getViolations()} a
-call on the API, to pick the queen that has he most violations.  By
-implementing \texttt{getQueenWithMaxViolation()} and looking ahead to
-see what moves are possible for a queen that would remove its
-violations with the implementing of \texttt{getMoves()}, this solver
-provides a viable planner implementation for a range of possible
-initial states.
+the rows and diagonals contain only one queen.
 
@@
 
+\eus Constraint Reasoning Engine or \texttt{CRE} can be used to compute the
+non-attack constraint violations, with \texttt{ce.getViolations()} a
+call on the \eu API. The \texttt{CRE} can also be used to pick the queen 
+that has he most violations in \texttt{getQueenWithMaxViolation()}, and to
+look ahead and see what moves are possible for a queen that would remove its
+violations in \texttt{getMoves()}.
+
 In similar fashion, the constraint violation and flaw reporting

### 2012-02-24 setobook.tex
Activated section numbering
Set section numbering depth to 1 (instead of initial -1)
updates to intro etc
heading changes based on Minh and discussion; also noted FP comments
minor fix to acknowledgements.

@@
 Rajan and Py are funded by a block grant from the David and Lucile
-Packard Foundation to MBARI. They thank their collaborators J. Ryan,
-J. Harvey, R. Vrijenhoek and J. Bellingham at MBARI. In addition they
-thank colleagues M. del Alto, D. Korsmeyer and S. Zornetzer at NASA
-Ames who made \eu available as open source. The authors are indebted
-to N. Muscettola for his ideas and early work at NASA Ames which forms
-the basis of this work.
+Packard Foundation to MBARI. They thank their collaborators John Ryan,
+Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. In
+addition they thank colleagues Martha del Alto, David Korsmeyer and
+Steven Zornetzer at NASA Ames who made \eu available as open
+source. The authors are indebted to Nicola Muscettola for his ideas
+and early efforts in Constraint-based Temporal Planning at NASA Ames
+which forms the basis of this work.
+The authors also want to thank everybdy who has contributed to the 
+development of \eu and its online documentation over the years, specially Ari Jonsson, Jeremy Frank, Conor McGann, Paul Morris,
+Michael Iatauro, Matthew Boyce and Tristan Smith.
 

### 2012-02-24 trunk/future.tex
added paragraph about europa's capabilities that could benefit t-rex in the future.

@@
 
-More specifically to \eu level we would like to improve \rx
+At the \eu level we would like to improve \rx
 integration of \eu ability to deliberate about resources. More
-specifically we need to ba able to update resource level as 
-modeled by \eu based on \rx observation.  As of today the 
-support of resource is limited inside the \eu solver and there's no
-direct way for \rx to update them dynamically based on recent
-observations. This result on a potential discrepeancy between measured
-given by, for example, the battery as modeled in \eu and the current
-battery level as read in a \rx timeline. Finding an automated way to
-link these two constructs would allow the plan to dynamically readapt
-when these two values differs -- for example if the vehicle looses a
-battery bank during a mission.
+specifically we need to be able to update resource level as 
+modeled by \eu based on \rx observation.  
+Keeping the link between these two constructs automatically updated 
+would allow the plan to dynamically adapt when these two values differ -- 
+for example if the vehicle looses a battery bank during a mission.
+Work is being done on \eu to provide enhanced modeling
+capabilities and more powerful domain-independent search, these
+would make it easier for \rx users to create domain models and to 
+perform more ambitious deliberation.
 

### 2012-02-24 trunk/arch.tex
updates to arch and results mostly cosmetic
final updates

@@
     with multiple possible clients ({\color{green}green} dashed
-    lines)]{\includegraphics[scale=0.45]{figs/AUV-agent.pdf}
+    lines).]{\includegraphics[scale=0.45]{figs/AUV-agent.pdf}
   \label{fig:agent}
@@
 
-It is important to note that, in T-REX architecture, the only critical
-task of a reactor is to successfully identify its state through
-synchronization as the execution frontier advances. Indeed, a reactor
-that fails to integrate its {\em external} state updates and/or
-identify its {\em internal} state at any tick express that it is not
-able anymore to explain the state of the world. In the agent point of
-view that means that any state information from this reactor is not to
-be trusted anymore and results on this reactor exclusion from the
-execution cycle and consequent destruction. As the agent destroy 
-the  faulty reactor is destroyed the agent produced on all of 
-its former {\em internal} timelines the special observation 
-\texttt{Failed} to all the reactors which did declare these state 
-variables as {\em external} informing
-them through this that this specific state variable is currently
-unavailable. On the other hand all goals exchanged between reactors
-are considered as rejectable by the reactor that manages them. As a
-result the failure to find a plan for these goals, while to be avoided,
-is not considered as a critical failure of the reactor and will only
-result on these goals being ignored until new goals are received.
-
-Take for example our AUV agent from Fig. \ref{fig:agent},
-and assume that the \texttt{Sampling} reactor failed to synchronize;
-in  such case this reactor is destroyed and both
-\texttt{Sampling\_policy} and \texttt{Relevant\_data} state variables
-produced the \texttt{Failed} observation. Therefore, the
-\texttt{Survey controller} will received this observation on its {\em
-  external} \texttt{Sampling\_policy} which indicates that this state
-variable is not managed anymore. At this point depending on this
-reactor implementation we have two possible outcomes:
-\begin{enumerate}
-\item \texttt{Survey  controller} model allows him to integrate this
-  \texttt{Failed} observation. For example it can identify that, from
-  now on, there's no way to take water samples. It still can 
-  navigate as \texttt{Pilot} is available which allow him to continue
-  this aspect of its current plan.
-\item \texttt{Survey controller} models is not able to handle this
-  unexpected failure resulting in turn on its failure to
-  synchronize. This result on this reactor being also destroyed and
-  the \texttt{Failed} observation is produced on the \texttt{survey}
-  timeline.
-\end{enumerate}
-Even on the worst case scenario presented on this second case, we
-note that a reactor failure only propagates up in the reactor graph. 
-In our example, even if all the reactors that depends on
-\texttt{Sampling} are entrained by its failure, we still keep the
-three reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
-  operation} allowing to  maintain the vehicle navigation control
-(\texttt{Pilot}) along with high level safety operation 
-(\texttt{Ship operation}). It illustrates how \rx can handle graceful
-system degradation despite the failure of one or more reactors. It
-also emphasize that it is critical to ensure that the low level
-reactors are properly designed in order to be as robust as 
-possible.
-
+\label{sec:delib-exec-as}
 \begin{figure}[!htbp]
@@
 
+\subsubsection{Handling Failure in \rx}
+\label{sec:rx-reactor-failure}
+
+The critical task of a \rx reactor is to successfully identify its
+state through synchronization as the execution frontier advances. When
+a reactor fails to integrate its {\em external} state update and/or
+identify its {\em internal} state at any tick, it implies that it is
+unable to explain the state of the world.  This in turn implies that
+any state information from this reactor is suspect, resulting in the
+exclusion of this reactor from the execution cycle and consequent
+destruction. As the agent destroys the faulty reactor, a special
+observation, \texttt{Failed}, is produced on all {\em internal}
+timelines. Reactors which declare these very state variables as {\em
+  external}, are informed of the failure of this timeline.  All goals
+exchanged between reactors are considered \emph{rejectable} by the
+reactor that manages them. As a result in such a situation, the
+failure to find a plan for these goals, is \emph{not} considered a
+critical failure of the reactor. It will only result on these goals
+being ignored until newer goals are received.
+
+Take for example our AUV agent from Fig. \ref{fig:agent}, and assume
+that the \texttt{Sampling} reactor failed to synchronize. In such a
+case this reactor is destroyed and both \texttt{Sampling\_policy} and
+\texttt{Relevant\_data} timelines generate the \texttt{Failed}
+observation as illustrated in Fig. \ref{fig:fail0}.  The
+\texttt{Survey controller} will receive this \texttt{Failed}
+observation on its {\em external} \texttt{Sampling\_policy} timeline
+which indicates that this state variable is no longer managed by its
+owner reactor. This leads to two potential outcomes based on how the
+reactor is implemented:
+
+\begin{enumerate}
+
+\item The \texttt{Survey controller} model allows it to integrate the
+  \texttt{Failed} observation. For example it can identify that,
+  subsequent use of the water sampler is no longer possible.  The \rx
+  agent can continue to navigate since \texttt{Pilot} is still
+  available. In such a case the agent structure remains as in
+  Fig. \ref{fig:fail0} with no consequent impact.
+
+\item The \texttt{Survey controller} model is unable to handle this
+  unexpected failure resulting in its failure to synchronize. This
+  results in the \texttt{Survey controller} also failing, with
+  subsequent destruction and flagging via the \texttt{Failed}
+  observation on its \texttt{survey} timeline as shown in
+  Fig. \ref{fig:fail1}.
+
+\end{enumerate}
+
+\begin{figure}[b]
+  \centering
+  \subfloat[\small When the \texttt{Sampling} reactor fails to
+  synchronize, it is removed from the reactor dependency graph along
+  with all its {\em internal} timelines while generating a \texttt{Failed} observation]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.pdf}
+  \label{fig:fail0}}
+  \qquad
+  \subfloat[\small In the worst case this \texttt{Failed} observation may imply
+  other reactors that depend on the \texttt{Sampling} reactor might fail to
+  synchronize leading to a bottom-up destruction of ``faulty'' reactors.]{
+    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1}\label{fig:fail1}}
+  \caption{\small Impact of the \texttt{Sampling} reactor failure in
+    the AUV agent shown in Fig. \ref{fig:agent}}
+\end{figure}
+
+Even in the worst of the two cases presented above, we note that a
+reactor failure only propagates \emph{up} in the reactor dependency
+graph.  In our example, even if all the reactors that depend on
+\texttt{Sampling} are entrained by its failure, \rx continues to keep
+the reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
+  operation} intact.  This ensures maintenance of vehicle control via
+the \texttt{Pilot}, in addition to an abstract commanding mode for
+safe operation via the \texttt{Ship operation} reactor, illustrating
+how \rx can handle system degradation despite the failure of one or
+more reactors, and doing so gracefully.
+
+Consequently, by ensuring that lower level reactors closest to the
+platform hardware are robustly designed, the \rx framework ensures
+that system designers can focus their attention on critical components
+while factoring in deliberative control of their platforms.
+
 \subsection{The  Deliberative Reactor}
@@
   that integrates new information in the plan. This signifies the
-  evolution of \emph{external} state variables and ensures that the
-  reactor propagates these to identify current reactor state as well
-  as to inform other reactors of any state change on its
-  \emph{internal} timelines. This solver is summoned at the beginning
-  of every single tick.
+  evolution of \emph{external} timelines and ensures that the reactor
+  propagates these to identify current reactor state as well as to
+  inform other reactors of any state change on its \emph{internal}
+  timelines. This solver is summoned at the beginning of every single
+  tick.
 
@@
 
-\begin{figure}[!htb]
+\begin{figure}[!b]
   \centering
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!b]
   \centering

### 2012-02-24 trunk/concepts.tex
final updates

@@
     token.]{\label{fig:europapr:activate}\includegraphics[scale=0.35]{figs/europa-pr-activate.pdf}}
-    \caption{\small Evolution of a timeline from the Shopping Agent example}
+    \caption{\small Evolution of a timeline from the Shopping Agent example.}
   \end{figure}
@@
 token variables.  The effect of merging is illustrated in
-Fig. \ref{fig:europapr:merge}
-
-% \begin{figure} 
-%   \centering
-%   \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
-%   \caption{\small State of timelines after satisfying a goal to be
-%     home by time $20$ on merging with initial state.}
-%   \label{fig:europapr:merge}
-% \end{figure}
-
-The goal to be home by time $20$ is satisfied by merging with the
-initial state that said that the Agent is home at time $0$.  The merge
-is possible because the tokens are of the same type \texttt{At}, and
-intersecting the domains for all of their variables doesn't yield an
-empty domain (which represents an inconsistency), in this case the
-token variables are the location, which is \texttt{Home} in both
-cases, and the temporal variables (start, end, duration), which can be
-intersected without causing an inconsistency. After the token
-representing the goal is merged onto the token representing the
-initial state, the resulting token represents not only the fact that
-the Agent must be \texttt{At(Home)} by time 0 but also
-\texttt{At(Home)} at time $20$, this is reflected in the new
-constraints on the token's start and end variables as depicted in
-Fig. \ref{fig:europapr:merge}.
+Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
+satisfied by merging with the initial state that said that the Agent
+is home at time $0$.  The merge is possible because the tokens are of
+the same type \texttt{At}, and intersecting the domains for all of
+their variables doesn't yield an empty domain (which represents an
+inconsistency), in this case the token variables are the location,
+which is \texttt{Home} in both cases, and the temporal variables
+(start, end, duration), which can be intersected without causing an
+inconsistency. After the token representing the goal is merged onto
+the token representing the initial state, the resulting token
+represents not only the fact that the Agent must be \texttt{At(Home)}
+by time 0 but also \texttt{At(Home)} at time $20$, this is reflected
+in the new constraints on the token's start and end variables as
+depicted in Fig. \ref{fig:europapr:merge}.
 
@@
 purchased before being able to buy it, and so on, until all domain
-rules are satisfied.
-
-% \begin{figure} \centering
-%   \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
-%   \caption{\small State of timelines after satisfying a goal to own a
-%     drill by time $10$ by activating corresponding token.}
-% \label{fig:europapr:activate}
-% \end{figure}
-
-As Fig. \ref{fig:europapr:activate} shows the decision to merge the
-goal to be \texttt{At(Home)} by time $20$ with the initial state was
-not appropriate since the Agent will have to get away from
-\texttt{Home} to be able to buy the drill which is inconsistent with
-the constraints that resulted after the merge. As a result that
-decision will have to be retracted at some point and the alternative
-to activate or merge with a different token will have to be
-considered.
+rules are satisfied. As Fig. \ref{fig:europapr:activate} shows the
+decision to merge the goal to be \texttt{At(Home)} by time $20$ with
+the initial state was not appropriate since the Agent will have to get
+away from \texttt{Home} to be able to buy the drill which is
+inconsistent with the constraints that resulted after the merge. As a
+result that decision will have to be retracted at some point and the
+alternative to activate or merge with a different token will have to
+be considered.
 
@@
   \includegraphics[scale=0.45]{figs/europa-pr-states.jpg}
-  \caption{\small Possible transitions for a Token State}
+  \caption{\small Possible transitions for a token state.}
 \label{fig:europapr:states}
@@
 
-  \begin{figure} 
+  \begin{figure}[t]
     \centering

### 2012-02-24 trunk/europa-basics.tex
final updates

@@
 {\scriptsize
-  \begin{quote}
-[Planning] \emph{is an abstract, explicit deliberation process that chooses and
-organizes actions by anticipating their expected outcomes. This
-deliberation aims at achieving as best as possible some prestated
-objectives. Automated planning is an area of Artificial Intelligence
-(AI) that studies this deliberation process computationally.} -- from
-\textbf{Automated Planning Theory and Practice} by Ghallab, Nau and
-Traverso \cite{ghallab04} 
+  \begin{quote} [Planning] \emph{is an abstract, explicit deliberation
+      process that chooses and organizes actions by anticipating their
+      expected outcomes. This deliberation aims at achieving as best
+      as possible some prestated objectives. Automated planning is an
+      area of Artificial Intelligence (AI) that studies this
+      deliberation process computationally.} -- from \textbf{Automated
+      Planning Theory and Practice} by Ghallab, Nau and Traverso
+    \cite{ghallab04}
 \end{quote}
@@
 
-\begin{figure} \centering
+Planning as an activity is ingrained in human decision-making
+\cite{berthoz} and its generalization into a computational framework
+demonstrating machine intelligence long targeted by pioneers in
+computer science \cite{computersthought}. Early work in this context
+focused on logical theorem proving by computer \cite{green69} which
+has formed the basis of much of the discipline of AI Planning. The
+emphasis in these early attempts was to logically derive the existence
+proof of a trajectory of actions for a hypothetical robot (moving
+blocks, chess pieces, furniture etc) to achieve a specific goal given
+specific initial conditions. Two important properties of
+\emph{Soundness}\footnote{A planning algorithm is sound if invoked on
+  a problem $P$ returns a plan which is a solution for $P$.} and
+\emph{Completeness}\footnote{A planning algorithm is complete if
+  invoked on a solvable problem $P$ is guaranteed to return a
+  solution.} along with derived results for algorithmic complexity
+\cite{corman} were a defining cornerstone for understanding how
+different computational paradigms could actually be tractable first in
+simulation and subsequently in actual robotic platforms. In this
+chapter our emphasis is to focus on embedded AI Planning and the
+execution of plans on an embedded robotic marine robotic
+platform. While the platform is domain specific, the techniques we
+highlight are general purpose as noted in Section \ref{sec:intro}.
+
+\begin{figure}[b]
+\centering
   \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
@@
 \centering
-\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
-\caption{\small N-Queens solution generated by \eu}
+\includegraphics[scale=0.3]{figs/Example-NQueens1.jpg}
+\caption{\small An $N$-Queens solution generated by \eu}
 \label{fig:nqueens-2}
-\vskip+0.1cm
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
-\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
-\caption{\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
-  problem.}
-\label{fig:rcpsp-1}
-\vskip+0.1cm
+\subfloat[\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
+  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=0.2]{figs/Example-UBO0.jpg}}\qquad
+\subfloat[\small A \eu solution to a shopping agent problem domain where
+  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=0.25]{figs/Example-Shopping0.jpg}}
+\caption{\small \eu based solutions to scheduling and planning problems.}
 \end{figure}
@@
 problem instance.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
-\caption{\small A \eu solution to a shopping agent problem domain where
-  the agent needs to buy Bananas, Milk and a Drill.}
-\label{fig:shopping-1}
-\vskip+0.1cm
-\end{figure}
 

### 2012-02-24 trunk/europa.tex
final updates

@@
 
-\begin{figure}
+\begin{figure}[b]
 \centering
@@
   integer variables to represent the built-in temporal parameters for
-  each token (activity duration, start and end times), string
+  each token (duration, start and end times), string
   variables to represent \textit{Location} and \textit{Product} names,
@@
 invoked to resolve those flaws as part of the deliberation
-process. Finally, constraints are restrictions on and between actions
+process. Finally, constraints are restrictions on and between facts
 and goals. In the example above, we use temporal constraints to
-specify when facts hold and when the goals should be
-accomplished. % As
-% always, all constraints in \eus library could be used.
+specify when facts hold and when the goals should be accomplished. In
+the model above, \texttt{before} and \texttt{after} are \texttt{Allen
+  Algebra} \cite{allen84} operators, which \eu supports to specify
+temporal relations between tokens (specifically between their start
+and end timepoints). The algebra is particularly effective in dealing
+with interval arithmetic; its elements are summarized in
+Fig. \ref{fig:allen-algebra}
 
-\begin{figure}[!htb]
+\begin{figure}[!t]
 \centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\includegraphics[scale=0.4]{figs/Allen-algebra.pdf}
 \caption{\small Temporal relations defined within the planner are
@@
 \end{figure}
-
-In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
-temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
-encoding of temporal constraint relationships between tokens.
-\kcomment{This para could do with a little embellishment especially
-  since it is the only material that actually talks about the kinds of
-  constraints and their origin.JB: I don't really know what else to
-  say given the table above.}
 
@@
 the search space must be weighted against the ability to search faster
-\cite{6007772}.
+\cite{6007772, morris11}.
 
@@
 any decisions that fall completely outside of the horizon can be
-ignored. This is particularly useful for recurrent activities;
-consider for instance a model where a medical checkup has to occur
-every $3$ months. The user can model a recurring activity and then use
-the horizon to ensure that a finite number of medical checkup
-activities are generated as part of a plan. This is also useful to
-ignore activities that may occur far in the future or in the past to
-be relevant for a particular plan, ensuring that the solver is making
-as few commitments as necessary and making planning tractable.
+ignored. This is useful to ignore activities that may occur far in the
+future or in the past to be relevant for a particular plan, ensuring
+that the solver is making as few commitments as necessary and reducing
+the complexity of plan generation.  The \emph{planning horizon} is
+also useful when dealing with recurrent activities; consider for
+instance a model where a medical checkup has to occur every $3$
+months. The user can model a recurring activity and then use the
+horizon to ensure that a finite number of medical checkup activities
+are generated as part of a plan.
 
@@
   by explicitly posted goals, or when a token is activated, rules may
-  activate the creation of \texttt{INACTIVE} slave tokens as described
+  cause the creation of \texttt{INACTIVE} slave tokens as described
   in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
@@
 
-  \item[\textbf{Merging}:] A free activity \kcomment{this could be
-      defined, since no def. for 'free' is provided} is merged with a
-    matching activity already in the plan. An activity $a$ is said to
-    match an activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
+  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
+    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
+    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
     and the temporal constraints involving $a$ are satisfied by
-    $a?$. Thus, $a$ and $a?$ can be considered the same activity
+    $a?$. Thus, $a$ and $a?$ can be considered the same token
     without the need to introduce $a$ in the plan. Consequently, the
@@
 
-  \item[\textbf{Activation}:] A new activity $a$ is introduced in the
+  \item[\textbf{Activation}:] A token $a$ is introduced in the
     current plan associating it with the appropriate timeline, without
@@
     associated with $a$ are applied and the subgoal activities
-    resulting from those domain rules are introduced as inactive
+    resulting from those domain rules are introduced as \texttt{INACTIVE}
     tokens.  This results in a number of open condition flaws,
@@
 partial plan and also for filtering infeasible values from
-consideration prior to commitment. % This allows in some cases a strong
-% look-ahead capability which is essential for tractable
-% search.
-Consistency testing is initiated by the \texttt{isInconsistent}
-procedure. The algorithm permits a heuristically controlled search by
+consideration prior to commitment. Consistency testing is initiated by
+the \texttt{isInconsistent} procedure.
+
+The algorithm permits a heuristically controlled search by
 applying orderings for \texttt{chooseFlaw} and
@@
       }
-      $checkSolution()$ \tcp*{See if we have a new best solution}
+      $checkSolution()$ \tcp*{Evaluate if a new best solution}
       $curIter$\texttt{++} \;
@@
 \end{algorithm}
+
+\eus Constraint Reasoning Engine or \texttt{CRE} can be used to compute the
+non-attack constraint violations, with \texttt{ce.getViolations()} a
+call on the \eu API. The \texttt{CRE} can also be used to pick the queen 
+that has he most violations in \texttt{getQueenWithMaxViolation()}, and to
+look ahead and see what moves are possible for a queen that would remove its
+violations in \texttt{getMoves()}.
 

### 2012-02-24 trunk/results.tex
final updates

@@
 
-A significant driver of the applicability of our work was the
-growing need for AUVs mission to be driven by scientific sensor 
-data.Thid trend is illustrated by the 
-implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
-$2$ litre syringe style containers $10$ of which are encased within
-the mid-body section of our Dorado vehicle as shown in Fig
-\ref{fig:gulper}. However, the original design and intended use of
-operation continued to dominated by a naive belief of collecting and
-returning water samples as traditional ship-borne methods often
-employ, namely using pre-scripted methods, albeit on a highly capable
-robot. With the installation of \rx onboard the Dorado with its
-capability to synthesize plans and replan on the fly, more imaginative
-methods of sample acquisition were targeted.
+A significant driver of the applicability of our work was the growing
+need for AUV missions to be driven by sensor data. This trend is
+illustrated by the implementation of MBARI's \emph{Gulper}
+water-samplers \cite{Bird07}, $2$ litre syringe style containers $10$
+of which are encased within the mid-body section of our Dorado vehicle
+as shown in Fig \ref{fig:gulper}. However, the original design and
+intended use of operation continued to be dominated by a naive belief
+of collecting and returning water samples as traditional ship-borne
+methods often employ, namely using pre-scripted methods. With the
+installation of \rx onboard the Dorado with its capability sense the
+environment and to appropriately synthesize plans and replan on the
+fly, more imaginative methods of sample acquisition were targeted.
 
@@
   particulate matter that originate from the sea floor
-  \cite{mcphee-shaw2006}} \cite{ryan10}. The location and scale of
+  \cite{mcphee-shaw2006}.} \cite{ryan10}. The location and scale of
 INLs in the water column, which depend upon seasonal and episodic

### 2012-02-24 future.tex
updates to arch and results mostly cosmetic
final updates

@@
 \label{sec:future}
-
-% Where do we see all this going esp. in the context of embedded AI Planning
-% for marine robots? This can be (and should be)
-% speculative. Shore-side/onboard autonomy coordination and control of
-% multiple vehicles is an important result of this activity.
-
-% 1.6 Future Work?
-% explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).
-
-% The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space
 
@@
 as \cite{kumar11,sergio12}. With the advent of the \od however,
-another direction our research is taking us is in Recommender systems
-\cite{Adomavicius05} as noted in Section \ref{sec:results}. We are
-working to design and deploy software infrastructure that will learn
-from user tagged data. To do so, we will use tagging techniques
-similar to what users of commercial image sharing sites like Flickr
-and Picassa do, to learn the context of the data (including remote
-sensing). In addition, the \od will integrate statistical Machine
-Learning techniques with incoming asynchronous data stream obtained by
-sensors and platforms. Identification models will be based on existing
-training sets for INLs and plumes; models for other features of
-scientific interests driven by \can requirements will be built in
-addition. Existing AUV archives for instance provide sufficient data
-for a learning base for a range of features of interest to \can; in
-addition we will build software that will learn to incorporate
-incoming data streams to provide an existence proof of a feature of
-interest.
+another direction our research is taking us is in \texttt{Recommender
+  Systems} \cite{Adomavicius05} as noted in Section
+\ref{sec:results}. We are working to design and deploy software
+infrastructure that will learn from user tagged data. To do so, we
+will use tagging techniques similar to what users of commercial image
+sharing sites like Flickr and Picassa do, to learn the context of the
+data (including remote sensing). In addition, the \od will integrate
+statistical Machine Learning techniques with incoming asynchronous
+data stream obtained by sensors and platforms. Identification models
+will be based on existing training sets for INLs and plumes; models
+for other features of scientific interests driven by \can requirements
+will be built in addition. Existing AUV archives for instance provide
+sufficient data for a learning base for a range of features of
+interest to \can; in addition we will build software that will learn
+to incorporate incoming data streams to provide an existence proof of
+a feature of interest.
+
+A number of issues remain to be explored improving \rx and \eu
+integration particularly in the context of engineering models for
+execution.
+
+Dispatchability of the {\em external} temporally flexible partial
+plans continues to be an open research problem. Substantial progress
+has been made \cite{vidal97,mus98a,tsam98,morris00} by analyzing the
+plan structure of static offline temporal plans. However in the
+context of embedded plan execution, balancing model design while
+dealing with execution uncertainty requires further research. For
+instance it should be possible to distinguish part of the plan that
+contributes to a reactor goal, which often needs to be dispatched as
+early as possible with other tokens describing world evolution with a
+preference to dispatch as late as possible. Making this distinction
+would improve the quality of agent behavior while avoiding extra model
+complexity.
+
+More abstractly at the architectural level, the ability to dynamically
+manipulate the reactor graph would be of importance while creating new
+reactors or modifying timelines. Doing so would provide more robust
+failure recovery mechanisms especially after a reactor has failed to
+synchronize. For example, an alternate reactor could take ownership of
+timelines left invalid by the destruction of another failed reactor
+ensuring a transparent implementation of component redundancy. While
+the current design broadly provides for such a capability, the \rx API
+has yet to implement it while ensuring other design elements within
+\rx and \eu are not compromised.
+
+Finally and importantly, integration of reasoning about resources
+within \rx and tying it to existing \eu capabilities needs to be
+demonstrated. Specifically \rx needs to be able to update resource
+levels as incoming observations.  Doing so would ensure plan
+adaptation occurs dynamically when planned and observed resource
+values diverge -- for instance if the vehicle looses a bank of
+batteries during the course of a mission.  Enhanced modeling
+capabilities and augmented domain-independent search within \eu would
+make it easier for \rx users to create such domain models and to
+perform more proactive deliberation.
+
+Along with all these potential extensions, we are also very cognizant
+that the learning curve associated with \rx and \eu for a general user
+in the marine robotics community needs to be tackled. At present, the
+complex system that underlies \rx and its underlying framework, even
+if well documented, can be intimidating for a naive first time user.
 

### 2012-02-24 intro.tex
final updates

@@
 and immediate impact for both civil and military
-applications. Autonomous Underwater Vehicles (AUVs) in particular have
-altered the way oceanographers are able to reach beyond the surface
-and make observations which are not only at higher resolution but have
-altered both the discipline and impacted the engineering methodologies
-behind Sampling, Control and Robotics itself
+applications. Autonomous Underwater Vehicles (AUVs) have not only
+altered the ability of oceanographers to reach beyond the surface and
+make high resolution observations, but have impacted the engineering
+methodologies behind sampling, control and robotics itself
 \cite{Brierley08032002,ryan05,Thomas06,Yoerger01012007,Incze2009,Rigby10}. Yet
@@
 adaptively. Further, over the last two decades substantial
-improvements in hardware including propulsion, battery technologies
-and hardware have not been kept abreast of in software especially in
-algorithmic methods in Sampling and Control and in systematic
-engineering and software engineering development.
+improvements in hardware including propulsion and battery technologies
+have outpaced software especially in algorithmic methods in sampling
+and control and in systematic software engineering development.
 
-This is one important reason that periodic calls towards sustained
-exploration and sampling \cite{rudnick03} are issued by scientists for
-whom the scope and scale of observations have become increasingly
-tractable using autonomous platforms both below and on the
-surface. The complexity of oceanic processes and the chronic
-under-sampling of the coastal oceans especially of dynamic features
-spread over large spatial scales (from meters to kilometers in extent)
-with dynamic biological activity across the temporal spectrum (in the
-order of hours to weeks), has made it challenging for robotic
-platforms to autonomously track and sample. A prominent example of
-such a process is coastal algal blooms which are patchy and could
-cover large coastal zones. Persistent observation of such dynamic
-events dictates that our robotic assets track and sample such patches
-which can evolve rapidly due to inherent bio-geochemical activity,
-advection as well as diffusion with the water mass it resides in.
+Large scale high resolution observations have become increasingly
+tractable using autonomous platforms both on and below the surface.
+This, in turn, has lead to periodic calls for sustained exploration
+and sampling \cite{rudnick03}, illustrating the high scientific value
+placed on such data.  The complexity of oceanic processes and the
+chronic under-sampling of the coastal oceans, especially of dynamic
+features studied over a variety of spatial and temporal scales, has
+made it challenging for robotic platforms to autonomously track and
+sample. A prominent example of such a process is coastal algal blooms
+which are patchy and could cover large coastal zones. Persistent
+observation of such dynamic events dictates that our robotic assets
+track and sample such patches which can evolve rapidly due to inherent
+bio-geochemical activity, as well as advection and diffusion of the
+water mass it resides in.
 
@@
 In this chapter, we propose an alternative view to decision-making
-using in-situ automated Planning \cite{ghallab04} and more important
-demonstrate that while earlier control paradigms primarily driven by
-\emph{reactive} approaches to control can be augmented to not only
-provide targeted observation and sampling, but also a more nuanced and
+using in-situ automated Planning \cite{ghallab04}. More importantly,
+we demonstrate that earlier control paradigms (primarily driven by
+\emph{reactive} approaches to control) can be augmented to provide
+targeted observation and sampling as well as a more nuanced and
 balanced consideration of mission objectives, environmental conditions
@@
 action space) is important to balance current needs of a robot with
-future desires or goals. And to do such planning in the context of
-time and available resources in-situ on the robot. Planning without
-time or action planning in of itself is not adequate; dealing with
-time ensures that the balance between now and the future is handled
+future desires or goals taking into account time and available
+resources in-situ on the robot. Planning without time or action
+planning in and of itself is not adequate; dealing with time ensures
+that the balance between now and the future is handled
 systematically. Yet, plans have to be tied to robotic action and in
@@
     useless, but planning is indispensable}'' and ``\emph{Failing to
-    plan is planning to fail}''}. Modeling of the marine environment
-to provide cues is hampered by poor understanding of the worlds ocean
-and poor predictive skill of ocean models and often little to no
-availability of a priori synoptic views of the survey area.
+    plan is planning to fail}''}. Use of environmental cues is often
+hampered by poor predicitve skill of ocean models and little to no
+availability of a priori synoptic views of the survey area
 
@@
 Subsumption-based approaches \cite{brooks86}. They have been adequate
-to doing routine straight-line survey tracts which have played a
+to doing routine straight-line survey trajectories which have played a
 valuable role in the acceptance of AUVs as a new tool for ocean
@@
 utilized; one consequence for instance, is any off-nominal condition
-results in a \emph{fail-safe} surface mode to radio for help. Further
-these approaches, words to the contrary are not only hard to maintain,
-are not 'adaptive' enough. Despite vigorous defense by Brooks early on
+results in a \emph{fail-safe} surface mode to radio for help. Software
+design using these reactive approaches, not only makes it hard for
+sustaining engineering development, these approaches are not
+adequately adaptive for emerging ocean science problems.  Despite
+vigorous defense by Brooks early on
 \cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep},
@@
 
-Persistent and early legitimate criticism of deliberation on the other
-hand also well articulated by Brooks
-\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
-was targeting the performance of in-situ deliberation. Terrestrial
-robotic platforms were well-known to frequently stop and ``plan''; 
-assuming in the meantime that the world will remain static. In
-the marine robotics world, such a situation would simply not be
-permissible given the need to make continuous observations without
-stopping an under-articulated robotic platform such as an AUV. However
-algorithmic and implementational advances by the AI community over the
-years has produced a range of planners with demonstrated embedded
-capabilities
+Criticism of AI planning, on the other hand, also well articulated by
+Brooks, was targeting the performance of in-situ
+deliberation. Terrestrial robotic platforms were well-known to
+frequently stop and ``plan''; assuming in the meantime that the world
+will remain static. In the marine robotics world, such a situation
+would simply not be permissible given the need to make continuous
+observations without stopping an under-articulated robotic platform
+such as an AUV. However, algorithmic and implementational advances by
+the AI community over the years have produced a range of planners with
+demonstrated embedded capabilities
 \cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
@@
 bring together in this chapter. Important lessons that were derived
-from this rich legacy involved the following:
+from this rich legacy included the following:
 
@@
 
-\item A model-based approach separating control formulation of the
-  platform which describes operational characteristics from the
+\item A model-based approach should separate control formulation of
+  the platform which describes operational characteristics, from the
   domain. The controller can then be rigorously tested and certified
@@
 within \rxe. A key portion is Section~\ref{sec:europa:pr} which
-details the internal plan representation essential to planning within
-a situated agent like \rxe. Modeling
-(Section~\ref{sec:europa:modeling}), Inference
+details the internal plan representation within a situated agent like
+\rxe. Modeling (Section~\ref{sec:europa:modeling}), Inference
 (Section~\ref{sec:europa:inference}) and Search
-(Section~\ref{sec:europa:search}) further detail the important details
-associated with \eu for the and relevant to how \rx works to the
-reader. With these foundational elements out of the way, we transition
-to \rx itself, in Section~\ref{sec:arch}, bring out architectural
-details in Section~\ref{sec:arch:trex}, deal with the execution cycle
-in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
-Section~\ref{sec:arch:europa} finally delving deeper into the key
+(Section~\ref{sec:europa:search}) articulates important details
+associated with \eu critical to \rxe.  With these foundational
+elements of deliberation out of the way, we transition to \rx itself
+in Section~\ref{sec:arch}, bring out architectural details in
+Section~\ref{sec:arch:trex}, deal with the execution cycle in
+Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
+Section~\ref{sec:arch:europa}, finally delving deeper into the key
 concept of synchronization in Section~\ref{sec:arch:synch}. Further

### 2012-04-20 trunk/arch.tex
updates

@@
 integrate in its evolving partial plan. For that reason, each reactor
-specifies the following two parameters that applied towards the scope
-of the agent lifetime:
+specifies the following two parameters that \kcomment{are} applied
+towards the scope of the agent lifetime:
 
@@
 have an explicit representation of time (such as but not limited to
-\eu) is often trivial.
+\eue) is often trivial.
 
@@
 the reactor is able to find a valid strategy in order to satisfy as
-many of its {\em internal} goals as possible. For \eu, this means that
-the solver is unable to identify any further flaws to be resolved in
-the plan.
+many of its {\em internal} goals as possible. For \eue, this means
+that the solver is unable to identify any further flaws to be resolved
+in the plan.
 
@@
 failure to find a plan for these goals, is \emph{not} considered a
-critical failure of the reactor. It will only result on these goals
-being ignored until newer goals are received.
+critical failure of the reactor. It will only result \kcomment{in}
+these goals being ignored until newer goals are received.
 
@@
 
-This suggests a deliberate non overlap between {\em reactive planner}
-and {\em deliberative planners} horizons. In \rxe, and more
-specifically in our \eu reactor the planning horizon of {\em
+This suggests a deliberate non overlap between \kcomment{the} {\em
+  reactive planner} and {\em deliberative planners} horizons. In \rxe,
+and more specifically in our \eu reactor the planning horizon of {\em
   synchronization} -- limited to the current tick -- is a subset of
@@
 conflict between the deliberation and synchronization by controlling
-the scheduling of these two process by the \rx agent.
+the scheduling of these two process\kcomment{es} by the \rx agent.
 
@@
 
-If \texttt{object.location} is {\em external} and \texttt{Agent} is an
+If \texttt{object.location\kcomment{.At}} is {\em external} and \texttt{Agent} is an
 {\em internal} timeline, a robot may encounter the situation where it
@@
 guarantees that the outcome of the \texttt{Going} will {\em always}
-result on ending at the traget location). However, given the very
-nature of the planning problem and the ability to search outcomes and
-to replan, the model should be written so that a future projection can
-be manipulated.
+result on ending at the \kcomment{target} location). However, given
+the very nature of the planning problem and the ability to search
+outcomes and to replan, the model should be written so that a future
+projection can be manipulated.
 
@@
 identified by the solver as any other option would lead immediately to
-n plan inconsistency. Since such a process is required by the \rx at
+plan inconsistency. Since such a process is required by the \rx at
 \emph{every} tick for \emph{every} reactor the model needs to be
@@
 acceptable for all the {\em internal} and {\em external} timelines of
-a \eu reactor.  By so doing we allow for graceful degradation where
-other reactors can readapt their plans on receipt of this observation.
+\kcomment{an} \eu reactor.  By so doing we allow for graceful
+degradation where other reactors can readapt their plans on receipt of
+this observation.
 
@@
   plan to identify its current {\em external} state. (This case is
-  developed further on the next section).
+  developed further \kcomment{in} the next section).
 
@@
 allows the planner to focus on its planning horizon and report
-decision on flaws that are beyond this scope. As, time advances, these
+decision on flaws that are beyond this scope. As time advances, these
 potential future flaws will eventually be part of the planning window

### 2012-04-20 future.tex
updates

@@
 Dearden and others \cite{wang09,ernits10,dearden11}. The second
-challenge has to implementing a policy within the \rx framework that
-explicitly deals with recovery since controller failure can be
-distributed across reactors. Since reactors are hierarchical in their
-relationship to one another, their well defined dependencies allow
-them to be removed from an agent systematically. Reactors more
-abstract in scope can be removed before those which are less abstract
-(and closer to the hardware). In the event a Navigation reactor, for
-example, has a fault, it can be removed safely with lower level
-functionalities (such as those in the executive). When the Navigator
-is removed, it will recursively remove more abstract reactors (such as
-a Mission Manager) that depend on it. Such a safety feature then
-allows the system designers to ensure that early on they can invest
-more in ensuring that reactors lower in the hierarchy are robust and
-potentially formally verified. It also means that there is a safe and
-sure way to take control of the vehicle at any level within the
-reactors hierarchy; as long as low level reactors are still active one
-will always been able to send goals/commands to them and execute what
-are traditionally known as runout sequences (or clean up scripts).
+challenge has to \kcomment{do with} implementing a policy within the
+\rx framework that explicitly deals with recovery since controller
+failure can be distributed across reactors. Since reactors are
+hierarchical in their relationship to one another, their well defined
+dependencies allow them to be removed from an agent
+systematically. Reactors more abstract in scope can be removed before
+those which are less abstract (and closer to the hardware). In the
+event a Navigation reactor, for example, has a fault, it can be
+removed safely with lower level functionalities (such as those in the
+executive). When the Navigator is removed, it will recursively remove
+more abstract reactors (such as a Mission Manager) that depend on
+it. Such a safety feature then allows the system designers to ensure
+that early on they can invest more in ensuring that reactors lower in
+the hierarchy are robust and potentially formally verified. It also
+means that there is a safe and sure way to take control of the vehicle
+at any level within the reactors hierarchy; as long as low level
+reactors are still active one will always been able to send
+goals/commands to them and execute what are traditionally known as
+runout sequences (or clean up scripts).
 
@@
 a prime motivation to move towards more semi-supervised methods such
-as \cite{kumar11,sergio12}. With the advent of the \od however,
+as \cite{kumar11}. With the advent of the \od however,
 another direction our research is taking us is in \texttt{Recommender

### 2012-04-20 results.tex
updates

@@
 
-A significant driver of the applicability of our work was the
-growing need for AUVs mission to be driven by scientific sensor 
-data.This trend is illustrated by the 
-implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
-$2$ litre syringe style containers $10$ of which are encased within
-the mid-body section of our Dorado vehicle as shown in Fig
-\ref{fig:gulper}. However, the original design and intended use of
-operation continued to dominated by a naive belief of collecting and
-returning water samples as traditional ship-borne methods often
-employ, namely using pre-scripted methods, albeit on a highly capable
-robot. With the installation of \rx onboard the Dorado with its
-capability to synthesize plans and replan on the fly, more imaginative
-methods of sample acquisition were targeted.
+A significant driver of the applicability of our work was the growing
+need for AUV missions to be driven by sensor data. This trend is
+illustrated by the implementation of MBARI's \emph{Gulper}
+water-samplers \cite{Bird07}, $2$ litre syringe style containers $10$
+of which are encased within the mid-body section of our Dorado vehicle
+as shown in Fig \ref{fig:gulper}. However, the original design and
+intended use of operation continued to be dominated by a naive belief
+of collecting and returning water samples as traditional ship-borne
+methods often employ, namely using pre-scripted methods. With the
+installation of \rx onboard the Dorado with its capability sense the
+environment and to appropriately synthesize plans and replan on the
+fly, more imaginative methods of sample acquisition were targeted.
 
@@
 Learned model with a deliberative planner is novel in the ocean
-sciences and is a continuing direction of investigation \cite{kumar11,
-  sergio12}.
+sciences and is a continuing direction of investigation
+\cite{kumar11}.
 
@@
 drifting instrument, the vehicle is expected to move around it. While
-our model suppirt multiples patterns(square, lawn-mower,
-transect(s)\dots) with any possible dimension factor, the
-majority of at-sea deployments have used a square pattern ($3X3$
+our model \kcomment{supports} multiples patterns (square, lawn-mower,
+transect(s)\dots) with any possible dimension factor, the majority of
+at-sea deployments have used a square pattern ($3X3$
 km\textsuperscript{2} or a $1X1$ km\textsuperscript{2})\footnote{Often

### 2012-04-20 setobook.tex
updates

@@
-\documentclass[graybox,12pt]{svmult}
+% \documentclass[a4paper, 12pt]{article}
+\documentclass[graybox,14pt]{svmult}
 
@@
   control architecture that incorporates in-situ decision making for
-  adaptation and control for autonomous underwater vehicles
-  (AUVs). The Teleo-Reactive EXecutive (\texttt{T-REX}) framework
-  deliberates about future states, plans for actions and executes
-  generated activities while monitoring plans for anomalous
-  conditions. Plans are no longer scripted a priori but synthesized
-  onboard with high-level directives instead of low level
+  autonomous underwater vehicles (AUVs). The Teleo-Reactive EXecutive
+  (\rx) framework deliberates about future states, plans for actions
+  and executes generated activities while monitoring plans for
+  anomalous conditions. Plans are no longer scripted a priori but
+  synthesized onboard with high-level directives instead of low level
   commands. Further, the architecture uses multiple control loops for
@@
 Steven Zornetzer at NASA Ames who made \eu available as open
-source. The authors are indebted to Nicola Muscettola for his ideas
-and early efforts in Constraint-based Temporal Planning at NASA Ames
-which forms the basis of this work.
-The authors also want to thank everybdy who has contributed to the 
-development of \eu and its online documentation over the years, specially Ari Jonsson, Jeremy Frank, Conor McGann, Paul Morris,
-Michael Iatauro, Matthew Boyce and Tristan Smith.
+source. The authors thank other members of the \eu team over the
+years, including Ari J\'onsson, Jeremy Frank, Conor McGann, Paul
+Morris, Michael Iatauro, Matthew Boyce and Tristan Smith.  Minh Bin Do
+(NASA Ames), William Yew Teck Tan (NUS, Singapore) and Rishi Graham
+(MBARI) provided comments on early drafts; we are grateful for their
+help in making this manuscript more accessible. Finally, the authors
+are indebted to Nicola Muscettola for his ideas and early efforts in
+Constraint-based Temporal Planning at NASA Ames which forms the basis
+of much of this work.
 

### 2012-04-23 concepts.tex
updates to reviews

@@
 \centering
-\includegraphics[scale=0.3]{figs/flexible-timelines.pdf}
+\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
   timepoints \cite{Dean88} to represent its start and end.  In such
-  representation, one writes \texttt{At($a$,$s$,$e$,$l$)} to indicate
-  that agent $a$ is at location $l$ from time $s$ to time $e$. In
-  fact, this pattern of using such predicates to describe both state
-  and behavior of objects is prevalent enough that \eu the token
-  representation. Token have built in variables to indicate the object
-  to which the statement principally applies as well as the timepoints
-  over which it holds. A token is an instance of a predicate that
-  represents an object's state or behavior and is defined over a
-  temporal extent. In \eu, all predicate instances are tokens. Also,
-  in the same way that objects are described generically by classes,
-  in \eu tokens can be described generically by token types. Every
-  token has five built-in variables:
+  \kcomment{a} representation, one writes \texttt{At($a$,$s$,$e$,$l$)}
+  to indicate that agent $a$ is at location $l$ from time $s$ to time
+  $e$. In fact, this pattern of \kcomment{usage} to describe both
+  state and behavior of objects is prevalent enough that \eu
+  \kcomment{encapsulates into} the token
+  representation. \kcomment{Tokens} have built in variables to
+  indicate the object to which \kcomment{it primarily} applies as well
+  as the timepoints over which it holds. A token, therefore, is an
+  instance of a predicate that represents an object's state or
+  behavior and is defined over a temporal extent. In \eu, all
+  predicate instances are tokens. \kcomment{In} the same way that
+  objects are described generically by classes, in \eu tokens can be
+  described generically by token types. Every token has five built-in
+  variables:
 
@@
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}.  \kcomment{add a figure}
+  overlap may occur among them \cite{mus94}. Fig. \ref{fig:europapr1}
+  below and several others after that illustrate how timelines are
+  used to represent state in a plan.
 
@@
 
-  \begin{figure} \centering
-    \includegraphics[scale=0.4]{figs/europa-pr-1.pdf}
-    \caption{\small Timelines representing initial state and goals for a Shopping Agent.}
-    \label{fig:europapr1}
+  \begin{figure}[t]
+    \centering 
+    \subfloat[\small Timelines representing initial state
+    and goals for a Shopping Agent.]{\label{fig:europapr1}\includegraphics[scale=0.4]{figs/europa-pr-1.pdf}}\qquad
+    \subfloat[\small State of timelines after satisfying a goal to be
+    home by time $20$ on merging with initial state.]{\label{fig:europapr:merge}\includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}}\qquad
+    \subfloat[\small State of timelines after satisfying a goal to own
+    a drill by time $10$ by activating corresponding
+    token.]{\label{fig:europapr:activate}\includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}} 
+    \caption{\small Evolution of a timeline from the Shopping Agent example.}
   \end{figure}
@@
   \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu must
-  first determine whether it can achieve those goal and how. When the
-  goals are posted and before any decisions are made, the
-  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an
-  \texttt{INACTIVE} state. \eu then considers the partial plan and
+  first determine whether it can achieve those \kcomment{goals} and
+  how. When the goals are posted and before any decisions are made,
+  the \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in
+  an \texttt{INACTIVE} state. \eu then considers the partial plan and
   evaluates two possible alternatives:
@@
 token variables.  The effect of merging is illustrated in
-Fig. \ref{fig:europapr:merge}
-
-\begin{figure} 
-  \centering
-  \includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}
-  \caption{\small State of timelines after satisfying a goal to be
-    home by time $20$ on merging with initial state.}
-  \label{fig:europapr:merge}
-\end{figure}
-
-The goal to be home by time $20$ is satisfied by merging with the
-initial state that said that the Agent is home at time $0$.  The merge
-is possible because the tokens are of the same type \texttt{At}, and
-intersecting the domains for all of their variables doesn't yield an
-empty domain (which represents an inconsistency), in this case the
-token variables are the location, which is \texttt{Home} in both
-cases, and the temporal variables (start, end, duration), which can be
-intersected without causing an inconsistency. After the token
-representing the goal is merged onto the token representing the
-initial state, the resulting token represents not only the fact that
-the Agent must be \texttt{At(Home)} by time 0 but also
-\texttt{At(Home)} at time $20$, this is reflected in the new
-constraints on the token's start and end variables as depicted in
-Fig. \ref{fig:europapr:merge}.
+Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
+satisfied by merging with the initial state that said that the Agent
+is home at time $0$.  The merge is possible because the tokens are of
+the same type \texttt{At}, and intersecting the domains for all of
+their variables doesn't yield an empty domain (which represents an
+inconsistency), in this case the token variables are the location,
+which is \texttt{Home} in both cases, and the temporal variables
+(start, end, duration), which can be intersected without causing an
+inconsistency. After the token representing the goal is merged onto
+the token representing the initial state, the resulting token
+represents not only the fact that the Agent must be \texttt{At(Home)}
+by time 0 but also \texttt{At(Home)} at time $20$, this is reflected
+in the new constraints on the token's start and end variables as
+depicted in Fig. \ref{fig:europapr:merge}.
 
@@
 purchased before being able to buy it, and so on, until all domain
-rules are satisfied.
-
-\begin{figure} \centering
-  \includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}
-  \caption{\small State of timelines after satisfying a goal to own a
-    drill by time $10$ by activating corresponding token.}
-\label{fig:europapr:activate}
-\end{figure}
-
-As Fig. \ref{fig:europapr:activate} shows the decision to merge the
-goal to be \texttt{At(Home)} by time $20$ with the initial state was
-not appropriate since the Agent will have to get away from
-\texttt{Home} to be able to buy the drill which is inconsistent with
-the constraints that resulted after the merge. As a result that
-decision will have to be retracted at some point and the alternative
-to activate or merge with a different token will have to be
-considered.
+rules are satisfied. As Fig. \ref{fig:europapr:activate} shows the
+decision to merge the goal to be \texttt{At(Home)} by time $20$ with
+the initial state was not appropriate since the Agent will have to get
+away from \texttt{Home} to be able to buy the drill which is
+inconsistent with the constraints that resulted after the merge. As a
+result that decision will have to be retracted at some point and the
+alternative to activate or merge with a different token will have to
+be considered.
 

### 2012-04-23 trunk/arch.tex
updates to reviews

@@
 planning and execution in a single representational and computational
-framework with a unified declarative model. Our approach with \rx
-builds on this design methodology albeit in a substantially systematic
-manner. The \rx architecture contributes to agent architectures over
-and beyond \texttt{IDEA} are in the following ways:
+framework with \kcomment{an} unified declarative model. Our approach
+with \rx builds on this design methodology albeit in a substantially
+systematic manner. The \rx architecture contributes to agent
+architectures over and beyond \texttt{IDEA} in the following
+\kcomment{significant} ways:
 
@@
   each reactor declares its planning look-ahead $\pi$ and expected
-  latency $\lambda$ before being able to produce plan to the agent
-  depending on where it is located in a reactor dependency graph such
-  as shown in Fig. \ref{fig:agent}. A reactor closer to the hardware
-  in the hierarchy, is expected to deliberate less and be highly
-  responsive to observations from the world. This implies lower values
-  of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex} below).
+  latency $\lambda$ before being able to produce \kcomment{a} plan to
+  the agent depending on where it is located in a reactor dependency
+  graph such as shown in Fig. \ref{fig:agent}. A reactor closer to the
+  hardware in the hierarchy, is expected to deliberate less and be
+  highly responsive to observations from the world. This implies lower
+  values of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex}
+  below).
 
@@
   process and world evolution to ensure that planning occurs
-  \emph{during continuous} state update in the {\em same} polan
-  database used for planning. By doing so, the planner is
-  better informed in its deliberation despite a larger latency than
+  \emph{during continuous} state update in the {\em same}
+  \kcomment{plan} database used for planning. By doing so, the planner
+  is better informed in its deliberation despite a larger latency than
   the changing world.
@@
   \caption{\small Impact of the \texttt{Sampling} reactor failure in
-    the AUV agent shown in Fig. \ref{fig:agent}}
+    the AUV agent \kcomment{from} Fig. \ref{fig:agent}}
 \end{figure}
@@
 
-Steps $2$ \& $3$ above are handled by the \eu solver from algorithm
+Steps $2$ \& $3$ above are handled by the \eu solver from Algorithm
 \ref{alg:europa:solve} which ensures that the {\em current state
-  flaws} are selected last so as to follow the sequence given
-above.
+  flaws} are selected last so as to follow the sequence given above.
 

### 2012-04-23 trunk/europa-basics.tex
updates to reviews

@@
 \rx is an adaptive, artificial intelligence based controller and
-provides general framework for building reasoning systems for
-real-world autonomous vehicles. At MBARI \rx is used for AUV control;
-another instantiation of the system is being used for control of a
-terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
+provides \kcomment{a} general framework for building reasoning systems
+for real-world autonomous vehicles. At MBARI \rx is used for AUV
+control; another instantiation of the system is being used for control
+of a terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
 development of \rx has been targeted at surveying a number of
@@
 chapter our emphasis is to focus on embedded AI Planning and the
-execution of plans on an embedded robotic marine robotic
+execution of plans on an embedded \kcomment{marine robotic}
 platform. While the platform is domain specific, the techniques we
@@
 \subfloat[\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
-  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=0.2]{figs/Example-UBO0.jpg}}\qquad
+  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=0.25]{figs/Example-UBO0.jpg}}\qquad
 \subfloat[\small A \eu solution to a shopping agent problem domain where
-  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=0.25]{figs/Example-Shopping0.jpg}}
+  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=0.3]{figs/Example-Shopping0.jpg}}
 \caption{\small \eu based solutions to scheduling and planning problems.}
@@
 
-To generate plans, the \texttt{CAIP} framework uses a plan-space search
-approach \cite{ghallab04}, where a partial plan is evolved until all
-open goals are achieved, and all inconsistencies are removed. The
-specific search algorithm used by \eu will be covered in detail below.
+To generate plans, the \texttt{CAIP} framework uses a plan-space
+search approach \cite{ghallab04}, where a \emph{partial plan} is
+evolved until all open goals are achieved and all inconsistencies are
+removed. \kcomment{A partial plan is a subsequence of a plan which can
+  be refined into a plan structure.} The specific search algorithm
+used by \eu will be covered in \kcomment{further} detail below.
 

### 2012-04-23 trunk/intro.tex
updates to reviews

@@
 In this chapter, we propose an alternative view to decision-making
-using in-situ automated Planning \cite{ghallab04} and more important
-demonstrate that while earlier control paradigms primarily driven by
-\emph{reactive} approaches to control can be augmented to not only
-provide targeted observation and sampling, but also a more nuanced and
+using in-situ automated Planning \cite{ghallab04}. More importantly,
+we demonstrate that earlier control paradigms (primarily driven by
+\emph{reactive} approaches to control) can be augmented to provide
+targeted observation and sampling as well as a more nuanced and
 balanced consideration of mission objectives, environmental conditions
 and available resources. Planning (or deliberation or projection in
-action space) is important to balance current needs of a robot with
-future desires or goals. And to do such planning in the context of
-time and available resources in-situ on the robot. Planning without
-time or action planning in of itself is not adequate; dealing with
-time ensures that the balance between now and the future is handled
-systematically. Yet, plans have to be tied to robotic action and in
-the real-world which is not static. Further, plans project future
-state which is not only likely to change, but is based on substantial
-uncertainty\footnote{It is important to remember that these principles
-  are broad much as Dwight Eisenhower is reputed to have said
-  ``\emph{In preparing for battle I have always found that plans are
-    useless, but planning is indispensable}'' and ``\emph{Failing to
-    plan is planning to fail}''}. Modeling of the marine environment
-to provide cues is hampered by poor understanding of the worlds ocean
-and poor predictive skill of ocean models and often little to no
-availability of a priori synoptic views of the survey area.
+action space \kcomment{and used interchangeably in this chapter}) is
+important to balance current needs of a robot with future desires or
+goals taking into account time and available resources in-situ on the
+robot. Planning without time or action planning in and of itself is
+not adequate; dealing with time ensures that the balance between now
+and the future is handled systematically. Yet, plans have to be tied
+to robotic action and in the real-world which is not static. Further,
+plans project future state which is not only likely to change, but is
+based on substantial uncertainty\footnote{It is important to remember
+  that these principles are broad much as Dwight Eisenhower is reputed
+  to have said ``\emph{In preparing for battle I have always found
+    that plans are useless, but planning is indispensable}'' and
+  ``\emph{Failing to plan is planning to fail}''}. Use of
+environmental cues is often hampered by poor \kcomment{predictive}
+skill \cite{anderson2009} of ocean models and little to no
+availability of a priori synoptic views of the survey
+area. \kcomment{Planning therefore provides a ``buffer'' against the
+  stochastic environment.}
 
@@
 We have developed, tested and deployed the Teleo-Reactive EXecutive
-(\texttt{T-REX}) an on-board adaptive control system that integrates
-AI based planning and %\sout{probabilistic} 
-state estimation in a hybrid executive\cite{mcgann08a,mcgann08b,py10}. 
-%  \sout{Probabilistic state estimation integrates a number of science observations to produce a
-% likelihood that the vehicle sensors perceive a feature of
-% interest}.\fcomment{\rx {\em does not} integrate probabilistic state
-%   estimation! ... The HMM estimation was coded on top of \eu model but
-%   neither \eu or \rx had explicit management of it} 
-State estimation allow the system to keep track of the world evolution
-as time advances based on a formal model distributed through the architecture.
-Onboard planning and execution enables adaptation of
-navigation and instrument control based on the state identified during
-state estimation.
-% \sout{the probability of having detected such a phenomenon} 
-It further enables goal-directed
-commanding within the context of projected mission state and allows
-for replanning for off-nominal situations and opportunistic science
-events. The framework in addition to being used on an AUV, is general
-enough be used for controlling a personal robot \cite{mcgann2009} and
-is deployed on a European planetary rover testbed
-\cite{goac11}. While it is possible to integrate probabilistic
-estimation on top of this framework \cite{mcgann08d}, such
-representation is not explicitly handled by the framework and is
-out of scope of this chapter.
-% Probabilistic state estimation through modeling within 
-% this framework can be done is discussed elsewhere
-% \cite{mcgann08d} and out of scope for this chapter. 
-Deliberation and reaction are integrated systematically over different temporal and
-functional scopes within a single agent and a single model that covers
-the needs of high-level mission management, low-level navigation,
-instrument control, and detection of unstructured and poorly
-understood phenomena. \rx is deployed on MBARI's Dorado AUV shown in
-Fig. \ref{fig:auv-fig}, which to the best of our knowledge is the only
-operational marine platform anywhere being used for routine scientific
-surveys with onboard plan synthesis.
+(\rxe) an on-board adaptive control system that integrates AI based
+planning and state estimation in a hybrid
+executive\cite{mcgann08a,mcgann08b,py10}.  State estimation allows the
+system to keep track of world evolution as time advances based on a
+formal model distributed through the architecture.  Onboard planning
+and execution enables adaptation of navigation and instrument control
+based on the state identified during state estimation.  It further
+enables goal-directed commanding within the context of projected
+mission state and allows for replanning for off-nominal situations and
+opportunistic science events. The framework in addition to being used
+on an AUV, is general enough \kcomment{to} be used for controlling a
+personal robot \cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed
+on a European planetary rover testbed \cite{goac11}. While it is
+possible to integrate probabilistic estimation on top of this
+framework \cite{mcgann08d}, such representation is not explicitly
+handled by the framework and is out of scope of this chapter.
+Deliberation and reaction are integrated systematically over different
+temporal and functional scopes within a single agent and a single
+model that covers the needs of high-level mission management,
+low-level navigation, instrument control, and detection of
+unstructured and poorly understood phenomena. \rx is deployed on
+MBARI's \emph{Dorado} AUV shown in Fig. \ref{fig:auv-fig}, which to
+the best of our knowledge is the only operational marine platform
+anywhere being used for routine scientific surveys with onboard plan
+synthesis.
 
@@
 deeper into the \eu framework itself, the basis of all deliberation
-within \rx. A key portion is Section~\ref{sec:europa:pr} which details
-the internal plan representation essential to planning within a
-situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
-Inference (Section~\ref{sec:europa:inference}) and Search
-(Section~\ref{sec:europa:search}) further detail the important details
-associated with \eu for the and relevant to how \rx works to the
-reader. With these foundational elements out of the way, we transition
-to \rx itself, in Section~\ref{sec:arch}, bring out architectural
-details in Section~\ref{sec:arch:trex}, deal with the execution cycle
-in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
-Section~\ref{sec:arch:europa} finally delving deeper into the key
+within \rxe. A key portion is Section~\ref{sec:europa:pr} which
+details the internal plan representation within a situated agent like
+\rxe. Modeling (Section~\ref{sec:europa:modeling}), Inference
+(Section~\ref{sec:europa:inference}) and Search
+(Section~\ref{sec:europa:search}) articulates important details
+associated with \eu critical to \rxe.  With these foundational
+elements of deliberation out of the way, we transition to \rx itself
+in Section~\ref{sec:arch}, bring out architectural details in
+Section~\ref{sec:arch:trex}, deal with the execution cycle in
+Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
+Section~\ref{sec:arch:europa}, finally delving deeper into the key
 concept of synchronization in Section~\ref{sec:arch:synch}. Further

### 2012-04-23 trunk/related.tex
updates to reviews

@@
 scalability and systematicity. \cite{Benjamin2006} is clear about the
-extant of adaptation in stating:
+\kcomment{extent} of adaptation in stating:
 
@@
 what would be expected of a typical well-rounded Computer Science
-graduate student.
+graduate student. Yet in our experience, the cost of model design in
+such systems, is well balanced against the cost of re-engineering new
+mission scripts for new deployment in more ``classical'' script-based
+controllers.
 
@@
 the field of robotics.
-\fcomment{Though it woulfd  be usefull to emphasize this ...}
-The cost of modeling put aduring the intial design, for example,
-should be balanced against the cost of redesigning
-a new mission script eate every new deployment in more ``classical''
-script-based controllers.
 
 \subsection{Evolution of AI-based Robotic Planning: A Necessary Digression}
+\label{sec:related:robotplans}
 
@@
 \texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
-\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). Early
-efforts were cognizant of if absent with, dealing with time and
-resources for such experiments.\fcomment{This sentence does not make
-  sence to me}
+\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). % Early
+% efforts were cognizant of if absent with, dealing with time and
+% resources for such experiments.
 
@@
 the planner is a monolithic computational environment. The rate of
-updates from the science drivers is roughly in the order of an earth
-orbit on the EO-1 vehicle. As a result, the ASE is only an incremental
-update on RAX and is not suitable for domains where complex automated
-reasoning and asynchronous environmental observations need to be
-factored in deliberation.
+updates from the science drivers is roughly \kcomment{on} the order of
+an earth orbit on the EO-1 vehicle. As a result, the ASE is only an
+incremental update on RAX and is not suitable for domains where
+complex automated reasoning and asynchronous environmental
+observations need to be factored in deliberation.
 
@@
 message passing mechanism. While this allows disparate code-bases to
-talk to each other thru IPC (a inter-process communication or message
-passing mechanism), failure in the centralized controller is
-tantamount to a system crash.
+talk to each other \kcomment{through} IPC (\kcomment{an} inter-process
+communication or message passing mechanism), failure in the
+centralized controller is tantamount to a system crash.
 
@@
 
-Our system, \texttt{T-REX}, the focus of this chapter, is an agent
-control structure formally defined as a composition of coordinated
-control loops where sensing, planning and acting are resulting from 
-concurrent deliberating tasks within a formal framework. \fcomment{Got
-  rid of sense-plan-act reference here. \rx plan/sense while sensing/planning.}
-\texttt{T-REX} was designed after carefully
-evaluating lessons learned from \texttt{IDEA} and owes substantially
-to the notion of partitioned problem solving from that effort. However
-this partitioning is systematic and methodologically principled where
-we manage the information flow within the partitioned structure to
-ensure consistency in order to direct the flow of goals and
+Our system, \rxe, the focus of this chapter, is an agent control
+structure formally defined as a composition of coordinated control
+loops where sensing, planning and acting are resulting from concurrent
+deliberating tasks within a formal framework.  \rx was designed after
+carefully evaluating lessons learned from \texttt{IDEA} to which it
+owes substantially the notion of partitioned problem solving. However
+partitioning in \rx is systematic and methodologically principled
+where we manage the information flow within the partitioned structure
+to ensure consistency in order to direct the flow of goals and
 observations in a timely manner. The resulting control structure

### 2012-04-23 trunk/results.tex
updates to reviews

@@
 fly, more imaginative methods of sample acquisition were targeted.
+\kcomment{The Dorado uses two onboard computers: a main vehicle computer which
+is a 244 Mhz PC/104 stack running the QNX real-time operating system,
+and a 367 MHz EPIC EPX-GX500 AMD Geode stack running Linux and \rxe. A
+socket-based communication protocol allows the exchange of goals and
+state updates between \rx and the lower level functional layer.}
 
@@
 \subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
-\caption{\small \ref{fig:odss-mtg} shows a 2010 \can meeting with the
-  \od in use. \ref{fig:odss-virtual} shows a drifter tracking tool
-  used during \can planning; it used a physical model of advected
+\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
+  the \od in use. Fig. \ref{fig:odss-virtual} shows a drifter tracking
+  tool used during \can planning; it used a physical model of advected
   surface currents from \texttt{CeNCOOS} \cite{cencoos} to plan AUV
-  deployments. \ref{fig:odss-ipad} shows \od use on an iPad.}
+  deployments. Fig. \ref{fig:odss-ipad} shows \od use on an iPad.}
 \label{fig:odss2}

### 2012-04-23 trunk/setobook.tex
added people from europa dev to acknowledgements section.
updates to reviews

@@
 Packard Foundation to MBARI. They thank their collaborators John Ryan,
-Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. In
-addition they thank colleagues Martha del Alto, David Korsmeyer and
-Steven Zornetzer at NASA Ames who made \eu available as open
-source. The authors are indebted to Nicola Muscettola for his ideas
-and early efforts in Constraint-based Temporal Planning at NASA Ames
-which forms the basis of this work.
+Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. They
+thank other members of the \eu team over the years, including Conor
+McGann, Ari J\'onsson, Jeremy Frank, Paul Morris, John Bresina,
+Michael Iatauro, Matthew Boyce and Tristan Smith.  The authors are
+grateful to Martha del Alto, David Korsmeyer and Steven Zornetzer all
+at NASA Ames \kcomment{who not only made \eu available as open source
+  but encouraged its use outside the agency}. Minh Bin Do (NASA Ames),
+William Yew Teck Tan (NUS, Singapore) and Rishi Graham (MBARI)
+provided valuable comments on early drafts; we are grateful for their
+help in making this manuscript more accessible. Finally, the authors
+are indebted to Nicola Muscettola \kcomment{(Google)} for his ideas
+and early efforts in Constraint-based Temporal Planning while at NASA
+Ames, and which forms the basis of much of this work.
 

### 2012-04-23 intro.tex
updates to reviews

@@
 and available resources. Planning (or deliberation or projection in
-action space) is important to balance current needs of a robot with
-future desires or goals taking into account time and available
-resources in-situ on the robot. Planning without time or action
-planning in and of itself is not adequate; dealing with time ensures
-that the balance between now and the future is handled
-systematically. Yet, plans have to be tied to robotic action and in
-the real-world which is not static. Further, plans project future
-state which is not only likely to change, but is based on substantial
-uncertainty\footnote{It is important to remember that these principles
-  are broad much as Dwight Eisenhower is reputed to have said
-  ``\emph{In preparing for battle I have always found that plans are
-    useless, but planning is indispensable}'' and ``\emph{Failing to
-    plan is planning to fail}''}. Use of environmental cues is often
-hampered by poor predicitve skill of ocean models and little to no
-availability of a priori synoptic views of the survey area
+action space \kcomment{and used interchangeably in this chapter}) is
+important to balance current needs of a robot with future desires or
+goals taking into account time and available resources in-situ on the
+robot. Planning without time or action planning in and of itself is
+not adequate; dealing with time ensures that the balance between now
+and the future is handled systematically. Yet, plans have to be tied
+to robotic action and in the real-world which is not static. Further,
+plans project future state which is not only likely to change, but is
+based on substantial uncertainty\footnote{It is important to remember
+  that these principles are broad much as Dwight Eisenhower is reputed
+  to have said ``\emph{In preparing for battle I have always found
+    that plans are useless, but planning is indispensable}'' and
+  ``\emph{Failing to plan is planning to fail}''}. Use of
+environmental cues is often hampered by poor \kcomment{predictive}
+skill \cite{anderson2009} of ocean models and little to no
+availability of a priori synoptic views of the survey
+area. \kcomment{Planning therefore provides a ``buffer'' against the
+  stochastic environment.}
 
@@
 would simply not be permissible given the need to make continuous
-observations without stopping an under-articulated robotic platform
-such as an AUV. However, algorithmic and implementational advances by
-the AI community over the years have produced a range of planners with
-demonstrated embedded capabilities
+observations without stopping a \kcomment{non-holonomic} robotic
+platform such as an AUV. However, algorithmic and implementational
+advances by the AI community over the years have produced a range of
+planners with demonstrated embedded capabilities
 \cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
@@
 We have developed, tested and deployed the Teleo-Reactive EXecutive
-(\rx) an on-board adaptive control system that integrates AI based
+(\rxe) an on-board adaptive control system that integrates AI based
 planning and state estimation in a hybrid
@@
 opportunistic science events. The framework in addition to being used
-on an AUV, is general enough be used for controlling a personal robot
-\cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed on a European
-planetary rover testbed \cite{goac11}. While it is possible to
-integrate probabilistic estimation on top of this framework
-\cite{mcgann08d}, such representation is not explicitly handled by the
-framework and is out of scope of this chapter.  Deliberation and
-reaction are integrated systematically over different temporal and
-functional scopes within a single agent and a single model that covers
-the needs of high-level mission management, low-level navigation,
-instrument control, and detection of unstructured and poorly
-understood phenomena. \rx is deployed on MBARI's \emph{Dorado} AUV
-shown in Fig. \ref{fig:auv-fig}, which to the best of our knowledge is
-the only operational marine platform anywhere being used for routine
-scientific surveys with onboard plan synthesis.
+on an AUV, is general enough \kcomment{to} be used for controlling a
+personal robot \cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed
+on a European planetary rover testbed \cite{goac11}. While it is
+possible to integrate probabilistic estimation on top of this
+framework \cite{mcgann08d}, such representation is not explicitly
+handled by the framework and is out of scope of this chapter.
+Deliberation and reaction are integrated systematically over different
+temporal and functional scopes within a single agent and a single
+model that covers the needs of high-level mission management,
+low-level navigation, instrument control, and detection of
+unstructured and poorly understood phenomena. \rx is deployed on
+MBARI's \emph{Dorado} AUV shown in Fig. \ref{fig:auv-fig}, which to
+the best of our knowledge is the only operational marine platform
+anywhere being used for routine scientific surveys with onboard plan
+synthesis.
 

### 2012-04-23 results.tex
updates to reviews

@@
 fly, more imaginative methods of sample acquisition were targeted.
+\kcomment{The Dorado uses two onboard computers: a main vehicle
+  computer which is a 244 Mhz PC/104 stack running the QNX real-time
+  operating system, and a 367 MHz EPIC EPX-GX500 AMD Geode stack
+  running Linux and \rxe. A socket-based communication protocol allows
+  the exchange of goals and state updates between \rx and the lower
+  level functional layer.}\comment{We should say something about the
+  peak load during re-planning.}
 

### 2012-04-25 trunk/europa-basics.tex
updates to review

@@
 \centering
-  \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
+  \includegraphics[scale=0.45]{figs/shopping_europa.pdf}
   \caption{\small A potential plan for a robot shopping for groceries.}
@@
 $d_1,..,d_N$; in this chapter we will deal only with discrete and
-finite domains. Given a defined conjunctive set of constraints on the
-variables: $C=\{c_1(x_1,..,x_N), ..., c_K(x_1,..,x_N)\}$, the
-objective is to find one or more value assignments to $V$ where all
-constraints are satisfied. To solve a problem, \textsf{CP} techniques
-use logical inference to perform Constraint Propagation, which
-consists of two operations:
+finite domains. Given a defined conjunctive set of constraints
+\kcomment{\cite{gen87}} on the variables: $C=\{c_1(x_1,..,x_N), ...,
+c_K(x_1,..,x_N)\}$, the objective is to find one or more value
+assignments to $V$ where all constraints are satisfied. To solve a
+problem, \textsf{CP} techniques use logical inference to perform
+Constraint Propagation, which consists of two operations:
 

### 2012-04-25 trunk/related.tex
updates to review

@@
 
-\begin{figure}[htpb]
+\begin{figure}[!t]
  \centering
- \includegraphics[scale=0.3]{figs/planner-arch.pdf}
+ \includegraphics[scale=0.4]{figs/planner-arch.pdf}
  \caption{\small A general architectural block diagram for an AI based
@@
 \cite{strips71} dominating the academic landscape. An early
-demonstration of planning as a situated agent was the highly
-influential \texttt{Shakey} the robot experiment
-\cite{shakey84}. Subsequent work driven largely by defense funding in
-the United States sequestered the discipline in academic labs
-principally applied for algorithmic development and if embedded then
-in testing within the confines of modest indoor settings. The leap to
-applying them for real world problem solving is relatively new and
-driven largely by NASA applications pioneered by \cite{mus94,mus98,
-  jonsson00, rajan00, chien05, bresina05}.
+demonstration of planning as a situated agent\footnote{\kcomment{By
+    \emph{situated} we emphasize that an agent is embedded within a
+    physical robot.}} was the highly influential \texttt{Shakey} the
+robot experiment \cite{shakey84}. Subsequent work driven largely by
+defense funding in the United States sequestered the discipline in
+academic labs principally applied for algorithmic development and if
+embedded then in testing within the confines of modest indoor
+settings. The leap to applying them for real world problem solving is
+relatively new and driven largely by NASA applications pioneered by
+\cite{mus94,mus98, jonsson00, rajan00, chien05, bresina05}.
 
@@
 It was with the demonstration of \texttt{RAX} $65$ Million miles in
-outer space, that temporal reasoning using methods came to the
-forefront of AI planning and for situated agents. Lessons learned from
-\texttt{RAX} led to the development of \texttt{IDEA}
+outer space, that temporal reasoning using \kcomment{planning} methods
+came to the forefront of AI Planning for situated agents. Lessons
+learned from \texttt{RAX} led to the development of \texttt{IDEA}
 \cite{mus02,mus04,Dias:2003ua,mus06} with the central theme of using

### 2012-04-25 arch.tex
updates to review

--- 2012-02-24 arch.tex
@@
   the world throughout all the reactors as time
-  advances.
+  advances. \kcomment{Observations can therefore be both sensor
+    measurements as well state passed up from other reactors. The
+    strict rules of timeline ownership enable a clear policy for
+    conflict resolution: observations dominate expectations.}
 
@@
 
-\subsubsection{Reactors failure handling in \rx}
+\subsubsection{Handling Failure in \rx}
 \label{sec:rx-reactor-failure}
-It is important to note that, in T-REX architecture, the only critical
-task of a reactor is to successfully identify its state through
-synchronization as the execution frontier advances. Indeed, a reactor
-that fails to integrate its {\em external} state updates and/or
-identify its {\em internal} state at any tick express that it is not
-able anymore to explain the state of the world. In the agent point of
-view that means that any state information from this reactor is not to
-be trusted anymore and results on this reactor exclusion from the
-execution cycle and consequent destruction. As the agent destroy 
-the  faulty reactor is destroyed the agent produced on all of 
-its former {\em internal} timelines the special observation 
-\texttt{Failed} to all the reactors which did declare these state 
-variables as {\em external} informing
-them through this that this specific state variable is currently
-unavailable. On the other hand all goals exchanged between reactors
-are considered as rejectable by the reactor that manages them. As a
-result the failure to find a plan for these goals, while to be avoided,
-is not considered as a critical failure of the reactor and will only
-result on these goals being ignored until new goals are received.
-
-Take for example our AUV agent from Fig. \ref{fig:agent},
-and assume that the \texttt{Sampling} reactor failed to synchronize;
-in  such case this reactor is destroyed and both
-\texttt{Sampling\_policy} and \texttt{Relevant\_data} state variables
-produced the \texttt{Failed} observation. Therefore, the
-\texttt{Survey controller} will received this observation on its {\em
-  external} \texttt{Sampling\_policy} which indicates that this state
-variable is not managed anymore. At this point depending on this
-reactor implementation we have two possible outcomes:
+
+The critical task of a \rx reactor is to successfully identify its
+state through synchronization as the execution frontier advances. When
+a reactor fails to integrate its {\em external} state update and/or
+identify its {\em internal} state at any tick, it implies that it is
+unable to explain the state of the world.  This in turn implies that
+any state information from this reactor is suspect, resulting in the
+exclusion of this reactor from the execution cycle and consequent
+destruction. As the agent destroys the faulty reactor, a special
+observation, \texttt{Failed}, is produced on all {\em internal}
+timelines. Reactors which declare these very state variables as {\em
+  external}, are informed of the failure of this timeline.  All goals
+exchanged between reactors are considered \emph{rejectable} by the
+reactor that manages them. As a result in such a situation, the
+failure to find a plan for these goals, is \emph{not} considered a
+critical failure of the reactor. It will only result \kcomment{in}
+these goals being ignored until newer goals are received.
+
+Take for example our AUV agent from Fig. \ref{fig:agent}, and assume
+that the \texttt{Sampling} reactor failed to synchronize. In such a
+case this reactor is destroyed and both \texttt{Sampling\_policy} and
+\texttt{Relevant\_data} timelines generate the \texttt{Failed}
+observation as illustrated in Fig. \ref{fig:fail0}.  The
+\texttt{Survey controller} will receive this \texttt{Failed}
+observation on its {\em external} \texttt{Sampling\_policy} timeline
+which indicates that this state variable is no longer managed by its
+owner reactor. This leads to two potential outcomes based on how the
+reactor is implemented:
+
 \begin{enumerate}
-\item \texttt{Survey  controller} model allows him to integrate this
-  \texttt{Failed} observation. For example it can identify that, from
-  now on, there's no way to take water samples. It still can 
-  navigate as \texttt{Pilot} is available which allow him to continue
-  this aspect of its current plan.
-\item \texttt{Survey controller} models is not able to handle this
-  unexpected failure resulting in turn on its failure to
-  synchronize. This result on this reactor being also destroyed and
-  the \texttt{Failed} observation is produced on the \texttt{survey}
-  timeline.
+
+\item The \texttt{Survey controller} model allows it to integrate the
+  \texttt{Failed} observation. For example it can identify that,
+  subsequent use of the water sampler is no longer possible.  The \rx
+  agent can continue to navigate since \texttt{Pilot} is still
+  available. In such a case the agent structure remains as in
+  Fig. \ref{fig:fail0} with no consequent impact.
+
+\item The \texttt{Survey controller} model is unable to handle this
+  unexpected failure resulting in its failure to synchronize. This
+  results in the \texttt{Survey controller} also failing, with
+  subsequent destruction and flagging via the \texttt{Failed}
+  observation on its \texttt{survey} timeline as shown in
+  Fig. \ref{fig:fail1}.
+
 \end{enumerate}
-Even on the worst case scenario presented on this second case, we
-note that a reactor failure only propagates up in the reactor graph. 
-In our example, even if all the reactors that depends on
-\texttt{Sampling} are entrained by its failure, we still keep the
-three reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
-  operation} allowing to  maintain the vehicle navigation control
-(\texttt{Pilot}) along with high level safety operation 
-(\texttt{Ship operation}). It illustrates how \rx can handle graceful
-system degradation despite the failure of one or more reactors. It
-also emphasize that it is critical to ensure that the low level
-reactors are properly designed in order to be as robust as 
-possible.
+
+\begin{figure}[b]
+  \centering
+  \subfloat[\small When the \texttt{Sampling} reactor fails to
+  synchronize, it is removed from the reactor dependency graph along
+  with all its {\em internal} timelines while generating a \texttt{Failed} observation]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.pdf}
+  \label{fig:fail0}}
+  \qquad
+  \subfloat[\small In the worst case this \texttt{Failed} observation may imply
+  other reactors that depend on the \texttt{Sampling} reactor might fail to
+  synchronize leading to a bottom-up destruction of ``faulty'' reactors.]{
+    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1}\label{fig:fail1}}
+  \caption{\small Impact of the \texttt{Sampling} reactor failure in
+    the AUV agent \kcomment{from} Fig. \ref{fig:agent}}
+\end{figure}
+
+Even in the worst of the two cases presented above, we note that a
+reactor failure only propagates \emph{up} in the reactor dependency
+graph.  In our example, even if all the reactors that depend on
+\texttt{Sampling} are entrained by its failure, \rx continues to keep
+the reactors \texttt{Vehicle}, \texttt{Pilot} and \texttt{Ship
+  operation} intact.  This ensures maintenance of vehicle control via
+the \texttt{Pilot}, in addition to an abstract commanding mode for
+safe operation via the \texttt{Ship operation} reactor, illustrating
+how \rx can handle system degradation despite the failure of one or
+more reactors, and doing so gracefully.
+
+Consequently, by ensuring that lower level reactors closest to the
+platform hardware are robustly designed, the \rx framework ensures
+that system designers can focus their attention on critical components
+while factoring in deliberative control of their platforms.
 

### 2012-04-25 concepts.tex
updates to review

@@
 \centering
-\includegraphics[scale=0.35]{figs/flexible-timelines.pdf}
+\includegraphics[scale=0.4]{figs/flexible-timelines.pdf}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
 samples 100 meters from an environmental hotspot while an AUV is
-performing a Yo-Yo sawtooth pattern in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
+performing a Yo-Yo sawtooth pattern\footnote{\kcomment{Variability in
+    the water column is along the vertical dimension. Since the Dorado
+    platform can only move forward, the Yo-Yo pattern is the most
+    efficient mechanism for studying water column properties.}}  in
+the water-column. Two samples are needed if the feature's signal is
+above a threshold; one otherwise. The token that is capturing the
+sensory threshold has a parametric constraints to the token which
+fires the requisite water sampler. In addition, the start time of the
+water sampling procedure is highly dependant on the variability of
+sub-sea currents and actual speed of the vehicle. Therefore a number
+of values are possible for the start times and duration of the
+sampling all of which are valid combinations for desired outcomes.
 

### 2012-04-25 trunk/arch.tex
updates to review

@@
 reactiveness and that the environment can change at a faster rate than
-the planner can plan for. In such a situation, the agent may thrash if
+the planner can plan for. In such a situation, the agent may thrash
+\kcomment{(\ie as it works to complete a goal)}if
 the internal state of the plan gets out of synch with the actual state

### 2012-04-25 trunk/europa.tex
updates to reviews
updates to review

@@
 \label{sec:europa:arch}
+
+\comment{I think we should expand on this section which, as the
+  reviewer notes, the inter-relationships between the elements in Fig
+  \ref{fig:europa-architecture} below is not fully captured. I think
+  an illustrative example of how these components work together might
+  be very useful. This might also help the sub-section end less
+  abruptly than it is currently.}
 
@@
 \centering
-\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
-\caption{\small The \eu software architecture.}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
+\caption{\small The \eu software architecture.\comment{Is there a
+    better block diagram we can use to respond to review comment on
+    page 2?}}
 \label{fig:europa-architecture}
@@
 \item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
-  reason about temporal constraints
+  reason about temporal constraints.
 
@@
 bound. Since maximum flow on a temporal network can be computed in
-polynomial time, \eu takes advantage of this algorithm to efficiently
-compute upper and lower bound envelopes on the resource
-level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
-computation.
+polynomial time \comment{definition needed?}, \eu takes advantage of
+this algorithm to efficiently compute upper and lower bound envelopes
+on the resource level. Fig. \ref{fig:resenvelopes} shows an example of
+such a resource computation.
 
@@
 \centering
-\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
+\includegraphics[scale=0.3]{figs/europa-resource-envelopes.png}
 \caption{\small An example of resource envelopes computed by \eu.}

### 2012-04-25 concepts.tex
updates to review

@@
 \eu uses a \emph{state variable} representation
-\cite{mus94,jonsson00,mcgann09} to describe the evolution of state
-over time. The instantiated history of such state variable evolution
-over a temporal horizon we call \emph{timelines} and which represent a
-single thread in the execution of a concurrent system. At any given
-time each thread can execute a single procedure.  Thus each timeline
-consists of a sequence of procedures which encapsulate and describe
-state evolution; we call these instantiated atomic entities
-\emph{tokens}.  A token therefore describes a procedure invocation,
-the state variables on which it can occur, the parameter values of the
-procedure, and the time values defining the interval. We allow
-encapsulation of uncertainty within these tokens with a range of start
-and end times and parameters, all of which are encoded as variables. A
-constraint solver in turn manipulates these variables defined in a \eu
-domain model. For example, consider a scientific need to take water
-samples 100 meters from an environmental hotspot while an AUV is
-performing a Yo-Yo sawtooth pattern\footnote{\kcomment{Variability in
-    the water column is along the vertical dimension. Since the Dorado
-    platform can only move forward, the Yo-Yo pattern is the most
-    efficient mechanism for studying water column properties.}}  in
-the water-column. Two samples are needed if the feature's signal is
-above a threshold; one otherwise. The token that is capturing the
-sensory threshold has a parametric constraints to the token which
-fires the requisite water sampler. In addition, the start time of the
-water sampling procedure is highly dependant on the variability of
-sub-sea currents and actual speed of the vehicle. Therefore a number
-of values are possible for the start times and duration of the
-sampling all of which are valid combinations for desired outcomes.
+\cite{mus94,jonsson00,py10} to describe the evolution of state over
+time. The instantiated history of such state variable evolution over a
+temporal horizon we call \kcomment{as a} \emph{timeline} and
+\kcomment{one which represents} a single thread in the execution of a
+concurrent system. At any given time each thread can execute a single
+procedure.  Thus each timeline consists of a sequence of procedures
+which encapsulate and describe state evolution; we call these
+instantiated atomic entities \emph{tokens}.  A token therefore
+describes a procedure invocation, the state variables on which it can
+occur, the parameter values of the procedure, and the time values
+defining the interval. We allow encapsulation of uncertainty within
+these tokens with a range of start and end times and parameters, all
+of which are encoded as variables. A constraint solver in turn
+manipulates these variables defined in a \eu domain model. For
+example, consider a scientific need to take water samples 100 meters
+from an environmental hotspot while an AUV is performing a Yo-Yo
+sawtooth pattern\footnote{\kcomment{Variability in the water column is
+    along the vertical dimension. Since the Dorado platform can only
+    move forward, the Yo-Yo pattern is the most efficient mechanism
+    for studying water column properties.}}  in the water-column. Two
+samples are needed if the feature's signal is above a threshold; one
+otherwise. The token that is capturing the sensory threshold has a
+parametric constraints to the token which fires the requisite water
+sampler. In addition, the start time of the water sampling procedure
+is highly dependant on the variability of sub-sea currents and actual
+speed of the vehicle. Therefore a number of values are possible for
+the start times and duration of the sampling all of which are valid
+combinations for desired outcomes.
 
@@
 \item[\textbf{Objects}] The items we wish to describe and refer to in
-  a domain are considered objects. As in the case with object-oriented
-  analysis and design, one can seek out the nouns in any domain
-  description to find likely objects. In the Shopping Agent example,
-  we might consider the Agent, Products and Locations all to be
-  objects. Objects have state and behavior, for example, a Shopping
-  Agent can have, its location, a bag that contains the products it
-  has already purchased (the bag can in turn be another object) and
-  behavior as in going to a location to look for a product, perform a
-  purchase, return home, etc.
-
-  Objects that have similar state and behavior can be described
+  a domain are considered objects. % As in the case with object-oriented
+  % analysis and design, one can seek out the nouns in any domain
+  % description to find likely objects. ---- this seemed confusing
+  In the Shopping Agent example, we might consider the Agent, Products
+  and Locations all to be objects. Objects have state and behavior.
+  For example, a Shopping Agent can have its location, a bag that
+  contains the products it has already purchased (the bag can in turn
+  be another object) and behavior, as in going to a location to look
+  for a product, perform a purchase, return home, etc. all as objects.
+  Those that have similar state and behavior can be described
   generically in terms of object types or classes. \eu allows the
@@
   languages, where a class can have attributes (represented by
-  variables), and classes can be arranged in a single-parent hierarchy
-  through inheritance.  However, one key aspect in \eu that is not
-  found in object-oriented programming: to describe state and behavior
-  for the purposes of planning we need to build on the formalism of
-  first order logic.
+  variables) and classes can be arranged in a single-parent hierarchy
+  through inheritance.  \kcomment{What is different however, is that
+    to describe state and behavior for planning one needs} to build on
+  the formalism of first order logic.
 
@@
   and \texttt{Going} token required by the master is referred to as a
-  \emph{slave}.  Application of a rule on a token can thus cause slave
-  tokens, variables and constraints to be introduced.  The capability
-  of a domain rule to cause a slave token to be created is a key
-  vehicle through which planning occurs. Semantically, a rule imposes
-  a requirement for supporting (slave) tokens to be in the plan in
-  order for the master token to be valid.
+  \emph{slave}.  Application of a rule on a token \kcomment{(\ie a
+    master)} can thus cause slave tokens, variables and constraints to
+  be introduced.  The capability of a domain rule to cause a slave
+  token to be created is a key vehicle through which planning
+  occurs. Semantically, a rule imposes a requirement for supporting
+  slave tokens to be in the plan in order for the master token to be
+  valid.
 
@@
     \centering
-    \includegraphics[scale=0.4]{figs/europa-pr-rules.pdf}
+    \includegraphics[scale=0.45]{figs/europa-pr-rules.pdf}
     \caption{\small Master and Slave tokens generated as a result of a domain rule.}

### 2012-04-30 trunk/arch.tex
updates

@@
 the planner can plan for. In such a situation, the agent may thrash
-\kcomment{(\ie as it works to complete a goal)}if
-the internal state of the plan gets out of synch with the actual state
-of the world. As a result the planning problem managed {\em in situ}
-has traditionally remained fairly detached from low-level
-control. Most adaptations at the lower-level reactive layers in
-contrast are managed by specialized components (for example
-\texttt{LAAS architecture} \cite{alami:1998p820,Ingrand07} and
+\footnote{\kcomment{In such an event as the planner works towards goal
+    satisfaction, the initial conditions leading to achieve that goal
+    are no longer valid. The planner then tries to achieve a different
+    goal which too has to be discarded similarly. And so on.}}  if the
+internal state of the plan gets out of synch with the actual state of
+the world.
+
+\kcomment{A important consequence of this has been that} the planning
+problem managed {\em in situ} has traditionally remained fairly
+detached from low-level control. Most adaptations at the lower-level
+reactive layers in contrast are managed by specialized components (for
+example \texttt{LAAS architecture} \cite{alami:1998p820,Ingrand07} and
 \texttt{CLARATy} \cite{Nesnas:2003do}) that rely on different

### 2012-04-30 europa.tex
added example to the architecture section.

@@
 
-\begin{figure}
+\begin{figure}[b]
 \centering
-\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
-\caption{\small The \eu software architecture.}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
+\caption{\small The \eu software architecture.\comment{Is there a
+    better block diagram we can use to respond to review comment on
+    page 2?}}
 \label{fig:europa-architecture}
@@
 \end{enumerate}
+
+Before diving into the details of each module, a simple example illustrates how they interact at a high level:
+
+\begin{verbatim}
+
+class LightBulb extends Timeline
+{
+    predicate On  {}
+    predicate Off {}
+}
+
+class LightSwitch extends Timeline
+{
+    LightBulb myBulb_;
+
+    LightSwitch(LightBulb b)
+    {
+        myBulb_ = b;
+    }
+
+    action turnOn  { duration=1; }
+    action turnOff { duration=1; }
+}
+
+LightSwitch::turnOn
+{
+    met_by(condition object.myBulb_.Off);  // Bulb must be Off to be turned On
+    meets(effect object.myBulb_.On);   // Must be turned on through the switch
+}
+
+LightSwitch::turnOff
+{
+    met_by(condition object.myBulb_.On);   // Bulb must be On to be turned Off
+    meets(effect object.myBulb_.Off);      // Must be turned off through the switch
+}
+
+// Problem instance : turning the light off
+LightBulb bulb1 = new LightBulb();
+LightSwitch switch1 = new LightSwitch(bulb1);
+
+// At time 0, the bulb is on
+fact(bulb1.On initialCondition);
+eq(initialCondition.start,0);
+
+// We want the bulb to be off by time 10
+goal(bulb1.Off goal1);
+lt(0,goal1.start);
+lt(goal1.start,10);
+
+\end{verbatim}
+
+The snippet above is a NDDL model of a light bulb and a switch that controls it (NDDL is covered in more detail below), along with a problem instance that requires the light bulb to be off by time=10, given an initial state with the bulb on at time=0.
+The NDDL module allows \eu to ingest this model and problem description.
+Once the NDDL statements are parsed, the different object types, object instances and constraints are instantiated inside the Plan Database and Constraint Engine modules where they are available for both the user and the solver module to query and manipulate programmatically.
+This instantiation of the initial state constitutes the initial plan that \eu holds for this particular problem instance.
+The user can then invoke functionality available in the solver module to modify or extend the initial plan in a way that eliminates any flaws (violated constraints, unsatisfied goals, oversubscribed resources) in the current plan.
+When actions are added to a plan, the Rules Engine module ensures that domain rules expressed in the model are enforced when \eu creates new action or predicate instances, for instance, if a LightSwitch::turnOn action is added to the plan, the RulesEngine  ensures that requirements for the existence of the conditions and effects of that action (that is, the corresponding LightBulb must be off before and on after the action) are posted through the PlanDatabase and ConstraintEngine modules.
 
@@
 and goals. In the example above, we use temporal constraints to
-specify when facts hold and when the goals should be
-accomplished. In the model above, \texttt{before} and \texttt{after} 
-are \texttt{Allen Algebra} \cite{allen84} operators, which \eu supports to specify
-temporal relations between tokens (specifically between their start and end timepoints)
-as summarized in Fig. \ref{fig:allen-algebra}
+specify when facts hold and when the goals should be accomplished. In
+the model above, \texttt{before} and \texttt{after} are \texttt{Allen
+  Algebra} \cite{allen84} operators, which \eu supports to specify
+temporal relations between tokens (specifically between their start
+and end timepoints). The algebra is particularly effective in dealing
+with interval arithmetic; its elements are summarized in
+Fig. \ref{fig:allen-algebra}
 
-\begin{figure}[!htb]
+\begin{figure}[!t]
 \centering
-\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
+\includegraphics[scale=0.4]{figs/Allen-algebra.pdf}
 \caption{\small Temporal relations defined within the planner are
@@
 the search space must be weighted against the ability to search faster
-\cite{6007772}, \comment{cite "Stable Grounded Inference in Flexible Resource Scheduling" by Morris, Bresina, Barreiro}.
+\cite{6007772, morris11}.
 
@@
 any decisions that fall completely outside of the horizon can be
-ignored. This is useful to ignore activities that may occur far in the future 
-or in the past to be relevant for a particular plan, ensuring that the solver is making
-as few commitments as necessary and reducing the complexity of plan generation.
-The \emph{planning horizon} is also useful when dealing with recurrent activities;
-consider for instance a model where a medical checkup has to occur
-every $3$ months. The user can model a recurring activity and then use
-the horizon to ensure that a finite number of medical checkup
-activities are generated as part of a plan. 
+ignored. This is useful to ignore activities that may occur far in the
+future or in the past to be relevant for a particular plan, ensuring
+that the solver is making as few commitments as necessary and reducing
+the complexity of plan generation.  The \emph{planning horizon} is
+also useful when dealing with recurrent activities; consider for
+instance a model where a medical checkup has to occur every $3$
+months. The user can model a recurring activity and then use the
+horizon to ensure that a finite number of medical checkup activities
+are generated as part of a plan.
+
 \subsubsection{Flaw Resolution}
@@
 partial plan and also for filtering infeasible values from
-consideration prior to commitment. Consistency testing is initiated by the \texttt{isInconsistent}
-procedure.
+consideration prior to commitment. Consistency testing is initiated by
+the \texttt{isInconsistent} procedure.
 

### 2012-05-01 arch.tex
updates

@@
 controller which could bring automated planning closer to low-level
-control of the vehicle while maintaining its reactivity.
-
-Traditionally planning has been deemed computationally expensive with
-planner performance restricting robot reactivity
+control of the vehicle while maintaining its reactivity. Traditionally
+planning has been deemed computationally expensive with planner
+performance restricting robot reactivity
 \cite{ghallab04,Dias:2003ua}. In control architectures where task
@@
 reactiveness and that the environment can change at a faster rate than
-the planner can plan for. In such a situation, the agent may thrash if
-the internal state of the plan gets out of synch with the actual state
-of the world. As a result the planning problem managed {\em in situ}
-has traditionally remained fairly detached from low-level
-control. Most adaptations at the lower-level reactive layers in
-contrast are managed by specialized components (for example
-\texttt{LAAS architecture} \cite{alami:1998p820,Ingrand07} and
+the planner can plan for. In such a situation, the agent may thrash
+\footnote{\kcomment{In such an event as the planner works towards goal
+    satisfaction, the initial conditions leading to achieve that goal
+    are no longer valid. The planner then tries to achieve a different
+    goal which too has to be discarded similarly. And so on.}}  if the
+internal state of the plan gets out of sync with the actual state of
+the world.
+
+\kcomment{A important consequence of this has been that} the planning
+problem managed {\em in situ} has traditionally remained fairly
+detached from low-level control. Most adaptations at the lower-level
+reactive layers in contrast are managed by specialized components (for
+example \texttt{LAAS architecture} \cite{alami:1998p820,Ingrand07} and
 \texttt{CLARATy} \cite{Nesnas:2003do}) that rely on different
@@
 with \rx builds on this design methodology albeit in a substantially
-systematic manner. The \rx architecture contributes to agent
-architectures over and beyond \texttt{IDEA} in the following
-\kcomment{significant} ways:
+systematic manner.
+
+The \rx architecture contributes to agent architectures over and
+beyond \texttt{IDEA} in the following \kcomment{significant} ways:
 
@@
   of high level waypoints into low-level commands) and temporally as
-  each reactor declares its planning look-ahead $\pi$ and expected
-  latency $\lambda$ before being able to produce \kcomment{a} plan to
-  the agent depending on where it is located in a reactor dependency
-  graph such as shown in Fig. \ref{fig:agent}. A reactor closer to the
-  hardware in the hierarchy, is expected to deliberate less and be
-  highly responsive to observations from the world. This implies lower
-  values of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex}
-  below).
+  each reactor declares its planning look-ahead ($\pi$) and expected
+  latency ($\lambda$) before being able to produce \kcomment{a} plan
+  to the agent depending on where it is located in a reactor
+  dependency graph such as shown in Fig. \ref{fig:agent}. A reactor
+  closer to the hardware in the hierarchy, is expected to deliberate
+  less and be highly responsive to observations from the world. This
+  implies lower values of $\pi$ and $\lambda$ (see Section
+  \ref{sec:arch:trex} below). \kcomment{In the current design, $\pi$
+    and $\lambda$ are application dependent and fixed at design time.}
 
@@
 Systematic partitioned inference is a major contribution of
-\rxe. However that is not the focus of this chapter with details
+\rxe. However that is not the focus of this chapter; details are
 covered in \cite{py10, rajan12}. Coupled interaction between planning
-and the world on the other hand, has a significant implication on how
+and the world on the other hand, has a substantial impact on how
 planning can be integrated in such frameworks. In this chapter we show
-how using the \eu planner can be coupled to our framework by first
-linking concepts in temporal planning with the \rx framework. Note
-that in the following treatise, we call the formal semantic framework
-of \rx as the 'agent'. Reactors constitute an entity within this agent
-and are standalone blocks of inference. The scope, detail and number
-of reactors are application specific and outside the scope of this
+how using the \eu planner can be coupled to the \rx framework by first
+linking concepts in temporal planning with the framework. Note that in
+the following treatise, we call the formal semantic framework of \rx
+as the 'agent'. Reactors constitute an entity within this agent and
+are standalone blocks of inference. The scope, detail and number of
+reactors are application specific and outside the scope of this
 chapter.
@@
 
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[scale=0.35]{figs/token-disptch.pdf}
+  \caption{\small \kcomment{Dispatching a token occurs only when the request
+    start time overlaps a window. \texttt{Goal$_1$} therefore gets
+    dispatched from \texttt{reactor$_1$} to \texttt{reactor$_2$} but
+    not \texttt{Goal$_2$} to avoid early commitment.}}
+  \label{fig:reactor-lookahead}
+\end{figure}
+
 This window is used by \rx to identify when to dispatch a request to a
@@
   external} state variable of a reactor while avoiding flooding the
-owner reactor of this state variable.
+owner reactor of this state variable and avoiding early commitment.
+\kcomment{Fig. \ref{fig:reactor-lookahead} illustrates an example
+  between two reactors where one token is dispatched when overlapping
+  the dispatch window while another is not.}
 
@@
 as this information is available and properly maintained for managing
-{\em internal} state information. This {\em internal} state in turn
-may be used by other reactors. Such a formalism also constrains
-reactor interaction to state information exchange which comes in two
-forms as illustrated in Fig. \ref{fig:trex:flow}:
+{\em internal} state information. % This {\em internal} state in turn
+% may be used by other reactors.
+Such a formalism also constrains reactor interaction to state
+information exchange which comes in two forms as illustrated in
+Fig. \ref{fig:trex:flow}:
 
@@
   advances. \kcomment{Observations can therefore be both sensor
-    measurements as well state passed up from other reactors. The
-    strict rules of timeline ownership enable a clear policy for
-    conflict resolution: observations dominate expectations.}
+    measurements as well as state information, passed up from other
+    reactors. The strict rules of timeline ownership enable a clear
+    policy for conflict resolution: observations dominate
+    expectations.}
 
@@
   \vskip-1pc
-  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
-  \caption{\small \rx reactor execution cycle: {\em Deliberation} is
+  \subfloat[\small \kcomment{A view of synchronization and deliberation within a
+  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle}} 
+  \qquad
+  \subfloat[\small A \rx reactor's execution cycle: {\em Deliberation} is
     interrupted by {\em synchronization} at the beginning of every
-    {\em tick} allowing integration of state information.}
-  \label{fig:tick-exec}
-  \vskip-0.8pc
+    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle}}
+  \caption{\small \kcomment{Detailed and abstracted views of
+      synchronization and deliberation within one tick cycle.}}
+  \label{fig:sync-plan-cycle}
 \end{figure}
@@
 during previous {\em deliberation} steps.  Such a relaxation can be
-viewed as a backtracking allowing the reactor to identify that its
-current plan has failed to execute triggering further deliberations
+viewed as a backtrack, allowing the reactor to identify that its
+current plan has failed to execute triggering further deliberation
 steps to find an alternate strategy.
@@
 Finally, when the deliberation process has identified a complete
-solution for its current {\em internal} goals, a part of its plan that
-applies to its {\em external} timelines is sent as a request to \rx to
-handle their dispatch to corresponding reactors.  Synchronization
-eventually receives {\em external } observations that confirms plan
-execution, in turn recording execution progress by having the part of
-the plan observed restricted to observation start times.
+solution for its current goals, a part of its plan that applies to its
+{\em external} timelines is sent as a request to \rx to handle their
+dispatch to corresponding reactors.  Synchronization eventually
+receives {\em external } observations that confirms plan execution, in
+turn recording execution progress by having the part of the plan
+observed restricted to observation start times.
 
@@
 in Section \ref{sec:europa:arch}. However this deliberation comes with
-some augmentation namely the reactor's deliberation latency and its
-plan look-ahead both expressed in ticks and that the agent needs to
-interleave deliberation with not just the synchronization but also
-with deliberation of other reactors\footnote{The current
-  implementation of \rx is running on a single process and it is the
-  responsibility of the agent itself to emulate reactor
-  multi-threading for deliberation and synchronization.}.  For this
-reason while the reactor relies on the \eu solver, it slices the
-execution of Algorithm \ref{alg:europa:solve} into atomic steps. This
-interruption occurs around the recursive call line \ref{li:recurse}
-while ensuring that the current partial plan is not inconsistant at
-any single step. This is managed by allowing the call to backtrack
-until a \textsf{DecisionPoInt} that is not exhausted (\ie
-$decision.hasNext()$ is \texttt{true}) is found in the call stack.
+some augmentation, namely the reactor's deliberation latency
+\kcomment{($\lambda$)} and its planning look-ahead \kcomment{($\pi$)}
+both expressed in ticks. \kcomment{In addition,} the agent needs to
+interleave deliberation with not just synchronization but also with
+the deliberation of other reactors\footnote{The current implementation
+  of \rx is running on a single process and it is the responsibility
+  of the agent itself to emulate reactor multi-threading for
+  deliberation and synchronization.}.  For this reason while the
+reactor relies on the \eu solver, it slices the execution of Algorithm
+\ref{alg:europa:solve} into atomic steps. This interruption occurs
+around the recursive call line \ref{li:recurse} while ensuring that
+the current partial plan is not inconsistant at any single step. This
+is managed by allowing the call to backtrack until a
+\textsf{DecisionPoInt} that is not exhausted (\ie $decision.hasNext()$
+is \texttt{true} \kcomment{and non empty}) is found in the call
+stack\footnote{\kcomment{The call stack refers to the sequence of
+    recursive function calls.}}.
 
@@
 step we either backtrack in the decision stack until a node of the
-tree is found with an alternate solution or the plan is relaxed since
-there is no other alternative solution. We do this by removing all
-goals while keeping only recent observations.
+\kcomment{search} tree is found with an alternate solution or the plan
+is relaxed since there is no other alternative solution. We do this by
+removing all goals while keeping only recent observations.
 
@@
 
-A key aspect of the \rx framework is interleaved plan and
+A key aspect of the \rx framework is interleaved planning and
 execution. Both {\em synchronization} -- keeping track of execution --
 and {\em deliberation} -- dealing with reactor planning -- of one
-reactor are sharing the same {\em single} plan database which 
-they modify alternatively as the \rx agent schedule their execution.
-
-As noted above, it is synchronization which drives planning, so the
-co-mingling between these two processes is critical to maintaining
-state within a robot especially when it is immersed in a dynamic
-environment. Further, it is important that \rx maintain
-% A core aspect of this reactor is that it takes advantage of the
-% specific sequencing enforced by the \rx architecture along with the
-% fact that only 
-one $plan$ structure shared between synchronization and planning. This
-allows these two processes to not only be interleaved for agent
-responsiveness, but ensures a viable mechanism to communicate.
-
-% but also impact
-% each other via plan manipulation.
-
-As noted earlier, synchronization propagates {\em
+reactor \kcomment{share} the same {\em single} plan database which
+they modify alternatively as the \rx agent schedules their execution.
+
+As \kcomment{shown in Section \ref{sec:arch:synch}}, it is
+synchronization which drives planning; so the co-mingling between
+these two processes is critical to maintaining state within a robot
+especially when it is immersed in a dynamic environment. Further, it
+is important that \rx maintain one $plan$ structure shared between
+synchronization and \kcomment{deliberation}. This allows these two
+processes to not only be interleaved for agent responsiveness, but
+ensures a viable mechanism to communicate.
+
+As \kcomment{also} noted earlier, synchronization propagates {\em
   external} observations when a new tick occurs. Earlier, we isolated
@@
 doing so we were able to develop the way we augment the \eu solver to
-propagate {\em external} observations to identify {\em
-  internal} state. This allowed us to focus on synchronization as a
-state evaluation process that can be resolved within the \eu framework
-with few extensions. Similarly we presented deliberation as an
-abstraction without discussing the disruptive nature of
-synchronization and how it impacts the planning process. We first
-analyze the case when synchronization occurs after planning has
-completed (\ie the last step of deliberation results in a plan that is
-considered complete for the current look-ahead). As synchronization
-starts with a plan, this deliberation will not only be a model-derived
-estimation uniquely based on {\em external} observations but will also
-include reactor intentions specified in the available partial plan.
+propagate {\em external} observations to identify {\em internal}
+state. This allowed us to focus on synchronization as a state
+evaluation process that can be resolved within the \eu framework with
+few extensions. Similarly we presented deliberation as an abstraction
+without discussing the disruptive nature of synchronization and how it
+impacts the planning process. \kcomment{That discussion occurs here in
+  this section.}
+
+We first analyze the case when synchronization occurs after planning
+has completed (\ie the last step of deliberation results in a plan
+that is considered complete for the current look-ahead). As
+synchronization starts with a plan, this deliberation will not only be
+a model-derived estimation uniquely based on {\em external}
+observations but will also include reactor intentions specified in the
+available partial plan.
 
@@
       indicate observations, tokens in {\color{blue}blue} the goal
-      that this plan attempts to solve for. Arrows indicate temporal
-      constraints. }
+      that this plan attempts to solve for \kcomment{and tokens in
+        {\color{grey}grey} the result of deliberation}. Arrows
+      indicate temporal  constraints \kcomment{with $==$ indicating 'equality' and $\geq$
+        indicating 'greater than equal to' in timepoints}.}
   \label{fig:shop:exec}
@@
 evaluate a solution that assumes that the observation is consistent
-with what was decided by the reactor.
-
-% \begin{figure}[!htb]
-%   \centering
-%   \includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t1}
-%   \caption{Shopping example : Result of synchronization after the {\em
-%       Going} observation was received by the reactor with the plan
-%     showed in Fig. \ref{fig:shop:exec0}.}
-%   \label{fig:shop:exec1}
-% \end{figure}
-
-This choice is directed not only by heuristic choices made in the
-solver but also enforced by the heuristic order of the decision point
-resolution choices for the {\em current state flaw} shown in
+with what was decided by the reactor. \kcomment{In the event of
+  inconsistency, the (plan) expectations do not match the reality;
+  then partial plan failure within the reactor is initiated.} This is
+directed not only by heuristic choices made in the solver but also
+enforced by the heuristic order of the decision point resolution
+choices for the {\em current state flaw} \kcomment{as} shown in
 Proposition \ref{prop:csf:resolve}. As noted, the first two choices
 are biased towards a conservative view with respect to the existing
-plan. The heuristics first attempt to extend the previous observation
-for this {\em internal} timeline to avoid early commitment. If this
-fails \rx attempts to start the next existing token in the plan to
-advance plan execution. Both choices attempt to resolve state with the
-least amount of plan perturbation by constraining further the start
-and end time of existing tokens generated by previous deliberations
-steps. By doing so, in a nominal situation, synchronization remains
-limited to restrict time points in the plan and propagates their
-impact. Most of the computational effort of creating new tokens in the
-plan and scheduling them onto timelines are managed during
-deliberation steps. Consequently, most of the time synchronization is
-limited to very few nodes in the \eu search space often requiring no
-backtracking. Fig. \ref{fig:shop:exec} shows how the existing plan can
-support the synchronization process. The observation \texttt{Going}
-occurring at $\tau'$ is fully compatible with the token already present
-in our plan. By updating the plan and propagating equality constraints,
-the resolution of the {\em current state flaw} for the {\em Agent}
-timeline becomes obvious as a solution, as the token \texttt{Go} from
-the plan presents itself.
+plan.
+
+\kcomment{\subsubsection{Dealing with partial plan failures}}
+\label{sec:arch:partialfailures}
+
+The \kcomment{solver's} heuristics first attempt to extend the
+previous observation for this {\em internal} timeline to avoid early
+commitment. If this fails \rx attempts to start the next token in the
+plan to advance plan execution. Both choices attempt to resolve state
+with the least amount of plan perturbation by constraining further the
+start and end time of existing tokens generated by previous
+deliberations steps. By doing so, in a nominal situation,
+synchronization remains limited to restricting time points in the plan
+and propagates their impact. Most of the computational effort of
+creating new tokens in the plan and scheduling them onto timelines are
+managed during deliberation steps. Consequently, most of the time
+synchronization is limited to very few nodes in the \eu search space
+often requiring no backtracking. Fig. \ref{fig:shop:exec} shows how
+the existing plan can support the synchronization process. The
+observation \texttt{Going} occurring at $\tau'$ is fully compatible
+with the token already present in our plan. By updating the plan and
+propagating equality constraints, the resolution of the {\em current
+  state flaw} for the {\em Agent} timeline becomes obvious as a
+solution, as the token \texttt{Go} from the plan presents itself.
 
@@
   \centering
-  \subfloat[Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
-  \subfloat[Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \subfloat[\small Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
+  \subfloat[\small Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
   \caption{\small Illustration of a conflict during synchronization

### 2012-05-01 europa.tex
updates

@@
 \label{sec:europa:arch}
-
-\comment{I think we should expand on this section which, as the
-  reviewer notes, the inter-relationships between the elements in Fig
-  \ref{fig:europa-architecture} below is not fully captured. I think
-  an illustrative example of how these components work together might
-  be very useful. This might also help the sub-section end less
-  abruptly than it is currently.}
 
@@
 object, variable and token instances. This constitutes the data
-foundation on which end user applications and automated problem solvers
- are built. The \textbf{Rules Engine} (\texttt{RE})
-modularly adds the semantics of domain rules that describe
-dependencies between tokens. The \textbf{Solver Module} provides the
-framework for flaw detection and resolution that is needed to
-implement search algorithms for planning. The \textbf{API Layer}
-provides access to services from all the different modules in a
-programmatic way so that \eu can be easily embedded into specialized
-applications.
+foundation on which end user applications and automated problem
+solvers are built. The \textbf{Rules Engine} (\texttt{RE}) modularly
+adds the semantics of domain rules that describe dependencies between
+tokens. The \textbf{Solver Module} provides the framework for flaw
+detection and resolution that is needed to implement search algorithms
+for planning. The \textbf{API Layer} provides access to services from
+all the different modules in a programmatic way so that \eu can be
+easily embedded into specialized applications.
 
@@
 languages that support those definitions in a much more natural and
-compact way. \texttt{NDDL} (New Domain Description Language
+compact way. \nd (New Domain Description Language
 [pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
@@
 
-Before diving into the details of each module, a simple example illustrates how they interact at a high level:
+\kcomment{Before going into the details of each module, a simple
+  example illustrates how they interact at a high level:}
 
@@
 
-The snippet above is a NDDL model of a light bulb and a switch that controls it (NDDL is covered in more detail below), along with a problem instance that requires the light bulb to be off by time=10, given an initial state with the bulb on at time=0.
-The NDDL module allows \eu to ingest this model and problem description.
-Once the NDDL statements are parsed, the different object types, object instances and constraints are instantiated inside the Plan Database and Constraint Engine modules where they are available for both the user and the solver module to query and manipulate programmatically.
-This instantiation of the initial state constitutes the initial plan that \eu holds for this particular problem instance.
-The user can then invoke functionality available in the solver module to modify or extend the initial plan in a way that eliminates any flaws (violated constraints, unsatisfied goals, oversubscribed resources) in the current plan.
-When actions are added to a plan, the Rules Engine module ensures that domain rules expressed in the model are enforced when \eu creates new action or predicate instances, for instance, if a LightSwitch::turnOn action is added to the plan, the RulesEngine  ensures that requirements for the existence of the conditions and effects of that action (that is, the corresponding LightBulb must be off before and on after the action) are posted through the PlanDatabase and ConstraintEngine modules.
+\kcomment{The snippet above is a \nd (covered in more detail in
+  Section \ref{sec:europa:modeling}) model of a light bulb and a
+  switch that controls it, along with a problem instance that requires
+  the light bulb to be off by time$=10$, given an initial state with
+  the bulb on at time$=0$. The \nd module allows \eu to ingest this
+  model and the problem description. Once the \nd statements are
+  parsed, different object types, object instances and constraints are
+  instantiated inside the \texttt{PDB} and \texttt{CRE} modules where
+  they are available for both the user and the solver module to query
+  and manipulate programmatically. This instantiation of the initial
+  state constitutes the initial plan that \eu holds for this
+  particular problem instance.
+
+  The user can then invoke functionality available in the solver
+  module to modify or extend this initial plan in a way that
+  eliminates any flaws (violated constraints, unsatisfied goals,
+  oversubscribed resources). When actions are added to a plan, the
+  \texttt{RE} module ensures that domain rules expressed in the model
+  are enforced when \eu creates new action or predicate instances. For
+  instance, if a \texttt{LightSwitch::turnOn} action is added to the
+  plan, the \texttt{RE} ensures that requirements for the existence of
+  the conditions and effects of that action (that is, the
+  corresponding \texttt{LightBulb} must be off before and on after the
+  action) are posted through the \texttt{PDB} and \texttt{CRE}
+  modules.  }
 
@@
 language for models, goals and plans is of great importance. For \eue,
-\texttt{NDDL} allows the user to specify the different components of a
-problem domain in a precise and concise manner. Among the main
-features that make \texttt{NDDL} a powerful tool for describing
-problem domains are that it is object oriented. \texttt{NDDL} supports
-classes and inheritance in similar fashion to popular Object-Oriented
-languages like Java and C++. It also supports polymorphism for some of
-the planning components. Using object classes and instances is a
-time-tested approach to naturally describe problem domains
-\cite{rumbaugh}. It offers declarative constructs to define
-constraints and causality dependencies between actions and effects and
-finally it offers procedural constructs to populate \eus \texttt{PDB}
-with specific problem instances expressed in terms of objects,
-variables, constraints, facts and goals. To articulate the most important
-elements of \texttt{NDDL}, we show how the Shopping Agent problem
-could be modeled:
+\nd allows the user to specify the different components of a problem
+domain in a precise and concise manner. Among the main features that
+make \nd a powerful tool for describing problem domains are that it is
+object oriented. \nd supports classes and inheritance in similar
+fashion to popular Object-Oriented languages like Java and C++. It
+also supports polymorphism for some of the planning components. Using
+object classes and instances is a time-tested approach to naturally
+describe problem domains \cite{rumbaugh}. It offers declarative
+constructs to define constraints and causality dependencies between
+actions and effects and finally it offers procedural constructs to
+populate \eus \texttt{PDB} with specific problem instances expressed
+in terms of objects, variables, constraints, facts and goals. To
+articulate the most important elements of \nde, we show how
+the Shopping Agent problem could be modeled:
 
@@
 
-\item[\textbf{Object Types}:] \texttt{NDDL} supports an object-oriented
+\item[\textbf{Object Types}:] \nd supports an object-oriented
   approach, where common data and behavior can be abstracted as a
@@
   \textit{AgentLocation} class to keep track of where the agent is at
-  any point in time. \texttt{NDDL} supports a single rooted class
+  any point in time. \nd supports a single rooted class
   hierarchy so that state and behavior can be extended through
@@
 
-Domain rules can be defined in \texttt{NDDL} in the context of a
-predicate or an action. In the example above we can see the main
-elements of a rule definition:
+Domain rules can be defined in \nd in the context of a predicate or an
+action. In the example above we can see the main elements of a rule
+definition:
 
@@
 \ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
-types included in \eus Constraint Library.
+types included in \eus Constraint Library.  Temporal and resource
+constraints are especially important to model real-world planning
+domains.
 
 \input{constraints-table}
-
-Temporal and resource constraints are especially important to model
-real-world planning domains.
 
@@
   \centering
-  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
-  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
+  \subfloat[\small A Simple Temporal Network (STN). \kcomment{Nodes correspond
+  to variables and their domains and arc's the \emph{constraint} or
+  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
+  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
   \caption{\small An illustration of a STN and its corresponding
@@
 actions that typically depend upon limited resources (\eg fuel,
-storage capacity, battery state of charge, data buffers).  \eu
-provides built-in support for resource computation. In \eus
-representation resource object types come in three forms with related
-constraints. \textbf{Reservoir} resources can have both
+storage capacity, battery state of charge, data buffers etc).  % \eu
+% provides built-in support for resource computation.
+In \eus representation, resource object types come in three forms with
+related constraints. \textbf{Reservoir} resources can have both
 \texttt{consume(resource,time,quantity)} and
@@
 Since \eu supports a flexible temporal representation, where time is
-represented by intervals, instead of single values, determining
-whether a resource is the source of a plan flaw or a constraint
-violation in general is substantially more difficult than just
-computing a profile of the resource and checking against some limits.
-With temporal flexibility, a number of resource profiles could be
-possible depending on precisely when each production/consumption event
-occurs out of a range of possible values that are in the domain of the
+represented by intervals, instead of singletons, determining whether a
+resource is the source of a plan flaw or a constraint violation in
+general is substantially more difficult than just computing a profile
+of the resource and checking against some limits.  With temporal
+flexibility, a number of resource profiles could be applicable
+depending on precisely when each production/consumption event occurs
+out of a range of possible values that are in the domain of the
 variables that represent the temporal extent for the event.
@@
 bound. Since maximum flow on a temporal network can be computed in
-polynomial time \comment{definition needed?}, \eu takes advantage of
-this algorithm to efficiently compute upper and lower bound envelopes
-on the resource level. Fig. \ref{fig:resenvelopes} shows an example of
-such a resource computation.
+polynomial time \cite{corman}, \eu takes advantage of this algorithm
+to efficiently compute upper and lower bound envelopes on the resource
+level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
+computation.
 
@@
 \centering
-\includegraphics[scale=0.3]{figs/europa-resource-envelopes.png}
+\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
 \caption{\small An example of resource envelopes computed by \eu.}
@@
   \BlankLine
-  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
+  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \comment{\{\emph{potential branching point(s)}\}}\;\label{li:decisionpt}
   \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
@@
     \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
-    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
+    \lElse{ $decision.undo()$ \comment{\{\emph{possible backtracking point}\}}\; } \label{li:backtrack}
    }
@@
  }
-\caption{$\mathrm{bool} ~ solve(plan)$}
+\caption{$\mathrm{bool} ~ solve(plan)$\kcomment{: To deliberate and
+    generate partial plans}}
 \label{alg:europa:solve}

### 2012-05-01 trunk/results.tex
updates

@@
 fly, more imaginative methods of sample acquisition were targeted.
-\kcomment{The Dorado uses two onboard computers: a main vehicle computer which
-is a 244 Mhz PC/104 stack running the QNX real-time operating system,
-and a 367 MHz EPIC EPX-GX500 AMD Geode stack running Linux and \rxe. A
-socket-based communication protocol allows the exchange of goals and
-state updates between \rx and the lower level functional layer.}
+\kcomment{The Dorado uses two onboard computers: a main vehicle
+  computer which is a 244 Mhz PC/104 stack running the QNX real-time
+  operating system, and a 367 MHz EPIC EPX-GX500 AMD Geode stack
+  running Linux and \rxe. A socket-based communication protocol allows
+  the exchange of goals and state updates between \rx and the lower
+  level functional layer. Event logs from field experiments show that
+  a peak load of up to $20\%$ occurs during re-planning. Other desktop
+  simulations \cite{mcgann09} show that partitioning scales with
+  problem size.  Reactor design and the number of partitions however
+  are still problem specific and determined empirically.}
 
@@
 identification. The temporal probabilistic transition then acts as a
-filter on top of the clustering model since it is not only derived
-from it, but it informs the planner of a potential trigger state only
-when the models' transition occurs to an appropriate state of
-recognition of the INL feature \cite{mcgann08b}. Not only was
-deliberation aided by direct perception of the state (of the INL) but
-temporal transition functions added a further level of validation
-between expected states in the INL model. This coupling of Machine
-Learned model with a deliberative planner is novel in the ocean
-sciences and is a continuing direction of investigation
-\cite{kumar11}.
+filter on top of the clustering model \kcomment{from \cite{fox2007},}
+since it is not only derived from it, but it informs the planner of a
+potential trigger state only when the models' transition occurs to an
+appropriate state of recognition of the INL feature
+\cite{mcgann08b}. Not only was deliberation aided by direct perception
+of the state (of the INL) but temporal transition functions added a
+further level of validation between expected states in the INL
+model. This coupling of Machine Learned model with a deliberative
+planner is novel in the ocean sciences and is a continuing direction
+of investigation \cite{kumar11}.
 
@@
 synthesis occurs frequently, but within the context of a drifting
-point generated by a drifter. Fig. \ref{fig:dta-setup} shows the
-general setup of our experiments which can be tailored with the use of
-various computational methods within the \texttt{DTA} app which is
-contained within the Oceanographic Decision Support System (\od; see
-below) \cite{das11}. GPS data from the free-floating drifter is sent
-via satellite and processed onshore to determine a projection (speed
-and bearing) of the drifter. This data is sent to the AUV; at the
-completion of the current survey, the vehicle receives this data as a
-high-level goal which is decomposed and synthesized as a survey
+point generated by a drifter. \kcomment{A drifter tracking application
+  (\texttt{DTA}) is used on-shore with an Oceanographic Decision
+  Support System (\ode; see below) \cite{das11}, to process the
+  drifter location from its onboard GPS and used as a target for the
+  Dorado vehicle to provide a configurable and dynamic track-line
+  enclosure to provide oceanographic context to the drifting body.}
+Fig. \ref{fig:dta-setup} shows the general setup of our experiments
+which can be tailored with the use of various computational methods
+within the \texttt{DTA} app. GPS data from the free-floating drifter
+is sent via satellite and processed onshore to determine a projection
+(speed and bearing) of the drifter. This data is sent to the AUV; at
+the completion of the current survey, the vehicle receives this data
+as a high-level goal which is decomposed and synthesized as a survey
 plan. Currently, the vehicle completes the survey without any further
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs/june11-drifter-follow-180-181.pdf}
+\includegraphics[width=0.45\textwidth]{figs/june11-drifter-follow-180-181.pdf}
 \caption{\small{A June 2011 drifter cruise shown in the Earth
@@
 \centering
-\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.38\textwidth]{figs/sep10-canon-cruise.png}}
-\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.35\textwidth]{figs/september-iterations.pdf}}
+\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.4\textwidth]{figs/sep10-canon-cruise.png}}
+\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.40\textwidth]{figs/september-iterations.pdf}}
   \caption{\small{The offshore \can ESP-drifter following cruise
@@
 
-These theoretic results were validated in a five-day long off-shore
-deployment carried out in September 2010, 160 kms off the California
-coast (Fig.~\ref{fig:sep10-ship}). A specialized drifter with a
-genomic sensor \cite{scholin09} hanging $20$ m below, performing
-in-situ identification of micro-organisms was the target. The
-experiment was supported by crews on two support vessels, the R/V
-\emph{Western Flyer} and the R/V \emph{Zephyr}. The \emph{Flyer}
-visited the drifter every four hours to carry out a series of
-ship-based sampling experiments and lab analysis on water samples to
-ground-truth the drifter sensor data. The \emph{Zephyr} was meanwhile
-focused on Lagrangian observation studies using the Dorado with the
-goal to monitor the nutrient budget at the perimeter of a $1$km $X
-1$km water patch around the advecting drifter while the AUV performed
-a transformed box pattern around the drifter. A number of logistical
-issues were kept in mind while designing and executing the
+These theoretic\kcomment{al} results were validated in a five-day long
+off-shore deployment carried out in September $2010$, $160$ kms off
+the California coast (Fig.~\ref{fig:sep10-ship}). A specialized
+drifter with a genomic sensor \cite{scholin09} hanging $20$ m below,
+performing in-situ identification of micro-organisms was the
+target. The experiment was supported by crews on two support vessels,
+the R/V \emph{Western Flyer} and the R/V \emph{Zephyr}. The
+\emph{Flyer} visited the drifter every four hours to carry out a
+series of ship-based sampling experiments and lab analysis on water
+samples to ground-truth the drifter sensor data. The \emph{Zephyr} was
+meanwhile focused on Lagrangian observation studies using the Dorado
+with the goal to monitor the nutrient budget at the perimeter of a
+$1$km $X 1$km water patch around the advecting drifter while the AUV
+performed a transformed box pattern around the drifter. A number of
+logistical issues were kept in mind while designing and executing the
 experiment. Each iteration began with the latest drifter update
@@
   Earth post-hoc image shows {\color{red}red} track lines of the \rx
-  commanded Dorado with the {\color{yellow}yellow} transects of the
-  LRAUV. {\color{yellow}Yellow} dots show the track line of a
-  GPS-enabled drifter used for seeding the experiment over the
-  Monterey Canyon with Santa Cruz, California to the right of the
+  commanded Dorado with the {\color{yellow}yellow} transects of a
+  separate AUV platform. {\color{yellow}Yellow} dots show the track
+  line of a GPS-enabled drifter used for seeding the experiment over
+  the Monterey Canyon with Santa Cruz, California to the right of the
   figure.}
@@
 \centering
-\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.3\textwidth]{figs/october-canon-meeting.png}}
-\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.33\textwidth]{figs/virtual-drifter-tool.png}}
-\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
+\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.325\textwidth]{figs/october-canon-meeting.png}}
+\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.37\textwidth]{figs/virtual-drifter-tool.png}}
+\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.325\textwidth]{figs/photo-ipad-2.jpg}}
 \caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
@@
 awareness database, incorporate model outputs projecting these outputs
-into the space-time region of interest to generates candidate
+into the space-time region of interest to generate candidate
 hypotheses with associated utility metrics. With the scientist's

### 2012-05-02 concepts.tex
updates

@@
 demonstrations for space telescope scheduling as noted in Section
-\ref{sec:rabeyond}. \texttt{HSTS} was also the
-\texttt{LISP}-based\footnote{This was the first (and to our knowledge
-  only) software ever to been written in \texttt{LISP} to be flown in
-  space.} planner that was flown as part of the Remote Agent
-\texttt{RAX} experiment. Concurrent to \texttt{RAX} deployment, a C++
-version of the planner was implemented which went on to be used for
-the \texttt{MAPGEN} \cite{bresina05} system and continues to be used
-to date, in the mission-critical uplink process for the
+\ref{sec:rabeyond}. \texttt{HSTS} was also the \texttt{LISP}-based
+planner that was flown as part of the Remote Agent \texttt{RAX}
+experiment. Concurrent to \texttt{RAX} deployment, a C++ version of
+the planner was implemented \kcomment{at NASA Ames. This} went on to
+be used for the \texttt{MAPGEN} \cite{bresina05} system and continues
+to be used to date, in the mission-critical uplink process for the
 command/control of the Mars Exploration Rovers (\texttt{MER})
@@
 
-\eu uses a \emph{domain model} written in a declarative language,
-together with initial conditions and goals (also in that specified
-language), to construct a set of temporal relations that must be true
-at start time. These models include assertions about the physics of
-the vehicle, i.e how it responds to external stimulus and internally
-driven goals. By propagating these relations forward using Simple
-Temporal Networks \cite{dechter91} and applying goal constraints, \eu
-can select a set of conditions that should be true in the future,
-where some of these conditions will correspond to actions the agent
-must take. The planner can backtrack and try another path during the
-search process if a goal cannot be reached while being capable of
-discarding unachievable goals.
+\eu uses a \emph{domain model} \kcomment{(see Fig. \ref{fig:planner})}
+written in a declarative language, together with initial conditions
+and goals (also in that specified language), to construct a set of
+temporal relations that must be true at start time. These models
+include assertions about the physics of the vehicle, i.e how it
+responds to external stimulus and internally driven goals. By
+propagating these relations forward using Simple Temporal Networks
+\cite{dechter91} and applying goal constraints, \eu can select a set
+of conditions that should be true in the future, where some of these
+conditions will correspond to actions the agent must take. The planner
+can backtrack and try another path during the search process if a goal
+cannot be reached while being capable of discarding unachievable
+goals.
 
@@
 interpreted a \textit{temporally flexible plan} which represents each
-start time as a variable and contains an explicit network of bounded
-delay constraints between such
-variables. Fig. \ref{fig:flex-timelines} shows an example of a
-flexible partial plan.
+start time as a \kcomment{flexible timepoint} variable
+\kcomment{backed by} an explicit network of bounded delay constraints
+between such variables. \kcomment{These timepoints \cite{Dean88}
+  represent temporal intervals signifying a change in state and are of
+  the form $[lb,ub]$ (where $lb, ub \in N$) are temporal lower and
+  upper bounds respectively. Instead of specifying a fixed integer,
+  time is represented as an interval}. Fig. \ref{fig:flex-timelines}
+shows an example of a flexible partial plan.
 
@@
 flexibility is critical when dealing with dynamic and uncertain
-conditions such as in the ocean, where precise timing of a robotic
-action might be indeterminate. Further, the advantage of flexibility
-can be contrasted with the consequences of the intrinsic inflexibility
-of traditional command sequences. Because they are inflexible,
-sequences must necessarily be designed considering worst case
-scenarios.
+conditions \kcomment{for robotic action in real-world environments},
+where precise timing of an action might be indeterminate. Further, the
+advantage of flexibility can be contrasted with the consequences of
+the intrinsic inflexibility of traditional command sequences. Because
+they are inflexible, sequences \kcomment{are brittle and therefore}
+must necessarily be designed considering worst case scenarios.
 
@@
 \cite{mus94,jonsson00,py10} to describe the evolution of state over
-time. The instantiated history of such state variable evolution over a
+time. The instantiated history of such \kcomment{an} evolution over a
 temporal horizon we call \kcomment{as a} \emph{timeline} and
@@
 concurrent system. At any given time each thread can execute a single
-procedure.  Thus each timeline consists of a sequence of procedures
-which encapsulate and describe state evolution; we call these
-instantiated atomic entities \emph{tokens}.  A token therefore
+procedure\kcomment{, with} each timeline consisting of a sequence of
+procedures which encapsulate and describe state evolution. We call
+these instantiated atomic entities \emph{tokens}.  A token therefore
 describes a procedure invocation, the state variables on which it can
-occur, the parameter values of the procedure, and the time values
-defining the interval. We allow encapsulation of uncertainty within
-these tokens with a range of start and end times and parameters, all
-of which are encoded as variables. A constraint solver in turn
-manipulates these variables defined in a \eu domain model. For
-example, consider a scientific need to take water samples 100 meters
-from an environmental hotspot while an AUV is performing a Yo-Yo
-sawtooth pattern\footnote{\kcomment{Variability in the water column is
-    along the vertical dimension. Since the Dorado platform can only
-    move forward, the Yo-Yo pattern is the most efficient mechanism
-    for studying water column properties.}}  in the water-column. Two
-samples are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
+occur, the parameter values of the procedure, and the
+\kcomment{timepoints} defining the interval \kcomment{(see
+  Fig. \ref{fig:flex-timelines})}. We allow encapsulation of
+uncertainty within these tokens with a range of start and end times
+and parameters, all of which are encoded as variables. A constraint
+solver in turn manipulates these variables defined in an \eu domain
+model. For example, consider a scientific need to take water samples
+100 meters from an environmental hotspot while an AUV is performing a
+Yo-Yo sawtooth pattern\footnote{\kcomment{Variability in the water
+    column is along the vertical dimension. Since the Dorado platform
+    can only move forward, the Yo-Yo pattern is the most efficient
+    mechanism for studying water column properties.}}  in the
+water-column. Two samples are needed if the feature's signal is above
+a threshold; one otherwise. The token that is capturing the sensory
+threshold has a parametric constraints to the token which
+\kcomment{triggers} the requisite water sampler. In addition, the
+start time of the water sampling procedure is highly dependant on the
+variability of sub-sea currents and actual speed of the
+vehicle. Therefore a number of values are possible for the start times
+and duration of the sampling all of which are valid combinations for
+desired outcomes.
 
@@
 
-\eu follows the representation outlined by the \texttt{CAIP} framework
-augmenting with elements that are necessary for an efficient
-implementation of constraint-based planning. Below we describe briefly
-the main elements, followed by a detailed description for each of
-these elements and how together they constitute a rich
-representational paradigm for inference in Sections
-\ref{sec:europa:modeling}, \ref{sec:europa:inference} and
-\ref{sec:europa:search}.
+\eu follows the representation outlined by \texttt{CAIP} augmented
+with elements that are necessary for an efficient implementation of
+constraint-based planning. We briefly describe the main elements,
+followed by a detailed description for each of these elements and how
+together they constitute a rich representational paradigm for
+inference in Sections \ref{sec:europa:modeling},
+\ref{sec:europa:inference} and \ref{sec:europa:search}.
 
@@
 \item \textbf{Tokens} Temporally scoped entities that correspond to
-  Intervals in the \texttt{CAIP} framework, are used to represent
-  state and actions in time. A\textbf{Token State model} is defined to
-  support an efficient implementation of plan-space search algorithms.
+  Intervals in \texttt{CAIP} that are used to represent state and
+  actions in time. A\textbf{Token State model} is defined to support
+  an efficient implementation of plan-space search algorithms.
 
@@
   modular and scaleable way. \textbf{Timelines and Resources} are
-  built in implementations of Object types that appear frequently in
+  built in implementations of object types that appear frequently in
   real-world domains.
@@
 \item[\textbf{Variables}] Values that need to be represented to
-  describe the problem domain and over which we may want to specify
-  constraints. In the Shopping Agent example, the times at which the
+  describe the problem domain \kcomment{over which constraints are
+    specified}.  In the Shopping Agent example, the times at which the
   agent needs to leave or be back, or executes a purchase, are all
-  instances where variables representing time is used.  As noted in
-  Section \ref{sec:europa:cp}, a variable can take values from a
-  domain. In \eu, domains are defined over a specific data type, the
-  primitive data types supported are, \textit{String},
+  instances where \emph{variables} representing time is used.  As
+  noted in Section \ref{sec:europa:cp}, a variable can take values
+  from a domain. In \eu, domains are defined over a specific data
+  type; the primitive data types supported are \textit{String},
   \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
   below). Further, two types of variable Domains can be defined over a
-  data type, \textit{Enumerated} and \textit{Interval} (used only for
-  numeric data types, they are specified by a [LowerBound,UpperBound]
-  pair, for instance: [1,10], [5.0,100.0]). It is this latter ability
-  to represent and reason over numeric intervals that forms the basis
-  of representing flexible plans.
+  data type \textit{Enumerated} and \textit{Interval} (used only for
+  numeric data types, they are specified by a
+  $[LowerBound,UpperBound]$ pair, for instance: $[1,10]$,
+  $[5.0,100.0]$). It is this latter ability to represent and reason
+  over numeric intervals that forms the basis of representing flexible
+  plans.
 
-\item[\textbf{Constraints}] valid plans in most problem domains have
-  to satisfy a number of restrictions, for instance, a Shopping Agent
+\item[\textbf{Constraints}] Valid plans in most problem domains have
+  to satisfy a number of restrictions for instance, a Shopping Agent
   may only be able to perform one task at a time, may have limited
-  time or an energy or fuel budget to perform a task. In \eu, these
-  restrictions are represented by constraints, which can be defined
-  over any combination of variables in a domain. \eu provides a
-  \emph{Constraint Library} that implements a number of useful
-  constraints (temporal, resource, relational), while allowing the
-  possibility to add new constraints if required by a particular
-  domain.
+  time or an energy or fuel budget to perform a task. These
+  restrictions are represented by \emph{constraints}, which can be
+  defined over any combination of variables in a domain. \eu provides
+  a \emph{Constraint Library} that implements a number of useful
+  constraints (temporal, resource, relational) while allowing the
+  possibility to \kcomment{augment} with new constraints if required.
 
 \item[\textbf{Objects}] The items we wish to describe and refer to in
-  a domain are considered objects. % As in the case with object-oriented
-  % analysis and design, one can seek out the nouns in any domain
-  % description to find likely objects. ---- this seemed confusing
-  In the Shopping Agent example, we might consider the Agent, Products
-  and Locations all to be objects. Objects have state and behavior.
-  For example, a Shopping Agent can have its location, a bag that
-  contains the products it has already purchased (the bag can in turn
-  be another object) and behavior, as in going to a location to look
-  for a product, perform a purchase, return home, etc. all as objects.
+  a domain are considered objects.  In the Shopping Agent example, we
+  might consider the Agent, Products and Locations all to be
+  \emph{objects}. Objects have state and behavior.  For example, a
+  Shopping Agent can have its location, a bag that contains the
+  products it has already purchased (the bag can in turn be another
+  object) and behavior, as in going to a location to look for a
+  product, perform a purchase, return home, etc. all as objects.
   Those that have similar state and behavior can be described
@@
   that agent $a$ is at location $l$, or a predicate
-  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking action to
-  buy product $p$. Note that $a$ is a variable which may have a number
-  of possible values in a problem with multiple agents. Similarly, $l$
+  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is acting to buy
+  product $p$. Note that $a$ is a variable which may have a number of
+  possible values in a problem with multiple agents. Similarly, $l$
   and $p$ are variables whose values are the set of possible positions
@@
   where the domains of variables $a$ and $l$ can have more than one
-  possible value. Or they can be grounded plans, where single values
-  will be specified for each variable as showing in the N-queen's
-  problem.
+  possible value. Or they can be grounded \kcomment{representations},
+  where single values will be specified for each variable as shown in
+  the $N$-Queens problem.
 
-  In general, for an executable plan it is insufficient to state
-  predicates that describe required state or behavior without also
-  specifying some temporal extent over which each of those predicates
-  hold. Predicates that are always true can be thought to hold from
-  the beginning to the end of the planning horizon. However, in
-  practice, the temporal extent of interest must be defined with
-  timepoints \cite{Dean88} to represent its start and end.  In such
+  In general, for an executable plan it is insufficient to
+  \kcomment{define} predicates that describe required state or
+  behavior without also specifying some temporal extent over which
+  each of those predicates hold. Predicates that are always true can
+  be thought to hold from the beginning to the end of the planning
+  horizon. However, in practice, the temporal extent of interest must
+  be defined with timepoints to represent its start and end.  In such
   \kcomment{a} representation, one writes \texttt{At($a$,$s$,$e$,$l$)}
@@
   overlap may occur among them \cite{mus94}. Fig. \ref{fig:europapr1}
-  below and several others after that illustrate how timelines are
+  and others \kcomment{in this chapter,} illustrate how timelines are
   used to represent state in a plan.
@@
 \item \textit{Resources}: Metric resources, e.g. the energy of a
-  battery or the capacity of a cargo hold, are objects with an
-  explicit quantitative state in time and with a circumscribed range
-  of changes that can occur to impact that state i.e. produce,
-  consume, use, change.  Resources are such a common requirement for \eu
-  users that built-in object types (with their corresponding token
-  types to denote production, consumption, etc) are provided.
-  Instances of classes derived from a resource will induce ordering
-  requirements on their tokens in order to ensure that the level of
-  the resource remains within specified limits.
+  battery or the capacity of a cargo hold are objects with an explicit
+  quantitative state in time and with a circumscribed range of changes
+  that can occur to impact that state i.e. produce, consume, use,
+  change.  Resources are a common enough requirement for \eu users
+  that built-in object types (with their corresponding token types to
+  denote production, consumption, etc) are provided.  Instances of
+  classes derived from a resource will induce ordering requirements on
+  their tokens in order to ensure that the level of the resource
+  remains within specified limits.
 
@@
   asks the Shopping Agent to own a specific product by a specific
-  time, or implicitly when subgoals are created as a result of domain
-  rules. For instance, when the Shopping Agent must get to a location
-  where a product is available before it is able to purchase it: in
-  this case the original goal of purchasing the product spawns a
-  subgoal of being at a specific location without which the purchase
-  will not occur as defined in the domain model.  Looking at how a
-  simple Shopping Agent goal would be addresed by \eu helps illustrate
-  how Token States support the planning process.  Let's assume the
-  planner starts with a partial plan that places the Shopping agent at
-  home at time $0$ with no products in possession. Let's also assume
-  that goals are posted for the Agent to own a drill by time $10$ and
-  to be home by time $20$ as shown in Fig. \ref{fig:europapr1}
+  time, or implicitly when subgoals are created as a consequence of
+  domain rules. For instance, the Shopping Agent must get to a
+  location where a product is available before it is able to purchase
+  it. In this case the original goal of purchasing the product spawns
+  a \emph{subgoal} of being at a specific location without which the
+  purchase will not occur as defined in the domain model.
+
+  Looking at how a simple Shopping Agent goal would be addresed by \eu
+  helps illustrate how Token States support the planning process.
+  Let's assume the planner starts with a partial plan that places the
+  Shopping agent at home at time $0$ with no products in
+  possession. Let's also assume that goals are posted for the Agent to
+  own a drill by time $10$ and to be home by time $20$ as shown in
+  Fig. \ref{fig:europapr1}
 
@@
   represent the goals are not immediately placed on the
-  \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu must
-  first determine whether it can achieve those \kcomment{goals} and
-  how. When the goals are posted and before any decisions are made,
-  the \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in
-  an \texttt{INACTIVE} state. \eu then considers the partial plan and
-  evaluates two possible alternatives:
+  \texttt{AgentBag} or \texttt{AgentLocation} timelines, since \eu
+  must first determine whether it can achieve those \kcomment{goals}
+  and how. When the goals are posted and before any decisions are
+  made, the \texttt{Own(Drill)} and \texttt{At(Home)} tokens are
+  created in an \texttt{INACTIVE} state. \eu then considers the
+  partial plan and evaluates two possible alternatives:
 
@@
 are transferred to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible. Two
-tokens are merge-compatible if they are instances of the same
-predicate and there are no empty intersections between corresponding
-token variables.  The effect of merging is illustrated in
-Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
-satisfied by merging with the initial state that said that the Agent
-is home at time $0$.  The merge is possible because the tokens are of
-the same type \texttt{At}, and intersecting the domains for all of
-their variables doesn't yield an empty domain (which represents an
-inconsistency), in this case the token variables are the location,
-which is \texttt{Home} in both cases, and the temporal variables
-(start, end, duration), which can be intersected without causing an
+token requires finding a target \texttt{ACTIVE} token that is
+compatible. Two tokens are merge-compatible if they are instances of
+the same predicate and there are no empty intersections between
+corresponding token variables.  The effect of merging is illustrated
+in Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
+satisfied by merging with the initial state that \kcomment{indicated}
+that the Agent is home at time $0$.  The merge is possible because the
+tokens are of the same \texttt{At} type and intersecting the domains
+for all of their variables doesn't yield an empty domain (representing
+an inconsistency). In this case the token variables are the location,
+which is \texttt{Home} in both cases and the temporal variables start,
+end and duration, which can be intersected without causing an
 inconsistency. After the token representing the goal is merged onto
 the token representing the initial state, the resulting token
-represents not only the fact that the Agent must be \texttt{At(Home)}
-by time 0 but also \texttt{At(Home)} at time $20$, this is reflected
-in the new constraints on the token's start and end variables as
-depicted in Fig. \ref{fig:europapr:merge}.
+represents not only that the Agent must be \texttt{At(Home)} by time
+$0$ but also \texttt{At(Home)} at time $20$, This is reflected in the
+new constraints on the token's start and end variables as depicted in
+Fig. \ref{fig:europapr:merge}.
 
@@
 inconsistent with the constraints that resulted after the merge. As a
-result that decision will have to be retracted at some point and the
-alternative to activate or merge with a different token will have to
-be considered.
+result that decision will have to be retracted and the alternative to
+activate or merge with a different token will need to be considered.
 
-If, after looking at all possibilities, neither activating nor merging
+If after looking at all possibilities, neither activating nor merging
 work, a token may be \texttt{REJECTED}. This is possible only for
@@
 token's state can go from \texttt{INACTIVE} to \texttt{ACTIVE, MERGED}
-or \texttt{REJECTED}, as shown in Fig. \ref{fig:europapr:states}.
+or \texttt{REJECTED} as shown in Fig. \ref{fig:europapr:states}.
 

### 2012-05-02 trunk/europa-basics.tex
updates

@@
 unpredictable. Typically this requires our AUVs to balance the goals
-of coverage spatially while opportunistically following features of
-scientific interest and to do so while being aware of its own
-limitations in terms of resources (\eg the battery state of charge)
-and overall proximity to other observational assets for obtaining
-scientific ground-truth. We are interested in representational
-frameworks that allow robots to pursue long-term objectives while
-choosing short-term gain and can gracefully deal with exogenous or
-endogenous change.
+of spatial coverage while opportunistically following features of
+scientific interest and to do so while being aware of its own resource
+limitations (\eg the battery state of charge) and overall proximity to
+other observational assets for obtaining scientific ground-truth. We
+are interested in representational frameworks that allow robots to
+pursue long-term objectives while choosing short-term gain and can
+gracefully deal with exogenous or endogenous change.
 
@@
 to be able to synthesize plans insitu and to re-plan without human
-intervention. To do so we use a temporal constraint-based planner with
-a demonstrated legacy of having flown on NASA space missions. Our
+intervention. \rx uses a temporal constraint-based planner with a
+demonstrated legacy of having flown on NASA space missions. Our
 autonomy architecture brings three key innovations for AUV control:
@@
 a sound theoretical basis and systematic compositional control with
-the use of partitioned networks doing so with a demonstrated
-capability at sea in novel ocean observation methods.
+the use of partitioned networks and doing so with a demonstrated
+capability at sea \kcomment{for} novel ocean observation methods.
 
 The next few sections will cover \eu in a bottom-up fashion, going
-from the generic theoretical framework that underlies \eue, to the
-specific representation, modeling and algorithms implemented by it.  A
-brief introduction to Constraint Programming and Constraint-Based
-Planning is provided followed by the specific representation that \eu
-uses to undertake Constraint-Based Planning.  Subsequently, we provide
-an overview of \eus architecture and how it can be embedded in a
-planning application such as \rxe. Finally, \eus modeling, inference
-and search mechanisms are covered in detail so that the reader can
-understand how \rx can perform deliberative planning and react to
-execution events by embedding and customizing \eu.
+from its \kcomment{general} theoretical framework, to the specific
+representation, modeling and algorithms implemented by it.  A brief
+introduction to Constraint Programming and Constraint-Based Planning
+is provided followed by the specific representation that \eu uses to
+undertake Constraint-Based Planning.  Subsequently, we provide an
+overview of \eus architecture and how it can be embedded in a planning
+application such as \rxe. Finally, \eus modeling, inference and search
+mechanisms are covered in detail so that the reader can
+\kcomment{appreciate} how \rx can perform \kcomment{deliberation while
+  being responsive to events in the real-world}.
 
@@
 \cite{berthoz} and its generalization into a computational framework
-demonstrating machine intelligence long targeted by pioneers in
-computer science \cite{computersthought}. Early work in this context
-focused on logical theorem proving by computer \cite{green69} which
-has formed the basis of much of the discipline of AI Planning. The
-emphasis in these early attempts was to logically derive the existence
-proof of a trajectory of actions for a hypothetical robot (moving
-blocks, chess pieces, furniture etc) to achieve a specific goal given
-specific initial conditions. Two important properties of
-\emph{Soundness}\footnote{A planning algorithm is sound if invoked on
-  a problem $P$ returns a plan which is a solution for $P$.} and
-\emph{Completeness}\footnote{A planning algorithm is complete if
-  invoked on a solvable problem $P$ is guaranteed to return a
-  solution.} along with derived results for algorithmic complexity
-\cite{corman} were a defining cornerstone for understanding how
-different computational paradigms could actually be tractable first in
-simulation and subsequently in actual robotic platforms. In this
-chapter our emphasis is to focus on embedded AI Planning and the
-execution of plans on an embedded \kcomment{marine robotic}
-platform. While the platform is domain specific, the techniques we
-highlight are general purpose as noted in Section \ref{sec:intro}.
+demonstrating machine intelligence \kcomment{has for long been}
+targeted by pioneers in computer science
+\cite{computersthought}. Early work in this context focused on logical
+theorem proving by computer \cite{green69} which has formed the basis
+of much of the discipline. The emphasis in these early efforts was to
+logically derive the existence proof of a trajectory of actions for a
+hypothetical robot (moving blocks, chess pieces, furniture etc) to
+achieve a specific goal given specific initial conditions. Two
+important properties of \emph{Soundness}\footnote{A planning algorithm
+  is sound if invoked on a problem $P$ returns a plan which is a
+  solution for $P$.} and \emph{Completeness}\footnote{A planning
+  algorithm is complete if invoked on a solvable problem $P$ is
+  guaranteed to return a solution.} along with derived results for
+algorithmic complexity \cite{gareyjohnson,corman} were a defining
+cornerstone for understanding how different computational paradigms
+could actually be tractable first in simulation and subsequently in
+actual robotic platforms. In this chapter our emphasis is to focus on
+embedded AI Planning and the execution of plans on an embedded
+\kcomment{marine robotic} platform. The techniques we highlight are
+general purpose as noted in Section \ref{sec:intro}.
 
@@
     In the near future, a personal robot sets out to buy a gallon of
-    milk This involves a number of tasks: obtain keys, obtain wallet,
+    milk. This involves a number of tasks: obtain keys, obtain wallet,
     start car, drive to store, find and obtain milk, purchase milk,
     etc.  The embedded planner has to have a ``model'' of the world in
-    which it lives and has to use the task primitives in this model to
-    structure the actions so it achieves its goal. Constraints control
-    when certain tasks can or cannot occur. For example the robot must
-    obtain the keys and wallet \emph{before} driving to the store and
-    pick up the milk \emph{before} purchasing it.
+    which it 'lives' and has to use the task primitives in this model
+    to structure the actions so it achieves its goal. Constraints
+    control when certain tasks can or cannot occur. For example the
+    robot must obtain the keys and wallet \emph{before} driving to the
+    store and pick up the milk \emph{before} purchasing it.
 \end{quote}
@@
 applications including recently, planning for the International Space
-Station \cite{barreiro09}. We motivate this section with some problem
-domains this planner can handle and then proceed to go into detail.
+Station \cite{barreiro09}. We first motivate this section with some
+problem domains this planner has handled.
 
 \paragraph{} {\em Constraint Satisfaction}: A canonical problem in
-dealing with constraints is the $N$-Queens problem in which chess queens
-must be placed on an  $N \times N$ chessboard so no queens attack the other. % Fig.
-% \ref{fig:nqueens-1} shows an example of a random positioning of queens
-% on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
-% are highlighted in red.
-% \begin{figure} \centering
-%   \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
-%   \caption{\small N-Queens problem. Queens in violation of the
-%     non-attack constraint are highlighted in red.}
-% \label{fig:nqueens-1}
-% \vskip+0.1cm
-% \end{figure}
+dealing with constraints is the $N$-Queens problem in which chess
+queens must be placed on an $N \times N$ chessboard so no queen can
+attack the other.
+
 If we define $N \times N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in
@@
 
-\paragraph{} \textit{Scheduling}: In the Resource Constrained Project
-Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, a project
-consisting of a set of activities must be scheduled in a way that
-satisfies minimum and/or maximum temporal separation constraints. The
-activity schedule must also respect fixed limits on the availability
-of resources required to perform each activity. In addition to
-satisfying temporal and resource constraints, it is common for the
-user to want to minimize makespan \cite{ghallab04} so that the entire
-project is finished as early as possible. Fig. \ref{fig:rcpsp-1} shows
-an example of a a solution provided by \eu for an \texttt{RCPSP}
-instance with 10 activities, 5 resources and 30 temporal constraints.
+\paragraph{} \textit{Scheduling}: The Resource Constrained Project
+Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, consists of a
+set of activities that must be scheduled in a way that satisfies
+minimum and/or maximum temporal separation constraints. The activity
+schedule must also respect fixed limits on the availability of
+resources required to perform each activity. In addition to satisfying
+temporal and resource constraints, it is common for the user to want
+to minimize makespan \cite{ghallab04} so that the entire project is
+finished as early as possible. Fig. \ref{fig:rcpsp-1} shows an example
+of a a solution provided by \eu for an \texttt{RCPSP} instance with
+$10$ activities, $5$ resources and $30$ temporal constraints.
 
@@
 \cite{russelnorvig} is a variation of the problem with which we
-started this chapter.  In this problem, an agent needs to purchase a
-set of products (milk, drill, etc) that are available at specific
-locations (supermarket, hardware store, etc). The agent is subject to
-temporal (must complete tasks by specific deadlines) and resource
-(fuel, carrying capacity, etc) constraints and it needs to synthesize
-actions that need to be performed to find and acquire the required
-items, as well as when to perform each of those actions.
+started this chapter.  An agent needs to purchase a set of products
+(milk, drill, etc) that are available at specific locations
+(supermarket, hardware store, etc). The agent is subject to temporal
+(must complete tasks by specific deadlines) and resource (fuel,
+carrying capacity, etc) constraints and it needs to synthesize actions
+that need to be performed to find and acquire the required items, as
+well as when to perform each of those actions.
 Fig. \ref{fig:shopping-1} shows a solution produced by \eu for such a
@@
 Constraint Satisfaction Programming (also known as Constraint
-Programming or \textsf{CP}) is a discipline that provides a generic
-framework for representing, solving and making logical inference on
-constraints. A complete treatment of this discipline is given in
-\cite{marriott98,apt03} with concise introductions provided in
-\cite{bartak99,lustig01}.
+Programming or \textsf{CP}) is \kcomment{an AI} discipline that
+provides a generic framework for representing, solving and making
+logical inference on constraints. A complete treatment of this
+discipline is given in \cite{marriott98,apt03} with concise
+introductions provided in \cite{bartak99,lustig01}.
 
@@
 
-\item \textbf{Bounds propagation} infers upper and lower variable
+\item \textbf{Bounds propagation} inferring upper and lower variable
   bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \& $x_i
@@
 
-\item \textbf{Domain reduction} infers a valid set of values for a
-  variable.  For example, for constraints allDifferent($x_1,x_2,x_3$),
-  $x_i \geq 0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can
-  infer that the valid and reduced domain for $x_3$ is $\{2,4\}$.
+\item \textbf{Domain reduction} inferring a valid set of values for a
+  variable.  For example, for constraints
+  \texttt{allDifferent}($x_1,x_2,x_3$), $x_i \geq 0$ \& $x_i \leq 4$,
+  if $x_1 = 1$ and $x_2 = 3$ we can infer that the valid and reduced
+  domain for $x_3$ is $\{2,4\}$.
 
@@
 constraints on $3$ variables ($x$, $y$ and $z$) represented as a
-constraint network, and their domains before arc consistency is
+constraint network and their domains before arc consistency is
 enforced.  After the \texttt{AC-3} algorithm is executed, the domains
@@
 \includegraphics[scale=0.3]{figs/constraint-net.pdf}
-\caption{\small Variables and Constraints represented as a Constraint Network}
+\caption{\small Variables and Constraints represented as a Constraint
+  Network \kcomment{before arc consistency is enforced}.}
 \label{fig:constraintnet}
@@
 be used to efficiently prove that a \textsf{CP} problem is
-unsatisfiable. However, they generally cannot prove that a \textsf{CP}
-problem is satisfiable and provide specific variable assignments that
-constitute solutions. To do so, search algorithms like global
-backtracking \cite{hooker05} or local search are used; these use
-consistency as an efficient way to prune the search space
+\emph{unsatisfiable}. However, they generally cannot prove that a
+\textsf{CP} problem is satisfiable and provide specific variable
+assignments that constitute solutions. To do so, search algorithms
+like global backtracking \cite{hooker05} or local search are used.
+These use consistency as an efficient way to prune the search space
 \cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
-\cite{ghallab04}, but often in practice a number of consistency and
-search algorithms can be used.
+\cite{ghallab04}, but often in practice \kcomment{heuristic methods
+  provide viable ways to generate efficient solutions}.
 
@@
 constraint that the user is able to reason about can be introduced
-into the system and, as long as the constraint propagation protocols
+into the system and as long as the constraint propagation protocols
 specified by the host \textsf{CP} system are enforced, it will be
@@
 
-The most common planning formulations use a propositional
-representation, where the state of the world is represented by a set
+The most common planning formulations use propositional
+\kcomment{logic}, where the state of the world is represented by a set
 of propositions (statements that can be true or false), and operators
 change the truth values of these propositions \cite{gen87}. Although
-these formulations are powerful and have allowed researchers to
-develop numerous contributions in automated planning, there are many
-classes of problem domains that are difficult to represent using this
-formalism. In particular, it is hard to represent time, resources,
-mutual exclusion and concurrency using propositions \cite{ghallab04}.
-Constraint Programming is a framework that lends itself well to
-represent and reason about all of those elements \cite{ghallab94}.
-Consequently, there has been substantial effort in building automated
-planners that take advantage of \textsf{CP} techniques.  The most
-common approach in constraint-based planning has been to encode the
-entire planning problem as a \textsf{CP} problem, where action choices
-and relationships are represented as variables and/or constraints
+these formulations are powerful \kcomment{for descriptive semantics}
+in automated planning, there are classes of problem domains that are
+difficult to represent. In particular, it is hard to represent time,
+resources, mutual exclusion and concurrency.  Constraint Programming
+is a framework that lends itself well to represent and reason about
+these elements \cite{ghallab94} with the consequence that there has
+been substantial effort in building automated planners that
+\kcomment{leverage} \textsf{CP} techniques.  The most common approach
+in constraint-based planning has been to encode the entire planning
+problem as a \textsf{CP} problem, where action choices and
+relationships are represented as variables and/or constraints
 \cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
 pure \textsf{CP} formulation is that standard solvers can be then used
-to solve the planning problem. However, this approach also has a
-couple of significant drawbacks:
+to solve the planning problem. However, this approach also has
+significant drawbacks:
 
@@
 
-These drawbacks have been addressed by a number of planning frameworks
-like \texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
+These drawbacks have been addressed by a number of frameworks like
+\texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
 \cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
-and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
-advantage of a \textsf{CP} representation to deal with time and
-resources, at the same time maintaining an explicit representation of
-the elements relevant to planning (action choices, causal
-relationships, etc). This combination results in a planning approach
-that can deal with the type of constraints found in real-world
-problems while allowing for intuitive problem descriptions and a
-natural way to express search algorithms and heuristics. \eu is one
-such implementation of the \texttt{CAIP} framework, in which planning
-problems are specified in terms of:
+and Attribute Planning) \cite{mus94,frank2003}.  They take advantage
+of a \textsf{CP} representation to deal with time and resources,
+\kcomment{while} maintaining an explicit representation of the
+elements relevant to planning (action choices, causal relationships,
+etc). This combination results in a planning approach that can deal
+with the \kcomment{kind} of constraints found in real-world problems
+while allowing for intuitive problem descriptions and a natural way to
+express search algorithms and heuristics. \eu is one such
+implementation of the \texttt{CAIP} framework, \kcomment{others
+  include \cite{chien00, iaai-CestaCFO09}}. In \eu planning problems
+are specified in terms of:
 
@@
 To generate plans, the \texttt{CAIP} framework uses a plan-space
-search approach \cite{ghallab04}, where a \emph{partial plan} is
-evolved until all open goals are achieved and all inconsistencies are
-removed. \kcomment{A partial plan is a subsequence of a plan which can
-  be refined into a plan structure.} The specific search algorithm
-used by \eu will be covered in \kcomment{further} detail below.
+search approach \cite{ghallab04}, where a \emph{partial
+  plan}\footnote{\kcomment{A partial plan is a subsequence of a plan
+    which can be refined into a plan structure.}} is evolved until all
+open goals are achieved and all inconsistencies are removed. The
+specific search algorithm used by \eu will be covered in
+\kcomment{further} detail below.
 

### 2012-05-02 trunk/intro.tex
updates

@@
 have outpaced software especially in algorithmic methods in sampling
-and control and in systematic software engineering development.
-
-Large scale high resolution observations have become increasingly
-tractable using autonomous platforms both on and below the surface.
-This, in turn, has lead to periodic calls for sustained exploration
-and sampling \cite{rudnick03}, illustrating the high scientific value
+and control and in systematic software engineering development. Large
+scale high resolution observations have become increasingly tractable
+using autonomous platforms both on and below the surface.  This, in
+turn, has lead to periodic calls for sustained exploration and
+sampling \cite{rudnick03}, illustrating the high scientific value
 placed on such data.  The complexity of oceanic processes and the
@@
 and available resources. Planning (or deliberation or projection in
-action space \kcomment{and used interchangeably in this chapter}) is
-important to balance current needs of a robot with future desires or
-goals taking into account time and available resources in-situ on the
+action space\footnote{\kcomment{We use the terms 'planning' and
+    'deliberation' interchangeably in this chapter.}}) is important to
+balance current needs of a robot with future desires or goals taking
+into account time and available resources in-situ on the
 robot. Planning without time or action planning in and of itself is
@@
 and the future is handled systematically. Yet, plans have to be tied
-to robotic action and in the real-world which is not static. Further,
-plans project future state which is not only likely to change, but is
-based on substantial uncertainty\footnote{It is important to remember
-  that these principles are broad much as Dwight Eisenhower is reputed
-  to have said ``\emph{In preparing for battle I have always found
-    that plans are useless, but planning is indispensable}'' and
-  ``\emph{Failing to plan is planning to fail}''}. Use of
+to robotic action and in dynamic real-world
+\kcomment{environments}. Further, plans \kcomment{not only} project
+future state which are likely to change, but \kcomment{need to factor
+  in} substantial environmental uncertainty\footnote{It is important
+  to remember that these principles are broad much as Dwight
+  Eisenhower is reputed to have said ``\emph{In preparing for battle I
+    have always found that plans are useless, but planning is
+    indispensable}'' and ``\emph{Failing to plan is planning to
+    fail}''.}. \kcomment{In the oceanographic context,} use of
 environmental cues is often hampered by poor \kcomment{predictive}
@@
 
-AUV control architectures have primarily been driven by reactive
-Subsumption-based approaches \cite{brooks86}. They have been adequate
-to doing routine straight-line survey trajectories which have played a
-valuable role in the acceptance of AUVs as a new tool for ocean
-science and DoD applications. However, scaling of problem domains in
-addition to dealing with event-response and discovery in the
-water-column challenges \emph{how} controllers are written in the
-first place, maintained and upgraded over the life-cycle of the
+\kcomment{To date} AUV control architectures have primarily been
+driven by reactive Subsumption-based approaches \cite{brooks86}. They
+have been adequate for routine straight-line survey trajectories which
+have played a valuable role in the acceptance of AUVs as a new tool
+for ocean science and \kcomment{military} applications. However,
+scaling of problem domains in addition to dealing with event-response
+and discovery in the water-column, challenges \emph{how} controllers
+are written, maintained and upgraded over the life-cycle of the
 platform. Dealing with off-nominal endogenous or exogenous conditions
@@
 utilized; one consequence for instance, is any off-nominal condition
-results in a \emph{fail-safe} surface mode to radio for help. Software
-design using these reactive approaches, not only makes it hard for
-sustaining engineering development, these approaches are not
-adequately adaptive for emerging ocean science problems.  Despite
-vigorous defense by Brooks early on
+results in a \emph{fail-safe} \kcomment{surfacing} mode to radio for
+help. Software design using these reactive approaches, not only makes
+it hard for sustaining engineering development, but these approaches
+are not adequately adaptive for emerging ocean science problems.
+Despite vigorous defense by Brooks early on
 \cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep},
-autonomous systems require a model if not of the world then of the
+autonomous systems do require a model if not of the world then of the
 \emph{physics} of the vehicle when it interacts with the world.
@@
 deliberation. Terrestrial robotic platforms were well-known to
-frequently stop and ``plan''; assuming in the meantime that the world
-will remain static. In the marine robotics world, such a situation
-would simply not be permissible given the need to make continuous
-observations without stopping an under-articulated robotic platform
-such as an AUV. However, algorithmic and implementational advances by
-the AI community over the years have produced a range of planners with
-demonstrated embedded capabilities
+frequently stop and ``plan'', assuming in the meantime that the world
+would remain static \cite{shakey84}. In the marine robotics world,
+such a situation is not permissible given the need to make continuous
+observations without stopping a \kcomment{non-holonomic} robotic
+platform such as an AUV. However, algorithmic and implementational
+advances by the AI community over the years have produced a range of
+planners with demonstrated embedded capabilities
 \cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
@@
 activities in the field of AI Planning and Plan Execution with
-demonstrated real-world and highly visible capabilities in the Space
-domain \cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons
-learned have been applied to the world of marine robotics which we
-bring together in this chapter. Important lessons that were derived
-from this rich legacy included the following:
+demonstrated real-world capabilities in the Space domain
+\cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons learned
+have been applied to the world of marine robotics which we bring
+together in this chapter. Important lessons that were derived from
+this rich legacy included the following:
 
@@
 \item A rich representation that can deal with co-temporal events for
-  modeling complex systems is a must.
+  modeling complex systems is \kcomment{important}.
 
@@
 \item Planning should be tightly inter-leaved with execution to enable
-  responsiveness to events which impact plan failure. This is also an
-  important factor leading to platform adaptivity.
+  responsiveness to events which impact plan failure. This is also
+  important \kcomment{for} platform adaptivity.
 
@@
   domain. The controller can then be rigorously tested and certified
-  even as the domain model evolves to fit diverse applications.
+  even as the domain model evolves to fit \kcomment{evolving
+    requirements and/or} diverse applications.
 
@@
 (\rxe) an on-board adaptive control system that integrates AI based
-planning and state estimation in a hybrid
-executive\cite{mcgann08a,mcgann08b,py10}.  State estimation allows the
-system to keep track of world evolution as time advances based on a
-formal model distributed through the architecture.  Onboard planning
-and execution enables adaptation of navigation and instrument control
-based on the state identified during state estimation.  It further
-enables goal-directed commanding within the context of projected
-mission state and allows for replanning for off-nominal situations and
-opportunistic science events. The framework in addition to being used
-on an AUV, is general enough \kcomment{to} be used for controlling a
-personal robot \cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed
-on a European planetary rover testbed \cite{goac11}. While it is
-possible to integrate probabilistic estimation on top of this
-framework \cite{mcgann08d}, such representation is not explicitly
-handled by the framework and is out of scope of this chapter.
-Deliberation and reaction are integrated systematically over different
-temporal and functional scopes within a single agent and a single
-model that covers the needs of high-level mission management,
-low-level navigation, instrument control, and detection of
-unstructured and poorly understood phenomena. \rx is deployed on
-MBARI's \emph{Dorado} AUV shown in Fig. \ref{fig:auv-fig}, which to
-the best of our knowledge is the only operational marine platform
-anywhere being used for routine scientific surveys with onboard plan
-synthesis.
+planning and state estimation in a hybrid executive
+\cite{mcgann08a,mcgann08b,py10}.  State estimation allows the system
+to keep track of world evolution as time advances based on a formal
+model distributed through the architecture.  Onboard planning and
+execution enables adaptation of navigation and instrument control
+based on \kcomment{estimated state}.  It further enables goal-directed
+commanding within the context of projected mission state and allows
+for replanning for off-nominal situations and opportunistic science
+events. The framework in addition to being used on an AUV, is general
+enough \kcomment{to} be used for controlling a personal robot
+\cite{pr2,Meeussen:2010dn,mcgann2009} and is deployed on a European
+planetary rover testbed \cite{goac11}. While it is possible to
+integrate probabilistic estimation on top of this framework
+\cite{mcgann08d}, such representation is not explicitly handled by the
+framework and is out of scope of this chapter.  Deliberation and
+reaction are integrated systematically over different temporal and
+functional scopes within a single agent and a single model that covers
+the needs of high-level mission management, low-level navigation,
+instrument control, and detection of unstructured and poorly
+understood phenomena. \rx is deployed on MBARI's \emph{Dorado} AUV
+shown in Fig. \ref{fig:auv-fig}, which to the best of our knowledge is
+the only operational marine platform anywhere being used for routine
+scientific surveys with onboard plan synthesis.
 
@@
 
-This chapter is organized as follows. We start with some foundational
-concepts in our Planning framework in Section~\ref{sec:concepts} which
-includes a brief overview of AI Planning in
-Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
-deeper into the \eu framework itself, the basis of all deliberation
-within \rxe. A key portion is Section~\ref{sec:europa:pr} which
-details the internal plan representation within a situated agent like
-\rxe. Modeling (Section~\ref{sec:europa:modeling}), Inference
+This chapter is organized as follows. We start \kcomment{by placing
+  this effort in the context of the overal research in AI Planning in
+  Section \ref{sec:related}} followed by some foundational concepts in
+our Planning framework in Section~\ref{sec:concepts} which includes a
+brief overview of AI Planning in Section~\ref{sec:planningfound}. In
+Section~\ref{sec:basics} we go deeper into the \eu framework itself,
+the basis of all deliberation within \rxe. A key portion is
+Section~\ref{sec:europa:pr} which details the internal plan
+representation within a situated agent like \rxe. Modeling
+(Section~\ref{sec:europa:modeling}), Inference
 (Section~\ref{sec:europa:inference}) and Search
-(Section~\ref{sec:europa:search}) articulates important details
+(Section~\ref{sec:europa:search}) articulate important details
 associated with \eu critical to \rxe.  With these foundational
@@
 details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
-how planning and execution are interleaved are shown in
+how planning and execution are interleaved are detailed in
 Section~\ref{sec:arch:intertwine}. We offer results from our

### 2012-05-02 trunk/related.tex
updates

@@
 the software development methodology leading to code bloat or worse.
-\cite{carreras06} provides a reasonably good overview of AUV control
-architectures that essentially is based on reactive control methods.
-
-\begin{figure}[!t]
- \centering
- \includegraphics[scale=0.4]{figs/planner-arch.pdf}
- \caption{\small A general architectural block diagram for an AI based
-   Planner.}
- \label{fig:planner}
-\end{figure}
-
-A word about generalized planning. Traditionally planning has been
-considered computationally intensive (the general planning problem can
-be worse than NP-complete \cite{ghallab04}). A major reason for this
-belief had to do with the role of the planner (which was assumed to be
-generative) and its place in the architecture (infrequently called to
-re-plan the \emph{entire} mission in off-nominal conditions) within a
-\emph{sense-plan-act} paradigm. This can limit the reactivity of a
-robot especially when the environment could change at a rate faster
-than the planner can plan.
-
-The architecture of a typical AI planner is along the lines of
-Fig. \ref{fig:planner}. Given a \emph{domain model} which encodes the
-platform constraints in a higher-level language, an \emph{initial
-  state}, a high-level objective or end \emph{goal}, the \emph{plan
-  database} (or \texttt{plandb}) is where all assertions are tracked
-and maintained. Theorem proving in the form of satisfaction of axioms
-embodied in the domain model occurs within the \texttt{plandb}. It is
-the job of the \emph{search engine} to provide the inferential
-mechanism for state space exploration potentially aided by heuristics
-encoded for \emph{search control}. In some complex domains
-\cite{mus98} \emph{external modules} can provide additional domain
-expertise which can augment the domain model. The end result is plan
-formulation using a multitude of approaches either using forward or
-backward search or using more generalized \emph{partial order} methods
-\cite{ghallab04}.
+\cite{carreras06} provides a good overview of AUV control
+architectures essentially based on reactive control methods.
 
@@
 \end{enumerate}
+
+\begin{figure}[!t]
+ \centering
+ \includegraphics[scale=0.4]{figs/planner-arch.pdf}
+ \caption{\small A general architectural block diagram for an AI based
+   Planner.}
+ \label{fig:planner}
+\end{figure}
+
+Traditionally generalized planning has been considered computationally
+intensive (the general planning problem can be worse than NP-complete
+\cite{ghallab04}). A major reason for this belief had to do with the
+role of the planner (which was assumed to be generative) and its place
+in the architecture (infrequently called to re-plan the \emph{entire}
+mission in off-nominal conditions) within a \emph{sense-plan-act}
+paradigm. This can limit the reactivity of a robot especially when the
+environment could change at a rate faster than the planner can plan.
+
+The architecture of a typical AI planner is along the lines of
+Fig. \ref{fig:planner}. Given a \emph{domain model} which encodes the
+platform constraints in a higher-level language, an \emph{initial
+  state}, a high-level objective or end \emph{goal}, the \emph{plan
+  database} (or \texttt{plandb}) is where all assertions are tracked
+and maintained. Theorem proving in the form of satisfaction of axioms
+embodied in the domain model occurs within the \texttt{plandb}. It is
+the job of the \emph{search engine} to provide the inferential
+mechanism for state space exploration potentially aided by heuristics
+encoded for \emph{search control}. In some complex domains
+\cite{mus98} \emph{external modules} can provide additional domain
+expertise which can augment the model. The end result is plan
+formulation using a multitude of approaches either using forward or
+backward search or using more generalized \emph{partial order} methods
+\cite{ghallab04}.
 
@@
     physical robot.}} was the highly influential \texttt{Shakey} the
-robot experiment \cite{shakey84}. Subsequent work driven largely by
-defense funding in the United States sequestered the discipline in
-academic labs principally applied for algorithmic development and if
-embedded then in testing within the confines of modest indoor
+robot \cite{shakey84}. Subsequent work driven largely by defense
+funding in the United States sequestered the discipline in academic
+labs principally applied for algorithmic development and if embedded
+then in testing within the confines of \kcomment{structured} indoor
 settings. The leap to applying them for real world problem solving is
@@
 $3$-layered architecture (see \cite{gat98} for a well thought out
-rationale), notable examples of which include \texttt{IPEM}
-\cite{AmbrosIngerson88}, \texttt{ROGUE} \cite{Haigh98}, the LAAS
-Architecture \cite{alami:1998p820}, the Remote Agent Experiment
-\texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
-\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). % Early
-% efforts were cognizant of if absent with, dealing with time and
-% resources for such experiments.
+rationale \kcomment{and survey}), notable examples of which include
+\texttt{IPEM} \cite{AmbrosIngerson88}, \texttt{ROGUE} \cite{Haigh98},
+the LAAS Architecture \cite{alami:1998p820}, the Remote Agent
+Experiment \texttt{RAX} \cite{mus98} and the Autonomous Spacecraft
+Experiment \texttt{ASE} \cite{chien99} (see \cite{Knight01} for a
+survey).
 
@@
 temporal database and the conceptualization of temporal
-intervals. Using these base-level concepts Muscettola and others
-\cite{mus92, ghallab94, laborie95, cesta96} developed the notion of
-plan/schedule envelopes using the notion of state variable
-instantiated timelines. Simultaneously work by Dechter et.al
-\cite{dechter91} resulted in efficient temporal constraint propagation
-which systematically defined the notion of Simple Temporal Networks
-(STNs) and path and arc consistency algorithms. This work neatly put
-together earlier efforts by Mackworth and Freuder \cite{mackworth77,
-  mack85} and the Constraints community on propagation algorithms. In
-parallel work in the UK as a result of US DARPA funding, resulted in
-the design of \texttt{O-Plan} \cite{currie91}, which leveraged the
-notion of constraints and metric time, but within a Hierarchical Task
-Network (HTN) representation. All these efforts were soon after Vere’s
-paper on planning and time \cite{vere83} which had a profound effect
-on the AI Planning community. In particular the work of Muscettola
-\cite{mus94} was embraced by NASA for telescope scheduling. Around the
-same time period, researchers at LAAS came up with an architecture
+intervals. Using these concepts Muscettola and others \cite{mus92,
+  ghallab94, laborie95, cesta96} developed the notion of plan/schedule
+envelopes using the notion of state variable instantiated
+timelines. Simultaneously work by Dechter et.al \cite{dechter91}
+resulted in efficient temporal constraint propagation which
+systematically defined the notion of Simple Temporal Networks (STNs)
+and path and arc consistency algorithms. This work neatly put together
+earlier efforts by Mackworth and Freuder \cite{mackworth77, mack85}
+and the Constraints community on propagation algorithms. In parallel
+work in the UK, resulted in the design of \texttt{O-Plan}
+\cite{currie91}, which leveraged the notion of constraints and metric
+time, but within a Hierarchical Task Network (HTN) representation.
+These efforts were soon after Vere’s paper on planning and time
+\cite{vere83} which had a profound effect on the AI Planning
+community. In particular the work of Muscettola \cite{mus94} was
+embraced by NASA for telescope scheduling. Around the same time
+period, researchers at LAAS came up with an architecture
 \cite{alami:1998p820} which encapsulated the \texttt{IxTeT} planner
-\cite{ghallab94}. It featured advanced concepts for a temporal planner
-with the notion of chronicles, plan recognition for partial plan
-fragments and early use of systematic resource
-constraints. Development on \texttt{IxTeT} has continued albeit at a
-slower pace with the contribution of various PhD students’ thesis.
+\cite{ghallab94} \kcomment{which dealt with time and resources}. It
+featured advanced concepts for a temporal planner with the notion of
+chronicles, plan recognition for partial plan fragments and early use
+of systematic resource constraints. % Development on \texttt{IxTeT} has
+% continued albeit at a slower pace with the contribution of various PhD
+% students’ thesis.
 
@@
 board and successfully commanded the spacecraft for a week in May
-1999. There were a number of significant software engineering lessons
-learned with the HSTS technology as integrated for the RAX:
+1999\footnote{This was the first (and to our knowledge only) software
+  ever to been written in \texttt{LISP} to be flown in space.}. There
+were a number of significant software engineering lessons learned with
+the HSTS technology as \kcomment{implemented} for the RAX
+\kcomment{experiment}:
 
@@
 \item Flexible temporal representation generating a range of plans as
-  against a single plan, were robust for plan execution.
+  against a single plan, were robust for \kcomment{robotic} plan
+  execution.
 
@@
 \cite{mus02,mus04,Dias:2003ua,mus06} with the central theme of using
-planners collectively for problem solving arose. Another critical
-theme was iteratively and incremental plan repair of an anytime plan
+planners collectively for problem solving. Another critical theme was
+iteratively and incremental plan repair of an anytime plan
 \cite{Zaimag96} as proposed by \texttt{CASPER} \cite{chien00}.
@@
 required substantial effort to customize for an application domain. It
-did move to a retooled version of \texttt{HSTS}, now called EUROPA
-\cite{frank2003, barreiro09} and ported to the C++ language.
+did move to a retooled version of \texttt{HSTS}, called \eu
+\cite{frank2003, barreiro09} while being ported to the C++ language.
 
@@
 earth analog rover deployments \cite{wetter05} a separate development
-was undertaken to deploy the EUROPA planner for NASA’s Mars
-Exploration Rovers (MER) mission. The \texttt{MAPGEN}
-(Mixed-initiative Activity Plan GENerator) planner \cite{bresina03,
-  aichang04, bresina05, bresina05a} as a decision support tool in the
-mixed-initiative mode; \texttt{MAPGEN} continues to be used to this
-day to command the twin rovers on the Red
-Planet. % While EUROPA2 was not
+was undertaken to deploy the \eu planner for NASA’s Mars Exploration
+Rovers (MER) mission. The \texttt{MAPGEN} (Mixed-initiative Activity
+Plan GENerator) planner \cite{bresina03, aichang04, bresina05,
+  bresina05a} as a decision support tool in the mixed-initiative mode;
+\texttt{MAPGEN} continues to be used to this day to command the twin
+rovers on the Red Planet. % While EUROPA2 was not
 % designed as an embedded planner, NASA undertook a bottom-up
@@
 a more observable vehicle, the EO-1 in Earth orbit. It encapsulated a
-classical 3-layered architecture like \texttt{RAX}. The fundamental
-functional difference was that goals could not only be sent by a
-ground segment, but also by an onboard science driver which could
-encapsulate a range of Machine Learned feature detectors which could
-trigger the planner. However ASE was not an integrated system; the
-\texttt{CASPER} planner \cite{chien00} would generate plans which were
-executed by a separate commercially available rule-based system
-\cite{icl}. \texttt{CASPER} touts a \emph{continuous} paradigm for
-planning; however this incremental modification simply takes the
-overall mission plan and selectively plans near-term at a more
-detailed level. The system also uses iterative-repair as a way to fix
-plans; however this too is derived from a model of the mission and
-spacecraft constraints which allow simple reconfiguration modes
-towards replanning. Robust software engineering is also not an
-important aspect targeted by \texttt{ASE} given the disparate models
-between the planner and the rule-based executive. With such a system,
-real-time updates from the environment cannot be accommodated since
-the planner is a monolithic computational environment. The rate of
-updates from the science drivers is roughly \kcomment{on} the order of
-an earth orbit on the EO-1 vehicle. As a result, the ASE is only an
-incremental update on RAX and is not suitable for domains where
-complex automated reasoning and asynchronous environmental
+classical 3-layered architecture like \texttt{RAX}. Goals could not
+only be sent by a ground segment, but also by an onboard science
+driver which could encapsulate a range of Machine Learned feature
+detectors which could trigger the planner. However \texttt{ASE} was
+not an integrated system; the \texttt{CASPER} planner \cite{chien00}
+would generate plans which were executed by a separate commercially
+available rule-based system \cite{icl}. \texttt{CASPER} demonstrates
+what is called as a \emph{continuous} paradigm for incremental
+planning and iterative plan repair; this methodology takes the overall
+mission plan and selectively plans \kcomment{for the} near-term at a
+more detailed level.
+% The system also uses iterative-repair as a way to fix
+% plans; however this too is derived from a model of the mission and
+% spacecraft constraints which allow simple reconfiguration modes
+% towards replanning. 
+Robust software engineering is also not an important aspect targeted
+by \texttt{ASE} given the disparate models between the planner and the
+rule-based executive. With such a system, real-time updates from the
+environment could not be accommodated since the planner is a
+monolithic computational environment. The rate of updates from the
+science drivers is roughly \kcomment{on} the order of an earth orbit
+on the EO-1 vehicle. As a result, the \texttt{ASE} is only an
+incremental update on \texttt{RAX} and is not suitable for domains
+where complex automated reasoning and asynchronous environmental
 observations need to be factored in deliberation.
 
-\texttt{TCA} \cite{simmons94} is another framework for robot control
-and has similar motivations to \texttt{RAX} and
-\texttt{ASE}. Reactivity and deliberation are also key to \texttt{TCA}
-and a systematic approach to development led to its design. However it
+\texttt{TCA} \cite{simmons94} another framework for robot control and
+has similar motivations to \texttt{RAX} and \texttt{ASE}. Reactivity
+and deliberation are also key to \texttt{TCA} \kcomment{along with} a
+systematic approach to development led to its design. However it
 suffers from three key weaknesses. First it has a weaker
@@
 
-The \texttt{ReSSAC} project at ONERA \cite{teichteil07} has similar
-motivations in controlling aerial UAV platforms for autonomous Search
+The \texttt{ReSSAC} project at ONERA \cite{teichteil07} \kcomment{is
+  motivated by} controlling aerial UAV platforms for autonomous Search
 and Rescue. Like \texttt{TCA}, the project uses a supervisor to
@@
 generate reachable goal states incrementally. This work however does
-not deal with metric time with a modestly simple model.
+not deal with metric time \kcomment{and comes} with a modestly simple
+model.
 
-\texttt{CIRCA} \cite{musliner95} is another effort to bring decision
-making for real-world problems with the augmented need to have
+\texttt{CIRCA} \cite{musliner95} is an effort to bring decision making
+for real-world problems with the augmented need to have
 \emph{verifiable} controllers synthesized in-situ to ensure that
-within explored states, there are no erroneous transitions to
-dangerous states by asynchronously generating Test-Action-Pairs
-(TAPs).  These are annotated production rules consisting of a set of
-tests (or preconditions) and a set of actions to take if all the tests
-return true. TAPs are synthesized and scheduled by \texttt{CIRCA} and
-provide a viable way to deal with a dynamically changing
-world. \texttt{CIRCA} however differentiates between reasoning about
-time and reasoning in real-time with the implication that reasoning
-necessarily requires substantial computation for state space
-exploration which negates the real-time (and thus real-world)
-impact.
+\kcomment{during state space exploration}, there are no erroneous
+transitions to dangerous states by asynchronously generating
+Test-Action-Pairs (TAPs).  These are annotated production rules
+consisting of a set of tests (or preconditions) and a set of actions
+to take if all the tests return true. TAPs are synthesized and
+scheduled by \texttt{CIRCA} and provide a viable way to deal with a
+dynamically changing world. \texttt{CIRCA} however differentiates
+between reasoning about time and reasoning in real-time with the
+implication that reasoning necessarily requires substantial
+computation for state space exploration which negates the real-time
+(and thus real-world) impact.
 
 A key and early concern that dominated planning was that of
-scalability.  The planning cycle in many approaches was monolithic
-often making fast reaction times impractical when necessary for
-embedded robotic agents. Many of these systems (which were
-$3$-layered) also utilized very different techniques for specifying
-each layer in the architecture resulting in duplication of effort and
-a diffusion of knowledge.  The work we highlight in this chapter
-builds on the approach used by \texttt{IDEA} \cite{mus02, mus04} in
-utilizing a collection of controllers, each interleaving planning and
-execution within a common framework. \texttt{IDEA} however, provided
-no support for conflict resolution between controllers, nor does it
-provide an efficient algorithm for integrating current state within a
-controller, relying instead on a possibly exponential planning
-algorithm. Efficient synchronization of state in a partitioned
-structure is fundamental to making the approach effective in practice.
+scalability.  The planning cycle in many approaches was \kcomment{(and
+  is)} monolithic, often making fast reaction times impractical when
+necessary for embedded robotic agents. Many of these systems (which
+were $3$-layered) also utilized very different techniques for
+specifying each layer in the architecture resulting in duplication of
+effort and a diffusion of knowledge.  The work we highlight in this
+chapter builds on the approach used by \texttt{IDEA} \cite{mus02,
+  mus04} in utilizing a collection of controllers, each interleaving
+planning and execution within a common framework. \texttt{IDEA}
+however, provided no support for conflict resolution between
+controllers, nor did it provide an efficient algorithm for integrating
+current state within a controller, relying instead on a possibly
+exponential planning algorithm. Efficient synchronization of state in
+a partitioned structure is fundamental to making the approach
+effective in practice.
 
@@
 structure formally defined as a composition of coordinated control
-loops where sensing, planning and acting are resulting from concurrent
-deliberating tasks within a formal framework.  \rx was designed after
-carefully evaluating lessons learned from \texttt{IDEA} to which it
-owes substantially the notion of partitioned problem solving. However
-partitioning in \rx is systematic and methodologically principled
-where we manage the information flow within the partitioned structure
-to ensure consistency in order to direct the flow of goals and
-observations in a timely manner. The resulting control structure
+loops where sensing, planning and acting \kcomment{result} from
+concurrent deliberating tasks within a formal framework.  \rx was
+designed after carefully evaluating lessons learned from \texttt{IDEA}
+to which it owes substantially the notion of partitioned problem
+solving. Partitioning in \rx is systematic and methodologically
+principled where we manage the information flow within the partitioned
+structure to ensure consistency in order to direct the flow of goals
+and observations in a timely manner. The resulting control structure
 improves scalability since many details of each controller can be
@@
 enable graceful system degradation, making this an effective
-\emph{divide-and-conquer} approach to the overall control problem.
+\emph{divide-and-conquer} approach to \kcomment{solving} the overall
+control problem.
 
@@
 derived from \rx is \texttt{ROAR} \cite{degroote11} where lower-level
-functional modules that control specific sensors are organized within
-a graph hierarchy. One reason they cite such a structure is essential
-is in dealing robustly to failure; an off-nominal condition will allow
-rapid graph traversal to identify alternative ways in which sensing
-tasks can be executed and therefore aiding in responsiveness. However,
-it is unclear whether to date this system actually deliberates and
-whether it is actually instantiated on a real platform for control.
+functional modules that control sensors are organized within a graph
+hierarchy. One reason they cite such a structure \kcomment{as being}
+essential is in dealing robustly to failure; an off-nominal condition
+will allow rapid graph traversal to identify alternative ways in which
+sensing tasks can be executed and therefore aiding \kcomment{platform}
+responsiveness. % However,
+% it is unclear whether to date this system actually deliberates and
+% whether it is actually instantiated on a real platform for control.
 

### 2012-05-02 trunk/setobook.tex
updates

@@
   autonomous underwater vehicles (AUVs). The Teleo-Reactive EXecutive
-  (\rx) framework deliberates about future states, plans for actions
+  (\rxe) framework deliberates about future states, plans for actions
   and executes generated activities while monitoring plans for
@@
   hardware becomes more robust allowing persistent ocean sampling and
-  observation. \rx is in routine operational use at MBARI, providing
-  scientists a new tool to sample and observe the dynamic coastal
-  ocean.}
+  observation. \rx is in \kcomment{regular} operational use at MBARI,
+  providing scientists a new tool to sample and observe the dynamic
+  coastal ocean.}
 

### 2012-05-03 arch.tex
updates

@@
 controller which could bring automated planning closer to low-level
-control of the vehicle while maintaining its reactivity. Traditionally
-planning has been deemed computationally expensive with planner
-performance restricting robot reactivity
+control of the vehicle while \kcomment{ensuring}
+reactivity. Traditionally planning has been deemed computationally
+expensive with planner performance restricting robot reactivity
 \cite{ghallab04,Dias:2003ua}. In control architectures where task
@@
 the planner can plan for. In such a situation, the agent may thrash
-\footnote{\kcomment{In such an event as the planner works towards goal
-    satisfaction, the initial conditions leading to achieve that goal
-    are no longer valid. The planner then tries to achieve a different
-    goal which too has to be discarded similarly. And so on.}}  if the
-internal state of the plan gets out of sync with the actual state of
-the world.
-
-\kcomment{A important consequence of this has been that} the planning
+\footnote{\kcomment{In such an event even as the planner works towards
+    goal satisfaction, the initial conditions leading to achieve that
+    goal are no longer valid. The planner then tries to achieve a
+    different goal which too has to be discarded similarly. And so
+    on.}}  if the internal state of the plan gets out of sync with the
+actual state of the world.
+
+\kcomment{An important consequence of this has been that} the planning
 problem managed {\em in situ} has traditionally remained fairly
@@
 framework with \kcomment{an} unified declarative model. Our approach
-with \rx builds on this design methodology albeit in a substantially
-systematic manner.
+builds on this design methodology albeit in a substantially systematic
+manner.
 
@@
 \item \emph{Partitioning} the global planning problem into multiple
-  decision loops, called {\em reactors}. Each reactor has its own
-  scope both functionally (one for example focusing on the conversion
-  of high level waypoints into low-level commands) and temporally as
-  each reactor declares its planning look-ahead ($\pi$) and expected
-  latency ($\lambda$) before being able to produce \kcomment{a} plan
-  to the agent depending on where it is located in a reactor
-  dependency graph such as shown in Fig. \ref{fig:agent}. A reactor
-  closer to the hardware in the hierarchy, is expected to deliberate
-  less and be highly responsive to observations from the world. This
-  implies lower values of $\pi$ and $\lambda$ (see Section
-  \ref{sec:arch:trex} below). \kcomment{In the current design, $\pi$
-    and $\lambda$ are application dependent and fixed at design time.}
+  decision loops called {\em reactors}. Each reactor has its own scope
+  both functionally (one for example focusing on the conversion of
+  high level waypoints into low-level commands) and temporally as each
+  reactor declares its planning look-ahead ($\pi$) and expected
+  latency ($\lambda$) before being able to produce \kcomment{a} plan.
+  \kcomment{These parameters depend} on where \kcomment{a reactor} is
+  located in a dependency graph such as \kcomment{the one} shown in
+  Fig. \ref{fig:agent}. A reactor closer to the hardware in the
+  hierarchy, is expected to deliberate less and be highly responsive
+  to observations from the world. This implies lower values of $\pi$
+  and $\lambda$ (see Section \ref{sec:arch:trex} below). \kcomment{In
+    the current design, $\pi$ and $\lambda$ are application dependent
+    and fixed at design time.}
 
 \item \emph{Coupling} a tighter integration between the planning
-  process and world evolution to ensure that planning occurs
-  \emph{during continuous} state update in the {\em same}
-  \kcomment{plan} database used for planning. By doing so, the planner
-  is better informed in its deliberation despite a larger latency than
-  the changing world.
+  process and world evolution \kcomment{with the use of the {\em same}
+    \kcomment{plan} database}. By doing so, the planner is better
+  informed in its deliberation despite a larger latency than the
+  changing world.
 \end{itemize}
@@
 Systematic partitioned inference is a major contribution of
-\rxe. However that is not the focus of this chapter; details are
+\rxe. However that is not the focus of this chapter with details
 covered in \cite{py10, rajan12}. Coupled interaction between planning

### 2012-05-03 europa.tex
updates

@@
 \eu is implemented as a library that encapsulates all elements
-described above making them accessible via modeling and Application
-Programming Interface (API) layers.  A user will typically use the
-modeling facilities to inject a problem domain description and the
-API to extract and manipulate plans and to obtain information about
-plan flaws and constraint violations at any stage of the planning
-process.
+\kcomment{previously described} and available via modeling and
+Application Programming Interfaces (APIs).  A user will typically use
+modeling facilities to inject a problem domain description and the API
+to extract and manipulate plans and to obtain information about plan
+flaws and constraint violations at any stage of the planning process.
 
@@
 \includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
-\caption{\small The \eu software architecture.\comment{Is there a
-    better block diagram we can use to respond to review comment on
-    page 2?}}
+\caption{\small The \eu software architecture.}
 \label{fig:europa-architecture}
@@
 constraints that define relationships among them. It also provides an
-efficient arc consistency mechanism \cite{mackworth77}. The
-\texttt{CRE} is designed so that specialized reasoning algorithms for
-new constraints can be easily and efficiently plugged in. The
-\textbf{Plan Database} (\texttt{PDB}) manages object and token
-types. It also maintains the state of partial plans in terms of
-object, variable and token instances. This constitutes the data
-foundation on which end user applications and automated problem
-solvers are built. The \textbf{Rules Engine} (\texttt{RE}) modularly
-adds the semantics of domain rules that describe dependencies between
-tokens. The \textbf{Solver Module} provides the framework for flaw
-detection and resolution that is needed to implement search algorithms
-for planning. The \textbf{API Layer} provides access to services from
-all the different modules in a programmatic way so that \eu can be
-easily embedded into specialized applications.
+efficient arc consistency mechanism \cite{mackworth77} designed so
+that specialized reasoning algorithms for new constraints can be
+easily and efficiently plugged in. The \textbf{Plan Database}
+(\texttt{PDB}) manages object and token types. It also maintains the
+state of partial plans in terms of object, variable and token
+instances. This constitutes the data foundation on which end user
+applications and automated problem solvers are built. The
+\textbf{Rules Engine} (\texttt{RE}) adds the semantics of domain rules
+that describe dependencies between tokens. The \textbf{Solver Module}
+provides the framework for flaw detection and resolution that is
+needed to implement search algorithms for planning. The \textbf{API
+  Layer} provides access to services from all the different modules in
+a programmatic way so that \eu can be easily embedded into specialized
+applications.
 
 Describing problem domains and problem instances in terms of objects,
-variables and domain rules in a programmatic way is cumbersome to the
-extent of being impractical; this motivated the creation of modeling
-languages that support those definitions in a much more natural and
-compact way. \nd (New Domain Description Language
-[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
-\eu and will be described in detail in the Modeling section
-below. Another, \texttt{ANML} is a language under development that
-provides a more intuitive syntax \cite{smith08}.
+variables and domain rules is cumbersome to the extent of being
+impractical; this motivated the creation of modeling languages that
+support those definitions in a much more natural and compact way. \nd
+(New Domain Description Language [pronounced ``noodle'']) \cite{NDDL}
+is the \kcomment{primary} modeling language for \eu described in
+Section \ref{sec:europa:modeling}. Another, \texttt{ANML}
+\cite{smith08} is a language under development that provides a more
+intuitive syntax \kcomment{and out of scope for this chapter}.
 
@@
 \item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
-  and Solver modules to provide reasoning and search for metric
-  resources.
+  and \textbf{Solver Module} to provide reasoning and search for
+  metric resources.
 
@@
   Problems that require challenging search may require specialized
-  solvers to be built as explained in Section \ref{sec:europa:search}.
+  solvers to be built as noted in Section \ref{sec:europa:search}.
 
@@
 \kcomment{Before going into the details of each module, a simple
-  example illustrates how they interact at a high level:}
+  example illustrates how they interact at an abstract level:}
 
@@
 
-\kcomment{The snippet above is a \nd (covered in more detail in
+\kcomment{The example above is a \nd (covered in more detail in
   Section \ref{sec:europa:modeling}) model of a light bulb and a
-  switch that controls it, along with a problem instance that requires
-  the light bulb to be off by time$=10$, given an initial state with
-  the bulb on at time$=0$. The \nd module allows \eu to ingest this
-  model and the problem description. Once the \nd statements are
-  parsed, different object types, object instances and constraints are
-  instantiated inside the \texttt{PDB} and \texttt{CRE} modules where
-  they are available for both the user and the solver module to query
-  and manipulate programmatically. This instantiation of the initial
-  state constitutes the initial plan that \eu holds for this
-  particular problem instance.
+  switch that controls it along with a problem instance that requires
+  the light bulb to be \texttt{off} by time$=10$, given an initial
+  state with the bulb \texttt{on} at time$=0$. The \nd module allows
+  \eu to ingest this model and the problem description. Once the \nd
+  statements are parsed, different object types, object instances and
+  constraints are instantiated inside the \texttt{PDB} and
+  \texttt{CRE} modules where they are available for the user and the
+  solver modules to query and manipulate programmatically. This
+  instantiation of the initial state constitutes the initial plan that
+  \eu holds for this particular problem instance.
 
@@
   oversubscribed resources). When actions are added to a plan, the
-  \texttt{RE} module ensures that domain rules expressed in the model
-  are enforced when \eu creates new action or predicate instances. For
+  \texttt{RE} ensures that domain rules expressed in the model are
+  enforced when \eu creates new action or predicate instances. For
   instance, if a \texttt{LightSwitch::turnOn} action is added to the
@@
 make \nd a powerful tool for describing problem domains are that it is
-object oriented. \nd supports classes and inheritance in similar
-fashion to popular Object-Oriented languages like Java and C++. It
-also supports polymorphism for some of the planning components. Using
-object classes and instances is a time-tested approach to naturally
-describe problem domains \cite{rumbaugh}. It offers declarative
-constructs to define constraints and causality dependencies between
-actions and effects and finally it offers procedural constructs to
-populate \eus \texttt{PDB} with specific problem instances expressed
-in terms of objects, variables, constraints, facts and goals. To
-articulate the most important elements of \nde, we show how
-the Shopping Agent problem could be modeled:
+object oriented. It supports classes and inheritance in similar
+fashion to popular object-oriented languages like Java and C++. It
+also supports polymorphism for some planning components. Using object
+classes and instances is a time-tested approach to naturally describe
+problem domains \cite{rumbaugh}. It offers declarative constructs to
+define constraints and causality dependencies between actions and
+effects and it offers procedural constructs to populate \eus
+\texttt{PDB} with specific problem instances expressed in terms of
+objects, variables, constraints, facts and goals.
+
+To articulate the most important elements of \nde, we show how the
+Shopping Agent problem could be modeled:
 
@@
 
-In the model above one can see how the main elements of a problem domain
-can be mapped to \eus representation:
+The model above the main elements of a problem domain can be mapped to
+\eus representation in the \kcomment{following manner}:
 
@@
   \textit{AgentLocation} class to keep track of where the agent is at
-  any point in time. \nd supports a single rooted class
-  hierarchy so that state and behavior can be extended through
-  inheritance. We take advantage of that feature by making
-  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
-  class which doesn't allow predicates to overlap in time for the same
-  object instance. This feature is used for instance, to ensure that
-  an agent can only be at one place at a time in our example.
+  any point in time. \nd supports a single rooted class hierarchy so
+  that state and behavior can be extended through inheritance. We take
+  advantage of that feature by making \textit{AgentLocation} inherit
+  from \eus built-in \textit{Timeline} class which does not allow
+  predicates to overlap in time for the same object instance. This
+  feature is used for instance, to ensure that an agent can only be at
+  one place at a time in our example.
 
@@
 Domain rules can be defined in \nd in the context of a predicate or an
-action. In the example above we can see the main elements of a rule
-definition:
+action. In the example above, the main elements of a rule definition
+\kcomment{are as follows}:
 
@@
 
-With that specification for the Shopping Agent domain, a particular
-problem instance could be stated as follows:
+With the specification for the Shopping Agent domain \kcomment{above},
+a particular problem instance could be stated as:
 
@@
 products, product-location associations and the Shopping Agent that we
-want to plan for. Facts are used to specify the current state of the
-world, do not need any support and are represented as tokens in the
-\texttt{PDB}. In this case we use facts to specify that the Shipping
-Agent is initially at home. Goals are also represented as tokens in
-the \texttt{PDB}. Different from facts, goals trigger inference in \eu
-which determines whether flaws (unresolved subgoals, overloaded
-resources, etc.) need to be resolved; if so, search modules can be
-invoked to resolve those flaws as part of the deliberation
-process. Finally, constraints are restrictions on and between facts
-and goals. In the example above, we use temporal constraints to
-specify when facts hold and when the goals should be accomplished. In
-the model above, \texttt{before} and \texttt{after} are \texttt{Allen
+want to plan for. Facts, represented as tokens in the \texttt{PDB},
+are used to specify the current state of the world \kcomment{and} do
+not need any support. In this \kcomment{example} we use facts to
+specify that the Shipping Agent is initially at home. Goals are also
+represented as tokens in the \texttt{PDB}. Different from facts, goals
+trigger inference in \eu which determine whether flaws (unresolved
+subgoals, overloaded resources, etc.) need to be resolved; if so,
+search modules can be invoked to resolve those flaws as part of the
+deliberation process. Finally, constraints are restrictions on and
+between facts and goals. In the example above, we use temporal
+constraints to specify when facts hold and when goals need to be
+accomplished. \texttt{before} and \texttt{after} are \texttt{Allen
   Algebra} \cite{allen84} operators, which \eu supports to specify
@@
 In \eue, all variables and constraints from a problem domain
-description are stored inside the \texttt{CRE}.  The \texttt{CRE}
-provides a framework where any number of constraint types can
-collaborate to perform inference and prune variable domains using
-Bounds Propagation and Domain Reduction techniques \cite{marriott98}. 
-\eu uses a Propagator pattern which keeps track of a set of
-constraint instances of the same type and is responsible for ensuring
-variables are updated through inference whenever a relevant change occurs
- (variable domain modifications or constraint addition/deletion).
+description are stored inside the \texttt{CRE}.  It provides a
+framework where any number of constraint types can collaborate to
+perform inference and prune variable domains using Bounds Propagation
+and Domain Reduction techniques \cite{marriott98}.  \eu uses a
+Propagator pattern which keeps track of a set of constraint instances
+of the same type and is responsible for ensuring variables are updated
+through inference whenever a relevant change occurs (variable domain
+modifications or constraint addition/deletion).
 
@@
 relating them to the rest of the plan through constraints, by changing
-the start or end time of an activity, etc), or from an automated planner (by
-expanding subgoals, introducing precedence constraints to resolve
-mutual exclusion conflicts, etc). When these changes occur, each
-propagator is only able to ensure consistency for the constraints that
-are associated with it.  To maintain global consistency, the
+the start or end time of an activity, etc), or from an automated
+planner (by expanding subgoals, introducing precedence constraints to
+resolve mutual exclusion conflicts, etc). When these changes occur,
+each propagator is only able to ensure consistency for the constraints
+that are associated with it.  To maintain global consistency, the
 \texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
-\cite{mackworth77} where propagators are invoked repeatedly until
-the constraint network returns to a quiescent state.
+\cite{mackworth77} where propagators are invoked repeatedly until the
+constraint network returns to a quiescent state.
+
+\input{constraints-table}
 
@@
 An \eu user can introduce new constraint types by implementing
-specialized constraint and propagator classes; however, \eu provides a
+specialized constraint and propagator classes. However \eu provides a
 built-in constraint library with a range of arithmetic, temporal, set
 and resource constraint types so that a large number of problem
-domains can be modeled without having to introduce new constraint
-types. Tables \ref{tab:calconst}, \ref{tab:setconst},
+domains can be modeled. Tables \ref{tab:calconst}, \ref{tab:setconst},
 \ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
@@
 
-\input{constraints-table}
-
-\eu uses variables to explicitly represent timepoints for plan
-actions and states. Constraints among timepoints provide a natural
-way to express domain axioms. For example, in order to state that
-activity $A$ must occur before activity $B$ we can say that the end
-timepoint of $A$ is $leq$ the start timepoint of $B$.
-\cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network
-(\texttt{STN}). Such a network can be transformed into a Distance
-Graph (\texttt{DG}) where the outward arc from a node represents the
-maximum distance from the source node to the target node.
-Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
-a single constraint along with the resulting \texttt{DG}.
-\cite{dechter91} also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative cycle:
+\eu uses variables to explicitly represent timepoints for plan actions
+and states. Constraints among timepoints provide a natural way to
+express domain axioms. For example, in order to state that activity
+$A$ must occur before activity $B$ we can say that the end timepoint
+of $A$ is $leq$ the start timepoint of $B$.  \cite{dechter91} proposed
+that constraints among timepoints can be grouped together to form a
+Simple Temporal Network (\texttt{STN}). Such a network can be
+transformed into a Distance Graph (\texttt{DG}) where the outward arc
+from a node represents the maximum distance from the source node to
+the target node.  \kcomment{They} also show that shortest path
+algorithms \kcomment{(see \cite{corman} for a review)} could be used
+to propagate values in the network and discover a negative cycle:
 namely a path from a node to itself that has a path length less than
-$0$. If such a cycle exists, the network is inconsistent. It was
-further shown that a single-source shortest path algorithm was
-sufficient to detect a negative cycle and provide sufficient
-propagation to yield a backtrack-free search. \eus propagator for
-temporal constraints is a straight implementation of this algorithm.
+$0$. If such a cycle exists, the network is \kcomment{provably}
+inconsistent. It was further shown that a single-source shortest path
+algorithm was sufficient to detect a negative cycle and provide
+sufficient propagation to yield a backtrack-free search. \eus
+propagator for temporal constraints is an implementation of such a
+single-source algorithm. Fig. \ref{fig:stn} illustrates an
+\texttt{STN} with $2$ variables and a single constraint along with the
+resulting \texttt{DG}.
 
@@
   \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
-  \caption{\small An illustration of a STN and its corresponding
-    distance graph.}
+  \caption{\small An illustration of a Simple Temporal Network
+    (\texttt{STN}) and its corresponding distance graph.}
   \label{fig:stn}
@@
 which are then subsequently released. \textbf {Unary} resources are
-reusables with capacity$ = 1$, which allows \eu to perform run time
+reusables with capacity$=1$, which \kcomment{accounts for} runtime
 savings in comparison to the generic reusable class. Configuration
@@
 violations when a resource is completely depleted, plan flaws can be
-generated whenever a resource level is outside a specific upper or
-lower bound, or when instantaneous production or consumption exceeds a
-specific threshold.
+generated whenever a resource level is outside a specific bound, or
+when instantaneous production or consumption exceeds a specific
+threshold.
 
-Since \eu supports a flexible temporal representation, where time is
-represented by intervals, instead of singletons, determining whether a
-resource is the source of a plan flaw or a constraint violation in
-general is substantially more difficult than just computing a profile
-of the resource and checking against some limits.  With temporal
-flexibility, a number of resource profiles could be applicable
-depending on precisely when each production/consumption event occurs
-out of a range of possible values that are in the domain of the
-variables that represent the temporal extent for the event.
-\cite{Muscettola04,Muscettola06} proposed an algorithm where the
+Determining whether a resource is the source of a plan flaw or a
+constraint violation in general is substantially more difficult in
+\eus flexible representation, than just computing a profile of the
+resource and checking against some limits.  With temporal flexibility,
+a number of resource profiles could be applicable depending on
+precisely when each production/consumption event occurs out of a range
+of possible values that are in the domain of the variables that
+represent the temporal extent for the event.
+\cite{Muscettola04,Muscettola06} propose an algorithm where the
 variables that represent the time for production/consumption events
@@
 temporal propagator; constraints on the resource can then be seen as
-quantities flowing through that network. Computing the maximum flow
-then, yields the maximum resource usage and thus a lower bound on the
-resource level; computing a minimum flow will similarly yield an upper
-bound. Since maximum flow on a temporal network can be computed in
-polynomial time \cite{corman}, \eu takes advantage of this algorithm
-to efficiently compute upper and lower bound envelopes on the resource
-level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
-computation.
+quantities \emph{flowing} through that network. Computing the maximum
+flow then, yields the maximum resource usage and thus a lower bound on
+the resource level; computing a minimum flow will similarly yield an
+upper bound. They show that such upper and lower bound envelopes on
+the resource level can be computed in polynomial time
+\cite{corman}. Resource envelopes can be useful when trying to
+evaluate plan brittleness in the face of changing activity durations
+or start times. They can also be useful when searching for plans;
+however they are relatively expensive to compute. Therefore their
+ability to prune the search space must be weighted against the ability
+to search faster \cite{6007772, morris11}. Fig. \ref{fig:resenvelopes}
+shows an example of such a resource computation.
 
@@
 \includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
-\caption{\small An example of resource envelopes computed by \eu.}
+\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eu.}
 \label{fig:resenvelopes}
 \end{figure}
-
-Resource envelopes can be useful when trying to evaluate plan
-brittleness in the face of changing activity durations or start
-times. They can also be useful when searching for plans; however they
-are relatively expensive to compute. Therefore their ability to prune
-the search space must be weighted against the ability to search faster
-\cite{6007772, morris11}.
 
@@
 Since temporal intervals may extend indefinitely in both positive and
-negative directions, for problems with a temporal dimension it is also
-common to specify a \emph{planning horizon} when invoking a solver;
-any decisions that fall completely outside of the horizon can be
-ignored. This is useful to ignore activities that may occur far in the
-future or in the past to be relevant for a particular plan, ensuring
-that the solver is making as few commitments as necessary and reducing
-the complexity of plan generation.  The \emph{planning horizon} is
-also useful when dealing with recurrent activities; consider for
-instance a model where a medical checkup has to occur every $3$
-months. The user can model a recurring activity and then use the
-horizon to ensure that a finite number of medical checkup activities
-are generated as part of a plan.
+negative directions \kcomment{(\eg $[-\infty,+\infty]$)}, for problems
+with a temporal dimension it is also common to specify a
+\emph{planning horizon} when invoking a solver. Any decisions that
+fall completely outside of the horizon can be ignored, ensuring that
+the solver is making as few commitments as necessary and reducing plan
+generation complexity.  The \emph{planning horizon} is also useful
+when dealing with recurrent activities. Consider for instance a model
+where a medical checkup has to occur every $3$ months. The user can
+model a recurring activity and then use the horizon to ensure that a
+finite number of medical checkup activities are generated as part of a
+plan.
 
@@
 
-The initial partial plan is the state of \eus Plan Database after the
+The initial partial plan is the state of \eus \texttt{PDB} after the
 initial state has been instantiated. This results in a set of
 variable, object and token instances. Inference then takes place to
-detect flaws in the partial plan; the three kinds of flaws that can be
+detect flaws in the partial plan. The three kinds of flaws that can be
 detected are:
@@
 \item[\textbf{Open Condition}:] an open condition is an
-  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
-  by explicitly posted goals, or when a token is activated, rules may
-  cause the creation of \texttt{INACTIVE} slave tokens as described
-  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
-  three ways:
+  \texttt{INACTIVE} token which can be generated by explicitly posted
+  goals or during token activation when rules may cause the creation
+  of \texttt{INACTIVE} slave tokens as described in Section
+  \ref{sec:europa:pr}. Open Conditions can be resolved in three ways:
 
@@
   \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
-    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
-    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
-    and the temporal constraints involving $a$ are satisfied by
-    $a?$. Thus, $a$ and $a?$ can be considered the same token
-    without the need to introduce $a$ in the plan. Consequently, the
-    domain rules associated with $a$ are not introduced into the
+    matching \texttt{ACTIVE} token already in the plan. A token $a$ is
+    said to match another token $a?$ iff $a$ and $a?$ unify
+    \cite{russelnorvig} and the temporal constraints involving $a$ are
+    satisfied by $a?$. Thus, $a$ and $a?$ can be considered the same
+    token without the need to introduce $a$ in the plan. Consequently,
+    the domain rules associated with $a$ are not introduced into the
     planning process, since they are already triggered when $a?$ was
-    introduced in the plan.
+    \kcomment{instantiated} in the plan.
 
-  \item[\textbf{Activation}:] A token $a$ is introduced in the
-    current plan associating it with the appropriate timeline, without
+  \item[\textbf{Activation}:] A token $a$ is introduced in the current
+    plan associating it with the appropriate timeline, without
     choosing a specific time slot; if its timing causes a problem a
@@
     associated with $a$ are applied and the subgoal activities
-    resulting from those domain rules are introduced as \texttt{INACTIVE}
-    tokens.  This results in a number of open condition flaws,
-    corresponding to the new subgoal activities.
+    resulting from those domain rules are introduced as
+    \texttt{INACTIVE} tokens.  This results in a number of open
+    condition flaws corresponding to the new subgoal activities.
 
@@
 decisions.  Furthermore, it aids in reducing the amount of work done
-by a solver so that only the necessary refinements are made leaving a
-partial plan with sufficient flexibility. For example, one can bypass
-committing unbound timepoint variables since threats will force a
-solver to impose ordering restrictions on these variables.  Thus the
-planning process may yield partially-ordered plans for which all
-possible extensions are provably valid.
+by a solver so that only the necessary refinements are made,
+\kcomment{ensuring} a partial plan with sufficient flexibility. For
+example, one can bypass committing unbound timepoint variables since
+threats will force a solver to impose ordering restrictions on these
+variables.  Thus the planning process may yield partially-ordered
+plans for which all possible extensions are provably valid.
 
-The Plan Space Planning algorithm described above can be implemented
-in multiple ways depending on the approach chosen for flaw and
-resolver selection and for backtracking. \eus built-in solver
-implements a chronological backtracking algorithm that is summarized
-in Algorithm \ref{alg:europa:solve}.
+The Plan Space Planning algorithm described can be implemented in
+multiple ways depending on the approach chosen for flaw and resolver
+selection and for backtracking. \eus built-in solver implements a
+chronological backtracking algorithm that is summarized in Algorithm
+\ref{alg:europa:solve}.
 
-\begin{algorithm}[H]
+\begin{algorithm}
 \KwIn{PartialPlan $plan$}
@@
 \item[Line \textbf{\ref{li:backtrack}}:] Otherwise, retract the last
-  refinement operation and move on to try the next one.
+  refinement operation and move on to try the next.
 
@@
 consideration prior to commitment. Consistency testing is initiated by
-the \texttt{isInconsistent} procedure.
-
-The algorithm permits a heuristically controlled search by
-applying orderings for \texttt{chooseFlaw} and
-\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
-search. Heuristics can be implemented by applying ordering/filtering
-policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
-instance, by dealing with earliest flaws first, or by making planning
-decisions (open condition flaws) before scheduling decision (threats).
+the \texttt{isInconsistent} procedure.  The algorithm permits a
+heuristically controlled search by applying orderings for
+\texttt{chooseFlaw} and \texttt{makeDecisionPoint} resulting in a
+chronologically-backtracked search. Heuristics can be implemented by
+applying ordering/filtering policies for \texttt{chooseFlaw} and
+\texttt{makeDecisionPoint}, for instance, by dealing with earliest
+flaws first, or by making planning decisions (open condition flaws)
+before scheduling decision (threats).
 
 It should be emphasized that while this algorithm is commonly employed
-by applications that embed  \eu, it is only one of many that could be
-implemented. As a simple example, consider the following solver for
-the $N$-Queens problem, where an integer variable V$_i$ represents the
-queen in column $i$; V$_1$ $= 3$ implies that the variable on column
-$1$ is placed on row $3$.  Non-attack constraints are posted as
-arithmetic constraints among the V$_i$ variables to ensure that all
-the rows and diagonals contain only one queen.
+by applications that embed \eue, it is only one of many that could be
+implemented. As an example, consider the solver in Algorithm
+\ref{algo:n-queens} for the $N$-Queens problem, where an integer
+variable V$_i$ represents the queen in column $i$; V$_1$ $= 3$ implies
+that the variable on column $1$ is placed on row $3$.  Non-attack
+constraints are posted as arithmetic constraints among the V$_i$
+variables to ensure that all the rows and diagonals contain only one
+queen. The \texttt{CRE} can be used to compute the non-attack
+constraint violations, with \texttt{ce.getViolations()} a call on the
+\eu API. The \texttt{CRE} can also be used to pick the queen that has
+the most violations in \texttt{getQueenWithMaxViolation()} and to look
+ahead and see what moves are possible for a queen that would remove
+its violations in \texttt{getMoves()}.
 
-\begin{algorithm}[H]
+In similar fashion, the constraint violation and flaw reporting
+mechanisms that \eu provides can be used as building blocks that
+specialized or general purpose search algorithms can use to construct
+plans.
+
+\begin{algorithm}
   \KwIn{ConstraintEngine $ce$}
@@
   }
-  \caption{$solveNQueens(ce, maxIter)$}
+  \caption{$solveNQueens(ce, maxIter)$\kcomment{: An $N$-Queens solver}}
+  \label{algo:n-queens}
 \end{algorithm}
-
-\eus Constraint Reasoning Engine or \texttt{CRE} can be used to compute the
-non-attack constraint violations, with \texttt{ce.getViolations()} a
-call on the \eu API. The \texttt{CRE} can also be used to pick the queen 
-that has he most violations in \texttt{getQueenWithMaxViolation()}, and to
-look ahead and see what moves are possible for a queen that would remove its
-violations in \texttt{getMoves()}.
-
-In similar fashion, the constraint violation and flaw reporting
-mechanisms that \eu provides can be used as building blocks that
-specialized or general purpose search algorithms can use to construct
-plans.
 

### 2012-05-03 trunk/concepts.tex
updates

@@
 \centering
-\includegraphics[scale=0.3]{figs/flexible-timelines.pdf}
+\includegraphics[scale=0.4]{figs/flexible-timelines.pdf}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
 flexibility is critical when dealing with dynamic and uncertain
-conditions such as in the ocean, where precise timing of a robotic
-action might be indeterminate. Further, the advantage of flexibility
-can be contrasted with the consequences of the intrinsic inflexibility
-of traditional command sequences. Because they are inflexible,
-sequences must necessarily be designed considering worst case
-scenarios.
+conditions \kcomment{for robotic action in real-world environments},
+where precise timing of an action might be indeterminate. Further, the
+advantage of flexibility can be contrasted with the consequences of
+the intrinsic inflexibility of traditional command sequences. Because
+they are inflexible, sequences \kcomment{are brittle and therefore}
+must necessarily be designed considering worst case scenarios.
 
 \eu uses a \emph{state variable} representation
-\cite{mus94,jonsson00,mcgann09} to describe the evolution of state
-over time. The instantiated history of such state variable evolution
-over a temporal horizon we call \emph{timelines} and which represent a
-single thread in the execution of a concurrent system. At any given
-time each thread can execute a single procedure.  Thus each timeline
-consists of a sequence of procedures which encapsulate and describe
-state evolution; we call these instantiated atomic entities
-\emph{tokens}.  A token therefore describes a procedure invocation,
-the state variables on which it can occur, the parameter values of the
-procedure, and the time values defining the interval. We allow
-encapsulation of uncertainty within these tokens with a range of start
-and end times and parameters, all of which are encoded as variables. A
-constraint solver in turn manipulates these variables defined in a \eu
-domain model. For example, consider a scientific need to take water
-samples 100 meters from an environmental hotspot while an AUV is
-performing a Yo-Yo sawtooth pattern in the water-column. Two samples
-are needed if the feature's signal is above a threshold; one
-otherwise. The token that is capturing the sensory threshold has a
-parametric constraints to the token which fires the requisite water
-sampler. In addition, the start time of the water sampling procedure
-is highly dependant on the variability of sub-sea currents and actual
-speed of the vehicle. Therefore a number of values are possible for
-the start times and duration of the sampling all of which are valid
-combinations for desired outcomes.
+\cite{mus94,jonsson00,py10} to describe the evolution of state over
+time. The instantiated history of such \kcomment{an} evolution over a
+temporal horizon we call \kcomment{as a} \emph{timeline} and
+\kcomment{one which represents} a single thread in the execution of a
+concurrent system. At any given time each thread can execute a single
+procedure\kcomment{, with} each timeline consisting of a sequence of
+procedures which encapsulate and describe state evolution. We call
+these instantiated atomic entities \emph{tokens}.  A token therefore
+describes a procedure invocation, the state variables on which it can
+occur, the parameter values of the procedure, and the
+\kcomment{timepoints} defining the interval \kcomment{(see
+  Fig. \ref{fig:flex-timelines})}. We allow encapsulation of
+uncertainty within these tokens with a range of start and end times
+and parameters, all of which are encoded as variables. A constraint
+solver in turn manipulates these variables defined in an \eu domain
+model. For example, consider a scientific need to take water samples
+100 meters from an environmental hotspot while an AUV is performing a
+Yo-Yo sawtooth pattern\footnote{\kcomment{Variability in the water
+    column is along the vertical dimension. Since the Dorado platform
+    can only move forward, the Yo-Yo pattern is the most efficient
+    mechanism for studying water column properties.}}  in the
+water-column. Two samples are needed if the feature's signal is above
+a threshold; one otherwise. The token that is capturing the sensory
+threshold has a parametric constraints to the token which
+\kcomment{triggers} the requisite water sampler. In addition, the
+start time of the water sampling procedure is highly dependant on the
+variability of sub-sea currents and actual speed of the
+vehicle. Therefore a number of values are possible for the start times
+and duration of the sampling all of which are valid combinations for
+desired outcomes.
 
@@
 \item[\textbf{Variables}] Values that need to be represented to
-  describe the problem domain and over which we may want to specify
-  constraints. In the Shopping Agent example, the times at which the
+  describe the problem domain \kcomment{over which constraints are
+    specified}.  In the Shopping Agent example, the times at which the
   agent needs to leave or be back, or executes a purchase, are all
-  instances where variables representing time is used.  As noted in
-  Section \ref{sec:europa:cp}, a variable can take values from a
-  domain. In \eu, domains are defined over a specific data type, the
-  primitive data types supported are, \textit{String},
+  instances where \emph{variables} representing time is used.  As
+  noted in Section \ref{sec:europa:cp}, a variable can take values
+  from a domain. In \eu, domains are defined over a specific data
+  type; the primitive data types supported are \textit{String},
   \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
   below). Further, two types of variable Domains can be defined over a
-  data type, \textit{Enumerated} and \textit{Interval} (used only for
-  numeric data types, they are specified by a [LowerBound,UpperBound]
-  pair, for instance: [1,10], [5.0,100.0]). It is this latter ability
-  to represent and reason over numeric intervals that forms the basis
-  of representing flexible plans.
+  data type \textit{Enumerated} and \textit{Interval} (used only for
+  numeric data types, they are specified by a
+  $[LowerBound,UpperBound]$ pair, for instance: $[1,10]$,
+  $[5.0,100.0]$). It is this latter ability to represent and reason
+  over numeric intervals that forms the basis of representing flexible
+  plans.
 
-\item[\textbf{Constraints}] valid plans in most problem domains have
-  to satisfy a number of restrictions, for instance, a Shopping Agent
+\item[\textbf{Constraints}] Valid plans in most problem domains have
+  to satisfy a number of restrictions for instance, a Shopping Agent
   may only be able to perform one task at a time, may have limited
-  time or an energy or fuel budget to perform a task. In \eu, these
-  restrictions are represented by constraints, which can be defined
-  over any combination of variables in a domain. \eu provides a
-  \emph{Constraint Library} that implements a number of useful
-  constraints (temporal, resource, relational), while allowing the
-  possibility to add new constraints if required by a particular
-  domain.
+  time or an energy or fuel budget to perform a task. These
+  restrictions are represented by \emph{constraints}, which can be
+  defined over any combination of variables in a domain. \eu provides
+  a \emph{Constraint Library} that implements a number of useful
+  constraints (temporal, resource, relational) while allowing the
+  possibility to \kcomment{augment} with new constraints if required.
 
 \item[\textbf{Objects}] The items we wish to describe and refer to in
-  a domain are considered objects. As in the case with object-oriented
-  analysis and design, one can seek out the nouns in any domain
-  description to find likely objects. In the Shopping Agent example,
-  we might consider the Agent, Products and Locations all to be
-  objects. Objects have state and behavior, for example, a Shopping
-  Agent can have, its location, a bag that contains the products it
-  has already purchased (the bag can in turn be another object) and
-  behavior as in going to a location to look for a product, perform a
-  purchase, return home, etc.
-
-  Objects that have similar state and behavior can be described
+  a domain are considered objects.  In the Shopping Agent example, we
+  might consider the Agent, Products and Locations all to be
+  \emph{objects}. Objects have state and behavior.  For example, a
+  Shopping Agent can have its location, a bag that contains the
+  products it has already purchased (the bag can in turn be another
+  object) and behavior, as in going to a location to look for a
+  product, perform a purchase, return home, etc. all as objects.
+  Those that have similar state and behavior can be described
   generically in terms of object types or classes. \eu allows the
@@
   where the domains of variables $a$ and $l$ can have more than one
-  possible value. Or they can be grounded plans, where single values
-  will be specified for each variable as showing in the N-queen's
-  problem.
+  possible value. Or they can be grounded \kcomment{representations},
+  where single values will be specified for each variable as shown in
+  the $N$-Queens problem.
 
-  In general, for an executable plan it is insufficient to state
-  predicates that describe required state or behavior without also
-  specifying some temporal extent over which each of those predicates
-  hold. Predicates that are always true can be thought to hold from
-  the beginning to the end of the planning horizon. However, in
-  practice, the temporal extent of interest must be defined with
-  timepoints \cite{Dean88} to represent its start and end.  In such
-  representation, one writes \texttt{At($a$,$s$,$e$,$l$)} to indicate
-  that agent $a$ is at location $l$ from time $s$ to time $e$. In
-  fact, this pattern of using such predicates to describe both state
-  and behavior of objects is prevalent enough that \eu the token
-  representation. Token have built in variables to indicate the object
-  to which the statement principally applies as well as the timepoints
-  over which it holds. A token is an instance of a predicate that
-  represents an object's state or behavior and is defined over a
-  temporal extent. In \eu, all predicate instances are tokens. Also,
-  in the same way that objects are described generically by classes,
-  in \eu tokens can be described generically by token types. Every
-  token has five built-in variables:
+  In general, for an executable plan it is insufficient to
+  \kcomment{define} predicates that describe required state or
+  behavior without also specifying some temporal extent over which
+  each of those predicates hold. Predicates that are always true can
+  be thought to hold from the beginning to the end of the planning
+  horizon. However, in practice, the temporal extent of interest must
+  be defined with timepoints to represent its start and end.  In such
+  \kcomment{a} representation, one writes \texttt{At($a$,$s$,$e$,$l$)}
+  to indicate that agent $a$ is at location $l$ from time $s$ to time
+  $e$. In fact, this pattern of \kcomment{usage} to describe both
+  state and behavior of objects is prevalent enough that \eu
+  \kcomment{encapsulates into} the token
+  representation. \kcomment{Tokens} have built in variables to
+  indicate the object to which \kcomment{it primarily} applies as well
+  as the timepoints over which it holds. A token, therefore, is an
+  instance of a predicate that represents an object's state or
+  behavior and is defined over a temporal extent. In \eu, all
+  predicate instances are tokens. \kcomment{In} the same way that
+  objects are described generically by classes, in \eu tokens can be
+  described generically by token types. Every token has five built-in
+  variables:
 
@@
   induce ordering requirements among its tokens to ensure no temporal
-  overlap may occur among them \cite{mus94}. Fig. \ref{fig:europapr1} below and several others
-  after that illustrate how timelines are used to represent state in a plan.
+  overlap may occur among them \cite{mus94}. Fig. \ref{fig:europapr1}
+  and others \kcomment{in this chapter,} illustrate how timelines are
+  used to represent state in a plan.
 
 \item \textit{Resources}: Metric resources, e.g. the energy of a
-  battery or the capacity of a cargo hold, are objects with an
-  explicit quantitative state in time and with a circumscribed range
-  of changes that can occur to impact that state i.e. produce,
-  consume, use, change.  Resources are such a common requirement for \eu
-  users that built-in object types (with their corresponding token
-  types to denote production, consumption, etc) are provided.
-  Instances of classes derived from a resource will induce ordering
-  requirements on their tokens in order to ensure that the level of
-  the resource remains within specified limits.
+  battery or the capacity of a cargo hold are objects with an explicit
+  quantitative state in time and with a circumscribed range of changes
+  that can occur to impact that state i.e. produce, consume, use,
+  change.  Resources are a common enough requirement for \eu users
+  that built-in object types (with their corresponding token types to
+  denote production, consumption, etc) are provided.  Instances of
+  classes derived from a resource will induce ordering requirements on
+  their tokens in order to ensure that the level of the resource
+  remains within specified limits.
 
@@
   asks the Shopping Agent to own a specific product by a specific
-  time, or implicitly when subgoals are created as a result of domain
-  rules. For instance, when the Shopping Agent must get to a location
-  where a product is available before it is able to purchase it: in
-  this case the original goal of purchasing the product spawns a
-  subgoal of being at a specific location without which the purchase
-  will not occur as defined in the domain model.  Looking at how a
-  simple Shopping Agent goal would be addresed by \eu helps illustrate
-  how Token States support the planning process.  Let's assume the
-  planner starts with a partial plan that places the Shopping agent at
-  home at time $0$ with no products in possession. Let's also assume
-  that goals are posted for the Agent to own a drill by time $10$ and
-  to be home by time $20$ as shown in Fig. \ref{fig:europapr1}
+  time, or implicitly when subgoals are created as a consequence of
+  domain rules. For instance, the Shopping Agent must get to a
+  location where a product is available before it is able to purchase
+  it. In this case the original goal of purchasing the product spawns
+  a \emph{subgoal} of being at a specific location without which the
+  purchase will not occur as defined in the domain model.
+
+  Looking at how a simple Shopping Agent goal would be addresed by \eu
+  helps illustrate how Token States support the planning process.
+  Let's assume the planner starts with a partial plan that places the
+  Shopping agent at home at time $0$ with no products in
+  possession. Let's also assume that goals are posted for the Agent to
+  own a drill by time $10$ and to be home by time $20$ as shown in
+  Fig. \ref{fig:europapr1}
 
   \begin{figure}[t]
-    \centering \subfloat[\small Timelines representing initial state
-    and goals for a Shopping
-    Agent.]{\label{fig:europapr1}\includegraphics[scale=0.35]{figs/europa-pr-1.pdf}}\qquad
+    \centering 
+    \subfloat[\small Timelines representing initial state
+    and goals for a Shopping Agent.]{\label{fig:europapr1}\includegraphics[scale=0.4]{figs/europa-pr-1.pdf}}\qquad
     \subfloat[\small State of timelines after satisfying a goal to be
-    home by time $20$ on merging with initial
-    state.]{\label{fig:europapr:merge}\includegraphics[scale=0.35]{figs/europa-pr-merge.pdf}}\qquad
+    home by time $20$ on merging with initial state.]{\label{fig:europapr:merge}\includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}}\qquad
     \subfloat[\small State of timelines after satisfying a goal to own
     a drill by time $10$ by activating corresponding
-    token.]{\label{fig:europapr:activate}\includegraphics[scale=0.35]{figs/europa-pr-activate.pdf}}
+    token.]{\label{fig:europapr:activate}\includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}} 
     \caption{\small Evolution of a timeline from the Shopping Agent example.}
@@
   represent the goals are not immediately placed on the
-  \texttt{AgentBag} or \texttt{AgentLocation} timelines since \eu must
-  first determine whether it can achieve those goal and how. When the
-  goals are posted and before any decisions are made, the
-  \texttt{Own(Drill)} and \texttt{At(Home)} tokens are created in an
-  \texttt{INACTIVE} state. \eu then considers the partial plan and
-  evaluates two possible alternatives:
+  \texttt{AgentBag} or \texttt{AgentLocation} timelines, since \eu
+  must first determine whether it can achieve those \kcomment{goals}
+  and how. When the goals are posted and before any decisions are
+  made, the \texttt{Own(Drill)} and \texttt{At(Home)} tokens are
+  created in an \texttt{INACTIVE} state. \eu then considers the
+  partial plan and evaluates two possible alternatives:
 
@@
 are transferred to the active token upon which it is merged. Merging a
-token requires finding a target active token that is compatible. Two
-tokens are merge-compatible if they are instances of the same
-predicate and there are no empty intersections between corresponding
-token variables.  The effect of merging is illustrated in
-Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
-satisfied by merging with the initial state that said that the Agent
-is home at time $0$.  The merge is possible because the tokens are of
-the same type \texttt{At}, and intersecting the domains for all of
-their variables doesn't yield an empty domain (which represents an
-inconsistency), in this case the token variables are the location,
-which is \texttt{Home} in both cases, and the temporal variables
-(start, end, duration), which can be intersected without causing an
+token requires finding a target \texttt{ACTIVE} token that is
+compatible. Two tokens are merge-compatible if they are instances of
+the same predicate and there are no empty intersections between
+corresponding token variables.  The effect of merging is illustrated
+in Fig. \ref{fig:europapr:merge}. The goal to be home by time $20$ is
+satisfied by merging with the initial state that \kcomment{indicated}
+that the Agent is home at time $0$.  The merge is possible because the
+tokens are of the same \texttt{At} type and intersecting the domains
+for all of their variables does not yield an empty domain (representing
+an inconsistency). In this case the token variables are the location,
+which is \texttt{Home} in both cases and the temporal variables start,
+end and duration, which can be intersected without causing an
 inconsistency. After the token representing the goal is merged onto
 the token representing the initial state, the resulting token
-represents not only the fact that the Agent must be \texttt{At(Home)}
-by time 0 but also \texttt{At(Home)} at time $20$, this is reflected
-in the new constraints on the token's start and end variables as
-depicted in Fig. \ref{fig:europapr:merge}.
+represents not only that the Agent must be \texttt{At(Home)} by time
+$0$ but also \texttt{At(Home)} at time $20$, This is reflected in the
+new constraints on the token's start and end variables as depicted in
+Fig. \ref{fig:europapr:merge}.
 

### 2012-05-03 trunk/arch.tex
updates

@@
 controller which could bring automated planning closer to low-level
-control of the vehicle while maintaining its reactivity.
-
-Traditionally planning has been deemed computationally expensive with
-planner performance restricting robot reactivity
+control of the vehicle while \kcomment{ensuring}
+reactivity. Traditionally planning has been deemed computationally
+expensive with planner performance restricting robot reactivity
 \cite{ghallab04,Dias:2003ua}. In control architectures where task
@@
 the planner can plan for. In such a situation, the agent may thrash
-\footnote{\kcomment{In such an event as the planner works towards goal
-    satisfaction, the initial conditions leading to achieve that goal
-    are no longer valid. The planner then tries to achieve a different
-    goal which too has to be discarded similarly. And so on.}}  if the
-internal state of the plan gets out of synch with the actual state of
-the world.
-
-\kcomment{A important consequence of this has been that} the planning
+\footnote{\kcomment{In such an event even as the planner works towards
+    goal satisfaction, the initial conditions leading to achieve that
+    goal are no longer valid. The planner then tries to achieve a
+    different goal which too has to be discarded similarly. And so
+    on.}}  if the internal state of the plan gets out of sync with the
+actual state of the world.
+
+\kcomment{An important consequence of this has been that} the planning
 problem managed {\em in situ} has traditionally remained fairly
@@
 framework with \kcomment{an} unified declarative model. Our approach
-with \rx builds on this design methodology albeit in a substantially
-systematic manner. The \rx architecture contributes to agent
-architectures over and beyond \texttt{IDEA} in the following
-\kcomment{significant} ways:
+builds on this design methodology albeit in a substantially systematic
+manner.
+
+The \rx architecture contributes to agent architectures over and
+beyond \texttt{IDEA} in the following \kcomment{significant} ways:
 
@@
 \item \emph{Partitioning} the global planning problem into multiple
-  decision loops, called {\em reactors}. Each reactor has its own
-  scope both functionally (one for example focusing on the conversion
-  of high level waypoints into low-level commands) and temporally as
-  each reactor declares its planning look-ahead $\pi$ and expected
-  latency $\lambda$ before being able to produce \kcomment{a} plan to
-  the agent depending on where it is located in a reactor dependency
-  graph such as shown in Fig. \ref{fig:agent}. A reactor closer to the
-  hardware in the hierarchy, is expected to deliberate less and be
-  highly responsive to observations from the world. This implies lower
-  values of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex}
-  below).
+  decision loops called {\em reactors}. Each reactor has its own scope
+  both functionally (one for example focusing on the conversion of
+  high level waypoints into low-level commands) and temporally as each
+  reactor declares its planning look-ahead ($\pi$) and expected
+  latency ($\lambda$) before being able to produce \kcomment{a} plan.
+  \kcomment{These parameters depend} on where \kcomment{a reactor} is
+  located in a dependency graph such as \kcomment{the one} shown in
+  Fig. \ref{fig:agent}. A reactor closer to the hardware in the
+  hierarchy, is expected to deliberate less and be highly responsive
+  to observations from the world. This implies lower values of $\pi$
+  and $\lambda$ (see Section \ref{sec:arch:trex} below). \kcomment{In
+    the current design, $\pi$ and $\lambda$ are application dependent
+    and fixed at design time.}
 
 \item \emph{Coupling} a tighter integration between the planning
-  process and world evolution to ensure that planning occurs
-  \emph{during continuous} state update in the {\em same}
-  \kcomment{plan} database used for planning. By doing so, the planner
-  is better informed in its deliberation despite a larger latency than
-  the changing world.
+  process and world evolution \kcomment{with the use of the {\em same}
+    \kcomment{plan} database}. By doing so, the planner is better
+  informed in its deliberation despite a larger latency than the
+  changing world.
 \end{itemize}
@@
 covered in \cite{py10, rajan12}. Coupled interaction between planning
-and the world on the other hand, has a significant implication on how
+and the world on the other hand, has a substantial impact on how
 planning can be integrated in such frameworks. In this chapter we show
-how using the \eu planner can be coupled to our framework by first
-linking concepts in temporal planning with the \rx framework. Note
-that in the following treatise, we call the formal semantic framework
-of \rx as the 'agent'. Reactors constitute an entity within this agent
-and are standalone blocks of inference. The scope, detail and number
-of reactors are application specific and outside the scope of this
+how using the \eu planner can be coupled to the \rx framework by first
+linking concepts in temporal planning with the framework. Note that in
+the following treatise, we call the formal semantic framework of \rx
+as the 'agent'. Reactors constitute an entity within this agent and
+are standalone blocks of inference. The scope, detail and number of
+reactors are application specific and outside the scope of this
 chapter.
@@
 subscribe to {\em external} state variables owned by other reactors,
-for which it will receive new updates when the owner reactor generates
-its {\em internal} state.
+for which it will receive updates when the owner reactor
+\kcomment{makes changes} to its {\em internal} state \kcomment{due to
+  deliberation or an exogenous state update}.
 
@@
 dependency graph from the least dependent -- ones with no {\em
-  external} state variables -- to the most, \ie which typically do not have
-{\em internal} state variables. Fig. \ref{fig:agent} shows an example
-of such a dependency graph implemented as a directed acyclic graph.
-
-Conversely each reactor can request a future change on one of its {\em
-  external} state variables which will be transformed as an {\em
-  internal} goal to the owner reactor(s).  Such \emph{goal
-  dispatching} is expected to produce a deliberation phase for the
-receiving reactor which can cyclically in turn, produce a plan
-impacting its {\em external} state variables' future that will
-propagate down following the same mechanism.
+  external} state variables -- to the most, \ie which typically do not
+have {\em internal} state variables. Conversely each reactor can
+request a future change on one of its {\em external} state variables
+which will be transformed as an {\em internal} goal to the owner
+reactor(s).  Such \emph{goal dispatching} is expected to produce a
+deliberation phase for the receiving reactor which can cyclically in
+turn, produce a plan impacting its {\em external} state variables'
+future that will propagate \kcomment{down the hierarchy} following the
+same mechanism. Fig. \ref{fig:agent} shows an example of such a
+dependency graph implemented as a directed acyclic graph.
+
+\begin{figure}[b]
+  \centering
+  \includegraphics[scale=0.45]{figs/token-disptch.pdf}
+  \caption{\small \kcomment{Dispatching a token occurs only when the
+      request start time overlaps a window. \texttt{Goal$_1$}
+      therefore gets dispatched from \texttt{Reactor$_1$} to
+      \texttt{Reactor$_2$} but not \texttt{Goal$_2$} to avoid early
+      commitment. Tokens from \texttt{Reactor$_1$} are 'dispatched'
+      for execution to \texttt{Reactor$_2$} only when necessary while
+      meeting the conditions in Definition \ref{def:dispatch}. $\tau$
+      signifies the execution frontier as time moves from left to
+      right in the illustration.}}
+  \label{fig:reactor-lookahead}
+\end{figure}
 
@@
   indicative of the number of ticks in the future this reactor is
-  looking ahead during deliberation.
+  looking \kcomment{towards} deliberation.
 
@@
   Where  $exts(r)$ is the set of the state variables which are
-  {\em externals} to $r$. 
+  {\em external} to $r$. 
 \end{definition}
@@
 critical at the agent level since it can impact the
-planning/deliberation process on {\em external} state variables of a
+planning/deliberation process on external state variables of a
 reactor. For simplicity we abstract it out as being equivalent to the
@@
 level this execution latency is used to define the {\em dispatch
-  window} of the same state variable.
+  window} of the same state variable \kcomment{(see \cite{py10} for formal
+proofs)}.
 
@@
   external} state variable of a reactor while avoiding flooding the
-owner reactor of this state variable.
-
-The choice of both the latency and look-ahead of a reactor can be
-delicate as one is dependent on the other. Needless to say, latency is
-highly dependent on exogenous paramaters such as the host computers
-computing power and the number of reactors needing to deliberate at
+owner reactor of this state variable \kcomment{and avoiding early
+  commitment.} The choice of both the latency and look-ahead of a
+reactor can be delicate as one is dependent on the other. Latency is
+dependent on exogenous paramaters such as the \kcomment{computer power
+  of the host} and the number of reactors needing to deliberate at
 once, among others. As the design stands, these variables should be
-evaluated carefully by the system designer balancing agent efficacy.
+evaluated carefully by the system designer balancing agent
+efficacy. Fig. \ref{fig:reactor-lookahead} illustrates an example
+between two reactors where one token is dispatched when overlapping
+the dispatch window while another is not.}
 
@@
 
-\item The future objectives a reactor has on \textsf{Internal} state
-  will likely imply sub-objectives to be reported to the owner of its
-  {\em external} state variables depending on their current
-  values. Should the \textsf{Pilot} want to visit a specific
+\item The future objectives a reactor has on its \emph{internal} state
+  variables will likely imply sub-objectives to be reported to the
+  owner of its {\em external} state variables depending on their
+  current values. Should the \textsf{Pilot} want to visit a specific
   \textsf{Waypoint} -- given its current position as provided by
@@
   \textsf{Nav\_cmd} states that should eventually help it reach this
-  location. %\comment{this explanation could do with a figure which I
-    %believe we have used previously}
+  location.
 
@@
 agnostic where its {\em external} state information is managed as long
-as this information is available and properly maintained for managing
-{\em internal} state information. This {\em internal} state in turn
-may be used by other reactors. Such a formalism also constrains
+as this information is available and maintained for managing {\em
+  internal} state information.  Such a formalism also constrains
 reactor interaction to state information exchange which comes in two
@@
 
-\item {\em Observation on current state values}: This information
-  propagates \emph{up} in the reactor dependency graph at the
+\item {\em Observation on current state values}: This propagates
+  \emph{up} in the reactor dependency graph \kcomment{during} the
   synchronization phase and occurs at the beginning of every
   tick. During this phase the agent provides updates to each reactor
-  on their {\em external} state variables for this tick so they can
-  compute their {\em internal} state variable values for this same
-  tick. By doing so we propagate a consistent view of that state of
-  the world throughout all the reactors as time
-  advances. \kcomment{Observations can therefore be both sensor
-    measurements as well state passed up from other reactors. The
-    strict rules of timeline ownership enable a clear policy for
-    conflict resolution: observations dominate expectations.}
+  on their {\em external} state variable(s) to \kcomment{enable
+    computation of} their {\em internal} state variable values for
+  this same tick. By doing so we propagate a consistent view of world
+  state throughout the reactors as time advances. \kcomment{Further,
+    observations can be both sensor measurements as well as state
+    information, passed up from other reactors. The strict rules of
+    timeline ownership enable a clear policy for conflict resolution:
+    observations dominate expectations.}
 
 \item {\em Goal request on future state}: This information propagates
-  \emph{down} in the reactor dependency graph. In order to satisfy own
-  {\em internal} objectives, a reactor may have a plan that relies on
-  future values of one of its {\em external} state. When such a plan
-  is identified the agent ensures that this part of the plan is
-  provided as a goal to the owner of the given {\em external} state
-  variable(s).
+  \emph{down} in the reactor dependency graph. In order to satisfy its
+  own {\em internal} objectives, a reactor may have a plan that relies
+  on future values of one \kcomment{(or more)} of its {\em external}
+  state. When such a plan is identified the agent ensures that this
+  part of the plan is provided as a goal to the owner of the given
+  {\em external} state variable(s).
 
@@
 
-The overall execution cycle of the agent is also abstracted out as a
-continuous planning/deliberation cycle between all reactors. At the
-agent interface level, each reactor provides only two abstract calls
-that conceptualize execution:
+% The overall execution cycle of the agent is also abstracted out as a
+\kcomment{Within \rx planning/deliberation is a continuous} cycle
+between all reactors. At the agent interface level, each reactor
+provides only two abstract calls that conceptualize execution:
 
@@
   deliberation. In the context of \eu for example, one deliberation
-  step corresponds to a single flaw resolution (Section
-  \ref{sec:europa:arch}) while ensuring that the resulting plan is not
-  inconsistent. Backtracking is included in the step should it be such
-  a case.  The returned value by the \texttt{step} function is a
-  boolean which indicates whether the reactor needs an extra step to
-  deliberate or if this last step resulted in a complete plan within
-  the reactor's look-ahead. The part of the plan that covers the
-  reactor's {\em external} timelines is provided to be dispatched as a
-  goal.
+  step corresponds \kcomment{to the resolution of} a single flaw
+  (Section \ref{sec:europa:arch}) while ensuring that the resulting
+  plan is not inconsistent. Backtracking is included in the step
+  should it be \kcomment{necessary}.  The returned value by the
+  \texttt{step} function is a boolean which indicates whether the
+  reactor needs an extra step to deliberate or if this last step
+  resulted in a complete plan within the reactor's
+  look-ahead. \kcomment{Subsequently} the part of the plan that covers
+  the reactor's {\em external} timelines is dispatched as a goal.
 
@@
 Both these functions have different scope and temporal constraints.
-The \texttt{synchronize} call is considered as atomic and will focus
-only on state inference for the current tick by identifying the
-current value of {\em internal} state variables from the latest
-updates on {\em external} state variables. This process propagates up
-in the reactor dependency graph at every tick to ensure that all
-reactors are aware of the current state of the world. \texttt{step}
-relates to a single decision step of the planning process focused on
-the future evolution of state variables; it can take multiple steps
-spanning multiple ticks for a reactor to produce a plan that is both
-complete and valid. How the reactor identifies that its plan is
-complete and valid is its responsibility and depends on the way each
-reactor resolves its plan at each {\em step}. The general idea is that
-the reactor is able to find a valid strategy in order to satisfy as
-many of its {\em internal} goals as possible. For \eue, this means
-that the solver is unable to identify any further flaws to be resolved
-in the plan.
-
-When such plan is identified, its {\em external} timeline(s) are then
-dispatched to the corresponding reactors which own these times and
-which in turn can start their own deliberation \texttt{step}s to
-satisfy these new objectives. Consequently execution occurs as these
-goals and plans propagate down in the reactor hierarchy until they
-eventually become goals for the lowest level reactor which -- instead
-of deliberating -- may just transform this goal into a command for
-execution in hardware.
-
-\label{sec:delib-exec-as}
+The \texttt{synchronize} call is considered as atomic and focuses on
+state inference for the current tick by identifying the current value
+of {\em internal} state variables from the latest updates on {\em
+  external} state variables. This process propagates up in the reactor
+dependency graph at every tick to ensure that all reactors are aware
+of current world state. \texttt{step} relates to a single decision
+step of the planning process focused on the future evolution of state
+variables; it can take multiple steps spanning multiple ticks for a
+reactor to produce a plan that is both complete and valid. How the
+reactor identifies that its plan is complete and valid is its
+responsibility and depends on the way each reactor resolves its plan
+at each {\em step}. The general idea is that the reactor is able to
+find a valid strategy in order to satisfy as many of its {\em
+  internal} goals as possible. For \eue, this means that the solver is
+unable to identify any further flaws to be resolved in the
+plan. \kcomment{Fig. \ref{fig:sync-cycle} illustrates the distinction
+  between a tick, \texttt{step} and synchronization within the tick.}
+
+When such a plan is identified, \kcomment{the reactors} {\em external}
+timeline(s) are then dispatched to their \kcomment{respective reactor
+  owners,} which in turn can start their own deliberation
+\texttt{step}s to satisfy these new objectives. Consequently execution
+occurs as these goals and plans propagate down in the reactor
+hierarchy until they eventually become goals for the lowest level
+reactor which -- instead of deliberating -- may just transform this
+goal into a command for execution in hardware.
+
+% \label{sec:delib-exec-as}
 \begin{figure}[!htbp]
@@
   \vskip-1pc
-  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
-  \caption{\small \rx reactor execution cycle: {\em Deliberation} is
+  \subfloat[\small \kcomment{A view of synchronization and deliberation within a
+  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle}} 
+  \qquad
+  \subfloat[\small A \rx reactor's execution cycle: {\em Deliberation} is
     interrupted by {\em synchronization} at the beginning of every
-    {\em tick} allowing integration of state information.}
-  \label{fig:tick-exec}
-  \vskip-0.8pc
+    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle}}
+    \caption{\small \kcomment{Detailed and abstracted views of
+        synchronization and deliberation within one and multiple tick
+        cycles.}}
+  \label{fig:sync-plan-cycle}
 \end{figure}
@@
 
-\subsubsection{Handling Failure in \rx}
+\subsubsection{Handling \kcomment{Reactor} Failure in \rx}
 \label{sec:rx-reactor-failure}
@@
   external}, are informed of the failure of this timeline.  All goals
-exchanged between reactors are considered \emph{rejectable} by the
-reactor that manages them. As a result in such a situation, the
-failure to find a plan for these goals, is \emph{not} considered a
-critical failure of the reactor. It will only result \kcomment{in}
-these goals being ignored until newer goals are received.
+exchanged between reactors \kcomment{and these timelines} are
+\kcomment{then} considered \emph{rejectable} by the reactor that
+manages them. As a result in such a situation, the failure to find a
+plan for these goals, is \emph{not} considered a critical failure of
+the reactor. It will only result \kcomment{in} these goals being
+ignored until newer goals are received.
+
+\begin{figure}[b]
+  \centering \subfloat[\small When the \texttt{Sampling} reactor fails
+  to synchronize, it is removed from the reactor dependency graph
+  along with all its {\em internal} timelines while generating a
+  \texttt{Failed} observation.]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.pdf}
+  \label{fig:fail0}}
+  \qquad
+  \subfloat[\small In the worst case this \texttt{Failed} observation may imply
+  other reactors that depend on the \texttt{Sampling} reactor might fail to
+  synchronize too leading to a bottom-up destruction of ``faulty'' reactors.]{
+    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1}\label{fig:fail1}}
+  \caption{\small Impact of the \texttt{Sampling} reactor failure in
+    the AUV agent \kcomment{from} Fig. \ref{fig:agent}.}
+\end{figure}
 
@@
 which indicates that this state variable is no longer managed by its
-owner reactor. This leads to two potential outcomes based on how the
-reactor is implemented:
+owner reactor. This leads to two \emph{potential} outcomes based on
+how the reactor is implemented:
 
@@
 
-\item The \texttt{Survey controller} model allows it to integrate the
-  \texttt{Failed} observation. For example it can identify that,
+\item The \texttt{Survey controller} model \kcomment{could} integrate
+  the \texttt{Failed} observation. For example it can identify that,
   subsequent use of the water sampler is no longer possible.  The \rx
@@
 \end{enumerate}
-
-\begin{figure}[b]
-  \centering
-  \subfloat[\small When the \texttt{Sampling} reactor fails to
-  synchronize, it is removed from the reactor dependency graph along
-  with all its {\em internal} timelines while generating a \texttt{Failed} observation]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.pdf}
-  \label{fig:fail0}}
-  \qquad
-  \subfloat[\small In the worst case this \texttt{Failed} observation may imply
-  other reactors that depend on the \texttt{Sampling} reactor might fail to
-  synchronize leading to a bottom-up destruction of ``faulty'' reactors.]{
-    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1}\label{fig:fail1}}
-  \caption{\small Impact of the \texttt{Sampling} reactor failure in
-    the AUV agent \kcomment{from} Fig. \ref{fig:agent}}
-\end{figure}
 
@@
   observation was received by the reactor with the plan showed in
-    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
+    Fig. \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
     \caption{\small The Shopping example: Tokens in {\color{red}red}
       indicate observations, tokens in {\color{blue}blue} the goal
-      that this plan attempts to solve for. Arrows indicate temporal
-      constraints. }
+      that this plan attempts to solve for \kcomment{and tokens in
+        {\color{grey}grey} the result of deliberation}. Arrows
+      indicate temporal  constraints \kcomment{with $==$ indicating 'equality' and $\geq$
+        indicating 'greater than equal to' in timepoints}.}
   \label{fig:shop:exec}
@@
   \centering
-  \subfloat[Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
-  \subfloat[Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \subfloat[\small Conflicting Observation.]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
+  \subfloat[\small Relaxed Partial Plan.]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
   \caption{\small Illustration of a conflict during synchronization

### 2012-05-03 trunk/europa.tex
updates

--- 2012-02-24 trunk/europa.tex
@@
 \centering
-\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
 \caption{\small The \eu software architecture.}
@@
 constraints that define relationships among them. It also provides an
-efficient arc consistency mechanism \cite{mackworth77}. The
-\texttt{CRE} is designed so that specialized reasoning algorithms for
-new constraints can be easily and efficiently plugged in. The
-\textbf{Plan Database} (\texttt{PDB}) manages object and token
-types. It also maintains the state of partial plans in terms of
-object, variable and token instances. This constitutes the data
-foundation on which end user applications and automated problem solvers
- are built. The \textbf{Rules Engine} (\texttt{RE})
-modularly adds the semantics of domain rules that describe
-dependencies between tokens. The \textbf{Solver Module} provides the
-framework for flaw detection and resolution that is needed to
-implement search algorithms for planning. The \textbf{API Layer}
-provides access to services from all the different modules in a
-programmatic way so that \eu can be easily embedded into specialized
+efficient arc consistency mechanism \cite{mackworth77} designed so
+that specialized reasoning algorithms for new constraints can be
+easily and efficiently plugged in. The \textbf{Plan Database}
+(\texttt{PDB}) manages object and token types. It also maintains the
+state of partial plans in terms of object, variable and token
+instances. This constitutes the data foundation on which end user
+applications and automated problem solvers are built. The
+\textbf{Rules Engine} (\texttt{RE}) adds the semantics of domain rules
+that describe dependencies between tokens. The \textbf{Solver Module}
+provides the framework for flaw detection and resolution that is
+needed to implement search algorithms for planning. The \textbf{API
+  Layer} provides access to services from all the different modules in
+a programmatic way so that \eu can be easily embedded into specialized
 applications.
@@
 Describing problem domains and problem instances in terms of objects,
-variables and domain rules in a programmatic way is cumbersome to the
-extent of being impractical; this motivated the creation of modeling
-languages that support those definitions in a much more natural and
-compact way. \texttt{NDDL} (New Domain Description Language
-[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
-\eu and will be described in detail in the Modeling section
-below. Another, \texttt{ANML} is a language under development that
-provides a more intuitive syntax \cite{smith08}.
+variables and domain rules is cumbersome to the extent of being
+impractical; this motivated the creation of modeling languages that
+support those definitions in a much more natural and compact way. \nd
+(New Domain Description Language [pronounced ``noodle'']) \cite{NDDL}
+is the \kcomment{primary} modeling language for \eu described in
+Section \ref{sec:europa:modeling}. Another, \texttt{ANML}
+\cite{smith08} is a language under development that provides a more
+intuitive syntax \kcomment{and out of scope for this chapter}.
 
@@
 \item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
-  reason about temporal constraints
+  reason about temporal constraints.
 
 \item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
-  and Solver modules to provide reasoning and search for metric
-  resources.
+  and \textbf{Solver Module} to provide reasoning and search for
+  metric resources.
 
@@
   Problems that require challenging search may require specialized
-  solvers to be built as explained in Section \ref{sec:europa:search}.
+  solvers to be built as noted in Section \ref{sec:europa:search}.
 
 \end{enumerate}
+
+\kcomment{Before going into the details of each module, a simple
+  example illustrates how they interact at an abstract level:}
+
+\begin{verbatim}
+
+class LightBulb extends Timeline
+{
+    predicate On  {}
+    predicate Off {}
+}
+
+class LightSwitch extends Timeline
+{
+    LightBulb myBulb_;
+
+    LightSwitch(LightBulb b)
+    {
+        myBulb_ = b;
+    }
+
+    action turnOn  { duration=1; }
+    action turnOff { duration=1; }
+}
+
+LightSwitch::turnOn
+{
+    met_by(condition object.myBulb_.Off);  // Bulb must be Off to be turned On
+    meets(effect object.myBulb_.On);   // Must be turned on through the switch
+}
+
+LightSwitch::turnOff
+{
+    met_by(condition object.myBulb_.On);   // Bulb must be On to be turned Off
+    meets(effect object.myBulb_.Off);      // Must be turned off through the switch
+}
+
+// Problem instance : turning the light off
+LightBulb bulb1 = new LightBulb();
+LightSwitch switch1 = new LightSwitch(bulb1);
+
+// At time 0, the bulb is on
+fact(bulb1.On initialCondition);
+eq(initialCondition.start,0);
+
+// We want the bulb to be off by time 10
+goal(bulb1.Off goal1);
+lt(0,goal1.start);
+lt(goal1.start,10);
+
+\end{verbatim}
+
+\kcomment{The example above is a \nd (covered in more detail in
+  Section \ref{sec:europa:modeling}) model of a light bulb and a
+  switch that controls it along with a problem instance that requires
+  the light bulb to be \texttt{off} by time$=10$, given an initial
+  state with the bulb \texttt{on} at time$=0$. The \nd module allows
+  \eu to ingest this model and the problem description. Once the \nd
+  statements are parsed, different object types, object instances and
+  constraints are instantiated inside the \texttt{PDB} and
+  \texttt{CRE} modules where they are available for the user and the
+  solver modules to query and manipulate programmatically. This
+  instantiation of the initial state constitutes the initial plan that
+  \eu holds for this particular problem instance.
+
+  The user can then invoke functionality available in the solver
+  module to modify or extend this initial plan in a way that
+  eliminates any flaws (violated constraints, unsatisfied goals,
+  oversubscribed resources). When actions are added to a plan, the
+  \texttt{RE} ensures that domain rules expressed in the model are
+  enforced when \eu creates new action or predicate instances. For
+  instance, if a \texttt{LightSwitch::turnOn} action is added to the
+  plan, the \texttt{RE} ensures that requirements for the existence of
+  the conditions and effects of that action (that is, the
+  corresponding \texttt{LightBulb} must be off before and on after the
+  action) are posted through the \texttt{PDB} and \texttt{CRE}
+  modules.  }
 
@@
 language for models, goals and plans is of great importance. For \eue,
-\texttt{NDDL} allows the user to specify the different components of a
-problem domain in a precise and concise manner. Among the main
-features that make \texttt{NDDL} a powerful tool for describing
-problem domains are that it is object oriented. \texttt{NDDL} supports
-classes and inheritance in similar fashion to popular Object-Oriented
-languages like Java and C++. It also supports polymorphism for some of
-the planning components. Using object classes and instances is a
-time-tested approach to naturally describe problem domains
-\cite{rumbaugh}. It offers declarative constructs to define
-constraints and causality dependencies between actions and effects and
-finally it offers procedural constructs to populate \eus \texttt{PDB}
-with specific problem instances expressed in terms of objects,
-variables, constraints, facts and goals. To articulate the most important
-elements of \texttt{NDDL}, we show how the Shopping Agent problem
-could be modeled:
+\nd allows the user to specify the different components of a problem
+domain in a precise and concise manner. Among the main features that
+make \nd a powerful tool for describing problem domains are that it is
+object oriented. It supports classes and inheritance in similar
+fashion to popular object-oriented languages like Java and C++. It
+also supports polymorphism for some planning components. Using object
+classes and instances is a time-tested approach to naturally describe
+problem domains \cite{rumbaugh}. It offers declarative constructs to
+define constraints and causality dependencies between actions and
+effects and it offers procedural constructs to populate \eus
+\texttt{PDB} with specific problem instances expressed in terms of
+objects, variables, constraints, facts and goals.
+
+To articulate the most important elements of \nde, we show how the
+Shopping Agent problem could be modeled:
 
@@
   \textit{AgentLocation} class to keep track of where the agent is at
-  any point in time. \texttt{NDDL} supports a single rooted class
-  hierarchy so that state and behavior can be extended through
-  inheritance. We take advantage of that feature by making
-  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
-  class which doesn't allow predicates to overlap in time for the same
-  object instance. This feature is used for instance, to ensure that
-  an agent can only be at one place at a time in our example.
+  any point in time. \nd supports a single rooted class hierarchy so
+  that state and behavior can be extended through inheritance. We take
+  advantage of that feature by making \textit{AgentLocation} inherit
+  from \eus built-in \textit{Timeline} class which does not allow
+  predicates to overlap in time for the same object instance. This
+  feature is used for instance, to ensure that an agent can only be at
+  one place at a time in our example.
 
@@
 
-Domain rules can be defined in \texttt{NDDL} in the context of a
-predicate or an action. In the example above we can see the main
-elements of a rule definition:
+Domain rules can be defined in \nd in the context of a predicate or an
+action. In the example above, the main elements of a rule definition
+\kcomment{are as follows}:
 
@@
 An \eu user can introduce new constraint types by implementing
-specialized constraint and propagator classes; however, \eu provides a
+specialized constraint and propagator classes. However \eu provides a
 built-in constraint library with a range of arithmetic, temporal, set
 and resource constraint types so that a large number of problem
-domains can be modeled without having to introduce new constraint
-types. Tables \ref{tab:calconst}, \ref{tab:setconst},
+domains can be modeled. Tables \ref{tab:calconst}, \ref{tab:setconst},
 \ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
-types included in \eus Constraint Library.
+types included in \eus Constraint Library.  Temporal and resource
+constraints are especially important to model real-world planning
+domains.
 
-\input{constraints-table}
-
-Temporal and resource constraints are especially important to model
-real-world planning domains.
-
-\eu uses variables to explicitly represent timepoints for plan
-actions and states. Constraints among timepoints provide a natural
-way to express domain axioms. For example, in order to state that
-activity $A$ must occur before activity $B$ we can say that the end
-timepoint of $A$ is $leq$ the start timepoint of $B$.
-\cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network
-(\texttt{STN}). Such a network can be transformed into a Distance
-Graph (\texttt{DG}) where the outward arc from a node represents the
-maximum distance from the source node to the target node.
-Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
-a single constraint along with the resulting \texttt{DG}.
-\cite{dechter91} also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative cycle:
+\eu uses variables to explicitly represent timepoints for plan actions
+and states. Constraints among timepoints provide a natural way to
+express domain axioms. For example, in order to state that activity
+$A$ must occur before activity $B$ we can say that the end timepoint
+of $A$ is $leq$ the start timepoint of $B$.  \cite{dechter91} proposed
+that constraints among timepoints can be grouped together to form a
+Simple Temporal Network (\texttt{STN}). Such a network can be
+transformed into a Distance Graph (\texttt{DG}) where the outward arc
+from a node represents the maximum distance from the source node to
+the target node.  \kcomment{They} also show that shortest path
+algorithms \kcomment{(see \cite{corman} for a review)} could be used
+to propagate values in the network and discover a negative cycle:
 namely a path from a node to itself that has a path length less than
-$0$. If such a cycle exists, the network is inconsistent. It was
-further shown that a single-source shortest path algorithm was
-sufficient to detect a negative cycle and provide sufficient
-propagation to yield a backtrack-free search. \eus propagator for
-temporal constraints is a straight implementation of this algorithm.
+$0$. If such a cycle exists, the network is \kcomment{provably}
+inconsistent. It was further shown that a single-source shortest path
+algorithm was sufficient to detect a negative cycle and provide
+sufficient propagation to yield a backtrack-free search. \eus
+propagator for temporal constraints is an implementation of such a
+single-source algorithm. Fig. \ref{fig:stn} illustrates an
+\texttt{STN} with $2$ variables and a single constraint along with the
+resulting \texttt{DG}.
 
@@
   \centering
-  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
-  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
-  \caption{\small An illustration of a STN and its corresponding
-    distance graph.}
+  \subfloat[\small A Simple Temporal Network (\texttt{STN}). \kcomment{Nodes correspond
+  to variables and their domains and arc's the \emph{constraint} or
+  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
+  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
+  \caption{\small An illustration of a \texttt{STN} and its
+    corresponding distance graph.}
   \label{fig:stn}
@@
 violations when a resource is completely depleted, plan flaws can be
-generated whenever a resource level is outside a specific upper or
-lower bound, or when instantaneous production or consumption exceeds a
-specific threshold.
+generated whenever a resource level is outside a specific bound, or
+when instantaneous production or consumption exceeds a specific
+threshold.
 
-Since \eu supports a flexible temporal representation, where time is
-represented by intervals, instead of single values, determining
-whether a resource is the source of a plan flaw or a constraint
-violation in general is substantially more difficult than just
-computing a profile of the resource and checking against some limits.
-With temporal flexibility, a number of resource profiles could be
-possible depending on precisely when each production/consumption event
-occurs out of a range of possible values that are in the domain of the
-variables that represent the temporal extent for the event.
-\cite{Muscettola04,Muscettola06} proposed an algorithm where the
+Determining whether a resource is the source of a plan flaw or a
+constraint violation in general is substantially more difficult in
+\eus flexible representation, than just computing a profile of the
+resource and checking against some limits.  With temporal flexibility,
+a number of resource profiles could be applicable depending on
+precisely when each production/consumption event occurs out of a range
+of possible values that are in the domain of the variables that
+represent the temporal extent for the event.
+\cite{Muscettola04,Muscettola06} propose an algorithm where the
 variables that represent the time for production/consumption events
@@
 temporal propagator; constraints on the resource can then be seen as
-quantities flowing through that network. Computing the maximum flow
-then, yields the maximum resource usage and thus a lower bound on the
-resource level; computing a minimum flow will similarly yield an upper
-bound. Since maximum flow on a temporal network can be computed in
-polynomial time, \eu takes advantage of this algorithm to efficiently
-compute upper and lower bound envelopes on the resource
-level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
-computation.
+quantities \emph{flowing} through that network. Computing the maximum
+flow then, yields the maximum resource usage and thus a lower bound on
+the resource level; computing a minimum flow will similarly yield an
+upper bound. They show that such upper and lower bound envelopes on
+the resource level can be computed in polynomial time
+\cite{corman}. Resource envelopes can be useful when trying to
+evaluate plan brittleness in the face of changing activity durations
+or start times. They can also be useful when searching for plans;
+however they are relatively expensive to compute. Therefore their
+ability to prune the search space must be weighted against the ability
+to search faster \cite{6007772, morris11}. Fig. \ref{fig:resenvelopes}
+shows an example of such a resource computation.
 
@@
 \includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
-\caption{\small An example of resource envelopes computed by \eu.}
+\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eue.}
 \label{fig:resenvelopes}
 \end{figure}
-
-Resource envelopes can be useful when trying to evaluate plan
-brittleness in the face of changing activity durations or start
-times. They can also be useful when searching for plans; however they
-are relatively expensive to compute. Therefore their ability to prune
-the search space must be weighted against the ability to search faster
-\cite{6007772, morris11}.
 

### 2012-05-03 Review-response.tex
updates

\documentclass[11pt,twoside]{article}
% \usepackage{subfigure}
% \usepackage{latexsym}
% \usepackage{graphicx}

% \usepackage{floatflt}
% \usepackage{float}

\input{preamble}

\renewcommand{\baselinestretch}{0.973}

\begin{document}
\thispagestyle{empty}
\pagestyle{empty}
\setlength{\parindent}{0pt} 
\setlength{\parskip}{2ex}

Kanna Rajan\\
Principal Researcher for Autonomy\\
Monterey Bay Aquarium Research Institute (MBARI)\\
Moss Landing, California\\
http://www.mbari.org/staff/kanna/\\
http://www.mbari.org/autonomy/index.htm \\
(831) 775-1833 (Work)\\

Reviewer and Editor for the manuscript \textsf{\underline{Towards Deliberative Control in Marine Robotics}}\\

Dear Mae,

First and foremost, we would like to thank the reviewer of our
manuscript above for the extensive and valuable comments provided. And
we thank you for your efforts to pull this all together.

We are heartened by the reviewers comments that the manuscript
``follows the intent and spirit'' of the book's theme. And we are
delighted in her/his belief that the manuscript should be accepted for
publication.

We have \emph{carefully} gone over all the comments provided. The
attached revised manuscript reflects those questions/comments from the
review and are highlighted in \kcomment{red} . In a number of
instances, we have taken the liberty to tighten the text for the
English and in one instance added some suplemental material which is
substantial, to explain how the \eu planner components work. We have
also added a couple of figures to better illustrate the concepts per
the reviewer's suggestions. The changes in sum are \underline{not}
substantial given the feedback. And as you would likely see, we've
worked hard to address all the comments while keeping the material
accessible.

Thank you once again for all your efforts on our collective behalf for
this book. We look forward to working with you to bring the book out
for educators at the earliest.

\vspace{+2cm}
\begin{tabbing}
\hspace{+10cm}\=Kanna Rajan\\
\>Corresponding Author\\
\end{tabbing}

\end{document}
\end{input}

### 2012-05-03 europa-basics.tex
updates

@@
 
+Planning as an activity is ingrained in human decision-making
+\cite{berthoz} and its generalization into a computational framework
+demonstrating machine intelligence \kcomment{has for long been}
+targeted by pioneers in computer science
+\cite{computersthought}. Early work in this context focused on logical
+theorem proving by computer \cite{green69} which has formed the basis
+of much of the discipline. The emphasis in these early efforts was to
+logically derive the existence proof of a trajectory of actions for a
+hypothetical robot (moving blocks, chess pieces, furniture etc) to
+achieve a specific goal given specific initial conditions. Two
+important properties of \emph{Soundness}\footnote{A planning algorithm
+  is sound if invoked on a problem $P$ returns a plan which is a
+  solution for $P$.} and \emph{Completeness}\footnote{A planning
+  algorithm is complete if invoked on a solvable problem $P$ is
+  guaranteed to return a solution.} along with derived results for
+algorithmic complexity \cite{gareyjohnson,corman} were a defining
+cornerstone for understanding how different computational paradigms
+could actually be tractable first in simulation and subsequently in
+actual robotic platforms. In this chapter our emphasis is to focus on
+embedded AI Planning and the execution of plans on an embedded
+\kcomment{marine robotic} platform. The techniques we highlight are
+general purpose as noted in Section \ref{sec:intro}.
+
 \begin{figure}[b]
 \centering
-  \includegraphics[scale=0.35]{figs/shopping_europa.pdf}
+  \includegraphics[scale=0.45]{figs/shopping_europa.pdf}
   \caption{\small A potential plan for a robot shopping for groceries.}
@@
 \includegraphics[scale=0.3]{figs/Example-NQueens1.jpg}
-\caption{\small An $N$-Queens solution generated by \eu}
+\caption{\small An $N$-Queens solution generated by \eue.}
 \label{fig:nqueens-2}
@@
 
-\paragraph{} \textit{Scheduling}: In the Resource Constrained Project
-Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, a project
-consisting of a set of activities must be scheduled in a way that
-satisfies minimum and/or maximum temporal separation constraints. The
-activity schedule must also respect fixed limits on the availability
-of resources required to perform each activity. In addition to
-satisfying temporal and resource constraints, it is common for the
-user to want to minimize makespan \cite{ghallab04} so that the entire
-project is finished as early as possible. Fig. \ref{fig:rcpsp-1} shows
-an example of a a solution provided by \eu for an \texttt{RCPSP}
-instance with 10 activities, 5 resources and 30 temporal constraints.
+\paragraph{} \textit{Scheduling}: The Resource Constrained Project
+Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, consists of a
+set of activities that must be scheduled in a way that satisfies
+minimum and/or maximum temporal separation constraints. The activity
+schedule must also respect fixed limits on the availability of
+resources required to perform each activity. In addition to satisfying
+temporal and resource constraints, it is common for the user to want
+to minimize makespan \cite{ghallab04} so that the entire project is
+finished as early as possible. Fig. \ref{fig:rcpsp-1} shows an example
+of a a solution provided by \eu for an \texttt{RCPSP} instance with
+$10$ activities, $5$ resources and $30$ temporal constraints.
 
-\begin{figure}
+\begin{figure}[t]
 \centering
-\includegraphics[scale=0.25]{figs/Example-UBO0.jpg}
-\caption{\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
-  problem.}
-\label{fig:rcpsp-1}
+\subfloat[\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
+  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=0.25]{figs/Example-UBO0.jpg}}\qquad
+\subfloat[\small A \eu solution to a shopping agent problem domain where
+  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=0.3]{figs/Example-Shopping0.jpg}}
+\caption{\small \eu based solutions to scheduling and planning problems.}
 \end{figure}
@@
 \cite{russelnorvig} is a variation of the problem with which we
-started this chapter.  In this problem, an agent needs to purchase a
-set of products (milk, drill, etc) that are available at specific
-locations (supermarket, hardware store, etc). The agent is subject to
-temporal (must complete tasks by specific deadlines) and resource
-(fuel, carrying capacity, etc) constraints and it needs to synthesize
-actions that need to be performed to find and acquire the required
-items, as well as when to perform each of those actions.
+started this chapter.  An agent needs to purchase a set of products
+(milk, drill, etc) that are available at specific locations
+(supermarket, hardware store, etc). The agent is subject to temporal
+(must complete tasks by specific deadlines) and resource (fuel,
+carrying capacity, etc) constraints and it needs to synthesize actions
+that need to be performed to find and acquire the required items, as
+well as when to perform each of those actions.
 Fig. \ref{fig:shopping-1} shows a solution produced by \eu for such a
 problem instance.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figs/Example-Shopping0.jpg}
-\caption{\small A \eu solution to a shopping agent problem domain where
-  the agent needs to buy Bananas, Milk and a Drill.}
-\label{fig:shopping-1}
-\end{figure}
 
@@
 
-To generate plans, the \texttt{CAIP} framework uses a plan-space search
-approach \cite{ghallab04}, where a partial plan is evolved until all
-open goals are achieved, and all inconsistencies are removed. The
-specific search algorithm used by \eu will be covered in detail below.
+To generate plans, the \texttt{CAIP} framework uses a plan-space
+search approach \cite{ghallab04}, where a \emph{partial
+  plan}\footnote{\kcomment{A partial plan is a subsequence of a plan
+    which can be refined into a plan structure.}} is evolved until all
+open goals are achieved and all inconsistencies are removed. The
+specific search algorithm used by \eu will be covered in
+\kcomment{further} detail below.
 

### 2012-05-04 Review-response.tex
updates

@@
 
-Kanna Rajan\\
-Principal Researcher for Autonomy\\
-Monterey Bay Aquarium Research Institute (MBARI)\\
-Moss Landing, California\\
-http://www.mbari.org/staff/kanna/\\
-http://www.mbari.org/autonomy/index.htm \\
-(831) 775-1833 (Work)\\
+\small{Kanna Rajan}\\
+\small{Principal Researcher for Autonomy}\\
+\small{Monterey Bay Aquarium Research Institute}\\
+\small{Moss Landing, California}\\
+\small{http://www.mbari.org/staff/kanna/}\\
+\small{http://www.mbari.org/autonomy/index.htm}\\
+\small{(831) 775-1833}\\
 
@@
 the reviewer's suggestions. The changes in sum are \underline{not}
-substantial given the feedback. And as you would likely see, we've
-worked hard to address all the comments while keeping the material
+substantial given the feedback. However the page count has, with the
+changes, gone up a little. And as you would likely see, we've worked
+hard to address all the comments while keeping the material
 accessible.
@@
 \begin{tabbing}
-\hspace{+10cm}\=Kanna Rajan\\
-\>Corresponding Author\\
+\hspace{+10cm}\=\small{Kanna Rajan}\\
+\>\small{Corresponding Author}\\
 \end{tabbing}

### 2012-05-04 arch.tex
updates

@@
 
-Systematic partitioned inference is a major contribution of
-\rxe. However that is not the focus of this chapter with details
-covered in \cite{py10, rajan12}. Coupled interaction between planning
-and the world on the other hand, has a substantial impact on how
-planning can be integrated in such frameworks. In this chapter we show
-how using the \eu planner can be coupled to the \rx framework by first
-linking concepts in temporal planning with the framework. Note that in
-the following treatise, we call the formal semantic framework of \rx
-as the 'agent'. Reactors constitute an entity within this agent and
-are standalone blocks of inference. The scope, detail and number of
+Systematic partitioned inference is a major contribution of \rx
+\kcomment{which is covered in more detail} in \cite{py10,
+  rajan12}. Coupled interaction between planning and the world on the
+other hand, has a substantial impact on how planning can be integrated
+in such frameworks. In this chapter we show how using the \eu planner
+can be coupled to the \rx framework by first linking concepts in
+temporal planning with the framework. Note that in the following
+treatise, we call the formal semantic framework of \rx as the
+'agent'. Reactors constitute an entity within this agent and are
+standalone blocks of inference. The scope, detail and number of
 reactors are application specific and outside the scope of this
@@
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[scale=0.35]{figs/token-disptch.pdf}
-  \caption{\small \kcomment{Dispatching a token occurs only when the request
-    start time overlaps a window. \texttt{Goal$_1$} therefore gets
-    dispatched from \texttt{reactor$_1$} to \texttt{reactor$_2$} but
-    not \texttt{Goal$_2$} to avoid early commitment.}}
-  \label{fig:reactor-lookahead}
-\end{figure}
-
 This window is used by \rx to identify when to dispatch a request to a
@@
   external} state variable of a reactor while avoiding flooding the
-owner reactor of this state variable and avoiding early commitment.
-\kcomment{Fig. \ref{fig:reactor-lookahead} illustrates an example
+owner reactor of this state variable \kcomment{and avoiding early
+  commitment.} The choice of both the latency and look-ahead of a
+reactor can be delicate as one is dependent on the other. Latency is
+dependent on exogenous paramaters such as the \kcomment{computer power
+  of the host} and the number of reactors needing to deliberate at
+once, among others. As the design stands, these variables should be
+evaluated carefully by the system designer balancing agent
+efficacy. Fig. \ref{fig:reactor-lookahead} illustrates an example
 between two reactors where one token is dispatched when overlapping
 the dispatch window while another is not.}
-
-The choice of both the latency and look-ahead of a reactor can be
-delicate as one is dependent on the other. Needless to say, latency is
-highly dependent on exogenous paramaters such as the host computers
-computing power and the number of reactors needing to deliberate at
-once, among others. As the design stands, these variables should be
-evaluated carefully by the system designer balancing agent efficacy.
 
@@
 agnostic where its {\em external} state information is managed as long
-as this information is available and properly maintained for managing
-{\em internal} state information. % This {\em internal} state in turn
-% may be used by other reactors.
-Such a formalism also constrains reactor interaction to state
-information exchange which comes in two forms as illustrated in
-Fig. \ref{fig:trex:flow}:
+as this information is available and maintained for managing {\em
+  internal} state information.  Such a formalism also constrains
+reactor interaction to state information exchange which comes in two
+forms as illustrated in Fig. \ref{fig:trex:flow}:
 
@@
 
-\item {\em Observation on current state values}: This information
-  propagates \emph{up} in the reactor dependency graph at the
+\item {\em Observation on current state values}: This propagates
+  \emph{up} in the reactor dependency graph \kcomment{during} the
   synchronization phase and occurs at the beginning of every
   tick. During this phase the agent provides updates to each reactor
-  on their {\em external} state variables for this tick so they can
-  compute their {\em internal} state variable values for this same
-  tick. By doing so we propagate a consistent view of that state of
-  the world throughout all the reactors as time
-  advances. \kcomment{Observations can therefore be both sensor
-    measurements as well as state information, passed up from other
-    reactors. The strict rules of timeline ownership enable a clear
-    policy for conflict resolution: observations dominate
-    expectations.}
+  on their {\em external} state variable(s) to \kcomment{enable
+    computation of} their {\em internal} state variable values for
+  this same tick. By doing so we propagate a consistent view of world
+  state throughout the reactors as time advances. \kcomment{Further,
+    observations can be both sensor measurements as well as state
+    information, passed up from other reactors. The strict rules of
+    timeline ownership enable a clear policy for conflict resolution:
+    observations dominate expectations.}
 
 \item {\em Goal request on future state}: This information propagates
-  \emph{down} in the reactor dependency graph. In order to satisfy own
-  {\em internal} objectives, a reactor may have a plan that relies on
-  future values of one of its {\em external} state. When such a plan
-  is identified the agent ensures that this part of the plan is
-  provided as a goal to the owner of the given {\em external} state
-  variable(s).
+  \emph{down} in the reactor dependency graph. In order to satisfy its
+  own {\em internal} objectives, a reactor may have a plan that relies
+  on future values of one \kcomment{(or more)} of its {\em external}
+  state. When such a plan is identified the agent ensures that this
+  part of the plan is provided as a goal to the owner of the given
+  {\em external} state variable(s).
 
@@
     \caption{\small \kcomment{Detailed and abstracted views of
-      synchronization and deliberation within one tick cycle.}}
+        synchronization and deliberation within one and multiple tick
+        cycles.}}
   \label{fig:sync-plan-cycle}
@@
 
-Within \rxe, an instance of an \eu reactor comes with the sole purpose
-to deliberate. A typical design of such a reactor takes its cues from
-\texttt{IDEA} \cite{mus02, mus06} where both planning and execution
-are based on a single unified model within a rich representation that
-provides support for temporal inference. In \rxe, the execution cycle
-consists of two deliberation processes which are implemented as two
-separate \eu solvers modifying the \emph{same} plan database.
+Within \rxe, \kcomment{the sole purpose of an instance of an \eu
+  reactor is to deliberate}. A typical reactor design takes its cues
+from \texttt{IDEA} \cite{mus02, mus06} where both planning and
+execution are based on a single unified model within a rich
+representation that provides support for temporal inference. In \rxe,
+the execution cycle consists of two deliberation processes which are
+implemented as two separate \eu solvers modifying the \emph{same} plan
+database.
 
@@
   evolution of \emph{external} timelines and ensures that the reactor
-  propagates these to identify current reactor state as well as to
-  inform other reactors of any state change on its \emph{internal}
-  timelines. This solver is summoned at the beginning of every single
-  tick.
-
-\item The \emph{deliberation solver} manages the deliberation process
-  of the reactor. It does so either to produce a new plan, to alter
-  its current plan as new goals are provided or when the
-  synchronization solver identifies a conflict between current state
-  and expectations from a previously generated plan. This process can
-  span multiple ticks and can therefore be interrupted by the
-  synchronization solver.
+  propagates \kcomment{changes} to identify current state as well as
+  to inform other reactors of any state change on its \emph{internal}
+  timelines. This solver is \kcomment{invoked} at the beginning of
+  every single tick.
+
+\item The \emph{deliberation solver} is used to either produce a new
+  plan, to alter its current plan as new goals are provided or when
+  the synchronization solver identifies a conflict between current
+  state and expectations from a previously generated plan. This
+  process can span multiple ticks and can therefore be interrupted by
+  the synchronization solver.
 
@@
 stops, its starting point is the last partial plan resulting from
-deliberation. This gives an initial frame to the problem that helps to
-target the synchronization problem by ensuring that the new {\em
-  external} state updates are compatible with this plan. If so,
+deliberation \kcomment{(see Fig. \ref{fig:sync-plan-cycle})}. This
+helps to target the synchronization problem by ensuring that the new
+{\em external} state updates are compatible with this plan. If so,
 synchronization propagates these events in the plan and can use it as
@@
 This assumption, while helpful to reduce the scope and complexity of
-the planning problem, becomes problematic for planning within a highly
+the planning problem, becomes problematic for planning within a
 dynamic environment. In a typical robot this implies that the only
@@
 planner an initial state and current objectives. During the planning,
-it is assumed that the agent can maintain the world as perceived by
-the planner. \cite{lemai04, lemai-chenevier2004} attempts to reduce
-the impact of planning by allowing local plan repair in the existing
-plan. In the eventuality that such repair is not feasible they require
-that their robot comes to a stop until (re-)planning is complete.
+it is assumed that the agent can \emph{maintain the world as perceived
+  by the planner}. \cite{lemai04, lemai-chenevier2004} attempts to
+reduce the impact of planning by allowing local plan repair in the
+existing plan. In the eventuality that such repair is not feasible
+they require that their robot comes to a stop until (re-)planning is
+complete.
 
@@
 \cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
--- similar to our {\em synchronization} solver -- was often the solver
-targeting the change at the execution frontier. 
-% Although, none of the papers we found were detailing how this was
-% handled in the architecture.
-\cite{Dias:2003ua} states:
+-- similar to our {\em synchronization} solver -- is the solver
+targeting change at the execution frontier.  \cite{Dias:2003ua}
+states:
 
@@
 This suggests a deliberate non overlap between \kcomment{the} {\em
-  reactive planner} and {\em deliberative planners} horizons. In \rxe,
-and more specifically in our \eu reactor the planning horizon of {\em
-  synchronization} -- limited to the current tick -- is a subset of
-the planning horizon of the deliberation steps -- starting at the
-current tick and with an extent based on latency $\lambda$ and
-look-ahead $\pi$ of the reactor. In this manner we avoid creating a
-conflict between the deliberation and synchronization by controlling
-the scheduling of these two process\kcomment{es} by the \rx agent.
-
-In the following sections we show how synchronization and deliberation
-are implemented independently to show how their interaction through
-the plan helps relax this assumption. It does so by integrating
-planning in the reactor execution cycle even while allowing for more
-informed deliberation.
+  reactive planner} and {\em deliberative planner}'s horizons. In
+\rxe, and more specifically in our \kcomment{deliberative} reactor,
+the planning horizon of {\em synchronization} -- limited to the
+current tick -- is a subset of the planning horizon of the
+deliberation steps -- starting at the current tick and with an extent
+based on latency $\lambda$ and look-ahead $\pi$ of the reactor. In
+this manner we avoid creating a conflict between the deliberation and
+synchronization by controlling the scheduling of these two
+process\kcomment{es} by the \rx agent.
+
+The following sections detail how synchronization and deliberation are
+implemented independently to show how their interaction through the
+plan helps relax this assumption. It does so by integrating planning
+in the reactor execution cycle even while allowing for more informed
+deliberation.
 
@@
 {\em internal} timelines, \rx makes sure that {\em external}
-observations are taken into account for every tick as shown in
+observations are taken into account for every tick as in
 \cite{py10}. Such a requirement is enforced to ensure that all
@@
 challenging part of synchronization is to decide the {\em internal}
-state values for the current tick. In the Shopping example for example,
-we have the following code snippet which helps identify when we can
-\texttt{Buy} a product:
+state values for the current tick. In the Shopping example from
+\kcomment{Section \ref{sec:europa:modeling}} for example, we have the
+following code snippet which helps identify when we can \texttt{Buy} a
+product:
 
@@
 Agent::Buy {
-  eq(duration, 10);
+  eq(10, duration);
   // [...]
@@
 
-If \texttt{object.location\kcomment{.At}} is {\em external} and \texttt{Agent} is an
-{\em internal} timeline, a robot may encounter the situation where it
-is at the location where it can buy one (possibly among others)
-product. This results in a decision-making problem related to the
-state value of the \texttt{Agent} timeline; the robot can either {\em
-  Buy} a specific product now, wait a little longer before doing so or
-just simply not buy any product at this location.
-
-In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
-new type of flaw which requires identification of all values of {\em
-  internal} state variables for the current tick. This flaw can be
-defined as:
+If \texttt{object.location\kcomment{.At}} is {\em external} and
+\texttt{Agent} is an {\em internal} timeline, a robot may encounter
+the situation where it is at the location where it can buy one
+(possibly among others) product. This results in a decision-making
+problem related to the state value of the \texttt{Agent} timeline; the
+robot can either {\em Buy} a specific product now, wait a little
+longer before doing so or just simply not buy any product at this
+location.
+
+In the \kcomment{deliberative} reactor this decision-making
+requirement is a new type of flaw which requires identification of all
+values of {\em internal} state variables for the current tick. This
+flaw can be defined as:
 
@@
     $\mathcal{M}$ -- resolve its {\em Internal} state that will then
-    be provided by the agent to other reactors using these state
-    variables.}
+    be provided \kcomment{in turn} by the agent to other reactors.}
   \label{fig:synch:flaw}
@@
 its current plan is properly executed. Finally if none of these two
-choices provides a valid solution -- \ie lead to an inconsistant state
--- the reactor tries all the possible state values remaining for the
-timeline until it finds one that brings plan consistency.
-
-% allows it to identify its state as required by \rx architecture.
-
-% The flaw forces the reactor to identify its internal state for the
-% current tick. By using this flaw one can describe the synchronization
-% process as:
+choices provides a valid solution -- \ie \kcomment{it leads} to an
+inconsistant state -- the reactor tries all the possible state values
+remaining for the timeline until it finds one that \kcomment{results
+  in} plan consistency.
 
@@
   external timeline into the plan database.
+
   \begin{itemize}
+
   \item New observations are integrated by creating the
     corresponding token as a fact in the plan database
+
   \item If no observations have been received at this tick the last
-    observation token restrict its \texttt{end} domaine to the
+    observation token restricts its \texttt{end} domain to the
     interval $[\tau+1, +\infty)$
+
   \end{itemize}
 
-\item Propagate this information in the plan database using model
-  $\mathcal{M}$ of this reactor.
-
-\item Resolve the current state value of each internal timelines based
-  on Proposition \ref{prop:csf:resolve}.
+\item \label{itm2} Propagate this information in the plan database
+  using reactor model $\mathcal{M}$.
+
+\item \label{itm3} Resolve the current state value of each internal
+  timeline based on Proposition \ref{prop:csf:resolve}.
 
@@
 
-Steps $2$ \& $3$ above are handled by the \eu solver from Algorithm
-\ref{alg:europa:solve} which ensures that the {\em current state
-  flaws} are selected last so as to follow the sequence given above.
+Steps \ref{itm2} \& \ref{itm3} above are handled by the \eu solver
+from Algorithm \ref{alg:europa:solve} which ensures that the {\em
+  current state flaws} are selected last so as to follow the sequence
+given above.
 
@@
 future inconsistency.  Such an assumption impacts the set of possible
-domains the reactor can support while remaining complete. Take for
-example this rule from our \texttt{Shopping} model:
+domains the reactor can support while remaining
+\kcomment{algorithmically} complete. Take for example this rule from
+our \texttt{Shopping} model:
 
@@
 \texttt{Go}. These cannot be retracted easily without a potential
-impact to the plan implying in terms of a cost in back-propagation.
+impact on computational cost.
 
@@
 One could therefore alter this part of the model to give the reactor
-more flexibility as follows:
+more flexibility by relaxing the constraint in line $5$ from a strict
+\texttt{equals} to a \texttt{contains} and to add the new constraint
+at line $11$ to ensure that the last \texttt{Going} ends up in the
+location the agent actually wants the robot to be at. This first level
+of relaxation will therefore allow the agent to use multiple
+\texttt{Going} tokens should the first token fail to reach the
+objective, as shown below:
 
@@
 
-We do so by relaxing the constraint in line $5$ from a strict
-\texttt{equals} to a \texttt{contains} and to add the new constraint
-at line $11$ which ensures that the last \texttt{Going} ends up in the
-location the agent actually wants the robot to be at. This first level
-of relaxation will therefore allow the agent to use multiple
-\texttt{Going} tokens should the first one fail to reach the
-objective.
-
 As the example above shows, synchronization involves a deliberation
@@
 
-One critical aspect to take into account is that from a \rx point of
-view, synchronization is the only critical task of a reactor. Such a
-process is used in order for the agent to identify its current state
-and maintain it consistently throughout the agent's lifetime. For a
+Synchronization is the only critical task of a reactor. Such a process
+is used in order for the agent to identify its current state and
+maintain it consistently throughout the agent's lifetime. For a
 reactor not being able to find a consistent solution during
 synchronization -- implying that after evaluating all possible
-solutions provided by a model it was unable to find a consistent plan
--- not only indicates that a reactor is unable to explain the state of
-the world, but more importantly it may jeopardize other reactors that
-rely on its {\em internal} state. For this reason a failure to
-synchronize immediately results in the \rx agent killing a reactor and
-notifying any reactor that depends on its {\em internal} state
-variables; it does so by posting the special observation
-\textsf{Failed}. This special state is a special predicate which is
-acceptable for all the {\em internal} and {\em external} timelines of
-\kcomment{an} \eu reactor.  By so doing we allow for graceful
-degradation where other reactors can readapt their plans on receipt of
-this observation.
-
-Finally, at the engine level careful choices were made in both the
-focus of the search -- limited only to tokens that can potentially
-overlap the current tick -- and heuristics where the {\em current
-  state flaw} is considered last during the search process. The later
-limits the amount of backtracking for a solution -- or at least how
-deep one needs to backtrack in a decision tree.  Similarly the
-sequence in which we evaluate the possible resolution for such a flaw
--- presented in Proposition \ref{prop:csf:resolve} -- was selected
-according to some simple assumptions, that any external observation
-that is compatible with an existing token in the current plan reflects
-the execution of the plan itself. In other words this means that our
+solutions provided by a model \kcomment{being} unable to find a
+consistent plan -- not only indicates that a reactor is unable to
+explain the state of the world, but more importantly it may jeopardize
+other reactors that rely on its {\em internal} state. For this reason
+a failure to synchronize immediately, results in the \rx agent killing
+a reactor and notifying any reactor that depends on its {\em internal}
+state variables; it does so by posting the special observation
+\textsf{Failed}. This special state is acceptable for all the {\em
+  internal} and {\em external} timelines of \kcomment{an} \eu reactor.
+By so doing we allow for graceful degradation where other reactors can
+readapt their plans on \kcomment{receiving} this observation.
+
+Finally, at the engine level careful choices \kcomment{are} made in
+both the focus of the search -- limited only to tokens that can
+potentially overlap the current tick -- and heuristics where the {\em
+  current state flaw} is considered last during the search
+process. The later limits the amount of backtracking for a solution --
+or at least how deep one needs to backtrack in a decision tree.
+Similarly the sequence in which we evaluate the possible resolution
+for such a flaw -- \kcomment{described} in Proposition
+\ref{prop:csf:resolve} -- \kcomment{is} selected according to
+\kcomment{the assumption}, that any external observation that is
+compatible with an existing token in the current plan reflects the
+execution of the plan itself. In other words this means that our
 solver always prefers to use tokens \emph{already existing} in the
@@
 \ref{alg:europa:solve} into atomic steps. This interruption occurs
-around the recursive call line \ref{li:recurse} while ensuring that
+around the recursive call, line \ref{li:recurse}, while ensuring that
 the current partial plan is not inconsistant at any single step. This
@@
 A critical aspect of this deliberation step is to ensure that at the
-exit of this call the $plan$ is not inconsistent. % Indeed, it is always
-% possible that the next call made by the agent is for synchronization
-% which could fail if the $plan$ is inconsistent before any attempt is
-% made for synchronization.
-If the partial plan is found to be inconsistent during a deliberation
-step we either backtrack in the decision stack until a node of the
-\kcomment{search} tree is found with an alternate solution or the plan
-is relaxed since there is no other alternative solution. We do this by
-removing all goals while keeping only recent observations.
+exit of this call the $plan$ is consistent.  If the partial plan is
+found to be inconsistent during a deliberation step we either
+backtrack in the decision stack until a node of the \kcomment{search}
+tree is found with an alternate solution or the plan is relaxed since
+there is no other alternative solution. We do this by removing all
+goals while keeping only recent observations.
 
@@
 subtle as its alteration of the plan at the execution frontier is more
-drastic. We can still give a non exhaustive list of possible
-alterations of the flaws directly related to {\em synchronization}:
+drastic. \kcomment{A possible list of alterations of the plan directly
+  related to {\em synchronization} could be}:
 
@@
 \item The {\em synchronization} solver may have created new flaws that
-  were outside of its scope but are now part of the scope of the {\em
-    deliberation} solver.
+  were outside of its scope but are now part of the {\em deliberation}
+  solver's scope.
 
@@
   plan to identify its current {\em external} state. (This case is
-  developed further \kcomment{in} the next section).
+  developed further \kcomment{in Section \ref{sec:arch:intertwine}}).
 
@@
 This filtering window complements the \rx {\em dispatch window} from
-Definition \ref{def:dispatch} which filters out at the agent level
+Definition \ref{def:dispatch} which filters out, at the agent level,
 those goals that are before the reactor execution latency or too far
@@
 allows the planner to focus on its planning horizon and report
-decision on flaws that are beyond this scope. As time advances, these
-potential future flaws will eventually be part of the planning window
-of the reactor and will be then resolved along with potentially new
-goals.  This in turn, results in an apparent \emph{continuous
-  planning} of the reactor. When there are no more flaws present in
-the current planning scope a partial plan solution is found. This has
-two effects:
+decisions on flaws that are beyond this scope \comment{::is this last
+  sentence correct?}. As time advances, these potential future flaws
+will eventually be part of the planning window of the reactor and will
+be then resolved along with potentially new goals.  This in turn,
+results in an apparent \emph{continuous planning}
+\kcomment{\cite{myers99}} of the reactor. When there are no more flaws
+present in the current planning scope a partial plan solution is
+found. This has two effects:
 
@@
 \item The token on the {\em external} timelines of this partial plan
-  can then be posted which eventually (at the end of the current tick
-  if these goals overlap the owner reactor's planning window) will be
-  dispatched as goals, in turn triggering further deliberation.
+  can then be posted. \kcomment{Eventually, at the end of the current
+    tick if this goal overlaps the owner reactor's planning window, it
+    will be dispatched, resulting in further deliberation.}
 
@@
 synchronization which drives planning; so the co-mingling between
-these two processes is critical to maintaining state within a robot
-especially when it is immersed in a dynamic environment. Further, it
-is important that \rx maintain one $plan$ structure shared between
-synchronization and \kcomment{deliberation}. This allows these two
-processes to not only be interleaved for agent responsiveness, but
+these two processes is critical to maintaining state within a robot.
+Further, it is important that \rx maintain one $plan$ structure shared
+between synchronization and \kcomment{deliberation}. This allows these
+two processes to not only be interleaved for agent responsiveness, but
 ensures a viable mechanism to communicate.
@@
 propagate {\em external} observations to identify {\em internal}
-state. This allowed us to focus on synchronization as a state
+state. This \kcomment{ensured} focus on synchronization as a state
 evaluation process that can be resolved within the \eu framework with
-few extensions. Similarly we presented deliberation as an abstraction
+few extensions. Similarly deliberation was presented as an abstraction
 without discussing the disruptive nature of synchronization and how it
@@
 We first analyze the case when synchronization occurs after planning
-has completed (\ie the last step of deliberation results in a plan
-that is considered complete for the current look-ahead). As
+has completed, \ie the last step of deliberation results in a plan
+that is considered complete for the current look-ahead. As
 synchronization starts with a plan, this deliberation will not only be
@@
 observations but will also include reactor intentions specified in the
-available partial plan.
-
-Consider for example the Shopping Agent model where the timeline 
+partial plan.
+
+Consider for example the Shopping Agent model \kcomment{from Section
+  \ref{sec:europa:modeling}}, where the timeline
 \texttt{Agent.position} is {\em external} while \texttt{Agent} is
-implemented a an {\em internal} timeline. As the reactor adds a goal to
-\texttt{Own} milk and the current {\em external} timeline indicates
+implemented as an {\em internal} timeline. As the reactor adds a goal
+to \texttt{Own} milk and the current {\em external} timeline indicates
 that the robot was \texttt{At(home)}, it produces a plan as shown in
@@
   observation was received by the reactor with the plan showed in
-    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
+    Fig. \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
     \caption{\small The Shopping example: Tokens in {\color{red}red}
@@
 agent location is able to change the location state variable to
-\texttt{Going}. \rx informs the reactor of this observation which is
-added to the partial plan as a fact -- starting exactly at $\tau'$ and
-lasting for an unknown duration which is at least one tick. As this
-new observation is integrated in the plan through synchronization, the
-solver can identify the similarity between this observation and the
-next token in the existing plan. Instead of initiating an exhaustive
-search of all possible implications, the synchronization process
-starts its execution with the assumption that these new observations
-are consistent with the currently maintained plan. As a consequence,
-at the end of synchronization the plan depicted in Fig.
-\ref{fig:shop:exec1} is likely. This is a result of propagation of
-information provided by the new observation in the existing plan while
-restricting the start time of \texttt{Going} and propagating temporal
-constraints, making \texttt{Go} to start {\em exactly} at $\tau'$. In
-doing so, the \eu solver's focus on the execution frontier ($\tau'$ in
-Fig. \ref{fig:shop:exec1}) takes advantage of the existing plan to
-evaluate a solution that assumes that the observation is consistent
-with what was decided by the reactor. \kcomment{In the event of
-  inconsistency, the (plan) expectations do not match the reality;
-  then partial plan failure within the reactor is initiated.} This is
-directed not only by heuristic choices made in the solver but also
-enforced by the heuristic order of the decision point resolution
-choices for the {\em current state flaw} \kcomment{as} shown in
-Proposition \ref{prop:csf:resolve}. As noted, the first two choices
-are biased towards a conservative view with respect to the existing
-plan.
+\texttt{Going}. \rx informs the reactor \kcomment{that owns the
+  \texttt{Agent} timeline,} of this observation which is added to
+\kcomment{its} partial plan as a fact -- starting exactly at $\tau'$
+and lasting for an unknown duration which is at least one tick. As
+this new observation is integrated in the plan through
+synchronization, the solver can identify the similarity between this
+observation and the next token in the existing plan. Instead of
+initiating an exhaustive search of all possible implications, the
+synchronization process starts its execution with the assumption that
+these new observations are consistent with the currently maintained
+plan. As a consequence, at the end of synchronization the plan
+depicted in Fig.  \ref{fig:shop:exec1} is likely. This is a result of
+propagation of information provided by the new observation in the
+existing plan while restricting the start time of \texttt{Going} and
+propagating temporal constraints, making token \texttt{Go} to start
+{\em exactly} at $\tau'$. In doing so, the \eu solver's focus on the
+execution frontier ($\tau'$ in Fig. \ref{fig:shop:exec1}) takes
+advantage of the existing plan to evaluate a solution that assumes
+that the observation is consistent with what was decided by the
+reactor. \kcomment{In the event of inconsistency, plan expectations do
+  not match reality; then partial plan failure within the reactor is
+  initiated.} This is directed not only by heuristic choices made in
+the solver but also enforced by the heuristic order of the decision
+point resolution choices for the {\em current state flaw}
+\kcomment{as} shown in Proposition \ref{prop:csf:resolve}. As noted,
+the first two choices are biased towards a conservative view with
+respect to the existing plan.
 
@@
 plan to advance plan execution. Both choices attempt to resolve state
-with the least amount of plan perturbation by constraining further the
+with the least amount of plan perturbation by further constraining the
 start and end time of existing tokens generated by previous
@@
 synchronization remains limited to restricting time points in the plan
-and propagates their impact. Most of the computational effort of
-creating new tokens in the plan and scheduling them onto timelines are
-managed during deliberation steps. Consequently, most of the time
-synchronization is limited to very few nodes in the \eu search space
-often requiring no backtracking. Fig. \ref{fig:shop:exec} shows how
-the existing plan can support the synchronization process. The
-observation \texttt{Going} occurring at $\tau'$ is fully compatible
-with the token already present in our plan. By updating the plan and
-propagating equality constraints, the resolution of the {\em current
-  state flaw} for the {\em Agent} timeline becomes obvious as a
-solution, as the token \texttt{Go} from the plan presents itself.
+and \kcomment{in turn} propagates their impact. Most of the
+computational effort of creating new tokens in the plan and scheduling
+them onto timelines are managed during deliberation
+steps. Consequently, most of the time synchronization is limited to
+very few nodes in the \eu search space often requiring no
+backtracking. Fig. \ref{fig:shop:exec} shows how the existing plan can
+support the synchronization process. The observation \texttt{Going}
+occurring at $\tau'$ is fully compatible with the token already
+present in our plan. By updating the plan and propagating equality
+constraints, the resolution of the {\em current state flaw} for the
+{\em Agent} timeline becomes obvious as a solution, as the token
+\texttt{Go} from the plan presents itself.
 
@@
 When invalid, identification occurs by synchronization failing to find
-a consistent solution. For example should the \texttt{Going}
+a consistent solution. For instance, should the \texttt{Going}
 observation received be to the \texttt{hardwareStore}, this would
@@
   \centering
-  \subfloat[\small Conflicting Observation]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
-  \subfloat[\small Relaxed Partial Plan]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \subfloat[\small Conflicting Observation.]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
+  \subfloat[\small Relaxed Partial Plan.]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
   \caption{\small Illustration of a conflict during synchronization
@@
 it by offering a new incomplete partial plan to be resolved in the
-next deliberation \texttt{steps}. % This on its own offer the core
-% functionality one should except to close the loop between planning and
-% execution in the sense that 
-Synchronization therefore, plays the role of what is typically termed
-as the {\em Executive} in other architectures \cite{gat98,
-  alami:1998p820, mus98, williams03, Nesnas:2003do}.
+next deliberation \texttt{step}s.  Synchronization therefore, plays
+the role of what is typically termed as the {\em Executive} in other
+architectures \cite{gat98, alami:1998p820, mus98, williams03,
+  Nesnas:2003do}.
 
@@
 step the \eu solver used for planning will have these new facts
-included in the partial plan which can introduce new flaws in the plan
-while potentially having resolved previously existing flaws. The
-deliberation process is therefore perturbed externally by {\em
-  synchronization} and consequently fully informed of the world state
-evolution as it is planning. This contrasts with other classical
-approaches, which avoid perturbing the planner during its search as it
-is not compatible with the ``off-line planning'' assumption, noted in
-Section \ref{sec:arch:europa}.
-
-By design choices made in the \rx framework -- both at the
-architectural level and the integration of the \eu framework for
-embedded deliberation -- we are able to implement a system that avoids
-this critical off-line planning limitation while enforcing a formal
+included in the partial plan which can introduce \kcomment{further}
+new flaws in the plan while potentially having resolved previously
+existing flaws. The deliberation process is therefore perturbed
+externally by {\em synchronization} and consequently fully informed of
+the \kcomment{evolving} world state as it is planning. This contrasts
+with other classical approaches, which avoid perturbing the planner
+during its search as it is not compatible with the ``off-line
+planning'' assumption, noted in Section \ref{sec:arch:europa}.
+
+Design choices made in the \rx framework -- both at the architectural
+level and the integration of the \eu framework for embedded
+deliberation -- allow the implementation of a system that avoids this
+critical off-line planning limitation while enforcing a formal
 framework that allows for deliberation and model-based agent control.
@@
 be more responsive to environmental change allowing informed
-decision-making within a tighter reaction loop and despite reactor
+decision-making within a tighter reaction loop despite reactor
 latency.

### 2012-05-04 trunk/future.tex
update

--- 2012-02-21 trunk/future.tex
@@
 algorithms and methods. In addition to augmenting traditional AUV
-surveys more advanced Lagrangian observations, mixed-initiative
-methods, our colleagues at MBARI have understood the longer term
-implications of our work. Our current and consequently future efforts
-are therefore grounded in important science problems in an operational
-oceanographic setting.
+surveys, more advanced Lagrangian observations and mixed-initiative
+methods, we believe our colleagues at MBARI have understood the longer
+term implications of our work. Our current and consequently future
+efforts are therefore grounded in important science problems in an
+operational oceanographic setting.
 
@@
 large part given the spatial and temporal scales of the
-bio-geochemical processes we are interested in in this project, it is
+bio-geochemical processes we are interested in, in this project, it is
 imperative that a key direction is in formulating a multi-robot
-adaptive sampling problem. In this context while superficially \rx can
-be considered as a multi-agent system and could therefore imply a
-direct extension, sporadic communication with high-latencies in
-surface communication, not to mention low throughput acoustic
-communications imply that the extension of \rx for distributed
-applications is non-trivial. Further, in order to efficiently execute
-coordinated observation of ocean processes with multiple autonomous
-sensing platforms, significant methodological as well as technical
-challenges must be addressed. We are working towards a systematic
-approach of using \rx as a backend planner for \can field
-experiments. The expectation is that \od users will plan initially
-individual vehicle surveys which are variants of standard
-straight-line transects. Lessons learned with human planners in the
-loop will be used towards planning larger surveys, also with simple
-fixed transects but using vehicle capabilities in ontologies similar
-to \cite{patron09}. Apportioning the planning task between shore-side
-components with substantial contextual information and compute power
-with robots with in-situ reasoning capabilities a la \rx is a
-challenge and an open research problem. Fig. \ref{fig:topics} shows
-what we believe is the scope and influence of the different topic
-areas we are focusing on.
+adaptive sampling problem. In this context while \rx can be considered
+as a 'multi-agent' system and \kcomment{could therefore imply its use
+  in a multi-vehicle setting as} a direct extension, sporadic
+communication with high-latencies in surface communication, not to
+mention low throughput acoustic communications imply that the
+extension of \rx for distributed applications is non-trivial. Further,
+in order to efficiently execute coordinated observations of ocean
+processes with multiple autonomous sensing platforms, significant
+methodological as well as technical challenges must be
+addressed.
+
+\kcomment{In this context} we are working towards a systematic
+approach of using \rx as a backend planner \kcomment{within the \od}
+for \can field experiments. The expectation is that \od users will
+\kcomment{initially be able to} plan individual vehicle surveys which
+\kcomment{would be} variants of straight-line transects. Lessons
+learned with human planners in the loop will be used towards planning
+larger surveys, also with simple fixed transects but using vehicle
+capabilities in ontologies similar to \cite{patron09}. \kcomment{Doing
+  so will likely enable the right vehicle to target an ocean feature
+  with appropriate sensors and sampling strategy}. Apportioning the
+planning task between shore-side components with substantial
+contextual information and compute power, with robots having in-situ
+reasoning capabilities similar to \rxe, is \kcomment{challenging in of
+  itself} and an open research problem. \kcomment{In pursuing this
+  line of research, we expect to touch on a range of associated
+  problems for decision-theoretic control of vehicles in the coastal
+  ocean}. Fig. \ref{fig:topics} shows what we believe is the scope and
+influence of the different topic areas we are \kcomment{currently}
+focusing on.
 
@@
 to graceful system degradation, making the overall approach effective
-towards increased robustness. There are two outstanding diagnosis and
-failure recovery challenges that we need to address; the first has to
-do with methodology related to model design of the instrument payload
-for a highly modular and configurable vehicle; this is local to a
-reactor. Doing so would enable (as a first step) using systematic
-Fault Diagnosis, Isolation and Recovery (FDIR) algorithms as
-demonstrated in the \texttt{RAX} experiment
-\cite{williams96,mus98,williams97} and more recently on an AUV by
-Dearden and others \cite{wang09,ernits10,dearden11}. The second
-challenge has to implementing a policy within the \rx framework that
-explicitly deals with recovery since controller failure can be
-distributed across reactors. Since reactors are hierarchical in their
-relationship to one another, their well defined dependencies allow
-them to be removed from an agent systematically. Reactors more
-abstract in scope can be removed before those which are less abstract
-(and closer to the hardware). In the event a Navigation reactor, for
-example, has a fault, it can be removed safely with lower level
-functionalities (such as those in the executive). When the Navigator
-is removed, it will recursively remove more abstract reactors (such as
-a Mission Manager) that depend on it. Such a safety feature then
-allows the system designers to ensure that early on they can invest
-more in ensuring that reactors lower in the hierarchy are robust and
-potentially formally verified. It also means that there is a safe and
-sure way to take control of the vehicle at any level within the
-reactors hierarchy; as long as low level reactors are still active one
-will always been able to send goals/commands to them and execute what
-are traditionally known as runout sequences (or clean up scripts).
+towards increased \kcomment{mission} robustness \kcomment{in the
+  context of fault diagnosis, isolation and recovery (FDIR)}. There
+are two outstanding diagnosis and failure recovery challenges that we
+need to address; the first has to do with methodology related to model
+design of the instrument payload for a highly modular and configurable
+vehicle; this is local to a reactor. Doing so would enable (as a first
+step) using systematic FDIR algorithms as demonstrated in the
+\texttt{RAX} experiment \cite{williams96,mus98,williams97} and more
+recently on AUVs \cite{wang09,ernits10,dearden11}. The second
+challenge has to \kcomment{do with} implementing a policy within the
+\rx framework that explicitly deals with recovery since controller
+failure can be distributed across reactors. Since reactors are
+hierarchical in their relationship to one another, their well defined
+dependencies allow them to be removed from an agent systematically as
+noted in Section \ref{sec:rx-reactor-failure}. Reactors more abstract
+in scope can be removed before those which are less abstract (and
+closer to the hardware). In the event a \texttt{Navigator} reactor,
+for example, has a fault, it can be removed safely with lower level
+functionalities (such as those in the executive). When the
+\texttt{Navigator} is removed, it will recursively remove more
+abstract reactors (such as a \texttt{Mission Manager}) that depend on
+it. Such a safety feature then allows the system designers to ensure
+that early on they can invest more in ensuring that reactors lower in
+the hierarchy are robust and potentially formally verified. It also
+means that there is a safe and sure way to take control of the vehicle
+at any level within the reactors hierarchy; as long as low level
+reactors are still active one will always been able to send
+goals/commands to them and execute what are traditionally known as
+runout sequences.
 
@@
 a prime motivation to move towards more semi-supervised methods such
-as \cite{kumar11,sergio12}. With the advent of the \od however,
-another direction our research is taking us is in Recommender systems
-\cite{Adomavicius05} as noted in Section \ref{sec:results}. We are
-working to design and deploy software infrastructure that will learn
-from user tagged data. To do so, we will use tagging techniques
-similar to what users of commercial image sharing sites like Flickr
-and Picassa do, to learn the context of the data (including remote
-sensing). In addition, the \od will integrate statistical Machine
-Learning techniques with incoming asynchronous data stream obtained by
-sensors and platforms. Identification models will be based on existing
-training sets for INLs and plumes; models for other features of
-scientific interests driven by \can requirements will be built in
+as \cite{kumar11}.
+
+With the advent of the \od however, another direction our research is
+taking us is in \texttt{Recommender Systems} \cite{Adomavicius05} as
+noted in Section \ref{sec:results}. We are working to design and
+deploy software infrastructure that will learn from user tagged
+data. To do so, we will use tagging techniques similar to what users
+of commercial image sharing sites like Flickr and Picassa do, to learn
+the context of the data (including remote sensing). In addition, the
+\od will integrate statistical Machine Learning techniques with
+incoming asynchronous data stream obtained by sensors and
+platforms. Identification models will be based on existing training
+sets for INLs and plumes; models for other features of scientific
+interests driven by \can requirements will be built in
 addition. Existing AUV archives for instance provide sufficient data
-for a learning base for a range of features of interest to \can; in
-addition we will build software that will learn to incorporate
+for a learning base for a range of features of interest to scientists;
+in addition we will build software that will learn to incorporate
 incoming data streams to provide an existence proof of a feature of
 interest.
+
+A number of issues remain to be explored improving \rx and \eu
+integration particularly in the context of engineering models for
+execution.
+
+Dispatchability of the {\em external} temporally flexible partial
+plans continues to be an open research problem. Substantial progress
+has been made \cite{vidal97,mus98a,tsam98,morris00} by analyzing the
+plan structure of static offline temporal plans. However in the
+context of embedded plan execution, balancing model design while
+dealing with execution uncertainty requires further research. For
+instance it should be possible to distinguish part of the plan that
+contributes to a reactor goal, which often needs to be dispatched as
+early as possible with other tokens describing world evolution with a
+preference to dispatch as late as possible. Making this distinction
+would improve the quality of agent behavior while avoiding extra model
+complexity.
+
+More abstractly at the architectural level, the ability to dynamically
+manipulate the reactor graph would be of importance while creating new
+reactors or modifying timelines. Doing so would provide more robust
+failure recovery mechanisms especially after a reactor has failed to
+synchronize. For example, an alternate reactor could take ownership of
+timelines left invalid by the destruction of another failed reactor
+ensuring a transparent implementation of component redundancy. While
+the current design broadly provides for such a capability, the \rx API
+has yet to implement it while ensuring other design elements within
+are not compromised.
+
+Finally and importantly, integration of reasoning about resources
+within \rx and tying it to existing \eu capabilities needs to be
+demonstrated. Specifically \rx needs to be able to update resource
+levels as incoming observations.  Doing so would ensure plan
+adaptation occurs dynamically when planned and observed resource
+values diverge -- for instance if the vehicle looses a bank of
+batteries during the course of a mission.  Enhanced modeling
+capabilities and augmented domain-independent search within \eu would
+make it easier for \rx users to create such domain models and to
+perform more proactive deliberation.
+
+Along with all these potential extensions, we are also very cognizant
+that the learning curve associated with \rx and \eu for a general user
+in the marine robotics community needs to be tackled. At present, the
+complex system that underlies \rx and its underlying framework, even
+if well documented, can be intimidating for a naive first time user.
 

### 2012-05-04 trunk/results.tex
updates
update

@@
+\vskip+0.5cm
 \section{Experimental Results}
@@
 Our early work focused on observing and sampling within Intermediate
-Nephaloid Layers (INLs) \footnote{INLs are fluid sheets of suspended
+Nephaloid Layers (INLs)\footnote{INLs are fluid sheets of suspended
   particulate matter that originate from the sea floor
@@
 \caption{\small An Intermediate Nepheloid Layer (INL) detected by an
-  AUV mission in the Monterey Bay in August 2005, showing the location
-  of the AUV transect in the bay (left) and optical backscatter image
-  at the 420 Nm frequency showing the INLs patchy nature
-  (right). \emph{Image courtesy: John Ryan, MBARI}}
+  AUV mission \kcomment{using a scripted mission plan} in the Monterey
+  Bay in August 2005. It shows the location of the AUV transect in the
+  bay (left) and optical backscatter image at the 420 Nm frequency
+  showing the INLs patchy nature (right). \emph{Image courtesy: John
+    Ryan, MBARI}}
 \label{fig:inl}
@@
 reactive, with the deliberation of an onboard temporal planner to
-actually trigger a sampler. This has two benefits. First, it goes to
-the core of having a deliberative approach to mission planning by
-balancing the immediate with that which is projected. Doing so allows
+trigger a sampler. This has two benefits. First, it goes to the core
+of having a deliberative approach to mission planning by balancing the
+immediate with that which is projected \cite{olaya11}. Doing so allows
 a more systematic approach in dealing with spatial factors in sampling
@@
 multiple features). These are critical not only for science but for
-informed decision making which pure reactive approaches simply cannot
-undertake.
+informed decision making which \kcomment{is challenging} for pure
+reactive approaches.
 
@@
 temporal dimension with a Hidden Markov Model (HMM) \cite{rabiner86,
-  Rabiner89atutorial}.  An HMM is a sequential state transition model
+  Rabiner89atutorial}.  A HMM is a sequential state transition model
 in which the current state is not directly visible, but an
@@
 appropriate state of recognition of the INL feature
-\cite{mcgann08b}. Not only was deliberation aided by direct perception
-of the state (of the INL) but temporal transition functions added a
+\cite{mcgann08b}. Not only is deliberation aided by direct perception
+of the state (of the INL) but temporal transition functions add a
 further level of validation between expected states in the INL
 model. This coupling of Machine Learned model with a deliberative
-planner is novel in the ocean sciences and is a continuing direction
-of investigation \cite{kumar11}.
+planner is novel in the ocean sciences and is a continuing target of
+investigation \cite{kumar11}.
 
@@
 the deliberating planner in-situ, the algorithm dynamically generates
-a policy on sample acquisition during the course of the mission. In so
+a policy on sample acquisition during the course of a mission. In so
 doing, utility functions can be used to balance sparcity or abundance
@@
 \emph{Search}: After the mapping phase and on receipt of this message
-on the vehicle T-REX evaluates the goal, automatically synthesizes a
-set of waypoints by generating a plan and then dispatches the plan to
+on the vehicle \rx evaluates the goal, automatically synthesizes a set
+of waypoints by generating a plan and then dispatches the plan to
 execute by having the vehicle swim back and forth at depth $d$ all the
@@
 account for vehicle control uncertainty), it then compares the
-temperature read with $t^{\circ}$. When the comparison value changes \rx
-assumes that the front is identified and switches into the next phase.
+temperature read with $t^{\circ}$. When the comparison value changes
+\rx assumes that the front is identified and switches into the next
+phase.
 
@@
 showing viably how a human being coupled with a goal-oriented
-controller could do problem solving jointly. The approach, is general
-and can be adapted to any feature or scalar field.
+controller could \kcomment{enable joint} problem solving. The
+approach, is general and can be adapted to any feature or scalar
+field.
 
@@
 controlling the Dorado AUV have centered on the \can (Controlled,
-Agile, and Novel Observing Network) program at MBARI. The fundamental
-science goals of the program are related to detecting and tracking of
-spatio-temporal coastal ocean features \cite{canon}. The dynamic
+Agile, and Novel Observing Network) program at MBARI \cite{canon}. The
+fundamental science goals of the program are related to detecting and
+tracking of spatio-temporal coastal ocean features. The dynamic
 features require a robotic platform to move with it, return water
@@
 strategies are tied to vehicle adaptation which is intrinsically tied
-to deliberation and difficult to accomplish otherwise.
+to deliberation and challenging to accomplish otherwise.
 
-Various adaptation methodologies in \can are still being
-investigated; however what has become a critical goal for the program
-has been to provide contextual environmental information about water
-patches. The form and content of such 'patches' varies; for instance
-initial attempts focused on Harmful Algal Blooms (HABs). They have
-large spatial ($>$ 50 km$^2$) and temporal (days to weeks) extent and
-are visible from space with visible coloring on the ocean surface. The
-direct impact of HABs is the introduction of toxins into the marine
-(and as a consequence human) food chain. In addition,
-\cite{anderson00} and \cite{hoagland06} estimate that the average
-economic impact resulting from HABs in the United States has been
+Various adaptation methodologies in \can are still being investigated;
+however what has become a critical goal for the program has been to
+provide contextual environmental information about water patches. The
+form and content of such 'patches' varies; for instance initial
+attempts focused on Harmful Algal Blooms (HABs). They have large
+spatial ($>$ 50 km$^2$) and temporal (days to weeks) extent and are
+visible from space with visible coloring on the ocean
+surface. \kcomment{The societal impact of HABs has been} in the
+introduction of toxins into the marine (and as a consequence human)
+food web, with \cite{anderson00} and \cite{hoagland06} estimating the
+consequent average economic impact in the United States to the tune of
 $\sim$\$75 million from 1987 to 2000. However, the processes behind
-bloom initiation, evolution and collapse is poorly understood in part
-due to the complex interactions between microbial communities and the
-surrounding environment. As a result, our capacity to assess the range
-of potential future scenarios that might result from ocean temperature
-changes, acidification, or nutrient shifts is highly limited.
+bloom initiation, evolution and collapse \kcomment{are} poorly
+understood in part due to the complex interactions between microbial
+communities and the surrounding environment. As a result, our capacity
+to assess the range of potential future scenarios that might result
+from ocean temperature changes, acidification, or nutrient shifts is
+highly limited.
 
@@
 quasi-lagrangian mode. As the water mass is being tracked with a
-drifting instrument, the vehicle is expected to move around it. While
-our model \kcomment{supports} multiples patterns (square, lawn-mower,
-transect(s)\dots) with any possible dimension factor, the majority of
-at-sea deployments have used a square pattern ($3X3$
+drifting instrument, the vehicle is expected to move around it
+\kcomment{gathering data to provide environmental context}. While our
+model \kcomment{supports} multiple patterns of observation (square,
+lawn-mower, transect(s)\dots) with any possible dimension factor, the
+majority of at-sea deployments have used a square pattern ($3X3$
 km\textsuperscript{2} or a $1X1$ km\textsuperscript{2})\footnote{Often
@@
   Dorado vehicle to provide a configurable and dynamic track-line
-  enclosure to provide oceanographic context to the drifting body.}
+  enclosure to provide context to the drifting body.}
 Fig. \ref{fig:dta-setup} shows the general setup of our experiments
@@
 as a high-level goal which is decomposed and synthesized as a survey
-plan. Currently, the vehicle completes the survey without any further
-dynamic updates from the drifter. In order to minimize the range of
-errors that the survey needs to encapsulate (data latency or surface
-or sub-surface currents for those drifters which have a drogue) the
-estimation is a projection in time and space of the drifters centroid
-sent with this goal from the shore-side app.
+plan \kcomment{and subsequently executed}. Currently, the vehicle
+completes the survey without any further dynamic updates from the
+drifter. In order to minimize the range of errors that the survey
+needs to encapsulate (data latency or surface or sub-surface currents
+for those drifters which have a drogue) the estimation is a projection
+in time and space of the drifters centroid sent with this goal from
+the shore-side app.
 
@@
 and uncertainty in the speed and bearing of the vehicle\footnote{In
-  our September 2010 experiment as the drifter moved over the Davidson
-  Seamount \cite{clague10} deviation in the California current
-  resulted in visible directional change that can be seen in
-  Fig. \ref{fig:sep10-transects} at the end of the mission.} exist;
-however, the observations of the survey to retain the environmental
-context require a change in reference frames so that the patch-center
-tagged by a GPS-tracked drifter remains within the survey perimeter at
-all times \cite{das11b}.
+  our September 2010 experiment \cite{das11b} as the drifter moved
+  over the Davidson Seamount \cite{clague10} deviation in the
+  California current, resulted in visible directional change that can
+  be seen in Fig. \ref{fig:sep10-transects} near the end of the
+  mission.} exist; moreover, the observations of the survey to retain
+the environmental context require a change in reference frames so that
+the patch-center tagged by a GPS-tracked drifter remains within the
+survey perimeter at all times \cite{das11b}.
 
@@
 \centering
-\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.4\textwidth]{figs/sep10-canon-cruise.png}}
+\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.44\textwidth]{figs/sep10-canon-cruise.png}}
 \subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.40\textwidth]{figs/september-iterations.pdf}}
@@
 
-These theoretic\kcomment{al} results were validated in a five-day long
+These theoretic\kcomment{al} results were validated in a five-day
 off-shore deployment carried out in September $2010$, $160$ kms off
@@
 velocity in the earth frame. Waypoints were computed once at the
-beginning of the survey with the AUV surfacing once for every survey,
-with each survey lasting $\sim1$ to $1.5$ hours. A total of 60 surveys
-were attempted over the course of the $5$
+beginning of the survey with the AUV surfacing once \kcomment{within
+  each} survey, with each survey lasting $\sim1$ to $1.5$ hours. A
+total of $60$ surveys were attempted over the course of $5$
 days. Fig. \ref{fig:sep10-transects} shows the overall transects over
 the course of the mission, with gaps in between days when the AUV was
-being recharged. Each deployment lasted $\sim12$ hours. The black dots
-in the figure show the beginning of each iteration when drifter
-updates were received by the AUV.
+being recharged on deck. Each deployment lasted $\sim12$ hours. The
+black dots in the figure show the beginning of each iteration when
+drifter updates were received by the AUV.
 
@@
 shore via the \od following a setup similar to
-Fig.\ref{fig:dta-setup}, to the Dorado running \rxe.
+Fig. \ref{fig:dta-setup}, to the Dorado running \rxe.
 
@@
 tracking more capable drifting platforms. In particular experiments
-with a profiling CTD instrument, the \texttt{Wirewalker}
-\cite{rainville01,pinkel11} are resulting in new modalities of making
-Lagrangian and Eulerian observations. Fig. \ref{fig:sept11-wirewalker}
-shows one such experiment in September 2011 with a GPS-enabled
-\texttt{Wirewalker} was tracked by the Dorado running \rxe. We are
-currently investigating spatio-temporal correlations \cite{gneiting07}
-of CTD data obtained by the \texttt{Wirewalker} and tying it to the
-observations made in the contextual environment by the Dorado's
-sensors.
+with a profiling CTD (Conductivity, Temperature, Depth) instrument,
+the \texttt{Wirewalker} \cite{pinkel11} are resulting in new
+modalities of making Lagrangian and Eulerian
+observations. Fig. \ref{fig:sept11-wirewalker} shows one such
+experiment in September 2011 with a GPS-enabled \texttt{Wirewalker}
+was tracked by the Dorado running \rxe. We are currently investigating
+spatio-temporal correlations \cite{gneiting07} of CTD data obtained by
+the \texttt{Wirewalker} and tying it to the observations made in the
+contextual environment by Dorado's sensors \cite{graham12}.
 
@@
 schedules.  Once a feature set has been identified for scientific
-observation and science goals have been collated, the appropriate
-robotic assets need to be assigned and deployed for observation and
+observation and science goals have been collated, appropriate robotic
+assets need to be assigned and deployed for observation and
 sampling. This asset allocation in turn depends on parameters like
@@
 deployments such as battery charge needed or distance/time to deploy,
-impact to long-range deployment plans among others. All drive the
+impact to long-range deployment plans, among others. All drive the
 strategy to balance short and long-term scientific and engineering
-needs. Our \can deliberations during the 2010 field season for example,
-were driven by all of the above \cite{das11}. Often these strategies
-were rebalanced during each decision-making cycle, especially in light
-of opportunistic science goals. For example, gliders running at $0.5$
-knots were often retargeted when surface support vessels noted the
-apparent dispersal of the bloom to subsurface layers when HABs were a
-\can target. However, their retargeting was carefully
-monitored to balance the long-term desire of obtaining time series
-within fixed tracks and with the need to ensure these assets were not
-exposed to strong currents. Powered AUVs were targeted to chase after
-advecting blooms, but the decision on their deployment locations and
-survey areas too close to the surf-zone were avoided. In
-patch-tracking experiments, the deployment of drifters close to shore
-and where depths of $20m$ or less were encountered, was avoided even
-when contrary to scientific intent. The need to obtain high-quality
-data for post-hoc analysis was also driven by the response time of
-data returned by experiment partners.
+needs. Our \can deliberations during the 2010 field season for
+example, were driven by all of the above \kcomment{factors}
+\cite{das11}. Often these strategies were rebalanced during each
+decision-making cycle, especially in light of opportunistic science
+goals. For example, gliders running at $0.5$ knots were often
+retargeted when surface support vessels noted the apparent dispersal
+of a bloom to subsurface layers when HABs were a \can target. However,
+their retargeting was carefully monitored to balance the long-term
+desire of obtaining time series within fixed tracks and with the need
+to ensure these assets were not exposed to strong currents. Powered
+AUVs were targeted after advecting blooms, but the decision on their
+deployment locations and survey areas too close to the surf-zone were
+avoided. In patch-tracking experiments, the deployment of drifters
+close to shore and where depths of $20m$ or less were encountered, was
+avoided even when contrary to scientific intent. The need to obtain
+high-quality data for post-hoc analysis was also driven by the
+response time of data returned by experiment partners.
 
@@
   system uses the same \eu planner used in \rxe. It continues to be
-  used routinely to this day on the MER mission.}.
+  used routinely to this day on mission-critical uplink process for
+  the MER mission.}.
 
@@
 making, onshore as well as on-board. We believe ultimately \od will
-combine capabilities inferential and data gathering and analysis
+combine inferential, data gathering, archiving and analysis
 capabilities and more uniquely, provide an event response capability
@@
 environment. The association between the signal and a prospective
-event will be undertaken through tagging by experts and machine
-learning techniques used in recommender systems
+event will be undertaken through tagging by experts and Machine
+Learning techniques used in recommender systems
 \cite{Adomavicius05}. Adaptive sampling strategies will then balance
-decision-making \situ with that done onshore with computational models
-and human agents, the subject of a new NSF project in which the
+decision-making \situ with \kcomment{those} onshore with computational
+models and human agents, the subject of a new NSF project in which the
 authors are participating.
 
-Operationally, the \od has become a software system that the
-scientist consults during a \can field campaign. S/he consults
-up-to-date model outputs, synoptic views of ocean color from remote
-sensing data and weather conditions for the next $24$-hour planning
-cycle and then queries the system to compute candidate sampling
-locations. In the near future, the \od will consult its situational
-awareness database, incorporate model outputs projecting these outputs
-into the space-time region of interest to generate candidate
-hypotheses with associated utility metrics. With the scientist's
-guidance the automated planner would then determine the most viable
-robotic option(s) to be dispatched (underwater, surface or aerial),
-given available resources, distance of travel, required payload and
-the environment. For instance, unpowered autonomous gliders would be
-chosen for longer-duration missions with smaller payload requirements
-in areas of moderate to weak ocean currents. The automated planner
-would balance these criteria to compute a viable trajectory plan for
-the selected vehicle. Vehicles with more controllability would be
-given a specified area to survey and the path generated \situ. Goals
-are communicated to the robot after analysis on shore for selective
-targeting of features by the conglomerate of vehicles in the ocean
-being tracked by the \od. Thus far, the \od has provided a simple
+Operationally, the \od has become a software system that the scientist
+consults during a \can field campaign. S/he consults up-to-date model
+outputs, synoptic views of ocean color from remote sensing data and
+weather conditions for the next $24$-hour planning cycle and then
+queries the system to compute candidate sampling locations. In the
+near future, the \od will consult its situational awareness database,
+incorporate model outputs projecting these outputs into the space-time
+region of interest to generate candidate hypotheses with associated
+utility metrics. With the scientist's guidance the automated planner
+would then determine the most viable robotic option(s) to be
+dispatched (underwater, surface or aerial), given available resources,
+distance of travel, required payload and the environment. For
+instance, gliders could be chosen for longer-duration missions with
+smaller payload requirements in areas of moderate to weak ocean
+currents. The automated planner would balance these criteria to
+compute a viable trajectory plan for the selected vehicle. Vehicles
+with more controllability would be given a specified area to survey
+and the path generated \situ. Goals are communicated to the robot
+after analysis on shore for selective targeting of features by the
+conglomerate of vehicles in the ocean \kcomment{that are} being
+tracked by the \od. \kcomment{To date}, the \od has provided a simple
 yet functional tool for scientists to use in the daily complex

### 2012-06-05 europa-basics.tex
updates

@@
 \centering
-  \includegraphics[scale=0.45]{figs/shopping_europa.pdf}
+  \includegraphics[scale=3]{figs/shopping_europa.jpeg}
   \caption{\small A potential plan for a robot shopping for groceries.}
@@
 \centering
-\includegraphics[scale=0.3]{figs/Example-NQueens1.jpg}
+\includegraphics[scale=2.0]{figs/Example-NQueens1.jpeg}
 \caption{\small An $N$-Queens solution generated by \eue.}
@@
 \subfloat[\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
-  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=0.25]{figs/Example-UBO0.jpg}}\qquad
+  problem.]{\label{fig:rcpsp-1}\includegraphics[scale=1.8]{figs/Example-UBO0.jpeg}}\qquad
 \subfloat[\small A \eu solution to a shopping agent problem domain where
-  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=0.3]{figs/Example-Shopping0.jpg}}
+  the agent needs to buy Bananas, Milk and a Drill.]{\label{fig:shopping-1}\includegraphics[scale=2.2]{figs/Example-Shopping0.jpeg}}
 \caption{\small \eu based solutions to scheduling and planning problems.}
@@
 \centering
-\includegraphics[scale=0.3]{figs/constraint-net.pdf}
+\includegraphics[scale=1.9]{figs/constraint-net.jpeg}
 \caption{\small Variables and Constraints represented as a Constraint

### 2012-06-05 europa.tex
updates

@@
 \centering
-\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
+\includegraphics[scale=2.5]{figs/EUROPA-Architecture.jpeg}
 \caption{\small The \eu software architecture.}
@@
 
-\kcomment{Before going into the details of each module, a simple
-  example illustrates how they interact at an abstract level:}
+Before going into the details of each module, a simple example
+illustrates how they interact at an abstract level.
 
+\begin{figure}[h!]
 \begin{verbatim}
@@
 }
-
 class LightSwitch extends Timeline
@@
     }
-
     action turnOn  { duration=1; }
@@
 }
-
 LightSwitch::turnOn
@@
 }
-
 LightSwitch::turnOff
@@
 }
-
 // Problem instance : turning the light off
@@
 lt(goal1.start,10);
+\end{verbatim}
+\caption{\small \nd modeling example for a light bulb and switch control.}
+\label{fig:nddllight}
+\end{figure}
 
-\end{verbatim}
-
-\kcomment{The example above is a \nd (covered in more detail in
+\kcomment{Fig. \ref{fig:nddllight} is a \nd (covered in more detail in
   Section \ref{sec:europa:modeling}) model of a light bulb and a
@@
 
-To articulate the most important elements of \nde, we show how the
-Shopping Agent problem could be modeled:
-
+\begin{figure}[tbp]
 \begin{verbatim}
-
 // Locations (Home, SuperMarket, etc.)
@@
 }
-
-// Products (Milk, Banana, etc.) and
+// Products (Milk, Banana, etc.)
 class Product {
@@
 }
-
 // ProductLocations (Banana can be found at SuperMarket, for example)
@@
 }
-
-// Use built-in Timeline functionality to enforce that an agent:
-// a) Can't be at more than one place at a time.
-// b) Can't Go more than one place at a time.
-// c) Can't Go somewhere and be At somewhere at the same time.
+// Use build-in timeline functionality to enforce that an agent cannot
+// be at more than one place at one time
+// Go to more than one place at one time
+// Go somewhere and be At somewhere at the same time 
 class AgentLocation extends Timeline{
@@
   }
-
   predicate Going {
@@
 }
+// In addition to having a location timeline, the agent can buy and
+// own things. Note that the actions and the location predicates can
+// be concurrent so they cannot be on the same timeline.
+class Agent {
 
-// In addition to having a location timeline, the agent can buy and own things.  
-// Note that the actions and the location predicates can be concurrent 
-// so they can't be on the same Timeline
-class Agent {
   AgentLocation location;
-
   Agent() {
@@
   }
-
   action Buy {
@@
   } 
-
   action Go {
@@
   }
-
   predicate Own {
@@
 \end{verbatim}
+\caption{\small \nd model of the Shopping Agent example.}
+\label{fig:nddlshop}
+\end{figure}
 
-The model above the main elements of a problem domain can be mapped to
-\eus representation in the \kcomment{following manner}:
+To articulate the most important elements of \nde, we show how the
+Shopping Agent problem could be modeled in
+Fig. \ref{fig:nddlshop}. The main elements of a problem domain can be
+mapped to \eus representation in the \kcomment{following manner}:
 
@@
 
+\begin{figure}[h!]
 \begin{verbatim}
 
-// Define the rules for our actions:
 Agent::Go {
@@
 \end{verbatim}
+\caption{\small \nd rules for the Shopping Agent.}
+\label{fig:nddlshop-rules}
+\end{figure}
 
 Domain rules can be defined in \nd in the context of a predicate or an
-action. In the example above, the main elements of a rule definition
-\kcomment{are as follows}:
+action. In Fig. \ref{fig:nddlshop-rules}, the main elements of a rule
+definition \kcomment{are as follows}:
 
@@
 With the specification for the Shopping Agent domain \kcomment{above},
-a particular problem instance could be stated as:
+a particular problem instance could be stated as in
+Fig. \ref{fig:nddlshop-instance}.
 
+\begin{figure}[h!]
 \begin{verbatim}
@@
 \end{verbatim}
+\caption{\small \nd instance and goal specification for the Shopping Agent.}
+\label{fig:nddlshop-instance}
+\end{figure}
 
@@
 with interval arithmetic; its elements are summarized in
-Fig. \ref{fig:allen-algebra}
+Fig. \ref{fig:allen-algebra}.
 
@@
 \centering
-\includegraphics[scale=0.4]{figs/Allen-algebra.pdf}
+\includegraphics[scale=2.3]{figs/Allen-algebra.jpeg}
 \caption{\small Temporal relations defined within the planner are
@@
   \centering
-  \subfloat[\small A Simple Temporal Network (STN). \kcomment{Nodes correspond
+  \subfloat[\small A Simple Temporal Network (\texttt{STN}). \kcomment{Nodes correspond
   to variables and their domains and arc's the \emph{constraint} or
-  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
-  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
-  \caption{\small An illustration of a Simple Temporal Network
-    (\texttt{STN}) and its corresponding distance graph.}
+  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.5\columnwidth]{figs/stn-graph.jpeg}}
+  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.5\columnwidth]{figs/distance-graph.pdf}}
+  \caption{\small An illustration of a \texttt{STN} and its
+    corresponding distance graph.}
   \label{fig:stn}
@@
 \centering
-\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
-\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eu.}
+\includegraphics[scale=2]{figs/europa-resource-envelopes.jpeg}
+\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eue.}
 \label{fig:resenvelopes}

### 2012-06-05 future.tex
updates

@@
   \centering
-  \includegraphics[scale=0.45]{figs/autonomy-topics.pdf}
+  \includegraphics[scale=0.45]{figs/autonomy-topics.jpeg}
   \caption{\small Overlapping topics in our research in deliberation
@@
 to graceful system degradation, making the overall approach effective
-towards increased robustness. There are two outstanding diagnosis and
-failure recovery challenges that we need to address; the first has to
-do with methodology related to model design of the instrument payload
-for a highly modular and configurable vehicle; this is local to a
-reactor. Doing so would enable (as a first step) using systematic
-Fault Diagnosis, Isolation and Recovery (FDIR) algorithms as
-demonstrated in the \texttt{RAX} experiment
-\cite{williams96,mus98,williams97} and more recently on an AUV by
-Dearden and others \cite{wang09,ernits10,dearden11}. The second
+towards increased \kcomment{mission} robustness \kcomment{in the
+  context of fault diagnosis, isolation and recovery (FDIR)}. There
+are two outstanding diagnosis and failure recovery challenges that we
+need to address; the first has to do with methodology related to model
+design of the instrument payload for a highly modular and configurable
+vehicle; this is local to a reactor. Doing so would enable (as a first
+step) using systematic FDIR algorithms as demonstrated in the
+\texttt{RAX} experiment \cite{williams96,mus98,williams97} and more
+recently on AUVs \cite{wang09,ernits10,dearden11}. The second
 challenge has to \kcomment{do with} implementing a policy within the
@@
 hierarchical in their relationship to one another, their well defined
-dependencies allow them to be removed from an agent
-systematically. Reactors more abstract in scope can be removed before
-those which are less abstract (and closer to the hardware). In the
-event a Navigation reactor, for example, has a fault, it can be
-removed safely with lower level functionalities (such as those in the
-executive). When the Navigator is removed, it will recursively remove
-more abstract reactors (such as a Mission Manager) that depend on
+dependencies allow them to be removed from an agent systematically as
+noted in Section \ref{sec:rx-reactor-failure}. Reactors more abstract
+in scope can be removed before those which are less abstract (and
+closer to the hardware). In the event a \texttt{Navigator} reactor,
+for example, has a fault, it can be removed safely with lower level
+functionalities (such as those in the executive). When the
+\texttt{Navigator} is removed, it will recursively remove more
+abstract reactors (such as a \texttt{Mission Manager}) that depend on
 it. Such a safety feature then allows the system designers to ensure
@@
 goals/commands to them and execute what are traditionally known as
-runout sequences (or clean up scripts).
+runout sequences.
 
@@
 a prime motivation to move towards more semi-supervised methods such
-as \cite{kumar11}. With the advent of the \od however,
-another direction our research is taking us is in \texttt{Recommender
-  Systems} \cite{Adomavicius05} as noted in Section
-\ref{sec:results}. We are working to design and deploy software
-infrastructure that will learn from user tagged data. To do so, we
-will use tagging techniques similar to what users of commercial image
-sharing sites like Flickr and Picassa do, to learn the context of the
-data (including remote sensing). In addition, the \od will integrate
-statistical Machine Learning techniques with incoming asynchronous
-data stream obtained by sensors and platforms. Identification models
-will be based on existing training sets for INLs and plumes; models
-for other features of scientific interests driven by \can requirements
-will be built in addition. Existing AUV archives for instance provide
-sufficient data for a learning base for a range of features of
-interest to \can; in addition we will build software that will learn
-to incorporate incoming data streams to provide an existence proof of
-a feature of interest.
+as \cite{kumar11}.
+
+With the advent of the \od however, another direction our research is
+taking us is in \texttt{Recommender Systems} \cite{Adomavicius05} as
+noted in Section \ref{sec:results}. We are working to design and
+deploy software infrastructure that will learn from user tagged
+data. To do so, we will use tagging techniques similar to what users
+of commercial image sharing sites like Flickr and Picassa do, to learn
+the context of the data (including remote sensing). In addition, the
+\od will integrate statistical Machine Learning techniques with
+incoming asynchronous data stream obtained by sensors and
+platforms. Identification models will be based on existing training
+sets for INLs and plumes; models for other features of scientific
+interests driven by \can requirements will be built in
+addition. Existing AUV archives for instance provide sufficient data
+for a learning base for a range of features of interest to scientists;
+in addition we will build software that will learn to incorporate
+incoming data streams to provide an existence proof of a feature of
+interest.
 
@@
 has yet to implement it while ensuring other design elements within
-\rx and \eu are not compromised.
+are not compromised.
 

### 2012-06-05 trunk/arch.tex
updates

@@
   \centering
-  \includegraphics[scale=0.45]{figs/token-disptch.pdf}
+  \includegraphics[scale=2.5]{figs/token-disptch.jpeg}
   \caption{\small \kcomment{Dispatching a token occurs only when the
@@
     with multiple possible clients ({\color{green}green} dashed
-    lines).]{\includegraphics[scale=0.45]{figs/AUV-agent.pdf}
+    lines).]{\includegraphics[scale=2.65]{figs/AUV-agent.jpeg}
   \label{fig:agent}
@@
   relates to desired future values of state variables.]{
-    \includegraphics[width=0.4\columnwidth]{figs/TREX_flow}
+    \includegraphics[width=0.4\columnwidth]{figs/TREX_flow.jpeg}
     \label{fig:trex:flow}
@@
   \subfloat[\small \kcomment{A view of synchronization and deliberation within a
-  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle}} 
+  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle.jpeg}} 
   \qquad
@@
     interrupted by {\em synchronization} at the beginning of every
-    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle}}
+    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle.jpeg}}
     \caption{\small \kcomment{Detailed and abstracted views of
@@
   along with all its {\em internal} timelines while generating a
-  \texttt{Failed} observation.]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.pdf}
+  \texttt{Failed} observation.]{\includegraphics[width=0.35\columnwidth]{figs/AUV-fail0.jpeg}
   \label{fig:fail0}}
@@
   synchronize too leading to a bottom-up destruction of ``faulty'' reactors.]{
-    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1}\label{fig:fail1}}
+    \includegraphics[width=0.35\columnwidth]{figs/AUV-fail1.jpeg}\label{fig:fail1}}
   \caption{\small Impact of the \texttt{Sampling} reactor failure in
@@
   \centering
-  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
+  \includegraphics[width=0.5\columnwidth]{figs/synch-relation.jpeg}
   \caption{\small Illustration of synchronization flaws in a
@@
 future inconsistency.  Such an assumption impacts the set of possible
-domains the reactor can support while remaining complete. Take for
-example this rule from our \texttt{Shopping} model:
-
+domains the reactor can support while remaining
+\kcomment{algorithmically} complete.
+
+\begin{figure}[h!]
 \begin{verbatim}
@@
 \end{verbatim}
-
-While this model is perfectly acceptable for planning it becomes
-problematic when execution is taken into account especially since it
-puts a strong emphasis between an evolving plan and an unknown future;
-\texttt{Going} to a location and the {\em expected} future outcome of
-ending at this location, ties it strongly to the constraint in line
-$5$ above. For instance, assume that the \texttt{Agent.location} is
-{\em external} to the reactor which receives the observation
-\texttt{Going(Home,SuperMarket)} which synchronization has resolved by
-producing the observation \texttt{Go(Home, SuperMarket)} in the
-\texttt{Agent} timeline. This would imply that the only possible
-outcome on the \texttt{Agent.location} timeline, is to be
-\texttt{At(SuperMarket)} in the directly foreseeable future.
+\label{fig:nddl-sync}
+\caption{\small \nd rule from the Shopping Agent example with impact
+  on a future state.}
+\end{figure}
+
+Take for example this rule from our \texttt{Shopping} model in
+Fig. \ref{fig:nddl-sync}. While this model is perfectly acceptable for
+planning it becomes problematic when execution is taken into account
+especially since it puts a strong emphasis between an evolving plan
+and an unknown future; \texttt{Going} to a location and the {\em
+  expected} future outcome of ending at this location, ties it
+strongly to the constraint in line $5$ above. For instance, assume
+that the \texttt{Agent.location} is {\em external} to the reactor
+which receives the observation \texttt{Going(Home,SuperMarket)} which
+synchronization has resolved by producing the observation
+\texttt{Go(Home, SuperMarket)} in the \texttt{Agent} timeline. This
+would imply that the only possible outcome on the
+\texttt{Agent.location} timeline, is to be \texttt{At(SuperMarket)} in
+the directly foreseeable future.
 
@@
 One could therefore alter this part of the model to give the reactor
-more flexibility as follows:
-
-% \begin{minipage}[t]{\textwidth}
-% \begin{figure}
+more flexibility by relaxing the constraint in line $5$ from a strict
+\texttt{equals} to a \texttt{contains} and to add the new constraint
+at line $11$ to ensure that the last \texttt{Going} ends up in the
+location the agent actually wants the robot to be at. This first level
+of relaxation will therefore allow the agent to use multiple
+\texttt{Going} tokens should the first token fail to reach the
+objective, as shown below:
+
+\begin{figure}[tbp]
 \begin{verbatim}
@@
 \end{verbatim}
-% \label{lab:model:shopping}
-% \caption{\small Model fragment to enforce a \texttt{Going} end condition}
-% \end{figure}
-% \end{minipage}
-
-We do so by relaxing the constraint in line $5$ from a strict
-\texttt{equals} to a \texttt{contains} and to add the new constraint
-at line $11$ which ensures that the last \texttt{Going} ends up in the
-location the agent actually wants the robot to be at. This first level
-of relaxation will therefore allow the agent to use multiple
-\texttt{Going} tokens should the first one fail to reach the
-objective.
-
-As the example above shows, synchronization involves a deliberation
-process. The risk is such that the search for a solution is not a
-simple sequence of decisions where the resolution can be easily
-identified by the solver as any other option would lead immediately to
-plan inconsistency. Since such a process is required by the \rx at
-\emph{every} tick for \emph{every} reactor the model needs to be
-designed carefully.
-
-One critical aspect to take into account is that from a \rx point of
-view, synchronization is the only critical task of a reactor. Such a
-process is used in order for the agent to identify its current state
-and maintain it consistently throughout the agent's lifetime. For a
+\caption{\small Model fragment to enforce a \texttt{Going} end condition}
+\label{lab:model:shopping}
+\end{figure}
+
+As the example in Fig. \ref{lab:model:shopping} shows, synchronization
+involves a deliberation process. The risk is such that the search for
+a solution is not a simple sequence of decisions where the resolution
+can be easily identified by the solver as any other option would lead
+immediately to plan inconsistency. Since such a process is required by
+the \rx at \emph{every} tick for \emph{every} reactor the model needs
+to be designed carefully.
+
+Synchronization is the only critical task of a reactor. Such a process
+is used in order for the agent to identify its current state and
+maintain it consistently throughout the agent's lifetime. For a
 reactor not being able to find a consistent solution during
 synchronization -- implying that after evaluating all possible
-solutions provided by a model it was unable to find a consistent plan
--- not only indicates that a reactor is unable to explain the state of
-the world, but more importantly it may jeopardize other reactors that
-rely on its {\em internal} state. For this reason a failure to
-synchronize immediately results in the \rx agent killing a reactor and
-notifying any reactor that depends on its {\em internal} state
-variables; it does so by posting the special observation
-\textsf{Failed}. This special state is a special predicate which is
-acceptable for all the {\em internal} and {\em external} timelines of
-\kcomment{an} \eu reactor.  By so doing we allow for graceful
-degradation where other reactors can readapt their plans on receipt of
-this observation.
-
-Finally, at the engine level careful choices were made in both the
-focus of the search -- limited only to tokens that can potentially
-overlap the current tick -- and heuristics where the {\em current
-  state flaw} is considered last during the search process. The later
-limits the amount of backtracking for a solution -- or at least how
-deep one needs to backtrack in a decision tree.  Similarly the
-sequence in which we evaluate the possible resolution for such a flaw
--- presented in Proposition \ref{prop:csf:resolve} -- was selected
-according to some simple assumptions, that any external observation
-that is compatible with an existing token in the current plan reflects
-the execution of the plan itself. In other words this means that our
+solutions provided by a model \kcomment{being} unable to find a
+consistent plan -- not only indicates that a reactor is unable to
+explain the state of the world, but more importantly it may jeopardize
+other reactors that rely on its {\em internal} state. For this reason
+a failure to synchronize immediately, results in the \rx agent killing
+a reactor and notifying any reactor that depends on its {\em internal}
+state variables; it does so by posting the special observation
+\textsf{Failed}. This special state is acceptable for all the {\em
+  internal} and {\em external} timelines of \kcomment{an} \eu reactor.
+By so doing we allow for graceful degradation where other reactors can
+readapt their plans on \kcomment{receiving} this observation.
+
+Finally, at the engine level careful choices \kcomment{are} made in
+both the focus of the search -- limited only to tokens that can
+potentially overlap the current tick -- and heuristics where the {\em
+  current state flaw} is considered last during the search
+process. The later limits the amount of backtracking for a solution --
+or at least how deep one needs to backtrack in a decision tree.
+Similarly the sequence in which we evaluate the possible resolution
+for such a flaw -- \kcomment{described} in Proposition
+\ref{prop:csf:resolve} -- \kcomment{is} selected according to
+\kcomment{the assumption}, that any external observation that is
+compatible with an existing token in the current plan reflects the
+execution of the plan itself. In other words this means that our
 solver always prefers to use tokens \emph{already existing} in the
@@
   \centering
-  \includegraphics[width=0.65\columnwidth]{figs/plan_window}
+  \includegraphics[width=0.65\columnwidth]{figs/plan_window.jpeg}
   \caption{\small The planning window of the deliberation solver and
@@
 alter their temporal scope to be outside of the planning window.  This
-allows the planner to focus on its planning horizon and report
-decision on flaws that are beyond this scope. As time advances, these
+allows the planner to focus on its planning horizon and not report
+decisions on flaws that are beyond this scope. As time advances, these
 potential future flaws will eventually be part of the planning window
@@
 goals.  This in turn, results in an apparent \emph{continuous
-  planning} of the reactor. When there are no more flaws present in
-the current planning scope a partial plan solution is found. This has
-two effects:
+  planning} \kcomment{\cite{myers99}} of the reactor. When there are
+no more flaws present in the current planning scope a partial plan
+solution is found. This has two effects:
 
@@
   \subfloat[\small A plan produced before
-  synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}} \\
+  synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0.jpeg}} \\
   \subfloat[\small Result of synchronization after the {\em Going}
   observation was received by the reactor with the plan showed in
-    Fig. \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
+    Fig. \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1.jpeg}}
     \caption{\small The Shopping example: Tokens in {\color{red}red}
@@
   \centering
-  \subfloat[\small Conflicting Observation.]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1}}\\
-  \subfloat[\small Relaxed Partial Plan.]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2}}
+  \subfloat[\small Conflicting Observation.]{\label{fig:shop:relax1}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-1.jpeg}}\\
+  \subfloat[\small Relaxed Partial Plan.]{\label{fig:shop:relax2}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_relax-2.jpeg}}
   \caption{\small Illustration of a conflict during synchronization

### 2012-06-05 trunk/concepts.tex
updates

@@
 \centering
-\includegraphics[scale=0.4]{figs/flexible-timelines.pdf}
+\includegraphics[scale=2.3]{figs/flexible-timelines.jpeg}
 \caption{\small Tokens with flexible temporal intervals and parametric
@@
     \subfloat[\small Timelines representing initial state
-    and goals for a Shopping Agent.]{\label{fig:europapr1}\includegraphics[scale=0.4]{figs/europa-pr-1.pdf}}\qquad
+    and goals for a Shopping Agent.]{\label{fig:europapr1}\includegraphics[scale=2.2]{figs/europa-pr-1.jpeg}}\qquad
     \subfloat[\small State of timelines after satisfying a goal to be
-    home by time $20$ on merging with initial state.]{\label{fig:europapr:merge}\includegraphics[scale=0.4]{figs/europa-pr-merge.pdf}}\qquad
+    home by time $20$ on merging with initial state.]{\label{fig:europapr:merge}\includegraphics[scale=2.2]{figs/europa-pr-merge.jpeg}}\qquad
     \subfloat[\small State of timelines after satisfying a goal to own
     a drill by time $10$ by activating corresponding
-    token.]{\label{fig:europapr:activate}\includegraphics[scale=0.4]{figs/europa-pr-activate.pdf}} 
+    token.]{\label{fig:europapr:activate}\includegraphics[scale=2.2]{figs/europa-pr-activate.jpeg}} 
     \caption{\small Evolution of a timeline from the Shopping Agent example.}
@@
 \begin{figure} \centering
-  \includegraphics[scale=0.45]{figs/europa-pr-states.jpg}
+  \includegraphics[scale=3.9]{figs/europa-pr-states.jpeg}
   \caption{\small Possible transitions for a token state.}
@@
     \centering
-    \includegraphics[scale=0.45]{figs/europa-pr-rules.pdf}
+    \includegraphics[scale=2.5]{figs/europa-pr-rules.jpeg}
     \caption{\small Master and Slave tokens generated as a result of a domain rule.}

### 2012-06-05 trunk/results.tex
updates

@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs/gulper.jpg}
+\includegraphics[width=0.45\textwidth]{figs/gulper.jpeg}
 \caption{\small{The Gulper water sampler \cite{Bird07} on MBARI's
@@
 \centering
-\includegraphics[scale=0.65]{figs/inl.jpg}
+\includegraphics[scale=0.65]{figs/inl.jpeg}
 \caption{\small An Intermediate Nepheloid Layer (INL) detected by an
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/front-mi.jpg}
+\includegraphics[width=0.8\textwidth]{figs/front-mi.jpeg}
 \caption{\small{Survey trackline of AUV following a thermal
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figs/dta-app.pdf}
+\includegraphics[width=0.5\textwidth]{figs/dta-app.jpeg}
 \caption{\small{Overall setup of \can lagrangian experiments with a
@@
 \centering
-\includegraphics[width=0.45\textwidth]{figs/june11-drifter-follow-180-181.pdf}
+\includegraphics[width=0.45\textwidth]{figs/june11-drifter-follow-180-181.jpeg}
 \caption{\small{A June 2011 drifter cruise shown in the Earth
@@
 \centering
-\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.44\textwidth]{figs/sep10-canon-cruise.png}}
-\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.40\textwidth]{figs/september-iterations.pdf}}
+\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.44\textwidth]{figs/sep10-canon-cruise.jpeg}}
+\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.41\textwidth]{figs/september-iterations.jpeg}}
   \caption{\small{The offshore \can ESP-drifter following cruise
@@
 \centering
-\includegraphics[width=0.7\textwidth]{figs/oct10-ballet.pdf}
+\includegraphics[width=0.7\textwidth]{figs/oct10-ballet.jpeg}
 \caption{\small An October 2010 robot ``ballet'' where a smaller
@@
 \centering 
-\subfloat[Temperature]{\label{fig:sep11-temp}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-temp.png}} 
-\subfloat[Salinity]{\label{fig:sep11-sal}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-salinity.png}}\\
-\subfloat[Fluorescence line height]{\label{fig:sep11-flh}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-flh.png}} 
-\subfloat[Nitrate]{\label{fig:sep11-n2}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-n2.png}} 
+\subfloat[\small Temperature]{\label{fig:sep11-temp}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-temp.jpeg}} 
+\subfloat[\small Salinity]{\label{fig:sep11-sal}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-salinity.jpeg}}\\
+\subfloat[\small Fluorescence line height]{\label{fig:sep11-flh}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-flh.jpeg}} 
+\subfloat[\small Nitrate]{\label{fig:sep11-n2}\includegraphics[width=0.5\textwidth]{figs/sep11-ww-n2.jpeg}} 
 \caption{\small Interpolated results of a September 2011 \rx mission
@@
   \subfloat[A conceptual view of the \od and how it fits into \can
-  field experiments]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
+  field experiments]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.jpeg}}\\
   \subfloat[Current implementation of \od for
-  \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
-\caption{\small The Oceanographic Decision Support System (\od) is a
+  \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.jpeg}}
+\caption{\small The Oceanographic Decision Support System (\ode) is a
   single portal for planning, visualization, situational awareness,
@@
 \centering
-\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.325\textwidth]{figs/october-canon-meeting.png}}
-\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.37\textwidth]{figs/virtual-drifter-tool.png}}
-\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.325\textwidth]{figs/photo-ipad-2.jpg}}
+\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.325\textwidth]{figs/october-canon-meeting.jpeg}}
+\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.37\textwidth]{figs/virtual-drifter-tool.jpeg}}
+\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.325\textwidth]{figs/photo-ipad-2.jpeg}}
 \caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with
@@
 conglomerate of vehicles in the ocean \kcomment{that are} being
-tracked by the \od. \kcomment{To date}, the \od has provided a simple
+tracked by the \ode. \kcomment{To date}, the \od has provided a simple
 yet functional tool for scientists to use in the daily complex

### 2012-06-05 final-version/DeliberativeControl.tex
Final Version to be sent to Publisher

--- 2012-05-02 trunk/setobook.tex
@@
 
-\tableofcontents{}
+% \tableofcontents{}
 
@@
 Rajan and Py are funded by a block grant from the David and Lucile
-Packard Foundation to MBARI. They thank their collaborators John Ryan,
-Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. They
-thank other members of the \eu team over the years, including Conor
-McGann, Ari J\'onsson, Jeremy Frank, Paul Morris, John Bresina,
-Michael Iatauro, Matthew Boyce and Tristan Smith.  The authors are
-grateful to Martha del Alto, David Korsmeyer and Steven Zornetzer all
-at NASA Ames \kcomment{who not only made \eu available as open source
-  but encouraged its use outside the agency}. Minh Bin Do (NASA Ames),
+Packard Foundation to MBARI. They thank their research collaborators
+John Ryan, Julio Harvey, Jim Bellingham and Robert Vrijenhoek at
+MBARI. The crew of the R/V \emph{Zephyr} and the DMO support team led
+by Hans Thomas has been invaluable. The authors also thank other
+members of the \eu team over the years, including Conor McGann, Ari
+J\'onsson, Jeremy Frank, Paul Morris, John Bresina, Michael Iatauro,
+Matthew Boyce and Tristan Smith.  They are grateful to Martha del
+Alto, David Korsmeyer and Steven Zornetzer all at NASA Ames
+\kcomment{who not only made \eu available as open source but
+  encouraged its use outside the agency}. Minh Bin Do (NASA Ames),
 William Yew Teck Tan (NUS, Singapore) and Rishi Graham (MBARI)
@@
 and early efforts in Constraint-based Temporal Planning while at NASA
-Ames, and which forms the basis of much of this work.
+Ames which forms the basis of much of this work.
 

### 2012-06-05 final-version/concepts.tex
Final Version to be sent to Publisher

--- 2012-06-05 trunk/concepts.tex
@@
 \cite{europapso} was named \eut and is the subject of this chapter and
-used by \rxe.  Note however, in using the name \eu, we refer to the
+used by \rxe.  Note however, in using the name \eue, we refer to the
 \emph{fundamental ideas as well as the planning paradigm} even if
@@
 
-\begin{figure}[!htb]
+\begin{figure}[!tb]
 \centering
@@
   noted in Section \ref{sec:europa:cp}, a variable can take values
-  from a domain. In \eu, domains are defined over a specific data
+  from a domain. In \eue, domains are defined over a specific data
   type; the primitive data types supported are \textit{String},
@@
 \item[\textbf{Tokens}] In first order logic, a predicate defines a
-  relation between objects and properties \cite{gen87}.  In \eu, we
+  relation between objects and properties \cite{gen87}.  In \eue, we
   define such relations between variables whose domains are sets of
@@
   instance of a predicate that represents an object's state or
-  behavior and is defined over a temporal extent. In \eu, all
+  behavior and is defined over a temporal extent. In \eue, all
   predicate instances are tokens. \kcomment{In} the same way that

### 2012-06-05 final-version/conclude.tex
Final Version to be sent to Publisher

\section{Conclusion}
\label{sec:conclusion}

% Final parting words summarizing the contribution of deliberation towards
% vehicle autonomy.

The Ocean Sciences are at a cusp, straddling traditional ship-based
exploration with newer observatory-based methods. In the United States
the NSF funded Ocean Observatories Initiative \cite{ooi} and regional
bodies such as \texttt{CeNCOOS} \cite{cencoos} are investing in new
observational methods and technologies which \kcomment{aim} to aid
oceanographers by producing substantially higher-resolution data
products. These will use moorings, cabled observatories and glider
fleets generating near real-time data from one or more locations as a
continuous stream. The tack chosen by \can has been to complement such
methods with the aim of developing methods for a \emph{portable
  mobile} observatory which is dependent on autonomous platforms. In
this context what we advocate in this chapter is an initial step in
making robotic platforms more adaptive, scaling them towards fleets of
vehicles (underwater \kcomment{and/}or on the surface) and most
importantly using them to solve the larger challenge of sampling
\kcomment{dynamic} spatio-temporal fields.

While our trajectory of research in the oceanographic domain follows
that in spacecraft autonomy, from onboard \cite{mus98} to
mixed-initiative methods \cite{bresina05}, the challenges are
significantly different with the environment playing a far larger
role. Moreover, prediction and projection of a future course of action
for a robot given the level of uncertainty calls for representational
and architectural methods which would enable adaptation at different
levels of abstraction, from mission-planning to low level actuation
\kcomment{and control}.

Our efforts were initially targeted to full robotic autonomy; while
this is an important goal that we continue to strive towards, it is
increasingly clear that human-in-the-loop methods also have a sizable
role to play in ocean exploration and by extension to \kcomment{the
  field of} Marine Robotics. Not only is an embedded
\kcomment{deliberative agent} an important element of the effort
towards smarter vehicles, but we are working towards an implementation
of \rx which will perform multi-vehicle plan synthesis for the \ode.

A key role that \rx has played and will continue to play, is towards
\emph{event response} scenarios. In Section \ref{sec:results} we show
the role Machine Learning techniques have played thus far in in-situ
plan adaptation. This interaction between Planning and Machine
Learning, we believe, will play a significant role going forward in
exploration and discovery. Interpretation of incoming sensor data to
deal with scientific surprise and \kcomment{re-adaptation} for
opportunistic science will be critical to dealing with the persistent
problem of undersampling the ocean\footnote{Walter Munk of the Scripps
  Institute of Oceanography has famously stated \emph{``Most of the
    previous century could be called a “century of undersampling”''}
  -- Testimony to the U.S. Commission On Ocean Policy, 18 April
  2002}.

Much of what we know of the ocean was derived by exploring in the
blind. In the near future, we anticipate longer duration robotic
vehicles with enough onboard computational capacity to run
deliberative agents such as \rxe, make water-column measurements and
reliably track and characterize dynamic features. While our
\kcomment{own} experiments have already demonstrated such
capabilities, the exploration Vs. exploitation tradeoff is better
determined when vehicles can make choices over longer time periods
(weeks and months) to study the impacts of bio-geochemical interaction
driven by longer duration physical forcing \kcomment{and climate
  change}.

Given our poor understanding of the coastal ocean in particular,
mixed-initiative methods of command and control for Sampling with
heterogenous assets is an important goal that is clearly at our
doorstep. We believe deliberation is, but a small yet critical part of
the solution to unfolding open research problems for AI and Marine
Robotics.

### 2012-06-05 final-version/europa-basics.tex
Final Version to be sent to Publisher

--- 2012-06-05 europa-basics.tex
@@
 important properties of \emph{Soundness}\footnote{A planning algorithm
-  is sound if invoked on a problem $P$ returns a plan which is a
-  solution for $P$.} and \emph{Completeness}\footnote{A planning
-  algorithm is complete if invoked on a solvable problem $P$ is
+  is \emph{sound} if invoked on a problem $P$ returns a plan which is
+  a solution for $P$.} and \emph{Completeness}\footnote{A planning
+  algorithm is \emph{complete} if invoked on a solvable problem $P$ is
   guaranteed to return a solution.} along with derived results for
@@
 These drawbacks have been addressed by a number of frameworks like
-\texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
+\texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTeT}
 \cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval

### 2012-06-05 final-version/nddlshop.tex
Final Version to be sent to Publisher

--- 2012-02-24 trunk/europa.tex
@@
-\subsection{Architecture}
-\label{sec:europa:arch}
-
-\eu is implemented as a library that encapsulates all elements
-described above making them accessible via modeling and Application
-Programming Interface (API) layers.  A user will typically use the
-modeling facilities to inject a problem domain description and the
-API to extract and manipulate plans and to obtain information about
-plan flaws and constraint violations at any stage of the planning
-process.
-
-\begin{figure}[b]
-\centering
-\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
-\caption{\small The \eu software architecture.}
-\label{fig:europa-architecture}
-\end{figure}
-
-Fig. \ref{fig:europa-architecture} shows the main architectural
-components in \eu and their relationships. The \textbf{Constraint
-  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
-constraints that define relationships among them. It also provides an
-efficient arc consistency mechanism \cite{mackworth77}. The
-\texttt{CRE} is designed so that specialized reasoning algorithms for
-new constraints can be easily and efficiently plugged in. The
-\textbf{Plan Database} (\texttt{PDB}) manages object and token
-types. It also maintains the state of partial plans in terms of
-object, variable and token instances. This constitutes the data
-foundation on which end user applications and automated problem solvers
- are built. The \textbf{Rules Engine} (\texttt{RE})
-modularly adds the semantics of domain rules that describe
-dependencies between tokens. The \textbf{Solver Module} provides the
-framework for flaw detection and resolution that is needed to
-implement search algorithms for planning. The \textbf{API Layer}
-provides access to services from all the different modules in a
-programmatic way so that \eu can be easily embedded into specialized
-applications.
-
-Describing problem domains and problem instances in terms of objects,
-variables and domain rules in a programmatic way is cumbersome to the
-extent of being impractical; this motivated the creation of modeling
-languages that support those definitions in a much more natural and
-compact way. \texttt{NDDL} (New Domain Description Language
-[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
-\eu and will be described in detail in the Modeling section
-below. Another, \texttt{ANML} is a language under development that
-provides a more intuitive syntax \cite{smith08}.
-
-In an effort to mimimize the amount of work that a user needs to
-perform to solve specific problems, \eu also bundles extension
-modules that have been found to be useful in many domains:
-
-\begin{enumerate}
-
-\item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
-  reason about temporal constraints
-
-\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
-  and Solver modules to provide reasoning and search for metric
-  resources.
-
-\item \textit{Chronological Backtracking Solver}: A general purpose
-  built-in solver to provide the ability to plan out-of-the-box.
-  Problems that require challenging search may require specialized
-  solvers to be built as explained in Section \ref{sec:europa:search}.
-
-\end{enumerate}
-
-\subsection{Modeling in \eu}
-\label{sec:europa:modeling}
-
-\eu can ingest descriptions of models, plans and goals and thereby can
-be applied to different domains and problems, simply by providing
-different models and goals. Consequently the expressiveness of the
-language for models, goals and plans is of great importance. For \eue,
-\texttt{NDDL} allows the user to specify the different components of a
-problem domain in a precise and concise manner. Among the main
-features that make \texttt{NDDL} a powerful tool for describing
-problem domains are that it is object oriented. \texttt{NDDL} supports
-classes and inheritance in similar fashion to popular Object-Oriented
-languages like Java and C++. It also supports polymorphism for some of
-the planning components. Using object classes and instances is a
-time-tested approach to naturally describe problem domains
-\cite{rumbaugh}. It offers declarative constructs to define
-constraints and causality dependencies between actions and effects and
-finally it offers procedural constructs to populate \eus \texttt{PDB}
-with specific problem instances expressed in terms of objects,
-variables, constraints, facts and goals. To articulate the most important
-elements of \texttt{NDDL}, we show how the Shopping Agent problem
-could be modeled:
-
+\begin{figure}[h!]
 \begin{verbatim}
@@
 // c) Can't Go somewhere and be At somewhere at the same time.
+
 class AgentLocation extends Timeline{
@@
 }
-
 // In addition to having a location timeline, the agent can buy and own things.  
@@
 // so they can't be on the same Timeline
+
 class Agent {
@@
   }
-
   action Go {
@@
 }
-\end{verbatim}
-
-In the model above one can see how the main elements of a problem domain
-can be mapped to \eus representation:
-
-\begin{description}
-
-\item[\textbf{Object Types}:] \texttt{NDDL} supports an object-oriented
-  approach, where common data and behavior can be abstracted as a
-  class. In this case, we have the \textit{Product} class represent
-  the products that the Shopping Agent needs, \textit{Location}
-  represents the places where products can be found,
-  \textit{ProductLocation} represents the association arising when a
-  product can be purchased at a location. The Shopping Agent itself is
-  represented by an \textit{Agent} class, which relies on the
-  \textit{AgentLocation} class to keep track of where the agent is at
-  any point in time. \texttt{NDDL} supports a single rooted class
-  hierarchy so that state and behavior can be extended through
-  inheritance. We take advantage of that feature by making
-  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
-  class which doesn't allow predicates to overlap in time for the same
-  object instance. This feature is used for instance, to ensure that
-  an agent can only be at one place at a time in our example.
-
-\item[\textbf{Variable Types and Instances}:] All object attributes
-  are represented as variables in this model. For instance, it uses
-  integer variables to represent the built-in temporal parameters for
-  each token (duration, start and end times), string
-  variables to represent \textit{Location} and \textit{Product} names,
-  and object reference variables to represent the structural
-  relationships between different objects (for instance, each Agent
-  maintains a variable to keep track of its \textit{AgentLocation}).
-
-\item[\textbf{Predicate Types}:] Some classes like \textit{Product},
-  \textit{Location} and \textit{ProductLocation} are static and so
-  don't have any predicate or action types associated with them. The
-  \textit{AgentLocation} class is used to keep track of state and so
-  has \texttt{At} and \texttt{Going} predicate types associated with
-  it. The \textit{Agent} class keeps track of the products it
-  purchases through the \texttt{Own} predicate; having a separate
-  \textit{AgentBag} class with its own predicates would have been an
-  equally valid modeling choice.
-
-\item[\textbf{Action Types}:] In our example, \textit{Agent} is solely
-  the class that has behavior associated with it. Consequently, it
-  defines action types to represent when the agent is taking action to
-  \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
-  \textit{Product}.
-
-\end{description}
-
-\begin{verbatim}
-
-// Define the rules for our actions:
-Agent::Go {
-  met_by(condition object.location.At origin);
-  eq(from, origin.loc);
-
-  equals(effect object.location.Going going);
-  eq(going.from, from);
-  eq(going.to, to);
-
-  meets(effect object.location.At destination);
-  eq(to, destination.loc);
-}
-
-Agent::Buy {
-  // A Buy takes 10 time units
-  eq(10, duration);
-
-  // initialized to all locations
-  ProductLocation possibleStores;
-
-  // limit possibleStores variables to ones that provide what we need to buy
-  eq(product, possibleStores.product);
-
-  // We must be At a location during our Buy, and that location must have the
-  // product we want available:
-  contained_by(condition object.location.At currLocation);
-  eq(currLocation.loc, possibleStores.location);
-
-  meets(effect Own purchase);
-  eq(purchase.product,product);
-}
-\end{verbatim}
-
-Domain rules can be defined in \texttt{NDDL} in the context of a
-predicate or an action. In the example above we can see the main
-elements of a rule definition:
-
-\begin{description}
-
-\item[\textbf{Variables}:] A token type (predicate or action) is part
-  of a class, therefore class attributes (see \texttt{object.location}
-  above), token type parameters (see \texttt{[from, to]} in
-  \texttt{Agent::Go}, or \texttt{[duration, product]} in
-  \texttt{Agent::Buy}) and locally declared variables (see
-  \texttt{possibleStores} in \texttt{Agent::Buy}) can all be used in
-  the definition of a rule.
-
-\item[\textbf{Constraints}:] Restrictions on values that variables can
-  take within a rule are expressed through constraints. In the example
-  above the \texttt{eq} constraint is used; in general, any constraint
-  available in \eus library can be used.
-
-\item[\textbf{Subgoals}:] In \texttt{Agent::Go}, subgoals are used to
-  specify that the Shopping Agent's location must transition from
-  origin to destination in a temporally consistent way. The
-  \texttt{[meets, equals, metby]} operators (see Section
-  \ref{sec:europa:inference}) are used to specify subgoals, along with
-  the temporal relationship each token representing a subgoal has with
-  the the token representing the \texttt{Agent::Go} action. Also,
-  \texttt{[condition, effect]} annotations can be used when defining a
-  subgoal inside an action. \eu does not intrinsicaly associate semantics
-  with these annotations; but they can be extremely useful for search
-  modules when creating plans.
-
-\end{description}
-
-With that specification for the Shopping Agent domain, a particular
-problem instance could be stated as follows:
-
-\begin{verbatim}
-
-// Allocate instances
-Location Home = new Location("Home");
-Location SuperMarket = new Location("SuperMarket");
-Location HardwareStore = new Location("HardwareStore");
-
-Product Banana = new Product("Banana");
-Product Milk = new Product("Milk");
-Product Drill = new Product("Drill");
-
-ProductLocation bananaLocation = new ProductLocation(SuperMarket, Banana);
-ProductLocation milkLocation = new ProductLocation(SuperMarket, Milk);
-ProductLocation drillLocation = new ProductLocation(HardwareStore, Drill);
-
-Agent agent = new Agent();
-
-// Indicate that the database is closed - no new objects can be created
-// this allows EUROPA to perform more efficient inference
-close();
-
-// We start the day at Home:
-fact(agent.location.At atHomeForBreakfast);
-atHomeForBreakfast.loc.specify(Home);
-
-// Goals for all of the agent's needs, buy if needed
-goal(agent.Own gotMilk);
-gotMilk.product.specify(Milk);
-goal(agent.Own gotBanana);
-gotBanana.product.specify(Banana);
-goal(agent.Own gotDrill);
-gotDrill.product.specify(Drill);
-
-// Make sure agent is home for dinner
-goal(agent.location.At atHomeForDinner);
-atHomeForDinner.loc.specify(Home);
-
-// Agent has all day to satisfy goals:
-gotMilk after atHomeForBreakfast;
-gotMilk before atHomeForDinner;
-
-gotBanana after atHomeForBreakfast;
-gotBanana before atHomeForDinner;
-
-gotDrill after atHomeForBreakfast;
-gotDrill before atHomeForDinner;
-
-// Force things to happen within our planning horizon:
-int horizonStart = 0;
-int horizonEnd = 100;
-leq(horizonStart, atHomeForBreakfast.start);
-leq(atHomeForDinner.end, horizonEnd);
 
 \end{verbatim}
-
-This problem instance consists of \textbf{Variables} and
-\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
-\textbf{Constraints}. In our example we define temporal variables that
-are used to specify the planning horizon (\texttt{horizonStart} and
-\texttt{horizonEnd}) and objects are used to specify locations,
-products, product-location associations and the Shopping Agent that we
-want to plan for. Facts are used to specify the current state of the
-world, do not need any support and are represented as tokens in the
-\texttt{PDB}. In this case we use facts to specify that the Shipping
-Agent is initially at home. Goals are also represented as tokens in
-the \texttt{PDB}. Different from facts, goals trigger inference in \eu
-which determines whether flaws (unresolved subgoals, overloaded
-resources, etc.) need to be resolved; if so, search modules can be
-invoked to resolve those flaws as part of the deliberation
-process. Finally, constraints are restrictions on and between facts
-and goals. In the example above, we use temporal constraints to
-specify when facts hold and when the goals should be accomplished. In
-the model above, \texttt{before} and \texttt{after} are \texttt{Allen
-  Algebra} \cite{allen84} operators, which \eu supports to specify
-temporal relations between tokens (specifically between their start
-and end timepoints). The algebra is particularly effective in dealing
-with interval arithmetic; its elements are summarized in
-Fig. \ref{fig:allen-algebra}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.4]{figs/Allen-algebra.pdf}
-\caption{\small Temporal relations defined within the planner are
-  based on \texttt{Allen Algebra} \cite{allen84} relations shown
-  above.}
-\label{fig:allen-algebra}
+\caption{\small \nd model of the Shopping Agent example.}
+\label{fig:nddlshop}
 \end{figure}
-
-\subsection{Automated Inference}
-\label{sec:europa:inference}
-
-In \eue, all variables and constraints from a problem domain
-description are stored inside the \texttt{CRE}.  The \texttt{CRE}
-provides a framework where any number of constraint types can
-collaborate to perform inference and prune variable domains using
-Bounds Propagation and Domain Reduction techniques \cite{marriott98}. 
-\eu uses a Propagator pattern which keeps track of a set of
-constraint instances of the same type and is responsible for ensuring
-variables are updated through inference whenever a relevant change occurs
- (variable domain modifications or constraint addition/deletion).
-
-In a planning problem, variable or constraint changes can come
-directly from an end user (by introducing new activities or goals and
-relating them to the rest of the plan through constraints, by changing
-the start or end time of an activity, etc), or from an automated planner (by
-expanding subgoals, introducing precedence constraints to resolve
-mutual exclusion conflicts, etc). When these changes occur, each
-propagator is only able to ensure consistency for the constraints that
-are associated with it.  To maintain global consistency, the
-\texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
-\cite{mackworth77} where propagators are invoked repeatedly until
-the constraint network returns to a quiescent state.
-
-\subsubsection{Constraint Formulation}
-\label{sec:europa:constraints}
-
-An \eu user can introduce new constraint types by implementing
-specialized constraint and propagator classes; however, \eu provides a
-built-in constraint library with a range of arithmetic, temporal, set
-and resource constraint types so that a large number of problem
-domains can be modeled without having to introduce new constraint
-types. Tables \ref{tab:calconst}, \ref{tab:setconst},
-\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
-types included in \eus Constraint Library.
-
-\input{constraints-table}
-
-Temporal and resource constraints are especially important to model
-real-world planning domains.
-
-\eu uses variables to explicitly represent timepoints for plan
-actions and states. Constraints among timepoints provide a natural
-way to express domain axioms. For example, in order to state that
-activity $A$ must occur before activity $B$ we can say that the end
-timepoint of $A$ is $leq$ the start timepoint of $B$.
-\cite{dechter91} proposed that constraints among timepoints can be
-grouped together to form a Simple Temporal Network
-(\texttt{STN}). Such a network can be transformed into a Distance
-Graph (\texttt{DG}) where the outward arc from a node represents the
-maximum distance from the source node to the target node.
-Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
-a single constraint along with the resulting \texttt{DG}.
-\cite{dechter91} also showed that shortest path algorithms could be
-used to propagate values in the network and discover a negative cycle:
-namely a path from a node to itself that has a path length less than
-$0$. If such a cycle exists, the network is inconsistent. It was
-further shown that a single-source shortest path algorithm was
-sufficient to detect a negative cycle and provide sufficient
-propagation to yield a backtrack-free search. \eus propagator for
-temporal constraints is a straight implementation of this algorithm.
-
-\begin{figure}[!htb]
-  \centering
-  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
-  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
-  \caption{\small An illustration of a STN and its corresponding
-    distance graph.}
-  \label{fig:stn}
-\end{figure}
-
-Most real-world planning problems need to deal with plans made up of
-actions that typically depend upon limited resources (\eg fuel,
-storage capacity, battery state of charge, data buffers).  \eu
-provides built-in support for resource computation. In \eus
-representation resource object types come in three forms with related
-constraints. \textbf{Reservoir} resources can have both
-\texttt{consume(resource,time,quantity)} and
-\texttt{produce(resource,time,quantity)} constraints; production and
-consumption is instantaneous. \textbf{Reusable} resources on the other
-hand, have constraints that use a quantity of the resource for a
-specific duration (\texttt{uses(resource,start,duration,quantity)}),
-which are then subsequently released. \textbf {Unary} resources are
-reusables with capacity$ = 1$, which allows \eu to perform run time
-savings in comparison to the generic reusable class. Configuration
-options are offered such that in addition to the usual constraint
-violations when a resource is completely depleted, plan flaws can be
-generated whenever a resource level is outside a specific upper or
-lower bound, or when instantaneous production or consumption exceeds a
-specific threshold.
-
-Since \eu supports a flexible temporal representation, where time is
-represented by intervals, instead of single values, determining
-whether a resource is the source of a plan flaw or a constraint
-violation in general is substantially more difficult than just
-computing a profile of the resource and checking against some limits.
-With temporal flexibility, a number of resource profiles could be
-possible depending on precisely when each production/consumption event
-occurs out of a range of possible values that are in the domain of the
-variables that represent the temporal extent for the event.
-\cite{Muscettola04,Muscettola06} proposed an algorithm where the
-variables that represent the time for production/consumption events
-are placed in a temporal network, similar to the network used by the
-temporal propagator; constraints on the resource can then be seen as
-quantities flowing through that network. Computing the maximum flow
-then, yields the maximum resource usage and thus a lower bound on the
-resource level; computing a minimum flow will similarly yield an upper
-bound. Since maximum flow on a temporal network can be computed in
-polynomial time, \eu takes advantage of this algorithm to efficiently
-compute upper and lower bound envelopes on the resource
-level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
-computation.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
-\caption{\small An example of resource envelopes computed by \eu.}
-\label{fig:resenvelopes}
-\end{figure}
-
-Resource envelopes can be useful when trying to evaluate plan
-brittleness in the face of changing activity durations or start
-times. They can also be useful when searching for plans; however they
-are relatively expensive to compute. Therefore their ability to prune
-the search space must be weighted against the ability to search faster
-\cite{6007772, morris11}.
-
-\subsection{Search in \eu}
-\label{sec:europa:search}
-
-Given the above, an automated problem solver can be created using a
-\emph{domain model} that describes the variable, object, predicate and
-action types that are relevant for the problem, a \emph{problem
-  instance} or initial state that consists of variable and object
-instances that exist for the entire planning horizon, \emph{temporally
-  scoped predicate and action} instances and \emph{temporally scoped
-  goals}. This information is retained in \eus \texttt{PDB} so that
-inference and search mechanisms can be used to look for a problem
-solution.
-
-Since temporal intervals may extend indefinitely in both positive and
-negative directions, for problems with a temporal dimension it is also
-common to specify a \emph{planning horizon} when invoking a solver;
-any decisions that fall completely outside of the horizon can be
-ignored. This is useful to ignore activities that may occur far in the
-future or in the past to be relevant for a particular plan, ensuring
-that the solver is making as few commitments as necessary and reducing
-the complexity of plan generation.  The \emph{planning horizon} is
-also useful when dealing with recurrent activities; consider for
-instance a model where a medical checkup has to occur every $3$
-months. The user can model a recurring activity and then use the
-horizon to ensure that a finite number of medical checkup activities
-are generated as part of a plan.
-
-\subsubsection{Flaw Resolution}
-\label{sec:europa:flaws}
-
-\eu provides a built-in solver that performs \emph{Plan Space
-  Planning} \cite{ghallab04}; the initial state is considered a
-partial plan that needs to be refined toward a solution that achieves
-stated goals. The operations to refine the partial plan \texttt{PP} at
-any time are:
-
-\begin{enumerate}
-
-\item Find flaws of \texttt{PP}, \ie conditions that prevent it from
-  being a solution plan.
-
-\item Select one such flaw.
-
-\item Select a resolver for the flaw.
-
-\item Refine \texttt{PP} by applying the resolver.
-
-\item If an inconsistency is found, try another resolver.
-
-\item If all possible resolvers for a particular flaw fail, return
-  failure, otherwise continue until resolving all flaws.
-
-\end{enumerate}
-
-The initial partial plan is the state of \eus Plan Database after the
-initial state has been instantiated. This results in a set of
-variable, object and token instances. Inference then takes place to
-detect flaws in the partial plan; the three kinds of flaws that can be
-detected are:
-
-\begin{description}
-
-\item[\textbf{Unbound Variable}:] a variable in the partial plan whose
-  domain is not a singleton. Unbound Variables are resolved by value
-  selection from their domain.
-
-\item[\textbf{Open Condition}:] an open condition is an
-  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
-  by explicitly posted goals, or when a token is activated, rules may
-  cause the creation of \texttt{INACTIVE} slave tokens as described
-  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
-  three ways:
-
-  \begin{description}
-
-  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
-    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
-    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
-    and the temporal constraints involving $a$ are satisfied by
-    $a?$. Thus, $a$ and $a?$ can be considered the same token
-    without the need to introduce $a$ in the plan. Consequently, the
-    domain rules associated with $a$ are not introduced into the
-    planning process, since they are already triggered when $a?$ was
-    introduced in the plan.
-
-  \item[\textbf{Activation}:] A token $a$ is introduced in the
-    current plan associating it with the appropriate timeline, without
-    choosing a specific time slot; if its timing causes a problem a
-    Threat flaw will be triggered (see below).  The domain rules
-    associated with $a$ are applied and the subgoal activities
-    resulting from those domain rules are introduced as \texttt{INACTIVE}
-    tokens.  This results in a number of open condition flaws,
-    corresponding to the new subgoal activities.
-
-  \item[\textbf{Rejection}:] Some goals may be optional; if so its
-    corresponding flaw may be resolved by discarding the
-    \texttt{INACTIVE} token.
-
-  \end{description}
-
-\item[\textbf{Threat}:] Once a token has been placed in the partial
-  plan it may impact other tokens indirectly through possible
-  overlapping requirements on objects. Recall for example that a token
-  may belong to objects (\eg timelines) which require a total order
-  over their tokens. If any two tokens could possibly overlap (though
-  not necessarily), then they pose a threat to each other in terms of
-  achieving an extension of the current partial plan which is complete
-  and consistent. Similarly, threats may arise where tokens share a
-  common resource and their current state might yield extensions of
-  the current partial plan which are inconsistent. Threats are
-  resolved by imposing ordering constraints among tokens.
-
-\end{description}
-
-Open conditions and threats allow flaw detection and resolution at a
-higher-level of abstraction (\ie in terms of objects and tokens) than
-that of binding variables as is common in Constraint Satisfaction
-Problems. This is advantageous when a solver applies heuristics for
-ordering choices since it provides a richer context in which to make
-decisions.  Furthermore, it aids in reducing the amount of work done
-by a solver so that only the necessary refinements are made leaving a
-partial plan with sufficient flexibility. For example, one can bypass
-committing unbound timepoint variables since threats will force a
-solver to impose ordering restrictions on these variables.  Thus the
-planning process may yield partially-ordered plans for which all
-possible extensions are provably valid.
-
-The Plan Space Planning algorithm described above can be implemented
-in multiple ways depending on the approach chosen for flaw and
-resolver selection and for backtracking. \eus built-in solver
-implements a chronological backtracking algorithm that is summarized
-in Algorithm \ref{alg:europa:solve}.
-
-\begin{algorithm}[H]
-\KwIn{PartialPlan $plan$}
-\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
-\Begin{
-  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
-  \BlankLine
-  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
-  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
-  \BlankLine
-  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
-  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
-    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
-    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
-    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
-   }
-   \Return \texttt{false} \; \label{li:noplan}
- }
-\caption{$\mathrm{bool} ~ solve(plan)$}
-\label{alg:europa:solve}
-\end{algorithm}
-
-The algorithm takes as input a partial plan $p$ and returns true if a
-complete and consistent refinement of $p$ could be found (or if $p$ is
-initially complete and consistent) and false otherwise. Briefly the
-algorithm is described as:
-
-\begin{description}
-
-\item[Line \textbf{\ref{li:propagate}}:] Propagate the constraints to
-  test for inconsistency. If inconsistent, then return false since no
-  refinements to $p$ can yield a consistent plan.
-
-\item[Line \textbf{\ref{li:selectflaw}}:] Choose a flaw from the set
-  of available flaws.
-
-\item[Line \textbf{\ref{li:noflaw}}:] If there are no flaws, then $p$
-  is complete and terminate with success.
-
-\item[Line \textbf{\ref{li:decisionpt}}:] Formulate a decision point
-  which is a branch in the search space. Each choice is a particular
-  refinement operation and the \texttt{DecisionPoint} collects all
-  possible refinement operations for the given flaw.
-
-\item[Line \textbf{\ref{li:decisionloop}}:] Keep trying until the
-  chosen flaw is resolved, or until there are no further resolvers to
-  try.
-
-\item[Line \textbf{\ref{li:decide}}:] A new partial plan is obtained
-  by application of a refinement operator. Note that the ordering over
-  refinement operators to select is a non-deterministic step.
-
-\item[Line \textbf{\ref{li:recurse}}:] Recursive call to solve the
-  new planning problem. If successful, then success.
-
-\item[Line \textbf{\ref{li:backtrack}}:] Otherwise, retract the last
-  refinement operation and move on to try the next one.
-
-\item[Line \textbf{\ref{li:noplan}}:] Arriving here implies all
-  options to resolve the flaw have been exhausted, including the case
-  where no options were available initially. Thus the problem is
-  insoluble.
-
-\end{description}
-
-Algorithm \ref{alg:europa:solve} provides for a sound and complete
-search, assuming that no flaws or available refinement operators are
-pruned unnecessarily.  Dead-ends in the search are discovered through
-constraint propagation, a vehicle for evaluating the consistency of a
-partial plan and also for filtering infeasible values from
-consideration prior to commitment. Consistency testing is initiated by
-the \texttt{isInconsistent} procedure.
-
-The algorithm permits a heuristically controlled search by
-applying orderings for \texttt{chooseFlaw} and
-\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
-search. Heuristics can be implemented by applying ordering/filtering
-policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
-instance, by dealing with earliest flaws first, or by making planning
-decisions (open condition flaws) before scheduling decision (threats).
-
-It should be emphasized that while this algorithm is commonly employed
-by applications that embed  \eu, it is only one of many that could be
-implemented. As a simple example, consider the following solver for
-the $N$-Queens problem, where an integer variable V$_i$ represents the
-queen in column $i$; V$_1$ $= 3$ implies that the variable on column
-$1$ is placed on row $3$.  Non-attack constraints are posted as
-arithmetic constraints among the V$_i$ variables to ensure that all
-the rows and diagonals contain only one queen.
-
-\begin{algorithm}[H]
-  \KwIn{ConstraintEngine $ce$}
-  \KwIn{int $maxIter$}
-  \Begin{
-    int $curIter \leftarrow 0$ \;
-    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
-        maxIter \right)$ }{
-      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
-      \;
-      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
-      curPos)$ \;
-      bool $moved \leftarrow $ \texttt{false} \;
-      \ForEach{int $newPos$ of $moves$}{
-        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
-        \lIf{$moved$}{ \textbf{break} \; }
-      }
-      \lIf{$! moved$}{
-        $forceMove(queenToMove, curPos, moves.first())$ \;
-      }
-      $checkSolution()$ \tcp*{Evaluate if a new best solution}
-      $curIter$\texttt{++} \;
-    }
-  }
-  \caption{$solveNQueens(ce, maxIter)$}
-\end{algorithm}
-
-\eus Constraint Reasoning Engine or \texttt{CRE} can be used to compute the
-non-attack constraint violations, with \texttt{ce.getViolations()} a
-call on the \eu API. The \texttt{CRE} can also be used to pick the queen 
-that has he most violations in \texttt{getQueenWithMaxViolation()}, and to
-look ahead and see what moves are possible for a queen that would remove its
-violations in \texttt{getMoves()}.
-
-In similar fashion, the constraint violation and flaw reporting
-mechanisms that \eu provides can be used as building blocks that
-specialized or general purpose search algorithms can use to construct
-plans.
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "setobook"
-%%% End:

### 2012-06-05 trunk/final-version/europa.tex
Final Version to be sent to Publisher

--- 2012-06-05 europa.tex
@@
 \centering
-\includegraphics[scale=2.5]{figs/EUROPA-Architecture.jpeg}
+\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpeg}
 \caption{\small The \eu software architecture.}
@@
 
-\kcomment{Fig. \ref{fig:nddllight} is a \nd (covered in more detail in
-  Section \ref{sec:europa:modeling}) model of a light bulb and a
+\kcomment{Fig. \ref{fig:nddllight} is a \nd model (covered in more
+  detail in Section \ref{sec:europa:modeling}) of a light bulb and a
   switch that controls it along with a problem instance that requires
-  the light bulb to be \texttt{off} by time$=10$, given an initial
-  state with the bulb \texttt{on} at time$=0$. The \nd module allows
+  the light bulb to be \texttt{off} by time $=10$, given an initial
+  state with the bulb \texttt{on} at time $=0$. The \nd module allows
   \eu to ingest this model and the problem description. Once the \nd
@@
   distance between them.}]{\label{fig:stn1}\includegraphics[width=0.5\columnwidth]{figs/stn-graph.jpeg}}
-  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.5\columnwidth]{figs/distance-graph.pdf}}
+  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.5\columnwidth]{figs/distance-graph.jpeg}}
   \caption{\small An illustration of a \texttt{STN} and its
@@
   \BlankLine
-  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \comment{\{\emph{potential branching point(s)}\}}\;\label{li:decisionpt}
+  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \tcp*{potential branching point(s)};\label{li:decisionpt}
   \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
@@
     \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
-    \lElse{ $decision.undo()$ \comment{\{\emph{possible backtracking point}\}}\; } \label{li:backtrack}
+    \lElse{ $decision.undo()$ } ;// \texttt{possible backtracking point}\ \label{li:backtrack}
    }

### 2012-06-06 final-version/concepts.tex
updates to final version

@@
 planner that was flown as part of the Remote Agent \texttt{RAX}
-experiment. Concurrent to \texttt{RAX} deployment, a C++ version of
-the planner was implemented \kcomment{at NASA Ames. This} went on to
-be used for the \texttt{MAPGEN} \cite{bresina05} system and continues
-to be used to date, in the mission-critical uplink process for the
-command/control of the Mars Exploration Rovers (\texttt{MER})
-mission. A vastly re-factored, higher performance open source version
-\cite{europapso} was named \eut and is the subject of this chapter and
-used by \rxe.  Note however, in using the name \eue, we refer to the
-\emph{fundamental ideas as well as the planning paradigm} even if
-there exists a specific implementation.
+experiment \cite{mus98}. Concurrent to the \texttt{RAX} deployment, a
+C++ version of the planner was implemented \kcomment{at NASA
+  Ames. This} went on to be used for the \texttt{MAPGEN} system and
+continues to be used to date, in the mission-critical uplink process
+for the command/control of the Mars Exploration Rovers (\texttt{MER})
+mission \cite{bresina05}. A vastly re-factored, higher performance
+open source version was named \eut \cite{europapso} and is the subject
+of this chapter and used by \rxe.  Note however, in using the name
+\eue, we refer to the \emph{fundamental ideas as well as the planning
+  paradigm}.
 

### 2012-06-06 final-version/intro.tex
updates to final version

--- 2012-05-02 trunk/intro.tex
@@
 \label{sec:intro}
+
+\setcounter{footnote}{0}
 
@@
 balanced consideration of mission objectives, environmental conditions
-and available resources. Planning (or deliberation or projection in
-action space\footnote{\kcomment{We use the terms 'planning' and
-    'deliberation' interchangeably in this chapter.}}) is important to
-balance current needs of a robot with future desires or goals taking
-into account time and available resources in-situ on the
-robot. Planning without time or action planning in and of itself is
-not adequate; dealing with time ensures that the balance between now
-and the future is handled systematically. Yet, plans have to be tied
-to robotic action and in dynamic real-world
-\kcomment{environments}. Further, plans \kcomment{not only} project
-future state which are likely to change, but \kcomment{need to factor
-  in} substantial environmental uncertainty\footnote{It is important
-  to remember that these principles are broad much as Dwight
-  Eisenhower is reputed to have said ``\emph{In preparing for battle I
-    have always found that plans are useless, but planning is
-    indispensable}'' and ``\emph{Failing to plan is planning to
-    fail}''.}. \kcomment{In the oceanographic context,} use of
-environmental cues is often hampered by poor \kcomment{predictive}
-skill \cite{anderson2009} of ocean models and little to no
-availability of a priori synoptic views of the survey
+and available resources. Planning\footnote{We use the terms 'planning'
+  and 'deliberation' interchangeably in this chapter.}  (or
+deliberation or projection in action space) is important to balance
+current needs of a robot with future desires or goals taking into
+account time and available resources in-situ on the robot. Planning
+without time or action planning in and of itself is not adequate;
+dealing with time ensures that the balance between now and the future
+is handled systematically. Yet, plans have to be tied to robotic
+action and in dynamic real-world \kcomment{environments}. Further,
+plans \kcomment{not only} project future state which are likely to
+change, but \kcomment{need to factor in} substantial environmental
+uncertainty\footnote{It is important to remember that these principles
+  are broad much as Dwight Eisenhower is reputed to have said
+  ``\emph{In preparing for battle I have always found that plans are
+    useless, but planning is indispensable}'' and ``\emph{Failing to
+    plan is planning to fail}''.}. \kcomment{In the oceanographic
+  context,} use of environmental cues is often hampered by poor
+\kcomment{predictive} skill \cite{anderson2009} of ocean models and
+little to no availability of a priori synoptic views of the survey
 area. \kcomment{Planning therefore provides a ``buffer'' against the
@@
   \centering \vskip-5pt
-  \includegraphics[scale=0.1]{figs/MBARI-AUV.jpg}
+  \includegraphics[scale=1.2]{figs/MBARI-AUV.jpeg}
   \caption{\small The MBARI \emph{Dorado} AUV on its support vessel
@@
 This chapter is organized as follows. We start \kcomment{by placing
-  this effort in the context of the overal research in AI Planning in
+  this effort in the context of the overall research in AI Planning in
   Section \ref{sec:related}} followed by some foundational concepts in
-our Planning framework in Section~\ref{sec:concepts} which includes a
+our planning framework in Section~\ref{sec:concepts} which includes a
 brief overview of AI Planning in Section~\ref{sec:planningfound}. In
@@
 Section~\ref{sec:arch:intertwine}. We offer results from our
-experiments at sea in Section~\ref{sec:results}, offer some insights
-into the near-term future goals of our research in
-Section~\ref{sec:future} and conclude with
-Section~\ref{sec:conclusion}.
+experiments at sea in Section~\ref{sec:results}, some insights into
+the near-term future goals of our research in Section~\ref{sec:future}
+and conclude with Section~\ref{sec:conclusion}.
 

### 2012-06-06 final-version/related.tex
updates to final version

--- 2012-05-02 trunk/related.tex
@@
 defense conflate the methodology of use (tethered vs. untethered) with
-their control. In this chapter, our notion of 'autonomy' deals
-explicitly not only with the notion of dealing with the \emph{sense,
-  plan and act} paradigm, but also the need to perform
-\emph{computational search} between different action outcomes, an idea
-central to the field of AI. Consequently, our literature survey here
-is targeted towards a more focused and (to this chapter) relevant part
-of the field of computation.
+their control. In this chapter, our notion of 'autonomy' refers not
+only to the notion of dealing with the \emph{sense, plan and act}
+paradigm in robotics, but also the need to perform \emph{computational
+  search} between different action outcomes, an idea central to the
+field of AI. Consequently, our literature survey here is targeted
+towards a more focused and (to this chapter) relevant part of the
+field of computation.
 
@@
  \centering
- \includegraphics[scale=0.4]{figs/planner-arch.pdf}
+ \includegraphics[scale=2]{figs/planner-arch.jpeg}
  \caption{\small A general architectural block diagram for an AI based

### 2012-06-06 final-version/europa-basics.tex
updates to final version

@@
 could actually be tractable first in simulation and subsequently in
-actual robotic platforms. In this chapter our emphasis is to focus on
-embedded AI Planning and the execution of plans on an embedded
-\kcomment{marine robotic} platform. The techniques we highlight are
-general purpose as noted in Section \ref{sec:intro}.
+actual robotic platforms. In this chapter our emphasis is on AI
+Planning and plan execution on an embedded \kcomment{marine robotic}
+platform. The techniques we highlight are general purpose as noted in
+Section \ref{sec:intro}.
 
-\begin{figure}[b]
+\begin{figure}[!t]
 \centering
@@
     etc.  The embedded planner has to have a ``model'' of the world in
-    which it 'lives' and has to use the task primitives in this model
-    to structure the actions so it achieves its goal. Constraints
-    control when certain tasks can or cannot occur. For example the
-    robot must obtain the keys and wallet \emph{before} driving to the
-    store and pick up the milk \emph{before} purchasing it.
+    which it ``lives'' and has to use the task primitives in this
+    model to structure the actions so it achieves its
+    goal. Constraints control when certain tasks can or cannot
+    occur. For example the robot must obtain the keys and wallet
+    \emph{before} driving to the store and pick up the milk
+    \emph{before} purchasing it.
 \end{quote}

### 2012-06-06 trunk/final-version/intro.tex
updates to final version

--- 2012-06-06 final-version/intro.tex
@@
 challenges remain, principally in making such robotic devices more
-attune to the environment, to sample large swaths at the Meso-scale
+attuned to the environment, to sample large swaths at the Meso-scale
 ($> 50$ km\textsuperscript{2}) and doing so systematically and
@@
 Execution. Current work is also looking at how other fields in Machine
-Intelligence especially Machine Learning can be leveraged to impact
+Intelligence, especially Machine Learning can be leveraged to impact
 the science of engineering systems in marine robotics and as a

### 2012-06-06 trunk/final-version/related.tex
Final Version to be sent to Publisher
updates to final version

--- 2012-05-02 trunk/related.tex
@@
 Such techniques have proved their mettle in the first round of use of
-AUVs; they've provided operators a simple way to handle control
+AUVs; they have provided operators a simple way to handle control
 complexity of the vehicle while returned data at far higher resolution
